### 2016-01-28 paper.tex
stub of paper, a few details added on the analysis

\documentclass[10pt]{sigplanconf}

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{stmaryrd}
\usepackage{color}
\usepackage{graphics}

\definecolor{darkgreen}{rgb}{0.0,0.5,0.0}
\definecolor{darkpurple}{rgb}{0.6,0.0,0.6}
\definecolor{orange}{rgb}{0.8,0.4,0.0}
\definecolor{darkorange}{rgb}{0.5,0.2,0.0}
\definecolor{marco}{rgb}{0.0,0.3,0.5}
\definecolor{gray}{rgb}{0.2,0.2,0.2}

\newcommand{\dnote}[1]{\textcolor{darkpurple}{Dom: #1}}

\newcounter{block}

\newtheorem{theorem}[block]{Theorem}
\newtheorem{lemma}[block]{Lemma}
\newtheorem{proposition}[block]{Proposition}
%\newtheorem{definition}[block]{Definition}

\theoremstyle{definition}
\newtheorem{remark}[block]{Remark}
\newtheorem{example}[block]{Example}
\newtheorem{definition}[block]{Definition}

% Writing macros
\newcommand{\ie}{\emph{i.e..}}
\newcommand{\eg}{\emph{e.g.}}

% Semantics related
\newcommand{\interp}[1]{\llbracket{#1}\rrbracket}

\title{Abstract Shape Specifications for Stencil Verification}
\authorinfo{}{}

\begin{document}
\maketitle

\begin{abstract}
\end{abstract}

\bibliographystyle{plain}

\section{Introduction}

\section{Specification language}

\begin{align*}
spec ::= & \; S  \; \texttt{,} \; spec \, \mid \, \epsilon \\ \\
%
S ::= &  \; \texttt{forward} \; [\textit{ixs}] \; [\textit{depth}] \\
& \; \mid \texttt{backward} \; [\textit{ixs}] \; [\textit{depth}] \\
& \; \mid \texttt{centered} \; [\textit{ixs}] \; [\textit{depth}] \\
& \; \mid \texttt{unspecified} \; [\textit{ixs}]
\end{align*}

\section{Static analysis}

\begin{definition}[Induction variable]

\end{definition}

\begin{definition}[Constant translations]
An indexing expression $e$ is a \emph{constant translation} if,
for an \emph{induction variable} $i$, then $e \equiv i + a$ or $e \equiv i - a$ 
where $a$ is a constant. The relation $\equiv$ is used which
represents term equality up-to commutativity of $+$ and the inverse
relation of $+$ and $-$ (\eg{}, $(-b) + i \equiv i - b$). 
In the following, we classify constant translation expressions 
using the predicate $\textsf{trans}$.
\end{definition}

\paragraph{Step 1: Data-access analysis}

\dnote{Might want to do 'stencil' (gather) and 'mould' (scatter)
  patterns separately}
\begin{align*}
\begin{array}{lll}
\interp{e_1 = e_2}    & = \interp{e_2} \\
\interp{e_1 e_2}      & = \interp{e_1} \cup \interp{e_2} \\
\interp{a(\tilde{e})} & = \{e | e \in \tilde{e}, \textsf{trans}(e)\} & \textit{where each $e \in
                                          \tilde{e}$ is a constant or constant translation}
\end{array}
\end{align*}

Index analysis is a \emph{coeffect analysis}.

\paragraph{Step 2: Coalesce contiguous indices into regions}

% fairly mechanical

\section{Evaluation}

\section{Discussion}

\bibliography{references}

\end{document}

### 2016-02-02 paper.tex
change to the spec grammar

@@
 \begin{align*}
-spec ::= & \; S  \; \texttt{,} \; spec \, \mid \, \epsilon \\ \\
+spec ::= & \; S  \; \texttt{,} \; spec \, \mid \, S \\[0.5em]
 %
-S ::= &  \; \texttt{forward} \; [\textit{ixs}] \; [\textit{depth}] \\
-& \; \mid \texttt{backward} \; [\textit{ixs}] \; [\textit{depth}] \\
-& \; \mid \texttt{centered} \; [\textit{ixs}] \; [\textit{depth}] \\
-& \; \mid \texttt{unspecified} \; [\textit{ixs}]
+S ::= &  \; \texttt{forward} \; \textit{depth} \; \textit{dims} \\
+& \; \mid \texttt{backward} \; \textit{depth} \; \textit{dims} \\
+& \; \mid \texttt{centered} \; \textit{depth} \; \textit{dims} \\
+& \; \mid \texttt{unspecified} \; [\textit{dims}] \\
+& \; \mid \texttt{unique} \; S \\[0.5em]
+\textit{depth} ::= & \; \mathbb{N} \\
+\textit{dims}  ::= & \; \textit{var} \; \texttt{,} \; \textit{dims} \,
+                      \mid \, \textit{var}
 \end{align*}

### 2016-02-18 paper.tex
expanding examples with different schemes

@@
 \usepackage{graphics}
+\usepackage{fancyvrb}
+\CustomVerbatimEnvironment{SpecVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.65cm,
+xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.97}
+\CustomVerbatimEnvironment{ExmVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.65cm,
+xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.97}
+\CustomVerbatimEnvironment{IVerbatim}{Verbatim}{fontsize=\relsize{-1},xleftmargin=0.65cm,
+xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.97}
 
@@
 %
-S ::= &  \; \texttt{forward} \; \textit{depth} \; \textit{dims} \\
-& \; \mid \texttt{backward} \; \textit{depth} \; \textit{dims} \\
-& \; \mid \texttt{centered} \; \textit{depth} \; \textit{dims} \\
-& \; \mid \texttt{unspecified} \; [\textit{dims}] \\
-& \; \mid \texttt{unique} \; S \\[0.5em]
+S ::= &  \; \texttt{forward} \; \texttt{depth=}\textit{depth} \; \texttt{dims=}\textit{dims} \\
+& \; \mid \texttt{backward} \; \texttt{depth=}\textit{depth} \; \texttt{dims=}\textit{dims} \\
+& \; \mid \texttt{centered} \; \texttt{depth=}\textit{depth} \; \texttt{dims=}\textit{dims} \\
+& \; \mid \texttt{unspecified} \; \texttt{dims=}[\textit{dims}] \\
+& \; \mid \texttt{read-once} \; S \\[0.5em]
 \textit{depth} ::= & \; \mathbb{N} \\
@@
 
+\subsection{Common patterns}
+
+\begin{example}[FTCS (Forward time, centered space)]
+
+  Otherwise known as the \emph{explicit method}, is a common
+  pattern since it tends to be more efficient than other approches
+  (see BTCS and CTCS below). However, its disadvantage is that it is
+  unstable if the resolution is not well chosen.
+
+\begin{ExmVerbatim}
+b(x) = a(x) + r*(a(x-1) - 2*a(x) + a(x+1))
+\end{ExmVerbatim}
+%
+For this statement, the inference provides the specification: 
+%
+\begin{SpecVerbatim}
+a: centered depth=1, dims=0
+\end{SpecVerbatim}
+
+\end{example}
+
+\paragraph{BTCS (Backward time, centered space) -- implicit method}
+
+\paragraph{CTCS (Centered time, centered space) - Crank-Nicolson method}
+
 \section{Static analysis}
 
-\begin{definition}[Induction variable]
-
-\end{definition}
+In the following, we defined
+\emph{base induction variables} to be the variables
+ defined by a ``for'' loop (\eg{}, that get incremented each time the
+loop is re-entered).
 
@@
 An indexing expression $e$ is a \emph{constant translation} if,
-for an \emph{induction variable} $i$, then $e \equiv i + a$ or $e \equiv i - a$ 
-where $a$ is a constant. The relation $\equiv$ is used which
-represents term equality up-to commutativity of $+$ and the inverse
+for a base induction variable $i$, then $e \equiv i + a$ or $e \equiv i - a$ 
+where $a$ is a constant. The relation $\equiv$ identifies terms
+up-to commutativity of $+$ and the inverse
 relation of $+$ and $-$ (\eg{}, $(-b) + i \equiv i - b$). 

### 2016-03-03 paper.tex
more intro; towards a definition of sub stencils

@@
 \usepackage{fancyvrb}
+\usepackage{subfigure}
+
 \CustomVerbatimEnvironment{SpecVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.65cm,
@@
 \CustomVerbatimEnvironment{ExmVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.65cm,
-xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.97}
+xrightmargin=0.2cm,baselinestretch=0.97,numbers=left}
 \CustomVerbatimEnvironment{IVerbatim}{Verbatim}{fontsize=\relsize{-1},xleftmargin=0.65cm,
@@
 
+\newcommand{\dimId}{\texttt{dim}}
+
 % Semantics related
@@
 
-\title{Abstract Shape Specifications for Stencil Verification}
+\title{Time and space specifications for stencils}
 \authorinfo{}{}
@@
 
-\section{Specification language}
+\emph{Stencils} are a ubiquitous programming pattern in many numerical
+applications. Informally, a stencil computation computes
+an array, where the value at each index $i$ of this array is
+calculated from a set of neighbouring values to $i$ in some input
+array(s), \eg{}, the Game of Life, convolutions in image processing, 
+approximations to differential equations. For example, the following
+psuedo-code iteratively applies the discrete Laplace
+operator (an approximation to a derivative):
+%
+\begin{ExmVerbatim}
+do iter = 0, itermax
+   do i = 1, (n-1)
+      b(i) = a(i-1) - 2*a(i) + a(i+1)
+   a = b
+\end{ExmVerbatim}
+%
+Line 3 is the core of the stencil computation, calculating
+the value at \texttt{b[i]} from a neighbourhood of elements about
+\texttt{a[i]}. Line 4 swaps 
+\texttt{a} and \texttt{b} between iterations, where \texttt{b} becomes the
+input for the next iteration.
 
+%This stencil computation exhibits statically decideable
+%spatial and temporal relationships between $\texttt{a}$ and
+%$\texttt{b}$. 
+In this simple example, the spatial and temporal relationship
+between \texttt{a} and \texttt{b} is easy to understand. However,
+more complex stencil computations can be much harder to understand
+and subsequently more prone to error. For example, Figure~\ref{ref:navier-stokes-fragment}
+shows a couple of lines from a Navier-Stokes fluid simulator in
+which two arrays are read from with different data access patterns,
+across two dimensions. The interaction is much harder to understand,
+with the potential for the developer to accidentally introduce an
+error, for example writing $\texttt{(i-1,j)}$ instead of
+$\texttt{(i+1,j)}$.
+
+We introduce a simple specification for stencil comptuations
+that abstracts the fine grained detail of their access patterns.
+In practise, most stencil computations have a regular shape that
+can be described simply and abstractly. In the case of our
+first Laplace example, our analysis tool infers the following specification:
+%
+\begin{SpecVerbatim}
+a : centered depth=1 dims=1
+b : backward depth=1\{a\} dims=t
+\end{SpecVerbatim}
+%
+The first line explains that \texttt{a} is read from
+with a symmetrical stencil pattern (``centered'') to a depth
+of one in each direction in its first (and only) dimension.
+The second line explains the temporal relationship between
+\texttt{b} and \texttt{a}: that the previous time step for \texttt{b}
+is actually provided by \texttt{a}. The dimension \texttt{t} is a
+reserved dimension identifiers for temporal specifications. In
+the case of the Navier-Stokes example, its inferred stencil is shown
+in Figure~\ref{ref:navier-stokes-fragment}(b).
+
+This paper makes the following contributions:
+%
+\begin{itemize}
+\item introduces a specification language for
+stencil computations that captures many common forms
+of stencil (Section~\ref{sec:lang});
+\item we detail specification inference and checking
+algorithsm (Section~\ref{sec:analysis});
+\item we evaluate our implementation of the approach
+in the CamFort tool for Fortran verification, studying
+a number of example programs to assess the usefulness
+of this approach.
+\dnote{insert results here}
+\end{itemize}
+%
+
+\begin{figure}
+\begin{ExmVerbatim}[firstnumber=20]
+du2dx = ((u(i,j)+u(i+1,j))*(u(i,j)+u(i+1,j))+ 
+    gamma*abs(u(i,j)+u(i+1,j))*(u(i,j)-u(i+1,j))- 
+    (u(i-1,j)+u(i,j))*(u(i-1,j)+u(i,j))- 
+    gamma*abs(u(i-1,j)+u(i,j))*(u(i-1,j)-u(i,j))) 
+    /(4.0*delx)
+
+duvdy = ((v(i,j)+v(i+1,j))*(u(i,j)+u(i,j+1))+ 
+   gamma*abs(v(i,j)+v(i+1,j))*(u(i,j)-u(i,j+1))- 
+   (v(i,j-1)+v(i+1,j-1))*(u(i,j-1)+u(i,j))-  
+   gamma*abs(v(i,j-1)+v(i+1,j-1))*(u(i,j-1)-u(i,j))) 
+   /(4.0*dely)
+
+laplu = (u(i+1,j)-2.0*u(i,j)+u(i-1,j))/delx/delx+ 
+          (u(i,j+1)-2.0*u(i,j)+u(i,j-1))/dely/dely
+\end{ExmVerbatim} 
+(a). Excerpt from the Fortran code, 
+showing highly-detailed stencil computations. \\
+
+\begin{SpecVerbatim}[xleftmargin=0.1cm]
+20-24  u: centered depth=1 dim=1
+26-30  u: centered depth=1 dim=2
+       v: forward depth=1 dim=1, backward depth=1 dim=1
+32-33  u: centered depth=1 dim=1,2
+\end{SpecVerbatim}
+(b). Inferred stencil specification from CamFort
+\caption{Fragment of Navier-Stokes fluid simulator and its specification}
+\label{ref:navier-stokes-fragment}
+\end{figure}
+
+\section{Stencil specification language}
+\label{sec:lang}
+
+\begin{figure}
 \begin{align*}
@@
 %
-S ::= &  \; \texttt{forward} \; \texttt{depth=}\textit{depth} \; \texttt{dims=}\textit{dims} \\
-& \; \mid \texttt{backward} \; \texttt{depth=}\textit{depth} \; \texttt{dims=}\textit{dims} \\
-& \; \mid \texttt{centered} \; \texttt{depth=}\textit{depth} \; \texttt{dims=}\textit{dims} \\
-& \; \mid \texttt{unspecified} \; \texttt{dims=}[\textit{dims}] \\
+S ::= &  \; \texttt{forward} \; [\texttt{depth=}\textit{depth}] \; \texttt{dims=}\textit{dims} \\
+& \; \mid \texttt{backward} \; [\texttt{depth=}\textit{depth}] \; \texttt{dims=}\textit{dims} \\
+& \; \mid \texttt{centered} \; [\texttt{depth=}\textit{depth}] \; \texttt{dims=}\textit{dims} \\
+& \; \mid \texttt{unspecified} \; [\texttt{dims=}\textit{dims}] \\
 & \; \mid \texttt{read-once} \; S \\[0.5em]
-\textit{depth} ::= & \; \mathbb{N} \\
-\textit{dims}  ::= & \; \textit{var} \; \texttt{,} \; \textit{dims} \,
-                      \mid \, \textit{var}
+\textit{depth} ::= & \; \mathbb{N} \, \mid \, \textit{temp-spec} \\ 
+\textit{temp-spec} ::= & \mathbb{N}\{var\} \\
+\textit{dims}  ::= & \;\mathbb{N}_{>0} \; \texttt{,} \; \textit{dims} \,
+                      \mid \, \mathbb{N}_{>0} \, \mid \, 
+                     \texttt{t}
 \end{align*}
+\caption{Syntax of specifications}
+\end{figure}
+
+\begin{figure}
+\begin{align*}
+(\texttt{a} : \; \texttt{bwd} \; \texttt{depth=1\{b\}} \;
+  \texttt{dim=t})
+\, & \equiv \,
+(\texttt{b} : \; \texttt{bwd} \; \texttt{depth=1\{a\}} \;
+  \texttt{dim=t})
+\\ 
+(\texttt{a} : \; \texttt{fwd} \; \texttt{depth=1\{b\}} \;
+  \texttt{dim=t})
+\, & \equiv \,
+(\texttt{b} : \; \texttt{fwd} \; \texttt{depth=1\{a\}} \;
+  \texttt{dim=t})
+\\
+(a : \texttt{read-once} \; S) \; & \equiv \; (a : \texttt{read-once} \;\,
+  \texttt{read-once} \; \, S)
+\end{align*}
+\caption{Equations on specifications}
+\end{figure}
+
+\subsection{Time specifications}
+
+\begin{ExmVerbatim}
+for t = 1, itermax:
+   for i = 1, (n-1):
+      a[t, i] = a[t-1, i-1] - 2*a[t-1, i] + a[t-1, i+1]
+\end{ExmVerbatim}
+
+\begin{SpecVerbatim}
+a(t=0) : bwd, depth=1
+a(1)   : centered, depth=1
+\end{SpecVerbatim}
+
+\begin{SpecVerbatim}
+a : centered depth=1 dim=1
+a : backward depth=1 dim=0
+\end{SpecVerbatim}
+
+\subsection{Safety and sub-specifications}
+
+\begin{figure}[t]
+{\framebox{
+\begin{minipage}{0.88\linewidth}
+\begin{align*}
+\hspace{-0.7em}
+\begin{array}{c}
+\dfrac{}{S <: S}(\textsc{refl}) \qquad \dfrac{R <: S \quad S <: T}{R <:
+  T}(\textsc{trans}) \\[1.5em]
+\dfrac{}{S <: \; \texttt{unspecified dims=}ds}(\top) 
+\\[1.5em]
+\dfrac{}{\texttt{read-once} \, S <: S}(\textsc{rep})
+\\[1.5em]
+\setlength{\arraycolsep}{0.1em}
+\dfrac{\hspace{3em} ds \subseteq es \; \wedge \; n \leq m \hspace{3em}}
+{\begin{array}{rl}
+\texttt{fwd} \; \texttt{depth=}n \; \texttt{dim=}ds & <: \texttt{fwd} \;
+  \texttt{depth=}m \; \texttt{dim=}es 
+%& \;\, ds \subseteq es \wedge n \leq m 
+\\
+\wedge \; \texttt{bwd} \; \texttt{depth=}n \; \texttt{dim=}ds & <: \texttt{bwd} \;
+  \texttt{depth=}m \; \texttt{dim=}es 
+%& \;\, ds \subseteq es \wedge n \leq m 
+\\
+\wedge \; \texttt{fwd} \; \texttt{depth=}n \; \texttt{dim=}ds & <: \texttt{sym} \;
+  \texttt{depth=}m \; \texttt{dim=}es 
+%& \;\, ds \subseteq es \wedge n \leq m 
+\\
+\wedge \; \texttt{bwd} \; \texttt{depth=}n \; \texttt{dim=}ds & <: \texttt{sym} \;
+  \texttt{depth=}m \; \texttt{dim=}es 
+\end{array}}{\footnotesize{(\textsc{cover})}}\hspace{-0.1em}
+\end{array}
+\end{align*}
+\end{minipage}}}
+\caption{Definition of sub-specification relation}
+\end{figure}
 
@@
 
-\section{Static analysis}
+\section{Specification inference and checking}
+\label{sec:analysis}
 
@@
 
+\subsection{Infering temporal specifications}
+
+The reserved dimension \dimId{t}
+
 \section{Evaluation}

### 2016-03-09 paper.tex
Concerns about operator overloading and sliding in Fortran

@@
 \newcommand{\dnote}[1]{\textcolor{darkpurple}{Dom: #1}}
+\newcommand{\mnote}[1]{\textcolor{darkgreen}{Mistral: #1}}
 
@@
 
+\mnote{Few cases that we might like to consider (there are examples
+    of each in \#camfort-main Slack channel): 
+  \begin{itemize}
+    \item Matrix operator overloading $a + b$, where both are arrays
+    \item Scalar operator overloading $a + b$, where a is an array b is a 
+      scalar. $b$ is added to all.
+    \item Sliding. $a = b$ where the dimensionsare $a(1:10)$, $b(2:11)$,
+      so the assignment causes shifting.
+  \end{itemize}
+}
+
 In the following, we defined

### 2016-03-10 paper.tex
working on description of equality and inequality
work on the multi-set model (missing temporal specs); and start of soundness proof

@@
 \usepackage{subfigure}
+\usepackage{amsthm}
 
@@
 
+\newcommand{\bn}{\mathbb{N}}
+
 \newcommand{\dnote}[1]{\textcolor{darkpurple}{Dom: #1}}
@@
 
-\newtheorem{theorem}[block]{Theorem}
 \newtheorem{lemma}[block]{Lemma}
@@
 \theoremstyle{definition}
+
+\newtheorem{theorem}[block]{Theorem} 
 \newtheorem{remark}[block]{Remark}
@@
 
-This paper makes the following contributions:
+In this paper, we make the following contributions:
 %
 \begin{itemize}
-\item introduces a specification language for
+\item we introduce a specification language for
 stencil computations that captures many common forms
-of stencil (Section~\ref{sec:lang});
-\item we detail specification inference and checking
-algorithsm (Section~\ref{sec:analysis});
+of data access pattern, both spatial and temporal (Section~\ref{sec:lang});
+\item we detail inference and checking
+algorithms for stencil specifications (Section~\ref{sec:analysis});
 \item we evaluate our implementation of the approach
@@
 
+Our specification system is based on the observation
+that most forms of array access in numberical code have 
+a fixed, statically-determined access pattern. For example, the 
+``\emph{five-point stencil}'' on a two-dimensional array reads from array
+indices $(i, j)$, $(i-1, j)$, $(i+1, j)$, $(i, j-1)$, and $(i, j+1)$
+for all $i, j$ within the inner boundary of the array (to avoid 
+out-of-bounds access at the edges). We revisit this hypothesis 
+in Section~\ref{sec:evaluation} where the inference of
+such regular stencil patterns on a corpus of numerical programs (both
+small and large). We found that indeed \dnote{..}.
+
+We outline the specification language here. Section~\ref{sec:syntax}
+oultines the syntax. Section~\ref{sec:semantics} defines its semantics
+via a simple multi-set interpretation over indices. Section~\ref{sec:eqs} provides an
+equational theory for specifications via relation ($\equiv$) and a
+theory of approximation via relation ($<:$). These equations are then
+proven sound with respecto the multi-set semantics of Section~\ref{sec:semantics}.
+
+\subsection{Syntax}
+\label{sec:syntax}
+
+Figure~\ref{fig:syntax}
+gives the syntax, which we introduces in stages below. 
+%%
+
 \begin{figure}
 \begin{align*}
-spec ::= & \; S  \; \texttt{,} \; spec \, \mid \, S \\[0.5em]
+\textit{specs} ::= & \; \textit{spec} \; \texttt{;} \; \textit{specs} \, \mid \, \textit{spec} \\
 %
+\textit{spec} :: = & \; \texttt{v} : S \\[0.5em]
 S ::= &  \; \texttt{forward} \; [\texttt{depth=}\textit{depth}] \; \texttt{dims=}\textit{dims} \\
@@
 \textit{depth} ::= & \; \mathbb{N} \, \mid \, \textit{temp-spec} \\ 
-\textit{temp-spec} ::= & \mathbb{N}\{var\} \\
-\textit{dims}  ::= & \;\mathbb{N}_{>0} \; \texttt{,} \; \textit{dims} \,
-                      \mid \, \mathbb{N}_{>0} \, \mid \, 
-                     \texttt{t}
+\textit{temp-spec} ::= & \; \mathbb{N}\{var\} \\
+\textit{dims} ::= & \; \mathbb{D} \; \texttt{,} \; \textit{dims} \, \mid
+                    \, \mathbb{D} \\
+\mathbb{D}  ::= & \; \mathbb{N}_{>0} \, \mid \, \texttt{t}
 \end{align*}
 \caption{Syntax of specifications}
+\label{fig:syntax}
 \end{figure}
 
-\begin{figure}
-\begin{align*}
-(\texttt{a} : \; \texttt{bwd} \; \texttt{depth=1\{b\}} \;
-  \texttt{dim=t})
-\, & \equiv \,
-(\texttt{b} : \; \texttt{bwd} \; \texttt{depth=1\{a\}} \;
-  \texttt{dim=t})
-\\ 
-(\texttt{a} : \; \texttt{fwd} \; \texttt{depth=1\{b\}} \;
-  \texttt{dim=t})
-\, & \equiv \,
-(\texttt{b} : \; \texttt{fwd} \; \texttt{depth=1\{a\}} \;
-  \texttt{dim=t})
-\\
-(a : \texttt{read-once} \; S) \; & \equiv \; (a : \texttt{read-once} \;\,
-  \texttt{read-once} \; \, S)
-\end{align*}
-\caption{Equations on specifications}
-\end{figure}
+\paragraph{Spatial specifications}
 
-\subsection{Time specifications}
+\paragraph{Time specifications}
+\dnote{The below is old so probably out of date}
 
@@
 
-\subsection{Safety and sub-specifications}
+\subsection{A semantic model for specifications}
+\label{sec:semantics}
+
+We give a model of our specifications which explains
+the range of indices which correspond to the specifications.
+The model is based on multi-sets since array indices my be repeated
+in an expression.
+
+Figure~\ref{fig:model} describes the model in terms of 
+an interpretation function $\interp{-}$ which maps specifications
+to multisets of relative indices.
+
+\begin{definition}Relative indices are drawn
+from $(\mathbb{Z}_{\bot})^{\mathbb{D}}$, essentially an infinite
+tuple of offsets. For example, $(i, j+1)$ corresponds to
+the relative index $(0, 1, \bot, \bot, \ldots)$
+\end{definition}
+
+The main interpretation function is overloaded on lists of 
+variable-spec pairs $\interp{\overline{v : S}}$, 
+returning multisets of variable-relative-index pairs. This provides
+the top-level definition of the model, with $\interp{-}$ overloaded
+on $S$, $\interp{S}$ mqpping to multisets of
+relative indices not associated to an array avariable.
+
+\begin{figure}
+\begin{align*}
+\hspace{-2em}
+\begin{array}{c}
+\hspace{-1em}
+\setlength{\arraycolsep}{0.1em}
+\begin{array}{rl}
+\interp{\textit{spec}; \textit{specs}} & = \interp{spec} \cup
+  \interp{\textit{specs}} \\
+%% TEMPORAL SPECS
+% \interp{\texttt{v} : \texttt{forward} \; \texttt{depth}=n\texttt{\{u\}} \;
+%  \texttt{dims=t}} & =  \{ (\texttt{u}, \textit{ix})^m \mid \forall m,
+%                     ix : \mathbb{Z}^{\mathbb{D}}, \textit{ix} \,
+%                     \texttt{t} = 0 \} \\[0.4em]
+%% SPATIAL SPECS
+\interp{\texttt{v} : S} & = \{(\texttt{v}, ix)^m \mid ix^m \in
+  \interp{S}\} \\
+\end{array}\\[1em]
+%\framebox{$\interp{S}$} \\[1em]
+\begin{array}{ll}
+& \interp{\texttt{reflexive}}
+  = \\
+& \hspace{0.9em}  \{ ix^m \mid \forall m, \, ix : \mathbb{Z}^{\mathbb{D}}, d : \mathbb{D}
+  \, . \, d \in \textit{ds} \Rightarrow ix \, d = 0 \} \\[0.4em]
+& \interp{\texttt{forward} \;\; \texttt{depth}=n \;\;
+  \texttt{dims}=\textit{ds}}  = \;\; (\texttt{t} \not\in ds) \\
+& \hspace{0.9em}  \{ ix^m \mid \forall m, \, ix : \mathbb{Z}^{\mathbb{D}}, d : \mathbb{D}
+  \, . \, d \in \textit{ds} \Rightarrow 0 \leq (ix \, d) \leq n \} \\[0.4em]
+& \interp{\texttt{backward} \;\; \texttt{depth}=n \;\; \texttt{dims}=\textit{ds}}
+  = \; \; (\texttt{t} \not\in ds) \\
+& \hspace{0.9em}  \{ ix^m \mid \forall m, \, ix : \mathbb{Z}^{\mathbb{D}}, d : \mathbb{D}
+  \, . \, d \in \textit{ds} \Rightarrow (-n) \leq (ix \, d) \leq 0 \}
+  \\[0.4em]
+& \interp{\texttt{read-once} \; S} = \{ ix^1 \mid ix^m \in \interp{S}
+  \} 
+\end{array}
+\end{array}
+\end{align*}
+\textit{$a^m \in A$ means there are $n$ copies of $a$ in
+  the multi-set $A$}
+\caption{Multi-set model of specifications}
+\label{fig:model}
+\end{figure}
+
+\subsection{Equational theory and approximations}
+\label{sec:eqs}
+
+Figure~\ref{fig:equations}
+
+\begin{figure}
+\begin{align*}
+\hspace{-1em}
+\setlength{\arraycolsep}{0.05em}
+\begin{array}{c}
+\framebox{$\overline{\texttt{v} : S} \equiv \overline{\texttt{u} : T}$} \\[1em]
+\begin{array}{rl}
+(\textsc{coalesceF}) \;\; & (\texttt{a} : \texttt{fwd} \;
+                       \texttt{dims}=ds \; \texttt{depth}=n; \\
+&  \;\, \texttt{a} :  \texttt{fwd} \; \texttt{dims}=ds \; \texttt{depth}=n+1)\\
+\equiv \; & \texttt{a} :  \texttt{fwd} \; \texttt{depth}=n+1,
+         \texttt{dims}=ds \\[0.75em]
+(\textsc{coalesceB}) \;\; & (\texttt{a} : \texttt{bwd} \;
+                       \texttt{dims}=ds \; \texttt{depth}=n; \\
+&  \;\, \texttt{a} :  \texttt{bwd} \; \texttt{dims}=ds \; \texttt{depth}=n+1) \\
+\equiv \; & \texttt{a} :  \texttt{bwd} \; \texttt{depth}=n+1,
+         \texttt{dims}=ds \\[0.75em]
+(\textsc{idem}) \;\; & (\texttt{a} : S; \texttt{a} : S) \equiv (\texttt{a}
+                  : S)
+\end{array} \\ \\
+\framebox{$\texttt{v} : S \equiv \texttt{v'} : S'$} \\[1em]
+\begin{array}{ll}
+(\texttt{a} : \; \texttt{bwd} \; \texttt{depth=1\{b\}} \;
+  \texttt{dim=t})
+\, & \equiv \,
+(\texttt{b} : \; \texttt{bwd} \; \texttt{depth=1\{a\}} \;
+  \texttt{dim=t})
+\\ 
+(\texttt{a} : \; \texttt{fwd} \; \texttt{depth=1\{b\}}\;
+  \texttt{dim=t})
+\, & \equiv \,
+(\texttt{b} : \; \texttt{fwd} \; \texttt{depth=1\{a\}} \;
+  \texttt{dim=t})
+\end{array}
+\\ \\
+\framebox{$S \equiv S'$} \\[0.75em]
+\begin{array}{ll}
+(a : \texttt{read-once} \; S) \; & \equiv \; (a : \texttt{read-once} \;\,
+  \texttt{read-once} \; \, S)
+\end{array}
+\end{array}
+\end{align*}
+\caption{Equations on specifications}
+\label{fig:equations}
+\end{figure}
+
+\begin{theorem}[Soundness]
+\[
+\overline{\texttt{v} : S}\equiv \overline{\texttt{u} : T}
+\; \Rightarrow \;
+\interp{\overline{\texttt{v} : S}} = \interp{\overline{\texttt{u} : T}}
+\]
+\end{theorem}
+
+\paragraph{Proof} (see Appendix)
+
+\subsubsection{Safety and sub-specifications}
+
+Figure~\ref{fig:inequations}
 
@@
   \texttt{depth=}m \; \texttt{dim=}es 
-\end{array}}{\footnotesize{(\textsc{cover})}}\hspace{-0.1em}
+\end{array}}
+%{\footnotesize{(\textsc{cover})}}
+%\hspace{-0.1em}
 \end{array}
@@
 \caption{Definition of sub-specification relation}
+\label{fig:inequations}
 \end{figure}
 
-\subsection{Common patterns}
+\begin{theorem}[Soundness]
+\[
+\overline{\texttt{v} : S} <: \overline{\texttt{u} : T}
+\; \Rightarrow \;
+\interp{\overline{\texttt{v} : S}} \subseteq \interp{\overline{\texttt{u} : T}}
+\]
+\end{theorem}
+
+\paragraph{Proof} (see Appendix)
+
+\section{Examples and common patterns}
 
@@
 \section{Evaluation}
+\label{sec:evaluation}
 
 \section{Discussion}
+\label{sec:discussion}
 
@@
 
+\onecolumn
+\appendix
+
+\section{Correctness}
+
+\begin{theorem}[Soundness]
+\[
+\overline{\texttt{v} : S}\equiv \overline{\texttt{u} : T}
+\; \Rightarrow \;
+\interp{\overline{\texttt{v} : S}} = \interp{\overline{\texttt{u} : T}}
+\]
+\end{theorem}
+
+\paragraph{Proof}
+\begin{itemize}
+\item \textsc{coalesceF}
+\begin{align*}
+\llbracket & \texttt{v} : \texttt{fwd} \; \texttt{dims}=ds \; \texttt{depth}=n; \,
+  \texttt{v} :  \texttt{fwd} \; \texttt{dims}=ds \;
+  \texttt{depth}=n+1)\rrbracket \\
+& =  \{ (\texttt{v}, ix)^m \mid \forall \, m, ix : \mathbb{Z}^{\mathbb{N}}, d : \mathbb{N}
+  \, . \, d \in \textit{ds} \Rightarrow ix \, d \leq (n+1) \} \; \cup \; \{ ix^m \mid \forall \, m, ix : \mathbb{Z}^{\mathbb{N}}, d : \mathbb{N}
+  \, . \, d \in \textit{ds} \Rightarrow ix \, d \leq n \}\\
+& =  \{ (\texttt{v}, ix)^m \mid \forall \, m, ix : \mathbb{Z}^{\mathbb{N}}, d : \mathbb{N}
+  \, . \, d \in \textit{ds} \Rightarrow ((ix \, d \leq (n+1)) \vee (ix \,
+  d \leq n)) \} \\
+& =  \{ (\texttt{v}, ix)^m \mid \forall \, m, ix : \mathbb{Z}^{\mathbb{N}}, d : \mathbb{N}
+  \, . \, d \in \textit{ds} \Rightarrow ix \, d \leq (n+1) \} \\
+& = \interp{\texttt{v} : \texttt{fwd} \; \texttt{dims}=ds \;
+  \texttt{depth}=n + 1} \qquad \Box
+\end{align*}
+\end{itemize}
+
 \end{document}

### 2016-03-10 paper.tex
corrected type of relative indices

@@
 
+\newcommand{\relix}{(\mathbb{Z}_\bot)^\mathbb{D}}
+
 We give a model of our specifications which explains
@@
   = \\
-& \hspace{0.9em}  \{ ix^m \mid \forall m, \, ix : \mathbb{Z}^{\mathbb{D}}, d : \mathbb{D}
+& \hspace{0.9em}  \{ ix^m \mid \forall m, \, ix : \relix, d : \mathbb{D}
   \, . \, d \in \textit{ds} \Rightarrow ix \, d = 0 \} \\[0.4em]
@@
   \texttt{dims}=\textit{ds}}  = \;\; (\texttt{t} \not\in ds) \\
-& \hspace{0.9em}  \{ ix^m \mid \forall m, \, ix : \mathbb{Z}^{\mathbb{D}}, d : \mathbb{D}
+& \hspace{0.9em}  \{ ix^m \mid \forall m, \, ix : \relix, d : \mathbb{D}
   \, . \, d \in \textit{ds} \Rightarrow 0 \leq (ix \, d) \leq n \} \\[0.4em]
@@
   = \; \; (\texttt{t} \not\in ds) \\
-& \hspace{0.9em}  \{ ix^m \mid \forall m, \, ix : \mathbb{Z}^{\mathbb{D}}, d : \mathbb{D}
+& \hspace{0.9em}  \{ ix^m \mid \forall m, \, ix : \relix, d : \mathbb{D}
   \, . \, d \in \textit{ds} \Rightarrow (-n) \leq (ix \, d) \leq 0 \}
@@
   \texttt{depth}=n+1)\rrbracket \\
-& =  \{ (\texttt{v}, ix)^m \mid \forall \, m, ix : \mathbb{Z}^{\mathbb{N}}, d : \mathbb{N}
+& =  \{ (\texttt{v}, ix)^m \mid \forall \, m, ix : \relix, d : \mathbb{N}
   \, . \, d \in \textit{ds} \Rightarrow ix \, d \leq (n+1) \} \; \cup \; \{ ix^m \mid \forall \, m, ix : \mathbb{Z}^{\mathbb{N}}, d : \mathbb{N}
   \, . \, d \in \textit{ds} \Rightarrow ix \, d \leq n \}\\
-& =  \{ (\texttt{v}, ix)^m \mid \forall \, m, ix : \mathbb{Z}^{\mathbb{N}}, d : \mathbb{N}
+& =  \{ (\texttt{v}, ix)^m \mid \forall \, m, ix : \relix, d : \mathbb{N}
   \, . \, d \in \textit{ds} \Rightarrow ((ix \, d \leq (n+1)) \vee (ix \,
   d \leq n)) \} \\
-& =  \{ (\texttt{v}, ix)^m \mid \forall \, m, ix : \mathbb{Z}^{\mathbb{N}}, d : \mathbb{N}
+& =  \{ (\texttt{v}, ix)^m \mid \forall \, m, ix : \relix, d : \mathbb{N}
   \, . \, d \in \textit{ds} \Rightarrow ix \, d \leq (n+1) \} \\

### 2016-03-17 paper.tex
product of stencils syntax

@@
 \textit{spec} :: = & \; \texttt{v} : S \\[0.5em]
-S ::= &  \; \texttt{forward} \; [\texttt{depth=}\textit{depth}] \; \texttt{dims=}\textit{dims} \\
-& \; \mid \texttt{backward} \; [\texttt{depth=}\textit{depth}] \; \texttt{dims=}\textit{dims} \\
-& \; \mid \texttt{centered} \; [\texttt{depth=}\textit{depth}] \; \texttt{dims=}\textit{dims} \\
-& \; \mid \texttt{unspecified} \; [\texttt{dims=}\textit{dims}] \\
+S ::= &  \; \texttt{forward}, \; [\texttt{depth=}\textit{depth}], \; \texttt{dims=}\textit{dims} \\
+& \; \mid \texttt{backward}, \; [\texttt{depth=}\textit{depth}], \; \texttt{dims=}\textit{dims} \\
+& \; \mid \texttt{centered}, \; [\texttt{depth=}\textit{depth}], \; \texttt{dims=}\textit{dims} \\
+& \; \mid \texttt{unspecified}, \; [\texttt{dims=}\textit{dims}] \\
+& \; \mid \textit{spec} \; \texttt{\&} \; \textit{spec} \\
 & \; \mid \texttt{read-once} \; S \\[0.5em]
@@
 \end{align*}
-\caption{Syntax of specifications}
+\caption{Specification syntax}
 \label{fig:syntax}

### 2016-05-17 paper.tex
new grammar proposal

--- 2016-03-10 paper.tex
@@
-\documentclass[10pt]{sigplanconf}
+\documentclass[9pt]{sigplanconf}
 
@@
 \begin{figure}
+\dnote{NEW grammar}
+\begin{align*}
+S_\odot ::= \; & \texttt{forward} \; [,\texttt{depth=}\mathbb{N}] \; 
+        , \texttt{dims}=\textit{dims}_\odot  \\
+\mid \; & \texttt{backward} \; [,\texttt{depth=}\mathbb{N}] \; , \texttt{dims=}\textit{dims}_\odot \\
+\mid \; & \texttt{centered} \; [,\texttt{depth=}\mathbb{N}] \; , \texttt{dims=}\textit{dims}_\odot \\ \\
+\textit{prodS} ::= \; & S_{\texttt{*}}  \; \texttt{*} \; Sprod \mid \epsilon \\
+\textit{sumS} ::= \; & \textit{prodS} \; \texttt{+} \; \textit{sumS} \mid S_{\texttt{+}} \; \texttt{+}
+          \; \textit{sumS} \mid \epsilon \\ 
+\textit{modS} ::= \; & 
+  [\texttt{read-once}] \\
+  & [, \texttt{reflexive} \; \texttt{dims=}\textit{dims}{\texttt{,}}] \\
+  & [, \texttt{irreflexive} \;
+    \texttt{dims=}\textit{dims}{\texttt{,}}] \; \textit{sumS} \\ 
+\textit{temporal} ::= \; & \ldots \\ 
+\textit{spec} ::= \; & \texttt{v :} \; [\textit{modS}] \; [\textit{temporal}]   \\ \\
+\textit{dims}_\odot ::= & \; \mathbb{D} \; \odot \; \textit{dims}_{\odot} \mid \mathbb{D}
+%
+%dims = dims+ | dims*
+%dims+ = D + dims+ | dims+
+%dims* = D * dims* | dims*%
+%
+%Smod = Linear Spa | NonLinear Spa
+\end{align*}
+
+\dnote{OLD grammar}
 \begin{align*}
@@
 \textit{spec} :: = & \; \texttt{v} : S \\[0.5em]
-S ::= &  \; \texttt{forward} \; [\texttt{depth=}\textit{depth}] \; \texttt{dims=}\textit{dims} \\
-& \; \mid \texttt{backward} \; [\texttt{depth=}\textit{depth}] \; \texttt{dims=}\textit{dims} \\
-& \; \mid \texttt{centered} \; [\texttt{depth=}\textit{depth}] \; \texttt{dims=}\textit{dims} \\
-& \; \mid \texttt{unspecified} \; [\texttt{dims=}\textit{dims}] \\
+S ::= &  \; \texttt{forward}, \; [\texttt{depth=}\textit{depth}], \; \texttt{dims=}\textit{dims} \\
+& \; \mid \texttt{backward}, \; [\texttt{depth=}\textit{depth}], \; \texttt{dims=}\textit{dims} \\
+& \; \mid \texttt{centered}, \; [\texttt{depth=}\textit{depth}], \; \texttt{dims=}\textit{dims} \\
+& \; \mid \texttt{unspecified}, \; [\texttt{dims=}\textit{dims}] \\
+& \; \mid \texttt{reflexive}\, \; 
+& \; \mid \texttt{irreflexive}, \; \textit{spec}
+& \; \mid \textit{spec} \; \texttt{\&} \; \textit{spec} \\
 & \; \mid \texttt{read-once} \; S \\[0.5em]

### 2016-05-20 paper.tex
revised DNF grammar; prior to removing it

--- 2016-03-10 paper.tex
@@
 \newcommand{\interp}[1]{\llbracket{#1}\rrbracket}
+
+% Syntax macros
+\newcommand{\stenFwd}[2]{\texttt{forward} (\texttt{depth}=#1,
+  \texttt{dim}=#2)}
+\newcommand{\stenBwd}[2]{\texttt{backward} (\texttt{depth}=#1,
+  \texttt{dim}=#2)}
+\newcommand{\stenCen}[2]{\texttt{backward} (\texttt{depth}=#1,
+  \texttt{dim}=#2)}
 
@@
 
-\begin{SpecVerbatim}[xleftmargin=0.1cm]
-20-24  u: centered depth=1 dim=1
-26-30  u: centered depth=1 dim=2
-       v: forward depth=1 dim=1, backward depth=1 dim=1
-32-33  u: centered depth=1 dim=1,2
+%\begin{SpecVerbatim}[xleftmargin=0.1cm]
+%20-24  u: centered depth=1 dim=1
+%6-30  u: centered depth=1 dim=2
+%      v: forward depth=1 dim=1, backward depth=1 dim=1
+%32-33  u: centered depth=1 dim=1,2
+%\end{SpecVerbatim}
+
+\begin{SpecVerbatim}[xleftmargin=0.3cm]
+%= stencil (centered, depth=1, dim=1)
+%=       + (centered, depth=1, dim=2) :: u
+
+%= stencil (forward, depth=1, dim=1)
+%=       + (backward, depth=1, dim=1) :: v
 \end{SpecVerbatim}
@@
 gives the syntax, which we introduces in stages below.
+The top-level is given by the \textit{spec} production
+which splits into either a \textit{spatial} specification or
+a \textit{temporal specification}.
 %%
 
-\begin{figure}
+\begin{figure}[t]
 \begin{align*}
-\textit{specs} ::= & \; \textit{spec} \; \texttt{;} \; \textit{specs} \, \mid \, \textit{spec} \\
-%
-\textit{spec} :: = & \; \texttt{v} : S \\[0.5em]
-S ::= &  \; \texttt{forward} \; [\texttt{depth=}\textit{depth}] \; \texttt{dims=}\textit{dims} \\
-& \; \mid \texttt{backward} \; [\texttt{depth=}\textit{depth}] \; \texttt{dims=}\textit{dims} \\
-& \; \mid \texttt{centered} \; [\texttt{depth=}\textit{depth}] \; \texttt{dims=}\textit{dims} \\
-& \; \mid \texttt{unspecified} \; [\texttt{dims=}\textit{dims}] \\
-& \; \mid \texttt{read-once} \; S \\[0.5em]
-\textit{depth} ::= & \; \mathbb{N} \, \mid \, \textit{temp-spec} \\ 
-\textit{temp-spec} ::= & \; \mathbb{N}\{var\} \\
-\textit{dims} ::= & \; \mathbb{D} \; \texttt{,} \; \textit{dims} \, \mid
-                    \, \mathbb{D} \\
-\mathbb{D}  ::= & \; \mathbb{N}_{>0} \, \mid \, \texttt{t}
+\def\arraystretch{1.2}
+\begin{array}{rl}
+\textit{spec} ::= \; & \textit{spatial} \mid \textit{temporal}   \\ \\
+\textit{spatial} ::= \; & [\textit{mod} \; \{, \textit{mod} \}] \; \textit{sumS} \\
+\textit{mod} ::= \; &
+ \texttt{read-once} \\
+& \mid \texttt{reflexive} \; \texttt{dim=}\textit{dim}{\texttt{,}} \\
+& \mid \texttt{irreflexive} \; \texttt{dim=}\textit{dim}{\texttt{,}} \\
+\textit{sumS} ::= \; & \textit{prodS} \;\, \texttt{+} \;\, \textit{sumS} \mid S_{\texttt{+}} \; \texttt{+}
+          \; \textit{sumS} \mid \epsilon \\
+\textit{prodS} ::= \; & S_{\texttt{*}}  \;\, \texttt{*} \;\, prodS
+                        \mid \epsilon \\ \\
+\textit{temporal} ::= \; & \texttt{dependency} \; (v \; \{ , v \})  \\ \\
+\multicolumn{2}{l}{(\textit{meta rules, where \textnormal{$\odot = \{\texttt{*}, \texttt{+}, \texttt{,}\}$}})} \\
+S_\odot ::= \; & \texttt{forward} \; [,\texttt{depth=}\mathbb{N}] \;
+        , \texttt{dims=}\textit{dims}_\odot  \\
+\mid \; & \texttt{backward} \; [,\texttt{depth=}\mathbb{N}] \; , \texttt{dims=}\textit{dims}_\odot \\
+\mid \; & \texttt{centered} \; [,\texttt{depth=}\mathbb{N}] \; ,
+          \texttt{dims=}\textit{dims}_\odot \\
+\textit{dims}_\odot ::= & \; \mathbb{N}_{>0} \; \odot \;
+                          \textit{dims}_{\odot} \; \mid \;
+                          \mathbb{N}_{> 0}
+\end{array}
 \end{align*}
-\caption{Syntax of specifications}
+\caption{Specification syntax}
 \label{fig:syntax}
@@
 \begin{figure}
-\begin{align*}
-\hspace{-2em}
-\begin{array}{c}
-\hspace{-1em}
-\setlength{\arraycolsep}{0.1em}
-\begin{array}{rl}
-\interp{\textit{spec}; \textit{specs}} & = \interp{spec} \cup
-  \interp{\textit{specs}} \\
-%% TEMPORAL SPECS
-% \interp{\texttt{v} : \texttt{forward} \; \texttt{depth}=n\texttt{\{u\}} \;
-%  \texttt{dims=t}} & =  \{ (\texttt{u}, \textit{ix})^m \mid \forall m,
-%                     ix : \mathbb{Z}^{\mathbb{D}}, \textit{ix} \,
-%                     \texttt{t} = 0 \} \\[0.4em]
-%% SPATIAL SPECS
-\interp{\texttt{v} : S} & = \{(\texttt{v}, ix)^m \mid ix^m \in
-  \interp{S}\} \\
-\end{array}\\[1em]
-%\framebox{$\interp{S}$} \\[1em]
-\begin{array}{ll}
-& \interp{\texttt{reflexive}}
-  = \\
-& \hspace{0.9em}  \{ ix^m \mid \forall m, \, ix : \relix, d : \mathbb{D}
-  \, . \, d \in \textit{ds} \Rightarrow ix \, d = 0 \} \\[0.4em]
-& \interp{\texttt{forward} \;\; \texttt{depth}=n \;\;
-  \texttt{dims}=\textit{ds}}  = \;\; (\texttt{t} \not\in ds) \\
-& \hspace{0.9em}  \{ ix^m \mid \forall m, \, ix : \relix, d : \mathbb{D}
-  \, . \, d \in \textit{ds} \Rightarrow 0 \leq (ix \, d) \leq n \} \\[0.4em]
-& \interp{\texttt{backward} \;\; \texttt{depth}=n \;\; \texttt{dims}=\textit{ds}}
-  = \; \; (\texttt{t} \not\in ds) \\
-& \hspace{0.9em}  \{ ix^m \mid \forall m, \, ix : \relix, d : \mathbb{D}
-  \, . \, d \in \textit{ds} \Rightarrow (-n) \leq (ix \, d) \leq 0 \}
-  \\[0.4em]
-& \interp{\texttt{read-once} \; S} = \{ ix^1 \mid ix^m \in \interp{S}
-  \} 
-\end{array}
-\end{array}
-\end{align*}
-\textit{$a^m \in A$ means there are $n$ copies of $a$ in
-  the multi-set $A$}
+\input{model.tex}
+%\textit{$a^m \in A$ means there are $n$ copies of $a$ in
+%  the multi-set $A$}
 \caption{Multi-set model of specifications}

### 2016-05-24 paper.tex
preliminary model def

@@
 \newtheorem{proposition}[block]{Proposition}
-%\newtheorem{definition}[block]{Definition}
+%\newtheorem{definition}[block]{Definisssstion}
 
@@
 % Syntax macros
-\newcommand{\stenFwd}[2]{\texttt{forward} (\texttt{depth}=#1,
-  \texttt{dim}=#2)}
-\newcommand{\stenBwd}[2]{\texttt{backward} (\texttt{depth}=#1,
-  \texttt{dim}=#2)}
-\newcommand{\stenCen}[2]{\texttt{backward} (\texttt{depth}=#1,
-  \texttt{dim}=#2)}
+\newcommand{\stenFwd}[2]{\texttt{forward}, \texttt{depth}=#1,
+  \texttt{dim}=#2}
+\newcommand{\stenBwd}[2]{\texttt{backward}, \texttt{depth}=#1,
+  \texttt{dim}=#2}
+\newcommand{\stenCen}[2]{\texttt{backward}, \texttt{depth}=#1,
+  \texttt{dim}=#2}
 
@@
 \begin{SpecVerbatim}
-a : centered depth=1 dims=1
-b : backward depth=1\{a\} dims=t
+%=  stencil centered, depth=1 dims=1 :: a
+%=  stencil dependency a, mutual    :: b
 \end{SpecVerbatim}
@@
 \def\arraystretch{1.2}
+\setlength{\arraycolsep}{0.2em}
+\newcommand{\dimTy}{\mathbb{D}}
 \begin{array}{rl}
-\textit{spec} ::= \; & \textit{spatial} \mid \textit{temporal}   \\ \\
-\textit{spatial} ::= \; & [\textit{mod} \; \{, \textit{mod} \}] \; \textit{sumS} \\
-\textit{mod} ::= \; &
+\textit{specification} ::= & [\textit{regionDec} \; \{\texttt{$\backslash$n} \;\;
+                             \textit{regionDec} \}] \;
+                             \textit{specDec} \\
+\textit{specDec} ::= & \texttt{stencil} \; \textit{spec} \mid \epsilon
+  \\
+\textit{regionDec} ::= &  \texttt{region} \; \textit{rvar} \; \texttt{=} \;
+                         \textit{region}\\
+\textit{spec} ::= & \textit{spatial} \mid \textit{temporal}   \\[1em]
+\textit{spatial} ::= & [\textit{mod} \; \{, \textit{mod} \}] \; \textit{region} \\
+\textit{mod} ::= &
  \texttt{read-once} \\
-& \mid \texttt{reflexive} \; \texttt{dim=}\textit{dim}{\texttt{,}} \\
-& \mid \texttt{irreflexive} \; \texttt{dim=}\textit{dim}{\texttt{,}} \\
-\textit{sumS} ::= \; & \textit{prodS} \;\, \texttt{+} \;\, \textit{sumS} \mid S_{\texttt{+}} \; \texttt{+}
-          \; \textit{sumS} \mid \epsilon \\
-\textit{prodS} ::= \; & S_{\texttt{*}}  \;\, \texttt{*} \;\, prodS
-                        \mid \epsilon \\ \\
-\textit{temporal} ::= \; & \texttt{dependency} \; (v \; \{ , v \})  \\ \\
-\multicolumn{2}{l}{(\textit{meta rules, where \textnormal{$\odot = \{\texttt{*}, \texttt{+}, \texttt{,}\}$}})} \\
-S_\odot ::= \; & \texttt{forward} \; [,\texttt{depth=}\mathbb{N}] \;
-        , \texttt{dims=}\textit{dims}_\odot  \\
-\mid \; & \texttt{backward} \; [,\texttt{depth=}\mathbb{N}] \; , \texttt{dims=}\textit{dims}_\odot \\
-\mid \; & \texttt{centered} \; [,\texttt{depth=}\mathbb{N}] \; ,
-          \texttt{dims=}\textit{dims}_\odot \\
-\textit{dims}_\odot ::= & \; \mathbb{N}_{>0} \; \odot \;
-                          \textit{dims}_{\odot} \; \mid \;
-                          \mathbb{N}_{> 0}
+\mid \; & \texttt{reflexive} \; \texttt{dims=}\dimTy \{\texttt{,} \; \dimTy\} \\
+\mid \; & \texttt{irreflexive} \; \texttt{dims=}\dimTy \{\texttt{,} \;
+          \dimTy\} \\[0.1em]
+\textit{region} ::=
+     \; & \texttt{forward} \; [,\texttt{depth=}\mathbb{N}] \; , \texttt{dim=}\dimTy  \\
+\mid \; & \texttt{backward} \; [,\texttt{depth=}\mathbb{N}] \; , \texttt{dim=}\dimTy \\
+\mid \; & \texttt{centered} \; [,\texttt{depth=}\mathbb{N}] \; , \texttt{dim=}\dimTy \\
+\mid \; & \textit{region} \; \texttt{+} \; \textit{region} \\
+\mid \; & \textit{region} \; \texttt{*} \; \textit{region} \\
+\mid \; & \textit{rvar}  \\[0.5em]
+\textit{temporal} ::= \; & \texttt{dependency} \; (v \; \{ , v \})
+  \\[0.5em]
+\dimTy ::= \; & \mathbb{N}_{>0} \\
+\textit{rvar} ::= \; & [\text{\texttt{a}-\texttt{z}$\,$\texttt{A}-\texttt{Z}$\,$\texttt{0}-\texttt{9}}]+
 \end{array}
 \end{align*}
-\caption{Specification syntax}
+\caption{Specification syntax (EBNF grammar)}
 \label{fig:syntax}
@@
 
-\paragraph{Time specifications}
-\dnote{The below is old so probably out of date}
-
-\begin{ExmVerbatim}
-for t = 1, itermax:
-   for i = 1, (n-1):
-      a[t, i] = a[t-1, i-1] - 2*a[t-1, i] + a[t-1, i+1]
-\end{ExmVerbatim}
-
-\begin{SpecVerbatim}
-a(t=0) : bwd, depth=1
-a(1)   : centered, depth=1
-\end{SpecVerbatim}
-
-\begin{SpecVerbatim}
-a : centered depth=1 dim=1
-a : backward depth=1 dim=0
-\end{SpecVerbatim}
+\paragraph{Temporal specifications}
 
@@
 \begin{SpecVerbatim}
-a: centered depth=1, dims=0
+%=  stencil centered, depth=1, dim=0 :: a
 \end{SpecVerbatim}

### 2016-06-15 paper.tex
updated EBNF grammar with approximation info
grammar tweaks
corrected the comment syntax
Adjust specification rule & intro term, nonterm macros.
Adjust specification rule to be regionDec | specDec.
Introduce \term and \nonterm macros to replace \texttt and \textit.
fixed multiline syntax in fig1
improving intro

@@
 \usepackage{amsthm}
+\usepackage{tikz}
 
@@
 % Syntax macros
-\newcommand{\stenFwd}[2]{\texttt{forward}, \texttt{depth}=#1,
-  \texttt{dim}=#2}
-\newcommand{\stenBwd}[2]{\texttt{backward}, \texttt{depth}=#1,
-  \texttt{dim}=#2}
-\newcommand{\stenCen}[2]{\texttt{backward}, \texttt{depth}=#1,
-  \texttt{dim}=#2}
+\newcommand{\nonterm}[1]{\textit{#1}}
+\newcommand{\term}[1]{\texttt{#1}}
+
+\newcommand{\stenFwd}[2]{\texttt{forward} \, (\texttt{depth=}#1,
+  \texttt{dim=}#2)}
+\newcommand{\stenBwd}[2]{\texttt{backward} \, (\texttt{depth=}#1,
+  \texttt{dim=}#2)}
+\newcommand{\stenCen}[2]{\texttt{centered} \, (\texttt{depth=}#1,
+  \texttt{dim=}#2)}
 
@@
 
-\emph{Stencils} are a ubiquitous programming pattern in many numerical
-applications. Informally, a stencil computation computes
-an array, where the value at each index $i$ of this array is
-calculated from a set of neighbouring values to $i$ in some input
-array(s), \eg{}, the Game of Life, convolutions in image processing,
-approximations to differential equations. For example, the following
-psuedo-code iteratively applies the discrete Laplace
-operator (an approximation to a derivative):
+\emph{Stencils} are a ubiquitous programming pattern, common in
+scientific and numerical computing applications. Informally, a stencil
+computation computes an array value, where the value at each index $i$ of
+this array is calculated from a \emph{neighbourhood} of values around $i$ in
+some input array(s), \eg{}, the Game of Life, convolutions in image
+processing, approximations to differential equations. For example, the
+following iteratively computes the one-dimensional discrete Laplace
+transform (an approximation to a derivative): 
 %
@@
 Line 3 is the core of the stencil computation, calculating
-the value at \texttt{b[i]} from a neighbourhood of elements about
-\texttt{a[i]}. Line 4 swaps
+the value at \texttt{b(i)} from a neighbourhood of elements about
+\texttt{a(i)}. Line 4 swaps
 \texttt{a} and \texttt{b} between iterations, where \texttt{b} becomes the
-input for the next iteration.
+input for the next iteration. (Note, we use the parentheses syntax
+\texttt{( )} for
+array subscripts rather than the bracket syntax \texttt{[ ]}.
 
@@
 %$\texttt{b}$.
-In this simple example, the spatial and temporal relationship
-between \texttt{a} and \texttt{b} is easy to understand. However,
-more complex stencil computations can be much harder to understand
-and subsequently more prone to error. For example, Figure~\ref{ref:navier-stokes-fragment}
-shows a couple of lines from a Navier-Stokes fluid simulator in
-which two arrays are read from with different data access patterns,
-across two dimensions. The interaction is much harder to understand,
-with the potential for the developer to accidentally introduce an
-error, for example writing $\texttt{(i-1,j)}$ instead of
-$\texttt{(i+1,j)}$.
+In this simple example, the dependencies between between \texttt{a}
+and \texttt{b} form a simple spatial relationship which is easily
+understood. This spatial relationship determines other aspects of the
+program and its efficient implementation: how much ``boundary'' is
+needed for the array, the most cache-efficient layout in memory,
+the partitioning shape for parallel implementations.
 
-We introduce a simple specification for stencil comptuations
-that abstracts the fine grained detail of their access patterns.
-In practise, most stencil computations have a regular shape that
-can be described simply and abstractly. In the case of our
-first Laplace example, our analysis tool infers the following specification:
+More complex stencil computations can be much harder to understand and
+subsequently more prone to error. For example,
+Figure~\ref{ref:navier-stokes-fragment} shows three lines from a
+Navier-Stokes fluid simulator in which two arrays are read from with
+different data access patterns, across two dimensions. The interaction
+is much harder to understand, with the potential for the developer to
+accidentally introduce an error via simple textual mistakes, for
+example writing $\texttt{(i-1,j)}$ instead of $\texttt{(i+1,j)}$.
+
+In this work, we introduce a simple specification language for the
+spatial and temporal properties of stencil comptuations. The
+specifications abstract over the fine grained detail of stencil access
+patterns.  In practise, most stencil computations have a regular shape
+that can be described simply and abstractly, with a small set of
+coarse-grained descriptions. In the case of our first Laplace example,
+our inference procedure provides the specification:
 %
 \begin{SpecVerbatim}
-%=  stencil centered, depth=1 dims=1 :: a
-%=  stencil dependency a, mutual    :: b
+!=  stencil centered (depth=1, dim=1) :: a
+!=  stencil dependency a, mutual       :: b
 \end{SpecVerbatim}
 %
-The first line explains that \texttt{a} is read from
-with a symmetrical stencil pattern (``centered'') to a depth
-of one in each direction in its first (and only) dimension.
-The second line explains the temporal relationship between
-\texttt{b} and \texttt{a}: that the previous time step for \texttt{b}
-is actually provided by \texttt{a}. The dimension \texttt{t} is a
-reserved dimension identifiers for temporal specifications. In
-the case of the Navier-Stokes example, its inferred stencil is shown
-in Figure~\ref{ref:navier-stokes-fragment}(b).
+The first line explains that \texttt{a} is read from with a
+symmetrical stencil pattern (``centered'') to a depth of one in each
+direction in its first (and only) dimension.  The second line explains
+the temporal relationship between \texttt{b} and \texttt{a}: that
+previous time step for \texttt{b} is actually provided by \texttt{a},
+and vice versa. This is explained as a mutual dependence between
+\texttt{a} and \texttt{b}. In the case of the Navier-Stokes example of
+Figure~\ref{ref:navier-stokes-fragment}, its inferred specification is shown
+in Figure~\ref{ref:navier-stokes-fragment}(b). The 
+specification explains that, over the whole fragment, \texttt{u} is
+read from with a centered pattern to depth of 1 in both dimensions
+(this is known as the \emph{five point stencil}) and \texttt{v}
+is read from in a neighbourhood bounded by forward to depth of $1$ in
+the first dimension and backward to a depth of $1$ in the second
+dimension. These can drawn pictorially as follows where the horizontal
+dimensions is \text{dim=1} and the vertical is \texttt{dim=2}:
+
+\begin{center}
+\begin{tikzpicture}
+\node at (1.25,1.22) {i, j};
+\draw (1,1) rectangle (1.5,1.5);
+\draw (1.5,1) rectangle (2,1.5);
+\draw (1,0.5) rectangle (1.5,1);
+\draw (0.5,1) rectangle (1,1.5);
+\draw (1,1.5) rectangle (1.5,2);
+%
+\node at (4.25,1.22) {i, j};
+\draw (4,1) rectangle (4.5,1.5);
+\draw (4.5,1) rectangle (5,1.5);
+\draw (4.5,1.5) rectangle (5,2);
+\draw (4,1.5) rectangle (4.5,2);
+\end{tikzpicture}
+\end{center}
 
@@
 \begin{SpecVerbatim}[xleftmargin=0.3cm]
-%=   stencil (centered, depth=1, dim=1)
-%=         + (centered, depth=1, dim=2) :: u
+!=   stencil centered(depth=1, dim=1)
+           + centered(depth=1, dim=2) :: u
 
-%=   stencil (forward, depth=1, dim=1)
-%=         + (backward, depth=1, dim=1) :: v
+!=   stencil forward(depth=1, dim=1)
+           * backward(depth=1, dim=2) :: v
 \end{SpecVerbatim}
@@
 \begin{array}{rl}
-\textit{specification} ::= & [\textit{regionDec} \; \{\texttt{$\backslash$n} \;\;
-                             \textit{regionDec} \}] \;
-                             \textit{specDec} \\
-\textit{specDec} ::= & \texttt{stencil} \; \textit{spec} \mid \epsilon
+\nonterm{specification} ::= & \nonterm{regionDec} \mid \nonterm{specDec} \\
+\nonterm{specDec} ::= & \term{stencil} \; \nonterm{spec}
   \\
-\textit{regionDec} ::= &  \texttt{region} \; \textit{rvar} \; \texttt{=} \;
-                         \textit{region}\\
-\textit{spec} ::= & \textit{spatial} \mid \textit{temporal}   \\[1em]
-\textit{spatial} ::= & [\textit{mod} \; \{, \textit{mod} \}] \; \textit{region} \\
-\textit{mod} ::= &
- \texttt{read-once} \\
-\mid \; & \texttt{reflexive} \; \texttt{dims=}\dimTy \{\texttt{,} \; \dimTy\} \\
-\mid \; & \texttt{irreflexive} \; \texttt{dims=}\dimTy \{\texttt{,} \;
-          \dimTy\} \\[0.1em]
-\textit{region} ::=
-     \; & \texttt{forward} \; [,\texttt{depth=}\mathbb{N}] \; , \texttt{dim=}\dimTy  \\
-\mid \; & \texttt{backward} \; [,\texttt{depth=}\mathbb{N}] \; , \texttt{dim=}\dimTy \\
-\mid \; & \texttt{centered} \; [,\texttt{depth=}\mathbb{N}] \; , \texttt{dim=}\dimTy \\
-\mid \; & \textit{region} \; \texttt{+} \; \textit{region} \\
-\mid \; & \textit{region} \; \texttt{*} \; \textit{region} \\
-\mid \; & \textit{rvar}  \\[0.5em]
-\textit{temporal} ::= \; & \texttt{dependency} \; (v \; \{ , v \})
+\nonterm{regionDec} ::= &  \texttt{region} \; \nonterm{rvar} \; \texttt{=} \;
+                         \nonterm{region}\\
+\nonterm{spec} ::= & \nonterm{spatial} \mid \nonterm{temporal}   \\[1em]
+\nonterm{spatial} ::= & [\nonterm{mod}] \;
+                       [\nonterm{approxMod},] \; [\nonterm{region}] \\
+\nonterm{mod} ::= & [\nonterm{modR},] \; [\nonterm{modRefl},] \; [\nonterm{modIrefl},] \\
+\nonterm{modR} ::= &
+ \term{readOnce} \\
+\nonterm{modRefl} ::= & \term{reflexive}, \; \term{dims=}\dimTy \{\term{,} \; \dimTy\} \\
+\nonterm{modIrrefl} ::= & \term{irreflexive}, \; \term{dims=}\dimTy \{\term{,}
+          \dimTy\}\\
+\nonterm{approxMod} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
+\nonterm{region} ::= & \term{forward}, \term{depth=}\mathbb{N}, \term{dim=}\dimTy  \\
+\mid \; & \term{backward}, \term{depth=}\mathbb{N}, \term{dim=}\dimTy \\
+\mid \; & \term{centered}, \term{depth=}\mathbb{N}, \term{dim=}\dimTy \\
+\mid \; & \nonterm{region} \; \term{+} \; \nonterm{region} \\
+\mid \; & \nonterm{region} \; \term{*} \; \nonterm{region} \\
+\mid \; & \nonterm{rvar}  \\[0.5em]
+\nonterm{temporal} ::= \; & \term{dependency} \; (v \; \{ , v \})
   \\[0.5em]
 \dimTy ::= \; & \mathbb{N}_{>0} \\
-\textit{rvar} ::= \; & [\text{\texttt{a}-\texttt{z}$\,$\texttt{A}-\texttt{Z}$\,$\texttt{0}-\texttt{9}}]+
+\nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+
 \end{array}
@@
 
+% Implicit vs explicit differentiation methods
+% (forward implicit needs 'solver')
+% backward - Euler
+% runge-kutta 2nd-order
+
 \begin{example}[FTCS (Forward time, centered space)]
@@
 \begin{SpecVerbatim}
-%=  stencil centered, depth=1, dim=0 :: a
+!=  stencil centered, depth=1, dim=0 :: a
 \end{SpecVerbatim}
@@
   \, . \, d \in \textit{ds} \Rightarrow ix \, d \leq (n+1) \} \\
-& = \interp{\texttt{v} : \texttt{fwd} \; \texttt{dims}=ds \;
+& = \interp{\texttt{v} : \texttt{fwd} \; \texttt{dim}=ds \;
   \texttt{depth}=n + 1} \qquad \Box

### 2016-06-15 paper.tex
tweak syntax in inference descriptions

@@
 
-\begin{definition}[Constant translations]
-An indexing expression $e$ is a \emph{constant translation} if,
-for a base induction variable $i$, then $e \equiv i + a$ or $e \equiv i - a$
+\begin{definition}[Neighbourhood indices]
+An array subscript $e$ is a \emph{neighbourhood index} if,
+for an induction variable $i$, then $e \equiv i + a$ or $e \equiv i - a$
 where $a$ is a constant. The relation $\equiv$ identifies terms
@@
 relation of $+$ and $-$ (\eg{}, $(-b) + i \equiv i - b$).
-In the following, we classify constant translation expressions
-using the predicate $\textsf{trans}$.
+In the following, we classify neighbourhood indices 
+using the predicate $\textsf{neigh}$.
 \end{definition}
@@
 \interp{e_1 e_2}      & = \interp{e_1} \cup \interp{e_2} \\
-\interp{a(\tilde{e})} & = \{e | e \in \tilde{e}, \textsf{trans}(e)\} & \textit{where each $e \in
+\interp{a(\tilde{e})} & = \{e | e \in \tilde{e}, \textsf{neigh}(e)\} & \textit{where each $e \in
                                           \tilde{e}$ is a constant or constant translation}
@@
 
-The reserved dimension \dimId{t}
+\begin{itemize}
+\item Extract from syntax finite map of variables
+to relative-index verctors representing the contributing
+array subscritps
+\end{itemize}
 

### 2016-06-16 paper.tex
inference

@@
 \usepackage{tikz}
+\usepackage{multirow}
 
 \CustomVerbatimEnvironment{SpecVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.65cm,
-xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.97}
+xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.98}
 \CustomVerbatimEnvironment{ExmVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.65cm,
-xrightmargin=0.2cm,baselinestretch=0.97,numbers=left}
+xrightmargin=0.2cm,baselinestretch=0.98,numbers=left}
 \CustomVerbatimEnvironment{IVerbatim}{Verbatim}{fontsize=\relsize{-1},xleftmargin=0.65cm,
-xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.97}
+xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.98}
 
@@
 % Writing macros
-\newcommand{\ie}{\emph{i.e..}}
+\newcommand{\ie}{\emph{i.e.}}
 \newcommand{\eg}{\emph{e.g.}}
@@
 
-\subsection{A semantic model for specifications}
-\label{sec:semantics}
-
-\newcommand{\relix}{(\mathbb{Z}_\bot)^\mathbb{D}}
-
-We give a model of our specifications which explains
-the range of indices which correspond to the specifications.
-The model is based on multi-sets since array indices my be repeated
-in an expression.
-
-Figure~\ref{fig:model} describes the model in terms of
-an interpretation function $\interp{-}$ which maps specifications
-to multisets of relative indices.
-
-\begin{definition}Relative indices are drawn
-from $(\mathbb{Z}_{\bot})^{\mathbb{D}}$, essentially an infinite
-tuple of offsets. For example, $(i, j+1)$ corresponds to
-the relative index $(0, 1, \bot, \bot, \ldots)$
-\end{definition}
-
-The main interpretation function is overloaded on lists of
-variable-spec pairs $\interp{\overline{v : S}}$,
-returning multisets of variable-relative-index pairs. This provides
-the top-level definition of the model, with $\interp{-}$ overloaded
-on $S$, $\interp{S}$ mqpping to multisets of
-relative indices not associated to an array avariable.
-
-\begin{figure}
-\input{model.tex}
-%\textit{$a^m \in A$ means there are $n$ copies of $a$ in
-%  the multi-set $A$}
-\caption{Multi-set model of specifications}
-\label{fig:model}
-\end{figure}
-
 \subsection{Equational theory and approximations}
@@
 
+\section{Semantics of specifications; a model}
+\label{sec:semantics}
+
+\newcommand{\relix}{(\mathbb{Z}_\bot)^\mathbb{D}}
+
+We give a model of our specifications which explains
+the range of indices which correspond to the specifications.
+The model is based on multi-sets since array indices my be repeated
+in an expression.
+
+Figure~\ref{fig:model} describes the model in terms of
+an interpretation function $\interp{-}$ which maps specifications
+to multisets of relative indices.
+
+\begin{definition}Relative indices are drawn
+from $(\mathbb{Z}_{\bot})^{\mathbb{D}}$, essentially an infinite
+tuple of offsets. For example, $(i, j+1)$ corresponds to
+the relative index $(0, 1, \bot, \bot, \ldots)$
+\end{definition}
+
+The main interpretation function is overloaded on lists of
+variable-spec pairs $\interp{\overline{v : S}}$,
+returning multisets of variable-relative-index pairs. This provides
+the top-level definition of the model, with $\interp{-}$ overloaded
+on $S$, $\interp{S}$ mqpping to multisets of
+relative indices not associated to an array avariable.
+
+\begin{figure}
+\input{model.tex}
+%\textit{$a^m \in A$ means there are $n$ copies of $a$ in
+%  the multi-set $A$}
+\caption{Multi-set model of specifications}
+\label{fig:model}
+\end{figure}
+
 \section{Examples and common patterns}
@@
 
-\section{Specification inference and checking}
+\section{Inference, checking, and synthesis}
 \label{sec:analysis}
@@
 
-In the following, we defined
-\emph{base induction variables} to be the variables
- defined by a ``for'' loop (\eg{}, that get incremented each time the
-loop is re-entered).
+\noindent
+We briefly sketch the algorithms for inferring and checking
+specifications and for synthesising programs from specifications.
+Checking and synthesis use the set-based semantic model of
+Section~\ref{sec:semantics}.
 
-\begin{definition}[Neighbourhood indices]
+% SYNTAX OPERATIONS AND PREDICATES
+\newcommand{\neigh}{\textsf{neigh}}
+\newcommand{\arrayTy}{\textsf{array}}
+\newcommand{\rhsExp}{\textsf{rhsExp}}
+\newcommand{\var}{\textsf{var}}
+
+%% VECTOR NOTATIONS
+\newcommand{\vect}[1]{\textbf{#1}}
+\newcommand{\vtwo}[2]{\setlength{\arraycolsep}{0em}
+\left[\begin{array}{l}#1\\#2\end{array}\right]}
+\newcommand{\stwo}[4]
+%{\vtwo{#1}{#2}\!\vtwo{#3}{#4}}
+{\setlength{\arraycolsep}{0.1em}
+\left[\begin{array}{rr}$#1$ & $#3$\\$#2$ & $#4$\end{array}\right]}
+
+%% OPERATIONS ON SPANS and VECTORS 
+\newcommand{\containedin}{\sqsubseteq}
+
+\paragraph{Basic syntax definitions} In the following, $v$
+  ranges over variables, $s$ over statements, and $e$ over expressions
+  (which may be impure).  By ``variable'', we mean in the imperative
+  sense (\ie{}, named binders to memory cells) instead of mathematical
+  variables. The function \rhsExp{} maps statements to a set of
+  expressions which occur in right-hand positions (\ie{}, not the
+  target of an assignment). The function \var{} maps expressions to a
+  set of the variables used in right-hand positions.
+
+\begin{definition}[Standard: base induction variable]
+An integer variable \textit{i} is a \emph{base induction variable} 
+within the scope of a ``for'' loop (or equivalent for the target
+language, \eg{}, \texttt{do} in Fortan for our implementation) if it
+is the control variable of the loop, incremented by $1$ per
+iteration. 
+\end{definition}
+
+\begin{definition}[Standard: array subscript]
+An \emph{array subscript} is an expression index an array, which in
+our abstract we write as $v(\bar{e})$ where $\bar{e}$ is shorthand for
+a syntactic list of indexing expressions. 
+\end{definition}
+
+\begin{definition}[Neighbourhood index]
 An array subscript $e$ is a \emph{neighbourhood index} if,
 for an induction variable $i$, then $e \equiv i + a$ or $e \equiv i - a$
-where $a$ is a constant. The relation $\equiv$ identifies terms
+where $a$ is a constant of integer type. The relation $\equiv$ identifies terms
 up-to commutativity of $+$ and the inverse
@@
 In the following, we classify neighbourhood indices 
-using the predicate $\textsf{neigh}$.
+using the predicate \neigh{}.
 \end{definition}
 
-\paragraph{Step 1: Data-access analysis}
+\subsection{Inference}
 
-\dnote{Might want to do 'stencil' (gather) and 'mould' (scatter)
-  patterns separately}
+We demonstate the main steps of the inference procedure over the
+following example program which computes the mean value 
+of a five-point stencil at each index of the input array:
+%%
+\begin{ExmVerbatim}
+do i = 1, (n-1)
+   do j = 1, (m-1)
+      x       = a(i-1, j) + a(i+1, j) 
+      y       = a(i, j-1) + a(i, j+1)
+      b(i, j) = (a(i, j) + x + y) / 5.0
+   end do
+end do
+\end{ExmVerbatim}
+%%
+\paragraph{Step 1: Standard control and dataflow analyses}
+The inference relies on some standard program analyses, computed
+before the main inference procedure:
+%
+\begin{enumerate}
+\item basic blocks (CFG);
+\item induction variables per basic block;
+\item (interprocedural) dataflow analysis, providing a \emph{flows to}
+  graph (as shorthand, the function
+  $\mathsf{flowTo}$ is used, implicitly parameterised by this graph,  
+  mapping an expression to the set of all expressions
+  with forwards dataflow to this expression, based on the transitive
+  closure of the flows graph);
+\item type information per variable, where we use the predicate
+\arrayTy{} to classify variables of array type. 
+\end{enumerate}
+%
+
+\paragraph{Step 2: Data-access analysis}
+
+For each assignment statement whose left-hand side is an array
+subscript on neighbourhood indices, a finite map is computed which
+maps array variables to a set of vectors representing array
+subscripts.  This finite map contains all array subscript expressions
+which flow to this statment. More formally, a function
+$\textsf{analyse}$ is applied to each statement in a program with the
+following clause:
+%
 \begin{align*}
-\begin{array}{lll}
-\interp{e_1 = e_2}    & = \interp{e_2} \\
-\interp{e_1 e_2}      & = \interp{e_1} \cup \interp{e_2} \\
-\interp{a(\tilde{e})} & = \{e | e \in \tilde{e}, \textsf{neigh}(e)\} & \textit{where each $e \in
-                                          \tilde{e}$ is a constant or constant translation}
+\begin{array}{lr}
+\textsf{analyse}(v(\overline{e_1}) = e_2)
+ := \qquad\qquad & \textit{where} \; \neigh(\overline{e_1}) \wedge \arrayTy(v)  \\[0.3em]
+\multicolumn{2}{l}{\qquad \bigcup\{v' \mapsto \{\textsf{convert}(\bar{e})\} \mid v'(\bar{e}) \leftarrow \mathsf{flowsTo}(e_2),
+  \arrayTy(v')\}}
 \end{array}
 \end{align*}
+%
+That is, we focus on assignments to an array subscript where each
+index indexing expression in $\bar{e}$ is a neighbourhood index.  For
+all array subscripts that flow to the right-hand side of this
+statement, a finite map is constructed from each array variable
+in this flow set to a representation of the subscripts, computed
+with \textsf{convert}.
 
-Index analysis is a \emph{coeffect analysis}.
+The \textsf{convert} function maps subscripts to a column vector
+representation with values drawn from $\mathbb{Z}_\top = \mathbb{Z} \cup \{\top\}$
+where $\mathbb{Z}$ represents the offset of a neighbourhood index
+and $\top$ represents any non-neighbourhood subscripts, defined:
+%
+\begin{align*}
+& \textsf{convert}((e_1, \ldots, e_n)) = [\textsf{conv}(e_1), \ldots,
+  \textsf{conv}(e_n)] \\
+& \textsf{conv}(e) \begin{cases}
+\pm a & \neigh(e) \wedge \, e \equiv i \pm a \\
+\top & \textit{otherwise} 
+\end{cases} 
+\end{align*}
+%
+For our example, the \textsf{analyse} function matches on
+line 5, with the following set for $\textsf{flowTo}(a(i, j) + x +
+  y)$:
+%
+\begin{align*}
+\{\texttt{a(i-1, j)}, \texttt{a(i+1, j)}, \texttt{a(i, j-1)},
+  \texttt{a(i, j+1)}, \texttt{a(i, j)}\}
+\end{align*}
+Subsequently the result of \textsf{analyse} on line 5 yields the map:
+\begin{align*}
+\texttt{a} \mapsto \{\vtwo{-1}{0}, \vtwo{1}{0},
+          \vtwo{0}{-1}, \vtwo{0}{1}, \vtwo{0}{0}\}
+\end{align*}
+%
 
-\paragraph{Step 2: Coalesce contiguous indices into regions}
+\paragraph{Step 3: Coalesce contiguous indices into regions}
 
-% fairly mechanical
+Let $M$ range over the finite maps generated by \textsf{analyse}.  For
+each $v \in \mathsf{dom}(M)$, the next step of the algorithm coalesces
+the set of relative index vectors $M(v)$ into contiguous regions within an
+$n$-dimensional space.
+
+Let $\vect{x}, \vect{y}, \vect{z}$ range over the column vectors of size
+$n$ whose values are drawn from $\mathbb{Z}_\top$.
+We write $\vect{x}_i$ for the $i$-th value of vector $\vect{x}$ (1
+indexed) and $\vect{x}_{i:j}$ for the subvector from entry $i$ to
+entry $j$ (inclusive).
+
+\begin{definition}[Span]
+  A \emph{span} represents an $n$-dimensional box (\emph{hyperrectangle}) as
+  a pair of $n$-dimensional vectors (represented as a $2 \times n$
+  matrix) giving the co-ordinates of the lower-bound vertex (first
+  column) and the upper-bound vertex (second column). We use the
+  notation $[\vect{x}^L \; \vect{x}^U]$ for such a span to mark out
+  the lower and upper bound vecvtors. 
+\end{definition}
+
+Firstly, for all $v \in \mathsf{dom}(M)$ then a new map $M'$ is
+created where each vector is mapped to the trivial \emph{unit span} is
+created by pairing a vector with itself:
+%
+\begin{align*}
+N(v) = \{[\vect{x} \; 
+  \vect{x}] \mid \vect{x} \in M(v)\}
+\end{align*}
+%
+For our example, this yields:
+%%
+\begin{align*}
+N(\texttt{a}) = \stwo{0}{0}{0}{0} \stwo{1}{0}{1}{0} \stwo{-1}{0}{-1}{0} \stwo{0}{1}{0}{1} \stwo{0}{-1}{0}{-1}
+\end{align*}
+%%
+A fixed point is then compute for the following
+procedure \textsf{regions} which, for per variable
+in the map, coalesces spans into contiguous regions in
+the $n$-dimensional spaace. That is, we compute
+ $(\mu \; \textsf{regions}) N$ where \textsf{regions}
+is defined as follows for each variable in the map:
+%
+\begin{enumerate}
+
+\item Compute all permutations on the column vectors in a span, \ie{},
+  $[\vect{x}^L \vect{x}^U] \mapsto [\pi\vect{x}^L \, \pi\vect{x}^U]$
+for a permutation $\pi$. For each permutation function $\pi^n_i$
+(the $i$-th permutation for vectors of size $n$, we pair the
+permutation function with the set of permuted spans so that 
+the spans can be unpermuted later. 
+%
+\[
+P(v) = \bigcup_{i \in !n} (\pi^n_{i} , \; \{[\pi^n_i
+\vect{x}^L \; \pi^n_i\vect{x}^U] \, \mid \, [\vect{x}^L \; \vect{x}^U]
+\leftarrow N(v)\}
+\]
+%
+For our example, this yields:
+%
+\begin{align*}
+P(\texttt{a}) = 
+\{&(\pi^2_1, \{ \stwo{0}{0}{0}{0}
+\stwo{1}{0}{1}{0} 
+\stwo{-1}{0}{-1}{0} 
+\stwo{0}{1}{0}{1} 
+\stwo{0}{-1}{0}{-1} \})
+\\
+&(\pi^2_2, \{ 
+ \stwo{0}{0}{0}{0}
+ \stwo{0}{1}{0}{1}
+ \stwo{0}{-1}{0}{-1}
+ \stwo{1}{0}{1}{0}
+ \stwo{-1}{0}{-1}{0}\})\}
+\end{align*}
+%
+where $\pi^2_1$ is the identity permutation and $\pi^2_2$ is the
+permutation flips the order of the two elements in the column
+vectors.
+
+\item Sort each permutation set into an ordered list, based on the ordering:
+\[
+\vect{x} \leq \vect{y} = \exists i \, . \; \vect{x}_i \leq \vect{y}_i 
+\; \wedge \; \vect{x}_{i:n} = \vect{y}_{i:n}
+\]
+%
+This yields:
+%%
+\begin{align*}
+\{&(\pi^2_1, [
+\stwo{0}{-1}{0}{-1}
+\stwo{-1}{0}{-1}{0} 
+\stwo{0}{0}{0}{0}
+\stwo{1}{0}{1}{0} 
+\stwo{0}{1}{0}{1}] )
+\\
+&(\pi^2_2, [
+\stwo{0}{-1}{0}{-1}
+\stwo{-1}{0}{-1}{0}
+\stwo{0}{0}{0}{0}
+\stwo{1}{0}{1}{0}
+\stwo{0}{1}{0}{1}])\}
+\end{align*}
+%%
+\item Fold each list pairwise by the following partial $\bullet$ which 
+coalescees and contiguous regions:
+%
+\begin{align*}
+& (\vect{x}^L,\vect{x}^U) \bullet (\vect{y}^L,\vect{y}^U) \\
+= &
+\begin{cases}
+(\vect{x}^L, \vect{y}^U) & \vect{x}^U_1 + 1 = \vect{y}^L_1 \; \wedge \;
+(\vect{x}^L_{1:n}, \vect{x}^U_{1:n}) = (\vect{y}^L_{1:n}, \vect{y}^U_{1:n}) \\
+\bot  & \textit{otherwise}
+\end{cases}
+\end{align*}
+For our example, this yields:
+%
+\begin{align*}
+Q(v) = \{&(\pi^2_1, [
+\stwo{0}{-1}{0}{-1}
+\stwo{-1}{0}{1}{0}, 
+\stwo{0}{1}{0}{1}, 
+]) \\
+&(\pi^2_2, [
+\stwo{0}{-1}{0}{-1},
+\stwo{-1}{0}{1}{0},
+\stwo{0}{1}{0}{1}])\}
+\end{align*}
+%
+\item Unpermute and union together, \ie{},
+%
+\[
+U(v) = \bigcup \{[\pi \vect{y}^L, \pi \vect{y}^U]
+ \mid [\vect{y}^L, \vect{y}^U] \leftarrow S, (\pi, S) \leftarrow Q(v)\}
+\]
+Which yields:
+%
+\begin{align*}
+U(\texttt{a}) = 
+\{\stwo{0}{-1}{0}{-1}
+\stwo{-1}{0}{1}{0}
+\stwo{0}{1}{0}{1}
+\stwo{-1}{0}{-1}{0}
+\stwo{0}{-1}{0}{1}
+\stwo{1}{0}{1}{0}\}
+\end{align*}
+%
+\item Filter by $\containedin$, defined: 
+%
+\begin{align*}
+(\vect{x}^L, \vect{x}^U) \containedin (\vect{y}^L, \vect{y}^U) = 
+\vect{y}^L_1 \leq \vect{x}^L_1 \wedge \vect{x}^U_1 \leq \vect{y}^U_1
+  \wedge (\vect{x}^L_{1:n}, \vect{x}^U_{1:n}) \containedin
+  (\vect{y}^L_{1:n}, \vect{y}^U_{1:n})
+\end{align*}
+
+Yields:
+\begin{align*}
+\stwo{-1}{0}{1}{0}, 
+\stwo{0}{-1}{0}{1}
+\end{align*}
+\end{enumerate}
+
+\begin{tikzpicture}
+\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
+\draw[->] (0.5,2) -- (2,2);
+\node at (1.25,1.22) {i, j};
+\draw[thick] (1,1) rectangle (1.5,1.5);
+\draw[thick] (1.5,1) rectangle (2,1.5);
+\draw[thick] (1,0.5) rectangle (1.5,1);
+\draw[thick] (0.5,1) rectangle (1,1.5);
+\draw[thick] (1,1.5) rectangle (1.5,2);
+\end{tikzpicture}
+%
+\begin{tikzpicture}
+\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
+\draw[->] (0.5,2) -- (2,2);
+\node at (1.25,1.22) {i, j};
+\draw[thick] (0.5,1) rectangle (2,1.5);
+\draw[thick] (1,0.5) rectangle (1.5,1);
+\draw[thick] (1,1.5) rectangle (1.5,2);
+\end{tikzpicture}
+%
+\begin{tikzpicture}
+\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
+\draw[->] (0.5,2) -- (0.5,0.5);
+\node at (1.25,1.22) {i, j};
+\draw[thick] (0.5,1) rectangle (2,1.5);
+\draw[thick] (1,0.5) rectangle (1.5,1);
+\draw[thick] (1,1.5) rectangle (1.5,2);
+\end{tikzpicture}
 

### 2016-06-16 paper.tex
amplifying informal intuition on inference procedure

@@
 \newcommand{\vtwo}[2]{\setlength{\arraycolsep}{0em}
-\left[\begin{array}{l}#1\\#2\end{array}\right]}
+\left[\begin{array}{l}$#1$\\$#2$\end{array}\right]}
 \newcommand{\stwo}[4]
@@
 
+\renewcommand*{\arraystretch}{0.8}
 \paragraph{Basic syntax definitions} In the following, $v$
@@
 Let $M$ range over the finite maps generated by \textsf{analyse}.  For
-each $v \in \mathsf{dom}(M)$, the next step of the algorithm coalesces
-the set of relative index vectors $M(v)$ into contiguous regions within an
-$n$-dimensional space.
+each $v \in \mathsf{dom}(M)$, the algorithm then constructs a group
+of regions which cover all contiguous groups of relative indices
+(from $M(v)$) in the $n$-dimensional space.
+
+Informally, the procedure proceeds as follows. First, relative
+indices are turned into unit regions in $n$-dimensions. For
+example, for the five point stencil there are five $1 \times 1$ regions,
+called \emph{spans}, which are essentially $0$-dimensional:
+%
+\begin{center}
+\begin{tikzpicture}
+%\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
+\node at (1.25,1.22) {\textnormal{\small{0,0}}};
+\draw[thick] (1,1) rectangle (1.5,1.5);
+\draw[thick] (1.5,1) rectangle (2,1.5);
+\draw[thick] (1,0.5) rectangle (1.5,1);
+\draw[thick] (0.5,1) rectangle (1,1.5);
+\draw[thick] (1,1.5) rectangle (1.5,2);
+\end{tikzpicture}
+\end{center}
+%
+Each dimension is traversed, coalescing consecutive spans
+which vary in $m$-dimensions into coalesced spans which vary in 
+$m+1$-dimensions. In this example, the $0$-dimensional spans
+ become $1$-dimensional spans (rows/columns):
+%
+\begin{align*}
+\begin{array}{cc}
+\begin{tikzpicture}
+%\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
+%\node at (1.25,1.22) {\textnormal{\small{0,0}}};
+\draw[thick] (0.5,1) rectangle (2,1.5);
+\draw[thick] (1,0.5) rectangle (1.5,1);
+\draw[thick] (1,1.5) rectangle (1.5,2);
+\end{tikzpicture}
+\qquad
+&
+\qquad
+\begin{tikzpicture}
+%\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
+%\node at (1.25,1.22) {\textnormal{\small{0,0}}};
+\draw[thick] (1.5,1) rectangle (2,1.5);
+\draw[thick] (0.5,1) rectangle (1,1.5);
+\draw[thick] (1,0.5) rectangle (1.5,2);
+\end{tikzpicture}
+\\
+\textit{dimension 1} \qquad & \qquad \textit{dimension 2}
+\end{array}
+\end{align*}
+%
+Any spans that are contained with any other span are deleted,
+leaving a minimal set of spans (which may overlap, but none of which
+fully contains another).
+\begin{center}
+\begin{tikzpicture}
+%\node at (1.25,1.22) {\textnormal{\small{0,0}}};
+\draw[thick] (1,0.5) rectangle (1.5,2);
+\node at (2,1.22) {+};
+\draw[thick] (2.5,1) rectangle (4,1.5);
+%\node at (2.75,1.22) {\textnormal{\small{0,0}}};
+\end{tikzpicture}
+\end{center}
+%
+The procedure is then iterated till a fixed point is reached. In this
+example this is reached in the first step. These final spans
+go on to form the basis of the spatial stencil specification, where
+each span is a region product \texttt{*} and multiple spans are
+combined with the region sum \texttt{+}.
+
+\paragraph{Span coalescing, formally}
 
@@
 
-\begin{definition}[Span]
+\begin{definition}[Spans]
   A \emph{span} represents an $n$-dimensional box (\emph{hyperrectangle}) as
@@
 
+The algorithm to create contiguous spans, covering the space of
+indices, then proceeds as follows. 
 Firstly, for all $v \in \mathsf{dom}(M)$ then a new map $M'$ is
@@
 %
-\item Filter by $\containedin$, defined: 
+\item Finally, filter by the region containment predicate $\containedin$, that
+  is, if any region is contained within another then remove the
+  smaller. The $\containedin$ predicate is defined:
 %
 \begin{align*}
-(\vect{x}^L, \vect{x}^U) \containedin (\vect{y}^L, \vect{y}^U) = 
-\vect{y}^L_1 \leq \vect{x}^L_1 \wedge \vect{x}^U_1 \leq \vect{y}^U_1
+& (\vect{x}^L, \vect{x}^U) \containedin (\vect{y}^L, \vect{y}^U) = \\
+& \vect{y}^L_1 \leq \vect{x}^L_1 \wedge \vect{x}^U_1 \leq \vect{y}^U_1
   \wedge (\vect{x}^L_{1:n}, \vect{x}^U_{1:n}) \containedin
@@
 \end{align*}
-
-Yields:
+For our example this then yields the final result of: 
 \begin{align*}
-\stwo{-1}{0}{1}{0}, 
-\stwo{0}{-1}{0}{1}
+\textsf{regions}(M) 
+= \{\stwo{-1}{0}{1}{0} \stwo{0}{-1}{0}{1}\}
 \end{align*}
+since $\stwo{0}{1}{0}{1} \sqsubseteq \stwo{0}{-1}{0}{1}$
+and $\stwo{1}{0}{1}{0} \sqsubseteq \stwo{-1}{0}{1}{0}$. 
 \end{enumerate}
-
-\begin{tikzpicture}
-\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
-\draw[->] (0.5,2) -- (2,2);
-\node at (1.25,1.22) {i, j};
-\draw[thick] (1,1) rectangle (1.5,1.5);
-\draw[thick] (1.5,1) rectangle (2,1.5);
-\draw[thick] (1,0.5) rectangle (1.5,1);
-\draw[thick] (0.5,1) rectangle (1,1.5);
-\draw[thick] (1,1.5) rectangle (1.5,2);
-\end{tikzpicture}
-%
-\begin{tikzpicture}
-\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
-\draw[->] (0.5,2) -- (2,2);
-\node at (1.25,1.22) {i, j};
-\draw[thick] (0.5,1) rectangle (2,1.5);
-\draw[thick] (1,0.5) rectangle (1.5,1);
-\draw[thick] (1,1.5) rectangle (1.5,2);
-\end{tikzpicture}
-%
-\begin{tikzpicture}
-\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
-\draw[->] (0.5,2) -- (0.5,0.5);
-\node at (1.25,1.22) {i, j};
-\draw[thick] (0.5,1) rectangle (2,1.5);
-\draw[thick] (1,0.5) rectangle (1.5,1);
-\draw[thick] (1,1.5) rectangle (1.5,2);
-\end{tikzpicture}
-
-\subsection{Infering temporal specifications}
-
-\begin{itemize}
-\item Extract from syntax finite map of variables
-to relative-index verctors representing the contributing
-array subscritps
-\end{itemize}
+For our example, applying \textsf{regions} again yields the same
+result, hence we have reached a fixed point.
 

### 2016-06-19 paper.tex
typo
redoing the equational theory

@@
   \texttt{dim=}#2)}
+\newcommand{\stenFwdS}[2]{\texttt{fwd} \, (\texttt{depth=}#1,
+  \texttt{dim=}#2)}
+\newcommand{\stenBwdS}[2]{\texttt{bwd} \, (\texttt{depth=}#1,
+  \texttt{dim=}#2)}
+\newcommand{\stenCenS}[2]{\texttt{cen} \, (\texttt{depth=}#1,
+  \texttt{dim=}#2)}
+
+% SYNTAX OPERATIONS AND PREDICATES
+\newcommand{\neigh}{\textsf{neigh}}
+\newcommand{\arrayTy}{\textsf{array}}
+\newcommand{\rhsExp}{\textsf{rhsExp}}
+\newcommand{\var}{\textsf{var}}
+
+%% VECTOR NOTATIONS
+\newcommand{\vect}[1]{\textbf{#1}}
+\newcommand{\vtwo}[2]{\setlength{\arraycolsep}{0em}
+\left[\begin{array}{l}$#1$\\$#2$\end{array}\right]}
+\newcommand{\vthree}[3]{\setlength{\arraycolsep}{0em}
+\left[\begin{array}{l}$#1$\\$#2$\\$#3$\end{array}\right]}
+\newcommand{\stwo}[4]
+%{\vtwo{#1}{#2}\!\vtwo{#3}{#4}}
+{\setlength{\arraycolsep}{0.1em}
+\left[\begin{array}{rr}$#1$ & $#3$\\$#2$ & $#4$\end{array}\right]}
+
+\newcommand{\singleEntry}[2]{\textbf{J}_{#2}^{#1}}
+
+%% OPERATIONS ON SPANS and VECTORS 
+\newcommand{\containedin}{\sqsubseteq}
 
@@
 \nonterm{approxMod} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
-\nonterm{region} ::= & \term{forward}, \term{depth=}\mathbb{N}, \term{dim=}\dimTy  \\
-\mid \; & \term{backward}, \term{depth=}\mathbb{N}, \term{dim=}\dimTy \\
-\mid \; & \term{centered}, \term{depth=}\mathbb{N}, \term{dim=}\dimTy \\
+\nonterm{region} ::= & \term{forward}, \term{depth=}\mathbb{N}_{>0}, \term{dim=}\dimTy  \\
+\mid \; & \term{backward}, \term{depth=}\mathbb{N}_{>0}, \term{dim=}\dimTy \\
+\mid \; & \term{centered}, \term{depth=}\mathbb{N}_{>0}, \term{dim=}\dimTy \\
 \mid \; & \nonterm{region} \; \term{+} \; \nonterm{region} \\
@@
 \mid \; & \nonterm{rvar}  \\[0.5em]
-\nonterm{temporal} ::= \; & \term{dependency} \; (v \; \{ , v \})
+\nonterm{temporal} ::= \; & \term{dependency} \; (v \; \{ , v \}) [, \texttt{mutual}]
   \\[0.5em]
@@
 \begin{align*}
-\hspace{-1em}
 \setlength{\arraycolsep}{0.05em}
 \begin{array}{c}
-\framebox{$\overline{\texttt{v} : S} \equiv \overline{\texttt{u} : T}$} \\[1em]
+\framebox{$\textit{region} \equiv \textit{region}'$} \\[1em]
+\begin{array}{rll}
+(\textsc{F\; \texttt{+} \;F}) \;\; &
+\stenFwdS{n \, \textsf{max} \, m}{d} \\
+ \equiv \; & \stenFwdS{n}{d} \; \texttt{+} \; \stenFwdS{m}{d} \\[1em]
+%
+(\textsc{B\; \texttt{+} \;B}) \;\; &
+\stenBwdS{n \, \textsf{max} \, m}{d} \\
+ \equiv \; & \stenBwdS{n}{d} \; \texttt{+} \; \stenBwdS{m}{d} \\[1em]
+%
+(\textsc{C\; \texttt{+} \;C}) \;\; &
+\stenCenS{n \, \textsf{max} \, m}{d} \\
+\equiv \; & \stenCenS{n}{d} \; \texttt{+} \; \stenCenS{m}{d} \\[1em]
+%
+(\textsc{C\; \texttt{+} \;F}) \;\; & \stenCenS{n}{d} \\
+\textit{$m \leq n$} \; \equiv \; & \stenCenS{n}{d} \; \texttt{+} \;
+                      \stenFwdS{m}{d} \\[1em]
+%
+(\textsc{C\; \texttt{+} \;B}) \;\; & 
+\stenCenS{n}{d} \\
+\textit{$m \leq n$} \; \equiv \; & \stenCenS{n}{d} \; \texttt{+} \;
+                      \stenBwdS{m}{d} \\[1em]
+(\textsc{B\; \texttt{+} \;F}) \;\; & 
+\stenCenS{n}{d} \\
+\equiv \; & \stenFwdS{n}{d} \; \texttt{+} \; \stenBwdS{n}{d} \\[1em]
+%
+(\texttt{+}\textsc{comm}) \;\; & S \; \texttt{+} \; T \; \equiv \; T \;
+                       \texttt{+} \; S \\[0.5em]
+%
+(\texttt{+}\textsc{assoc}) \;\; & R \; \texttt{+} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
+                       \texttt{+} \; S) \; \texttt{+} \; T \\[0.5em]
+(\texttt{*}\textsc{comm}) \;\; & S \; \texttt{*} \; T \; \equiv \; T \;
+                       \texttt{*} \; S \\[0.5em]
+%
+(\texttt{*}\textsc{assoc}) \;\; & R \; \texttt{*} \; (S \; \texttt{*} \; T) \; \equiv \; (R \;
+                       \texttt{*} \; S) \; \texttt{*} \; T \\[0.5em] 
+(\textsc{dist}) \;\; & R \; \texttt{*} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
+                       \texttt{*} \; S) \; \texttt{+} \; (R
+                       \; \texttt{*} \; T) \\[0.5em] 
+(\textsc{inter}) \;\; & (S \; \texttt{+} \; S') \; \texttt{*} \;
+                              (T \; \texttt{+} \; T') \\
+ \equiv \; & (S \;  \texttt{*} \; S) 
+ \; \texttt{+} \; (S \; \texttt{*} \; T') 
+ \; \texttt{+} \; (S' \; \texttt{*} \; T)
+ \; \texttt{+} \; (S' \; \texttt{*} T')  \\[0.5em] 
+\end{array} \\ \\
+%
+\framebox{$\overline{\textit{specDec} :: \texttt{v}} \equiv 
+\overline{\textit{specDec'} :: \texttt{v}}$} \\[1em]
 \begin{array}{rl}
-(\textsc{coalesceF}) \;\; & (\texttt{a} : \texttt{fwd} \;
-                       \texttt{dims}=ds \; \texttt{depth}=n; \\
-&  \;\, \texttt{a} :  \texttt{fwd} \; \texttt{dims}=ds \; \texttt{depth}=n+1)\\
-\equiv \; & \texttt{a} :  \texttt{fwd} \; \texttt{depth}=n+1,
-         \texttt{dims}=ds \\[0.75em]
-(\textsc{coalesceB}) \;\; & (\texttt{a} : \texttt{bwd} \;
-                       \texttt{dims}=ds \; \texttt{depth}=n; \\
-&  \;\, \texttt{a} :  \texttt{bwd} \; \texttt{dims}=ds \; \texttt{depth}=n+1) \\
-\equiv \; & \texttt{a} :  \texttt{bwd} \; \texttt{depth}=n+1,
-         \texttt{dims}=ds \\[0.75em]
-(\textsc{idem}) \;\; & (\texttt{a} : S; \texttt{a} : S) \equiv (\texttt{a}
-                  : S)
-\end{array} \\ \\
-\framebox{$\texttt{v} : S \equiv \texttt{v'} : S'$} \\[1em]
-\begin{array}{ll}
-(\texttt{a} : \; \texttt{bwd} \; \texttt{depth=1\{b\}} \;
-  \texttt{dim=t})
-\, & \equiv \,
-(\texttt{b} : \; \texttt{bwd} \; \texttt{depth=1\{a\}} \;
-  \texttt{dim=t})
+(\textsc{mutual}) \; & 
+\texttt{stencil} \; \texttt{dependency} (\texttt{v}), \texttt{mutual}
+  :: \texttt{w} 
 \\
-(\texttt{a} : \; \texttt{fwd} \; \texttt{depth=1\{b\}}\;
-  \texttt{dim=t})
-\, & \equiv \,
-(\texttt{b} : \; \texttt{fwd} \; \texttt{depth=1\{a\}} \;
-  \texttt{dim=t})
-\end{array}
-\\ \\
-\framebox{$S \equiv S'$} \\[0.75em]
-\begin{array}{ll}
-(a : \texttt{read-once} \; S) \; & \equiv \; (a : \texttt{read-once} \;\,
-  \texttt{read-once} \; \, S)
+\equiv \; & \texttt{stencil} \; \texttt{dependency} (\texttt{w}), \texttt{mutual} ::
+  \texttt{v} \\[0.5em]
+(\textsc{coalesce}) \; & 
+\texttt{stencil} \; \texttt{dependency} (\bar{v}) 
+  :: \texttt{v} \\
+& ; \texttt{stencil} \; \texttt{dependency} (\bar{w}) 
+  :: \texttt{v} 
+\\
+\equiv \; & \texttt{stencil} \; \texttt{dependency} (\bar{v}, \bar{w}) ::
+  \texttt{v} \\[0.5em]
+(\textsc{spatial}) \; &
+(\textit{region} \equiv \textit{region'}) \\
+\Rightarrow \; & (\texttt{stencil} \; [\textit{mod}] \; [\textit{approxMod},] \;
+\textit{region} \\
+& \!\!\equiv \; \texttt{stencil} \; [\textit{mod}] \;
+            [\textit{approxMod},] \;
+\textit{region}') \\
 \end{array}
@@
 \end{figure}
-
-\begin{theorem}[Soundness]
-\[
-\overline{\texttt{v} : S}\equiv \overline{\texttt{u} : T}
-\; \Rightarrow \;
-\interp{\overline{\texttt{v} : S}} = \interp{\overline{\texttt{u} : T}}
-\]
-\end{theorem}
-
-\paragraph{Proof} (see Appendix)
 
@@
 
-\begin{theorem}[Soundness]
-\[
-\overline{\texttt{v} : S} <: \overline{\texttt{u} : T}
-\; \Rightarrow \;
-\interp{\overline{\texttt{v} : S}} \subseteq \interp{\overline{\texttt{u} : T}}
-\]
-\end{theorem}
-
-\paragraph{Proof} (see Appendix)
-
 \section{Semantics of specifications; a model}
@@
 \end{figure}
+
+\begin{theorem}[Soundness]
+\[
+\overline{\texttt{v} : S}\equiv \overline{\texttt{u} : T}
+\; \Rightarrow \;
+\interp{\overline{\texttt{v} : S}} = \interp{\overline{\texttt{u} : T}}
+\]
+\end{theorem}
+
+\paragraph{Proof} (see Appendix)
+
+\begin{theorem}[Soundness]
+\[
+\overline{\texttt{v} : S} <: \overline{\texttt{u} : T}
+\; \Rightarrow \;
+\interp{\overline{\texttt{v} : S}} \subseteq \interp{\overline{\texttt{u} : T}}
+\]
+\end{theorem}
+
+\paragraph{Proof} (see Appendix)
 
@@
 Section~\ref{sec:semantics}.
-
-% SYNTAX OPERATIONS AND PREDICATES
-\newcommand{\neigh}{\textsf{neigh}}
-\newcommand{\arrayTy}{\textsf{array}}
-\newcommand{\rhsExp}{\textsf{rhsExp}}
-\newcommand{\var}{\textsf{var}}
-
-%% VECTOR NOTATIONS
-\newcommand{\vect}[1]{\textbf{#1}}
-\newcommand{\vtwo}[2]{\setlength{\arraycolsep}{0em}
-\left[\begin{array}{l}$#1$\\$#2$\end{array}\right]}
-\newcommand{\stwo}[4]
-%{\vtwo{#1}{#2}\!\vtwo{#3}{#4}}
-{\setlength{\arraycolsep}{0.1em}
-\left[\begin{array}{rr}$#1$ & $#3$\\$#2$ & $#4$\end{array}\right]}
-
-%% OPERATIONS ON SPANS and VECTORS 
-\newcommand{\containedin}{\sqsubseteq}
 
@@
 %%
-\item Fold each list pairwise by the following partial $\bullet$ which 
-coalescees and contiguous regions:
+\item Fold each list pairwise by the following partial operation
+ $\bullet$ which coalescees contiguous regions:
 %

### 2016-06-19 paper.tex
expanding section on explaining the syntax

@@
 
-\CustomVerbatimEnvironment{SpecVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.65cm,
-xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.98}
-\CustomVerbatimEnvironment{ExmVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.65cm,
-xrightmargin=0.2cm,baselinestretch=0.98,numbers=left}
-\CustomVerbatimEnvironment{IVerbatim}{Verbatim}{fontsize=\relsize{-1},xleftmargin=0.65cm,
-xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.98}
+\CustomVerbatimEnvironment{SpecVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.5cm,
+xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.98,numbersep=0.9em}
+\CustomVerbatimEnvironment{ExmVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.5cm,
+xrightmargin=0.2cm,baselinestretch=0.98,numbers=left,numbersep=0.9em}
+\CustomVerbatimEnvironment{IVerbatim}{Verbatim}{fontsize=\relsize{-1},xleftmargin=0.5cm,
+xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.98,numbersep=0.9em}
 
@@
 
-\newcommand{\stenFwd}[2]{\texttt{forward} \, (\texttt{depth=}#1,
-  \texttt{dim=}#2)}
-\newcommand{\stenBwd}[2]{\texttt{backward} \, (\texttt{depth=}#1,
-  \texttt{dim=}#2)}
-\newcommand{\stenCen}[2]{\texttt{centered} \, (\texttt{depth=}#1,
-  \texttt{dim=}#2)}
-\newcommand{\stenFwdS}[2]{\texttt{fwd} \, (\texttt{depth=}#1,
-  \texttt{dim=}#2)}
-\newcommand{\stenBwdS}[2]{\texttt{bwd} \, (\texttt{depth=}#1,
-  \texttt{dim=}#2)}
-\newcommand{\stenCenS}[2]{\texttt{cen} \, (\texttt{depth=}#1,
-  \texttt{dim=}#2)}
+\newcommand{\stenFwd}[2]{\term{forward} \, (\term{depth=}#1,
+  \term{dim=}#2)}
+\newcommand{\stenBwd}[2]{\term{backward} \, (\term{depth=}#1,
+  \term{dim=}#2)}
+\newcommand{\stenCen}[2]{\term{centered} \, (\term{depth=}#1,
+  \term{dim=}#2)}
+\newcommand{\stenFwdS}[2]{\term{fwd} \, (\term{depth=}#1,
+  \term{dim=}#2)}
+\newcommand{\stenBwdS}[2]{\term{bwd} \, (\term{depth=}#1,
+  \term{dim=}#2)}
+\newcommand{\stenCenS}[2]{\term{cen} \, (\term{depth=}#1,
+  \term{dim=}#2)}
 
@@
 
-\subsection{Syntax}
+\subsection{Notation and convention}
+\label{sec:notation}
+
+\renewcommand*{\arraystretch}{0.8}
+\paragraph{Target language syntax} Throughout, for the source language
+syntax, $v$ ranges over program variables, $s$ over statements, and
+$e$ over expressions (which may be impure).  By ``variable'', we mean
+in the imperative sense (\ie{}, named binders to memory cells) instead
+of mathematical variables. We make clear a number of standard notions
+used throughout.
+
+\begin{definition}[Base induction variable]
+  An variable \textit{v} of integer type is a \emph{base induction
+    variable} if it is the control variable of a ``for''
+  loop\footnote{Or equivalent for the target language, \eg{},
+    \texttt{do} in Fortan for our implementation}, incremented by $1$ per
+iteration. The variable is marked as a base induction variable
+only within the scope of the loop, and such variables will be ranged
+over by $i$, $j$, $k$. Since we do not use non-base induction
+variables (which are traditionally affine transformations of induction
+variables) we often simply say \emph{induction variable}. 
+\end{definition}
+
+\begin{definition}[Array subscript]
+  An \emph{array subscript} is an expression that indexes an array
+  (for the purpose or reading or writing an element), which we denote
+  by $v(\bar{e})$ for the source language where $\bar{e}$ is shorthand
+  for a syntactic list of indexing expressions. A \emph{relative
+    index} is a list $\bar{e}$ where each $e \in \bar{e}$ is defined
+  in terms of some base induction variable.
+\end{definition}
+
+\begin{definition}[Neighbourhood index and neighbourhood subscripts]
+  For an array subscript $v(\bar{e})$, we say say that $e \in \bar{e}$
+  is a \emph{neighbourhood index} if it is of the form 
+  $e \equiv i$, $e \equiv i + a$ or $e \equiv i - a$ where $a$ is a
+  constant of integer type. That is, a neighbourhood index is a
+  relative index defined as a constant translation of a base induction
+  variable. The relation $\equiv$ identifies terms up-to commutativity
+  of $+$ and the inverse relation of $+$ and $-$ (\eg{},
+  $(-b) + i \equiv i - b$).  We classify neighbourhood indices using
+  the predicate \neigh{}. An array subscript $v(\var{e})$ where every
+ $e \in \bar{e}$ is a neighbourhood index is called a 
+ \emph{neighbourhood subscript}. 
+\end{definition}
+
+\subsection{Specification syntax}
 \label{sec:syntax}
 
-Figure~\ref{fig:syntax}
-gives the syntax, which we introduces in stages below.
-The top-level is given by the \textit{spec} production
-which splits into either a \textit{spatial} specification or
-a \textit{temporal specification}.
-%%
+Figure~\ref{fig:syntax} gives the syntax of stencil specifications,
+which we introduces in stages below.  The top-level is given by the
+\nonterm{specification} production which splits into either a
+\nonterm{regionDec} (region declaration) or a \nonterm{specDec}
+(specification declaration). Specification declarations associate a
+specification to one or more program variables $\bar{v}$ (here given
+using the Fortran style syntax of $\ldots \; \texttt{::} \bar{v}$).
+Specification declarations are either \nonterm{spatial} or
+\nonterm{temporal} specifications and describe, for spatial
+specifcations how an array variable $v$ is read, and for temporal
+specifications what variables are used to define the array variable $v$.
+%%
+
+\subsubsection{Spatial specification syntax}
+
+\emph{Regions} are the central building blocks of spatial
+specifications. Regions can either be declared along with a
+\nonterm{regionDec}, assigning a region specification \nonterm{region} to
+a region variable \nonterm{rvar} or given directly within a
+\nonterm{spatial} specification.
+
+\paragraph{Region constants}
+
+Regions have as terminals the \emph{region constants}, denoted by 
+\term{forward}, \term{backward}, or \term{centered}. Each
+region constant is given a depth parameter $n$ (natural number greater
+than 0) and a dimension identifier $d$ (also a natural number greater than
+0). Region constants specify that an array is read from by a collection
+of indices which in the $d^{th}$ dimension are 
+neighbourhood indices ranging from $i + 0$ up to $i + n$ inclusively, \eg{}, the
+following is valid stencil, reading from \term{a} and writing to \term{b}:
+%%
+\begin{ExmVerbatim}
+!= stencil forward(depth=2, dim=1) :: a
+b(i, 0) = a(i, 0) + a(i+1, 0) + a(i+2, 0)
+\end{ExmVerbatim}
+%%
+were \texttt{i} is an induction variable.  The stencil specification
+is associated to the array variable \texttt{a}. Note that the second
+dimension is fixed at a constant here.
+
+A \term{backward} stencil
+is similar but has negative neighbourhood indices, \eg{}, 
+%
+\begin{ExmVerbatim}
+!= stencil backward(depth=2, dim=1) :: a
+b(i+1) = a(i) + a(i-1) + a(i-2)
+\end{ExmVerbatim}
+%
+Note that the left-hand subscript on \term{b} does not affect
+the stencil specification. The requirement on the left-hand side
+is merely that it is an array subscript with at least one neighbourhood index,
+but this may be on different induction variables to the right-hand
+side, which is the stencil we are concerened with.
+
+A \texttt{centered} stencil is a combination of \texttt{forward}
+and \texttt{backward} stencils to the same depth, \eg{},
+%%
+\begin{ExmVerbatim}
+!= stencil centered(depth=1, dim=1) :: a
+b(i) = ( a(i) + a(i-1) + a(i+1) ) / 3.0
+\end{ExmVerbatim}
 
@@
 \nonterm{specification} ::= & \nonterm{regionDec} \mid \nonterm{specDec} \\
-\nonterm{specDec} ::= & \term{stencil} \; \nonterm{spec}
+\nonterm{specDec} ::= & \term{stencil} \; \nonterm{spec} \;
+                        \texttt{::} \; v
   \\
@@
 \nonterm{approxMod} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
-\nonterm{region} ::= & \term{forward}, \term{depth=}\mathbb{N}_{>0}, \term{dim=}\dimTy  \\
-\mid \; & \term{backward}, \term{depth=}\mathbb{N}_{>0}, \term{dim=}\dimTy \\
-\mid \; & \term{centered}, \term{depth=}\mathbb{N}_{>0}, \term{dim=}\dimTy \\
+\nonterm{region} ::= & \stenFwd{\mathbb{N}_{>0}}{\dimTy}  \\
+\mid \; & \stenBwd{\mathbb{N}_{>0}}{\dimTy} \\
+\mid \; & \stenCen{\mathbb{N}_{>0}}{\dimTy} \\
 \mid \; & \nonterm{region} \; \term{+} \; \nonterm{region} \\
@@
 
-\paragraph{Spatial specifications}
-
-\paragraph{Temporal specifications}
+\paragraph{Sum and product of regions}
+%%
+Region terms can be combined using the sum operator
+\term{+} or the product operator \term{*}.
+
+The sum of two regions $r \texttt{+} r'$ specifies that an array is
+read using the neighbourhood indices described by both $r$ and
+$r'$. It can be thought of as a conjunction of speicifcations. For
+example, the following gives the specification for the five-point
+stencil which is the sum of two \texttt{centered} regions in each
+dimension:
+%%
+\begin{ExmVerbatim}
+!= stencil centered(depth=1, dim=1) 
+         + centered(depth=1, dim=2) :: a
+
+b(i,j) = -4*a(i,j) + a(i-1,j) + a(i+1,j) 
+                   + a(i,j-1) + a(i,j+1) 
+\end{ExmVerbatim}
+%%
+The product of two regions $r \texttt{*} r'$ specifies that the array
+is read with neighbourhood indices drawn from the bounding box created
+by the regions two regions $r$ and $r'$. For example, the following
+code (defining a nine-point stencil) has a specification as the
+product of two \texttt{centered} regions in each dimension:
+%%
+\begin{ExmVerbatim}
+!= stencil centered(depth=1, dim=1) 
+         * centered(depth=1, dim=2) :: a
+
+x = a(i, j)   + a(i-1, j)   + a(i+1, j)
+y = a(i, j-1) + a(i-1, j-1) + a(i+1, j-1)
+z = a(i, j+1) + a(i-1, j+1) + a(i+1, j+1)
+b(i, j) = (x + y + z) / 9.0
+\end{ExmVerbatim}
+%%
+Note that there are a number of intermediate assignments over which 
+the whole specification ranges.
+
+\paragraph{Region declarations and variables}
+
+Region specifications can be assigned to region variables via
+a region declaration, which can be used later to form a spatial
+specification. For example, the speicifcation from the previous
+ example can be restated as:
+%%
+\begin{ExmVerbatim}
+!= region r1 = centered(depth=1, dim=1) 
+!= region r2 = centered(depth=1, dim=2)
+!= stencil r1*r2 :: a
+\end{ExmVerbatim}
+%%
+\paragraph{Spatial modifiers}
+%%
+There are three kinds of modifier that can be given to spatial 
+specifications: \texttt{readOnce}, \texttt{reflexive}, 
+\texttt{irreflexive}.
+
+The \texttt{readOnce} modifier enforces that no index appears
+more than once. For example, in all of the previous examples the
+\texttt{readOnce} modifier could be added, \eg{}
+%
+\begin{ExmVerbatim}
+!= stencil readOnce, backward(depth=2, dim=1) :: a
+b(i+1) = a(i) + a(i-1) + a(i-2)
+\end{ExmVerbatim}
+%
+However, this would be an invalid specification if any of the 
+array subscripts was reapeated. This modifier provides a way to
+ rule out any accidental repetition of array subscripts. 
+The notion correponds to that of \emph{linearity}, though we
+opt for the more informative and easily understood name
+\texttt{readOnce}. This modifier is optional, so it need not
+be present even if the stencil is linear.
+
+The \texttt{irreflexive} modifier is parameteresed by a
+comma-separated list of dimensions, indicating that for those
+dimensions there is no neighbour index with $0$ offset, \eg{}, 
+%
+\begin{ExmVerbatim}
+!= region r = centered(depth=1, dim=1)
+!= stencil irreflexive(dim=1), r :: a
+b(i) = a(i+1) + a(i-1)
+\end{ExmVerbatim}
+%
+or with just a \texttt{forward} region an example would be: 
+%
+\begin{ExmVerbatim}
+!= region r = forward(depth=1, dim=1)
+!= stencil irreflexive(dim=1), r :: a
+b(i) = a(i+1)
+\end{ExmVerbatim}
+
+The \term{reflexive} modifier is dual to \term{irreflexive},
+indicating dimensions in which there is a neighbour index with $0$
+offset. The region constants \term{forward}, \term{backward},
+\term{centered} implicitly include the \term{reflexive} modifier,
+but it is useful when there is a dimension in which there is reflexive
+indexing but not a region constant, \eg{},
+%%
+\begin{ExmVerbatim}
+!= region r = centered(depth=1, dim=1)
+!= stencil reflexive(dims=2), r :: a
+b(i,j) = a(i-1,j) + a(i+1,j) + a(i, j) 
+\end{ExmVerbatim}
+%%
+Overlapping \term{irreflexive} and \term{reflexive} modifiers
+cannot occur together-- this is an invalid specification.
+
+\paragraph{Under and over-approximation}
+
+In some cases, it is useful to give a lower and/or upper bound for a
+stencil. This can be done using either the \term{atMost} or
+\term{atLeast} modifiers. This is particularly useful in situations
+where there is a non-contiguous stencil, which does not fit the rest
+of the specification syntax.  For example:
+%%
+\begin{ExmVerbatim}
+!= stencil atMost, forward(depth=2, dim=1) :: a
+b(i) = a(i) + a(i+2) 
+\end{ExmVerbatim}
+%%
+
+\subsubsection{Temporal specifications}
 
@@
 
-\renewcommand*{\arraystretch}{0.8}
-\paragraph{Basic syntax definitions} In the following, $v$
-  ranges over variables, $s$ over statements, and $e$ over expressions
-  (which may be impure).  By ``variable'', we mean in the imperative
-  sense (\ie{}, named binders to memory cells) instead of mathematical
-  variables. The function \rhsExp{} maps statements to a set of
-  expressions which occur in right-hand positions (\ie{}, not the
+\paragraph{Syntax functions}
+The function \rhsExp{} maps statements to a
+set of expressions which occur in right-hand positions (\ie{}, not the
 target of an assignment). The function \var{} maps expressions to a
 set of the variables used in right-hand positions.
-
-\begin{definition}[Standard: base induction variable]
-An integer variable \textit{i} is a \emph{base induction variable} 
-within the scope of a ``for'' loop (or equivalent for the target
-language, \eg{}, \texttt{do} in Fortan for our implementation) if it
-is the control variable of the loop, incremented by $1$ per
-iteration. 
-\end{definition}
-
-\begin{definition}[Standard: array subscript]
-An \emph{array subscript} is an expression index an array, which in
-our abstract we write as $v(\bar{e})$ where $\bar{e}$ is shorthand for
-a syntactic list of indexing expressions. 
-\end{definition}
-
-\begin{definition}[Neighbourhood index]
-An array subscript $e$ is a \emph{neighbourhood index} if,
-for an induction variable $i$, then $e \equiv i + a$ or $e \equiv i - a$
-where $a$ is a constant of integer type. The relation $\equiv$ identifies terms
-up-to commutativity of $+$ and the inverse
-relation of $+$ and $-$ (\eg{}, $(-b) + i \equiv i - b$).
-In the following, we classify neighbourhood indices 
-using the predicate \neigh{}.
-\end{definition}
 

### 2016-06-20 paper.tex
section on equations and inequations

@@
 \usepackage{multirow}
+
+\definecolor{darkblue}{rgb}{0.0,0.0,0.5}
+\definecolor{darkgreen}{rgb}{0.0,0.4,0.0}
+
+\usepackage[]{hyperref}
+\hypersetup{
+    unicode=false,          % non-Latin characters in Acrobat's bookmarks
+    pdftoolbar=true,        % show Acrobat toolbar?
+    pdfmenubar=true,        % show Acrobat menu?
+    pdffitwindow=false,      % page fit to window when opened
+    pdftitle={},    % title
+    pdfauthor={}
+    pdfsubject={},   % subject of the document
+    pdfnewwindow=true,      % links in new window
+    pdfkeywords={keywords}, % list of keywords
+    colorlinks=true,       % false: boxed links; true: colored links
+    linkcolor=darkblue,          % color of internal links
+    citecolor=darkblue,        % color of links to bibliography
+    filecolor=green,      % color of file links
+    urlcolor=blue,          % color of external links
+}
 
@@
 
-\bibliographystyle{plain}
+\bibliographystyle{abbrvnat}
 
@@
 \begin{ExmVerbatim}
+!= stencil atLeast, reflexive(dims=1)      :: a
 != stencil atMost, forward(depth=2, dim=1) :: a
@@
 %%
-
 \subsubsection{Temporal specifications}
+
+\subsection{Note on the design}
+
+The names ``forward'', ``backward'' and ``centered''
+are inspired by standard terminology in numerical analysis
+for the shape of discretisation schemes. For example,
+the standard ``explicit method'' for approximation
+PDEs is the \emph{Forward Time, Centered Space} (FTCS)
+scheme~\cite{dawson1991finite}. For
+the one-dimensional heat equations, an FTCS discretisation 
+provides approximation code with the following stencil~\cite{recktenwald2004finite}:
+%%
+\begin{ExmVerbatim}
+do i=2:n-1
+  u(i) = r*v(i-1) + r2*v(i) + r*v(i+1)
+end do
+\end{ExmVerbatim}
+%%
+where \texttt{r} and \texttt{r2} are some constants. 
+A valid specification for this stencil is then:
+%%
+\begin{ExmVerbatim}
+!= stencil centered(depth=1, dim=1) 
+         + centered(depth=1, dim=2) :: v
+\end{ExmVerbatim}
+%%
+Such a specification can be inferred, or can be inserted by a user
+and checked against the code.
 
@@
 \label{sec:eqs}
-
-Figure~\ref{fig:equations}
 
@@
 %
-\framebox{$\overline{\textit{specDec} :: \texttt{v}} \equiv 
-\overline{\textit{specDec'} :: \texttt{v}}$} \\[1em]
+\framebox{${\textit{specDec} :: \texttt{v}} \equiv 
+{\textit{specDec'} :: \texttt{v}}$} \\[1em]
 \begin{array}{rl}
@@
   \texttt{v} \\[0.5em]
-(\textsc{coalesce}) \; & 
-\texttt{stencil} \; \texttt{dependency} (\bar{v}) 
-  :: \texttt{v} \\
-& ; \texttt{stencil} \; \texttt{dependency} (\bar{w}) 
-  :: \texttt{v} 
-\\
-\equiv \; & \texttt{stencil} \; \texttt{dependency} (\bar{v}, \bar{w}) ::
-  \texttt{v} \\[0.5em]
+%(\textsc{coalesce}) \; & 
+%\texttt{stencil} \; \texttt{dependency} (\bar{v}) 
+%  :: \texttt{v} \\
+%& ; \texttt{stencil} \; \texttt{dependency} (\bar{w}) 
+%  :: \texttt{v} 
+%\\
+%\equiv \; & \texttt{stencil} \; \texttt{dependency} (\bar{v}, \bar{w}) ::
+%  \texttt{v} \\[0.5em]
 (\textsc{spatial}) \; &
@@
 
-\subsubsection{Safety and sub-specifications}
-
-Figure~\ref{fig:inequations}
+\paragraph{Equations}
+Figure~\ref{fig:equations} lists the equational theory for
+specifications. This is broken down into equations on regions only,
+given by the equivalence relation $\textit{region} \equiv \textit{region}'$ and on
+on specification declarations assigned to variables via the equivalence relation
+$\textit{specDec} :: \texttt{v} \equiv
+\textit{specDec'} :: \texttt{v}$.
+
+To save space we use abbreviations 
+\term{fwd}, \term{bwd}, and \term{cen} for \term{forward},
+\term{backward}, and \term{centered} regions respectively.  The
+equations reveal something of their semantics.  The first six
+equations (with labels of the form $\ldots \texttt{+} \ldots$) 
+explain the behaviour of overlapping regions with sum
+\term{+}. The next six equations explain the algebraic behaviour of
+\term{+} and \term{*}.  Together, we see that \term{+} is an
+associative, commutative, idempotent binary operator that distributes
+with \term{*}, which is associative and commutative. This distribution
+of \term{*} over \term{+} is key as it is used internally to give a
+normal form for stencil specifications akin to Disjunctive Normal Form
+(DNF) (where \term{+} is taken as ``disjunction'' and \term{*} as the
+``conjunction''). We revisit this normalisation in Section~\ref{}.
 
 \begin{figure}[t]
-{\framebox{
 \begin{minipage}{0.88\linewidth}
@@
   T}(\textsc{trans}) \\[1.5em]
-\dfrac{}{S <: \; \texttt{unspecified dims=}ds}(\top)
 \\[1.5em]
@@
 \setlength{\arraycolsep}{0.1em}
-\dfrac{\hspace{3em} ds \subseteq es \; \wedge \; n \leq m \hspace{3em}}
+\dfrac{\hspace{3em} m \leq n \hspace{3em}}
 {\begin{array}{rl}
-\texttt{fwd} \; \texttt{depth=}n \; \texttt{dim=}ds & <: \texttt{fwd} \;
-  \texttt{depth=}m \; \texttt{dim=}es
+\stenFwdS{m}{d} & \stenFwdS{n}{d}
 %& \;\, ds \subseteq es \wedge n \leq m
 \\
-\wedge \; \texttt{bwd} \; \texttt{depth=}n \; \texttt{dim=}ds & <: \texttt{bwd} \;
-  \texttt{depth=}m \; \texttt{dim=}es
+\wedge \stenBwdS{m}{d} & <: \stenBwdS{n}{d}
 %& \;\, ds \subseteq es \wedge n \leq m
 \\
-\wedge \; \texttt{fwd} \; \texttt{depth=}n \; \texttt{dim=}ds & <: \texttt{sym} \;
-  \texttt{depth=}m \; \texttt{dim=}es
+\wedge \; \stenCenS{m}{d} & <: \stenCenS{n}{d}
 %& \;\, ds \subseteq es \wedge n \leq m
 \\
-\wedge \; \texttt{bwd} \; \texttt{depth=}n \; \texttt{dim=}ds & <: \texttt{sym} \;
-  \texttt{depth=}m \; \texttt{dim=}es
+\wedge \; \stenBwdS{m}{d} & <: \stenCenS{n}{d} \\
+\wedge \; \stenFwdS{m}{d} & <: \stenCenS{n}{d} \\
 \end{array}}
@@
 \end{align*}
-\end{minipage}}}
-\caption{Definition of sub-specification relation}
+\end{minipage}
+\caption{Inequations on specifications}
 \label{fig:inequations}
 \end{figure}
+
+\paragraph{Inequations: sub-specifications}
+
+Figure~\ref{fig:inequations}
 
@@
 \paragraph{Proof} (see Appendix)
-
-\section{Examples and common patterns}
-
-% Implicit vs explicit differentiation methods
-% (forward implicit needs 'solver')
-% backward - Euler
-% runge-kutta 2nd-order
-
-\begin{example}[FTCS (Forward time, centered space)]
-
-  Otherwise known as the \emph{explicit method}, is a common
-  pattern since it tends to be more efficient than other approches
-  (see BTCS and CTCS below). However, its disadvantage is that it is
-  unstable if the resolution is not well chosen.
-
-\begin{ExmVerbatim}
-b(x) = a(x) + r*(a(x-1) - 2*a(x) + a(x+1))
-\end{ExmVerbatim}
-%
-For this statement, the inference provides the specification:
-%
-\begin{SpecVerbatim}
-!=  stencil centered, depth=1, dim=0 :: a
-\end{SpecVerbatim}
-
-\end{example}
-
-\paragraph{BTCS (Backward time, centered space) -- implicit method}
-
-\paragraph{CTCS (Centered time, centered space) - Crank-Nicolson method}
 

### 2016-06-20 paper.tex
stripped out some of temporal parts

@@
 specification to one or more program variables $\bar{v}$ (here given
-using the Fortran style syntax of $\ldots \; \texttt{::} \bar{v}$).
-Specification declarations are either \nonterm{spatial} or
-\nonterm{temporal} specifications and describe, for spatial
-specifcations how an array variable $v$ is read, and for temporal
-specifications what variables are used to define the array variable $v$.
-%%
-
-\subsubsection{Spatial specification syntax}
+using the Fortran style syntax of $\ldots \; \texttt{::} \bar{v}$),
+describing how the array variables $\bar{v}$ are acessed.
+%Specification declarations are either \nonterm{spatial} or
+%\nonterm{temporal} specifications and describe, for spatial
+%specifcations,
+%Specifications declarations describe 
+%how an array variable $v$ is read
+%, and for temporal
+%specifications what variables are used to define the array variable $v$.
+%%
+
+%\subsubsection{Spatial specification syntax}
 
@@
 \nonterm{regionDec} ::= &  \texttt{region} \; \nonterm{rvar} \; \texttt{=} \;
-                         \nonterm{region}\\
-\nonterm{spec} ::= & \nonterm{spatial} \mid \nonterm{temporal}   \\[1em]
-\nonterm{spatial} ::= & [\nonterm{mod}] \;
+                         \nonterm{region}\\[1em]
+%\nonterm{spec} ::= & \nonterm{spatial} \mid \nonterm{temporal}   \\[1em]
+\nonterm{spec} ::= & [\nonterm{mod}] \;
                        [\nonterm{approxMod},] \; [\nonterm{region}] \\
@@
  \term{readOnce} \\
-\nonterm{modRefl} ::= & \term{reflexive}, \; \term{dims=}\dimTy \{\term{,} \; \dimTy\} \\
-\nonterm{modIrrefl} ::= & \term{irreflexive}, \; \term{dims=}\dimTy \{\term{,}
-          \dimTy\}\\
+\nonterm{modRefl} ::= & \term{reflexive} \, (\term{dims=}\dimTy \{\term{,} \; \dimTy\}) \\
+\nonterm{modIrrefl} ::= & \term{irreflexive} \, (\term{dims=}\dimTy \{\term{,}
+          \dimTy\}) \\
 \nonterm{approxMod} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
@@
 \mid \; & \nonterm{rvar}  \\[0.5em]
-\nonterm{temporal} ::= \; & \term{dependency} \; (v \; \{ , v \}) [, \texttt{mutual}]
-  \\[0.5em]
+%\nonterm{temporal} ::= \; & \term{dependency} \; (v \; \{ , v \}) [, \texttt{mutual}]
+%  \\[0.5em]
 \dimTy ::= \; & \mathbb{N}_{>0} \\
@@
 %%
-\subsubsection{Temporal specifications}
+%\subsubsection{Temporal specifications}
 
@@
 \begin{array}{rl}
-(\textsc{mutual}) \; & 
-\texttt{stencil} \; \texttt{dependency} (\texttt{v}), \texttt{mutual}
-  :: \texttt{w} 
-\\
-\equiv \; & \texttt{stencil} \; \texttt{dependency} (\texttt{w}), \texttt{mutual} ::
-  \texttt{v} \\[0.5em]
+% MUTUAL 
+%(\textsc{mutual}) \; & 
+%\texttt{stencil} \; \texttt{dependency} (\texttt{v}), \texttt{mutual}
+%  :: \texttt{w} 
+%\\
+%\equiv \; & \texttt{stencil} \; \texttt{dependency} (\texttt{w}), \texttt{mutual} ::
+%  \texttt{v} \\[0.5em]
 %(\textsc{coalesce}) \; & 
@@
 \begin{figure}[t]
-\begin{minipage}{0.88\linewidth}
-\begin{align*}
-\hspace{-0.7em}
+\begin{align*}
 \begin{array}{c}
@@
   T}(\textsc{trans}) \\[1.5em]
-\\[1.5em]
 \dfrac{}{\texttt{read-once} \, S <: S}(\textsc{rep})
@@
 {\begin{array}{rl}
-\stenFwdS{m}{d} & \stenFwdS{n}{d}
+\stenFwdS{m}{d} & <: \stenFwdS{n}{d}
 %& \;\, ds \subseteq es \wedge n \leq m
@@
 \end{align*}
-\end{minipage}
 \caption{Inequations on specifications}

### 2016-06-22 paper.tex
title change
abstract

@@
 
-\title{Time and space specifications for stencils}
-\authorinfo{}{}
+\title{Abstract Spatial Specification of Stencil Computations}
+\authorinfo{Dominic Orchard \and Mistral Contrastin 
+\and Matthew Danish \and Andrew Rice}{University of Cambridge}{}
 
@@
 \begin{abstract}
+  Verifying the correctness of numerical computations is hard because
+  potential specifications for their behaviour tend to be highly
+  decouple from the code, defined in terms of the originating
+  mathematical model. For example, a system of PDEs specifies a
+  relationship between variables, but tools for encoding such a
+  specification, and pertinent properties such as convergence and
+  stability, are rare.  Furthermore, checking the correctness of some
+  implementation of a discrete approximation against such
+  specifications is extremely challenging. We seek intermediate
+  systems of specification and verification between lower-level
+  aspects (\eg{}, out-of-bounds errors) and full correctness
+  with-respect to a numerical model. In this paper, we focus on
+  \emph{stencil computations}-- a common idiom in numerical code.  Our
+  conjecture is that the vast majority of stencil computations have a
+  simple, regular, static shape that can be captured by a simple
+  specification language. We propose a language for this, along with
+  an inference and checking procedure. We test our conjecture against
+  a corpus of numerical code, and show that we can give static
+  specifications to $X\%$ of identifiable stencil computations, with a
+  mean ratio of $Y:Z$ specification-to-program code.  This paper
+  details our stencil specification language, its static semantics,
+  inference, checking, and program synthesis procedures, our
+  implementation, and our evaluation study.
 \end{abstract}
+
+\keywords{program verification, specifications, static analysis,
+  stencil computation}
 
@@
   T}(\textsc{trans}) \\[1.5em]
-\dfrac{}{\texttt{read-once} \, S <: S}(\textsc{rep})
+\dfrac{}{\texttt{readOnce} \, S <: S}(\textsc{rep})
 \\[1.5em]
@@
 \wedge \; \stenBwdS{m}{d} & <: \stenCenS{n}{d} \\
-\wedge \; \stenFwdS{m}{d} & <: \stenCenS{n}{d} \\
-\end{array}}
-%{\footnotesize{(\textsc{cover})}}
-%\hspace{-0.1em}
+\wedge \; \stenFwdS{m}{d} & <: \stenCenS{n}{d} \\[1em]
+\end{array}} \\[2.5em]
+\hspace{-0.5em}
+\dfrac{S_1 <: T_1 \quad S_2 <: T_2} 
+      {S_1 \, \texttt{+} \, S_2 <: T_1 \, \texttt{+} \, T_2}
+(\textsc{cong}\texttt{+}) \;\;\;
+\dfrac{S_1 <: T_1 \quad S_2 <: T_2} 
+      {S_1 \, \texttt{*} \, S_2 <: T_1 \, \texttt{*} \, T_2}
+(\textsc{cong}\texttt{*})
 \end{array}
 \end{align*}
-\caption{Inequations on specifications}
+\caption{Inequations on regions}
 \label{fig:inequations}

### 2016-06-22 paper.tex
abs tweak

@@
   with-respect to a numerical model. In this paper, we focus on
-  \emph{stencil computations}-- a common idiom in numerical code.  Our
-  conjecture is that the vast majority of stencil computations have a
-  simple, regular, static shape that can be captured by a simple
-  specification language. We propose a language for this, along with
-  an inference and checking procedure. We test our conjecture against
-  a corpus of numerical code, and show that we can give static
-  specifications to $X\%$ of identifiable stencil computations, with a
-  mean ratio of $Y:Z$ specification-to-program code.  This paper
-  details our stencil specification language, its static semantics,
-  inference, checking, and program synthesis procedures, our
-  implementation, and our evaluation study.
+  \emph{stencil computations}-- a common idiom in numerical code, but
+  one which is prone to error due to fine-grained programming
+  mistakes.  Our conjecture is that the vast majority of stencil
+  computations have a simple, regular, static shape that can be
+  described by a simple specification language. We propose a language
+  for this, along with an inference and checking procedure. We test
+  our conjecture against a corpus of numerical Fortran code, and show
+  that static specifications can be given to $X\%$ of identifiable
+  stencil computations, with a mean ratio of $Y:Z$
+  specification-to-program code. This paper details our stencil
+  specification language, its static semantics, inference, checking,
+  and program synthesis procedures, our implementation, and our
+  evaluation study.
 \end{abstract}

### 2016-06-22 paper.tex
intro update

@@
    a = b
+   end do
+end do
 \end{ExmVerbatim}
@@
 \texttt{a} and \texttt{b} between iterations, where \texttt{b} becomes the
-input for the next iteration. (Note, we use the parentheses syntax
-\texttt{( )} for
-array subscripts rather than the bracket syntax \texttt{[ ]}.
+input for the next iteration. Parentheses 
+\texttt{( )} are used here for 
+array subscripts rather than bracket syntax \texttt{[ ]}.
 
@@
 %$\texttt{b}$.
-In this simple example, the dependencies between between \texttt{a}
-and \texttt{b} form a simple spatial relationship which is easily
+In this simple example, the dependency between \texttt{a}
+and \texttt{b} on line 3 forms a simple spatial relationship which is easily
 understood. This spatial relationship determines other aspects of the
@@
 In this work, we introduce a simple specification language for the
-spatial and temporal properties of stencil comptuations. The
+spatial properties of stencil comptuations. The
 specifications abstract over the fine grained detail of stencil access
@@
 \begin{SpecVerbatim}
-!=  stencil centered (depth=1, dim=1) :: a
-!=  stencil dependency a, mutual       :: b
+!=  stencil centered(depth=1, dim=1) :: a
 \end{SpecVerbatim}
 %
-The first line explains that \texttt{a} is read from with a
+This explains that \texttt{a} is read from with a
 symmetrical stencil pattern (``centered'') to a depth of one in each
-direction in its first (and only) dimension.  The second line explains
-the temporal relationship between \texttt{b} and \texttt{a}: that
-previous time step for \texttt{b} is actually provided by \texttt{a},
-and vice versa. This is explained as a mutual dependence between
-\texttt{a} and \texttt{b}. In the case of the Navier-Stokes example of
+direction in its first (and only) dimension.  
+%The second line explains
+%the temporal relationship between \texttt{b} and \texttt{a}: that
+%previous time step for \texttt{b} is actually provided by \texttt{a},
+%and vice versa. This is explained as a mutual dependence between
+%\texttt{a} and \texttt{b}.
+In the case of the Navier-Stokes example of
 Figure~\ref{ref:navier-stokes-fragment}, its inferred specification is shown
@@
 the first dimension and backward to a depth of $1$ in the second
-dimension. These can drawn pictorially as follows where the horizontal
-dimensions is \text{dim=1} and the vertical is \texttt{dim=2}:
-
-\begin{center}
-\begin{tikzpicture}
-\node at (1.25,1.22) {i, j};
-\draw (1,1) rectangle (1.5,1.5);
-\draw (1.5,1) rectangle (2,1.5);
-\draw (1,0.5) rectangle (1.5,1);
-\draw (0.5,1) rectangle (1,1.5);
-\draw (1,1.5) rectangle (1.5,2);
-%
-\node at (4.25,1.22) {i, j};
-\draw (4,1) rectangle (4.5,1.5);
-\draw (4.5,1) rectangle (5,1.5);
-\draw (4.5,1.5) rectangle (5,2);
-\draw (4,1.5) rectangle (4.5,2);
-\end{tikzpicture}
-\end{center}
+dimension. Figure~\ref{ref:navier-stokes-fragment}(c) describes these
+pictorially.
 
@@
 
-\begin{figure}
+\begin{figure}[t]
 \begin{ExmVerbatim}[firstnumber=20]
@@
 \end{ExmVerbatim}
-(a). Excerpt from the Fortran code,
+(a). Excerpt from Fortran code for a Navier-Stoke fluid simulator,
 showing highly-detailed stencil computations. \\
@@
 (b). Inferred stencil specification from CamFort
+
+\begin{center}
+\begin{tikzpicture}
+\node at (1.25,1.22) {i, j};
+\draw (1,1) rectangle (1.5,1.5);
+\draw (1.5,1) rectangle (2,1.5);
+\draw (1,0.5) rectangle (1.5,1);
+\draw (0.5,1) rectangle (1,1.5);
+\draw (1,1.5) rectangle (1.5,2);
+%
+\node at (4.25,1.22) {i, j};
+\draw (4,1) rectangle (4.5,1.5);
+\draw (4.5,1) rectangle (5,1.5);
+\draw (4.5,1.5) rectangle (5,2);
+\draw (4,1.5) rectangle (4.5,2);
+\end{tikzpicture}
+\end{center}
+(c). Pictoral representation of the two stencil specifications.
+The horizontal dimensions is \text{dim=1} and the vertical is \texttt{dim=2}:
 \caption{Fragment of Navier-Stokes fluid simulator and its specification}

### 2016-06-23 paper.tex
explanation on the model

@@
 \newcommand{\vect}[1]{\textbf{#1}}
+\newcommand{\vtwoh}[2]{\setlength{\arraycolsep}{0em}
+\left[\begin{array}{cc}#1 \; & \; #2\end{array}\right]}
+\newcommand{\vthreeh}[3]{\setlength{\arraycolsep}{0em}
+\left[\begin{array}{ccc}#1 \; & \; #2 \; & \; #3\end{array}\right]}
 \newcommand{\vtwo}[2]{\setlength{\arraycolsep}{0em}
@@
 
-We give a model of our specifications which explains
-the range of indices which correspond to the specifications.
-The model is based on multi-sets since array indices my be repeated
-in an expression.
-
-Figure~\ref{fig:model} describes the model in terms of
-an interpretation function $\interp{-}$ which maps specifications
-to multisets of relative indices.
-
-\begin{definition}Relative indices are drawn
-from $(\mathbb{Z}_{\bot})^{\mathbb{D}}$, essentially an infinite
-tuple of offsets. For example, $(i, j+1)$ corresponds to
-the relative index $(0, 1, \bot, \bot, \ldots)$
+We provide a semantic model of the stencil specification language as
+the basis for building correct implementations, showing the soundness
+of the equational theory and approximations, and as a basis for the
+inference, checking, and program synthesis algorithms.
+The semantics of stencil specifications can be given a denotational
+model as multisets of vectors, represent array indexing.
+
+\begin{definition}
+  A specification of maximum dimensionality $n$, is modelled by a
+  multi-set of column vectors of size $n$  ($n$-dimensional vectors /
+  $n$-vectors for short)
+  with values drawn from $\mathbb{Z} \cup \{\infty\}$, representing
+  relative indices. The $\infty$ element represents non-relative
+  indices (\eg{}, constants), which are seen as infinite with respect
+  to the frame of reference of a relative offset from an induction
+  variable.
+
+  For example, the array index $(i, j+1)$ corresponds to the vector
+  $\vtwoh{0}{1}$, whereas, the $(0, j)$ corresponds to the
+  vector $\vtwoh{\infty}{0}$ since $0$ is a non-relative index. 
 \end{definition}
 
-The main interpretation function is overloaded on lists of
-variable-spec pairs $\interp{\overline{v : S}}$,
-returning multisets of variable-relative-index pairs. This provides
-the top-level definition of the model, with $\interp{-}$ overloaded
-on $S$, $\interp{S}$ mqpping to multisets of
-relative indices not associated to an array avariable.
+Figure~\ref{fig:model} outlines the core rules for regions
+via the interpretation function $\interp{-}$.  We explain the
+intermediate notations and definitions:
+
+\begin{definition}A \emph{single-entry vector} of size $n$, denoted 
+$\singleEntry{r}{n}$, is a vector 
+where the $r^{th}$ entry is $1$ and all others are $0$, \eg{},
+$\singleEntry{1}{2} = \vtwoh{1}{0}$.
+\end{definition}
+
+\begin{definition}The product of two models $\otimes$ is defined:
+%%
+\begin{align*}
+M \otimes N & =
+  \{(u \bowtie v)_i \mid i \in \{0, \ldots, 2^n \}, (u, v) \in
+              M \times N\}^1
+\end{align*}
+The $\bowtie$ operation, which we call \emph{pairwise permutation}
+(described below), constructs a $2^n \times n$ matrix ($2^n$-vector of
+$n$-vectors).  Using this, the $\otimes$ operation constructs the set
+of every pairwise permutation for every pair $(u, v)$ of $n$-vectors in the
+Cartesian product of $M \times N$. Note the $^1$ superscript here 
+on the set-builder notation which denotes that this construction
+constructs a degenerate multiset (where all the multiplicities are 1).
+This is important for the function of $\otimes$ in the model (see below).
+
+The pairwise permutation $u \bowtie v$ takes two vectors and builds the matrix
+where the rows are all possible $n$-vectors generated by 
+non-determinstically picking each $i^{th}$ entry from either the
+$i^{th}$ entry in $u$ or the $i^{th}$ entry in $j$. For example:  
+%
+\begin{equation*}
+\vthreeh{0}{1}{2} \bowtie \vthreeh{3}{4}{5} = 
+\setlength{\arraycolsep}{0.5em}
+\left[\begin{array}{ccc}
+0 & 1 & 2 \\
+0 & 1 & 5 \\
+0 & 4 & 2 \\
+0 & 4 & 5 \\
+3 & 1 & 2 \\
+3 & 1 & 5 \\
+3 & 4 & 2 \\
+3 & 4 & 5 
+\end{array}
+\right]
+\end{equation*}
+%
+The $2^n$ unique choices for $\bowtie$ on $n$-vectors 
+corresponds to taking all bit-strings of length $n$ and
+selecting from $u$ for $1$ and $v$ for $0$. Thus, we can
+defined the pairwise pemutation matrix as:
+%
+\begin{align*}
+(u \bowtie v)_{i,j} & = b_{i,j} u_i + \neg b_{i,j} v_i
+\end{align*}
+%
+where $b$ is a logical matrix where $b_{i,j}$ is the $j$-th bit of the integer $i$. 
+Recall that the vectors are defined over the domain
+$\mathbb{Z} \cup \{\infty\}$. 
+For addition and multiplication, we take 
+ $\infty$ to be an absorbing element. Thus, if either
+vector in a pairwise permutation contains $\infty$ in
+the $j^{th}$ entry, then $j^{th}$ entry of every pairwise
+permutation is $\infty$, \eg{}, 
+%%
+\begin{equation*}
+\vtwoh{0}{1} \bowtie \vtwoh{2}{\infty} = 
+\setlength{\arraycolsep}{0.5em}
+\left[\begin{array}{cc}
+0 & \infty \\
+0 & \infty \\ 
+2 & \infty \\
+2 & \infty
+\end{array}
+\right]
+\end{equation*}
+
+%%
+\end{definition}
+
+%The main interpretation function is overloaded on lists of
+%variable-spec pairs $\interp{\overline{v : S}}$,
+%returning multisets of variable-relative-index pairs. This provides
+%the top-level definition of the model, with $\interp{-}$ overloaded
+%on $S$, $\interp{S}$ mqpping to multisets of
+%relative indices not associated to an array avariable.
 
@@
 
+\paragraph{Step 4: Convert spans into region AST}
+
+Given the set $\textsf{regions}(M)$ of spans covering the indexing
+space, the next step converts these into abstract syntax tree
+for the specification language.
+
 \section{Evaluation}

### 2016-06-23 paper.tex
First use of abbreviation is expanded

@@
   decouple from the code, defined in terms of the originating
-  mathematical model. For example, a system of PDEs specifies a
-  relationship between variables, but tools for encoding such a
-  specification, and pertinent properties such as convergence and
+  mathematical model. For example, a system of partial differential equations
+  (PDEs) specifies a relationship between variables, but tools for encoding such
+  a specification, and pertinent properties such as convergence and
   stability, are rare.  Furthermore, checking the correctness of some

### 2016-06-23 paper.tex
Clarify bracket syntax and note a possible error in 1D Laplace transform

@@
 following iteratively computes the one-dimensional discrete Laplace
-transform (an approximation to a derivative): 
+transform (an approximation to a derivative) in Fortran:
+
+\mnote{Line 4 does not swap  between iterations it does within. Is this 
+correct? I think it needs to be on Line 5? If it is meant to be in Line 4, it is
+wrongly indented.}
 %
@@
 \texttt{a} and \texttt{b} between iterations, where \texttt{b} becomes the
-input for the next iteration. Parentheses 
-\texttt{( )} are used here for 
-array subscripts rather than bracket syntax \texttt{[ ]}.
+input for the next iteration. In Fortran, parentheses \texttt{( )} are used
+for array subscripts rather than the more familiar bracket syntax \texttt{[ ]}.
 

### 2016-06-23 paper.tex
typo

@@
 \begin{definition}[Base induction variable]
-  An variable \textit{v} of integer type is a \emph{base induction
+  A variable \textit{v} of integer type is a \emph{base induction
     variable} if it is the control variable of a ``for''

### 2016-06-24 paper.tex
fixed first example
Remove unnecessary note

@@
 transform (an approximation to a derivative) in Fortran: 
-
-\mnote{Line 4 does not swap  between iterations it does within. Is this 
-correct? I think it needs to be on Line 5? If it is meant to be in Line 4, it is
-wrongly indented.}
 %
@@
       b(i) = a(i-1) - 2*a(i) + a(i+1)
+   end do
    a = b
-   end do
 end do

### 2016-06-24 paper.tex
fixed spurious example and its discussion which was wrong

@@
 transform (an approximation to a derivative) in Fortran:
+
+\mnote{Line 4 does not swap  between iterations it does within. Is this 
+correct? I think it needs to be on Line 5? If it is meant to be in Line 4, it is
+wrongly indented.}
 %
@@
 != stencil backward(depth=2, dim=1) :: a
-b(i+1) = a(i) + a(i-1) + a(i-2)
-\end{ExmVerbatim}
-%
-Note that the left-hand subscript on \term{b} does not affect
-the stencil specification. The requirement on the left-hand side
-is merely that it is an array subscript with at least one neighbourhood index,
-but this may be on different induction variables to the right-hand
-side, which is the stencil we are concerened with.
-
+b(i) = a(i) + a(i-1) + a(i-2)
+\end{ExmVerbatim}
+%
 A \texttt{centered} stencil is a combination of \texttt{forward}
@@
 \begin{align*}
-(u \bowtie v)_{i,j} & = b_{i,j} u_i + \neg b_{i,j} v_i
+(u \bowtie v)_{i,j} & = b_{i,j} u_j + \neg b_{i,j} v_j
 \end{align*}
@@
 
-\section{Correctness}
-
-\begin{theorem}[Soundness]
-\[
-\overline{\texttt{v} : S}\equiv \overline{\texttt{u} : T}
-\; \Rightarrow \;
-\interp{\overline{\texttt{v} : S}} = \interp{\overline{\texttt{u} : T}}
-\]
-\end{theorem}
-
-\paragraph{Proof}
-\begin{itemize}
-\item \textsc{coalesceF}
-\begin{align*}
-\llbracket & \texttt{v} : \texttt{fwd} \; \texttt{dims}=ds \; \texttt{depth}=n; \,
-  \texttt{v} :  \texttt{fwd} \; \texttt{dims}=ds \;
-  \texttt{depth}=n+1)\rrbracket \\
-& =  \{ (\texttt{v}, ix)^m \mid \forall \, m, ix : \relix, d : \mathbb{N}
-  \, . \, d \in \textit{ds} \Rightarrow ix \, d \leq (n+1) \} \; \cup \; \{ ix^m \mid \forall \, m, ix : \mathbb{Z}^{\mathbb{N}}, d : \mathbb{N}
-  \, . \, d \in \textit{ds} \Rightarrow ix \, d \leq n \}\\
-& =  \{ (\texttt{v}, ix)^m \mid \forall \, m, ix : \relix, d : \mathbb{N}
-  \, . \, d \in \textit{ds} \Rightarrow ((ix \, d \leq (n+1)) \vee (ix \,
-  d \leq n)) \} \\
-& =  \{ (\texttt{v}, ix)^m \mid \forall \, m, ix : \relix, d : \mathbb{N}
-  \, . \, d \in \textit{ds} \Rightarrow ix \, d \leq (n+1) \} \\
-& = \interp{\texttt{v} : \texttt{fwd} \; \texttt{dim}=ds \;
-  \texttt{depth}=n + 1} \qquad \Box
-\end{align*}
-\end{itemize}
+\input{proof.tex}
 

### 2016-06-24 paper.tex
Remove occurance of multi-set

--- 2016-06-24 paper.tex
@@
 oultines the syntax. Section~\ref{sec:semantics} defines its semantics
-via a simple multi-set interpretation over indices. Section~\ref{sec:eqs} provides an
+via a simple set interpretation over indices. Section~\ref{sec:eqs} provides an
 equational theory for specifications via relation ($\equiv$) and a
 theory of approximation via relation ($<:$). These equations are then
-proven sound with respecto the multi-set semantics of Section~\ref{sec:semantics}.
+proven sound with respect to the set semantics of Section~\ref{sec:semantics}.
 
@@
   $(-b) + i \equiv i - b$).  We classify neighbourhood indices using
-  the predicate \neigh{}. An array subscript $v(\var{e})$ where every
+  the predicate \neigh{}. An array subscript $v(\bar{e})$ where every
  $e \in \bar{e}$ is a neighbourhood index is called a 
@@
 The semantics of stencil specifications can be given a denotational
-model as multisets of vectors, represent array indexing.
+model as sets of vectors, represent array indexing.
 
@@
   A specification of maximum dimensionality $n$, is modelled by a
-  multi-set of column vectors of size $n$  ($n$-dimensional vectors /
+  set of column vectors of size $n$  ($n$-dimensional vectors /
   $n$-vectors for short)
@@
 of every pairwise permutation for every pair $(u, v)$ of $n$-vectors in the
-Cartesian product of $M \times N$. Note the $^1$ superscript here 
-on the set-builder notation which denotes that this construction
-constructs a degenerate multiset (where all the multiplicities are 1).
-This is important for the function of $\otimes$ in the model (see below).
+Cartesian product of $M \times N$.
 
@@
 \begin{align*}
-(u \bowtie v)_{i,j} & = b_{i,j} u_i + \neg b_{i,j} v_i
+(u \bowtie v)_{i,j} & = b_{i,j} u_i + \neg b_{i,j} v_j
 \end{align*}
@@
 %  the multi-set $A$}
-\caption{Multi-set model of specifications}
+\caption{Set model of specifications}
 \label{fig:model}

### 2016-06-24 paper.tex

@@
 \begin{align*}
-(u \bowtie v)_{i,j} & = b_{i,j} u_i + \neg b_{i,j} v_j
+(u \bowtie v)_{i,j} & = b_{i,j} u_j + \neg b_{i,j} v_j
 \end{align*}

### 2016-06-24 paper.tex
Remove multi-set related things, add some of the proofs
clean whitespace
explanation of mapping regions to syntax

@@
   \{(u \bowtie v)_i \mid i \in \{0, \ldots, 2^n \}, (u, v) \in
-              M \times N\}^1
+              M \times N\}
 \end{align*}
@@
 Given the set $\textsf{regions}(M)$ of spans covering the indexing
-space, the next step converts these into abstract syntax tree
+space, the next step converts these into abstract syntax trees 
 for the specification language.
+
+For each span $S = [\vect{l} \; \vect{u}]$, 
+for each dimension, each pair of entries in the lower- and upper-bound 
+vectors is mapped to a concrete \textit{spec}, as defined by the
+$\mathsf{conv}$ function: 
+%%
+\begin{align*}
+& \mathsf{conv}_i\; [\vect{l} \; \vect{u}] = \\
+& \begin{cases}
+%% FWD
+ \stenFwd{\vect{u}_i}{i}  & \vect{l}_i = 0 \; \wedge \; \vect{u}_i > 0 \\[0.7em]
+ \texttt{irreflexive(dims=$i$)}, & \\
+ \qquad \stenFwd{\vect{u}_i}{i}  & \vect{l}_i = 1 \; \wedge \; \vect{u}_i > 0 \\[0.7em]
+%% BWD
+ \stenBwd{|\vect{l}_i|}{i}  & \vect{l}_i < 0 \; \wedge \; \vect{u}_i = 0 \\[0.7em]
+ \texttt{irreflexive(dims=$i$)}, & \\
+ \quad\;\; \stenBwd{|\vect{l}_i|}{i}  & \vect{l}_i < 0 \; \wedge \; \vect{u}_i = -1 \\[0.7em]
+%% CEN
+ \stenCen{\vect{u}_i}{i}  &\vect{l}_i < 0 \; \wedge \; \vect{u}_i > 0
+ \; \\
+ & \qquad\quad \wedge \; |\vect{l}_i| = \vect{u}_i \\[0.7em]
+ \stenBwd{|\vect{l}_i|}{i} & \vect{l}_i < 0 \; \wedge \; \vect{u}_i >
+ 0 \\
+\quad \texttt{+} \; \stenFwd{\vect{u}_i}{i}  & \qquad\quad \wedge \; |\vect{l}_i| \neq \vect{u}_i \\[0.7em]
+% ABSOLUTE REP
+ \texttt{irreflexive(dims=$i$)} & \vect{l}_i = \infty \; \vee \; \vect{u}_i
+ = \infty \\[0.7em]
+% BOUND
+ \texttt{atMost}, \stenFwd{\vect{u}_i}{i} & \vect{l}_i > 1 \\[0.7em]
+ \texttt{atMost}, \stenBwd{|\vect{l}_i|}{i} \quad & \vect{l}_i < (-1) \\[0.7em]
+  \end{cases}
+\end{align*}
+%%
+The first two cases map to \texttt{forward} stencils since the lower
+bound is $0$ or $1$, with the depth determined by the upper bound of
+the span. If the lower bound is $1$, then the 
+$0$-offset point is missing so the specification is given an
+\texttt{irreflexive} modifier. The second two cases are similar for
+\texttt{backward} but now the depth of the region is determined by the
+absolute of the lower bound. In the fifth-case, when the lower bound
+is negative and the upper bound is positive and they have the same
+magnitude, then a \texttt{centered} region is constructed with
+the depth as the common magnitude. If the magnitudes are not the same
+(sixth case) then separated \texttt{forward} and \texttt{backward}
+regions are summed, with their corresponding depths.
+
+In the case where either bound is $\infty$, representing a non
+neighbourhood index, then a simple \texttt{irreflexive} specification
+is given with no regions, indicating that there is not even reflexive
+0-offset access in this dimension.
+
+Finally in the last two cases, if the lower bound is greater than 1 
+(which implies the upper bound is greater than 1 too) then 
+this represents a region that is not at the origin (or next to) the
+origin and thus not representable in the specificaiton language. 
+Thus, an upper bound is instead returned showing the maximum extent
+of the region. The last case is the dual, when the lower bound is less
+than $-1$, giving an upper bound on a \texttt{backward} specification.
+
+\begin{lemma}
+$\mathsf{conv}_i$ is total given the pre-condition that the lower
+bound of the input span is less than or equal to its upper bound. 
+\end{lemma}
 

### 2016-06-30 paper.tex
removed INTER axiom
Proof of * commutativity and matrix notation fix
split modifieres in the grammar
butchered the abstract

@@
 \begin{abstract}
-  Verifying the correctness of numerical computations is hard because
-  potential specifications for their behaviour tend to be highly
-  decouple from the code, defined in terms of the originating
-  mathematical model. For example, a system of partial differential equations
-  (PDEs) specifies a relationship between variables, but tools for encoding such
-  a specification, and pertinent properties such as convergence and
-  stability, are rare.  Furthermore, checking the correctness of some
-  implementation of a discrete approximation against such
-  specifications is extremely challenging. We seek intermediate
+Verifying the correctness of numerical computations relies on a specification of their behaviour.
+A natural way to do this would be in terms of the original mathematical model but this approach yields specifications which are highly decoupled from the code.
+Checking the correctness of code which implements a discrete approximation of such specifications is extremely challenging. 
+Instead we have been investigating intermediate
   systems of specification and verification between lower-level
   aspects (\eg{}, out-of-bounds errors) and full correctness
-  with-respect to a numerical model. In this paper, we focus on
-  \emph{stencil computations}-- a common idiom in numerical code, but
-  one which is prone to error due to fine-grained programming
-  mistakes.  Our conjecture is that the vast majority of stencil
-  computations have a simple, regular, static shape that can be
-  described by a simple specification language. We propose a language
-  for this, along with an inference and checking procedure. We test
-  our conjecture against a corpus of numerical Fortran code, and show
-  that static specifications can be given to $X\%$ of identifiable
+  based on a numerical model. In this paper, we focus on
+  \emph{stencil computations}---a common idiom in numerical code, but
+  one which is error-prone due to fine-grained indexing errors.  
+%
+We describe a stencil specification language which can be directly embedded as annotations in source code, along with an inference and checking procedure. 
+%
+We evaluate our language against a corpus of numerical Fortran code, and show
+  that precise specifications can be given to $X\%$ of identifiable
   stencil computations, with a mean ratio of $Y:Z$
-  specification-to-program code. This paper details our stencil
+  specification-to-program code and that the vast majority of stencil computations have a simple, regular, static shape.
+%
+This paper details our stencil
   specification language, its static semantics, inference, checking,
@@
   evaluation study.
+%
+We provide an implementation of stencil inference and validation in \emph{Camfort}, an open-source program analysis and refactoring tool.
 \end{abstract}
@@
 \nonterm{specification} ::= & \nonterm{regionDec} \mid \nonterm{specDec} \\
-\nonterm{specDec} ::= & \term{stencil} \; \nonterm{spec} \;
+\nonterm{specDec} ::= & \term{stencil} \; \nonterm{specTop} \;
                         \texttt{::} \; v
@@
                          \nonterm{region}\\[1em]
-%\nonterm{spec} ::= & \nonterm{spatial} \mid \nonterm{temporal}   \\[1em]
-\nonterm{spec} ::= & [\nonterm{mod}] \;
-                       [\nonterm{approxMod},] \; [\nonterm{region}] \\
+%\nonterm{spec} ::= & \nonterm{spatial} \mid \nonterm{temporal}
+%\\[1em]
+\nonterm{specTop} ::= & [\nonterm{approxMod},] \; \nonterm{spec} \\
+\nonterm{spec} ::= & [\nonterm{mod}] \; [\nonterm{region}] \\
 \nonterm{mod} ::= & [\nonterm{modR},] \; [\nonterm{modRefl},] \; [\nonterm{modIrefl},] \\
@@
                        \; \texttt{*} \; T) \\[0.5em]
-(\textsc{inter}) \;\; & (S \; \texttt{+} \; S') \; \texttt{*} \;
-                              (T \; \texttt{+} \; T') \\
- \equiv \; & (S \;  \texttt{*} \; S)
- \; \texttt{+} \; (S \; \texttt{*} \; T')
- \; \texttt{+} \; (S' \; \texttt{*} \; T)
- \; \texttt{+} \; (S' \; \texttt{*} T')  \\[0.5em]
 \end{array} \\ \\
@@
 M \otimes N & =
-  \{(u \bowtie v)_i \mid i \in \{0, \ldots, 2^n \}, (u, v) \in
-              M \times N\}
+  \{(u \bowtie v)_i \mid i \in \{1, \ldots, 2^n\}, (u, v) \in M \times N\}
 \end{align*}
@@
 (described below), constructs a $2^n \times n$ matrix ($2^n$-vector of
-$n$-vectors).  Using this, the $\otimes$ operation constructs the set
-of every pairwise permutation for every pair $(u, v)$ of $n$-vectors in the
-Cartesian product of $M \times N$.
+$n$-vectors). Using this, the $\otimes$ operation constructs the set of every
+pairwise permutation for every pair $(u, v)$ of $n$-vectors in the Cartesian
+product of $M \times N$.
 
@@
 %
-where $b$ is a logical matrix where $b_{i,j}$ is the $j$-th bit of the integer $i$.
-Recall that the vectors are defined over the domain
+where $b$ is a logical matrix where $b_{i,j}$ is the $j$-th bit of the integer
+$i - 1$. Recall that the vectors are defined over the domain
 $\mathbb{Z} \cup \{\infty\}$.

### 2016-07-01 paper.tex
Start adding spatial model

--- 2016-06-24 paper.tex
@@
                          \nonterm{region}\\[1em]
-%\nonterm{spec} ::= & \nonterm{spatial} \mid \nonterm{temporal}   \\[1em]
-\nonterm{spec} ::= & [\nonterm{mod}] \;
-                       [\nonterm{approxMod},] \; [\nonterm{region}] \\
-\nonterm{mod} ::= & [\nonterm{modR},] \; [\nonterm{modRefl},] \; [\nonterm{modIrefl},] \\
-\nonterm{modR} ::= &
- \term{readOnce} \\
+%\nonterm{spec} ::= & \nonterm{spatial} \mid \nonterm{temporal}
+%\\[1em]
+\nonterm{specTop} ::= & [\nonterm{approxMod},] \; \nonterm{spec} \\
+\nonterm{spec} ::= & [\nonterm{modR}] \; \nonterm{qualifiedRegion} \\
+\nonterm{qualifiedRegion} ::= & [\nonterm{modRefl},] \; [\nonterm{modIrrefl},] \; \nonterm{region}\\
+\nonterm{modR} ::= & \term{readOnce} \\
 \nonterm{modRefl} ::= & \term{reflexive} \, (\term{dims=}\dimTy \{\term{,} \; \dimTy\}) \\
@@
 
-Figure~\ref{fig:model} outlines the core rules for regions
+Figure~\ref{fig:region-model} outlines the core rules for regions
 via the interpretation function $\interp{-}$.  We explain the
@@
 \begin{figure}
-\input{model.tex}
+\input{region-model.tex}
 %\textit{$a^m \in A$ means there are $n$ copies of $a$ in
 %  the multi-set $A$}
-\caption{Set model of specifications}
-\label{fig:model}
+\caption{Set model of regions}
+\label{fig:region-model}
+\end{figure}
+
+\begin{figure}
+\input{spatial-model.tex}
+\caption{Model of specifications}
+\label{fig:spatial-model}
 \end{figure}

### 2016-07-03 paper.tex
Mathematical model nearly finished

--- 2016-06-30 paper.tex
@@
 
+\newcommand{\stenRefl}[1]{\term{reflexive} \, (\term{dim=}#1)}
 \newcommand{\stenFwd}[2]{\term{forward} \, (\term{depth=}#1,
@@
   \term{dim=}#2)}
+\newcommand{\stenReflS}[1]{\term{refl} \, (\term{dim=}#1)}
 \newcommand{\stenFwdS}[2]{\term{fwd} \, (\term{depth=}#1,
@@
 \newcommand{\singleEntry}[2]{\textbf{J}_{#2}^{#1}}
+\newcommand{\zeroEntry}[2]{\textbf{K}_{#2}^{#1}}
 
@@
 \nonterm{specTop} ::= & [\nonterm{approxMod},] \; \nonterm{spec} \\
-\nonterm{spec} ::= & [\nonterm{mod}] \; [\nonterm{region}] \\
-\nonterm{mod} ::= & [\nonterm{modR},] \; [\nonterm{modRefl},] \; [\nonterm{modIrefl},] \\
-\nonterm{modR} ::= &
- \term{readOnce} \\
-\nonterm{modRefl} ::= & \term{reflexive} \, (\term{dims=}\dimTy \{\term{,} \; \dimTy\}) \\
-\nonterm{modIrrefl} ::= & \term{irreflexive} \, (\term{dims=}\dimTy \{\term{,}
+\nonterm{spec} ::= & [\nonterm{modMult},] \; \nonterm{refRegion} \\
+\nonterm{refRegion} ::= & [\nonterm{modRefl},] \; \nonterm{region}\\
+\nonterm{modMult} ::= & \term{readOnce} \\
+\nonterm{modRefl} ::= & \term{irreflexive} \, (\term{dims=}\dimTy \{\term{,}
           \dimTy\}) \\
 \nonterm{approxMod} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
-\nonterm{region} ::= & \stenFwd{\mathbb{N}_{>0}}{\dimTy}  \\
+\nonterm{region} ::=
+        & \stenFwd{\mathbb{N}_{>0}}{\dimTy}  \\
 \mid \; & \stenBwd{\mathbb{N}_{>0}}{\dimTy} \\
 \mid \; & \stenCen{\mathbb{N}_{>0}}{\dimTy} \\
+\mid \; & \stenRefl{\dimTy}  \\
 \mid \; & \nonterm{region} \; \term{+} \; \nonterm{region} \\
@@
 To save space we use abbreviations
-\term{fwd}, \term{bwd}, and \term{cen} for \term{forward},
-\term{backward}, and \term{centered} regions respectively.  The
-equations reveal something of their semantics.  The first six
+\term{refl}, \term{fwd}, \term{bwd}, and \term{cen} for \term{reflexive},
+\term{forward}, \term{backward}, and \term{centered} regions respectively.
+The equations reveal something of their semantics.  The first six
 equations (with labels of the form $\ldots \texttt{+} \ldots$)
@@
 $\singleEntry{r}{n}$, is a vector
-where the $r^{th}$ entry is $1$ and all others are $0$, \eg{},
-$\singleEntry{1}{2} = \vtwoh{1}{0}$.
+where the $r^{th}$ entry is $1$ and all others are $\infty$, \eg{},
+$\singleEntry{1}{2} = \vtwoh{1}{\infty}$.
 \end{definition}
 
+\begin{definition}A \emph{zero-entry vector} of size $n$, denoted
+$\zeroEntry{r}{n}$, is a vector where the $r^{th}$ entry is $0$ and all others
+are $\infty$, \eg{}, $\zeroEntry{1}{2} = \vtwoh{0}{\infty}$.
+\end{definition}
+
 \begin{definition}The product of two models $\otimes$ is defined:
 %%
-\begin{align*}
-M \otimes N & =
-  \{(u \bowtie v)_i \mid i \in \{1, \ldots, 2^n\}, (u, v) \in M \times N\}
-\end{align*}
+\begin{alignat*}{2}
+& dims(M_m,N_n) =
+  \bigg\{ x \; \bigg| \;
+    \parbox{4.1cm}{
+      $x \in \{1, \ldots, m\} \wedge M_x \neq \infty \; \vee \;$ \\
+      $x \in \{1, \ldots, n\} \wedge N_x \neq \infty$
+                  } \bigg\} \\
+& M_n \otimes N_n =
+  \Bigg\{x \; \Bigg| \;
+    \parbox{4.8cm}{$i \in \{1, \ldots, 2^n\},$
+                   $(u, v) \in M_n \times N_n,$ \\
+                   $x = (u \bowtie v)_i,$ \\
+                   $\forall j \in dims(M_n,N_n) (x_j \neq \infty) $
+                  } \Bigg\}
+\end{alignat*}
 The $\bowtie$ operation, which we call \emph{pairwise permutation}
@@
 where $b$ is a logical matrix where $b_{i,j}$ is the $j$-th bit of the integer
-$i - 1$. Recall that the vectors are defined over the domain
-$\mathbb{Z} \cup \{\infty\}$.
-For addition and multiplication, we take
- $\infty$ to be an absorbing element. Thus, if either
-vector in a pairwise permutation contains $\infty$ in
-the $j^{th}$ entry, then $j^{th}$ entry of every pairwise
-permutation is $\infty$, \eg{},
-%%
-\begin{equation*}
-\vtwoh{0}{1} \bowtie \vtwoh{2}{\infty} =
-\setlength{\arraycolsep}{0.5em}
-\left[\begin{array}{cc}
-0 & \infty \\
-0 & \infty \\
-2 & \infty \\
-2 & \infty
-\end{array}
-\right]
-\end{equation*}
+$i - 1$. Recall that the vectors are defined over the domain $\mathbb{Z} \cup
+\{\infty\}$. In pairwise permutation $\infty$
 
@@
 \begin{figure}
-\input{model.tex}
+\input{region-model.tex}
 %\textit{$a^m \in A$ means there are $n$ copies of $a$ in
 %  the multi-set $A$}
-\caption{Set model of specifications}
-\label{fig:model}
+\caption{Set model of regions}
+\label{fig:region-model}
+\end{figure}
+
+\begin{figure}
+\input{spatial-model.tex}
+\caption{Model of specifications}
+\label{fig:spatial-model}
+\end{figure}
+
+\begin{figure}
+\input{consistency.tex}
+\caption{Consistency function for a model and a set of indices}
+\label{fig:spatial-model}
 \end{figure}

### 2016-07-03 paper.tex
Sexier name
Separate coalesced region to make understanding easier
Fix math typos in inference & better partial order on matrices
added some words about the corpus and scripts to pull in the results and stats
remove infer-output-july3.txt and replace it with the next results file...

@@
 
-\title{Abstract Spatial Specification of Stencil Computations}
+\include{results}
+\include{corpussize}
+
+\title{Verification of Stencil Computations through Spatial Specifications}
 \authorinfo{Dominic Orchard \and Mistral Contrastin
@@
 \draw[thick] (0.5,1) rectangle (2,1.5);
-\draw[thick] (1,0.5) rectangle (1.5,1);
-\draw[thick] (1,1.5) rectangle (1.5,2);
+\draw[thick] (1,0.40) rectangle (1.5,0.90);
+\draw[thick] (1,1.60) rectangle (1.5,2.10);
 \end{tikzpicture}
@@
 %\node at (1.25,1.22) {\textnormal{\small{0,0}}};
-\draw[thick] (1.5,1) rectangle (2,1.5);
-\draw[thick] (0.5,1) rectangle (1,1.5);
+\draw[thick] (1.6,1) rectangle (2.1,1.5);
+\draw[thick] (0.4,1) rectangle (0.9,1.5);
 \draw[thick] (1,0.5) rectangle (1.5,2);
@@
 \label{sec:evaluation}
+We now present an evaluation of the stencil language. We first confirm
+that stencils are common in scientific code and further that our
+stencil language can express a great many of these.
+
+\subsection{Software Corpus}
+\begin{figure}
+\begin{tabular}{|l|l|l|}
+\hline
+Package       & Files & Lines of code \\
+\hline
+UM            & \umFiles  & \umLoC       \\
+E3MG          & \ethreemgeaFiles   & \ethreemgeaLoC        \\
+BLAS          & \blasFiles   & \blasLoC        \\
+Hybrid4       & \hybridfourFiles    & \hybridfourLoC          \\
+GEOS-Chem     & \geoschemFiles   & \geoschemLoC       \\
+Navier        & \navierFiles     & \navierLoC           \\
+CP            & \computationalphysicstwoFiles    & \computationalphysicstwoLoC           \\
+\hline
+\textbf{Total}& \overallFiles  & \overallLoC     \\
+\hline 
+\end{tabular}
+\caption{Summary of software packages used for evaluation\label{fig:corpus}}
+\end{figure}
+
+Figure \ref{fig:corpus} shows summary statistics of the software packages used in our evaluation. In total we analysed 1.1 million lines of code from 7 packages.
+
+\textbf{The Unified Model}\footnote{\url{http://www.metoffice.gov.uk/research/modelling-systems/unified-model}} is a weather forecasting and climate modelling tool developed by the Met Office in the United Kingdom.
+It is used by research organisations and meterological services around the world.
+We use the development branch (trunk) of the model.
+The Met Office runs a comprehensive code quality system incorporating dedicated commiters (we counted 11) for particular parts of the model.
+We counted 120 additional contributors whose submissions are reviewed and tested before being accepted into the code base.
+
+% e3mg-ea
+\textbf{E3MG} (An Energy-Environment-Economy (E3) Model at the Global Level) is a macroeconomic model used for asssessment of environmental policy~\cite{RePEc:aen:journl:2006se-a12}.
+This model was developed by Cambridge Econometrics, an independent consultancy company.
+
+\textbf{BLAS} (Basic Linear Algebra Subprograms)\footnote{http://www.netlib.org/blas/} provides efficient and portable routines for vector and matrix operations.
+These routines feature in many other libraries (including LAPACK).
+We used version 3.6.0.
+
+\textbf{Hybrid4} is a vegetation and biomass model for simulating carbon, water and nitrogen flows~\cite{GBC:GBC635}.
+
+\textbf{GEOS-Chem}\footnote{\url{http://acmg.seas.harvard.edu/geos/}} is a three-dimensional model of tropospheric chemistry developed at Harvard and used by approximately 70 universities and research institutions world-wide.
+We examined version 10-01.
+
+\textbf{Navier} is an implementation of a Navier-Stokes fluid dynamics computation~\cite{griebel1997numerical}.
+
+\textbf{CP} consists of the example code from the book ``Computational Physics''~\cite{nicholas2006computational} introducing numerical techniques and their application to modern physics problems such as fields, waves, statistical mechanics and quantum mechanics.
+
+% atLeast: number of stencils with an atLeast modifier
+% atMost: number of stencils with an atMost modifier
+% boundedBoth: number of stencils with both atLeast and atMost modifier
+%% atLeast == atMost == boundedBoth - whenever we infer one we infer the other but users might write just one
+
+% backward: number of backward stencils
+% forward: number of forward stencils
+% centereed: number of centered stencils
+
+% justReflexive: number of only reflexive stencils
+% irreflexive: number of irreflexive stencils
+% readOnce: number of readOnce stencils
+
+% depthX: number of stencils with depth X
+% dimsX: number of stencils with dimensions=X (check this!)
+
+% lexFailed: number of files we failed to lex
+% parseFailed: number of files we failed to parse
+% lexOrParseFailed: number files we failed to lex and parse
+% parseOk: number of files parsed successfully
+%% lexFailed + parseFailed = lexOrParseFailed
+%% lexOrParseFailed + parseOk = number of files in source repo
+
+% tickAssign: number of array subscripts with a LHS which are all neighbour indices or constants
+% tickAssignSuccess: number of tickAssign's we generated some specification for
+%% tickAssign < tickAssignSuccess
 

### 2016-07-03 resultsToTex.py

@@ -1,0 +1,22 @@
+#!/usr/bin/python
+
+import sys
+import re
+
+def sanitize(s):
+    s = s.strip()
+    s = re.sub("_","",s)
+    for (i,j)  in enumerate(["zero","one","two","three","four","five","six","seven","eight","nine","ten"]):
+        s = re.sub("%d" % i,j,s)
+    return s
+
+with open(sys.argv[1]) as f:
+    for l in f:
+        if l.strip():
+            if l[0:7] == "overall":
+                corpus = "overall"
+            elif l[0:6] == "corpus":
+                corpus = l.split(":")[1].strip()
+            else:
+                (key,value) = map(lambda x:x.strip(),l.split(":"))
+                print "\\newcommand{\\%s}{%s}" % (sanitize(corpus+key),value)

### 2016-07-03 paper.tex
added details on successful parsing and explanation of failures

@@
 \[
-P(v) = \bigcup_{i \in !n} (\pi^n_{i} , \; \{[\pi^n_i
+P(v) = \bigcup_{i \in n!} (\pi^n_{i} , \; \{[\pi^n_i
 \vect{x}^L \; \pi^n_i\vect{x}^U] \, \mid \, [\vect{x}^L \; \vect{x}^U]
@@
 P(\texttt{a}) =
-\{&(\pi^2_1, \{ \stwo{0}{0}{0}{0}
-\stwo{1}{0}{1}{0}
-\stwo{-1}{0}{-1}{0}
-\stwo{0}{1}{0}{1}
+\{&(\pi^2_1, \{ \stwo{0}{0}{0}{0},
+\stwo{1}{0}{1}{0},
+\stwo{-1}{0}{-1}{0},
+\stwo{0}{1}{0}{1},
 \stwo{0}{-1}{0}{-1} \})
@@
 &(\pi^2_2, \{
- \stwo{0}{0}{0}{0}
- \stwo{0}{1}{0}{1}
- \stwo{0}{-1}{0}{-1}
- \stwo{1}{0}{1}{0}
+ \stwo{0}{0}{0}{0},
+ \stwo{0}{1}{0}{1},
+ \stwo{0}{-1}{0}{-1},
+ \stwo{1}{0}{1}{0},
  \stwo{-1}{0}{-1}{0}\})\}
@@
 \item Sort each permutation set into an ordered list, based on the ordering:
-\[
-\vect{x} \leq \vect{y} = \exists i \, . \; \vect{x}_i \leq \vect{y}_i
-\; \wedge \; \vect{x}_{i:n} = \vect{y}_{i:n}
-\]
+\begin{equation*}
+  \vect{x} \leq \vect{y} =
+      \exists i \, . \; \vect{x}_{i,1} \leq \vect{y}_{i,1} \; \wedge \;
+        (i = n \vee \vect{x}_{i+1:n} = \vect{y}_{i+1:n})
+\end{equation*}
 %
@@
 \{&(\pi^2_1, [
-\stwo{0}{-1}{0}{-1}
-\stwo{-1}{0}{-1}{0}
-\stwo{0}{0}{0}{0}
-\stwo{1}{0}{1}{0}
+\stwo{0}{-1}{0}{-1},
+\stwo{-1}{0}{-1}{0},
+\stwo{0}{0}{0}{0},
+\stwo{1}{0}{1}{0},
 \stwo{0}{1}{0}{1}] )
@@
 &(\pi^2_2, [
-\stwo{0}{-1}{0}{-1}
-\stwo{-1}{0}{-1}{0}
-\stwo{0}{0}{0}{0}
-\stwo{1}{0}{1}{0}
+\stwo{0}{-1}{0}{-1},
+\stwo{-1}{0}{-1}{0},
+\stwo{0}{0}{0}{0},
+\stwo{1}{0}{1}{0},
 \stwo{0}{1}{0}{1}])\}
@@
 (\vect{x}^L, \vect{y}^U) & \vect{x}^U_1 + 1 = \vect{y}^L_1 \; \wedge \;
-(\vect{x}^L_{1:n}, \vect{x}^U_{1:n}) = (\vect{y}^L_{1:n}, \vect{y}^U_{1:n}) \\
+\vect{x}_{2:n} = \vect{y}_{2:n}) \\
 \bot  & \textit{otherwise}
@@
 Q(v) = \{&(\pi^2_1, [
-\stwo{0}{-1}{0}{-1}
+\stwo{0}{-1}{0}{-1},
 \stwo{-1}{0}{1}{0},
-\stwo{0}{1}{0}{1},
+\stwo{0}{1}{0}{1}
 ]) \\
@@
 \[
-U(v) = \bigcup \{[\pi \vect{y}^L, \pi \vect{y}^U]
- \mid [\vect{y}^L, \vect{y}^U] \leftarrow S, (\pi, S) \leftarrow Q(v)\}
+U(v) = \bigcup \{[\pi \vect{x}^L, \pi \vect{x}^U]
+ \mid [\vect{x}^L, \vect{x}^U] \leftarrow S, (\pi, S) \leftarrow Q(v)\}
 \]
@@
 U(\texttt{a}) =
-\{\stwo{0}{-1}{0}{-1}
-\stwo{-1}{0}{1}{0}
-\stwo{0}{1}{0}{1}
-\stwo{-1}{0}{-1}{0}
-\stwo{0}{-1}{0}{1}
+\{\stwo{0}{-1}{0}{-1},
+\stwo{-1}{0}{1}{0},
+\stwo{0}{1}{0}{1},
+\stwo{-1}{0}{-1}{0},
+\stwo{0}{-1}{0}{1},
 \stwo{1}{0}{1}{0}\}
@@
 & \vect{y}^L_1 \leq \vect{x}^L_1 \wedge \vect{x}^U_1 \leq \vect{y}^U_1
-  \wedge (\vect{x}^L_{1:n}, \vect{x}^U_{1:n}) \containedin
-  (\vect{y}^L_{1:n}, \vect{y}^U_{1:n})
+  \wedge (\vect{x}^L_{2:n}, \vect{x}^U_{2:n}) \containedin
+  (\vect{y}^L_{2:n}, \vect{y}^U_{2:n})
 \end{align*}
@@
 \label{sec:evaluation}
-We now present an evaluation of the stencil language. We first confirm
-that stencils are common in scientific code and further that our
-stencil language can express a great many of these.
+We evaluated our stencil language on a corpus of scientific code. We first give an overview of the corpus itself before presenting our results.
 
@@
 \begin{figure}
-\begin{tabular}{|l|l|l|}
+\begin{tabular}{|l|l|l|l|}
 \hline
-Package       & Files & Lines of code \\
+Package       & Files & Lines of code & Files analyzed \\
 \hline
-UM            & \umFiles  & \umLoC       \\
-E3MG          & \ethreemgeaFiles   & \ethreemgeaLoC        \\
-BLAS          & \blasFiles   & \blasLoC        \\
-Hybrid4       & \hybridfourFiles    & \hybridfourLoC          \\
-GEOS-Chem     & \geoschemFiles   & \geoschemLoC       \\
-Navier        & \navierFiles     & \navierLoC           \\
-CP            & \computationalphysicstwoFiles    & \computationalphysicstwoLoC           \\
+UM            & \umFiles & \umLoC & \umparseOk  \\
+E3MG          & \ethreemgeaFiles   & \ethreemgeaLoC & \ethreemgeaparseOk  \\
+BLAS          & \blasFiles   & \blasLoC & \blasparseOk        \\
+Hybrid4       & \hybridfourFiles    & \hybridfourLoC & \hybridfourparseOk         \\
+GEOS-Chem     & \geoschemFiles   & \geoschemLoC & \geoschemparseOk       \\
+Navier        & \navierFiles     & \navierLoC & \navierparseOk           \\
+CP            & \computationalphysicstwoFiles    & \computationalphysicstwoLoC & \computationalphysicstwoparseOk           \\
 \hline
-\textbf{Total}& \overallFiles  & \overallLoC     \\
+\textbf{Total}& \overallFiles  & \overallLoC   & \overallparseOk  \\
 \hline 
@@
 
-Figure \ref{fig:corpus} shows summary statistics of the software packages used in our evaluation. In total we analysed 1.1 million lines of code from 7 packages.
+Figure \ref{fig:corpus} shows summary statistics of the software packages used in our evaluation, all of which are written in Fortran 90 or Fortran 77.
+In total we analysed 1.1 million lines of code from 7 packages.
+The ``Files analyzed'' column shows the number of files in each corpus that we were able to analyze with Camfort.
+The most common reasons for Camfort rejecting a file were either use of a C-preprocessor, or illegal use of language features from a more modern Fortran variant.
 

### 2016-07-03 paper.tex
Spelling corrections

@@
 processing, approximations to differential equations. For example, the
-following iteratively computes the one-dimensional discrete Laplace
+following iteration computes the one-dimensional discrete Laplace
 transform (an approximation to a derivative) in Fortran:
@@
 In this work, we introduce a simple specification language for the
-spatial properties of stencil comptuations. The
+spatial properties of stencil computations. The
 specifications abstract over the fine grained detail of stencil access
@@
 \end{center}
-(c). Pictoral representation of the two stencil specifications.
+(c). Pictorial representation of the two stencil specifications.
 The horizontal dimensions is \text{dim=1} and the vertical is \texttt{dim=2}:
@@
 Our specification system is based on the observation
-that most forms of array access in numberical code have
+that most forms of array access in numerical code have
 a fixed, statically-determined access pattern. For example, the
@@
 We outline the specification language here. Section~\ref{sec:syntax}
-oultines the syntax. Section~\ref{sec:semantics} defines its semantics
+outlines the syntax. Section~\ref{sec:semantics} defines its semantics
 via a simple set interpretation over indices. Section~\ref{sec:eqs} provides an
@@
   loop\footnote{Or equivalent for the target language, \eg{},
-    \texttt{do} in Fortan for our implementation}, incremented by $1$ per
+    \texttt{do} in Fortran for our implementation}, incremented by $1$ per
 iteration. The variable is marked as a base induction variable
@@
 \begin{definition}[Neighbourhood index and neighbourhood subscripts]
-  For an array subscript $v(\bar{e})$, we say say that $e \in \bar{e}$
+  For an array subscript $v(\bar{e})$, we say that $e \in \bar{e}$
   is a \emph{neighbourhood index} if it is of the form
@@
 using the Fortran style syntax of $\ldots \; \texttt{::} \bar{v}$),
-describing how the array variables $\bar{v}$ are acessed.
+describing how the array variables $\bar{v}$ are accessed.
 %Specification declarations are either \nonterm{spatial} or
 %\nonterm{temporal} specifications and describe, for spatial
-%specifcations,
+%specifications,
 %Specifications declarations describe
@@
 read using the neighbourhood indices described by both $r$ and
-$r'$. It can be thought of as a conjunction of speicifcations. For
+$r'$. It can be thought of as a conjunction of specifications. For
 example, the following gives the specification for the five-point
@@
 a region declaration, which can be used later to form a spatial
-specification. For example, the speicifcation from the previous
+specification. For example, the specification from the previous
  example can be restated as:
@@
 However, this would be an invalid specification if any of the
-array subscripts was reapeated. This modifier provides a way to
+array subscripts was repeated. This modifier provides a way to
  rule out any accidental repetition of array subscripts.
-The notion correponds to that of \emph{linearity}, though we
+The notion corresponds to that of \emph{linearity}, though we
 opt for the more informative and easily understood name
@@
 
-The \texttt{irreflexive} modifier is parameteresed by a
+The \texttt{irreflexive} modifier is parameterised by a
 comma-separated list of dimensions, indicating that for those
@@
 specifications. This is broken down into equations on regions only,
-given by the equivalence relation $\textit{region} \equiv \textit{region}'$ and on
+given by the equivalence relation $\textit{region} \equiv \textit{region}'$ and
 on specification declarations assigned to variables via the equivalence relation
@@
 where the rows are all possible $n$-vectors generated by
-non-determinstically picking each $i^{th}$ entry from either the
+non-deterministic picking each $i^{th}$ entry from either the
 $i^{th}$ entry in $u$ or the $i^{th}$ entry in $j$. For example:
@@
 selecting from $u$ for $1$ and $v$ for $0$. Thus, we can
-defined the pairwise pemutation matrix as:
+defined the pairwise permutation matrix as:
 %
@@
 %the top-level definition of the model, with $\interp{-}$ overloaded
-%on $S$, $\interp{S}$ mqpping to multisets of
-%relative indices not associated to an array avariable.
+%on $S$, $\interp{S}$ mapping to multisets of
+%relative indices not associated to an array variable.
 
@@
       scalar. $b$ is added to all.
-    \item Sliding. $a = b$ where the dimensionsare $a(1:10)$, $b(2:11)$,
+    \item Sliding. $a = b$ where the dimensions are $a(1:10)$, $b(2:11)$,
       so the assignment causes shifting.
@@
 
-\subsection{Inference}
-
-We demonstate the main steps of the inference procedure over the
+\subsection{Inference\label{sec:inference}}
+
+We demonstrate the main steps of the inference procedure over the
 following example program which computes the mean value
@@
 %%
-\paragraph{Step 1: Standard control and dataflow analyses}
+\paragraph{Step 1: Standard control and data-flow analyses}
 The inference relies on some standard program analyses, computed
@@
 \item induction variables per basic block;
-\item (interprocedural) dataflow analysis, providing a \emph{flows to}
+\item (interprocedural) data-flow analysis, providing a \emph{flows to}
   graph (as shorthand, the function
@@
   mapping an expression to the set of all expressions
-  with forwards dataflow to this expression, based on the transitive
+  with forwards data-flow to this expression, based on the transitive
   closure of the flows graph);
@@
 subscripts.  This finite map contains all array subscript expressions
-which flow to this statment. More formally, a function
+which flow to this statement. More formally, a function
 $\textsf{analyse}$ is applied to each statement in a program with the
@@
 We write $\vect{x}_i$ for the $i$-th value of vector $\vect{x}$ (1
-indexed) and $\vect{x}_{i:j}$ for the subvector from entry $i$ to
+indexed) and $\vect{x}_{i:j}$ for the sub-vector from entry $i$ to
 entry $j$ (inclusive).
@@
 \begin{definition}[Spans]
-  A \emph{span} represents an $n$-dimensional box (\emph{hyperrectangle}) as
+  A \emph{span} represents an $n$-dimensional box (\emph{hyper-rectangle}) as
   a pair of $n$-dimensional vectors (represented as a $2 \times n$
@@
   notation $[\vect{x}^L \; \vect{x}^U]$ for such a span to mark out
-  the lower and upper bound vecvtors.
+  the lower and upper bound vectors.
 \end{definition}
@@
 in the map, coalesces spans into contiguous regions in
-the $n$-dimensional spaace. That is, we compute
+the $n$-dimensional space. That is, we compute
  $(\mu \; \textsf{regions}) N$ where \textsf{regions}
@@
 permutation function with the set of permuted spans so that
-the spans can be unpermuted later.
+the spans can be un-permuted later.
 %
@@
 \item Fold each list pairwise by the following partial operation
- $\bullet$ which coalescees contiguous regions:
+ $\bullet$ which coalesces contiguous regions:
 %
@@
 %
-\item Unpermute and union together, \ie{},
+\item Un-permute and union together, \ie{},
 %
@@
 this represents a region that is not at the origin (or next to) the
-origin and thus not representable in the specificaiton language.
+origin and thus cannot be represented in the specification language.
 Thus, an upper bound is instead returned showing the maximum extent
@@
 \begin{lemma}
-$\mathsf{conv}_i$ is total given the pre-condition that the lower
+$\mathsf{conv}_i$ is total given the precondition that the lower
 bound of the input span is less than or equal to its upper bound.
@@
 \hline
-Package       & Files & Lines of code & Files analyzed \\
+Package       & Files & Lines of code & Files analysed \\
 \hline
@@
 In total we analysed 1.1 million lines of code from 7 packages.
-The ``Files analyzed'' column shows the number of files in each corpus that we were able to analyze with Camfort.
+The ``Files analysed'' column shows the number of files in each corpus that we were able to analyse with Camfort.
 The most common reasons for Camfort rejecting a file were either use of a C-preprocessor, or illegal use of language features from a more modern Fortran variant.
@@
 \textbf{The Unified Model}\footnote{\url{http://www.metoffice.gov.uk/research/modelling-systems/unified-model}} is a weather forecasting and climate modelling tool developed by the Met Office in the United Kingdom.
-It is used by research organisations and meterological services around the world.
+It is used by research organisations and meteorological services around the world.
 We use the development branch (trunk) of the model.
-The Met Office runs a comprehensive code quality system incorporating dedicated commiters (we counted 11) for particular parts of the model.
+The Met Office runs a comprehensive code quality system incorporating dedicated committers (we counted 11) for particular parts of the model.
 We counted 120 additional contributors whose submissions are reviewed and tested before being accepted into the code base.
@@
 % e3mg-ea
-\textbf{E3MG} (An Energy-Environment-Economy (E3) Model at the Global Level) is a macroeconomic model used for asssessment of environmental policy~\cite{RePEc:aen:journl:2006se-a12}.
+\textbf{E3MG} (An Energy-Environment-Economy (E3) Model at the Global Level) is a macroeconomic model used for assessment of environmental policy~\cite{RePEc:aen:journl:2006se-a12}.
 This model was developed by Cambridge Econometrics, an independent consultancy company.
 
-\textbf{BLAS} (Basic Linear Algebra Subprograms)\footnote{http://www.netlib.org/blas/} provides efficient and portable routines for vector and matrix operations.
+\textbf{BLAS} (Basic Linear Algebra Subprograms)\footnote{\url{http://www.netlib.org/blas/}} provides efficient and portable routines for vector and matrix operations.
 These routines feature in many other libraries (including LAPACK).
@@
 \end{document}
+
+%%  LocalWords:  refactoring affine parameterised irreflexive atMost
+%%  LocalWords:  centered readOnce atLeast discretisation Equational
+%%  LocalWords:  equational disjunction denotational dimensionality
+%%  LocalWords:  interprocedural Fortran Camfort preprocessor
+%%  LocalWords:  committers

### 2016-07-03 resultsToTex.py
updated python script to generate derived statistics

@@ -10,13 +10,39 @@ def sanitize(s):
         s = re.sub("%d" % i,j,s)
     return s
 
+files = {}
+
+with open(sys.argv[2]) as f:
+    for l in f:
+        (k,v) = re.match("\\\\newcommand{\\\\(.*)}{(.*)}",l.strip()).groups()
+        files[k] = v
+
+corpora = []
+
 with open(sys.argv[1]) as f:
     for l in f:
         if l.strip():
             if l[0:7] == "overall":
                 corpus = "overall"
             elif l[0:6] == "corpus":
-                corpus = l.split(":")[1].strip()
+                corpus = sanitize(l.split(":")[1])
+                corpora.append(corpus)
             else:
                 (key,value) = map(lambda x:x.strip(),l.split(":"))
-                print "\\newcommand{\\%s}{%s}" % (sanitize(corpus+key),value)
+                print "\\newcommand{\\%s}{%s}" % (corpus+sanitize(key),value)
+                files[sanitize(corpus+key)] = value
+
+(totalTickAssign, totalTickAssignSuccess, totalLoC) = (0,0,0)
+for corpus in corpora:
+    if corpus != "ethreemgmodernise":
+        loc = files[corpus+"LoC"]
+        tickAssign = files[corpus+"tickAssign"]
+        tickAssignSuccess = files[corpus+"tickAssignSuccess"]
+        totalTickAssign += int(tickAssign)
+        totalTickAssignSuccess += int(tickAssignSuccess)
+        totalLoC += int(loc)
+        print "\\newcommand{\\%s}{%.2g}" % (corpus+"tickAssignPercent", float(tickAssign) / float(loc) * 100)
+        print "\\newcommand{\\%s}{%.2g}" % (corpus+"tickAssignSuccessPercent", float(tickAssignSuccess) / float(loc) * 100)
+
+print "\\newcommand{\\%s}{%.2g}" % ("overalltickAssignPercent", float(totalTickAssign) / float(totalLoC) * 100)
+print "\\newcommand{\\%s}{%.2g}" % ("overalltickAssignSuccessPercent", float(totalTickAssignSuccess) / float(totalLoC) * 100)

### 2016-07-03 paper.tex
removed duplicate label
added table on tickAssign rates

@@
 \caption{Consistency function for a model and a set of indices}
-\label{fig:spatial-model}
+\label{fig:consistency}
 \end{figure}
@@
 
+\subsection{Occurrences of stencils}
+
+We first examined how frequently stencil computations occur in our
+corpus. We define a \emph{potential} statement as one whose left-hand
+side is an array subscript on neighbourhood indices. These statements
+are the inputs to Step 2 of the Inference process described
+Section~\ref{sec:inference}. Each potential statement for which we
+eventually generate some stencil specifications for is also counted as
+an \emph{actual} statement.
+
+\begin{figure}
+\begin{tabular}{|l|l|l|}
+\hline
+Package       & \% potential & \% actual   \\
+\hline
+UM            & \umtickAssignPercent &  \umtickAssignSuccessPercent   \\
+E3MG          & \ethreemgeatickAssignPercent & \ethreemgeatickAssignSuccessPercent\\
+BLAS          & \blastickAssignPercent & \blastickAssignSuccessPercent \\
+Hybrid4       & \hybridfourtickAssignPercent & \hybridfourtickAssignPercent \\
+GEOS-Chem     & \geoschemtickAssignPercent & \geoschemtickAssignPercent \\
+Navier        & \naviertickAssignPercent & \naviertickAssignPercent \\
+CP            & \computationalphysicstwotickAssignPercent & \computationalphysicstwotickAssignSuccessPercent\\
+\hline
+\textbf{Total}& \overalltickAssignPercent  & \overalltickAssignSuccessPercent    \\
+\hline 
+\end{tabular}
+\caption{Stencil assignment rates\label{fig:tickAssign}}
+\end{figure}
+
 \section{Discussion}

### 2016-07-03 resultsToTex.py
output a default value of 0 for keys not present

@@ -2,6 +2,7 @@
 
 import sys
 import re
+from collections import defaultdict
 
 def sanitize(s):
     s = s.strip()
@@ -17,8 +18,8 @@ with open(sys.argv[2]) as f:
         (k,v) = re.match("\\\\newcommand{\\\\(.*)}{(.*)}",l.strip()).groups()
         files[k] = v
 
-corpora = []
-
+corpora = defaultdict(lambda:{})
+keys = {}
 with open(sys.argv[1]) as f:
     for l in f:
         if l.strip():
@@ -26,15 +27,20 @@ with open(sys.argv[1]) as f:
                 corpus = "overall"
             elif l[0:6] == "corpus":
                 corpus = sanitize(l.split(":")[1])
-                corpora.append(corpus)
             else:
                 (key,value) = map(lambda x:x.strip(),l.split(":"))
-                print "\\newcommand{\\%s}{%s}" % (corpus+sanitize(key),value)
+                key = sanitize(key);
+                keys[key] = 1
+                corpora[corpus][key] = 1
+                print "\\newcommand{\\%s}{%s}" % (corpus+key,value)
                 files[sanitize(corpus+key)] = value
 
 (totalTickAssign, totalTickAssignSuccess, totalLoC) = (0,0,0)
-for corpus in corpora:
-    if corpus != "ethreemgmodernise":
+for corpus in corpora.keys():
+    if corpus != "overall" and corpus != "ethreemgmodernise":
+        for key in keys.keys():
+            if not corpora[corpus].has_key(key):
+                print "\\newcommand{\\%s}{0}" % (corpus+key)
         loc = files[corpus+"LoC"]
         tickAssign = files[corpus+"tickAssign"]
         tickAssignSuccess = files[corpus+"tickAssignSuccess"]
@@ -46,3 +52,4 @@ for corpus in corpora:
 
 print "\\newcommand{\\%s}{%.2g}" % ("overalltickAssignPercent", float(totalTickAssign) / float(totalLoC) * 100)
 print "\\newcommand{\\%s}{%.2g}" % ("overalltickAssignSuccessPercent", float(totalTickAssignSuccess) / float(totalLoC) * 100)
+

### 2016-07-03 paper.tex
related work on PLDI'16 paper, and examples

--- 2016-07-03 paper.tex
@@
 \definecolor{darkgreen}{rgb}{0.0,0.4,0.0}
+\definecolor{darkdarkgreen}{rgb}{0.0,0.35,0.0}
+
+\newcommand{\sten}[1]{\textcolor{darkdarkgreen}{\!#1}}
 
@@
 \CustomVerbatimEnvironment{ExmVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.5cm,
-xrightmargin=0.2cm,baselinestretch=0.98,numbers=left,numbersep=0.9em}
+xrightmargin=0.2cm,baselinestretch=0.98,numbers=left,numbersep=0.9em,commandchars=\\\{\}}
 \CustomVerbatimEnvironment{IVerbatim}{Verbatim}{fontsize=\relsize{-1},xleftmargin=0.5cm,
@@
 
-\newcommand{\stenRefl}[1]{\term{reflexive} \, (\term{dim=}#1)}
 \newcommand{\stenFwd}[2]{\term{forward} \, (\term{depth=}#1,
@@
   \term{dim=}#2)}
-\newcommand{\stenReflS}[1]{\term{refl} \, (\term{dim=}#1)}
 \newcommand{\stenFwdS}[2]{\term{fwd} \, (\term{depth=}#1,
@@
 \newcommand{\singleEntry}[2]{\textbf{J}_{#2}^{#1}}
-\newcommand{\zeroEntry}[2]{\textbf{K}_{#2}^{#1}}
 
@@
 
-\include{results}
-\include{corpussize}
-
-\title{Verification of Stencil Computations through Spatial Specifications}
+\title{Abstract Spatial Specification of Stencil Computations}
 \authorinfo{Dominic Orchard \and Mistral Contrastin
@@
 \nonterm{specTop} ::= & [\nonterm{approxMod},] \; \nonterm{spec} \\
-\nonterm{spec} ::= & [\nonterm{modMult},] \; \nonterm{refRegion} \\
-\nonterm{refRegion} ::= & [\nonterm{modRefl},] \; \nonterm{region}\\
-\nonterm{modMult} ::= & \term{readOnce} \\
-\nonterm{modRefl} ::= & \term{irreflexive} \, (\term{dims=}\dimTy \{\term{,}
+\nonterm{spec} ::= & [\nonterm{mod}] \; [\nonterm{region}] \\
+\nonterm{mod} ::= & [\nonterm{modR},] \; [\nonterm{modRefl},] \; [\nonterm{modIrefl},] \\
+\nonterm{modR} ::= &
+ \term{readOnce} \\
+\nonterm{modRefl} ::= & \term{reflexive} \, (\term{dims=}\dimTy \{\term{,} \; \dimTy\}) \\
+\nonterm{modIrrefl} ::= & \term{irreflexive} \, (\term{dims=}\dimTy \{\term{,}
           \dimTy\}) \\
 \nonterm{approxMod} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
-\nonterm{region} ::=
-        & \stenFwd{\mathbb{N}_{>0}}{\dimTy}  \\
+\nonterm{region} ::= & \stenFwd{\mathbb{N}_{>0}}{\dimTy}  \\
 \mid \; & \stenBwd{\mathbb{N}_{>0}}{\dimTy} \\
 \mid \; & \stenCen{\mathbb{N}_{>0}}{\dimTy} \\
-\mid \; & \stenRefl{\dimTy}  \\
 \mid \; & \nonterm{region} \; \term{+} \; \nonterm{region} \\
@@
 %%
+Such a stencil computation is a common image convolution (for example
+for edge detection). 
 Note that there are a number of intermediate assignments over which
 the whole specification ranges.
+
+As another example, the following code and specification
+implements the \emph{Roberts cross}
+edge-detection convolution~\cite{davis1975survey}:
+%%
+\begin{ExmVerbatim}
+do j=0, jmax-1
+    do i=0, imax-1
+\sten{      != region r1 = forward(depth=1, dim=1)}
+\sten{      != region r2 = forward(depth=1, dim=2)}
+\sten{      != stencil readOnce, r1*r2 :: a}
+      x = a(i,j) - a(i+1,j+1)
+      y = a(i+1,j) - a(i,j+1)
+    b(i,j) = sqrt(x*x+y*y)
+   end do
+end do
+\end{ExmVerbatim}
 
@@
 To save space we use abbreviations
-\term{refl}, \term{fwd}, \term{bwd}, and \term{cen} for \term{reflexive},
-\term{forward}, \term{backward}, and \term{centered} regions respectively.
-The equations reveal something of their semantics.  The first six
+\term{fwd}, \term{bwd}, and \term{cen} for \term{forward},
+\term{backward}, and \term{centered} regions respectively.  The
+equations reveal something of their semantics.  The first six
 equations (with labels of the form $\ldots \texttt{+} \ldots$)
@@
 
-Figure~\ref{fig:region-model} outlines the core rules for regions
+Figure~\ref{fig:model} outlines the core rules for regions
 via the interpretation function $\interp{-}$.  We explain the
@@
 $\singleEntry{r}{n}$, is a vector
-where the $r^{th}$ entry is $1$ and all others are $\infty$, \eg{},
-$\singleEntry{1}{2} = \vtwoh{1}{\infty}$.
+where the $r^{th}$ entry is $1$ and all others are $0$, \eg{},
+$\singleEntry{1}{2} = \vtwoh{1}{0}$.
 \end{definition}
 
-\begin{definition}A \emph{zero-entry vector} of size $n$, denoted
-$\zeroEntry{r}{n}$, is a vector where the $r^{th}$ entry is $0$ and all others
-are $\infty$, \eg{}, $\zeroEntry{1}{2} = \vtwoh{0}{\infty}$.
-\end{definition}
-
 \begin{definition}The product of two models $\otimes$ is defined:
 %%
-\begin{alignat*}{2}
-& dims(M_m,N_n) =
-  \bigg\{ x \; \bigg| \;
-    \parbox{4.1cm}{
-      $x \in \{1, \ldots, m\} \wedge M_x \neq \infty \; \vee \;$ \\
-      $x \in \{1, \ldots, n\} \wedge N_x \neq \infty$
-                  } \bigg\} \\
-& M_n \otimes N_n =
-  \Bigg\{x \; \Bigg| \;
-    \parbox{4.8cm}{$i \in \{1, \ldots, 2^n\},$
-                   $(u, v) \in M_n \times N_n,$ \\
-                   $x = (u \bowtie v)_i,$ \\
-                   $\forall j \in dims(M_n,N_n) (x_j \neq \infty) $
-                  } \Bigg\}
-\end{alignat*}
+\begin{align*}
+M \otimes N & =
+  \{(u \bowtie v)_i \mid i \in \{1, \ldots, 2^n\}, (u, v) \in M \times N\}
+\end{align*}
 The $\bowtie$ operation, which we call \emph{pairwise permutation}
@@
 where $b$ is a logical matrix where $b_{i,j}$ is the $j$-th bit of the integer
-$i - 1$. Recall that the vectors are defined over the domain $\mathbb{Z} \cup
-\{\infty\}$. In pairwise permutation $\infty$
+$i - 1$. Recall that the vectors are defined over the domain
+$\mathbb{Z} \cup \{\infty\}$.
+For addition and multiplication, we take
+ $\infty$ to be an absorbing element. Thus, if either
+vector in a pairwise permutation contains $\infty$ in
+the $j^{th}$ entry, then $j^{th}$ entry of every pairwise
+permutation is $\infty$, \eg{},
+%%
+\begin{equation*}
+\vtwoh{0}{1} \bowtie \vtwoh{2}{\infty} =
+\setlength{\arraycolsep}{0.5em}
+\left[\begin{array}{cc}
+0 & \infty \\
+0 & \infty \\
+2 & \infty \\
+2 & \infty
+\end{array}
+\right]
+\end{equation*}
 
@@
 \begin{figure}
-\input{region-model.tex}
+\input{model.tex}
 %\textit{$a^m \in A$ means there are $n$ copies of $a$ in
 %  the multi-set $A$}
-\caption{Set model of regions}
-\label{fig:region-model}
-\end{figure}
-
-\begin{figure}
-\input{spatial-model.tex}
-\caption{Model of specifications}
-\label{fig:spatial-model}
-\end{figure}
-
-\begin{figure}
-\input{consistency.tex}
-\caption{Consistency function for a model and a set of indices}
-\label{fig:spatial-model}
+\caption{Set model of specifications}
+\label{fig:model}
 \end{figure}
@@
 \draw[thick] (0.5,1) rectangle (2,1.5);
-\draw[thick] (1,0.40) rectangle (1.5,0.90);
-\draw[thick] (1,1.60) rectangle (1.5,2.10);
+\draw[thick] (1,0.5) rectangle (1.5,1);
+\draw[thick] (1,1.5) rectangle (1.5,2);
 \end{tikzpicture}
@@
 %\node at (1.25,1.22) {\textnormal{\small{0,0}}};
-\draw[thick] (1.6,1) rectangle (2.1,1.5);
-\draw[thick] (0.4,1) rectangle (0.9,1.5);
+\draw[thick] (1.5,1) rectangle (2,1.5);
+\draw[thick] (0.5,1) rectangle (1,1.5);
 \draw[thick] (1,0.5) rectangle (1.5,2);
@@
 \[
-P(v) = \bigcup_{i \in n!} (\pi^n_{i} , \; \{[\pi^n_i
+P(v) = \bigcup_{i \in !n} (\pi^n_{i} , \; \{[\pi^n_i
 \vect{x}^L \; \pi^n_i\vect{x}^U] \, \mid \, [\vect{x}^L \; \vect{x}^U]
@@
 P(\texttt{a}) =
-\{&(\pi^2_1, \{ \stwo{0}{0}{0}{0},
-\stwo{1}{0}{1}{0},
-\stwo{-1}{0}{-1}{0},
-\stwo{0}{1}{0}{1},
+\{&(\pi^2_1, \{ \stwo{0}{0}{0}{0}
+\stwo{1}{0}{1}{0}
+\stwo{-1}{0}{-1}{0}
+\stwo{0}{1}{0}{1}
 \stwo{0}{-1}{0}{-1} \})
@@
 &(\pi^2_2, \{
- \stwo{0}{0}{0}{0},
- \stwo{0}{1}{0}{1},
- \stwo{0}{-1}{0}{-1},
- \stwo{1}{0}{1}{0},
+ \stwo{0}{0}{0}{0}
+ \stwo{0}{1}{0}{1}
+ \stwo{0}{-1}{0}{-1}
+ \stwo{1}{0}{1}{0}
  \stwo{-1}{0}{-1}{0}\})\}
@@
 \item Sort each permutation set into an ordered list, based on the ordering:
-\begin{equation*}
-  \vect{x} \leq \vect{y} =
-      \exists i \, . \; \vect{x}_{i,1} \leq \vect{y}_{i,1} \; \wedge \;
-        (i = n \vee \vect{x}_{i+1:n} = \vect{y}_{i+1:n})
-\end{equation*}
+\[
+\vect{x} \leq \vect{y} = \exists i \, . \; \vect{x}_i \leq \vect{y}_i
+\; \wedge \; \vect{x}_{i:n} = \vect{y}_{i:n}
+\]
 %
@@
 \{&(\pi^2_1, [
-\stwo{0}{-1}{0}{-1},
-\stwo{-1}{0}{-1}{0},
-\stwo{0}{0}{0}{0},
-\stwo{1}{0}{1}{0},
+\stwo{0}{-1}{0}{-1}
+\stwo{-1}{0}{-1}{0}
+\stwo{0}{0}{0}{0}
+\stwo{1}{0}{1}{0}
 \stwo{0}{1}{0}{1}] )
@@
 &(\pi^2_2, [
-\stwo{0}{-1}{0}{-1},
-\stwo{-1}{0}{-1}{0},
-\stwo{0}{0}{0}{0},
-\stwo{1}{0}{1}{0},
+\stwo{0}{-1}{0}{-1}
+\stwo{-1}{0}{-1}{0}
+\stwo{0}{0}{0}{0}
+\stwo{1}{0}{1}{0}
 \stwo{0}{1}{0}{1}])\}
@@
 (\vect{x}^L, \vect{y}^U) & \vect{x}^U_1 + 1 = \vect{y}^L_1 \; \wedge \;
-\vect{x}_{2:n} = \vect{y}_{2:n}) \\
+(\vect{x}^L_{1:n}, \vect{x}^U_{1:n}) = (\vect{y}^L_{1:n}, \vect{y}^U_{1:n}) \\
 \bot  & \textit{otherwise}
@@
 Q(v) = \{&(\pi^2_1, [
-\stwo{0}{-1}{0}{-1},
+\stwo{0}{-1}{0}{-1}
 \stwo{-1}{0}{1}{0},
-\stwo{0}{1}{0}{1}
+\stwo{0}{1}{0}{1},
 ]) \\
@@
 \[
-U(v) = \bigcup \{[\pi \vect{x}^L, \pi \vect{x}^U]
- \mid [\vect{x}^L, \vect{x}^U] \leftarrow S, (\pi, S) \leftarrow Q(v)\}
+U(v) = \bigcup \{[\pi \vect{y}^L, \pi \vect{y}^U]
+ \mid [\vect{y}^L, \vect{y}^U] \leftarrow S, (\pi, S) \leftarrow Q(v)\}
 \]
@@
 U(\texttt{a}) =
-\{\stwo{0}{-1}{0}{-1},
-\stwo{-1}{0}{1}{0},
-\stwo{0}{1}{0}{1},
-\stwo{-1}{0}{-1}{0},
-\stwo{0}{-1}{0}{1},
+\{\stwo{0}{-1}{0}{-1}
+\stwo{-1}{0}{1}{0}
+\stwo{0}{1}{0}{1}
+\stwo{-1}{0}{-1}{0}
+\stwo{0}{-1}{0}{1}
 \stwo{1}{0}{1}{0}\}
@@
 & \vect{y}^L_1 \leq \vect{x}^L_1 \wedge \vect{x}^U_1 \leq \vect{y}^U_1
-  \wedge (\vect{x}^L_{2:n}, \vect{x}^U_{2:n}) \containedin
-  (\vect{y}^L_{2:n}, \vect{y}^U_{2:n})
+  \wedge (\vect{x}^L_{1:n}, \vect{x}^U_{1:n}) \containedin
+  (\vect{y}^L_{1:n}, \vect{y}^U_{1:n})
 \end{align*}
@@
 \label{sec:evaluation}
-We evaluated our stencil language on a corpus of scientific code. We first give an overview of the corpus itself before presenting our results.
-
-\subsection{Software Corpus}
-\begin{figure}
-\begin{tabular}{|l|l|l|l|}
-\hline
-Package       & Files & Lines of code & Files analyzed \\
-\hline
-UM            & \umFiles & \umLoC & \umparseOk  \\
-E3MG          & \ethreemgeaFiles   & \ethreemgeaLoC & \ethreemgeaparseOk  \\
-BLAS          & \blasFiles   & \blasLoC & \blasparseOk        \\
-Hybrid4       & \hybridfourFiles    & \hybridfourLoC & \hybridfourparseOk         \\
-GEOS-Chem     & \geoschemFiles   & \geoschemLoC & \geoschemparseOk       \\
-Navier        & \navierFiles     & \navierLoC & \navierparseOk           \\
-CP            & \computationalphysicstwoFiles    & \computationalphysicstwoLoC & \computationalphysicstwoparseOk           \\
-\hline
-\textbf{Total}& \overallFiles  & \overallLoC   & \overallparseOk  \\
-\hline 
-\end{tabular}
-\caption{Summary of software packages used for evaluation\label{fig:corpus}}
-\end{figure}
-
-Figure \ref{fig:corpus} shows summary statistics of the software packages used in our evaluation, all of which are written in Fortran 90 or Fortran 77.
-In total we analysed 1.1 million lines of code from 7 packages.
-The ``Files analyzed'' column shows the number of files in each corpus that we were able to analyze with Camfort.
-The most common reasons for Camfort rejecting a file were either use of a C-preprocessor, or illegal use of language features from a more modern Fortran variant.
-
-\textbf{The Unified Model}\footnote{\url{http://www.metoffice.gov.uk/research/modelling-systems/unified-model}} is a weather forecasting and climate modelling tool developed by the Met Office in the United Kingdom.
-It is used by research organisations and meterological services around the world.
-We use the development branch (trunk) of the model.
-The Met Office runs a comprehensive code quality system incorporating dedicated commiters (we counted 11) for particular parts of the model.
-We counted 120 additional contributors whose submissions are reviewed and tested before being accepted into the code base.
-
-% e3mg-ea
-\textbf{E3MG} (An Energy-Environment-Economy (E3) Model at the Global Level) is a macroeconomic model used for asssessment of environmental policy~\cite{RePEc:aen:journl:2006se-a12}.
-This model was developed by Cambridge Econometrics, an independent consultancy company.
-
-\textbf{BLAS} (Basic Linear Algebra Subprograms)\footnote{http://www.netlib.org/blas/} provides efficient and portable routines for vector and matrix operations.
-These routines feature in many other libraries (including LAPACK).
-We used version 3.6.0.
-
-\textbf{Hybrid4} is a vegetation and biomass model for simulating carbon, water and nitrogen flows~\cite{GBC:GBC635}.
-
-\textbf{GEOS-Chem}\footnote{\url{http://acmg.seas.harvard.edu/geos/}} is a three-dimensional model of tropospheric chemistry developed at Harvard and used by approximately 70 universities and research institutions world-wide.
-We examined version 10-01.
-
-\textbf{Navier} is an implementation of a Navier-Stokes fluid dynamics computation~\cite{griebel1997numerical}.
-
-\textbf{CP} consists of the example code from the book ``Computational Physics''~\cite{nicholas2006computational} introducing numerical techniques and their application to modern physics problems such as fields, waves, statistical mechanics and quantum mechanics.
-
-% atLeast: number of stencils with an atLeast modifier
-% atMost: number of stencils with an atMost modifier
-% boundedBoth: number of stencils with both atLeast and atMost modifier
-%% atLeast == atMost == boundedBoth - whenever we infer one we infer the other but users might write just one
-
-% backward: number of backward stencils
-% forward: number of forward stencils
-% centereed: number of centered stencils
-
-% justReflexive: number of only reflexive stencils
-% irreflexive: number of irreflexive stencils
-% readOnce: number of readOnce stencils
-
-% depthX: number of stencils with depth X
-% dimsX: number of stencils with dimensions=X (check this!)
-
-% lexFailed: number of files we failed to lex
-% parseFailed: number of files we failed to parse
-% lexOrParseFailed: number files we failed to lex and parse
-% parseOk: number of files parsed successfully
-%% lexFailed + parseFailed = lexOrParseFailed
-%% lexOrParseFailed + parseOk = number of files in source repo
-
-% tickAssign: number of array subscripts with a LHS which are all neighbour indices or constants
-% tickAssignSuccess: number of tickAssign's we generated some specification for
-%% tickAssign < tickAssignSuccess
 
@@
 
+\subsection{Related work}
+
+Various prior approaches to specifying and verifying the behaviour
+of stencils have been based on fine-grained representations of the
+code. These essentially duplicate the indexing scheme of a stencil
+computation in their specifications.
+
+\citet{kamil2016verified} propose a technique they call
+\emph{verified lifting} for mapping low-level, possibly
+greatly optimised, stencil computations into a high-level
+predicate language using inductive program synthesis. 
+The aim is that these high-level abstract specifications can
+then be translated into high-performance DSLs, easing the
+use of high-performance DSLs. Compared to our approach,
+their specification language is more fine grained, capturing
+the exact indexing pattern of a stencil. For example, the following
+is an example post-condition in their language, generated by their
+compiler \textsc{stng} from a Fortran stencil computation~\cite [p.3]{kamil2016verified}:
+%
+\begin{align*}
+& \textit{post}(\texttt{a}, \texttt{b}) \equiv \forall \texttt{imin}+1 
+\leq \texttt{i} \leq \texttt{imax}, \texttt{jmin} \leq \texttt{j} \leq
+\texttt{jmax} \\
+& \qquad \texttt{a}(\texttt{i},\texttt{j}) = 
+\texttt{b}(\texttt{i}-1,\texttt{j}) + \texttt{b}(\texttt{i},\texttt{j})
+\end{align*}
+%
+Noticeable, the bounds of the stencil and the exact indexing pattern
+are captured. Our approach aims to be much more abstract, to
+facilitate human reasoning, and avoid low-level lexical
+errors. However, \textsf{stng} specifications are generated, \ie{}, 
+are never written by a programmer, avoiding this
+problem. Their inductive program synthesis approach allows
+specifications to extracted even from optimised and unrolled code. For
+example, the above post condition is generated from the code~\cite [p.3]{kamil2016verified}:
+%
+\begin{ExmVerbatim}
+do j=jmin, jmax 
+    t = b(imin, j)
+    do i=imin+1,imax
+        q = b(i, j)
+        a(i, j) = q + t
+        t = q
+    end do
+end do
+\end{ExmVerbatim}
+%
+Thus, \textsc{stng} can understand cross-loop data-dependencies
+in order to assign the above post-condition. CamFort is not currently
+able to ascribe a specification in the above case as the use of
+\texttt{t} on line 5 maps only to the definition on line 2, and no on
+line 6. Extending our approach to follow loop-carried dependencies is
+future work. An alternate approach for us would be to build ontop of
+the work of Kamil~\emph{et al.} by inferring our specifications from 
+\textsc{stng} (as an intermediate step).
+
+In an alternate approach,~\citet{abe2013model}
+
+There are many DSLs for stencil comptuation aimed at either
+performance or verification, or both. Many elimiante out-of-bounds
+errors, \eg{},~\cite{DBLP:journals/corr/abs-1109-0777}.
+
 \bibliography{references}

### 2016-07-03 paper.tex
abstract tweak

@@
 \begin{abstract}
-Verifying the correctness of numerical computations relies on a specification of their behaviour.
-A natural way to do this would be in terms of the original mathematical model but this approach yields specifications which are highly decoupled from the code.
-Checking the correctness of code which implements a discrete approximation of such specifications is extremely challenging.
-Instead we have been investigating intermediate
-  systems of specification and verification between lower-level
-  aspects (\eg{}, out-of-bounds errors) and full correctness
-  based on a numerical model. In this paper, we focus on
-  \emph{stencil computations}---a common idiom in numerical code, but
-  one which is error-prone due to fine-grained indexing errors.
-%
-We describe a stencil specification language which can be directly embedded as annotations in source code, along with an inference and checking procedure.
-%
-We evaluate our language against a corpus of numerical Fortran code, and show
-  that precise specifications can be given to $X\%$ of identifiable
-  stencil computations, with a mean ratio of $Y:Z$
-  specification-to-program code and that the vast majority of stencil computations have a simple, regular, static shape.
-%
-This paper details our stencil
-  specification language, its static semantics, inference, checking,
-  and program synthesis procedures, our implementation, and our
-  evaluation study.
-%
-We provide an implementation of stencil inference and validation in \emph{Camfort}, an open-source program analysis and refactoring tool.
+  Verifying the correctness of numerical computations relies on first
+  giving a specification of their behaviour. In many cases, a
+  numerical computation is derived from an underlying mathematical
+  model, but this is often difficult to use as a specification as it
+  is highly decoupled from the implementation. We have been studying
+  specification techniques in middle ground between low-level and
+  high-level specifications for numerical code. In this paper, we
+  focus on \emph{stencil computations}---a common idiom in numerical code, but
+  one which is error-prone due to fine-grained indexing errors. We
+  hypothesise that in practise, stencil computations tend to have a
+  regular shape which is amenable to static analysis and simple
+  specification. We describe a abstract specification language for
+  stencil computations that can be directly embedded as annotations in
+  source code, along with an inference and checking procedure. We
+  evaluate our language against a corpus of numerical Fortran code,
+  and show that precise specifications can be given to roughly half of
+  the identifiable stencil computations, and that the vast majority of
+  stencil computations indeed have a simple, regular, static
+  shape. This paper details our stencil specification language, its
+  static semantics, inference, checking, and program synthesis
+  procedures, our implementation, and our evaluation study. We provide
+  an implementation of stencil inference and validation in Camfort, an
+  open-source program analysis and refactoring tool.
 \end{abstract}

### 2016-07-03 paper.tex
started to count bounded/exact

--- 2016-07-03 paper.tex
@@
 \definecolor{darkgreen}{rgb}{0.0,0.4,0.0}
-\definecolor{darkdarkgreen}{rgb}{0.0,0.35,0.0}
-
-\newcommand{\sten}[1]{\textcolor{darkdarkgreen}{\!#1}}
 
@@
 \CustomVerbatimEnvironment{ExmVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.5cm,
-xrightmargin=0.2cm,baselinestretch=0.98,numbers=left,numbersep=0.9em,commandchars=\\\{\}}
+xrightmargin=0.2cm,baselinestretch=0.98,numbers=left,numbersep=0.9em}
 \CustomVerbatimEnvironment{IVerbatim}{Verbatim}{fontsize=\relsize{-1},xleftmargin=0.5cm,
@@
 \newcommand{\mnote}[1]{\textcolor{darkgreen}{Mistral: #1}}
+\newcommand{\anote}[1]{\textcolor{red}{Andy: #1}}
 
@@
 %%
-Such a stencil computation is a common image convolution (for example
-for edge detection). 
 Note that there are a number of intermediate assignments over which
 the whole specification ranges.
-
-As another example, the following code and specification
-implements the \emph{Roberts cross}
-edge-detection convolution~\cite{davis1975survey}:
-%%
-\begin{ExmVerbatim}
-do j=0, jmax-1
-    do i=0, imax-1
-\sten{      != region r1 = forward(depth=1, dim=1)}
-\sten{      != region r2 = forward(depth=1, dim=2)}
-\sten{      != stencil readOnce, r1*r2 :: a}
-      x = a(i,j) - a(i+1,j+1)
-      y = a(i+1,j) - a(i,j+1)
-    b(i,j) = sqrt(x*x+y*y)
-   end do
-end do
-\end{ExmVerbatim}
 
@@
 \begin{figure}
-\begin{tabular}{|l|l|l|}
+\begin{tabular}{|l|l|l|l|l|}
 \hline
-Package       & \% potential & \% actual   \\
+Package       & \% potential & \% actual & Total & Total    \\
+              & stencils     & stencils  & exact  & bounded \\
 \hline
-UM            & \umtickAssignPercent &  \umtickAssignSuccessPercent   \\
-E3MG          & \ethreemgeatickAssignPercent & \ethreemgeatickAssignSuccessPercent\\
-BLAS          & \blastickAssignPercent & \blastickAssignSuccessPercent \\
-Hybrid4       & \hybridfourtickAssignPercent & \hybridfourtickAssignPercent \\
-GEOS-Chem     & \geoschemtickAssignPercent & \geoschemtickAssignPercent \\
-Navier        & \naviertickAssignPercent & \naviertickAssignPercent \\
-CP            & \computationalphysicstwotickAssignPercent & \computationalphysicstwotickAssignSuccessPercent\\
+UM            & \umtickAssignPercent & \umtickAssignSuccessPercent & \umnumStencilLines & \umboundedBoth  \\
+E3MG          & \ethreemgeatickAssignPercent & \ethreemgeatickAssignSuccessPercent & \ethreemgeanumStencilLines & \ethreemgeaboundedBoth \\
+BLAS          & \blastickAssignPercent & \blastickAssignSuccessPercent & \blasnumStencilLines & \blasboundedBoth \\
+Hybrid4       & \hybridfourtickAssignPercent & \hybridfourtickAssignPercent & \hybridfournumStencilLines &  \hybridfourboundedBoth\\
+GEOS-Chem     & \geoschemtickAssignPercent & \geoschemtickAssignPercent & \geoschemnumStencilLines & \geoschemboundedBoth \\
+Navier        & \naviertickAssignPercent & \naviertickAssignPercent & \naviernumStencilLines & \navierboundedBoth \\
+CP            & \computationalphysicstwotickAssignPercent & \computationalphysicstwotickAssignSuccessPercent & \computationalphysicstwonumStencilLines & \computationalphysicstwoboundedBoth \\
 \hline
-\textbf{Total}& \overalltickAssignPercent  & \overalltickAssignSuccessPercent    \\
+\textbf{Total}& \overalltickAssignPercent  & \overalltickAssignSuccessPercent & \overallnumStencilLines & \overallboundedBoth    \\
 \hline 
@@
 
+The first two columns of Figure \ref{fig:tickAssign} shows the percentage of lines which were
+assigned potential and actual stencils.
+\anote{Add explanation about why our results are low. Include an example from BLAS of a tickAssign'd statement that is not a stencil}
+\anote{Add discussion about the rate of stencils in the code. Comparison with rate of assignment statements would be nice}
+
+If the inference process cannot determine an exact stencil for a statement it will instead produce an upper (\term{atMost}) and lower (\term{atLeast}) bound.
+The numbers of exact and bounded stencils are shown in columns three and four of Figure \ref{fig:tickAssign}.
+We can see that our stencil language is capable of precisely specifying the overwhelming majority of stencils.
+Most of the bounded stencils apply to some form of data copying.
+For example (taken from E3MG):
+\begin{ExmVerbatim}
+!- stencil atLeast, readOnce, irreflexive(dims=2), (reflexive(dim=1)) :: sfda
+!- stencil atMost, readOnce, (forward(depth=8, dim=2))*(reflexive(dim=1)) :: sfdt
+ZZ1(I,J) = SFDT(I,J+8)
+\end{ExmVerbatim}
+
 \section{Discussion}
 \label{sec:discussion}
-
-\subsection{Related work}
-
-Various prior approaches to specifying and verifying the behaviour
-of stencils have been based on fine-grained representations of the
-code. These essentially duplicate the indexing scheme of a stencil
-computation in their specifications.
-
-\citet{kamil2016verified} propose a technique they call
-\emph{verified lifting} for mapping low-level, possibly
-greatly optimised, stencil computations into a high-level
-predicate language using inductive program synthesis. 
-The aim is that these high-level abstract specifications can
-then be translated into high-performance DSLs, easing the
-use of high-performance DSLs. Compared to our approach,
-their specification language is more fine grained, capturing
-the exact indexing pattern of a stencil. For example, the following
-is an example post-condition in their language, generated by their
-compiler \textsc{stng} from a Fortran stencil computation~\cite [p.3]{kamil2016verified}:
-%
-\begin{align*}
-& \textit{post}(\texttt{a}, \texttt{b}) \equiv \forall \texttt{imin}+1 
-\leq \texttt{i} \leq \texttt{imax}, \texttt{jmin} \leq \texttt{j} \leq
-\texttt{jmax} \\
-& \qquad \texttt{a}(\texttt{i},\texttt{j}) = 
-\texttt{b}(\texttt{i}-1,\texttt{j}) + \texttt{b}(\texttt{i},\texttt{j})
-\end{align*}
-%
-Noticeable, the bounds of the stencil and the exact indexing pattern
-are captured. Our approach aims to be much more abstract, to
-facilitate human reasoning, and avoid low-level lexical
-errors. However, \textsf{stng} specifications are generated, \ie{}, 
-are never written by a programmer, avoiding this
-problem. Their inductive program synthesis approach allows
-specifications to extracted even from optimised and unrolled code. For
-example, the above post condition is generated from the code~\cite [p.3]{kamil2016verified}:
-%
-\begin{ExmVerbatim}
-do j=jmin, jmax 
-    t = b(imin, j)
-    do i=imin+1,imax
-        q = b(i, j)
-        a(i, j) = q + t
-        t = q
-    end do
-end do
-\end{ExmVerbatim}
-%
-Thus, \textsc{stng} can understand cross-loop data-dependencies
-in order to assign the above post-condition. CamFort is not currently
-able to ascribe a specification in the above case as the use of
-\texttt{t} on line 5 maps only to the definition on line 2, and no on
-line 6. Extending our approach to follow loop-carried dependencies is
-future work. An alternate approach for us would be to build ontop of
-the work of Kamil~\emph{et al.} by inferring our specifications from 
-\textsc{stng} (as an intermediate step).
-
-In an alternate approach,~\citet{abe2013model}
-
-There are many DSLs for stencil comptuation aimed at either
-performance or verification, or both. Many elimiante out-of-bounds
-errors, \eg{},~\cite{DBLP:journals/corr/abs-1109-0777}.
 

### 2016-07-03 paper.tex

--- 2016-07-03 paper.tex
@@
 \end{figure}
+
+The first two columns of Figure \ref{fig:tickAssign} shows the percentage of lines which were
+assigned potential and actual stencils.
+\anote{Add explanation about why our results are low. Include an example from BLAS of a tickAssign'd statement that is not a stencil}
+\anote{Add discussion about the rate of stencils in the code. Comparison with rate of assignment statements would be nice}
+
+If the inference process cannot determine an exact stencil for a statement it will instead produce an upper (\term{atMost}) and lower (\term{atLeast}) bound.
+The numbers of exact and bounded stencils are shown in columns three and four of Figure \ref{fig:tickAssign}.
+We can see that our stencil language is capable of precisely specifying the overwhelming majority of stencils.
+Most of the bounded stencils apply to some form of data copying.
+For example (taken from E3MG):
+\begin{ExmVerbatim}
+!- stencil atLeast, readOnce, irreflexive(dims=2), (reflexive(dim=1)) :: sfda
+!- stencil atMost, readOnce, (forward(depth=8, dim=2))*(reflexive(dim=1)) :: sfdt
+ZZ1(I,J) = SFDT(I,J+8)
+\end{ExmVerbatim}
 

### 2016-07-03 paper.tex
fixed the definition of effective dimensions
merge-conflict

@@
 specifications abstract over the fine grained detail of stencil access
-patterns.  In practise, most stencil computations have a regular shape
+patterns. In practice, most stencil computations have a regular shape
 that can be described simply and abstractly, with a small set of
@@
 \end{ExmVerbatim}
-(a). Excerpt from Fortran code for a Navier-Stoke fluid simulator,
+(a) Excerpt from Fortran code for a Navier-Stoke fluid simulator,
 showing highly-detailed stencil computations. \\
@@
 
-\begin{SpecVerbatim}[xleftmargin=0.3cm]
-!=   stencil centered(depth=1, dim=1)
-           + centered(depth=1, dim=2) :: u
+\begin{SpecVerbatim}[xleftmargin=0cm]
+!= stencil centered(depth=1, dim=1)*reflexive(dim=2)
+         + centered(depth=1, dim=2)*reflexive(dim=1) :: u
 
@@
 \end{SpecVerbatim}
-(b). Inferred stencil specification from CamFort
+(b) Inferred stencil specification from CamFort
 
@@
 \end{center}
-(c). Pictorial representation of the two stencil specifications.
+(c) Pictorial representation of the two stencil specifications.
 The horizontal dimensions is \text{dim=1} and the vertical is \texttt{dim=2}:
@@
 Regions have as terminals the \emph{region constants}, denoted by
-\term{forward}, \term{backward}, or \term{centered}. Each
-region constant is given a depth parameter $n$ (natural number greater
-than 0) and a dimension identifier $d$ (also a natural number greater than
-0). Region constants specify that an array is read from by a collection
-of indices which in the $d^{th}$ dimension are
-neighbourhood indices ranging from $i + 0$ up to $i + n$ inclusively, \eg{}, the
-following is valid stencil, reading from \term{a} and writing to \term{b}:
+\term{reflexive}, \term{forward}, \term{backward}, or \term{centered}. Each
+region constant except \term{reflexive} is given a depth parameter $n$ (natural
+number greater than 0). They all receive a dimension identifier $d$ (also a
+natural number greater than 0). Region constants specify that an array is read
+from by a collection of indices which in the $d^{th}$ dimension are
+neighbourhood indices ranging from $i + 0$ up to $i \pm n$ inclusively,
+\eg{}, the following is valid stencil, reading from \term{a} and writing to
+\term{b}:
 %%
@@
 The sum of two regions $r \texttt{+} r'$ specifies that an array is
-read using the neighbourhood indices described by both $r$ and
-$r'$. It can be thought of as a conjunction of specifications. For
+read using the neighbourhood indices described by either $r$ or
+$r'$. It can be thought of as a disjunction of specifications. For
 example, the following gives the specification for the five-point
@@
 \begin{ExmVerbatim}
-!= stencil centered(depth=1, dim=1)
-         + centered(depth=1, dim=2) :: a
+!= stencil centered(depth=1, dim=1)*reflexive(dim=2)
+         + centered(depth=1, dim=2)*reflexive(dim=1) :: a
 

### 2016-07-04 paper.tex
fix to effective dims
correct effective dimensions def
Make consistent with new syn & sem. Better syntax highlighting.

@@
 \usepackage{multirow}
+\usepackage{minted}
+
+\fvset{
+  linenos=true,
+  fontsize=\footnotesize,
+  breaklines=true,
+  breakafter=+*,
+  xleftmargin=\parindent
+}
 
@@
 %
-\begin{ExmVerbatim}
+\begin{minted}{fortran}
 do iter = 0, itermax
@@
 end do
-\end{ExmVerbatim}
+\end{minted}
 %
@@
 %
-\begin{SpecVerbatim}
+\begin{minted}{fortran}
 !=  stencil centered(depth=1, dim=1) :: a
-\end{SpecVerbatim}
+\end{minted}
 %
@@
 \begin{figure}[t]
-\begin{ExmVerbatim}[firstnumber=20]
-du2dx = ((u(i,j)+u(i+1,j))*(u(i,j)+u(i+1,j))+
-    gamma*abs(u(i,j)+u(i+1,j))*(u(i,j)-u(i+1,j))-
-    (u(i-1,j)+u(i,j))*(u(i-1,j)+u(i,j))-
-    gamma*abs(u(i-1,j)+u(i,j))*(u(i-1,j)-u(i,j)))
+  \begin{minted}[firstnumber=20]{fortran}
+du2dx = ((u(i,j)+u(i+1,j))*(u(i,j)+u(i+1,j))+ &
+  gamma*abs(u(i,j)+u(i+1,j))*(u(i,j)-u(i+1,j))- &
+  (u(i-1,j)+u(i,j))*(u(i-1,j)+u(i,j))- &
+  gamma*abs(u(i-1,j)+u(i,j))*(u(i-1,j)-u(i,j))) &
   /(4.0*delx)
 
-duvdy = ((v(i,j)+v(i+1,j))*(u(i,j)+u(i,j+1))+
-   gamma*abs(v(i,j)+v(i+1,j))*(u(i,j)-u(i,j+1))-
-   (v(i,j-1)+v(i+1,j-1))*(u(i,j-1)+u(i,j))-
-   gamma*abs(v(i,j-1)+v(i+1,j-1))*(u(i,j-1)-u(i,j)))
+duvdy = ((v(i,j)+v(i+1,j))*(u(i,j)+u(i,j+1))+ &
+  gamma*abs(v(i,j)+v(i+1,j))*(u(i,j)-u(i,j+1))- &
+  (v(i,j-1)+v(i+1,j-1))*(u(i,j-1)+u(i,j))- &
+  gamma*abs(v(i,j-1)+v(i+1,j-1))*(u(i,j-1)-u(i,j))) &
   /(4.0*dely)
 
-laplu = (u(i+1,j)-2.0*u(i,j)+u(i-1,j))/delx/delx+
+laplu = (u(i+1,j)-2.0*u(i,j)+u(i-1,j))/delx/delx+ &
   (u(i,j+1)-2.0*u(i,j)+u(i,j-1))/dely/dely
-\end{ExmVerbatim}
+\end{minted}
 (a) Excerpt from Fortran code for a Navier-Stoke fluid simulator,
@@
 
-\begin{SpecVerbatim}[xleftmargin=0cm]
-!= stencil centered(depth=1, dim=1)*reflexive(dim=2)
-         + centered(depth=1, dim=2)*reflexive(dim=1) :: u
-
-!= stencil forward(depth=1, dim=1)
-         * backward(depth=1, dim=2) :: v
-\end{SpecVerbatim}
+\begin{minted}{fortran}
+!= stencil centered(depth=1,dim=1)*reflexive(dim=2) + centered(depth=1,dim=2)*reflexive(dim=1) :: u
+
+!= stencil forward(depth=1,dim=1) * backward(depth=1,dim=2) :: v
+\end{minted}
 (b) Inferred stencil specification from CamFort
@@
 %%
-\begin{ExmVerbatim}
+\begin{minted}{fortran}
 != stencil forward(depth=2, dim=1) :: a
 b(i, 0) = a(i, 0) + a(i+1, 0) + a(i+2, 0)
-\end{ExmVerbatim}
+\end{minted}
 %%
@@
 %
-\begin{ExmVerbatim}
+\begin{minted}{fortran}
 != stencil backward(depth=2, dim=1) :: a
 b(i) = a(i) + a(i-1) + a(i-2)
-\end{ExmVerbatim}
+\end{minted}
 %
@@
 %%
-\begin{ExmVerbatim}
+\begin{minted}{fortran}
 != stencil centered(depth=1, dim=1) :: a
 b(i) = ( a(i) + a(i-1) + a(i+1) ) / 3.0
-\end{ExmVerbatim}
+\end{minted}
 
@@
 example, the following gives the specification for the five-point
-stencil which is the sum of two \texttt{centered} regions in each
-dimension:
-%%
-\begin{ExmVerbatim}
-!= stencil centered(depth=1, dim=1)*reflexive(dim=2)
-         + centered(depth=1, dim=2)*reflexive(dim=1) :: a
+stencil which is the sum of two compound \texttt{reflexive} and
+\texttt{centered} regions in each dimension:
+%%
+\begin{minted}{fortran}
+!= stencil centered(depth=1, dim=1)*reflexive(dim=2) + centered(depth=1, dim=2)*reflexive(dim=1) :: a
 
@@
                    + a(i,j-1) + a(i,j+1)
-\end{ExmVerbatim}
-%%
+\end{minted}
+%%
+Here the left hand side of \texttt{+} says when the second dimension
+(induction variable $j$) is fixed at the origin, the first dimension
+(induction variable $i$) accesses immediate vicinity of the origin.
+The right hand side of \texttt{+} is similar but the dimensions are reversed.
+This reflects the symmetry under rotation in the fivepoint stencil.
+
 The product of two regions $r \texttt{*} r'$ specifies that the array
@@
 %%
-\begin{ExmVerbatim}
-!= stencil centered(depth=1, dim=1)
-         * centered(depth=1, dim=2) :: a
-
+\begin{minted}[breakindent=2.9em]{fortran}
 x = a(i, j)   + a(i-1, j)   + a(i+1, j)
@@
 z = a(i, j+1) + a(i-1, j+1) + a(i+1, j+1)
+
+!= stencil centered(depth=1, dim=1) * centered(depth=1, dim=2) :: a
 b(i, j) = (x + y + z) / 9.0
-\end{ExmVerbatim}
+\end{minted}
 %%
@@
 
+The region constructor \texttt{*} can also be interpreted as conjunction of
+two specifications since they need to hold simultaneously for their
+corresponding dimensions.
+
 As another example, the following code and specification
@@
 %%
-\begin{ExmVerbatim}
+\begin{minted}[breakindent=5.5em]{fortran}
 do j=0, jmax-1
     do i=0, imax-1
-\sten{      != region r1 = forward(depth=1, dim=1)}
-\sten{      != region r2 = forward(depth=1, dim=2)}
-\sten{      != stencil readOnce, r1*r2 :: a}
       x = a(i,j) - a(i+1,j+1)
       y = a(i+1,j) - a(i,j+1)
+      != stencil forward(depth=1,dim=1) * forward(depth=1,dim=2) :: a
       b(i,j) = sqrt(x*x+y*y)
@@
 end do
-\end{ExmVerbatim}
+\end{minted}
 
@@
 %%
-\begin{ExmVerbatim}
+\begin{minted}{fortran}
 != region r1 = centered(depth=1, dim=1)
 != region r2 = centered(depth=1, dim=2)
-!= stencil r1*r2 :: a
-\end{ExmVerbatim}
+!= region robertsCross = r1*r2
+!= stencil robertsCross :: a
+\end{minted}
+This is especially useful for common stencils such as Robert's cross
+as the region can be defined once and used by region name many times.
 %%
@@
 %%
-There are three kinds of modifier that can be given to spatial
-specifications: \texttt{readOnce}, \texttt{reflexive},
-\texttt{irreflexive}.
-
-The \texttt{readOnce} modifier enforces that no index appears
-more than once. For example, in all of the previous examples the
+There are two modifiers one used at specification the other at region
+level to provide better descriptions for stencil computations.
+
+At the specification level, the \texttt{readOnce} modifier enforces that no 
+index appears more than once. For example, in all of the previous examples the
 \texttt{readOnce} modifier could be added, \eg{}
 %
-\begin{ExmVerbatim}
+\begin{minted}{fortran}
 != stencil readOnce, backward(depth=2, dim=1) :: a
 b(i+1) = a(i) + a(i-1) + a(i-2)
-\end{ExmVerbatim}
+\end{minted}
 %
@@
 rule out any accidental repetition of array subscripts.
-The notion corresponds to that of \emph{linearity}, though we
-opt for the more informative and easily understood name
+The notion corresponds to that of \emph{linearity} in type systems,
+though we opt for the more informative and easily understood name
 \texttt{readOnce}. This modifier is optional, so it need not
@@
 
-The \texttt{irreflexive} modifier is parameterised by a
-comma-separated list of dimensions, indicating that for those
-dimensions there is no neighbour index with $0$ offset, \eg{},
-%
-\begin{ExmVerbatim}
-!= region r = centered(depth=1, dim=1)
-!= stencil irreflexive(dim=1), r :: a
+At the region level, \texttt{irreflexive} modifier qualifies base
+regions of \texttt{forward}, \texttt{backward}, and \texttt{centered}.
+It indicates that in the dimension of the region there is no neighbour index
+with $0$ offset, \eg{},
+%
+\begin{minted}{fortran}
+!= stencil centered(depth=1, dim=1, irreflexive) :: a
 b(i) = a(i+1) + a(i-1)
-\end{ExmVerbatim}
+\end{minted}
 %
@@
 %
-\begin{ExmVerbatim}
-!= region r = forward(depth=1, dim=1)
-!= stencil irreflexive(dim=1), r :: a
+\begin{minted}{fortran}
+!= stencil forward(depth=1, dim=1, irreflexive) :: a
 b(i) = a(i+1)
-\end{ExmVerbatim}
-
-The \term{reflexive} modifier is dual to \term{irreflexive},
-indicating dimensions in which there is a neighbour index with $0$
-offset. The region constants \term{forward}, \term{backward},
-\term{centered} implicitly include the \term{reflexive} modifier,
-but it is useful when there is a dimension in which there is reflexive
-indexing but not a region constant, \eg{},
-%%
-\begin{ExmVerbatim}
-!= region r = centered(depth=1, dim=1)
-!= stencil reflexive(dims=2), r :: a
-b(i,j) = a(i-1,j) + a(i+1,j) + a(i, j)
-\end{ExmVerbatim}
-%%
-Overlapping \term{irreflexive} and \term{reflexive} modifiers
-cannot occur together-- this is an invalid specification.
+\end{minted}
+
+Naturally, \term{irreflexive} qualifier cannot be used on \term{reflexive}
+regions.
 
@@
 %%
-\begin{ExmVerbatim}
-!= stencil atLeast, reflexive(dims=1)      :: a
+\begin{minted}{fortran}
+!= stencil atLeast, reflexive(dim=1)      :: a
 != stencil atMost, forward(depth=2, dim=1) :: a
 b(i) = a(i) + a(i+2)
-\end{ExmVerbatim}
+\end{minted}
 %%
@@
 %%
-\begin{ExmVerbatim}
-do i=2:n-1
+\begin{minted}{fortran}
+do i=2, n-1
   u(i) = r*v(i-1) + r2*v(i) + r*v(i+1)
 end do
-\end{ExmVerbatim}
+\end{minted}
 %%
@@
 %%
-\begin{ExmVerbatim}
-!= stencil centered(depth=1, dim=1)
-         + centered(depth=1, dim=2) :: v
-\end{ExmVerbatim}
+\begin{minted}{fortran}
+!= stencil centered(depth=1, dim=1) :: v
+\end{minted}
 %%
@@
 \equiv \; & \stenFwdS{n}{d} \; \texttt{+} \; \stenBwdS{n}{d} \\[1em]
+%
+(\texttt{+}\textsc{idem}) \;\; & S \; \texttt{+} \; S \; \equiv \; S \\[0.5em]
 %
@@
 
-\begin{definition}The product of two models $\otimes$ is defined:
-%%
+\begin{definition}The product of two models of dimensionality
+$n$ is given by the tensor $\otimes_n$ defined:
+%%
+\newcommand{\effdims}[2]{\mathit{dims}(#1)_{#2}}
 \begin{alignat*}{2}
-& dims(M_m,N_n) =
-  \bigg\{ x \; \bigg| \;
-    \parbox{4.1cm}{
-      $x \in \{1, \ldots, m\} \wedge M_x \neq \infty \; \vee \;$ \\
-      $x \in \{1, \ldots, n\} \wedge N_x \neq \infty$
-                  } \bigg\} \\
-& M_n \otimes N_n =
+& \effdims{M}{n} =
+\bigcup_{u \in M} \{x \mid x \in \{1,...,n\}, u_x \neq \infty\} \\
+& M \otimes_n N =
   \Bigg\{x \; \Bigg| \;
-    \parbox{4.8cm}{$i \in \{1, \ldots, 2^n\},$
-                   $(u, v) \in M_n \times N_n,$ \\
+    \parbox{5.6cm}{$i \in \{1, \ldots, 2^n\},$
+                   $(u, v) \in M \times N,$ \\
                    $x = (u \bowtie v)_i,$ \\
-                   $\forall j \in dims(M_n,N_n) (x_j \neq \infty) $
+                   $j \in (\effdims{M}{n} \cup \effdims{N}{n}),$ \\
+                   $x_j \neq \infty$
                   } \Bigg\}
@@
 %%
-\begin{ExmVerbatim}
+\begin{minted}{fortran}
 do i = 1, (n-1)
@@
 end do
-\end{ExmVerbatim}
+\end{minted}
 %%
@@
 For example (taken from E3MG):
-\begin{ExmVerbatim}
+\begin{minted}{fortran}
 !- stencil atLeast, readOnce, irreflexive(dims=2), (reflexive(dim=1)) :: sfda
@@
 ZZ1(I,J) = SFDT(I,J+8)
-\end{ExmVerbatim}
+\end{minted}
 
@@
 %
-\begin{ExmVerbatim}
+\begin{minted}{fortran}
 do j=jmin, jmax

### 2016-07-04 paper.tex
Change minted style
refactoring grammar to have irreflexive as attribute
spelling corrections

@@
 }
+
+\usemintedstyle{vs}
 
@@
 %
-Noticeable, the bounds of the stencil and the exact indexing pattern
+Notably, the bounds of the stencil and the exact indexing pattern
 are captured. Our approach aims to be much more abstract, to
@@
 line 6. Extending our approach to follow loop-carried dependencies is
-future work. An alternate approach for us would be to build ontop of
+future work. An alternate approach for us would be to build on top of
 the work of Kamil~\emph{et al.} by inferring our specifications from
@@
 
-There are many DSLs for stencil comptuation aimed at either
-performance or verification, or both. Many elimiante out-of-bounds
-errors, \eg{},~\cite{DBLP:journals/corr/abs-1109-0777}.
+There are many DSLs for stencil computation aimed at either
+performance or verification, or both. Many eliminate out-of-bounds
+errors, \eg{},~\citet{DBLP:journals/corr/abs-1109-0777}.
 

### 2016-07-04 paper.tex
related work: Blom 2014, Solar-Lezama 2007, and Tang 2011.

@@
 
-In an alternate approach,~\citet{abe2013model}
+% Abe, 2013, Model checking stencil computations
+% -- Unable to obtain copy
+% In an alternate approach,~\citet{Abe:2013:IPDPSW}
 
@@
 errors, \eg{},~\citet{DBLP:journals/corr/abs-1109-0777}.
+
+% GPGPU
+This work shares some similarities with efforts to verify kernels
+written for General-Purpose Graphics Processing Unit (GPGPU)
+programming, such as in \citet{Blom:2014:SoCP}. Stencils are a form of
+kernel, and GPGPU programming can be viewed as a massively parallel
+method of transforming a large matrix. However, that work focuses
+mainly on the synchronisation of such kernels, and the avoidance of
+data races, while we are interested in the correctness of stencils
+embedded within a more typical general-purpose programming language.
+
+% Sketching Stencils
+\citet{Solar-Lezama:2007:PLDI} give specifications of stencils using
+unoptimised ``reference'' stencils, coupled with partial
+implementations of stencils, which are then completed by a code
+generation tool. These kinds of specifications are just simple
+implementations, so this tool is useful for hand-written, optimised
+stencils. The primary purpose of this tool is optimisation rather than
+correctness, and the language of specification is much different and
+more elaborate than we envision for our system.
+
+% Pochoir
+\citet{Tang:2011:SPAA} defines a specification language for writing
+stencils embedded in C++ (with Cilk extensions) that are then compiled
+into fast parallel programs based on trapezoidal decompositions with
+hyperspace cuts. The Pochoir specification language is used for
+describing the computing kernel, boundary conditions and shape of the
+stencil. Programs are first compiled in C++ with a template library
+that checks whether the annotations are used correctly, and then a
+Pochoir preprocessor may be applied to generate the high-performance
+parallel version that takes advantage of the Cilk compiler.  Pochoir
+is aimed primarily at assisting programmers who may be reluctant to
+implement the tricky cache-oblivious ``hypertrapezoidal'' algorithms
+that are significantly faster than the typical looping implementations
+of stencil computations.
+
+% 3D stencil codes
+\citet{Zhang:2012:CGO}
+
+% Ypnos
+\citet{Orchard:2010:DAMP}
 

### 2016-07-04 paper.tex
typo
small steps in the evaluation

@@
 \hline
-Package       & \% potential & \% actual & Total & Total    \\
-              & stencils     & stencils  & exact  & bounded \\
+Package       & Potential & Actual & Exact & Inexact    \\
 \hline
-UM            & \umtickAssignPercent & \umtickAssignSuccessPercent & \umnumStencilLines & \umboundedBoth  \\
-E3MG          & \ethreemgeatickAssignPercent & \ethreemgeatickAssignSuccessPercent & \ethreemgeanumStencilLines & \ethreemgeaboundedBoth \\
-BLAS          & \blastickAssignPercent & \blastickAssignSuccessPercent & \blasnumStencilLines & \blasboundedBoth \\
-Hybrid4       & \hybridfourtickAssignPercent & \hybridfourtickAssignPercent & \hybridfournumStencilLines &  \hybridfourboundedBoth\\
-GEOS-Chem     & \geoschemtickAssignPercent & \geoschemtickAssignPercent & \geoschemnumStencilLines & \geoschemboundedBoth \\
-Navier        & \naviertickAssignPercent & \naviertickAssignPercent & \naviernumStencilLines & \navierboundedBoth \\
-CP            & \computationalphysicstwotickAssignPercent & \computationalphysicstwotickAssignSuccessPercent & \computationalphysicstwonumStencilLines & \computationalphysicstwoboundedBoth \\
+UM            & \umtickAssign\ (\umtickAssignPercent\%) & 
+                \umtickAssignSuccess\ (\umtickAssignSuccessPercent\%) & 
+                \umnumStencilLines & 
+                \umboundedBoth  \\
+E3MG          & \ethreemgeatickAssign\ (\ethreemgeatickAssignPercent\%) & 
+                \ethreemgeatickAssignSuccess\ (\ethreemgeatickAssignSuccessPercent\%) & 
+                \ethreemgeanumStencilLines & 
+                \ethreemgeaboundedBoth \\
+BLAS          & \blastickAssign\ (\blastickAssignPercent\%) & 
+                \blastickAssignSuccess\ (\blastickAssignSuccessPercent\%) & 
+                \blasnumStencilLines & 
+                \blasboundedBoth \\
+Hybrid4       & \hybridfourtickAssign\ (\hybridfourtickAssignPercent\%) & 
+                \hybridfourtickAssignSuccess\ (\hybridfourtickAssignSuccessPercent\%) & 
+                \hybridfournumStencilLines &  
+                \hybridfourboundedBoth\\
+GEOS-Chem     & \geoschemtickAssign\ (\geoschemtickAssignPercent\%) & 
+                \geoschemtickAssignSuccess\ (\geoschemtickAssignSuccessPercent\%) & 
+                \geoschemnumStencilLines & 
+                \geoschemboundedBoth \\
+Navier        & \naviertickAssign\ (\naviertickAssignPercent\%) & 
+                \naviertickAssignSuccess\ (\naviertickAssignPercent\%) & 
+                \naviernumStencilLines & 
+                \navierboundedBoth \\
+CP            & \computationalphysicstwotickAssign\ (\computationalphysicstwotickAssignPercent\%) & 
+                \computationalphysicstwotickAssignSuccess\ (\computationalphysicstwotickAssignSuccessPercent\%) & 
+                \computationalphysicstwonumStencilLines & 
+                \computationalphysicstwoboundedBoth \\
 \hline
-\textbf{Total}& \overalltickAssignPercent  & \overalltickAssignSuccessPercent & \overallnumStencilLines & \overallboundedBoth    \\
+\textbf{Total}& \overalltickAssign\ (\overalltickAssignPercent\%)  & 
+                \overalltickAssignSuccess\ (\overalltickAssignSuccessPercent\%) & 
+                \overallnumStencilLines & 
+                \overallboundedBoth    \\
 \hline
@@
 
-The first two columns of Figure \ref{fig:tickAssign} shows the percentage of lines which were
-assigned potential and actual stencils.
+The first two columns of Figure \ref{fig:tickAssign} shows the number of potential and actual stencils and their rate as a percentage of the total number of assignment statements in the code.
 \anote{Add explanation about why our results are low. Include an example from BLAS of a tickAssign'd statement that is not a stencil}
@@
 If the inference process cannot determine an exact stencil for a statement it will instead produce an upper (\term{atMost}) and lower (\term{atLeast}) bound.
-The numbers of exact and bounded stencils are shown in columns three and four of Figure \ref{fig:tickAssign}.
+The number of exact and inexact stencils are shown in columns three and four of Figure \ref{fig:tickAssign}.
 We can see that our stencil language is capable of precisely specifying the overwhelming majority of stencils.
-Most of the bounded stencils apply to some form of data copying.
+Most of the inexact stencils apply to some form of data copying.
 For example (taken from E3MG):
@@
 \end{minted}
+
+\subsection{Stencil categories}
+
+% examine the various types of stencil that we found
+
+\subsection{Detecting programming errors}
+
+% != stencil readOnce, (forward(depth=1, dim=3, irreflexive))*(backward(depth=1, dim=1))*(backward(depth=1, dim=2)) + (forward(depth=1, dim=3))*(backward(depth=1, dim=1))*(reflexive(dim=2)) + (forward(depth=1, dim=3))*(backward(depth=1, dim=1, irreflexive))*(backward(depth=1, dim=2, irreflexive)) + (forward(depth=1, dim=3))*(backward(depth=1, dim=2))*(reflexive(dim=1)) + (backward(depth=1, dim=1))*(backward(depth=1, dim=2, irreflexive))*(reflexive(dim=3)) + (backward(depth=1, dim=1, irreflexive))*(backward(depth=1, dim=2))*(reflexive(dim=3)) :: cx_rho_centre
+% /um/trunk/src/atmosphere/diffusion_and_filtering/turb_smagorinsky.F90
+% Tried introducing indexing errors replacing one of the indices of cx_rho_instance with -1,0,+1
+% 42 cases were caught
+% 6 cases were not caught
 

### 2016-07-04 paper.tex
add further equations and expanded the description; start on the model section

--- 2016-07-04 paper.tex
@@
 \definecolor{darkdarkgreen}{rgb}{0.0,0.35,0.0}
-
-\newcommand{\sten}[1]{\textcolor{darkdarkgreen}{\!#1}}
 
@@
 \newcommand{\stenRefl}[1]{\term{reflexive} \, (\term{dim=}#1)}
-\newcommand{\stenFwd}[2]{\term{forward} \, (\term{depth=}#1,
-  \term{dim=}#2)}
-\newcommand{\stenBwd}[2]{\term{backward} \, (\term{depth=}#1,
-  \term{dim=}#2)}
-\newcommand{\stenCen}[2]{\term{centered} \, (\term{depth=}#1,
-  \term{dim=}#2)}
+\newcommand{\stenFwd}[3]{\term{forward} \, (\term{depth=}#1,
+  \term{dim=}#2{#3})}
+\newcommand{\stenBwd}[3]{\term{backward} \, (\term{depth=}#1,
+  \term{dim=}#2{#3})}
+\newcommand{\stenCen}[3]{\term{centered} \, (\term{depth=}#1,
+  \term{dim=}#2{#3})}
+\newcommand{\irrefl}{\texttt{irreflexive}}
+
 \newcommand{\stenReflS}[1]{\term{refl} \, (\term{dim=}#1)}
@@
   \term{dim=}#2)}
+\newcommand{\irreflS}{\texttt{irrefl}}
+
+\newcommand{\stenFwdSR}[3]{\term{fwd} (\term{depth=}#1,
+  \term{dim=}#2, #3)}
+\newcommand{\stenBwdSR}[3]{\term{bwd} (\term{depth=}#1,
+  \term{dim=}#2, #3)}
+\newcommand{\stenCenSR}[3]{\term{cen} (\term{depth=}#1,
+  \term{dim=}#2, #3)}
+\newcommand{\stenReflSR}[1]{\term{refl} (\term{dim=}#1)}
 
@@
 \nonterm{specTop} ::= & [\nonterm{approxMod},] \; \nonterm{spec} \\
-\nonterm{spec} ::= & [\nonterm{modMult},] \; \nonterm{refRegion} \\
-\nonterm{refRegion} ::= & [\nonterm{modRefl},] \; \nonterm{region}\\
-\nonterm{modMult} ::= & \term{readOnce} \\
-\nonterm{modRefl} ::= & \term{irreflexive} \, (\term{dims=}\dimTy \{\term{,}
-          \dimTy\}) \\
+\nonterm{spec} ::= & [\nonterm{multMod},] \; \nonterm{region} \\
+\nonterm{multMod} ::= & \term{readOnce} \\
 \nonterm{approxMod} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
 \nonterm{region} ::= 
-        & \stenFwd{\mathbb{N}_{>0}}{\dimTy}  \\
-\mid \; & \stenBwd{\mathbb{N}_{>0}}{\dimTy} \\
-\mid \; & \stenCen{\mathbb{N}_{>0}}{\dimTy} \\
+        & \stenFwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]} \\
+\mid \; & \stenBwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]} \\
+\mid \; & \stenCen{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]} \\
 \mid \; & \stenRefl{\dimTy}  \\
@@
 \begin{array}{rll}
-(\textsc{F\; \texttt{+} \;F}) \;\; &
-\stenFwdS{n \, \textsf{max} \, m}{d} \\
- \equiv \; & \stenFwdS{n}{d} \; \texttt{+} \; \stenFwdS{m}{d} \\[1em]
-%
-(\textsc{B\; \texttt{+} \;B}) \;\; &
-\stenBwdS{n \, \textsf{max} \, m}{d} \\
- \equiv \; & \stenBwdS{n}{d} \; \texttt{+} \; \stenBwdS{m}{d} \\[1em]
-%
-(\textsc{C\; \texttt{+} \;C}) \;\; &
-\stenCenS{n \, \textsf{max} \, m}{d} \\
-\equiv \; & \stenCenS{n}{d} \; \texttt{+} \; \stenCenS{m}{d} \\[1em]
-%
-(\textsc{C\; \texttt{+} \;F}) \;\; & \stenCenS{n}{d} \\
-\textit{$m \leq n$} \; \equiv \; & \stenCenS{n}{d} \; \texttt{+} \;
-                      \stenFwdS{m}{d} \\[1em]
-%
-(\textsc{C\; \texttt{+} \;B}) \;\; &
+(\textsc{F\, \texttt{+} \,F}) \;\; &
+\stenFwdSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
+ \equiv \; & \stenFwdSR{n}{d}{r} \; \texttt{+} \; \stenFwdSR{m}{d}{s} \\[1em]
+%
+%
+(\textsc{B\, \texttt{+} \,B}) \;\; &
+\stenBwdSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
+ \equiv \; & \stenBwdS{n}{d}{r} \; \texttt{+} \; \stenBwdSR{m}{d}{s} \\[1em]
+%
+%
+(\textsc{C\, \texttt{+} \,C}) \;\; &
+\stenCenSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
+\equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \; \stenCenSR{m}{d}{s} \\[1em]
+%
+%
+(\textsc{C\, \texttt{+} \,F}) \;\; & \stenCenSR{n}{d}{r \vee s} \\
+\textit{$m \leq n$} \; \equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \;
+                      \stenFwdSR{m}{d}{s} \\[1em]
+%
+%
+(\textsc{C\, \texttt{+} \,B}) \;\; &
+\stenCenSR{n}{d}{r \vee s} \\
+\textit{$m \leq n$} \; \equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \;
+                      \stenBwdSR{m}{d}{s} \\[1em]
+%
+%
+(\textsc{B\, \texttt{+} \,F}) \;\; &
+\stenCenSR{n}{d}{r \vee s} \\
+\equiv \; & \stenFwdSR{n}{d}{r} \; \texttt{+} \; \stenBwdSR{n}{d}{s}
+  \\[1em]
+%
+%
+(\textsc{R\, \texttt{+} \,F}) \;\; &
+\stenFwdS{n}{d} \\
+\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenFwdSR{n}{d}{r} \\[1em]
+%
+%
+(\textsc{R\, \texttt{+} \,B}) \;\; &
+\stenBwdS{n}{d} \\
+\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenBwdSR{n}{d}{r} \\[1em]
+%
+%
+(\textsc{R\, \texttt{+} \,C}) \;\; &
 \stenCenS{n}{d} \\
-\textit{$m \leq n$} \; \equiv \; & \stenCenS{n}{d} \; \texttt{+} \;
-                      \stenBwdS{m}{d} \\[1em]
-(\textsc{B\; \texttt{+} \;F}) \;\; &
-\stenCenS{n}{d} \\
-\equiv \; & \stenFwdS{n}{d} \; \texttt{+} \; \stenBwdS{n}{d} \\[1em]
+\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenCenSR{n}{d}{r} \\[1em]
+%
+%
+(\textsc{irrefl}) \;\; & (R \, \term{*} \, S^\irreflS) \, \term{+} \,
+                         (R^\irreflS \, \term{*} \, S) \; 
+\equiv \; R \, \term{*} \, S \\[0.5em]
+%
 %
@@
                        \texttt{*} \; S) \; \texttt{+} \; (R
-                       \; \texttt{*} \; T) \\[0.5em]
-\end{array} \\ \\
-%
-\framebox{${\textit{specDec} :: \texttt{v}} \equiv
-{\textit{specDec'} :: \texttt{v}}$} \\[1em]
-\begin{array}{rl}
+                       \; \texttt{*} \; T)
+\end{array}
+%
+%\framebox{${\textit{specDec} :: \texttt{v}} \equiv
+%{\textit{specDec'} :: \texttt{v}}$} \\[1em]
+%\begin{array}{rl}
 % MUTUAL
@@
 %  \texttt{v} \\[0.5em]
-(\textsc{spatial}) \; &
-(\textit{region} \equiv \textit{region'}) \\
-\Rightarrow \; & (\texttt{stencil} \; [\textit{mod}] \; [\textit{approxMod},] \;
-\textit{region} \\
-& \!\!\equiv \; \texttt{stencil} \; [\textit{mod}] \;
-            [\textit{approxMod},] \;
-\textit{region}') \\
-\end{array}
+%\end{array}
 \end{array}
@@
 
-\paragraph{Equations}
 Figure~\ref{fig:equations} lists the equational theory for
-specifications. This is broken down into equations on regions only,
-given by the equivalence relation $\textit{region} \equiv \textit{region}'$ and
-on specification declarations assigned to variables via the equivalence relation
-$\textit{specDec} :: \texttt{v} \equiv
-\textit{specDec'} :: \texttt{v}$.
-
-To save space we use abbreviations
+specifications which is defined by the binary relation $\equiv$
+between region terms. Stencil declarations are considered
+equal when they have the same modifiers and $\equiv$ equates
+their regions, \ie{}:
+\begin{align*}
+& \textit{region} \equiv \textit{region'} \\
+& \quad \Rightarrow \;\; \texttt{stencil} \; [\textit{approxMod},] \; [\textit{multMod},] \;
+\textit{region} \\[-0.4em]
+& \quad\quad \;\; \equiv \;\texttt{stencil} \; [\textit{approxMod},] \;
+            [\textit{multMod},] \; \textit{region}'
+\end{align*}
+%
+To save space in the equations of Figure~\ref{fig:equations}, we use abbreviations
 \term{refl}, \term{fwd}, \term{bwd}, and \term{cen} for \term{reflexive},
-\term{forward}, \term{backward}, and \term{centered} regions respectively.
-The equations reveal something of their semantics.  The first six
+\term{forward}, \term{backward}, and \term{centered} regions
+respectively. Futhermore, the abbreviations include syntactic sugar
+for quantifying over the abseence or presence of the \irrefl{}
+attribute on region constants as a boolean. For example, 
+the syntactic sugar in our equations for \term{fwd} is defined:
+%%
+\begin{align*}
+& \stenFwdSR{n}{d}{r} 
+\; := \; \\[-0.4em]
+& \qquad \begin{cases}
+\stenFwd{n}{d}{, \irrefl} & \textit{if} \;\, \neg \, r \\
+\stenFwd{n}{d}{}  & \textit{if} \;\,  r
+\end{cases}
+\end{align*}
+%
+and similarly for \term{bwd} and \term{cen}.
+We also frequently abbreviate \irrefl{} to \irreflS{}.
+
+The equations reveal a partial semantics for stencil
+specifications, which we briefly summarise.
+
+\paragraph{Overlapping regions with +}
+The first nine
 equations (with labels of the form $\ldots \texttt{+} \ldots$)
-explain the behaviour of overlapping regions with sum
-\term{+}. The next six equations explain the algebraic behaviour of
-\term{+} and \term{*}.  Together, we see that \term{+} is an
-associative, commutative, idempotent binary operator that distributes
-with \term{*}, which is associative and commutative. This distribution
-of \term{*} over \term{+} is key as it is used internally to give a
-normal form for stencil specifications akin to Disjunctive Normal Form
-(DNF) (where \term{+} is taken as ``disjunction'' and \term{*} as the
-``conjunction''). We revisit this normalisation in Section~\ref{}.
+explain the notion of \emph{overlapping} regions with sum
+\term{+}. Note that in the first six, $r$ and $s$ range 
+over booleans denoting the absence (true) or presence (false)
+of \irrefl{} which are combined via disjunction. For example,
+the rule $(\textsc{B} \, \term{+} \,
+\textsc{F})$ explains that the addition of \term{backward}
+and \term{forward} regions in the same dimension and the same
+depth is equivalent to a \term{centered} region of that depth.
+If the \term{forward} part is irrefleixve (\ie{}, $r =
+\textit{false}$) but \term{backward} is not (\ie{}, $s =
+\textit{true}$) then their addition is equal to a \term{centered}
+region with no \irrefl{} modifier by the disjunction ($r \vee s =
+\textit{true}$), \ie{}:
+%%
+\begin{align*}
+& \stenCenS{n}{d}  \\[-0.2em]
+\equiv \;\; & \stenFwdSR{n}{d}{\irreflS} \; \texttt{+} \; \stenBwdS{n}{d}
+\end{align*}
+%%
+That is, the \term{fwd} region is irreflexive describing code which is missing
+a zero-offset neighbourhood index (\eg{}, missing $\texttt{a(i)}$), but the addition of
+the \term{bwd} has no irreflexive attribute and thus provides the
+missing zero offset index.
+
+The equation (\textsc{irrefl}) explains an annihilation of 
+\irrefl{} using a further syntactic sugar.
+We write $R^{\irreflS}$ to denote a region constant with an
+irreflexive attribute, \eg{}, 
+%%
+\begin{align*}
+\stenFwdS{n}{d}^{\irreflS} \, := \,
+\stenFwdSR{n}{d}{\irreflS}
+\end{align*}
+%% 
+Thus, (\textsc{irrefl}) explains the sum of a 
+product of two regions $R$ and $S$ with itself,
+where in each component of the sum one of $R$
+or $S$ has the irreflexive attribute. These cancel to give
+just the specification $R \, \term{*} \, S$.
+
+\paragraph{Interaction between \term{+} and \term{*}}
+
+The last six equations (from (\textsc{\term{+}idem}) inclusive), explain
+the algebraic behaviour of \term{+} and \term{*}.  Together, we see
+that \term{+} is an associative, commutative, idempotent binary
+operator that distributes with \term{*}, which is associative and
+commutative. This distribution of \term{*} over \term{+} is key as it
+is used internally to give a normal form for stencil specifications
+akin to Disjunctive Normal Form (DNF) (where \term{+} is taken as
+``disjunction'' and \term{*} as the ``conjunction''). We revisit this
+normalisation in Section~\ref{}.
 
@@
 
-Figure~\ref{fig:inequations}
+Figure~\ref{fig:inequations}\dnote{TODO: this might
+need to include notions of consistency}
 
@@
 
-We provide a semantic model of the stencil specification language as
+We define a semantic model of our stencil specification language as
 the basis for building correct implementations, showing the soundness
@@
   A specification of maximum dimensionality $n$, is modelled by a
-  set of column vectors of size $n$  ($n$-dimensional vectors /
+  set of vectors of size $n$ ($n$-dimensional vectors or 
   $n$-vectors for short)
   with values drawn from $\mathbb{Z} \cup \{\infty\}$, representing
-  relative indices. The $\infty$ element represents non-relative
-  indices (\eg{}, constants), which are seen as infinite with respect
+  relative indices. The element $\infty$ represents non-relative
+  indices, such as constants, which are seen as infinite with respect
   to the frame of reference of a relative offset from an induction
@@
 
-  For example, the array index $(i, j+1)$ corresponds to the vector
-  $\vtwoh{0}{1}$, whereas, the $(0, j)$ corresponds to the
-  vector $\vtwoh{\infty}{0}$ since $0$ is a non-relative index.
+  For example, the array index term $(i, j+1)$ corresponds to the vector
+  $\vtwoh{0}{1}$, whereas, the index term $(3, j)$ corresponds to the
+  vector $\vtwoh{\infty}{0}$ since $3$ is a non-relative index.
 \end{definition}
 
-Figure~\ref{fig:region-model} outlines the core rules for regions
-via the interpretation function $\interp{-}$.  We explain the
-intermediate notations and definitions:
+We define the interpretation function $\inter{-}$ overloaded on regions
+in Figure~\ref{fig:region-model} and on the top-level of 
+a specification \textit{specTop} in
+Figure~\ref{fig:spatial-model}. Some
+intermediate notations and definitions are used.
 
@@
 $\singleEntry{r}{n}$, is a vector
-where the $r^{th}$ entry is $1$ and all others are $\infty$, \eg{},
-$\singleEntry{1}{2} = \vtwoh{1}{\infty}$.
+where the $r^{th}$ entry is $1$ and all others are $\infty$. For
+example, 
+$\singleEntry{1}{2} = \vtwoh{\infty}{1}$ and $\singleEntry{2}{3} = 
+\vthreeh{1}{\infty}{1}$. 
 \end{definition}
@@
 $\zeroEntry{r}{n}$, is a vector where the $r^{th}$ entry is $0$ and all others
-are $\infty$, \eg{}, $\zeroEntry{1}{2} = \vtwoh{0}{\infty}$.
+are $\infty$, \eg{}, $\zeroEntry{1}{2} = \vtwoh{\infty}{0}$.
 \end{definition}
@@
 %  the multi-set $A$}
-\caption{Set model of regions}
+\caption{Model of regions, $\interp{\textit{region}}$}
 \label{fig:region-model}
@@
 \input{spatial-model.tex}
-\caption{Model of specifications}
+\caption{Model of specifications, $\interp{\textit{specTop}}$}
 \label{fig:spatial-model}
@@
 %% FWD
- \stenFwd{\vect{u}_i}{i}  & \vect{l}_i = 0 \; \wedge \; \vect{u}_i > 0 \\[0.7em]
- \texttt{irreflexive(dims=$i$)}, & \\
- \qquad \stenFwd{\vect{u}_i}{i}  & \vect{l}_i = 1 \; \wedge \; \vect{u}_i > 0 \\[0.7em]
+ \stenFwd{\vect{u}_i}{i}{}  & \vect{l}_i = 0 \; \wedge \; \vect{u}_i > 0 \\[0.7em]
+ \stenFwd{\vect{u}_i}{i}{, \irreflS}  & \vect{l}_i = 1 \; \wedge \; \vect{u}_i > 0 \\[0.7em]
 %% BWD
- \stenBwd{|\vect{l}_i|}{i}  & \vect{l}_i < 0 \; \wedge \; \vect{u}_i = 0 \\[0.7em]
- \texttt{irreflexive(dims=$i$)}, & \\
- \quad\;\; \stenBwd{|\vect{l}_i|}{i}  & \vect{l}_i < 0 \; \wedge \; \vect{u}_i = -1 \\[0.7em]
+ \stenBwd{|\vect{l}_i|}{i}{}  & \vect{l}_i < 0 \; \wedge \; \vect{u}_i = 0 \\[0.7em]
+\stenBwd{|\vect{l}_i|}{i}{, \irreflS} & \vect{l}_i < 0 \; \wedge \; \vect{u}_i = -1 \\[0.7em]
 %% CEN
- \stenCen{\vect{u}_i}{i}  &\vect{l}_i < 0 \; \wedge \; \vect{u}_i > 0
+ \stenCen{\vect{u}_i}{i}{}  &\vect{l}_i < 0 \; \wedge \; \vect{u}_i > 0
  \; \\
  & \qquad\quad \wedge \; |\vect{l}_i| = \vect{u}_i \\[0.7em]
- \stenBwd{|\vect{l}_i|}{i} & \vect{l}_i < 0 \; \wedge \; \vect{u}_i >
+ \stenBwd{|\vect{l}_i|}{i}{} & \vect{l}_i < 0 \; \wedge \; \vect{u}_i >
  0 \\
-\quad \texttt{+} \; \stenFwd{\vect{u}_i}{i}  & \qquad\quad \wedge \; |\vect{l}_i| \neq \vect{u}_i \\[0.7em]
+\quad \texttt{+} \; \stenFwd{\vect{u}_i}{i}{}  & \qquad\quad \wedge \; |\vect{l}_i| \neq \vect{u}_i \\[0.7em]
 % ABSOLUTE REP
- \texttt{irreflexive(dims=$i$)} & \vect{l}_i = \infty \; \vee \; \vect{u}_i
+ \epsilon & \vect{l}_i = \infty \; \vee \; \vect{u}_i
  = \infty \\[0.7em]
 % BOUND
- \texttt{atMost}, \stenFwd{\vect{u}_i}{i} & \vect{l}_i > 1 \\[0.7em]
- \texttt{atMost}, \stenBwd{|\vect{l}_i|}{i} \quad & \vect{l}_i < (-1) \\[0.7em]
+ \texttt{atMost}, \stenFwd{\vect{u}_i}{i}{} & \vect{l}_i > 1 \\[0.7em]
+ \texttt{atMost}, \stenBwd{|\vect{l}_i|}{i}{} \quad & \vect{l}_i < (-1) \\[0.7em]
   \end{cases}

### 2016-07-04 paper.tex
correcting the model and extending the proofs to new equations

@@
 \nonterm{approxMod} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
-\nonterm{region} ::= 
-        & \stenFwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]} \\
-\mid \; & \stenBwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]} \\
-\mid \; & \stenCen{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]} \\
-\mid \; & \stenRefl{\dimTy}  \\
-\mid \; & \nonterm{region} \; \term{+} \; \nonterm{region} \\
-\mid \; & \nonterm{region} \; \term{*} \; \nonterm{region} \\
-\mid \; & \nonterm{rvar}  \\[0.5em]
+\nonterm{region} ::= & \\
+\multicolumn{2}{l}{\qquad\qquad\;\;\; \stenFwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]}} \\
+\multicolumn{2}{l}{\qquad\qquad \mid \; \stenBwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]}} \\
+\multicolumn{2}{l}{\qquad\qquad \mid \; \stenCen{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]}} \\
+\multicolumn{2}{l}{\qquad\qquad \mid \; \stenRefl{\dimTy}}  \\
+\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{region} \; \term{+} \; \nonterm{region}} \\
+\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{region} \; \term{*} \; \nonterm{region}} \\
+\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{rvar}}  \\[0.5em]
 %\nonterm{temporal} ::= \; & \term{dependency} \; (v \; \{ , v \}) [, \texttt{mutual}]
@@
 \stenFwd{n}{d}{}  & \textit{if} \;\,  r
-\end{cases}
+\end{cases} \\
+\textit{or} \;\;
+& \stenFwdS{n}{d} := \stenFwd{n}{d}{}
 \end{align*}
@@
 
-We define the interpretation function $\inter{-}$ overloaded on regions
-in Figure~\ref{fig:region-model} and on the top-level of 
-a specification \textit{specTop} in
-Figure~\ref{fig:spatial-model}. Some
+We define the interpretation function $\interp{-}_n$ where $n$ is the
+maximum dimensionality of the specification being modelled, overloaded
+on regions in Figure~\ref{fig:region-model} and on the top-level of a
+specification \textit{specTop} in Figure~\ref{fig:spatial-model}. Some
 intermediate notations and definitions are used.
@@
 \end{definition}
+
+The first four equations of Figure~\ref{fig:region-model}
+give the meanings of the region constants as sets of indices.
 

### 2016-07-05 resultsToTex.py
updated with latest results

@@ -27,6 +27,8 @@ with open(sys.argv[1]) as f:
                 corpus = "overall"
             elif l[0:6] == "corpus":
                 corpus = sanitize(l.split(":")[1])
+            elif l.strip().startswith("numStencilSpecs"):
+                pass
             else:
                 (key,value) = map(lambda x:x.strip(),l.split(":"))
                 key = sanitize(key);

### 2016-07-05 paper.tex

@@
 
+%% MODEL
+\newcommand{\effdims}[2]{\mathit{dims}(#1)_{#2}}
+
 \include{results}
@@
 
-\title{Verification of Stencil Computations through Spatial Specifications}
-\authorinfo{Dominic Orchard \and Mistral Contrastin
-\and Matthew Danish \and Andrew Rice}{University of Cambridge}{}
+\title{Verification of Stencil Computations through Spatial
+  Specifications}
+\authorinfo{Blind}{}{}
+%\authorinfo{Dominic Orchard \and Mistral Contrastin
+%\and Matthew Danish \and Andrew Rice}{University of Cambridge}{}
 
@@
   procedures, our implementation, and our evaluation study. We provide
-  an implementation of stencil inference and validation in Camfort, an
-  open-source program analysis and refactoring tool.
+  an implementation of ontop of CamFort, an
+  open-source program analysis and refactoring tool for Fortran corpus.
 \end{abstract}
@@
 %%
-\newcommand{\effdims}[2]{\mathit{dims}(#1)_{#2}}
 \begin{alignat*}{2}
@@
 
-\onecolumn
-\appendix
-
-\input{proof.tex}
-
 \end{document}

### 2016-07-05 paper.tex
proof of IRREFL
progress on explaining the specification algebra used in the inference

@@
 & \effdims{M}{n} =
-\bigcup_{u \in M} \{x \mid x \in \{1,...,n\}, u_x \neq \infty\} \\
+\bigcup_{u \in M} \{x \mid x \in \{1,\ldots,n\}, u_x \neq \infty\} \\
 & M \otimes_n N =
@@
 space, the next step converts these into abstract syntax trees
-for the specification language.
-
-For each span $S = [\vect{l} \; \vect{u}]$,
-for each dimension, each pair of entries in the lower- and upper-bound
-vectors is mapped to a concrete \textit{spec}, as defined by the
-$\mathsf{conv}$ function:
-%%
-\begin{align*}
-& \mathsf{conv}_i\; [\vect{l} \; \vect{u}] = \\
-& \begin{cases}
+for the specification language. This conversion is defined in terms
+of an algebra on \textit{specTop} syntax which mirrors the shape
+of the syntax with additive and multiplicative binary operators
+$\oplus$ and $\odot$: 
+%%
+\begin{align*}
+\oplus & : \textit{specTop} \rightarrow \textit{specTop} \rightarrow
+  \textit{specTop} \\
+\odot & : \textit{specTop} \rightarrow \textit{specTop} \rightarrow
+        \textit{specTop} \\
+\mathsf{toSpec} & : (\mathbb{Z} \cup \{\infty\}) \rightarrow (\mathbb{Z} \cup
+  \{\infty\})  \rightarrow \textit{specTop}
+\end{align*}
+%%
+The operation $\mathsf{toSpec}$ takes a pair of lower and upper bounds
+values and produces a specification which contains a one-dimensional
+region. The
+
+$\oplus$ and $\odot$ respectively (defined in Definition~\ref{}
+and Definition~\ref{} below).
+
+The proceeds by converting each span
+into a specification by converting ea
+ is mapped to a specification,
+and then summing specifications. This relies on product and sum
+operations lifted to specification syntax (which is more than just
+building regions out of \term{*} and \term{+}).
+
+\begin{equation*}
+\sum\limits_{\vect{x} \in \textsf{regions}(M)}
+\prod\limits_{i \in \{1, \ldots, n\}} \, \mathsf{toSpec}(\vect{x}^L_i, \vect{x}^U_i)
+\end{equation*}
+
+For each span, and each dimension within it, each pair of lower and
+upper bounds is mapped into a specification. That is,
+for all $\vect{x}$ of size $n$, and for all $i \in \{1,\ldots,n\}$
+then $\mathsf{toSpec} (\vect{x}^L_i, \vect{x}^U_i)$ maps each
+pair of lower and upper bounds in each dimension to a piece of syntax 
+of \textit{specTop} syntax, defined:
+%%
+%\newcommand{\llower}{\vect{x}^L_i}
+%\newcommand{\lupper}{\vect{x}^U_i}
+\newcommand{\llower}{l}
+\newcommand{\lupper}{u}
+\begin{align*}
+&\mathsf{toSpec} \; (\llower, \lupper) = \\
+&\begin{cases}
 %% FWD
- \stenFwd{\vect{u}_i}{i}{}  & \vect{l}_i = 0 \; \wedge \; \vect{u}_i > 0 \\[0.7em]
- \stenFwd{\vect{u}_i}{i}{, \irreflS}  & \vect{l}_i = 1 \; \wedge \; \vect{u}_i > 0 \\[0.7em]
+ \stenFwdS{\lupper}{i}  & \llower = 0 \, \wedge \, \lupper > 0 \\[0.3em]
+ \stenFwdSR{\lupper}{i}{\irreflS}  & \llower = 1 \, \wedge \, \lupper > 0 \\[0.3em]
 %% BWD
- \stenBwd{|\vect{l}_i|}{i}{}  & \vect{l}_i < 0 \; \wedge \; \vect{u}_i = 0 \\[0.7em]
-\stenBwd{|\vect{l}_i|}{i}{, \irreflS} & \vect{l}_i < 0 \; \wedge \; \vect{u}_i = -1 \\[0.7em]
+ \stenBwdS{|\llower|}{i}  & \llower < 0 \, \wedge \, \lupper = 0 \\[0.3em]
+\stenBwdSR{|\llower|}{i}{\irreflS} & \llower < 0 \, \wedge \, \lupper = -1 \\[0.3em]
 %% CEN
- \stenCen{\vect{u}_i}{i}{}  &\vect{l}_i < 0 \; \wedge \; \vect{u}_i > 0
- \; \\
- & \qquad\quad \wedge \; |\vect{l}_i| = \vect{u}_i \\[0.7em]
- \stenBwd{|\vect{l}_i|}{i}{} & \vect{l}_i < 0 \; \wedge \; \vect{u}_i >
- 0 \\
-\quad \texttt{+} \; \stenFwd{\vect{u}_i}{i}{}  & \qquad\quad \wedge \; |\vect{l}_i| \neq \vect{u}_i \\[0.7em]
+ \stenCenS{\lupper}{i}  & \llower < 0 \, \wedge \, \lupper > 0
+ \, \wedge \, |\llower| = \lupper \\[0.3em]
+ \stenBwdS{|\llower|}{i} & \llower < 0 \, \wedge \, \lupper > 0  \, \wedge \, |\llower| \neq \lupper \\
+\quad \texttt{+} \, \stenFwdS{\lupper}{i}  & \qquad\quad \\[0.3em]
 % ABSOLUTE REP
- \epsilon & \vect{l}_i = \infty \; \vee \; \vect{u}_i
- = \infty \\[0.7em]
+ \epsilon & \llower = \infty \, \vee \, \lupper
+ = \infty \\[0.3em]
 % BOUND
- \texttt{atMost}, \stenFwd{\vect{u}_i}{i}{} & \vect{l}_i > 1 \\[0.7em]
- \texttt{atMost}, \stenBwd{|\vect{l}_i|}{i}{} \quad & \vect{l}_i < (-1) \\[0.7em]
+ \texttt{atMost}, \stenFwdS{\lupper}{i} & \llower > 1 \\[0.3em]
+ \texttt{atMost}, \stenBwdS{|\llower|}{i} \quad & \llower < (-1) \\[0.3em]
   \end{cases}
@@
 the span. If the lower bound is $1$, then the
-$0$-offset point is missing so the specification is given an
-\texttt{irreflexive} modifier. The second two cases are similar for
+$0$-offset point is missing so the region is given an
+\texttt{irreflexive} attribute. The second two cases are similar for
 \texttt{backward} but now the depth of the region is determined by the
-absolute of the lower bound. In the fifth-case, when the lower bound
+absolute of the lower bound. In the fifth-case, where the lower bound
 is negative and the upper bound is positive and they have the same
@@
 In the case where either bound is $\infty$, representing a non
-neighbourhood index, then a simple \texttt{irreflexive} specification
-is given with no regions, indicating that there is not even reflexive
+neighbourhood index, then an empty specification is given
+ with no regions, indicating that there is not even reflexive
 0-offset access in this dimension.
@@
 \begin{lemma}
-$\mathsf{conv}_i$ is total given the precondition that the lower
+$\mathsf{toSpec}_i$ is total (\ie{}, defined for all spans) given the precondition that the lower
 bound of the input span is less than or equal to its upper bound.
 \end{lemma}
+
+\begin{equation*}
+{\Large{\bullet}}_{i \in \{1, \ldots, n\}} \, \mathsf{toSpec}(\vect{x}^L_i, \vect{x}^U_i)
+\end{equation*}
 

### 2016-07-05 paper.tex

@@
 of the syntax with additive and multiplicative binary operators
-$\oplus$ and $\odot$: 
+$\oplus$ and $\odot$ and a constructor for \textit{specTop}:
 %%
@@
 The operation $\mathsf{toSpec}$ takes a pair of lower and upper bounds
-values and produces a specification which contains a one-dimensional
-region. The
-
-$\oplus$ and $\odot$ respectively (defined in Definition~\ref{}
-and Definition~\ref{} below).
-
-The proceeds by converting each span
-into a specification by converting ea
- is mapped to a specification,
-and then summing specifications. This relies on product and sum
-operations lifted to specification syntax (which is more than just
-building regions out of \term{*} and \term{+}).
-
+values and constructs a specification which contains a one-dimensional
+region. This step of the inference then proceeds by converting each
+span into a $n$-dimensional specification by converting each
+pair of bounds in a span and multiplying by $\otimes$
+then summing the specifications resulting specification by $\oplus$.
+That is: 
 \begin{equation*}
+\textit{spec}(M) = 
 \sum\limits_{\vect{x} \in \textsf{regions}(M)}
@@
 \end{lemma}
-
-\begin{equation*}
-{\Large{\bullet}}_{i \in \{1, \ldots, n\}} \, \mathsf{toSpec}(\vect{x}^L_i, \vect{x}^U_i)
-\end{equation*}
 

### 2016-07-05 paper.tex
bit more work on eval

--- 2016-07-05 paper.tex
@@
 The first two columns of Figure \ref{fig:tickAssign} shows the number of potential and actual stencils and their rate as a percentage of the total number of assignment statements in the code.
-\anote{Add explanation about why our results are low. Include an example from BLAS of a tickAssign'd statement that is not a stencil}
-\anote{Add discussion about the rate of stencils in the code. Comparison with rate of assignment statements would be nice}
-
-If the inference process cannot determine an exact stencil for a statement it will instead produce an upper (\term{atMost}) and lower (\term{atLeast}) bound.
+There are a variety of reasons why we might not infer an actual stencil from a potential one: 1) \textbf{Non-subset induction variables} occurs when the induction variables on the RHS are not a subset of the LHS.
+These cases are not stencils by our definition.
+The degenereate case of this is to have only constant indices on the LHS.
+We see lots of examples of this in loops as accumulators e.g.
+computing the sum over some other array; 2) \textbf{Derived induction variables} where the indexing variable (\mintinline{fortran}{x}) is derived from an induction variable \mintinline{fortran}{i} e.g.
+\mintinline{fortran}{x = len - i}; 3) \textbf{Inconsistent induction dimensions} occur when an induction variable is used to specify more than one array dimension on the RHS.
+These are common in matrix operations such as LU-decomposition which contains assignments such as \mintinline{fortran}{a(l) = a(l) - a(m) * b(l, m)}.
+
+ %% This example is from geos-chem/GeosRad/rrtmg_sw_spcvmc.F90:609
+ %%                      pnifddir(ikl) = pnifddir(ikl) + zincflx(iw)*ztdbt(jk)
+ %% ikl is a derived induction variable (ikl = rows - jk) - we can't deal with that at the moment
+
+  %% This example is from geos-chem/GTMM/sort_pick_veg.F90
+  %% DO k=1,n_veg
+  %%    DO j=2,n/n_veg
+  %%       a=arr(k,j)
+  %%       b=ind(k,j)
+  %%       DO i=j-1, 1, -1
+  %%          IF (arr(k,i) >=a) exit
+  %%          arr(k,i+1)=arr(k,i)
+  %%          ind(k,i+1)=ind(k,i)
+  %%       END DO
+  %%       ! NO STENCIL - inconsistentIV
+  %%       ind(k,i+1)=b
+  %%       arr(k,i+1)=a
+  %%    END DO
+  %% END DO
+  %% ind(k,j) flows into b but the LHS doesn't use j. This is a valid 'access' stencil but not a valid comonad stencil
+
+%% From geos-chem/GTMM/loadCASAinput.F90:241
+%%          ppt_mo(1,i)=sum(ppt(:,:,i))
+%% If you expand the loop for the sum
+%% do x=1,lenDim1
+%%  do y=1,lenDim2
+%%   acc += ppt(x,y,i)  ! can't be a stencil but could be an access pattern
+%%  enddo
+%% enddo
+%% ppt_mo(1,i) = acc  ! should be reflexive(dim=3) :: ppt
+%% This should be fixable before the paper deadline
+
+  %% From blas/ssyrk.f:337
+  %%            DO 210 J = 1,N
+  %%                 DO 200 I = 1,J
+  %%                     TEMP = ZERO
+  %%                     DO 190 L = 1,K
+  %%                         TEMP = TEMP + A(L,I)*A(L,J)
+  %% 190                 CONTINUE
+  %%                     IF (BETA.EQ.ZERO) THEN
+  %%                         C(I,J) = ALPHA*TEMP  ! No stencil
+  %%                     ELSE
+  %%                         C(I,J) = ALPHA*TEMP + BETA*C(I,J)
+  %%                     END IF
+  %% 200             CONTINUE
+  %% 210         CONTINUE
+  %% In this case A does flow into TEMP and thus into C but its a reduction. There's an 'access' stencil above TEMP but that's all
+
+   %% From blas/sspr2.f:231
+   %%                    DO 10 I = 1,J
+   %%                        AP(K) = AP(K) + X(I)*TEMP1 + Y(I)*TEMP2
+   %%                        K = K + 1
+   %% 10                 CONTINUE
+   %% I don't think we notice that K is an induction variable in these cases
+
+  %% From computational-physics-2/fortran90/millikan.f90:73
+  %% DO I = 1, N+1
+  %%   STMP = U(1,I)*U(1,I)    ! access, not a stencil
+  %%   S(1) = S(1) + STMP      ! access, not a stencil
+  %%   G(1) = G(1) + K(I)*STMP ! access, not a stencil
+  %%   U(1,N+2) = U(1,N+2) + U(1,I)*F(I)  ! access, not a stencil
+  %% END DO
+
+%% From UM: mym_solve_simeq_lud.F90:148
+%% DO m = 1, n - 1
+%%   DO l = m + 1, n
+%%     bvec(l) = bvec(l) - bvec(m) * amat(l, m)
+%%   END DO
+%% END DO
+%% We just don't deal with this - Lower upper decomposition solver
+
+If the statement meets the stencil critera but the inference process cannot determine an exact stencil for a statement it will instead produce an upper (\term{atMost}) and lower (\term{atLeast}) bound.
 The number of exact and inexact stencils are shown in columns three and four of Figure \ref{fig:tickAssign}.
@@
 \subsection{Stencil categories}
+
+\textbf{All reflexive} The vast majority (X\%) of stencils were reflexive in every dimension of the array.
+These most often correspond to applying point-wise transformations to data such as scaling e.g.
+\minted{fortran}{ZZ1(I,K) = .000041868*ZZ1(I,K)}.
 

### 2016-07-05 paper.tex
stencil categories

@@
 
-\textbf{All reflexive} The vast majority (X\%) of stencils were reflexive in every dimension of the array.
+We grouped stencils together into four classes based on similar levels of complexity. These four classes account for 99.7\% of all the stencils we found in the corpus.
+
+\textbf{All-reflexive} stencils are reflexive in every dimension of the array.
 These most often correspond to applying point-wise transformations to data such as scaling e.g.
-\minted{fortran}{ZZ1(I,K) = .000041868*ZZ1(I,K)}.
-
-% examine the various types of stencil that we found
+\mintinline{fortran}{ZZ1(I,K) = .000041868*ZZ1(I,K)}.
+Arrays with this class of stencil must be indexed by all the induction variables used in the LHS without any offsets.
+
+\textbf{Some-reflexive} stencils consist of only reflexive terms but leave some dimensions unspecified.
+Examples in this class include \mintinline{fortran}{leafpools(:,age_class)=leafpool(:,1)} (which contains an implicit loop), and \mintinline{fortran}{RVM(J,I2) = VRFO(ivar,J)} where \mintinline{fortran}{J} is the induction variable.
+Arrays with this class of stencil can be indexed by a subset of the LHS induction variables but again without any offsets.
+
+\textbf{Single-action} stencils consist of a single \term{backward},\term{forward}, or \term{centered} term and some number of reflexive terms.
+% /home/mrd45/src/corpus/geos-chem/GTMM/CasaRegridModule.F90:645
+\begin{minted}{fortran}
+do j=2,jm-ig
+   do i=1,im
+      != stencil readOnce, (backward(depth=1, dim=2))*(reflexive(dim=1)) :: dm, q
+      al(i,j) = 0.5e+0_fp*(q(i,j-1)+q(i,j)) + r3*(dm(i,j-1) - dm(i,j))
+   enddo
+enddo
+\end{minted}
+
+\textbf{Single-action-irreflexive} stencils consist of a single \term{backward},\term{forward}, or \term{centered} irreflexive term and some number of reflexive terms.
+% /home/mrd45/src/corpus/geos-chem/GeosRad/rrtmg_lw_rtrnmc.F90:355
+\begin{minted}{fortran}
+do lev = nlayers, 1, -1
+    ! ...
+    radld = radld - radld * (atrans(lev) + &
+       efclfrac(lev,igc) * (1. - atrans(lev))) + &
+       gassrc + cldfmc(igc,lev) * &
+       (bbdtot * atot(lev) - gassrc)
+    != stencil readOnce, (forward(depth=1, dim=2, irreflexive)) :: cldfmc
+    drad(lev-1) = drad(lev-1) + radld
+\end{minted}
+Notice in this example how the read from \mintinline{fortran}{cldfmc} flows through onto the line with the stencil.
 

### 2016-07-05 paper.tex
Consistency and checking nearly explained

@@
 \usepackage{minted}
+\usepackage{breqn}
 
@@
 
+\newcommand\mdef{\stackrel{\mathclap{\normalfont\mbox{def}}}{=}}
+
 \newcommand{\bn}{\mathbb{N}}
@@
 \newcommand{\vtwo}[2]{\setlength{\arraycolsep}{0em}
-\left[\begin{array}{l}$#1$\\$#2$\end{array}\right]}
+\left[\begin{array}{l}#1\\#2\end{array}\right]}
 \newcommand{\vthree}[3]{\setlength{\arraycolsep}{0em}
@@
 \nonterm{specification} ::= & \nonterm{regionDec} \mid \nonterm{specDec} \\
-\nonterm{specDec} ::= & \term{stencil} \; \nonterm{specTop} \;
+\nonterm{specDec} ::= & \term{stencil} \; \nonterm{spec} \;
                         \texttt{::} \; v
@@
 %\\[1em]
-\nonterm{specTop} ::= & [\nonterm{approxMod},] \; \nonterm{spec} \\
-\nonterm{spec} ::= & [\nonterm{multMod},] \; \nonterm{region} \\
-\nonterm{multMod} ::= & \term{readOnce} \\
-\nonterm{approxMod} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
+\nonterm{spec} ::= & [\nonterm{approx},] \; [\nonterm{mult},] \; \nonterm{region} \\
+\nonterm{mult} ::= & \term{readOnce} \\
+\nonterm{approx} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
 \nonterm{region} ::= & \\
@@
 & \textit{region} \equiv \textit{region'} \\
-& \quad \Rightarrow \;\; \texttt{stencil} \; [\textit{approxMod},] \; [\textit{multMod},] \;
+& \quad \Rightarrow \;\; \texttt{stencil} \; [\textit{approx},] \; [\textit{mult},] \;
 \textit{region} \\[-0.4em]
-& \quad\quad \;\; \equiv \;\texttt{stencil} \; [\textit{approxMod},] \;
-            [\textit{multMod},] \; \textit{region}'
+& \quad\quad \;\; \equiv \;\texttt{stencil} \; [\textit{approx},] \;
+            [\textit{mult},] \; \textit{region}'
 \end{align*}
@@
 The semantics of stencil specifications can be given a denotational
-model as sets of vectors, represent array indexing.
+model as sets of unifiers for array indexing.
 
@@
   A specification of maximum dimensionality $n$, is modelled by a
-  set of vectors of size $n$ ($n$-dimensional vectors or 
-  $n$-vectors for short)
-  with values drawn from $\mathbb{Z} \cup \{\infty\}$, representing
-  relative indices. The element $\infty$ represents non-relative
-  indices, such as constants, which are seen as infinite with respect
-  to the frame of reference of a relative offset from an induction
-  variable.
-
-  For example, the array index term $(i, j+1)$ corresponds to the vector
-  $\vtwoh{0}{1}$, whereas, the index term $(3, j)$ corresponds to the
-  vector $\vtwoh{\infty}{0}$ since $3$ is a non-relative index.
+  set of unifiers. Here, a unifier is represented as a vector of size $n$
+  whose values are drawn from $\mathbb{Z} \cup \{\infty\}$. The purpose
+  of the unifier is to accept or reject relativised indices of an array
+  subscript. Formal definition of unification in this context is given
+  later.
+
+  The element $\infty$ represents the general acceptor meaning it would unify
+  with any index in the same dimension including constants or any offset from
+  an induction variable.
 \end{definition}
+
+\begin{definition} Unification relation, $\sim$, establishes whether
+  a particular relativised subscript, $u$, of a group of subscripts involved
+  in a stencil computation conforms with a unifier, $v$, of the model for
+  the stencil computation.
+
+  Formally, the unification relation can be defined as follows:
+  $$u \sim_n v \triangleq
+  \{ (u,v) \; \mid \; \forall i \in \{ 1, \ldots, n \}, v_i = \infty \vee u_i = v_i \}$$
+\end{definition}
+
+For example, consider a unifier of the form $\vtwoh{\infty}{1}$.
+
+An array index term $(i, j+1)$ would corresponds to the vector
+$\vtwoh{0}{1}$ after relativisation. This would unify with the
+$\vtwoh{\infty}{1}$ as the first dimension of the unifier have a
+general acceptor and the second dimension matches the relative index.
+
+Another index term $(3, j)$ corresponds to the vector $\vtwoh{\infty}{0}$
+after relativisation since $3$ is a non-relative index. This would fail
+to unify. In the first dimension non-relative index would unify with
+the general acceptor but in the second dimension 1 wouldn't unify with 3.
 
@@
 on regions in Figure~\ref{fig:region-model} and on the top-level of a
-specification \textit{specTop} in Figure~\ref{fig:spatial-model}. Some
+specification \textit{spec} in Figure~\ref{fig:spatial-model}. Some
 intermediate notations and definitions are used.
@@
 \begin{definition}A \emph{single-entry vector} of size $n$, denoted
-$\singleEntry{r}{n}$, is a vector
-where the $r^{th}$ entry is $1$ and all others are $\infty$. For
-example, 
-$\singleEntry{1}{2} = \vtwoh{\infty}{1}$ and $\singleEntry{2}{3} = 
-\vthreeh{1}{\infty}{1}$. 
+$\singleEntry{r}{n}$, is a vector where the $r^{th}$ entry is $1$
+and all others are $\infty$, \eg{}, $\singleEntry{1}{2} =
+\vtwoh{1}{\infty}$ and $\singleEntry{2}{3} = \vthreeh{\infty}{1}{\infty}$.
+
+Here the meaning of a scaled single-entry vector is that at the
+dimension it is not $\infty$, it provides a unifier for consistent
+relativised array subscript indices. In particular the constrained
+requires a non-zero offset from the induction variable.
 \end{definition}
@@
 $\zeroEntry{r}{n}$, is a vector where the $r^{th}$ entry is $0$ and all others
-are $\infty$, \eg{}, $\zeroEntry{1}{2} = \vtwoh{\infty}{0}$.
+are $\infty$, \eg{}, $\zeroEntry{1}{2} = \vtwoh{0}{\infty}$.
+
+Similar to the single-entry vector, zero-entry vector also imposes
+a constaint on relativised indices. The only difference is the indexing
+expression crosses the origin with respect to the induction variable.
 \end{definition}
@@
 The first four equations of Figure~\ref{fig:region-model}
-give the meanings of the region constants as sets of indices.
+give the meanings of the region constants as sets of unifiers.
+The next two shows how regions are composed in the model.
 
@@
 \end{alignat*}
+In plain language, tensor product combines non-$\infty$ constraints
+of the unifiers in different dimensions into less accepting unifiers.
+This means an access pattern unifying with this unifier would require
+both models to be satisfied. Use of $dims$ gives the dimensions that
+constitutent models talk about and is used in the product to eliminate
+general unifier appearing in these dimensions. Allowing general unifier
+in these dimensions would prevent both models to be satisfied
+simultaneously.
+
 The $\bowtie$ operation, which we call \emph{pairwise permutation}
@@
 where $b$ is a logical matrix where $b_{i,j}$ is the $j$-th bit of the integer
-$i - 1$. Recall that the vectors are defined over the domain $\mathbb{Z} \cup
-\{\infty\}$. In pairwise permutation $\infty$
-
+$i - 1$.
 %%
@@
 \input{spatial-model.tex}
-\caption{Model of specifications, $\interp{\textit{specTop}}$}
+\caption{Model of specifications, $\interp{\textit{spec}}$}
 \label{fig:spatial-model}
-\end{figure}
-
-\begin{figure}
-\input{consistency.tex}
-\caption{Consistency function for a model and a set of indices}
-\label{fig:consistency}
 \end{figure}
@@
 
-\section{Inference, checking, and synthesis}
+\section{Inference and checking}
 \label{sec:analysis}
-
-\mnote{Few cases that we might like to consider (there are examples
-    of each in \#camfort-main Slack channel):
-  \begin{itemize}
-    \item Matrix operator overloading $a + b$, where both are arrays
-    \item Scalar operator overloading $a + b$, where a is an array b is a
-      scalar. $b$ is added to all.
-    \item Sliding. $a = b$ where the dimensions are $a(1:10)$, $b(2:11)$,
-      so the assignment causes shifting.
-  \end{itemize}
-}
 
@@
 
-\subsection{Inference\label{sec:inference}}
+\subsection{Inference\label{subsec:inference}}
 
@@
 
+\subsection{Checking\label{subsec:checking}}
+
+Verification occurs when a user supplied specification is compared to the
+actual access pattern. The mathematical model developed in
+Section~\ref{sec:semantics} is used this comparison.
+
+A specification is first synthesised into its model. Then access patterns
+are checked against the model for consistency as defined in
+Figure~\ref{fig:consistency}. The consistency predicate is defined recursively.
+It first checks multiplicity if $\textsf{once}$ is at the top level, \ie{}
+specification in question involves \texttt{readOnce} modifier, then array
+accesses are checked for uniqueness using $unique$ predicate and the problem
+is reduced to $\textsf{mult}$, which does not impose any extra constraints.
+Next, consistency predicate considers whether the stencil is an approximation.
+Semantically, a lower bound is achieved if every unifier in the model
+correspond to some relativised vector of indices as that means all cells
+described in the specification are part of the access pattern. Conversely,
+an upper bound means all observed access patterns have a unifier. Hence,
+the actual cells are a subset of the specified ones. In combination an
+exact specification, \ie{} one without any approximation modifiers, is
+one that is both a lower and an upper bound.
+
+\begin{figure}
+\input{consistency.tex}
+\caption{Consistency function of a model against stencil computation}
+\label{fig:consistency}
+\end{figure}
+
+We will consider both a consistent and an inconsistent example to show how
+checking works. The following is an example with a looser but still consistent
+version of the \textit{fivepoint} stencil specification inferred previously in
+the Subsection~\ref{subsec:inference}:
+%
+\begin{minted}[breakindent=8.1em]{fortran}
+!= region fivepoint = centered(depth=1,dim=1) + centered(depth=1,dim=2)
+!= stencil readOnce, fivepoint :: b
+a(i,j) = b(i, j-1) + b(i,j) + b(i,j+1) + b(i-1,j) + b(i+1,j)
+\end{minted}
+%
+\begin{enumerate}
+\item Generate model from the specification as described in
+  Section~\ref{sec:semantics}.
+%
+\begin{dmath*}
+\interp{\texttt{stencil readOnce, fivepoint}}_2 =
+  \textsf{once}(
+    \textsf{exact}(
+      \{ \vtwo{-1}{\infty},
+          \vtwo{0}{\infty},
+          \vtwo{1}{\infty},
+          \vtwo{\infty}{-1},
+          \vtwo{\infty}{0},
+          \vtwo{\infty}{1}\}))
+\end{dmath*}
+%
+\item Now $cons$ function is applied with this model along side with the
+  assignment on line three and array $b$. For brevity the assignment is
+  referred as $L3$ and the set of unifiers defined in the previous step as
+  $m$.
+
+  The predicate is reduced to
+  $cons(\textsf{once}(\textsf{exact}(m)),L3,b)$.
+%
+\item $cons$ predicate from the previous step evaluates to
+  $$unique(L3,b) \; \wedge \; cons(\textsf{mult}(\textsf{exact}(m)),L3,b).$$
+%
+  The former condition holds as no access pattern is repeated in $L3$ by simple
+  inspection.
+%
+\item We further expand the function to 
+  \begin{equation*}
+    cons(\textsf{mult}(\textsf{up}(m)),L3,b) \wedge 
+    cons(\textsf{mult}(\textsf{low}(m),L3,b).
+  \end{equation*}
+%
+\item To evaluate the two predicates above, we need relativised indices for the
+  array in question. This is obtained by $analyse$ function from
+  Subsection~\ref{subsec:inference}.
+
+  For the example above the relativised indices are as follows:
+%
+\begin{dmath*}
+analyse(L3,b) =
+  \{ \vtwo{0}{-1},
+     \vtwo{0}{0},
+     \vtwo{0}{1},
+     \vtwo{-1}{0},
+     \vtwo{1}{0}
+  \}
+\end{dmath*}
+%
+\item Below we present two match ups of relativised access patterns and unifiers
+  to show both the lower and the upper bound holds.
+
+  The unifications for the lower bound:
+  \begin{align*}
+     \vtwoh{0}{-1} & \sim_2 \vtwoh{0}{\infty} \\
+     \vtwoh{0}{0} & \sim_2 \vtwoh{0}{\infty} \\
+     \vtwoh{0}{1} & \sim_2 \vtwoh{0}{\infty} \\
+     \vtwoh{-1}{0} & \sim_2 \vtwoh{\infty}{0} \\
+     \vtwoh{1}{0} & \sim_2 \vtwoh{\infty}{0}
+  \end{align*}
+
+  The unifications for the upper bound:
+  \begin{align*}
+     \vtwoh{-1}{0} & \sim_2 \vtwoh{-1}{\infty} \\
+     \vtwoh{0}{0} & \sim_2 \vtwoh{0}{\infty} \\
+     \vtwoh{1}{0} & \sim_2 \vtwoh{1}{\infty}  \\
+     \vtwoh{0}{-1} & \sim_2 \vtwoh{\infty}{-1} \\
+     \vtwoh{0}{0} & \sim_2 \vtwoh{\infty}{0} \\
+     \vtwoh{0}{1} & \sim_2 \vtwoh{\infty}{1}
+  \end{align*}
+
+  Since both predicates are satisfied the array accesses specified on line three
+  are consistent with the \texttt{fivepoint} specification.
+\end{enumerate}
+
 \section{Evaluation}
@@
 are the inputs to Step 2 of the Inference process described
-Section~\ref{sec:inference}. Each potential statement for which we
+Section~\ref{subsec:inference}. Each potential statement for which we
 eventually generate some stencil specifications for is also counted as

### 2016-07-05 paper.tex
removing spectop references

@@
 (a) Excerpt from Fortran code for a Navier-Stoke fluid simulator,
-showing highly-detailed stencil computations. \\
-
-%\begin{SpecVerbatim}[xleftmargin=0.1cm]
-%20-24  u: centered depth=1 dim=1
-%6-30  u: centered depth=1 dim=2
-%      v: forward depth=1 dim=1, backward depth=1 dim=1
-%32-33  u: centered depth=1 dim=1,2
-%\end{SpecVerbatim}
-
+showing highly-detailed stencil computations. 
+\vspace{0.5em}
+\hrule
+%
 \begin{minted}{fortran}
@@
 \end{minted}
-(b) Inferred stencil specification from CamFort
-
+(b) Inferred stencil specification from CamFort.
+\vspace{0.5em}
+\hrule
+%
 \begin{center}
@@
 \end{center}
+\vspace{-0.3em}
 (c) Pictorial representation of the two stencil specifications.
-The horizontal dimensions is \text{dim=1} and the vertical is \texttt{dim=2}:
-\caption{Fragment of Navier-Stokes fluid simulator and its specification}
+The horizontal dimension is \texttt{dim=1} and the vertical is \texttt{dim=2}.
+\caption{Fragment of a Navier-Stokes fluid simulator and its
+  specification in our language}
 \label{ref:navier-stokes-fragment}
@@
 ``disjunction'' and \term{*} as the ``conjunction''). We revisit this
-normalisation in Section~\ref{}.
+normalisation in Section~\ref{sec:inf-step4}.
 
@@
 %%
-\paragraph{Step 1: Standard control and data-flow analyses}
+\subsubsection{Step 1: Standard control and data-flow analyses}
+\label{sec:inf-step1}
+
 The inference relies on some standard program analyses, computed
@@
 
-\paragraph{Step 2: Data-access analysis}
+\subsubsection{Step 2: Data-access analysis}
+\label{sec:inf-step2}
 
@@
 
-\paragraph{Step 3: Coalesce contiguous indices into regions}
+\subsubsection{Step 3: Coalesce contiguous indices into regions}
+\label{sec:inf-step3}
 
@@
 
-\paragraph{Step 4: Convert spans into region AST}
+\subsubsection{Step 4: Convert spans into specifications}
+\label{sec:inf-step4}
+
+\newcommand{\bplus}{\operatornamewithlimits{\term{+}}}
+\newcommand{\tySum}[1]{#1^{\term{+}}}
+\newcommand{\tyProd}[1]{#1^{\term{*}}}
+\newcommand{\specDNF}{\textbf{spec}}
 
@@
 for the specification language. This conversion is defined in terms
-of an algebra on \textit{specTop} syntax which mirrors the shape
-of the syntax with additive and multiplicative binary operators
-$\oplus$ and $\odot$ and a constructor for \textit{specTop}:
-%%
-\begin{align*}
-\oplus & : \textit{specTop} \rightarrow \textit{specTop} \rightarrow
-  \textit{specTop} \\
-\odot & : \textit{specTop} \rightarrow \textit{specTop} \rightarrow
-        \textit{specTop} \\
+of an algebra on \textit{spec} syntax in disjunctive-normal form
+(which we denote \specDNF{}). The algebra mirrors the shape
+of the region syntax with additive and multiplicative binary operators
+$\oplus$ and $\odot$ and a constructor for \specDNF{}:
+%%
+\begin{align*}
+\oplus & : \specDNF \rightarrow \specDNF \rightarrow
+  \specDNF \\
+\odot & : \specDNF \rightarrow \specDNF \rightarrow
+        \specDNF \\
 \mathsf{toSpec} & : (\mathbb{Z} \cup \{\infty\}) \rightarrow (\mathbb{Z} \cup
-  \{\infty\})  \rightarrow \textit{specTop}
-\end{align*}
-%%
-The operation $\mathsf{toSpec}$ takes a pair of lower and upper bounds
-values and constructs a specification which contains a one-dimensional
-region. This step of the inference then proceeds by converting each
-span into a $n$-dimensional specification by converting each
-pair of bounds in a span and multiplying by $\otimes$
-then summing the specifications resulting specification by $\oplus$.
-That is: 
+  \{\infty\})  \rightarrow \specDNF
+\end{align*}
+%%
+The $\mathsf{toSpec}$ operation takes a pair of lower and upper bounds
+values (one-dimensional) and constructs a specification (in the
+\textit{spec} syntax) which contains a one-dimensional
+region. An entire span is convereted then into a $n$-dimensional
+specification by applying $\mathsf{toSpec}$ to each
+pair of bounds in a span, per dimension, and multiplying these
+specifications by $\odot$. The resulting specifications for
+each span are then summed by by $\oplus$, that is:
 \begin{equation*}
@@
 \end{equation*}
-
-For each span, and each dimension within it, each pair of lower and
-upper bounds is mapped into a specification. That is,
-for all $\vect{x}$ of size $n$, and for all $i \in \{1,\ldots,n\}$
-then $\mathsf{toSpec} (\vect{x}^L_i, \vect{x}^U_i)$ maps each
-pair of lower and upper bounds in each dimension to a piece of syntax 
-of \textit{specTop} syntax, defined:
+(where $\oplus$ is the additive and $\odot$ is the multiplicative used
+within $\Sigma$ and $\Pi$ respectively here).
+
+\begin{definition}[DNF specifications]
+Let $r_1, ..., r_n$ range over elements of the \textit{region}
+syntax (see Figure~\ref{fig:syntax}, p.~\pageref{fig:syntax})
+The domain of \specDNF{} specifications
+in disjunctive normal form ranges over terms of the form:
+%%
+(r_1 \term{*} \ldots \term{*} r_n)\term{+} \ldots 
+\term{+} (r_1 \term{*} \ldots \term{*} r_n)
+%%
+\end{definition}
+
+\begin{definition} The $\mathsf{toSpec}$ constructor is defined:
 %%
@@
 \begin{align*}
-&\mathsf{toSpec} \; (\llower, \lupper) = \\
+&\mathsf{toSpec} \; \llower \; \lupper = \\
 &\begin{cases}
 %% FWD
- \stenFwdS{\lupper}{i}  & \llower = 0 \, \wedge \, \lupper > 0 \\[0.3em]
- \stenFwdSR{\lupper}{i}{\irreflS}  & \llower = 1 \, \wedge \, \lupper > 0 \\[0.3em]
+ \stenFwdS{\lupper}{i}  & \llower = 0 \, \wedge \, \lupper > 0 \\[0.2em]
+ \stenFwdSR{\lupper}{i}{\irreflS}  & \llower = 1 \, \wedge \, \lupper > 0 \\[0.2em]
 %% BWD
- \stenBwdS{|\llower|}{i}  & \llower < 0 \, \wedge \, \lupper = 0 \\[0.3em]
-\stenBwdSR{|\llower|}{i}{\irreflS} & \llower < 0 \, \wedge \, \lupper = -1 \\[0.3em]
+ \stenBwdS{|\llower|}{i}  & \llower < 0 \, \wedge \, \lupper = 0 \\[0.2em]
+\stenBwdSR{|\llower|}{i}{\irreflS} & \llower < 0 \, \wedge \, \lupper = -1 \\[0.2em]
 %% CEN
  \stenCenS{\lupper}{i}  & \llower < 0 \, \wedge \, \lupper > 0
- \, \wedge \, |\llower| = \lupper \\[0.3em]
+ \, \wedge \, |\llower| = \lupper \\[0.2em]
  \stenBwdS{|\llower|}{i} & \llower < 0 \, \wedge \, \lupper > 0  \, \wedge \, |\llower| \neq \lupper \\
-\quad \texttt{+} \, \stenFwdS{\lupper}{i}  & \qquad\quad \\[0.3em]
+\quad \texttt{+} \, \stenFwdS{\lupper}{i}  & \qquad\quad \\[0.2em]
+% BOUND
+ \texttt{atMost}, \stenFwdS{\lupper}{i} & \llower > 1 \\[0.2em]
+ \texttt{atMost}, \stenBwdS{|\llower|}{i} \quad & \llower < (-1)
+ \\[0.2em]
 % ABSOLUTE REP
- \epsilon & \llower = \infty \, \vee \, \lupper
- = \infty \\[0.3em]
-% BOUND
- \texttt{atMost}, \stenFwdS{\lupper}{i} & \llower > 1 \\[0.3em]
- \texttt{atMost}, \stenBwdS{|\llower|}{i} \quad & \llower < (-1) \\[0.3em]
+ \epsilon & \llower = \infty \, \vee \, \lupper = \infty
   \end{cases}
@@
 
-In the case where either bound is $\infty$, representing a non
-neighbourhood index, then an empty specification is given
- with no regions, indicating that there is not even reflexive
-0-offset access in this dimension.
-
-Finally in the last two cases, if the lower bound is greater than 1
+If the lower bound is greater than 1
 (which implies the upper bound is greater than 1 too) then
@@
 Thus, an upper bound is instead returned showing the maximum extent
-of the region. The last case is the dual, when the lower bound is less
+of the region. The penultimate case is the dual, when the lower bound is less
 than $-1$, giving an upper bound on a \texttt{backward} specification.
 
+Finally, in the case where either bound is $\infty$, representing a non
+neighbourhood index, then an empty specification is given
+ with no regions, indicating that there is not even reflexive
+0-offset access in this dimension.
+\end{definition}
+
 \begin{lemma}
-$\mathsf{toSpec}_i$ is total (\ie{}, defined for all spans) given the precondition that the lower
+$\mathsf{toSpec}$ is total (\ie{}, defined for all inputs)
+ given the precondition that the lower
 bound of the input span is less than or equal to its upper bound.
 \end{lemma}
+
+%\newcommand{\mkBig}[2]{\overset{\LARGE{\term{#1}}}{#2}}
+\begin{definition} The $\odot$ operation is defined:
+\begin{align*}
+\tySum{r_1} \odot \tySum{r_2} = 
+ \bplus_{(\tyProd{s_1}, \tyProd{s_2}) \in (\tySum{r_1} \times
+   \tySum{r_2})} (\tyProd{s_1} \; \term{*} \; \tyProd{s_2})
+%\begin{cases}%
+%\end{cases}
+\end{align*}
+\end{definition}
 
@@
 
+\textbf{All reflexive} The vast majority (X\%) of stencils were
+reflexive in every dimension of the array. These most often correspond
+to applying point-wise transformations to data such as scaling
+e.g. \begin{minted}{fortran}
+ZZ1(I,K) = .000041868*ZZ1(I,K)
+\end{minted}
+
 \textbf{All-reflexive} stencils are reflexive in every dimension of the array.

### 2016-07-06 paper.tex
working on the model

@@
 %% MODEL
-\newcommand{\effdims}[2]{\mathit{dims}(#1)_{#2}}
+\newcommand{\effdims}[2]{\mathit{constr}(#1)_{#2}}
 
@@
  \emph{neighbourhood subscript}.
+\label{def:neighbour}
 \end{definition}
@@
 
-\newcommand{\relix}{(\mathbb{Z}_\bot)^\mathbb{D}}
-
-We define a semantic model of our stencil specification language as
-the basis for building correct implementations, showing the soundness
-of the equational theory and approximations, and as a basis for the
-inference, checking, and program synthesis algorithms.
-The semantics of stencil specifications can be given a denotational
-model as sets of unifiers for array indexing.
-
-\begin{definition}
-  A specification of maximum dimensionality $n$, is modelled by a
-  set of unifiers. Here, a unifier is represented as a vector of size $n$
-  whose values are drawn from $\mathbb{Z} \cup \{\infty\}$. The purpose
-  of the unifier is to accept or reject relativised indices of an array
-  subscript. Formal definition of unification in this context is given
-  later.
-
-  The element $\infty$ represents the general acceptor meaning it would unify
-  with any index in the same dimension including constants or any offset from
-  an induction variable.
+\newcommand{\domainVal}{\mathbb{Z}_\infty}
+
+We define a semantic model for our stencil specification language.
+This model has a number of purposes: (1) it is used in the inference,
+checking, and program synthesis algorithms (introduced in
+Section~\ref{sec:analysis}); (2) gives a basis for the equational
+theory and approximations of the previous section; and (3) can be used
+to guide correct implementations. The model is a denotational
+semantics, in terms of sets of vectors which we call \emph{index
+  schemes}. We introduce first the notion of indexing schemes
+and \emph{unification} which gives the domain of our model
+and forms the basis of the rest of the algorithsm for our
+specificaiton. We then define the denotational model
+
+\subsection{Indexing schemes and unification}
+
+\begin{definition}[Index scheme]
+An \emph{index scheme} is a vector of size $n$ (an $n$-vector)
+with values drawn from $\domainVal = \mathbb{Z} \cup \{\infty\}$. 
+Values in $\mathbb{Z}$ represent offsets 
+of a neighbourhood index (indexing
+expressions of the form $i + a$ where $i$ is an induction variable
+and $a$ is a constant offset, Definition~\ref{def:neighbour},
+p.~\pageref{def:neighbour}). The additional value
+$\infty$ represents any indexing behaviour--- it is a 
+wildcard.
+Throughout $u, v, w$ will range over index schemes.
 \end{definition}
 
-\begin{definition} Unification relation, $\sim$, establishes whether
-  a particular relativised subscript, $u$, of a group of subscripts involved
-  in a stencil computation conforms with a unifier, $v$, of the model for
-  the stencil computation.
-
-  Formally, the unification relation can be defined as follows:
-  $$u \sim_n v \triangleq
-  \{ (u,v) \; \mid \; \forall i \in \{ 1, \ldots, n \}, v_i = \infty \vee u_i = v_i \}$$
+Indexing schemes give an abstract
+representation of index expressions in the source language.
+Our model represents the meaning of stencil specifications
+as indexing schemes, which explain the range of possible
+indexing behaviours allowed under the specification.
+The relationship between indexing schemes and source language
+terms is given precisely by the partial function, \textsf{schematic}, which
+maps syntactic array index terms to indexing schemes. This
+operation forms the basis of the analysis required for the inference
+and the checking procedures in Section~\ref{sec:analysis}.
+
+\begin{definition}[Schematic]
+The partial function $\textsf{schematic}$ maps 
+source-language indexing terms $(e_1, \ldots, e_n)$
+into $n$-dimensional indexing schemes 
+($\in \domainVal^n$), defined:
+%
+\begin{align*}
+\mathsf{schematic}(e_1, \ldots, e_n) & = 
+[\downarrow\!e_1 \ldots \downarrow\!e_n ] 
+\quad \textit{iff} \; \forall i . \downarrow\!e_i \neq \bot 
+\end{align*}
+%
+The $\downarrow$ operator is a partial\footnote{We represent
+  partiality explicitly in our types with the set $A + \bot$.} function 
+ $\downarrow : \textit{expr} \rightarrow (\domainVal + \bot)$
+which maps each individal indexing expression to a component of
+the indexing scheme, defined: 
+%%
+\begin{align*}
+\downarrow\!e
+ & =  \begin{cases}
+a & e \equiv i + a \; \wedge \; \mathsf{neigh}(i + a) \\
+-a & e \equiv i - a \; \wedge \; \mathsf{neigh}(i - a) \\
+\infty & \textit{$\mathsf{IV}(e) = \emptyset$} \\
+\bot   & \textit{otherwise}
+\end{cases}
+\end{align*}
+%%
+where $\mathsf{IV}(e)$ is the set of free induction variables
+appearing in an expression, and $\neigh{}$ classifies expressions which 
+are constant offsets from an induction variable
+(Defn.~\ref{def:neighbour}). In the third case, 
+if the expression is not a neighbourhood index but is not
+a relative index either (\ie{} is not defined in terms
+of an induction variable) then it is mapped to $\infty$, otherwise
+$\downarrow\!e$ is undefined. The $\mathsf{schematic}$ function
+is undefined if $\downarrow$ of any of its components $e_i$ is
+undefined.
+
+For example $\textsf{schematic}(i, j+1) = \vtwoh{0}{1}$ assuming
+$i$ and $j$ are induction variables and $\textsf{schematic}(i-1, 2, j)
+ = \vthreeh{-1}{\infty}{0}$. However, $\textsf{schematic}(i*2, j)$, for
+ example, is
+ undefined as its first index is a non-neighbour relative index, which 
+ is outside the scope of our specification language.
 \end{definition}
 
-For example, consider a unifier of the form $\vtwoh{\infty}{1}$.
-
-An array index term $(i, j+1)$ would corresponds to the vector
-$\vtwoh{0}{1}$ after relativisation. This would unify with the
-$\vtwoh{\infty}{1}$ as the first dimension of the unifier have a
-general acceptor and the second dimension matches the relative index.
-
-Another index term $(3, j)$ corresponds to the vector $\vtwoh{\infty}{0}$
-after relativisation since $3$ is a non-relative index. This would fail
-to unify. In the first dimension non-relative index would unify with
-the general acceptor but in the second dimension 1 wouldn't unify with 3.
-
-We define the interpretation function $\interp{-}_n$ where $n$ is the
-maximum dimensionality of the specification being modelled, overloaded
-on regions in Figure~\ref{fig:region-model} and on the top-level of a
-specification \textit{spec} in Figure~\ref{fig:spatial-model}. Some
-intermediate notations and definitions are used.
+\begin{definition}[Unification]
+Two schemes $u, v$ of dimensionality $n$
+are said to be \emph{unifiable}, written
+$u \sim_n v$, iff the following holds:
+%%
+\[
+u \sim_n v \; \Leftrightarrow  \;
+  \forall i\!\in\!\{ 1,.., n \} . \, (u_i = \infty
+\, \vee \, v_i = \infty \, \vee \, u_i = v_i)
+\]
+%%
+That is, two schemes are related by the binary relation
+$\sim_n$ over $(\mathbb{Z}_{\infty})^n$ if in each dimension
+either is $\infty$ or they are equal. 
+Thus, the element $\infty$ represents the ``general acceptor''
+ meaning it unifies with any index in the same dimension.
+
+The subscript on $\sim_n$ is sometimes dropped when the
+intended dimensionality is clear or inconsequential.
+\end{definition}
+
+%Unification relation, $\sim$, establishes whether
+%a particular relativised subscript, $u$, of a group of subscripts involved
+%in a stencil computation conforms with a unifier, $v$, of the model for
+%the stencil computation.
+
+\begin{example}
+Take three array indexing terms
+$(i, j+1)$, $(1, j+1)$ and $(3, j)$.
+As given by \textsf{schematic}, these have the 
+corresponding schemes: $\vtwoh{0}{1}$, $\vtwoh{\infty}{1}$
+and $\vtwoh{\infty}{0}$ respectively. The following
+relationship of unification (or lack of) then holds between
+them:
+%%
+\begin{equation*}
+\vtwoh{\infty}{0} \; \not\sim_2 \; \vtwoh{0}{1} \; \sim_2 \;
+\vtwoh{\infty}{1} \; \not\sim_2 \; \vtwoh{\infty}{0}
+\end{equation*}
+%%
+The middle pair unifies as $0$ and $\infty$ unify, 
+whereas the left and right pairs do not unify as
+$0 \not\sim_1 1$. 
+\end{example}
+
+Our terminology of \emph{indexing schemes} and
+\emph{unification} are by analogy
+with the \emph{type schemes} and \emph{type unification} of
+ML~\cite{milner1978theory}, though our notion is much more rigid.
+
+\subsection{Denotational model}
+
+The model is given by the interpretation function $\interp{-}_n$
+(where $n$ is the maximum dimensionality of the specification being
+modelled) mapping closed\footnote{That is, we assume that there are no
+  occurences of \textit{rvar} in a specification being modelled.  Any
+  \emph{open} specification containing region variables can be made
+  closed by straightforward syntactic subtitution with a (closed) 
+  \textit{region}.} specifications to sets of
+$n$-dimensional indexing schemes. The interperation is overloaded on
+\emph{regions} in Figure~\ref{fig:region-model} and on the top-level
+of a specification \textit{spec} in Figure~\ref{fig:spatial-model}.
+
+We first define some intermediate notations and definitions.
 
@@
 \vtwoh{1}{\infty}$ and $\singleEntry{2}{3} = \vthreeh{\infty}{1}{\infty}$.
-
-Here the meaning of a scaled single-entry vector is that at the
-dimension it is not $\infty$, it provides a unifier for consistent
-relativised array subscript indices. In particular the constrained
-requires a non-zero offset from the induction variable.
+Thus, a single-entry vector is an indexing scheme which unifies with
+all other schemes which have $1$ as their $r^{th}$ entry,
+describing a neighbour offset of $1$. 
+%DAO don't understand this sentence:
+%In particular the constrained
+%requires a non-zero offset from the induction variable.
 \end{definition}
@@
 
-Similar to the single-entry vector, zero-entry vector also imposes
-a constaint on relativised indices. The only difference is the indexing
-expression crosses the origin with respect to the induction variable.
+Similarly to the single-entry vector, zero-entry vectors 
+are also indexing schemes which unify with all other schemes
+as long as the $r^{th}$ entry is $0$, representing indices at
+the ``origin'' in dimension $r$. 
 \end{definition}
 
+\begin{figure}
+\input{region-model.tex}
+%\textit{$a^m \in A$ means there are $n$ copies of $a$ in
+%  the multi-set $A$}
+\caption{Model of regions, $\interp{\textit{region}}$}
+\label{fig:region-model}
+\end{figure}
+
+\noindent
 The first four equations of Figure~\ref{fig:region-model}
-give the meanings of the region constants as sets of unifiers.
-The next two shows how regions are composed in the model.
-
-\begin{definition}The product of two models of dimensionality
-$n$ is given by the tensor $\otimes_n$ defined:
+give the meanings of the region constants as sets of indexing
+schemes. The next two give the meaning of combining regions
+by $\term{+}$ and $\term{*}$.
+
+\begin{example}
+For a $2$-dimensional situation, 
+\begin{align*}
+& \interp{\stenCenS{2}{1}}_2 \\
+= \; & \{i\singleEntry{1}{2} \,|\, i \in \{\text{-}2,..., 2\}\!\setminus\!0 \}
+  \cup  \{ \zeroEntry{1}{2} \,|\, \textit{true} \} \\
+= \; & \{-2\singleEntry{1}{2}, \, -\singleEntry{1}{2}, \,
+  \singleEntry{1}{2}, \, 2\singleEntry{1}{2}\} \cup
+  \{\zeroEntry{1}{2}\} \\
+= \; & \{\vtwoh{-2}{\infty}, \vtwoh{-1}{\infty}, \vtwoh{0}{\infty}, 
+\vtwoh{1}{\infty}, \vtwoh{2}{\infty}\} 
+\end{align*}
+Note the absboring behaviour of $\infty$ with respect to
+multiplication. 
+\end{example}
+\noindent
+The model of $\term{+}$ is the straightforward union of 
+the models of subterms. The model of $\term{*}$ is uses
+an intermediate tensor $\otimes$ which requires some intermediate
+definitions.
+
+\begin{definition}[Constrained dimensions]
+Given an $n$-dimensional model $M$, the \emph{constrained dimensions}
+of that model is a subset of $\{1, \ldots, n\}$ giving
+the dimensions for which at least one scheme $u \in M$ has
+a non-$\infty$ value in that dimension. That is:
+%
+\begin{align*}
+\effdims{M}{n} =
+\bigcup_{u \in M} \{i \mid i \in \{1,\ldots,n\} \, \wedge \, u_i \neq
+  \infty\}
+\end{align*}
+For example, $\effdims{\{\vtwoh{0}{\infty},\vtwoh{\infty}{\infty}\}}{2}
+ = \{1\}$.
+\end{definition}
+
+\begin{definition}The tensor $\otimes_n$ of $n$-dimensional
+models is defined:
+%
+%\begin{align*}
+%\{\vtwoh{0}{0}, \vtwoh{1,1}\}
+% \otimes \{\vtwoh{0,\infty}\}
+%\end{align*}
+%
+%
 %%
 \begin{alignat*}{2}
-& \effdims{M}{n} =
-\bigcup_{u \in M} \{x \mid x \in \{1,\ldots,n\}, u_x \neq \infty\} \\
 & M \otimes_n N =
   \Bigg\{x \; \Bigg| \;
-    \parbox{5.6cm}{$i \in \{1, \ldots, 2^n\},$
-                   $(u, v) \in M \times N,$ \\
-                   $x = (u \bowtie v)_i,$ \\
-                   $j \in (\effdims{M}{n} \cup \effdims{N}{n}),$ \\
-                   $x_j \neq \infty$
+    \parbox{6cm}{$(u, v) \in M \times N $ \\[0.1em]
+                 $\;\; \wedge \;\, j \in (\effdims{M}{n} \cup
+                   \effdims{N}{n})$\\[0.1em]
+                  $\;\; \wedge \;\, i \in \{1, \ldots, 2^n\} \, \wedge
+                  \, x = (u \bowtie v)_i$ \\[0.1em]                   
+                   $\;\; \wedge \;\, x_j \neq \infty$
                   } \Bigg\} 
 \end{alignat*}
-In plain language, tensor product combines non-$\infty$ constraints
-of the unifiers in different dimensions into less accepting unifiers.
-This means an access pattern unifying with this unifier would require
-both models to be satisfied. Use of $dims$ gives the dimensions that
-constitutent models talk about and is used in the product to eliminate
-general unifier appearing in these dimensions. Allowing general unifier
-in these dimensions would prevent both models to be satisfied
-simultaneously.
+The intuition for the tensor is that it gives all possible
+ways of combining indexing schemes in two models
+such that for every scheme
+$u \in (M \otimes_n N)$ then $w \sim u$ for some $w$
+\emph{iff} $\;\exists v \in M . w \sim v$ and
+$\exists v' \in N . w \sim v'$. Thus it acts as a kind of intersection
+for indexing schemes. We unpack its definition in pieces.
+
+We start with the first guard taking the Cartesian product of
+the two models and quantifying over all pairings $(u, v)$. 
+The second guard says that $j$ is in the set of constrainted
+dimensions for $M$ and the constrained dimensions for $N$.
+That is, every $j$ corresponds to a dimension in which 
+either $M$ or $N$ has a scheme $u$ where $u_j \neq \infty$,
+and thus is constraining.
+
+%for all possible
+%pairs of schemes $(u, v)$ in the Cartesian product of the
+%two models $M \times N$, then
+
+% combines non-$\infty$ constraints
+%of the unifiers in different dimensions into less accepting unifiers.
+%This means an access pattern unifying with this unifier would require
+%both models to be satisfied. Use of $dims$ gives the dimensions that
+%constitutent models talk about and is used in the product to eliminate
+%general unifier appearing in these dimensions. Allowing general unifier
+%in these dimensions would prevent both models to be satisfied
+%simultaneously.
 
@@
 (described below), constructs a $2^n \times n$ matrix ($2^n$-vector of
-$n$-vectors). Using this, the $\otimes$ operation constructs the set of every
-pairwise permutation for every pair $(u, v)$ of $n$-vectors in the Cartesian
-product of $M \times N$.
+$n$-vectors), which is indexed by $i$ above
+and then $j$. Using this, the $\otimes$ operation constructs the set
+of schemes which are the 
+pairwise permutations of every pair $(u, v)$ of schemes in the Cartesian
+product of $M \times N$. These permuted schemes are only those
+which have non-$\infty$ components in the constrained dimensions
+given by each model.
 
@@
 %%
+\todo{Complete this}
 \end{definition}
@@
 \begin{figure}
-\input{region-model.tex}
-%\textit{$a^m \in A$ means there are $n$ copies of $a$ in
-%  the multi-set $A$}
-\caption{Model of regions, $\interp{\textit{region}}$}
-\label{fig:region-model}
-\end{figure}
-
-\begin{figure}
 \input{spatial-model.tex}
@@
 
-\begin{theorem}[Soundness]
+\paragraph{Model of \textit{spec}}
+
+\todo{Explain the injections and the domain here}
+
+\subsection{Soundness}
+
+Given the model defined in this section we can
+consider soundness for the equational theory of
+specifications and the notion of spatial approximation
+on specifications. The proofs are given in our supplement.
+
+\begin{theorem}[Soundness of equational theory]
 \[
@@
 
-\paragraph{Proof} (see Appendix)
-
-\begin{theorem}[Soundness]
+\begin{theorem}[Soundness of approximations]
 \[
@@
 \end{theorem}
-
-\paragraph{Proof} (see Appendix)
 
@@
 
+\newcommand{\spanOp}{\textsf{spans}}
 \paragraph{Span coalescing, formally}
@@
 A fixed point is then compute for the following
-procedure \textsf{regions} which, for per variable
+procedure \spanOp which, for per variable
 in the map, coalesces spans into contiguous regions in
 the $n$-dimensional space. That is, we compute
- $(\mu \; \textsf{regions}) N$ where \textsf{regions}
+ $(\mu \; \spanOp) N$ where \spanOp
 is defined as follows for each variable in the map:
@@
 \begin{align*}
-\textsf{regions}(M)
+\spanOp(M)
 = \{\stwo{-1}{0}{1}{0} \stwo{0}{-1}{0}{1}\}
@@
 \end{enumerate}
-For our example, applying \textsf{regions} again yields the same
+For our example, applying \spanOp again yields the same
 result, hence we have reached a fixed point.
@@
 \newcommand{\specDNF}{\textbf{spec}}
-
-Given the set $\textsf{regions}(M)$ of spans covering the indexing
-space, the next step converts these into abstract syntax trees
+\newcommand{\result}[1]{\lfloor{#1}\rceil}
+
+Next, the set of spans $\spanOp(M)$ covering the indexing
+space is converted into abstract syntax trees 
 for the specification language. This conversion is defined in terms
 of an algebra on \textit{spec} syntax in disjunctive-normal form
-(which we denote \specDNF{}). The algebra mirrors the shape
-of the region syntax with additive and multiplicative binary operators
-$\oplus$ and $\odot$ and a constructor for \specDNF{}:
+(which we denote \specDNF{}).
+
+\begin{align*}
+\result{\specDNF{}} = \specDNF + ((\specDNF{} + 1) \times
+  (\specDNF{} + 1))
+\end{align*}
+
+\begin{align*}
+\result{\specDNF{}} & ::= \textsf{exact} \, \specDNF{}
+\mid \textsf{bounds} ([\specDNF{}], [\specDNF{}]) \\
+\end{align*}
+
+The algebra mirrors the shape
+of the region combinators for addition and multiplication, 
+with operators $\oplus$ and $\odot$ and a constructor for \specDNF{}:
 %%
@@
 The $\mathsf{toSpec}$ operation takes a pair of lower and upper bounds
-values (one-dimensional) and constructs a specification (in the
-\textit{spec} syntax) which contains a one-dimensional
-region. An entire span is convereted then into a $n$-dimensional
+values (one-dimensional) and constructs a specification 
+which contains a one-dimensional region. 
+An entire span is converted into an $n$-dimensional
 specification by applying $\mathsf{toSpec}$ to each
-pair of bounds in a span, per dimension, and multiplying these
+pair of lower and upper bounds in a span, per dimension, and multiplying these
 specifications by $\odot$. The resulting specifications for
-each span are then summed by by $\oplus$, that is:
+each span are then summed by by $\oplus$, that is, this step is given by:
 \begin{equation*}
-\textit{spec}(M) = 
-\sum\limits_{\vect{x} \in \textsf{regions}(M)}
+\textsf{spec}(M) = 
+\sum\limits_{\vect{x} \in \spanOp(M)}
 \prod\limits_{i \in \{1, \ldots, n\}} \, \mathsf{toSpec}(\vect{x}^L_i, \vect{x}^U_i)
@@
 within $\Sigma$ and $\Pi$ respectively here).
-
-\begin{definition}[DNF specifications]
+We unpack the definitions of this algebra.
+
+\begin{definition}[Specifications in DNF]
 Let $r_1, ..., r_n$ range over elements of the \textit{region}
-syntax (see Figure~\ref{fig:syntax}, p.~\pageref{fig:syntax})
-The domain of \specDNF{} specifications
-in disjunctive normal form ranges over terms of the form:
-%%
-(r_1 \term{*} \ldots \term{*} r_n)\term{+} \ldots 
-\term{+} (r_1 \term{*} \ldots \term{*} r_n)
-%%
+syntax (see Figure~\ref{fig:syntax}, p.~\pageref{fig:syntax}).
+The set \specDNF{} is a subset of syntax trees \textit{spec}
+of specifications in disjunctive normal form over regions 
+where the disjunct is \term{+} (since its model is in terms of set
+union) and the conjunct is \term{*} (since its model is a form of tensor).
+That is elements of \specDNF{} are of the form:
+%%
+\[
+(r^1_{1} \; \term{*} \ldots \term{*} \; r^1_n)\; \term{+} \; \ldots 
+\term{+} \; (r^m_1 \, \term{*} \; \ldots \term{*} \; r^m_p)
+\]
+%%
+More formally, $\specDNF \subseteq \textit{spec}$, defined:
+%%
+\begin{align*}
+\specDNF{} & ::= [\textit{mult},] \;
+\tySum{\specDNF} \\
+\tySum{\specDNF} & ::= \tyProd{\specDNF} \; \term{+} \; \tySum{\specDNF} \mid
+  \tyProd{\specDNF} \\
+\tyProd{\specDNF} & ::= \textit{region} \; \term{*} \; \tyProd{\specDNF} \mid
+   \textit{region}
+\end{align*}
+%%
+In the following, $s, t$ range over elements of
+\specDNF{}, $\tySum{s}, \tySum{t}$ range over elements of 
+$\tySum{\specDNF}$ and $\tyProd{s}, \tyProd{t}$ range over elements
+of $\tyProd{\specDNF}$ (for example, $\tyProd{s} = r^1_{1} \; \term{*}
+\ldots \term{*} \; r^1_n$).
 \end{definition}
@@
 %\newcommand{\mkBig}[2]{\overset{\LARGE{\term{#1}}}{#2}}
-\begin{definition} The $\odot$ operation is defined:
-\begin{align*}
-\tySum{r_1} \odot \tySum{r_2} = 
- \bplus_{(\tyProd{s_1}, \tyProd{s_2}) \in (\tySum{r_1} \times
-   \tySum{r_2})} (\tyProd{s_1} \; \term{*} \; \tyProd{s_2})
+\begin{definition} The $\odot$ operation on \specDNF{} is defined
+\begin{align*}
+& ([\textit{approx}_1] \; [\textit{mult}_1] \; 
+\tySum{s})
+\odot
+([\textit{approx}_2] \; [\textit{mult}_2] \;
+\tySum{t}) = \\
+& \begin{cases}
+\tySum{s} \tySum{\odot} \tySum{t} & \textit{approx}_1 =
+\textit{approx_2} = \varepsilon
+\\
+\
+& \textit{approx}_1 = \texttt{atLeast}, \textit{approx}_2 = \varepsilon
+\end{cases}
+\\
+\end{align*}
+\begin{align*}
+s \tySum{\odot} t & =
+ \bplus_{(\tyProd{s_1}, \tyProd{s_2}) \in (s \times
+   t)} (\tyProd{s_1} \; \term{*} \; \tyProd{s_2})
 %\begin{cases}%

### 2016-07-06 paper.tex
consisntent of section 4 wrt section 3

@@
 wildcard.
-Throughout $u, v, w$ will range over index schemes.
+Throughout $u, v, w$ range over index schemes, and $u_i$ denotes
+the $i^{th}$ element of $u$. 
 \end{definition}
@@
  := \qquad\qquad & \textit{where} \; \neigh(\overline{e_1}) \wedge \arrayTy(v)  \\[0.3em]
-\multicolumn{2}{l}{\qquad \bigcup\{v' \mapsto \{\textsf{convert}(\bar{e})\} \mid v'(\bar{e}) \leftarrow \mathsf{flowsTo}(e_2),
+\multicolumn{2}{l}{\qquad \bigcup\{v' \mapsto \{\textsf{schematic}(\bar{e})\} \mid v'(\bar{e}) \leftarrow \mathsf{flowsTo}(e_2),
   \arrayTy(v')\}}
@@
 That is, we focus on assignments to an array subscript where each
-index indexing expression in $\bar{e}$ is a neighbourhood index.  For
+indexing expression in $\bar{e}$ is a neighbourhood index.  For
 all array subscripts that flow to the right-hand side of this
-statement, a finite map is constructed from each array variable
-in this flow set to a representation of the subscripts, computed
-with \textsf{convert}.
-
-The \textsf{convert} function maps subscripts to a column vector
-representation with values drawn from $\mathbb{Z}_\top = \mathbb{Z} \cup \{\top\}$
-where $\mathbb{Z}$ represents the offset of a neighbourhood index
-and $\top$ represents any non-neighbourhood subscripts, defined:
-%
-\begin{align*}
-& \textsf{convert}((e_1, \ldots, e_n)) = [\textsf{conv}(e_1), \ldots,
-  \textsf{conv}(e_n)] \\
-& \textsf{conv}(e) \begin{cases}
-\pm a & \neigh(e) \wedge \, e \equiv i \pm a \\
-\top & \textit{otherwise}
-\end{cases}
-\end{align*}
-%
+statement, a finite map is constructed, mapping each array variable
+in this flow a set of indexing schemes for its subscripts, computed
+with \textsf{schematic}.
+
 For our example, the \textsf{analyse} function matches on
-line 5, with the following set for $\textsf{flowTo}(a(i, j) + x +
-  y)$:
+line 5, with the following set for $\textsf{flowsTo}(\texttt{a(i, j) + x +
+  y})$:
 %
@@
 \begin{align*}
-\texttt{a} \mapsto \{\vtwo{-1}{0}, \vtwo{1}{0},
-          \vtwo{0}{-1}, \vtwo{0}{1}, \vtwo{0}{0}\}
+\texttt{a} \mapsto \{\vtwoh{-1}{0}, \vtwoh{1}{0},
+          \vtwoh{0}{-1}, \vtwoh{0}{1}, \vtwoh{0}{0}\}
 \end{align*}
@@
 each $v \in \mathsf{dom}(M)$, the algorithm then constructs a group
-of regions which cover all contiguous groups of relative indices
+of regions (called \emph{spans}) which cover all contiguous groups of relative indices
 (from $M(v)$) in the $n$-dimensional space.
-
-Informally, the procedure proceeds as follows. First, relative
-indices are turned into unit regions in $n$-dimensions. For
-example, for the five point stencil there are five $1 \times 1$ regions,
-called \emph{spans}, which are essentially $0$-dimensional:
+Informally, indexing schemes are converted to pairs $u \mapsto (u,
+u)$, defining the lower and upper bound of a unit span 
+in $n$-dimensions. For
+example, for the five point stencil there are five $1 \times 1$ spans.
 %
@@
 %
-The procedure is then iterated till a fixed point is reached. In this
-example this is reached in the first step. These final spans
-go on to form the basis of the spatial stencil specification, where
-each span is a region product \texttt{*} and multiple spans are
+This procedure is iterated till a fixed point is reached. In this
+example this is reached in the first step. The final spans
+are used to construct stencil specification, where
+each span is product \texttt{*} of regions and multiple spans are
 combined with the region sum \texttt{+}.
@@
 \newcommand{\spanOp}{\textsf{spans}}
-\paragraph{Span coalescing, formally}
 
 Let $\vect{x}, \vect{y}, \vect{z}$ range over the column vectors of size
-$n$ whose values are drawn from $\mathbb{Z}_\top$.
-We write $\vect{x}_i$ for the $i$-th value of vector $\vect{x}$ (1
-indexed) and $\vect{x}_{i:j}$ for the sub-vector from entry $i$ to
+$n$ whose values are drawn from $\domainVal{}$.
+We write $\vect{x}_{i:j}$ for the sub-vector from entry $i$ to
 entry $j$ (inclusive).
@@
   column) and the upper-bound vertex (second column). We use the
-  notation $[\vect{x}^L \; \vect{x}^U]$ for such a span to mark out
-  the lower and upper bound vectors.
+  notation $\vect{x} = [\vect{x}^L \; \vect{x}^U]$, where $\vect{x}^L$
+  and $\vect{x}^U$ give the lower and upper bounds of the span.
 \end{definition}
@@
 
-\textbf{The Unified Model}\footnote{\url{http://www.metoffice.gov.uk/research/modelling-systems/unified-model}} is a weather forecasting and climate modelling tool developed by the Met Office in the United Kingdom.
-It is used by research organisations and meteorological services around the world.
-We use the development branch (trunk) of the model.
+\textbf{The Unified
+  Model}\footnote{\url{http://www.metoffice.gov.uk/research/modelling-systems/unified-model}}
+is a weather forecasting and climate modelling tool developed by the
+Met Office in the United Kingdom. It is used by research organisations
+and meteorological services around the world. We use the development
+branch (trunk) of the model. The code based in closed source but
+institutional licenses are available for research purposes.
 The Met Office runs a comprehensive code quality system incorporating dedicated committers (we counted 11) for particular parts of the model.
@@
 
-\textbf{Navier} is an implementation of a Navier-Stokes fluid dynamics computation~\cite{griebel1997numerical}.
+\textbf{Navier} is a smaller projecter, giving a discrete
+approximation to the two-dimensional Navier-Stokes fluid equations,
+based on the book of~\citet{griebel1997numerical}.
 

### 2016-07-06 paper.tex
improving tensor description

@@
 The model of $\term{+}$ is the straightforward union of 
-the models of subterms. The model of $\term{*}$ is uses
-an intermediate tensor $\otimes$ which requires some intermediate
-definitions.
-
+the models of subterms. The model of $\term{*}$ is
+the tensor $\otimes$ of subterm models. The tensor 
+is non-trivial; we give some informal intutition first.
+
+\paragraph{$\otimes$ of models}
+
+The intuition behind $\otimes$ is that it takes
+all possible pairs of indexing schemes, treating them as the
+lower and upper bounds of an $n$-dimensional rectangle from which the 
+remaining vertices of the rectangle are generated (like a bounding
+box). \eg{}:
+%%
+\begin{align*}
+\{\vtwoh{1}{2}, \vtwoh{5}{6}\} \otimes \{\vtwoh{3}{4}\} = 
+\{&\vtwoh{1}{2}, \vtwoh{3}{2}, \vtwoh{1}{4}, \vtwoh{3}{4}, \\
+  & \vtwoh{3}{6}, \vtwoh{5}{4}, \vtwoh{5}{6}\}
+\end{align*}
+%%
+Internally $\otimes$ takes a Cartesian product of the indexing
+schemes (all pairs) and then combines each pair using the 
+\emph{pairwise permutation} written $\bowtie$ (defined later)
+which generates all possible interleavings of two vectors
+(\ie{}, all possible non-deterministic choices between $u_i$ and
+$v_i$ in each dimension). Some care is then taken over
+how this interacts with $\infty$: pariwse permtuation
+corresponds to non-determinstic choice, in each dimension $i$,
+between two values $u_i$ and $v_i$, but if either 
+is $\infty$ then the resultying pairwise permutation is 
+filtered out if the dimension $i$ has \emph{any other scheme
+in the model where $u_i \neq \infty$}, \eg{}:
+\begin{align*}
+\{\vtwoh{1}{\infty}\} \otimes \{\vtwoh{3}{4}\} = 
+\{\vtwoh{1}{4}, \vtwoh{3}{4}\}
+\end{align*}
+% 
+where $\vtwoh{1}{\infty}$ and $\vtwoh{3}{\infty}$ have been filtered.
+With respect to the notion of unification, the 
+%
+%
+% combines non-$\infty$ constraints
+%of the unifiers in different dimensions into less accepting unifiers.
+%This means an access pattern unifying with this unifier would require
+%both models to be satisfied. Use of $dims$ gives the dimensions that
+%constitutent models talk about and is used in the product to eliminate
+%general unifier appearing in these dimensions. Allowing general unifier
+%in these dimensions would prevent both models to be satisfied
+%simultaneously.
+%
+The notion of which dimensioons $i$ have that \emph{any scheme in the model
+has $u_i \neq \infty$} is represented by the \emph{constrained
+dimensions} intermediate function:
+%
 \begin{definition}[Constrained dimensions]
 Given an $n$-dimensional model $M$, the \emph{constrained dimensions}
-of that model is a subset of $\{1, \ldots, n\}$ giving
+of that model is a subset of $\{1, \ldots, n\}$ of 
 the dimensions for which at least one scheme $u \in M$ has
@@
 models is defined:
-%
-%\begin{align*}
-%\{\vtwoh{0}{0}, \vtwoh{1,1}\}
-% \otimes \{\vtwoh{0,\infty}\}
-%\end{align*}
-%
-%
-%%
 \begin{alignat*}{2}
@@
 \end{alignat*}
-The intuition for the tensor is that it gives all possible
-ways of combining indexing schemes in two models
-such that for every scheme
-$u \in (M \otimes_n N)$ then $w \sim u$ for some $w$
-\emph{iff} $\;\exists v \in M . w \sim v$ and
-$\exists v' \in N . w \sim v'$. Thus it acts as a kind of intersection
-for indexing schemes. We unpack its definition in pieces.
-
-We start with the first guard taking the Cartesian product of
-the two models and quantifying over all pairings $(u, v)$. 
-The second guard says that $j$ is in the set of constrainted
-dimensions for $M$ and the constrained dimensions for $N$.
+The first guard takes the Cartesian product of
+the two models and quantifies over all pairings $(u, v)$. 
+The second guard says that $j$ ranges of the constrained
+dimensions for $M$ unioned with the constrained dimensions for $N$.
 That is, every $j$ corresponds to a dimension in which 
 either $M$ or $N$ has a scheme $u$ where $u_j \neq \infty$,
-and thus is constraining.
-
-%for all possible
-%pairs of schemes $(u, v)$ in the Cartesian product of the
-%two models $M \times N$, then
-
-% combines non-$\infty$ constraints
-%of the unifiers in different dimensions into less accepting unifiers.
-%This means an access pattern unifying with this unifier would require
-%both models to be satisfied. Use of $dims$ gives the dimensions that
-%constitutent models talk about and is used in the product to eliminate
-%general unifier appearing in these dimensions. Allowing general unifier
-%in these dimensions would prevent both models to be satisfied
-%simultaneously.
-
-The $\bowtie$ operation, which we call \emph{pairwise permutation}
-(described below), constructs a $2^n \times n$ matrix ($2^n$-vector of
+and thus is constrained.
+
+The $\bowtie$ \emph{pairwise permutation} 
+constructs a $2^n \times n$ matrix ($2^n$-vector of
 $n$-vectors), which is indexed by $i$ above
@@
 which have non-$\infty$ components in the constrained dimensions
-given by each model.
+given by each model, given by the guard $x_j \neq \infty$.
 
@@
 %%
-\todo{Complete this}
 \end{definition}

### 2016-07-06 paper.tex
reformat
typo
tweak

@@
 and forms the basis of the rest of the algorithsm for our
-specificaiton. We then define the denotational model
+specificaiton. We then define the denotational model.
 
@@
 source-language indexing terms $(e_1, \ldots, e_n)$
-into $n$-dimensional indexing schemes 
-($\in \domainVal^n$), defined:
+into $n$-dimensional index schemes 
+$\in (\domainVal)^n$, defined:
 %
@@
 \vthreeh{0}{1}{2} \bowtie \vthreeh{3}{4}{5} =
-\setlength{\arraycolsep}{0.5em}
-\left[\begin{array}{ccc}
-0 & 1 & 2 \\
-0 & 1 & 5 \\
-0 & 4 & 2 \\
-0 & 4 & 5 \\
-3 & 1 & 2 \\
-3 & 1 & 5 \\
-3 & 4 & 2 \\
-3 & 4 & 5
+\left[
+\setlength{\arraycolsep}{0.42em}
+\begin{array}{cccccccc}
+0 & 0 & 0 & 0 & 3 & 3 & 3 & 3 \\ 
+1 & 1 & 4 & 4 & 1 & 1 & 4 & 4 \\
+2 & 5 & 2 & 5 & 2 & 5 & 2 & 5
 \end{array}
 \right]
+% \\
+%\setlength{\arraycolsep}{0.5em}
+%\left[\begin{array}{ccc}
+%0 & 1 & 2 \\
+%0 & 1 & 5 \\
+%0 & 4 & 2 \\
+%0 & 4 & 5 \\
+%3 & 1 & 2 \\
+%3 & 1 & 5 \\
+%3 & 4 & 2 \\
+%3 & 4 & 5
+%\end{array}
+%\right]
 \end{equation*}
@@
 %%
+\dnote{I didn't understand mistral's description of unification
+wrt to $\tensor$. I can't work it out either! Mistral have another
+go given my above explanation}
 \end{definition}

### 2016-07-06 paper.tex
expand on ML

@@
 with the \emph{type schemes} and \emph{type unification} of
-ML~\cite{milner1978theory}, though our notion is much more rigid.
+polymorphic types in ML~\cite{milner1978theory}. Our notion is
+much more rigid however, there is only one way to unify two
+index schemes in our model.
 

### 2016-07-06 paper.tex
Updated log/output files and removed need for corpussize.tex
Those variables are now updated by examining the output of analyse; latest
camfort includes a 'linesTotal' output as well.

--- 2016-07-06 paper.tex
@@
 \include{results}
-\include{corpussize}
 
@@
 and forms the basis of the rest of the algorithsm for our
-specificaiton. We then define the denotational model.
+specificaiton. We then define the denotational model
 
@@
 source-language indexing terms $(e_1, \ldots, e_n)$
-into $n$-dimensional index schemes 
-$\in (\domainVal)^n$, defined:
+into $n$-dimensional indexing schemes 
+($\in \domainVal^n$), defined:
 %

### 2016-07-06 resultsToTex.py

@@ -13,11 +13,6 @@ def sanitize(s):
 
 files = {}
 
-with open(sys.argv[2]) as f:
-    for l in f:
-        (k,v) = re.match("\\\\newcommand{\\\\(.*)}{(.*)}",l.strip()).groups()
-        files[k] = v
-
 corpora = defaultdict(lambda:{})
 keys = {}
 with open(sys.argv[1]) as f:
@@ -37,13 +32,20 @@ with open(sys.argv[1]) as f:
                 print "\\newcommand{\\%s}{%s}" % (corpus+key,value)
                 files[sanitize(corpus+key)] = value
 
+for corpus in corpora.keys():
+    numFiles = int(files.get(corpus+"parseOk", 0)) + int(files.get(corpus+"lexOrParseFailed", 0))
+    print "\\newcommand{\\%s}{%s}" % (corpus+"Files",numFiles)
+    numLoC = files[corpus+"linesTotal"]
+    print "\\newcommand{\\%s}{%s}" % (corpus+"LoC",numLoC)
+
+
 (totalTickAssign, totalTickAssignSuccess, totalLoC) = (0,0,0)
 for corpus in corpora.keys():
     if corpus != "overall" and corpus != "ethreemgmodernise":
         for key in keys.keys():
             if not corpora[corpus].has_key(key):
                 print "\\newcommand{\\%s}{0}" % (corpus+key)
-        loc = files[corpus+"LoC"]
+        loc = files[corpus+"linesTotal"]
         tickAssign = files[corpus+"tickAssign"]
         tickAssignSuccess = files[corpus+"tickAssignSuccess"]
         totalTickAssign += int(tickAssign)
@@ -54,4 +56,3 @@ for corpus in corpora.keys():
 
 print "\\newcommand{\\%s}{%.2g}" % ("overalltickAssignPercent", float(totalTickAssign) / float(totalLoC) * 100)
 print "\\newcommand{\\%s}{%.2g}" % ("overalltickAssignSuccessPercent", float(totalTickAssignSuccess) / float(totalLoC) * 100)
-

### 2016-07-06 paper.tex
use siunitx to pretty print numbers in the evaluation
also change figure environment -> table environment for the evaluation table

--- 2016-07-06 paper.tex
@@
 \usepackage{breqn}
+\usepackage{siunitx} % For pretty-printing numeric values and SI units
+                     % of measure. e.g., the tabular column type S is
+                     % used to print nice-looking tables of numbers.
+\sisetup{group-separator={,},group-minimum-digits={3},output-decimal-marker={.}}
 
@@
 \subsection{Software Corpus}
-\begin{figure}
-\begin{tabular}{|l|l|l|l|}
+\begin{table}
+\centering
+\sisetup{
+  table-format = 7,
+  table-number-alignment = right
+}
+\begin{tabular}{|l|S|S|S|}
 \hline
-Package       & Files & Lines of code & Files analysed \\
+Package       & {Files} & {Lines of code} & {Files analysed} \\ % braces prevent headers from being formatted as numbers
 \hline
@@
 \end{tabular}
-\caption{Summary of software packages used for evaluation\label{fig:corpus}}
-\end{figure}
-
-Figure \ref{fig:corpus} shows summary statistics of the software packages used in our evaluation, all of which are written in Fortran 90 or Fortran 77.
-In total we analysed 1.1 million lines of code from 7 packages.
-The ``Files analysed'' column shows the number of files in each corpus that we were able to analyse with Camfort.
-The most common reasons for Camfort rejecting a file were either use of a C-preprocessor, or illegal use of language features from a more modern Fortran variant.
+\caption{Summary of software packages used for evaluation\label{tab:corpus}}
+\end{table}
+
+Table~\ref{tab:corpus} shows summary statistics of the software
+packages used in our evaluation, all of which are written in Fortran
+90 or Fortran 77. In total we analysed 1.1 million lines of code from
+7 packages. The ``Files analysed'' column shows the number of files in
+each corpus that we were able to analyse with Camfort. The most common
+reasons for Camfort rejecting a file were either use of a
+C-preprocessor, or illegal use of language features from a more modern
+Fortran variant.
 

### 2016-07-06 paper.tex
Fix minor latex mistakes
figure env -> table env again

@@
 \hline
-Package       & {Files} & {Lines of code} & {Files analysed} \\ % braces prevent headers from being formatted as numbers
+Package   & {Files} & {Files analysed} & {Lines of code}  \\ % braces prevent headers from being formatted as numbers
 \hline
-UM            & \umFiles & \umLoC & \umparseOk  \\
-E3MG          & \ethreemgeaFiles   & \ethreemgeaLoC & \ethreemgeaparseOk  \\
-BLAS          & \blasFiles   & \blasLoC & \blasparseOk        \\
-Hybrid4       & \hybridfourFiles    & \hybridfourLoC & \hybridfourparseOk         \\
-GEOS-Chem     & \geoschemFiles   & \geoschemLoC & \geoschemparseOk       \\
-Navier        & \navierFiles     & \navierLoC & \navierparseOk           \\
-CP            & \computationalphysicstwoFiles    & \computationalphysicstwoLoC & \computationalphysicstwoparseOk           \\
+UM        & \umFiles                       & \umparseOk                      & \umLoC                     \\
+E3MG      & \ethreemgeaFiles               & \ethreemgeaparseOk              & \ethreemgeaLoC             \\
+BLAS      & \blasFiles                     & \blasparseOk                    & \blasLoC                   \\
+Hybrid4   & \hybridfourFiles               & \hybridfourparseOk              & \hybridfourLoC             \\
+GEOS-Chem & \geoschemFiles                 & \geoschemparseOk                & \geoschemLoC               \\
+Navier    & \navierFiles                   & \navierparseOk                  & \navierLoC                 \\
+CP        & \computationalphysicstwoFiles  & \computationalphysicstwoparseOk & \computationalphysicstwoLoC\\
 \hline
-\textbf{Total}& \overallFiles  & \overallLoC   & \overallparseOk  \\
+\textbf{Total} & \overallFiles  & \overallparseOk & \overallLoC  \\
 \hline
@@
 
-\begin{figure}
+\begin{table}
 \begin{tabular}{|l|l|l|l|l|}
 \hline
-Package       & Potential & Actual & Exact & Inexact    \\
+{Package}     & {Potential} & {Actual} & {Exact} & {Inexact}    \\
 \hline
@@
 \end{tabular}
-\caption{Stencil assignment rates\label{fig:tickAssign}}
-\end{figure}
-
-The first two columns of Figure \ref{fig:tickAssign} shows the number of potential and actual stencils and their rate as a percentage of the total number of assignment statements in the code.
+\caption{Stencil assignment rates\label{tab:tickAssign}}
+\end{table}
+
+The first two columns of Table~\ref{tab:tickAssign} shows the number of potential and actual stencils and their rate as a percentage of the total number of assignment statements in the code.
 There are a variety of reasons why we might not infer an actual stencil from a potential one: 1) \textbf{Non-subset induction variables} occurs when the induction variables on the RHS are not a subset of the LHS.

### 2016-07-06 paper.tex
update additional ref to tab:tickAssign

@@
 
+\newcommand{\todo}[1]{\textcolor{blue}{Dom: #1}}
 \newcommand{\dnote}[1]{\textcolor{darkpurple}{Dom: #1}}
@@
 \dnote{I didn't understand mistral's description of unification
-wrt to $\tensor$. I can't work it out either! Mistral have another
+wrt to tensor. I can't work it out either! Mistral have another
 go given my above explanation}
@@
 If the statement meets the stencil critera but the inference process cannot determine an exact stencil for a statement it will instead produce an upper (\term{atMost}) and lower (\term{atLeast}) bound.
-The number of exact and inexact stencils are shown in columns three and four of Figure \ref{fig:tickAssign}.
+The number of exact and inexact stencils are shown in columns three and four of Figure \ref{tab:tickAssign}.
 We can see that our stencil language is capable of precisely specifying the overwhelming majority of stencils.

### 2016-07-06 paper.tex
Fixing the text in the Stencils categories subsection.

@@
 
-\textbf{All reflexive} The vast majority (X\%) of stencils were
+\textbf{All-reflexive} The vast majority (X\%) of stencils were
 reflexive in every dimension of the array. These most often correspond
@@
 \end{minted}
-
-\textbf{All-reflexive} stencils are reflexive in every dimension of the array.
-These most often correspond to applying point-wise transformations to data such as scaling e.g.
-\mintinline{fortran}{ZZ1(I,K) = .000041868*ZZ1(I,K)}.
-Arrays with this class of stencil must be indexed by all the induction variables used in the LHS without any offsets.
-
-\textbf{Some-reflexive} stencils consist of only reflexive terms but leave some dimensions unspecified.
-Examples in this class include \mintinline{fortran}{leafpools(:,age_class)=leafpool(:,1)} (which contains an implicit loop), and \mintinline{fortran}{RVM(J,I2) = VRFO(ivar,J)} where \mintinline{fortran}{J} is the induction variable.
-Arrays with this class of stencil can be indexed by a subset of the LHS induction variables but again without any offsets.
-
-\textbf{Single-action} stencils consist of a single \term{backward},\term{forward}, or \term{centered} term and some number of reflexive terms.
+Arrays with this class of stencil must be indexed by all the induction
+variables used in the LHS without any offsets.
+
+\textbf{Some-reflexive} stencils consist of only reflexive terms but
+leave some dimensions unspecified. Two examples in this class are listed:
+\begin{minted}{fortran}
+leafpools(:,age_class)=leafpool(:,1)
+RVM(J,I2) = VRFO(ivar,J)
+\end{minted}
+where for the first example \mintinline{fortran}{leafpools} contains an
+implicit loop, and for the second example \mintinline{fortran}{J} is
+the induction variable. Arrays with this class of stencil can be
+indexed by a subset of the LHS induction variables but again without
+any offsets.
+
+\textbf{Single-action} stencils consist of a single \term{backward},
+\term{forward}, or \term{centered} term and some number of reflexive
+terms.
 % /home/mrd45/src/corpus/geos-chem/GTMM/CasaRegridModule.F90:645
@@
 
-\textbf{Single-action-irreflexive} stencils consist of a single \term{backward},\term{forward}, or \term{centered} irreflexive term and some number of reflexive terms.
+\textbf{Single-action-irreflexive} stencils consist of a single
+\term{backward}, \term{forward}, or \term{centered} irreflexive term
+and some number of reflexive terms.
 % /home/mrd45/src/corpus/geos-chem/GeosRad/rrtmg_lw_rtrnmc.F90:355
@@
 \end{minted}
-Notice in this example how the read from \mintinline{fortran}{cldfmc} flows through onto the line with the stencil.
+Notice that, in this last example, the read from
+\mintinline{fortran}{cldfmc} flows through onto the line with the
+stencil.
 

### 2016-07-06 paper.tex
fixed broken merge

@@
 \usepackage{amsthm}
-\usepackage{tikz}
 \usepackage{multirow}
@@
 \sisetup{group-separator={,},group-minimum-digits={3},output-decimal-marker={.}}
+\usepackage{tikz}
+\usepackage{xypic}
+
+\tikzset{round left paren/.style={ncbar=0.4cm,out=110,in=-110}}
+\tikzset{round right paren/.style={ncbar=0.4cm,out=70,in=-70}}
 
@@
 \begin{array}{c}
-\framebox{$\textit{region} \equiv \textit{region}'$} \\[1em]
+\framebox{$\textit{region} \equiv \textit{region}'$} \\[0.7em]
 \begin{array}{rll}
@@
 \stenFwdSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
- \equiv \; & \stenFwdSR{n}{d}{r} \; \texttt{+} \; \stenFwdSR{m}{d}{s} \\[1em]
+ \equiv \; & \stenFwdSR{n}{d}{r} \; \texttt{+} \; \stenFwdSR{m}{d}{s} \\[0.7em]
 %
@@
 \stenBwdSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
- \equiv \; & \stenBwdS{n}{d}{r} \; \texttt{+} \; \stenBwdSR{m}{d}{s} \\[1em]
+ \equiv \; & \stenBwdS{n}{d}{r} \; \texttt{+} \; \stenBwdSR{m}{d}{s} \\[0.7em]
 %
@@
 \stenCenSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
-\equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \; \stenCenSR{m}{d}{s} \\[1em]
+\equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \; \stenCenSR{m}{d}{s} \\[0.7em]
 %
@@
 \textit{$m \leq n$} \; \equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \;
-                      \stenFwdSR{m}{d}{s} \\[1em]
+                      \stenFwdSR{m}{d}{s} \\[0.7em]
 %
@@
 \textit{$m \leq n$} \; \equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \;
-                      \stenBwdSR{m}{d}{s} \\[1em]
+                      \stenBwdSR{m}{d}{s} \\[0.7em]
 %
@@
 \equiv \; & \stenFwdSR{n}{d}{r} \; \texttt{+} \; \stenBwdSR{n}{d}{s}
-  \\[1em]
+  \\[0.7em]
 %
@@
 \stenFwdS{n}{d} \\
-\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenFwdSR{n}{d}{r} \\[1em]
+\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenFwdSR{n}{d}{r} \\[0.7em]
 %
@@
 \stenBwdS{n}{d} \\
-\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenBwdSR{n}{d}{r} \\[1em]
+\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenBwdSR{n}{d}{r} \\[0.7em]
 %
@@
 \stenCenS{n}{d} \\
-\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenCenSR{n}{d}{r} \\[1em]
+\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenCenSR{n}{d}{r} \\[0.7em]
 %
@@
 %\framebox{${\textit{specDec} :: \texttt{v}} \equiv
-%{\textit{specDec'} :: \texttt{v}}$} \\[1em]
+%{\textit{specDec'} :: \texttt{v}}$} \\[0.7em]
 %\begin{array}{rl}
@@
 semantics, in terms of sets of vectors which we call \emph{index
-  schemes}. We introduce first the notion of indexing schemes
-and \emph{unification} which gives the domain of our model
-and forms the basis of the rest of the algorithsm for our
-specificaiton. We then define the denotational model.
-
-\subsection{Indexing schemes and unification}
+  schemes}. We introduces these first then define the denotational semantics.
+
+\subsection{Indexing schemes and generalisation}
 
@@
 
-\begin{definition}[Unification]
-Two schemes $u, v$ of dimensionality $n$
-are said to be \emph{unifiable}, written
-$u \sim_n v$, iff the following holds:
+\begin{definition}[Generalisation]
+Given two schemes $v, w$ of dimensionality $n$,
+the $w$ is said to \emph{generalise} $v$ 
+(or conversely $v$ \emph{specialises} $w$),
+written $v \preceq_n w$, iff the following holds:
 %%
 \[
-u \sim_n v \; \Leftrightarrow  \;
-  \forall i\!\in\!\{ 1,.., n \} . \, (u_i = \infty
-\, \vee \, v_i = \infty \, \vee \, u_i = v_i)
+v \preceq_n w \;\; \Leftrightarrow  \;\;
+  \forall i\!\in\!\{ 1, \ldots, n \} . \, (w_i = \infty \, \vee \, v_i = w_i)
 \]
 %%
-That is, two schemes are related by the binary relation
-$\sim_n$ over $(\mathbb{Z}_{\infty})^n$ if in each dimension
-either is $\infty$ or they are equal. 
-Thus, the element $\infty$ represents the ``general acceptor''
- meaning it unifies with any index in the same dimension.
-
-The subscript on $\sim_n$ is sometimes dropped when the
+That is, two schemes are related by the (reflexive and transitive) binary relation
+$\preceq_n$ over $(\mathbb{Z}_{\infty})^n$ if in each dimension
+the two schemes are equal or $w_i = \infty$ 
+in place of some $\mathbb{Z}$ in $v_i$ for $v \preceq_n w$. 
+Thus, the element $\infty$ 
+generalising any index in the same dimension.
+
+The subscript on $\preceq_n$ is sometimes dropped when the
 intended dimensionality is clear or inconsequential.
@@
 and $\vtwoh{\infty}{0}$ respectively. The following
-relationship of unification (or lack of) then holds between
-them:
+relationship of generalises then holds:
+%\[
+%\xymatrix@R=0.2em@C=2em{
+%[\infty \;\; 0] \ar@{-}[dr]^{\preceq_n} &  \\
+%& [0 \;\; 1]  \\
+%[\infty \;\; 1]  \ar@{-}[ur]_{\preceq_n} & 
+%}
+%\]
 %%
 \begin{equation*}
-\vtwoh{\infty}{0} \; \not\sim_2 \; \vtwoh{0}{1} \; \sim_2 \;
-\vtwoh{\infty}{1} \; \not\sim_2 \; \vtwoh{\infty}{0}
+\vtwoh{\infty}{0} \; \not\preceq_2 \; \vtwoh{0}{1} 
+\qquad
+\vtwoh{\infty}{1} \; \preceq_2 \; \vtwoh{0}{1} 
 \end{equation*}
 %%
-The middle pair unifies as $0$ and $\infty$ unify, 
-whereas the left and right pairs do not unify as
-$0 \not\sim_1 1$. 
 \end{example}
@@
 example, for the five point stencil there are five $1 \times 1$ spans.
-%
-\begin{center}
+Figure~\ref{fig:inference-steps-informal}(a) gives an illustration
+of these five unit spans.
+
+\begin{figure}[t]
+\vspace{-0.3em}
+\hspace{-0.5em}
+\begin{minipage}{0.16\linewidth}
+\begin{align*}
+\setlength{\arraycolsep}{0em}
+\begin{array}{c}
+\quad \\[0.2em]
+{\scalebox{0.75}{
 \begin{tikzpicture}
@@
 \end{tikzpicture} 
-\end{center}
+}} \\[0.36em]
+(a)
+\end{array}
+\end{align*}
+\end{minipage} $\rightarrow$ \begin{minipage}{0.43\linewidth}
+\begin{align*}
+\setlength{\arraycolsep}{0em}
+\begin{array}{c}
+{\scalebox{0.75}{
+\begin{tikzpicture}
+%\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
+%\node at (1.25,1.22) {\textnormal{\small{0,0}}};
+\draw [black, thin] (0.35,0) to [round left paren ] (0.35,2);
+\draw[thick] (0.5,1) rectangle (2,1.5);
+\draw[thick] (1,0.40) rectangle (1.5,0.90);
+\draw[thick] (1,1.60) rectangle (1.5,2.10);
+\node at (2.3,1.22) {+};
+\draw[thick] (3.8,1) rectangle (4.3,1.5);
+\draw[thick] (2.6,1) rectangle (3.1,1.5);
+\draw[thick] (3.2,0.5) rectangle (3.7,2);
+\draw [black, thin] (4.4,0) to [round right paren ] (4.4,2);
+\end{tikzpicture}
+}}
+ \\ (b)
+\end{array}
+\end{align*}
+\end{minipage}$\rightarrow$ \begin{minipage}{0.25\linewidth}
+\begin{align*}
+\setlength{\arraycolsep}{0em}
+\begin{array}{c}
+{\scalebox{0.75}{
+\begin{tikzpicture}
+%\node at (1.25,1.22) {\textnormal{\small{0,0}}};
+\draw [black, thin] (0.8,0) to [round left paren ] (0.8,2);
+\draw[thick] (1,0.5) rectangle (1.5,2);
+\node at (1.8,1.22) {+};
+\draw[thick] (2.1,1) rectangle (3.6,1.5);
+\draw [black, thin] (3.7,0) to [round right paren ] (3.7,2);
+%\node at (2.75,1.22) {\textnormal{\small{0,0}}};
+\end{tikzpicture}
+}}  \\ (c)
+\end{array}
+\end{align*}
+\end{minipage}
+\caption{Illustration of grouping spans in 
+specification inference}
+\label{fig:inference-steps-informal}
+\end{figure}
+
 %
@@
 $m+1$-dimensions. In this example, the $0$-dimensional spans
- become $1$-dimensional spans (rows/columns):
-%
-\begin{align*}
-\begin{array}{cc}
-\begin{tikzpicture}
-%\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
-%\node at (1.25,1.22) {\textnormal{\small{0,0}}};
-\draw[thick] (0.5,1) rectangle (2,1.5);
-\draw[thick] (1,0.40) rectangle (1.5,0.90);
-\draw[thick] (1,1.60) rectangle (1.5,2.10);
-\end{tikzpicture}
-\qquad
-&
-\qquad
-\begin{tikzpicture}
-%\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
-%\node at (1.25,1.22) {\textnormal{\small{0,0}}};
-\draw[thick] (1.6,1) rectangle (2.1,1.5);
-\draw[thick] (0.4,1) rectangle (0.9,1.5);
-\draw[thick] (1,0.5) rectangle (1.5,2);
-\end{tikzpicture}
-\\
-\textit{dimension 1} \qquad & \qquad \textit{dimension 2}
-\end{array}
-\end{align*}
-%
+ become $1$-dimensional spans (rows/columns), illustrated
+in Figure~\ref{fig:inference-steps-informal}(b).
 Any spans that are contained with any other span are deleted,
 leaving a minimal set of spans (which may overlap, but none of which
-fully contains another).
-\begin{center}
-\begin{tikzpicture}
-%\node at (1.25,1.22) {\textnormal{\small{0,0}}};
-\draw[thick] (1,0.5) rectangle (1.5,2);
-\node at (2,1.22) {+};
-\draw[thick] (2.5,1) rectangle (4,1.5);
-%\node at (2.75,1.22) {\textnormal{\small{0,0}}};
-\end{tikzpicture}
-\end{center}
-%
+fully contains another) (Figure~\ref{fig:inference-steps-informal}(c).
+
 This procedure is iterated till a fixed point is reached. In this

### 2016-07-06 paper.tex
purged unification from section 3

@@
 \begin{figure}
+\vspace{-0.2em}
 \begin{align*}
@@
 \end{example}
-
+\noindent
 Our terminology of \emph{indexing schemes} and
-\emph{unification} are by analogy
-with the \emph{type schemes} and \emph{type unification} of
+\emph{generalisation} are by analogy
+with the \emph{type schemes} and \emph{type generalisation} of
 polymorphic types in ML~\cite{milner1978theory}. Our notion is
-much more rigid however, there is only one way to unify two
-index schemes in our model.
+much more rigid however, as our schemes do not bind names
+and thus generalisation of does not require substitution; $\infty$ acts
+as a positional placeholder.
 
@@
 \vtwoh{1}{\infty}$ and $\singleEntry{2}{3} = \vthreeh{\infty}{1}{\infty}$.
-Thus, a single-entry vector is an indexing scheme which unifies with
-all other schemes which have $1$ as their $r^{th}$ entry,
+Thus, a single-entry vector is an indexing scheme which generalises 
+all other schemes which at least have $1$ as their $r^{th}$ entry,
 describing a neighbour offset of $1$. 
@@
 Similarly to the single-entry vector, zero-entry vectors 
-are also indexing schemes which unify with all other schemes
-as long as the $r^{th}$ entry is $0$, representing indices at
+are also indexing schemes which generalising all other schemes
+which at least have the $r^{th}$ entry is $0$, representing indices at
 the ``origin'' in dimension $r$. 
@@
 where $\vtwoh{1}{\infty}$ and $\vtwoh{3}{\infty}$ have been filtered.
-With respect to the notion of unification, the 
+%With respect to the notion of unification, the 
 %
@@
 %of the unifiers in different dimensions into less accepting unifiers.
-%This means an access pattern unifying with this unifier would require
+%This means an access pattern generalises with this scheme would require
 %both models to be satisfied. Use of $dims$ gives the dimensions that
@@
 %simultaneously.
+%  u in M * N   u < w
+% mplies that 
+%  u_1 in M  u1 < w
+%  u_2 in M  u2 < w
 %
@@
 %%
-\dnote{I didn't understand mistral's description of unification
-wrt to tensor. I can't work it out either! Mistral have another
-go given my above explanation}
 \end{definition}

### 2016-07-06 paper.tex
Complete negative example for checking and amend consistency function

@@
 
-We will consider both a consistent and an inconsistent example to show how
-checking works. The following is an example with a looser but still consistent
-version of the \textit{fivepoint} stencil specification inferred previously in
-the Subsection~\ref{subsec:inference}:
+We will consider an extended example of a stencil computation that is
+consistent with its annotation to show how checking works. The following is an
+example with a looser but still consistent version of the \textit{fivepoint}
+stencil specification inferred previously in the
+Subsection~\ref{subsec:inference}:
 %
@@
 
-  The unifications for the lower bound:
+  The unifications for the upper bound:
   \begin{align*}
@@
 
-  The unifications for the upper bound:
+  The unifications for the lower bound:
   \begin{align*}
@@
 \end{enumerate}
+
+Now when $L3$ is slightly changed with an additional term \texttt{b(i+2,j)}, the
+new assignment $L3'$ becomes:
+%
+\begin{minted}[firstnumber=3]{fortran}
+a(i,j) = b(i, j-1) + b(i,j) + b(i,j+1) + b(i-1,j) + b(i+1,j) + b(i+2,j+2)
+\end{minted}
+%
+The first four steps will proceed the same way and in the fifth step result
+of $\mathit{analyse}(L3',b)$ will have the extra term $\vtwoh{2}{2}$. In sixth
+step, since the model has not changed and the relativised indices are only
+augmented with a new element, the unification for lower bound still hold.
+However, there are not elements in the model, $m$, that can unify with
+$\vtwoh{2}{2}$, hence the model fails to hold as an upper bound.
+
+Note that if the specification was as follows:
+%
+\begin{minted}{fortran}
+!= stencil atLeast, readOnce, fivepoint :: b}
+\end{minted}
+%
+The consistency function would have held as
+$\mathit{cons}(\textsf{mult}(\textsf{up}(m)),L3',b)$ for the upper bound would
+have never been needed. This semantically makes sense since the region
+$\texttt{fivepoint}$ is indeed a lower bound for the access pattern of $b$.
 

### 2016-07-06 paper.tex
work on spec model; fixing figure 7

@@
 \usepackage{xypic}
+
+\usetikzlibrary{calc}
+
+%% http://tex.stackexchange.com/questions/55068/is-there-a-tikz-equivalent-to-the-pstricks-ncbar-command
+\tikzset{
+    ncbar angle/.initial=90,
+    ncbar/.style={
+        to path=(\tikztostart)
+        -- ($(\tikztostart)!#1!\pgfkeysvalueof{/tikz/ncbar angle}:(\tikztotarget)$)
+        -- ($(\tikztotarget)!($(\tikztostart)!#1!\pgfkeysvalueof{/tikz/ncbar angle}:(\tikztotarget)$)!\pgfkeysvalueof{/tikz/ncbar angle}:(\tikztostart)$)
+        -- (\tikztotarget)
+    },
+    ncbar/.default=0.5cm,
+}
 
@@
 Similarly to the single-entry vector, zero-entry vectors 
-are also indexing schemes which generalising all other schemes
+are generalise all other schemes
 which at least have the $r^{th}$ entry is $0$, representing indices at
@@
 %  the multi-set $A$}
-\caption{Model of regions, $\interp{\textit{region}}$}
+\caption{Model of regions, 
+$\interp{-}_n : \textit{region} \rightarrow \mathcal{P}(\domainVal{}^n)$}
 \label{fig:region-model}
@@
 %%
-Internally $\otimes$ takes a Cartesian product of the indexing
-schemes (all pairs) and then combines each pair using the 
-\emph{pairwise permutation} written $\bowtie$ (defined later)
-which generates all possible interleavings of two vectors
+The tensor takes a Cartesian product of the two models,
+giving all pairs of schemes which are combined using the 
+\emph{pairwise permutation} written $\bowtie$ (defined later).
+Pairwise permutation generates all possible interleavings of two vectors
 (\ie{}, all possible non-deterministic choices between $u_i$ and
 $v_i$ in each dimension). Some care is then taken over
-how this interacts with $\infty$: pariwse permtuation
+how this interacts with $\infty$: pairwise permtuation
 corresponds to non-determinstic choice, in each dimension $i$,
@@
 corresponds to taking all bit-strings of length $n$ and
-selecting from $u$ for $1$ and $v$ for $0$. Thus, we can
-defined the pairwise permutation matrix as:
+selecting from $u$ for $1$ and $v$ for $0$. The pairwise permutation
+matrix  s thus defined in terms of the logical
+matrix $b$ where $b_{i,j}$ is the $j$-th bit of the integer
+$i - 1$ as:
 %
@@
 %
-where $b$ is a logical matrix where $b_{i,j}$ is the $j$-th bit of the integer
-$i - 1$.
-%%
 \end{definition}
@@
 
+\paragraph{Model of \textit{spec}}
+
+Thus far we have explained the model of regions as sets of index
+schemes, that is
+$\interp{\textit{region}}_n \in \mathcal{P}(\domainVal{}^n)$.
+Figure~\ref{fig:spatial-model} defines the top-level model of the
+\textit{spec} syntax, which includes the multiplicity and
+approximation modifiers.
+
 \begin{figure}
 \input{spatial-model.tex}
-\caption{Model of specifications, $\interp{\textit{spec}}$}
+\caption{\textit{spec} model, 
+$\interp{-}_n : \textit{spec} \rightarrow \textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$}
 \label{fig:spatial-model}
 \end{figure}
-
-\paragraph{Model of \textit{spec}}
-
-\todo{Explain the injections and the domain here}
 
@@
 \quad \\[0.2em]
-{\scalebox{0.75}{
-\begin{tikzpicture}
+\begin{tikzpicture}[scale=0.75]
 %\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
@@
 \draw[thick] (1,1.5) rectangle (1.5,2);
-\end{tikzpicture} 
-}} \\[0.36em]
+\end{tikzpicture} \\[0.36em]
 (a)
@@
 \begin{array}{c}
-{\scalebox{0.75}{
-\begin{tikzpicture}
+\begin{tikzpicture}[scale=0.75]
 %\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
@@
 \draw [black, thin] (4.4,0) to [round right paren ] (4.4,2);
-\end{tikzpicture}
-}}
- \\ (b)
+\end{tikzpicture} \\ (b)
 \end{array}
@@
 \begin{array}{c}
-{\scalebox{0.75}{
-\begin{tikzpicture}
+\begin{tikzpicture}[scale=0.75]
 %\node at (1.25,1.22) {\textnormal{\small{0,0}}};
@@
 %\node at (2.75,1.22) {\textnormal{\small{0,0}}};
-\end{tikzpicture}
-}}  \\ (c)
+\end{tikzpicture} \\ (c)
 \end{array}

### 2016-07-06 paper.tex
reformat tab:corpus

@@
                      % used to print nice-looking tables of numbers.
-\sisetup{group-separator={,},group-minimum-digits={3},output-decimal-marker={.}}
+\sisetup{ % defaults
+  group-separator={,},
+  group-minimum-digits={3},
+  output-decimal-marker={.},
+  table-format = 6
+}
+
 \usepackage{tikz}
@@
 
-Thus far we have explained the model of regions as sets of index
+We have explained the model of regions as sets of index
 schemes, that is
@@
 \textit{spec} syntax, which includes the multiplicity and
-approximation modifiers.
+approximation modifiers. The domain of this interepretation
+is $\textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$
+where $\textsf{Mult}$ and $\textsf{Approx}$ are labelled variants
+labelling our set-basd model with additional information.
+
+\begin{definition} The \textsf{Mult}
+and \textsf{Approx} are parametric labelled variant (sum) types 
+with injection function given by their definition:
+%
+\begin{align*}
+\textsf{Mult} \, a & = \textsf{mult} \, a \,\mid\, \textsf{only} \, a \\
+\textsf{Approx} \, a & = \textsf{exact} \, a \,\mid\, \textsf{lower} \, a \,\mid\,
+\textsf{upper} \, a \,\mid\, \textsf{both} \, a \, a
+\end{align*}
+\ie{}, $\mathsf{lower}$ is an injection $\mathsf{lower} : a \rightarrow \mathsf{Approx} \, a$, and so
+on. These are used for labelled models, and other components later in
+Section~\ref{sec:analysis} with additional information.
+\label{def:mult-and-approx}
+\end{definition}
+%
 
@@
 \centering
-\sisetup{
-  table-format = 7,
-  table-number-alignment = right
-}
-\begin{tabular}{|l|S|S|S|}
+\footnotesize
+\begin{tabular}{lSSSS}
 \hline
-Package   & {Files} & {Files analysed} & {Lines of code}  \\ % braces prevent headers from being formatted as numbers
+& \multicolumn{2}{c}{Number of files} & \multicolumn{2}{c}{Lines of code} \\
+
+Package   & {Total} & {Parsed} & {Total} & {Parsed}  \\ % braces prevent headers from being formatted as numbers
 \hline
-UM        & \umFiles                       & \umparseOk                      & \umLoC                     \\
-E3MG      & \ethreemgeaFiles               & \ethreemgeaparseOk              & \ethreemgeaLoC             \\
-BLAS      & \blasFiles                     & \blasparseOk                    & \blasLoC                   \\
-Hybrid4   & \hybridfourFiles               & \hybridfourparseOk              & \hybridfourLoC             \\
-GEOS-Chem & \geoschemFiles                 & \geoschemparseOk                & \geoschemLoC               \\
-Navier    & \navierFiles                   & \navierparseOk                  & \navierLoC                 \\
-CP        & \computationalphysicstwoFiles  & \computationalphysicstwoparseOk & \computationalphysicstwoLoC\\
+UM        & \umFiles                       & \umparseOk                      & \umLoC                      & \umlinesParsed \\
+E3MG      & \ethreemgeaFiles               & \ethreemgeaparseOk              & \ethreemgeaLoC              & \ethreemgealinesParsed \\
+BLAS      & \blasFiles                     & \blasparseOk                    & \blasLoC                    & \blaslinesParsed \\
+Hybrid4   & \hybridfourFiles               & \hybridfourparseOk              & \hybridfourLoC              & \hybridfourlinesParsed \\
+GEOS-Chem & \geoschemFiles                 & \geoschemparseOk                & \geoschemLoC                & \geoschemlinesParsed \\
+Navier    & \navierFiles                   & \navierparseOk                  & \navierLoC                  & \navierlinesParsed \\
+CP        & \computationalphysicstwoFiles  & \computationalphysicstwoparseOk & \computationalphysicstwoLoC & \computationalphysicstwolinesParsed \\
 \hline
-\textbf{Total} & \overallFiles  & \overallparseOk & \overallLoC  \\
+\textbf{Total} & \overallFiles  & \overallparseOk & \overallLoC & \overalllinesParsed \\
 \hline
@@
 packages used in our evaluation, all of which are written in Fortran
-90 or Fortran 77. In total we analysed 1.1 million lines of code from
-7 packages. The ``Files analysed'' column shows the number of files in
-each corpus that we were able to analyse with Camfort. The most common
-reasons for Camfort rejecting a file were either use of a
-C-preprocessor, or illegal use of language features from a more modern
-Fortran variant.
+90 or Fortran 77. In total we analysed \SI{\overallLoC} lines of code
+from 7 packages, of which we successfully parsed
+\SI{\overalllinesParsed} lines. The ``Number of files'' column shows
+how many files in each corpus that we were able to analyse with
+Camfort. The most common reasons for Camfort rejecting a file were
+either use of a C-preprocessor, or illegal use of language features
+from a more modern Fortran variant.
 

### 2016-07-06 resultsToTex.py
change lines-of-code summing to use 'linesParsed' instead of 'linesTotal'.

@@ -45,7 +45,7 @@ for corpus in corpora.keys():
         for key in keys.keys():
             if not corpora[corpus].has_key(key):
                 print "\\newcommand{\\%s}{0}" % (corpus+key)
-        loc = files[corpus+"linesTotal"]
+        loc = files[corpus+"linesParsed"]
         tickAssign = files[corpus+"tickAssign"]
         tickAssignSuccess = files[corpus+"tickAssignSuccess"]
         totalTickAssign += int(tickAssign)

### 2016-07-06 paper.tex
make labelled variants for Mult and Approx

--- 2016-07-06 paper.tex
@@
 \textit{spec} syntax, which includes the multiplicity and
-approximation modifiers.
+approximation modifiers. The domain of this interepretation
+is $\textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$
+where $\textsf{Mult}$ and $\textsf{Approx}$ are labelled variants
+labelling our set-basd model with additional information.
+
+\begin{definition} The \textsf{Mult}
+and \textsf{Approx} are parametric labelled variant types 
+with injection function given by their definition:
+%
+\begin{align*}
+\textsf{Mult} \, a & = \textsf{mult} \, a \mid \textsf{only} \, a \\
+\textsf{Approx} \, a & = \textsf{exact} \, a \mid \textsf{lower} \, a \mid
+\textsf{upper} \, a \mid \textsf{both} \, a \, a
+\end{align*}
+\ie{}, $\mathsf{lower}$ is an injection $\mathsf{lower} : a \rightarrow \mathsf{Approx} \, a$, and so
+on. 
+\label{def:mult-and-approx}
+\end{definition}
+%
 

### 2016-07-06 paper.tex
tweak to variants

@@
 \begin{definition} The \textsf{Mult}
-and \textsf{Approx} are parametric labelled variant types 
+and \textsf{Approx} are parametric labelled variant (sum) types 
 with injection function given by their definition:
@@
 \begin{align*}
-\textsf{Mult} \, a & = \textsf{mult} \, a \mid \textsf{only} \, a \\
-\textsf{Approx} \, a & = \textsf{exact} \, a \mid \textsf{lower} \, a \mid
-\textsf{upper} \, a \mid \textsf{both} \, a \, a
+\textsf{Mult} \, a & = \textsf{mult} \, a \,\mid\, \textsf{only} \, a \\
+\textsf{Approx} \, a & = \textsf{exact} \, a \,\mid\, \textsf{lower} \, a \,\mid\,
+\textsf{upper} \, a \,\mid\, \textsf{both} \, a \, a
 \end{align*}
 \ie{}, $\mathsf{lower}$ is an injection $\mathsf{lower} : a \rightarrow \mathsf{Approx} \, a$, and so
-on. 
+on. These are used for labelled models, and other components later in
+Section~\ref{sec:analysis} with additional information.
 \label{def:mult-and-approx}

### 2016-07-06 paper.tex

--- 2016-07-06 paper.tex
@@
 \begin{table}
-\begin{tabular}{|l|l|l|l|l|}
+\sisetup{table-format=4,table-text-alignment=right}
+\footnotesize
+\centering
+\begin{tabular}{lS@{~~~}rS@{~~~}rSS[table-format=1]}
 \hline
-{Package}     & {Potential} & {Actual} & {Exact} & {Inexact}    \\
+  & \multicolumn{2}{c}{Potential} & \multicolumn{2}{c}{Actual} & {Exact} & {Inexact}    \\
+{Package}     & {\#} & \si{\percent} & {\#} & \si{\percent} \\
 \hline
-UM            & \umtickAssign\ (\umtickAssignPercent\%) &
-                \umtickAssignSuccess\ (\umtickAssignSuccessPercent\%) &
+UM            & \umtickAssign & \SI{\umtickAssignPercent}{\percent} &
+                \umtickAssignSuccess & \SI{\umtickAssignSuccessPercent}{\percent} &
                 \umnumStencilLines &
                 \umboundedBoth  \\
-E3MG          & \ethreemgeatickAssign\ (\ethreemgeatickAssignPercent\%) &
-                \ethreemgeatickAssignSuccess\ (\ethreemgeatickAssignSuccessPercent\%) &
+E3MG          & \ethreemgeatickAssign & \SI{\ethreemgeatickAssignPercent}{\percent} &
+                \ethreemgeatickAssignSuccess & \SI{\ethreemgeatickAssignSuccessPercent}{\percent} &
                 \ethreemgeanumStencilLines &
                 \ethreemgeaboundedBoth \\
-BLAS          & \blastickAssign\ (\blastickAssignPercent\%) &
-                \blastickAssignSuccess\ (\blastickAssignSuccessPercent\%) &
+BLAS          & \blastickAssign & \SI{\blastickAssignPercent}{\percent} &
+                \blastickAssignSuccess & \SI{\blastickAssignSuccessPercent}{\percent} &
                 \blasnumStencilLines &
                 \blasboundedBoth \\
-Hybrid4       & \hybridfourtickAssign\ (\hybridfourtickAssignPercent\%) &
-                \hybridfourtickAssignSuccess\ (\hybridfourtickAssignSuccessPercent\%) &
+Hybrid4       & \hybridfourtickAssign & \SI{\hybridfourtickAssignPercent}{\percent} &
+                \hybridfourtickAssignSuccess & \SI{\hybridfourtickAssignSuccessPercent}{\percent} &
                 \hybridfournumStencilLines &
                 \hybridfourboundedBoth\\
-GEOS-Chem     & \geoschemtickAssign\ (\geoschemtickAssignPercent\%) &
-                \geoschemtickAssignSuccess\ (\geoschemtickAssignSuccessPercent\%) &
+GEOS-Chem     & \geoschemtickAssign & \SI{\geoschemtickAssignPercent}{\percent} &
+                \geoschemtickAssignSuccess & \SI{\geoschemtickAssignSuccessPercent}{\percent} &
                 \geoschemnumStencilLines &
                 \geoschemboundedBoth \\
-Navier        & \naviertickAssign\ (\naviertickAssignPercent\%) &
-                \naviertickAssignSuccess\ (\naviertickAssignPercent\%) &
+Navier        & \naviertickAssign & \SI{\naviertickAssignPercent}{\percent} &
+                \naviertickAssignSuccess & \SI{\naviertickAssignPercent}{\percent} &
                 \naviernumStencilLines &
                 \navierboundedBoth \\
-CP            & \computationalphysicstwotickAssign\ (\computationalphysicstwotickAssignPercent\%) &
-                \computationalphysicstwotickAssignSuccess\ (\computationalphysicstwotickAssignSuccessPercent\%) &
+CP            & \computationalphysicstwotickAssign & \SI{\computationalphysicstwotickAssignPercent}{\percent} &
+                \computationalphysicstwotickAssignSuccess & \SI{\computationalphysicstwotickAssignSuccessPercent}{\percent} &
                 \computationalphysicstwonumStencilLines &
@@
 \hline
-\textbf{Total}& \overalltickAssign\ (\overalltickAssignPercent\%)  &
-                \overalltickAssignSuccess\ (\overalltickAssignSuccessPercent\%) &
+\textbf{Total}& \overalltickAssign & \SI{\overalltickAssignPercent}{\percent}  &
+                \overalltickAssignSuccess & \SI{\overalltickAssignSuccessPercent}{\percent} &
                 \overallnumStencilLines &
@@
 
-The first two columns of Table~\ref{tab:tickAssign} shows the number of potential and actual stencils and their rate as a percentage of the total number of assignment statements in the code.
-There are a variety of reasons why we might not infer an actual stencil from a potential one: 1) \textbf{Non-subset induction variables} occurs when the induction variables on the RHS are not a subset of the LHS.
-These cases are not stencils by our definition.
-The degenereate case of this is to have only constant indices on the LHS.
-We see lots of examples of this in loops as accumulators e.g.
-computing the sum over some other array; 2) \textbf{Derived induction variables} where the indexing variable (\mintinline{fortran}{x}) is derived from an induction variable \mintinline{fortran}{i} e.g.
-\mintinline{fortran}{x = len - i}; 3) \textbf{Inconsistent induction dimensions} occur when an induction variable is used to specify more than one array dimension on the RHS.
-These are common in matrix operations such as LU-decomposition which contains assignments such as \mintinline{fortran}{a(l) = a(l) - a(m) * b(l, m)}.
+The first two columns of Table~\ref{tab:tickAssign} shows the number
+of potential and actual stencils and their rate as a percentage of the
+total number of successfully parsed lines of code. There are a variety
+of reasons why we might not infer an actual stencil from a potential
+one: 1) \textbf{Non-subset induction variables} occurs when the
+induction variables on the RHS are not a subset of the LHS. These
+cases are not stencils by our definition. The degenereate case of this
+is to have only constant indices on the LHS. We see lots of examples
+of this in loops as accumulators e.g. computing the sum over some
+other array; 2) \textbf{Derived induction variables} where the
+indexing variable (\mintinline{fortran}{x}) is derived from an
+induction variable \mintinline{fortran}{i} e.g. \mintinline{fortran}{x
+  = len - i}; 3) \textbf{Inconsistent induction dimensions} occur when
+an induction variable is used to specify more than one array dimension
+on the RHS. These are common in matrix operations such as
+LU-decomposition which contains assignments such as
+\mintinline{fortran}{a(l) = a(l) - a(m) * b(l, m)}.
 

### 2016-07-06 paper.tex
eliminate an overflow
slight adjustment to tab:corpus

@@
 %
-The consistency function would have held as
-$\mathit{cons}(\textsf{mult}(\textsf{up}(m)),L3',b)$ for the upper bound would
-have never been needed. This semantically makes sense since the region
-$\texttt{fivepoint}$ is indeed a lower bound for the access pattern of $b$.
+The consistency function would have held because for the upper bound
+$\mathit{cons}(\textsf{mult}(\textsf{up}(m)),L3',b)$ would never have
+been needed. This semantically makes sense since the region
+$\texttt{fivepoint}$ is indeed a lower bound for the access pattern of
+$b$.
 
@@
 \footnotesize
-\begin{tabular}{lSSSS}
+\begin{tabular}{lS[table-format=4]S[table-format=4]SS}
 \hline

### 2016-07-06 paper.tex
Make checking consistent with new terminology & add consistency clause for 'both'

@@
 
-\section{Inference and checking}
+\section{Inference, checking, and synthesis}
 \label{sec:analysis}
@@
 
-Verification occurs when a user supplied specification is compared to the
-actual access pattern. The mathematical model developed in
-Section~\ref{sec:semantics} is used this comparison.
-
-A specification is first synthesised into its model. Then access patterns
-are checked against the model for consistency as defined in
-Figure~\ref{fig:consistency}. The consistency predicate is defined recursively.
-It first checks multiplicity if $\textsf{once}$ is at the top level, \ie{}
-specification in question involves \texttt{readOnce} modifier, then array
-accesses are checked for uniqueness using $unique$ predicate and the problem
-is reduced to $\textsf{mult}$, which does not impose any extra constraints.
+In checking mode, the specification is verified with respect to the indexing
+scheme of the actual access pattern. The mathematical model of the specification
+developed in Section~\ref{sec:semantics} is used this process.
+
+A specification is first synthesised into its model. Then occurrences of an
+array in an assignment are checked against the model for consistency as defined
+in Figure~\ref{fig:consistency}. The consistency predicate is defined
+recursively.
+
+It first checks multiplicity if $\textsf{once}$ is at the top
+level, \ie{} specification in question involves \texttt{readOnce} modifier, then
+the array accesses are checked for uniqueness using $unique$ predicate and the
+problem is reduced to $\textsf{mult}$, which does not impose any extra
+constraints.
+
 Next, consistency predicate considers whether the stencil is an approximation.
-Semantically, a lower bound is achieved if every unifier in the model
-correspond to some relativised vector of indices as that means all cells
+Semantically, a lower bound is achieved if every indexing scheme in the
+model generalises some relativised access pattern that means all cells
 described in the specification are part of the access pattern. Conversely,
-an upper bound means all observed access patterns have a unifier. Hence,
-the actual cells are a subset of the specified ones. In combination an
-exact specification, \ie{} one without any approximation modifiers, is
-one that is both a lower and an upper bound.
+an upper bound means all observed access patterns can be generalised with
+the indexing schemes of the model. Hence, the actual cells are a subset of the
+specified ones. In combination an exact specification, \ie{} one without any
+approximation modifiers, is one that is both a lower and an upper bound.
 
@@
 %
-\item To evaluate the two predicates above, we need relativised indices for the
+\item To evaluate the two predicates above, we need indexing schemes for the
   array in question. This is obtained by $analyse$ function from
@@
 
-  For the example above the relativised indices are as follows:
+  For the example above the indexing schemes are as follows:
 %
@@
 %
-\item Below we present two match ups of relativised access patterns and unifiers
-  to show both the lower and the upper bound holds.
-
-  The unifications for the upper bound:
+\item Below we present two sets of generalisations to show both the lower and
+  the upper bound holds.
+
+  The generalisations for the upper bound:
   \begin{align*}
@@
 
-  The unifications for the lower bound:
+  The generalisations for the lower bound:
   \begin{align*}
@@
 
-  Since both predicates are satisfied the array accesses specified on line three
-  are consistent with the \texttt{fivepoint} specification.
+  Since both predicates are satisfied the array accesses specified on line
+  three, the stencil comutation is consistent with the \texttt{fivepoint}
+  specification.
 \end{enumerate}
 
-Now when $L3$ is slightly changed with an additional term \texttt{b(i+2,j)}, the
-new assignment $L3'$ becomes:
+Now when $L3$ is slightly changed with an additional term \texttt{b(i+2,j+2)},
+the new assignment $L3'$ becomes:
 %
@@
 %
-The first four steps will proceed the same way and in the fifth step result
-of $\mathit{analyse}(L3',b)$ will have the extra term $\vtwoh{2}{2}$. In sixth
-step, since the model has not changed and the relativised indices are only
-augmented with a new element, the unification for lower bound still hold.
-However, there are not elements in the model, $m$, that can unify with
+The first four steps will proceed the same way. In the fifth step, result
+of $\mathit{analyse}(L3',b)$ will have the extra term $\vtwoh{2}{2}$. In the
+sixth step, since the model has not changed and the indexing schemes are only
+augmented with a new element, the generalisations for lower bound still hold.
+However, there are no elements in the model, $m$, that can generalise
 $\vtwoh{2}{2}$, hence the model fails to hold as an upper bound.
@@
 \begin{minted}{fortran}
-!= stencil atLeast, readOnce, fivepoint :: b}
-\end{minted}
-%
-The consistency function would have held because for the upper bound
-$\mathit{cons}(\textsf{mult}(\textsf{up}(m)),L3',b)$ would never have
+!= stencil atLeast, readOnce, fivepoint :: b
+\end{minted}
+%
+The consistency predicate would have held because for the upper bound
+$\mathit{cons}(\textsf{mult}(\textsf{up}(m)),L3',b)$ would not have
 been needed. This semantically makes sense since the region
-$\texttt{fivepoint}$ is indeed a lower bound for the access pattern of
-$b$.
+$\texttt{fivepoint}$ is indeed contained in the regions touched by $b$.
+
+\subsection{Synthesis\label{subsec:synthesis}}
 

### 2016-07-06 paper.tex
Forgot to change sim to preceq
rest of algebra

--- 2016-07-06 paper.tex
@@
 \begin{definition} The \textsf{Mult}
-and \textsf{Approx} are parametric labelled variant (sum) types 
-with injection function given by their definition:
+and \textsf{Approx} are parametric labelled variant types 
+with injection functions given by their definition:
 %
@@
 \ie{}, $\mathsf{lower}$ is an injection $\mathsf{lower} : a \rightarrow \mathsf{Approx} \, a$, and so
-on. These are used for labelled models, and other components later in
+on. The $\textsf{Mult}$ type corresponds to the presence or absence of the
+\term{readOnce} modifier as shown in Figure~\ref{fig:spatial-model}. 
+The $\textsf{Approx}$ type corresponds to the presence or absence of
+the spatial approximation modifier, with \textsf{exact} when there is
+no such modifier and \textsf{lower} and \textsf{upper} for
+\term{atLeast} and \term{atMost}. Later, it will be useful to model
+pairs of lower and upper bounds which is provided by \textsf{both} 
+(but which isn't used here).
+
+These are used for labelled models, and other components later in
 Section~\ref{sec:analysis} with additional information.
@@
 
-\begin{figure}
+\begin{figure}[t]
 \input{spatial-model.tex}
@@
 \label{fig:spatial-model}
+\vspace{-1em}
 \end{figure}
@@
 consider soundness for the equational theory of
-specifications and the notion of spatial approximation
-on specifications. The proofs are given in our supplement.
-
-\begin{theorem}[Soundness of equational theory]
+specifications and the notion of spatial approximation. The proofs are given in our supplement.
+
+\begin{theorem}[Soundness of equational theory (Figure~\ref{fig:equations})]
 \[
-\overline{\texttt{v} : S}\equiv \overline{\texttt{u} : T}
-\; \Rightarrow \;
-\interp{\overline{\texttt{v} : S}} = \interp{\overline{\texttt{u} : T}}
+S \equiv T \; \Rightarrow \;
+\interp{S}_n = \interp{T}_n
 \]
@@
 \begin{align*}
-& (\vect{x}^L,\vect{x}^U) \bullet (\vect{y}^L,\vect{y}^U) \\
+\vect{x} \bullet \vect{y}
 = &
@@
 
-\newcommand{\bplus}{\operatornamewithlimits{\term{+}}}
+\newcommand{\bplus}{\operatornamewithlimits{\term{\huge{+}}}}
 \newcommand{\tySum}[1]{#1^{\term{+}}}
@@
 \newcommand{\specDNF}{\textbf{spec}}
-\newcommand{\result}[1]{\lfloor{#1}\rceil}
 
@@
 space is converted into abstract syntax trees 
-for the specification language. This conversion is defined in terms
+of the specification language. This conversion is defined in terms
 of an algebra on \textit{spec} syntax in disjunctive-normal form
-(which we denote \specDNF{}).
-
-\begin{align*}
-\result{\specDNF{}} = \specDNF + ((\specDNF{} + 1) \times
-  (\specDNF{} + 1))
-\end{align*}
-
-\begin{align*}
-\result{\specDNF{}} & ::= \textsf{exact} \, \specDNF{}
-\mid \textsf{bounds} ([\specDNF{}], [\specDNF{}]) \\
-\end{align*}
-
-The algebra mirrors the shape
+(which we denote \specDNF{}), and which is wrapped
+with the $\textsf{Approx}$ type (see
+Definition~\ref{def:mult-and-approx},
+p.~\pageref{def:mult-and-approx}), which we write as $\mathsf{A}$
+for brevity.  The algebra mirrors the shape
 of the region combinators for addition and multiplication, 
@@
 \begin{align*}
-\oplus & : \specDNF \rightarrow \specDNF \rightarrow
-  \specDNF \\
-\odot & : \specDNF \rightarrow \specDNF \rightarrow
-        \specDNF \\
+\oplus & : \mathsf{A}(\specDNF) \rightarrow \mathsf{A}(\specDNF) \rightarrow
+  \mathsf{A}(\specDNF) \\
+\odot & : \mathsf{A}(\specDNF) \rightarrow \mathsf{A}(\specDNF) \rightarrow
+        \mathsf{A}(\specDNF) \\
 \mathsf{toSpec} & : (\mathbb{Z} \cup \{\infty\}) \rightarrow (\mathbb{Z} \cup
-  \{\infty\})  \rightarrow \specDNF
+  \{\infty\})  \rightarrow \mathsf{A}(\specDNF)
 \end{align*}
@@
 %%
-More formally, $\specDNF \subseteq \textit{spec}$, defined:
-%%
-\begin{align*}
-\specDNF{} & ::= [\textit{mult},] \;
-\tySum{\specDNF} \\
+More formally, $\specDNF \subseteq (\textit{spec} \cup \{\epsilon\})$, defined:
+%%
+\begin{align*}
+\specDNF{} & ::= \tySum{\specDNF} \mid \epsilon \\
 \tySum{\specDNF} & ::= \tyProd{\specDNF} \; \term{+} \; \tySum{\specDNF} \mid
@@
 %%
+We incldue the empty element $\epsilon$ which is not present in the
+syntax. 
 In the following, $s, t$ range over elements of
@@
 
+\newcommand{\exactR}{\mathsf{exact}\;}
+\newcommand{\exact}[1]{\exactR(#1)}
+\newcommand{\upperB}[1]{\mathsf{upper}\;(#1)}
+
 \begin{definition} The $\mathsf{toSpec}$ constructor is defined:
@@
 &\mathsf{toSpec} \; \llower \; \lupper = \\
-&\begin{cases}
+&\hspace{-0.4em}\begin{cases}
 %% FWD
- \stenFwdS{\lupper}{i}  & \llower = 0 \, \wedge \, \lupper > 0 \\[0.2em]
- \stenFwdSR{\lupper}{i}{\irreflS}  & \llower = 1 \, \wedge \, \lupper > 0 \\[0.2em]
+\exact{\stenFwdS{\lupper}{i}}  & \llower = 0 \, \wedge \, \lupper > 0 \\[0.2em]
+\exact{\stenFwdSR{\lupper}{i}{\irreflS}}\!\!\!\! & \llower = 1 \, \wedge \, \lupper > 0 \\[0.2em]
 %% BWD
- \stenBwdS{|\llower|}{i}  & \llower < 0 \, \wedge \, \lupper = 0 \\[0.2em]
-\stenBwdSR{|\llower|}{i}{\irreflS} & \llower < 0 \, \wedge \, \lupper = -1 \\[0.2em]
+\exact{\stenBwdS{|\llower|}{i}}  & \llower < 0 \, \wedge \, \lupper = 0 \\[0.2em]
+\exact{\stenBwdSR{|\llower|}{i}{\irreflS}}\!\!\!\! & \llower < 0 \, \wedge \, \lupper = -1 \\[0.2em]
 %% CEN
- \stenCenS{\lupper}{i}  & \llower < 0 \, \wedge \, \lupper > 0
- \, \wedge \, |\llower| = \lupper \\[0.2em]
- \stenBwdS{|\llower|}{i} & \llower < 0 \, \wedge \, \lupper > 0  \, \wedge \, |\llower| \neq \lupper \\
-\quad \texttt{+} \, \stenFwdS{\lupper}{i}  & \qquad\quad \\[0.2em]
+\exact{\stenCenS{\lupper}{i}}& \llower < 0 \wedge \lupper > 0 \wedge |\llower| = \lupper \\[0.2em]
+\exactR(\stenBwdS{|\llower|}{i} & \llower < 0 \wedge \lupper > 0  \wedge |\llower| \neq \lupper \\
+\qquad\qquad \texttt{+} \, \stenFwdS{\lupper}{i}) & \\[0.2em]
 % BOUND
- \texttt{atMost}, \stenFwdS{\lupper}{i} & \llower > 1 \\[0.2em]
- \texttt{atMost}, \stenBwdS{|\llower|}{i} \quad & \llower < (-1)
+\upperB{\stenFwdS{\lupper}{i}} & \llower > 1 \\[0.2em]
+\upperB{\stenBwdS{|\llower|}{i}} \quad & \llower < (-1)
  \\[0.2em]
 % ABSOLUTE REP
- \epsilon & \llower = \infty \, \vee \, \lupper = \infty
+\exact{\epsilon} & \llower = \infty \, \vee \, \lupper = \infty
   \end{cases}
@@
 
-Finally, in the case where either bound is $\infty$, representing a non
+Finally, if either bound is $\infty$, representing a non
 neighbourhood index, then an empty specification is given
@@
 \begin{lemma}
-$\mathsf{toSpec}$ is total (\ie{}, defined for all inputs)
+$\mathsf{toSpec}$ is total 
  given the precondition that the lower
@@
 %\newcommand{\mkBig}[2]{\overset{\LARGE{\term{#1}}}{#2}}
-\begin{definition} The $\odot$ operation on \specDNF{} is defined
-\begin{align*}
-& ([\textit{approx}_1] \; [\textit{mult}_1] \; 
-\tySum{s})
-\odot
-([\textit{approx}_2] \; [\textit{mult}_2] \;
-\tySum{t}) = \\
-& \begin{cases}
-\tySum{s} \tySum{\odot} \tySum{t} & \textit{approx}_1 =
-\textit{approx_2} = \varepsilon
+\begin{definition} The $\odot$ operation on $\textsf{A}(\specDNF{})$
+  is defined in terms of the intermediate $\hat{\odot}$ on \specDNF{}:
+%%
+\begin{align*}
+S \,\hat{\odot}\, \epsilon = S \quad
+\epsilon \,\hat{\odot}\, S = S \qquad
+S \,\hat{\odot}\, T = \bplus_{(\tyProd{s}, \tyProd{t}) \in (S \times
+   T)} (\tyProd{s} \; \term{*} \; \tyProd{t})
+\end{align*}
+%%
+That is, the $\hat{\odot}$ product of two specifications $S$ and $T$ in DNF form 
+is the $\term{+}$ sum of all pairwise $\term{*}$ products for every
+pair of $\tyProd{spec}$ drawn from $S$ and $T$. The result
+is indeed in DNF-form. The operation is
+commutative and associative, and sound with respect to our
+model (see Lemma~\ref{lem:alg-soundness} below).
+The $\hat{\odot}$ operation is then lifted to all combinations of
+\textsf{Approx} for $\odot$:
+\begin{align*}
+\textsf{exact}(S) \odot \textsf{exact}(T) = \; & \textsf{exact}(S
+                                              \hat{\odot} T) \\
+\textsf{lower}(S) \odot \textsf{lower}(T) = \; & 
+      \textsf{lower} \, (S \hat{\odot} T) \\
+\textsf{upper}(S) \odot \textsf{upper}(T) = \; &
+      \textsf{upper} \, (S \hat{\odot} T) \\
+\textsf{lower}(S) \odot \textsf{exact}(T) = \; &
+       %\textsf{exact}(T) \odot \textsf{lower}(S) = \;
+       \textsf{both} \, (S \hat{\odot} T, \; T) \\
+\textsf{upper}(S) \odot \textsf{exact}(T) = \; & 
+       %\textsf{exact}(T) \odot \textsf{upper}(S) =
+      \; \textsf{both} \, (T, \; S \hat{\odot} T) \\
+\textsf{lower}(S) \odot \textsf{both}(T_l,T_u) = \; &
+        %\textsf{both}(T_l,T_u) \odot \textsf{lower}(S) = \;
+        \textsf{both} \, (S \hat{\odot} T_l, \; T_u) \\
+\textsf{upper}(S) \odot \textsf{both}(T_l,T_u) = \; & \textsf{both} \,
+                                                      (T_l, \; S \hat{\odot} T_u) \\
+\textsf{both}(S_l,S_u) \odot
+\textsf{both}(T_l,T_u) & = \textsf{both} (S_l \hat{\odot} T_l, \; S_u
+                         \hat{\odot} T_u)
+\end{align*}
+For brevity we omit the cases where the parameters above are flipped
+as $\odot$ is commutative.
+\end{definition}
+
+\begin{definition}The $\oplus$ operation on $\textsf{A}(\specDNF)$ 
+is defined in a similar way to $\odot$, with $\hat{\oplus}$ on
+$\specDNF$ given by:
+%%
+\begin{equation*}
+S \,\hat{\oplus}\, \epsilon = S \quad
+\epsilon \,\hat{\oplus}\, S = S \quad
+S \, \hat{\oplus} \, T = S \, \term{+} \, T
+\end{equation*}
+%%
+Thus, if neither specification is empty, their $\oplus$ is just
+the syntatic sum $\term{+}$. Then $\oplus$ is the lifting of
+$\hat{\oplus}$ to boudns with the shape as $\odot$ in the previous
+definition.
+\end{definition}
+
+\begin{lemma}[$\hat{\odot}$, $\hat{\oplus}$ soundness] $\forall S,
+  T \in \specDNF$ where $S \neq \epsilon \wedge T \neq \epsilon$:
+\begin{align*}
+\interp{S \; \hat{\odot} \; T}_n & = \interp{S}_n \otimes \interp{T}_n
   \\
-\
-& \textit{approx}_1 = \texttt{atLeast}, \textit{approx}_2 = \varepsilon
-\end{cases}
-\\
-\end{align*}
-\begin{align*}
-s \tySum{\odot} t & =
- \bplus_{(\tyProd{s_1}, \tyProd{s_2}) \in (s \times
-   t)} (\tyProd{s_1} \; \term{*} \; \tyProd{s_2})
-%\begin{cases}%
-%\end{cases}
-\end{align*}
-\end{definition}
+\interp{S \; \hat{\oplus} \; T}_n & = \interp{S}_n \cup \interp{T}_n
+\end{align*}
+\label{lem:alg-soundness}
+\end{lemma}
+%
 

### 2016-07-06 paper.tex

@@
 %
-\item Below we present two match ups of relativised access patterns and unifiers
-  to show both the lower and the upper bound holds.
-
-  The unifications for the upper bound:
-  \begin{align*}
-     \vtwoh{0}{-1} & \sim_2 \vtwoh{0}{\infty} \\
-     \vtwoh{0}{0} & \sim_2 \vtwoh{0}{\infty} \\
-     \vtwoh{0}{1} & \sim_2 \vtwoh{0}{\infty} \\
-     \vtwoh{-1}{0} & \sim_2 \vtwoh{\infty}{0} \\
-     \vtwoh{1}{0} & \sim_2 \vtwoh{\infty}{0}
-  \end{align*}
-
-  The unifications for the lower bound:
-  \begin{align*}
-     \vtwoh{-1}{0} & \sim_2 \vtwoh{-1}{\infty} \\
-     \vtwoh{0}{0} & \sim_2 \vtwoh{0}{\infty} \\
-     \vtwoh{1}{0} & \sim_2 \vtwoh{1}{\infty}  \\
-     \vtwoh{0}{-1} & \sim_2 \vtwoh{\infty}{-1} \\
-     \vtwoh{0}{0} & \sim_2 \vtwoh{\infty}{0} \\
-     \vtwoh{0}{1} & \sim_2 \vtwoh{\infty}{1}
-  \end{align*}
-
-  Since both predicates are satisfied the array accesses specified on line three
-  are consistent with the \texttt{fivepoint} specification.
+\item Below we present two sets of generalisations to show both the lower and
+  the upper bound holds.
+
+  The generalisations for the upper bound:
+  \begin{align*}
+     \vtwoh{0}{-1} & \preceq_2 \vtwoh{0}{\infty} \\
+     \vtwoh{0}{0} & \preceq_2 \vtwoh{0}{\infty} \\
+     \vtwoh{0}{1} & \preceq_2 \vtwoh{0}{\infty} \\
+     \vtwoh{-1}{0} & \preceq_2 \vtwoh{\infty}{0} \\
+     \vtwoh{1}{0} & \preceq_2 \vtwoh{\infty}{0}
+  \end{align*}
+
+  The generalisations for the lower bound:
+  \begin{align*}
+     \vtwoh{-1}{0} & \preceq_2 \vtwoh{-1}{\infty} \\
+     \vtwoh{0}{0} & \preceq_2 \vtwoh{0}{\infty} \\
+     \vtwoh{1}{0} & \preceq_2 \vtwoh{1}{\infty}  \\
+     \vtwoh{0}{-1} & \preceq_2 \vtwoh{\infty}{-1} \\
+     \vtwoh{0}{0} & \preceq_2 \vtwoh{\infty}{0} \\
+     \vtwoh{0}{1} & \preceq_2 \vtwoh{\infty}{1}
+  \end{align*}
+
+  Since both predicates are satisfied the array accesses specified on line
+  three, the stencil comutation is consistent with the \texttt{fivepoint}
+  specification.
 \end{enumerate}
 
-Now when $L3$ is slightly changed with an additional term \texttt{b(i+2,j)}, the
-new assignment $L3'$ becomes:
+Now when $L3$ is slightly changed with an additional term \texttt{b(i+2,j+2)},
+the new assignment $L3'$ becomes:
 %
@@
 \begin{minted}{fortran}
-!= stencil atLeast, readOnce, fivepoint :: b}
-\end{minted}
-%
-The consistency function would have held as
-$\mathit{cons}(\textsf{mult}(\textsf{up}(m)),L3',b)$ for the upper bound would
-have never been needed. This semantically makes sense since the region
-$\texttt{fivepoint}$ is indeed a lower bound for the access pattern of $b$.
+!= stencil atLeast, readOnce, fivepoint :: b
+\end{minted}
+%
+The consistency predicate would have held because for the upper bound
+$\mathit{cons}(\textsf{mult}(\textsf{up}(m)),L3',b)$ would not have
+been needed. This semantically makes sense since the region
+$\texttt{fivepoint}$ is indeed contained in the regions touched by $b$.
+
+\subsection{Synthesis\label{subsec:synthesis}}
 

### 2016-07-06 paper.tex
full set of inequations

@@
 \newcommand{\effdims}[2]{\mathit{constr}(#1)_{#2}}
+
+\newcommand{\trule}[1]{\footnotesize{#1}}
 
@@
 \begin{array}{c}
-\framebox{$\textit{region} \equiv \textit{region}'$} \\[0.7em]
+%\framebox{$\textit{region} \equiv \textit{region}'$} \\[0.7em]
 \begin{array}{rll}
@@
 \end{align*}
-\caption{Equations on specifications}
+\caption{Equations on regions, $\textit{region} \equiv \textit{region}'$}
 \label{fig:equations}
@@
 \begin{array}{c}
-\dfrac{}{S <: S}(\textsc{refl}) \qquad \dfrac{R <: S \quad S <: T}{R <:
-  T}(\textsc{trans}) \\[1.5em]
-\dfrac{}{\texttt{readOnce} \, S <: S}(\textsc{rep})
+%%
+%% INEQUATIONS ON REGION
+%% 
+%\framebox{$\textit{region} <: \textit{region}'$} \\[0.85em]
+%
+\dfrac{}{R <:^r R}\trule{\textsc{refl}} \qquad \dfrac{P <:^r R \quad R <:^r Q}{P <:^r
+  Q}\trule{\textsc{trans}} 
+\qquad
+\dfrac{R \equiv S}{R <:^r S}\trule{\textsc{eq}} 
 \\[1.5em]
 \setlength{\arraycolsep}{0.1em}
-\dfrac{\hspace{3em} m \leq n \hspace{3em}}
+\dfrac{\hspace{3em} m \leq n \qquad r \Rightarrow s \hspace{3em}}
 {\begin{array}{rl}
-\stenFwdS{m}{d} & <: \stenFwdS{n}{d}
+\stenFwdSR{m}{d}{r} & <:^r \stenFwdSR{n}{d}{s}
 %& \;\, ds \subseteq es \wedge n \leq m
 \\
-\wedge \stenBwdS{m}{d} & <: \stenBwdS{n}{d}
+\wedge \; \stenBwdSR{m}{d}{r} & <:^r \stenBwdSR{n}{d}{s}
 %& \;\, ds \subseteq es \wedge n \leq m
 \\
-\wedge \; \stenCenS{m}{d} & <: \stenCenS{n}{d}
+\wedge \; \stenCenSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s}
 %& \;\, ds \subseteq es \wedge n \leq m
 \\
-\wedge \; \stenBwdS{m}{d} & <: \stenCenS{n}{d} \\
-\wedge \; \stenFwdS{m}{d} & <: \stenCenS{n}{d} \\[1em]
-\end{array}} \\[2.5em]
+\wedge \; \stenBwdSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s} \\
+\wedge \; \stenFwdSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s} \\[1em]
+\end{array}}\trule{\textsc{over}} \\
 \hspace{-0.5em}
-\dfrac{S_1 <: T_1 \quad S_2 <: T_2}
-      {S_1 \, \texttt{+} \, S_2 <: T_1 \, \texttt{+} \, T_2}
-(\textsc{cong}\texttt{+}) \;\;\;
-\dfrac{S_1 <: T_1 \quad S_2 <: T_2}
-      {S_1 \, \texttt{*} \, S_2 <: T_1 \, \texttt{*} \, T_2}
-(\textsc{cong}\texttt{*})
+\dfrac{P_1 <:^r R_1 \quad P_2 <:^r R_2}
+      {P_1 \, \texttt{+} \, P_2 <:^r R_1 \, \texttt{+} \, R_2}
+\trule{\textsc{cong}\texttt{+}} \;\;\;
+\dfrac{P_1 <:^r R_1 \quad P_2 <:^r R_2}
+      {P_1 \, \texttt{*} \, P_2 <:^r R_1 \, \texttt{*} \, R_2}
+\trule{\textsc{cong}\texttt{*}} \\[1.5em]
+%%
+%% INEQUATIONS ON SPEC
+%%
+%\framebox{$\textit{spec} <: \textit{spec}'$} \\[0.8em]
+%
+\dfrac{}{\term{readOnce} \, R <: R}\trule{\textsc{rep}}
+\quad
+\dfrac{}{\term{atLeast} \, R <: \term{atMost} \,
+  R}\trule{\textsc{bound}} \\[1.5em]
+%%
+\dfrac{P <:^r R}{\term{atLeast} \, P <: \term{atLeast} \,
+  R}\trule{\textsc{cong}\texttt{L}}
+\ 
+\dfrac{P <:^r R}{\term{atMost} \, P <: \term{atMost} \,
+  R}\trule{\textsc{cong}\texttt{M}} \\[1.5em]
+\dfrac{P <:^r R}{\term{readOnce} \, P <: \term{readOnce} \,
+  R}\trule{\textsc{cong}\texttt{R}} 
+%%
 \end{array}
 \end{align*}
-\caption{Inequations on regions}
+\caption{Inequations $<:^r$ on \textit{regions} and
+$<:$ on \textit{specs}.}
 \label{fig:inequations}
@@
 
-Figure~\ref{fig:inequations}\dnote{TODO: this might
-need to include notions of consistency}
+Figure~\ref{fig:inequations} defines a notion of sub-specification
+via the relation $<:^r$ on \textit{region} 
+syntax and then $<:$ on \textit{spec} syntax inductively. 
+Both relations are reflexive and transitive (we omit 
+these equations for $<:$ on \textit{spec}), and congruent
+with respect to \term{+}, \term{*}, \term{atLeast}, \term{atMost},
+and \term{readOnce}. The (\trule{\textsc{eq}}) axiom connects
+equations with inequations. The (\trule{\textsc{over}}) explains
+the notion of spatial over approximation via overlapping regions.
+The (\trule{\textsc{rule}}) allows the modifier that
+enforces that every index is read only once to be dropped
+as an approxiamtion. The (\trule{\textsc{bound}}) gives
+a connection between the \term{atLeast} and \term{atMost}
+approximation modifiers.
 
@@
 specifications and for synthesising programs from specifications.
-Checking and synthesis use the set-based semantic model of
-Section~\ref{sec:semantics}.
-
-\paragraph{Syntax functions}
-The function \rhsExp{} maps statements to a
+These build on the model of Section~\ref{sec:semantics}.
+
+\paragraph{Syntax helper functions}
+Let \rhsExp{} map source language statements to a
 set of expressions which occur in right-hand positions (\ie{}, not the
-target of an assignment). The function \var{} maps expressions to a
+target of an assignment). Let \var{} map expressions to a
 set of the variables used in right-hand positions.

### 2016-07-06 paper.tex
use subcaption instead of obsolete package subfigure
added subsubsection: "Example: 2-D Jacobi iteration"

@@
 \usepackage{fancyvrb}
-\usepackage{subfigure}
 \usepackage{amsthm}
@@
 \usepackage{breqn}
+\usepackage{subcaption}
 \usepackage{siunitx} % For pretty-printing numeric values and SI units
@@
 
+\subsubsection{Example: 2-D Jacobi iteration}
+
+One relatively basic example of a stencil program is two dimensional
+Jacobi iteration that repeatedly goes through each cell in a matrix
+and computes the average value of the four adjacent cells. The kernel
+can be expressed in a line of Fortran code,
+\begin{minted}{fortran}
+  a(i,j) = (a(i-1,j)+a(i+1,j)+a(i,j+1)+a(i,j-1))/4
+\end{minted}
+and Camfort is able to infer a precise stencil specification for this
+line,
+\begin{minted}{fortran}
+  != stencil reflexive(dim=1)*centered(depth=1, dim=2, irreflexive) + reflexive(dim=2)*centered(depth=1, dim=1, irreflexive) :: a
+\end{minted}
+which when inserted back into the program can be checked by Camfort as
+well. We then tested the error-detecting ability of Camfort by
+generating a family of 6,561 Jacobi iteration programs by perturbing
+the indices used in the following manner: each access of the array was
+templated as \mintinline{fortran}{a(i+ki,j+kj)} and then all possible
+combinations of \mintinline{fortran}{ki} and \mintinline{fortran}{kj}
+were drawn from the set $\{-1,0,1\}$. There are four uses of
+\mintinline{fortran}{ki} and four of \mintinline{fortran}{kj}, hence
+$3^8=6561$ possible perturbations of this kind. This particular setup
+tests the class of errors in which the programmer typed an incorrect
+number but at least got the ordering of the indices correct, although
+Camfort would also detect a mixup of variables in this case as well.
+
+Camfort checked all 6,561 programs and found that only 24 of them were
+valid implementations of the stencil specification. This was the
+expected result because there are $4!$ possible correct permutations
+of the four terms in the kernel. All of the other programs were found
+to violate the specification in some way. Another way to test the
+programs is through visual examination of their output in graphical
+format. Figure~\ref{fig:jacobi} shows some of the results of running
+the Jacobi iteration programs, rendered into image files. All of the
+correct variants emitted images indistinguishable from
+Figure~\ref{fig:jacobicorrect}, while all of the incorrect variants
+had wildly different image outputs. Both the ``visual test'' and a
+byte-by-byte comparison showed results that completely concurred with
+the results of the Camfort test.
+
+\begin{figure}[t]
+  \centering
+  \subcaptionbox{\label{fig:jacobicorrect}}{\includegraphics[width=0.1\textwidth]{jacobi-correct.png}}
+  \subcaptionbox{}{\includegraphics[width=0.1\textwidth]{jacobi-incorrect1.png}}
+  \subcaptionbox{}{\includegraphics[width=0.1\textwidth]{jacobi-incorrect2.png}}
+  \subcaptionbox{}{\includegraphics[width=0.1\textwidth]{jacobi-incorrect3.png}}
+  \caption{Greyscale output of several perturbations of a Jacobi
+    iteration program. The correct output is shown in
+    (\subref{fig:jacobicorrect}).}
+  \label{fig:jacobi}
+\end{figure}
+
 % != stencil readOnce, (forward(depth=1, dim=3, irreflexive))*(backward(depth=1, dim=1))*(backward(depth=1, dim=2)) + (forward(depth=1, dim=3))*(backward(depth=1, dim=1))*(reflexive(dim=2)) + (forward(depth=1, dim=3))*(backward(depth=1, dim=1, irreflexive))*(backward(depth=1, dim=2, irreflexive)) + (forward(depth=1, dim=3))*(backward(depth=1, dim=2))*(reflexive(dim=1)) + (backward(depth=1, dim=1))*(backward(depth=1, dim=2, irreflexive))*(reflexive(dim=3)) + (backward(depth=1, dim=1, irreflexive))*(backward(depth=1, dim=2))*(reflexive(dim=3)) :: cx_rho_centre

### 2016-07-06 paper.tex
completed soundness subsection

@@
 where $\mathsf{IV}(e)$ is the set of free induction variables
-appearing in an expression, and $\neigh{}$ classifies expressions which 
-are constant offsets from an induction variable
+ in an expression and $\neigh{}$ classifies expressions which 
+are constant offsets on an induction variable
 (Defn.~\ref{def:neighbour}). In the third case, 
@@
 
-For example $\textsf{schematic}(i, j+1) = \vtwoh{0}{1}$ assuming
+For example, assuming
 $i$ and $j$ are induction variables and $\textsf{schematic}(i-1, 2, j)
@@
 \begin{example}
-Take three array indexing terms
-$(i, j+1)$, $(1, j+1)$ and $(3, j)$.
-As given by \textsf{schematic}, these have the 
-corresponding schemes: $\vtwoh{0}{1}$, $\vtwoh{\infty}{1}$
+Three array indexing terms
+$(i, j+1)$, $(1, j+1)$ and $(3, j)$
+are mapped by \textsf{schematic} to the 
+index schemes: $\vtwoh{0}{1}$, $\vtwoh{\infty}{1}$
 and $\vtwoh{\infty}{0}$ respectively. The following
@@
 \begin{align*}
-& \interp{\stenCenS{2}{1}}_2 \\
-= \; & \{i\singleEntry{1}{2} \,|\, i \in \{\text{-}2,..., 2\}\!\setminus\!0 \}
+\interp{\stenCenS{2}{1}}_2 = \; & \{i\singleEntry{1}{2} \,|\, i \in \{\text{-}2,..., 2\}\!\setminus\!0 \}
   \cup  \{ \zeroEntry{1}{2} \,|\, \textit{true} \} \\
@@
 
-\begin{theorem}[Soundness of approximations]
+\begin{theorem}[Soundness of inequations
+  (Figure~\ref{fig:inequations})]
+
+Let $\leq$ be an ordering on
+$\textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$
+and $\leq'$ be an ordering on
+$\textsf{Approx}(\mathcal{P}(\domainVal{}^n))$ defined by the
+following caluses
+\begin{align*}
+(M \leq' N \Rightarrow \textit{inj} \, M \leq \textit{inj} \, N) &
+\qquad 
+(\textsf{once} \, M \leq' \textsf{mult} \, M) \\
+%%
+%%
+(M \subseteq N \Rightarrow \textit{inj}' M \leq' \textit{inj}' N)
+& \;\;\; (\textsf{exact} M \leq  \textsf{lower} M)
+\;\;\; (\textsf{lower} M \leq  \textsf{upper} M)
+\end{align*}
+%%
+where $\textit{inj}$ ranges over the unary injections
+of $\textsf{Approx}$ and $\textit{inj}'$ ranges over the injections of
+$\textsf{Mult}$. Then, soundness of $<:$ is that:
+%%
 \[
-\overline{\texttt{v} : S} <: \overline{\texttt{u} : T}
+S <: T
 \; \Rightarrow \;
-\interp{\overline{\texttt{v} : S}} \subseteq \interp{\overline{\texttt{u} : T}}
+\interp{S}_n \leq \interp{T}_n
 \]

### 2016-07-06 paper.tex
actually its line 5 that swaps a and b...
fixed a typo
typos

@@
 the value at \texttt{b(i)} from a neighbourhood of elements about
-\texttt{a(i)}. Line 4 swaps
+\texttt{a(i)}. Line 5 swaps
 \texttt{a} and \texttt{b} between iterations, where \texttt{b} becomes the
@@
 %%
-There are two modifiers one used at specification the other at region
+There are two modifiers: one used at specification and the other at region
 level to provide better descriptions for stencil computations.
@@
 However, this would be an invalid specification if any of the
-array subscripts was repeated. This modifier provides a way to
+array subscripts were repeated. This modifier provides a way to
 rule out any accidental repetition of array subscripts.
@@
 over booleans denoting the absence (true) or presence (false)
-of \irrefl{} which are combined via disjunction. For example,
+of \irrefl{}. These are combined via disjunction. For example,
 the rule $(\textsc{B} \, \term{+} \,
@@
 
-Figure~\ref{fig:inequations} defines a notion of sub-specification
+Figure~\ref{fig:inequations} defines a notion of sub-specifications
 via the relation $<:^r$ on \textit{region} 
@@
 
+\anote{Can we say something here about why these equations are useful i.e. we use them to do simplification of stencils?}
+
 \section{Semantics of specifications; a model}
@@
 semantics, in terms of sets of vectors which we call \emph{index
-  schemes}. We introduces these first then define the denotational semantics.
+  schemes}. We introduce these first and then define the denotational semantics.
 
@@
 terms is given precisely by the partial function, \textsf{schematic}, which
-maps syntactic array index terms to indexing schemes. This
+maps syntactic array-index terms to indexing schemes. This
 operation forms the basis of the analysis required for the inference
@@
 Our terminology of \emph{indexing schemes} and
-\emph{generalisation} are by analogy
+\emph{generalisation} is by analogy
 with the \emph{type schemes} and \emph{type generalisation} of
@@
 much more rigid however, as our schemes do not bind names
-and thus generalisation of does not require substitution; $\infty$ acts
+and thus generalisation does not require substitution; $\infty$ acts
 as a positional placeholder.
@@
 selecting from $u$ for $1$ and $v$ for $0$. The pairwise permutation
-matrix  s thus defined in terms of the logical
+matrix is thus defined in terms of the logical
 matrix $b$ where $b_{i,j}$ is the $j$-th bit of the integer

### 2016-07-06 paper.tex
fix macro and some editing to intro

@@
 
-\newcommand{\trule}[1]{\footnotesize{#1}}
+\newcommand{\trule}[1]{{\footnotesize{#1}}}
 
@@
   is highly decoupled from the implementation. We have been studying
-  specification techniques in middle ground between low-level and
+  specification techniques in the middle ground between low-level and
   high-level specifications for numerical code. In this paper, we
@@
   regular shape which is amenable to static analysis and simple
-  specification. We describe a abstract specification language for
-  stencil computations that can be directly embedded as annotations in
+  specification. We describe a abstract spatial specification language for
+  stencil computations that can be embedded as annotations in
   source code, along with an inference and checking procedure. We
@@
   static semantics, inference, checking, and program synthesis
-  procedures, our implementation, and our evaluation study. We provide
-  an implementation of ontop of CamFort, an
-  open-source program analysis and refactoring tool for Fortran corpus.
+  procedures, our implementation, and our evaluation study. 
 \end{abstract}
@@
 scientific and numerical computing applications. Informally, a stencil
-computation computes an array value, where the value at each index $i$ of
+computation outputs an array, where the value at each index $i$ of
 this array is calculated from a \emph{neighbourhood} of values around $i$ in
@@
 processing, approximations to differential equations. For example, the
-following iteration computes the one-dimensional discrete Laplace
+following computes the one-dimensional discrete Laplace
 transform (an approximation to a derivative) in Fortran:
@@
 %$\texttt{b}$.
-In this simple example, the dependency between \texttt{a}
-and \texttt{b} on line 3 forms a simple spatial relationship which is easily
-understood. This spatial relationship determines other aspects of the
+The shape of the data access pattern on line 3 determines other aspects of the
 program and its efficient implementation: how much ``boundary'' is
 needed for the array, the most cache-efficient layout in memory,
-the partitioning shape for parallel implementations.
-
-More complex stencil computations can be much harder to understand and
-subsequently more prone to error. For example,
+and the partitioning shape for parallel implementations.
+In this example, the access pattern is simple, static, and easily
+understood. More complex stencil computations 
+are much more prone to error from simple lexical mistakes and
+harder to understand. For example,
 Figure~\ref{ref:navier-stokes-fragment} shows three lines from a
@@
 algorithms for stencil specifications (Section~\ref{sec:analysis});
-\item we evaluate our implementation of the approach
-in the CamFort tool for Fortran verification, studying
-a number of example programs to assess the usefulness
-of this approach.
+\item we evaluate our approach, providing an 
+implementation as an extension of CamFort,
+an open-source program analysis and refactoring tool for Fortran
+corpus. We use this implementation to evaluate our approach against
+a corpus of scientific applications, totalling roughly one million
+lines of code, and we study some verification case studies.
 \dnote{insert results here}

### 2016-07-06 paper.tex
typos

--- 2016-07-06 paper.tex
@@
 
-This procedure is iterated till a fixed point is reached. In this
+This procedure is iterated until a fixed point is reached. In this
 example this is reached in the first step. The final spans
@@
 The algorithm to create contiguous spans, covering the space of
-indices, then proceeds as follows.
-Firstly, for all $v \in \mathsf{dom}(M)$ then a new map $M'$ is
+indices, now proceeds as follows.
+Firstly, for all $v \in \mathsf{dom}(M)$ a new map $M'$ is
 created where each vector is mapped to the trivial \emph{unit span} is
@@
 %%
-A fixed point is then compute for the following
-procedure \spanOp which, for per variable
+A fixed point is then computed for the following
+procedure \spanOp which, per variable
 in the map, coalesces spans into contiguous regions in
@@
 for a permutation $\pi$. For each permutation function $\pi^n_i$
-(the $i$-th permutation for vectors of size $n$, we pair the
+(the $i$-th permutation for vectors of size $n$) we pair the
 permutation function with the set of permuted spans so that
@@
 specifications by $\odot$. The resulting specifications for
-each span are then summed by by $\oplus$, that is, this step is given by:
+each span are then summed by $\oplus$:
 \begin{equation*}
@@
 %%
-We incldue the empty element $\epsilon$ which is not present in the
+We include the empty element $\epsilon$ which is not present in the
 syntax. 
@@
 
-A specification is first synthesised into its model. Then occurrences of an
+A specification is first synthesised into its model. Then the occurrences of an
 array in an assignment are checked against the model for consistency as defined
@@
 
-Next, consistency predicate considers whether the stencil is an approximation.
+Next, a consistency predicate considers whether the stencil is an approximation.
 Semantically, a lower bound is achieved if every indexing scheme in the
@@
 %
-\item Now $cons$ function is applied with this model along side with the
+\item Now the $cons$ predicate is applied with this model alongside the
   assignment on line three and array $b$. For brevity the assignment is
-  referred as $L3$ and the set of unifiers defined in the previous step as
+  referred to as $L3$ and the set of unifiers defined in the previous step as
   $m$.
@@
 %
-  The former condition holds as no access pattern is repeated in $L3$ by simple
+  This condition holds as no access pattern is repeated in $L3$ by simple
   inspection.
@@
 \item To evaluate the two predicates above, we need indexing schemes for the
-  array in question. This is obtained by $analyse$ function from
+  array in question. This is obtained by the $analyse$ function from
   Subsection~\ref{subsec:inference}.
@@
 
-Now when $L3$ is slightly changed with an additional term \texttt{b(i+2,j+2)},
+Now when $L3$ is modified with an additional term \texttt{b(i+2,j+2)},
 the new assignment $L3'$ becomes:
@@
 %
-The first four steps will proceed the same way. In the fifth step, result
-of $\mathit{analyse}(L3',b)$ will have the extra term $\vtwoh{2}{2}$. In the
+The first four steps proceed as before. In the fifth step, the result
+of $\mathit{analyse}(L3',b)$ will have an extra term $\vtwoh{2}{2}$. In the
 sixth step, since the model has not changed and the indexing schemes are only
@@
 
-Note that if the specification was as follows:
+Note that if the specification were as follows:
 %

### 2016-07-06 paper.tex
added more detail to evaluation

@@
 \label{sec:evaluation}
-We evaluated our stencil language on a corpus of scientific code. We first give an overview of the corpus itself before presenting our results.
+We evaluated the stencil language on a corpus of scientific code. We first give an overview of the corpus itself before presenting our results.
 
@@
 
-\textbf{Navier} is a smaller projecter, giving a discrete
+\textbf{Navier} is a small project, giving a discrete
 approximation to the two-dimensional Navier-Stokes fluid equations,
@@
 
-\textbf{CP} consists of the example code from the book ``Computational Physics''~\cite{nicholas2006computational} introducing numerical techniques and their application to modern physics problems such as fields, waves, statistical mechanics and quantum mechanics.
+\textbf{CP} consists of the example code from the second edition of the book ``Computational Physics''~\cite{nicholas2006computational} introducing numerical techniques and their application to modern physics problems such as fields, waves, statistical mechanics and quantum mechanics.
 
@@
 Section~\ref{subsec:inference}. Each potential statement for which we
-eventually generate some stencil specifications for is also counted as
+eventually generate some stencil specifications is also counted as
 an \emph{actual} statement.
@@
 indexing variable (\mintinline{fortran}{x}) is derived from an
-induction variable \mintinline{fortran}{i} e.g. \mintinline{fortran}{x
+induction variable (\mintinline{fortran}{i}) as in \mintinline{fortran}{x
   = len - i}; 3) \textbf{Inconsistent induction dimensions} occur when
@@
 
-We grouped stencils together into four classes based on similar levels of complexity. These four classes account for 99.7\% of all the stencils we found in the corpus.
-
-\textbf{All-reflexive} The vast majority (X\%) of stencils were
+We grouped stencils together into four classes based on similar levels of complexity. These four classes account for 99.4\% of all the stencils we found in the corpus.
+
+\textbf{All-reflexive} The vast majority (66\%) of stencils were
 reflexive in every dimension of the array. These most often correspond
@@
 
-\textbf{Some-reflexive} stencils consist of only reflexive terms but
-leave some dimensions unspecified. Two examples in this class are listed:
+\textbf{Some-reflexive} stencils (24\%) consist of only reflexive terms but
+leave some dimensions unspecified. Here are two examples:
 \begin{minted}{fortran}
@@
 implicit loop, and for the second example \mintinline{fortran}{J} is
-the induction variable. Arrays with this class of stencil can be
-indexed by a subset of the LHS induction variables but again without
+the induction variable. This stencil permits array indexing by a subset of the LHS induction variables but again without
 any offsets.
 
-\textbf{Single-action} stencils consist of a single \term{backward},
-\term{forward}, or \term{centered} term and some number of reflexive
-terms.
+\textbf{Single-action} stencils (6\%) consist of a single \term{backward},
+\term{forward}, or \term{centered} region constant and some number of \term{reflexive}
+region constants.
 % /home/mrd45/src/corpus/geos-chem/GTMM/CasaRegridModule.F90:645
@@
 \end{minted}
-
-\textbf{Single-action-irreflexive} stencils consist of a single
+This stencil requires a neighbourhood index including zero-offset in one dimension only. The remaining dimensions must either have no offset or not pertain to an induction variable.
+
+\textbf{Single-action-irreflexive} stencils (3\%) ) consist of a single
 \term{backward}, \term{forward}, or \term{centered} irreflexive term
@@
 \mintinline{fortran}{cldfmc} flows through onto the line with the
-stencil.
+stencil. Stencils of this type required a neightbourhood index without a zero-offset in one dimension only. The remaining dimensions must either have no offset or not pertain to an induction variable.
+
+The remaining stencil shapes (less than 1\%) consisted primarily of multiple \term{forward}, \term{backward}, or \term{centered} dimensions.
+We note that only 0.2\% of stencils found required use of the union (\term{+}) operator.
+Despite this low frequency these stencils are arguably the most important since they represent more complicated code and thus greater potential for programming errors.
 
@@
 \label{sec:discussion}
+
+\anote{Any discussion left to do? - something on checking perhaps? access stencils?}
 

### 2016-07-06 paper.tex
context about science added

@@
 \usepackage{subcaption}
+%\usepackage{subfigure}
 \usepackage{siunitx} % For pretty-printing numeric values and SI units
@@
   high-level specifications for numerical code. In this paper, we
-  focus on \emph{stencil computations}---a common idiom in numerical code, but
-  one which is error-prone due to fine-grained indexing errors. We
-  hypothesise that in practise, stencil computations tend to have a
-  regular shape which is amenable to static analysis and simple
-  specification. We describe a abstract spatial specification language for
-  stencil computations that can be embedded as annotations in
-  source code, along with an inference and checking procedure. We
-  evaluate our language against a corpus of numerical Fortran code,
-  and show that precise specifications can be given to roughly half of
-  the identifiable stencil computations, and that the vast majority of
+  focus on \emph{stencil computations}---a common idiom in numerical
+  code, but one which is error-prone due to fine-grained indexing
+  errors. We hypothesise that in practise, stencil computations tend
+  to have a regular shape which is amenable to static analysis and
+  simple specification. We describe a abstract spatial specification
+  language for stencil computations that can be embedded as
+  annotations in source code, along with inference, checking, and
+  specification and program synthesis procedures. We evaluate our
+  language against a corpus of numerical Fortran code, and show that
+  precise specifications can be given to roughly half of the
+  identifiable stencil computations, and that the vast majority of
   stencil computations indeed have a simple, regular, static
@@
 
-In this work, we introduce a simple specification language for the
-spatial properties of stencil computations. The
-specifications abstract over the fine grained detail of stencil access
-patterns. In practice, most stencil computations have a regular shape
-that can be described simply and abstractly, with a small set of
-coarse-grained descriptions. In the case of our first Laplace example,
-our inference procedure provides the specification:
-%
-\begin{minted}{fortran}
+In practice, most stencil computations have a regular shape that can
+be described simply and abstractly with a small set of coarse-grained
+spatial descriptions. We introduce a simple specification language of
+such descriptions, abstracting over the fine-grained detail of stencil
+access patterns. For our first Laplace example, our
+inference procedure provides the following specification to line 3:
+%
+\begin{minted}[linenos=false]{fortran}
 !=  stencil centered(depth=1, dim=1) :: a
@@
 This explains that \texttt{a} is read from with a
-symmetrical stencil pattern (``centered'') to a depth of one in each
+symmetrical access pattern (``centered'') to a depth of one in each
 direction in its first (and only) dimension.
@@
 the first dimension and backward to a depth of $1$ in the second
-dimension. Figure~\ref{ref:navier-stokes-fragment}(c) describes these
-pictorially.
+dimension. Figure~\ref{ref:navier-stokes-fragment}(c) illustrates
+the specifications pictorially.
 
@@
 \begin{itemize}
-\item we introduce a specification language for
+\item we introduce our specification language for
 stencil computations that captures many common forms
-of data access pattern, both spatial and temporal (Section~\ref{sec:lang});
-\item we detail inference and checking
-algorithms for stencil specifications (Section~\ref{sec:analysis});
+of data access pattern (Section~\ref{sec:lang})
+along with a denotational model (Section~\ref{sec:semantics});
+
+\item we detail inference, checking, specification synthesis, 
+and program synthesis algorithms for our language
+(Section~\ref{sec:analysis}), which relies on the semantic model;
+
 \item we evaluate our approach, providing an 
@@
 \begin{figure}[t]
-\begin{minted}[firstnumber=20]{fortran}
+\begin{minted}[firstnumber=20,xleftmargin=2em]{fortran}
 du2dx = ((u(i,j)+u(i+1,j))*(u(i,j)+u(i+1,j))+ &
@@
   (v(i,j-1)+v(i+1,j-1))*(u(i,j-1)+u(i,j))- &
-  gamma*abs(v(i,j-1)+v(i+1,j-1))*(u(i,j-1)-u(i,j))) &
-  /(4.0*dely)
+  gamma*abs(v(i,j-1)+v(i+1,j-1))*(u(i,j-1)-u(i,j))) & /(4.0*dely)
 
@@
 %
-\begin{minted}{fortran}
+\begin{minted}[linenos=false,xleftmargin=1em]{fortran}
 != stencil centered(depth=1,dim=1)*reflexive(dim=2) + centered(depth=1,dim=2)*reflexive(dim=1) :: u
@@
 \begin{center}
-\begin{tikzpicture}
+\begin{tikzpicture}[scale=0.85]
 \node at (1.25,1.22) {i, j};
@@
 
+\paragraph{Context}
+
+In recent years, there has been an increased awareness 
+of the need for formal verification techniques in 
+science~\cite{post2005computational,oberkampf2010verification,orchard2014computational}. 
+Our specification language is an approach in this direction, proving 
+a system of lighweight specification for numerical code.
+The aim is to not only provide a way of exposing errors, but also
+for specifying program behaviour to future-proof the code and ease
+its maintenance. In order to increase the likelihood of impact, we aim
+to integrate with existing practises smoothly, hence our
+specifications are integrated as comments and our tool can synthesis
+specifications for the user.
+
+Much of this paper concerns the technical aspects of the language,
+its semantics, its inference, checking, and synthesis procedures.
+The evaluation focuses on its application to real scientific code base
+used actively in research.
+
 \section{Stencil specification language}
@@
 
-Our specification system is based on the observation
+Our specification language is based on the observation
 that most forms of array access in numerical code have
@@
 
-We outline the specification language here. Section~\ref{sec:syntax}
-outlines the syntax. Section~\ref{sec:semantics} defines its semantics
-via a simple set interpretation over indices. Section~\ref{sec:eqs} provides an
-equational theory for specifications via relation ($\equiv$) and a
-theory of approximation via relation ($<:$). These equations are then
-proven sound with respect to the set semantics of Section~\ref{sec:semantics}.
+Section~\ref{sec:syntax} outlines the syntax of specifications. 
+ Section~\ref{sec:eqs} provides an equational
+theory for specifications via relation ($\equiv$) and a notion of
+sub-specification via relation ($<:$).  Section~\ref{sec:semantics}
+gives a semantic model via a set-based interpretation, for which the
+equality and sub-specification relation are proven sound.
 
@@
 in the imperative sense (\ie{}, named binders to memory cells) instead
-of mathematical variables. We make clear a number of standard notions
-used throughout.
+of mathematical variables.
 
@@
     variable} if it is the control variable of a ``for''
-  loop\footnote{Or equivalent for the target language, \eg{},
-    \texttt{do} in Fortran for our implementation}, incremented by $1$ per
+  loop,\footnote{Or equivalent for the target language, \eg{},
+    \texttt{do} in Fortran for our implementation} incremented by $1$ per
 iteration. The variable is marked as a base induction variable
-only within the scope of the loop, and such variables will be ranged
-over by $i$, $j$, $k$. Since we do not use non-base induction
-variables (which are traditionally affine transformations of induction
-variables) we often simply say \emph{induction variable}.
+only within the scope of the loop. Induction variables are ranged
+over by $i$, $j$, $k$ throughout. Since we do not use non-base
+(derived) induction variables we often simply say \emph{induction variable}.
 \end{definition}
 
-\begin{definition}[Array subscript]
+\begin{definition}[Array subscript and index]
   An \emph{array subscript} is an expression that indexes an array
-  (for the purpose or reading or writing an element), which we denote
+  (for the purpose of reading or writing an element), which we denote
   by $v(\bar{e})$ for the source language where $\bar{e}$ is shorthand
@@
     index} is a list $\bar{e}$ where each $e \in \bar{e}$ is defined
-  in terms of some base induction variable.
+  in terms of base induction variables.
 \end{definition}
 
-\begin{definition}[Neighbourhood index and neighbourhood subscripts]
+\begin{definition}[Neighbourhood index and subscript]
   For an array subscript $v(\bar{e})$, we say that $e \in \bar{e}$
   is a \emph{neighbourhood index} if it is of the form
-  $e \equiv i$, $e \equiv i + a$ or $e \equiv i - a$ where $a$ is a
-  constant of integer type. That is, a neighbourhood index is a
+  $e \equiv i$, $e \equiv i + a$ or $e \equiv i - a$ where $a$ is an
+  integer constant. That is, a neighbourhood index is a
   relative index defined as a constant translation of a base induction

### 2016-07-06 paper.tex
conclusion needed...
added some words about introducing errors in the um stencil

@@
 
-% != stencil readOnce, (forward(depth=1, dim=3, irreflexive))*(backward(depth=1, dim=1))*(backward(depth=1, dim=2)) + (forward(depth=1, dim=3))*(backward(depth=1, dim=1))*(reflexive(dim=2)) + (forward(depth=1, dim=3))*(backward(depth=1, dim=1, irreflexive))*(backward(depth=1, dim=2, irreflexive)) + (forward(depth=1, dim=3))*(backward(depth=1, dim=2))*(reflexive(dim=1)) + (backward(depth=1, dim=1))*(backward(depth=1, dim=2, irreflexive))*(reflexive(dim=3)) + (backward(depth=1, dim=1, irreflexive))*(backward(depth=1, dim=2))*(reflexive(dim=3)) :: cx_rho_centre
+\subsubsection{Turbulence calculation}
+
+The Unified Model contains an implementation of the Smagorinsky subgrid-scale model for calculating turbulence.
+The source file in question yields 38 stencils from 340 lines of code.
+We selected one as an example:
+
+\begin{minted}[breakindent=8.1em]{fortran}
+ != stencil readOnce, (forward(depth=1, dim=3, irreflexive))*(backward(depth=1, dim=1))*(backward(depth=1, dim=2)) + (forward(depth=1, dim=3))*(backward(depth=1, dim=1))*(reflexive(dim=2)) + (forward(depth=1, dim=3))*(backward(depth=1, dim=1, irreflexive))*(backward(depth=1, dim=2, irreflexive)) + (forward(depth=1, dim=3))*(backward(depth=1, dim=2))*(reflexive(dim=1)) + (backward(depth=1, dim=1))*(backward(depth=1, dim=2, irreflexive))*(reflexive(dim=3)) + (backward(depth=1, dim=1, irreflexive))*(backward(depth=1, dim=2))*(reflexive(dim=3)) :: x
+\end{minted}
+
+This stencil specifies the access pattern to a 3-dimensional array (which we have renamed to be called \texttt{x}).
+The array is accessed 8 times in the computation.
+We tried introducing indexing errors by changing the offset in one dimension of one of the references to \texttt{x}.
+For example changing \mintinline{fortran}{x(i+1,j,k)} to \mintinline{fortran}{x(i,j,k)} or \mintinline{fortran}{x(i-1,j,k)}.
+There are 48 possible errors of this type of which we found only 6 were not detected by the stencil check.
 % /um/trunk/src/atmosphere/diffusion_and_filtering/turb_smagorinsky.F90
@@
 
+\section{Conclusion}
+
+\anote{We need a conclusion!}
+
 \bibliography{references}

### 2016-07-06 paper.tex
simplifying edits

--- 2016-07-06 paper.tex
@@
     variable} if it is the control variable of a ``for''
-  loop,\footnote{Or equivalent for the target language, \eg{},
+  loop,\footnote{Or equivalent in the source language, \eg{},
     \texttt{do} in Fortran for our implementation} incremented by $1$ per
@@
 
-\begin{definition}[Neighbourhood index and subscript]
+\begin{definition}[Neighbourhood index]
   For an array subscript $v(\bar{e})$, we say that $e \in \bar{e}$
@@
   integer constant. That is, a neighbourhood index is a
-  relative index defined as a constant translation of a base induction
-  variable. The relation $\equiv$ identifies terms up-to commutativity
-  of $+$ and the inverse relation of $+$ and $-$ (\eg{},
+  relative index defined as constant translations of base induction
+  variables. (The relation $\equiv$ hre identifies terms up-to commutativity
+  of $+$ and the inverse relation of $+$ and $-$ \eg{},
   $(-b) + i \equiv i - b$).  We classify neighbourhood indices using
-  the predicate \neigh{}. An array subscript $v(\bar{e})$ where every
- $e \in \bar{e}$ is a neighbourhood index is called a
- \emph{neighbourhood subscript}.
+  the predicate \neigh{}. %An array subscript $v(\bar{e})$ where every
+% $e \in \bar{e}$ is a neighbourhood index is called a
+% \emph{neighbourhood subscript}.
 \label{def:neighbour}
@@
 Figure~\ref{fig:syntax} gives the syntax of stencil specifications,
-which we introduces in stages below.  The top-level is given by the
+which we introduce in stages below.  The top-level is given by the
 \nonterm{specification} production which splits into either a
 \nonterm{regionDec} (region declaration) or a \nonterm{specDec}
-(specification declaration). Specification declarations associate a
-specification to one or more program variables $\bar{v}$ (here given
-using the Fortran style syntax of $\ldots \; \texttt{::} \bar{v}$),
-describing how the array variables $\bar{v}$ are accessed.
+(specification declaration). Specification declarations associate
+(via the operator $\texttt{::}$) a specification to one or more program variables
+$\bar{v}$, describing how the array variables $\bar{v}$ are accessed.
 %Specification declarations are either \nonterm{spatial} or
@@
 region constant except \term{reflexive} is given a depth parameter $n$ (natural
-number greater than 0). They all receive a dimension identifier $d$ (also a
+number greater than 0). Each receives a dimension identifier $d$ (also a
 natural number greater than 0). Region constants specify that an array is read
@@
 neighbourhood indices ranging from $i + 0$ up to $i \pm n$ inclusively,
-\eg{}, the following is valid stencil, reading from \term{a} and writing to
+\eg{}, the following is a valid stencil, reading from \term{a} and writing to
 \term{b}:
@@
 dimension is fixed at a constant here.
-
 A \term{backward} stencil

### 2016-07-06 paper.tex
first pass at a conclusion

--- 2016-07-06 paper.tex
@@
 \textbf{Single-action-irreflexive} stencils (3\%) ) consist of a single
-\term{backward}, \term{forward}, or \term{centered} irreflexive term
-and some number of reflexive terms.
+\term{backward}, \term{forward}, or \term{centered} irreflexive region constant
+and some number of reflexive region constants.
 % /home/mrd45/src/corpus/geos-chem/GeosRad/rrtmg_lw_rtrnmc.F90:355
@@
 
-\section{Conclusion}
-
-\anote{We need a conclusion!}
+\section{Conclusion and future work}
+
+In this paper we have described an annotation language which can be
+used to specify the access pattern of a stencil computation. We
+defined a denotational semantic model for our language and proved its
+soundness. This gave a basis for an equational theory of
+specifications and also justifies our inference and checking
+procedure. We evaluated our approach by implementing the stencil inference and
+checking procedure in Camfort, an open-source analysis tool\footnote{\url{https://github.com/camfort/camfort}} for
+Fortran programs and running the inference process over a corpus of
+1.1 million lines of Fortran code.
+
+Our motivation for this work was to provide an intermediate
+specification language to allow scientists to verify numerical
+computations. The most common types of stencils we found consisted only of
+reflexive region constants which are unlikely to give rise to
+programming errors. However, around 10\% of stencils covered more
+complicated program structures for which indexing errors are much more
+plausible. We showed two examples of more complicated stencils for
+which CamFort was able to detect a significant majority of possible
+indexing errors.
+
+Our definition of a stencil computation meant that we generated no
+specification for a large number of iterated array accesses in the corpus. For
+future work we will be looking at expanding the definition to allow us
+to capture more of these examples.
 

### 2016-07-06 paper.tex
intro editing; subfigure

@@
 \usepackage{breqn}
-\usepackage{subcaption}
-%\usepackage{subfigure}
+%\usepackage{subcaption}
+\usepackage{subfigure}
 \usepackage{siunitx} % For pretty-printing numeric values and SI units
@@
     variable} if it is the control variable of a ``for''
-  loop,\footnote{Or equivalent in the source language, \eg{},
-    \texttt{do} in Fortran for our implementation} incremented by $1$ per
+  loop,\footnote{Or equivalent in the source language \eg{} 
+    \texttt{do} in Fortran for our implementation.} incremented by $1$ per
 iteration. The variable is marked as a base induction variable
@@
 region constant except \term{reflexive} is given a depth parameter $n$ (natural
-number greater than 0). Each receives a dimension identifier $d$ (also a
+number greater than 0). Every region constant receives a dimension identifier $d$ (also a
 natural number greater than 0). Region constants specify that an array is read
@@
 %%
-were \texttt{i} is an induction variable.  The stencil specification
+were \texttt{i} is an induction variable.  The specification
 is associated to the array variable \texttt{a}. Note that the second
-dimension is fixed at a constant here.
-A \term{backward} stencil
-is similar but has negative neighbourhood indices, \eg{},
+dimension is constant.
+
+The \term{forward},
+\term{backward}, and \term{centered} regions can all receive an
+additional optional attribute \term{irreflexive} which marks
+that the region does not include the origin (offset 0).
+For example, the following is a \term{backward} stencil, which is similar
+to \term{forward} but has negative neighbourhood indices, and which
+is irreflexive:
 %
 \begin{minted}{fortran}
-!= stencil backward(depth=2, dim=1) :: a
-b(i) = a(i) + a(i-1) + a(i-2)
+!= stencil backward(depth=2, dim=1, irreflexive) :: a
+b(i) = a(i-1) + a(i-2)
 \end{minted}
@@
 \nonterm{specDec} ::= & \term{stencil} \; \nonterm{spec} \;
-                        \texttt{::} \; v
-  \\
+                        \texttt{::} \; v \\
 \nonterm{regionDec} ::= &  \texttt{region} \; \nonterm{rvar} \; \texttt{=} \;
-                         \nonterm{region}\\[1em]
+                         \nonterm{region}\\[0.4em]
 %\nonterm{spec} ::= & \nonterm{spatial} \mid \nonterm{temporal}
@@
 \dimTy ::= \; & \mathbb{N}_{>0} \\
-\nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+
+\nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+\\[-1em]
 \end{array}
@@
 
-The sum of two regions $r \texttt{+} r'$ specifies that an array is
-read using the neighbourhood indices described by either $r$ or
-$r'$. It can be thought of as a disjunction of specifications. For
-example, the following gives the specification for the five-point
-stencil which is the sum of two compound \texttt{reflexive} and
-\texttt{centered} regions in each dimension:
-%%
-\begin{minted}{fortran}
-!= stencil centered(depth=1, dim=1)*reflexive(dim=2) + centered(depth=1, dim=2)*reflexive(dim=1) :: a
-
-b(i,j) = -4*a(i,j) + a(i-1,j) + a(i+1,j)
-                   + a(i,j-1) + a(i,j+1)
-\end{minted}
-%%
-Here the left hand side of \texttt{+} says when the second dimension
-(induction variable $j$) is fixed at the origin, the first dimension
-(induction variable $i$) accesses immediate vicinity of the origin.
-The right hand side of \texttt{+} is similar but the dimensions are reversed.
-This reflects the symmetry under rotation in the fivepoint stencil.
-
-The product of two regions $r \texttt{*} r'$ specifies that the array
+The product of two regions $r \texttt{*} r'$ specifies that an array
 is read with neighbourhood indices drawn from the bounding box created
 by the regions two regions $r$ and $r'$. For example, the following
-code (defining a nine-point stencil) has a specification as the
+code (defining a \emph{nine-point stencil}) has a specification given by the
 product of two \texttt{centered} regions in each dimension:
@@
 %%
-Such a stencil computation is a common image convolution (for example
-for edge detection).
-Note that there are a number of intermediate assignments over which
-the whole specification ranges.
-
-The region constructor \texttt{*} can also be interpreted as conjunction of
-two specifications since they need to hold simultaneously for their
-corresponding dimensions.
-
-As another example, the following code and specification
-implements the \emph{Roberts cross}
+This pattern is a common image convolution (for example
+for edge detection). The specification ranges over the
+values that flow the array subscript on the left-hand side,
+and so ranges over the intermediate assignments to \term{x},
+\term{y}, and \term{z}.
+
+The region constructor \texttt{*} can also be interpreted as a
+kind of conjunction on specifications since they constraint
+indices to be within both regions simultaneously.
+
+The sum of two regions $r \texttt{+} r'$ specifies that an array is
+read using the neighbourhood indices described by either $r$ or
+$r'$; it acts as a disjunction on specifications. For
+example, the following gives the specification of a five-point
+stencil which is the sum of two compound \texttt{reflexive} and
+\texttt{centered} regions in each dimension:
+%%
+\begin{minted}{fortran}
+!= stencil centered(depth=1, dim=1)*reflexive(dim=2) + centered(depth=1, dim=2)*reflexive(dim=1) :: a
+b(i,j) = -4*a(i,j) + a(i-1,j) + a(i+1,j)
+                   + a(i,j-1) + a(i,j+1)
+\end{minted}
+%%
+Here the left hand side of \texttt{+} says that when the second dimension
+(induction variable $j$) is fixed at the origin, the first dimension
+(induction variable $i$) accesses the immediate vicinity of the origin
+(to depth of one). The right hand side of \texttt{+} is similar but the dimensions are reversed.
+This reflects the symmetry under rotation of the five-point stencil.
+
+As another example, the following 
+implements and specifies the \emph{Roberts cross}
 edge-detection convolution~\cite{davis1975survey}:
@@
 \begin{minted}{fortran}
-!= region r1 = centered(depth=1, dim=1)
-!= region r2 = centered(depth=1, dim=2)
+!= region r1 = forward(depth=1, dim=1)
+!= region r2 = forward(depth=1, dim=2)
 != region robertsCross = r1*r2
@@
 \end{minted}
-This is especially useful for common stencils such as Robert's cross
-as the region can be defined once and used by region name many times.
+This is especially useful for common stencil patterns, such as Roberts cross,
+as the region can be defined once and reused.
 %%
@@
 \stenBwdSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
- \equiv \; & \stenBwdS{n}{d}{r} \; \texttt{+} \; \stenBwdSR{m}{d}{s} \\[0.7em]
+ \equiv \; & \stenBwdSR{n}{d}{r} \; \texttt{+} \; \stenBwdSR{m}{d}{s} \\[0.7em]
 %
@@
   \centering
-  \subcaptionbox{\label{fig:jacobicorrect}}{\includegraphics[width=0.1\textwidth]{jacobi-correct.png}}
-  \subcaptionbox{}{\includegraphics[width=0.1\textwidth]{jacobi-incorrect1.png}}
-  \subcaptionbox{}{\includegraphics[width=0.1\textwidth]{jacobi-incorrect2.png}}
-  \subcaptionbox{}{\includegraphics[width=0.1\textwidth]{jacobi-incorrect3.png}}
+  %\subcaptionbox{\label{fig:jacobicorrect}}{\includegraphics[width=0.1\textwidth]{jacobi-correct.png}}
+  %\subcaptionbox{}{\includegraphics[width=0.1\textwidth]{jacobi-incorrect1.png}}
+  %\subcaptionbox{}{\includegraphics[width=0.1\textwidth]{jacobi-incorrect2.png}}
+  %\subcaptionbox{}{\includegraphics[width=0.1\textwidth]{jacobi-incorrect3.png}}
+  \subfigure[]{\label{fig:jacobicorrect}\includegraphics[width=0.1\textwidth]{jacobi-correct.png}}
+  \subfigure[]{\includegraphics[width=0.1\textwidth]{jacobi-incorrect1.png}}
+  \subfigure[]{\includegraphics[width=0.1\textwidth]{jacobi-incorrect2.png}}
+  \subfigure[]{\includegraphics[width=0.1\textwidth]{jacobi-incorrect3.png}}
   \caption{Greyscale output of several perturbations of a Jacobi

### 2016-07-06 paper.tex
tweaks to abstract and intro

--- 2016-07-06 paper.tex
@@
   language against a corpus of numerical Fortran code, and show that
-  precise specifications can be given to roughly half of the
-  identifiable stencil computations, and that the vast majority of
-  stencil computations indeed have a simple, regular, static
-  shape. This paper details our stencil specification language, its
-  static semantics, inference, checking, and program synthesis
-  procedures, our implementation, and our evaluation study.
+  the vast majority of stencil computations indeed have a simple,
+  regular, static shape. Around 10\% of the stencils found evoke a
+  programming pattern for which indexing errors are plausible. We
+  investigate two stencils in detail and find that we are able to
+  detect the vast majority of potential errors. This paper details our
+  stencil specification language, its static semantics, inference,
+  checking, and program synthesis procedures, our implementation, and
+  our evaluation study.
 \end{abstract}
@@
 of data access pattern (Section~\ref{sec:lang})
-along with a denotational model (Section~\ref{sec:semantics});
+along with a sound denotational model (Section~\ref{sec:semantics});
 
@@
 a corpus of scientific applications, totalling roughly one million
-lines of code, and we study some verification case studies.
-\dnote{insert results here}
+lines of code, and we study some verification case studies;
+
+\item we show that around 10\% of stencils are in a class that
+  represent a plausible scenario for a programming error;
+
+\item we consider two particular stencils in detail and show that we
+  are able to detect the vast majority of (synthesised) possible
+  indexing errors.
 \end{itemize}
@@
 such regular stencil patterns on a corpus of numerical programs (both
-small and large). We found that indeed \dnote{..}.
+small and large).
 

### 2016-07-06 paper.tex
Revised checking, consistency, and worked example

@@
 \usepackage{fancyvrb}
+\usepackage{subfigure}
 \usepackage{amsthm}
@@
 \begin{array}{c}
-%\framebox{$\textit{region} \equiv \textit{region}'$} \\[0.7em]
+\framebox{$\textit{region} \equiv \textit{region}'$} \\[0.7em]
 \begin{array}{rll}
@@
 \end{align*}
-\caption{Equations on regions, $\textit{region} \equiv \textit{region}'$}
+\caption{Equations on specifications}
 \label{fig:equations}
@@
 \begin{array}{c}
-%%
-%% INEQUATIONS ON REGION
-%% 
-%\framebox{$\textit{region} <: \textit{region}'$} \\[0.85em]
-%
-\dfrac{}{R <:^r R}\trule{\textsc{refl}} \qquad \dfrac{P <:^r R \quad R <:^r Q}{P <:^r
-  Q}\trule{\textsc{trans}} 
-\qquad
-\dfrac{R \equiv S}{R <:^r S}\trule{\textsc{eq}} 
+\dfrac{}{S <: S}(\textsc{refl}) \qquad \dfrac{R <: S \quad S <: T}{R <:
+  T}(\textsc{trans}) \\[1.5em]
+\dfrac{}{\texttt{readOnce} \, S <: S}(\textsc{rep})
 \\[1.5em]
 \setlength{\arraycolsep}{0.1em}
-\dfrac{\hspace{3em} m \leq n \qquad r \Rightarrow s \hspace{3em}}
+\dfrac{\hspace{3em} m \leq n \hspace{3em}}
 {\begin{array}{rl}
-\stenFwdSR{m}{d}{r} & <:^r \stenFwdSR{n}{d}{s}
+\stenFwdS{m}{d} & <: \stenFwdS{n}{d}
 %& \;\, ds \subseteq es \wedge n \leq m
 \\
-\wedge \; \stenBwdSR{m}{d}{r} & <:^r \stenBwdSR{n}{d}{s}
+\wedge \stenBwdS{m}{d} & <: \stenBwdS{n}{d}
 %& \;\, ds \subseteq es \wedge n \leq m
 \\
-\wedge \; \stenCenSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s}
+\wedge \; \stenCenS{m}{d} & <: \stenCenS{n}{d}
 %& \;\, ds \subseteq es \wedge n \leq m
 \\
-\wedge \; \stenBwdSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s} \\
-\wedge \; \stenFwdSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s} \\[1em]
-\end{array}}\trule{\textsc{over}} \\
+\wedge \; \stenBwdS{m}{d} & <: \stenCenS{n}{d} \\
+\wedge \; \stenFwdS{m}{d} & <: \stenCenS{n}{d} \\[1em]
+\end{array}} \\[2.5em]
 \hspace{-0.5em}
-\dfrac{P_1 <:^r R_1 \quad P_2 <:^r R_2}
-      {P_1 \, \texttt{+} \, P_2 <:^r R_1 \, \texttt{+} \, R_2}
-\trule{\textsc{cong}\texttt{+}} \;\;\;
-\dfrac{P_1 <:^r R_1 \quad P_2 <:^r R_2}
-      {P_1 \, \texttt{*} \, P_2 <:^r R_1 \, \texttt{*} \, R_2}
-\trule{\textsc{cong}\texttt{*}} \\[1.5em]
-%%
-%% INEQUATIONS ON SPEC
-%%
-%\framebox{$\textit{spec} <: \textit{spec}'$} \\[0.8em]
-%
-\dfrac{}{\term{readOnce} \, R <: R}\trule{\textsc{rep}}
-\quad
-\dfrac{}{\term{atLeast} \, R <: \term{atMost} \,
-  R}\trule{\textsc{bound}} \\[1.5em]
-%%
-\dfrac{P <:^r R}{\term{atLeast} \, P <: \term{atLeast} \,
-  R}\trule{\textsc{cong}\texttt{L}}
-\ 
-\dfrac{P <:^r R}{\term{atMost} \, P <: \term{atMost} \,
-  R}\trule{\textsc{cong}\texttt{M}} \\[1.5em]
-\dfrac{P <:^r R}{\term{readOnce} \, P <: \term{readOnce} \,
-  R}\trule{\textsc{cong}\texttt{R}} 
-%%
+\dfrac{S_1 <: T_1 \quad S_2 <: T_2}
+      {S_1 \, \texttt{+} \, S_2 <: T_1 \, \texttt{+} \, T_2}
+(\textsc{cong}\texttt{+}) \;\;\;
+\dfrac{S_1 <: T_1 \quad S_2 <: T_2}
+      {S_1 \, \texttt{*} \, S_2 <: T_1 \, \texttt{*} \, T_2}
+(\textsc{cong}\texttt{*})
 \end{array}
 \end{align*}
-\caption{Inequations $<:^r$ on \textit{regions} and
-$<:$ on \textit{specs}.}
+\caption{Inequations on regions}
 \label{fig:inequations}
@@
 where $\mathsf{IV}(e)$ is the set of free induction variables
- in an expression and $\neigh{}$ classifies expressions which 
-are constant offsets on an induction variable
+appearing in an expression, and $\neigh{}$ classifies expressions which
+are constant offsets from an induction variable
 (Defn.~\ref{def:neighbour}). In the third case,
@@
 
-For example, assuming
+For example $\textsf{schematic}(i, j+1) = \vtwoh{0}{1}$ assuming
 $i$ and $j$ are induction variables and $\textsf{schematic}(i-1, 2, j)
@@
 \begin{example}
-Three array indexing terms
-$(i, j+1)$, $(1, j+1)$ and $(3, j)$
-are mapped by \textsf{schematic} to the 
-index schemes: $\vtwoh{0}{1}$, $\vtwoh{\infty}{1}$
+Take three array indexing terms
+$(i, j+1)$, $(1, j+1)$ and $(3, j)$.
+As given by \textsf{schematic}, these have the
+corresponding schemes: $\vtwoh{0}{1}$, $\vtwoh{\infty}{1}$
 and $\vtwoh{\infty}{0}$ respectively. The following
@@
 \begin{align*}
-\interp{\stenCenS{2}{1}}_2 = \; & \{i\singleEntry{1}{2} \,|\, i \in \{\text{-}2,..., 2\}\!\setminus\!0 \}
+& \interp{\stenCenS{2}{1}}_2 \\
+= \; & \{i\singleEntry{1}{2} \,|\, i \in \{\text{-}2,..., 2\}\!\setminus\!0 \}
   \cup  \{ \zeroEntry{1}{2} \,|\, \textit{true} \} \\
@@
 \begin{definition} The \textsf{Mult}
-and \textsf{Approx} are parametric labelled variant types 
-with injection functions given by their definition:
+and \textsf{Approx} are parametric labelled variant (sum) types
+with injection function given by their definition:
 %
@@
 \ie{}, $\mathsf{lower}$ is an injection $\mathsf{lower} : a \rightarrow \mathsf{Approx} \, a$, and so
-on. The $\textsf{Mult}$ type corresponds to the presence or absence of the
-\term{readOnce} modifier as shown in Figure~\ref{fig:spatial-model}. 
-The $\textsf{Approx}$ type corresponds to the presence or absence of
-the spatial approximation modifier, with \textsf{exact} when there is
-no such modifier and \textsf{lower} and \textsf{upper} for
-\term{atLeast} and \term{atMost}. Later, it will be useful to model
-pairs of lower and upper bounds which is provided by \textsf{both} 
-(but which isn't used here).
-
-These are used for labelled models, and other components later in
+on. These are used for labelled models, and other components later in
 Section~\ref{sec:analysis} with additional information.
@@
 
-\begin{figure}[t]
+\begin{figure}
 \input{spatial-model.tex}
@@
 \label{fig:spatial-model}
-\vspace{-1em}
 \end{figure}
@@
 consider soundness for the equational theory of
-specifications and the notion of spatial approximation. The proofs are given in our supplement.
-
-\begin{theorem}[Soundness of equational theory (Figure~\ref{fig:equations})]
+specifications and the notion of spatial approximation
+on specifications. The proofs are given in our supplement.
+
+\begin{theorem}[Soundness of equational theory]
 \[
-S \equiv T \; \Rightarrow \;
-\interp{S}_n = \interp{T}_n
+\overline{\texttt{v} : S}\equiv \overline{\texttt{u} : T}
+\; \Rightarrow \;
+\interp{\overline{\texttt{v} : S}} = \interp{\overline{\texttt{u} : T}}
 \]
@@
 
-\begin{theorem}[Soundness of inequations
-  (Figure~\ref{fig:inequations})]
-
-Let $\leq$ be an ordering on
-$\textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$
-and $\leq'$ be an ordering on
-$\textsf{Approx}(\mathcal{P}(\domainVal{}^n))$ defined by the
-following caluses
-\begin{align*}
-(M \leq' N \Rightarrow \textit{inj} \, M \leq \textit{inj} \, N) &
-\qquad 
-(\textsf{once} \, M \leq' \textsf{mult} \, M) \\
-%%
-%%
-(M \subseteq N \Rightarrow \textit{inj}' M \leq' \textit{inj}' N)
-& \;\;\; (\textsf{exact} M \leq  \textsf{lower} M)
-\;\;\; (\textsf{lower} M \leq  \textsf{upper} M)
-\end{align*}
-%%
-where $\textit{inj}$ ranges over the unary injections
-of $\textsf{Approx}$ and $\textit{inj}'$ ranges over the injections of
-$\textsf{Mult}$. Then, soundness of $<:$ is that:
-%%
+\begin{theorem}[Soundness of approximations]
 \[
-S <: T
+\overline{\texttt{v} : S} <: \overline{\texttt{u} : T}
 \; \Rightarrow \;
-\interp{S}_n \leq \interp{T}_n
+\interp{\overline{\texttt{v} : S}} \subseteq \interp{\overline{\texttt{u} : T}}
 \]
@@
 specifications and for synthesising programs from specifications.
-These build on the model of Section~\ref{sec:semantics}.
-
-\paragraph{Syntax helper functions}
-Let \rhsExp{} map source language statements to a
+Checking and synthesis use the set-based semantic model of
+Section~\ref{sec:semantics}.
+
+\paragraph{Syntax functions}
+The function \rhsExp{} maps statements to a
 set of expressions which occur in right-hand positions (\ie{}, not the
-target of an assignment). Let \var{} map expressions to a
+target of an assignment). The function \var{} maps expressions to a
 set of the variables used in right-hand positions.
@@
 \begin{align*}
-\vect{x} \bullet \vect{y}
+& (\vect{x}^L,\vect{x}^U) \bullet (\vect{y}^L,\vect{y}^U) \\
 = &
@@
 
-\newcommand{\bplus}{\operatornamewithlimits{\term{\huge{+}}}}
+\newcommand{\bplus}{\operatornamewithlimits{\term{+}}}
 \newcommand{\tySum}[1]{#1^{\term{+}}}
@@
 \newcommand{\specDNF}{\textbf{spec}}
+\newcommand{\result}[1]{\lfloor{#1}\rceil}
 
@@
 space is converted into abstract syntax trees
-of the specification language. This conversion is defined in terms
+for the specification language. This conversion is defined in terms
 of an algebra on \textit{spec} syntax in disjunctive-normal form
-(which we denote \specDNF{}), and which is wrapped
-with the $\textsf{Approx}$ type (see
-Definition~\ref{def:mult-and-approx},
-p.~\pageref{def:mult-and-approx}), which we write as $\mathsf{A}$
-for brevity.  The algebra mirrors the shape
+(which we denote \specDNF{}).
+
+\begin{align*}
+\result{\specDNF{}} = \specDNF + ((\specDNF{} + 1) \times
+  (\specDNF{} + 1))
+\end{align*}
+
+\begin{align*}
+\result{\specDNF{}} & ::= \textsf{exact} \, \specDNF{}
+\mid \textsf{bounds} ([\specDNF{}], [\specDNF{}]) \\
+\end{align*}
+
+The algebra mirrors the shape
 of the region combinators for addition and multiplication,
@@
 \begin{align*}
-\oplus & : \mathsf{A}(\specDNF) \rightarrow \mathsf{A}(\specDNF) \rightarrow
-  \mathsf{A}(\specDNF) \\
-\odot & : \mathsf{A}(\specDNF) \rightarrow \mathsf{A}(\specDNF) \rightarrow
-        \mathsf{A}(\specDNF) \\
+\oplus & : \specDNF \rightarrow \specDNF \rightarrow
+  \specDNF \\
+\odot & : \specDNF \rightarrow \specDNF \rightarrow
+        \specDNF \\
 \mathsf{toSpec} & : (\mathbb{Z} \cup \{\infty\}) \rightarrow (\mathbb{Z} \cup
-  \{\infty\})  \rightarrow \mathsf{A}(\specDNF)
+  \{\infty\})  \rightarrow \specDNF
 \end{align*}
@@
 %%
-More formally, $\specDNF \subseteq (\textit{spec} \cup \{\epsilon\})$, defined:
-%%
-\begin{align*}
-\specDNF{} & ::= \tySum{\specDNF} \mid \epsilon \\
+More formally, $\specDNF \subseteq \textit{spec}$, defined:
+%%
+\begin{align*}
+\specDNF{} & ::= [\textit{mult},] \;
+\tySum{\specDNF} \\
 \tySum{\specDNF} & ::= \tyProd{\specDNF} \; \term{+} \; \tySum{\specDNF} \mid
@@
 
-\newcommand{\exactR}{\mathsf{exact}\;}
-\newcommand{\exact}[1]{\exactR(#1)}
-\newcommand{\upperB}[1]{\mathsf{upper}\;(#1)}
-
 \begin{definition} The $\mathsf{toSpec}$ constructor is defined:
@@
 &\mathsf{toSpec} \; \llower \; \lupper = \\
-&\hspace{-0.4em}\begin{cases}
+&\begin{cases}
 %% FWD
-\exact{\stenFwdS{\lupper}{i}}  & \llower = 0 \, \wedge \, \lupper > 0 \\[0.2em]
-\exact{\stenFwdSR{\lupper}{i}{\irreflS}}\!\!\!\! & \llower = 1 \, \wedge \, \lupper > 0 \\[0.2em]
+ \stenFwdS{\lupper}{i}  & \llower = 0 \, \wedge \, \lupper > 0 \\[0.2em]
+ \stenFwdSR{\lupper}{i}{\irreflS}  & \llower = 1 \, \wedge \, \lupper > 0 \\[0.2em]
 %% BWD
-\exact{\stenBwdS{|\llower|}{i}}  & \llower < 0 \, \wedge \, \lupper = 0 \\[0.2em]
-\exact{\stenBwdSR{|\llower|}{i}{\irreflS}}\!\!\!\! & \llower < 0 \, \wedge \, \lupper = -1 \\[0.2em]
+ \stenBwdS{|\llower|}{i}  & \llower < 0 \, \wedge \, \lupper = 0 \\[0.2em]
+\stenBwdSR{|\llower|}{i}{\irreflS} & \llower < 0 \, \wedge \, \lupper = -1 \\[0.2em]
 %% CEN
-\exact{\stenCenS{\lupper}{i}}& \llower < 0 \wedge \lupper > 0 \wedge |\llower| = \lupper \\[0.2em]
-\exactR(\stenBwdS{|\llower|}{i} & \llower < 0 \wedge \lupper > 0  \wedge |\llower| \neq \lupper \\
-\qquad\qquad \texttt{+} \, \stenFwdS{\lupper}{i}) & \\[0.2em]
+ \stenCenS{\lupper}{i}  & \llower < 0 \, \wedge \, \lupper > 0
+ \, \wedge \, |\llower| = \lupper \\[0.2em]
+ \stenBwdS{|\llower|}{i} & \llower < 0 \, \wedge \, \lupper > 0  \, \wedge \, |\llower| \neq \lupper \\
+\quad \texttt{+} \, \stenFwdS{\lupper}{i}  & \qquad\quad \\[0.2em]
 % BOUND
-\upperB{\stenFwdS{\lupper}{i}} & \llower > 1 \\[0.2em]
-\upperB{\stenBwdS{|\llower|}{i}} \quad & \llower < (-1)
+ \texttt{atMost}, \stenFwdS{\lupper}{i} & \llower > 1 \\[0.2em]
+ \texttt{atMost}, \stenBwdS{|\llower|}{i} \quad & \llower < (-1)
  \\[0.2em]
 % ABSOLUTE REP
-\exact{\epsilon} & \llower = \infty \, \vee \, \lupper = \infty
+ \epsilon & \llower = \infty \, \vee \, \lupper = \infty
   \end{cases}
@@
 
-Finally, if either bound is $\infty$, representing a non
+Finally, in the case where either bound is $\infty$, representing a non
 neighbourhood index, then an empty specification is given
@@
 \begin{lemma}
-$\mathsf{toSpec}$ is total 
+$\mathsf{toSpec}$ is total (\ie{}, defined for all inputs)
  given the precondition that the lower
@@
 %\newcommand{\mkBig}[2]{\overset{\LARGE{\term{#1}}}{#2}}
-\begin{definition} The $\odot$ operation on $\textsf{A}(\specDNF{})$
-  is defined in terms of the intermediate $\hat{\odot}$ on \specDNF{}:
-%%
-\begin{align*}
-S \,\hat{\odot}\, \epsilon = S \quad
-\epsilon \,\hat{\odot}\, S = S \qquad
-S \,\hat{\odot}\, T = \bplus_{(\tyProd{s}, \tyProd{t}) \in (S \times
-   T)} (\tyProd{s} \; \term{*} \; \tyProd{t})
-\end{align*}
-%%
-That is, the $\hat{\odot}$ product of two specifications $S$ and $T$ in DNF form 
-is the $\term{+}$ sum of all pairwise $\term{*}$ products for every
-pair of $\tyProd{spec}$ drawn from $S$ and $T$. The result
-is indeed in DNF-form. The operation is
-commutative and associative, and sound with respect to our
-model (see Lemma~\ref{lem:alg-soundness} below).
-The $\hat{\odot}$ operation is then lifted to all combinations of
-\textsf{Approx} for $\odot$:
-\begin{align*}
-\textsf{exact}(S) \odot \textsf{exact}(T) = \; & \textsf{exact}(S
-                                              \hat{\odot} T) \\
-\textsf{lower}(S) \odot \textsf{lower}(T) = \; & 
-      \textsf{lower} \, (S \hat{\odot} T) \\
-\textsf{upper}(S) \odot \textsf{upper}(T) = \; &
-      \textsf{upper} \, (S \hat{\odot} T) \\
-\textsf{lower}(S) \odot \textsf{exact}(T) = \; &
-       %\textsf{exact}(T) \odot \textsf{lower}(S) = \;
-       \textsf{both} \, (S \hat{\odot} T, \; T) \\
-\textsf{upper}(S) \odot \textsf{exact}(T) = \; & 
-       %\textsf{exact}(T) \odot \textsf{upper}(S) =
-      \; \textsf{both} \, (T, \; S \hat{\odot} T) \\
-\textsf{lower}(S) \odot \textsf{both}(T_l,T_u) = \; &
-        %\textsf{both}(T_l,T_u) \odot \textsf{lower}(S) = \;
-        \textsf{both} \, (S \hat{\odot} T_l, \; T_u) \\
-\textsf{upper}(S) \odot \textsf{both}(T_l,T_u) = \; & \textsf{both} \,
-                                                      (T_l, \; S \hat{\odot} T_u) \\
-\textsf{both}(S_l,S_u) \odot
-\textsf{both}(T_l,T_u) & = \textsf{both} (S_l \hat{\odot} T_l, \; S_u
-                         \hat{\odot} T_u)
-\end{align*}
-For brevity we omit the cases where the parameters above are flipped
-as $\odot$ is commutative.
+\begin{definition} The $\odot$ operation on \specDNF{} is defined
+\begin{align*}
+& ([\textit{approx}_1] \; [\textit{mult}_1] \;
+\tySum{s})
+\odot
+([\textit{approx}_2] \; [\textit{mult}_2] \;
+\tySum{t}) = \\
+& \begin{cases}
+\tySum{s} \tySum{\odot} \tySum{t} & \textit{approx}_1 =
+\textit{approx_2} = \varepsilon
+\\
+\
+& \textit{approx}_1 = \texttt{atLeast}, \textit{approx}_2 = \varepsilon
+\end{cases}
+\\
+\end{align*}
+\begin{align*}
+s \tySum{\odot} t & =
+ \bplus_{(\tyProd{s_1}, \tyProd{s_2}) \in (s \times
+   t)} (\tyProd{s_1} \; \term{*} \; \tyProd{s_2})
+%\begin{cases}%
+%\end{cases}
+\end{align*}
 \end{definition}
-
-\begin{definition}The $\oplus$ operation on $\textsf{A}(\specDNF)$ 
-is defined in a similar way to $\odot$, with $\hat{\oplus}$ on
-$\specDNF$ given by:
-%%
-\begin{equation*}
-S \,\hat{\oplus}\, \epsilon = S \quad
-\epsilon \,\hat{\oplus}\, S = S \quad
-S \, \hat{\oplus} \, T = S \, \term{+} \, T
-\end{equation*}
-%%
-Thus, if neither specification is empty, their $\oplus$ is just
-the syntatic sum $\term{+}$. Then $\oplus$ is the lifting of
-$\hat{\oplus}$ to boudns with the shape as $\odot$ in the previous
-definition.
-\end{definition}
-
-\begin{lemma}[$\hat{\odot}$, $\hat{\oplus}$ soundness] $\forall S,
-  T \in \specDNF$ where $S \neq \epsilon \wedge T \neq \epsilon$:
-\begin{align*}
-\interp{S \; \hat{\odot} \; T}_n & = \interp{S}_n \otimes \interp{T}_n
-  \\
-\interp{S \; \hat{\oplus} \; T}_n & = \interp{S}_n \cup \interp{T}_n
-\end{align*}
-\label{lem:alg-soundness}
-\end{lemma}
-%
 
@@
 
-A specification is first synthesised into its model. Then the occurrences of an
-array in an assignment are checked against the model for consistency as defined
-in Figure~\ref{fig:consistency}. The consistency predicate is defined
-recursively.
-
-It first checks multiplicity if $\textsf{once}$ is at the top
-level, \ie{} specification in question involves \texttt{readOnce} modifier, then
-the array accesses are checked for uniqueness using $unique$ predicate and the
-problem is reduced to $\textsf{mult}$, which does not impose any extra
-constraints.
+Checking procedure synthesises syntactic specification into its model and
+generates an injection from indexing schemes to the multiplicity of observed
+access patterns, this is formally defined in Figure~\ref{fig:consistency}.
+Then, the consistency of the injection from the indexing schemes with the model
+is checked.
+
+It the multiplicity of the access pattern and the model is matched and are
+$\mathit{once}$, then the problem is reduced to checking consistency in
+models in which duplicate access patterns are allowed.
 
@@
 \begin{enumerate}
+\item From now on assignment on line three is referred as $L3$ for brevity.
+  Verification is done by computing the following:
+  $$check(\texttt{stencil readOnce, fivepoint},L3,b)$$
+%
 \item Generate model from the specification as described in
-  Section~\ref{sec:semantics}.
-%
-\begin{dmath*}
-\interp{\texttt{stencil readOnce, fivepoint}}_2 =
-  \textsf{once}(
-    \textsf{exact}(
-      \{ \vtwo{-1}{\infty},
+  Section~\ref{sec:semantics}. We refer to the indexing
+  schemes within the model as set $m$.
+%
+\[ m= \{  \vtwo{-1}{\infty},
           \vtwo{0}{\infty},
@@
           \vtwo{\infty}{0},
-          \vtwo{\infty}{1}\}))
+        \vtwo{\infty}{1} \} \]
+\begin{dmath*}
+\interp{\texttt{stencil readOnce, fivepoint}}_2 =
+  \{\textsf{once}(\textsf{exact}(m)) \}
 \end{dmath*}
 %
-\item Now the $cons$ predicate is applied with this model alongside the
-  assignment on line three and array $b$. For brevity the assignment is
-  referred to as $L3$ and the set of unifiers defined in the previous step as
-  $m$.
-
-  The predicate is reduced to
-  $cons(\textsf{once}(\textsf{exact}(m)),L3,b)$.
-%
-\item $cons$ predicate from the previous step evaluates to
-  $$unique(L3,b) \; \wedge \; cons(\textsf{mult}(\textsf{exact}(m)),L3,b).$$
-%
-  This condition holds as no access pattern is repeated in $L3$ by simple
-  inspection.
-%
-\item We further expand the function to 
-  \begin{equation*}
-    cons(\textsf{mult}(\textsf{up}(m)),L3,b) \wedge 
-    cons(\textsf{mult}(\textsf{low}(m),L3,b).
-  \end{equation*}
-%
 \item To evaluate the two predicates above, we need indexing schemes for the
-  array in question. This is obtained by the $analyse$ function from
-  Subsection~\ref{subsec:inference}.
-
-  For the example above the indexing schemes are as follows:
-%
-\begin{dmath*}
-analyse(L3,b) =
-  \{ \vtwo{0}{-1},
+  array in question. This is obtained by $modelise$ function. For brevity the
+  set of indexing schemes is referred as $m'$.
+
+  For the example above the observed model is as follows:
+%
+\[
+m' = \{
+    \vtwo{0}{-1},
     \vtwo{0}{0},
@@
   \}
+\]
+\[modelise(L3,b) = \textsf{once}(m')\]
+%
+\item Now $cons$ predicate is applied with this model along side with the
+  assignment on line three and array $b$.
+
+  The predicate is reduced to
+  $cons(\textsf{once}(\textsf{exact}(m)),\textsf{mult}(m'))$.
+%
+\item $cons$ predicate from the previous step evaluates to
+  $$cons(\textsf{mult}(\textsf{exact}(m)),\textsf{once}(m')).$$
+%
+\item We further evaluate the predicate to
+  \begin{dmath*}
+    cons(\textsf{mult}(\textsf{up}(m)),\textsf{mult}(m')) \wedge
+    cons(\textsf{mult}(\textsf{low}(m),\textsf{mult}(m')).
   \end{dmath*}
@@
   Since both predicates are satisfied the array accesses specified on line
-  three, the stencil comutation is consistent with the \texttt{fivepoint}
+  three, the stencil computation is consistent with the \texttt{fivepoint}
   specification.
@@
 %
-The first four steps proceed as before. In the fifth step, the result
-of $\mathit{analyse}(L3',b)$ will have an extra term $\vtwoh{2}{2}$. In the
-sixth step, since the model has not changed and the indexing schemes are only
-augmented with a new element, the generalisations for lower bound still hold.
-However, there are no elements in the model, $m$, that can generalise
-$\vtwoh{2}{2}$, hence the model fails to hold as an upper bound.
+In the first six steps, evaluation proceeds the same way except for step three.
+In the third step, result of $\mathit{modelise}(L3',b)$ will have the extra term
+$\vtwoh{2}{2}$. In the seventh step, since the model has not changed and the
+indexing schemes are only augmented with a new element, the generalisations for
+lower bound still hold. However, there are no elements in the model, $m$, that
+can generalise $\vtwoh{2}{2}$, hence the model fails to hold as an upper bound.
 
@@
 The consistency predicate would have held because for the upper bound
-$\mathit{cons}(\textsf{mult}(\textsf{up}(m)),L3',b)$ would not have
-been needed. This semantically makes sense since the region
-$\texttt{fivepoint}$ is indeed contained in the regions touched by $b$.
+$\mathit{cons}(\textsf{mult}(\textsf{up}(m)),\textsf{mult}(m'))$ would not have
+been needed. This semantically makes sense since the region $\texttt{fivepoint}$
+is indeed contained in the regions touched by $b$.
+
+\subsubsection{A note on region approximations}
 

### 2016-07-06 paper.tex
section 2.3 edits

@@
 \label{ref:navier-stokes-fragment}
+\vspace{-0.8em}
 \end{figure}
@@
     variable} if it is the control variable of a ``for''
-  loop,\footnote{Or equivalent in the source language \eg{}
-    \texttt{do} in Fortran for our implementation.} incremented by $1$ per
+  loop,\footnote{Or equivalent, \eg{}, \texttt{do} in Fortran, for our implementation.} incremented by $1$ per
 iteration. The variable is marked as a base induction variable
@@
 %
+The \term{irreflexive} attribute cannot be used on \term{reflexive}
+regions.
+
 A \texttt{centered} stencil is a combination of \texttt{forward}
@@
 \label{fig:syntax}
+\vspace{-0.5em}
 \end{figure}
@@
 %%
-\paragraph{Spatial modifiers}
-%%
-There are two modifiers: one used at specification and the other at region
-level to provide better descriptions for stencil computations.
-
-At the specification level, the \texttt{readOnce} modifier enforces that no
-index appears more than once. For example, in all of the previous examples the
-\texttt{readOnce} modifier could be added, \eg{}
+\paragraph{Modifiers}
+%%
+Region specifications can be modified 
+by \emph{approximation} and \emph{multiplicity} information 
+(in the \textit{spec} rule of Fig.~\ref{fig:syntax}).
+
+The \texttt{readOnce} modifier enforces that no index appears more
+than once (that is, its multiplicity is one). For example, in all of
+the previous examples the \texttt{readOnce} modifier could be added,
+\eg{}
 %
@@
 %
-However, this would be an invalid specification if any of the
+This specification would be invalid if any of the
 array subscripts were repeated. This modifier provides a way to
 rule out any accidental repetition of array subscripts.
-The notion corresponds to that of \emph{linearity} in type systems,
-though we opt for the more informative and easily understood name
+The notion is similar to that of \emph{linearity} in linear type
+systems~\cite{wadler1990linear}, where a value must be used
+exactly once. We opt for the more informative and easily understood name
 \texttt{readOnce}. This modifier is optional, so it need not
@@
 
-At the region level, \texttt{irreflexive} modifier qualifies base
-regions of \texttt{forward}, \texttt{backward}, and \texttt{centered}.
-It indicates that in the dimension of the region there is no neighbour index
-with $0$ offset, \eg{},
-%
-\begin{minted}{fortran}
-!= stencil centered(depth=1, dim=1, irreflexive) :: a
-b(i) = a(i+1) + a(i-1)
-\end{minted}
-%
-or with just a \texttt{forward} region an example would be:
-%
-\begin{minted}{fortran}
-!= stencil forward(depth=1, dim=1, irreflexive) :: a
-b(i) = a(i+1)
-\end{minted}
-
-Naturally, \term{irreflexive} qualifier cannot be used on \term{reflexive}
-regions.
-
-\paragraph{Under and over-approximation}
+%At the region level, \texttt{irreflexive} modifier qualifies base
+%regions of \texttt{forward}, \texttt{backward}, and \texttt{centered}.
+%It indicates that in the dimension of the region there is no neighbour index
+%with $0$ offset, \eg{},
+%
+%\begin{minted}{fortran}
+%!= stencil centered(depth=1, dim=1, irreflexive) :: a
+%b(i) = a(i+1) + a(i-1)
+%\end{minted}
+%
+%or with just a \texttt{forward} region an example would be:
+%%
+%\begin{minted}{fortran}
+%!= stencil forward(depth=1, dim=1, irreflexive) :: a
+%b(i) = a(i+1)
+%\end{minted}
+
+%\paragraph{Under and over-approximation}
 
@@
 \term{atLeast} modifiers. This is particularly useful in situations
-where there is a non-contiguous stencil, which does not fit the rest
-of the specification syntax. For example:
+where there is a non-contiguous stencil pattern, which cannot be expressed
+precisely in our specification syntax. For example:
 %%
@@
 for the shape of discretisation schemes. For example,
-the standard ``explicit method'' for approximation
-PDEs is the \emph{Forward Time, Centered Space} (FTCS)
+the standard \emph{explicit method} for approximating
+PDEs is known as the \emph{Forward Time, Centered Space} (FTCS)
 scheme~\cite{dawson1991finite}. For
-the one-dimensional heat equations, an FTCS discretisation
+the one-dimensional heat equation, an FTCS discretisation
 provides approximation code with the following stencil~\cite{recktenwald2004finite}:
@@
 %%
-where \texttt{r} and \texttt{r2} are some constants.
-A valid specification for this stencil is then:
+where \texttt{r} and \texttt{r2} are constants.
+A valid specification for this is:
 %%

### 2016-07-06 paper.tex
now we find 39 stencils
edits to section 3

@@
   breaklines=true,
-  breakafter=+*,
+  breakafter=),
   xleftmargin=\parindent
@@
 \end{center}
-\vspace{-0.3em}
+\vspace{-0.5em}
 (c) Pictorial representation of the two stencil specifications.
@@
 \begin{figure}[t]
+\vspace{-0.1em}
 \begin{align*}
@@
 \setlength{\arraycolsep}{0.2em}
-\newcommand{\dimTy}{\mathbb{D}}
+\newcommand{\dimTy}{\mathbb{N}_{>0}}
 \begin{array}{rl}
@@
 \nonterm{approx} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
-\nonterm{region} ::= & \\
-\multicolumn{2}{l}{\qquad\qquad\;\;\; \stenFwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]}} \\
+\nonterm{region} ::= & \nonterm{rvar} \; \\
+\multicolumn{2}{l}{\qquad\qquad \mid \; \stenRefl{\dimTy}} \\
+\multicolumn{2}{l}{\qquad\qquad \mid \;\stenFwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]}} \\
 \multicolumn{2}{l}{\qquad\qquad \mid \; \stenBwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]}} \\
 \multicolumn{2}{l}{\qquad\qquad \mid \; \stenCen{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]}} \\
-\multicolumn{2}{l}{\qquad\qquad \mid \; \stenRefl{\dimTy}}  \\
 \multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{region} \; \term{+} \; \nonterm{region}} \\
-\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{region} \; \term{*} \; \nonterm{region}} \\
-\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{rvar}}  \\[0.5em]
+\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{region} \; \term{*} \; \nonterm{region}} \\[0.5em]
+%\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{rvar}}  
+%\\[0.5em]
 %\nonterm{temporal} ::= \; & \term{dependency} \; (v \; \{ , v \}) [, \texttt{mutual}]
 %  \\[0.5em]
-\dimTy ::= \; & \mathbb{N}_{>0} \\
 \nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+\\[-1em]
@@
                          (R^\irreflS \, \term{*} \, S) \;
-\equiv \; R \, \term{*} \, S \\[0.5em]
-%
-%
-(\texttt{+}\textsc{idem}) \;\; & S \; \texttt{+} \; S \; \equiv \; S \\[0.5em]
+\equiv \; R \, \term{*} \, S \\[0.4em]
+%
+%
+(\texttt{+}\textsc{idem}) \;\; & S \; \texttt{+} \; S \; \equiv \; S \\[0.4em]
 %
 (\texttt{+}\textsc{comm}) \;\; & S \; \texttt{+} \; T \; \equiv \; T \;
-                       \texttt{+} \; S \\[0.5em]
+                       \texttt{+} \; S \\[0.4em]
 %
 (\texttt{+}\textsc{assoc}) \;\; & R \; \texttt{+} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
-                       \texttt{+} \; S) \; \texttt{+} \; T \\[0.5em]
+                       \texttt{+} \; S) \; \texttt{+} \; T \\[0.4em]
 (\texttt{*}\textsc{comm}) \;\; & S \; \texttt{*} \; T \; \equiv \; T \;
-                       \texttt{*} \; S \\[0.5em]
+                       \texttt{*} \; S \\[0.4em]
 %
 (\texttt{*}\textsc{assoc}) \;\; & R \; \texttt{*} \; (S \; \texttt{*} \; T) \; \equiv \; (R \;
-                       \texttt{*} \; S) \; \texttt{*} \; T \\[0.5em]
+                       \texttt{*} \; S) \; \texttt{*} \; T \\[0.4em]
 (\textsc{dist}) \;\; & R \; \texttt{*} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
@@
 \end{align*}
-\caption{Equations on specifications}
+\caption{Equations on specifications (regions)}
+\vspace{-1em}
 \label{fig:equations}
@@
 \begin{align*}
-& \textit{region} \equiv \textit{region'} \\
-& \quad \Rightarrow \;\; \texttt{stencil} \; [\textit{approx},] \; [\textit{mult},] \;
+(\textit{region} \equiv \textit{region'}) 
+ \Rightarrow \quad\;\; & \texttt{stencil} \; [\textit{approx},] \; [\textit{mult},] \;
 \textit{region} \\[-0.4em]
-& \quad\quad \;\; \equiv \;\texttt{stencil} \; [\textit{approx},] \;
+\quad \equiv \; & \texttt{stencil} \; [\textit{approx},] \;
             [\textit{mult},] \; \textit{region}'
@@
 attribute on region constants as a boolean. For example,
-the syntactic sugar in our equations for \term{fwd} is defined:
-%%
-\begin{align*}
+the syntactic sugar for \term{fwd} is defined:
+%%
+\begin{align*}
+\hspace{-0.25em}
 & \stenFwdSR{n}{d}{r}
@@
 & \qquad \begin{cases}
-\stenFwd{n}{d}{, \irrefl} & \textit{if} \;\, \neg \, r \\
-\stenFwd{n}{d}{}  & \textit{if} \;\,  r
+\stenFwd{n}{d}{, \irrefl} & \;\;\textit{if} \;\, \neg \, r \\
+\stenFwd{n}{d}{}  & \;\;\textit{if} \;\,  r
 \end{cases} \\
@@
 The Unified Model contains an implementation of the Smagorinsky subgrid-scale model for calculating turbulence.
-The source file in question yields 38 stencils from 340 lines of code.
+The source file in question yields 39 stencils from 340 lines of code.
 We selected one as an example:
 
-\begin{minted}[breakindent=8.1em]{fortran}
- != stencil readOnce, (forward(depth=1, dim=3, irreflexive))*(backward(depth=1, dim=1))*(backward(depth=1, dim=2)) + (forward(depth=1, dim=3))*(backward(depth=1, dim=1))*(reflexive(dim=2)) + (forward(depth=1, dim=3))*(backward(depth=1, dim=1, irreflexive))*(backward(depth=1, dim=2, irreflexive)) + (forward(depth=1, dim=3))*(backward(depth=1, dim=2))*(reflexive(dim=1)) + (backward(depth=1, dim=1))*(backward(depth=1, dim=2, irreflexive))*(reflexive(dim=3)) + (backward(depth=1, dim=1, irreflexive))*(backward(depth=1, dim=2))*(reflexive(dim=3)) :: x
+\begin{minted}[breakindent=0em,linenos=false,xleftmargin=0em]{fortran}
+!=stencil readOnce,(forward(depth=1,dim=3,irreflexive))*(backward(depth=1,dim=1))*(backward(depth=1,dim=2))+(forward(depth=1,dim=3))*(backward(depth=1,dim=1))*(reflexive(dim=2))+(forward(depth=1,dim=3))*(backward(depth=1,dim=1,irreflexive))*(backward(depth=1,dim=2,irreflexive))+(forward(depth=1,dim=3))*(backward(depth=1,dim=2))*(reflexive(dim=1))+(backward(depth=1,dim=1))*(backward(depth=1,dim=2,irreflexive))*(reflexive(dim=3))+(backward(depth=1,dim=1,irreflexive))*(backward(depth=1,dim=2))*(reflexive(dim=3))::x
 \end{minted}
@@
 % 3D stencil codes
-\citet{Zhang:2012:CGO}
-
-% Ypnos
-\citet{Orchard:2010:DAMP}
+%\citet{Zhang:2012:CGO}
 

### 2016-07-06 paper.tex
putting the inequations back
putting back lots of stuff that was somehow lost

@@
 %
-To save space in the equations of Figure~\ref{fig:equations}, we use abbreviations
-\term{refl}, \term{fwd}, \term{bwd}, and \term{cen} for \term{reflexive},
-\term{forward}, \term{backward}, and \term{centered} regions
-respectively. Futhermore, the abbreviations include syntactic sugar
+The equations of Figure~\ref{fig:equations} use abbreviations
+\term{refl}, \term{fwd}, \term{bwd}, \term{cen}, and \irreflS{}
+for \term{reflexive}, \term{forward}, \term{backward},
+\term{centered}, and \irrefl{} respectively. Futhermore, the abbreviations include syntactic sugar
 for quantifying over the abseence or presence of the \irrefl{}
@@
 and similarly for \term{bwd} and \term{cen}.
-We also frequently abbreviate \irrefl{} to \irreflS{}.
 
@@
 equations (with labels of the form $\ldots \texttt{+} \ldots$)
-explain the notion of \emph{overlapping} regions with sum
-\term{+}. Note that in the first six, $r$ and $s$ range
+explain the notion of \emph{region overlapping} with sum
+\term{+}. In the first six, $r$ and $s$ range
 over booleans denoting the absence (true) or presence (false)
@@
 the rule $(\textsc{B} \, \term{+} \,
-\textsc{F})$ explains that the addition of \term{backward}
+\textsc{F})$ explains that the sum of \term{backward}
 and \term{forward} regions in the same dimension and the same
 depth is equivalent to a \term{centered} region of that depth.
-If the \term{forward} part is irrefleixve (\ie{}, $r =
-\textit{false}$) but \term{backward} is not (\ie{}, $s =
-\textit{true}$) then their addition is equal to a \term{centered}
-region with no \irrefl{} modifier by the disjunction ($r \vee s =
+If the \term{forward} part is irrefleixve ($r =
+\textit{false}$) but \term{backward} is not ($s =
+\textit{true}$) then their sum is equal to a \term{centered}
+region with no \irreflS{} modifier by the disjunction ($r \vee s =
 \textit{true}$), \ie{}:
@@
 
-The equation (\textsc{irrefl}) explains an annihilation of
-\irrefl{} using a further syntactic sugar.
+The equation (\textsc{irrefl}) provides an annihilation of
+\irreflS{} using a further syntactic sugar.
 We write $R^{\irreflS}$ to denote a region constant with an
@@
 \begin{align*}
-\stenFwdS{n}{d}^{\irreflS} \, := \,
+\stenFwdS{n}{d}^{\irreflS}  := \,
 \stenFwdSR{n}{d}{\irreflS}
@@
 %%
-Thus, (\textsc{irrefl}) explains the sum of a
-product of two regions $R$ and $S$ with itself,
+The left-hand side of (\textsc{irrefl}) has a sum of
+products of two regions $R$ and $S$, 
 where in each component of the sum one of $R$
 or $S$ has the irreflexive attribute. These cancel to give
-just the specification $R \, \term{*} \, S$.
+just $R \, \term{*} \, S$.
 
@@
 \begin{align*}
+\hspace{-0.5em}
 \begin{array}{c}
-\dfrac{}{S <: S}(\textsc{refl}) \qquad \dfrac{R <: S \quad S <: T}{R <:
-  T}(\textsc{trans}) \\[1.5em]
-\dfrac{}{\texttt{readOnce} \, S <: S}(\textsc{rep})
-\\[1.5em]
+%%
+%% INEQUATIONS ON REGION
+%% 
+%\framebox{$\textit{region} <: \textit{region}'$} \\[0.85em]
+%
+%\dfrac{}{R <:^r R}\trule{\textsc{refl}} \qquad \dfrac{P <:^r R \quad R <:^r Q}{P <:^r
+%  Q}\trule{\textsc{trans}} 
+%\qquad
+%\\[1.5em]
+\vspace{-0.4em}
 \setlength{\arraycolsep}{0.1em}
-\dfrac{\hspace{3em} m \leq n \hspace{3em}}
+\dfrac{\hspace{3em} m \leq n \qquad\qquad r \Rightarrow s \hspace{3em}}
 {\begin{array}{rl}
-\stenFwdS{m}{d} & <: \stenFwdS{n}{d}
+\stenFwdSR{m}{d}{r} & <:^r \stenFwdSR{n}{d}{s}
 %& \;\, ds \subseteq es \wedge n \leq m
 \\
-\wedge \stenBwdS{m}{d} & <: \stenBwdS{n}{d}
+\wedge \; \stenBwdSR{m}{d}{r} & <:^r \stenBwdSR{n}{d}{s}
 %& \;\, ds \subseteq es \wedge n \leq m
 \\
-\wedge \; \stenCenS{m}{d} & <: \stenCenS{n}{d}
+\wedge \; \stenCenSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s}
 %& \;\, ds \subseteq es \wedge n \leq m
 \\
-\wedge \; \stenBwdS{m}{d} & <: \stenCenS{n}{d} \\
-\wedge \; \stenFwdS{m}{d} & <: \stenCenS{n}{d} \\[1em]
-\end{array}} \\[2.5em]
+\wedge \; \stenBwdSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s} \\
+\wedge \; \stenFwdSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s} \\[1.2em]
+\end{array}}\trule{\textsc{over}} \\
 \hspace{-0.5em}
-\dfrac{S_1 <: T_1 \quad S_2 <: T_2}
-      {S_1 \, \texttt{+} \, S_2 <: T_1 \, \texttt{+} \, T_2}
-(\textsc{cong}\texttt{+}) \;\;\;
-\dfrac{S_1 <: T_1 \quad S_2 <: T_2}
-      {S_1 \, \texttt{*} \, S_2 <: T_1 \, \texttt{*} \, T_2}
-(\textsc{cong}\texttt{*})
+\dfrac{P_1 <:^r R_1 \quad P_2 <:^r R_2}
+      {P_1 \, \texttt{+} \, P_2 <:^r R_1 \, \texttt{+} \, R_2}
+\trule{\textsc{cong}\texttt{+}} \;\;\;
+\dfrac{P_1 <:^r R_1 \quad P_2 <:^r R_2}
+      {P_1 \, \texttt{*} \, P_2 <:^r R_1 \, \texttt{*} \, R_2}
+\trule{\textsc{cong}\texttt{*}} \\[1.5em]
+%%
+%% INEQUATIONS ON SPEC
+%%
+%\framebox{$\textit{spec} <: \textit{spec}'$} \\[0.8em]
+%
+\dfrac{}{\term{readOnce} \, R <: R}\trule{\textsc{rep}}
+\quad
+\dfrac{}{\term{atLeast} \, R <: \term{atMost} \,
+  R}\trule{\textsc{bound}} \\[1.5em]
+%%
+\dfrac{P <:^r R}{\term{atLeast}  P <: \term{atLeast} 
+  R}\trule{\textsc{cong}\texttt{L}}
+\ 
+\dfrac{P <:^r R}{\term{atMost}  P <: \term{atMost} 
+  R}\trule{\textsc{cong}\texttt{M}} \\[1.5em]
+\dfrac{P <:^r R}{\term{readOnce} \, P <: \term{readOnce} \,
+  R}\trule{\textsc{cong}\texttt{R}} 
+\quad
+\dfrac{R \equiv S}{R <:^r S}\trule{\textsc{eq}} 
+%%
 \end{array}
 \end{align*}
-\caption{Inequations on regions}
+\caption{Inequations $<:^r$ on \textit{regions} and
+$<:$ on \textit{specs} (omitting reflexivity
+and transitivity). }
 \label{fig:inequations}
+\label{-0.5em}
 \end{figure}
@@
 %\newcommand{\mkBig}[2]{\overset{\LARGE{\term{#1}}}{#2}}
-\begin{definition} The $\odot$ operation on \specDNF{} is defined
-\begin{align*}
-& ([\textit{approx}_1] \; [\textit{mult}_1] \;
-\tySum{s})
-\odot
-([\textit{approx}_2] \; [\textit{mult}_2] \;
-\tySum{t}) = \\
-& \begin{cases}
-\tySum{s} \tySum{\odot} \tySum{t} & \textit{approx}_1 =
-\textit{approx_2} = \varepsilon
+\begin{definition} The $\odot$ operation on $\textsf{A}(\specDNF{})$
+  is defined in terms of the intermediate $\hat{\odot}$ on \specDNF{}:
+%%
+\begin{align*}
+S \,\hat{\odot}\, \epsilon = S \quad
+\epsilon \,\hat{\odot}\, S = S \qquad
+S \,\hat{\odot}\, T = \bplus_{(\tyProd{s}, \tyProd{t}) \in (S \times
+   T)} (\tyProd{s} \; \term{*} \; \tyProd{t})
+\end{align*}
+%%
+That is, the $\hat{\odot}$ product of two specifications $S$ and $T$ in DNF form 
+is the $\term{+}$ sum of all pairwise $\term{*}$ products for every
+pair of $\tyProd{spec}$ drawn from $S$ and $T$. The result
+is indeed in DNF-form. The operation is
+commutative and associative, and sound with respect to our
+model (see Lemma~\ref{lem:alg-soundness} below).
+The $\hat{\odot}$ operation is then lifted to all combinations of
+\textsf{Approx} for $\odot$:
+\begin{align*}
+\textsf{exact}(S) \odot \textsf{exact}(T) = \; & \textsf{exact}(S
+                                              \hat{\odot} T) \\
+\textsf{lower}(S) \odot \textsf{lower}(T) = \; & 
+      \textsf{lower} \, (S \hat{\odot} T) \\
+\textsf{upper}(S) \odot \textsf{upper}(T) = \; &
+      \textsf{upper} \, (S \hat{\odot} T) \\
+\textsf{lower}(S) \odot \textsf{exact}(T) = \; &
+       %\textsf{exact}(T) \odot \textsf{lower}(S) = \;
+       \textsf{both} \, (S \hat{\odot} T, \; T) \\
+\textsf{upper}(S) \odot \textsf{exact}(T) = \; & 
+       %\textsf{exact}(T) \odot \textsf{upper}(S) =
+      \; \textsf{both} \, (T, \; S \hat{\odot} T) \\
+\textsf{lower}(S) \odot \textsf{both}(T_l,T_u) = \; &
+        %\textsf{both}(T_l,T_u) \odot \textsf{lower}(S) = \;
+        \textsf{both} \, (S \hat{\odot} T_l, \; T_u) \\
+\textsf{upper}(S) \odot \textsf{both}(T_l,T_u) = \; & \textsf{both} \,
+                                                      (T_l, \; S \hat{\odot} T_u) \\
+\textsf{both}(S_l,S_u) \odot
+\textsf{both}(T_l,T_u) & = \textsf{both} (S_l \hat{\odot} T_l, \; S_u
+                         \hat{\odot} T_u)
+\end{align*}
+For brevity we omit the cases where the parameters above are flipped
+as $\odot$ is commutative.
+\end{definition}
+
+\begin{definition}The $\oplus$ operation on $\textsf{A}(\specDNF)$ 
+is defined in a similar way to $\odot$, with $\hat{\oplus}$ on
+$\specDNF$ given by:
+%%
+\begin{equation*}
+S \,\hat{\oplus}\, \epsilon = S \quad
+\epsilon \,\hat{\oplus}\, S = S \quad
+S \, \hat{\oplus} \, T = S \, \term{+} \, T
+\end{equation*}
+%%
+Thus, if neither specification is empty, their $\oplus$ is just
+the syntatic sum $\term{+}$. Then $\oplus$ is the lifting of
+$\hat{\oplus}$ to boudns with the shape as $\odot$ in the previous
+definition.
+\end{definition}
+
+\begin{lemma}[$\hat{\odot}$, $\hat{\oplus}$ soundness] $\forall S,
+  T \in \specDNF$ where $S \neq \epsilon \wedge T \neq \epsilon$:
+\begin{align*}
+\interp{S \; \hat{\odot} \; T}_n & = \interp{S}_n \otimes \interp{T}_n
   \\
-\
-& \textit{approx}_1 = \texttt{atLeast}, \textit{approx}_2 = \varepsilon
-\end{cases}
-\\
-\end{align*}
-\begin{align*}
-s \tySum{\odot} t & =
- \bplus_{(\tyProd{s_1}, \tyProd{s_2}) \in (s \times
-   t)} (\tyProd{s_1} \; \term{*} \; \tyProd{s_2})
-%\begin{cases}%
-%\end{cases}
-\end{align*}
-\end{definition}
+\interp{S \; \hat{\oplus} \; T}_n & = \interp{S}_n \cup \interp{T}_n
+\end{align*}
+\label{lem:alg-soundness}
+\end{lemma}
+%
 

### 2016-07-06 paper.tex
Note on stencil approximations

@@
 
-\subsubsection{A note on region approximations}
+\subsubsection{A note on stencil approximations}
+
+As noted above, the \textsf{fivepoint} region defined above is more accepting
+than that used in Section~\ref{sec:lang}. This introduces a new notion of
+approximation beyond that is achieved in Section~\ref{sec:semantics}
+using upper and lower bounds because the in both cases the specifications are
+used to generate \emph{exact} models. Formally, we say a specification $s'$
+approximates a more specific specification $s$ \emph{iff} $\forall m \in
+\textsf{Mult}(\domainVal) . \mathit{cons}(s,m) \implies \mathit{cons}(s',m)$.
+
+Equations provided on regions in Section~\ref{sec:lang} when all modifiers
+equal provide such approximations, but it goes further than that. The
+regions describing the fivepoint stencil for example cannot be shown equal
+using these equations, but satisfies the approximation predicate as defined
+above.
+
+The checking algorithm though happily accepts an approximation in place of the
+most specific specification. Although this decreases the likelihood of catching
+stencil errors (as more access patterns conform), it reduces the space and
+effort it takes to write the specification.
 

### 2016-07-06 paper.tex
edits to section 3

@@
 
-\subsection{Equational theory and approximations}
+\subsection{Equational theory and sub-specifications}
 \label{sec:eqs}
@@
 normalisation in Section~\ref{sec:inf-step4}.
+
+The equational theory provides not only an axiomatic semantics
+for the language, which the user can use to rewrite their
+specification, but it is also used extensively in the inference
+procedure (Section~\ref{subsec:inference}) to simplify specifications.
 
@@
 \label{fig:inequations}
-\label{-0.5em}
+\vspace{-0.8em}
 \end{figure}
@@
 Both relations are reflexive and transitive (we omit
-these equations for $<:$ on \textit{spec}), and congruent
+these equations) and congruent
 with respect to \term{+}, \term{*}, \term{atLeast}, \term{atMost},
-and \term{readOnce}. The (\trule{\textsc{eq}}) axiom connects
-equations with inequations. The (\trule{\textsc{over}}) explains
-the notion of spatial over approximation via overlapping regions.
-The (\trule{\textsc{rule}}) allows the modifier that
-enforces that every index is read only once to be dropped
-as an approxiamtion. The (\trule{\textsc{bound}}) gives
+and \term{readOnce}. The (\trule{\textsc{eq}}) rule connects region
+equations with inequations. The (\trule{\textsc{over}}) rule explains
+the notion of spatial over-approximation via overlapping regions.
+The (\trule{\textsc{rep}}) allows the modifier that
+enforces that every index is read just once to be dropped
+as an approxiamtion; (\trule{\textsc{bound}}) gives
 a connection between the \term{atLeast} and \term{atMost}
@@
 
-\anote{Can we say something here about why these equations are useful i.e. we use them to do simplification of stencils?}
-
 \section{Semantics of specifications; a model}
@@
 
-We define a semantic model for our stencil specification language.
-This model has a number of purposes: (1) it is used in the inference,
-checking, and program synthesis algorithms (introduced in
-Section~\ref{sec:analysis}); (2) gives a basis for the equational
+We define a denotational model of the semantics of our 
+stencil specification language. This model has a number of purposes: (1) it is used in the inference,
+checking, and synthesis algorithms (Section~\ref{sec:analysis}); (2)
+justifies the equational
 theory and approximations of the previous section; and (3) can be used
-to guide correct implementations. The model is a denotational
-semantics, in terms of sets of vectors which we call \emph{index
-  schemes}. We introduce these first and then define the denotational semantics.
+to guide correct implementations. The model domain is over sets of vectors which we call \emph{index
+  schemes}, which we introduce first before the model in Section~\ref{subsec:model}.
 
@@
 
-\begin{definition}[Index scheme]
+\begin{definition}
 An \emph{index scheme} is a vector of size $n$ (an $n$-vector)
@@
 expressions of the form $i + a$ where $i$ is an induction variable
-and $a$ is a constant offset, Definition~\ref{def:neighbour},
-p.~\pageref{def:neighbour}). The additional value
+and $a$ is a constant offset, Def.\ref{def:neighbour},
+p.\pageref{def:neighbour}). The additional value
 $\infty$ represents any indexing behaviour--- it is a
 wildcard.
-Throughout $u, v, w$ range over index schemes, and $u_i$ denotes
-the $i^{th}$ element of $u$.
+Throughout $u, v, w$ range over index schemes and $u_i$ denotes
+the $i^{th}$ element of $u$. We write $(\domainVal)^n$ for
+the set of $n$-dimensional schemes.
 \end{definition}
@@
 maps syntactic array-index terms to indexing schemes. This
-operation forms the basis of the analysis required for the inference
-and the checking procedures in Section~\ref{sec:analysis}.
-
-\begin{definition}[Schematic]
+operation is used by the inference
+and the checking procedures.
+
+\begin{definition}%[Schematic]
 The partial function $\textsf{schematic}$ maps
-source-language indexing terms $(e_1, \ldots, e_n)$
-into $n$-dimensional index schemes
-$\in (\domainVal)^n$, defined:
+source indexing terms $(e_1, .., e_n)$
+to $n$-dimensional index schemes, defined:
 %
@@
 [\downarrow\!e_1 \ldots \downarrow\!e_n ]
-\quad \textit{iff} \; \forall i . \downarrow\!e_i \neq \bot
+\quad \textit{iff} \;\; \forall i . \downarrow\!e_i \neq \bot
 \end{align*}
@@
 The $\downarrow$ operator is a partial\footnote{We represent
-  partiality explicitly in our types with the set $A + \bot$.} function
+  partiality in our signatures with the set $A + \bot$.} 
+function
  $\downarrow : \textit{expr} \rightarrow (\domainVal + \bot)$
-which maps each individal indexing expression to a component of
-the indexing scheme, defined:
+mapping individal indexing expressions to 
+an indexing scheme component, defined:
 %%
@@
 if the expression is not a neighbourhood index but is not
-a relative index either (\ie{} is not defined in terms
+a relative index either (\ie{}, not defined in terms
 of an induction variable) then it is mapped to $\infty$, otherwise
@@
 
-For example $\textsf{schematic}(i, j+1) = \vtwoh{0}{1}$ assuming
-$i$ and $j$ are induction variables and $\textsf{schematic}(i-1, 2, j)
+For example, assuming $i$ and $j$ are induction variables and $\textsf{schematic}(i-1, 2, j)
  = \vthreeh{-1}{\infty}{0}$. However, $\textsf{schematic}(i*2, j)$, for
@@
 \subsection{Denotational model}
+\label{subsec:model}
 
@@
 pair of $\tyProd{spec}$ drawn from $S$ and $T$. The result
-is indeed in DNF-form. The operation is
+is indeed in DNF-form, where $\odot$ is essentially applying
+ the \trule{\textsc{dist}} rule of our equational theory
+ (Figure~\ref{fig:equations}). The operation is
 commutative and associative, and sound with respect to our

### 2016-07-06 paper.tex
edits to model

@@
 \newcommand{\vect}[1]{\textbf{#1}}
+
+\newcommand{\vtwohs}[2]{\setlength{\arraycolsep}{0em}
+\left[\begin{array}{cc}#1 \, & \, #2\end{array}\right]\!}
+
 \newcommand{\vtwoh}[2]{\setlength{\arraycolsep}{0em}
 \left[\begin{array}{cc}#1 \; & \; #2\end{array}\right]}
+
 \newcommand{\vthreeh}[3]{\setlength{\arraycolsep}{0em}
@@
 %
-The $\downarrow$ operator is a partial\footnote{We represent
+The $\downarrow$ operator is partial\footnote{We represent
   partiality in our signatures with the set $A + \bot$.} 
-function
  $\downarrow : \textit{expr} \rightarrow (\domainVal + \bot)$
@@
 Given two schemes $v, w$ of dimensionality $n$,
-the $w$ is said to \emph{generalise} $v$
+then $w$ is said to \emph{generalise} $v$
 (or conversely $v$ \emph{specialises} $w$),
@@
 in place of some $\mathbb{Z}$ in $v_i$ for $v \preceq_n w$.
-Thus, the element $\infty$
-generalising any index in the same dimension.
-
-The subscript on $\preceq_n$ is sometimes dropped when the
-intended dimensionality is clear or inconsequential.
+Thus, the element $\infty$ generalises any index in the same dimension.
+%
+%The subscript on $\preceq_n$ is sometimes dropped when the
+%intended dimensionality is clear or inconsequential.
 \end{definition}
@@
 \begin{example}
-Take three array indexing terms
-$(i, j+1)$, $(1, j+1)$ and $(3, j)$.
-As given by \textsf{schematic}, these have the
-corresponding schemes: $\vtwoh{0}{1}$, $\vtwoh{\infty}{1}$
+The following three array indexing terms
+$(i, j+1)$, $(1, j+1)$ and $(3, j)$ 
+are given by \textsf{schematic} the 
+index schemes $\vtwoh{0}{1}$, $\vtwoh{\infty}{1}$
 and $\vtwoh{\infty}{0}$ respectively. The following
@@
 %%
-\begin{equation*}
-\vtwoh{\infty}{0} \; \not\preceq_2 \; \vtwoh{0}{1}
-\qquad
-\vtwoh{\infty}{1} \; \preceq_2 \; \vtwoh{0}{1}
-\end{equation*}
+$\vtwoh{\infty}{0} \; \not\preceq_2 \; \vtwoh{0}{1}$
+and
+$\vtwoh{\infty}{1} \; \preceq_2 \; \vtwoh{0}{1}$. 
 %%
@@
 are $\infty$, \eg{}, $\zeroEntry{1}{2} = \vtwoh{0}{\infty}$.
-
 Similarly to the single-entry vector, zero-entry vectors
@@
 
-\begin{figure}
+\begin{figure}[t]
 \input{region-model.tex}
@@
 \label{fig:region-model}
+\vspace{-0.7em}
 \end{figure}
@@
 \noindent
-The first four equations of Figure~\ref{fig:region-model}
-give the meanings of the region constants as sets of indexing
-schemes. The next two give the meaning of combining regions
-by $\term{+}$ and $\term{*}$.
+The first four equations of Fig.~\ref{fig:region-model}
+give the model of the region constants as sets of indexing
+schemes. We illustrate with an example.
 
 \begin{example}
-For a $2$-dimensional situation,
-\begin{align*}
-& \interp{\stenCenS{2}{1}}_2 \\
-= \; & \{i\singleEntry{1}{2} \,|\, i \in \{\text{-}2,..., 2\}\!\setminus\!0 \}
-  \cup  \{ \zeroEntry{1}{2} \,|\, \textit{true} \} \\
-= \; & \{-2\singleEntry{1}{2}, \, -\singleEntry{1}{2}, \,
+For a $2$-dimensional stencil computation, then
+\begin{align*}
+& \interp{\stenCenS{2}{1}}_2 \\[-0.1em]
+& \qquad = \; \{i\singleEntry{1}{2} \,|\, i \in \{\text{-}2,..., 2\}\!\setminus\!0 \}
+  \cup  \{ \zeroEntry{1}{2} \,|\, \textit{true} \} \\[-0.2em]
+& \qquad = \; \{-2\singleEntry{1}{2}, \, -\singleEntry{1}{2}, \,
   \singleEntry{1}{2}, \, 2\singleEntry{1}{2}\} \cup
-  \{\zeroEntry{1}{2}\} \\
-= \; & \{\vtwoh{-2}{\infty}, \vtwoh{-1}{\infty}, \vtwoh{0}{\infty},
+  \{\zeroEntry{1}{2}\} \\[-0.2em]
+& \qquad = \; \{\vtwoh{-2}{\infty}, \vtwoh{-1}{\infty}, \vtwoh{0}{\infty},
 \vtwoh{1}{\infty}, \vtwoh{2}{\infty}\}
 \end{align*}
-Note the absboring behaviour of $\infty$ with respect to
+Note the absorbing behaviour of $\infty$ with respect to
 multiplication.
@@
 \noindent
-The model of $\term{+}$ is the straightforward union of
-the models of subterms. The model of $\term{*}$ is
-the tensor $\otimes$ of subterm models. The tensor
-is non-trivial; we give some informal intutition first.
+The last two equations of Fig.~\ref{fig:region-model} give the
+models of $\term{+}$ and $\term{*}$. For $\term{+}$ this is the 
+straightforward union of the models of subterms. The model of
+$\term{*}$ takes the tensor $\otimes$ of subterm models. The tensor is
+non-trivial; we give some informal intutition first.
 
@@
 remaining vertices of the rectangle are generated (like a bounding
-box). \eg{}:
-%%
-\begin{align*}
-\{\vtwoh{1}{2}, \vtwoh{5}{6}\} \otimes \{\vtwoh{3}{4}\} =
-\{&\vtwoh{1}{2}, \vtwoh{3}{2}, \vtwoh{1}{4}, \vtwoh{3}{4}, \\
-  & \vtwoh{3}{6}, \vtwoh{5}{4}, \vtwoh{5}{6}\}
+box) \eg{}:
+%%
+\begin{align*}
+\{\vtwohs{1}{2}, \vtwohs{5}{6}\,\} \otimes \{\vtwohs{3}{4}\} =
+\{\vtwohs{1}{2}, \vtwohs{3}{2}, \vtwohs{1}{4}, \vtwohs{3}{4}, 
+   \vtwohs{3}{6}, \vtwohs{5}{4}, \vtwohs{5}{6}\,\}
 \end{align*}

### 2016-07-06 paper.tex
edits to section 3 and 4

@@
 \begin{definition}%[Schematic]
+\label{def:schematic}
 The partial function $\textsf{schematic}$ maps
@@
 between two values $u_i$ and $v_i$, but if either
-is $\infty$ then the resultying pairwise permutation is
+is $\infty$ then the resulting pairwise permutation is
 filtered out if the dimension $i$ has \emph{any other scheme
@@
 %
-The notion of which dimensioons $i$ have that \emph{any scheme in the model
+The notion of which dimensions $i$ have that \emph{any scheme in the model
 has $u_i \neq \infty$} is represented by the \emph{constrained
@@
 %
-\begin{definition}[Constrained dimensions]
-Given an $n$-dimensional model $M$, the \emph{constrained dimensions}
-of that model is a subset of $\{1, \ldots, n\}$ of
-the dimensions for which at least one scheme $u \in M$ has
+\begin{definition}%[Constrained dimensions]
+For an $n$-dimensional model $M$, its \emph{constrained dimensions}
+is the set of dimension identifiers for which at least one scheme $u \in M$ has
 a non-$\infty$ value in that dimension. That is:
@@
   \Bigg\{x \; \Bigg| \;
-    \parbox{6cm}{$(u, v) \in M \times N $ \\[0.1em]
+    \parbox{5.2cm}{$(u, v) \in M \times N $ \\[0.1em]
                  $\;\; \wedge \;\, j \in (\effdims{M}{n} \cup
@@
 The first guard takes the Cartesian product of
-the two models and quantifies over all pairings $(u, v)$.
-The second guard says that $j$ ranges of the constrained
-dimensions for $M$ unioned with the constrained dimensions for $N$.
+the models and quantifies over all pairings $(u, v)$.
+The second guard binds $j$ which ranges over the constrained
+dimensions for $M$ union with the constrained dimensions for $N$.
 That is, every $j$ corresponds to a dimension in which
@@
 
-The $\bowtie$ \emph{pairwise permutation}
-constructs a $2^n \times n$ matrix ($2^n$-vector of
-$n$-vectors), which is indexed by $i$ above
-and then $j$. Using this, the $\otimes$ operation constructs the set
-of schemes which are the
-pairwise permutations of every pair $(u, v)$ of schemes in the Cartesian
-product of $M \times N$. These permuted schemes are only those
-which have non-$\infty$ components in the constrained dimensions
-given by each model, given by the guard $x_j \neq \infty$.
-
-The pairwise permutation $u \bowtie v$ takes two vectors and builds the matrix
-where the rows are all possible $n$-vectors generated by
-non-deterministic picking each $i^{th}$ entry from either the
-$i^{th}$ entry in $u$ or the $i^{th}$ entry in $j$. For example:
+The \emph{pairwise permutation} $u \bowtie v$ 
+builds a $2^n \times n$ matrix 
+of all possible $n$-vectors generated by
+non-deterministically picking for each $j^{th}$ entry either 
+$u_j$ or $v_j$ \eg{}
 %
@@
 %
+The permutated schemes $x = (u \bowtie v)_i$ added to the model 
+$M \otimes N$ are only those which have do not have $\infty$ in 
+the constrained dimensions given by each model ($x_j \neq
+\infty$). This filters out all permutations which have
+$\infty$ in a constrained dimension position. 
 \end{definition}
@@
 
-\begin{definition} The \textsf{Mult}
-and \textsf{Approx} are parametric labelled variant (sum) types
+\begin{definition} \textsf{Mult}
+and \textsf{Approx} are parametric labelled variant types
 with injection function given by their definition:
@@
 \ie{}, $\mathsf{lower}$ is an injection $\mathsf{lower} : a \rightarrow \mathsf{Approx} \, a$, and so
-on. These are used for labelled models, and other components later in
-Section~\ref{sec:analysis} with additional information.
+on. The $\textsf{Mult}$ type corresponds to the presence or absence of the
+\term{readOnce} modifier as shown in Figure~\ref{fig:spatial-model}. 
+The $\textsf{Approx}$ type corresponds to the presence or absence of
+the spatial approximation modifier, with \textsf{exact} when there is
+no such modifier and \textsf{lower} and \textsf{upper} for
+\term{atLeast} and \term{atMost}. Later, it will be useful to model
+pairs of lower and upper bounds which is provided by \textsf{both} 
+(but which isn't used here). These are used for labelled models, and other components later in
+Section~\ref{sec:analysis}.% with additional information.
 \label{def:mult-and-approx}
 \end{definition}
-%
 
@@
 \label{fig:spatial-model}
+\vspace{-0.75em}
 \end{figure}
@@
 
-Given the model defined in this section we can
+Given the above model, we can
 consider soundness for the equational theory of
@@
 $\textsf{Approx}(\mathcal{P}(\domainVal{}^n))$ defined by the
-following caluses
-\begin{align*}
+following clauses:
+\begin{align*}
+\def\arraystretch{1.2}
+\begin{array}{c}
+\begin{array}{cc}
 (M \leq' N \Rightarrow \textit{inj} \, M \leq \textit{inj} \, N) &
-\qquad 
 (\textsf{once} \, M \leq' \textsf{mult} \, M) \\
@@
 (M \subseteq N \Rightarrow \textit{inj}' M \leq' \textit{inj}' N)
-& \;\;\; (\textsf{exact} M \leq  \textsf{lower} M)
-\;\;\; (\textsf{lower} M \leq  \textsf{upper} M)
+& (\textsf{exact} M \leq  \textsf{lower} M)
+\end{array} \\ 
+(\textsf{lower} M \leq  \textsf{upper} M)
+\end{array}
 \end{align*}
@@
 \noindent
-We briefly sketch the algorithms for inferring and checking
-specifications and for synthesising programs from specifications.
-Checking and synthesis use the set-based semantic model of
+We briefly sketch the algorithms for inferring and synthesising
+specifications (Sect.~\ref{subsec:inference}), for checking code
+against specifications (Sect.~\ref{subsec:checking}) and for
+synthesising programs from specifications
+(Sect.~\ref{subsec:synthesis}).  These utilise the semantic model of
 Section~\ref{sec:semantics}.
 
-\paragraph{Syntax functions}
-The function \rhsExp{} maps statements to a
+\paragraph{Syntax helper functions}
+The function \rhsExp{} maps source statements to a
 set of expressions which occur in right-hand positions (\ie{}, not the
@@
 
-\subsection{Inference\label{subsec:inference}}
-
-We demonstrate the main steps of the inference procedure over the
+\subsection{Inference and specification synthesis\label{subsec:inference}}
+
+We demonstrate the inference procedure over the
 following example program which computes the mean value
-of a five-point stencil at each index of the input array:
+of a five-point stencil:%at each index of the input array:
 %%
@@
 \begin{align*}
+\hspace{-0.5em}
 \begin{array}{lr}
@@
  := \qquad\qquad & \textit{where} \; \neigh(\overline{e_1}) \wedge \arrayTy(v)  \\[0.3em]
-\multicolumn{2}{l}{\qquad \bigcup\{v' \mapsto \{\textsf{schematic}(\bar{e})\} \mid v'(\bar{e}) \leftarrow \mathsf{flowsTo}(e_2),
+\multicolumn{2}{l}{\qquad \bigcup\{v' \mapsto \{
+\textsf{schematic}(\bar{e})\} \mid v'(\bar{e}) \leftarrow \mathsf{flowsTo}(e_2),
   \arrayTy(v')\}}
@@
 %
-That is, we focus on assignments to an array subscript where each
-indexing expression in $\bar{e}$ is a neighbourhood index.  For
+That is, we focus on assignments to an array subscript where the LHS 
+indexing expression $\overline{e_1}$ is a neighbourhood index.  For
 all array subscripts that flow to the right-hand side of this
 statement, a finite map is constructed, mapping each array variable
-in this flow a set of indexing schemes for its subscripts, computed
-with \textsf{schematic}.
+to a set of indexing schemes for its subscripts, computed
+with \textsf{schematic} (Defn.~\ref{def:schematic},
+p.~\pageref{def:schematic}).
+ Note that \textsf{schematic} is undefined if
+$\bar{e}$ contains relative indices which are not neighbourhood
+indices (\ie{}, these are not stencil computations we can handle).
 
@@
 %
+If the LHS $\overline{e_1}$ contains non-$0$ neighour offsets 
+then all RHS schemes are \emph{relativized} by
+$\textsf{schematic}(\overline{e_1})$. For example, 
+if \texttt{a(i+1) = b(i)} then relativisation of the RHS by the LHS
+produces an analysis which is equivalent to the analysis of \texttt{a(i) = b(i-1)}.
 

### 2016-07-07 paper.tex
fix use of subfigure
more related work

@@
 %\usepackage{subcaption}
-\usepackage{subfigure}
 \usepackage{siunitx} % For pretty-printing numeric values and SI units
@@
     iteration program. The correct output is shown in
-    (\subref{fig:jacobicorrect}).}
+    \subref{fig:jacobicorrect}.}
   \label{fig:jacobi}
@@
 
-% Abe, 2013, Model checking stencil computations
-% -- Unable to obtain copy
-% In an alternate approach,~\citet{Abe:2013:IPDPSW}
-
 There are many DSLs for stencil computation aimed at either
@@
 data races, while we are interested in the correctness of stencils
-embedded within a more typical general-purpose programming language.
+embedded within a more typical general-purpose programming
+language. Other work on GPGPU computation, such as
+\citet{Zhang:2012:CGO}, has been primarily focused on generating
+optimised code based upon on relatively simple specifications: to
+provide performance while keeping the programmer's effort within
+reason.
 
@@
 that are significantly faster than the typical looping implementations
-of stencil computations.
-
-% 3D stencil codes
-%\citet{Zhang:2012:CGO}
+of stencil computations. Like much of the related work, the main goal
+is optimisation rather than catching bugs and verification of
+program correctness.
+
+By contrast, the work of \citet{Abe:2013:IPDPSW} has looked at the
+correctness problem explicitly by bringing a form of model-checking to
+verify certain stencil computations in partitioned global address
+space languages. These are scenarios where an array is divided into
+subarrays on multiple processors but each has the ability to access
+the others' memory. The authors avoid the state explosion dilemma of
+model-checking by relying upon the fact that accesses to the
+``boundary elements'' will require a different method by virtue of the
+non-locality of that memory. Since this access is performed in a
+different manner, those can be identified automatically by static
+analysis. To achieve this, the authors have designed a new language to
+be used for writing stencil computations. Much of the specification
+effort goes towards describing the nuts and bolts of distributing the
+computation over multiple processors. The code for the stencil kernel
+is generated from a relatively high-level specification. This approach
+differs from Camfort because we are interested in integrating directly
+into existing, legacy codebases and long-established languages such as
+older versions of Fortran. We infer stencil specifications from
+Fortran code and check annotations on stencil computations within
+Fortran code.
 

### 2016-07-07 paper.tex
improving inference part

@@
 
-\subsubsection{Step 3: Coalesce contiguous indices into regions}
+\subsubsection{Step 3: Coalesce contiguous schemes into spans}
 \label{sec:inf-step3}
 
-Let $M$ range over the finite maps generated by \textsf{analyse}.  For
-each $v \in \mathsf{dom}(M)$, the algorithm then constructs a group
-of regions (called \emph{spans}) which cover all contiguous groups of relative indices
-(from $M(v)$) in the $n$-dimensional space.
-Informally, indexing schemes are converted to pairs $u \mapsto (u,
-u)$, defining the lower and upper bound of a unit span
-in $n$-dimensions. For
-example, for the five point stencil there are five $1 \times 1$ spans.
-Figure~\ref{fig:inference-steps-informal}(a) gives an illustration
-of these five unit spans.
+Let $U$ range over the finite maps generated by \textsf{analyse}.  For
+each $x \in \mathsf{dom}(U)$, the next step of the algorithm
+constructs a set of $n$-dimensional rectangles covering all contiguous groups of schemes
+in $U(x)$.
+
+\newcommand{\spanOp}{\textsf{spans}}
+
+\begin{definition}[Spans]
+  A \emph{span} represents an $n$-dimensional box (\emph{hyper-rectangle}) as
+  a pair of $n$-dimensional vectors (or represented as a $2 \times n$
+  matrix) with values drawn from $\domainVal{}$. This  
+   gives the co-ordinates of a lower-bound vertex (first
+  column) and the upper-bound vertex (second column). We use the
+  notation $\vect{x} = [\vect{x}^L \; \vect{x}^U]$, where $\vect{x}^L$
+  and $\vect{x}^U$ give the lower and upper bounds of the span.
+In the rest, $\vect{x}, \vect{y}$ range over spans. 
+For a span $\vect{x}$, we write $\vect{x}_{i:j}$ for the sub-span
+comprises the subvectors $[\vect{x}^L_{i:j} \; \vect{x}^U_{i:j}]$ from entry $i$ to
+entry $j$ (inclusive).
+\end{definition}
+
+\noindent
+The indexing schemes of $U(x)$ are converted to \emph{unit} spans $u \mapsto [u \; 
+u]$, defining the lower and upper bound.
+ For the five-point stencil example there are five $1 \times 1$ spans.
+Figure~\ref{fig:inference-steps-informal}(a) illustrates these.
 
@@
 \begin{array}{c}
-\quad \\[0.2em]
-\begin{tikzpicture}[scale=0.75]
+\quad \\[0em]
+\begin{tikzpicture}[scale=0.8]
 %\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
-\node at (1.25,1.22) {\textnormal{\small{0,0}}};
+\node at (1.25,1.22) {\textnormal{\footnotesize{0,0}}};
 \draw[thick] (1,1) rectangle (1.5,1.5);
@@
 \draw[thick] (1,1.5) rectangle (1.5,2);
-\end{tikzpicture} \\[0.36em]
+\end{tikzpicture} \\[0em]
 (a)
@@
 \begin{array}{c}
-\begin{tikzpicture}[scale=0.75]
+\begin{tikzpicture}[scale=0.72]
 %\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
@@
 \draw [black, thin] (4.4,0) to [round right paren ] (4.4,2);
-\end{tikzpicture} \\ (b)
+\end{tikzpicture} \\[-0.5em] (b)
 \end{array}
@@
 \begin{array}{c}
-\begin{tikzpicture}[scale=0.75]
+\begin{tikzpicture}[scale=0.72]
 %\node at (1.25,1.22) {\textnormal{\small{0,0}}};
@@
 %\node at (2.75,1.22) {\textnormal{\small{0,0}}};
-\end{tikzpicture} \\ (c)
+\end{tikzpicture} \\[-0.5em] (c)
 \end{array}
@@
 \label{fig:inference-steps-informal}
+\vspace{-0.75em}
 \end{figure}
 
-%
-Each dimension is traversed, coalescing consecutive spans
-which vary in $m$-dimensions into coalesced spans which vary in
-$m+1$-dimensions. In this example, the $0$-dimensional spans
- become $1$-dimensional spans (rows/columns), illustrated
-in Figure~\ref{fig:inference-steps-informal}(b).
-Any spans that are contained with any other span are deleted,
+% taking permustations is a systematic way of producing all rotations
+% of the n-dimensional space - forward ref to fold
+The next step of the algorithm repeatedly coalesces any contiguous spans
+until a fixed point is reached, resulting in a set of spans covering
+over the original space of schemes. Two $n$-dimensional spans $\vect{x}$ and $\vect{y}$
+are contiguous if for all but one $j \in \{1, \ldots, n\}$ then
+$\vect{x}_j = \vect{y}_j$ and for one $j$ then $\vect{x}^U_1 + 1 =
+\vect{y}^U$. For example, unit spans
+$\left[\vtwoh{-1}{0}\; \vtwoh{-1}{0}\right]$ and $[\vtwoh{0}{0} \; \vtwoh{0}{0}]$ 
+are contiguous in the first dimension and can be coalesced
+into the span $[\vtwoh{-1}{0} \; \vtwoh{0}{0}]$.
+
+\begin{definition}\label{def:span-coalesc} Our operator for span
+  coalescing $\bullet$ is defined for spans contiguous in the first dimension:
+%%
+\begin{align*}
+\vect{x} \bullet \vect{y}
+= &
+\begin{cases}
+(\vect{x}^L, \vect{y}^U) & \vect{x}^U_1 + 1 = \vect{y}^L_1 \; \wedge \;
+\vect{x}_{2:n} = \vect{y}_{2:n}) \\
+\bot  & \textit{otherwise}
+\end{cases}
+% $m$-dimensions into coalesced spans which vary in
+% $m+1$-dimensions.
+\end{align*}
+%%
+\end{definition}
+In order to coalesce regions which are contiguous in other dimensions
+other than the first, we generate all rotations of the spans in
+$n$-dimensional space by permutation, and fold together contiguous
+regions using the above $\bullet$ operation. For our example,
+from the unit spans in Figure~\ref{fig:inference-steps-informal}(a)
+the contiguous spans in Figure~\ref{fig:inference-steps-informal}(b)
+are generated. Then any spans that are contained with any other span are deleted,
 leaving a minimal set of spans (which may overlap, but none of which
@@
 example this is reached in the first step. The final spans
-are used to construct stencil specification, where
+are used to synthesis concrete stencil specification syntax, where
 each span is product \texttt{*} of regions and multiple spans are
@@
 
-\newcommand{\spanOp}{\textsf{spans}}
-
-Let $\vect{x}, \vect{y}, \vect{z}$ range over the column vectors of size
-$n$ whose values are drawn from $\domainVal{}$.
-We write $\vect{x}_{i:j}$ for the sub-vector from entry $i$ to
-entry $j$ (inclusive).
-
-\begin{definition}[Spans]
-  A \emph{span} represents an $n$-dimensional box (\emph{hyper-rectangle}) as
-  a pair of $n$-dimensional vectors (represented as a $2 \times n$
-  matrix) giving the co-ordinates of the lower-bound vertex (first
-  column) and the upper-bound vertex (second column). We use the
-  notation $\vect{x} = [\vect{x}^L \; \vect{x}^U]$, where $\vect{x}^L$
-  and $\vect{x}^U$ give the lower and upper bounds of the span.
-\end{definition}
-
-The algorithm to create contiguous spans, covering the space of
+More formally, the algorithm to create contiguous spans, covering the space of
 indices, now proceeds as follows.
@@
 %
-\begin{align*}
-& (\vect{x}^L,\vect{x}^U) \bullet (\vect{y}^L,\vect{y}^U) \\
-= &
-\begin{cases}
-(\vect{x}^L, \vect{y}^U) & \vect{x}^U_1 + 1 = \vect{y}^L_1 \; \wedge \;
-\vect{x}_{2:n} = \vect{y}_{2:n}) \\
-\bot  & \textit{otherwise}
-\end{cases}
-\end{align*}
+
 For our example, this yields:

### 2016-07-07 paper.tex
section 4 edits

@@
 \usepackage{breqn}
+\usepackage{enumitem}
 %\usepackage{subcaption}
@@
 %  \\[0.5em]
-\nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+\\[-1em]
+\nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+\\[-0.5em]
 \end{array}
@@
 \label{fig:syntax}
-\vspace{-0.5em}
+\vspace{-0.8em}
 \end{figure}
@@
 \begin{figure}
-\vspace{-0.2em}
+%\vspace{-0.2em}
 \begin{align*}
@@
 
-Let $U$ range over the finite maps generated by \textsf{analyse}.  For
-each $x \in \mathsf{dom}(U)$, the next step of the algorithm
-constructs a set of $n$-dimensional rectangles covering all contiguous groups of schemes
+Let $U$ range over the finite maps computed by \textsf{analyse}.  For
+each $x \in \mathsf{dom}(U)$, the algorithm 
+constructs next a set of $n$-dimensional rectangles covering all contiguous groups of schemes
 in $U(x)$.
@@
 \begin{definition}[Spans]
-  A \emph{span} represents an $n$-dimensional box (\emph{hyper-rectangle}) as
-  a pair of $n$-dimensional vectors (or represented as a $2 \times n$
-  matrix) with values drawn from $\domainVal{}$. This  
-   gives the co-ordinates of a lower-bound vertex (first
-  column) and the upper-bound vertex (second column). We use the
-  notation $\vect{x} = [\vect{x}^L \; \vect{x}^U]$, where $\vect{x}^L$
-  and $\vect{x}^U$ give the lower and upper bounds of the span.
-In the rest, $\vect{x}, \vect{y}$ range over spans. 
-For a span $\vect{x}$, we write $\vect{x}_{i:j}$ for the sub-span
-comprises the subvectors $[\vect{x}^L_{i:j} \; \vect{x}^U_{i:j}]$ from entry $i$ to
-entry $j$ (inclusive).
+  A \emph{span} represents an $n$-dimensional box
+  (\emph{hyper-rectangle}) as a pair of $n$-dimensional vectors (or
+  represented as a $2 \times n$ matrix) with values drawn from
+  $\domainVal{}$. This gives the co-ordinates of a lower-bound vertex
+  (first comonent) and the upper-bound (second component). We write 
+  spans as $\vect{x} = [\vect{x}^L \; \vect{x}^U]$, where $\vect{x}^L$
+  and $\vect{x}^U$ give the lower and upper bounds.  In
+  the rest, $\vect{x}, \vect{y}$ range over spans.  For a span
+  $\vect{x}$, we write $\vect{x}_{i:j}$ for the sub-span comprises the
+  subvectors $[\vect{x}^L_{i:j} \; \vect{x}^U_{i:j}]$ from entry $i$
+  to entry $j$ (inclusive).
 \end{definition}
@@
 \noindent
-The indexing schemes of $U(x)$ are converted to \emph{unit} spans $u \mapsto [u \; 
+Each indexing scheme in $U(x)$ is converted to a \emph{unit} span $u \mapsto [u \; 
 u]$, defining the lower and upper bound.
@@
 \label{fig:inference-steps-informal}
-\vspace{-0.75em}
+\vspace{-0.8em}
 \end{figure}
@@
 % of the n-dimensional space - forward ref to fold
-The next step of the algorithm repeatedly coalesces any contiguous spans
-until a fixed point is reached, resulting in a set of spans covering
-over the original space of schemes. Two $n$-dimensional spans $\vect{x}$ and $\vect{y}$
+The next step repeatedly coalesces any contiguous spans
+until a fixed point is reached, resulting in a set of (possibly
+overlapping) spans covering the original space of schemes. Two $n$-dimensional spans $\vect{x}$ and $\vect{y}$
 are contiguous if for all but one $j \in \{1, \ldots, n\}$ then
 $\vect{x}_j = \vect{y}_j$ and for one $j$ then $\vect{x}^U_1 + 1 =
-\vect{y}^U$. For example, unit spans
-$\left[\vtwoh{-1}{0}\; \vtwoh{-1}{0}\right]$ and $[\vtwoh{0}{0} \; \vtwoh{0}{0}]$ 
+\vect{y}^U$. For example, spans
+$\left[\vtwoh{-1}{0}\; \vtwoh{-1}{0}\right]$ and $[\vtwoh{0}{0} \; \vtwoh{1}{0}]$ 
 are contiguous in the first dimension and can be coalesced
-into the span $[\vtwoh{-1}{0} \; \vtwoh{0}{0}]$.
-
-\begin{definition}\label{def:span-coalesc} Our operator for span
+into the span $[\vtwoh{-1}{0} \; \vtwoh{1}{0}]$.
+
+\begin{definition}\label{def:span-coalesc} Span
   coalescing $\bullet$ is defined for spans contiguous in the first dimension:
@@
 \end{definition}
-In order to coalesce regions which are contiguous in other dimensions
+\noindent
+To coalesce regions which are contiguous in other dimensions
 other than the first, we generate all rotations of the spans in
-$n$-dimensional space by permutation, and fold together contiguous
-regions using the above $\bullet$ operation. For our example,
-from the unit spans in Figure~\ref{fig:inference-steps-informal}(a)
-the contiguous spans in Figure~\ref{fig:inference-steps-informal}(b)
-are generated. Then any spans that are contained with any other span are deleted,
+$n$-dimensional space by permutation on spans, and fold together contiguous
+regions using $\bullet$ (above). This 
+generates all possible ways of coalescing regions. For our example,
+the unit spans in Figure~\ref{fig:inference-steps-informal}(a)
+are colaesced into the contiguous spans in
+Figure~\ref{fig:inference-steps-informal}(b).
+This procedure is iterated until a fixed point is reached. In this
+example this is reached in the first step. 
+ Then any spans that are completely contained by any other span are deleted,
 leaving a minimal set of spans (which may overlap, but none of which
-fully contains another) (Figure~\ref{fig:inference-steps-informal}(c).
-
-This procedure is iterated until a fixed point is reached. In this
-example this is reached in the first step. The final spans
-are used to synthesis concrete stencil specification syntax, where
-each span is product \texttt{*} of regions and multiple spans are
-combined with the region sum \texttt{+}.
-
-More formally, the algorithm to create contiguous spans, covering the space of
-indices, now proceeds as follows.
-Firstly, for all $v \in \mathsf{dom}(M)$ a new map $M'$ is
-created where each vector is mapped to the trivial \emph{unit span} is
-created by pairing a vector with itself:
-%
-\begin{align*}
-N(v) = \{[\vect{x} \;
-  \vect{x}] \mid \vect{x} \in M(v)\}
-\end{align*}
-%
-For our example, this yields:
-%%
-\begin{align*}
-N(\texttt{a}) = \stwo{0}{0}{0}{0} \stwo{1}{0}{1}{0} \stwo{-1}{0}{-1}{0} \stwo{0}{1}{0}{1} \stwo{0}{-1}{0}{-1}
-\end{align*}
-%%
-A fixed point is then computed for the following
-procedure \spanOp which, per variable
-in the map, coalesces spans into contiguous regions in
-the $n$-dimensional space. That is, we compute
- $(\mu \; \spanOp) N$ where \spanOp
-is defined as follows for each variable in the map:
-%
-\begin{enumerate}
-
-\item Compute all permutations on the column vectors in a span, \ie{},
+fully contains another), Figure~\ref{fig:inference-steps-informal}(c).
+The final spans are used to synthesis concrete stencil specification
+syntax (Section~\ref{sec:inf-step4}), where
+each span becomes a product \texttt{*} of regions which are 
+combined by the sum \texttt{+}.
+
+More formally the algorithm proceeds as follows.
+For all $x \in \mathsf{dom}(U)$, let $M = U(x)$. Every scheme in $M$ 
+is mapped to a unit span
+by $\textsf{unitSpans}(M) = \{[u \; u] \mid u\!\in\!M\}$. For our example:
+%%
+\begin{align*}
+N(U(\texttt{a})) = \stwo{0}{0}{0}{0} \stwo{1}{0}{1}{0} \stwo{-1}{0}{-1}{0} \stwo{0}{1}{0}{1} \stwo{0}{-1}{0}{-1}
+\end{align*}
+%%
+In the following we let $M_0 = N(U(\texttt{a}))$ for our example.
+
+A fixed point of the following function, \spanOp{}, is taken (per variable
+in the map U) to coalesces spans into contiguous regions, \ie{}, we compute
+ $(\mu \, \spanOp) \; N(M)$ where \spanOp{} is defined by: \\[-0.5em]
+
+\noindent
+\textit{1).\;} Compute all permutations on the column vectors in a span, \ie{},
   $[\vect{x}^L \vect{x}^U] \mapsto [\pi\vect{x}^L \, \pi\vect{x}^U]$
@@
 %
-\[
-P(v) = \bigcup_{i \in n!} (\pi^n_{i} , \; \{[\pi^n_i
+\begin{equation*}
+P(M) = \bigcup_{i \in n!} (\pi^n_{i} , \; \{[\pi^n_i
 \vect{x}^L \; \pi^n_i\vect{x}^U] \, \mid \, [\vect{x}^L \; \vect{x}^U]
-\leftarrow N(v)\}
-\]
-%
-For our example, this yields:
-%
-\begin{align*}
-P(\texttt{a}) =
+\leftarrow M\}
+\end{equation*}
+%
+This corresponds to all rotations of the space.  For our example:
+%
+\begin{align*}
+P(M_0) =
 \{&(\pi^2_1, \{ \stwo{0}{0}{0}{0},
@@
 where $\pi^2_1$ is the identity permutation and $\pi^2_2$ is the
-permutation flips the order of the two elements in the column
-vectors.
-
-\item Sort each permutation set into an ordered list, based on the ordering:
+permutation flips the order of the two elements in the 
+vectors. \\
+
+\noindent
+\textbf{2).} Sort each permutation set into an ordered list, by the ordering:
 \begin{equation*}
   \vect{x} \leq \vect{y} =
-      \exists i \, . \; \vect{x}_{i,1} \leq \vect{y}_{i,1} \; \wedge \;
+      \exists i \, . \; \vect{x}^L_{i} \leq \vect{y}^L_{i} \; \wedge \;
         (i = n \vee \vect{x}_{i+1:n} = \vect{y}_{i+1:n})
@@
 %
-This yields:
-%%
-\begin{align*}
+giving the sorting function $\textsf{sort}(M)$. For our example:
+%%
+\begin{align*}
+\textsf{sort}(P(M_0)) = 
 \{&(\pi^2_1, [
@@
 \end{align*}
-%%
-\item Fold each list pairwise by the following partial operation
- $\bullet$ which coalesces contiguous regions:
-%
-
-For our example, this yields:
-%
-\begin{align*}
-Q(v) = \{&(\pi^2_1, [
+\noindent
+\textbf{3).} Reduce each list pairwise by $\bullet$ operation 
+(Defn.~\ref{def:span-coalesc}) to coalesce contiguous regions,
+given by $\textsf{fold}\bullet$. For our example, this yields:
+%
+\begin{align*}
+\{&(\pi^2_1, [
 \stwo{0}{-1}{0}{-1},
@@
 \stwo{0}{1}{0}{1}
-]) \\
-&(\pi^2_2, [
+]), (\pi^2_2, [
 \stwo{0}{-1}{0}{-1},
@@
 %
-\item Un-permute and union together, \ie{},
+\textbf{4).} Un-permute and union together, \ie{},
 %
 \[
-U(v) = \bigcup \{[\pi \vect{x}^L, \pi \vect{x}^U]
- \mid [\vect{x}^L, \vect{x}^U] \leftarrow S, (\pi, S) \leftarrow Q(v)\}
+U(M) = \bigcup \{[\pi \vect{x}^L, \pi \vect{x}^U]
+ \mid [\vect{x}^L, \vect{x}^U] \leftarrow S, (\pi, S) \leftarrow M\}
 \]
-Which yields:
-%
-\begin{align*}
-U(\texttt{a}) =
+For our example:
+%
+\begin{align*}
+U(M) =
 \{\stwo{0}{-1}{0}{-1},
@@
 %
-\item Finally, filter by the region containment predicate $\containedin$, that
+\textbf{5).} Filter by the region containment predicate $\containedin$, that
   is, if any region is contained within another then remove the
-  smaller. The $\containedin$ predicate is defined:
-%
-\begin{align*}
-& (\vect{x}^L, \vect{x}^U) \containedin (\vect{y}^L, \vect{y}^U) = \\
-& \vect{y}^L_1 \leq \vect{x}^L_1 \wedge \vect{x}^U_1 \leq \vect{y}^U_1
+  smaller, defined:
+%
+\begin{align*}
+& \vect{x} \containedin \vect{y} = \vect{y}^L_1 \leq \vect{x}^L_1 \wedge \vect{x}^U_1 \leq \vect{y}^U_1
   \wedge (\vect{x}^L_{2:n}, \vect{x}^U_{2:n}) \containedin
@@
 and $\stwo{1}{0}{1}{0} \sqsubseteq \stwo{-1}{0}{1}{0}$.
-\end{enumerate}
-For our example, applying \spanOp again yields the same
-result, hence we have reached a fixed point.
-
-\subsubsection{Step 4: Convert spans into specifications}
+Applying \spanOp{} again yields the same
+result for our example; the fixed point is reached.
+
+\subsubsection{Step 4: Synthesise specifications from spans}
 \label{sec:inf-step4}
 
-\newcommand{\bplus}{\operatornamewithlimits{\term{\huge{+}}}}
+\newcommand{\bplus}{\operatornamewithlimits{\term{\Large{+}}}}
 \newcommand{\tySum}[1]{#1^{\term{+}}}
@@
 
-Next, the set of spans $\spanOp(M)$ covering the indexing
-space is converted into abstract syntax trees 
-of the specification language. This conversion is defined in terms
+Abstract syntax trees of the specification language
+are synthesised from the set $\spanOp(M)$ of coverings spans. 
+This conversion is defined in terms
 of an algebra on \textit{spec} syntax in disjunctive-normal form
-(which we denote \specDNF{}), and which is wrapped
-with the $\textsf{Approx}$ type (see
-Definition~\ref{def:mult-and-approx},
+(which we denote \specDNF{}) and which is wrapped
+in the $\textsf{Approx}$ type (see
+Defn.~\ref{def:mult-and-approx},
 p.~\pageref{def:mult-and-approx}), which we write as $\mathsf{A}$
 for brevity.  The algebra mirrors the shape
-of the region combinators for addition and multiplication, 
+of the region combinators for sum and product, 
 with operators $\oplus$ and $\odot$ and a constructor for \specDNF{}:
@@
 %%
-The $\mathsf{toSpec}$ operation takes a pair of lower and upper bounds
-values (one-dimensional) and constructs a specification 
-which contains a one-dimensional region. 
+The $\mathsf{toSpec}$ functions maps a pair of lower and upper bounds
+values (one-dimensional) to a specification with a one-dimensional region. 
 An entire span is converted into an $n$-dimensional
@@
 syntax (see Figure~\ref{fig:syntax}, p.~\pageref{fig:syntax}).
-The set \specDNF{} is a subset of syntax trees \textit{spec}
-of specifications in disjunctive normal form over regions 
+The set \specDNF{} is a subset of syntax trees \textit{spec},
+containing specifications in disjunctive normal form over regions 
 where the disjunct is \term{+} (since its model is in terms of set
@@
 %%
-\[
+$
 (r^1_{1} \; \term{*} \ldots \term{*} \; r^1_n)\; \term{+} \; \ldots 
 \term{+} \; (r^m_1 \, \term{*} \; \ldots \term{*} \; r^m_p)
-\]
-%%
+$.
+
 More formally, $\specDNF \subseteq (\textit{spec} \cup \{\epsilon\})$, defined:
@@
 syntax. 
-In the following, $s, t$ range over elements of
-\specDNF{}, $\tySum{s}, \tySum{t}$ range over elements of 
-$\tySum{\specDNF}$ and $\tyProd{s}, \tyProd{t}$ range over elements
+In the following, $S, T$ range over elements of
+\specDNF{} and $\tyProd{s}, \tyProd{t}$ range over elements
 of $\tyProd{\specDNF}$ (for example, $\tyProd{s} = r^1_{1} \; \term{*}
@@
 \begin{definition} The $\mathsf{toSpec}$ constructor is defined:
-%%
+
 %\newcommand{\llower}{\vect{x}^L_i}
@@
 \newcommand{\lupper}{u}
+{\scalebox{0.97}{
+\hspace{-1.8em}
+\begin{minipage}{1\linewidth}
 \begin{align*}
@@
 \end{align*}
-%%
+\end{minipage}
+}}\\
+
+\noindent
 The first two cases map to \texttt{forward} stencils since the lower
@@
 0-offset access in this dimension.
+
+Note, $\mathsf{toSpec} \, l \, u$ is always defined 
+ if $l$ less than or equal to $u$. 
 \end{definition}
-
-\begin{lemma}
-$\mathsf{toSpec}$ is total 
- given the precondition that the lower
-bound of the input span is less than or equal to its upper bound.
-\end{lemma}
 
@@
 \begin{definition} The $\odot$ operation on $\textsf{A}(\specDNF{})$
-  is defined in terms of the intermediate $\hat{\odot}$ on \specDNF{}:
+  is defined in terms of the intermediate operator $\hat{\odot}$ on
+  \specDNF{}, defined:
 %%
@@
 pair of $\tyProd{spec}$ drawn from $S$ and $T$. The result
-is indeed in DNF-form, where $\odot$ is essentially applying
- the \trule{\textsc{dist}} rule of our equational theory
- (Figure~\ref{fig:equations}). The operation is
+is indeed in DNF-form, where $\hat{\odot}$ is essentially applying
+ the (\trule{\textsc{dist}}) rule of our equational theory
+ (Fig.~\ref{fig:equations},p.~\pageref{fig:equations}). The operation is
 commutative and associative, and sound with respect to our
@@
 \begin{align*}
-\textsf{exact}(S) \odot \textsf{exact}(T) = \; & \textsf{exact}(S
-                                              \hat{\odot} T) \\
-\textsf{lower}(S) \odot \textsf{lower}(T) = \; & 
-      \textsf{lower} \, (S \hat{\odot} T) \\
-\textsf{upper}(S) \odot \textsf{upper}(T) = \; &
-      \textsf{upper} \, (S \hat{\odot} T) \\
+%\textsf{lower}(S) \odot \textsf{lower}(T) = \; & 
+%      \textsf{lower} \, (S \hat{\odot} T) \\
+%\textsf{upper}(S) \odot \textsf{upper}(T) = \; &
+%      \textsf{upper} \, (S \hat{\odot} T) \\
 \textsf{lower}(S) \odot \textsf{exact}(T) = \; &
@@
 \textsf{both}(T_l,T_u) & = \textsf{both} (S_l \hat{\odot} T_l, \; S_u
-                         \hat{\odot} T_u)
-\end{align*}
-For brevity we omit the cases where the parameters above are flipped
+                         \hat{\odot} T_u) \\
+\textit{inj}(S) \odot \textit{inj}(T) = \; & \textit{inj}(S
+                                              \hat{\odot} T)
+\end{align*}
+where in the last case $\textit{inj}$ corresponds to the unary
+injections of \textsf{Approx}. For brevity we omit the cases where the parameters above are flipped
 as $\odot$ is commutative.
@@
 \begin{definition}The $\oplus$ operation on $\textsf{A}(\specDNF)$ 
-is defined in a similar way to $\odot$, with $\hat{\oplus}$ on
+is defined in a similar way to $\odot$, with the intermediate $\hat{\oplus}$ on
 $\specDNF$ given by:

### 2016-07-07 paper.tex
checking editing

@@
 %%
-Here the left hand side of \texttt{+} says that when the second dimension
+Here the left-hand side of \texttt{+} says that when the second dimension
 (induction variable $j$) is fixed at the origin, the first dimension
@@
 \subsection{Indexing schemes and generalisation}
+\label{subsec:schemes-and-gen}
 
@@
 which flow to this statement. More formally, a function
-$\textsf{analyse}$ is applied to each statement in a program with the
+$\textit{analyse}$ is applied to each statement in a program with the
 following clause:
@@
 \begin{array}{lr}
-\textsf{analyse}(v(\overline{e_1}) = e_2)
+\textit{analyse}(v(\overline{e_1}) = e_2)
  := \qquad\qquad & \textit{where} \; \neigh(\overline{e_1}) \wedge \arrayTy(v)  \\[0.3em]
@@
 
-For our example, the \textsf{analyse} function matches on
+For our example, the \textit{analyse} function matches on
 line 5, with the following set for $\textsf{flowsTo}(\texttt{a(i, j) + x +
@@
 \end{align*}
-Subsequently the result of \textsf{analyse} on line 5 yields the map:
+Subsequently the result of \textit{analyse} on line 5 yields the map:
 \begin{align*}
@@
 
-Let $U$ range over the finite maps computed by \textsf{analyse}.  For
+Let $U$ range over the finite maps computed by \textit{analyse}.  For
 each $x \in \mathsf{dom}(U)$, the algorithm 
@@
 %%
-Thus, if neither specification is empty, their $\oplus$ is just
+Thus, if neither specification is empty, their $\hat{\oplus}$ is just
 the syntatic sum $\term{+}$. Then $\oplus$ is the lifting of
-$\hat{\oplus}$ to boudns with the shape as $\odot$ in the previous
+$\hat{\oplus}$ to \textsf{Approx} with the shape as $\odot$ in the previous
 definition.
@@
 \begin{align*}
-\interp{S \; \hat{\odot} \; T}_n & = \interp{S}_n \otimes \interp{T}_n
-  \\
-\interp{S \; \hat{\oplus} \; T}_n & = \interp{S}_n \cup \interp{T}_n
+\interp{S \; \hat{\odot} \; T}_n = \interp{S}_n \otimes \interp{T}_n
+  \qquad
+\interp{S \; \hat{\oplus} \; T}_n = \interp{S}_n \cup \interp{T}_n
 \end{align*}
 \label{lem:alg-soundness}
+\vspace{-2em}
 \end{lemma}
 %
-
-\subsection{Checking\label{subsec:checking}}
-
-In checking mode, the specification is verified with respect to the indexing
-scheme of the actual access pattern. The mathematical model of the specification
-developed in Section~\ref{sec:semantics} is used this process.
-
-Checking procedure synthesises syntactic specification into its model and
-generates an injection from indexing schemes to the multiplicity of observed
-access patterns, this is formally defined in Figure~\ref{fig:consistency}.
-Then, the consistency of the injection from the indexing schemes with the model
-is checked.
-
-It the multiplicity of the access pattern and the model is matched and are
-$\mathit{once}$, then the problem is reduced to checking consistency in
-models in which duplicate access patterns are allowed.
-
-Next, a consistency predicate considers whether the stencil is an approximation.
-Semantically, a lower bound is achieved if every indexing scheme in the
-model generalises some relativised access pattern that means all cells
-described in the specification are part of the access pattern. Conversely,
-an upper bound means all observed access patterns can be generalised with
-the indexing schemes of the model. Hence, the actual cells are a subset of the
-specified ones. In combination an exact specification, \ie{} one without any
-approximation modifiers, is one that is both a lower and an upper bound.
+\noindent
+Thus our model validates the specification synthesis process which concludes the
+inference procedure. For our example, we infer and then synthesise the
+5-point specification:
+%%
+\begin{minted}[linenos=false]{fortran}
+!= stencil centered(depth=1, dim=1)*reflexive(dim=2) + centered(depth=1, dim=2)*reflexive(dim=1) :: a
+\end{minted}
+%%
+
+\subsection{Specification checking\label{subsec:checking}}
+
+In checking mode, the access pattern of a stencil computation in the
+source language is verified against a specification. The model of
+Section~\ref{sec:semantics} again aids this process.  Checking
+proceeds by generating a model $M$ from a specification and generating a set $N$ of indexing
+schemes from the source code. The source code is then valid with
+respect to the specification if $N$ is \emph{consistent} with $M$. The
+notion of consistency is built out of index scheme 
+\emph{generalisation} (Sect.~\ref{subsec:schemes-and-gen}). 
+Consistency is defined in Figure~\ref{fig:consistency},
+but we first set up some intermediate definitions before unpacking its definition.
+
+\begin{definition}[Checking]
+Let \texttt{a} be an array variable and let
+$S \in \textit{spec}$ be a specification with
+the \textit{specDec} \texttt{!= stencil
+  $S$ :: a} in our syntax.
+For a array subscript
+assignment statement $\textit{assg} = v(\overline{e_1}) = e_2$ in our source language
+(with all the appropriate data flow information in place)
+where $e_2$ has at most $n$-dimensional array access, 
+then the assignment conforms to the specification $S$ if 
+the predicate $\mathit{check}_n(S, \textit{assg}, \texttt{a})$ holds, 
+defined:
+\begin{align*}
+& \mathit{check_n}(S, \textit{assg}, x) =
+  \mathit{cons_n}(\interp{S}_n, \, \textit{modelise}(\textit{assg},
+  x)) \\
+& \mathit{modelise} : \textit{statement} \times \textit{variable}
+  \rightarrow \textsf{Mult}((\domainVal)^n) \\ 
+& \mathit{modelise} (\textit{assg}, x) =
+  \begin{cases}
+    \textsf{once}(\mathit{analyse}(\textit{assg})(x)) & \textit{unique}(\textit{assg},x) \\
+    \textsf{mult}(\mathit{analyse}(\textit{assg})(x)) & otherwise
+  \end{cases}
+\end{align*}
+Thus, checking proceeds by reusing the \textit{analyse} function of
+inference (Section~\ref{subsec:inference}) to extract indexing
+schemes for stencil computations flowing to an assignment. 
+This is then augmented with multiplicity information
+by $\mathit{modelise}$ where $\mathit{unique}(\textit{assg},x)$ is a
+ predicate indicating whether indices to array $x$ in an
+assignment $\textit{assg}$ are unique or not. 
+\end{definition}
+
+Subsequently the binary predicate for consistency, at the heart of checking
+is defined over the sets: 
+%
+\begin{equation*}
+\textit{cons}_n \subseteq \textsf{Mult}(\textsf{Approx}(\domainVal{}^n))
+\times \textsf{Mult}(\domainVal{}^N)
+\end{equation*}
+%
+where first component is a model of a specification,
+and the second component is a set of indexing schemes generated from
+the array subscripts in a program, with added multiplicity
+information. Giving \textsf{Approx} information to the
+index schemes inferred from the program is not necessary, nor
+desirable-- we have an exact set of schemes. We refer to the first
+component of the predicate as \emph{spec model} and the second as \emph{code model}.
+
+\begin{definition}
+Consistency is defined in Figure~\ref{fig:consistency}. We let
+$M_s, M_c$ range over elements of $\domainVal{}^n$ where $M_s$ comes
+from the spec model and $M_c$ comes from \textit{modelise} on the
+source code. Let $N_c$ range over elements of
+$\textsf{Mult}(\domainVal{}^n)$ coming from the code model, and $A_s$
+over elements of $\textsf{Mult}(\textsf{Approx}(\domainVal{}^n))$ from the spec model.
+
+The first equation states that, if the multiplicity of the spec and code
+model are both $\mathsf{once}$, then the problem is reduced to checking consistency of
+models in which non-unique indexing is allowed, labelled with
+$\mathsf{mult}$.
+
+The second equation considers the case where the specification is
+an \term{atMost} upper-bound approximation (labelled with \textsf{up}
+as shorthand for \textsf{upper} here). 
+An upper bound specification means that for every scheme in the code
+model, modelling a source-level index, $\forall u \in M_c$ there exists
+a generalising scheme in the spec model $\exists v \in M$, \ie{}, $u
+\preceq_n v$. That is, every source-level index is generalised by the model.
+
+The third equation is the converse of an \term{atLeast},
+lower approximation (labelled \textsf{low} as shorthand for
+\textsf{lower}). Consistency is dual to the above: 
+for every scheme in the spec model $\forall v \in M_s$ there
+exists a scheme in the code model $\exists u \in M_c$ such that $u
+\preceq_n v$. That is, every model scheme is generalised by source indexing.
+
+These two are combined for consistency on an \textsf{exact}
+specification (the fourth equation) such that the spec model
+provides both a consistent lower and consistent upper bound on
+the source code. Thus, for every scheme in the code model there
+is a scheme in the spec model that generalises it, and for every
+scheme in the spec model there is a scheme in the source model that
+specialises it.
+
+Lastly, if the spec model has both an upper and lower bound
+consistency is defined inductively similar to the \textsf{exact} case.
+\end{definition}
 
@@
 \input{consistency.tex}
-\caption{Consistency function of a model against stencil computation}
+\caption{Consistency between a model and indexing schemes from a stencil computation}
 \label{fig:consistency}
+\vspace{-0.9em}
 \end{figure}
 
-We will consider an extended example of a stencil computation that is
-consistent with its annotation to show how checking works. The following is an
-example with a looser but still consistent version of the \textit{fivepoint}
-stencil specification inferred previously in the
-Subsection~\ref{subsec:inference}:
+\begin{example}
+We give an extended example of verifying a stencil computation against
+its specification. We consider a more general, but still consistent, version of the \textit{five-point}
+stencil specification inferred previously in Section~\ref{subsec:inference}:
 %
@@
 != stencil readOnce, fivepoint :: b
-a(i,j) = b(i, j-1) + b(i,j) + b(i,j+1) + b(i-1,j) + b(i+1,j)
+a(i,j) = b(i,j-1) + b(i,j) + b(i,j+1) + b(i-1,j) + b(i+1,j)
 \end{minted}
 %
+We refer to the assignment on line three as $l3$ for brevity.
+
+Verification of the code against the spec is then given by 
+$check(({\textcolor{darkgreen}{\texttt{readOnce, fivepoint}}}),l3,\texttt{b})$. 
+We explain the steps.
 \begin{enumerate}  
-\item From now on assignment on line three is referred as $L3$ for brevity.
-  Verification is done by computing the following:
-  $$check(\texttt{stencil readOnce, fivepoint},L3,b)$$
-%
-\item Generate model from the specification as described in
-  Section~\ref{sec:semantics}. We refer to the indexing
-  schemes within the model as set $m$.
-%
-\[ m= \{  \vtwo{-1}{\infty},
+%
+\item Generate the model from the specification as in
+  Section~\ref{sec:semantics}. 
+%
+\begin{align*}
+ M_s= \{  \vtwo{-1}{\infty},
           \vtwo{0}{\infty},
@@
           \vtwo{\infty}{0},
-        \vtwo{\infty}{1} \} \]
-\begin{dmath*}
-\interp{\texttt{stencil readOnce, fivepoint}}_2 =
-  \{\textsf{once}(\textsf{exact}(m)) \}
-\end{dmath*}
-%
-\item To evaluate the two predicates above, we need indexing schemes for the
-  array in question. This is obtained by $modelise$ function. For brevity the
-  set of indexing schemes is referred as $m'$.
-
-  For the example above the observed model is as follows:
-%
-\[
-m' = \{
+        \vtwo{\infty}{1} \}  \\[0.4em]
+\interp{{\textcolor{darkgreen}{\texttt{readOnce, fivepoint}}}}_2 =
+  \textsf{once}(\textsf{exact} (M_s))
+\end{align*}
+%
+\item Obtained the code model via $\textit{modelise}$, giving:
+%
+\begin{align*}
+M_c = \{
     \vtwo{0}{-1},
@@
   \}
-\]
-\[modelise(L3,b) = \textsf{once}(m')\]
-%
-\item Now $cons$ predicate is applied with this model along side with the
-  assignment on line three and array $b$.
-
-  The predicate is reduced to
-  $cons(\textsf{once}(\textsf{exact}(m)),\textsf{mult}(m'))$.
-%
-\item $cons$ predicate from the previous step evaluates to
-  $$cons(\textsf{mult}(\textsf{exact}(m)),\textsf{once}(m')).$$
-%
-\item We further evaluate the predicate to
-  \begin{dmath*}
-    cons(\textsf{mult}(\textsf{up}(m)),\textsf{mult}(m')) \wedge
-    cons(\textsf{mult}(\textsf{low}(m),\textsf{mult}(m')).
-  \end{dmath*}
-%
-\item Below we present two sets of generalisations to show both the lower and
-  the upper bound holds.
-
-  The generalisations for the upper bound:
-  \begin{align*}
-     \vtwoh{0}{-1} & \preceq_2 \vtwoh{0}{\infty} \\
-     \vtwoh{0}{0} & \preceq_2 \vtwoh{0}{\infty} \\
-     \vtwoh{0}{1} & \preceq_2 \vtwoh{0}{\infty} \\
-     \vtwoh{-1}{0} & \preceq_2 \vtwoh{\infty}{0} \\
-     \vtwoh{1}{0} & \preceq_2 \vtwoh{\infty}{0}
-  \end{align*}
-
-  The generalisations for the lower bound:
-  \begin{align*}
-     \vtwoh{-1}{0} & \preceq_2 \vtwoh{-1}{\infty} \\
-     \vtwoh{0}{0} & \preceq_2 \vtwoh{0}{\infty} \\
-     \vtwoh{1}{0} & \preceq_2 \vtwoh{1}{\infty}  \\
-     \vtwoh{0}{-1} & \preceq_2 \vtwoh{\infty}{-1} \\
-     \vtwoh{0}{0} & \preceq_2 \vtwoh{\infty}{0} \\
-     \vtwoh{0}{1} & \preceq_2 \vtwoh{\infty}{1}
-  \end{align*}
-
-  Since both predicates are satisfied the array accesses specified on line
-  three, the stencil computation is consistent with the \texttt{fivepoint}
-  specification.
+\\[0.4em]
+\textit{modelise}(l3, \texttt{b}) = \textsf{once} (M_c)
+\end{align*}
+%
+\item Consistency $cons_2(\textsf{once}(\textsf{exact}(M_s)),
+  \textsf{once}(M_c))$ holds when 
+  $\textit{cons}_2(\textsf{mult}(\textsf{exact}(M_s)),\textsf{mult}(M_c))$ (Fig.~\ref{fig:consistency}) 
+  which holds when: 
+  \begin{align*}
+     & \textit{cons}_2(\textsf{mult}(\textsf{up}(M_s)),\textsf{mult}(M_c)) \\
+   \wedge \; & \textit{cons}_2(\textsf{mult}(\textsf{low}(M_s),\textsf{mult}(M_c))
+  \end{align*}
+%
+\item This condition holds by the following sets of scheme
+  generalisations: for the upper bound, $\forall u \in M_c,
+  \exists v \in M_s . u \preceq_2 v$: 
+  \begin{align*}
+    \begin{array}{l}
+     \vtwoh{0}{-1} \preceq_2 \vtwoh{0}{\infty} \quad
+     \vtwoh{0}{0} \preceq_2 \vtwoh{0}{\infty} \quad
+     \vtwoh{0}{1} \preceq_2 \vtwoh{0}{\infty} \\[0.5em]
+     \vtwoh{-1}{0} \preceq_2 \vtwoh{\infty}{0} \quad
+     \vtwoh{1}{0} \preceq_2 \vtwoh{\infty}{0}
+     \end{array}
+  \end{align*}
+  % 
+  and for the lower bound: $\forall v \in M_s, \exists u \in M_c . u
+  \preceq_n v$: 
+  \begin{align*}
+    \begin{array}{l}
+     \vtwoh{-1}{0} \preceq_2 \vtwoh{-1}{\infty} \quad
+     \vtwoh{0}{0} \preceq_2 \vtwoh{0}{\infty} \quad
+     \vtwoh{1}{0} \preceq_2 \vtwoh{1}{\infty}  \\
+     \vtwoh{0}{-1}  \preceq_2 \vtwoh{\infty}{-1} \quad
+     \vtwoh{0}{0}  \preceq_2 \vtwoh{\infty}{0} \quad
+     \vtwoh{0}{1}  \preceq_2 \vtwoh{\infty}{1}
+     \end{array}
+  \end{align*}
+
+  Since both the consistency predicate is satisfied, the array
+  accesses flowing to line three in the stencil computation are valid
+  with respect to the {\textcolor{darkgreen}{\texttt{fivepoint}}} specification.
 \end{enumerate}
-
+\noindent
 Now when $L3$ is modified with an additional term \texttt{b(i+2,j+2)},
@@
 is indeed contained in the regions touched by $b$.
-
-\subsubsection{A note on stencil approximations}
+\end{example}
 

### 2016-07-07 paper.tex
section 4 edits

@@
 \begin{minted}[linenos=false]{fortran}
-!= stencil centered(depth=1, dim=1)*reflexive(dim=2) + centered(depth=1, dim=2)*reflexive(dim=1) :: a
+!= stencil centered(depth=1, dim=1)*reflexive(dim=2) + centered(depth=1, dim=2)*reflexive(dim=1):: a
 \end{minted}
@@
 its specification. We consider a more general, but still consistent, version of the \textit{five-point}
-stencil specification inferred previously in Section~\ref{subsec:inference}:
+stencil specification inferred previously in
+Section~\ref{subsec:inference} (see top-left):
 %
@@
   \exists v \in M_s . u \preceq_2 v$: 
-  \begin{align*}
+  \begin{align}
     \begin{array}{l}
@@
      \end{array}
-  \end{align*}
+  \label{eq:check-example-upper}
+  \end{align}
   % 
@@
   \preceq_n v$: 
-  \begin{align*}
+  \begin{align}
+    \hspace{-1em}
     \begin{array}{l}
@@
      \end{array}
-  \end{align*}
-
+    \label{eq:check-example-lower}
+  \end{align}
   Since both the consistency predicate is satisfied, the array
@@
 \end{enumerate}
-\noindent
-Now when $L3$ is modified with an additional term \texttt{b(i+2,j+2)},
-the new assignment $L3'$ becomes:
+\vspace{-0.7em}
+\end{example}
+\begin{example} 
+We consider now an erroneous user modification to $l3$ with an
+additional term \texttt{b(i+2,j+2)} such that $l3'$ is:
 %
 \begin{minted}[firstnumber=3]{fortran}
-a(i,j) = b(i, j-1) + b(i,j) + b(i,j+1) + b(i-1,j) + b(i+1,j) + b(i+2,j+2)
+a(i,j) = b(i,j-1) + b(i,j) + b(i,j+1) + b(i-1,j) + b(i+1,j) + b(i+2,j+2)
 \end{minted}
 %
-In the first six steps, evaluation proceeds the same way except for step three.
-In the third step, result of $\mathit{modelise}(L3',b)$ will have the extra term
-$\vtwoh{2}{2}$. In the seventh step, since the model has not changed and the
-indexing schemes are only augmented with a new element, the generalisations for
-lower bound still hold. However, there are no elements in the model, $m$, that
-can generalise $\vtwoh{2}{2}$, hence the model fails to hold as an upper bound.
+The checking procedure proceeds in the same way as the first three
+steps of the previous example, except that the result of
+$\mathit{modelise}(l3',\texttt{b})$ generates an additional scheme
+$\vtwoh{2}{2}$. Since the spec model has not changed and the code
+model is has only the additional scheme $\vtwoh{2}{2}$, the generalisations for
+the lower bound \eqref{eq:check-example-lower} 
+of step 4 above still holds since its consistency condition
+universally quanitifes over the \emph{spec model}. However, for
+the upper bound, the consistency condition universally quantifies
+over scheme $u$ of the \emph{code model} for which there exists
+a scheme $v \in M_s$ such that $u \preceq_2 v$. Thus, we require a $v
+\in M_s$ such that $\vtwoh{2}{2} \preceq_2 v$ but no such generalising
+scheme existing in the spec model. Hence the model fails to hold as an upper bound.
 
@@
 %
-The consistency predicate would have held because for the upper bound
-$\mathit{cons}(\textsf{mult}(\textsf{up}(m)),\textsf{mult}(m'))$ would not have
-been needed. This semantically makes sense since the region $\texttt{fivepoint}$
-is indeed contained in the regions touched by $b$.
+then consistency holds as it no longer requires the failing upper-bound case 
+$\mathit{cons}(\textsf{mult}(\textsf{up}(M_s)),\textsf{mult}(M_c))$.
+ This makes sense semantically since the region ${\textcolor{darkgreen}{\texttt{fivepoint}}}$
+is indeed contained in the regions touched by $\texttt{b}$ in $l3'$. 
 \end{example}
 
-As noted above, the \textsf{fivepoint} region defined above is more accepting
-than that used in Section~\ref{sec:lang}. This introduces a new notion of
-approximation beyond that is achieved in Section~\ref{sec:semantics}
-using upper and lower bounds because the in both cases the specifications are
-used to generate \emph{exact} models. Formally, we say a specification $s'$
-approximates a more specific specification $s$ \emph{iff} $\forall m \in
-\textsf{Mult}(\domainVal) . \mathit{cons}(s,m) \implies \mathit{cons}(s',m)$.
-
-Equations provided on regions in Section~\ref{sec:lang} when all modifiers
-equal provide such approximations, but it goes further than that. The
-regions describing the fivepoint stencil for example cannot be shown equal
-using these equations, but satisfies the approximation predicate as defined
-above.
-
-The checking algorithm though happily accepts an approximation in place of the
-most specific specification. Although this decreases the likelihood of catching
+As noted, the {\textcolor{darkgreen}{\term{fivepoint}}} region
+ above is more accepting than that used in
+Section~\ref{sec:lang}. This introduces a new notion of approximation
+beyond what is given by the syntactic theory of sub-specifications
+$<:$ in Section~\ref{sec:eqs}. Consistency leads to a notion of
+approximation that is a \textit{semantic sub-specification} relation:
+%using upper and lower bounds because the in both cases the specifications are
+%used to generate \emph{exact} models.
+
+\begin{definition}[Approximation]
+A specification $S'$
+\emph{approximates} a more specific specification $S$, written 
+$S \sqsubseteq S'$, \emph{iff}:
+$$
+\forall M_c \in \textsf{Mult}(\domainVal{}^n) . \mathit{cons}_n(\interp{S}_n,M_c) \implies
+\mathit{cons}(\interp{S'}_n,M_c)
+$$
+\end{definition}
+%
+For the syntactic sub-specification equations on regions $<:^r$, it follows that
+these imply semantic approximation:
+%
+\begin{lemma}[Sub-specification $<:^r$ coincides with approx 
+$\sqsubseteq$]
+\begin{equation*}
+\forall S, S' . \; S <:^r S' \Rightarrow S \sqsubseteq S'
+\end{equation*}
+\end{lemma}
+%
+However, the converse direction does not hold. 
+The regions describing the more acceptiong five-point stencil
+above do not follow from the definition of $<:$ on the previous
+more specific specification.%, that is:
+%${\textcolor{darkgreen}{\term{centered(depth=1, dim=1)*reflexive(dim=2)
+%      + centered(depth=1, dim=2)*reflexive(dim=1)}}}
+%\not{<:} {\textcolor{darkgreen}{\term{centered(depth=1,dim=1) +
+%      centered(depth=1,dim=2)}}}$.
+ However, consitency (and thus $\sqsubseteq$) shows
+that we can check the five-point stencil code against the more general
+specifcation of {\textcolor{darkgreen}{\term{fivepoint}}}. Thus, we
+can always use an approximation in place of the
+more specific specification. Although this decreases the likelihood of catching
 stencil errors (as more access patterns conform), it reduces the space and

### 2016-07-07 paper.tex
intro edits

@@
 
-\keywords{program verification, specifications, static analysis,
-  stencil computation}
+\keywords{program verification, specification, stencils}
 
@@
 %
-This explains that \texttt{a} is read from with a
-symmetrical access pattern (``centered'') to a depth of one in each
+This describes that \texttt{a} is accessed with a
+symmetrical pattern (``centered'') to a depth of one in each
 direction in its first (and only) dimension.
@@
 %\texttt{a} and \texttt{b}.
-In the case of the Navier-Stokes example of
+In the Navier-Stokes example of
 Figure~\ref{ref:navier-stokes-fragment}, its inferred specification is shown
-in Figure~\ref{ref:navier-stokes-fragment}(b). The
+in Fig.~\ref{ref:navier-stokes-fragment}(b). The
 specification explains that, over the whole fragment, \texttt{u} is
 read from with a centered pattern to depth of 1 in both dimensions
-(this is known as the \emph{five point stencil}) and \texttt{v}
+(this is known as the \emph{five-point stencil}) and \texttt{v}
 is read from in a neighbourhood bounded by forward to depth of $1$ in
 the first dimension and backward to a depth of $1$ in the second
-dimension. Figure~\ref{ref:navier-stokes-fragment}(c) illustrates
-the specifications pictorially.
+dimension. Fig.~\ref{ref:navier-stokes-fragment}(c) illustrates
+the specification pictorially.
 
@@
 %
-\begin{itemize}
-\item we introduce our specification language for
-stencil computations that captures many common forms
-of data access pattern (Section~\ref{sec:lang})
-along with a sound denotational model (Section~\ref{sec:semantics});
-
-\item we detail inference, checking, specification synthesis,
-and program synthesis algorithms for our language
-(Section~\ref{sec:analysis}), which relies on the semantic model;
-
-\item we evaluate our approach, providing an
+\begin{itemize}[leftmargin=0.8em]
+\item we introduce a specification language for
+stencil computation that captures many common forms
+of data access pattern (Section~\ref{sec:lang}), providing
+ an equational theory and denotational model (Sect.~\ref{sec:semantics});
+
+\item we detail inference, checking, and specification synthesis 
+algorithms for our language (Sect.~\ref{sec:analysis}), leveraging the semantic model;
+
+\item we evaluate our approach with an
 implementation as an extension of CamFort,
-an open-source program analysis and refactoring tool for Fortran
-corpus. We use this implementation to evaluate our approach against
+an open-source program analysis tool for Fortran
+programs. We use our implementation to evaluate our approach against
 a corpus of scientific applications, totalling roughly one million
-lines of code, and we study some verification case studies;
+lines of code;
 
@@
 
-\item we consider two particular stencils in detail and show that we
+\item we provide a more detailed verification case study of two
+  particular stencil computation and show that we
   are able to detect the vast majority  of (synthesised) possible
@@
   (v(i,j-1)+v(i+1,j-1))*(u(i,j-1)+u(i,j))-      &
-  gamma*abs(v(i,j-1)+v(i+1,j-1))*(u(i,j-1)-u(i,j))) & /(4.0*dely)
+  gamma*abs(v(i,j-1)+v(i+1,j-1))*(u(i,j-1)-     & 
+  u(i,j))) / (4.0*dely)
 
@@
 \end{minted}
-(a) Excerpt from Fortran code for a Navier-Stoke fluid simulator,
-showing highly-detailed stencil computations.
+(a) Excerpt of a Navier-Stokes fluid simulator
+(based on~\citet{griebel1997numerical}) showing highly-detailed stencil computation.
 \vspace{0.5em}
@@
 \end{minted}
-(b) Inferred stencil specification from CamFort.
+(b) Inferred and synthesised specification for (a) by CamFort.
 \vspace{0.5em}
@@
 science~\cite{post2005computational,oberkampf2010verification,orchard2014computational}.
-Our specification language is an approach in this direction, proving
+Our specification language is an approach in this direction, providing
 a system of lighweight specification for numerical code.
@@
 
-Much of this paper concerns the technical aspects of the language,
-its semantics, its inference, checking, and synthesis procedures.
-The evaluation focuses on its application to real scientific code base
-used actively in research.
+This paper considers the technical aspects of the language,
+its semantics, inference, checking, and synthesis procedures.
+We also evaluate our implementation on real scientific code 
+used actively in research. Our case studies include the Unified Weather 
+model used by the UK Met Office~\cite{um-website} ($\approx{}600\textit{kloc}$) and other
+research-based numerical models, ranging from small to large.
 
@@
 out-of-bounds access at the edges). We revisit this hypothesis
-in Section~\ref{sec:evaluation} where the inference of
-such regular stencil patterns on a corpus of numerical programs (both
-small and large).
-
-Section~\ref{sec:syntax} outlines the syntax of specifications.
+in Section~\ref{sec:evaluation} with the inference of
+such regular stencil patterns on our corpus of numerical programs.
+
+Section~\ref{sec:syntax} outlines the syntax.
  Section~\ref{sec:eqs} provides an equational
-theory for specifications via relation ($\equiv$) and a notion of
-sub-specification via relation ($<:$).  Section~\ref{sec:semantics}
+theory for specifications and a notion of
+sub-specification.  Section~\ref{sec:semantics}
 gives a semantic model via a set-based interpretation, for which the
@@
 \paragraph{Target language syntax} Throughout, for the source language
-syntax, $v$ ranges over program variables, $s$ over statements, and
-$e$ over expressions (which may be impure).  By ``variable'', we mean
+syntax, $e$ ranges over its expressions (which may be impure).  By ``variable'', we mean
 in the imperative sense (\ie{}, named binders to memory cells) instead
@@
 \end{align*}
-Thus, checking proceeds by reusing the \textit{analyse} function of
-inference (Section~\ref{subsec:inference}) to extract indexing
-schemes for stencil computations flowing to an assignment. 
+Thus, checking proceeds by reusing the \textit{analyse} function from 
+inference (Sec.~\ref{subsec:inference}) to extract index 
+schemes from source code. 
 This is then augmented with multiplicity information
@@
 \textbf{The Unified
-  Model}\footnote{\url{http://www.metoffice.gov.uk/research/modelling-systems/unified-model}}
+  Model}~\cite{um-website}
 is a weather forecasting and climate modelling tool developed by the

### 2016-07-07 paper.tex
Revise corpus references

@@
 \textbf{The Unified
-  Model}~\cite{um-website}
+  Model}~\cite{um}
 is a weather forecasting and climate modelling tool developed by the
@@
 
-\textbf{BLAS} (Basic Linear Algebra Subprograms)\footnote{\url{http://www.netlib.org/blas/}} provides efficient and portable routines for vector and matrix operations.
+\textbf{BLAS}~\cite{blas} (Basic Linear Algebra Subprograms) provides efficient and portable routines for vector and matrix operations.
 These routines feature in many other libraries (including LAPACK).
@@
 
-\textbf{GEOS-Chem}\footnote{\url{http://acmg.seas.harvard.edu/geos/}} is a three-dimensional model of tropospheric chemistry developed at Harvard and used by approximately 70 universities and research institutions world-wide.
+\textbf{GEOS-Chem}~\cite{geos-chem} is a three-dimensional model of tropospheric chemistry developed at Harvard and used by approximately 70 universities and research institutions world-wide.
 We examined version 10-01.

### 2016-07-07 paper.tex
edits to eval and intro

@@
   annotations in source code, along with inference, checking, and
-  specification and program synthesis procedures. We evaluate our
+  specification synthesis procedures. We evaluate our
   language against a corpus of numerical Fortran code, and show that
@@
   stencil specification language, its static semantics, inference,
-  checking, and program synthesis procedures, our implementation, and
+  checking, and synthesis procedures, our implementation, and
   our evaluation study.
@@
 \caption{Fragment of a Navier-Stokes fluid simulator and its
-  specification in our language}
+  specification in our language.}
 \label{ref:navier-stokes-fragment}
@@
 \renewcommand*{\arraystretch}{0.8}
-\paragraph{Target language syntax} Throughout, for the source language
-syntax, $e$ ranges over its expressions (which may be impure).  By ``variable'', we mean
-in the imperative sense (\ie{}, named binders to memory cells) instead
-of mathematical variables.
+For the target language
+syntax, $e$ ranges over its expressions (may be impure) and $v$
+overs its (imperative) variables. %By
+%``variable'', we mean in the imperative sense (named binders to memory cells).
 
@@
   loop,\footnote{Or equivalent, \eg{}, \texttt{do} in Fortran, for our implementation.} incremented by $1$ per
-iteration. The variable is marked as a base induction variable
-only within the scope of the loop. Induction variables are ranged
-over by $i$, $j$, $k$ throughout. Since we do not use non-base
-(derived) induction variables we often simply say \emph{induction variable}.
+iteration. The variable is marked as an induction variable
+only within the body of the loop. Since we ignore non-base
+(derived) induction variables, we simply say \emph{induction
+  variable}. Induction variables are ranged over by $i$, $j$, $k$ throughout. 
 \end{definition}
@@
 \begin{figure}[t]
+\vspace{-0.9em}
 \input{region-model.tex}
@@
 \end{definition}
+\vspace{-0.5em}
 
 \begin{figure}
+\vspace{-0.5em}
 \input{spatial-model.tex}
@@
 \subsection{Soundness}
-
+%
 Given the above model, we can
@@
 %
+\noindent
 However, the converse direction does not hold. 
@@
 
-\subsection{Synthesis\label{subsec:synthesis}}
-
 \section{Evaluation}
 \label{sec:evaluation}
-We evaluated the stencil language on a corpus of scientific code. We first give an overview of the corpus itself before presenting our results.
+We evaluated our specificataion language on a corpus of scientific
+code.  We overview the corpus itself before presenting our results.
 
@@
 \end{tabular}
-\caption{Summary of software packages used for evaluation\label{tab:corpus}}
+\caption{Summary of software packages used for
+  evaluation\label{tab:corpus}}
+\vspace{-1em}
 \end{table}
@@
 how many files in each corpus that we were able to analyse with
-Camfort. The most common reasons for Camfort rejecting a file were
+CamFort. The most common reasons for CamFort rejecting a file were
 either use of a C-preprocessor, or illegal use of language features
-from a more modern Fortran variant.
+from a modern Fortran variant.
 
@@
 and meteorological services around the world. We use the development
-branch (trunk) of the model. The code based in closed source but
+branch (trunk) of the model. The code base in closed source but
 institutional licenses are available for research purposes.
@@
 \textbf{E3MG} (An Energy-Environment-Economy (E3) Model at the Global Level) is a macroeconomic model used for assessment of environmental policy~\cite{RePEc:aen:journl:2006se-a12}.
-This model was developed by Cambridge Econometrics, an independent consultancy company.
-
-\textbf{BLAS}~\cite{blas} (Basic Linear Algebra Subprograms) provides efficient and portable routines for vector and matrix operations.
-These routines feature in many other libraries (including LAPACK).
-We used version 3.6.0.
+This was developed by Cambridge Econometrics, an independent consultancy company.
+
+\textbf{BLAS}~\cite{blas} (Basic Linear Algebra Subprograms) is a popular
+library providing efficient and portable routines for vector and
+matrix operations. These routines feature in many other libraries
+(including LAPACK). We used version 3.6.0. We chose to include this
+package for breadth, as it provides general numerical functions rather
+than a specialised scientific model.
 
@@
 
-\textbf{GEOS-Chem}~\cite{geos-chem} is a three-dimensional model of tropospheric chemistry developed at Harvard and used by approximately 70 universities and research institutions world-wide.
+\textbf{GEOS-Chem}~\cite{geos-chem} is a three-dimensional model of tropospheric chemistry developed at Harvard and used by $\sim$70 universities and research institutions world-wide.
 We examined version 10-01.
 
-\textbf{Navier} is a small project, giving a discrete
+\textbf{Navier} is a small numerical simulation, giving a discrete
 approximation to the two-dimensional Navier-Stokes fluid equations,
@@
 We first examined how frequently stencil computations occur in our
-corpus. We define a \emph{potential} statement as one whose left-hand
+corpus. This was carried out by running our implementation in 
+whole-code-base inference mode. We define a \emph{potential} statement as one whose left-hand
 side is an array subscript on neighbourhood indices. These statements
-are the inputs to Step 2 of the Inference process described
+are the inputs to Step 2 of the Inference process described in 
 Section~\ref{subsec:inference}. Each potential statement for which we
-eventually generate some stencil specifications is also counted as
+eventually generate some specification is also counted as
 an \emph{actual} statement.
@@
 \caption{Stencil assignment rates\label{tab:tickAssign}}
+\vspace{-0.8em}
 \end{table}
 
-The first two columns of Table~\ref{tab:tickAssign} shows the number
+The first two columns of Table~\ref{tab:tickAssign} show the number
 of potential and actual stencils and their rate as a percentage of the
-total number of successfully parsed lines of code. There are a variety
-of reasons why we might not infer an actual stencil from a potential
-one: 1) \textbf{Non-subset induction variables} occurs when the
-induction variables on the RHS are not a subset of the LHS. These
+total number of successfully parsed lines of code. The ratio of
+actual to potential stencils overall is thus $32\%$, \ie{}, we give
+a specification to roughly a third of the potential stencil specifications.
+
+There are a variety of reasons why we might not infer a actual specification from a potential
+stencil computation: \\
+\noindent 
+1) \textbf{Non-subset induction variables} occur when the
+induction variables on the RHS are not a subset of those in the LHS. These
 cases are not stencils by our definition. The degenereate case of this
 is to have only constant indices on the LHS. We see lots of examples
-of this in loops as accumulators e.g. computing the sum over some
-other array; 2) \textbf{Derived induction variables} where the
+of this in loops as accumulators \eg{} computing the sum over an array; \\
+\noindent
+ 2) \textbf{Derived induction variables} where the
 indexing variable (\mintinline{fortran}{x}) is derived from an
-induction variable (\mintinline{fortran}{i}) as in \mintinline{fortran}{x
-  = len - i}; 3) \textbf{Inconsistent induction dimensions} occur when
+induction variable (\mintinline{fortran}{i}) as in
+\mintinline{fortran}{x = len - i};  \\
+\noindent
+3) \textbf{Inconsistent induction dimensions} occur when
 an induction variable is used to specify more than one array dimension
-on the RHS. These are common in matrix operations such as
-LU-decomposition which contains assignments such as
+on the RHS or multiple induction variables are used for the same
+dimension on the RHS. These are common in matrix operations such as
+LU-decomposition with assignments such as
 \mintinline{fortran}{a(l) = a(l) - a(m) * b(l, m)}.
@@
 
-If the statement meets the stencil critera but the inference process cannot determine an exact stencil for a statement it will instead produce an upper (\term{atMost}) and lower (\term{atLeast}) bound.
-The number of exact and inexact stencils are shown in columns three and four of Figure \ref{tab:tickAssign}.
-We can see that our stencil language is capable of precisely specifying the overwhelming majority of stencils.
+If the statement meets the stencil critera but the inference process
+cannot determine an exact stencil for a statement it will instead
+produce an upper (\term{atMost}) and lower (\term{atLeast}) bound. The
+number of exact and inexact stencils are shown in columns three and
+four of Table~\ref{tab:tickAssign}. Note that the total for ``Exact``
+is higher than ``Actual`` because for some lines we produce multiple
+specifications (perhaps for different array variables).
+We see that our specification language is capable of precisely specifying the overwhelming majority of stencils.
 Most of the inexact stencils apply to some form of data copying.
@@
 
-We grouped stencils together into four classes based on similar levels of complexity. These four classes account for 99.4\% of all the stencils we found in the corpus.
+We grouped our inferred specifications together into four classes based on similar levels of complexity. These four classes account for 99.4\% of all the stencils we found in the corpus.
 
@@
 \end{minted}
-and Camfort is able to infer a precise stencil specification for this
+and CamFort is able to infer a precise stencil specification for this
 line,
@@
 \end{minted}
-which when inserted back into the program can be checked by Camfort as
-well. We then tested the error-detecting ability of Camfort by
+which when inserted back into the program can be checked by CamFort as
+well. We then tested the error-detecting ability of CamFort by
 generating a family of 6,561 Jacobi iteration programs by perturbing
@@
 number but at least got the ordering of the indices correct, although
-Camfort would also detect a mixup of variables in this case as well.
-
-Camfort checked all 6,561 programs and found that only 24 of them were
+CamFort would also detect a mixup of variables in this case as well.
+
+CamFort checked all 6,561 programs and found that only 24 of them were
 valid implementations of the stencil specification. This was the
@@
 byte-by-byte comparison showed results that completely concurred with
-the results of the Camfort test.
+the results of the CamFort test.
 
@@
 \label{sec:discussion}
+
+\paragraph{Program synthesis}
 
@@
 is generated from a relatively high-level specification. This approach
-differs from Camfort because we are interested in integrating directly
+differs from CamFort because we are interested in integrating directly
 into existing, legacy codebases and long-established languages such as
@@
 procedure. We evaluated our approach by implementing the stencil inference and
-checking procedure in Camfort, an open-source analysis tool\footnote{\url{https://github.com/camfort/camfort}} for
+checking procedure in CamFort, an open-source analysis tool\footnote{\url{https://github.com/camfort/camfort}} for
 Fortran programs and running the inference process over a corpus of
@@
 %%  LocalWords:  equational disjunction denotational dimensionality
-%%  LocalWords:  interprocedural Fortran Camfort preprocessor
+%%  LocalWords:  interprocedural Fortran CamFort preprocessor
 %%  LocalWords:  committers

### 2016-07-07 paper.tex
eval section edits

@@
 institutional licenses are available for research purposes.
-The Met Office runs a comprehensive code quality system incorporating dedicated committers (we counted 11) for particular parts of the model.
+%The Met Office runs a comprehensive code quality system incorporating dedicated committers (we counted 11) for particular parts of the model.
 We counted 120 additional contributors whose submissions are reviewed and tested before being accepted into the code base.
@@
 
-\textbf{GEOS-Chem}~\cite{geos-chem} is a three-dimensional model of tropospheric chemistry developed at Harvard and used by $\sim$70 universities and research institutions world-wide.
-We examined version 10-01.
+\textbf{GEOS-Chem}~\cite{geos-chem} is a three-dimensional model of
+tropospheric chemistry developed at Harvard and used by $\sim$70
+universities and research institutions world-wide. We use v.10-01.
 
@@
 Most of the inexact stencils apply to some form of data copying.
-For example (taken from E3MG):
+For example (taken from \textbf{E3MG}):
 \begin{minted}{fortran}
@@
 
-\textbf{All-reflexive} The vast majority (66\%) of stencils were
-reflexive in every dimension of the array. These most often correspond
-to applying point-wise transformations to data such as scaling
-e.g. \begin{minted}{fortran}
+\textbf{All-reflexive} The vast majority (66\%) of specification are
+reflexive in every dimension. These most often correspond
+to point-wise transformations on data such as scaling, \eg{} from \textbf{E3MG}: \begin{minted}{fortran}
 ZZ1(I,K) = .000041868*ZZ1(I,K)
 \end{minted}
-Arrays with this class of stencil must be indexed by all the induction
+Arrays with this class of stencil are indexed by all the induction
 variables used in the LHS without any offsets.
 
-\textbf{Some-reflexive} stencils (24\%) consist of only reflexive terms but
-leave some dimensions unspecified. Here are two examples:
+\textbf{Some-reflexive} specifications (24\%) consist of only reflexive terms but
+leave some dimensions unspecified. These two examples are from
+\textbf{GEOS-Chem} and \textbf{E3MG}: 
 \begin{minted}{fortran}
@@
 where for the first example \mintinline{fortran}{leafpools} contains an
-implicit loop, and for the second example \mintinline{fortran}{J} is
+implicit loop due to the range term \texttt{:}, and for the second \mintinline{fortran}{J} is
 the induction variable. This stencil permits array indexing by a subset of the LHS induction variables but again without
@@
 
-\textbf{Single-action} stencils (6\%) consist of a single \term{backward},
-\term{forward}, or \term{centered} region constant and some number of \term{reflexive}
-region constants.
+\textbf{Single-action} specifications (6\%) consist of a single \term{backward},
+\term{forward}, or \term{centered} region and some number of \term{reflexive}
+region constants. For example, taken from \textbf{GEOS-Chem}: 
 % /home/mrd45/src/corpus/geos-chem/GTMM/CasaRegridModule.F90:645
-\begin{minted}{fortran}
+\begin{minted}[breakafter=*]{fortran}
 do j=2,jm-ig
    do i=1,im
-      != stencil readOnce, (backward(depth=1, dim=2))*(reflexive(dim=1)) :: dm, q
+      != stencil readOnce, (backward(depth=1, dim=2)) *(reflexive(dim=1)) :: dm, q
       al(i,j) = 0.5e+0_fp*(q(i,j-1)+q(i,j)) + r3*(dm(i,j-1) - dm(i,j))
-   enddo
-enddo
 \end{minted}
-This stencil requires a neighbourhood index including zero-offset in one dimension only. The remaining dimensions must either have no offset or not pertain to an induction variable.
-
-\textbf{Single-action-irreflexive} stencils (3\%) ) consist of a single
-\term{backward}, \term{forward}, or \term{centered} irreflexive region constant
-and some number of reflexive region constants.
+% TODO: A BIT UNCLEAR
+%This stencil requires a neighbourhood index including zero-offset in one dimension only. The remaining dimensions must either have no offset or not pertain to an induction variable.
+These kinds of specification require zero offsets in one dimension and 
+and neighbourhood offsets (from 0 to $-1$ above) in the other.
+The remaining dimensions must either have no offset or not pertain to
+an induction variable.
+
+\textbf{Single-action-irreflexive} specifications (3\%) consist of a single
+\term{backward}, \term{forward}, or \term{centered} irreflexive region 
+and some number of reflexive region constants, \eg{}, from \textbf{GEOS-Chem}:
 % /home/mrd45/src/corpus/geos-chem/GeosRad/rrtmg_lw_rtrnmc.F90:355
@@
        (bbdtot * atot(lev) - gassrc)
-    != stencil readOnce, (forward(depth=1, dim=2, irreflexive)) :: cldfmc
+    != stencil readOnce, forward(depth=1, dim=2, irreflexive) :: cldfmc
     drad(lev-1) = drad(lev-1) + radld
 \end{minted}
-Notice that, in this last example, the read from
-\mintinline{fortran}{cldfmc} flows through onto the line with the
-stencil. Stencils of this type required a neightbourhood index without a zero-offset in one dimension only. The remaining dimensions must either have no offset or not pertain to an induction variable.
-
-The remaining stencil shapes (less than 1\%) consisted primarily of multiple \term{forward}, \term{backward}, or \term{centered} dimensions.
-We note that only 0.2\% of stencils found required use of the union (\term{+}) operator.
-Despite this low frequency these stencils are arguably the most important since they represent more complicated code and thus greater potential for programming errors.
-
-\subsection{Detecting programming errors}
-
-\subsubsection{Example: 2-D Jacobi iteration}
-
-One relatively basic example of a stencil program is two dimensional
+Note that the read from 
+\mintinline{fortran}{cldfmc} on line 5 flows to line 8. Stencils of
+the above kind require a neightbourhood index without a zero-offset in
+one dimension only. The remaining dimensions must either have no
+offset or not pertain to an induction variable.
+
+The remaining specification shapes (less than 1\%) consisted
+primarily of multiple \term{forward}, \term{backward}, or
+\term{centered} dimensions. We note that only $\sim$0.4\% of inferred
+stencil specifications required use of the \term{+} operator.
+Despite their low frequency these stencils are arguably the most important since they represent more complicated code and thus greater potential for programming errors.
+
+\subsection{Verification case study 1: 2-D Jacobi iteration}
+
+One relatively basic example of a stencil computation is two-dimensional
 Jacobi iteration that repeatedly goes through each cell in a matrix
 and computes the average value of the four adjacent cells. The kernel
-can be expressed in a line of Fortran code,
+can be expressed in a line of Fortran code:
 \begin{minted}{fortran}
@@
 \end{minted}
-and CamFort is able to infer a precise stencil specification for this
-line,
-\begin{minted}{fortran}
-  != stencil reflexive(dim=1)*centered(depth=1, dim=2, irreflexive) + reflexive(dim=2)*centered(depth=1, dim=1, irreflexive) :: a
+and CamFort is able to infer a precise specification for this:
+\begin{minted}[breakafter=+:,breakindent=-0.6em,breaksymbolsep=0.4em,linenos=false,xleftmargin=-0.5em]{fortran}
+  != stencil reflexive(dim=1)*centered(depth=1,dim=2,irreflexive)+ reflexive(dim=2)*centered(depth=1,dim=1,irreflexive) :: a
 \end{minted}
-which when inserted back into the program can be checked by CamFort as
-well. We then tested the error-detecting ability of CamFort by
+We then tested the error-detecting ability of our approach by 
 generating a family of 6,561 Jacobi iteration programs by perturbing
@@
 
-CamFort checked all 6,561 programs and found that only 24 of them were
+We assigned the specification inferred above to the generated Jacobi
+stencils. CamFort checked all 6,561 programs against this spec and found that only 24 of them were
 valid implementations of the stencil specification. This was the
-expected result because there are $4!$ possible correct permutations
+expected result as there are $4!$ possible correct permutations
 of the four terms in the kernel. All of the other programs were found
@@
 
-\subsubsection{Turbulence calculation}
+\subsection{Verification case study 2: Turbulence calculation}
 

### 2016-07-07 paper.tex
eval edits

@@
 We assigned the specification inferred above to the generated Jacobi
-stencils. CamFort checked all 6,561 programs against this spec and found that only 24 of them were
+stencils programs. CamFort checked all 6,561 programs against this specification and found that only 24 of them were
 valid implementations of the stencil specification. This was the
@@
   \label{fig:jacobi}
+  \vspace{-0.7em}
 \end{figure}
@@
 
-The Unified Model contains an implementation of the Smagorinsky subgrid-scale model for calculating turbulence.
-The source file in question yields 39 stencils from 340 lines of code.
-We selected one as an example:
-
-\begin{minted}[breakindent=0em,linenos=false,xleftmargin=0em]{fortran}
+The Unified Model has an implementation of the Smagorinsky
+subgrid-scale model for calculating turbulence on which our
+inference yields 39 specs from 340 lines of code. One example is:
+
+\begin{minted}[breakindent=0em,linenos=false,xleftmargin=0em,breakafter=)]{fortran}
 !=stencil readOnce,(forward(depth=1,dim=3,irreflexive))*(backward(depth=1,dim=1))*(backward(depth=1,dim=2))+(forward(depth=1,dim=3))*(backward(depth=1,dim=1))*(reflexive(dim=2))+(forward(depth=1,dim=3))*(backward(depth=1,dim=1,irreflexive))*(backward(depth=1,dim=2,irreflexive))+(forward(depth=1,dim=3))*(backward(depth=1,dim=2))*(reflexive(dim=1))+(backward(depth=1,dim=1))*(backward(depth=1,dim=2,irreflexive))*(reflexive(dim=3))+(backward(depth=1,dim=1,irreflexive))*(backward(depth=1,dim=2))*(reflexive(dim=3))::x
 \end{minted}
-
-This stencil specifies the access pattern to a 3-dimensional array (which we have renamed to be called \texttt{x}).
+%
+This specifies the access pattern to a 3-dimensional array (which we have renamed to be called \texttt{x}).
 The array is accessed 8 times in the computation.
@@
 For example changing \mintinline{fortran}{x(i+1,j,k)} to \mintinline{fortran}{x(i,j,k)} or \mintinline{fortran}{x(i-1,j,k)}.
-There are 48 possible errors of this type of which we found only 6 were not detected by the stencil check.
+There are 48 possible errors of this type of which we found only 6 were not detected by the specification checking.
 % /um/trunk/src/atmosphere/diffusion_and_filtering/turb_smagorinsky.F90

### 2016-07-07 paper.tex
Extend case study 2

@@
 For example changing \mintinline{fortran}{x(i+1,j,k)} to \mintinline{fortran}{x(i,j,k)} or \mintinline{fortran}{x(i-1,j,k)}.
-There are 48 possible errors of this type of which we found only 6 were not detected by the specification checking.
+
+There are 48 possible errors of this type of which we found only 6 were not 
+detected by the specification checking. The reason is that the inferred
+specification describes a family of access patterns. Some offset errors
+produce a stencil computation that belongs to the same family and goes
+undetected.
+
+One of the reasons this specification can catch many offset errors is that it is
+a \texttt{readOnce} specification. This prevents offset errors leading to
+duplicate access patterns be caught even if the resulting set of accesses are
+part of the same family.
 % /um/trunk/src/atmosphere/diffusion_and_filtering/turb_smagorinsky.F90
@@
 % 6 cases were not caught
-
 \section{Discussion}

### 2016-07-07 paper.tex
space saving

--- 2016-07-07 paper.tex
@@
 \begin{array}{c}
-%\framebox{$\textit{region} \equiv \textit{region}'$} \\[0.7em]
+%\framebox{$\textit{region} \equiv \textit{region}'$} \\[0.6em]
 \begin{array}{rll}
@@
 \stenFwdSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
- \equiv \; & \stenFwdSR{n}{d}{r} \; \texttt{+} \; \stenFwdSR{m}{d}{s} \\[0.7em]
+ \equiv \; & \stenFwdSR{n}{d}{r} \; \texttt{+} \; \stenFwdSR{m}{d}{s} \\[0.6em]
 %
@@
 \stenBwdSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
- \equiv \; & \stenBwdSR{n}{d}{r} \; \texttt{+} \; \stenBwdSR{m}{d}{s} \\[0.7em]
+ \equiv \; & \stenBwdSR{n}{d}{r} \; \texttt{+} \; \stenBwdSR{m}{d}{s} \\[0.6em]
 %
@@
 \stenCenSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
-\equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \; \stenCenSR{m}{d}{s} \\[0.7em]
+\equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \; \stenCenSR{m}{d}{s} \\[0.6em]
 %
@@
 \textit{$m \leq n$} \; \equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \;
-                      \stenFwdSR{m}{d}{s} \\[0.7em]
+                      \stenFwdSR{m}{d}{s} \\[0.6em]
 %
@@
 \textit{$m \leq n$} \; \equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \;
-                      \stenBwdSR{m}{d}{s} \\[0.7em]
+                      \stenBwdSR{m}{d}{s} \\[0.6em]
 %
@@
 \equiv \; & \stenFwdSR{n}{d}{r} \; \texttt{+} \; \stenBwdSR{n}{d}{s}
-  \\[0.7em]
+  \\[0.6em]
 %
@@
 \stenFwdS{n}{d} \\
-\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenFwdSR{n}{d}{r} \\[0.7em]
+\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenFwdSR{n}{d}{r} \\[0.6em]
 %
@@
 \stenBwdS{n}{d} \\
-\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenBwdSR{n}{d}{r} \\[0.7em]
+\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenBwdSR{n}{d}{r} \\[0.6em]
 %
@@
 \stenCenS{n}{d} \\
-\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenCenSR{n}{d}{r} \\[0.7em]
+\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenCenSR{n}{d}{r} \\[0.6em]
 %
@@
 \end{align*}
-\caption{Equations on specifications (regions)}
+\caption{Equations on specifications (\emph{regions})}
 \vspace{-1em}
@@
 
-\begin{definition} The $\mathsf{toSpec}$ constructor is defined:
-
+\begin{definition} The $\mathsf{toSpec}$ constructor is defined
+in Figure~\ref{fig:to-spec}
+
+\begin{figure}[t]
 %\newcommand{\llower}{\vect{x}^L_i}
 %\newcommand{\lupper}{\vect{x}^U_i}
+\vspace{-0.7em}
 \newcommand{\llower}{l}
@@
 {\scalebox{0.97}{
-\hspace{-1.8em}
 \begin{minipage}{1\linewidth}
@@
 \end{minipage}
-}}\\
+}}
+\caption{\textsf{toSpec} constructor of $\mathsf{A}(\specDNF)$ values.}
+\label{fig:tospec}
+\vspace{-0.9em}
+\end{figure}
 
@@
 
-\begin{lemma}[$\hat{\odot}$, $\hat{\oplus}$ soundness] $\forall S,
-  T \in \specDNF$ where $S \neq \epsilon \wedge T \neq \epsilon$:
+\begin{lemma}[$\hat{\odot}$, $\hat{\oplus}$ soundness]$\forall S,
+  T \in \specDNF$ where $S\!\neq \epsilon \wedge T\!\neq \epsilon$:
 \begin{align*}
@@
 \emph{generalisation} (Sect.~\ref{subsec:schemes-and-gen}). 
-Consistency is defined in Figure~\ref{fig:consistency},
-but we first set up some intermediate definitions before unpacking its definition.
+%Consistency is defined in Figure~\ref{fig:consistency},
+We first set up some intermediate definitions.% before unpacking its definition.
 
@@
   evaluation\label{tab:corpus}}
-\vspace{-1em}
+\vspace{-0.7em}
 \end{table}
@@
 \end{tabular}
-\caption{Stencil assignment rates\label{tab:tickAssign}}
+\caption{Stencil specification assignment rates\label{tab:tickAssign}}
 \vspace{-0.8em}
@@
 \begin{minted}[breakafter=+:,breakindent=-0.6em,breaksymbolsep=0.4em,linenos=false,xleftmargin=-0.5em]{fortran}
-  != stencil reflexive(dim=1)*centered(depth=1,dim=2,irreflexive)+ reflexive(dim=2)*centered(depth=1,dim=1,irreflexive) :: a
+  != stencil reflexive(dim=1)*centered(depth=1,dim=2,irreflexive)+reflexive(dim=2)*centered(depth=1,dim=1,irreflexive)
 \end{minted}
+assigned to \term{a}. 
 We then tested the error-detecting ability of our approach by 
@@
 predicate language using inductive program synthesis.
-The aim is that these high-level abstract specifications can
-then be translated into high-performance DSLs, easing the
-use of high-performance DSLs. Compared to our approach,
+The aim is that the high-level abstract specifications can
+be automatically translated into high-performance DSLs. Compared to our approach,
 their specification language is more fine grained, capturing
 the exact indexing pattern of a stencil. For example, the following
-is an example post-condition in their language, generated by their
+is a post-condition generated by their
 compiler \textsc{stng} from a Fortran stencil computation~\cite [p.3]{kamil2016verified}:
@@
 facilitate human reasoning, and avoid low-level lexical
-errors. However, \textsf{stng} specifications are generated, \ie{},
+errors. However, \textsc{stng} specifications are generated, \ie{},
 are never written by a programmer, avoiding this
@@
 able to ascribe a specification in the above case as the use of
-\texttt{t} on line 5 maps only to the definition on line 2, and no on
-line 6. Extending our approach to follow loop-carried dependencies is
+\texttt{t} on line 5 maps only to the definition on line 2, and not to
+that on line 6. Extending our approach to follow loop-carried dependencies is
 future work. An alternate approach for us would be to build on top of
@@
 \textsc{stng} (as an intermediate step).
-
-There are many DSLs for stencil computation aimed at either
-performance or verification, or both. Many eliminate out-of-bounds
-errors, \eg{},~\citet{DBLP:journals/corr/abs-1109-0777}.
 
@@
 
+There are many DSLs for stencil computation aimed at either
+performance or verification, or both. Many eliminate out-of-bounds
+errors, \eg{},~\citet{DBLP:journals/corr/abs-1109-0777}.
+
 \section{Conclusion and future work}

### 2016-07-07 paper.tex
Minor fortran typo

@@
 != stencil centered(depth=1, dim=1)*reflexive(dim=2) + centered(depth=1, dim=2)*reflexive(dim=1) :: a
-b(i,j) = -4*a(i,j) + a(i-1,j) + a(i+1,j)
+b(i,j) = -4*a(i,j) + a(i-1,j) + a(i+1,j) &
                    + a(i,j-1) + a(i,j+1)

### 2016-07-07 paper.tex
Gather lines into same line

@@
   do i=0, imax-1
-      x = a(i,j) - a(i+1,j+1)
-      y = a(i+1,j) - a(i,j+1)
+    x = a(i,j)-a(i+1,j+1); y = a(i+1,j)-a(i,j+1)
     != stencil forward(depth=1,dim=1) * forward(depth=1,dim=2) :: a
@@
   do j = 1, (m-1)
-      x       = a(i-1, j) + a(i+1, j)
-      y       = a(i, j-1) + a(i, j+1)
+    x = a(i-1, j)+a(i+1, j); y = a(i, j-1)+a(i, j+1)
     b(i, j) = (a(i, j) + x + y) / 5.0
@@
 != stencil readOnce, fivepoint :: b
-a(i,j) = b(i,j-1) + b(i,j) + b(i,j+1) + b(i-1,j) + b(i+1,j)
+a(i,j) = b(i,j-1)+b(i,j)+b(i,j+1)+b(i-1,j)+b(i+1,j)
 \end{minted}

### 2016-07-07 paper.tex
conclusion edits

--- 2016-07-07 paper.tex
@@
 
-\begin{definition} The $\mathsf{toSpec}$ constructor is defined
-in Figure~\ref{fig:to-spec}
-
 \begin{figure}[t]
@@
 \caption{\textsf{toSpec} constructor of $\mathsf{A}(\specDNF)$ values.}
-\label{fig:tospec}
-\vspace{-0.9em}
+\label{fig:to-spec}
+\vspace{-0.8em}
 \end{figure}
 
-\noindent
+\begin{definition} The $\mathsf{toSpec}$ constructor is defined
+in Figure~\ref{fig:to-spec}.
 The first two cases map to \texttt{forward} stencils since the lower
@@
 % 6 cases were not caught
-\section{Discussion}
+\section{Related work}
 \label{sec:discussion}
-
-\paragraph{Program synthesis}
-
-\anote{Any discussion left to do? - something on checking perhaps? access stencils?}
-
-\subsection{Related work}
 
@@
     do i=imin+1,imax
-        q = b(i, j)
-        a(i, j) = q + t
+        q = b(i, j);  a(i, j) = q + t
         t = q
@@
 % GPGPU
-This work shares some similarities with efforts to verify kernels
+Our work has some similarities with efforts to verify kernels
 written for General-Purpose Graphics Processing Unit (GPGPU)
@@
 method of transforming a large matrix. However, that work focuses
-mainly on the synchronisation of such kernels, and the avoidance of
+mainly on the synchronisation of such kernels and the avoidance of
 data races, while we are interested in the correctness of stencils
@@
 language. Other work on GPGPU computation, such as
-\citet{Zhang:2012:CGO}, has been primarily focused on generating
-optimised code based upon on relatively simple specifications: to
+\citet{Zhang:2012:CGO}, has focused primarily on generating
+optimised code based on relatively simple specifications: to
 provide performance while keeping the programmer's effort within
@@
 % Pochoir
-\citet{Tang:2011:SPAA} defines a specification language for writing
+\citet{Tang:2011:SPAA} define a specification language for writing
 stencils embedded in C++ (with Cilk extensions) that are then compiled
@@
 hyperspace cuts. The Pochoir specification language is used for
-describing the computing kernel, boundary conditions and shape of the
+describing the kernel, boundary conditions and shape of the
 stencil. Programs are first compiled in C++ with a template library
@@
 Pochoir preprocessor may be applied to generate the high-performance
-parallel version that takes advantage of the Cilk compiler.  Pochoir
+parallel versions using Cilk.  Pochoir
 is aimed primarily at assisting programmers who may be reluctant to
-implement the tricky cache-oblivious ``hypertrapezoidal'' algorithms
-that are significantly faster than the typical looping implementations
-of stencil computations. Like much of the related work, the main goal
-is optimisation rather than catching bugs and verification of
-program correctness.
+implement the high-performance tricky cache-oblivious
+``hypertrapezoidal'' algorithms for stencils. 
+%that are significantly faster than the typical looping implementations
+%of stencil computations.
+Like much of the related work, the main goal 
+is optimisation rather than verifying program correctness.
 
@@
 correctness problem explicitly by bringing a form of model-checking to
-verify certain stencil computations in partitioned global address
-space languages. These are scenarios where an array is divided into
-subarrays on multiple processors but each has the ability to access
-the others' memory. The authors avoid the state explosion dilemma of
-model-checking by relying upon the fact that accesses to the
-``boundary elements'' will require a different method by virtue of the
-non-locality of that memory. Since this access is performed in a
-different manner, those can be identified automatically by static
-analysis. To achieve this, the authors have designed a new language to
+verify certain stencil computations. This is in the context of
+decomposing stencils for parallel computation on multiple processors
+ in partitioned global address
+space languages. %These are scenarios where an array is divided into
+%subarrays on multiple processors but each has the ability to access
+%the others' memory. The authors avoid the state explosion dilemma of
+%model-checking by relying upon the fact that accesses to the
+%``boundary elements'' will require a different method by virtue of the
+%non-locality of that memory. Since this access is performed in a
+%different manner, those can be identified automatically by static
+%analysis. To achieve this, 
+The authors have designed a new language to
 be used for writing stencil computations. Much of the specification
@@
 
-There are many DSLs for stencil computation aimed at either
-performance or verification, or both. Many eliminate out-of-bounds
-errors, \eg{},~\citet{DBLP:journals/corr/abs-1109-0777}.
+%There are many DSLs for stencil computation aimed at either
+%performance or verification, or both. Many eliminate out-of-bounds
+%errors, \eg{},~\citet{DBLP:journals/corr/abs-1109-0777}.
 
@@
 
-In this paper we have described an annotation language which can be
-used to specify the access pattern of a stencil computation. We
-defined a denotational semantic model for our language and proved its
-soundness. This gave a basis for an equational theory of
-specifications and also justifies our inference and checking
-procedure. We evaluated our approach by implementing the stencil inference and
-checking procedure in CamFort, an open-source analysis tool\footnote{\url{https://github.com/camfort/camfort}} for
-Fortran programs and running the inference process over a corpus of
-1.1 million lines of Fortran code.
-
-Our motivation for this work was to provide an intermediate
-specification language to allow scientists to verify numerical
-computations. The most common types of stencils we found consisted only of
-reflexive region constants which are unlikely to give rise to
-programming errors. However, around 10\% of stencils covered more
-complicated program structures for which indexing errors are much more
-plausible. We showed two examples of more complicated stencils for
-which CamFort was able to detect a significant majority of possible
-indexing errors.
+In this paper, we described an abstract specification language 
+for the access pattern of a stencil computation. Compared with the
+related work (discussed above), our language is novel in that it does
+not aim to replicate indexing in the specification, but instead
+provides a set of abstract spatial combinators which capture a large number
+of common patterns.
+
+We evaluated our approach by implementing the stencil inference and
+checking procedure ontop of CamFort, an open-source analysis
+tool\footnote{\url{https://github.com/camfort/camfort}} for Fortran
+programs and running the inference process over a corpus of 1.1
+million lines of Fortran code.  Our motivation for this work was to
+provide an intermediate specification language to allow scientists to
+verify numerical computations.  Our evaluation showed that relatively
+simple stencil comptuations are the norm, for which we can specify a
+large proportion, comprising mainly reflexive region constants which
+are less likely to give rise to programming errors.  However, around
+10\% of stencils covered more complicated program structures for which
+indexing errors are much more plausible. In the more detailed cases,
+with larger specifications, the importance and impact of such
+specifications grows.  We showed two examples of more complicated
+stencils, for which CamFort was able to detect a significant majority
+of possible indexing errors in our case studies.
 
 Our definition of a stencil computation meant that we generated no
-specification for a large number of iterated array accesses in the corpus. For
-future work we will be looking at expanding the definition to allow us
-to capture more of these examples.
+specification for a large number of iterated array accesses in the
+corpus, such as stencils that incorporates reductions. For
+future work, we will be looking into expanding the flexibility of 
+our inference approach to more of these examples. Furthermore, our specification
+language can be used more flexibly to give descriptions of data access
+patterns in any context, not just within a stencil.
 

### 2016-07-07 paper.tex
edits

@@
 We also evaluate our implementation on real scientific code 
-used actively in research. Our case studies include the Unified Weather 
-model used by the UK Met Office~\cite{um-website} ($\approx{}600\textit{kloc}$) and other
-research-based numerical models, ranging from small to large.
+used actively in research, ranging from small ($\sim$700loc) to large
+($\sim$600kloc). 
+ %Our case studies include the Unified Weather 
+%model used by the UK Met Office~\cite{um-website} ($\approx{}600\textit{kloc}$) and other
+%research-based numerical models, ranging from small to large.
 
@@
 z = a(i, j+1) + a(i-1, j+1) + a(i+1, j+1)
-
 != stencil centered(depth=1, dim=1) * centered(depth=1, dim=2) :: a
@@
 
-\subsection{Note on the design}
+\paragraph{Note on the design}
 
@@
 the one-dimensional heat equation, an FTCS discretisation
-provides approximation code with the following stencil~\cite{recktenwald2004finite}:
+provides code with the following stencil~\cite{recktenwald2004finite}:
 %%
@@
 specifications which is defined by the binary relation $\equiv$
-between region terms. Stencil declarations are considered
-equal when they have the same modifiers and $\equiv$ equates
-their regions, \ie{}:
-\begin{align*}
-(\textit{region} \equiv \textit{region'}) 
- \Rightarrow \quad\;\; & \texttt{stencil} \; [\textit{approx},] \; [\textit{mult},] \;
-\textit{region} \\[-0.4em]
-\quad \equiv \; & \texttt{stencil} \; [\textit{approx},] \;
-            [\textit{mult},] \; \textit{region}'
-\end{align*}
-%
-The equations of Figure~\ref{fig:equations} use abbreviations
+between region terms. The equations use abbreviations
 \term{refl}, \term{fwd}, \term{bwd}, \term{cen}, and \irreflS{}
@@
 
-The equations reveal a partial semantics for stencil
-specifications, which we briefly summarise.
+Stencil specifications in \textit{spec} are considered
+equal when they have the same modifiers and $\equiv$ equates
+their regions, \ie{}:
+\begin{align*}
+(\textit{region} \equiv \textit{region'}) 
+ \Rightarrow \quad\;\; & \texttt{stencil} \; [\textit{approx},] \; [\textit{mult},] \;
+\textit{region} \\[-0.4em]
+\quad \equiv \; & \texttt{stencil} \; [\textit{approx},] \;
+            [\textit{mult},] \; \textit{region}'
+\end{align*}
+
+The equations reveal a part of the semantics for our
+specification language, which we briefly unpack.
 
@@
 the set of $n$-dimensional schemes.
-\end{definition}
 
@@
 and the checking procedures.
-
+\end{definition}
+%
 \begin{definition}%[Schematic]
@@
 %
-The $\downarrow$ operator is partial\footnote{We represent
-  partiality in our signatures with the set $A + \bot$.} 
+The $\downarrow$ operator is partial
  $\downarrow : \textit{expr} \rightarrow (\domainVal + \bot)$
@@
 where $\mathsf{IV}(e)$ is the set of free induction variables
-appearing in an expression, and $\neigh{}$ classifies expressions which
+in an expression, and $\neigh{}$ classifies expressions which
 are constant offsets from an induction variable
@@
 of an induction variable) then it is mapped to $\infty$, otherwise
-$\downarrow\!e$ is undefined. The $\mathsf{schematic}$ function
+$\downarrow\!e$ is undefined (returns $\bot$). The $\mathsf{schematic}$ function
 is undefined if $\downarrow$ of any of its components $e_i$ is
@@
 
-\begin{figure}[t]
-\vspace{-0.3em}
-\hspace{-0.5em}
-\begin{minipage}{0.16\linewidth}
-\begin{align*}
-\setlength{\arraycolsep}{0em}
-\begin{array}{c}
-\quad \\[0em]
-\begin{tikzpicture}[scale=0.8]
-%\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
-\node at (1.25,1.22) {\textnormal{\footnotesize{0,0}}};
-\draw[thick] (1,1) rectangle (1.5,1.5);
-\draw[thick] (1.5,1) rectangle (2,1.5);
-\draw[thick] (1,0.5) rectangle (1.5,1);
-\draw[thick] (0.5,1) rectangle (1,1.5);
-\draw[thick] (1,1.5) rectangle (1.5,2);
-\end{tikzpicture} \\[0em]
-(a)
-\end{array}
-\end{align*}
-\end{minipage} $\rightarrow$ \begin{minipage}{0.43\linewidth}
-\begin{align*}
-\setlength{\arraycolsep}{0em}
-\begin{array}{c}
-\begin{tikzpicture}[scale=0.72]
-%\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
-%\node at (1.25,1.22) {\textnormal{\small{0,0}}};
-\draw [black, thin] (0.35,0) to [round left paren ] (0.35,2);
-\draw[thick] (0.5,1) rectangle (2,1.5);
-\draw[thick] (1,0.40) rectangle (1.5,0.90);
-\draw[thick] (1,1.60) rectangle (1.5,2.10);
-\node at (2.3,1.22) {+};
-\draw[thick] (3.8,1) rectangle (4.3,1.5);
-\draw[thick] (2.6,1) rectangle (3.1,1.5);
-\draw[thick] (3.2,0.5) rectangle (3.7,2);
-\draw [black, thin] (4.4,0) to [round right paren ] (4.4,2);
-\end{tikzpicture} \\[-0.5em] (b)
-\end{array}
-\end{align*}
-\end{minipage}$\rightarrow$ \begin{minipage}{0.25\linewidth}
-\begin{align*}
-\setlength{\arraycolsep}{0em}
-\begin{array}{c}
-\begin{tikzpicture}[scale=0.72]
-%\node at (1.25,1.22) {\textnormal{\small{0,0}}};
-\draw [black, thin] (0.8,0) to [round left paren ] (0.8,2);
-\draw[thick] (1,0.5) rectangle (1.5,2);
-\node at (1.8,1.22) {+};
-\draw[thick] (2.1,1) rectangle (3.6,1.5);
-\draw [black, thin] (3.7,0) to [round right paren ] (3.7,2);
-%\node at (2.75,1.22) {\textnormal{\small{0,0}}};
-\end{tikzpicture} \\[-0.5em] (c)
-\end{array}
-\end{align*}
-\end{minipage}
-\caption{Illustration of grouping spans in
-specification inference}
-\label{fig:inference-steps-informal}
-\vspace{-0.8em}
-\end{figure}
-
 % taking permustations is a systematic way of producing all rotations
@@
  $(\mu \, \spanOp) \; N(M)$ where \spanOp{} is defined by: \\[-0.5em]
+
+\begin{figure}[t]
+\vspace{-0.6em}
+\hspace{-0.5em}
+\begin{minipage}{0.16\linewidth}
+\begin{align*}
+\setlength{\arraycolsep}{0em}
+\begin{array}{c}
+\quad \\[0em]
+\begin{tikzpicture}[scale=0.8]
+%\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
+\node at (1.25,1.22) {\textnormal{\footnotesize{0,0}}};
+\draw[thick] (1,1) rectangle (1.5,1.5);
+\draw[thick] (1.5,1) rectangle (2,1.5);
+\draw[thick] (1,0.5) rectangle (1.5,1);
+\draw[thick] (0.5,1) rectangle (1,1.5);
+\draw[thick] (1,1.5) rectangle (1.5,2);
+\end{tikzpicture} \\[0em]
+(a)
+\end{array}
+\end{align*}
+\end{minipage} $\rightarrow$ \begin{minipage}{0.43\linewidth}
+\begin{align*}
+\setlength{\arraycolsep}{0em}
+\begin{array}{c}
+\begin{tikzpicture}[scale=0.72]
+%\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
+%\node at (1.25,1.22) {\textnormal{\small{0,0}}};
+\draw [black, thin] (0.35,0) to [round left paren ] (0.35,2);
+\draw[thick] (0.5,1) rectangle (2,1.5);
+\draw[thick] (1,0.40) rectangle (1.5,0.90);
+\draw[thick] (1,1.60) rectangle (1.5,2.10);
+\node at (2.3,1.22) {+};
+\draw[thick] (3.8,1) rectangle (4.3,1.5);
+\draw[thick] (2.6,1) rectangle (3.1,1.5);
+\draw[thick] (3.2,0.5) rectangle (3.7,2);
+\draw [black, thin] (4.4,0) to [round right paren ] (4.4,2);
+\end{tikzpicture} \\[-0.5em] (b)
+\end{array}
+\end{align*}
+\end{minipage}$\rightarrow$ \begin{minipage}{0.25\linewidth}
+\begin{align*}
+\setlength{\arraycolsep}{0em}
+\begin{array}{c}
+\begin{tikzpicture}[scale=0.72]
+%\node at (1.25,1.22) {\textnormal{\small{0,0}}};
+\draw [black, thin] (0.8,0) to [round left paren ] (0.8,2);
+\draw[thick] (1,0.5) rectangle (1.5,2);
+\node at (1.8,1.22) {+};
+\draw[thick] (2.1,1) rectangle (3.6,1.5);
+\draw [black, thin] (3.7,0) to [round right paren ] (3.7,2);
+%\node at (2.75,1.22) {\textnormal{\small{0,0}}};
+\end{tikzpicture} \\[-0.5em] (c)
+\end{array}
+\end{align*}
+\end{minipage}
+\caption{Illustration of grouping spans in
+specification inference}
+\label{fig:inference-steps-informal}
+\vspace{-0.8em}
+\end{figure}
 
@@
 \begin{figure}
+\vspace{-0.4em}
 \input{consistency.tex}
@@
   \label{fig:jacobi}
-  \vspace{-0.7em}
+  \vspace{-2em}
 \end{figure}
@@
 undetected.
-
 One of the reasons this specification can catch many offset errors is that it is
@@
 
-In this paper, we described an abstract specification language 
-for the access pattern of a stencil computation. Compared with the
+We presented in this work an specification language 
+for the access pattern of a stencil computation. Compared to the
 related work (discussed above), our language is novel in that it does
 not aim to replicate indexing in the specification, but instead
-provides a set of abstract spatial combinators which capture a large number
+provides abstract spatial combinators which capture a large number
 of common patterns.
@@
 checking procedure ontop of CamFort, an open-source analysis
-tool\footnote{\url{https://github.com/camfort/camfort}} for Fortran
+tool\footnotemark for Fortran
 programs and running the inference process over a corpus of 1.1
@@
 
+\footnotetext{\url{https://github.com/camfort/camfort}}
 \bibliography{references}

### 2016-07-07 paper.tex
edits

@@
-\documentclass[9pt]{sigplanconf}
+\documentclass[9pt,preprint]{sigplanconf}
 
@@
 
-\item we evaluate our approach with an
-implementation as an extension of CamFort,
-an open-source program analysis tool for Fortran
-programs. We use our implementation to evaluate our approach against
-a corpus of scientific applications, totalling roughly one million
-lines of code;
+\item we provide an implementation of our approach as an extension to
+  CamFort, an open-source program analysis tool for Fortran programs.
+  We use this for a quantiative evaluation of our approach over a corpus of
+  scientific applications, totalling roughly one million lines of
+  code;
 
@@
 
-\subsection{Notation and convention}
-\label{sec:notation}
+\paragraph{Notation and convention}
+%\label{sec:notation}
 
 \renewcommand*{\arraystretch}{0.8}
-For the target language
-syntax, $e$ ranges over its expressions (may be impure) and $v$
+For the target language, $e$ ranges over expressions (may be impure) and $v$
 overs its (imperative) variables. %By
@@
 \label{fig:region-model}
-\vspace{-0.7em}
+\vspace{-1em}
 \end{figure}
@@
 \begin{figure}
-\vspace{-0.5em}
 \input{spatial-model.tex}
@@
 specifications (Sect.~\ref{subsec:inference}), for checking code
-against specifications (Sect.~\ref{subsec:checking}) and for
-synthesising programs from specifications
-(Sect.~\ref{subsec:synthesis}).  These utilise the semantic model of
+against specifications (Sect.~\ref{subsec:checking}). These utilise the semantic model of
 Section~\ref{sec:semantics}.
@@
 %
-\begin{enumerate}
-\item basic blocks (CFG);
-\item induction variables per basic block;
-\item (interprocedural) data-flow analysis, providing a \emph{flows to}
+\textbf{(1)} basic blocks (CFG);
+\textbf{(2)} induction variables per basic block;
+\textbf{(3)} (interprocedural) data-flow analysis, providing a \emph{flows to}
   graph (as shorthand, the function
@@
   closure of the flows graph);
-\item type information per variable, where we use the predicate
+\textbf{(4)} type information per variable, where we use the predicate
 \arrayTy{} to classify variables of array type.
-\end{enumerate}
 %
@@
 \begin{figure}[t]
-\vspace{-0.6em}
+\vspace{-0.8em}
 \hspace{-0.5em}
@@
 permutation flips the order of the two elements in the 
-vectors. \\
+vectors. \\[-1em]
 
@@
 \begin{align*}
-%\textsf{lower}(S) \odot \textsf{lower}(T) = \; & 
-%      \textsf{lower} \, (S \hat{\odot} T) \\
-%\textsf{upper}(S) \odot \textsf{upper}(T) = \; &
-%      \textsf{upper} \, (S \hat{\odot} T) \\
 \textsf{lower}(S) \odot \textsf{exact}(T) = \; &
@@
        %\textsf{exact}(T) \odot \textsf{upper}(S) =
-      \; \textsf{both} \, (T, \; S \hat{\odot} T) \\
+       \textsf{both} \, (T, \; S \hat{\odot} T) \\
 \textsf{lower}(S) \odot \textsf{both}(T_l,T_u) = \; &
@@
 \textbf{Total} & \overallFiles  & \overallparseOk & \overallLoC & \overalllinesParsed \\
-\hline
 \end{tabular}
@@
                 \overallnumStencilLines &
-                \overallboundedBoth    \\
-\hline
+                \overallboundedBoth   
 \end{tabular}
 \caption{Stencil specification assignment rates\label{tab:tickAssign}}
-\vspace{-0.8em}
+\vspace{-1em}
 \end{table}
@@
 
-The Unified Model has an implementation of the Smagorinsky
-subgrid-scale model for calculating turbulence on which our
-inference yields 39 specs from 340 lines of code. One example is:
-
+\begin{figure}[t]
 \begin{minted}[breakindent=0em,linenos=false,xleftmargin=0em,breakafter=)]{fortran}
@@
 \end{minted}
-%
+\caption{A complex specification inferred from
+  \textbf{Unified Model}\label{fig:smagorinsky}}
+\vspace{-1em}
+\end{figure}
+
+\noindent
+The Unified Model has an implementation of the Smagorinsky
+subgrid-scale model for calculating turbulence on which our
+inference yields 39 specs from 340 lines of code. We show
+one example in Figure~\ref{fig:smagorinsky}. 
 This specifies the access pattern to a 3-dimensional array (which we have renamed to be called \texttt{x}).
@@
 
-There are 48 possible errors of this type of which we found only 6 were not 
-detected by the specification checking. The reason is that the inferred
+There are 48 possible errors of this type of which only 6 were not 
+detected by the checker. The reason is that the inferred
 specification describes a family of access patterns. Some offset errors

### 2016-07-07 paper.tex
tweaks and spelling

@@
   code, but one which is error-prone due to fine-grained indexing
-  errors. We hypothesise that in practise, stencil computations tend
+  errors. We hypothesise that in practice, stencil computations tend
   to have a regular shape which is amenable to static analysis and
@@
   CamFort, an open-source program analysis tool for Fortran programs.
-  We use this for a quantiative evaluation of our approach over a corpus of
+  We use this for a quantitative evaluation of our approach over a corpus of
   scientific applications, totalling roughly one million lines of
@@
 Our specification language is an approach in this direction, providing
-a system of lighweight specification for numerical code.
+a system of lightweight specification for numerical code.
 The aim is to not only provide a way of exposing errors, but also
@@
 its maintenance. In order to increase the likelihood of impact, we aim
-to integrate with existing practises smoothly, hence our
+to integrate with existing practices smoothly, hence our
 specifications are integrated as comments and our tool can synthesis
@@
   relative index defined as constant translations of base induction
-  variables. (The relation $\equiv$ hre identifies terms up-to commutativity
+  variables. (The relation $\equiv$ here identifies terms up-to commutativity
   of $+$ and the inverse relation of $+$ and $-$ \eg{},
@@
 for \term{reflexive}, \term{forward}, \term{backward},
-\term{centered}, and \irrefl{} respectively. Futhermore, the abbreviations include syntactic sugar
-for quantifying over the abseence or presence of the \irrefl{}
+\term{centered}, and \irrefl{} respectively. Furthermore, the abbreviations include syntactic sugar
+for quantifying over the absence or presence of the \irrefl{}
 attribute on region constants as a boolean. For example,
@@
 depth is equivalent to a \term{centered} region of that depth.
-If the \term{forward} part is irrefleixve ($r =
+If the \term{forward} part is irreflexive ($r =
 \textit{false}$) but \term{backward} is not ($s =
@@
 enforces that every index is read just once to be dropped
-as an approxiamtion; (\trule{\textsc{bound}}) gives
+as an approximation; (\trule{\textsc{bound}}) gives
 a connection between the \term{atLeast} and \term{atMost}
@@
  $\downarrow : \textit{expr} \rightarrow (\domainVal + \bot)$
-mapping individal indexing expressions to 
+mapping individual indexing expressions to 
 an indexing scheme component, defined:
@@
 modelled) mapping closed\footnote{That is, we assume that there are no
-  occurences of \textit{rvar} in a specification being modelled.  Any
+  occurrences of \textit{rvar} in a specification being modelled.  Any
   \emph{open} specification containing region variables can be made
-  closed by straightforward syntactic subtitution with a (closed)
+  closed by straightforward syntactic substitution with a (closed)
   \textit{region}.} specifications to sets of
-$n$-dimensional indexing schemes. The interperation is overloaded on
+$n$-dimensional indexing schemes. The interpretation is overloaded on
 \emph{regions} in Figure~\ref{fig:region-model} and on the top-level
@@
 $\term{*}$ takes the tensor $\otimes$ of subterm models. The tensor is
-non-trivial; we give some informal intutition first.
+non-trivial; we give some informal intuition first.
 
@@
 $v_i$ in each dimension). Some care is then taken over
-how this interacts with $\infty$: pairwise permtuation
-corresponds to non-determinstic choice, in each dimension $i$,
+how this interacts with $\infty$: pairwise permutation
+corresponds to non-deterministic choice, in each dimension $i$,
 between two values $u_i$ and $v_i$, but if either
@@
 %both models to be satisfied. Use of $dims$ gives the dimensions that
-%constitutent models talk about and is used in the product to eliminate
+%constituent models talk about and is used in the product to eliminate
 %general unifier appearing in these dimensions. Allowing general unifier
@@
 %  u in M * N   u < w
-% mplies that
+% implies that
 %  u_1 in M  u1 < w
@@
 %
-The permutated schemes $x = (u \bowtie v)_i$ added to the model 
+The permuted schemes $x = (u \bowtie v)_i$ added to the model 
 $M \otimes N$ are only those which have do not have $\infty$ in 
@@
 \textit{spec} syntax, which includes the multiplicity and
-approximation modifiers. The domain of this interepretation
+approximation modifiers. The domain of this interpretation
 is $\textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$
 where $\textsf{Mult}$ and $\textsf{Approx}$ are labelled variants
-labelling our set-basd model with additional information.
+labelling our set-based model with additional information.
 
@@
 %
-If the LHS $\overline{e_1}$ contains non-$0$ neighour offsets 
-then all RHS schemes are \emph{relativized} by
+If the LHS $\overline{e_1}$ contains non-$0$ neighbour offsets 
+then all RHS schemes are \emph{relativised} by
 $\textsf{schematic}(\overline{e_1})$. For example, 
@@
   $\domainVal{}$. This gives the co-ordinates of a lower-bound vertex
-  (first comonent) and the upper-bound (second component). We write 
+  (first component) and the upper-bound (second component). We write 
   spans as $\vect{x} = [\vect{x}^L \; \vect{x}^U]$, where $\vect{x}^L$
@@
 
-% taking permustations is a systematic way of producing all rotations
+% taking permutations is a systematic way of producing all rotations
 % of the n-dimensional space - forward ref to fold
@@
 the unit spans in Figure~\ref{fig:inference-steps-informal}(a)
-are colaesced into the contiguous spans in
+are coalesced into the contiguous spans in
 Figure~\ref{fig:inference-steps-informal}(b).
@@
 Thus, if neither specification is empty, their $\hat{\oplus}$ is just
-the syntatic sum $\term{+}$. Then $\oplus$ is the lifting of
+the syntactic sum $\term{+}$. Then $\oplus$ is the lifting of
 $\hat{\oplus}$ to \textsf{Approx} with the shape as $\odot$ in the previous
@@
 of step 4 above still holds since its consistency condition
-universally quanitifes over the \emph{spec model}. However, for
+universally quantifies over the \emph{spec model}. However, for
 the upper bound, the consistency condition universally quantifies
@@
 However, the converse direction does not hold. 
-The regions describing the more acceptiong five-point stencil
+The regions describing the more accepting five-point stencil
 above do not follow from the definition of $<:$ on the previous
@@
 %      centered(depth=1,dim=2)}}}$.
- However, consitency (and thus $\sqsubseteq$) shows
+ However, consistency (and thus $\sqsubseteq$) shows
 that we can check the five-point stencil code against the more general
-specifcation of {\textcolor{darkgreen}{\term{fivepoint}}}. Thus, we
+specification of {\textcolor{darkgreen}{\term{fivepoint}}}. Thus, we
 can always use an approximation in place of the
@@
 \label{sec:evaluation}
-We evaluated our specificataion language on a corpus of scientific
+We evaluated our specification language on a corpus of scientific
 code.  We overview the corpus itself before presenting our results.
@@
 induction variables on the RHS are not a subset of those in the LHS. These
-cases are not stencils by our definition. The degenereate case of this
+cases are not stencils by our definition. The degenerate case of this
 is to have only constant indices on the LHS. We see lots of examples
@@
 
-If the statement meets the stencil critera but the inference process
+If the statement meets the stencil criteria but the inference process
 cannot determine an exact stencil for a statement it will instead
@@
 \mintinline{fortran}{cldfmc} on line 5 flows to line 8. Stencils of
-the above kind require a neightbourhood index without a zero-offset in
+the above kind require a neighbourhood index without a zero-offset in
 one dimension only. The remaining dimensions must either have no
@@
 verify numerical computations.  Our evaluation showed that relatively
-simple stencil comptuations are the norm, for which we can specify a
+simple stencil computations are the norm, for which we can specify a
 large proportion, comprising mainly reflexive region constants which

### 2016-07-07 paper.tex
tweak

@@
 (where $n$ is the maximum dimensionality of the specification being
-modelled) mapping closed\footnote{That is, we assume that there are no
+modelled) mapping closed\footnote{That is, we assume there are no
   occurrences of \textit{rvar} in a specification being modelled.  Any

### 2016-07-07 paper.tex
tweak
title
changed wording of 10% result in tinro

@@
 
-\title{Verification of Stencil Computations through Spatial
+\title{Verification of Stencil Computations with Spatial
   Specifications}
@@
 \item we provide an implementation of our approach as an extension to
-  CamFort, an open-source program analysis tool for Fortran programs.
-  We use this for a quantitative evaluation of our approach over a corpus of
-  scientific applications, totalling roughly one million lines of
-  code;
-
-\item we show that around 10\% of stencils are in a class that
-  represent a plausible scenario for a programming error;
+  CamFort, an open-source program analysis tool for Fortran programs;
+
+\item we identified 35,000 stencils in one million lines of code. Approximately 
+10\% of the stencils we found correspond to code where an indexing error is plausible;
 
@@
 \multicolumn{2}{l}{\qquad\qquad \mid \; \stenCen{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]}} \\
-\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{region} \; \term{+} \; \nonterm{region}} \\
-\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{region} \; \term{*} \; \nonterm{region}} \\[0.5em]
+\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{region} \; \term{+}
+  \; \nonterm{region} \; \mid \; \nonterm{region} \; \term{*} \; \nonterm{region}} \\[0.5em]
 %\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{rvar}}  

### 2016-07-07 paper.tex
secion 7

--- 2016-07-07 paper.tex
@@
   \subfigure[]{\includegraphics[width=0.1\textwidth]{jacobi-incorrect3.png}}
-  \caption{Greyscale output of several perturbations of a Jacobi
-    iteration program. The correct output is shown in
-    \subref{fig:jacobicorrect}.}
+  \caption{Perturbations of Jacobi code; \subref{fig:jacobicorrect} is reference output.}
   \label{fig:jacobi}
-  \vspace{-2em}
+  \vspace{-1em}
 \end{figure}

### 2016-07-07 paper.tex
edit

@@
 Line 3 is the core of the stencil computation, calculating
-the value at \texttt{b(i)} from a neighbourhood of elements about
+values at \texttt{b(i)} from a neighbourhood of elements about
 \texttt{a(i)}. Line 5 swaps
 \texttt{a} and \texttt{b} between iterations, where \texttt{b} becomes the
-input for the next iteration. In Fortran, parentheses \texttt{( )} are used
-for array subscripts rather than the more familiar bracket syntax \texttt{[ ]}.
-
+input for the next iteration. %In Fortran, parentheses \texttt{( )} are used
+%for array subscripts rather than the more familiar bracket syntax \texttt{[ ]}.
+%
 %This stencil computation exhibits statically decideable
@@
 and the partitioning shape for parallel implementations.
+
 In this example, the access pattern is simple, static, and easily
@@
   CamFort, an open-source program analysis tool for Fortran programs.
-  We use this for a quantitative evaluation of our approach over a corpus of
-  scientific applications, totalling roughly one million lines of
-  code;
 

### 2016-07-07 paper.tex
sten def

@@
 
-\usepackage[]{hyperref}
+\usepackage[draft]{hyperref}
 \hypersetup{
@@
   numerical computation is derived from an underlying mathematical
-  model, but this is often difficult to use as a specification as it
-  is highly decoupled from the implementation. We have been studying
-  specification techniques in the middle ground between low-level and
-  high-level specifications for numerical code. In this paper, we
-  focus on \emph{stencil computations}---a common idiom in numerical
+  model, but this is often infeasible as a specification since it
+  is highly decoupled from the implementation. In this paper, we consider a 
+  lightweight specification technique between low-level and
+  high-level specifications for numerical code. We
+  target \emph{stencil computations}---a common idiom in numerical
   code, but one which is error-prone due to fine-grained indexing
-  errors. We hypothesise that in practice, stencil computations tend
+  errors. We show that in practice, stencil computations tend
   to have a regular shape which is amenable to static analysis and
-  simple specification. We describe a abstract spatial specification
+  simple specification. We describe an abstract spatial specification
   language for stencil computations that can be embedded as
   annotations in source code, along with inference, checking, and
-  specification synthesis procedures. We evaluate our
+  specification synthesis. We evaluate our
   language against a corpus of numerical Fortran code, and show that
@@
   regular, static shape. Around 10\% of the stencils found evoke a
-  programming pattern for which indexing errors are plausible. We
-  investigate two stencils in detail and find that we are able to
-  detect the vast majority of potential errors. This paper details our
-  stencil specification language, its static semantics, inference,
+  programming pattern for which indexing errors are plausible. 
+  We simulate relevant programming errors in two stencil examples, 
+  and show that we are able to detect the vast majority of potential
+  errors. This paper details our stencil specification language, a
+  denotational model, inference,
   checking, and synthesis procedures, our implementation, and
-  our evaluation study.
+  our evaluation studies.
 \end{abstract}
@@
 such descriptions, abstracting over the fine-grained detail of stencil
-access patterns. For our first Laplace example, our
+access patterns. For the Laplace example, our
 inference procedure provides the following specification to line 3:
@@
 \begin{itemize}[leftmargin=0.8em]
-\item we introduce a specification language for
-stencil computation that captures many common forms
-of data access pattern (Section~\ref{sec:lang}), providing
- an equational theory and denotational model (Sect.~\ref{sec:semantics});
+\item we introduce a specification language for stencils 
+  that captures many common forms of data access pattern
+  (Section~\ref{sec:lang});
 
 \item we detail inference, checking, and specification synthesis 
-algorithms for our language (Sect.~\ref{sec:analysis}), leveraging the semantic model;
+algorithms for our language (Sect.~\ref{sec:analysis}), derived from
+the a denotational model for the language (Sect.~\ref{sec:semantics});
 
 \item we provide an implementation of our approach as an extension to
-  CamFort, an open-source program analysis tool for Fortran programs.
-
-\item we show that around 10\% of stencils are in a class that
-  represent a plausible scenario for a programming error;
-
-\item we provide a more detailed verification case study of two
+  CamFort, an open-source program analysis tool for Fortran;
+
+\item we give a quantiative study of
+  stencil computations in a corpus of numerical Fortran code,
+  totalling one million lines of code. Our tool identifies and
+  synthesises specifications for 35,000 stencil computations in the
+  corpus. Approximately 10\% of the stencils we found are non-trivial,
+  corresponding to code which is a possible source of errors;
+
+\item we give a more detailed verification case study of two
   particular stencil computation and show that we
@@
   of $+$ and the inverse relation of $+$ and $-$ \eg{},
-  $(-b) + i \equiv i - b$).  We classify neighbourhood indices using
-  the predicate \neigh{}. %An array subscript $v(\bar{e})$ where every
-% $e \in \bar{e}$ is a neighbourhood index is called a
-% \emph{neighbourhood subscript}.
+  $(-b) + i \equiv i - b$).  We classify neighbourhood indices 
+  by the predicate $\neigh_I$ where $I$ is a set of induction
+  variables. We omit $I$ when implicit from the context.
 \label{def:neighbour}
@@
 \label{def:schematic}
-The partial function $\textsf{schematic}$ maps
-source indexing terms $(e_1, .., e_n)$
+The partial function $\textsf{schematic}_I$ (parameterised by a set of
+induction variables $I$) maps source indexing terms $(e_1, .., e_n)$
 to $n$-dimensional index schemes, defined:
@@
 \begin{align*}
-\mathsf{schematic}(e_1, \ldots, e_n) & =
-[\downarrow\!e_1 \ldots \downarrow\!e_n ]
-\quad \textit{iff} \;\; \forall i . \downarrow\!e_i \neq \bot
-\end{align*}
-%
-The $\downarrow$ operator is partial
- $\downarrow : \textit{expr} \rightarrow (\domainVal + \bot)$
-mapping individual indexing expressions to 
-an indexing scheme component, defined:
-%%
-\begin{align*}
-\downarrow\!e
+\mathsf{schematic}_I(e_1, ..., e_n) & =
+[\mathsf{s}_I(e_1) ... \mathsf{s}_I(e_n) ]
+\quad \textit{iff} \;\; \forall x . \mathsf{s}_I(e_x) \neq \bot
+\end{align*}
+%
+where $\mathsf{s}_I$ is partial and 
+maps individual indexing expressions to 
+ indexing scheme components in $\domainVal{}$ defined:
+%%
+\begin{align*}
+\mathsf{s}_I(e)
  & =  \begin{cases}
-a & e \equiv i + a \; \wedge \; \mathsf{neigh}(i + a) \\
--a & e \equiv i - a \; \wedge \; \mathsf{neigh}(i - a) \\
+a & e \equiv i + a \; \wedge \; \neigh_I(e) \\
+-a & e \equiv i - a \; \wedge \; \neigh_I(e) \\
 \infty & \textit{$\mathsf{IV}(e) = \emptyset$} \\
@@
 where $\mathsf{IV}(e)$ is the set of free induction variables
-in an expression, and $\neigh{}$ classifies expressions which
-are constant offsets from an induction variable
+in an expression, and $\neigh_I$ classifies expressions which
+are constant offsets from an induction variable $i \in I$
 (Defn.~\ref{def:neighbour}). In the third case,
@@
 of an induction variable) then it is mapped to $\infty$, otherwise
-$\downarrow\!e$ is undefined (returns $\bot$). The $\mathsf{schematic}$ function
-is undefined if $\downarrow$ of any of its components $e_i$ is
-undefined.
-
-For example, assuming $i$ and $j$ are induction variables and $\textsf{schematic}(i-1, 2, j)
- = \vthreeh{-1}{\infty}{0}$. However, $\textsf{schematic}(i*2, j)$, for
- example, is
- undefined as its first index is a non-neighbour relative index, which
- is outside the scope of our specification language.
+$\mathsf{s}_I(e)$ is undefined (returns $\bot$). The $\mathsf{schematic}$ function
+is undefined if $\mathsf{s}_I$ of any of its components $e_i$ is
+undefined. We omit the parameter set of induction variables $I$ to
+$\mathsf{schematic}$ when clear from the context. 
 \end{definition}
@@
 $\vtwoh{\infty}{1} \; \preceq_2 \; \vtwoh{0}{1}$.
-%%
+
+As an example of undefined \textsf{schematic}, 
+ $\textsf{schematic}(i*2, j) = \bot$ as its first index is a non-neighbour relative index, which
+ is outside the scope of our specification language.
 \end{example}
@@
 as a positional placeholder.
+
+From \textsf{schematic} we now give a
+formal characterisation of stencil computations for our purposes:
+%%
+\begin{definition}[Stencil computations]
+  Let $a$ be an array of dimensionality $n$ and $\overline{b}$ a
+  collection of arrays of arbitrary (possible differing)
+  dimensionalities.  
+  A \emph{stencil computation} comprises an iteration by a set of at
+  most $n$ induction variables $I$ 
+  over a subset of the index space of $a$. Elements of $a$ are determined by an
+  assignment $a(\bar{e}) = e_r$ where $\bar{e}$ is an array index such
+  that  $\mathsf{schematic}_I(\bar{e}) \neq \bot$ and for all array
+  subscripts $b(\bar{e}') \in e_r$ then
+  $\mathsf{schematic}_I(\bar{e}') \neq \bot$. 
+\label{def:stencil}
+\end{definition}
+%%
 
@@
 induction variables on the RHS are not a subset of those in the LHS. These
-cases are not stencils by our definition. The degenerate case of this
+cases are not stencils by our definition (Defn.~\ref{def:stencil},p.~\pageref{def:stencil}). The degenerate case of this
 is to have only constant indices on the LHS. We see lots of examples

### 2016-07-07 paper.tex
dispersing the 'context' section

@@
 
-\usepackage[draft]{hyperref}
+\usepackage{hyperref}
 \hypersetup{
@@
 
-\item we give a quantiative study of
-  stencil computations in a corpus of numerical Fortran code,
-  totalling one million lines of code. Our tool identifies and
+\item we perform a quantitative study of
+  stencil computations in a corpus of numerical Fortran programs
+  (ranging from small to large),  totalling one million lines of code. Our tool identifies and
   synthesises specifications for 35,000 stencil computations in the
@@
 
-\item we give a more detailed verification case study of two
-  particular stencil computation and show that we
-  are able to detect the vast majority of (synthesised) possible
+\item we give a detailed verification case study of two
+  particular stencil computation, simulating programming errors.
+  Our approach is able to detect the vast majority of possible
   indexing errors.
@@
 
-\paragraph{Context}
-
-In recent years, there has been an increased awareness
-of the need for formal verification techniques in
-science~\cite{post2005computational,oberkampf2010verification,orchard2014computational}.
-Our specification language is an approach in this direction, providing
-a system of lightweight specification for numerical code.
-The aim is to not only provide a way of exposing errors, but also
-for specifying program behaviour to future-proof the code and ease
-its maintenance. In order to increase the likelihood of impact, we aim
-to integrate with existing practices smoothly, hence our
-specifications are integrated as comments and our tool can synthesis
-specifications for the user.
-
-This paper considers the technical aspects of the language,
-its semantics, inference, checking, and synthesis procedures.
-We also evaluate our implementation on real scientific code 
-used actively in research, ranging from small ($\sim$700loc) to large
-($\sim$600kloc). 
+%This paper considers the technical aspects of the language,
+%its semantics, inference, checking, and synthesis procedures.
+%We also evaluate our implementation on real scientific code 
+%used actively in research, ranging from small ($\sim$700loc) to large
+%($\sim$600kloc). 
  %Our case studies include the Unified Weather 
@@
 \begin{figure}[t]
-\vspace{-0.1em}
+\vspace{-0.5em}
 \begin{align*}
@@
 %
-(\texttt{+}\textsc{idem}) \;\; & S \; \texttt{+} \; S \; \equiv \; S \\[0.4em]
-%
-(\texttt{+}\textsc{comm}) \;\; & S \; \texttt{+} \; T \; \equiv \; T \;
+(\texttt{+}\textsc{idem}) \;\; & S \; \texttt{+} \; S \; \equiv \; S \quad
+%
+(\texttt{+}\textsc{comm}) \;\; S \; \texttt{+} \; T \; \equiv \; T \;
                        \texttt{+} \; S \\[0.4em]
@@
 Figure~\ref{fig:equations} lists the equational theory for
-specifications which is defined by the binary relation $\equiv$
-between region terms. The equations use abbreviations
+specifications given by the binary relation $\equiv$
+on \emph{region} terms. The equations use abbreviations
 \term{refl}, \term{fwd}, \term{bwd}, \term{cen}, and \irreflS{}
@@
 attribute on region constants as a boolean. For example,
-the syntactic sugar for \term{fwd} is defined:
+\term{fwd} syntactic sugar is defined:
 %%
@@
 & \qquad \begin{cases}
-\stenFwd{n}{d}{, \irrefl} & \;\;\textit{if} \;\, \neg \, r \\
+\stenFwd{n}{d}{, \irrefl} & \;\;\textit{if} \;\, \neg \, r \\[-0.1em]
 \stenFwd{n}{d}{}  & \;\;\textit{if} \;\,  r
@@
 \end{align*}
-
+%
 The equations reveal a part of the semantics for our
@@
 \wedge \; \stenFwdSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s} \\[1.2em]
-\end{array}}\trule{\textsc{over}} \\
+\end{array}}\trule{\textsc{over}} \\[-0.2em]
 \hspace{-0.5em}
@@
 \textsf{both}(S_l,S_u) \odot
-\textsf{both}(T_l,T_u) & = \textsf{both} (S_l \hat{\odot} T_l, \; S_u
+\textsf{both}(T_l,T_u)  = \; & \textsf{both} (S_l \hat{\odot} T_l, \; S_u
                          \hat{\odot} T_u) \\
@@
 where in the last case $\textit{inj}$ corresponds to the unary
-injections of \textsf{Approx}. For brevity we omit the cases where the parameters above are flipped
+injections of \textsf{Approx}. For brevity we omit the cases where the arguments above are flipped
 as $\odot$ is commutative.
@@
   %\subcaptionbox{}{\includegraphics[width=0.1\textwidth]{jacobi-incorrect3.png}}
-  \subfigure[]{\label{fig:jacobicorrect}\includegraphics[width=0.1\textwidth]{jacobi-correct.png}}
-  \subfigure[]{\includegraphics[width=0.1\textwidth]{jacobi-incorrect1.png}}
-  \subfigure[]{\includegraphics[width=0.1\textwidth]{jacobi-incorrect2.png}}
+  \subfigure[]{\label{fig:jacobicorrect}\includegraphics[width=0.1\textwidth]{jacobi-correct.png}}\;\;
+  \subfigure[]{\includegraphics[width=0.1\textwidth]{jacobi-incorrect1.png}}\;\;
+  \subfigure[]{\includegraphics[width=0.1\textwidth]{jacobi-incorrect2.png}}\;\;
   \subfigure[]{\includegraphics[width=0.1\textwidth]{jacobi-incorrect3.png}}
@@
 
-We presented in this work an specification language 
-for the access pattern of a stencil computation. Compared to the
-related work (discussed above), our language is novel in that it does
+In recent years, there has been an increased awareness
+of the need for formal verification techniques in
+science~\cite{post2005computational,oberkampf2010verification,orchard2014computational}.
+Our specification language is an approach in this direction, providing
+a system of lightweight specification for numerical code.
+Compared to other specification approaches (discussed above), our language is novel in that it does
 not aim to replicate indexing in the specification, but instead
@@
 
+Our aim is to not only provide a way of exposing errors, but also
+for specifying program behaviour to future-proof the code and ease
+its maintenance. To increase the likelihood of impact, we 
+integrate smoothly with existing practices, providing 
+specifications as comments and specification synthesis
+specifications in the the tool.
+
 We evaluated our approach by implementing the stencil inference and
-checking procedure ontop of CamFort, an open-source analysis
+checking procedure on top of CamFort, an open-source analysis
 tool\footnotemark for Fortran

### 2016-07-07 paper.tex
shorten caption
andy edits

@@
   Specifications}
-\authorinfo{Blind}{}{}
+\authorinfo{}{}{}
 %\authorinfo{Dominic Orchard \and Mistral Contrastin
@@
 understood. More complex stencil computations
-are much more prone to error from simple lexical mistakes and
-harder to understand. For example,
+are much more prone to error from simple lexical mistakes. For example,
 Figure~\ref{ref:navier-stokes-fragment} shows three lines from a
-Navier-Stokes fluid simulator in which two arrays are read from with
+Navier-Stokes fluid simulator in which two arrays are read with
 different data access patterns, across two dimensions. The interaction
 is much harder to understand, with the potential for the developer to
-accidentally introduce an error via simple textual mistakes, for
+introduce an error via simple textual mistakes, for
 example writing $\texttt{(i-1,j)}$ instead of $\texttt{(i+1,j)}$.
@@
 %\texttt{a} and \texttt{b}.
-In the Navier-Stokes example of
-Figure~\ref{ref:navier-stokes-fragment}, its inferred specification is shown
+The inferred specification for the Navier-Stokes example is shown 
 in Fig.~\ref{ref:navier-stokes-fragment}(b). The
-specification explains that, over the whole fragment, \texttt{u} is
-read from with a centered pattern to depth of 1 in both dimensions
+specification requires that, over the whole fragment, \texttt{u} is
+accessed with a centered pattern to depth of 1 in both dimensions
 (this is known as the \emph{five-point stencil}) and \texttt{v}
-is read from in a neighbourhood bounded by forward to depth of $1$ in
+is accessed in a neighbourhood bounded by forward to depth of $1$ in
 the first dimension and backward to a depth of $1$ in the second
@@
 \item we detail inference, checking, and specification synthesis 
-algorithms for our language (Sect.~\ref{sec:analysis}), derived from
-the a denotational model for the language (Sect.~\ref{sec:semantics});
+algorithms for our language (Section~\ref{sec:analysis}), derived from
+a denotational model for the language (Section~\ref{sec:semantics});
 
@@
 
-\item we perform a quantitative study of
-  stencil computations in a corpus of numerical Fortran programs
+\item we report on a quantitative study of
+  stencil computations on a corpus of numerical Fortran programs
   (ranging from small to large),  totalling one million lines of code. Our tool identifies and
@@
 
-Our specification language is based on the observation
+Our specification language is based on the hypothesis
 that most forms of array access in numerical code have
 a fixed, statically-determined access pattern. For example, the
-``\emph{five-point stencil}'' on a two-dimensional array reads from array
+\emph{five-point stencil} on a two-dimensional array reads from array
 indices $(i, j)$, $(i-1, j)$, $(i+1, j)$, $(i, j-1)$, and $(i, j+1)$
@@
 For the target language, $e$ ranges over expressions (may be impure) and $v$
-overs its (imperative) variables. %By
+over its (imperative) variables. %By
 %``variable'', we mean in the imperative sense (named binders to memory cells).
@@
 \begin{definition}[Base induction variable]
-  A variable \textit{v} of integer type is a \emph{base induction
+  A integer variable \textit{v} is a \emph{base induction
     variable} if it is the control variable of a ``for''
@@
 \begin{definition}[Array subscript and index]
-  An \emph{array subscript} is an expression that indexes an array
-  (for the purpose of reading or writing an element), which we denote
+  An \emph{array subscript} is an expression that indexes an array, which we denote
   by $v(\bar{e})$ for the source language where $\bar{e}$ is shorthand
@@
   by the predicate $\neigh_I$ where $I$ is a set of induction
-  variables. We omit $I$ when implicit from the context.
+  variables. We omit $I$ when clear from the context.
 \label{def:neighbour}
@@
 \emph{Regions} are the central building blocks of spatial
-specifications. Regions can either be declared along with a
+specifications. Regions can either be declared alone with a
 \nonterm{regionDec}, assigning a region specification \nonterm{region} to
@@
 %%
-were \texttt{i} is an induction variable.  The specification
+where \texttt{i} is an induction variable.  The specification
 is associated to the array variable \texttt{a}. Note that the second
@@
 %%
-This pattern is a common image convolution (for example
+This pattern is common in image convolution applicaitons (for example
 for edge detection). The specification ranges over the
-values that flow the array subscript on the left-hand side,
+values that flow to the array subscript on the left-hand side,
 and so ranges over the intermediate assignments to \term{x},
@@
 The region constructor \texttt{*} can also be interpreted as a
-kind of conjunction on specifications since they constraint
+kind of conjunction on specifications since they constrain
 indices to be within both regions simultaneously.
@@
 %
-(\texttt{+}\textsc{idem}) \;\; & S \; \texttt{+} \; S \; \equiv \; S \quad
-%
-(\texttt{+}\textsc{comm}) \;\; S \; \texttt{+} \; T \; \equiv \; T \;
+(\texttt{+}\textsc{idem}) \;\; & S \; \texttt{+} \; S \; \equiv \; S \\[0.4em]
+%
+(\texttt{+}\textsc{comm}) \;\; & S \; \texttt{+} \; T \; \equiv \; T \;
                        \texttt{+} \; S \\[0.4em]
@@
 \caption{Equations on specifications (\emph{regions})}
-\vspace{-1em}
+\vspace{-0.5em}
 \label{fig:equations}
@@
 \setlength{\arraycolsep}{0.1em}
-\dfrac{\hspace{3em} m \leq n \qquad\qquad r \Rightarrow s \hspace{3em}}
+\dfrac{\hspace{3em} m \leq n \qquad\qquad r \implies s \hspace{3em}}
 {\begin{array}{rl}
@@
 $(i, j+1)$, $(1, j+1)$ and $(3, j)$ 
-are given by \textsf{schematic} the 
+are mapped by \textsf{schematic} to the 
 index schemes $\vtwoh{0}{1}$, $\vtwoh{\infty}{1}$
 and $\vtwoh{\infty}{0}$ respectively. The following
-relationship of generalises then holds:
+relationships of generalisation then holds:
 %\[
@@
 polymorphic types in ML~\cite{milner1978theory}. Our notion is
-much more rigid however, as our schemes do not bind names
-and thus generalisation does not require substitution; $\infty$ acts
-as a positional placeholder.
+much more rigid however, as our schemes do not bind names.
+%and thus generalisation does not require substitution; $\infty$ acts
+%as a positional placeholder.
 
@@
   dimensionalities.  
-  A \emph{stencil computation} comprises an iteration by a set of at
-  most $n$ induction variables $I$ 
+  A \emph{stencil computation} comprises an iteration by a set $I$ of at
+  most $n$ induction variables 
   over a subset of the index space of $a$. Elements of $a$ are determined by an
   assignment $a(\bar{e}) = e_r$ where $\bar{e}$ is an array index such
-  that  $\mathsf{schematic}_I(\bar{e}) \neq \bot$ and for all array
-  subscripts $b(\bar{e}') \in e_r$ then
+  that  $\mathsf{schematic}_I(\bar{e}) \neq \bot$ and for all $b \in
+  \bar{b}$ and all subscripts $b(\bar{e}') \in e_r . \, $
   $\mathsf{schematic}_I(\bar{e}') \neq \bot$. 
@@
 \noindent
-The last two equations of Fig.~\ref{fig:region-model} give the
+The final two equations of Fig.~\ref{fig:region-model} give the
 models of $\term{+}$ and $\term{*}$. For $\term{+}$ this is the 
@@
 We briefly sketch the algorithms for inferring and synthesising
-specifications (Sect.~\ref{subsec:inference}), for checking code
-against specifications (Sect.~\ref{subsec:checking}). These utilise the semantic model of
+specifications (Section~\ref{subsec:inference}), for checking code
+against specifications (Section~\ref{subsec:checking}). These utilise the semantic model of
 Section~\ref{sec:semantics}.
@@
 %%
-\subsubsection{Step 1: Standard control and data-flow analyses}
+\subsubsection{Step 1: Analyse standard control and data-flow}
 \label{sec:inf-step1}
@@
 
-\subsubsection{Step 2: Data-access analysis}
+\subsubsection{Step 2: Analyse data-access}
 \label{sec:inf-step2}
@@
   and $\vect{x}^U$ give the lower and upper bounds.  In
-  the rest, $\vect{x}, \vect{y}$ range over spans.  For a span
+  the following, $\vect{x}, \vect{y}$ range over spans.  For a span
   $\vect{x}$, we write $\vect{x}_{i:j}$ for the sub-span comprises the
@@
 fully contains another), Figure~\ref{fig:inference-steps-informal}(c).
-The final spans are used to synthesis concrete stencil specification
+The final spans are used to synthesise concrete stencil specification
 syntax (Section~\ref{sec:inf-step4}), where
@@
 
-A fixed point of the following function, \spanOp{}, is taken (per variable
-in the map U) to coalesces spans into contiguous regions, \ie{}, we compute
- $(\mu \, \spanOp) \; N(M)$ where \spanOp{} is defined by: \\[-0.5em]
+Each set of unit spans is applied to the fixed point of the \spanOp{} function
+(defined shortly), which coalesces spans into contiguous regions. That is, we compute
+ $(\mu \, \spanOp)$ on input $N(M)$, where \spanOp{} is defined by the steps:
 
@@
 \noindent
-\textit{1).\;} Compute all permutations on the column vectors in a span, \ie{},
+\textbf{1)\,} Compute all permutations on the column vectors in a span, \ie{},
   $[\vect{x}^L \vect{x}^U] \mapsto [\pi\vect{x}^L \, \pi\vect{x}^U]$
@@
 \noindent
-\textbf{2).} Sort each permutation set into an ordered list, by the ordering:
+\textbf{2)\,} Sort each permutation set into an ordered list, by the ordering:
 \begin{equation*}
@@
 \noindent
-\textbf{3).} Reduce each list pairwise by $\bullet$ operation 
+\textbf{3)\,} Reduce each list pairwise by $\bullet$ operation 
 (Defn.~\ref{def:span-coalesc}) to coalesce contiguous regions,
@@
 %
-\textbf{4).} Un-permute and union together, \ie{},
+\textbf{4)\,} Un-permute and union together, \ie{},
 %
@@
 %
-\textbf{5).} Filter by the region containment predicate $\containedin$, that
+\textbf{5)\,} Filter by the region containment predicate $\containedin$, that
   is, if any region is contained within another then remove the
@@
 
+\newcommand{\oplusbig}{\operatornamewithlimits{\term{\Large{$\oplus$}}}}
+\newcommand{\odobig}{\operatornamewithlimits{\term{\Large{$\odot$}}}}
 \newcommand{\bplus}{\operatornamewithlimits{\term{\Large{+}}}}
@@
 %%
-The $\mathsf{toSpec}$ functions maps a pair of lower and upper bounds
+The $\mathsf{toSpec}$ function maps a pair of lower and upper bound
 values (one-dimensional) to a specification with a one-dimensional region. 
@@
 \textsf{spec}(M) = 
-\sum\limits_{\vect{x} \in \spanOp(M)}
-\prod\limits_{i \in \{1, \ldots, n\}} \, \mathsf{toSpec}(\vect{x}^L_i, \vect{x}^U_i)
+\oplusbig\limits_{\vect{x} \in \spanOp(M)} \,
+\odobig\limits_{i \in \{1, \ldots, n\}} \, \mathsf{toSpec}(\vect{x}^L_i, \vect{x}^U_i)
 \end{equation*}
-(where $\oplus$ is the additive and $\odot$ is the multiplicative used
-within $\Sigma$ and $\Pi$ respectively here).
 We unpack the definitions of this algebra.
@@
 \input{consistency.tex}
-\caption{Consistency between a model and indexing schemes from a stencil computation}
+\caption{Consistency between a model and indexing schemes}
 \label{fig:consistency}
@@
 
+\subsection{Approximations and checking}
+
 As noted, the {\textcolor{darkgreen}{\term{fivepoint}}} region
@@
 $$
-\forall M_c \in \textsf{Mult}(\domainVal{}^n) . \mathit{cons}_n(\interp{S}_n,M_c) \implies
+\forall M_c \in \textsf{Mult}(\domainVal{}^n) . \;\, \mathit{cons}_n(\interp{S}_n,M_c) \implies
 \mathit{cons}(\interp{S'}_n,M_c)
@@
 \subsection{Software Corpus}
+
+Table~\ref{tab:corpus} shows summary statistics of the software
+packages used in our evaluation, all of which are written in Fortran
+90 or Fortran 77. In total we analysed \SI{\overallLoC} lines of code
+from 7 packages, of which we successfully parsed
+\SI{\overalllinesParsed} lines. The ``Number of files'' column shows
+how many files in each corpus that we were able to analyse with
+CamFort. The most common reasons for CamFort rejecting a file were
+either use of a C-preprocessor, or illegal use of language features
+from a modern Fortran variant.
+
+\textbf{The Unified
+  Model}~\cite{um}
+is a weather forecasting and climate modelling tool developed by the
+Met Office in the United Kingdom. It is used by research organisations
+and meteorological services around the world. We use the development
+branch (trunk) of the model. The code base in closed source but
+institutional licenses are available for research purposes.
+%The Met Office runs a comprehensive code quality system incorporating dedicated committers (we counted 11) for particular parts of the model.
+We counted 120 additional contributors whose submissions are reviewed and tested before being accepted into the code base.
+
+% e3mg-ea
+\textbf{E3MG} (An Energy-Environment-Economy (E3) Model at the Global Level) is a macroeconomic model used for assessment of environmental policy~\cite{RePEc:aen:journl:2006se-a12}.
+This was developed by Cambridge Econometrics, an independent consultancy company.
+
+\textbf{BLAS}~\cite{blas} (Basic Linear Algebra Subprograms) is a popular
+library providing efficient and portable routines for vector and
+matrix operations. These routines feature in many other libraries
+(including LAPACK). We used version 3.6.0. We chose to include this
+package for breadth, as it provides general numerical functions rather
+than a specialised scientific model.
+
+\textbf{Hybrid4} is a vegetation and biomass model for simulating carbon, water and nitrogen flows~\cite{GBC:GBC635}.
+
+\textbf{GEOS-Chem}~\cite{geos-chem} is a three-dimensional model of
+tropospheric chemistry developed at Harvard and used by $\sim$70
+universities and research institutions world-wide. We use v.10-01.
+
+\textbf{Navier} is a small numerical simulation, giving a discrete
+approximation to the two-dimensional Navier-Stokes fluid equations,
+based on the book of~\citet{griebel1997numerical}.
+
+\textbf{CP} consists of the example code from the second edition of the book ``Computational Physics''~\cite{nicholas2006computational} introducing numerical techniques and their application to modern physics problems such as fields, waves, statistical mechanics and quantum mechanics.
+
+% atLeast: number of stencils with an atLeast modifier
+% atMost: number of stencils with an atMost modifier
+% boundedBoth: number of stencils with both atLeast and atMost modifier
+%% atLeast == atMost == boundedBoth - whenever we infer one we infer the other but users might write just one
+
+% backward: number of backward stencils
+% forward: number of forward stencils
+% centereed: number of centered stencils
+
+% justReflexive: number of only reflexive stencils
+% irreflexive: number of irreflexive stencils
+% readOnce: number of readOnce stencils
+
+% depthX: number of stencils with depth X
+% dimsX: number of stencils with dimensions=X (check this!)
+
+% lexFailed: number of files we failed to lex
+% parseFailed: number of files we failed to parse
+% lexOrParseFailed: number files we failed to lex and parse
+% parseOk: number of files parsed successfully
+%% lexFailed + parseFailed = lexOrParseFailed
+%% lexOrParseFailed + parseOk = number of files in source repo
+
+% tickAssign: number of array subscripts with a LHS which are all neighbour indices or constants
+% tickAssignSuccess: number of tickAssign's we generated some specification for
+%% tickAssign < tickAssignSuccess
+
+\subsection{Occurrences of stencils}
+
+We first examined how frequently stencil computations occur in our
+corpus. This was carried out by running our implementation in 
+whole-code-base inference mode. We define a \emph{potential} statement as one whose left-hand
+side is an array subscript on neighbourhood indices. These statements
+are the inputs to Step 2 of the Inference process described in 
+Section~\ref{subsec:inference}. Each potential statement for which we
+eventually generate some specification is also counted as
+an \emph{actual} statement.
+
 \begin{table}
@@
   evaluation\label{tab:corpus}}
-\vspace{-0.7em}
+\vspace{-1em}
 \end{table}
-
-Table~\ref{tab:corpus} shows summary statistics of the software
-packages used in our evaluation, all of which are written in Fortran
-90 or Fortran 77. In total we analysed \SI{\overallLoC} lines of code
-from 7 packages, of which we successfully parsed
-\SI{\overalllinesParsed} lines. The ``Number of files'' column shows
-how many files in each corpus that we were able to analyse with
-CamFort. The most common reasons for CamFort rejecting a file were
-either use of a C-preprocessor, or illegal use of language features
-from a modern Fortran variant.
-
-\textbf{The Unified
-  Model}~\cite{um}
-is a weather forecasting and climate modelling tool developed by the
-Met Office in the United Kingdom. It is used by research organisations
-and meteorological services around the world. We use the development
-branch (trunk) of the model. The code base in closed source but
-institutional licenses are available for research purposes.
-%The Met Office runs a comprehensive code quality system incorporating dedicated committers (we counted 11) for particular parts of the model.
-We counted 120 additional contributors whose submissions are reviewed and tested before being accepted into the code base.
-
-% e3mg-ea
-\textbf{E3MG} (An Energy-Environment-Economy (E3) Model at the Global Level) is a macroeconomic model used for assessment of environmental policy~\cite{RePEc:aen:journl:2006se-a12}.
-This was developed by Cambridge Econometrics, an independent consultancy company.
-
-\textbf{BLAS}~\cite{blas} (Basic Linear Algebra Subprograms) is a popular
-library providing efficient and portable routines for vector and
-matrix operations. These routines feature in many other libraries
-(including LAPACK). We used version 3.6.0. We chose to include this
-package for breadth, as it provides general numerical functions rather
-than a specialised scientific model.
-
-\textbf{Hybrid4} is a vegetation and biomass model for simulating carbon, water and nitrogen flows~\cite{GBC:GBC635}.
-
-\textbf{GEOS-Chem}~\cite{geos-chem} is a three-dimensional model of
-tropospheric chemistry developed at Harvard and used by $\sim$70
-universities and research institutions world-wide. We use v.10-01.
-
-\textbf{Navier} is a small numerical simulation, giving a discrete
-approximation to the two-dimensional Navier-Stokes fluid equations,
-based on the book of~\citet{griebel1997numerical}.
-
-\textbf{CP} consists of the example code from the second edition of the book ``Computational Physics''~\cite{nicholas2006computational} introducing numerical techniques and their application to modern physics problems such as fields, waves, statistical mechanics and quantum mechanics.
-
-% atLeast: number of stencils with an atLeast modifier
-% atMost: number of stencils with an atMost modifier
-% boundedBoth: number of stencils with both atLeast and atMost modifier
-%% atLeast == atMost == boundedBoth - whenever we infer one we infer the other but users might write just one
-
-% backward: number of backward stencils
-% forward: number of forward stencils
-% centereed: number of centered stencils
-
-% justReflexive: number of only reflexive stencils
-% irreflexive: number of irreflexive stencils
-% readOnce: number of readOnce stencils
-
-% depthX: number of stencils with depth X
-% dimsX: number of stencils with dimensions=X (check this!)
-
-% lexFailed: number of files we failed to lex
-% parseFailed: number of files we failed to parse
-% lexOrParseFailed: number files we failed to lex and parse
-% parseOk: number of files parsed successfully
-%% lexFailed + parseFailed = lexOrParseFailed
-%% lexOrParseFailed + parseOk = number of files in source repo
-
-% tickAssign: number of array subscripts with a LHS which are all neighbour indices or constants
-% tickAssignSuccess: number of tickAssign's we generated some specification for
-%% tickAssign < tickAssignSuccess
-
-\subsection{Occurrences of stencils}
-
-We first examined how frequently stencil computations occur in our
-corpus. This was carried out by running our implementation in 
-whole-code-base inference mode. We define a \emph{potential} statement as one whose left-hand
-side is an array subscript on neighbourhood indices. These statements
-are the inputs to Step 2 of the Inference process described in 
-Section~\ref{subsec:inference}. Each potential statement for which we
-eventually generate some specification is also counted as
-an \emph{actual} statement.
-
 \begin{table}
+\vspace{-0.5em}
 \sisetup{table-format=4,table-text-alignment=right}
@@
 
-In recent years, there has been an increased awareness
-of the need for formal verification techniques in
+There is an increasing awareness
+of the need for verification techniques in
 science~\cite{post2005computational,oberkampf2010verification,orchard2014computational}.
@@
 Compared to other specification approaches (discussed above), our language is novel in that it does
-not aim to replicate indexing in the specification, but instead
-provides abstract spatial combinators which capture a large number
-of common patterns.
-
-Our aim is to not only provide a way of exposing errors, but also
-for specifying program behaviour to future-proof the code and ease
-its maintenance. To increase the likelihood of impact, we 
-integrate smoothly with existing practices, providing 
-specifications as comments and specification synthesis
-specifications in the the tool.
+not aim to reify indexing in the specification, but instead
+provides abstract spatial descriptions which capture a large number
+of common patterns. Our tool is useful for both detecting
+errors and for specifying program behaviour to future-proof code and ease
+ maintenance.
+
+%To increase the likelihood of impact, we 
+%integrate smoothly with existing practices, providing 
+%specifications as comments and specification synthesis
+%specifications in the the tool.
 
@@
 checking procedure on top of CamFort, an open-source analysis
-tool\footnotemark for Fortran
+tool for Fortran
 programs and running the inference process over a corpus of 1.1
@@
 
-\footnotetext{\url{https://github.com/camfort/camfort}}
+%\footnotetext{\url{https://github.com/camfort/camfort}}
 \bibliography{references}

### 2016-07-07 paper.tex
last edits

@@
   assignment $a(\bar{e}) = e_r$ where $\bar{e}$ is an array index such
-  that  $\mathsf{schematic}_I(\bar{e}) \neq \bot$ and for all $b \in
-  \bar{b}$ and all subscripts $b(\bar{e}') \in e_r . \, $
+  that  $\mathsf{schematic}_I(\bar{e}) \neq \bot$ and %for all $b \in  \bar{b}$ and
+  for all subscripts $b(\bar{e}') \in e_r . \, $
   $\mathsf{schematic}_I(\bar{e}') \neq \bot$. 
@@
 model, modelling a source-level index, $\forall u \in M_c$ there exists
-a generalising scheme in the spec model $\exists v \in M$, \ie{}, $u
+a generalising scheme in the spec model $\exists v \in M_s$, \ie{}, $u
 \preceq_n v$. That is, every source-level index is generalised by the model.
@@
 stencil specification inferred previously in
-Section~\ref{subsec:inference} (see top-left):
+Section~\ref{subsec:inference}:
 %
@@
 
-\subsection{Software Corpus}
+%\subsection{Software Corpus}
+\paragraph{Software Corpus}
 
@@
 \begin{table}
-\vspace{-0.5em}
+\vspace{-1em}
 \sisetup{table-format=4,table-text-alignment=right}
@@
 
-\begin{figure}[t]
-\begin{minted}[breakindent=0em,linenos=false,xleftmargin=0em,breakafter=)]{fortran}
+\begin{figure}[t]\begin{minted}[fontsize=\scriptsize,breakindent=0em,linenos=false,xleftmargin=0em,breakafter=)]{fortran}
 !=stencil readOnce,(forward(depth=1,dim=3,irreflexive))*(backward(depth=1,dim=1))*(backward(depth=1,dim=2))+(forward(depth=1,dim=3))*(backward(depth=1,dim=1))*(reflexive(dim=2))+(forward(depth=1,dim=3))*(backward(depth=1,dim=1,irreflexive))*(backward(depth=1,dim=2,irreflexive))+(forward(depth=1,dim=3))*(backward(depth=1,dim=2))*(reflexive(dim=1))+(backward(depth=1,dim=1))*(backward(depth=1,dim=2,irreflexive))*(reflexive(dim=3))+(backward(depth=1,dim=1,irreflexive))*(backward(depth=1,dim=2))*(reflexive(dim=3))::x
@@
 % 6 cases were not caught
-\section{Related work}
+\section{Related work and conclusions}
 \label{sec:discussion}
@@
     do i=imin+1,imax
-        q = b(i, j);  a(i, j) = q + t
-        t = q
+        q = b(i, j);  a(i, j) = q + t; t = q
     end do
@@
 able to ascribe a specification in the above case as the use of
-\texttt{t} on line 5 maps only to the definition on line 2, and not to
-that on line 6. Extending our approach to follow loop-carried dependencies is
+\texttt{t} on line 4 maps only to the definition on line 2, and not to
+that on line 4. Extending our approach to follow loop-carried dependencies is
 future work. An alternate approach for us would be to build on top of
@@
 
-\section{Conclusion and future work}
+\paragraph{Conclusion and future work}
 
@@
 provides abstract spatial descriptions which capture a large number
-of common patterns. Our tool is useful for both detecting
-errors and for specifying program behaviour to future-proof code and ease
- maintenance.
+of common patterns. %Our tool is useful for both detecting
+%errors and for specifying program behaviour to future-proof code and ease
+% maintenance.
 
@@
 
-We evaluated our approach by implementing the stencil inference and
-checking procedure on top of CamFort, an open-source analysis
+We evaluated our approach on top of CamFort, an open-source analysis
 tool for Fortran
-programs and running the inference process over a corpus of 1.1
-million lines of Fortran code.  Our motivation for this work was to
+programs, running the inference process over a corpus of 1.1
+million lines of Fortran code. 
+ Our motivation for this work was to
 provide an intermediate specification language to allow scientists to
 verify numerical computations.  Our evaluation showed that relatively
-simple stencil computations are the norm, for which we can specify a
-large proportion, comprising mainly reflexive region constants which
+simple stencil computations are the norm comprising mainly reflexive region constants which
 are less likely to give rise to programming errors.  However, around

### 2016-07-14 paper.tex
submitted version of the paper
Modify submitted version to add DO NOT DISTRIBUTE in title banner and footer
added missing last line to navier example
update region declaration syntax in the paper

@@
   Specifications}
-\authorinfo{}{}{}
-%\authorinfo{Dominic Orchard \and Mistral Contrastin
-%\and Matthew Danish \and Andrew Rice}{University of Cambridge}{}
+\titlebanner{DRAFT -- Do not distribute}
+%\authorinfo{}{}{}
+\authorinfo{Dominic Orchard \and Mistral Contrastin
+\and Matthew Danish \and Andrew Rice}{University of Cambridge}{}
+
+\preprintfooter{DO NOT DISTRIBUTE}
 
@@
   (u(i,j+1)-2.0*u(i,j)+u(i,j-1))/dely/dely
+
+f(i,j) = u(i,j)+del_t*(laplu/Re-du2dx-duvdy)
 \end{minted}
@@
 \nonterm{regionDec}, assigning a region specification \nonterm{region} to
-a region variable \nonterm{rvar} or given directly within a
+a region variable \nonterm{rvar} (via $\texttt{::}$) or given directly within a
 \nonterm{spatial} specification.
@@
                         \texttt{::} \; v \\
-\nonterm{regionDec} ::= &  \texttt{region} \; \nonterm{rvar} \; \texttt{=} \;
+\nonterm{regionDec} ::= &  \texttt{region} \; \texttt{::} \; \nonterm{rvar} \; \texttt{=} \;
                          \nonterm{region}\\[0.4em]
@@
 \begin{minted}{fortran}
-!= region r1 = forward(depth=1, dim=1)
-!= region r2 = forward(depth=1, dim=2)
-!= region robertsCross = r1*r2
+!= region :: r1 = forward(depth=1, dim=1)
+!= region :: r2 = forward(depth=1, dim=2)
+!= region :: robertsCross = r1*r2
 != stencil robertsCross :: a
@@
 \begin{minted}[breakindent=8.1em]{fortran}
-!= region fivepoint = centered(depth=1,dim=1) + centered(depth=1,dim=2)
+!= region :: fivepoint = centered(depth=1,dim=1) + centered(depth=1,dim=2)
 != stencil readOnce, fivepoint :: b

### 2016-08-23 paper.tex
Extend consistency definition to make readOnce optional in the spec
Add small explanation of eliminating all infinity vectors

@@
 models in which non-unique indexing is allowed, labelled with
-$\mathsf{mult}$.
+$\mathsf{mult}$. Similarly, the second rule ensures that absence of
+\texttt{readOnce} modifier in the specification does not prohibit a set of
+unique stencil accesses from being consistent.
 
@@
 \end{example}
+
+One way implementation may deviate from the model is elimination of all $\infty$
+offset vector that occurs when all indices are in non-neighbourhood form \eg{}
+\mintinline{fortran}{b(42,0)}. The only generalisation scheme that can unify
+with this pattern is the most general one, meaning it requires outright
+rejection of an otherwise fine stencil computation. See the following example,
+where not discarding all $\infty$ access pattern prevents verification:
+%
+\begin{minted}{fortran}
+!= stencil forward(dim=1,depth=1) :: a
+a(i) = a(i) + a(i+1) + a(1)
+\end{minted}
 

### 2016-10-03 new-intro.tex
attempt at a new, more clear intro

\documentclass{article}

\usepackage{enumitem}
\usepackage{minted}
\newcommand{\ie}{\emph{i.e.}}
\newcommand{\eg}{\emph{e.g.}}
\usemintedstyle{vs}
\fvset{
  linenos=true,
  fontsize=\small,
  breaklines=true,
  breakafter=),
  xleftmargin=4em
}

\begin{document}

\emph{Stencils} are a ubiquitous programming pattern, common in
scientific and numerical computing applications. Informally, a stencil
computation computes an array whose value at each index $i$ is
calculated from a \emph{neighbourhood} of values around $i$ in some
input array(s), \eg{}, the Game of Life, convolutions in image
processing, approximations to differential equations. For example, the
following computes the one-dimensional discrete Laplace transform (an
approximation to a derivative) in Fortran:
%
\begin{minted}{fortran}
do i = 1, (n-1)
   b(i) = a(i-1) - 2*a(i) + a(i+1)
end do
\end{minted}
%
Values \texttt{b(i)} are calculated from a neighbourhood of elements
around \texttt{i} in the input array \texttt{a}.  In this example, the
access pattern is simple and easily understood. More complex
stencil computations are more prone to errors from simple lexical
mistakes. For example, the following is a three-line snippet from a
Navier-Stokes fluid simulator in which two arrays are read with
different data access patterns, across two dimensions.
%%
\begin{minted}[firstnumber=20,xleftmargin=4em]{fortran}
du2dx = ((u(i,j)+u(i+1,j))*(u(i,j)+u(i+1,j))+   &
  gamma*abs(u(i,j)+u(i+1,j))*(u(i,j)-u(i+1,j))- &
  (u(i-1,j)+u(i,j))*(u(i-1,j)+u(i,j))-          &
  gamma*abs(u(i-1,j)+u(i,j))*(u(i-1,j)-u(i,j))) &
  /(4.0*delx)

duvdy = ((v(i,j)+v(i+1,j))*(u(i,j)+u(i,j+1))+   &
  gamma*abs(v(i,j)+v(i+1,j))*(u(i,j)-u(i,j+1))- &
  (v(i,j-1)+v(i+1,j-1))*(u(i,j-1)+u(i,j))-      &
  gamma*abs(v(i,j-1)+v(i+1,j-1))*(u(i,j-1)-     &
  u(i,j))) / (4.0*dely)

laplu = (u(i+1,j)-2.0*u(i,j)+u(i-1,j))/delx/delx+ &
  (u(i,j+1)-2.0*u(i,j)+u(i,j-1))/dely/dely

f(i,j) = u(i,j)+del_t*(laplu/Re-du2dx-duvdy)
\end{minted}
%%
The access pattern is much harder to understand than the
one-dimensional Laplace.  The miasma of indexing expressions of the
form \texttt{var(i$\pm$a, j$\pm$b)} is not only hard to read, but is
prone to simple textual input mistakes, \eg{}, swapping \texttt{-} and
\texttt{+}, missing an indexing term, or transforming the wrong
variable \eg{} \texttt{(i+1,j)} instead of \texttt{(i,j+1)}.

In practice, we find that the usual development procedure for complex
stencil computations involves some ad hoc testing to ensure that no
transcription mistakes have been made. This may involve visual
inspection of the results, based on a rendered image. Typically such
testing is then discarded once the code is seen to be correct.

This is not the only information that is typically discarded. The
shape of the indexing pattern is often a result of choices made in the
numerical-analysis discretisation procedure used to convert a set of
continuous equations into a discrete approximation. Rarely are these
decisions captured in the source code, yet the shape of access is
usually uniform and is given a perspicuous and concise description in
numerical analysis literature \eg{}, ``\emph{centered in space, to a
  depth of 1}'' referring to indexing terms \texttt{a(i)},
\texttt{a(i-1)} and \texttt{a(i+1)}.

To support the development of correct stencil computations, we propose
a simple, abstract specification language for the data access pattern
of stencils. The shape of the stencil computation in the
Laplace example is specifed in our language as:
%
\begin{minted}[linenos=false]{fortran}
!=  stencil centered(depth=1, dim=1) :: a
\end{minted}
%
This describes that \texttt{a} is accessed with a symmetrical pattern
(``centered'') to a depth of one in each direction in its first
dimension.  The stencil computation shape in the Navier-Stokes example
is specified as:
%
\begin{minted}[linenos=false,xleftmargin=4em,breakindent=3em]{fortran}
!= stencil centered(depth=1,dim=1)*reflexive(dim=2) + centered(depth=1,dim=2)*reflexive(dim=1) :: u

!= stencil forward(depth=1,dim=1) * backward(depth=1,dim=2) :: v
\end{minted}
%
The specification requires that, over the whole fragment, \texttt{u}
is accessed with a centered pattern to depth of 1 in both dimensions
(this is known as the \emph{five-point stencil}) and \texttt{v} is
accessed in a neighbourhood bounded forwards to depth of $1$ in the
first dimension and backward to a depth of $1$ in the second
dimension.

We provide a verification tool for such specification which can check
the correctness of code against the specification, and can also infer
specifications, synthesising these as comments in the
source code. The main features and benefits of our approach are as follows:
%
\begin{itemize}
\item The main use case is that a programmers writes a specification
  first (perhaps informed by their numerical analysis), then writes
  the corresponding source code, and lastly has the tool check its
  conformance to the specifications. This specify-and-check approach
  removes the usually ad hoc testing effort, and provides the
  opportunity for easy re-verification in the event of any refactoring
  of software maintenance activities.
%
\item The specification provides a concise description which can be
  preserved in the source code to provide documentation to future
  developers, contrasting with ad hoc tests which are usually
  discarded or specialised to the development environment. Separate from the
  verification procedure, the infer-and-synthesise mode of the tool
  provides a way to automatically generate additional useful
  documentation for future users.
%
\item Specifications are integrated into the code as comments for easy
integration with existing tools, as well as
  documentation systems (\eg{}, we support Doxygen syntax).
%
\item Our specification format is deliberately abstract, with a small
  number of combinators that \emph{do not involve any indexing
    expressions}, \eg{} does not involve expressions like
  \texttt{a(i+1,j-1)}. This contrasts with other stencil specification
  approaches. For example, the work of Kamil et al. provides
  full-functionality specifications of stencil computations from which
  equivalent DSL code can be generated.  If a user wishes to specify,
  say, the Naiver-Stokes stencil above, they must write a
  specification which is very similar, involving textually dense
  array-indexing expressions. Therefore any low-level mistakes that
  could be made programming complex indexing code could also be made
  when writing a specification. Our specifications are much more
  abstract, and also more lightweight with the aim of aiding adoption
  by scientists.
\end{itemize}

It should be noted that our verification tool does not deal in the
class of bugs that can be detected automatically (\emph{push-button
  verification}). Instead, stencil computation bugs must largely be
identified relative to a specification of the intended behaviour.

In this paper, we make the following contributions:
%
\begin{itemize}
\item we introduce our specification language for stencils
  that captures many common forms of data access pattern
  (Section~\ref{sec:lang});

\item we detail checking and inference algorithms for our
  specifications (Section~\ref{sec:analysis}), derived from a
  denotational model for the specificaiton language
  (Section~\ref{sec:semantics});

\item we provide an implementation of our approach as an extension to
  CamFort, an open-source program analysis tool for Fortran;

\item we report on a quantitative study of stencil computations on a
  corpus of numerical Fortran programs (ranging from small to large),
  totalling one million lines of code. Our tool identifies and infers
  specifications for 35,000 stencil computations in the
  corpus. Approximately 10\% of the stencils we found are non-trivial,
  corresponding to code which is a possible source of errors. This
  validates our hypothesis that the majority of stencil computations have a
  regular shape, and validates the design of our language in its
  capability to capture many core patterns.

\item We give a detailed verification case study of two
  particular stencil computation, simulating programming errors.
  Our approach is able to detect the all possible
  indexing errors.
\end{itemize}
%

\end{document}

### 2016-10-15 new-intro.tex
Add citation to Kamil et al.

@@
   \texttt{a(i+1,j-1)}. This contrasts with other stencil specification
-  approaches. For example, the work of Kamil et al. provides
-  full-functionality specifications of stencil computations from which
-  equivalent DSL code can be generated.  If a user wishes to specify,
-  say, the Naiver-Stokes stencil above, they must write a
-  specification which is very similar, involving textually dense
-  array-indexing expressions. Therefore any low-level mistakes that
-  could be made programming complex indexing code could also be made
-  when writing a specification. Our specifications are much more
-  abstract, and also more lightweight with the aim of aiding adoption
-  by scientists.
+  approaches. For example, the work of Kamil et al.~\cite{kamil2016verified}
+  provides full-functionality specifications of stencil computations from which
+  equivalent DSL code can be generated.  If a user wishes to specify, say, the
+  Naiver-Stokes stencil above, they must write a specification which is very
+  similar, involving textually dense array-indexing expressions. Therefore any
+  low-level mistakes that could be made programming complex indexing code could
+  also be made when writing a specification. Our specifications are much more
+  abstract, and also more lightweight with the aim of aiding adoption by
+  scientists.
 \end{itemize}

### 2016-10-15 new-intro.tex
Specify that we handle FORD syntax

@@
 integration with existing tools, as well as
-  documentation systems (\eg{}, we support Doxygen syntax).
+  documentation systems (\eg{}, we support Doxygen and FORD syntax).
 %

### 2016-10-15 new-intro.tex
Reference to evaluation section in contributions

@@
 \item we report on a quantitative study of stencil computations on a
-  corpus of numerical Fortran programs (ranging from small to large),
-  totalling one million lines of code. Our tool identifies and infers
-  specifications for 35,000 stencil computations in the
+  corpus of numerical Fortran programs (ranging from small to large)
+  (Section~\ref{sec:evaluation}), totalling one million lines of code. Our tool
+  identifies and infers specifications for 35,000 stencil computations in the
   corpus. Approximately 10\% of the stencils we found are non-trivial,
-  corresponding to code which is a possible source of errors. This
-  validates our hypothesis that the majority of stencil computations have a
-  regular shape, and validates the design of our language in its
-  capability to capture many core patterns.
+  corresponding to code which is a possible source of errors. This validates
+  our hypothesis that the majority of stencil computations have a regular
+  shape, and validates the design of our language in its capability to capture
+  many core patterns.
 

### 2016-10-15 new-intro.tex
Emphesise that case studies are not synthetic
Fix typo in contributions

@@
 \item We give a detailed verification case study of two
-  particular stencil computation, simulating programming errors.
-  Our approach is able to detect the all possible
-  indexing errors.
+  particular stencil computation commonly used in scientifc computing,
+  simulating programming errors. Our approach is able to detect all
+  possible indexing errors for these computations.
 \end{itemize}

### 2016-10-15 new-intro.tex
Replace occurrences of (Sub)section with \S

@@
   that captures many common forms of data access pattern
-  (Section~\ref{sec:lang});
+  (\S~\ref{sec:lang});
 
 \item we detail checking and inference algorithms for our
-  specifications (Section~\ref{sec:analysis}), derived from a
+  specifications (\S~\ref{sec:analysis}), derived from a
   denotational model for the specificaiton language
-  (Section~\ref{sec:semantics});
+  (\S~\ref{sec:semantics});
 
@@
   corpus of numerical Fortran programs (ranging from small to large)
-  (Section~\ref{sec:evaluation}), totalling one million lines of code. Our tool
+  (\S~\ref{sec:evaluation}), totalling one million lines of code. Our tool
   identifies and infers specifications for 35,000 stencil computations in the

### 2016-10-15 paper.tex
revert to sigplan format
Replace occurrences of (Sub)section with \S

@@
 \usepackage{subfigure}
-\usepackage{amsthm}
 \usepackage{multirow}
@@
 \usepackage{xypic}
+\usepackage{natbib}
 
@@
 \newtheorem{proposition}[block]{Proposition}
-%\newtheorem{definition}[block]{Definisssstion}
 
@@
 \titlebanner{DRAFT -- Do not distribute}
-%\authorinfo{}{}{}
+\authorinfo{}{}{}
 \authorinfo{Dominic Orchard \and Mistral Contrastin
@@
   that captures many common forms of data access pattern
-  (Section~\ref{sec:lang});
+  (\S~\ref{sec:lang});
 
 \item we detail inference, checking, and specification synthesis 
-algorithms for our language (Section~\ref{sec:analysis}), derived from
-a denotational model for the language (Section~\ref{sec:semantics});
+algorithms for our language (\S~\ref{sec:analysis}), derived from
+a denotational model for the language (\S~\ref{sec:semantics});
 
@@
 out-of-bounds access at the edges). We revisit this hypothesis
-in Section~\ref{sec:evaluation} with the inference of
+in \S~\ref{sec:evaluation} with the inference of
 such regular stencil patterns on our corpus of numerical programs.
 
-Section~\ref{sec:syntax} outlines the syntax.
- Section~\ref{sec:eqs} provides an equational
+\S~\ref{sec:syntax} outlines the syntax.
+ \S~\ref{sec:eqs} provides an equational
 theory for specifications and a notion of
-sub-specification.  Section~\ref{sec:semantics}
+sub-specification.  \S~\ref{sec:semantics}
 gives a semantic model via a set-based interpretation, for which the
@@
 ``disjunction'' and \term{*} as the ``conjunction''). We revisit this
-normalisation in Section~\ref{sec:inf-step4}.
+normalisation in \S~\ref{sec:inf-step4}.
 
@@
 specification, but it is also used extensively in the inference
-procedure (Section~\ref{subsec:inference}) to simplify specifications.
+procedure (\S~\ref{subsec:inference}) to simplify specifications.
 
@@
 stencil specification language. This model has a number of purposes: (1) it is used in the inference,
-checking, and synthesis algorithms (Section~\ref{sec:analysis}); (2)
+checking, and synthesis algorithms (\S~\ref{sec:analysis}); (2)
 justifies the equational
@@
 to guide correct implementations. The model domain is over sets of vectors which we call \emph{index
-  schemes}, which we introduce first before the model in Section~\ref{subsec:model}.
+  schemes}, which we introduce first before the model in \S~\ref{subsec:model}.
 
@@
 (but which isn't used here). These are used for labelled models, and other components later in
-Section~\ref{sec:analysis}.% with additional information.
+\S~\ref{sec:analysis}.% with additional information.
 \label{def:mult-and-approx}
@@
 We briefly sketch the algorithms for inferring and synthesising
-specifications (Section~\ref{subsec:inference}), for checking code
-against specifications (Section~\ref{subsec:checking}). These utilise the semantic model of
-Section~\ref{sec:semantics}.
+specifications (\S~\ref{subsec:inference}), for checking code
+against specifications (\S~\ref{subsec:checking}). These utilise the semantic model of
+\S~\ref{sec:semantics}.
 
@@
 The final spans are used to synthesise concrete stencil specification
-syntax (Section~\ref{sec:inf-step4}), where
+syntax (\S~\ref{sec:inf-step4}), where
 each span becomes a product \texttt{*} of regions which are 
@@
 source language is verified against a specification. The model of
-Section~\ref{sec:semantics} again aids this process.  Checking
+\S~\ref{sec:semantics} again aids this process.  Checking
 proceeds by generating a model $M$ from a specification and generating a set $N$ of indexing
@@
 stencil specification inferred previously in
-Section~\ref{subsec:inference}:
+\S~\ref{subsec:inference}:
 %
@@
 \item Generate the model from the specification as in
-  Section~\ref{sec:semantics}. 
+  \S~\ref{sec:semantics}. 
 %
@@
  above is more accepting than that used in
-Section~\ref{sec:lang}. This introduces a new notion of approximation
+\S~\ref{sec:lang}. This introduces a new notion of approximation
 beyond what is given by the syntactic theory of sub-specifications
-$<:$ in Section~\ref{sec:eqs}. Consistency leads to a notion of
+$<:$ in \S~\ref{sec:eqs}. Consistency leads to a notion of
 approximation that is a \textit{semantic sub-specification} relation:
@@
 are the inputs to Step 2 of the Inference process described in 
-Section~\ref{subsec:inference}. Each potential statement for which we
+\S~\ref{subsec:inference}. Each potential statement for which we
 eventually generate some specification is also counted as

### 2016-11-02 new-intro.tex
POPL version

@@
 
+\title{Spatial Verification of Stencil Computations with Lightweight
+   Specifications}
+
 \begin{document}
+\maketitle
 
@@
 It should be noted that our verification tool does not deal in the
-class of bugs that can be detected automatically (\emph{push-button
+class of bugs that can be detected automatic
+ally (\emph{push-button
   verification}). Instead, stencil computation bugs must largely be
@@
   that captures many common forms of data access pattern
-  (\S~\ref{sec:lang});
+  (\S~\ref{sec:lang}); we explain the syntax and describe informally
+  its semantics before giving a concrete denotational model
+  which makes the meaning precise (\S~\ref{sec:model});
 
 \item we detail checking and inference algorithms for our
-  specifications (\S~\ref{sec:analysis}), derived from a
-  denotational model for the specificaiton language
-  (\S~\ref{sec:semantics});
+  specifications (\S~\ref{sec:analysis}), derived from the
+  denotational model;
 
@@
   identifies and infers specifications for 35,000 stencil computations in the
-  corpus. Approximately 10\% of the stencils we found are non-trivial,
-  corresponding to code which is a possible source of errors. This validates
+  corpus. Approximately 10\% of the stencils we found are non-trivial;
+  we quantify the different classes of pattern found in our corpus. This validates
   our hypothesis that the majority of stencil computations have a regular
@@
 
-\item We give a detailed verification case study of two
+\item to test our implementation, we give a verification case study of two
   particular stencil computation commonly used in scientifc computing,
   simulating programming errors. Our approach is able to detect all
-  possible indexing errors for these computations.
+  possible indexing errors for these computations (\S~\ref{sec:casestudies}).
 \end{itemize}

### 2016-11-02 paper.tex
POPL version
starting on PLDI version

--- 2016-07-14 paper.tex
@@
 \titlebanner{DRAFT -- Do not distribute}
-%\authorinfo{}{}{}
-\authorinfo{Dominic Orchard \and Mistral Contrastin
-\and Matthew Danish \and Andrew Rice}{University of Cambridge}{}
+\authorinfo{}{}{}
 
@@
 
-\begin{abstract}
-  Verifying the correctness of numerical computations relies on first
-  giving a specification of their behaviour. In many cases, a
-  numerical computation is derived from an underlying mathematical
-  model, but this is often infeasible as a specification since it
-  is highly decoupled from the implementation. In this paper, we consider a 
-  lightweight specification technique between low-level and
-  high-level specifications for numerical code. We
-  target \emph{stencil computations}---a common idiom in numerical
-  code, but one which is error-prone due to fine-grained indexing
-  errors. We show that in practice, stencil computations tend
-  to have a regular shape which is amenable to static analysis and
-  simple specification. We describe an abstract spatial specification
-  language for stencil computations that can be embedded as
-  annotations in source code, along with inference, checking, and
-  specification synthesis. We evaluate our
-  language against a corpus of numerical Fortran code, and show that
-  the vast majority of stencil computations indeed have a simple,
-  regular, static shape. Around 10\% of the stencils found evoke a
-  programming pattern for which indexing errors are plausible. 
-  We simulate relevant programming errors in two stencil examples, 
-  and show that we are able to detect the vast majority of potential
-  errors. This paper details our stencil specification language, a
-  denotational model, inference,
-  checking, and synthesis procedures, our implementation, and
-  our evaluation studies.
-\end{abstract}
-
-\keywords{program verification, specification, stencils}
-
-\bibliographystyle{abbrvnat}
-
-\section{Introduction}
-
 \emph{Stencils} are a ubiquitous programming pattern, common in
 scientific and numerical computing applications. Informally, a stencil
-computation outputs an array, where the value at each index $i$ of
-this array is calculated from a \emph{neighbourhood} of values around $i$ in
-some input array(s), \eg{}, the Game of Life, convolutions in image
+computation computes an array whose value at each index $i$ is
+calculated from a \emph{neighbourhood} of values around $i$ in some
+input array(s), \eg{}, the Game of Life, convolutions in image
 processing, approximations to differential equations. For example, the
-following computes the one-dimensional discrete Laplace
-transform (an approximation to a derivative) in Fortran:
+following computes the one-dimensional discrete Laplace transform (an
+approximation to a derivative) in Fortran:
 %
 \begin{minted}{fortran}
-do iter = 0, itermax
 do i = 1, (n-1)
@@
 end do
-   a = b
-end do
 \end{minted}
 %
-Line 3 is the core of the stencil computation, calculating
-values at \texttt{b(i)} from a neighbourhood of elements about
-\texttt{a(i)}. Line 5 swaps
-\texttt{a} and \texttt{b} between iterations, where \texttt{b} becomes the
-input for the next iteration. %In Fortran, parentheses \texttt{( )} are used
-%for array subscripts rather than the more familiar bracket syntax \texttt{[ ]}.
-%
-%This stencil computation exhibits statically decideable
-%spatial and temporal relationships between $\texttt{a}$ and
-%$\texttt{b}$.
-The shape of the data access pattern on line 3 determines other aspects of the
-program and its efficient implementation: how much ``boundary'' is
-needed for the array, the most cache-efficient layout in memory,
-and the partitioning shape for parallel implementations.
-
-In this example, the access pattern is simple, static, and easily
-understood. More complex stencil computations
-are much more prone to error from simple lexical mistakes. For example,
-Figure~\ref{ref:navier-stokes-fragment} shows three lines from a
+Values \texttt{b(i)} are calculated from a neighbourhood of elements
+around \texttt{i} in the input array \texttt{a}.  In this example, the
+access pattern is simple and easily understood. More complex
+stencil computations are more prone to errors from simple lexical
+mistakes. For example, the following is a three-line snippet from a
 Navier-Stokes fluid simulator in which two arrays are read with
-different data access patterns, across two dimensions. The interaction
-is much harder to understand, with the potential for the developer to
-introduce an error via simple textual mistakes, for
-example writing $\texttt{(i-1,j)}$ instead of $\texttt{(i+1,j)}$.
-
-In practice, most stencil computations have a regular shape that can
-be described simply and abstractly with a small set of coarse-grained
-spatial descriptions. We introduce a simple specification language of
-such descriptions, abstracting over the fine-grained detail of stencil
-access patterns. For the Laplace example, our
-inference procedure provides the following specification to line 3:
-%
-\begin{minted}[linenos=false]{fortran}
-!=  stencil centered(depth=1, dim=1) :: a
-\end{minted}
-%
-This describes that \texttt{a} is accessed with a
-symmetrical pattern (``centered'') to a depth of one in each
-direction in its first (and only) dimension.
-%The second line explains
-%the temporal relationship between \texttt{b} and \texttt{a}: that
-%previous time step for \texttt{b} is actually provided by \texttt{a},
-%and vice versa. This is explained as a mutual dependence between
-%\texttt{a} and \texttt{b}.
-The inferred specification for the Navier-Stokes example is shown 
-in Fig.~\ref{ref:navier-stokes-fragment}(b). The
-specification requires that, over the whole fragment, \texttt{u} is
-accessed with a centered pattern to depth of 1 in both dimensions
-(this is known as the \emph{five-point stencil}) and \texttt{v}
-is accessed in a neighbourhood bounded by forward to depth of $1$ in
-the first dimension and backward to a depth of $1$ in the second
-dimension. Fig.~\ref{ref:navier-stokes-fragment}(c) illustrates
-the specification pictorially.
-
-In this paper, we make the following contributions:
-%
-\begin{itemize}[leftmargin=0.8em]
-\item we introduce a specification language for stencils 
-  that captures many common forms of data access pattern
-  (Section~\ref{sec:lang});
-
-\item we detail inference, checking, and specification synthesis 
-algorithms for our language (Section~\ref{sec:analysis}), derived from
-a denotational model for the language (Section~\ref{sec:semantics});
-
-\item we provide an implementation of our approach as an extension to
-  CamFort, an open-source program analysis tool for Fortran;
-
-\item we report on a quantitative study of
-  stencil computations on a corpus of numerical Fortran programs
-  (ranging from small to large),  totalling one million lines of code. Our tool identifies and
-  synthesises specifications for 35,000 stencil computations in the
-  corpus. Approximately 10\% of the stencils we found are non-trivial,
-  corresponding to code which is a possible source of errors;
-
-\item we give a detailed verification case study of two
-  particular stencil computation, simulating programming errors.
-  Our approach is able to detect the vast majority of possible
-  indexing errors.
-\end{itemize}
-%
-
-\begin{figure}[t]
-\begin{minted}[firstnumber=20,xleftmargin=2em]{fortran}
+different data access patterns, across two dimensions.
+%%
+\begin{minted}[firstnumber=20,xleftmargin=4em]{fortran}
 du2dx = ((u(i,j)+u(i+1,j))*(u(i,j)+u(i+1,j))+   &
@@
 \end{minted}
-(a) Excerpt of a Navier-Stokes fluid simulator
-(based on~\citet{griebel1997numerical}) showing highly-detailed stencil computation.
-\vspace{0.5em}
-\hrule
-%
-\begin{minted}[linenos=false,xleftmargin=1em]{fortran}
+%%
+The access pattern is much harder to understand than the
+one-dimensional Laplace.  The miasma of indexing expressions of the
+form \texttt{var(i$\pm$a, j$\pm$b)} is not only hard to read, but is
+prone to simple textual input mistakes, \eg{}, swapping \texttt{-} and
+\texttt{+}, missing an indexing term, or transforming the wrong
+variable \eg{} \texttt{(i+1,j)} instead of \texttt{(i,j+1)}.
+
+In practice, we find that the usual development procedure for complex
+stencil computations involves some ad hoc testing to ensure that no
+transcription mistakes have been made. This may involve visual
+inspection of the results, based on a rendered image. Typically such
+testing is then discarded once the code is seen to be correct.
+
+This is not the only information that is typically discarded. The
+shape of the indexing pattern is often a result of choices made in the
+numerical-analysis discretisation procedure used to convert a set of
+continuous equations into a discrete approximation. Rarely are these
+decisions captured in the source code, yet the shape of access is
+usually uniform and is given a perspicuous and concise description in
+numerical analysis literature \eg{}, ``\emph{centered in space, to a
+  depth of 1}'' referring to indexing terms \texttt{a(i)},
+\texttt{a(i-1)} and \texttt{a(i+1)}.
+
+To support the development of correct stencil computations, we propose
+a simple, abstract specification language for the data access pattern
+of stencils. The shape of the stencil computation in the
+Laplace example is specifed in our language as:
+%
+\begin{minted}[linenos=false]{fortran}
+!=  stencil centered(depth=1, dim=1) :: a
+\end{minted}
+%
+This describes that \texttt{a} is accessed with a symmetrical pattern
+(``centered'') to a depth of one in each direction in its first
+dimension.  The stencil computation shape in the Navier-Stokes example
+is specified as:
+%
+\begin{minted}[linenos=false,xleftmargin=4em,breakindent=3em]{fortran}
 != stencil centered(depth=1,dim=1)*reflexive(dim=2) + centered(depth=1,dim=2)*reflexive(dim=1) :: u
@@
 \end{minted}
-(b) Inferred and synthesised specification for (a) by CamFort.
-\vspace{0.5em}
-\hrule
-%
-\begin{center}
-\begin{tikzpicture}[scale=0.85]
-\node at (1.25,1.22) {i, j};
-\draw (1,1) rectangle (1.5,1.5);
-\draw (1.5,1) rectangle (2,1.5);
-\draw (1,0.5) rectangle (1.5,1);
-\draw (0.5,1) rectangle (1,1.5);
-\draw (1,1.5) rectangle (1.5,2);
-%
-\node at (4.25,1.22) {i, j};
-\draw (4,1) rectangle (4.5,1.5);
-\draw (4.5,1) rectangle (5,1.5);
-\draw (4.5,1.5) rectangle (5,2);
-\draw (4,1.5) rectangle (4.5,2);
-\end{tikzpicture}
-\end{center}
-\vspace{-0.5em}
-(c) Pictorial representation of the two stencil specifications.
-The horizontal dimension is \texttt{dim=1} and the vertical is \texttt{dim=2}.
-\caption{Fragment of a Navier-Stokes fluid simulator and its
-  specification in our language.}
-\label{ref:navier-stokes-fragment}
-\vspace{-0.8em}
-\end{figure}
-
-%This paper considers the technical aspects of the language,
-%its semantics, inference, checking, and synthesis procedures.
-%We also evaluate our implementation on real scientific code 
-%used actively in research, ranging from small ($\sim$700loc) to large
-%($\sim$600kloc). 
- %Our case studies include the Unified Weather 
-%model used by the UK Met Office~\cite{um-website} ($\approx{}600\textit{kloc}$) and other
-%research-based numerical models, ranging from small to large.
-
-\section{Stencil specification language}
-\label{sec:lang}
-
-Our specification language is based on the hypothesis
-that most forms of array access in numerical code have
-a fixed, statically-determined access pattern. For example, the
-\emph{five-point stencil} on a two-dimensional array reads from array
-indices $(i, j)$, $(i-1, j)$, $(i+1, j)$, $(i, j-1)$, and $(i, j+1)$
-for all $i, j$ within the inner boundary of the array (to avoid
-out-of-bounds access at the edges). We revisit this hypothesis
-in Section~\ref{sec:evaluation} with the inference of
-such regular stencil patterns on our corpus of numerical programs.
-
-Section~\ref{sec:syntax} outlines the syntax.
- Section~\ref{sec:eqs} provides an equational
-theory for specifications and a notion of
-sub-specification.  Section~\ref{sec:semantics}
-gives a semantic model via a set-based interpretation, for which the
-equality and sub-specification relation are proven sound.
-
-\paragraph{Notation and convention}
-%\label{sec:notation}
-
-\renewcommand*{\arraystretch}{0.8}
-For the target language, $e$ ranges over expressions (may be impure) and $v$
-over its (imperative) variables. %By
-%``variable'', we mean in the imperative sense (named binders to memory cells).
-
-\begin{definition}[Base induction variable]
-  A integer variable \textit{v} is a \emph{base induction
-    variable} if it is the control variable of a ``for''
-  loop,\footnote{Or equivalent, \eg{}, \texttt{do} in Fortran, for our implementation.} incremented by $1$ per
-iteration. The variable is marked as an induction variable
-only within the body of the loop. Since we ignore non-base
-(derived) induction variables, we simply say \emph{induction
-  variable}. Induction variables are ranged over by $i$, $j$, $k$ throughout. 
-\end{definition}
-
-\begin{definition}[Array subscript and index]
-  An \emph{array subscript} is an expression that indexes an array, which we denote
-  by $v(\bar{e})$ for the source language where $\bar{e}$ is shorthand
-  for a syntactic list of indexing expressions. A \emph{relative
-    index} is a list $\bar{e}$ where each $e \in \bar{e}$ is defined
-  in terms of base induction variables.
-\end{definition}
-
-\begin{definition}[Neighbourhood index]
-  For an array subscript $v(\bar{e})$, we say that $e \in \bar{e}$
-  is a \emph{neighbourhood index} if it is of the form
-  $e \equiv i$, $e \equiv i + a$ or $e \equiv i - a$ where $a$ is an
-  integer constant. That is, a neighbourhood index is a
-  relative index defined as constant translations of base induction
-  variables. (The relation $\equiv$ here identifies terms up-to commutativity
-  of $+$ and the inverse relation of $+$ and $-$ \eg{},
-  $(-b) + i \equiv i - b$).  We classify neighbourhood indices 
-  by the predicate $\neigh_I$ where $I$ is a set of induction
-  variables. We omit $I$ when clear from the context.
-\label{def:neighbour}
-\end{definition}
-
-\subsection{Specification syntax}
-\label{sec:syntax}
-
-Figure~\ref{fig:syntax} gives the syntax of stencil specifications,
-which we introduce in stages below.  The top-level is given by the
-\nonterm{specification} production which splits into either a
-\nonterm{regionDec} (region declaration) or a \nonterm{specDec}
-(specification declaration). Specification declarations associate
-(via the operator $\texttt{::}$) a specification to one or more program variables
-$\bar{v}$, describing how the array variables $\bar{v}$ are accessed.
-%Specification declarations are either \nonterm{spatial} or
-%\nonterm{temporal} specifications and describe, for spatial
-%specifications,
-%Specifications declarations describe
-%how an array variable $v$ is read
-%, and for temporal
-%specifications what variables are used to define the array variable $v$.
-%%
-
-%\subsubsection{Spatial specification syntax}
-
-\emph{Regions} are the central building blocks of spatial
-specifications. Regions can either be declared alone with a
-\nonterm{regionDec}, assigning a region specification \nonterm{region} to
-a region variable \nonterm{rvar} (via $\texttt{::}$) or given directly within a
-\nonterm{spatial} specification.
-
-\paragraph{Region constants}
-
-Regions have as terminals the \emph{region constants}, denoted by
-\term{reflexive}, \term{forward}, \term{backward}, or \term{centered}. Each
-region constant except \term{reflexive} is given a depth parameter $n$ (natural
-number greater than 0). Every region constant receives a dimension identifier $d$ (also a
-natural number greater than 0). Region constants specify that an array is read
-from by a collection of indices which in the $d^{th}$ dimension are
-neighbourhood indices ranging from $i + 0$ up to $i \pm n$ inclusively,
-\eg{}, the following is a valid stencil, reading from \term{a} and writing to
-\term{b}:
-%%
-\begin{minted}{fortran}
-!= stencil forward(depth=2, dim=1) :: a
-b(i, 0) = a(i, 0) + a(i+1, 0) + a(i+2, 0)
-\end{minted}
-%%
-where \texttt{i} is an induction variable.  The specification
-is associated to the array variable \texttt{a}. Note that the second
-dimension is constant.
-
-The \term{forward},
-\term{backward}, and \term{centered} regions can all receive an
-additional optional attribute \term{irreflexive} which marks
-that the region does not include the origin (offset 0).
-For example, the following is a \term{backward} stencil, which is similar
-to \term{forward} but has negative neighbourhood indices, and which
-is irreflexive:
-%
-\begin{minted}{fortran}
-!= stencil backward(depth=2, dim=1, irreflexive) :: a
-b(i) = a(i-1) + a(i-2)
-\end{minted}
-%
-The \term{irreflexive} attribute cannot be used on \term{reflexive}
-regions.
-
-A \texttt{centered} stencil is a combination of \texttt{forward}
-and \texttt{backward} stencils to the same depth, \eg{},
-%%
-\begin{minted}{fortran}
-!= stencil centered(depth=1, dim=1) :: a
-b(i) = ( a(i) + a(i-1) + a(i+1) ) / 3.0
-\end{minted}
-
-\begin{figure}[t]
-\vspace{-0.5em}
-\begin{align*}
-\def\arraystretch{1.2}
-\setlength{\arraycolsep}{0.2em}
-\newcommand{\dimTy}{\mathbb{N}_{>0}}
-\begin{array}{rl}
-\nonterm{specification} ::= & \nonterm{regionDec} \mid \nonterm{specDec} \\
-\nonterm{specDec} ::= & \term{stencil} \; \nonterm{spec} \;
-                        \texttt{::} \; v \\
-\nonterm{regionDec} ::= &  \texttt{region} \; \texttt{::} \; \nonterm{rvar} \; \texttt{=} \;
-                         \nonterm{region}\\[0.4em]
-%\nonterm{spec} ::= & \nonterm{spatial} \mid \nonterm{temporal}
-%\\[1em]
-\nonterm{spec} ::= & [\nonterm{approx},] \; [\nonterm{mult},] \; \nonterm{region} \\
-\nonterm{mult} ::= & \term{readOnce} \\
-\nonterm{approx} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
-\nonterm{region} ::= & \nonterm{rvar} \; \\
-\multicolumn{2}{l}{\qquad\qquad \mid \; \stenRefl{\dimTy}} \\
-\multicolumn{2}{l}{\qquad\qquad \mid \;\stenFwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]}} \\
-\multicolumn{2}{l}{\qquad\qquad \mid \; \stenBwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]}} \\
-\multicolumn{2}{l}{\qquad\qquad \mid \; \stenCen{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]}} \\
-\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{region} \; \term{+}
-  \; \nonterm{region} \; \mid \; \nonterm{region} \; \term{*} \; \nonterm{region}} \\[0.5em]
-%\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{rvar}}  
-%\\[0.5em]
-%\nonterm{temporal} ::= \; & \term{dependency} \; (v \; \{ , v \}) [, \texttt{mutual}]
-%  \\[0.5em]
-\nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+\\[-0.5em]
-\end{array}
-\end{align*}
-\caption{Specification syntax (EBNF grammar)}
-\label{fig:syntax}
-\vspace{-0.8em}
-\end{figure}
-
-\paragraph{Sum and product of regions}
-%%
-Region terms can be combined using the sum operator
-\term{+} or the product operator \term{*}.
-
-The product of two regions $r \texttt{*} r'$ specifies that an array
-is read with neighbourhood indices drawn from the bounding box created
-by the regions two regions $r$ and $r'$. For example, the following
-code (defining a \emph{nine-point stencil}) has a specification given by the
-product of two \texttt{centered} regions in each dimension:
-%%
-\begin{minted}[breakindent=2.9em]{fortran}
-x = a(i, j)   + a(i-1, j)   + a(i+1, j)
-y = a(i, j-1) + a(i-1, j-1) + a(i+1, j-1)
-z = a(i, j+1) + a(i-1, j+1) + a(i+1, j+1)
-!= stencil centered(depth=1, dim=1) * centered(depth=1, dim=2) :: a
-b(i, j) = (x + y + z) / 9.0
-\end{minted}
-%%
-This pattern is common in image convolution applicaitons (for example
-for edge detection). The specification ranges over the
-values that flow to the array subscript on the left-hand side,
-and so ranges over the intermediate assignments to \term{x},
-\term{y}, and \term{z}.
-
-The region constructor \texttt{*} can also be interpreted as a
-kind of conjunction on specifications since they constrain
-indices to be within both regions simultaneously.
-
-The sum of two regions $r \texttt{+} r'$ specifies that an array is
-read using the neighbourhood indices described by either $r$ or
-$r'$; it acts as a disjunction on specifications. For
-example, the following gives the specification of a five-point
-stencil which is the sum of two compound \texttt{reflexive} and
-\texttt{centered} regions in each dimension:
-%%
-\begin{minted}{fortran}
-!= stencil centered(depth=1, dim=1)*reflexive(dim=2) + centered(depth=1, dim=2)*reflexive(dim=1) :: a
-b(i,j) = -4*a(i,j) + a(i-1,j) + a(i+1,j) &
-                   + a(i,j-1) + a(i,j+1)
-\end{minted}
-%%
-Here the left-hand side of \texttt{+} says that when the second dimension
-(induction variable $j$) is fixed at the origin, the first dimension
-(induction variable $i$) accesses the immediate vicinity of the origin
-(to depth of one). The right hand side of \texttt{+} is similar but the dimensions are reversed.
-This reflects the symmetry under rotation of the five-point stencil.
-
-As another example, the following
-implements and specifies the \emph{Roberts cross}
-edge-detection convolution~\cite{davis1975survey}:
-%%
-\begin{minted}[breakindent=5.5em]{fortran}
-do j=0, jmax-1
-  do i=0, imax-1
-    x = a(i,j)-a(i+1,j+1); y = a(i+1,j)-a(i,j+1)
-    != stencil forward(depth=1,dim=1) * forward(depth=1,dim=2) :: a
-    b(i,j) = sqrt(x*x+y*y)
-  end do
-end do
-\end{minted}
-
-\paragraph{Region declarations and variables}
-
-Region specifications can be assigned to region variables via
-a region declaration, which can be used later to form a spatial
-specification. For example, the specification from the previous
- example can be restated as:
-%%
-\begin{minted}{fortran}
-!= region :: r1 = forward(depth=1, dim=1)
-!= region :: r2 = forward(depth=1, dim=2)
-!= region :: robertsCross = r1*r2
-!= stencil robertsCross :: a
-\end{minted}
-This is especially useful for common stencil patterns, such as Roberts cross,
-as the region can be defined once and reused.
-%%
-\paragraph{Modifiers}
-%%
-Region specifications can be modified 
-by \emph{approximation} and \emph{multiplicity} information 
-(in the \textit{spec} rule of Fig.~\ref{fig:syntax}).
-
-The \texttt{readOnce} modifier enforces that no index appears more
-than once (that is, its multiplicity is one). For example, in all of
-the previous examples the \texttt{readOnce} modifier could be added,
-\eg{}
-%
-\begin{minted}{fortran}
-!= stencil readOnce, backward(depth=2, dim=1) :: a
-b(i+1) = a(i) + a(i-1) + a(i-2)
-\end{minted}
-%
-This specification would be invalid if any of the
-array subscripts were repeated. This modifier provides a way to
-rule out any accidental repetition of array subscripts.
-The notion is similar to that of \emph{linearity} in linear type
-systems~\cite{wadler1990linear}, where a value must be used
-exactly once. We opt for the more informative and easily understood name
-\texttt{readOnce}. This modifier is optional, so it need not
-be present even if the stencil is linear.
-
-%At the region level, \texttt{irreflexive} modifier qualifies base
-%regions of \texttt{forward}, \texttt{backward}, and \texttt{centered}.
-%It indicates that in the dimension of the region there is no neighbour index
-%with $0$ offset, \eg{},
-%
-%\begin{minted}{fortran}
-%!= stencil centered(depth=1, dim=1, irreflexive) :: a
-%b(i) = a(i+1) + a(i-1)
-%\end{minted}
-%
-%or with just a \texttt{forward} region an example would be:
-%%
-%\begin{minted}{fortran}
-%!= stencil forward(depth=1, dim=1, irreflexive) :: a
-%b(i) = a(i+1)
-%\end{minted}
-
-%\paragraph{Under and over-approximation}
-
-In some cases, it is useful to give a lower and/or upper bound for a
-stencil. This can be done using either the \term{atMost} or
-\term{atLeast} modifiers. This is particularly useful in situations
-where there is a non-contiguous stencil pattern, which cannot be expressed
-precisely in our specification syntax. For example:
-%%
-\begin{minted}{fortran}
-!= stencil atLeast, reflexive(dim=1)       :: a
-!= stencil atMost, forward(depth=2, dim=1) :: a
-b(i) = a(i) + a(i+2)
-\end{minted}
-%%
-%\subsubsection{Temporal specifications}
-
-\paragraph{Note on the design}
-
-The names ``forward'', ``backward'' and ``centered''
-are inspired by standard terminology in numerical analysis
-for the shape of discretisation schemes. For example,
-the standard \emph{explicit method} for approximating
-PDEs is known as the \emph{Forward Time, Centered Space} (FTCS)
-scheme~\cite{dawson1991finite}. For
-the one-dimensional heat equation, an FTCS discretisation
-provides code with the following stencil~\cite{recktenwald2004finite}:
-%%
-\begin{minted}{fortran}
-do i=2, n-1
-  u(i) = r*v(i-1) + r2*v(i) + r*v(i+1)
-end do
-\end{minted}
-%%
-where \texttt{r} and \texttt{r2} are constants.
-A valid specification for this is:
-%%
-\begin{minted}{fortran}
-!= stencil centered(depth=1, dim=1) :: v
-\end{minted}
-%%
-Such a specification can be inferred, or can be inserted by a user
-and checked against the code.
-
-\subsection{Equational theory and sub-specifications}
-\label{sec:eqs}
-
-\begin{figure}
-%\vspace{-0.2em}
-\begin{align*}
-\setlength{\arraycolsep}{0.05em}
-\begin{array}{c}
-%\framebox{$\textit{region} \equiv \textit{region}'$} \\[0.6em]
-\begin{array}{rll}
-(\textsc{F\, \texttt{+} \,F}) \;\; &
-\stenFwdSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
- \equiv \; & \stenFwdSR{n}{d}{r} \; \texttt{+} \; \stenFwdSR{m}{d}{s} \\[0.6em]
-%
-%
-(\textsc{B\, \texttt{+} \,B}) \;\; &
-\stenBwdSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
- \equiv \; & \stenBwdSR{n}{d}{r} \; \texttt{+} \; \stenBwdSR{m}{d}{s} \\[0.6em]
-%
-%
-(\textsc{C\, \texttt{+} \,C}) \;\; &
-\stenCenSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
-\equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \; \stenCenSR{m}{d}{s} \\[0.6em]
-%
-%
-(\textsc{C\, \texttt{+} \,F}) \;\; & \stenCenSR{n}{d}{r \vee s} \\
-\textit{$m \leq n$} \; \equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \;
-                      \stenFwdSR{m}{d}{s} \\[0.6em]
-%
-%
-(\textsc{C\, \texttt{+} \,B}) \;\; &
-\stenCenSR{n}{d}{r \vee s} \\
-\textit{$m \leq n$} \; \equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \;
-                      \stenBwdSR{m}{d}{s} \\[0.6em]
-%
-%
-(\textsc{B\, \texttt{+} \,F}) \;\; &
-\stenCenSR{n}{d}{r \vee s} \\
-\equiv \; & \stenFwdSR{n}{d}{r} \; \texttt{+} \; \stenBwdSR{n}{d}{s}
-  \\[0.6em]
-%
-%
-(\textsc{R\, \texttt{+} \,F}) \;\; &
-\stenFwdS{n}{d} \\
-\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenFwdSR{n}{d}{r} \\[0.6em]
-%
-%
-(\textsc{R\, \texttt{+} \,B}) \;\; &
-\stenBwdS{n}{d} \\
-\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenBwdSR{n}{d}{r} \\[0.6em]
-%
-%
-(\textsc{R\, \texttt{+} \,C}) \;\; &
-\stenCenS{n}{d} \\
-\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenCenSR{n}{d}{r} \\[0.6em]
-%
-%
-(\textsc{irrefl}) \;\; & (R \, \term{*} \, S^\irreflS) \, \term{+} \,
-                         (R^\irreflS \, \term{*} \, S) \;
-\equiv \; R \, \term{*} \, S \\[0.4em]
-%
-%
-(\texttt{+}\textsc{idem}) \;\; & S \; \texttt{+} \; S \; \equiv \; S \\[0.4em]
-%
-(\texttt{+}\textsc{comm}) \;\; & S \; \texttt{+} \; T \; \equiv \; T \;
-                       \texttt{+} \; S \\[0.4em]
-%
-(\texttt{+}\textsc{assoc}) \;\; & R \; \texttt{+} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
-                       \texttt{+} \; S) \; \texttt{+} \; T \\[0.4em]
-(\texttt{*}\textsc{comm}) \;\; & S \; \texttt{*} \; T \; \equiv \; T \;
-                       \texttt{*} \; S \\[0.4em]
-%
-(\texttt{*}\textsc{assoc}) \;\; & R \; \texttt{*} \; (S \; \texttt{*} \; T) \; \equiv \; (R \;
-                       \texttt{*} \; S) \; \texttt{*} \; T \\[0.4em]
-(\textsc{dist}) \;\; & R \; \texttt{*} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
-                       \texttt{*} \; S) \; \texttt{+} \; (R
-                       \; \texttt{*} \; T)
-\end{array}
-%
-%\framebox{${\textit{specDec} :: \texttt{v}} \equiv
-%{\textit{specDec'} :: \texttt{v}}$} \\[0.7em]
-%\begin{array}{rl}
-% MUTUAL
-%(\textsc{mutual}) \; &
-%\texttt{stencil} \; \texttt{dependency} (\texttt{v}), \texttt{mutual}
-%  :: \texttt{w}
-%\\
-%\equiv \; & \texttt{stencil} \; \texttt{dependency} (\texttt{w}), \texttt{mutual} ::
-%  \texttt{v} \\[0.5em]
-%(\textsc{coalesce}) \; &
-%\texttt{stencil} \; \texttt{dependency} (\bar{v})
-%  :: \texttt{v} \\
-%& ; \texttt{stencil} \; \texttt{dependency} (\bar{w})
-%  :: \texttt{v}
-%\\
-%\equiv \; & \texttt{stencil} \; \texttt{dependency} (\bar{v}, \bar{w}) ::
-%  \texttt{v} \\[0.5em]
-%\end{array}
-\end{array}
-\end{align*}
-\caption{Equations on specifications (\emph{regions})}
-\vspace{-0.5em}
-\label{fig:equations}
-\end{figure}
-
-Figure~\ref{fig:equations} lists the equational theory for
-specifications given by the binary relation $\equiv$
-on \emph{region} terms. The equations use abbreviations
-\term{refl}, \term{fwd}, \term{bwd}, \term{cen}, and \irreflS{}
-for \term{reflexive}, \term{forward}, \term{backward},
-\term{centered}, and \irrefl{} respectively. Furthermore, the abbreviations include syntactic sugar
-for quantifying over the absence or presence of the \irrefl{}
-attribute on region constants as a boolean. For example,
-\term{fwd} syntactic sugar is defined:
-%%
-\begin{align*}
-\hspace{-0.25em}
-& \stenFwdSR{n}{d}{r}
-\; := \; \\[-0.4em]
-& \qquad \begin{cases}
-\stenFwd{n}{d}{, \irrefl} & \;\;\textit{if} \;\, \neg \, r \\[-0.1em]
-\stenFwd{n}{d}{}  & \;\;\textit{if} \;\,  r
-\end{cases} \\
-\textit{or} \;\;
-& \stenFwdS{n}{d} := \stenFwd{n}{d}{}
-\end{align*}
-%
-and similarly for \term{bwd} and \term{cen}.
-
-Stencil specifications in \textit{spec} are considered
-equal when they have the same modifiers and $\equiv$ equates
-their regions, \ie{}:
-\begin{align*}
-(\textit{region} \equiv \textit{region'}) 
- \Rightarrow \quad\;\; & \texttt{stencil} \; [\textit{approx},] \; [\textit{mult},] \;
-\textit{region} \\[-0.4em]
-\quad \equiv \; & \texttt{stencil} \; [\textit{approx},] \;
-            [\textit{mult},] \; \textit{region}'
-\end{align*}
-%
-The equations reveal a part of the semantics for our
-specification language, which we briefly unpack.
-
-\paragraph{Overlapping regions with +}
-The first nine
-equations (with labels of the form $\ldots \texttt{+} \ldots$)
-explain the notion of \emph{region overlapping} with sum
-\term{+}. In the first six, $r$ and $s$ range
-over booleans denoting the absence (true) or presence (false)
-of \irrefl{}. These are combined via disjunction. For example,
-the rule $(\textsc{B} \, \term{+} \,
-\textsc{F})$ explains that the sum of \term{backward}
-and \term{forward} regions in the same dimension and the same
-depth is equivalent to a \term{centered} region of that depth.
-If the \term{forward} part is irreflexive ($r =
-\textit{false}$) but \term{backward} is not ($s =
-\textit{true}$) then their sum is equal to a \term{centered}
-region with no \irreflS{} modifier by the disjunction ($r \vee s =
-\textit{true}$), \ie{}:
-%%
-\begin{align*}
-& \stenCenS{n}{d}  \\[-0.2em]
-\equiv \;\; & \stenFwdSR{n}{d}{\irreflS} \; \texttt{+} \; \stenBwdS{n}{d}
-\end{align*}
-%%
-That is, the \term{fwd} region is irreflexive describing code which is missing
-a zero-offset neighbourhood index (\eg{}, missing $\texttt{a(i)}$), but the addition of
-the \term{bwd} has no irreflexive attribute and thus provides the
-missing zero offset index.
-
-The equation (\textsc{irrefl}) provides an annihilation of
-\irreflS{} using a further syntactic sugar.
-We write $R^{\irreflS}$ to denote a region constant with an
-irreflexive attribute, \eg{},
-%%
-\begin{align*}
-\stenFwdS{n}{d}^{\irreflS}  := \,
-\stenFwdSR{n}{d}{\irreflS}
-\end{align*}
-%%
-The left-hand side of (\textsc{irrefl}) has a sum of
-products of two regions $R$ and $S$, 
-where in each component of the sum one of $R$
-or $S$ has the irreflexive attribute. These cancel to give
-just $R \, \term{*} \, S$.
-
-\paragraph{Interaction between \term{+} and \term{*}}
-
-The last six equations (from (\textsc{\term{+}idem}) inclusive), explain
-the algebraic behaviour of \term{+} and \term{*}.  Together, we see
-that \term{+} is an associative, commutative, idempotent binary
-operator that distributes with \term{*}, which is associative and
-commutative. This distribution of \term{*} over \term{+} is key as it
-is used internally to give a normal form for stencil specifications
-akin to Disjunctive Normal Form (DNF) (where \term{+} is taken as
-``disjunction'' and \term{*} as the ``conjunction''). We revisit this
-normalisation in Section~\ref{sec:inf-step4}.
-
-The equational theory provides not only an axiomatic semantics
-for the language, which the user can use to rewrite their
-specification, but it is also used extensively in the inference
-procedure (Section~\ref{subsec:inference}) to simplify specifications.
-
-\begin{figure}[t]
-\begin{align*}
-\hspace{-0.5em}
-\begin{array}{c}
-%%
-%% INEQUATIONS ON REGION
-%% 
-%\framebox{$\textit{region} <: \textit{region}'$} \\[0.85em]
-%
-%\dfrac{}{R <:^r R}\trule{\textsc{refl}} \qquad \dfrac{P <:^r R \quad R <:^r Q}{P <:^r
-%  Q}\trule{\textsc{trans}} 
-%\qquad
-%\\[1.5em]
-\vspace{-0.4em}
-\setlength{\arraycolsep}{0.1em}
-\dfrac{\hspace{3em} m \leq n \qquad\qquad r \implies s \hspace{3em}}
-{\begin{array}{rl}
-\stenFwdSR{m}{d}{r} & <:^r \stenFwdSR{n}{d}{s}
-%& \;\, ds \subseteq es \wedge n \leq m
-\\
-\wedge \; \stenBwdSR{m}{d}{r} & <:^r \stenBwdSR{n}{d}{s}
-%& \;\, ds \subseteq es \wedge n \leq m
-\\
-\wedge \; \stenCenSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s}
-%& \;\, ds \subseteq es \wedge n \leq m
-\\
-\wedge \; \stenBwdSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s} \\
-\wedge \; \stenFwdSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s} \\[1.2em]
-\end{array}}\trule{\textsc{over}} \\[-0.2em]
-\hspace{-0.5em}
-\dfrac{P_1 <:^r R_1 \quad P_2 <:^r R_2}
-      {P_1 \, \texttt{+} \, P_2 <:^r R_1 \, \texttt{+} \, R_2}
-\trule{\textsc{cong}\texttt{+}} \;\;\;
-\dfrac{P_1 <:^r R_1 \quad P_2 <:^r R_2}
-      {P_1 \, \texttt{*} \, P_2 <:^r R_1 \, \texttt{*} \, R_2}
-\trule{\textsc{cong}\texttt{*}} \\[1.5em]
-%%
-%% INEQUATIONS ON SPEC
-%%
-%\framebox{$\textit{spec} <: \textit{spec}'$} \\[0.8em]
-%
-\dfrac{}{\term{readOnce} \, R <: R}\trule{\textsc{rep}}
-\quad
-\dfrac{}{\term{atLeast} \, R <: \term{atMost} \,
-  R}\trule{\textsc{bound}} \\[1.5em]
-%%
-\dfrac{P <:^r R}{\term{atLeast}  P <: \term{atLeast} 
-  R}\trule{\textsc{cong}\texttt{L}}
-\ 
-\dfrac{P <:^r R}{\term{atMost}  P <: \term{atMost} 
-  R}\trule{\textsc{cong}\texttt{M}} \\[1.5em]
-\dfrac{P <:^r R}{\term{readOnce} \, P <: \term{readOnce} \,
-  R}\trule{\textsc{cong}\texttt{R}} 
-\quad
-\dfrac{R \equiv S}{R <:^r S}\trule{\textsc{eq}} 
-%%
-\end{array}
-\end{align*}
-\caption{Inequations $<:^r$ on \textit{regions} and
-$<:$ on \textit{specs} (omitting reflexivity
-and transitivity). }
-\label{fig:inequations}
-\vspace{-0.8em}
-\end{figure}
-
-\paragraph{Inequations: sub-specifications}
-
-Figure~\ref{fig:inequations} defines a notion of sub-specifications
-via the relation $<:^r$ on \textit{region}
-syntax and then $<:$ on \textit{spec} syntax inductively.
-Both relations are reflexive and transitive (we omit
-these equations) and congruent
-with respect to \term{+}, \term{*}, \term{atLeast}, \term{atMost},
-and \term{readOnce}. The (\trule{\textsc{eq}}) rule connects region
-equations with inequations. The (\trule{\textsc{over}}) rule explains
-the notion of spatial over-approximation via overlapping regions.
-The (\trule{\textsc{rep}}) allows the modifier that
-enforces that every index is read just once to be dropped
-as an approximation; (\trule{\textsc{bound}}) gives
-a connection between the \term{atLeast} and \term{atMost}
-approximation modifiers.
-
-\section{Semantics of specifications; a model}
-\label{sec:semantics}
-
-\newcommand{\domainVal}{\mathbb{Z}_\infty}
-
-We define a denotational model of the semantics of our 
-stencil specification language. This model has a number of purposes: (1) it is used in the inference,
-checking, and synthesis algorithms (Section~\ref{sec:analysis}); (2)
-justifies the equational
-theory and approximations of the previous section; and (3) can be used
-to guide correct implementations. The model domain is over sets of vectors which we call \emph{index
-  schemes}, which we introduce first before the model in Section~\ref{subsec:model}.
-
-\subsection{Indexing schemes and generalisation}
-\label{subsec:schemes-and-gen}
-
-\begin{definition}
-An \emph{index scheme} is a vector of size $n$ (an $n$-vector)
-with values drawn from $\domainVal = \mathbb{Z} \cup \{\infty\}$.
-Values in $\mathbb{Z}$ represent offsets
-of a neighbourhood index (indexing
-expressions of the form $i + a$ where $i$ is an induction variable
-and $a$ is a constant offset, Def.\ref{def:neighbour},
-p.\pageref{def:neighbour}). The additional value
-$\infty$ represents any indexing behaviour--- it is a
-wildcard.
-Throughout $u, v, w$ range over index schemes and $u_i$ denotes
-the $i^{th}$ element of $u$. We write $(\domainVal)^n$ for
-the set of $n$-dimensional schemes.
-
-Indexing schemes give an abstract
-representation of index expressions in the source language.
-Our model represents the meaning of stencil specifications
-as indexing schemes, which explain the range of possible
-indexing behaviours allowed under the specification.
-The relationship between indexing schemes and source language
-terms is given precisely by the partial function, \textsf{schematic}, which
-maps syntactic array-index terms to indexing schemes. This
-operation is used by the inference
-and the checking procedures.
-\end{definition}
-%
-\begin{definition}%[Schematic]
-\label{def:schematic}
-The partial function $\textsf{schematic}_I$ (parameterised by a set of
-induction variables $I$) maps source indexing terms $(e_1, .., e_n)$
-to $n$-dimensional index schemes, defined:
-%
-\begin{align*}
-\mathsf{schematic}_I(e_1, ..., e_n) & =
-[\mathsf{s}_I(e_1) ... \mathsf{s}_I(e_n) ]
-\quad \textit{iff} \;\; \forall x . \mathsf{s}_I(e_x) \neq \bot
-\end{align*}
-%
-where $\mathsf{s}_I$ is partial and 
-maps individual indexing expressions to 
- indexing scheme components in $\domainVal{}$ defined:
-%%
-\begin{align*}
-\mathsf{s}_I(e)
- & =  \begin{cases}
-a & e \equiv i + a \; \wedge \; \neigh_I(e) \\
--a & e \equiv i - a \; \wedge \; \neigh_I(e) \\
-\infty & \textit{$\mathsf{IV}(e) = \emptyset$} \\
-\bot   & \textit{otherwise}
-\end{cases}
-\end{align*}
-%%
-where $\mathsf{IV}(e)$ is the set of free induction variables
-in an expression, and $\neigh_I$ classifies expressions which
-are constant offsets from an induction variable $i \in I$
-(Defn.~\ref{def:neighbour}). In the third case,
-if the expression is not a neighbourhood index but is not
-a relative index either (\ie{}, not defined in terms
-of an induction variable) then it is mapped to $\infty$, otherwise
-$\mathsf{s}_I(e)$ is undefined (returns $\bot$). The $\mathsf{schematic}$ function
-is undefined if $\mathsf{s}_I$ of any of its components $e_i$ is
-undefined. We omit the parameter set of induction variables $I$ to
-$\mathsf{schematic}$ when clear from the context. 
-\end{definition}
-
-\begin{definition}[Generalisation]
-Given two schemes $v, w$ of dimensionality $n$,
-then $w$ is said to \emph{generalise} $v$
-(or conversely $v$ \emph{specialises} $w$),
-written $v \preceq_n w$, iff the following holds:
-%%
-\[
-v \preceq_n w \;\; \Leftrightarrow  \;\;
-  \forall i\!\in\!\{ 1, \ldots, n \} . \, (w_i = \infty \, \vee \, v_i = w_i)
-\]
-%%
-That is, two schemes are related by the (reflexive and transitive) binary relation
-$\preceq_n$ over $(\mathbb{Z}_{\infty})^n$ if in each dimension
-the two schemes are equal or $w_i = \infty$
-in place of some $\mathbb{Z}$ in $v_i$ for $v \preceq_n w$.
-Thus, the element $\infty$ generalises any index in the same dimension.
-%
-%The subscript on $\preceq_n$ is sometimes dropped when the
-%intended dimensionality is clear or inconsequential.
-\end{definition}
-
-%Unification relation, $\sim$, establishes whether
-%a particular relativised subscript, $u$, of a group of subscripts involved
-%in a stencil computation conforms with a unifier, $v$, of the model for
-%the stencil computation.
-
-\begin{example}
-The following three array indexing terms
-$(i, j+1)$, $(1, j+1)$ and $(3, j)$ 
-are mapped by \textsf{schematic} to the 
-index schemes $\vtwoh{0}{1}$, $\vtwoh{\infty}{1}$
-and $\vtwoh{\infty}{0}$ respectively. The following
-relationships of generalisation then holds:
-%\[
-%\xymatrix@R=0.2em@C=2em{
-%[\infty \;\; 0] \ar@{-}[dr]^{\preceq_n} &  \\
-%& [0 \;\; 1]  \\
-%[\infty \;\; 1]  \ar@{-}[ur]_{\preceq_n} &
-%}
-%\]
-%%
-$\vtwoh{\infty}{0} \; \not\preceq_2 \; \vtwoh{0}{1}$
-and
-$\vtwoh{\infty}{1} \; \preceq_2 \; \vtwoh{0}{1}$.
-
-As an example of undefined \textsf{schematic}, 
- $\textsf{schematic}(i*2, j) = \bot$ as its first index is a non-neighbour relative index, which
- is outside the scope of our specification language.
-\end{example}
-\noindent
-Our terminology of \emph{indexing schemes} and
-\emph{generalisation} is by analogy
-with the \emph{type schemes} and \emph{type generalisation} of
-polymorphic types in ML~\cite{milner1978theory}. Our notion is
-much more rigid however, as our schemes do not bind names.
-%and thus generalisation does not require substitution; $\infty$ acts
-%as a positional placeholder.
-
-From \textsf{schematic} we now give a
-formal characterisation of stencil computations for our purposes:
-%%
-\begin{definition}[Stencil computations]
-  Let $a$ be an array of dimensionality $n$ and $\overline{b}$ a
-  collection of arrays of arbitrary (possible differing)
-  dimensionalities.  
-  A \emph{stencil computation} comprises an iteration by a set $I$ of at
-  most $n$ induction variables 
-  over a subset of the index space of $a$. Elements of $a$ are determined by an
-  assignment $a(\bar{e}) = e_r$ where $\bar{e}$ is an array index such
-  that  $\mathsf{schematic}_I(\bar{e}) \neq \bot$ and %for all $b \in  \bar{b}$ and
-  for all subscripts $b(\bar{e}') \in e_r . \, $
-  $\mathsf{schematic}_I(\bar{e}') \neq \bot$. 
-\label{def:stencil}
-\end{definition}
-%%
-
-\subsection{Denotational model}
-\label{subsec:model}
-
-The model is given by the interpretation function $\interp{-}_n$
-(where $n$ is the maximum dimensionality of the specification being
-modelled) mapping closed\footnote{That is, we assume there are no
-  occurrences of \textit{rvar} in a specification being modelled.  Any
-  \emph{open} specification containing region variables can be made
-  closed by straightforward syntactic substitution with a (closed)
-  \textit{region}.} specifications to sets of
-$n$-dimensional indexing schemes. The interpretation is overloaded on
-\emph{regions} in Figure~\ref{fig:region-model} and on the top-level
-of a specification \textit{spec} in Figure~\ref{fig:spatial-model}.
-
-We first define some intermediate notations and definitions.
-
-\begin{definition}A \emph{single-entry vector} of size $n$, denoted
-$\singleEntry{r}{n}$, is a vector where the $r^{th}$ entry is $1$
-and all others are $\infty$, \eg{}, $\singleEntry{1}{2} =
-\vtwoh{1}{\infty}$ and $\singleEntry{2}{3} = \vthreeh{\infty}{1}{\infty}$.
-Thus, a single-entry vector is an indexing scheme which generalises
-all other schemes which at least have $1$ as their $r^{th}$ entry,
-describing a neighbour offset of $1$.
-%DAO don't understand this sentence:
-%In particular the constrained
-%requires a non-zero offset from the induction variable.
-\end{definition}
-
-\begin{definition}A \emph{zero-entry vector} of size $n$, denoted
-$\zeroEntry{r}{n}$, is a vector where the $r^{th}$ entry is $0$ and all others
-are $\infty$, \eg{}, $\zeroEntry{1}{2} = \vtwoh{0}{\infty}$.
-Similarly to the single-entry vector, zero-entry vectors
-are generalise all other schemes
-which at least have the $r^{th}$ entry is $0$, representing indices at
-the ``origin'' in dimension $r$.
-\end{definition}
-
-\begin{figure}[t]
-\vspace{-0.9em}
-\input{region-model.tex}
-%\textit{$a^m \in A$ means there are $n$ copies of $a$ in
-%  the multi-set $A$}
-\caption{Model of regions,
-$\interp{-}_n : \textit{region} \rightarrow \mathcal{P}(\domainVal{}^n)$}
-\label{fig:region-model}
-\vspace{-1em}
-\end{figure}
-
-\noindent
-The first four equations of Fig.~\ref{fig:region-model}
-give the model of the region constants as sets of indexing
-schemes. We illustrate with an example.
-
-\begin{example}
-For a $2$-dimensional stencil computation, then
-\begin{align*}
-& \interp{\stenCenS{2}{1}}_2 \\[-0.1em]
-& \qquad = \; \{i\singleEntry{1}{2} \,|\, i \in \{\text{-}2,..., 2\}\!\setminus\!0 \}
-  \cup  \{ \zeroEntry{1}{2} \,|\, \textit{true} \} \\[-0.2em]
-& \qquad = \; \{-2\singleEntry{1}{2}, \, -\singleEntry{1}{2}, \,
-  \singleEntry{1}{2}, \, 2\singleEntry{1}{2}\} \cup
-  \{\zeroEntry{1}{2}\} \\[-0.2em]
-& \qquad = \; \{\vtwoh{-2}{\infty}, \vtwoh{-1}{\infty}, \vtwoh{0}{\infty},
-\vtwoh{1}{\infty}, \vtwoh{2}{\infty}\}
-\end{align*}
-Note the absorbing behaviour of $\infty$ with respect to
-multiplication.
-\end{example}
-\noindent
-The final two equations of Fig.~\ref{fig:region-model} give the
-models of $\term{+}$ and $\term{*}$. For $\term{+}$ this is the 
-straightforward union of the models of subterms. The model of
-$\term{*}$ takes the tensor $\otimes$ of subterm models. The tensor is
-non-trivial; we give some informal intuition first.
-
-\paragraph{$\otimes$ of models}
-
-The intuition behind $\otimes$ is that it takes
-all possible pairs of indexing schemes, treating them as the
-lower and upper bounds of an $n$-dimensional rectangle from which the
-remaining vertices of the rectangle are generated (like a bounding
-box) \eg{}:
-%%
-\begin{align*}
-\{\vtwohs{1}{2}, \vtwohs{5}{6}\,\} \otimes \{\vtwohs{3}{4}\} =
-\{\vtwohs{1}{2}, \vtwohs{3}{2}, \vtwohs{1}{4}, \vtwohs{3}{4}, 
-   \vtwohs{3}{6}, \vtwohs{5}{4}, \vtwohs{5}{6}\,\}
-\end{align*}
-%%
-The tensor takes a Cartesian product of the two models,
-giving all pairs of schemes which are combined using the
-\emph{pairwise permutation} written $\bowtie$ (defined later).
-Pairwise permutation generates all possible interleavings of two vectors
-(\ie{}, all possible non-deterministic choices between $u_i$ and
-$v_i$ in each dimension). Some care is then taken over
-how this interacts with $\infty$: pairwise permutation
-corresponds to non-deterministic choice, in each dimension $i$,
-between two values $u_i$ and $v_i$, but if either
-is $\infty$ then the resulting pairwise permutation is
-filtered out if the dimension $i$ has \emph{any other scheme
-in the model where $u_i \neq \infty$}, \eg{}:
-\begin{align*}
-\{\vtwoh{1}{\infty}\} \otimes \{\vtwoh{3}{4}\} =
-\{\vtwoh{1}{4}, \vtwoh{3}{4}\}
-\end{align*}
-%
-where $\vtwoh{1}{\infty}$ and $\vtwoh{3}{\infty}$ have been filtered.
-%With respect to the notion of unification, the
-%
-%
-% combines non-$\infty$ constraints
-%of the unifiers in different dimensions into less accepting unifiers.
-%This means an access pattern generalises with this scheme would require
-%both models to be satisfied. Use of $dims$ gives the dimensions that
-%constituent models talk about and is used in the product to eliminate
-%general unifier appearing in these dimensions. Allowing general unifier
-%in these dimensions would prevent both models to be satisfied
-%simultaneously.
-%  u in M * N   u < w
-% implies that
-%  u_1 in M  u1 < w
-%  u_2 in M  u2 < w
-%
-The notion of which dimensions $i$ have that \emph{any scheme in the model
-has $u_i \neq \infty$} is represented by the \emph{constrained
-dimensions} intermediate function:
-%
-\begin{definition}%[Constrained dimensions]
-For an $n$-dimensional model $M$, its \emph{constrained dimensions}
-is the set of dimension identifiers for which at least one scheme $u \in M$ has
-a non-$\infty$ value in that dimension. That is:
-%
-\begin{align*}
-\effdims{M}{n} =
-\bigcup_{u \in M} \{i \mid i \in \{1,\ldots,n\} \, \wedge \, u_i \neq
-  \infty\}
-\end{align*}
-For example, $\effdims{\{\vtwoh{0}{\infty},\vtwoh{\infty}{\infty}\}}{2}
- = \{1\}$.
-\end{definition}
-
-\begin{definition}The tensor $\otimes_n$ of $n$-dimensional
-models is defined:
-\begin{alignat*}{2}
-& M \otimes_n N =
-  \Bigg\{x \; \Bigg| \;
-    \parbox{5.2cm}{$(u, v) \in M \times N $ \\[0.1em]
-                 $\;\; \wedge \;\, j \in (\effdims{M}{n} \cup
-                   \effdims{N}{n})$\\[0.1em]
-                  $\;\; \wedge \;\, i \in \{1, \ldots, 2^n\} \, \wedge
-                  \, x = (u \bowtie v)_i$ \\[0.1em]
-                   $\;\; \wedge \;\, x_j \neq \infty$
-                  } \Bigg\}
-\end{alignat*}
-The first guard takes the Cartesian product of
-the models and quantifies over all pairings $(u, v)$.
-The second guard binds $j$ which ranges over the constrained
-dimensions for $M$ union with the constrained dimensions for $N$.
-That is, every $j$ corresponds to a dimension in which
-either $M$ or $N$ has a scheme $u$ where $u_j \neq \infty$,
-and thus is constrained.
-
-The \emph{pairwise permutation} $u \bowtie v$ 
-builds a $2^n \times n$ matrix 
-of all possible $n$-vectors generated by
-non-deterministically picking for each $j^{th}$ entry either 
-$u_j$ or $v_j$ \eg{}
-%
-\begin{equation*}
-\vthreeh{0}{1}{2} \bowtie \vthreeh{3}{4}{5} =
-\left[
-\setlength{\arraycolsep}{0.42em}
-\begin{array}{cccccccc}
-0 & 0 & 0 & 0 & 3 & 3 & 3 & 3 \\
-1 & 1 & 4 & 4 & 1 & 1 & 4 & 4 \\
-2 & 5 & 2 & 5 & 2 & 5 & 2 & 5
-\end{array}
-\right]
-% \\
-%\setlength{\arraycolsep}{0.5em}
-%\left[\begin{array}{ccc}
-%0 & 1 & 2 \\
-%0 & 1 & 5 \\
-%0 & 4 & 2 \\
-%0 & 4 & 5 \\
-%3 & 1 & 2 \\
-%3 & 1 & 5 \\
-%3 & 4 & 2 \\
-%3 & 4 & 5
-%\end{array}
-%\right]
-\end{equation*}
-%
-The $2^n$ unique choices for $\bowtie$ on $n$-vectors
-corresponds to taking all bit-strings of length $n$ and
-selecting from $u$ for $1$ and $v$ for $0$. The pairwise permutation
-matrix is thus defined in terms of the logical
-matrix $b$ where $b_{i,j}$ is the $j$-th bit of the integer
-$i - 1$ as:
-%
-\begin{align*}
-(u \bowtie v)_{i,j} & = b_{i,j} u_j + \neg b_{i,j} v_j
-\end{align*}
-%
-The permuted schemes $x = (u \bowtie v)_i$ added to the model 
-$M \otimes N$ are only those which have do not have $\infty$ in 
-the constrained dimensions given by each model ($x_j \neq
-\infty$). This filters out all permutations which have
-$\infty$ in a constrained dimension position. 
-\end{definition}
-
-%The main interpretation function is overloaded on lists of
-%variable-spec pairs $\interp{\overline{v : S}}$,
-%returning multisets of variable-relative-index pairs. This provides
-%the top-level definition of the model, with $\interp{-}$ overloaded
-%on $S$, $\interp{S}$ mapping to multisets of
-%relative indices not associated to an array variable.
-
-\paragraph{Model of \textit{spec}}
-
-We have explained the model of regions as sets of index
-schemes, that is
-$\interp{\textit{region}}_n \in \mathcal{P}(\domainVal{}^n)$.
-Figure~\ref{fig:spatial-model} defines the top-level model of the
-\textit{spec} syntax, which includes the multiplicity and
-approximation modifiers. The domain of this interpretation
-is $\textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$
-where $\textsf{Mult}$ and $\textsf{Approx}$ are labelled variants
-labelling our set-based model with additional information.
-
-\begin{definition} \textsf{Mult}
-and \textsf{Approx} are parametric labelled variant types
-with injection function given by their definition:
-%
-\begin{align*}
-\textsf{Mult} \, a & = \textsf{mult} \, a \,\mid\, \textsf{only} \, a \\
-\textsf{Approx} \, a & = \textsf{exact} \, a \,\mid\, \textsf{lower} \, a \,\mid\,
-\textsf{upper} \, a \,\mid\, \textsf{both} \, a \, a
-\end{align*}
-\ie{}, $\mathsf{lower}$ is an injection $\mathsf{lower} : a \rightarrow \mathsf{Approx} \, a$, and so
-on. The $\textsf{Mult}$ type corresponds to the presence or absence of the
-\term{readOnce} modifier as shown in Figure~\ref{fig:spatial-model}. 
-The $\textsf{Approx}$ type corresponds to the presence or absence of
-the spatial approximation modifier, with \textsf{exact} when there is
-no such modifier and \textsf{lower} and \textsf{upper} for
-\term{atLeast} and \term{atMost}. Later, it will be useful to model
-pairs of lower and upper bounds which is provided by \textsf{both} 
-(but which isn't used here). These are used for labelled models, and other components later in
-Section~\ref{sec:analysis}.% with additional information.
-\label{def:mult-and-approx}
-\end{definition}
-\vspace{-0.5em}
-
-\begin{figure}
-\input{spatial-model.tex}
-\caption{\textit{spec} model,
-$\interp{-}_n : \textit{spec} \rightarrow \textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$}
-\label{fig:spatial-model}
-\vspace{-0.75em}
-\end{figure}
-
-\subsection{Soundness}
-%
-Given the above model, we can
-consider soundness for the equational theory of
-specifications and the notion of spatial approximation. The proofs are given in our supplement.
-
-\begin{theorem}[Soundness of equational theory (Figure~\ref{fig:equations})]
-\[
-S \equiv T \; \Rightarrow \;
-\interp{S}_n = \interp{T}_n
-\]
-\end{theorem}
-
-\begin{theorem}[Soundness of inequations
-  (Figure~\ref{fig:inequations})]
-
-Let $\leq$ be an ordering on
-$\textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$
-and $\leq'$ be an ordering on
-$\textsf{Approx}(\mathcal{P}(\domainVal{}^n))$ defined by the
-following clauses:
-\begin{align*}
-\def\arraystretch{1.2}
-\begin{array}{c}
-\begin{array}{cc}
-(M \leq' N \Rightarrow \textit{inj} \, M \leq \textit{inj} \, N) &
-(\textsf{once} \, M \leq' \textsf{mult} \, M) \\
-%%
-%%
-(M \subseteq N \Rightarrow \textit{inj}' M \leq' \textit{inj}' N)
-& (\textsf{exact} M \leq  \textsf{lower} M)
-\end{array} \\ 
-(\textsf{lower} M \leq  \textsf{upper} M)
-\end{array}
-\end{align*}
-%%
-where $\textit{inj}$ ranges over the unary injections
-of $\textsf{Approx}$ and $\textit{inj}'$ ranges over the injections of
-$\textsf{Mult}$. Then, soundness of $<:$ is that:
-%%
-\[
-S <: T
-\; \Rightarrow \;
-\interp{S}_n \leq \interp{T}_n
-\]
-\end{theorem}
-
-\section{Inference, checking, and synthesis}
-\label{sec:analysis}
-
-\noindent
-We briefly sketch the algorithms for inferring and synthesising
-specifications (Section~\ref{subsec:inference}), for checking code
-against specifications (Section~\ref{subsec:checking}). These utilise the semantic model of
-Section~\ref{sec:semantics}.
-
-\paragraph{Syntax helper functions}
-The function \rhsExp{} maps source statements to a
-set of expressions which occur in right-hand positions (\ie{}, not the
-target of an assignment). The function \var{} maps expressions to a
-set of the variables used in right-hand positions.
-
-\subsection{Inference and specification synthesis\label{subsec:inference}}
-
-We demonstrate the inference procedure over the
-following example program which computes the mean value
-of a five-point stencil:%at each index of the input array:
-%%
-\begin{minted}{fortran}
-do i = 1, (n-1)
-  do j = 1, (m-1)
-    x = a(i-1, j)+a(i+1, j); y = a(i, j-1)+a(i, j+1)
-    b(i, j) = (a(i, j) + x + y) / 5.0
-  end do
-end do
-\end{minted}
-%%
-\subsubsection{Step 1: Analyse standard control and data-flow}
-\label{sec:inf-step1}
-
-The inference relies on some standard program analyses, computed
-before the main inference procedure:
-%
-\textbf{(1)} basic blocks (CFG);
-\textbf{(2)} induction variables per basic block;
-\textbf{(3)} (interprocedural) data-flow analysis, providing a \emph{flows to}
-  graph (as shorthand, the function
-  $\mathsf{flowTo}$ is used, implicitly parameterised by this graph,
-  mapping an expression to the set of all expressions
-  with forwards data-flow to this expression, based on the transitive
-  closure of the flows graph);
-\textbf{(4)} type information per variable, where we use the predicate
-\arrayTy{} to classify variables of array type.
-%
-
-\subsubsection{Step 2: Analyse data-access}
-\label{sec:inf-step2}
-
-For each assignment statement whose left-hand side is an array
-subscript on neighbourhood indices, a finite map is computed which
-maps array variables to a set of vectors representing array
-subscripts.  This finite map contains all array subscript expressions
-which flow to this statement. More formally, a function
-$\textit{analyse}$ is applied to each statement in a program with the
-following clause:
-%
-\begin{align*}
-\hspace{-0.5em}
-\begin{array}{lr}
-\textit{analyse}(v(\overline{e_1}) = e_2)
- := \qquad\qquad & \textit{where} \; \neigh(\overline{e_1}) \wedge \arrayTy(v)  \\[0.3em]
-\multicolumn{2}{l}{\qquad \bigcup\{v' \mapsto \{
-\textsf{schematic}(\bar{e})\} \mid v'(\bar{e}) \leftarrow \mathsf{flowsTo}(e_2),
-  \arrayTy(v')\}}
-\end{array}
-\end{align*}
-%
-That is, we focus on assignments to an array subscript where the LHS 
-indexing expression $\overline{e_1}$ is a neighbourhood index.  For
-all array subscripts that flow to the right-hand side of this
-statement, a finite map is constructed, mapping each array variable
-to a set of indexing schemes for its subscripts, computed
-with \textsf{schematic} (Defn.~\ref{def:schematic},
-p.~\pageref{def:schematic}).
- Note that \textsf{schematic} is undefined if
-$\bar{e}$ contains relative indices which are not neighbourhood
-indices (\ie{}, these are not stencil computations we can handle).
-
-For our example, the \textit{analyse} function matches on
-line 5, with the following set for $\textsf{flowsTo}(\texttt{a(i, j) + x +
-  y})$:
-%
-\begin{align*}
-\{\texttt{a(i-1, j)}, \texttt{a(i+1, j)}, \texttt{a(i, j-1)},
-  \texttt{a(i, j+1)}, \texttt{a(i, j)}\}
-\end{align*}
-Subsequently the result of \textit{analyse} on line 5 yields the map:
-\begin{align*}
-\texttt{a} \mapsto \{\vtwoh{-1}{0}, \vtwoh{1}{0},
-          \vtwoh{0}{-1}, \vtwoh{0}{1}, \vtwoh{0}{0}\}
-\end{align*}
-%
-If the LHS $\overline{e_1}$ contains non-$0$ neighbour offsets 
-then all RHS schemes are \emph{relativised} by
-$\textsf{schematic}(\overline{e_1})$. For example, 
-if \texttt{a(i+1) = b(i)} then relativisation of the RHS by the LHS
-produces an analysis which is equivalent to the analysis of \texttt{a(i) = b(i-1)}.
-
-\subsubsection{Step 3: Coalesce contiguous schemes into spans}
-\label{sec:inf-step3}
-
-Let $U$ range over the finite maps computed by \textit{analyse}.  For
-each $x \in \mathsf{dom}(U)$, the algorithm 
-constructs next a set of $n$-dimensional rectangles covering all contiguous groups of schemes
-in $U(x)$.
-
-\newcommand{\spanOp}{\textsf{spans}}
-
-\begin{definition}[Spans]
-  A \emph{span} represents an $n$-dimensional box
-  (\emph{hyper-rectangle}) as a pair of $n$-dimensional vectors (or
-  represented as a $2 \times n$ matrix) with values drawn from
-  $\domainVal{}$. This gives the co-ordinates of a lower-bound vertex
-  (first component) and the upper-bound (second component). We write 
-  spans as $\vect{x} = [\vect{x}^L \; \vect{x}^U]$, where $\vect{x}^L$
-  and $\vect{x}^U$ give the lower and upper bounds.  In
-  the following, $\vect{x}, \vect{y}$ range over spans.  For a span
-  $\vect{x}$, we write $\vect{x}_{i:j}$ for the sub-span comprises the
-  subvectors $[\vect{x}^L_{i:j} \; \vect{x}^U_{i:j}]$ from entry $i$
-  to entry $j$ (inclusive).
-\end{definition}
-
-\noindent
-Each indexing scheme in $U(x)$ is converted to a \emph{unit} span $u \mapsto [u \; 
-u]$, defining the lower and upper bound.
- For the five-point stencil example there are five $1 \times 1$ spans.
-Figure~\ref{fig:inference-steps-informal}(a) illustrates these.
-
-% taking permutations is a systematic way of producing all rotations
-% of the n-dimensional space - forward ref to fold
-The next step repeatedly coalesces any contiguous spans
-until a fixed point is reached, resulting in a set of (possibly
-overlapping) spans covering the original space of schemes. Two $n$-dimensional spans $\vect{x}$ and $\vect{y}$
-are contiguous if for all but one $j \in \{1, \ldots, n\}$ then
-$\vect{x}_j = \vect{y}_j$ and for one $j$ then $\vect{x}^U_1 + 1 =
-\vect{y}^U$. For example, spans
-$\left[\vtwoh{-1}{0}\; \vtwoh{-1}{0}\right]$ and $[\vtwoh{0}{0} \; \vtwoh{1}{0}]$ 
-are contiguous in the first dimension and can be coalesced
-into the span $[\vtwoh{-1}{0} \; \vtwoh{1}{0}]$.
-
-\begin{definition}\label{def:span-coalesc} Span
-  coalescing $\bullet$ is defined for spans contiguous in the first dimension:
-%%
-\begin{align*}
-\vect{x} \bullet \vect{y}
-= &
-\begin{cases}
-(\vect{x}^L, \vect{y}^U) & \vect{x}^U_1 + 1 = \vect{y}^L_1 \; \wedge \;
-\vect{x}_{2:n} = \vect{y}_{2:n}) \\
-\bot  & \textit{otherwise}
-\end{cases}
-% $m$-dimensions into coalesced spans which vary in
-% $m+1$-dimensions.
-\end{align*}
-%%
-\end{definition}
-\noindent
-To coalesce regions which are contiguous in other dimensions
-other than the first, we generate all rotations of the spans in
-$n$-dimensional space by permutation on spans, and fold together contiguous
-regions using $\bullet$ (above). This 
-generates all possible ways of coalescing regions. For our example,
-the unit spans in Figure~\ref{fig:inference-steps-informal}(a)
-are coalesced into the contiguous spans in
-Figure~\ref{fig:inference-steps-informal}(b).
-This procedure is iterated until a fixed point is reached. In this
-example this is reached in the first step. 
- Then any spans that are completely contained by any other span are deleted,
-leaving a minimal set of spans (which may overlap, but none of which
-fully contains another), Figure~\ref{fig:inference-steps-informal}(c).
-The final spans are used to synthesise concrete stencil specification
-syntax (Section~\ref{sec:inf-step4}), where
-each span becomes a product \texttt{*} of regions which are 
-combined by the sum \texttt{+}.
-
-More formally the algorithm proceeds as follows.
-For all $x \in \mathsf{dom}(U)$, let $M = U(x)$. Every scheme in $M$ 
-is mapped to a unit span
-by $\textsf{unitSpans}(M) = \{[u \; u] \mid u\!\in\!M\}$. For our example:
-%%
-\begin{align*}
-N(U(\texttt{a})) = \stwo{0}{0}{0}{0} \stwo{1}{0}{1}{0} \stwo{-1}{0}{-1}{0} \stwo{0}{1}{0}{1} \stwo{0}{-1}{0}{-1}
-\end{align*}
-%%
-In the following we let $M_0 = N(U(\texttt{a}))$ for our example.
-
-Each set of unit spans is applied to the fixed point of the \spanOp{} function
-(defined shortly), which coalesces spans into contiguous regions. That is, we compute
- $(\mu \, \spanOp)$ on input $N(M)$, where \spanOp{} is defined by the steps:
-
-\begin{figure}[t]
-\vspace{-0.8em}
-\hspace{-0.5em}
-\begin{minipage}{0.16\linewidth}
-\begin{align*}
-\setlength{\arraycolsep}{0em}
-\begin{array}{c}
-\quad \\[0em]
-\begin{tikzpicture}[scale=0.8]
-%\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
-\node at (1.25,1.22) {\textnormal{\footnotesize{0,0}}};
-\draw[thick] (1,1) rectangle (1.5,1.5);
-\draw[thick] (1.5,1) rectangle (2,1.5);
-\draw[thick] (1,0.5) rectangle (1.5,1);
-\draw[thick] (0.5,1) rectangle (1,1.5);
-\draw[thick] (1,1.5) rectangle (1.5,2);
-\end{tikzpicture} \\[0em]
-(a)
-\end{array}
-\end{align*}
-\end{minipage} $\rightarrow$ \begin{minipage}{0.43\linewidth}
-\begin{align*}
-\setlength{\arraycolsep}{0em}
-\begin{array}{c}
-\begin{tikzpicture}[scale=0.72]
-%\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
-%\node at (1.25,1.22) {\textnormal{\small{0,0}}};
-\draw [black, thin] (0.35,0) to [round left paren ] (0.35,2);
-\draw[thick] (0.5,1) rectangle (2,1.5);
-\draw[thick] (1,0.40) rectangle (1.5,0.90);
-\draw[thick] (1,1.60) rectangle (1.5,2.10);
-\node at (2.3,1.22) {+};
-\draw[thick] (3.8,1) rectangle (4.3,1.5);
-\draw[thick] (2.6,1) rectangle (3.1,1.5);
-\draw[thick] (3.2,0.5) rectangle (3.7,2);
-\draw [black, thin] (4.4,0) to [round right paren ] (4.4,2);
-\end{tikzpicture} \\[-0.5em] (b)
-\end{array}
-\end{align*}
-\end{minipage}$\rightarrow$ \begin{minipage}{0.25\linewidth}
-\begin{align*}
-\setlength{\arraycolsep}{0em}
-\begin{array}{c}
-\begin{tikzpicture}[scale=0.72]
-%\node at (1.25,1.22) {\textnormal{\small{0,0}}};
-\draw [black, thin] (0.8,0) to [round left paren ] (0.8,2);
-\draw[thick] (1,0.5) rectangle (1.5,2);
-\node at (1.8,1.22) {+};
-\draw[thick] (2.1,1) rectangle (3.6,1.5);
-\draw [black, thin] (3.7,0) to [round right paren ] (3.7,2);
-%\node at (2.75,1.22) {\textnormal{\small{0,0}}};
-\end{tikzpicture} \\[-0.5em] (c)
-\end{array}
-\end{align*}
-\end{minipage}
-\caption{Illustration of grouping spans in
-specification inference}
-\label{fig:inference-steps-informal}
-\vspace{-0.8em}
-\end{figure}
-
-\noindent
-\textbf{1)\,} Compute all permutations on the column vectors in a span, \ie{},
-  $[\vect{x}^L \vect{x}^U] \mapsto [\pi\vect{x}^L \, \pi\vect{x}^U]$
-for a permutation $\pi$. For each permutation function $\pi^n_i$
-(the $i$-th permutation for vectors of size $n$) we pair the
-permutation function with the set of permuted spans so that
-the spans can be un-permuted later.
-%
-\begin{equation*}
-P(M) = \bigcup_{i \in n!} (\pi^n_{i} , \; \{[\pi^n_i
-\vect{x}^L \; \pi^n_i\vect{x}^U] \, \mid \, [\vect{x}^L \; \vect{x}^U]
-\leftarrow M\}
-\end{equation*}
-%
-This corresponds to all rotations of the space.  For our example:
-%
-\begin{align*}
-P(M_0) =
-\{&(\pi^2_1, \{ \stwo{0}{0}{0}{0},
-\stwo{1}{0}{1}{0},
-\stwo{-1}{0}{-1}{0},
-\stwo{0}{1}{0}{1},
-\stwo{0}{-1}{0}{-1} \})
-\\
-&(\pi^2_2, \{
- \stwo{0}{0}{0}{0},
- \stwo{0}{1}{0}{1},
- \stwo{0}{-1}{0}{-1},
- \stwo{1}{0}{1}{0},
- \stwo{-1}{0}{-1}{0}\})\}
-\end{align*}
-%
-where $\pi^2_1$ is the identity permutation and $\pi^2_2$ is the
-permutation flips the order of the two elements in the 
-vectors. \\[-1em]
-
-\noindent
-\textbf{2)\,} Sort each permutation set into an ordered list, by the ordering:
-\begin{equation*}
-  \vect{x} \leq \vect{y} =
-      \exists i \, . \; \vect{x}^L_{i} \leq \vect{y}^L_{i} \; \wedge \;
-        (i = n \vee \vect{x}_{i+1:n} = \vect{y}_{i+1:n})
-\end{equation*}
-%
-giving the sorting function $\textsf{sort}(M)$. For our example:
-%%
-\begin{align*}
-\textsf{sort}(P(M_0)) = 
-\{&(\pi^2_1, [
-\stwo{0}{-1}{0}{-1},
-\stwo{-1}{0}{-1}{0},
-\stwo{0}{0}{0}{0},
-\stwo{1}{0}{1}{0},
-\stwo{0}{1}{0}{1}] )
-\\
-&(\pi^2_2, [
-\stwo{0}{-1}{0}{-1},
-\stwo{-1}{0}{-1}{0},
-\stwo{0}{0}{0}{0},
-\stwo{1}{0}{1}{0},
-\stwo{0}{1}{0}{1}])\}
-\end{align*}
-\noindent
-\textbf{3)\,} Reduce each list pairwise by $\bullet$ operation 
-(Defn.~\ref{def:span-coalesc}) to coalesce contiguous regions,
-given by $\textsf{fold}\bullet$. For our example, this yields:
-%
-\begin{align*}
-\{&(\pi^2_1, [
-\stwo{0}{-1}{0}{-1},
-\stwo{-1}{0}{1}{0},
-\stwo{0}{1}{0}{1}
-]), (\pi^2_2, [
-\stwo{0}{-1}{0}{-1},
-\stwo{-1}{0}{1}{0},
-\stwo{0}{1}{0}{1}])\}
-\end{align*}
-%
-\textbf{4)\,} Un-permute and union together, \ie{},
-%
-\[
-U(M) = \bigcup \{[\pi \vect{x}^L, \pi \vect{x}^U]
- \mid [\vect{x}^L, \vect{x}^U] \leftarrow S, (\pi, S) \leftarrow M\}
-\]
-For our example:
-%
-\begin{align*}
-U(M) =
-\{\stwo{0}{-1}{0}{-1},
-\stwo{-1}{0}{1}{0},
-\stwo{0}{1}{0}{1},
-\stwo{-1}{0}{-1}{0},
-\stwo{0}{-1}{0}{1},
-\stwo{1}{0}{1}{0}\}
-\end{align*}
-%
-\textbf{5)\,} Filter by the region containment predicate $\containedin$, that
-  is, if any region is contained within another then remove the
-  smaller, defined:
-%
-\begin{align*}
-& \vect{x} \containedin \vect{y} = \vect{y}^L_1 \leq \vect{x}^L_1 \wedge \vect{x}^U_1 \leq \vect{y}^U_1
-  \wedge (\vect{x}^L_{2:n}, \vect{x}^U_{2:n}) \containedin
-  (\vect{y}^L_{2:n}, \vect{y}^U_{2:n})
-\end{align*}
-For our example this then yields the final result of:
-\begin{align*}
-\spanOp(M)
-= \{\stwo{-1}{0}{1}{0} \stwo{0}{-1}{0}{1}\}
-\end{align*}
-since $\stwo{0}{1}{0}{1} \sqsubseteq \stwo{0}{-1}{0}{1}$
-and $\stwo{1}{0}{1}{0} \sqsubseteq \stwo{-1}{0}{1}{0}$.
-Applying \spanOp{} again yields the same
-result for our example; the fixed point is reached.
-
-\subsubsection{Step 4: Synthesise specifications from spans}
-\label{sec:inf-step4}
-
-\newcommand{\oplusbig}{\operatornamewithlimits{\term{\Large{$\oplus$}}}}
-\newcommand{\odobig}{\operatornamewithlimits{\term{\Large{$\odot$}}}}
-\newcommand{\bplus}{\operatornamewithlimits{\term{\Large{+}}}}
-\newcommand{\tySum}[1]{#1^{\term{+}}}
-\newcommand{\tyProd}[1]{#1^{\term{*}}}
-\newcommand{\specDNF}{\textbf{spec}}
-
-Abstract syntax trees of the specification language
-are synthesised from the set $\spanOp(M)$ of coverings spans. 
-This conversion is defined in terms
-of an algebra on \textit{spec} syntax in disjunctive-normal form
-(which we denote \specDNF{}) and which is wrapped
-in the $\textsf{Approx}$ type (see
-Defn.~\ref{def:mult-and-approx},
-p.~\pageref{def:mult-and-approx}), which we write as $\mathsf{A}$
-for brevity.  The algebra mirrors the shape
-of the region combinators for sum and product, 
-with operators $\oplus$ and $\odot$ and a constructor for \specDNF{}:
-%%
-\begin{align*}
-\oplus & : \mathsf{A}(\specDNF) \rightarrow \mathsf{A}(\specDNF) \rightarrow
-  \mathsf{A}(\specDNF) \\
-\odot & : \mathsf{A}(\specDNF) \rightarrow \mathsf{A}(\specDNF) \rightarrow
-        \mathsf{A}(\specDNF) \\
-\mathsf{toSpec} & : (\mathbb{Z} \cup \{\infty\}) \rightarrow (\mathbb{Z} \cup
-  \{\infty\})  \rightarrow \mathsf{A}(\specDNF)
-\end{align*}
-%%
-The $\mathsf{toSpec}$ function maps a pair of lower and upper bound
-values (one-dimensional) to a specification with a one-dimensional region. 
-An entire span is converted into an $n$-dimensional
-specification by applying $\mathsf{toSpec}$ to each
-pair of lower and upper bounds in a span, per dimension, and multiplying these
-specifications by $\odot$. The resulting specifications for
-each span are then summed by $\oplus$:
-\begin{equation*}
-\textsf{spec}(M) = 
-\oplusbig\limits_{\vect{x} \in \spanOp(M)} \,
-\odobig\limits_{i \in \{1, \ldots, n\}} \, \mathsf{toSpec}(\vect{x}^L_i, \vect{x}^U_i)
-\end{equation*}
-We unpack the definitions of this algebra.
-
-\begin{definition}[Specifications in DNF]
-Let $r_1, ..., r_n$ range over elements of the \textit{region}
-syntax (see Figure~\ref{fig:syntax}, p.~\pageref{fig:syntax}).
-The set \specDNF{} is a subset of syntax trees \textit{spec},
-containing specifications in disjunctive normal form over regions 
-where the disjunct is \term{+} (since its model is in terms of set
-union) and the conjunct is \term{*} (since its model is a form of tensor).
-That is elements of \specDNF{} are of the form:
-%%
-$
-(r^1_{1} \; \term{*} \ldots \term{*} \; r^1_n)\; \term{+} \; \ldots 
-\term{+} \; (r^m_1 \, \term{*} \; \ldots \term{*} \; r^m_p)
-$.
-
-More formally, $\specDNF \subseteq (\textit{spec} \cup \{\epsilon\})$, defined:
-%%
-\begin{align*}
-\specDNF{} & ::= \tySum{\specDNF} \mid \epsilon \\
-\tySum{\specDNF} & ::= \tyProd{\specDNF} \; \term{+} \; \tySum{\specDNF} \mid
-  \tyProd{\specDNF} \\
-\tyProd{\specDNF} & ::= \textit{region} \; \term{*} \; \tyProd{\specDNF} \mid
-   \textit{region}
-\end{align*}
-%%
-We include the empty element $\epsilon$ which is not present in the
-syntax. 
-In the following, $S, T$ range over elements of
-\specDNF{} and $\tyProd{s}, \tyProd{t}$ range over elements
-of $\tyProd{\specDNF}$ (for example, $\tyProd{s} = r^1_{1} \; \term{*}
-\ldots \term{*} \; r^1_n$).
-\end{definition}
-
-\newcommand{\exactR}{\mathsf{exact}\;}
-\newcommand{\exact}[1]{\exactR(#1)}
-\newcommand{\upperB}[1]{\mathsf{upper}\;(#1)}
-
-\begin{figure}[t]
-%\newcommand{\llower}{\vect{x}^L_i}
-%\newcommand{\lupper}{\vect{x}^U_i}
-\vspace{-0.7em}
-\newcommand{\llower}{l}
-\newcommand{\lupper}{u}
-{\scalebox{0.97}{
-\begin{minipage}{1\linewidth}
-\begin{align*}
-&\mathsf{toSpec} \; \llower \; \lupper = \\
-&\hspace{-0.4em}\begin{cases}
-%% FWD
-\exact{\stenFwdS{\lupper}{i}}  & \llower = 0 \, \wedge \, \lupper > 0 \\[0.2em]
-\exact{\stenFwdSR{\lupper}{i}{\irreflS}}\!\!\!\! & \llower = 1 \, \wedge \, \lupper > 0 \\[0.2em]
-%% BWD
-\exact{\stenBwdS{|\llower|}{i}}  & \llower < 0 \, \wedge \, \lupper = 0 \\[0.2em]
-\exact{\stenBwdSR{|\llower|}{i}{\irreflS}}\!\!\!\! & \llower < 0 \, \wedge \, \lupper = -1 \\[0.2em]
-%% CEN
-\exact{\stenCenS{\lupper}{i}}& \llower < 0 \wedge \lupper > 0 \wedge |\llower| = \lupper \\[0.2em]
-\exactR(\stenBwdS{|\llower|}{i} & \llower < 0 \wedge \lupper > 0  \wedge |\llower| \neq \lupper \\
-\qquad\qquad \texttt{+} \, \stenFwdS{\lupper}{i}) & \\[0.2em]
-% BOUND
-\upperB{\stenFwdS{\lupper}{i}} & \llower > 1 \\[0.2em]
-\upperB{\stenBwdS{|\llower|}{i}} \quad & \llower < (-1)
- \\[0.2em]
-% ABSOLUTE REP
-\exact{\epsilon} & \llower = \infty \, \vee \, \lupper = \infty
-  \end{cases}
-\end{align*}
-\end{minipage}
-}}
-\caption{\textsf{toSpec} constructor of $\mathsf{A}(\specDNF)$ values.}
-\label{fig:to-spec}
-\vspace{-0.8em}
-\end{figure}
-
-\begin{definition} The $\mathsf{toSpec}$ constructor is defined
-in Figure~\ref{fig:to-spec}.
-The first two cases map to \texttt{forward} stencils since the lower
-bound is $0$ or $1$, with the depth determined by the upper bound of
-the span. If the lower bound is $1$, then the
-$0$-offset point is missing so the region is given an
-\texttt{irreflexive} attribute. The second two cases are similar for
-\texttt{backward} but now the depth of the region is determined by the
-absolute of the lower bound. In the fifth-case, where the lower bound
-is negative and the upper bound is positive and they have the same
-magnitude, then a \texttt{centered} region is constructed with
-the depth as the common magnitude. If the magnitudes are not the same
-(sixth case) then separated \texttt{forward} and \texttt{backward}
-regions are summed, with their corresponding depths.
-
-If the lower bound is greater than 1
-(which implies the upper bound is greater than 1 too) then
-this represents a region that is not at the origin (or next to) the
-origin and thus cannot be represented in the specification language.
-Thus, an upper bound is instead returned showing the maximum extent
-of the region. The penultimate case is the dual, when the lower bound is less
-than $-1$, giving an upper bound on a \texttt{backward} specification.
-
-Finally, if either bound is $\infty$, representing a non
-neighbourhood index, then an empty specification is given
- with no regions, indicating that there is not even reflexive
-0-offset access in this dimension.
-
-Note, $\mathsf{toSpec} \, l \, u$ is always defined 
- if $l$ less than or equal to $u$. 
-\end{definition}
-
-%\newcommand{\mkBig}[2]{\overset{\LARGE{\term{#1}}}{#2}}
-\begin{definition} The $\odot$ operation on $\textsf{A}(\specDNF{})$
-  is defined in terms of the intermediate operator $\hat{\odot}$ on
-  \specDNF{}, defined:
-%%
-\begin{align*}
-S \,\hat{\odot}\, \epsilon = S \quad
-\epsilon \,\hat{\odot}\, S = S \qquad
-S \,\hat{\odot}\, T = \bplus_{(\tyProd{s}, \tyProd{t}) \in (S \times
-   T)} (\tyProd{s} \; \term{*} \; \tyProd{t})
-\end{align*}
-%%
-That is, the $\hat{\odot}$ product of two specifications $S$ and $T$ in DNF form 
-is the $\term{+}$ sum of all pairwise $\term{*}$ products for every
-pair of $\tyProd{spec}$ drawn from $S$ and $T$. The result
-is indeed in DNF-form, where $\hat{\odot}$ is essentially applying
- the (\trule{\textsc{dist}}) rule of our equational theory
- (Fig.~\ref{fig:equations},p.~\pageref{fig:equations}). The operation is
-commutative and associative, and sound with respect to our
-model (see Lemma~\ref{lem:alg-soundness} below).
-The $\hat{\odot}$ operation is then lifted to all combinations of
-\textsf{Approx} for $\odot$:
-\begin{align*}
-\textsf{lower}(S) \odot \textsf{exact}(T) = \; &
-       %\textsf{exact}(T) \odot \textsf{lower}(S) = \;
-       \textsf{both} \, (S \hat{\odot} T, \; T) \\
-\textsf{upper}(S) \odot \textsf{exact}(T) = \; & 
-       %\textsf{exact}(T) \odot \textsf{upper}(S) =
-       \textsf{both} \, (T, \; S \hat{\odot} T) \\
-\textsf{lower}(S) \odot \textsf{both}(T_l,T_u) = \; &
-        %\textsf{both}(T_l,T_u) \odot \textsf{lower}(S) = \;
-        \textsf{both} \, (S \hat{\odot} T_l, \; T_u) \\
-\textsf{upper}(S) \odot \textsf{both}(T_l,T_u) = \; & \textsf{both} \,
-                                                      (T_l, \; S \hat{\odot} T_u) \\
-\textsf{both}(S_l,S_u) \odot
-\textsf{both}(T_l,T_u)  = \; & \textsf{both} (S_l \hat{\odot} T_l, \; S_u
-                         \hat{\odot} T_u) \\
-\textit{inj}(S) \odot \textit{inj}(T) = \; & \textit{inj}(S
-                                              \hat{\odot} T)
-\end{align*}
-where in the last case $\textit{inj}$ corresponds to the unary
-injections of \textsf{Approx}. For brevity we omit the cases where the arguments above are flipped
-as $\odot$ is commutative.
-\end{definition}
-
-\begin{definition}The $\oplus$ operation on $\textsf{A}(\specDNF)$ 
-is defined in a similar way to $\odot$, with the intermediate $\hat{\oplus}$ on
-$\specDNF$ given by:
-%%
-\begin{equation*}
-S \,\hat{\oplus}\, \epsilon = S \quad
-\epsilon \,\hat{\oplus}\, S = S \quad
-S \, \hat{\oplus} \, T = S \, \term{+} \, T
-\end{equation*}
-%%
-Thus, if neither specification is empty, their $\hat{\oplus}$ is just
-the syntactic sum $\term{+}$. Then $\oplus$ is the lifting of
-$\hat{\oplus}$ to \textsf{Approx} with the shape as $\odot$ in the previous
-definition.
-\end{definition}
-
-\begin{lemma}[$\hat{\odot}$, $\hat{\oplus}$ soundness]$\forall S,
-  T \in \specDNF$ where $S\!\neq \epsilon \wedge T\!\neq \epsilon$:
-\begin{align*}
-\interp{S \; \hat{\odot} \; T}_n = \interp{S}_n \otimes \interp{T}_n
-  \qquad
-\interp{S \; \hat{\oplus} \; T}_n = \interp{S}_n \cup \interp{T}_n
-\end{align*}
-\label{lem:alg-soundness}
-\vspace{-2em}
-\end{lemma}
-%
-\noindent
-Thus our model validates the specification synthesis process which concludes the
-inference procedure. For our example, we infer and then synthesise the
-5-point specification:
-%%
-\begin{minted}[linenos=false]{fortran}
-!= stencil centered(depth=1, dim=1)*reflexive(dim=2) + centered(depth=1, dim=2)*reflexive(dim=1):: a
-\end{minted}
-%%
-
-\subsection{Specification checking\label{subsec:checking}}
-
-In checking mode, the access pattern of a stencil computation in the
-source language is verified against a specification. The model of
-Section~\ref{sec:semantics} again aids this process.  Checking
-proceeds by generating a model $M$ from a specification and generating a set $N$ of indexing
-schemes from the source code. The source code is then valid with
-respect to the specification if $N$ is \emph{consistent} with $M$. The
-notion of consistency is built out of index scheme 
-\emph{generalisation} (Sect.~\ref{subsec:schemes-and-gen}). 
-%Consistency is defined in Figure~\ref{fig:consistency},
-We first set up some intermediate definitions.% before unpacking its definition.
-
-\begin{definition}[Checking]
-Let \texttt{a} be an array variable and let
-$S \in \textit{spec}$ be a specification with
-the \textit{specDec} \texttt{!= stencil
-  $S$ :: a} in our syntax.
-For a array subscript
-assignment statement $\textit{assg} = v(\overline{e_1}) = e_2$ in our source language
-(with all the appropriate data flow information in place)
-where $e_2$ has at most $n$-dimensional array access, 
-then the assignment conforms to the specification $S$ if 
-the predicate $\mathit{check}_n(S, \textit{assg}, \texttt{a})$ holds, 
-defined:
-\begin{align*}
-& \mathit{check_n}(S, \textit{assg}, x) =
-  \mathit{cons_n}(\interp{S}_n, \, \textit{modelise}(\textit{assg},
-  x)) \\
-& \mathit{modelise} : \textit{statement} \times \textit{variable}
-  \rightarrow \textsf{Mult}((\domainVal)^n) \\ 
-& \mathit{modelise} (\textit{assg}, x) =
-  \begin{cases}
-    \textsf{once}(\mathit{analyse}(\textit{assg})(x)) & \textit{unique}(\textit{assg},x) \\
-    \textsf{mult}(\mathit{analyse}(\textit{assg})(x)) & otherwise
-  \end{cases}
-\end{align*}
-Thus, checking proceeds by reusing the \textit{analyse} function from 
-inference (Sec.~\ref{subsec:inference}) to extract index 
-schemes from source code. 
-This is then augmented with multiplicity information
-by $\mathit{modelise}$ where $\mathit{unique}(\textit{assg},x)$ is a
- predicate indicating whether indices to array $x$ in an
-assignment $\textit{assg}$ are unique or not. 
-\end{definition}
-
-Subsequently the binary predicate for consistency, at the heart of checking
-is defined over the sets: 
-%
-\begin{equation*}
-\textit{cons}_n \subseteq \textsf{Mult}(\textsf{Approx}(\domainVal{}^n))
-\times \textsf{Mult}(\domainVal{}^N)
-\end{equation*}
-%
-where first component is a model of a specification,
-and the second component is a set of indexing schemes generated from
-the array subscripts in a program, with added multiplicity
-information. Giving \textsf{Approx} information to the
-index schemes inferred from the program is not necessary, nor
-desirable-- we have an exact set of schemes. We refer to the first
-component of the predicate as \emph{spec model} and the second as \emph{code model}.
-
-\begin{definition}
-Consistency is defined in Figure~\ref{fig:consistency}. We let
-$M_s, M_c$ range over elements of $\domainVal{}^n$ where $M_s$ comes
-from the spec model and $M_c$ comes from \textit{modelise} on the
-source code. Let $N_c$ range over elements of
-$\textsf{Mult}(\domainVal{}^n)$ coming from the code model, and $A_s$
-over elements of $\textsf{Mult}(\textsf{Approx}(\domainVal{}^n))$ from the spec model.
-
-The first equation states that, if the multiplicity of the spec and code
-model are both $\mathsf{once}$, then the problem is reduced to checking consistency of
-models in which non-unique indexing is allowed, labelled with
-$\mathsf{mult}$.
-
-The second equation considers the case where the specification is
-an \term{atMost} upper-bound approximation (labelled with \textsf{up}
-as shorthand for \textsf{upper} here). 
-An upper bound specification means that for every scheme in the code
-model, modelling a source-level index, $\forall u \in M_c$ there exists
-a generalising scheme in the spec model $\exists v \in M_s$, \ie{}, $u
-\preceq_n v$. That is, every source-level index is generalised by the model.
-
-The third equation is the converse of an \term{atLeast},
-lower approximation (labelled \textsf{low} as shorthand for
-\textsf{lower}). Consistency is dual to the above: 
-for every scheme in the spec model $\forall v \in M_s$ there
-exists a scheme in the code model $\exists u \in M_c$ such that $u
-\preceq_n v$. That is, every model scheme is generalised by source indexing.
-
-These two are combined for consistency on an \textsf{exact}
-specification (the fourth equation) such that the spec model
-provides both a consistent lower and consistent upper bound on
-the source code. Thus, for every scheme in the code model there
-is a scheme in the spec model that generalises it, and for every
-scheme in the spec model there is a scheme in the source model that
-specialises it.
-
-Lastly, if the spec model has both an upper and lower bound
-consistency is defined inductively similar to the \textsf{exact} case.
-\end{definition}
-
-\begin{figure}
-\vspace{-0.4em}
-\input{consistency.tex}
-\caption{Consistency between a model and indexing schemes}
-\label{fig:consistency}
-\vspace{-0.9em}
-\end{figure}
-
-\begin{example}
-We give an extended example of verifying a stencil computation against
-its specification. We consider a more general, but still consistent, version of the \textit{five-point}
-stencil specification inferred previously in
-Section~\ref{subsec:inference}:
-%
-\begin{minted}[breakindent=8.1em]{fortran}
-!= region :: fivepoint = centered(depth=1,dim=1) + centered(depth=1,dim=2)
-!= stencil readOnce, fivepoint :: b
-a(i,j) = b(i,j-1)+b(i,j)+b(i,j+1)+b(i-1,j)+b(i+1,j)
-\end{minted}
-%
-We refer to the assignment on line three as $l3$ for brevity.
-
-Verification of the code against the spec is then given by 
-$check(({\textcolor{darkgreen}{\texttt{readOnce, fivepoint}}}),l3,\texttt{b})$. 
-We explain the steps.
-\begin{enumerate}  
-%
-\item Generate the model from the specification as in
-  Section~\ref{sec:semantics}. 
-%
-\begin{align*}
- M_s= \{  \vtwo{-1}{\infty},
-          \vtwo{0}{\infty},
-          \vtwo{1}{\infty},
-          \vtwo{\infty}{-1},
-          \vtwo{\infty}{0},
-        \vtwo{\infty}{1} \}  \\[0.4em]
-\interp{{\textcolor{darkgreen}{\texttt{readOnce, fivepoint}}}}_2 =
-  \textsf{once}(\textsf{exact} (M_s))
-\end{align*}
-%
-\item Obtained the code model via $\textit{modelise}$, giving:
-%
-\begin{align*}
-M_c = \{
-    \vtwo{0}{-1},
-    \vtwo{0}{0},
-    \vtwo{0}{1},
-    \vtwo{-1}{0},
-    \vtwo{1}{0}
-  \}
-\\[0.4em]
-\textit{modelise}(l3, \texttt{b}) = \textsf{once} (M_c)
-\end{align*}
-%
-\item Consistency $cons_2(\textsf{once}(\textsf{exact}(M_s)),
-  \textsf{once}(M_c))$ holds when 
-  $\textit{cons}_2(\textsf{mult}(\textsf{exact}(M_s)),\textsf{mult}(M_c))$ (Fig.~\ref{fig:consistency}) 
-  which holds when: 
-  \begin{align*}
-     & \textit{cons}_2(\textsf{mult}(\textsf{up}(M_s)),\textsf{mult}(M_c)) \\
-   \wedge \; & \textit{cons}_2(\textsf{mult}(\textsf{low}(M_s),\textsf{mult}(M_c))
-  \end{align*}
-%
-\item This condition holds by the following sets of scheme
-  generalisations: for the upper bound, $\forall u \in M_c,
-  \exists v \in M_s . u \preceq_2 v$: 
-  \begin{align}
-    \begin{array}{l}
-     \vtwoh{0}{-1} \preceq_2 \vtwoh{0}{\infty} \quad
-     \vtwoh{0}{0} \preceq_2 \vtwoh{0}{\infty} \quad
-     \vtwoh{0}{1} \preceq_2 \vtwoh{0}{\infty} \\[0.5em]
-     \vtwoh{-1}{0} \preceq_2 \vtwoh{\infty}{0} \quad
-     \vtwoh{1}{0} \preceq_2 \vtwoh{\infty}{0}
-     \end{array}
-  \label{eq:check-example-upper}
-  \end{align}
-  % 
-  and for the lower bound: $\forall v \in M_s, \exists u \in M_c . u
-  \preceq_n v$: 
-  \begin{align}
-    \hspace{-1em}
-    \begin{array}{l}
-     \vtwoh{-1}{0} \preceq_2 \vtwoh{-1}{\infty} \quad
-     \vtwoh{0}{0} \preceq_2 \vtwoh{0}{\infty} \quad
-     \vtwoh{1}{0} \preceq_2 \vtwoh{1}{\infty}  \\
-     \vtwoh{0}{-1}  \preceq_2 \vtwoh{\infty}{-1} \quad
-     \vtwoh{0}{0}  \preceq_2 \vtwoh{\infty}{0} \quad
-     \vtwoh{0}{1}  \preceq_2 \vtwoh{\infty}{1}
-     \end{array}
-    \label{eq:check-example-lower}
-  \end{align}
-  Since both the consistency predicate is satisfied, the array
-  accesses flowing to line three in the stencil computation are valid
-  with respect to the {\textcolor{darkgreen}{\texttt{fivepoint}}} specification.
-\end{enumerate}
-\vspace{-0.7em}
-\end{example}
-\begin{example} 
-We consider now an erroneous user modification to $l3$ with an
-additional term \texttt{b(i+2,j+2)} such that $l3'$ is:
-%
-\begin{minted}[firstnumber=3]{fortran}
-a(i,j) = b(i,j-1) + b(i,j) + b(i,j+1) + b(i-1,j) + b(i+1,j) + b(i+2,j+2)
-\end{minted}
-%
-The checking procedure proceeds in the same way as the first three
-steps of the previous example, except that the result of
-$\mathit{modelise}(l3',\texttt{b})$ generates an additional scheme
-$\vtwoh{2}{2}$. Since the spec model has not changed and the code
-model is has only the additional scheme $\vtwoh{2}{2}$, the generalisations for
-the lower bound \eqref{eq:check-example-lower} 
-of step 4 above still holds since its consistency condition
-universally quantifies over the \emph{spec model}. However, for
-the upper bound, the consistency condition universally quantifies
-over scheme $u$ of the \emph{code model} for which there exists
-a scheme $v \in M_s$ such that $u \preceq_2 v$. Thus, we require a $v
-\in M_s$ such that $\vtwoh{2}{2} \preceq_2 v$ but no such generalising
-scheme existing in the spec model. Hence the model fails to hold as an upper bound.
-
-Note that if the specification were as follows:
-%
-\begin{minted}{fortran}
-!= stencil atLeast, readOnce, fivepoint :: b
-\end{minted}
-%
-then consistency holds as it no longer requires the failing upper-bound case 
-$\mathit{cons}(\textsf{mult}(\textsf{up}(M_s)),\textsf{mult}(M_c))$.
- This makes sense semantically since the region ${\textcolor{darkgreen}{\texttt{fivepoint}}}$
-is indeed contained in the regions touched by $\texttt{b}$ in $l3'$. 
-\end{example}
-
-\subsection{Approximations and checking}
-
-As noted, the {\textcolor{darkgreen}{\term{fivepoint}}} region
- above is more accepting than that used in
-Section~\ref{sec:lang}. This introduces a new notion of approximation
-beyond what is given by the syntactic theory of sub-specifications
-$<:$ in Section~\ref{sec:eqs}. Consistency leads to a notion of
-approximation that is a \textit{semantic sub-specification} relation:
-%using upper and lower bounds because the in both cases the specifications are
-%used to generate \emph{exact} models.
-
-\begin{definition}[Approximation]
-A specification $S'$
-\emph{approximates} a more specific specification $S$, written 
-$S \sqsubseteq S'$, \emph{iff}:
-$$
-\forall M_c \in \textsf{Mult}(\domainVal{}^n) . \;\, \mathit{cons}_n(\interp{S}_n,M_c) \implies
-\mathit{cons}(\interp{S'}_n,M_c)
-$$
-\end{definition}
-%
-For the syntactic sub-specification equations on regions $<:^r$, it follows that
-these imply semantic approximation:
-%
-\begin{lemma}[Sub-specification $<:^r$ coincides with approx 
-$\sqsubseteq$]
-\begin{equation*}
-\forall S, S' . \; S <:^r S' \Rightarrow S \sqsubseteq S'
-\end{equation*}
-\end{lemma}
-%
-\noindent
-However, the converse direction does not hold. 
-The regions describing the more accepting five-point stencil
-above do not follow from the definition of $<:$ on the previous
-more specific specification.%, that is:
-%${\textcolor{darkgreen}{\term{centered(depth=1, dim=1)*reflexive(dim=2)
-%      + centered(depth=1, dim=2)*reflexive(dim=1)}}}
-%\not{<:} {\textcolor{darkgreen}{\term{centered(depth=1,dim=1) +
-%      centered(depth=1,dim=2)}}}$.
- However, consistency (and thus $\sqsubseteq$) shows
-that we can check the five-point stencil code against the more general
-specification of {\textcolor{darkgreen}{\term{fivepoint}}}. Thus, we
-can always use an approximation in place of the
-more specific specification. Although this decreases the likelihood of catching
-stencil errors (as more access patterns conform), it reduces the space and
-effort it takes to write the specification.
-
-\section{Evaluation}
-\label{sec:evaluation}
-We evaluated our specification language on a corpus of scientific
-code.  We overview the corpus itself before presenting our results.
-
-%\subsection{Software Corpus}
-\paragraph{Software Corpus}
-
-Table~\ref{tab:corpus} shows summary statistics of the software
-packages used in our evaluation, all of which are written in Fortran
-90 or Fortran 77. In total we analysed \SI{\overallLoC} lines of code
-from 7 packages, of which we successfully parsed
-\SI{\overalllinesParsed} lines. The ``Number of files'' column shows
-how many files in each corpus that we were able to analyse with
-CamFort. The most common reasons for CamFort rejecting a file were
-either use of a C-preprocessor, or illegal use of language features
-from a modern Fortran variant.
-
-\textbf{The Unified
-  Model}~\cite{um}
-is a weather forecasting and climate modelling tool developed by the
-Met Office in the United Kingdom. It is used by research organisations
-and meteorological services around the world. We use the development
-branch (trunk) of the model. The code base in closed source but
-institutional licenses are available for research purposes.
-%The Met Office runs a comprehensive code quality system incorporating dedicated committers (we counted 11) for particular parts of the model.
-We counted 120 additional contributors whose submissions are reviewed and tested before being accepted into the code base.
-
-% e3mg-ea
-\textbf{E3MG} (An Energy-Environment-Economy (E3) Model at the Global Level) is a macroeconomic model used for assessment of environmental policy~\cite{RePEc:aen:journl:2006se-a12}.
-This was developed by Cambridge Econometrics, an independent consultancy company.
-
-\textbf{BLAS}~\cite{blas} (Basic Linear Algebra Subprograms) is a popular
-library providing efficient and portable routines for vector and
-matrix operations. These routines feature in many other libraries
-(including LAPACK). We used version 3.6.0. We chose to include this
-package for breadth, as it provides general numerical functions rather
-than a specialised scientific model.
-
-\textbf{Hybrid4} is a vegetation and biomass model for simulating carbon, water and nitrogen flows~\cite{GBC:GBC635}.
-
-\textbf{GEOS-Chem}~\cite{geos-chem} is a three-dimensional model of
-tropospheric chemistry developed at Harvard and used by $\sim$70
-universities and research institutions world-wide. We use v.10-01.
-
-\textbf{Navier} is a small numerical simulation, giving a discrete
-approximation to the two-dimensional Navier-Stokes fluid equations,
-based on the book of~\citet{griebel1997numerical}.
-
-\textbf{CP} consists of the example code from the second edition of the book ``Computational Physics''~\cite{nicholas2006computational} introducing numerical techniques and their application to modern physics problems such as fields, waves, statistical mechanics and quantum mechanics.
-
-% atLeast: number of stencils with an atLeast modifier
-% atMost: number of stencils with an atMost modifier
-% boundedBoth: number of stencils with both atLeast and atMost modifier
-%% atLeast == atMost == boundedBoth - whenever we infer one we infer the other but users might write just one
-
-% backward: number of backward stencils
-% forward: number of forward stencils
-% centereed: number of centered stencils
-
-% justReflexive: number of only reflexive stencils
-% irreflexive: number of irreflexive stencils
-% readOnce: number of readOnce stencils
-
-% depthX: number of stencils with depth X
-% dimsX: number of stencils with dimensions=X (check this!)
-
-% lexFailed: number of files we failed to lex
-% parseFailed: number of files we failed to parse
-% lexOrParseFailed: number files we failed to lex and parse
-% parseOk: number of files parsed successfully
-%% lexFailed + parseFailed = lexOrParseFailed
-%% lexOrParseFailed + parseOk = number of files in source repo
-
-% tickAssign: number of array subscripts with a LHS which are all neighbour indices or constants
-% tickAssignSuccess: number of tickAssign's we generated some specification for
-%% tickAssign < tickAssignSuccess
-
-\subsection{Occurrences of stencils}
-
-We first examined how frequently stencil computations occur in our
-corpus. This was carried out by running our implementation in 
-whole-code-base inference mode. We define a \emph{potential} statement as one whose left-hand
-side is an array subscript on neighbourhood indices. These statements
-are the inputs to Step 2 of the Inference process described in 
-Section~\ref{subsec:inference}. Each potential statement for which we
-eventually generate some specification is also counted as
-an \emph{actual} statement.
-
-\begin{table}
-\centering
-\footnotesize
-\begin{tabular}{lS[table-format=4]S[table-format=4]SS}
-\hline
-& \multicolumn{2}{c}{Number of files} & \multicolumn{2}{c}{Lines of code} \\
-
-Package   & {Total} & {Parsed} & {Total} & {Parsed}  \\ % braces prevent headers from being formatted as numbers
-\hline
-UM        & \umFiles                       & \umparseOk                      & \umLoC                      & \umlinesParsed \\
-E3MG      & \ethreemgeaFiles               & \ethreemgeaparseOk              & \ethreemgeaLoC              & \ethreemgealinesParsed \\
-BLAS      & \blasFiles                     & \blasparseOk                    & \blasLoC                    & \blaslinesParsed \\
-Hybrid4   & \hybridfourFiles               & \hybridfourparseOk              & \hybridfourLoC              & \hybridfourlinesParsed \\
-GEOS-Chem & \geoschemFiles                 & \geoschemparseOk                & \geoschemLoC                & \geoschemlinesParsed \\
-Navier    & \navierFiles                   & \navierparseOk                  & \navierLoC                  & \navierlinesParsed \\
-CP        & \computationalphysicstwoFiles  & \computationalphysicstwoparseOk & \computationalphysicstwoLoC & \computationalphysicstwolinesParsed \\
-\hline
-\textbf{Total} & \overallFiles  & \overallparseOk & \overallLoC & \overalllinesParsed \\
-\end{tabular}
-\caption{Summary of software packages used for
-  evaluation\label{tab:corpus}}
-\vspace{-1em}
-\end{table}
-\begin{table}
-\vspace{-1em}
-\sisetup{table-format=4,table-text-alignment=right}
-\footnotesize
-\centering
-\begin{tabular}{lS@{~~~}rS@{~~~}rSS[table-format=1]}
-\hline
-  & \multicolumn{2}{c}{Potential} & \multicolumn{2}{c}{Actual} & {Exact} & {Inexact}    \\
-{Package}     & {\#} & \si{\percent} & {\#} & \si{\percent} \\
-\hline
-UM            & \umtickAssign & \SI{\umtickAssignPercent}{\percent} &
-                \umtickAssignSuccess & \SI{\umtickAssignSuccessPercent}{\percent} &
-                \umnumStencilLines &
-                \umboundedBoth  \\
-E3MG          & \ethreemgeatickAssign & \SI{\ethreemgeatickAssignPercent}{\percent} &
-                \ethreemgeatickAssignSuccess & \SI{\ethreemgeatickAssignSuccessPercent}{\percent} &
-                \ethreemgeanumStencilLines &
-                \ethreemgeaboundedBoth \\
-BLAS          & \blastickAssign & \SI{\blastickAssignPercent}{\percent} &
-                \blastickAssignSuccess & \SI{\blastickAssignSuccessPercent}{\percent} &
-                \blasnumStencilLines &
-                \blasboundedBoth \\
-Hybrid4       & \hybridfourtickAssign & \SI{\hybridfourtickAssignPercent}{\percent} &
-                \hybridfourtickAssignSuccess & \SI{\hybridfourtickAssignSuccessPercent}{\percent} &
-                \hybridfournumStencilLines &
-                \hybridfourboundedBoth\\
-GEOS-Chem     & \geoschemtickAssign & \SI{\geoschemtickAssignPercent}{\percent} &
-                \geoschemtickAssignSuccess & \SI{\geoschemtickAssignSuccessPercent}{\percent} &
-                \geoschemnumStencilLines &
-                \geoschemboundedBoth \\
-Navier        & \naviertickAssign & \SI{\naviertickAssignPercent}{\percent} &
-                \naviertickAssignSuccess & \SI{\naviertickAssignPercent}{\percent} &
-                \naviernumStencilLines &
-                \navierboundedBoth \\
-CP            & \computationalphysicstwotickAssign & \SI{\computationalphysicstwotickAssignPercent}{\percent} &
-                \computationalphysicstwotickAssignSuccess & \SI{\computationalphysicstwotickAssignSuccessPercent}{\percent} &
-                \computationalphysicstwonumStencilLines &
-                \computationalphysicstwoboundedBoth \\
-\hline
-\textbf{Total}& \overalltickAssign & \SI{\overalltickAssignPercent}{\percent}  &
-                \overalltickAssignSuccess & \SI{\overalltickAssignSuccessPercent}{\percent} &
-                \overallnumStencilLines &
-                \overallboundedBoth   
-\end{tabular}
-\caption{Stencil specification assignment rates\label{tab:tickAssign}}
-\vspace{-1em}
-\end{table}
-
-The first two columns of Table~\ref{tab:tickAssign} show the number
-of potential and actual stencils and their rate as a percentage of the
-total number of successfully parsed lines of code. The ratio of
-actual to potential stencils overall is thus $32\%$, \ie{}, we give
-a specification to roughly a third of the potential stencil specifications.
-
-There are a variety of reasons why we might not infer a actual specification from a potential
-stencil computation: \\
-\noindent 
-1) \textbf{Non-subset induction variables} occur when the
-induction variables on the RHS are not a subset of those in the LHS. These
-cases are not stencils by our definition (Defn.~\ref{def:stencil},p.~\pageref{def:stencil}). The degenerate case of this
-is to have only constant indices on the LHS. We see lots of examples
-of this in loops as accumulators \eg{} computing the sum over an array; \\
-\noindent
-2) \textbf{Derived induction variables} where the
-indexing variable (\mintinline{fortran}{x}) is derived from an
-induction variable (\mintinline{fortran}{i}) as in
-\mintinline{fortran}{x = len - i};  \\
-\noindent
-3) \textbf{Inconsistent induction dimensions} occur when
-an induction variable is used to specify more than one array dimension
-on the RHS or multiple induction variables are used for the same
-dimension on the RHS. These are common in matrix operations such as
-LU-decomposition with assignments such as
-\mintinline{fortran}{a(l) = a(l) - a(m) * b(l, m)}.
-
- %% This example is from geos-chem/GeosRad/rrtmg_sw_spcvmc.F90:609
- %%                      pnifddir(ikl) = pnifddir(ikl) + zincflx(iw)*ztdbt(jk)
- %% ikl is a derived induction variable (ikl = rows - jk) - we can't deal with that at the moment
-
-  %% This example is from geos-chem/GTMM/sort_pick_veg.F90
-  %% DO k=1,n_veg
-  %%    DO j=2,n/n_veg
-  %%       a=arr(k,j)
-  %%       b=ind(k,j)
-  %%       DO i=j-1, 1, -1
-  %%          IF (arr(k,i) >=a) exit
-  %%          arr(k,i+1)=arr(k,i)
-  %%          ind(k,i+1)=ind(k,i)
-  %%       END DO
-  %%       ! NO STENCIL - inconsistentIV
-  %%       ind(k,i+1)=b
-  %%       arr(k,i+1)=a
-  %%    END DO
-  %% END DO
-  %% ind(k,j) flows into b but the LHS doesn't use j. This is a valid 'access' stencil but not a valid comonad stencil
-
-%% From geos-chem/GTMM/loadCASAinput.F90:241
-%%          ppt_mo(1,i)=sum(ppt(:,:,i))
-%% If you expand the loop for the sum
-%% do x=1,lenDim1
-%%  do y=1,lenDim2
-%%   acc += ppt(x,y,i)  ! can't be a stencil but could be an access pattern
-%%  enddo
-%% enddo
-%% ppt_mo(1,i) = acc  ! should be reflexive(dim=3) :: ppt
-%% This should be fixable before the paper deadline
-
-  %% From blas/ssyrk.f:337
-  %%            DO 210 J = 1,N
-  %%                 DO 200 I = 1,J
-  %%                     TEMP = ZERO
-  %%                     DO 190 L = 1,K
-  %%                         TEMP = TEMP + A(L,I)*A(L,J)
-  %% 190                 CONTINUE
-  %%                     IF (BETA.EQ.ZERO) THEN
-  %%                         C(I,J) = ALPHA*TEMP  ! No stencil
-  %%                     ELSE
-  %%                         C(I,J) = ALPHA*TEMP + BETA*C(I,J)
-  %%                     END IF
-  %% 200             CONTINUE
-  %% 210         CONTINUE
-  %% In this case A does flow into TEMP and thus into C but its a reduction. There's an 'access' stencil above TEMP but that's all
-
-   %% From blas/sspr2.f:231
-   %%                    DO 10 I = 1,J
-   %%                        AP(K) = AP(K) + X(I)*TEMP1 + Y(I)*TEMP2
-   %%                        K = K + 1
-   %% 10                 CONTINUE
-   %% I don't think we notice that K is an induction variable in these cases
-
-  %% From computational-physics-2/fortran90/millikan.f90:73
-  %% DO I = 1, N+1
-  %%   STMP = U(1,I)*U(1,I)    ! access, not a stencil
-  %%   S(1) = S(1) + STMP      ! access, not a stencil
-  %%   G(1) = G(1) + K(I)*STMP ! access, not a stencil
-  %%   U(1,N+2) = U(1,N+2) + U(1,I)*F(I)  ! access, not a stencil
-  %% END DO
-
-%% From UM: mym_solve_simeq_lud.F90:148
-%% DO m = 1, n - 1
-%%   DO l = m + 1, n
-%%     bvec(l) = bvec(l) - bvec(m) * amat(l, m)
-%%   END DO
-%% END DO
-%% We just don't deal with this - Lower upper decomposition solver
-
-If the statement meets the stencil criteria but the inference process
-cannot determine an exact stencil for a statement it will instead
-produce an upper (\term{atMost}) and lower (\term{atLeast}) bound. The
-number of exact and inexact stencils are shown in columns three and
-four of Table~\ref{tab:tickAssign}. Note that the total for ``Exact``
-is higher than ``Actual`` because for some lines we produce multiple
-specifications (perhaps for different array variables).
-We see that our specification language is capable of precisely specifying the overwhelming majority of stencils.
-Most of the inexact stencils apply to some form of data copying.
-For example (taken from \textbf{E3MG}):
-\begin{minted}{fortran}
-!- stencil atLeast, readOnce, irreflexive(dims=2), (reflexive(dim=1)) :: sfda
-!- stencil atMost, readOnce, (forward(depth=8, dim=2))*(reflexive(dim=1)) :: sfdt
-ZZ1(I,J) = SFDT(I,J+8)
-\end{minted}
-
-\subsection{Stencil categories}
-
-We grouped our inferred specifications together into four classes based on similar levels of complexity. These four classes account for 99.4\% of all the stencils we found in the corpus.
-
-\textbf{All-reflexive} The vast majority (66\%) of specification are
-reflexive in every dimension. These most often correspond
-to point-wise transformations on data such as scaling, \eg{} from \textbf{E3MG}: \begin{minted}{fortran}
-ZZ1(I,K) = .000041868*ZZ1(I,K)
-\end{minted}
-Arrays with this class of stencil are indexed by all the induction
-variables used in the LHS without any offsets.
-
-\textbf{Some-reflexive} specifications (24\%) consist of only reflexive terms but
-leave some dimensions unspecified. These two examples are from
-\textbf{GEOS-Chem} and \textbf{E3MG}: 
-\begin{minted}{fortran}
-leafpools(:,age_class)=leafpool(:,1)
-RVM(J,I2) = VRFO(ivar,J)
-\end{minted}
-where for the first example \mintinline{fortran}{leafpools} contains an
-implicit loop due to the range term \texttt{:}, and for the second \mintinline{fortran}{J} is
-the induction variable. This stencil permits array indexing by a subset of the LHS induction variables but again without
-any offsets.
-
-\textbf{Single-action} specifications (6\%) consist of a single \term{backward},
-\term{forward}, or \term{centered} region and some number of \term{reflexive}
-region constants. For example, taken from \textbf{GEOS-Chem}: 
-% /home/mrd45/src/corpus/geos-chem/GTMM/CasaRegridModule.F90:645
-\begin{minted}[breakafter=*]{fortran}
-do j=2,jm-ig
-   do i=1,im
-      != stencil readOnce, (backward(depth=1, dim=2)) *(reflexive(dim=1)) :: dm, q
-      al(i,j) = 0.5e+0_fp*(q(i,j-1)+q(i,j)) + r3*(dm(i,j-1) - dm(i,j))
-\end{minted}
-% TODO: A BIT UNCLEAR
-%This stencil requires a neighbourhood index including zero-offset in one dimension only. The remaining dimensions must either have no offset or not pertain to an induction variable.
-These kinds of specification require zero offsets in one dimension and 
-and neighbourhood offsets (from 0 to $-1$ above) in the other.
-The remaining dimensions must either have no offset or not pertain to
-an induction variable.
-
-\textbf{Single-action-irreflexive} specifications (3\%) consist of a single
-\term{backward}, \term{forward}, or \term{centered} irreflexive region 
-and some number of reflexive region constants, \eg{}, from \textbf{GEOS-Chem}:
-% /home/mrd45/src/corpus/geos-chem/GeosRad/rrtmg_lw_rtrnmc.F90:355
-\begin{minted}{fortran}
-do lev = nlayers, 1, -1
-    ! ...
-    radld = radld - radld * (atrans(lev) + &
-       efclfrac(lev,igc) * (1. - atrans(lev))) + &
-       gassrc + cldfmc(igc,lev) * &
-       (bbdtot * atot(lev) - gassrc)
-    != stencil readOnce, forward(depth=1, dim=2, irreflexive) :: cldfmc
-    drad(lev-1) = drad(lev-1) + radld
-\end{minted}
-Note that the read from 
-\mintinline{fortran}{cldfmc} on line 5 flows to line 8. Stencils of
-the above kind require a neighbourhood index without a zero-offset in
-one dimension only. The remaining dimensions must either have no
-offset or not pertain to an induction variable.
-
-The remaining specification shapes (less than 1\%) consisted
-primarily of multiple \term{forward}, \term{backward}, or
-\term{centered} dimensions. We note that only $\sim$0.4\% of inferred
-stencil specifications required use of the \term{+} operator.
-Despite their low frequency these stencils are arguably the most important since they represent more complicated code and thus greater potential for programming errors.
-
-\subsection{Verification case study 1: 2-D Jacobi iteration}
-
-One relatively basic example of a stencil computation is two-dimensional
-Jacobi iteration that repeatedly goes through each cell in a matrix
-and computes the average value of the four adjacent cells. The kernel
-can be expressed in a line of Fortran code:
-\begin{minted}{fortran}
-  a(i,j) = (a(i-1,j)+a(i+1,j)+a(i,j+1)+a(i,j-1))/4
-\end{minted}
-and CamFort is able to infer a precise specification for this:
-\begin{minted}[breakafter=+:,breakindent=-0.6em,breaksymbolsep=0.4em,linenos=false,xleftmargin=-0.5em]{fortran}
-  != stencil reflexive(dim=1)*centered(depth=1,dim=2,irreflexive)+reflexive(dim=2)*centered(depth=1,dim=1,irreflexive)
-\end{minted}
-assigned to \term{a}. 
-We then tested the error-detecting ability of our approach by 
-generating a family of 6,561 Jacobi iteration programs by perturbing
-the indices used in the following manner: each access of the array was
-templated as \mintinline{fortran}{a(i+ki,j+kj)} and then all possible
-combinations of \mintinline{fortran}{ki} and \mintinline{fortran}{kj}
-were drawn from the set $\{-1,0,1\}$. There are four uses of
-\mintinline{fortran}{ki} and four of \mintinline{fortran}{kj}, hence
-$3^8=6561$ possible perturbations of this kind. This particular setup
-tests the class of errors in which the programmer typed an incorrect
-number but at least got the ordering of the indices correct, although
-CamFort would also detect a mixup of variables in this case as well.
-
-We assigned the specification inferred above to the generated Jacobi
-stencils programs. CamFort checked all 6,561 programs against this specification and found that only 24 of them were
-valid implementations of the stencil specification. This was the
-expected result as there are $4!$ possible correct permutations
-of the four terms in the kernel. All of the other programs were found
-to violate the specification in some way. Another way to test the
-programs is through visual examination of their output in graphical
-format. Figure~\ref{fig:jacobi} shows some of the results of running
-the Jacobi iteration programs, rendered into image files. All of the
-correct variants emitted images indistinguishable from
-Figure~\ref{fig:jacobicorrect}, while all of the incorrect variants
-had wildly different image outputs. Both the ``visual test'' and a
-byte-by-byte comparison showed results that completely concurred with
-the results of the CamFort test.
-
-\begin{figure}[t]
-  \centering
-  %\subcaptionbox{\label{fig:jacobicorrect}}{\includegraphics[width=0.1\textwidth]{jacobi-correct.png}}
-  %\subcaptionbox{}{\includegraphics[width=0.1\textwidth]{jacobi-incorrect1.png}}
-  %\subcaptionbox{}{\includegraphics[width=0.1\textwidth]{jacobi-incorrect2.png}}
-  %\subcaptionbox{}{\includegraphics[width=0.1\textwidth]{jacobi-incorrect3.png}}
-  \subfigure[]{\label{fig:jacobicorrect}\includegraphics[width=0.1\textwidth]{jacobi-correct.png}}\;\;
-  \subfigure[]{\includegraphics[width=0.1\textwidth]{jacobi-incorrect1.png}}\;\;
-  \subfigure[]{\includegraphics[width=0.1\textwidth]{jacobi-incorrect2.png}}\;\;
-  \subfigure[]{\includegraphics[width=0.1\textwidth]{jacobi-incorrect3.png}}
-  \caption{Perturbations of Jacobi code; \subref{fig:jacobicorrect} is reference output.}
-  \label{fig:jacobi}
-  \vspace{-1em}
-\end{figure}
-
-\subsection{Verification case study 2: Turbulence calculation}
-
-\begin{figure}[t]\begin{minted}[fontsize=\scriptsize,breakindent=0em,linenos=false,xleftmargin=0em,breakafter=)]{fortran}
-!=stencil readOnce,(forward(depth=1,dim=3,irreflexive))*(backward(depth=1,dim=1))*(backward(depth=1,dim=2))+(forward(depth=1,dim=3))*(backward(depth=1,dim=1))*(reflexive(dim=2))+(forward(depth=1,dim=3))*(backward(depth=1,dim=1,irreflexive))*(backward(depth=1,dim=2,irreflexive))+(forward(depth=1,dim=3))*(backward(depth=1,dim=2))*(reflexive(dim=1))+(backward(depth=1,dim=1))*(backward(depth=1,dim=2,irreflexive))*(reflexive(dim=3))+(backward(depth=1,dim=1,irreflexive))*(backward(depth=1,dim=2))*(reflexive(dim=3))::x
-\end{minted}
-\caption{A complex specification inferred from
-  \textbf{Unified Model}\label{fig:smagorinsky}}
-\vspace{-1em}
-\end{figure}
-
-\noindent
-The Unified Model has an implementation of the Smagorinsky
-subgrid-scale model for calculating turbulence on which our
-inference yields 39 specs from 340 lines of code. We show
-one example in Figure~\ref{fig:smagorinsky}. 
-This specifies the access pattern to a 3-dimensional array (which we have renamed to be called \texttt{x}).
-The array is accessed 8 times in the computation.
-We tried introducing indexing errors by changing the offset in one dimension of one of the references to \texttt{x}.
-For example changing \mintinline{fortran}{x(i+1,j,k)} to \mintinline{fortran}{x(i,j,k)} or \mintinline{fortran}{x(i-1,j,k)}.
-
-There are 48 possible errors of this type of which only 6 were not 
-detected by the checker. The reason is that the inferred
-specification describes a family of access patterns. Some offset errors
-produce a stencil computation that belongs to the same family and goes
-undetected.
-One of the reasons this specification can catch many offset errors is that it is
-a \texttt{readOnce} specification. This prevents offset errors leading to
-duplicate access patterns be caught even if the resulting set of accesses are
-part of the same family.
-% /um/trunk/src/atmosphere/diffusion_and_filtering/turb_smagorinsky.F90
-% Tried introducing indexing errors replacing one of the indices of cx_rho_instance with -1,0,+1
-% 42 cases were caught
-% 6 cases were not caught
-\section{Related work and conclusions}
-\label{sec:discussion}
-
-Various prior approaches to specifying and verifying the behaviour
-of stencils have been based on fine-grained representations of the
-code. These essentially duplicate the indexing scheme of a stencil
-computation in their specifications.
-
-\citet{kamil2016verified} propose a technique they call
-\emph{verified lifting} for mapping low-level, possibly
-greatly optimised, stencil computations into a high-level
-predicate language using inductive program synthesis.
-The aim is that the high-level abstract specifications can
-be automatically translated into high-performance DSLs. Compared to our approach,
-their specification language is more fine grained, capturing
-the exact indexing pattern of a stencil. For example, the following
-is a post-condition generated by their
-compiler \textsc{stng} from a Fortran stencil computation~\cite [p.3]{kamil2016verified}:
-%
-\begin{align*}
-& \textit{post}(\texttt{a}, \texttt{b}) \equiv \forall \texttt{imin}+1
-\leq \texttt{i} \leq \texttt{imax}, \texttt{jmin} \leq \texttt{j} \leq
-\texttt{jmax} \\
-& \qquad \texttt{a}(\texttt{i},\texttt{j}) =
-\texttt{b}(\texttt{i}-1,\texttt{j}) + \texttt{b}(\texttt{i},\texttt{j})
-\end{align*}
-%
-Notably, the bounds of the stencil and the exact indexing pattern
-are captured. Our approach aims to be much more abstract, to
-facilitate human reasoning, and avoid low-level lexical
-errors. However, \textsc{stng} specifications are generated, \ie{},
-are never written by a programmer, avoiding this
-problem. Their inductive program synthesis approach allows
-specifications to extracted even from optimised and unrolled code. For
-example, the above post condition is generated from the code~\cite [p.3]{kamil2016verified}:
-%
-\begin{minted}{fortran}
-do j=jmin, jmax
-    t = b(imin, j)
-    do i=imin+1,imax
-        q = b(i, j);  a(i, j) = q + t; t = q
-    end do
-end do
-\end{minted}
-%
-Thus, \textsc{stng} can understand cross-loop data-dependencies
-in order to assign the above post-condition. CamFort is not currently
-able to ascribe a specification in the above case as the use of
-\texttt{t} on line 4 maps only to the definition on line 2, and not to
-that on line 4. Extending our approach to follow loop-carried dependencies is
-future work. An alternate approach for us would be to build on top of
-the work of Kamil~\emph{et al.} by inferring our specifications from
-\textsc{stng} (as an intermediate step).
-
-% GPGPU
-Our work has some similarities with efforts to verify kernels
-written for General-Purpose Graphics Processing Unit (GPGPU)
-programming, such as in \citet{Blom:2014:SoCP}. Stencils are a form of
-kernel, and GPGPU programming can be viewed as a massively parallel
-method of transforming a large matrix. However, that work focuses
-mainly on the synchronisation of such kernels and the avoidance of
-data races, while we are interested in the correctness of stencils
-embedded within a more typical general-purpose programming
-language. Other work on GPGPU computation, such as
-\citet{Zhang:2012:CGO}, has focused primarily on generating
-optimised code based on relatively simple specifications: to
-provide performance while keeping the programmer's effort within
-reason.
-
-% Sketching Stencils
-\citet{Solar-Lezama:2007:PLDI} give specifications of stencils using
-unoptimised ``reference'' stencils, coupled with partial
-implementations of stencils, which are then completed by a code
-generation tool. These kinds of specifications are just simple
-implementations, so this tool is useful for hand-written, optimised
-stencils. The primary purpose of this tool is optimisation rather than
-correctness, and the language of specification is much different and
-more elaborate than we envision for our system.
-
-% Pochoir
-\citet{Tang:2011:SPAA} define a specification language for writing
-stencils embedded in C++ (with Cilk extensions) that are then compiled
-into fast parallel programs based on trapezoidal decompositions with
-hyperspace cuts. The Pochoir specification language is used for
-describing the kernel, boundary conditions and shape of the
-stencil. Programs are first compiled in C++ with a template library
-that checks whether the annotations are used correctly, and then a
-Pochoir preprocessor may be applied to generate the high-performance
-parallel versions using Cilk.  Pochoir
-is aimed primarily at assisting programmers who may be reluctant to
-implement the high-performance tricky cache-oblivious
-``hypertrapezoidal'' algorithms for stencils. 
-%that are significantly faster than the typical looping implementations
-%of stencil computations.
-Like much of the related work, the main goal 
-is optimisation rather than verifying program correctness.
-
-By contrast, the work of \citet{Abe:2013:IPDPSW} has looked at the
-correctness problem explicitly by bringing a form of model-checking to
-verify certain stencil computations. This is in the context of
-decomposing stencils for parallel computation on multiple processors
- in partitioned global address
-space languages. %These are scenarios where an array is divided into
-%subarrays on multiple processors but each has the ability to access
-%the others' memory. The authors avoid the state explosion dilemma of
-%model-checking by relying upon the fact that accesses to the
-%``boundary elements'' will require a different method by virtue of the
-%non-locality of that memory. Since this access is performed in a
-%different manner, those can be identified automatically by static
-%analysis. To achieve this, 
-The authors have designed a new language to
-be used for writing stencil computations. Much of the specification
-effort goes towards describing the nuts and bolts of distributing the
-computation over multiple processors. The code for the stencil kernel
-is generated from a relatively high-level specification. This approach
-differs from CamFort because we are interested in integrating directly
-into existing, legacy codebases and long-established languages such as
-older versions of Fortran. We infer stencil specifications from
-Fortran code and check annotations on stencil computations within
-Fortran code.
-
-%There are many DSLs for stencil computation aimed at either
-%performance or verification, or both. Many eliminate out-of-bounds
-%errors, \eg{},~\citet{DBLP:journals/corr/abs-1109-0777}.
-
-\paragraph{Conclusion and future work}
-
-There is an increasing awareness
-of the need for verification techniques in
-science~\cite{post2005computational,oberkampf2010verification,orchard2014computational}.
-Our specification language is an approach in this direction, providing
-a system of lightweight specification for numerical code.
-Compared to other specification approaches (discussed above), our language is novel in that it does
-not aim to reify indexing in the specification, but instead
-provides abstract spatial descriptions which capture a large number
-of common patterns. %Our tool is useful for both detecting
-%errors and for specifying program behaviour to future-proof code and ease
-% maintenance.
-
-%To increase the likelihood of impact, we 
-%integrate smoothly with existing practices, providing 
-%specifications as comments and specification synthesis
-%specifications in the the tool.
-
-We evaluated our approach on top of CamFort, an open-source analysis
-tool for Fortran
-programs, running the inference process over a corpus of 1.1
-million lines of Fortran code. 
- Our motivation for this work was to
-provide an intermediate specification language to allow scientists to
-verify numerical computations.  Our evaluation showed that relatively
-simple stencil computations are the norm comprising mainly reflexive region constants which
-are less likely to give rise to programming errors.  However, around
-10\% of stencils covered more complicated program structures for which
-indexing errors are much more plausible. In the more detailed cases,
-with larger specifications, the importance and impact of such
-specifications grows.  We showed two examples of more complicated
-stencils, for which CamFort was able to detect a significant majority
-of possible indexing errors in our case studies.
-
-Our definition of a stencil computation meant that we generated no
-specification for a large number of iterated array accesses in the
-corpus, such as stencils that incorporates reductions. For
-future work, we will be looking into expanding the flexibility of 
-our inference approach to more of these examples. Furthermore, our specification
-language can be used more flexibly to give descriptions of data access
-patterns in any context, not just within a stencil.
-
-%\footnotetext{\url{https://github.com/camfort/camfort}}
+%
+The specification requires that, over the whole fragment, \texttt{u}
+is accessed with a centered pattern to depth of 1 in both dimensions
+(this is known as the \emph{five-point stencil}) and \texttt{v} is
+accessed in a neighbourhood bounded forwards to depth of $1$ in the
+first dimension and backward to a depth of $1$ in the second
+dimension.
+
+We provide a verification tool for such specification which can check
+the correctness of code against the specification, and can also infer
+specifications, synthesising these as comments in the
+source code. The main features and benefits of our approach are as follows:
+%
+\begin{itemize}
+\item The main use case is that a programmers writes a specification
+  first (perhaps informed by their numerical analysis), then writes
+  the corresponding source code, and lastly has the tool check its
+  conformance to the specifications. This specify-and-check approach
+  removes the usually ad hoc testing effort, and provides the
+  opportunity for easy re-verification in the event of any refactoring
+  of software maintenance activities.
+%
+\item The specification provides a concise description which can be
+  preserved in the source code to provide documentation to future
+  developers, contrasting with ad hoc tests which are usually
+  discarded or specialised to the development environment. Separate from the
+  verification procedure, the infer-and-synthesise mode of the tool
+  provides a way to automatically generate additional useful
+  documentation for future users.
+%
+\item Specifications are integrated into the code as comments for easy
+integration with existing tools, as well as
+  documentation systems (\eg{}, we support Doxygen and FORD syntax).
+%
+\item Our specification format is deliberately abstract, with a small
+  number of combinators that \emph{do not involve any indexing
+    expressions}, \eg{} does not involve expressions like
+  \texttt{a(i+1,j-1)}. This contrasts with other stencil specification
+  approaches. For example, the work of Kamil et al.~\cite{kamil2016verified}
+  provides full-functionality specifications of stencil computations from which
+  equivalent DSL code can be generated.  If a user wishes to specify, say, the
+  Naiver-Stokes stencil above, they must write a specification which is very
+  similar, involving textually dense array-indexing expressions. Therefore any
+  low-level mistakes that could be made programming complex indexing code could
+  also be made when writing a specification. Our specifications are much more
+  abstract, and also more lightweight with the aim of aiding adoption by
+  scientists.
+\end{itemize}
+
+It should be noted that our verification tool does not deal in the
+class of bugs that can be detected automatically (\emph{push-button
+  verification}). Instead, stencil computation bugs must largely be
+identified relative to a specification of the intended behaviour.
+
+In this paper, we make the following contributions:
+%
+\begin{itemize}
+\item we introduce our specification language for stencils
+  that captures many common forms of data access pattern
+  (\S~\ref{sec:lang});
+
+\item we detail checking and inference algorithms for our
+  specifications (\S~\ref{sec:analysis}), derived from a
+  denotational model for the specificaiton language
+  (\S~\ref{sec:semantics});
+
+\item we provide an implementation of our approach as an extension to
+  CamFort, an open-source program analysis tool for Fortran;
+
+\item we report on a quantitative study of stencil computations on a
+  corpus of numerical Fortran programs (ranging from small to large)
+  (\S~\ref{sec:evaluation}), totalling one million lines of code. Our tool
+  identifies and infers specifications for 35,000 stencil computations in the
+  corpus. Approximately 10\% of the stencils we found are non-trivial,
+  corresponding to code which is a possible source of errors. This validates
+  our hypothesis that the majority of stencil computations have a regular
+  shape, and validates the design of our language in its capability to capture
+  many core patterns.
+
+\item We give a detailed verification case study of two
+  particular stencil computation commonly used in scientifc computing,
+  simulating programming errors. Our approach is able to detect all
+  possible indexing errors for these computations.
+\end{itemize}
+%
+
 \bibliography{references}

### 2016-11-02 paper.tex
tweaking a bit the formatting

@@
 \newcommand{\trule}[1]{{\footnotesize{#1}}}
-
-\include{results}
+%\include{results}
 
@@
   Specifications}
-\titlebanner{DRAFT -- Do not distribute}
+%\titlebanner{DRAFT -- Do not distribute}
 \authorinfo{}{}{}
-
-\preprintfooter{DO NOT DISTRIBUTE}
+%\preprintfooter{DO NOT DISTRIBUTE}
 
@@
 \maketitle
+
+\keywords{program verification, specification, stencils}
+
+\bibliographystyle{abbrvnat}
+
+\section{Introduction}
 
@@
 %%
-\begin{minted}[firstnumber=20,xleftmargin=4em]{fortran}
+\begin{minted}[firstnumber=20,xleftmargin=2em]{fortran}
 du2dx = ((u(i,j)+u(i+1,j))*(u(i,j)+u(i+1,j))+   &

### 2016-11-02 paper.tex
starting new early section on the model

@@
 
+\newcommand{\domainVal}{\mathbb{Z}_\infty}
+\section{Stencil specifications}
+
+\section{The language}
+
+We define a simple set-theoretic denotational model of the semantics
+of our stencil specification language. This has a number of purposes:
+(1) it serves to explain exactly the meaning of our specifications
+(2) it is 
+used as the basis of the inference, checking, and synthesis algorithms
+(\S~\ref{sec:analysis}); (3) justifies the equational theory and
+approximations of the previous section; and (4) can be used to guide
+correct implementations. The model domain is over sets of vectors
+which we call \emph{index schemes}, which we introduce first before
+the model in \S~\ref{subsec:model}.
+
 \bibliography{references}

### 2016-11-08 paper.tex
rebuilding section on model

@@
 % SYNTAX OPERATIONS AND PREDICATES
-\newcommand{\neigh}{\textsf{neigh}}
+%\newcommand{\neigh}{\textsf{neigh}}
 \newcommand{\arrayTy}{\textsf{array}}
@@
 
+\begin{definition}[Base induction variable]
+  An integer variable is a \emph{base induction variable} if is the
+  control variable of a ``for'' loop (or equivalent, \eg{},
+  \texttt{do} in Fortran for our implementation) incremented by $1$
+  per iteration. The variable is interpreted as a based induction
+variable only within the scope of the loop body. For brevity we
+say \emph{induction variable}, which are ranged over by $i, j, k$
+throughout.
+%TODO: consider mentioning derived, non-based induction variables
+% as we did before.
+\end{definition}
+
+\begin{definition}[Array subscripts and indices]
+  An \emph{array subscript}, denoted $a(\bar{e})$, is an expression
+  which reads from an array $a$ at an \emph{index} specified by a
+  comma-separated sequence of integer expressions denoted $\bar{e}$ or 
+  in expanded form as $(e_1, \ldots, e_n)$. An
+  \emph{index} $\bar{e}$ is called \emph{relative} if it is defined in
+  terms of induction variables.
+\end{definition}
+
+\begin{definition}[Neighbourhood index]
+\label{def:neighbour}
+  For an array subscript $a(\bar{e})$ an index $\bar{e}$ is called a
+  \emph{neighbourhood index} if every $e \in \bar{e}$ is of the form
+  $e \equiv i$ or $e \equiv i + c$ or $e \equiv i - c$ where $c$ is an
+  integer constant. That is, a neighbourhood index is a relative index
+  where each component is a constant translation of a base induction
+  variable. (The relation $\equiv$ here identifies terms up-to
+  commutativity of $+$ and the inverse relationship of $+$ and $-$
+  \eg{}, $(-b) + i \equiv i - b$).  %We classify neighbourhood indices
+%  by the predicate $\neigh_I$ where $I$ is a set of induction
+%  variables. We omit $I$ when clear from the context.
+\end{definition}
+
 \section{The language}
@@
 
+\begin{definition}
+  An \emph{index scheme} is a vector of size $n$ (an $n$-vector) with
+  values drawn from $\domainVal = \mathbb{Z} \cup \{\infty\}$.  Values
+  in $\mathbb{Z}$ represent constant offsets of neighbourhood indices
+  (expressions of the form $i \pm c$ where $i$ is an induction
+  variable and $c$ is a constant integer, Def.~\ref{def:neighbour},
+  p.~\pageref{def:neighbour}). $\infty$ represents any indexing
+  behaviour--- it is a wildcard.  Throughout $u, v, w$ range over
+  index schemes and $u_i$ denotes the $i^{th}$ element of $u$. We
+  write $(\domainVal)^n$ for the set of $n$-dimensional schemes.
+\end{definition}
+
+Indexing schemes give an abstract
+representation of array subscripts in the source language.
+Our model represents the meaning of stencil specifications
+as indexing schemes, which explain the range of possible
+indexing behaviours which are consistent with the specification.
+The relationship between indexing schemes and source language
+terms is given precisely by the following function \textsf{schematic} which
+maps syntactic index expressions to indexing schemes. This
+is used in inference and checking.
+
+\newcommand{\schemeOne}{\textsf{s}_I}
+%\newcommand{\schemeOne}{\textsf{schema$^1_I$}}
+\newcommand{\schemeN}{\textsf{schematic}_I}
+%\newcommand{\schemeN}{\textsf{schema$^n_I$}}
+\begin{definition}%[Schematic]
+\label{def:schematic}
+Let $\schemeOne$ be a partial function, parameterised by a set of
+induction variables $I$, mapping expression syntax to
+$\domainVal$ values:
+%%
+\begin{align*}
+\schemeOne(e)
+ & =  \begin{cases}
+c & e \equiv i + c \;\; \wedge \;\; i \in I \\
+-c & e \equiv i - c \;\; \wedge \;\; i \in I \\
+\infty & \textit{$I \not\subseteq \textit{FV}(e)$} \\
+\bot   & \textit{otherwise}
+\end{cases}
+\end{align*}
+where $\textit{FV}(e)$ is the set of free variables in $e$. In the
+first two cases, expressions that are a constant offset of an 
+induction variable (\ie{}, neighbourhood indices) are mapped to the
+offset $c$ or $-c$ $\in \domainVal$. In the third case, if the
+expression is not a neighbourhood index, but neither is it a relative
+index (\ie{}, it is not defined in terms of an induction variable) then it
+is mapped to $\infty \in \domainVal{}$. Otherwise, $\schemeOne(e)$
+is undefined (returns $\bot$).
+
+The partial function $\schemeN$, also parameterised by a set of
+induction variables $I$, uses the above operation $\schemeOne$ to
+ maps from the syntax of indices to index schemes, \ie{} values of 
+$\domainVal^n$. It is defined on the syntax of index
+expressions $\bar{e} = (e_1, ..., e_n)$ as:
+\begin{align*}
+\schemeN(e_1, ..., e_n) & =
+[\schemeOne(e_1) ... \schemeOne(e_n) ]
+\quad \textit{iff} \;\; \forall x . \schemeOne(e_x) \neq \bot
+\end{align*}
+%
+Thus, for an indexing expression, if each component is a neighbourhood
+index, then $\schemeN$ maps it to an indexing scheme.
+\end{definition}
+
 \bibliography{references}

### 2016-11-11 paper.tex
introduction on stencil language

@@
 
-\newcommand{\stenRefl}[1]{\term{reflexive} \, (\term{dim=}#1)}
+\newcommand{\stenRefl}[1]{\term{pointwise} \, (\term{dim=}#1)}
 \newcommand{\stenFwd}[3]{\term{forward} \, (\term{depth=}#1,
@@
   \term{dim=}#2{#3})}
-\newcommand{\irrefl}{\texttt{irreflexive}}
+\newcommand{\irrefl}{\texttt{nonpointwise}}
 
-\newcommand{\stenReflS}[1]{\term{refl} \, (\term{dim=}#1)}
+\newcommand{\stenReflS}[1]{\term{point} \, (\term{dim=}#1)}
 \newcommand{\stenFwdS}[2]{\term{fwd} \, (\term{depth=}#1,
@@
   \term{dim=}#2)}
-\newcommand{\irreflS}{\texttt{irrefl}}
+\newcommand{\irreflS}{\texttt{nonpoint}}
 
@@
   \term{dim=}#2, #3)}
-\newcommand{\stenReflSR}[1]{\term{refl} (\term{dim=}#1)}
+\newcommand{\stenReflSR}[1]{\term{point} (\term{dim=}#1)}
 
@@
 \begin{minted}[linenos=false,xleftmargin=4em,breakindent=3em]{fortran}
-!= stencil centered(depth=1,dim=1)*reflexive(dim=2) + centered(depth=1,dim=2)*reflexive(dim=1) :: u
+!= stencil centered(depth=1,dim=1) * pointwise(dim=2) + centered(depth=1,dim=2) * pointwise(dim=1) :: u
 
@@
 
+\paragraph{Notation and convention}
+%\label{sec:notation}
+
+\renewcommand*{\arraystretch}{0.8}
+For the target language, $e$ ranges over expressions (may be impure) and $v$
+over its (imperative) variables. %By
+%``variable'', we mean in the imperative sense (named binders to memory cells).
+
 \begin{definition}[Base induction variable]
@@
 
-\section{The language}
+\dnote{Define: Stencil kernel, or maybe stencil assignment,
+ and stencil much earlier than previously?}
+
+Specification declarations associate (via
+the operator $\texttt{::}$) a specification to one or more array
+variables $\bar{v}$.% specifying the shape of indexing expressions for
+%arrays $\bar{v}$ that are simultaenously permitted and expected within
+%the code to which the specification is given.
+
+\begin{definition}[Consistency, informal]
+  A stencil specification defines a \emph{region} which, informally,
+  is an n-dimensional rectangle which is possibly infinite in some
+  dimensions. A stencil kernel in code is \emph{consistent} with a
+  specification on array variable $a$ if every array index on $a$
+\end{definition}
+
+\subsection{Specification syntax}
+\label{sec:syntax}
+
+Figure~\ref{fig:syntax} gives the syntax of stencil specifications,
+which we introduce in stages below.  The top-level is given by the
+\nonterm{specification} production which splits into either a
+\nonterm{regionDec} (region declaration) or a \nonterm{specDec}
+(specification declaration).
+
+Regions comprise \emph{region constants} which are 
+combined via operators \term{+} and \term{*}.
+
+\paragraph{Region constants}
+
+Region constants specify a pattern of continguous array accesses in a
+single dimension.  Region constants are either \term{pointwise},
+\term{forward}, \term{backward}, or \term{centered}.  Each region
+constant has a dimension identifier $d$ given by a natural number
+greater than 0. Each constant except \term{pointwise} is given a depth
+parameter $n$, a natural number greater than 0; \term{pointwise}
+regions implicitly have a depth of $0$.
+
+A \term{forward} region of depth $n$ specifies continguous
+neighbourhood indices in dimension $d$ ranging from $i$ to $i + n$
+inclusively for some induction variable $i$.  Similarly, a
+\term{backward} region of depth $n$ specifies contiguous neighbour
+indices in dimension $d$ from $i$ to $i - n$, and \term{centered}
+of depth $n$ from $i - n$ to $i + n$. A \term{pointwise}
+stencil specifies a neighbour index $i$. For example the
+following shows three specifications with three consistent stencil
+kernels reading from arrays \term{a}, \term{b}, and \term{c} respectively:
+%%
+\begin{minted}{fortran}
+!= stencil forward(depth=2, dim=1) :: a
+d(i, 0) = a(i, 0) + a(i+1, 0) + a(i+2, 0)
+!= stencil backward(depth=2, dim=1) :: b
+d(i, j) = b(i, j) + b(i-1, j) + b(i-2, j)
+!= stencil centered(depth=1, dim=2) :: c
+d(i, j) = (c(j-1) + c(j) + c(j+1))/3.0
+\end{minted}
+%%
+Note that every dimension need not be specified for; in the above, the
+specifications on lines $1$ and $5$ are the principal specifications
+for code on lines $2$ and $6$, but the specification on line
+$3$ is not principal for line $4$;
+we return to the notion of principality later.
+
+The \term{forward}, \term{backward}, and \term{centered} regions may
+all have an additional attribute \term{nonpointwise} which marks that
+the region does not include indices which are purely base induction
+variables, \ie{}, offset by $0$.  For example, the following is a
+\term{nonpointwise} \term{backward} stencil
+%
+\begin{minted}{fortran}
+!= stencil backward(depth=2, dim=1, nonpointwise) :: a
+b(i) = a(i-1) + a(i-2)
+\end{minted}
+%
+The \term{nonpointwise} attribute cannot be used on \term{pointwise}
+regions.
+
+\section{A model}
 
@@
 
+\begin{definition}[Generalisation]
+Given two schemes $v, w$ of dimensionality $n$,
+then $w$ is said to \emph{generalise} $v$
+(or conversely $v$ \emph{specialises} $w$),
+written $v \preceq_n w$, iff the following holds:
+%%
+\[
+v \preceq_n w \;\; \Leftrightarrow  \;\;
+  \forall i\!\in\!\{ 1, \ldots, n \} . \, (w_i = \infty \, \vee \, v_i = w_i)
+\]
+%%
+That is, two schemes are related by the (reflexive and transitive) binary relation
+$\preceq_n$ over $(\mathbb{Z}_{\infty})^n$ if in each dimension
+the two schemes are equal or $w_i = \infty$
+in place of some $\mathbb{Z}$ in $v_i$ for $v \preceq_n w$.
+Thus, the element $\infty$ generalises any index in the same dimension.
+%
+\end{definition}
+
+\begin{example}
+Consider the following three indexing schemes
+corresponding to array indexing terms
+$(i, j+1)$, $(1, j+1)$ and $(3, j)$:
+%
+\begin{align*}
+\begin{array}{ll}
+\schemeN{(i, j+1)} & =\vtwoh{0}{1} \\
+\schemeN{(1, j+1)} & = \vtwoh{\infty}{1} \\
+\schemeN{(3, j)} & = \vtwoh{\infty}{0}
+\end{array}
+\end{align*}
+There is one notion of generalisation between these
+schemes, namely that $\vtwoh{0}{1} \preceq_2 \vtwoh{\infty}{1}$.
+At the level of syntax, this translates to saying
+$(i, j+1)$ can be instantiated (specialised) to $(3, j+1)$.
+
+As an example of undefined \textsf{schematic},
+$\textsf{schematic}(i*2, j) = \bot$ as its first index is a
+non-neighbour relative index, which is outside the scope of our
+specification language.
+\end{example}
+
+Our terminology of \emph{indexing schemes} and
+\emph{generalisation} is by analogy
+with the \emph{type schemes} and \emph{type generalisation} of
+polymorphic types in ML~\cite{milner1978theory}. Our notion is
+much more rigid however, as our schemes do not bind names.
+
+\todo{Consider what to do regarding stencil model
+- can give previous model, or can try to base on the work cited
+on Wikipeadia that we discussed.}
+
+\subsection{Denotational model}
+
+Our model is given by the interpretation function $\interp{-}_n$
+(where $n$ is the maximum dimensionality of the specification being
+modelled) mapping closed\footnote{That is, we assume there are no
+  occurrences of \textit{rvar} in a specification being modelled.  Any
+  \emph{open} specification containing region variables can be made
+  closed by straightforward syntactic substitution with a (closed)
+  \textit{region}.} specifications to sets of
+$n$-dimensional indexing schemes. The interpretation is overloaded on
+\emph{regions} in Figure~\ref{fig:region-model} and on the top-level
+of a specification \textit{spec} in Figure~\ref{fig:spatial-model}.
+
+We first define some intermediate notations and definitions.
+
+\begin{definition}A \emph{single-entry vector} of size $n$, denoted
+$\singleEntry{r}{n}$, is a vector where the $r^{th}$ entry is $1$
+and all others are $\infty$, \eg{}, $\singleEntry{1}{2} =
+\vtwoh{1}{\infty}$ and $\singleEntry{2}{3} = \vthreeh{\infty}{1}{\infty}$.
+Thus, a single-entry vector is an indexing scheme which generalises
+all other schemes that have $1$ as their $r^{th}$ entry,
+describing a neighbour offset of $1$.
+\end{definition}
+
+\begin{definition}
+A \emph{zeroed single-entry vector} of size $n$, denoted
+$\zeroEntry{r}{n}$, is a vector where the $r^{th}$ entry is $0$ and all others
+are $\infty$, \eg{}, $\zeroEntry{1}{2} = \vtwoh{0}{\infty}$.
+Similarly to the single-entry vector, zeroed single-entry vectors
+generalise all other schemes
+that have $0$ as their $r^{th}$, representing indices at
+the ``origin'' in dimension $r$.
+\end{definition}
+
+\begin{figure}[t]
+\vspace{-0.9em}
+\input{region-model.tex}
+%\textit{$a^m \in A$ means there are $n$ copies of $a$ in
+%  the multi-set $A$}
+\caption{Model of regions,
+$\interp{-}_n : \textit{region} \rightarrow \mathcal{P}(\domainVal{}^n)$}
+\label{fig:region-model}
+\vspace{-1em}
+\end{figure}
+
+\noindent
+The first four equations of Fig.~\ref{fig:region-model}
+give the model of the region constants as sets of indexing
+schemes. We illustrate with an example.
+
+\begin{example}
+For a $2$-dimensional stencil computation, then
+\begin{align*}
+& \interp{\stenCenS{2}{1}}_2 \\[-0.1em]
+& \qquad = \; \{i\singleEntry{1}{2} \,|\, i \in \{\text{-}2,..., 2\}\!\setminus\!0 \}
+  \cup  \{ \zeroEntry{1}{2} \,|\, \textit{true} \} \\[-0.2em]
+& \qquad = \; \{-2\singleEntry{1}{2}, \, -\singleEntry{1}{2}, \,
+  \singleEntry{1}{2}, \, 2\singleEntry{1}{2}\} \cup
+  \{\zeroEntry{1}{2}\} \\[-0.2em]
+& \qquad = \; \{\vtwoh{-2}{\infty}, \vtwoh{-1}{\infty}, \vtwoh{0}{\infty},
+\vtwoh{1}{\infty}, \vtwoh{2}{\infty}\}
+\end{align*}
+Note the absorbing behaviour of $\infty$ with respect to
+multiplication.
+\end{example}
+\noindent
+The final two equations of Fig.~\ref{fig:region-model} give the models
+of $\term{+}$ and $\term{*}$. For $\term{+}$ this is the
+straightforward union of the models of subterms. The model of
+$\term{*}$ takes the tensor $\otimes$ of subterm models. The tensor is
+more involved; we give some informal intuition first.
+
+% r+s means that indices must be consistent with either r or s 
+% r*s means that indices must be consistent with both r and s at the
+% same time
+
+% specifications correspond to a set of schemes
+
+% r+s means an index must be generalised by a scheme in either r or s
+%       and every scheme in r must have an index which specialises it
+%       and every scheme in s must have an index which specialises it
+
+% r*s means an index must be generalised by a scheme in r and a scheme in s
+%       and every pairs of schemes in r and s must have an index which 
+%           specialises both 
+%
+
+% r+s means indices must be consistent with either r or s
+%           and there are indices why are generalised by r and s
+% (there are indices which are consistent with every part of r or s)
+% (every part of r and s independently have an index which is generalised by it)
+
+% r*s means indices must be consistent simultaenously with both r and s
+%           and there are indices which are generalised 
+%
+
+%  forward(depth=1,dim=2) & pointwise(dim=1) 
+% | forward(depth=1,dim=1) & pointwise(dim=2)
+
+% (o, p) \preceq [[ ]]
+
+\paragraph{$\otimes$ of models}
+
+The intuition behind $\otimes$ is that it takes
+all possible pairs of indexing schemes,
+and treats each pair as lower and upper bounds of an 
+$n$-dimensional rectangle. From these lower and upper bounds
+
+ treating them as the
+lower and upper bounds of an $n$-dimensional rectangle from which the
+remaining vertices of the rectangle are generated (like a bounding
+box) \eg{}:
+%%
+\begin{align*}
+\{\vtwohs{1}{2}, \vtwohs{5}{6}\,\} \otimes \{\vtwohs{3}{4}\} =
+\{\vtwohs{1}{2}, \vtwohs{3}{2}, \vtwohs{1}{4}, \vtwohs{3}{4}, 
+   \vtwohs{3}{6}, \vtwohs{5}{4}, \vtwohs{5}{6}\,\}
+\end{align*}
+%%
+The tensor takes a Cartesian product of the two models,
+giving all pairs of schemes which are combined using the
+\emph{pairwise permutation} written $\bowtie$ (defined later).
+
+\section{Checking and inference}
+
+\subsection{Specification checking}
+
+\subsection{Specification inference}
+
 \bibliography{references}
@@
 
-%%  LocalWords:  refactoring affine parameterised irreflexive atMost
+%%  LocalWords:  refactoring affine parameterised nonpointwise atMost
 %%  LocalWords:  centered readOnce atLeast discretisation Equational

### 2016-11-11 paper.tex
on an informal definition of consistency

@@
   specification on array variable $a$ if every array index on $a$
+  is a neighbourhood index within the $n$-dimensional rectangle
+  specified by the 
 \end{definition}

### 2016-11-11 paper.tex
giving stencil definition earlier

--- 2016-11-11 paper.tex
@@
 \begin{definition}[Base induction variable]
-  An integer variable is a \emph{base induction variable} if is the
+  An integer variable is a \emph{base induction variable} (or simply
+\emph{induction variable} for brevity) if it is the
   control variable of a ``for'' loop (or equivalent, \eg{},
@@
   per iteration. The variable is interpreted as a based induction
-variable only within the scope of the loop body. For brevity we
-say \emph{induction variable}, which are ranged over by $i, j, k$
-throughout.
+variable only within the scope of the loop body. Induction varaibles 
+are ranged over by $i, j, k$ throughout.
 %TODO: consider mentioning derived, non-based induction variables
@@
   comma-separated sequence of integer expressions denoted $\bar{e}$ or
-  in expanded form as $(e_1, \ldots, e_n)$. An
-  \emph{index} $\bar{e}$ is called \emph{relative} if it is defined in
-  terms of induction variables.
+  in expanded form as $(e_1, \ldots, e_n)$. An \emph{index} $\bar{e}$
+  is called \emph{relative} if each $e \in \bar{e}$ is defined in
+  terms of an induction variable.
+
+  Relative indices contrast with absolute indices where each $e \in
+  \bar{e}$ is a constant integer expression relative to the containing
+  loop, \ie{}, $e$ is not defined in terms of an induction variable.
 \end{definition}
@@
   commutativity of $+$ and the inverse relationship of $+$ and $-$
-  \eg{}, $(-b) + i \equiv i - b$).  %We classify neighbourhood indices
-%  by the predicate $\neigh_I$ where $I$ is a set of induction
-%  variables. We omit $I$ when clear from the context.
+  \eg{}, $(-b) + i \equiv i - b$). 
 \end{definition}
 
-\dnote{Define: Stencil kernel, or maybe stencil assignment,
- and stencil much earlier than previously?}
+\begin{definition}[Stencil computations]
+  Let $a$ be an array of dimensionality $n$ and $\overline{b}$ a
+  collection of arrays of arbitrary (possible differing)
+  dimensionalities.  
+  A \emph{stencil computation} comprises an iteration by a set $I$ of at
+  most $n$ induction variables 
+  over a subset of the index space of $a$. Elements of $a$ are determined by an
+  assignment $a(\bar{e}) = e_r$ where each $e \in
+  \bar{e}$ is either a neighbourhood or constant index.
+  For all arrays $b \in \bar{b}$ and array subscripts $b(\var{e'})$ in
+  $e_r$ and that flow to $e_r$, then each $e' \in \bar{e'}$ is also
+  either a neighbourhood or constant index.
 
-Specification declarations associate (via
-the operator $\texttt{::}$) a specification to one or more array
-variables $\bar{v}$.% specifying the shape of indexing expressions for
+  We refer to the assignment $a(\bar{e}) = e_r$ and all the associated
+  statements with dataflow to $e_r$ as the \emph{stencil kernel}.
+\label{def:stencil}
+\end{definition}
+
+Specification declarations associate to a stencil kernel 
+a specification for the array access on one or more array
+variables $\bar{v}$ within that kernel.
+
+% specifying the shape of indexing expressions for
 %arrays $\bar{v}$ that are simultaenously permitted and expected within
@@
 \begin{definition}[Consistency, informal]
-  A stencil specification defines a \emph{region} which, informally,
-  is an n-dimensional rectangle which is possibly infinite in some
-  dimensions. A stencil kernel in code is \emph{consistent} with a
-  specification on array variable $a$ if every array index on $a$
+  A stencil specification defines a \emph{region} associated to an
+  array variable $a$ against which a stencil kernel should be
+  \emph{consistent}. Informally, the region defined by a specification
+  is an $n$-dimensional rectangle crossing the origin.
+  Consistency is then a two way requirement.
+  On the one hand, all array subscripts contributing to the kernel
+  should be consistent with the region; on the other, all parts of the
+  region should have an index which fits the space. 
 \end{definition}
+\todo{Improve the above- haven't quite introduce enough yet to say
+more}
+
+\begin{figure}[t]
+\vspace{-0.5em}
+\begin{align*}
+\def\arraystretch{1.2}
+\setlength{\arraycolsep}{0.2em}
+\newcommand{\dimTy}{\mathbb{N}_{>0}}
+\begin{array}{rl}
+\nonterm{specification} ::= & \nonterm{regionDec} \mid \nonterm{specDec} \\
+\nonterm{specDec} ::= & \term{stencil} \; \nonterm{spec} \;
+                        \texttt{::} \; v \\
+\nonterm{regionDec} ::= &  \texttt{region} \; \texttt{::} \; \nonterm{rvar} \; \texttt{=} \;
+                         \nonterm{region}\\[0.4em]
+%\nonterm{spec} ::= & \nonterm{spatial} \mid \nonterm{temporal}
+%\\[1em]
+\nonterm{spec} ::= & [\nonterm{mult},] \; [\nonterm{approx},] \; \nonterm{region} \\
+\nonterm{mult} ::= & \term{readOnce} \\
+\nonterm{approx} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
+\nonterm{region} ::= & \nonterm{rvar} \; \\
+\multicolumn{2}{l}{\qquad\qquad \mid \; \stenRefl{\dimTy}} \\
+\multicolumn{2}{l}{\qquad\qquad \mid \;\stenFwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{nonpointwise}]}} \\
+\multicolumn{2}{l}{\qquad\qquad \mid \; \stenBwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{nonpointwise}]}} \\
+\multicolumn{2}{l}{\qquad\qquad \mid \; \stenCen{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{nonpointwise}]}} \\
+\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{region} \, \term{+}
+  \, \nonterm{region} \; \mid \; \nonterm{region} \; \term{*} \; \nonterm{region}} \\[0.5em]
+%\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{rvar}}  
+%\\[0.5em]
+%\nonterm{temporal} ::= \; & \term{dependency} \; (v \; \{ , v \}) [, \texttt{mutual}]
+%  \\[0.5em]
+\nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+\\[-0.5em]
+\end{array}
+\end{align*}
+\caption{Specification syntax (EBNF grammar)}
+\label{fig:syntax}
+\vspace{-0.8em}
+\end{figure}
 
@@
 stencil specifies a neighbour index $i$. For example the
-following shows three specifications with three consistent stencil
-kernels reading from arrays \term{a}, \term{b}, and \term{c} respectively:
+following shows four specifications with four consistent stencil
+kernels reading from arrays \term{a}, \term{b}, \term{c} and \term{d} respectively:
 %%
@@
 != stencil forward(depth=2, dim=1) :: a
-d(i, 0) = a(i, 0) + a(i+1, 0) + a(i+2, 0)
+e(i, 0) = a(i, 0) + a(i+1, 0) + a(i+2, 0)
 != stencil backward(depth=2, dim=1) :: b
-d(i, j) = b(i, j) + b(i-1, j) + b(i-2, j)
+e(i, j) = b(i, j) + b(i-1, j) + b(i-2, j)
 != stencil centered(depth=1, dim=2) :: c
-d(i, j) = (c(j-1) + c(j) + c(j+1))/3.0
+e(i, j) = (c(j-1) + c(j) + c(j+1))/3.0
+!= stencil pointwise(dim=3) :: d
+e(i, j) = d(0, 0, i)
 \end{minted}
@@
 Note that every dimension need not be specified for; in the above, the
-specifications on lines $1$ and $5$ are the principal specifications
-for code on lines $2$ and $6$, but the specification on line
+specifications on lines $1$ and $5$ and $7$ are the principal specifications
+for code on lines $2$ and $6$ and $8$, but the specification on line
 $3$ is not principal for line $4$;
@@
 %
-The \term{nonpointwise} attribute cannot be used on \term{pointwise}
-regions.
+The \term{pointwise} regions cannot have a \term{nonpointwise} attribute.
 

### 2016-11-11 paper.tex

@@
 Regions comprise \emph{region constants} which are 
-combined via operators \term{+} and \term{*}.
+combined via region operators \term{+} and \term{*}.
 
@@
 
+\paragraph{Combining regions}
+
+The region operators \term{+} and \term{*} provide ways to combine
+regions in a disjunctive and conjuctive fashion respectively.
+
+The conjuction of two regions $r \term{*} s$ means that any indices
+in the specified code must be consistent with both $r$ and $s$ at
+the same time. Dually, there must be an index which is consitent
+with every component of region $r$ and
+
 \section{A model}
@@
   values drawn from $\domainVal = \mathbb{Z} \cup \{\infty\}$.  Values
-  in $\mathbb{Z}$ represent constant offsets of neighbourhood indices
-  (expressions of the form $i \pm c$ where $i$ is an induction
-  variable and $c$ is a constant integer, Def.~\ref{def:neighbour},
-  p.~\pageref{def:neighbour}). $\infty$ represents any indexing
-  behaviour--- it is a wildcard.  Throughout $u, v, w$ range over
+  in $\mathbb{Z}$ represent the offsets of neighbourhood indices
+  (\ie{}, the constant integer $c$ in an indexing expression of the
+  form $i \pm c$, Def.~\ref{def:neighbour},
+  p.~\pageref{def:neighbour}) and $\infty$ represents any constant
+  inddex. Throughout $u, v, w$ range over
   index schemes and $u_i$ denotes the $i^{th}$ element of $u$. We
@@
 
-\todo{Consider what to do regarding stencil model
-- can give previous model, or can try to base on the work cited
-on Wikipeadia that we discussed.}
+\begin{definition}[Quotienting schemes]
+A scheme $u$ can be quotiented by a set of dimension identifies
+by the operator $/_{n,I}$ where $I$ is a set of dimension
+identifiers drawn from $\mathbb{N}_{\leq n}$ where 
+\begin{equation*}
+u/_{n,I} = [ ]
+\end{equation*}
+\end{definition}
 
@@
 
-The intuition behind $\otimes$ is that it takes
-all possible pairs of indexing schemes,
-and treats each pair as lower and upper bounds of an 
-$n$-dimensional rectangle. From these lower and upper bounds
-
- treating them as the
-lower and upper bounds of an $n$-dimensional rectangle from which the
-remaining vertices of the rectangle are generated (like a bounding
-box) \eg{}:
+The intuition behind $\otimes$ is that it takes all possible pairs of
+indexing schemes, and treats each pair as lower and upper bounds of an
+$n$-dimensional rectangle. From these lower and upper bounds, 
+the remaining vertices of the $n$-dimensional rectangle are
+generated (like a bounding box) \eg{}:
 %%
@@
 \emph{pairwise permutation} written $\bowtie$ (defined later).
+Pairwise permutation generates all possible interleavings of two
+vectors. Some care is then taken over
+how this interacts with $\infty$: pairwise permutation
+corresponds to non-deterministic choice, in each dimension $i$,
+between two values $u_i$ and $v_i$, but if either
+is $\infty$ then the resulting pairwise permutation is
+filtered out if the dimension $i$ has \emph{any other scheme
+in the model where $u_i \neq \infty$}, \eg{}:
+\begin{align*}
+\{\vtwoh{1}{\infty}\} \otimes \{\vtwoh{3}{4}\} =
+\{\vtwoh{1}{4}, \vtwoh{3}{4}\}
+\end{align*}
+%
+where $\vtwoh{1}{\infty}$ and $\vtwoh{3}{\infty}$ have been filtered.
+
+The notion of which dimensions $i$ have \emph{any scheme in the model
+where $u_i \neq \infty$} is represented by the \emph{constrained
+dimensions} intermediate function:
+\begin{definition}%[Constrained dimensions]
+For an $n$-dimensional model $M$, its \emph{constrained dimensions}
+is the set of dimension identifiers for which at least one scheme $u \in M$ has
+a non-$\infty$ value in that dimension. That is:
+%
+\begin{align*}
+\effdims{M}{n} =
+\bigcup_{u \in M} \{i \mid i \in \{1,\ldots,n\} \, \wedge \, u_i \neq
+  \infty\}
+\end{align*}
+For example, $\effdims{\{\vtwoh{0}{\infty},\vtwoh{\infty}{\infty}\}}{2}
+ = \{1\}$.
+\end{definition}
+\begin{definition}The tensor $\otimes_n$ of $n$-dimensional
+models is defined:
+\begin{alignat*}{2}
+& M \otimes_n N =
+  \Bigg\{x \; \Bigg| \;
+    \parbox{5.2cm}{$(u, v) \in M \times N $ \\[0.1em]
+                 $\;\; \wedge \;\, j \in (\effdims{M}{n} \cup
+                   \effdims{N}{n})$\\[0.1em]
+                  $\;\; \wedge \;\, i \in \{1, \ldots, 2^n\} \, \wedge
+                  \, x = (u \bowtie v)_i$ \\[0.1em]
+                   $\;\; \wedge \;\, x_j \neq \infty$
+                  } \Bigg\}
+\end{alignat*}
+The first guard takes the Cartesian product of
+the models and quantifies over all pairings $(u, v)$.
+The second guard binds $j$ which ranges over the constrained
+dimensions for $M$ unioned with the constrained dimensions for $N$.
+That is, every $j$ corresponds to a dimension in which
+either $M$ or $N$ has a scheme $u$ where $u_j \neq \infty$,
+and thus is constrained.
+The \emph{pairwise permutation} $u \bowtie v$ 
+builds a $2^n \times n$ matrix 
+of all possible $n$-vectors generated by
+non-deterministically picking for each $j^{th}$ entry either 
+$u_j$ or $v_j$ \eg{}
+%
+\begin{equation*}
+\vthreeh{0}{1}{2} \bowtie \vthreeh{3}{4}{5} =
+\left[
+\setlength{\arraycolsep}{0.42em}
+\begin{array}{cccccccc}
+0 & 0 & 0 & 0 & 3 & 3 & 3 & 3 \\
+1 & 1 & 4 & 4 & 1 & 1 & 4 & 4 \\
+2 & 5 & 2 & 5 & 2 & 5 & 2 & 5
+\end{array}
+\right]
+\end{equation*}
+%
+The $2^n$ unique choices for $\bowtie$ on $n$-vectors
+corresponds to taking all bit-strings of length $n$ and
+selecting from $u$ for $1$ and $v$ for $0$. The pairwise permutation
+matrix is thus defined in terms of the logical
+matrix $b$ where $b_{i,j}$ is the $j$-th bit of the integer
+$i - 1$ as:
+%
+\begin{align*}
+(u \bowtie v)_{i,j} & = b_{i,j} u_j + \neg b_{i,j} v_j
+\end{align*}
+%
+The permuted schemes $x = (u \bowtie v)_i$ added to the model 
+$M \otimes N$ are only those which have do not have $\infty$ in 
+the constrained dimensions given by each model ($x_j \neq
+\infty$). This filters out all permutations which have
+$\infty$ in a constrained dimension position. 
+\end{definition}
 
@@
 \subsection{Specification checking}
+
+%\newcommand{\cons}[3]{\mathbf{C}_{#3}(#1, #2)}
+\newcommand{\cons}[3]{#2 \;\, \textbf{consistent} \,\, #1}
+\begin{align*}
+\dfrac{\forall u \in C, \exists v \in S .  \, u \preceq v}
+      {\cons{\mathsf{up}(S)}{C}{n}}
+\qquad
+\dfrac{\forall v \in S, \exists u \in C . \, u \preceq v}
+      {\cons{\mathsf{low}(S)}{C}{n}} \\[0.75em]
+\dfrac{\cons{\mathsf{down}(S)}{C}{n} \qquad \cons{\mathsf{up}(S)}{C}{n}}
+  {\cons{\mathsf{exact}(S)}{C}{n}} \\[0.75em]
+\dfrac{\cons{\mathsf{down}(S_1)}{C}{n} \qquad \cons{\mathsf{up}(S_2)}{C}{n}}
+  {\cons{\mathsf{both}(S_1, S_2)}{C}{n}}
+\end{align*}
 

### 2016-11-12 paper.tex
representing consistency

@@
 
-\newcommand{\trule}[1]{{\footnotesize{#1}}}
+\newcommand{\trule}[1]{{\footnotesize{(\text{#1})}}}
 %\include{results}
@@
 
+%\newcommand{\cons}[3]{\mathbf{C}_{#3}(#1, #2)}
+\newcommand{\cons}[3]{#2 \;\, \textbf{cons} \,\, #1}
+\newcommand{\consA}[3]{#2 \;\, \textbf{cons}_M \,\, #1}
+
 \subsection{Specification checking}
 
-%\newcommand{\cons}[3]{\mathbf{C}_{#3}(#1, #2)}
-\newcommand{\cons}[3]{#2 \;\, \textbf{consistent} \,\, #1}
+\begin{definition}
+Consistency is defined in terms of the binary predicate
+$\textbf{cons}$ in Figure~\ref{fig:consistency}.
+Consistency relates models of source code and models 
+of specifications (which are bounded), thus
+ $\textbf{cons} \subseteq \mathcal{P}(\domainVal{}^n)
+\times \textsf{Approx}(\mathcal{P}(\domainVal{}^n))$. 
+We denote sets of indexing schemes $\mathcal{P}(\domainVal{}^n)$
+that come from a specification model by $S$ and from source code model
+by $C$.
+
+Rules (\emph{up}) and (\emph{low}) are the core rules relating models of
+specifications and models of code. The (up) rule 
+considers the case where the specification is
+an \term{atMost} upper-bound approximation.
+An upper bound specification means that for every scheme in the code
+model, modelling a source-level index, $\forall u \in C$ there exists
+a generalising scheme in the spec model $\exists v \in S$, \ie{}, $u
+\preceq v$. That is, every source-level index scheme is generalised by the specification.
+The (down) rule provides the case of 
+\term{atLeast} lower-bound specifications. Consistency is dual to the above: 
+for every scheme in the specification model $\forall v \in S$ there
+exists a scheme in the code model $\exists u \in C$ such that $u
+\preceq v$. That is, every specification scheme is generalised by some source
+index scheme.
+
+The (\emph{up}) and (\emph{down}) rules are combined for an \textsf{exact}
+specification (\textit{exact}) such that a specification's model
+provides both a consistent lower and consistent upper bound on
+for the source code. Thus, for every scheme in the code model there
+is a scheme in the spec model that generalises it, and for every
+scheme in the spec model there is a scheme in the source model that
+specialises it.  The (\textit{both}) rule is similar, decomposing
+a specification which has both a lower and upper bound into
+consistency of these bounds.
+
+Rules $(1)$ and $(*)$ extend the $\textbf{cons}$
+predicate to $\textbf{cons}_M \subseteq
+\textsf{Mult}(\mathcal{P}(\domainVal{}^n)) 
+\times \textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$
+which inludes multiplicity information. Both rules 
+state that consistent models remain consistent if they
+have the same multiplicity. Relatedly, rule $(1 < *)$ ensures
+that absence of a \texttt{readOnce} modifier in the specification
+does not prohibit a set of unique stencil accesses in the
+code from being consistent.
+\end{definition}
+
+\begin{figure}[t]
 \begin{align*}
-\dfrac{\forall u \in C, \exists v \in S .  \, u \preceq v}
-      {\cons{\mathsf{up}(S)}{C}{n}}
+\begin{array}{c}
+%%
+\trule{up}\dfrac{\forall u \in C \, \exists v \in S .  \, u \preceq v}
+      {\cons{\textsf{upper}(S)}{C}{n}}
 \qquad
-\dfrac{\forall v \in S, \exists u \in C . \, u \preceq v}
-      {\cons{\mathsf{low}(S)}{C}{n}} \\[0.75em]
-\dfrac{\cons{\mathsf{down}(S)}{C}{n} \qquad \cons{\mathsf{up}(S)}{C}{n}}
-  {\cons{\mathsf{exact}(S)}{C}{n}} \\[0.75em]
-\dfrac{\cons{\mathsf{down}(S_1)}{C}{n} \qquad \cons{\mathsf{up}(S_2)}{C}{n}}
-  {\cons{\mathsf{both}(S_1, S_2)}{C}{n}}
+\trule{low}\dfrac{\forall v \in S \, \exists u \in C . \, u \preceq v}
+      {\cons{\textsf{lower}(S)}{C}{n}} \\[1.5em]
+%%
+\trule{exact}\dfrac{\cons{\textsf{lower}(S)}{C}{n} \qquad \cons{\textsf{upper}(S)}{C}{n}}
+  {\cons{\textsf{exact}(S)}{C}{n}} \\[1.5em]
+%%
+\trule{both}\dfrac{\cons{\textsf{lower}(S_1)}{C}{n} \qquad \cons{\textsf{upper}(S_2)}{C}{n}}
+  {\cons{\textsf{both}(S_1, S_2)}{C}{n}}\\[1.5em]
+%%
+\trule{$1$}
+\dfrac{\cons{S}{C}{n}}{\consA{\textsf{once}(S)}{\textsf{once}(C)}{n}}
+\;\;
+\trule{$*$}
+\dfrac{\cons{S}{C}{n}}{\consA{\textsf{mult}(S)}{\textsf{mult}(C)}{n}}
+  \\[1.5em]
+\trule{$1 < *$}
+\dfrac{\cons{S}{C}{n}}{\consA{\textsf{mult}(S)}{\textsf{once}(C)}{n}}
+\end{array}
 \end{align*}
+\caption{Consistency between models of code and specification}
+\label{fig:consistency}
+\vspace{-0.9em}
+\end{figure}
 

### 2016-11-12 paper.tex
reworking the checking procedure

@@
 
-\section{Checking and inference}
+\section{Analysis, checking, and inference}
+\label{sec:analysis}
 
 %\newcommand{\cons}[3]{\mathbf{C}_{#3}(#1, #2)}
-\newcommand{\cons}[3]{#2 \;\, \textbf{cons} \,\, #1}
-\newcommand{\consA}[3]{#2 \;\, \textbf{cons}_M \,\, #1}
+\newcommand{\consName}{\textbf{cons0}}
+\newcommand{\cons}[2]{#2 \;\, \consName{} \,\, #1}
+\newcommand{\consAName}{\textbf{cons}}
+\newcommand{\consA}[2]{#2 \;\, \consAName{} \,\, #1}
+
+\noindent
+We outline here the procecures for checking the conformance
+of source code against specifications (\S~\ref{subsec:checking}) 
+and for infering specifications from code (\S~\ref{subsec:inference}).
+Both rely on the analysis of source code, converting array indexing
+expressions into sets of index schemes. We outline this analysis
+first (\S~\ref{subsec:analysis}). Note that the analysis procedure
+can be made arbitrarily more complex and wide-ranging independent
+of the checking and inference procedures.
+
+\subsection{Analysis of array-based code}
+\label{subsec:analysis}
+
+\newcommand{\neigh}{\textsf{neigh}}
+
+\paragraph{Syntax helper functions}
+The function \rhsExp{} maps source statements to a
+set of expressions which occur in right-hand positions (\ie{}, not the
+target of an assignment). The function \var{} maps expressions to a
+set of the variables used in right-hand positions.
+
+\paragraph{Analysis pre-requisites}
+The analysis builds on standard program analyses:
+%
+\textbf{(1)} basic blocks (CFG);
+\textbf{(2)} induction variables per basic block;
+\textbf{(3)} (interprocedural) data-flow analysis, providing a \emph{flows to}
+  graph (as shorthand, the function
+  $\mathsf{flowTo}$ is used, implicitly parameterised by this graph,
+  mapping an expression to the set of all expressions
+  with forwards data-flow to this expression, based on the transitive
+  closure of the flows graph);
+\textbf{(4)} type information per variable, where we use the predicate
+\arrayTy{} to classify variables of array type.
+
+\paragraph{Core analysis}
+For each assignment statement \textit{assg} within a loop whose left-hand side is an array
+subscript on neighbourhood indices, a finite map is computed mapping
+array variables to \emph{multisets} of indexing schemes.
+This is captured by the function $\textit{analyse}$:
+%
+\begin{align*}
+\hspace{-0.5em}
+\begin{array}{lr}
+\textit{analyse}(a(\overline{e_1}) = e_2)
+ := \qquad\qquad\qquad & \textit{where} \; \neigh(\overline{e_1}) \\[0.3em]
+\multicolumn{2}{l}{\qquad \bigcup\{a' \mapsto \{
+\textsf{schematic}(\bar{e})\} \mid a'(\bar{e}) \leftarrow \mathsf{flowsTo}(e_2),
+  \arrayTy(a')\}}
+\end{array}
+\end{align*}
+%
+The predicate $\neigh$ classifies neighbourhood indices, based on
+the induction variables of the loop (implicit here).
+Thus, we focus on assignments to an array subscript where the LHS 
+index $\overline{e_1}$ is a neighbourhood index.  For
+all array subscripts that flow to the right-hand side of this
+statement, a finite map is constructed, mapping each array variable
+to a \emph{multiset} of indexing schemes for its subscripts, computed
+with \textsf{schematic} (Defn.~\ref{def:schematic},
+p.~\pageref{def:schematic}).
+
+Note that \textsf{schematic} is undefined ($\bot$) if
+$\bar{e}$ contains relative indices which are not neighbourhood
+indices. The \textit{analysis} function feeds into the
+\textit{modelise} function which filters out those arrays 
+which contain undefined indexing schemes and adds multiplicity
+information:
+%
+\begin{align*}
+& \mathit{modelise} (\textit{assg}) 
+= a \mapsto 
+\begin{cases}
+\bot & \bot \in M \\
+\textsf{only}(|M|) & \forall u \in M . \#(M, u) = 1 \\
+\textsf{mult}(|M|) & \forall u \in M . \#(M, u) = n
+\end{cases} \\
+& \quad \textit{where} \; M = \mathit{analyse}(\textit{assg}) (a)
+\end{align*}
+where $\#(M, u)$ denotes the multiplicity of an element $u$ of the
+multiset $M$ and $|M|$ denotes the set of elements in the multiset
+$M$.
+
+Thus, the analysis maps source code to index schemes, which is then
+augmented with multiplicity information by $\mathit{modelise}$,
+\ie{} values in $\mathsf{Mult}(\mathcal{P}((\domainVal{})^n))$.
+
+If the LHS $\overline{e_1}$ contains non-$0$ neighbour offsets 
+then all RHS schemes are \emph{relativised} by
+$\textsf{schematic}(\overline{e_1})$. For example, 
+if \texttt{a(i+1) = b(i)} then relativisation of the RHS by the LHS
+produces an analysis which is equivalent to the analysis of \texttt{a(i) = b(i-1)}. 
+\dnote{Somewhere we need a discussion about syntactic vs semantic
+analysis...}
+
+\begin{example}
+We demonstrate the analysis and modelise procedures over the
+following example program which computes the mean value
+of a five-point stencil:
+%%
+\begin{minted}{fortran}
+do i = 1, (n-1)
+  do j = 1, (m-1)
+    x = a(i-1, j)+a(i+1, j); y = a(i, j-1)+a(i, j+1)
+    b(i, j) = (a(i, j) + x + y) / 5.0
+  end do
+end do
+\end{minted}
+%
+The \textit{analyse} function matches on
+line 4, with the following set for $\textsf{flowsTo}(\texttt{a(i, j) + x +
+  y})$:
+%
+\begin{align*}
+\{\texttt{a(i-1, j)}, \texttt{a(i+1, j)}, \texttt{a(i, j-1)},
+  \texttt{a(i, j+1)}, \texttt{a(i, j)}\}
+\end{align*}
+Subsequently the result of \textit{analyse} on line 4 yields the map:
+\begin{align*}
+\texttt{a} \mapsto \{\vtwoh{-1}{0}^1, \vtwoh{1}{0}^1,
+          \vtwoh{0}{-1}^1, \vtwoh{0}{1}^1, \vtwoh{0}{0}^1\}
+\end{align*}
+%
+where multiplicities are denoted by superscripts. The \emph{modelise}
+function then yields the following since all multiplicities in the model are one:
+%%
+\begin{align*}
+\texttt{a} \mapsto \mathsf{only}(\{\vtwoh{-1}{0}, \vtwoh{1}{0},
+          \vtwoh{0}{-1}, \vtwoh{0}{1}, \vtwoh{0}{0}\})
+\end{align*}
+%%
+\end{example}
+
+%Given an assignment $\textit{assg}$ to an array subscript in the source language
+%and an array variable $a$, then $\mathit{modelise}(\textit{assg}, a)$ performs
+%an analysis to find all array subscripts of $a$ converting them to
+%index schemes. For the moment, we leave this analysis abstract
+%as the function $\emph{analyse}$ in the following definition of $\emph{modelise}$:
+%
 
 \subsection{Specification checking}
+\label{subsec:checking}
+
+In checking mode, the access pattern of stencil computations in the
+source language is verified against any specification associated with
+them in the code. The model of
+\S~\ref{sec:semantics} guides this process.  Checking
+proceeds by generating a model from a specification and generating
+a model from the source code, comparing them for
+\emph{consistency}. The notion of consistency is built out of index scheme 
+\emph{generalisation} (Sect.~\ref{subsec:schemes-and-gen}).
+
+\begin{definition}[Checking]
+Let $a$ be an array variable and let
+$S \in \textit{spec}$ be a specification associated to $a$ 
+via \texttt{!= stencil $S$ :: a}. A stencil kernel 
+ $v(\overline{e_1}) = e_2$ in the source language
+conforms to $S$ if the predicate $\textbf{check}_n(S, v(\overline{e_1}) = e_2, a)$ holds, 
+defined:
+\begin{align*}
+& \textbf{check}_n(S, \textit{assg}, a) =
+  \consA_n{\interp{S}_n}{\textit{modelise}(\textit{assg})(a)}
+&
+\end{align*}
+Thus, checking proceeds by using \textit{modelise} 
+(\S~\ref{subsec:analyis}) to extract index schemes from source code
+augmented with multiplicitly for the whole set of schemes. 
+This is compared with the model of the specification $\interp{S}_n$ by
+the consistency predicate $\consAName$.
+\end{definition}
 
 \begin{definition}
-Consistency is defined in terms of the binary predicate
-$\textbf{cons}$ in Figure~\ref{fig:consistency}.
-Consistency relates models of source code and models 
-of specifications (which are bounded), thus
- $\textbf{cons} \subseteq \mathcal{P}(\domainVal{}^n)
-\times \textsf{Approx}(\mathcal{P}(\domainVal{}^n))$. 
-We denote sets of indexing schemes $\mathcal{P}(\domainVal{}^n)$
-that come from a specification model by $S$ and from source code model
-by $C$.
+Consistency relates models of source code to models 
+of specifications (which are bounded) via two predicates:
+\begin{align*}
+\consName_n & \subseteq \mathcal{P}(\domainVal{}^n)
+\times \textsf{Approx}(\mathcal{P}(\domainVal{}^n)) \\
+\consAName_n & \subseteq \textsf{Mult}(\mathcal{P}(\domainVal{}^n))
+\times \textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))
+\end{align*}
+%
+where $\consName$ relates models without multiplicity.
+Figure~\ref{fig:consistency} defines the two predicates, where for
+brevity we make implicit the subscript $n$ defining the
+dimensionality.  We denote sets of indexing schemes
+$\mathcal{P}((\domainVal{})^n)$ that come from a specification model
+by $M$ and from source code model by $N$.
 
@@
 Rules $(1)$ and $(*)$ extend the $\textbf{cons}$
-predicate to $\textbf{cons}_M \subseteq
+predicate to $\textbf{cons}_m \subseteq
 \textsf{Mult}(\mathcal{P}(\domainVal{}^n)) 
@@
 %%
-\trule{up}\dfrac{\forall u \in C \, \exists v \in S .  \, u \preceq v}
-      {\cons{\textsf{upper}(S)}{C}{n}}
+\trule{up}\dfrac{\forall u \in M \, \exists v \in N .  \, u \preceq v}
+      {\cons{\textsf{upper}(N)}{M}}
 \qquad
-\trule{low}\dfrac{\forall v \in S \, \exists u \in C . \, u \preceq v}
-      {\cons{\textsf{lower}(S)}{C}{n}} \\[1.5em]
-%%
-\trule{exact}\dfrac{\cons{\textsf{lower}(S)}{C}{n} \qquad \cons{\textsf{upper}(S)}{C}{n}}
-  {\cons{\textsf{exact}(S)}{C}{n}} \\[1.5em]
-%%
-\trule{both}\dfrac{\cons{\textsf{lower}(S_1)}{C}{n} \qquad \cons{\textsf{upper}(S_2)}{C}{n}}
-  {\cons{\textsf{both}(S_1, S_2)}{C}{n}}\\[1.5em]
+\trule{low}\dfrac{\forall v \in N \, \exists u \in M . \, u \preceq v}
+      {\cons{\textsf{lower}(N)}{M}} \\[1.5em]
+%%
+\trule{exact}\dfrac{\cons{\textsf{lower}(N)}{M} \qquad \cons{\textsf{upper}(N)}{M}}
+  {\cons{\textsf{exact}(N)}{M}} \\[1.5em]
+%%
+\trule{both}\dfrac{\cons{\textsf{lower}(M_1)}{M} \qquad \cons{\textsf{upper}(M_2)}{M}}
+  {\cons{\textsf{both}(M_1, M_2)}{M}}\\[1.5em]
 %%
 \trule{$1$}
-\dfrac{\cons{S}{C}{n}}{\consA{\textsf{once}(S)}{\textsf{once}(C)}{n}}
+\dfrac{\cons{N}{M}}{\consA{\textsf{once}(N)}{\textsf{once}(M)}}
 \;\;
 \trule{$*$}
-\dfrac{\cons{S}{C}{n}}{\consA{\textsf{mult}(S)}{\textsf{mult}(C)}{n}}
+\dfrac{\cons{N}{M}}{\consA{\textsf{mult}(N)}{\textsf{mult}(M)}}
   \\[1.5em]
 \trule{$1 < *$}
-\dfrac{\cons{S}{C}{n}}{\consA{\textsf{mult}(S)}{\textsf{once}(C)}{n}}
+\dfrac{\cons{N}{M}}{\consA{\textsf{mult}(N)}{\textsf{once}(M)}}
 \end{array}
@@
 \end{figure}
+
+\begin{example}
+  We give an extended example of verifying a stencil computation
+  against its specification. We consider a more general, but still
+  consistent, version of the \textit{five-point} stencil specification
+  inferred previously in \S~\ref{subsec:inference}:
+%
+\begin{minted}[breakindent=8.1em]{fortran}
+!= region :: fivepoint = centered(depth=1,dim=1) + centered(depth=1,dim=2)
+!= stencil readOnce, fivepoint :: b
+a(i,j) = b(i,j-1)+b(i,j)+b(i,j+1)+b(i-1,j)+b(i+1,j)
+\end{minted}
+%
+We refer to the assignment on line three as $l3$ for brevity.
+
+Verification of the code against the spec is then given by 
+$check(({\textcolor{darkgreen}{\texttt{readOnce, fivepoint}}}),l3,\texttt{b})$. 
+We explain the steps.
+\begin{enumerate}  
+%
+\item Generate the model from the specification as in
+  \S~\ref{sec:semantics}. 
+%
+\begin{align*}
+ N_s= \{  \vtwo{-1}{\infty},
+          \vtwo{0}{\infty},
+          \vtwo{1}{\infty},
+          \vtwo{\infty}{-1},
+          \vtwo{\infty}{0},
+        \vtwo{\infty}{1} \}  \\[0.4em]
+\interp{{\textcolor{darkgreen}{\texttt{readOnce, fivepoint}}}}_2 =
+  \textsf{once}(\textsf{exact} (N_s))
+\end{align*}
+%
+\item Obtained the code model via $\textit{modelise}$, giving:
+%
+\begin{align*}
+M_c = \{
+    \vtwo{0}{-1},
+    \vtwo{0}{0},
+    \vtwo{0}{1},
+    \vtwo{-1}{0},
+    \vtwo{1}{0}
+  \}
+\\[0.4em]
+\textit{modelise}(l3, \texttt{b}) = \textsf{once} (M_c)
+\end{align*}
+%
+\item Consistency $cons_2(\textsf{once}(\textsf{exact}(N_s)),
+  \textsf{once}(M_c))$ holds when 
+  $\textit{cons}_2(\textsf{mult}(\textsf{exact}(N_s)),\textsf{mult}(M_c))$ (Fig.~\ref{fig:consistency}) 
+  which holds when: 
+  \begin{align*}
+     & \textit{cons}_2(\textsf{mult}(\textsf{up}(N_s)),\textsf{mult}(M_c)) \\
+   \wedge \; & \textit{cons}_2(\textsf{mult}(\textsf{low}(N_s),\textsf{mult}(M_c))
+  \end{align*}
+%
+\item This condition holds by the following sets of scheme
+  generalisations: for the upper bound, $\forall u \in M_c,
+  \exists v \in N_s . u \preceq_2 v$: 
+  \begin{align}
+    \begin{array}{l}
+     \vtwoh{0}{-1} \preceq_2 \vtwoh{0}{\infty} \quad
+     \vtwoh{0}{0} \preceq_2 \vtwoh{0}{\infty} \quad
+     \vtwoh{0}{1} \preceq_2 \vtwoh{0}{\infty} \\[0.5em]
+     \vtwoh{-1}{0} \preceq_2 \vtwoh{\infty}{0} \quad
+     \vtwoh{1}{0} \preceq_2 \vtwoh{\infty}{0}
+     \end{array}
+  \label{eq:check-example-upper}
+  \end{align}
+  % 
+  and for the lower bound: $\forall v \in N_s, \exists u \in M_c . u
+  \preceq_n v$: 
+  \begin{align}
+    \hspace{-1em}
+    \begin{array}{l}
+     \vtwoh{-1}{0} \preceq_2 \vtwoh{-1}{\infty} \quad
+     \vtwoh{0}{0} \preceq_2 \vtwoh{0}{\infty} \quad
+     \vtwoh{1}{0} \preceq_2 \vtwoh{1}{\infty}  \\
+     \vtwoh{0}{-1}  \preceq_2 \vtwoh{\infty}{-1} \quad
+     \vtwoh{0}{0}  \preceq_2 \vtwoh{\infty}{0} \quad
+     \vtwoh{0}{1}  \preceq_2 \vtwoh{\infty}{1}
+     \end{array}
+    \label{eq:check-example-lower}
+  \end{align}
+  Since both the consistency predicate is satisfied, the array
+  accesses flowing to line three in the stencil computation are valid
+  with respect to the {\textcolor{darkgreen}{\texttt{fivepoint}}} specification.
+\end{enumerate}
+\vspace{-0.7em}
+\end{example}
+\begin{example} 
+We consider now an erroneous user modification to $l3$ with an
+additional term \texttt{b(i+2,j+2)} such that $l3'$ is:
+%
+\begin{minted}[firstnumber=3]{fortran}
+a(i,j) = b(i,j-1) + b(i,j) + b(i,j+1) + b(i-1,j) + b(i+1,j) + b(i+2,j+2)
+\end{minted}
+%
+The checking procedure proceeds in the same way as the first three
+steps of the previous example, except that the result of
+$\mathit{modelise}(l3',\texttt{b})$ generates an additional scheme
+$\vtwoh{2}{2}$. Since the spec model has not changed and the code
+model is has only the additional scheme $\vtwoh{2}{2}$, the generalisations for
+the lower bound \eqref{eq:check-example-lower} 
+of step 4 above still holds since its consistency condition
+universally quantifies over the \emph{spec model}. However, for
+the upper bound, the consistency condition universally quantifies
+over scheme $u$ of the \emph{code model} for which there exists
+a scheme $v \in N_s$ such that $u \preceq_2 v$. Thus, we require a $v
+\in N_s$ such that $\vtwoh{2}{2} \preceq_2 v$ but no such generalising
+scheme existing in the spec model. Hence the model fails to hold as an upper bound.
+
+Note that if the specification were as follows:
+%
+\begin{minted}{fortran}
+!= stencil atLeast, readOnce, fivepoint :: b
+\end{minted}
+%
+then consistency holds as it no longer requires the failing upper-bound case 
+$\mathit{cons}(\textsf{mult}(\textsf{up}(N_s)),\textsf{mult}(M_c))$.
+ This makes sense semantically since the region ${\textcolor{darkgreen}{\texttt{fivepoint}}}$
+is indeed contained in the regions touched by $\texttt{b}$ in $l3'$. 
+\end{example}
+
+\dnote{Build this into the definitions earlier I think}
+One way the implementation may deviate from the model is elimination of all $\infty$
+offset vector that occurs when all indices are in non-neighbourhood form \eg{}
+\mintinline{fortran}{b(42,0)}. The only generalisation scheme that can unify
+with this pattern is the most general one, meaning it requires outright
+rejection of an otherwise fine stencil computation. See the following example,
+where not discarding all $\infty$ access pattern prevents verification:
+%
+\begin{minted}{fortran}
+!= stencil forward(dim=1,depth=1) :: a
+a(i) = a(i) + a(i+1) + a(1)
+\end{minted}
 

### 2016-11-12 paper.tex
introduction to the syntax

@@
   breaklines=true,
-  breakafter=),
+ breakafter=w),
   xleftmargin=\parindent
@@
 
-\newcommand{\stenRefl}[1]{\term{pointwise} \, (\term{dim=}#1)}
+\newcommand{\stenRefl}[1]{\term{pointed} \, (\term{dim=}#1)}
 \newcommand{\stenFwd}[3]{\term{forward} \, (\term{depth=}#1,
@@
   \term{dim=}#2{#3})}
-\newcommand{\irrefl}{\texttt{nonpointwise}}
+\newcommand{\irrefl}{\texttt{nonpointed}}
 
@@
 stencil computations are more prone to errors from simple lexical
-mistakes. For example, the following is a three-line snippet from a
-Navier-Stokes fluid simulator in which two arrays are read with
+mistakes. For example, the following is a small snippet from a
+Navier-Stokes fluid simulator (based on~\citet{griebel1997numerical}) 
+ in which two arrays are read with
 different data access patterns, across two dimensions.
@@
 
-In practice, we find that the usual development procedure for complex
+In practice, the typical development procedure for complex
 stencil computations involves some ad hoc testing to ensure that no
@@
 inspection of the results, based on a rendered image. Typically such
-testing is then discarded once the code is seen to be correct.
-
-This is not the only information that is typically discarded. The
-shape of the indexing pattern is often a result of choices made in the
-numerical-analysis discretisation procedure used to convert a set of
+testing is discarded once the code is seen to be correct.
+
+This is not the only information that is often discarded. The
+``shape'' of the indexing pattern is usually the result of choices made in the
+numerical-analysis procedure to convert some 
 continuous equations into a discrete approximation. Rarely are these
 decisions captured in the source code, yet the shape of access is
-usually uniform and is given a perspicuous and concise description in
+usually uniform and is has a perspicuous and concise description in
 numerical analysis literature \eg{}, ``\emph{centered in space, to a
@@
 (``centered'') to a depth of one in each direction in its first
-dimension.  The stencil computation shape in the Navier-Stokes example
+dimension.  The shape of the Navier-Stokes example
 is specified as:
 %
-\begin{minted}[linenos=false,xleftmargin=4em,breakindent=3em]{fortran}
-!= stencil centered(depth=1,dim=1) * pointwise(dim=2) + centered(depth=1,dim=2) * pointwise(dim=1) :: u
+\begin{minted}[linenos=false,xleftmargin=0em,xrightmargin=1.1em,breakindent=1.9em]{fortran}
+!= stencil centered(depth=1,dim=1)*pointed(dim=2)     + centered(depth=1,dim=2)*pointed(dim=1) :: u
 
@@
 first dimension and backward to a depth of $1$ in the second
-dimension.
-
-We provide a verification tool for such specification which can check
+dimension. Note the relatively small size of the specification
+compared to the source code, and its high-level of abstraction.
+
+We provide a verification tool for such specifications which can check
 the correctness of code against the specification, and can also infer
@@
     expressions}, \eg{} does not involve expressions like
-  \texttt{a(i+1,j-1)}. This contrasts with other stencil specification
-  approaches. For example, the work of Kamil et al.~\cite{kamil2016verified}
-  provides full-functionality specifications of stencil computations from which
-  equivalent DSL code can be generated.  If a user wishes to specify, say, the
-  Naiver-Stokes stencil above, they must write a specification which is very
-  similar, involving textually dense array-indexing expressions. Therefore any
-  low-level mistakes that could be made programming complex indexing code could
+  \texttt{a(i+1,j-1)}. This contrasts with other specification
+  approaches.% For example, the work of Kamil et al.~\cite{kamil2016verified}
+  %provides full-functionality specifications of stencil computations
+  %in the form of a domain-specific language.  
+  If a user wishes to specify, say, the
+  Naiver-Stokes stencil above, in a deductive verification languages
+  like ACSL, they must write a specification which is very
+  similar to the code, involving textually dense array-indexing expressions. Therefore any
+  low-level mistakes that could be made whilst programming complex indexing code could
   also be made when writing a specification. Our specifications are much more
@@
 class of bugs that can be detected automatically (\emph{push-button
-  verification}). Instead, stencil computation bugs must largely be
+  verification}). Instead, stencil computation bugs must be
 identified relative to a specification of the intended behaviour.
@@
 
+Our specification language is based on the hypothesis
+that most forms of array access in numerical code have
+a fixed, statically-determined access pattern. For example, the
+\emph{five-point stencil} on a two-dimensional array reads from array
+indices $(i, j)$, $(i-1, j)$, $(i+1, j)$, $(i, j-1)$, and $(i, j+1)$
+for all $i, j$ within the inner boundary of the array (to avoid
+out-of-bounds access at the edges). We revisit this hypothesis
+in \S~\ref{sec:evaluation} with the inference of
+such regular stencil patterns on our corpus of numerical programs.
+
+We first set up various definitions used in this paper.
+\S~\ref{sec:syntax} then outlines the syntax of stencil specifications
+and gives an informal introduction. \S~\ref{sec:semantics} then
+gives a semantic model via a set-based interpretation which makes 
+concrete the meaning of specifications. \S~\ref{sec:eqs} then provides
+a syntactic equational theory for specifications which is 
+sound with respect to the model.
+%sub-specification for which the
+%equality and sub-specification relation are proven sound.
+
 \paragraph{Notation and convention}
@@
   \texttt{do} in Fortran for our implementation) incremented by $1$
-  per iteration. The variable is interpreted as a based induction
+  per iteration. A variable is interpreted as a base induction
 variable only within the scope of the loop body. Induction varaibles 
@@
   comma-separated sequence of integer expressions denoted $\bar{e}$ or
-  in expanded form as $(e_1, \ldots, e_n)$. An \emph{index} $\bar{e}$
-  is called \emph{relative} if each $e \in \bar{e}$ is defined in
-  terms of an induction variable.
+  in expanded form as $(e_1, \ldots, e_n)$. An index $\bar{e}$
+  is called \emph{relative} if each $e \in \bar{e}$ involves an induction variable.
 
@@
   most $n$ induction variables 
-  over a subset of the index space of $a$. Elements of $a$ are determined by an
+  over a subset of the index space of $a$. Each step of the interation
+  determs the elements of $a$ by an 
   assignment $a(\bar{e}) = e_r$ where each $e \in
   \bar{e}$ is either a neighbourhood or constant index.
-  For all arrays $b \in \bar{b}$ and array subscripts $b(\var{e'})$ in
-  $e_r$ and that flow to $e_r$, then each $e' \in \bar{e'}$ is also
+  For all arrays $b \in \bar{b}$ and array subscripts $b(\bar{e}')$ in
+  $e_r$ and that flow to $e_r$, then each $e' \in \bar{e}'$ is also
   either a neighbourhood or constant index.
@@
 
-Specification declarations associate to a stencil kernel 
-a specification for the array access on one or more array
-variables $\bar{v}$ within that kernel.
+\noindent
+Specifications are associated to a stencil kernel 
+and are declared for one or more array variables
+$\bar{v}$ within that kernel.
 
@@
   Consistency is then a two way requirement.
-  On the one hand, all array subscripts contributing to the kernel
+  On the one hand, all array subscripts on $a$ contributing to the kernel
   should be consistent with the region; on the other, all parts of the
-  region should have an index which fits the space. 
+  region should have an index corresponding to that part.
 \end{definition}
@@
 \multicolumn{2}{l}{\qquad\qquad \mid \; \stenRefl{\dimTy}} \\
-\multicolumn{2}{l}{\qquad\qquad \mid \;\stenFwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{nonpointwise}]}} \\
-\multicolumn{2}{l}{\qquad\qquad \mid \; \stenBwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{nonpointwise}]}} \\
-\multicolumn{2}{l}{\qquad\qquad \mid \; \stenCen{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{nonpointwise}]}} \\
+\multicolumn{2}{l}{\qquad\qquad \mid \;\stenFwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{nonpointed}]}} \\
+\multicolumn{2}{l}{\qquad\qquad \mid \; \stenBwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{nonpointed}]}} \\
+\multicolumn{2}{l}{\qquad\qquad \mid \; \stenCen{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{nonpointed}]}} \\
 \multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{region} \, \term{+}
@@
 
-Region constants specify a pattern of continguous array accesses in a
-single dimension.  Region constants are either \term{pointwise},
+Region constants specify a finite contiguous region in a
+single dimension relative to the origin and are either \term{pointed},
 \term{forward}, \term{backward}, or \term{centered}.  Each region
-constant has a dimension identifier $d$ given by a natural number
-greater than 0. Each constant except \term{pointwise} is given a depth
-parameter $n$, a natural number greater than 0; \term{pointwise}
+constant has a dimension identifier $d$ given as a natural number
+greater than 0. Each constant except \term{pointed} has a depth
+parameter $n$ given as a natural number greater than 0; \term{pointed}
 regions implicitly have a depth of $0$.
 
-A \term{forward} region of depth $n$ specifies continguous
-neighbourhood indices in dimension $d$ ranging from $i$ to $i + n$
+A \term{forward} region of depth $n$ specifies a continguous
+region in dimension $d$ starting at the origin; this corresponds
+to specifying neighbourhood indices in dimension $d$ ranging from $i$ to $i + n$
 inclusively for some induction variable $i$.  Similarly, a
-\term{backward} region of depth $n$ specifies contiguous neighbour
-indices in dimension $d$ from $i$ to $i - n$, and \term{centered}
-of depth $n$ from $i - n$ to $i + n$. A \term{pointwise}
+\term{backward} region of depth $n$ corresponds to contiguous indices
+in dimension $d$ from $i$ to $i - n$, and \term{centered}
+of depth $n$ from $i - n$ to $i + n$. A \term{pointed}
 stencil specifies a neighbour index $i$. For example the
@@
 e(i, j) = (c(j-1) + c(j) + c(j+1))/3.0
-!= stencil pointwise(dim=3) :: d
+!= stencil pointed(dim=3) :: d
 e(i, j) = d(0, 0, i)
@@
 %%
-Note that every dimension need not be specified for; in the above, the
-specifications on lines $1$ and $5$ and $7$ are the principal specifications
-for code on lines $2$ and $6$ and $8$, but the specification on line
-$3$ is not principal for line $4$;
-we return to the notion of principality later.
+Not every dimension needs to be specified. In the above, the
+specifications on line $1$, $3$, and $7$ leaves some dimensions unspecified.
 
 The \term{forward}, \term{backward}, and \term{centered} regions may
-all have an additional attribute \term{nonpointwise} which marks that
-the region does not include indices which are purely base induction
-variables, \ie{}, offset by $0$.  For example, the following is a
-\term{nonpointwise} \term{backward} stencil
+all have an additional attribute \term{nonpointed} which marks that
+the region does not include the origin.  For example, the following is a
+\term{nonpointed} \term{backward} stencil
 %
 \begin{minted}{fortran}
-!= stencil backward(depth=2, dim=1, nonpointwise) :: a
+!= stencil backward(depth=2, dim=1, nonpointed) :: a
 b(i) = a(i-1) + a(i-2)
@@
 %
-The \term{pointwise} regions cannot have a \term{nonpointwise} attribute.
+The \term{pointed} regions cannot have a \term{nonpointed} attribute.
 
@@
 
+% r+s means that indices must be consistent with either r or s 
+% r*s means that indices must be consistent with both r and s at the
+% same time
+
+% specifications correspond to a set of schemes
+
+% r+s means an index must be generalised by a scheme in either r or s
+%       and every scheme in r must have an index which specialises it
+%       and every scheme in s must have an index which specialises it
+
+% r*s means an index must be generalised by a scheme in r and a scheme in s
+%       and every pairs of schemes in r and s must have an index which 
+%           specialises both 
+%
+
 The region operators \term{+} and \term{*} provide ways to combine
-regions in a disjunctive and conjuctive fashion respectively.
+regions disjunctively and conjuctively respectively.
 
 The conjuction of two regions $r \term{*} s$ means that any indices
-in the specified code must be consistent with both $r$ and $s$ at
-the same time. Dually, there must be an index which is consitent
-with every component of region $r$ and
+in the specified code must be consistent with both $r$ and $s$
+simultaenously. Dually, for every pair of components in the regions
+$r$ and $s$ there must be an index consistent with both.
+For example, the following code (defining a \emph{nine-point stencil}) 
+has a specification given by the product of two \texttt{centered}
+regions in each dimension:
+%%
+\begin{minted}[breakindent=2.9em]{fortran}
+x = a(i, j)   + a(i-1, j)   + a(i+1, j)
+y = a(i, j-1) + a(i-1, j-1) + a(i+1, j-1)
+z = a(i, j+1) + a(i-1, j+1) + a(i+1, j+1)
+!= stencil centered(depth=1, dim=1) * centered(depth=1, dim=2) :: a
+b(i, j) = (x + y + z) / 9.0
+\end{minted}
+%%
+This pattern is common in image convolution applicaitons (for example
+for edge detection). The specification ranges over the
+values that flow to the array subscript on the left-hand side,
+and so ranges over the intermediate assignments to \term{x},
+\term{y}, and \term{z}. Each index in the code is consitent
+with both specifications simultaneously, \eg{}, \texttt{a(i-1, j+1)}
+is within the centered region in dimension $1$ and the centered region
+in dimension $2$.
+
+The product $r \term{*} s$ can also be thought of as a bounding box
+over the two regions $r$ and $s$.
+
+The disjunction of two regions $r \term{+} s$ means that any indices
+in the spcified code must be consistent with either of $r$ or $s$. 
+Dually, for every component in the region $r$ there must be an index
+within it and every component in the region $s$ must have an index
+within it. For
+example, the following gives the specification of a five-point
+stencil which is the sum of two compound \texttt{pointed} and
+\texttt{centered} regions in each dimension:
+%
+\begin{minted}{fortran}
+!= stencil centered(depth=1, dim=1)*pointed(dim=2)    + centered(depth=1, dim=2)*pointed(dim=1) :: a
+b(i,j) = -4*a(i,j) + a(i-1,j) + a(i+1,j) &
+                   + a(i,j-1) + a(i,j+1)
+\end{minted}
+%%
+Here the left-hand side of \texttt{+} says that when the second dimension
+(induction variable $j$) is fixed at the origin, the first dimension
+(induction variable $i$) accesses the immediate vicinity of the origin
+(to depth of one). The right hand side of \texttt{+} is similar but the dimensions are reversed.
+This reflects the symmetry under rotation of the five-point stencil.
+
+As another example, the following
+implements and specifies the \emph{Roberts cross}
+edge-detection convolution~\cite{davis1975survey}:
+%%
+\begin{minted}[breakindent=5.5em]{fortran}
+do j=0, jmax-1
+  do i=0, imax-1
+    x = a(i,j)-a(i+1,j+1); y = a(i+1,j)-a(i,j+1)
+    != stencil forward(depth=1,dim=1) * forward(depth=1,dim=2) :: a
+    b(i,j) = sqrt(x*x+y*y)
+  end do
+end do
+\end{minted}
 
@@
 
-% r+s means that indices must be consistent with either r or s 
-% r*s means that indices must be consistent with both r and s at the
-% same time
-
 % specifications correspond to a set of schemes
@@
 %
-
-%  forward(depth=1,dim=2) & pointwise(dim=1) 
-% | forward(depth=1,dim=1) & pointwise(dim=2)
-
-% (o, p) \preceq [[ ]]
 
@@
 
-%%  LocalWords:  refactoring affine parameterised nonpointwise atMost
+%%  LocalWords:  refactoring affine parameterised nonpointed atMost
 %%  LocalWords:  centered readOnce atLeast discretisation Equational

### 2016-11-13 paper.tex
reinstating and editing equational theory

@@
 
-\newcommand{\stenReflS}[1]{\term{point} \, (\term{dim=}#1)}
+\newcommand{\stenReflS}[1]{\term{pointed} \, (\term{dim=}#1)}
 \newcommand{\stenFwdS}[2]{\term{fwd} \, (\term{depth=}#1,
@@
   \term{dim=}#2)}
-\newcommand{\irreflS}{\texttt{nonpoint}}
+\newcommand{\irreflS}{\texttt{np}}
 
@@
 
+\marginpar{\dnote{Is it worth having this informal consistency here?}}
 \begin{definition}[Consistency, informal]
@@
 \end{definition}
-\todo{Improve the above- haven't quite introduce enough yet to say
-more}
 
@@
 
-\section{A model}
+\paragraph{Region declarations and variables}
+
+Region specifications can be assigned to region variables via
+a region declaration, which can be used later in another
+specification. For example, the last specification 
+ example above can be restated as:
+%%
+\begin{minted}{fortran}
+!= region :: r1 = forward(depth=1, dim=1)
+!= region :: r2 = forward(depth=1, dim=2)
+!= region :: robertsCross = r1*r2
+!= stencil robertsCross :: a
+\end{minted}
+This is especially useful for common stencil patterns, such as Roberts cross,
+as the region can be defined once and reused.
+%%
+\paragraph{Modifiers}
+%%
+Region specifications can be modified 
+by \emph{approximation} and \emph{multiplicity} information 
+(in the \textit{spec} rule of Fig.~\ref{fig:syntax}).
+
+The \texttt{readOnce} modifier enforces that no index appears more
+than once (that is, its multiplicity is one). For example, in all of
+the previous examples the \texttt{readOnce} modifier could be added,
+\eg{}
+%
+\begin{minted}{fortran}
+!= stencil readOnce, backward(depth=2, dim=1) :: a
+b(i+1) = a(i) + a(i-1) + a(i-2)
+\end{minted}
+%
+This specification would be invalid if any of the
+array subscripts were repeated. This modifier provides a way to
+rule out any accidental repetition of array subscripts.
+The notion is similar to that of \emph{linearity} in linear type
+systems~\cite{wadler1990linear}, where a value must be used
+exactly once. We opt for the more informative and easily understood name
+\texttt{readOnce}. This modifier is optional, so it need not
+be present even if the stencil is linear.
+
+In some cases, it is useful to give a lower and/or upper bound for a
+stencil. This can be done using either the \term{atMost} or
+\term{atLeast} modifiers. This is particularly useful in situations
+where there is a non-contiguous stencil pattern, which cannot be expressed
+precisely in our specification syntax. For example:
+%
+\begin{minted}{fortran}
+!= stencil atLeast, pointed(dim=1)       :: a
+!= stencil atMost, forward(depth=4, dim=1) :: a
+b(i) = a(i) + a(i+4)
+\end{minted}
+
+\section{A model for specifications}
+\label{sec:semantics}
 
 We define a simple set-theoretic denotational model of the semantics
-of our stencil specification language. This has a number of purposes:
-(1) it serves to explain exactly the meaning of our specifications
-(2) it is 
-used as the basis of the inference, checking, and synthesis algorithms
-(\S~\ref{sec:analysis}); (3) justifies the equational theory and
-approximations of the previous section; and (4) can be used to guide
+of our specification language. This has a number of purposes:
+(1) it serves to explain the meaning of our specifications
+(2) it is used in the inference and checking algorithms
+(Section~\ref{sec:analysis}); (3) justifies an equational theory for
+specifications in the next section; and (4) can be used to guide
 correct implementations. The model domain is over sets of vectors
-which we call \emph{index schemes}, which we introduce first before
-the model in \S~\ref{subsec:model}.
+which we call \emph{index schemes}.
 
@@
   p.~\pageref{def:neighbour}) and $\infty$ represents any constant
-  inddex. Throughout $u, v, w$ range over
+  index. Throughout $u, v, w$ range over
   index schemes and $u_i$ denotes the $i^{th}$ element of $u$. We
-  write $(\domainVal)^n$ for the set of $n$-dimensional schemes.
+  write $(\domainVal)^n$ for the set of all $n$-dimensional schemes.
 \end{definition}
@@
 indexing behaviours which are consistent with the specification.
-The relationship between indexing schemes and source language
-terms is given precisely by the following function \textsf{schematic} which
-maps syntactic index expressions to indexing schemes. This
-is used in inference and checking.
+The relationship between source language
+terms and indexing terms
+ is given precisely by the following function \textsf{schematic},
+ which forms part of checking and inference (\S\ref{sec:analysis}).
 
@@
 induction variables $I$, uses the above operation $\schemeOne$ to
- maps from the syntax of indices to index schemes, \ie{} values of 
-$\domainVal^n$. It is defined on the syntax of index
-expressions $\bar{e} = (e_1, ..., e_n)$ as:
+ map from the syntax of indices to index schemes, \ie{} values of 
+$(\domainVal)^n$. It is defined on the syntax of indices $\bar{e} = (e_1, ..., e_n)$ as:
 \begin{align*}
@@
 %
-Thus, for an indexing expression, if each component is a neighbourhood
+Thus, if each expression in an index is a neighbourhood or constant
 index, then $\schemeN$ maps it to an indexing scheme.
@@
 \begin{example}
-Consider the following three indexing schemes
+Consider the following indexing schemes
 corresponding to array indexing terms
-$(i, j+1)$, $(1, j+1)$ and $(3, j)$:
+$(i, j+1)$, $(1, j+1)$, $(3, j)$ and $(i*2, j)$:
 %
@@
 \schemeN{(1, j+1)} & = \vtwoh{\infty}{1} \\
-\schemeN{(3, j)} & = \vtwoh{\infty}{0}
+\schemeN{(3, j)} & = \vtwoh{\infty}{0} \\
+\schemeN{(i*2, j)} & = \bot
 \end{array}
 \end{align*}
-There is one notion of generalisation between these
+There is one generalisation between these
 schemes, namely that $\vtwoh{0}{1} \preceq_2 \vtwoh{\infty}{1}$.
 At the level of syntax, this translates to saying
-$(i, j+1)$ can be instantiated (specialised) to $(3, j+1)$.
-
-As an example of undefined \textsf{schematic},
-$\textsf{schematic}(i*2, j) = \bot$ as its first index is a
-non-neighbour relative index, which is outside the scope of our
-specification language.
+$(i, j+1)$ can be instantiated (specialised) to $(1, j+1)$. 
+Note the last example is undefined as its first expression is a
+non-neighbour relative index.
 \end{example}
@@
 
-\begin{definition}[Quotienting schemes]
-A scheme $u$ can be quotiented by a set of dimension identifies
-by the operator $/_{n,I}$ where $I$ is a set of dimension
-identifiers drawn from $\mathbb{N}_{\leq n}$ where 
-\begin{equation*}
-u/_{n,I} = [ ]
-\end{equation*}
-\end{definition}
-
 \subsection{Denotational model}
 
-Our model is given by the interpretation function $\interp{-}_n$
+The model is defined by the interpretation function $\interp{-}_n$
 (where $n$ is the maximum dimensionality of the specification being
@@
 
+\paragraph{Model of \textit{spec}}
+
+Regions are modelled as sets of index
+schemes, that is
+$\interp{\textit{region}}_n \in \mathcal{P}(\domainVal{}^n)$.
+Figure~\ref{fig:spatial-model} defines the top-level model of the
+\textit{spec} syntax, which includes the multiplicity and
+approximation modifiers. The domain of this interpretation
+is $\textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$
+where $\textsf{Mult}$ and $\textsf{Approx}$ are labelled variants
+labelling our set-based model with additional information.
+
+\begin{definition} \textsf{Mult}
+and \textsf{Approx} are parametric labelled variant types
+with injection function given by their definition:
+%
+\begin{align*}
+\textsf{Mult} \, a & = \textsf{mult} \, a \,\mid\, \textsf{only} \, a \\
+\textsf{Approx} \, a & = \textsf{exact} \, a \,\mid\, \textsf{lower} \, a \,\mid\,
+\textsf{upper} \, a \,\mid\, \textsf{both} \, a \, a
+\end{align*}
+\ie{}, $\mathsf{lower}$ is an injection $\mathsf{lower} : a \rightarrow \mathsf{Approx} \, a$, and so
+on. The $\textsf{Mult}$ type corresponds to the presence or absence of the
+\term{readOnce} modifier as shown in Figure~\ref{fig:spatial-model}. 
+The $\textsf{Approx}$ type corresponds to the presence or absence of
+the spatial approximation modifier, with \textsf{exact} when there is
+no such modifier and \textsf{lower} and \textsf{upper} for
+\term{atLeast} and \term{atMost}. Later, it will be useful to model
+pairs of lower and upper bounds which is provided by \textsf{both} 
+but which is not used here.% These are used for labelled models, and other components later in.
+\label{def:mult-and-approx}
+\end{definition}
+
+\begin{figure}
+\input{spatial-model.tex}
+\caption{\textit{spec} model,
+$\interp{-}_n : \textit{spec} \rightarrow \textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$}
+\label{fig:spatial-model}
+\vspace{-0.75em}
+\end{figure}
+
+\section{Equational theory}
+\label{sec:eqs}
+
+Specifications are subject to a set of equations which can be used to
+understand when specifications are equivalent. Furthermore,
+specifications are subject to a notion of approximation. We outline
+the equational theory and notion of approximation here.
+
+\begin{figure}
+%\vspace{-0.2em}
+\begin{align*}
+\setlength{\arraycolsep}{0.05em}
+\begin{array}{c}
+%\framebox{$\textit{region} \equiv \textit{region}'$} \\[0.6em]
+\begin{array}{rll}
+(\textsc{F\, \texttt{+} \,F}) \;\; &
+\stenFwdSR{n \, \textsf{max} \, m}{d}{p \vee q} \\
+ \equiv \; & \stenFwdSR{n}{d}{p} \; \texttt{+} \; \stenFwdSR{m}{d}{q} \\[0.6em]
+%
+%
+(\textsc{B\, \texttt{+} \,B}) \;\; &
+\stenBwdSR{n \, \textsf{max} \, m}{d}{p \vee q} \\
+ \equiv \; & \stenBwdSR{n}{d}{p} \; \texttt{+} \; \stenBwdSR{m}{d}{q} \\[0.6em]
+%
+%
+(\textsc{C\, \texttt{+} \,C}) \;\; &
+\stenCenSR{n \, \textsf{max} \, m}{d}{p \vee q} \\
+\equiv \; & \stenCenSR{n}{d}{p} \; \texttt{+} \; \stenCenSR{m}{d}{q} \\[0.6em]
+%
+%
+(\textsc{C\, \texttt{+} \,F}) \;\; & \stenCenSR{n}{d}{p \vee q} \\
+\textit{$m \leq n$} \; \equiv \; & \stenCenSR{n}{d}{p} \; \texttt{+} \;
+                      \stenFwdSR{m}{d}{q} \\[0.6em]
+%
+%
+(\textsc{C\, \texttt{+} \,B}) \;\; &
+\stenCenSR{n}{d}{p \vee q} \\
+\textit{$m \leq n$} \; \equiv \; & \stenCenSR{n}{d}{p} \; \texttt{+} \;
+                      \stenBwdSR{m}{d}{q} \\[0.6em]
+%
+%
+(\textsc{B\, \texttt{+} \,F}) \;\; &
+\stenCenSR{n}{d}{p \vee q} \\
+\equiv \; & \stenFwdSR{n}{d}{p} \; \texttt{+} \; \stenBwdSR{n}{d}{q}
+  \\[0.6em]
+%
+%
+(\textsc{P\, \texttt{+} \,F}) \;\; &
+\stenFwdS{n}{d} \\
+\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenFwdSR{n}{d}{p} \\[0.6em]
+%
+%
+(\textsc{P\, \texttt{+} \,B}) \;\; &
+\stenBwdS{n}{d} \\
+\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenBwdSR{n}{d}{p} \\[0.6em]
+%
+%
+(\textsc{P\, \texttt{+} \,C}) \;\; &
+\stenCenS{n}{d} \\
+\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenCenSR{n}{d}{p} \\[0.6em]
+%
+%
+(\textsc{irrefl}) \;\; & (R \, \term{*} \, S^\irreflS) \, \term{+} \,
+                         (R^\irreflS \, \term{*} \, S) \;
+\equiv \; R \, \term{*} \, S \\[0.4em]
+%
+(\textsc{$\ast$Dim}) \;\; & R^d \, \term{$\ast$} \, S^d \equiv R^d
+                             \, \term{+} \, S^d \\[0.4em]
+%
+(\texttt{+}\textsc{idem}) \;\; & S \; \texttt{+} \; S \; \equiv \; S \\[0.4em]
+%
+(\texttt{+}\textsc{comm}) \;\; & S \; \texttt{+} \; T \; \equiv \; T \;
+                       \texttt{+} \; S \\[0.4em]
+%
+(\texttt{+}\textsc{assoc}) \;\; & R \; \texttt{+} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
+                       \texttt{+} \; S) \; \texttt{+} \; T \\[0.4em]
+(\texttt{*}\textsc{comm}) \;\; & S \; \texttt{*} \; T \; \equiv \; T \;
+                       \texttt{*} \; S \\[0.4em]
+%
+(\texttt{*}\textsc{assoc}) \;\; & R \; \texttt{*} \; (S \; \texttt{*} \; T) \; \equiv \; (R \;
+                       \texttt{*} \; S) \; \texttt{*} \; T \\[0.4em]
+(\textsc{dist}) \;\; & R \; \texttt{*} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
+                       \texttt{*} \; S) \; \texttt{+} \; (R
+                       \; \texttt{*} \; T)
+\end{array}
+\end{array}
+\end{align*}
+\caption{Equations on specifications (\emph{regions})}
+\vspace{-0.5em}
+\label{fig:equations}
+\end{figure}
+
+Figure~\ref{fig:equations} lists the equational theory for
+specifications given by the binary relation $\equiv$
+on \emph{region} terms. The equations use abbreviations
+\term{point}, \term{fwd}, \term{bwd}, \term{cen}
+for \term{nonpointed}, \term{forward}, \term{backward},
+\term{centered} respectively. Furthermore, the abbreviations include syntactic sugar
+for quantifying over the absence or presence of the \irrefl{}
+attribute on region constants as a boolean. For example,
+\term{fwd} syntactic sugar is defined:
+%%
+\begin{align*}
+\hspace{-0.25em}
+& \stenFwdSR{n}{d}{r}
+\; := \; \\[-0.4em]
+& \qquad \begin{cases}
+\stenFwd{n}{d}{, \irrefl} & \;\;\textit{if} \;\, \neg \, r \\[-0.1em]
+\stenFwd{n}{d}{}  & \;\;\textit{if} \;\,  r
+\end{cases} \\
+\textit{or} \;\;
+& \stenFwdS{n}{d} := \stenFwd{n}{d}{}
+\end{align*}
+%
+and similarly for \term{bwd} and \term{cen}.
+
+Stencil specifications in \textit{spec} are considered
+equal when they have the same modifiers and $\equiv$ equates
+their regions, \ie{}:
+\begin{align*}
+(\textit{region} \equiv \textit{region'}) 
+ \Rightarrow \quad\;\; & \texttt{stencil} \; [\textit{approx},] \; [\textit{mult},] \;
+\textit{region} \\[-0.4em]
+\quad \equiv \; & \texttt{stencil} \; [\textit{approx},] \;
+            [\textit{mult},] \; \textit{region}'
+\end{align*}
+%
+
+\paragraph{Overlapping regions with +}
+The first nine
+equations (with labels of the form $\ldots \texttt{+} \ldots$)
+explain the notion of \emph{region overlapping} with sum
+\term{+}. In the first six, $p$ and $q$ range
+over booleans denoting the absence (true) or presence (false)
+of \irrefl{}. These are combined via disjunction. For example,
+the rule $(\textsc{B} \, \term{+} \,
+\textsc{F})$ explains that the sum of \term{backward}
+and \term{forward} regions in the same dimension and the same
+depth is equivalent to a \term{centered} region of that depth.
+If the \term{forward} part is non-pointed ($p =
+\textit{false}$) but \term{backward} is pointed ($q =
+\textit{true}$) then their sum is equal to a \term{centered}
+region with no \irreflS{} modifier by the disjunction ($p \vee q =
+\textit{true}$), \ie{}:
+%%
+\begin{align*}
+& \stenCenS{n}{d}  \\[-0.2em]
+\equiv \;\; & \stenFwdSR{n}{d}{\irrefl} \; \texttt{+} \; \stenBwdS{n}{d}
+\end{align*}
+%%
+That is, the \term{fwd} region is irreflexive describing code which is missing
+a zero-offset neighbourhood index (\eg{}, missing $\texttt{a(i)}$), but the addition of
+the \term{bwd} has no irreflexive attribute and thus provides the
+missing zero offset index.
+
+The equation (\textsc{irrefl}) provides an annihilation of
+\irrefl{} using a further syntactic sugar.
+We write $R^{\irreflS}$ to denote a region constant with an
+non-pointed attribute, \eg{},
+%%
+\begin{align*}
+\stenFwdS{n}{d}^{\irreflS}  := \,
+\stenFwdSR{n}{d}{\irrefl}
+\end{align*}
+%%
+The left-hand side of (\textsc{irrefl}) has a sum of
+products of two regions $R$ and $S$, 
+where in each component of the sum one of $R$
+or $S$ has the irreflexive attribute. These cancel to give
+just $R \, \term{*} \, S$.
+
+The equation (\textsc{$\ast$Dim}) uses syntactic sugar
+where $R^d$ and $S^d$ denote regions whose dimension is $d$.
+Thus, the rule shows that a product of two regions on 
+the same dimension is equal to their addition.
+
 \section{Analysis, checking, and inference}

### 2016-11-13 paper.tex
equations and inequations

@@
 %\include{results}
+
+%\newcommand{\cons}[3]{\mathbf{C}_{#3}(#1, #2)}
+\newcommand{\consName}{\textbf{cons0}}
+\newcommand{\cons}[2]{#2 \;\, \consName{} \,\, #1}
+\newcommand{\consAName}{\textbf{cons}}
+\newcommand{\consA}[2]{#2 \;\, \consAName{} \,\, #1}
 
@@
 
+\paragraph{Interaction between \term{+} and \term{*}}
+
+The last seven equations (from (\textsc{\term{*}Dim}) inclusive), explain
+the algebraic behaviour of \term{+} and \term{*}.  Together, we see
+that \term{+} is an associative, commutative, idempotent binary
+operator that distributes with \term{*}, which is associative and
+commutative. This distribution of \term{*} over \term{+} is key as it
+is used internally to give a normal form for stencil specifications
+akin to Disjunctive Normal Form (DNF) (where \term{+} is 
+disjunction and \term{*} is conjunction). We revisit this
+normalisation in \S~\ref{sec:inf-step4}.
+
 The equation (\textsc{$\ast$Dim}) uses syntactic sugar
 where $R^d$ and $S^d$ denote regions whose dimension is $d$.
-Thus, the rule shows that a product of two regions on 
+This equation shows that a product of two regions on 
 the same dimension is equal to their addition.
+
+The equational theory provides not only an axiomatic semantics
+for the language, which the user can use to rewrite their
+specification, but it is also used extensively in the inference
+procedure (\S~\ref{subsec:inference}) to simplify specifications.
+
+\begin{theorem}[Soundness of equational theory]
+\begin{equation*}
+\forall S, T, n . \quad
+S \equiv T \; \Rightarrow \;
+\interp{S}_n = \interp{T}_n
+\end{equation*}
+\end{theorem}
+
+\subsection{Approximation}
+
+Figure~\ref{fig:inequations} defines a notion of sub-specifications
+via the relation $<:^r$ on \textit{region}
+syntax and then $<:$ on \textit{spec} syntax inductively.
+Both relations are reflexive and transitive (we omit
+these equations) and congruent
+with respect to \term{+}, \term{*}, \term{atLeast}, \term{atMost},
+and \term{readOnce}. The (\trule{\textsc{eq}}) rule connects region
+equations with inequations. The (\trule{\textsc{over}}) rule explains
+the notion of spatial over-approximation via overlapping regions.
+The (\trule{\textsc{rep}}) allows the modifier that
+enforces that every index is read just once to be dropped
+as an approximation; (\trule{\textsc{bound}}) gives
+a connection between the \term{atLeast} and \term{atMost}
+approximation modifiers.
+
+\begin{theorem}[Soundness of approximation]
+\begin{equation*}
+\forall S, T, M, n . \quad
+S <: T \; \wedge (\consA{\interp{S}_n}{M})
+\Rightarrow (\consA{\interp{T}_n}{M})
+\end{equation*}
+\end{theorem}
+
+\begin{figure}[t]
+\begin{align*}
+\hspace{-0.5em}
+\begin{array}{c}
+%%
+%% INEQUATIONS ON REGION
+%% 
+%\framebox{$\textit{region} <: \textit{region}'$} \\[0.85em]
+%
+%\dfrac{}{R <:^r R}\trule{\textsc{refl}} \qquad \dfrac{P <:^r R \quad R <:^r Q}{P <:^r
+%  Q}\trule{\textsc{trans}} 
+%\qquad
+%\\[1.5em]
+\vspace{-0.4em}
+\setlength{\arraycolsep}{0.1em}
+\dfrac{\hspace{3em} m \leq n \qquad\qquad r \implies s \hspace{3em}}
+{\begin{array}{rl}
+\stenFwdSR{m}{d}{r} & <:^r \stenFwdSR{n}{d}{s}
+%& \;\, ds \subseteq es \wedge n \leq m
+\\
+\wedge \; \stenBwdSR{m}{d}{r} & <:^r \stenBwdSR{n}{d}{s}
+%& \;\, ds \subseteq es \wedge n \leq m
+\\
+\wedge \; \stenCenSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s}
+%& \;\, ds \subseteq es \wedge n \leq m
+\\
+\wedge \; \stenBwdSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s} \\
+\wedge \; \stenFwdSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s} \\[1.2em]
+\end{array}}\trule{\textsc{over}} \\[-0.2em]
+\hspace{-0.5em}
+\dfrac{R_1 <:^r S_1 \quad R_2 <:^r S_2}
+      {R_1 \, \texttt{+} \, R_2 <:^r S_1 \, \texttt{+} \, S_2}
+\trule{\textsc{cong}\texttt{+}} 
+\dfrac{R_1 <:^r S_1 \quad R_2 <:^r S_2}
+      {R_1 \, \texttt{*} \, R_2 <:^r S_1 \, \texttt{*} \, S_2}
+\trule{\textsc{cong}\texttt{*}} \\[1.5em]
+%%
+%% INEQUATIONS ON SPEC
+%%
+%\framebox{$\textit{spec} <: \textit{spec}'$} \\[0.8em]
+%
+%
+\dfrac{R <:^r S}{\term{atLeast}  S <: \term{atLeast} 
+  R}\trule{\texttt{L}}
+\;\;
+\dfrac{R <:^r S}{\term{atMost}  R <: \term{atMost} 
+  S}\trule{\texttt{M}} \\[1.5em]
+\dfrac{R <:^r S}{R <: \term{atMost} \, S}\trule{\textsc{shrink}} \quad
+\dfrac{R <:^r S}{S <: \term{atLeast} \, R}\trule{\textsc{grow}} \quad
+ \\[1.5em]
+%%
+\dfrac{\mathsf{dims}(R) \cap \mathsf{dims}(S) = \emptyset}{R \ast S <: R}
+\trule{\textsc{Gen$\ast$}}
+\quad
+\dfrac{R \equiv S}{R <: S}\trule{\textsc{eq}} 
+%%
+\\[1.5em]
+\dfrac{R <: S}{\term{readOnce} \, R <: \term{readOnce} \,
+  S}\trule{\textsc{cong}\texttt{R}}
+\;\;
+\dfrac{}{\term{readOnce} \, R <: R}\trule{\textsc{rep}}
+%\dfrac{}{R <: \term{atMost} R} 
+%\quad
+%\dfrac{}{R <: \term{atLeast} R}\\[1.5em]
+\end{array}
+\end{align*}
+\caption{Inequations $<:^r$ on \textit{regions} and
+$<:$ on \textit{specs} (omitting reflexivity
+and transitivity). }
+\label{fig:inequations}
+\vspace{-0.8em}
+\end{figure}
 
@@
 \label{sec:analysis}
-
-%\newcommand{\cons}[3]{\mathbf{C}_{#3}(#1, #2)}
-\newcommand{\consName}{\textbf{cons0}}
-\newcommand{\cons}[2]{#2 \;\, \consName{} \,\, #1}
-\newcommand{\consAName}{\textbf{cons}}
-\newcommand{\consA}[2]{#2 \;\, \consAName{} \,\, #1}
 
@@
 %%
-\trule{up}\dfrac{\forall u \in M \, \exists v \in N .  \, u \preceq v}
+\trule{up}\dfrac{\forall u \in M, \, \exists v \in N .  \, u \preceq v}
       {\cons{\textsf{upper}(N)}{M}}
 \qquad
-\trule{low}\dfrac{\forall v \in N \, \exists u \in M . \, u \preceq v}
+\trule{low}\dfrac{\forall v \in Nm \, \exists u \in M . \, u \preceq v}
       {\cons{\textsf{lower}(N)}{M}} \\[1.5em]

### 2016-11-13 paper.tex
reinstanting inference and fixing a few things

@@
 %%
-That is, the \term{fwd} region is irreflexive describing code which is missing
+That is, the \term{fwd} region is non-pointed describing code which is missing
 a zero-offset neighbourhood index (\eg{}, missing $\texttt{a(i)}$), but the addition of
-the \term{bwd} has no irreflexive attribute and thus provides the
+the \term{bwd} has no non-pointed attribute and thus provides the
 missing zero offset index.
@@
 where in each component of the sum one of $R$
-or $S$ has the irreflexive attribute. These cancel to give
+or $S$ has the non-pointed attribute. These cancel to give
 just $R \, \term{*} \, S$.
@@
 \begin{example}
+\label{exm:checking}
 We demonstrate the analysis and modelise procedures over the
@@
 \subsection{Specification inference} 
+\label{subsec:inference}
+
+We demonstrate the inference procedure over the
+ program from Exmaple~\ref{exm:checking} (p.~\pageref{exm:checking})
+ computing the mean of a five-point stencil:
+%%
+\begin{minted}{fortran}
+do i = 1, (n-1)
+  do j = 1, (m-1)
+    x = a(i-1, j)+a(i+1, j); y = a(i, j-1)+a(i, j+1)
+    b(i, j) = (a(i, j) + x + y) / 5.0
+  end do
+end do
+\end{minted}
+%
+\subsubsection{Step 1: Standard analysis and data access analysis}
+\label{sec:inf-step1}
+
+The first step of the inference is to perform the 
+analysis of Section~\ref{subsec:analysis} to produce a 
+model of the code in terms of indexing schemes.
+The \emph{analyse} and \emph{modelise} procedures before yielded the finite map:
+%
+\begin{align*}
+\texttt{a} \mapsto \mathsf{only}(\{\vtwoh{-1}{0}, \vtwoh{1}{0},
+          \vtwoh{0}{-1}, \vtwoh{0}{1}, \vtwoh{0}{0}\})
+\end{align*}
+%
+
+\subsubsection{Step 2: Coalesce contiguous schemes into spans}
+\label{sec:inf-step3}
+
+Let $U$ range over the finite maps computed by \textit{modelise}.  For
+each array variable $a \in \mathsf{dom}(U)$, the algorithm constructs next a set of
+$n$-dimensional rectangles covering all contiguous groups of schemes
+in $U(a)$.
+
+\newcommand{\spanOp}{\textsf{spans}}
+
+\begin{definition}[Spans]
+  A \emph{span} represents an $n$-dimensional rectangle
+  as a pair of $n$-dimensional vectors (or
+  represented as a $2 \times n$ matrix) with values drawn from
+  $\domainVal{}$. This gives the co-ordinates of a lower-bound vertex
+  (first component) and the upper-bound (second component). We write 
+  spans as $\vect{x} = [\vect{x}^L \; \vect{x}^U]$, where $\vect{x}^L$
+  and $\vect{x}^U$ give the lower and upper bounds.  In
+  the following, $\vect{x}, \vect{y}$ range over spans.  For a span
+  $\vect{x}$, we write $\vect{x}_{i:j}$ for the sub-span comprising the
+  subvectors $[\vect{x}^L_{i:j} \; \vect{x}^U_{i:j}]$ from entry $i$
+  to entry $j$ (inclusive).
+\end{definition}
+
+\noindent
+Each indexing scheme $u \in U(a)$ is converted to a \emph{unit} span
+$u \mapsto [u \; u]$, defining the lower and upper bound of an
+$n$-dimensional cube of side length $1$.  For the five-point stencil
+example there are five unit spans.
+Figure~\ref{fig:inference-steps-informal}(a) illustrates these.
+
+The next step repeatedly coalesces any contiguous spans until a fixed
+point is reached, resulting in a set of (possibly overlapping) spans
+covering the original space of schemes. Two $n$-dimensional spans
+$\vect{x}$ and $\vect{y}$ are contiguous if for all but one
+$j \in \{1, \ldots, n\}$ then $\vect{x}_j = \vect{y}_j$ and for one
+$j$ then $\vect{x}^U_1 + 1 = \vect{y}^U$. For example, spans
+$\left[\vtwoh{-1}{0}\; \vtwoh{-1}{0}\right]$ and
+$[\vtwoh{0}{0} \; \vtwoh{1}{0}]$ are contiguous in the first dimension
+and can be coalesced into the span $[\vtwoh{-1}{0} \; \vtwoh{1}{0}]$.
+
+\begin{definition}\label{def:span-coalesc} 
+  Span coalescing $\bullet$ is defined for spans contiguous in the
+  first dimension:
+%%
+\begin{align*}
+\vect{x} \bullet \vect{y}
+= &
+\begin{cases}
+(\vect{x}^L, \vect{y}^U) & \vect{x}^U_1 + 1 = \vect{y}^L_1 \; \wedge \;
+\vect{x}_{2:n} = \vect{y}_{2:n}) \\
+\bot  & \textit{otherwise}
+\end{cases}
+% $m$-dimensions into coalesced spans which vary in
+% $m+1$-dimensions.
+\end{align*}
+%%
+\end{definition}
+\noindent
+To coalesce regions which are contiguous in other dimensions
+other than the first, we generate all rotations of the spans in
+$n$-dimensional space by permutation on spans, and fold together contiguous
+regions using $\bullet$ (above). This 
+generates all possible ways of coalescing regions. For our example,
+the unit spans in Figure~\ref{fig:inference-steps-informal}(a)
+are coalesced into the contiguous spans in
+Figure~\ref{fig:inference-steps-informal}(b).
+This procedure is iterated until a fixed point is reached. In this
+example this is reached in the first step. 
+ Then any spans that are completely contained by any other span are deleted,
+leaving a minimal set of spans (which may overlap, but none of which
+fully contains another), Figure~\ref{fig:inference-steps-informal}(c).
+The final spans are used to synthesise concrete stencil specification
+syntax (\S~\ref{sec:inf-step4}), where
+each span becomes a product \texttt{*} of regions which are 
+combined by the sum \texttt{+}.
+
+More formally the algorithm proceeds as follows.
+For all $a \in \mathsf{dom}(U)$, let $M = U(a)$. Every scheme in $M$ 
+is mapped to a unit span
+by $\textsf{unitSpans}(M) = \{[u \; u] \mid u\!\in\!M\}$. For our example:
+%%
+\begin{align*}
+  N(U(\texttt{a})) = \stwo{0}{0}{0}{0} \stwo{1}{0}{1}{0} \stwo{-1}{0}{-1}{0} \stwo{0}{1}{0}{1} \stwo{0}{-1}{0}{-1}
+\end{align*}
+%%
+In the following we let $M_0 = N(U(\texttt{a}))$ for our example.
+
+Each set of unit spans is applied to the fixed point of the \spanOp{} function
+(defined shortly), which coalesces spans into contiguous regions. That is, we compute
+ $(\mu \, \spanOp)$ on input $N(M)$, where \spanOp{} is defined by the steps:
+
+\begin{figure}[t]
+\vspace{-0.8em}
+\hspace{-0.5em}
+\begin{minipage}{0.16\linewidth}
+\begin{align*}
+\setlength{\arraycolsep}{0em}
+\begin{array}{c}
+\quad \\[0em]
+\begin{tikzpicture}[scale=0.8]
+%\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
+\node at (1.25,1.22) {\textnormal{\footnotesize{0,0}}};
+\draw[thick] (1,1) rectangle (1.5,1.5);
+\draw[thick] (1.5,1) rectangle (2,1.5);
+\draw[thick] (1,0.5) rectangle (1.5,1);
+\draw[thick] (0.5,1) rectangle (1,1.5);
+\draw[thick] (1,1.5) rectangle (1.5,2);
+\end{tikzpicture} \\[0em]
+(a)
+\end{array}
+\end{align*}
+\end{minipage} $\rightarrow$ \begin{minipage}{0.43\linewidth}
+\begin{align*}
+\setlength{\arraycolsep}{0em}
+\begin{array}{c}
+\begin{tikzpicture}[scale=0.72]
+%\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
+%\node at (1.25,1.22) {\textnormal{\small{0,0}}};
+\draw [black, thin] (0.35,0) to [round left paren ] (0.35,2);
+\draw[thick] (0.5,1) rectangle (2,1.5);
+\draw[thick] (1,0.40) rectangle (1.5,0.90);
+\draw[thick] (1,1.60) rectangle (1.5,2.10);
+\node at (2.3,1.22) {+};
+\draw[thick] (3.8,1) rectangle (4.3,1.5);
+\draw[thick] (2.6,1) rectangle (3.1,1.5);
+\draw[thick] (3.2,0.5) rectangle (3.7,2);
+\draw [black, thin] (4.4,0) to [round right paren ] (4.4,2);
+\end{tikzpicture} \\[-0.5em] (b)
+\end{array}
+\end{align*}
+\end{minipage}$\rightarrow$ \begin{minipage}{0.25\linewidth}
+\begin{align*}
+\setlength{\arraycolsep}{0em}
+\begin{array}{c}
+\begin{tikzpicture}[scale=0.72]
+%\node at (1.25,1.22) {\textnormal{\small{0,0}}};
+\draw [black, thin] (0.8,0) to [round left paren ] (0.8,2);
+\draw[thick] (1,0.5) rectangle (1.5,2);
+\node at (1.8,1.22) {+};
+\draw[thick] (2.1,1) rectangle (3.6,1.5);
+\draw [black, thin] (3.7,0) to [round right paren ] (3.7,2);
+%\node at (2.75,1.22) {\textnormal{\small{0,0}}};
+\end{tikzpicture} \\[-0.5em] (c)
+\end{array}
+\end{align*}
+\end{minipage}
+\caption{Illustration of grouping spans in
+specification inference}
+\label{fig:inference-steps-informal}
+\vspace{-0.8em}
+\end{figure}
+
+\noindent
+\textbf{1)\,} Compute all permutations on the column vectors in a span, \ie{},
+  $[\vect{x}^L \vect{x}^U] \mapsto [\pi\vect{x}^L \, \pi\vect{x}^U]$
+for a permutation $\pi$. For each permutation function $\pi^n_i$
+(the $i$-th permutation for vectors of size $n$) we pair the
+permutation function with the set of permuted spans so that
+the spans can be un-permuted later.
+%
+\begin{equation*}
+P(M) = \bigcup_{i \in n!} (\pi^n_{i} , \; \{[\pi^n_i
+\vect{x}^L \; \pi^n_i\vect{x}^U] \, \mid \, [\vect{x}^L \; \vect{x}^U]
+\leftarrow M\}
+\end{equation*}
+%
+This corresponds to all rotations of the space.  For our example:
+%
+\begin{align*}
+P(M_0) =
+\{&(\pi^2_1, \{ \stwo{0}{0}{0}{0},
+\stwo{1}{0}{1}{0},
+\stwo{-1}{0}{-1}{0},
+\stwo{0}{1}{0}{1},
+\stwo{0}{-1}{0}{-1} \})
+\\
+&(\pi^2_2, \{
+ \stwo{0}{0}{0}{0},
+ \stwo{0}{1}{0}{1},
+ \stwo{0}{-1}{0}{-1},
+ \stwo{1}{0}{1}{0},
+ \stwo{-1}{0}{-1}{0}\})\}
+\end{align*}
+%
+where $\pi^2_1$ is the identity permutation and $\pi^2_2$ is the
+permutation flips the order of the two elements in the 
+vectors.
+
+\noindent
+\textbf{2)\,} Sort each permutation set into an ordered list, by the ordering:
+\begin{equation*}
+  \vect{x} \leq \vect{y} =
+      \exists i \, . \; \vect{x}^L_{i} \leq \vect{y}^L_{i} \; \wedge \;
+        (i = n \vee \vect{x}_{i+1:n} = \vect{y}_{i+1:n})
+\end{equation*}
+%
+giving the sorting function $\textsf{sort}(M)$. For our example:
+%%
+\begin{align*}
+\textsf{sort}(P(M_0)) = 
+\{&(\pi^2_1, [
+\stwo{0}{-1}{0}{-1},
+\stwo{-1}{0}{-1}{0},
+\stwo{0}{0}{0}{0},
+\stwo{1}{0}{1}{0},
+\stwo{0}{1}{0}{1}] )
+\\
+&(\pi^2_2, [
+\stwo{0}{-1}{0}{-1},
+\stwo{-1}{0}{-1}{0},
+\stwo{0}{0}{0}{0},
+\stwo{1}{0}{1}{0},
+\stwo{0}{1}{0}{1}])\}
+\end{align*}
+\noindent
+\textbf{3)\,} Reduce each list pairwise by $\bullet$ operation 
+(Defn.~\ref{def:span-coalesc}) to coalesce contiguous regions,
+given by $\textsf{fold}\bullet$. For our example, this yields:
+%
+\begin{align*}
+\{&(\pi^2_1, [
+\stwo{0}{-1}{0}{-1},
+\stwo{-1}{0}{1}{0},
+\stwo{0}{1}{0}{1}
+]), (\pi^2_2, [
+\stwo{0}{-1}{0}{-1},
+\stwo{-1}{0}{1}{0},
+\stwo{0}{1}{0}{1}])\}
+\end{align*}
+%
+\textbf{4)\,} Un-permute and union together, \ie{},
+%
+\[
+U(M) = \bigcup \{[\pi \vect{x}^L, \pi \vect{x}^U]
+ \mid [\vect{x}^L, \vect{x}^U] \leftarrow S, (\pi, S) \leftarrow M\}
+\]
+For our example:
+%
+\begin{align*}
+U(M) =
+\{\stwo{0}{-1}{0}{-1},
+\stwo{-1}{0}{1}{0},
+\stwo{0}{1}{0}{1},
+\stwo{-1}{0}{-1}{0},
+\stwo{0}{-1}{0}{1},
+\stwo{1}{0}{1}{0}\}
+\end{align*}
+%
+\textbf{5)\,} Filter by the region containment predicate $\containedin$, that
+  is, if any region is contained within another then remove the
+  smaller, defined:
+%
+\begin{align*}
+& \vect{x} \containedin \vect{y} = \vect{y}^L_1 \leq \vect{x}^L_1 \wedge \vect{x}^U_1 \leq \vect{y}^U_1
+  \wedge (\vect{x}^L_{2:n}, \vect{x}^U_{2:n}) \containedin
+  (\vect{y}^L_{2:n}, \vect{y}^U_{2:n})
+\end{align*}
+For our example this then yields the final result of:
+\begin{align*}
+\spanOp(M)
+= \{\stwo{-1}{0}{1}{0} \stwo{0}{-1}{0}{1}\}
+\end{align*}
+since $\stwo{0}{1}{0}{1} \sqsubseteq \stwo{0}{-1}{0}{1}$
+and $\stwo{1}{0}{1}{0} \sqsubseteq \stwo{-1}{0}{1}{0}$.
+Applying \spanOp{} again yields the same
+result for our example; the fixed point is reached.
+
+\subsubsection{Step 4: Synthesise specifications from spans}
+\label{sec:inf-step4}
+
+\newcommand{\oplusbig}{\operatornamewithlimits{\term{\Large{$\oplus$}}}}
+\newcommand{\odobig}{\operatornamewithlimits{\term{\Large{$\odot$}}}}
+\newcommand{\bplus}{\operatornamewithlimits{\term{\Large{+}}}}
+\newcommand{\tySum}[1]{#1^{\term{+}}}
+\newcommand{\tyProd}[1]{#1^{\term{*}}}
+\newcommand{\specDNF}{\textbf{spec}}
+
+Abstract syntax trees of the specification language
+are synthesised from the set $\spanOp(M)$ of coverings spans. 
+This conversion is defined in terms
+of an algebra on \textit{spec} syntax in disjunctive-normal form
+(which we denote \specDNF{}) and which is wrapped
+in the $\textsf{Approx}$ type (see
+Defn.~\ref{def:mult-and-approx},
+p.~\pageref{def:mult-and-approx}).  The algebra mirrors the shape
+of the region combinators for sum and product, 
+with operators $\oplus$ and $\odot$ and a constructor for \specDNF{}:
+%%
+\begin{align*}
+\oplus & : \mathsf{Approx}(\specDNF) \rightarrow \mathsf{Approx}(\specDNF) \rightarrow
+  \mathsf{Approx}(\specDNF) \\
+\odot & : \mathsf{Approx}(\specDNF) \rightarrow \mathsf{Approx}(\specDNF) \rightarrow
+        \mathsf{Approx}(\specDNF) \\
+\mathsf{toSpec} & : \mathbb{N}_{>0} \rightarrow (\domainVal{} \times \domainVal{}) \rightarrow \mathsf{Approx}(\specDNF)
+\end{align*}
+%%
+The $\mathsf{toSpec}$ function maps a dimension identifier
+and a pair of lower and upper bound
+values (one-dimensional) to a specification with a one-dimensional region. 
+An entire span is converted into an $n$-dimensional
+specification by applying $\mathsf{toSpec}$ to each
+pair of lower and upper bounds in a span, per dimension, and multiplying these
+specifications by $\odot$. The resulting specifications for
+each span are then summed by $\oplus$:
+\begin{equation*}
+\textsf{spec}(M) = 
+\oplusbig\limits_{\vect{x} \in \spanOp(M)} \,
+\odobig\limits_{i \in \{1, \ldots, n\}} \, \mathsf{toSpec} \; i \; (\vect{x}^L_i, \vect{x}^U_i)
+\end{equation*}
+We unpack the definitions of this algebra.
+
+\begin{definition}[Specifications in DNF]
+Let $r_1, ..., r_n$ range over elements of the \textit{region}
+syntax (see Figure~\ref{fig:syntax}, p.~\pageref{fig:syntax}).
+The set \specDNF{} is a subset of syntax trees \textit{spec},
+containing specifications in disjunctive normal form over regions 
+where the disjunct is \term{+} (since its model is in terms of set
+union) and the conjunct is \term{*} (since its model is a form of tensor).
+That is elements of \specDNF{} are of the form:
+%%
+$
+(r^1_{1} \; \term{*} \ldots \term{*} \; r^1_n)\; \term{+} \; \ldots 
+\term{+} \; (r^m_1 \, \term{*} \; \ldots \term{*} \; r^m_p)
+$.
+
+More formally, $\specDNF \subseteq (\textit{spec} \cup \{\epsilon\})$, defined:
+%%
+\begin{align*}
+\specDNF{} & ::= \tySum{\specDNF} \mid \epsilon \\
+\tySum{\specDNF} & ::= \tyProd{\specDNF} \; \term{+} \; \tySum{\specDNF} \mid
+  \tyProd{\specDNF} \\
+\tyProd{\specDNF} & ::= \textit{region} \; \term{*} \; \tyProd{\specDNF} \mid
+   \textit{region}
+\end{align*}
+%%
+We include the empty element $\epsilon$ which is not present in the
+syntax. 
+In the following, $S, T$ range over elements of
+\specDNF{} and $\tyProd{s}, \tyProd{t}$ range over elements
+of $\tyProd{\specDNF}$ (for example, $\tyProd{s} = r^1_{1} \; \term{*}
+\ldots \term{*} \; r^1_n$).
+\end{definition}
+
+\newcommand{\exactR}{\mathsf{exact}\;}
+\newcommand{\exact}[1]{\exactR(#1)}
+\newcommand{\upperB}[1]{\mathsf{upper}\;(#1)}
+
+\begin{figure}[t]
+%\newcommand{\llower}{\vect{x}^L_i}
+%\newcommand{\lupper}{\vect{x}^U_i}
+\vspace{-0.7em}
+\newcommand{\llower}{l}
+\newcommand{\lupper}{u}
+{\scalebox{0.97}{
+\begin{minipage}{1\linewidth}
+\begin{align*}
+&\mathsf{toSpec} \; d \; (\llower, \lupper) = \\
+&\hspace{-0.4em}\begin{cases}
+%% POINT
+\exact{\stenReflS{d}} & \llower = 0 \, \wedge \, \lupper = 0\\[0.2em]
+%% FWD
+\exact{\stenFwdS{\lupper}{d}}  & \llower = 0 \, \wedge \, \lupper > 0 \\[0.2em]
+\exact{\stenFwdSR{\lupper}{d}{\irreflS}}\!\!\!\! & \llower = 1 \, \wedge \, \lupper > 0 \\[0.2em]
+%% BWD
+\exact{\stenBwdS{|\llower|}{d}}  & \llower < 0 \, \wedge \, \lupper = 0 \\[0.2em]
+\exact{\stenBwdSR{|\llower|}{d}{\irreflS}}\!\!\!\! & \llower < 0 \, \wedge \, \lupper = -1 \\[0.2em]
+%% CEN
+\exact{\stenCenS{\lupper}{d}}& \llower < 0 \wedge \lupper > 0 \wedge |\llower| = \lupper \\[0.2em]
+\exactR(\stenBwdS{|\llower|}{d} & \llower < 0 \wedge \lupper > 0  \wedge |\llower| \neq \lupper \\
+\qquad\qquad \texttt{+} \, \stenFwdS{\lupper}{d}) & \\[0.2em]
+% BOUND
+\upperB{\stenFwdS{\lupper}{d}} & \llower > 1 \\[0.2em]
+\upperB{\stenBwdS{|\llower|}{d}} \quad & \llower < (-1)
+ \\[0.2em]
+% ABSOLUTE REP
+\exact{\epsilon} & \llower = \infty \, \vee \, \lupper = \infty
+  \end{cases}
+\end{align*}
+\end{minipage}
+}}
+\caption{\textsf{toSpec} constructor of $\mathsf{Approx}(\specDNF)$ values.}
+\label{fig:to-spec}
+\vspace{-0.8em}
+\end{figure}
+
+\begin{definition} The $\mathsf{toSpec}$ constructor is defined
+in Figure~\ref{fig:to-spec}.
+The first two cases map to \texttt{forward} stencils since the lower
+bound is $0$ or $1$, with the depth determined by the upper bound of
+the span. If the lower bound is $1$, then the
+$0$-offset point is missing so the region is given a
+\texttt{nonpointed} attribute. The second two cases are similar for
+\texttt{backward} but now the depth of the region is determined by the
+absolute of the lower bound. In the fifth-case, where the lower bound
+is negative and the upper bound is positive and they have the same
+magnitude, then a \texttt{centered} region is constructed with
+the depth as the common magnitude. If the magnitudes are not the same
+(sixth case) then separated \texttt{forward} and \texttt{backward}
+regions are summed, with their corresponding depths.
+
+If the lower bound is greater than 1
+(which implies the upper bound is greater than 1 too) then
+this represents a region that is not at the origin (or next to) the
+origin and thus cannot be represented in the specification language.
+Thus, an upper bound is instead returned showing the maximum extent
+of the region. The penultimate case is the dual, when the lower bound is less
+than $-1$, giving an upper bound on a \texttt{backward} specification.
+
+Finally, if either bound is $\infty$, representing a non
+neighbourhood index, then an empty specification is given
+ with no regions, indicating that there is not even a 
+pointed 0-offset access in this dimension.
+
+Note, $\mathsf{toSpec} \, d \, (l, u)$ is always defined 
+ if $l$ less than or equal to $u$. 
+\end{definition}
+
+\begin{definition} The $\odot$ operation on $\textsf{Approx}(\specDNF{})$
+  is defined in terms of the intermediate operator $\hat{\odot}$ on
+  \specDNF{}, defined:
+%%
+\begin{align*}
+S \,\hat{\odot}\, \epsilon = S \quad
+\epsilon \,\hat{\odot}\, S = S \qquad
+S \,\hat{\odot}\, T = \bplus_{(\tyProd{s}, \tyProd{t}) \in (S \times
+   T)} (\tyProd{s} \; \term{*} \; \tyProd{t})
+\end{align*}
+%%
+That is, the $\hat{\odot}$ product of two specifications $S$ and $T$ in DNF form 
+is the $\term{+}$ sum of all pairwise $\term{*}$ products for every
+pair of $\tyProd{spec}$ drawn from $S$ and $T$. The result
+is indeed in DNF-form, where $\hat{\odot}$ is essentially applying
+ the (\trule{\textsc{dist}}) rule of our equational theory
+ (Fig.~\ref{fig:equations},p.~\pageref{fig:equations}). The operation is
+commutative and associative, and sound with respect to our
+model (see Lemma~\ref{lem:alg-soundness} below).
+The $\hat{\odot}$ operation is then lifted to all combinations of
+\textsf{Approx} for $\odot$:
+\begin{align*}
+\textsf{lower}(S) \odot \textsf{exact}(T) = \; &
+       %\textsf{exact}(T) \odot \textsf{lower}(S) = \;
+       \textsf{both} \, (S \hat{\odot} T, \; T) \\
+\textsf{upper}(S) \odot \textsf{exact}(T) = \; & 
+       %\textsf{exact}(T) \odot \textsf{upper}(S) =
+       \textsf{both} \, (T, \; S \hat{\odot} T) \\
+\textsf{lower}(S) \odot \textsf{both}(T_l,T_u) = \; &
+        %\textsf{both}(T_l,T_u) \odot \textsf{lower}(S) = \;
+        \textsf{both} \, (S \hat{\odot} T_l, \; T_u) \\
+\textsf{upper}(S) \odot \textsf{both}(T_l,T_u) = \; & \textsf{both} \,
+                                                      (T_l, \; S \hat{\odot} T_u) \\
+\textsf{both}(S_l,S_u) \odot
+\textsf{both}(T_l,T_u)  = \; & \textsf{both} (S_l \hat{\odot} T_l, \; S_u
+                         \hat{\odot} T_u) \\
+\textit{inj}(S) \odot \textit{inj}(T) = \; & \textit{inj}(S
+                                              \hat{\odot} T)
+\end{align*}
+where in the last case $\textit{inj}$ corresponds to the unary
+injections of \textsf{Approx}. For brevity we omit the cases where the arguments above are flipped
+as $\odot$ is commutative.
+\end{definition}
+
+\begin{definition}The $\oplus$ operation on $\textsf{Approx}(\specDNF)$ 
+is defined in a similar way to $\odot$, with the intermediate $\hat{\oplus}$ on
+$\specDNF$ given by:
+%%
+\begin{equation*}
+S \,\hat{\oplus}\, \epsilon = S \quad
+\epsilon \,\hat{\oplus}\, S = S \quad
+S \, \hat{\oplus} \, T = S \, \term{+} \, T
+\end{equation*}
+%%
+Thus, if neither specification is empty, their $\hat{\oplus}$ is just
+the syntactic sum $\term{+}$. Then $\oplus$ is the lifting of
+$\hat{\oplus}$ to \textsf{Approx} with the shape as $\odot$ in the previous
+definition.
+\end{definition}
+
+\begin{lemma}[$\hat{\odot}$, $\hat{\oplus}$ soundness]$\forall S,
+  T \in \specDNF$ where $S\!\neq \epsilon \wedge T\!\neq \epsilon$:
+\begin{align*}
+\interp{S \; \hat{\odot} \; T}_n = \interp{S}_n \otimes \interp{T}_n
+  \qquad
+\interp{S \; \hat{\oplus} \; T}_n = \interp{S}_n \cup \interp{T}_n
+\end{align*}
+\label{lem:alg-soundness}
+\vspace{-2em}
+\end{lemma}
+%
+\noindent
+Thus our model validates the specification synthesis process which concludes the
+inference procedure. For our example, we infer and then synthesise the
+5-point specification:
+%%
+\begin{minted}[linenos=false]{fortran}
+!= stencil centered(depth=1, dim=1)*pointed(dim=2) + centered(depth=1, dim=2)*pointed(dim=1):: a
+\end{minted}
+%%
+
+\section{Evaluation}
 

### 2016-11-13 paper.tex
small tweaks

@@
 At the level of syntax, this translates to saying
-$(i, j+1)$ can be instantiated (specialised) to $(1, j+1)$. 
+$(i, j+1)$ can be instantiated to $(1, j+1)$. 
 Note the last example is undefined as its first expression is a
@@
 %
+
+\begin{figure}
+\input{spatial-model.tex}
+\caption{\textit{spec} model,
+$\interp{-}_n : \textit{spec} \rightarrow \textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$}
+\label{fig:spatial-model}
+\vspace{-0.75em}
+\end{figure}
 
@@
 \end{align*}
-\ie{}, $\mathsf{lower}$ is an injection $\mathsf{lower} : a \rightarrow \mathsf{Approx} \, a$, and so
-on. The $\textsf{Mult}$ type corresponds to the presence or absence of the
+\eg{}, $\mathsf{lower}$ is an injection
+$\mathsf{lower} : a \rightarrow \mathsf{Approx} \, a$ etc. The
+$\textsf{Mult}$ type corresponds to the presence or absence of the
 \term{readOnce} modifier as shown in Figure~\ref{fig:spatial-model}.
@@
 \term{atLeast} and \term{atMost}. Later, it will be useful to model
-pairs of lower and upper bounds which is provided by \textsf{both} 
-but which is not used here.% These are used for labelled models, and other components later in.
+pairs of lower and upper bounds provided by \textsf{both} but which is
+not used
+here.% These are used for labelled models, and other components later in.
 \label{def:mult-and-approx}
 \end{definition}
-
-\begin{figure}
-\input{spatial-model.tex}
-\caption{\textit{spec} model,
-$\interp{-}_n : \textit{spec} \rightarrow \textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$}
-\label{fig:spatial-model}
-\vspace{-0.75em}
-\end{figure}
 
@@
 \begin{figure}
-%\vspace{-0.2em}
+\vspace{-0.3em}
 \begin{align*}

### 2016-11-14 paper.tex
Reduce verbosity of main features of the tool

@@
 
-We provide a verification tool for such specifications which can check
-the correctness of code against the specification, and can also infer
-specifications, synthesising these as comments in the
-source code. The main features and benefits of our approach are as follows:
+We provide a verification tool which can check the correctness of stencil code
+against its specification, infer such specifications from bare source, and
+synthese inferred specifications as comments and insert them into the source
+code. The main features and benefits of our approach are as follows:
 %
 \begin{itemize}
-\item The main use case is that a programmers writes a specification
-  first (perhaps informed by their numerical analysis), then writes
-  the corresponding source code, and lastly has the tool check its
-  conformance to the specifications. This specify-and-check approach
-  removes the usually ad hoc testing effort, and provides the
-  opportunity for easy re-verification in the event of any refactoring
-  of software maintenance activities.
-%
-\item The specification provides a concise description which can be
-  preserved in the source code to provide documentation to future
-  developers, contrasting with ad hoc tests which are usually
-  discarded or specialised to the development environment. Separate from the
-  verification procedure, the infer-and-synthesise mode of the tool
-  provides a way to automatically generate additional useful
-  documentation for future users.
+%\item The main use case is that a programmer writes a specification
+%  first (perhaps informed by their numerical analysis), then writes
+%  the corresponding source code, and lastly has the tool check its
+%  conformance to the specifications. This specify-and-check approach
+%  removes the usually ad hoc testing effort, and provides the
+%  opportunity for easy re-verification in the event of any refactoring
+%  of software maintenance activities.
+\item The primary use case is that a programmer writes a specification first,
+  then the corresponding source code and the tool checks conformance between
+  the two. This specify-and-check approach reduces the testing effort and
+  future-proofs against bugs introduced during refactoring and maintanence.
+%
+%\item The specification provides a concise description which can be
+%  preserved in the source code to provide documentation to future
+%  developers, contrasting with ad hoc tests which are usually
+%  discarded or specialised to the development environment. Separate from the
+%  verification procedure, the infer-and-synthesise mode of the tool
+%  provides a way to automatically generate additional useful
+%  documentation for future users.
+
+% Emphasis here is on documentation generation, so it makes little sense to
+% compare this aspect of CamFort to ad hoc testing. Apples and oranges.
+\item A specification concisely captures the access pattern and turns it into
+  documentation. Inference algorithm efficiently produces specifications with
+  no programmer intervention, while synthesis mechanism inserts the
+  specifications at appropriate places in the source code automatically.
 %
 \item Specifications are integrated into the code as comments for easy
-integration with existing tools, as well as
-  documentation systems (\eg{}, we support Doxygen and FORD syntax).
-%
-\item Our specification format is deliberately abstract, with a small
-  number of combinators that \emph{do not involve any indexing
-    expressions}, \eg{} does not involve expressions like
-  \texttt{a(i+1,j-1)}. This contrasts with other specification
-  approaches.% For example, the work of Kamil et al.~\cite{kamil2016verified}
-  %provides full-functionality specifications of stencil computations
-  %in the form of a domain-specific language.  
-  If a user wishes to specify, say, the
-  Naiver-Stokes stencil above, in a deductive verification languages
-  like ACSL, they must write a specification which is very
-  similar to the code, involving textually dense array-indexing expressions. Therefore any
-  low-level mistakes that could be made whilst programming complex indexing code could
-  also be made when writing a specification. Our specifications are much more
-  abstract, and also more lightweight with the aim of aiding adoption by
-  scientists.
+  integration with existing tools, as well as documentation systems (\eg{}, we
+  support Doxygen and FORD syntax).
+%
+\item Our specification format is deliberately abstract, with a small number of
+  combinators that \emph{do not involve any indexing expressions}, \eg{}
+  \texttt{a(i+1,j-1)}. This contrasts with other specification approaches. If a
+  user specifies, say, the Naiver-Stokes stencil above, in a deductive
+  verification language such as ACSL, they must write a specification involving
+  textually dense array-indexing expressions. Therefore, any low-level mistakes
+  that could be made whilst programming complex indexing code could also be
+  made when writing its specification. Our specifications are more abstract and
+  lightweight with the aim of aiding adoption by scientists and preventing
+  indexing errors.
 \end{itemize}

### 2016-11-14 paper.tex
Rectangle -> shape
Use agree as consistency predicate instead of cons

@@
 %\newcommand{\cons}[3]{\mathbf{C}_{#3}(#1, #2)}
-\newcommand{\consName}{\textbf{cons0}}
-\newcommand{\cons}[2]{#2 \;\, \consName{} \,\, #1}
-\newcommand{\consAName}{\textbf{cons}}
-\newcommand{\consA}[2]{#2 \;\, \consAName{} \,\, #1}
+\newcommand{\consName}{\textbf{agree\textprime}}
+\newcommand{\cons}[2]{#1 \,\, \consName{} \,\, #2}
+\newcommand{\consAName}{\textbf{agree}}
+\newcommand{\consA}[2]{#1 \,\, \consAName{} \,\, #2}
+\newcommand{\consSub}[3]{#2 \,\, \consAName{}_{#1} \,\,#3}
 
@@
   \emph{consistent}. Informally, the region defined by a specification
-  is an $n$-dimensional rectangle crossing the origin.
-  Consistency is then a two way requirement.
+  is an $n$-dimensional shape crossing the origin with respect to the induction
+  variable. Consistency is then a two way requirement.
   On the one hand, all array subscripts on $a$ contributing to the kernel
@@
 \forall S, T, M, n . \quad
-S <: T \; \wedge (\consA{\interp{S}_n}{M})
-\Rightarrow (\consA{\interp{T}_n}{M})
+S <: T \; \wedge (\consA{M}{\interp{S}_n})
+\Rightarrow (\consA{M}{\interp{T}_n})
 \end{equation*}
@@
 & \textbf{check}_n(S, \textit{assg}, a) =
-  \consA_n{\interp{S}_n}{\textit{modelise}(\textit{assg})(a)}
+  \consSub{n}{\textit{modelise}(\textit{assg})(a)}{\interp{S}_n}
 &
@@
 
-Rules $(1)$ and $(*)$ extend the $\textbf{cons}$
-predicate to $\textbf{cons}_m \subseteq
+Rules $(1)$ and $(*)$ extend the \consAName{}
+predicate to $\consAName{}_m \subseteq
 \textsf{Mult}(\mathcal{P}(\domainVal{}^n))
@@
 \trule{up}\dfrac{\forall u \in M, \, \exists v \in N .  \, u \preceq v}
-      {\cons{\textsf{upper}(N)}{M}}
-\qquad
+      {\cons{M}{\textsf{upper}(N)}}
+\;\;
 \trule{low}\dfrac{\forall v \in Nm \, \exists u \in M . \, u \preceq v}
-      {\cons{\textsf{lower}(N)}{M}} \\[1.5em]
-%%
-\trule{exact}\dfrac{\cons{\textsf{lower}(N)}{M} \qquad \cons{\textsf{upper}(N)}{M}}
-  {\cons{\textsf{exact}(N)}{M}} \\[1.5em]
-%%
-\trule{both}\dfrac{\cons{\textsf{lower}(M_1)}{M} \qquad \cons{\textsf{upper}(M_2)}{M}}
-  {\cons{\textsf{both}(M_1, M_2)}{M}}\\[1.5em]
+      {\cons{M}{\textsf{lower}(N)}} \\[1.5em]
+%%
+\trule{exact}\dfrac{\cons{M}{\textsf{lower}(N)} \qquad \cons{M}{\textsf{upper}(N)}}
+  {\cons{M}{\textsf{exact}(N)}} \\[1.5em]
+%%
+\trule{both}\dfrac{\cons{M}{\textsf{lower}(M_1)} \qquad \cons{M}{\textsf{upper}(M_2)}}
+  {\cons{M}{\textsf{both}(M_1, M_2)}}\\[1.5em]
 %%
 \trule{$1$}
-\dfrac{\cons{N}{M}}{\consA{\textsf{once}(N)}{\textsf{once}(M)}}
+\dfrac{\cons{M}{N}}{\consA{\textsf{once}(M)}{\textsf{once}(N)}}
 \;\;
 \trule{$*$}
-\dfrac{\cons{N}{M}}{\consA{\textsf{mult}(N)}{\textsf{mult}(M)}}
+\dfrac{\cons{M}{N}}{\consA{\textsf{mult}(M)}{\textsf{mult}(N)}}
   \\[1.5em]
 \trule{$1 < *$}
-\dfrac{\cons{N}{M}}{\consA{\textsf{mult}(N)}{\textsf{once}(M)}}
+\dfrac{\cons{M}{N}}{\consA{\textsf{once}(M)}{\textsf{mult}(N)}}
 \end{array}
@@
 %
-\item Consistency $cons_2(\textsf{once}(\textsf{exact}(N_s)),
-  \textsf{once}(M_c))$ holds when
-  $\textit{cons}_2(\textsf{mult}(\textsf{exact}(N_s)),\textsf{mult}(M_c))$ (Fig.~\ref{fig:consistency})
+\item Model $\textsf{once}(\textsf{exact}(N_s))$ is consistent with
+  $\textsf{once}(M_c)$ when
+  $\textsf{mult}(\textsf{exact}(N_s))$ is consistent with
+  $\textsf{mult}(M_c)$ (Fig.~\ref{fig:consistency})
   which holds when:
   \begin{align*}
-     & \textit{cons}_2(\textsf{mult}(\textsf{up}(N_s)),\textsf{mult}(M_c)) \\
-   \wedge \; & \textit{cons}_2(\textsf{mult}(\textsf{low}(N_s),\textsf{mult}(M_c))
+    & \consSub{2}{\textsf{mult}(\textsf{up}(N_s))}{\textsf{mult}(M_c)} \\
+    \wedge \; & \consSub{2}{\textsf{mult}(\textsf{low}(N_s)}{\textsf{mult}(M_c)}
   \end{align*}
@@
 %
-then consistency holds as it no longer requires the failing upper-bound case
-$\mathit{cons}(\textsf{mult}(\textsf{up}(N_s)),\textsf{mult}(M_c))$.
+then consistency holds as it no longer requires the failing upper-bound case:
+$\consSub{2}{\textsf{mult}(\textsf{up}(N_s))}{\textsf{mult}(M_c)}$.
  This makes sense semantically since the region ${\textcolor{darkgreen}{\texttt{fivepoint}}}$

### 2016-11-15 paper.tex
tweak
some grammar edit; and remaining fixes from mistral's comments

@@
 
-We provide a verification tool which can check the correctness of stencil code
-against its specification, infer such specifications from bare source, and
-synthese inferred specifications as comments and insert them into the source
-code. The main features and benefits of our approach are as follows:
+We provide a verification tool which can check the spatial correctness
+of stencil computations against a specification. Furthermore, the tool
+can automatically infer specifications and insert these as comments
+into the source code. The main features and benefits of our approach
+are as follows:
 %
@@
 \item The primary use case is that a programmer writes a specification first,
-  then the corresponding source code and the tool checks conformance between
-  the two. This specify-and-check approach reduces the testing effort and
+  then writes the corresponding source code. Our tool then checks conformance between
+  the two. This specify-and-check approach reduces testing effort and
   future-proofs against bugs introduced during refactoring and maintanence.
@@
 % compare this aspect of CamFort to ad hoc testing. Apples and oranges.
-\item A specification concisely captures the access pattern and turns it into
-  documentation. Inference algorithm efficiently produces specifications with
-  no programmer intervention, while synthesis mechanism inserts the
-  specifications at appropriate places in the source code automatically.
+\item A specification concisely captures the access pattern of a
+  stencil and turns it into documentation. The inference procedure
+  efficiently produces specifications with no programmer intervention,
+  automatically inserting specifications at appropriate places in the
+  source code.
 %
@@
   \emph{consistent}. Informally, the region defined by a specification
-  is an $n$-dimensional shape crossing the origin with respect to the induction
+  is an $n$-dimensional rectilinear shape in a discrete space crossing the origin with respect to the induction
   variable. Consistency is then a two way requirement.
@@
   form $i \pm c$, Def.~\ref{def:neighbour},
-  p.~\pageref{def:neighbour}) and $\infty$ represents any constant
-  index. Throughout $u, v, w$ range over
+  p.~\pageref{def:neighbour}) and $\infty$ is a wild-card
+  representing both constant indices and un-specified indices. Throughout $u, v, w$ range over
   index schemes and $u_i$ denotes the $i^{th}$ element of $u$. We
@@
 schemes, namely that $\vtwoh{0}{1} \preceq_2 \vtwoh{\infty}{1}$.
-At the level of syntax, this translates to saying
-$(i, j+1)$ can be instantiated to $(1, j+1)$.
 Note the last example is undefined as its first expression is a
@@
 
-Rules $(1)$ and $(*)$ extend the \consAName{}
-predicate to $\consAName{}_m \subseteq
+Rules $(1)$ and $(*)$ extend the \consName{}
+predicate to $\consAName{} \subseteq
 \textsf{Mult}(\mathcal{P}(\domainVal{}^n))

### 2016-11-15 paper.tex
evaluation stubs

@@
 
-\title{Verification of Stencil Computations with Spatial
-  Specifications}
+\title{Specification and Verification of Stencil Computation Shape}
 %\titlebanner{DRAFT -- Do not distribute}
@@
 
+\subsection{Core hypotheses}
+
+\paragraph{Hypothesis 1: Stencil computations
+(as in Definition~\ref{def:stencil}) are a common programming pattern
+in numerical code}
+
+% Number of stencils (tickAssign) / number of loops
+% Percentage of a code base that is taken up by stencils
+
+\paragraph{Hypothesis 2: Most stencil computations
+have a regular pattern which can be captured with a simple
+language of contiguous regions}
+
+% Number of stencils we give a spec to (where a spec declaration to n
+% variables is counted n times) as percentage of number of stencils
+
+%
+\subsection{Analysing the design of our specifications}
+
+% 1. How may are just reflexive and how many some reflexive
+
+% 2. Break down of "interesting specifications" (contain more than
+% just reflexive)
+%   - Single-action, histogram on depths
+%   - Single-action + irreflexives, histogram on depths
+%   - Bounded
+%   - Multi-action combined only by *
+%   - Multi-action combined also by +, histogram of tree depths
+
+% Present not just percentages but concrete numbers
+
 \bibliography{references}

### 2016-11-15 paper.tex
tweak to evaluation stub
various small fixes and tweaks

@@
     \parbox{5.2cm}{$(u, v) \in M \times N $ \\[0.1em]
+                  $\;\; \wedge \;\, i \in \{1, \ldots, 2^n\} \, \wedge
+                  \, x = (u \bowtie v)_i$ \\[0.1em]
                   $\;\; \wedge \;\, j \in (\effdims{M}{n} \cup
                    \effdims{N}{n})$\\[0.1em]
-                  $\;\; \wedge \;\, i \in \{1, \ldots, 2^n\} \, \wedge
-                  \, x = (u \bowtie v)_i$ \\[0.1em]
                    $\;\; \wedge \;\, x_j \neq \infty$
@@
 %
-(\textsc{irrefl}) \;\; & (R \, \term{*} \, S^\irreflS) \, \term{+} \,
+(\textsc{npoint}) \;\; & (R \, \term{*} \, S^\irreflS) \, \term{+} \,
                          (R^\irreflS \, \term{*} \, S) \;
@@
 
-The equation (\textsc{irrefl}) provides an annihilation of
+The equation (\textsc{point}) provides an annihilation of
 \irrefl{} using a further syntactic sugar.
@@
 %%
-The left-hand side of (\textsc{irrefl}) has a sum of
+The left-hand side of (\textsc{point}) has a sum of
 products of two regions $R$ and $S$,
@@
 with respect to \term{+}, \term{*}, \term{atLeast}, \term{atMost},
-and \term{readOnce}. The (\trule{\textsc{eq}}) rule connects region
-equations with inequations. The (\trule{\textsc{over}}) rule explains
+and \term{readOnce}. The \trule{\textsc{eq}} rule connects region
+equations with inequations. The \trule{\textsc{over}} rule explains
 the notion of spatial over-approximation via overlapping regions.
-The (\trule{\textsc{rep}}) allows the modifier that
+The \trule{\textsc{rep}} allows the modifier that
 enforces that every index is read just once to be dropped
-as an approximation; (\trule{\textsc{bound}}) gives
+as an approximation; \trule{\textsc{bound}} gives
 a connection between the \term{atLeast} and \term{atMost}
@@
   S}\trule{\texttt{M}} \\[1.5em]
-\dfrac{R <:^r S}{R <: \term{atMost} \, S}\trule{\textsc{shrink}} \quad
-\dfrac{R <:^r S}{S <: \term{atLeast} \, R}\trule{\textsc{grow}} \quad
+\dfrac{R <:^r S}{R <: \term{atMost} \, S}\trule{\textsc{grow}} \quad
+\dfrac{R <:^r S}{S <: \term{atLeast} \, R}\trule{\textsc{shrink}} \quad
  \\[1.5em]
@@
 
+% Corpus details
+% specfem3d ``simulates acoustic (fluid), elastic (solid), coupled
+% acoustic/elastic, poroelastic or seismic wave propagation in any
+% type of conforming mesh of hexahedra (structured or not).''
+% https://github.com/geodynamics/specfem3d
+
+% arpack-ng - Fortran 77 routines designed to solve large scaled
+% eigenvalue problems - https://github.com/opencollab/arpack-ng
+
 \subsection{Core hypotheses}
@@
 %   - Single-action + irreflexives, histogram on depths
-%   - Bounded
+%   - Bounded specs (how many lower, upper, or both)
 %   - Multi-action combined only by *

### 2016-11-15 paper.tex

@@
 
-\section{Introduction}
+\section{Introduction}\label{sec:intro}
 
@@
 
+\section{Related work}\label{sec:related-work}
+
+\citet{kamil2016verified} propose verified lifting to extract a mathematical
+representation of low-level, and potentially optimised, stencils in Fortran
+code. Problem domain wise their work is the closest to ours. However, since
+their aim is to use the extracted predicate to pipe to the high-performance
+compiler \textsf{Halide}~\citep{ragan2013halide}, they need to capture the full
+meaning of the stencil computation.
+
+This led them to develop \textsf{STNG}, a loop invariant and postcondition
+finder through syntax-guided synthesis. Using this approach they restrict the
+search space of postconditions to predicates of a certain form, one that is
+compatible with \textsf{Halide}. They then look at concrete loop iterations and
+form an hypothesis about the invariant, which is later confirmed or put back
+into the system using a SMT solver.
+
+CamFort differs from this work significantly here. The aim of CamFort is not a
+full representation of the stencil, but enough information to capture the access
+pattern. Hence, it does not require SMT solving or search of loop invariants. It
+suffices to do comparatively simple data-flow analysis.
+
+Due to the sheer amount of analysis involved, Kamil et al. reports that some
+degenerate kernels take up to 17 hours to analyse and others require programmer
+intervention for correct invariants to be inferred. Since CamFort mostly does
+syntactic analysis, inference procedure always terminates quickly in all real
+world examples and never requires programmer intervention. Downside of our
+approach is that if optimisation causes stencil computation to be heavily
+obfuscated, \textsf{STNG} would capture the access behaviour better.
+
+More fundamentally, unlike CamFort, Kamil et al. does not provide a syntactic
+representation of their postcondition, so there is no way of future-proofing the
+code for future edits. Even if they were to provide a syntactic representation,
+for complex stencils such as Navier-Stokes from \S~\ref{sec:intro}, it would be
+as verbose as the code itself making it difficult for programmer to understand
+the overall shape of the indexing behaviour.
+
+\section{Conclusion}\label{sec:conclusion}
+
 \bibliography{references}

### 2016-11-15 paper.tex

--- 2016-11-15 paper.tex
@@
 % Present not just percentages but concrete numbers
+
+\section{Related work}\label{sec:related-work}
+
+\citet{kamil2016verified} propose verified lifting to extract a mathematical
+representation of low-level, and potentially optimised, stencils in Fortran
+code. Problem domain wise their work is the closest to ours. However, since
+their aim is to use the extracted predicate to pipe to the high-performance
+compiler \textsf{Halide}~\citep{ragan2013halide}, they need to capture the full
+meaning of the stencil computation.
+
+This led them to develop \textsf{STNG}, a loop invariant and postcondition
+finder through syntax-guided synthesis. Using this approach they restrict the
+search space of postconditions to predicates of a certain form, one that is
+compatible with \textsf{Halide}. They then look at concrete loop iterations and
+form an hypothesis about the invariant, which is later confirmed or put back
+into the system using a SMT solver.
+
+CamFort differs from this work significantly here. The aim of CamFort is not a
+full representation of the stencil, but enough information to capture the access
+pattern. Hence, it does not require SMT solving or search of loop invariants. It
+suffices to do comparatively simple data-flow analysis.
+
+Due to the sheer amount of analysis involved, Kamil et al. reports that some
+degenerate kernels take up to 17 hours to analyse and others require programmer
+intervention for correct invariants to be inferred. Since CamFort mostly does
+syntactic analysis, inference procedure always terminates quickly in all real
+world examples and never requires programmer intervention. Downside of our
+approach is that if optimisation causes stencil computation to be heavily
+obfuscated, \textsf{STNG} would capture the access behaviour better.
+
+More fundamentally, unlike CamFort, Kamil et al. does not provide a syntactic
+representation of their postcondition, so there is no way of future-proofing the
+code for future edits. Even if they were to provide a syntactic representation,
+for complex stencils such as Navier-Stokes from \S~\ref{sec:intro}, it would be
+as verbose as the code itself making it difficult for programmer to understand
+the overall shape of the indexing behaviour.
+
+\section{Conclusion}\label{sec:conclusion}
 

### 2016-11-15 paper.tex
abstract

@@
 \bibliographystyle{abbrvnat}
+
+\begin{abstract}
+Stencil computations are a common programming pattern in numerical
+  code and typically form the core of graphics and scientific
+  computing applications. However, they are often complex and
+  unwieldy, involving substantial fine-grained manipulation of array
+  indices across multiple arrays and dimensions. This requires careful
+  testing, which is almost always done manually and then discarded
+  during development. In this paper, we propose an alternate approach:
+  a specification language and automated tool for verifying the
+  spatial behaviour of stencils. The specification language is based
+  on the hypothesis that stencil computations read from arrays with a
+  regular, fixed pattern which can be captured abstractly by a simple
+  set of combinators in our specification language. These
+  specifications can be embedded as annotations in source code,
+  against which code can be checked for conformance. For legacy code,
+  spatial specifications can be inferred, providing documentation and
+  aiding future software maintenance. We evaluate our language against
+  a corpus of numerical Fortran code, and show that the vast majority
+  of stencil computations indeed have a simple, regular, static shape
+  validating the design of our language.  We introduce the
+  language, checking, and inference procedures, and its evaluation.
+\end{abstract}
 

### 2016-11-15 paper.tex
Typos

--- 2016-11-15 paper.tex
@@
 
-It should be noted that our verification tool does not deal in the
+It should be noted that our verification tool does not target the
 class of bugs that can be detected automatically (\emph{push-button
@@
 \item We give a detailed verification case study of two
-  particular stencil computation commonly used in scientifc computing,
+  particular stencil computations commonly used in scientifc computing,
   simulating programming errors. Our approach is able to detect all

### 2016-11-15 paper.tex
Typos

@@
 %%
-This pattern is common in image convolution applicaitons (for example
+This pattern is common in image convolution applications (for example
 for edge detection). The specification ranges over the
@@
 and so ranges over the intermediate assignments to \term{x},
-\term{y}, and \term{z}. Each index in the code is consitent
+\term{y}, and \term{z}. Each index in the code is consistent
 with both specifications simultaneously, \eg{}, \texttt{a(i-1, j+1)}

### 2016-11-15 paper.tex
new title

--- 2016-11-15 paper.tex
@@
 
-\title{Specification and Verification of Stencil Computation Shape}
+\title{Specifying and Verifying the Spatial Properties of Stencil Computations}
 %\titlebanner{DRAFT -- Do not distribute}
@@
 \maketitle
-
-\keywords{program verification, specification, stencils}
 
@@
 
+\keywords{program verification, specification, stencils}
+
 \section{Introduction}\label{sec:intro}

### 2016-11-15 paper.tex
10pt
erroneous comma removal

@@
-\documentclass[9pt,preprint]{sigplanconf}
+\documentclass[10pt,preprint]{sigplanconf}
 
@@
   validating the design of our language.  We introduce the
-  language, checking, and inference procedures, and its evaluation.
+  language, checking and inference procedures, and its evaluation.
 \end{abstract}
@@
 representation of low-level, and potentially optimised, stencils in Fortran
-code. Problem domain wise their work is the closest to ours. However, since
-their aim is to use the extracted predicate to pipe to the high-performance
-compiler \textsf{Halide}~\citep{ragan2013halide}, they need to capture the full
-meaning of the stencil computation.
-
-This led them to develop \textsf{STNG}, a loop invariant and postcondition
-finder through syntax-guided synthesis. Using this approach they restrict the
-search space of postconditions to predicates of a certain form, one that is
-compatible with \textsf{Halide}. They then look at concrete loop iterations and
-form an hypothesis about the invariant, which is later confirmed or put back
-into the system using a SMT solver.
-
-CamFort differs from this work significantly here. The aim of CamFort is not a
+code. Their aim is to use the extracted predicate to pipe to the \textsf{Halide}~\citep{ragan2013halide} high-performance
+compiler and so they need to capture the full meaning of the stencil computation.
+
+% ACR-I think we could skip this in the interests of space
+%This led them to develop \textsf{STNG}, a loop invariant and postcondition
+%finder through syntax-guided synthesis. Using this approach they restrict the
+%search space of postconditions to predicates of a certain form, one that is
+%compatible with \textsf{Halide}. They then look at concrete loop iterations and
+%form an hypothesis about the invariant, which is later confirmed or put back
+%into the system using a SMT solver.
+
+CamFort differs from this significantly. The aim of CamFort is not a
 full representation of the stencil, but enough information to capture the access
-pattern. Hence, it does not require SMT solving or search of loop invariants. It
+pattern. Hence, we do not require SMT solving or search of loop invariants. It
 suffices to do comparatively simple data-flow analysis.
 
-Due to the sheer amount of analysis involved, Kamil et al. reports that some
+Due to the sheer amount of analysis involved, Kamil et al. report that some
 degenerate kernels take up to 17 hours to analyse and others require programmer
 intervention for correct invariants to be inferred. Since CamFort mostly does
-syntactic analysis, inference procedure always terminates quickly in all real
-world examples and never requires programmer intervention. Downside of our
-approach is that if optimisation causes stencil computation to be heavily
+syntactic analysis, the inference procedure terminates quickly and never requires programmer intervention. The downside of our
+approach is that if optimisation causes the stencil computation to be heavily
 obfuscated, \textsf{STNG} would capture the access behaviour better.
 
-More fundamentally, unlike CamFort, Kamil et al. does not provide a syntactic
+Most fundamentally, unlike CamFort, Kamil et al. does not provide a syntactic
 representation of their postcondition, so there is no way of future-proofing the

### 2016-11-15 paper.tex
Remove Doxygen FORD integration feature (space constraints)
lightweight subsumes abstract

@@
 %
-\item Specifications are integrated into the code as comments for easy
-  integration with existing tools, as well as documentation systems (\eg{}, we
-  support Doxygen and FORD syntax).
+%\item Specifications are integrated into the code as comments for easy
+%  integration with existing tools, as well as documentation systems (\eg{}, we
+%  support Doxygen and FORD syntax).
 %
@@
   that could be made whilst programming complex indexing code could also be
-  made when writing its specification. Our specifications are more abstract and
-  lightweight with the aim of aiding adoption by scientists and preventing
-  indexing errors.
+  made when writing its specification. Our specifications are more lightweight
+  with the aim of aiding adoption by scientists and preventing indexing errors.
 \end{itemize}

### 2016-11-15 paper.tex
Make our weakness (no push-button verif.) less verbose and more clear

@@
 
-It should be noted that our verification tool does not target the
+Our verification tool does not target the
 class of bugs that can be detected automatically (\emph{push-button
   verification}). Instead, stencil computation bugs must be
-identified relative to a specification of the intended behaviour.
+identified relative to a specification of the intended access pattern.
 

### 2016-11-15 paper.tex
Capitalise contributions

@@
 \begin{itemize}
-\item we introduce our specification language for stencils
+\item We introduce our specification language for stencils
   that captures many common forms of data access pattern
@@
 
-\item we detail checking and inference algorithms for our
+\item We detail checking and inference algorithms for our
   specifications (\S~\ref{sec:analysis}), derived from a
@@
 
-\item we report on a quantitative study of stencil computations on a
+\item We report on a quantitative study of stencil computations on a
   corpus of numerical Fortran programs (ranging from small to large)

### 2016-11-15 paper.tex
Make contributions less verbose

@@
 \begin{itemize}
-\item We introduce our specification language for stencils
-  that captures many common forms of data access pattern
+\item We introduce a specification language for stencils
+  that captures common forms of data access patterns
   (\S~\ref{sec:lang});
@@
 \item We report on a quantitative study of stencil computations on a
-  corpus of numerical Fortran programs (ranging from small to large)
+  corpus of numerical Fortran program
   (\S~\ref{sec:evaluation}), totalling one million lines of code. Our tool
@@
 \item We give a detailed verification case study of two
-  particular stencil computations commonly used in scientifc computing,
-  simulating programming errors. Our approach is able to detect all
+  stencil computations commonly used in scientifc computing,
+  simulating programming errors. Our approach detects all
   possible indexing errors for these computations.

### 2016-11-15 paper.tex
Typos and compressed kamil a bit

--- 2016-11-15 paper.tex
@@
   validating the design of our language.  We introduce the
-  language, checking and inference procedures, and its evaluation.
+  language, checking, and inference procedures, and its evaluation.
 \end{abstract}
@@
 %
-%\item Specifications are integrated into the code as comments for easy
-%  integration with existing tools, as well as documentation systems (\eg{}, we
-%  support Doxygen and FORD syntax).
+\item Specifications are integrated into the code as comments for easy
+  integration with existing tools, as well as documentation systems (\eg{}, we
+  support Doxygen and FORD syntax).
 %
@@
   that could be made whilst programming complex indexing code could also be
-  made when writing its specification. Our specifications are more lightweight
-  with the aim of aiding adoption by scientists and preventing indexing errors.
+  made when writing its specification. Our specifications are more abstract and
+  lightweight with the aim of aiding adoption by scientists and preventing
+  indexing errors.
 \end{itemize}
 
-Our verification tool does not target the
+It should be noted that our verification tool does not target the
 class of bugs that can be detected automatically (\emph{push-button
   verification}). Instead, stencil computation bugs must be
-identified relative to a specification of the intended access pattern.
+identified relative to a specification of the intended behaviour.
 

### 2016-11-15 paper.tex
Change the title to 'Specifying and Verifying the Spatial Properties of Stencil Computations'
Simplify induction variable definition

--- 2016-11-15 paper.tex
@@
 
-\title{Specifying and Verifying the Spatial Properties of Stencil Computations}
+\title{Verifying Spatial Properties of Stencil Computation}
 %\titlebanner{DRAFT -- Do not distribute}
@@
 
-\begin{definition}[Base induction variable]
-  An integer variable is a \emph{base induction variable} (or simply
-\emph{induction variable} for brevity) if it is the
-  control variable of a ``for'' loop (or equivalent, \eg{},
-  \texttt{do} in Fortran for our implementation) incremented by $1$
-  per iteration. A variable is interpreted as a base induction
-variable only within the scope of the loop body. Induction varaibles
-are ranged over by $i, j, k$ throughout.
+\begin{definition}[Induction variable]
+  An integer variable is an \emph{induction variable} if it is the control
+  variable of a \texttt{for} loop incremented by $1$ per iteration. A variable
+  is interpreted as an induction variable only within the scope of the loop
+  body. Induction varaibles range over $i, j, k$ throughout.
 %TODO: consider mentioning derived, non-based induction variables

### 2016-11-15 paper.tex
space saving in figures

--- 2016-11-15 paper.tex
@@
 
-\title{Specifying and Verifying the Spatial Properties of Stencil Computations}
+\title{Verifying Spatial Properties of Stencil Computations}
 %\titlebanner{DRAFT -- Do not distribute}
@@
 
-\keywords{program verification, specification, stencils}
+%\keywords{program verification, specification, stencils}
 
@@
 %
-\begin{minted}[linenos=false,xleftmargin=0em,xrightmargin=1.1em,breakindent=1.9em]{fortran}
+\begin{minted}[linenos=false,xleftmargin=0em,xrightmargin=0.8em,breakindent=1.9em]{fortran}
 != stencil centered(depth=1,dim=1)*pointed(dim=2)     + centered(depth=1,dim=2)*pointed(dim=1) :: u
@@
 \label{def:stencil}
+
 \end{definition}
@@
 \noindent
+
+% specifying the shape of indexing expressions for
+%arrays $\bar{v}$ that are simultaenously permitted and expected within
+%the code to which the specification is given.
+
+%\marginpar{\dnote{Is it worth having this informal consistency here?}}
+\begin{definition}[Consistency, informal]
 Specifications are associated to a stencil kernel
@@
 $\bar{v}$ within that kernel.
-
-% specifying the shape of indexing expressions for
-%arrays $\bar{v}$ that are simultaenously permitted and expected within
-%the code to which the specification is given.
-
-\marginpar{\dnote{Is it worth having this informal consistency here?}}
-\begin{definition}[Consistency, informal]
   A stencil specification defines a \emph{region} associated to an
@@
 \begin{figure}[t]
-\vspace{-0.5em}
-\begin{align*}
-\def\arraystretch{1.2}
+\vspace{-0.9em}
+\begin{align*}
+\def\arraystretch{1}
 \setlength{\arraycolsep}{0.2em}
@@
 \multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{region} \, \term{+}
-  \, \nonterm{region} \; \mid \; \nonterm{region} \; \term{*} \; \nonterm{region}} \\[0.5em]
+  \, \nonterm{region} \; \mid \; \nonterm{region} \; \term{*} \; \nonterm{region}} \\[0.25em]
 %\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{rvar}}
@@
 %  \\[0.5em]
-\nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+\\[-0.5em]
+\nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+\\[-1em]
 \end{array}
@@
 by \emph{approximation} and \emph{multiplicity} information
-(in the \textit{spec} rule of Fig.~\ref{fig:syntax}).
+(in \textit{spec} in Fig.~\ref{fig:syntax}).
 

### 2016-11-15 paper.tex
small shortening at start of sec 2

@@
 
-We first set up various definitions used in this paper.
-\S~\ref{sec:syntax} then outlines the syntax of stencil specifications
-and gives an informal introduction. \S~\ref{sec:semantics} then
+We give various definitions used in the paper then 
+\S~\ref{sec:syntax} outlines the syntax of stencil specifications
+with an informal introduction. \S~\ref{sec:semantics} 
 gives a semantic model via a set-based interpretation which makes
@@
 
-\paragraph{Notation and convention}
+\paragraph{Notation}
 %\label{sec:notation}

### 2016-11-15 paper.tex
some of the evaluation results, rough rough draft

@@
 
+We have processed nearly one million lines of Fortran code using our
+CamFort tool and examined the output to try and determine how useful
+and applicable it may be. The body of Fortran code is collected from
+many computational science projects for which we believe our tool will
+be potentially helpful.
+
 % Corpus details
@@
 
+%    numStencilSpecs: 60525
+% 4338 files, linesTotal: 1357377, linesParsed: 959427
 % Number of stencils (tickAssign) / number of loops
 % Percentage of a code base that is taken up by stencils
+
+Out of 1.35 million lines of code that we fed into our program, and
+the 959,427 lines that we were able to parse, we identified 60,525
+cases that fit our criteria for stencil specification.
 
@@
 
+    % tickAssign: 97439
+    % tickAssignSuccess: 29402 % <- number of times tickAssign is accompanied by a successful stencil inference
+                               % counts only by +1 each time that tickAssign finds a stencil
+
+In the analysed body of code we found that we were able to deduce
+some form of stencil specification from approximately 30\% of the
+potential stencils examined.
+
 %
@@
 % 1. How may are just reflexive and how many some reflexive
+% justReflexive: 55504
+% someReflexive: 15823
+
+55,504 specifications involved only reflexive regions, of which 15,823
+only had specifications for some of the dimensions, while 39,681 were
+reflexive on all dimensions.
+
+% (/ 55504.0 60525) = 91.7%
+% (/ 15823.0 60525) = 26.1%
+% 91.7\% of the specifications involved only reflexive regions. 26.1\%
 
@@
 %   - Bounded specs (how many lower, upper, or both)
+
+Bounded specifications were fairly unusual, with only 157
+\texttt{atMost} and 36 \texttt{atLeast}, the latter of which were
+always also bounded on both sides.
+    % atLeast: 36
+    % atMost: 157
+    % boundedBoth: 36
+
 %   - Multi-action combined only by *

### 2016-11-15 paper.tex
Simplify array subscript definition
Remove spacing between definition 4 and 5

@@
   in expanded form as $(e_1, \ldots, e_n)$. An index $\bar{e}$
-  is called \emph{relative} if each $e \in \bar{e}$ involves an induction variable.
-
-  Relative indices contrast with absolute indices where each $e \in
-  \bar{e}$ is a constant integer expression relative to the containing
-  loop, \ie{}, $e$ is not defined in terms of an induction variable.
+  is called \emph{relative} if each $e \in \bar{e}$ involves an induction
+  variable.
+
+  Absolute index is a constant integer expression relative to the enclosing
+  loop.
 \end{definition}
@@
 \begin{definition}[Stencil computations]
+\label{def:stencil}
   Let $a$ be an array of dimensionality $n$ and $\overline{b}$ a
@@
   statements with dataflow to $e_r$ as the \emph{stencil kernel}.
-\label{def:stencil}
-
 \end{definition}
-
-\noindent
-
+%
 % specifying the shape of indexing expressions for
@@
 %the code to which the specification is given.
-
+%
 %\marginpar{\dnote{Is it worth having this informal consistency here?}}

### 2016-11-15 paper.tex
making figure 6 not hoverflow
single action
single-action-irreflexive
saving space in section 5example and making it consistent with new consistency definition

@@
 \begin{align*}
+\hspace{-0.5em}
 \begin{array}{c}
 %%
-\trule{up}\dfrac{\forall u \in M, \, \exists v \in N .  \, u \preceq v}
+\trule{up}\dfrac{\forall u \in M, \exists v \in N .  \, u \preceq v}
       {\cons{M}{\textsf{upper}(N)}}
-\;\;
-\trule{low}\dfrac{\forall v \in Nm \, \exists u \in M . \, u \preceq v}
+\,
+\trule{low}\dfrac{\forall v \in N, \exists u \in M . \, u \preceq v}
       {\cons{M}{\textsf{lower}(N)}} \\[1.5em]
@@
 \end{align*}
-\caption{Consistency between models of code and specification}
+\vspace{-0.5em}
+\caption{Consistency of models (code vs. specification)}
 \label{fig:consistency}
@@
 %
-We refer to the assignment on line three as $l3$ for brevity.
-
-Verification of the code against the spec is then given by
-$check(({\textcolor{darkgreen}{\texttt{readOnce, fivepoint}}}),l3,\texttt{b})$.
-We explain the steps.
+We refer to line three as $l3$ for brevity,
+which is checked via 
+$check(({\textcolor{darkgreen}{\texttt{readOnce, fivepoint}}}),l3,\texttt{b})$
+with steps:
 \begin{enumerate}
@@
           \vtwo{\infty}{0},
-        \vtwo{\infty}{1} \}  \\[0.4em]
+        \vtwo{\infty}{1} \}  \\
 \interp{{\textcolor{darkgreen}{\texttt{readOnce, fivepoint}}}}_2 =
@@
 %
-\item Obtained the code model via $\textit{modelise}$, giving:
-%
-\begin{align*}
-M_c = \{
+\item Obtain the code model via $\textit{modelise}$, giving:
+%
+\begin{align*}
+M_c = & \{
     \vtwo{0}{-1},
@@
   \}
-\\[0.4em]
-\textit{modelise}(l3, \texttt{b}) = \textsf{once} (M_c)
-\end{align*}
-%
-\item Model $\textsf{once}(\textsf{exact}(N_s))$ is consistent with
-  $\textsf{once}(M_c)$ when
-  $\textsf{mult}(\textsf{exact}(N_s))$ is consistent with
-  $\textsf{mult}(M_c)$ (Fig.~\ref{fig:consistency})
-  which holds when:
-  \begin{align*}
-    & \consSub{2}{\textsf{mult}(\textsf{up}(N_s))}{\textsf{mult}(M_c)} \\
-    \wedge \; & \consSub{2}{\textsf{mult}(\textsf{low}(N_s)}{\textsf{mult}(M_c)}
-  \end{align*}
-%
-\item This condition holds by the following sets of scheme
+\\
+\textit{modelise}(l3, \texttt{b}) = & \textsf{once} (M_c)
+\end{align*}
+%
+\item By the definition of consistency (Fig.~\ref{fig:consistency})
+ $\textsf{once}(\textsf{exact}(N_s))$ is consistent with
+  $\textsf{once}(M_c)$ when:
+  \begin{align*}
+    & \consSub{2}{\textsf{up}(N_s)}{M_c} \; 
+    \wedge \; \consSub{2}{\textsf{low}(N_s)}{M_c}
+  \end{align*}
+%
+\item This holds by the following sets of scheme
   generalisations: for the upper bound, $\forall u \in M_c,
@@
   \begin{align}
+    \hspace{-1em}
     \begin{array}{l}
+   \hspace{-1.5em}
      \vtwoh{0}{-1} \preceq_2 \vtwoh{0}{\infty} \quad
@@
      \vtwoh{0}{1} \preceq_2 \vtwoh{0}{\infty} \\[0.5em]
-     \vtwoh{-1}{0} \preceq_2 \vtwoh{\infty}{0} \quad
+        \hspace{-1.5em}\vtwoh{-1}{0} \preceq_2 \vtwoh{\infty}{0} \quad
      \vtwoh{1}{0} \preceq_2 \vtwoh{\infty}{0}
@@
     \begin{array}{l}
-     \vtwoh{-1}{0} \preceq_2 \vtwoh{-1}{\infty} \quad
+     \hspace{-1.3em}\vtwoh{-1}{0} \preceq_2 \vtwoh{-1}{\infty} \quad
      \vtwoh{0}{0} \preceq_2 \vtwoh{0}{\infty} \quad
      \vtwoh{1}{0} \preceq_2 \vtwoh{1}{\infty}  \\
-     \vtwoh{0}{-1}  \preceq_2 \vtwoh{\infty}{-1} \quad
+     \hspace{-1.3em}\vtwoh{0}{-1}  \preceq_2 \vtwoh{\infty}{-1} \quad
      \vtwoh{0}{0}  \preceq_2 \vtwoh{\infty}{0} \quad
@@
   \end{align}
-  Since both the consistency predicate is satisfied, the array
+  Since the consistency predicate is satisfied, the array
   accesses flowing to line three in the stencil computation are valid
@@
 \begin{example}
-We consider now an erroneous user modification to $l3$ with an
+Consider an erroneous user modification to $l3$ with an
 additional term \texttt{b(i+2,j+2)} such that $l3'$ is:

### 2016-11-15 paper.tex
5.3 space saving

@@
 
-We demonstrate the inference procedure over the
- program from Exmaple~\ref{exm:checking} (p.~\pageref{exm:checking})
- computing the mean of a five-point stencil:
-%%
-\begin{minted}{fortran}
-do i = 1, (n-1)
-  do j = 1, (m-1)
-    x = a(i-1, j)+a(i+1, j); y = a(i, j-1)+a(i, j+1)
-    b(i, j) = (a(i, j) + x + y) / 5.0
-  end do
-end do
-\end{minted}
-%
-\subsubsection{Step 1: Standard analysis and data access analysis}
+We illustrate each step of inference using the
+ five-point stencil program from Example~\ref{exm:checking} (p.~\pageref{exm:checking}).
+
+\subsubsection{Step 1: Data access analysis}
 \label{sec:inf-step1}
 
-The first step of the inference is to perform the
-analysis of Section~\ref{subsec:analysis} to produce a
+The first step performs the
+analysis of \S~\ref{subsec:analysis} to produce a
 model of the code in terms of indexing schemes.

### 2016-11-15 paper.tex
Fix and simplify def. of Neighbourhood index

@@
 \label{def:neighbour}
-  For an array subscript $a(\bar{e})$ an index $\bar{e}$ is called a
-  \emph{neighbourhood index} if every $e \in \bar{e}$ is of the form
-  $e \equiv i$ or $e \equiv i + c$ or $e \equiv i - c$ where $c$ is an
-  integer constant. That is, a neighbourhood index is a relative index
-  where each component is a constant translation of a base induction
-  variable. (The relation $\equiv$ here identifies terms up-to
-  commutativity of $+$ and the inverse relationship of $+$ and $-$
-  \eg{}, $(-b) + i \equiv i - b$).
+  For an array subscript $a(\bar{e})$ an index $e \in \bar{e}$ is a
+  \emph{neighbourhood index} if $e$ is of the form $e \equiv i$, $e \equiv i +
+  c$, or $e \equiv i - c$, where $c$ is an integer constant. That is, a
+  neighbourhood index is a constant translation of an induction variable. (The
+  relation $\equiv$ identifies terms up-to commutativity of $+$ and the inverse
+  relationship of $+$ and $-$ \eg{}, $(-b) + i \equiv i - b$).
 \end{definition}
@@
 %   - Single-action, histogram on depths
+
+% singleAction: 4837
+% singleActionIrr: 1532
+A single-action specification consists of no more than one forward,
+backward, or centered region and can be combined with any number of
+reflexive regions. We identified 4,837 single-action specifications,
+of which 1,532 were single-action with an \texttt{irreflexive}
+modifier.
+
 %   - Single-action + irreflexives, histogram on depths

### 2016-11-15 paper.tex
multi-action
Simplify Stencil Computation definition

@@
 
-  Absolute index is a constant integer expression relative to the enclosing
-  loop.
+  \emph{Absolute index} is a constant integer expression relative to the
+  enclosing loop.
 \end{definition}
@@
   Let $a$ be an array of dimensionality $n$ and $\overline{b}$ a
-  collection of arrays of arbitrary (possible differing)
+  collection of arrays of arbitrary (possibly differing)
   dimensionalities.
@@
   most $n$ induction variables
-  over a subset of the index space of $a$. Each step of the interation
-  determs the elements of $a$ by an
+  over a subset of the index space of $a$. Each iteration
+  determines the elements of $a$ by an
   assignment $a(\bar{e}) = e_r$ where each $e \in
   \bar{e}$ is either a neighbourhood or constant index.
-  For all arrays $b \in \bar{b}$ and array subscripts $b(\bar{e}')$ in
-  $e_r$ and that flow to $e_r$, then each $e' \in \bar{e}'$ is also
-  either a neighbourhood or constant index.
+  For all $b \in \bar{b}$ and array subscripts $b(\bar{e}')$
+  that flow to $e_r$, each $e' \in \bar{e}'$ is
+  either a neighbourhood or absolute index.
 
@@
 %   - Single-action, histogram on depths
+%   - Single-action + irreflexives, histogram on depths
 
@@
 
-%   - Single-action + irreflexives, histogram on depths
+%   - Multi-action combined only by *
+%   - Multi-action combined also by +, histogram of tree depths
+
+    % multiAction: 156
+    % multiActionMulOnly: 146
+A multi-action specification consists of at least two uses of forward,
+backward, or centered regions, and can be combined with any number of
+reflexive regions. We identified 156 multi-action specifications out
+of which 146 had regions combined only with $\term{*}$.
+
 %   - Bounded specs (how many lower, upper, or both)
@@
     % boundedBoth: 36
-
-%   - Multi-action combined only by *
-%   - Multi-action combined also by +, histogram of tree depths
 

### 2016-11-15 paper.tex
Informal consistency description

@@
 %\marginpar{\dnote{Is it worth having this informal consistency here?}}
-\begin{definition}[Consistency, informal]
-Specifications are associated to a stencil kernel
-and are declared for one or more array variables
-$\bar{v}$ within that kernel.
-  A stencil specification defines a \emph{region} associated to an
-  array variable $a$ against which a stencil kernel should be
-  \emph{consistent}. Informally, the region defined by a specification
-  is an $n$-dimensional rectilinear shape in a discrete space crossing the origin with respect to the induction
-  variable. Consistency is then a two way requirement.
-  On the one hand, all array subscripts on $a$ contributing to the kernel
-  should be consistent with the region; on the other, all parts of the
-  region should have an index corresponding to that part.
-\end{definition}
+
+Now we informally describe what it means for a stencil kernel to be consistent
+with a specification. This will will be formalised in \S~\ref{sec:semantics} and
+algorithmically described in  and \S~\ref{sec:analysis}.
+Specifications are associated to a stencil kernel and are declared for one or
+more array variables $\bar{v}$ within that kernel.  A stencil specification
+defines a \emph{region} associated to an array variable $a$ against which a
+stencil kernel should be \emph{consistent}. The region defined by a
+specification is an $n$-dimensional rectilinear shape in a discrete space
+crossing the origin with respect to the induction variable. Consistency is then
+a two way requirement. All array subscripts on $a$
+contributing to the kernel should be consistent with the region. Conversely,
+all parts of the region should have an index corresponding to that part.
 

### 2016-11-15 paper.tex
Simplify syntax entry

@@
 
-Figure~\ref{fig:syntax} gives the syntax of stencil specifications,
-which we introduce in stages below.  The top-level is given by the
-\nonterm{specification} production which splits into either a
-\nonterm{regionDec} (region declaration) or a \nonterm{specDec}
-(specification declaration).
+Figure~\ref{fig:syntax} gives the syntax of stencil specifications, which is
+detailed below. The entry point is the \nonterm{specification} production which
+splits into either a \emph{region declaration} or a \emph{specification
+declaration}.
 

### 2016-11-15 paper.tex
Simplify Region Constants description

@@
 \term{forward}, \term{backward}, or \term{centered}.  Each region
-constant has a dimension identifier $d$ given as a natural number
-greater than 0. Each constant except \term{pointed} has a depth
-parameter $n$ given as a natural number greater than 0; \term{pointed}
+constant has a dimension identifier $d$ given as a positive natural number.
+Each constant except \term{pointed} has a depth
+parameter $n$ given as a positive natural number; \term{pointed}
 regions implicitly have a depth of $0$.
@@
 A \term{forward} region of depth $n$ specifies a continguous
-region in dimension $d$ starting at the origin; this corresponds
+region in dimension $d$ starting at the origin. This corresponds
 to specifying neighbourhood indices in dimension $d$ ranging from $i$ to $i + n$
-inclusively for some induction variable $i$.  Similarly, a
+for some induction variable $i$. Similarly, a
 \term{backward} region of depth $n$ corresponds to contiguous indices
-in dimension $d$ from $i$ to $i - n$, and \term{centered}
+from $i$ to $i - n$ and \term{centered}
 of depth $n$ from $i - n$ to $i + n$. A \term{pointed}
-stencil specifies a neighbour index $i$. For example the
+stencil specifies a neighbour index $i$. For example, the
 following shows four specifications with four consistent stencil
@@
 The \term{forward}, \term{backward}, and \term{centered} regions may
-all have an additional attribute \term{nonpointed} which marks that
-the region does not include the origin.  For example, the following is a
+all have an additional attribute \term{nonpointed} which marks absence
+of the origin.  For example, the following is a
 \term{nonpointed} \term{backward} stencil
@@
 != stencil backward(depth=2, dim=1, nonpointed) :: a
-b(i) = a(i-1) + a(i-2)
+b(i) = a(i-1) + 10*a(i-2)
 \end{minted}
-%
-The \term{pointed} regions cannot have a \term{nonpointed} attribute.
 

### 2016-11-15 paper.tex
Simplify region combinations subsubsection

@@
 $r$ and $s$ there must be an index consistent with both.
-For example, the following code (defining a \emph{nine-point stencil})
+For example, the following \emph{nine-point stencil}
 has a specification given by the product of two \texttt{centered}
@@
 %%
-This pattern is common in image convolution applications (for example
-for edge detection). The specification ranges over the
+This pattern is common in image convolution applications.
+The specification ranges over the
 values that flow to the array subscript on the left-hand side,
@@
 This reflects the symmetry under rotation of the five-point stencil.
-
-As another example, the following
-implements and specifies the \emph{Roberts cross}
-edge-detection convolution~\cite{davis1975survey}:
-%%
-\begin{minted}[breakindent=5.5em]{fortran}
-do j=0, jmax-1
-  do i=0, imax-1
-    x = a(i,j)-a(i+1,j+1); y = a(i+1,j)-a(i,j+1)
-    != stencil forward(depth=1,dim=1) * forward(depth=1,dim=2) :: a
-    b(i,j) = sqrt(x*x+y*y)
-  end do
-end do
-\end{minted}
 

### 2016-11-15 paper.tex
shrunk section 4 considerably

--- 2016-11-15 paper.tex
@@
 
-  \emph{Absolute index} is a constant integer expression relative to the
-  enclosing loop.
+  Absolute index is a constant integer expression relative to the enclosing
+  loop.
 \end{definition}
@@
   Let $a$ be an array of dimensionality $n$ and $\overline{b}$ a
-  collection of arrays of arbitrary (possibly differing)
+  collection of arrays of arbitrary (possible differing)
   dimensionalities.
@@
   most $n$ induction variables
-  over a subset of the index space of $a$. Each iteration
-  determines the elements of $a$ by an
+  over a subset of the index space of $a$. Each step of the interation
+  determs the elements of $a$ by an
   assignment $a(\bar{e}) = e_r$ where each $e \in
   \bar{e}$ is either a neighbourhood or constant index.
-  For all $b \in \bar{b}$ and array subscripts $b(\bar{e}')$
-  that flow to $e_r$, each $e' \in \bar{e}'$ is
-  either a neighbourhood or absolute index.
+  For all arrays $b \in \bar{b}$ and array subscripts $b(\bar{e}')$ in
+  $e_r$ and that flow to $e_r$, then each $e' \in \bar{e}'$ is also
+  either a neighbourhood or constant index.
 
@@
 %\marginpar{\dnote{Is it worth having this informal consistency here?}}
-
-Now we informally describe what it means for a stencil kernel to be consistent
-with a specification. This will will be formalised in \S~\ref{sec:semantics} and
-algorithmically described in  and \S~\ref{sec:analysis}.
-Specifications are associated to a stencil kernel and are declared for one or
-more array variables $\bar{v}$ within that kernel.  A stencil specification
-defines a \emph{region} associated to an array variable $a$ against which a
-stencil kernel should be \emph{consistent}. The region defined by a
-specification is an $n$-dimensional rectilinear shape in a discrete space
-crossing the origin with respect to the induction variable. Consistency is then
-a two way requirement. All array subscripts on $a$
-contributing to the kernel should be consistent with the region. Conversely,
-all parts of the region should have an index corresponding to that part.
+\begin{definition}[Consistency, informal]
+Specifications are associated to a stencil kernel
+and are declared for one or more array variables
+$\bar{v}$ within that kernel.
+  A stencil specification defines a \emph{region} associated to an
+  array variable $a$ against which a stencil kernel should be
+  \emph{consistent}. Informally, the region defined by a specification
+  is an $n$-dimensional rectilinear shape in a discrete space crossing the origin with respect to the induction
+  variable. Consistency is then a two way requirement.
+  On the one hand, all array subscripts on $a$ contributing to the kernel
+  should be consistent with the region; on the other, all parts of the
+  region should have an index corresponding to that part.
+\end{definition}
 
@@
 
-Figure~\ref{fig:syntax} gives the syntax of stencil specifications, which is
-detailed below. The entry point is the \nonterm{specification} production which
-splits into either a \emph{region declaration} or a \emph{specification
-declaration}.
+Figure~\ref{fig:syntax} gives the syntax of stencil specifications,
+which we introduce in stages below.  The top-level is given by the
+\nonterm{specification} production which splits into either a
+\nonterm{regionDec} (region declaration) or a \nonterm{specDec}
+(specification declaration).
 
@@
 
-Specifications are subject to a set of equations which can be used to
-understand when specifications are equivalent. Furthermore,
-specifications are subject to a notion of approximation. We outline
-the equational theory and notion of approximation here.
-
-\begin{figure}
-\vspace{-0.3em}
-\begin{align*}
-\setlength{\arraycolsep}{0.05em}
-\begin{array}{c}
-%\framebox{$\textit{region} \equiv \textit{region}'$} \\[0.6em]
-\begin{array}{rll}
-(\textsc{F\, \texttt{+} \,F}) \;\; &
-\stenFwdSR{n \, \textsf{max} \, m}{d}{p \vee q} \\
- \equiv \; & \stenFwdSR{n}{d}{p} \; \texttt{+} \; \stenFwdSR{m}{d}{q} \\[0.6em]
-%
-%
-(\textsc{B\, \texttt{+} \,B}) \;\; &
-\stenBwdSR{n \, \textsf{max} \, m}{d}{p \vee q} \\
- \equiv \; & \stenBwdSR{n}{d}{p} \; \texttt{+} \; \stenBwdSR{m}{d}{q} \\[0.6em]
-%
-%
-(\textsc{C\, \texttt{+} \,C}) \;\; &
-\stenCenSR{n \, \textsf{max} \, m}{d}{p \vee q} \\
-\equiv \; & \stenCenSR{n}{d}{p} \; \texttt{+} \; \stenCenSR{m}{d}{q} \\[0.6em]
-%
-%
-(\textsc{C\, \texttt{+} \,F}) \;\; & \stenCenSR{n}{d}{p \vee q} \\
-\textit{$m \leq n$} \; \equiv \; & \stenCenSR{n}{d}{p} \; \texttt{+} \;
-                      \stenFwdSR{m}{d}{q} \\[0.6em]
-%
-%
-(\textsc{C\, \texttt{+} \,B}) \;\; &
-\stenCenSR{n}{d}{p \vee q} \\
-\textit{$m \leq n$} \; \equiv \; & \stenCenSR{n}{d}{p} \; \texttt{+} \;
-                      \stenBwdSR{m}{d}{q} \\[0.6em]
-%
-%
-(\textsc{B\, \texttt{+} \,F}) \;\; &
-\stenCenSR{n}{d}{p \vee q} \\
-\equiv \; & \stenFwdSR{n}{d}{p} \; \texttt{+} \; \stenBwdSR{n}{d}{q}
-  \\[0.6em]
-%
-%
-(\textsc{P\, \texttt{+} \,F}) \;\; &
-\stenFwdS{n}{d} \\
-\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenFwdSR{n}{d}{p} \\[0.6em]
-%
-%
-(\textsc{P\, \texttt{+} \,B}) \;\; &
-\stenBwdS{n}{d} \\
-\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenBwdSR{n}{d}{p} \\[0.6em]
-%
-%
-(\textsc{P\, \texttt{+} \,C}) \;\; &
-\stenCenS{n}{d} \\
-\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenCenSR{n}{d}{p} \\[0.6em]
-%
-%
-(\textsc{npoint}) \;\; & (R \, \term{*} \, S^\irreflS) \, \term{+} \,
-                         (R^\irreflS \, \term{*} \, S) \;
-\equiv \; R \, \term{*} \, S \\[0.4em]
-%
-(\textsc{$\ast$Dim}) \;\; & R^d \, \term{$\ast$} \, S^d \equiv R^d
-                             \, \term{+} \, S^d \\[0.4em]
-%
-(\texttt{+}\textsc{idem}) \;\; & S \; \texttt{+} \; S \; \equiv \; S \\[0.4em]
-%
-(\texttt{+}\textsc{comm}) \;\; & S \; \texttt{+} \; T \; \equiv \; T \;
-                       \texttt{+} \; S \\[0.4em]
-%
-(\texttt{+}\textsc{assoc}) \;\; & R \; \texttt{+} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
-                       \texttt{+} \; S) \; \texttt{+} \; T \\[0.4em]
-(\texttt{*}\textsc{comm}) \;\; & S \; \texttt{*} \; T \; \equiv \; T \;
-                       \texttt{*} \; S \\[0.4em]
-%
-(\texttt{*}\textsc{assoc}) \;\; & R \; \texttt{*} \; (S \; \texttt{*} \; T) \; \equiv \; (R \;
-                       \texttt{*} \; S) \; \texttt{*} \; T \\[0.4em]
-(\textsc{dist}) \;\; & R \; \texttt{*} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
+Specifications have a theory of equality $\equiv$ as well
+as approximation $<:$ which we define here.
+For brevity we use abbreviations $\term{fwd}$, $\term{bwd}$, and
+$\term{cen}$ for region constants here.
+
+The equality relation $\equiv$ is defined over \emph{region}
+syntax which gives the following axioms on regions:
+%
+\begin{itemize}
+\item $\term{+}$ is idempotent, commutative, and associative;
+\item $\term{*}$ is commutative, associative, and distributes over
+$\term{+}$, \ie{},  $R \; \texttt{*} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                        \texttt{*} \; S) \; \texttt{+} \; (R
-                       \; \texttt{*} \; T)
-\end{array}
-\end{array}
-\end{align*}
-\caption{Equations on specifications (\emph{regions})}
-\vspace{-0.5em}
-\label{fig:equations}
-\end{figure}
-
-Figure~\ref{fig:equations} lists the equational theory for
-specifications given by the binary relation $\equiv$
-on \emph{region} terms. The equations use abbreviations
-\term{point}, \term{fwd}, \term{bwd}, \term{cen}
-for \term{nonpointed}, \term{forward}, \term{backward},
-\term{centered} respectively. Furthermore, the abbreviations include syntactic sugar
-for quantifying over the absence or presence of the \irrefl{}
-attribute on region constants as a boolean. For example,
-\term{fwd} syntactic sugar is defined:
-%%
-\begin{align*}
-\hspace{-0.25em}
-& \stenFwdSR{n}{d}{r}
-\; := \; \\[-0.4em]
-& \qquad \begin{cases}
-\stenFwd{n}{d}{, \irrefl} & \;\;\textit{if} \;\, \neg \, r \\[-0.1em]
-\stenFwd{n}{d}{}  & \;\;\textit{if} \;\,  r
-\end{cases} \\
-\textit{or} \;\;
-& \stenFwdS{n}{d} := \stenFwd{n}{d}{}
-\end{align*}
-%
-and similarly for \term{bwd} and \term{cen}.
-
+                       \; \texttt{*} \; T)$. 
+Distributivity is 
+used later to give a normal form for specifications
+akin to Disjunctive Normal Form (\S~\ref{sec:inf-step4}).
+
+\item (\emph{overlapping}) Given two region constants $R_1$ and $R_2$ 
+on the same dimension, then some combinations of region constants
+can be coalesced into a single region constant $R_3$ if $R_1$ completely
+overlaps $R_2$. There are five possible instantiations of the
+following rule $R_1 + R_2 \equiv R_3$ which we denote: $\term{fwd + fwd} \equiv \term{fwd}$, $\term{bwd + bwd} \equiv \term{bwd}$, 
+$\term{cen + cen} \equiv \term{cen}$, $\term{cen + fwd} \equiv \term{cen}$ and
+$\term{cen + bwd} \equiv \term{cen}$
+%
+\newcommand{\stenGR}[4]{\small{\textit{#4}(\term{depth=}#1,
+  \term{dim=}#2, #3)}}
+\newcommand{\stenG}[3]{\small{\textit{#3}(\term{depth=}#1,
+  \term{dim=}#2)}}
+\begin{align*}
+\hspace{-1em} & \stenGR{n}{d}{p \vee q}{$R_3$} \\[-0.4em]
+\hspace{-1em} \equiv \; & \stenGR{n}{d}{p}{$R_1$} \; \texttt{+} \; \stenGR{m}{d}{q}{$R_2$}
+\end{align*}
+where $n \geq m$ and $p$ and $q$ range
+over booleans representing the absence (true) or presence (false)
+of \irrefl{} which are combined via disjunction.
+
+\item (\textit{overlapping pointed}) \term{pointed} regions coalesce with other regions of the
+same dimension:
+%
+\begin{align*}
+& \stenG{n}{d}{region} \\[-0.4em]
+\equiv \; & \stenGR{n}{d}{p}{region} \; \texttt{+} \;
+                          \stenReflS{d}
+\end{align*}
+%
+\item \textit{(centered)} a forward and backward region of the same
+  depth and dimension coalseces into a centered region:
+\begin{align*}
+\hspace{-1em} & {\small{\stenCen{n}{d}{, p \vee q}}} \\[-0.4em]
+\hspace{-1em} \equiv \; & {\small{\stenFwdSR{n}{d}{p} \; \texttt{+} \;
+                          \stenBwdSR{n}{d}{q}}}
+\end{align*}
+%%
+\item $(\textsc{point}) \;\; (R \, \term{*} \, S^\irreflS) \, \term{+} \,
+                         (R^\irreflS \, \term{*} \, S) \; \equiv \; R
+                         \, \term{*} \, S$
+meaning that for the sum of products of two regions $R$ and $S$,
+where in each component one of $R$ or $S$ has the non-pointed
+attribute denoted by $R^{\term{np}}$ and $S^{\term{op}}$,
+the non-pointed attributes cancel to give $R \, \term{*} \, S$.
+%
+\item $(\textsc{$\ast$Dim}) \;\; R^d \, \term{$\ast$} \, S^d \equiv R^d
+                             \, \term{+} \, S^d$
+%
+where $R^d$ and $S^d$ denote regions whose dimension is $d$.
+Thus a product of two regions on
+the same dimension is equal to their addition.
+\end{itemize}
+%
 Stencil specifications in \textit{spec} are considered
 equal when they have the same modifiers and $\equiv$ equates
-their regions, \ie{}:
-\begin{align*}
-(\textit{region} \equiv \textit{region'})
- \Rightarrow \quad\;\; & \texttt{stencil} \; [\textit{approx},] \; [\textit{mult},] \;
-\textit{region} \\[-0.4em]
-\quad \equiv \; & \texttt{stencil} \; [\textit{approx},] \;
-            [\textit{mult},] \; \textit{region}'
-\end{align*}
-%
-
-\paragraph{Overlapping regions with +}
-The first nine
-equations (with labels of the form $\ldots \texttt{+} \ldots$)
-explain the notion of \emph{region overlapping} with sum
-\term{+}. In the first six, $p$ and $q$ range
-over booleans denoting the absence (true) or presence (false)
-of \irrefl{}. These are combined via disjunction. For example,
-the rule $(\textsc{B} \, \term{+} \,
-\textsc{F})$ explains that the sum of \term{backward}
-and \term{forward} regions in the same dimension and the same
-depth is equivalent to a \term{centered} region of that depth.
-If the \term{forward} part is non-pointed ($p =
-\textit{false}$) but \term{backward} is pointed ($q =
-\textit{true}$) then their sum is equal to a \term{centered}
-region with no \irreflS{} modifier by the disjunction ($p \vee q =
-\textit{true}$), \ie{}:
-%%
-\begin{align*}
-& \stenCenS{n}{d}  \\[-0.2em]
-\equiv \;\; & \stenFwdSR{n}{d}{\irrefl} \; \texttt{+} \; \stenBwdS{n}{d}
-\end{align*}
-%%
-That is, the \term{fwd} region is non-pointed describing code which is missing
-a zero-offset neighbourhood index (\eg{}, missing $\texttt{a(i)}$), but the addition of
-the \term{bwd} has no non-pointed attribute and thus provides the
-missing zero offset index.
-
-The equation (\textsc{point}) provides an annihilation of
-\irrefl{} using a further syntactic sugar.
-We write $R^{\irreflS}$ to denote a region constant with an
-non-pointed attribute, \eg{},
-%%
-\begin{align*}
-\stenFwdS{n}{d}^{\irreflS}  := \,
-\stenFwdSR{n}{d}{\irrefl}
-\end{align*}
-%%
-The left-hand side of (\textsc{point}) has a sum of
-products of two regions $R$ and $S$,
-where in each component of the sum one of $R$
-or $S$ has the non-pointed attribute. These cancel to give
-just $R \, \term{*} \, S$.
-
-\paragraph{Interaction between \term{+} and \term{*}}
-
-The last seven equations (from (\textsc{\term{*}Dim}) inclusive), explain
-the algebraic behaviour of \term{+} and \term{*}.  Together, we see
-that \term{+} is an associative, commutative, idempotent binary
-operator that distributes with \term{*}, which is associative and
-commutative. This distribution of \term{*} over \term{+} is key as it
-is used internally to give a normal form for stencil specifications
-akin to Disjunctive Normal Form (DNF) (where \term{+} is
-disjunction and \term{*} is conjunction). We revisit this
-normalisation in \S~\ref{sec:inf-step4}.
-
-The equation (\textsc{$\ast$Dim}) uses syntactic sugar
-where $R^d$ and $S^d$ denote regions whose dimension is $d$.
-This equation shows that a product of two regions on
-the same dimension is equal to their addition.
+their regions.
 
@@
 
-\subsection{Approximation}
-
 Figure~\ref{fig:inequations} defines a notion of sub-specifications
@@
 Both relations are reflexive and transitive (we omit
-these equations) and congruent
-with respect to \term{+}, \term{*}, \term{atLeast}, \term{atMost},
-and \term{readOnce}. The \trule{\textsc{eq}} rule connects region
-equations with inequations. The \trule{\textsc{over}} rule explains
-the notion of spatial over-approximation via overlapping regions.
-The \trule{\textsc{rep}} allows the modifier that
-enforces that every index is read just once to be dropped
-as an approximation; \trule{\textsc{bound}} gives
-a connection between the \term{atLeast} and \term{atMost}
-approximation modifiers.
+these equations). The $<:^r$ relation is congruent
+with respect to \term{+} and \term{*} $<:$ is congruent
+with repect to \term{readOnce}, the equations of which we omit for brevity.
+
+The \trule{\textsc{eq}} rule connects region
+equations to approximation; \trule{\textsc{over}} explains
+the notion of spatial over-approximation via overlapping regions; 
+\trule{\textsc{rep}} allows the \term{readOnly} modifier that
+to be dropped. The \trule{\textsc{shrink}} 
+and \trule{\textsc{grow}} rules turn the spatial approxiamation on
+regions into approximation on specifications via the \term{atLeast} and \term{atMost}
+modifiers. Finally, \trule{\textsc{Gen$\ast$}} shows
+that specifications can be generalised by removing parts of a product
+$\ast$ which have disjoint dimensions to the rest of the region
+product (where $\mathsf{dims}$ denotes the set of dimensions specified
+in a region).
 
 \begin{theorem}[Soundness of approximation]
+A piece of code consistent with a specification $S$
+ is consistent with a generalisation $T$ of the specification.
+That is for all models $M$ of a piece of code:
 \begin{equation*}
-\forall S, T, M, n . \quad
-S <: T \; \wedge (\consA{M}{\interp{S}_n})
-\Rightarrow (\consA{M}{\interp{T}_n})
+S <: T \; \wedge (\consA{M}{\interp{S}_n}) \;
+\Rightarrow \; (\consA{M}{\interp{T}_n})
 \end{equation*}
+where $\consAName$ is the consistency predicate used in checking (\S~\ref{subsec:checking}).
 \end{theorem}
@@
 \begin{figure}[t]
+\vspace{-0.5em}
 \begin{align*}
 \hspace{-0.5em}
+{\small{
 \begin{array}{c}
@@
 \end{array}}\trule{\textsc{over}} \\[-0.2em]
-\hspace{-0.5em}
-\dfrac{R_1 <:^r S_1 \quad R_2 <:^r S_2}
-      {R_1 \, \texttt{+} \, R_2 <:^r S_1 \, \texttt{+} \, S_2}
-\trule{\textsc{cong}\texttt{+}}
-\dfrac{R_1 <:^r S_1 \quad R_2 <:^r S_2}
-      {R_1 \, \texttt{*} \, R_2 <:^r S_1 \, \texttt{*} \, S_2}
-\trule{\textsc{cong}\texttt{*}} \\[1.5em]
+%\hspace{-0.5em}
+%\dfrac{R_1 <:^r S_1 \quad R_2 <:^r S_2}
+%      {R_1 \, \texttt{+} \, R_2 <:^r S_1 \, \texttt{+} \, S_2}
+%\trule{\textsc{cong}\texttt{+}}
+%\dfrac{R_1 <:^r S_1 \quad R_2 <:^r S_2}
+%      {R_1 \, \texttt{*} \, R_2 <:^r S_1 \, \texttt{*} \, S_2}
+%\trule{\textsc{cong}\texttt{*}} \\[1.5em]
 %%
@@
   S}\trule{\texttt{M}} \\[1.5em]
-\dfrac{R <:^r S}{R <: \term{atMost} \, S}\trule{\textsc{grow}} \quad
-\dfrac{R <:^r S}{S <: \term{atLeast} \, R}\trule{\textsc{shrink}} \quad
+\dfrac{R <:^r S}{R <: \term{atMost} \, S}\trule{\textsc{grow}} \,
+\dfrac{R <:^r S}{S <: \term{atLeast} \, R}\trule{\textsc{shrink}} \,
+\dfrac{R \equiv S}{R <:^r S}\trule{\textsc{eq}}
  \\[1.5em]
@@
 \quad
-\dfrac{R \equiv S}{R <: S}\trule{\textsc{eq}}
-%%
-\\[1.5em]
-\dfrac{R <: S}{\term{readOnce} \, R <: \term{readOnce} \,
-  S}\trule{\textsc{cong}\texttt{R}}
-\;\;
+%%
+%\\[1.5em]
+%\dfrac{R <: S}{\term{readOnce} \, R <: \term{readOnce} \,
+%  S}\trule{\textsc{cong}\texttt{R}}
+%\;\;
+\;
 \dfrac{}{\term{readOnce} \, R <: R}\trule{\textsc{rep}}
@@
 %\dfrac{}{R <: \term{atLeast} R}\\[1.5em]
-\end{array}
+\end{array}}}
 \end{align*}
 \caption{Inequations $<:^r$ on \textit{regions} and
-$<:$ on \textit{specs} (omitting reflexivity
-and transitivity). }
+$<:$ on \textit{specs} (omitting reflexivity, transitivity, and congruences). }
 \label{fig:inequations}

### 2016-11-15 paper.tex
Simplify and chang example in 'Region declarations and variables''

@@
 Region specifications can be assigned to region variables via
-a region declaration, which can be used later in another
-specification. For example, the last specification
+region declarations. For example, the last specification
  example above can be restated as:
@@
 \begin{minted}{fortran}
-!= region :: r1 = forward(depth=1, dim=1)
-!= region :: r2 = forward(depth=1, dim=2)
-!= region :: robertsCross = r1*r2
-!= stencil robertsCross :: a
+!= region :: r1 = centered(depth=1, dim=1) * pointed(dim=2)
+!= region :: r2 = centered(depth=1, dim=2) * pointed(dim=1)
+!= region :: fivePoint = r1 + r2
+!= stencil fivePoint :: a
 \end{minted}
-This is especially useful for common stencil patterns, such as Roberts cross,
-as the region can be defined once and reused.
+This is especially useful for common stencil patterns, such as five point
+stencil, as the region can be defined once and reused many times later.
 %%

### 2016-11-15 paper.tex
small rearrangement

@@
 overlaps $R_2$. There are five possible instantiations of the
-following rule $R_1 + R_2 \equiv R_3$ which we denote: $\term{fwd + fwd} \equiv \term{fwd}$, $\term{bwd + bwd} \equiv \term{bwd}$, 
-$\term{cen + cen} \equiv \term{cen}$, $\term{cen + fwd} \equiv \term{cen}$ and
-$\term{cen + bwd} \equiv \term{cen}$
+following rule $R_1 + R_2 \equiv R_3$:
 %
@@
 of \irrefl{} which are combined via disjunction.
+We denote the possibilities as $\term{fwd + fwd} \equiv \term{fwd}$, $\term{bwd + bwd} \equiv \term{bwd}$, 
+$\term{cen + cen} \equiv \term{cen}$, $\term{cen + fwd} \equiv \term{cen}$ and
+$\term{cen + bwd} \equiv \term{cen}$.
 

### 2016-11-15 paper.tex
Fix example alignment in 'Modifiers'
further space iprovements in 4

@@
 of \irrefl{} which are combined via disjunction.
-We denote the possibilities as $\term{fwd + fwd} \equiv \term{fwd}$, $\term{bwd + bwd} \equiv \term{bwd}$, 
-$\term{cen + cen} \equiv \term{cen}$, $\term{cen + fwd} \equiv \term{cen}$ and
-$\term{cen + bwd} \equiv \term{cen}$.
+We denote the possibilities as $\term{fwd+fwd} \equiv \term{fwd}$, $\term{bwd+bwd} \equiv \term{bwd}$, 
+$\term{cen+cen} \equiv \term{cen}$, $\term{cen+fwd} \equiv \term{cen}$ and
+$\term{cen+bwd} \equiv \term{cen}$. %Commutative of $\term{+}$
+%provides symmetric variants.
 
@@
 
-\begin{theorem}[Soundness of approximation]
-A piece of code consistent with a specification $S$
- is consistent with a generalisation $T$ of the specification.
-That is for all models $M$ of a piece of code:
+\begin{theorem}[Approximation soundness]
+Code consistent with a specification $S$
+ is consistent with a more general specification $T$, 
+\ie{}, for all models $M$ of a piece of code:
 \begin{equation*}
@@
 \end{equation*}
-where $\consAName$ is the consistency predicate used in checking (\S~\ref{subsec:checking}).
+where $\consAName$ implements consistency in checking (\S~\ref{subsec:checking}).
 \end{theorem}

### 2016-11-15 paper.tex
Remove pagerefs and use Def. abbreviation consistently

@@
   (\ie{}, the constant integer $c$ in an indexing expression of the
-  form $i \pm c$, Def.~\ref{def:neighbour},
-  p.~\pageref{def:neighbour}) and $\infty$ is a wild-card
+  form $i \pm c$, Def.~\ref{def:neighbour}) and $\infty$ is a wild-card
   representing both constant indices and un-specified indices. Throughout $u, v, w$ range over
@@
 to a \emph{multiset} of indexing schemes for its subscripts, computed
-with \textsf{schematic} (Defn.~\ref{def:schematic},
-p.~\pageref{def:schematic}).
+with \textsf{schematic} (Def.~\ref{def:schematic}).
 
@@
 We illustrate each step of inference using the
- five-point stencil program from Example~\ref{exm:checking} (p.~\pageref{exm:checking}).
+ five-point stencil program from Example~\ref{exm:checking}.
 
@@
 \textbf{3)\,} Reduce each list pairwise by $\bullet$ operation
-(Defn.~\ref{def:span-coalesc}) to coalesce contiguous regions,
+(Def.~\ref{def:span-coalesc}) to coalesce contiguous regions,
 given by $\textsf{fold}\bullet$. For our example, this yields:
@@
 in the $\textsf{Approx}$ type (see
-Defn.~\ref{def:mult-and-approx},
-p.~\pageref{def:mult-and-approx}).  The algebra mirrors the shape
+Def.~\ref{def:mult-and-approx}).  The algebra mirrors the shape
 of the region combinators for sum and product,
@@
 Let $r_1, ..., r_n$ range over elements of the \textit{region}
-syntax (see Figure~\ref{fig:syntax}, p.~\pageref{fig:syntax}).
+syntax (see Figure~\ref{fig:syntax}).
 The set \specDNF{} is a subset of syntax trees \textit{spec},
@@
  the (\trule{\textsc{dist}}) rule of our equational theory
- (Fig.~\ref{fig:equations},p.~\pageref{fig:equations}). The operation is
+ (Fig.~\ref{fig:equations}). The operation is
 commutative and associative, and sound with respect to our

### 2016-11-15 paper.tex
roberts cross and section 2 shortening

--- 2016-11-15 paper.tex
@@
  breakafter=w),
-  xleftmargin=\parindent
+  xleftmargin=\parindent,
+  numbersep=1em
 }
@@
   An integer variable is an \emph{induction variable} if it is the control
-  variable of a \texttt{for} loop incremented by $1$ per iteration. A variable
+  variable of a ``\texttt{for}'' loop incremented by $1$ per iteration. A variable
   is interpreted as an induction variable only within the scope of the loop
@@
 Region specifications can be assigned to region variables via
-region declarations. For example, the last specification
- example above can be restated as:
+region declarations. For example, the shape of a 
+``\emph{Roberts cross}'' edge-detection convolution~\cite{davis1975survey}
+can be stated:
 %%
 \begin{minted}{fortran}
-!= region :: r1 = centered(depth=1, dim=1) * pointed(dim=2)
-!= region :: r2 = centered(depth=1, dim=2) * pointed(dim=1)
-!= region :: fivePoint = r1 + r2
-!= stencil fivePoint :: a
+!= region :: r1 = forward(depth=1, dim=1)
+!= region :: r2 = forward(depth=1, dim=2)
+!= region :: robertsCross = r1*r2
+!= stencil robertsCross :: a
 \end{minted}
-This is especially useful for common stencil patterns, such as five point
-stencil, as the region can be defined once and reused many times later.
+This is useful for common stencil patterns, such as the five-point
+pattern, as the region can be defined once and reused.
 %%

### 2016-11-15 paper.tex
Typos in index scheme definition.

--- 2016-11-15 paper.tex
@@
   form $i \pm c$, Def.~\ref{def:neighbour}) and $\infty$ is a wild-card
-  representing both constant indices and un-specified indices. Throughout $u, v, w$ range over
+  representing both constant indices and unspecified indices. Throughout, $u, v,
+  w$ range over
   index schemes and $u_i$ denotes the $i^{th}$ element of $u$. We

### 2016-11-16 paper.tex
Further fix array subscript def

--- 2016-11-15 paper.tex
@@
   comma-separated sequence of integer expressions denoted $\bar{e}$ or
-  in expanded form as $(e_1, \ldots, e_n)$. An index $\bar{e}$
-  is called \emph{relative} if each $e \in \bar{e}$ involves an induction
+  in expanded form as $(e_1, \ldots, e_n)$. An index $e_i$
+  is called \emph{relative} if the expression involves an induction
   variable.

### 2016-11-16 paper.tex
Simplify index scheme def, fix s_I def

--- 2016-11-15 paper.tex
@@
 
-Indexing schemes give an abstract
-representation of array subscripts in the source language.
+Indexing schemes give an abstracts array subscripts.
 Our model represents the meaning of stencil specifications
@@
 -c & e \equiv i - c \;\; \wedge \;\; i \in I \\
-\infty & \textit{$I \not\subseteq \textit{FV}(e)$} \\
+\infty & \textit{$I \cap \textit{FreeVariables}(e) = \emptyset $} \\
 \bot   & \textit{otherwise}
@@
 \end{align*}
-where $\textit{FV}(e)$ is the set of free variables in $e$. In the
+In the
 first two cases, expressions that are a constant offset of an

### 2016-11-16 paper.tex
Simplify s_I explanation

@@
 offset $c$ or $-c$ $\in \domainVal$. In the third case, if the
-expression is not a neighbourhood index, but neither is it a relative
-index (\ie{}, it is not defined in terms of an induction variable) then it
+expression is not defined in terms of an induction variable, it
 is mapped to $\infty \in \domainVal{}$. Otherwise, $\schemeOne(e)$

### 2016-11-16 paper.tex
Gloves are off. Remove courtesy notation explanation. Typo fix.
Clean up generalisation

@@
   form $i \pm c$, Def.~\ref{def:neighbour}) and $\infty$ is a wild-card
-  representing both constant indices and unspecified indices. Throughout, $u, v,
-  w$ range over
-  index schemes and $u_i$ denotes the $i^{th}$ element of $u$. We
+  representing both constant indices and unspecified indices.
+  For an index scheme $u$, $u_i$ denotes the $i^{th}$ element of $u$. We
   write $(\domainVal)^n$ for the set of all $n$-dimensional schemes.
@@
 
-Indexing schemes give an abstracts array subscripts.
+Indexing schemes abstract array subscripts.
 Our model represents the meaning of stencil specifications
@@
 %%
-That is, two schemes are related by the (reflexive and transitive) binary relation
-$\preceq_n$ over $(\mathbb{Z}_{\infty})^n$ if in each dimension
-the two schemes are equal or $w_i = \infty$
-in place of some $\mathbb{Z}$ in $v_i$ for $v \preceq_n w$.
-Thus, the element $\infty$ generalises any index in the same dimension.
+Two schemes are related by the (reflexive and transitive) binary relation
+$\preceq_n$ over $(\mathbb{Z}_{\infty})^n \times (\mathbb{Z}_{\infty})^n$.
+The element $\infty$ generalises any index in the same dimension.
 %

### 2016-11-16 paper.tex
Cleanup inxeding scheme conversion examples

@@
 \begin{example}
-Consider the following indexing schemes
-corresponding to array indexing terms
-$(i, j+1)$, $(1, j+1)$, $(3, j)$ and $(i*2, j)$:
+Consider the following array subscript to indexing schemes
+conversions:
 %
@@
 schemes, namely that $\vtwoh{0}{1} \preceq_2 \vtwoh{\infty}{1}$.
-Note the last example is undefined as its first expression is a
+The last example is undefined as its first expression is a
 non-neighbour relative index.

### 2016-11-16 paper.tex
section 5 polishing / shrinking

--- 2016-11-16 paper.tex
@@
 a simple, abstract specification language for the data access pattern
-of stencils. The shape of the stencil computation in the
-Laplace example is specifed in our language as:
+of stencils. The shape of the
+Laplace example stencil is specifed in our language as:
 %
@@
 %
-This describes that \texttt{a} is accessed with a symmetrical pattern
+This describes \texttt{a} is accessed with a symmetrical pattern
 (``centered'') to a depth of one in each direction in its first
@@
   An integer variable is an \emph{induction variable} if it is the control
-  variable of a \texttt{for} loop incremented by $1$ per iteration. A variable
+  variable of a ``\texttt{for}'' loop incremented by $1$ per iteration. A variable
   is interpreted as an induction variable only within the scope of the loop
-  body. Induction varaibles range over $i, j, k$ throughout.
+  body. Throuhgout, $i, j, k$ range over induction variables.
 %TODO: consider mentioning derived, non-based induction variables
@@
   variable.
-
-  \emph{Absolute index} is a constant integer expression relative to the
+  An \emph{absolute index} is a constant integer expression relative to the
   enclosing loop.
@@
 Now we informally describe what it means for a stencil kernel to be consistent
-with a specification. This will will be formalised in \S~\ref{sec:semantics} and
+with a specification. This will be formalised in \S~\ref{sec:semantics} and
 algorithmically described in  and \S~\ref{sec:analysis}.
@@
 contributing to the kernel should be consistent with the region. Conversely,
-all parts of the region should have an index corresponding to that part.
+all parts of the region should have an index corresponding it.
 
@@
 %%
-Not every dimension needs to be specified. In the above, the
-specifications on line $1$, $3$, and $7$ leaves some dimensions unspecified.
+Not every dimension needs to be specified, \eg{},
+specifications on lines $1$, $3$, and $7$ leave some dimensions unspecified.
 
@@
 
-The region operators \term{+} and \term{*} provide ways to combine
-regions disjunctively and conjuctively respectively.
+The region operators \term{+} and \term{*} respectively combine
+regions disjunctively and conjuctively.
 
@@
 Region specifications can be assigned to region variables via
-region declarations. For example, the last specification
- example above can be restated as:
+region declarations. For example, the shape of a
+``\emph{Roberts cross}'' edge-detection convolution~\cite{davis1975survey}
+can be stated:
 %%
 \begin{minted}{fortran}
-!= region :: r1 = centered(depth=1, dim=1) * pointed(dim=2)
-!= region :: r2 = centered(depth=1, dim=2) * pointed(dim=1)
-!= region :: fivePoint = r1 + r2
-!= stencil fivePoint :: a
+!= region :: r1 = forward(depth=1, dim=1)
+!= region :: r2 = forward(depth=1, dim=2)
+!= region :: robertsCross = r1*r2
+!= stencil robertsCross :: a
 \end{minted}
-This is especially useful for common stencil patterns, such as five point
-stencil, as the region can be defined once and reused many times later.
+This is useful for common stencil patterns, such as the five-point
+pattern, as the region can be defined once and reused.
 %%
@@
 The \texttt{readOnce} modifier enforces that no index appears more
-than once (that is, its multiplicity is one). For example, in all of
-the previous examples the \texttt{readOnce} modifier could be added,
-\eg{}
+than once (that is, its multiplicity is one). For example, all of
+the previous examples could have \texttt{readOnce} added:
 %
@@
 
-Indexing schemes give an abstracts array subscripts.
+Indexing schemes give an abstract
+representation of array subscripts in the source language.
 Our model represents the meaning of stencil specifications
@@
 -c & e \equiv i - c \;\; \wedge \;\; i \in I \\
-\infty & \textit{$I \cap \textit{FreeVariables}(e) = \emptyset $} \\
+\infty & \textit{$I \not\subseteq \textit{FV}(e)$} \\
 \bot   & \textit{otherwise}
@@
 \end{align*}
-In the
+where $\textit{FV}(e)$ is the set of free variables in $e$. In the
 first two cases, expressions that are a constant offset of an
@@
 offset $c$ or $-c$ $\in \domainVal$. In the third case, if the
-expression is not defined in terms of an induction variable, it
+expression is not a neighbourhood index, but neither is it a relative
+index (\ie{}, it is not defined in terms of an induction variable) then it
 is mapped to $\infty \in \domainVal{}$. Otherwise, $\schemeOne(e)$
@@
 as approximation $<:$ which we define here.
-For brevity we use abbreviations $\term{fwd}$, $\term{bwd}$, and
-$\term{cen}$ for region constants here.
+For brevity here we use abbreviations $\term{fwd}$, $\term{bwd}$, and
+$\term{cen}$ for region constants.
 
 The equality relation $\equiv$ is defined over \emph{region}
-syntax which gives the following axioms on regions:
+syntax which gives the following axioms:
 %
@@
 \end{align*}
-\caption{Inequations $<:^r$ on \textit{regions} and
+\vspace{-0.5em}
+\caption{Approximation $<:^r$ on \textit{regions} and
 $<:$ on \textit{specs} (omitting reflexivity, transitivity, and congruences). }
 \label{fig:inequations}
-\vspace{-0.8em}
+\vspace{-0.5em}
 \end{figure}
@@
 \noindent
-We outline here the procecures for checking the conformance
+We outline here the procedures for checking conformance
 of source code against specifications (\S~\ref{subsec:checking})
 and for infering specifications from code (\S~\ref{subsec:inference}).
-Both rely on the analysis of source code, converting array indexing
+Both rely on a program analysis that converts array indexing
 expressions into sets of index schemes. We outline this analysis
@@
 can be made arbitrarily more complex and wide-ranging independent
-of the checking and inference procedures.
-
-\subsection{Analysis of array-based code}
+of the checking and inference procedures. At the moment, the analysis
+is largely \emph{syntactic}, with only a small amount of
+semantic interpretation of the code.
+
+\subsection{Analysis of array accesses}
 \label{subsec:analysis}
@@
 
-\paragraph{Syntax helper functions}
-The function \rhsExp{} maps source statements to a
-set of expressions which occur in right-hand positions (\ie{}, not the
-target of an assignment). The function \var{} maps expressions to a
-set of the variables used in right-hand positions.
-
-\paragraph{Analysis pre-requisites}
 The analysis builds on standard program analyses:
@@
   closure of the flows graph);
-\textbf{(4)} type information per variable, where we use the predicate
-\arrayTy{} to classify variables of array type.
-
-\paragraph{Core analysis}
+\textbf{(4)} type information per variable.% where we use the predicate
+%\arrayTy{} to classify variables of array type.
+
+The core analysis, captured by the function $\textit{analyse}$ below, then proceeds as follows.
 For each assignment statement \textit{assg} within a loop whose left-hand side is an array
 subscript on neighbourhood indices, a finite map is computed mapping
-array variables to \emph{multisets} of indexing schemes.
-This is captured by the function $\textit{analyse}$:
+array variables to \emph{multisets} of indexing schemes:
 %
@@
 \multicolumn{2}{l}{\qquad \bigcup\{a' \mapsto \{
-\textsf{schematic}(\bar{e})\} \mid a'(\bar{e}) \leftarrow \mathsf{flowsTo}(e_2),
-  \arrayTy(a')\}}
+\textsf{schematic}(\bar{e})\} \mid a'(\bar{e}) \leftarrow \mathsf{flowsTo}(e_2)\}}
 \end{array}
@@
 %
-The predicate $\neigh$ classifies neighbourhood indices, based on
-the induction variables of the loop (implicit here).
+The predicate $\neigh(\bar{e_1})$ classifies indices where each
+$e \in \bar{e_1}$ is either a
+neighbourhood index (based on
+the induction variables of the loop, implicit here)
+or a constant index.\footnote{If all indices $e \in \bar{e}$ are constants, then
+  $\neigh(\bar{e})$ does not hold. Instead $a(\bar{e}$) is
+  considered a constant by our analysis, rather
+than an array subscript.}% contributing to a stencil.}
 Thus, we focus on assignments to an array subscript where the LHS
-index $\overline{e_1}$ is a neighbourhood index.  For
+indices $\overline{e_1}$ are neighbourhood indidces.  For
 all array subscripts that flow to the right-hand side of this
 statement, a finite map is constructed, mapping each array variable
-to a \emph{multiset} of indexing schemes for its subscripts, computed
+to a multiset of indexing schemes for its subscripts, computed
 with \textsf{schematic} (Def.~\ref{def:schematic}).
-
-Note that \textsf{schematic} is undefined ($\bot$) if
+Note \textsf{schematic} is undefined ($\bot$) if
 $\bar{e}$ contains relative indices which are not neighbourhood
-indices. The \textit{analysis} function feeds into the
+indices. The \textit{analysis} function is used by the
 \textit{modelise} function which filters out those arrays
@@
 & \mathit{modelise} (\textit{assg})
-= a \mapsto
+= \textit{let} \;\; M = \mathit{analyse}(\textit{assg}) (a) \;\; \textit{in}
+\\
+& a \mapsto
 \begin{cases}
 \bot & \bot \in M \\
-\textsf{only}(|M|) & \forall u \in M . \#(M, u) = 1 \\
-\textsf{mult}(|M|) & \forall u \in M . \#(M, u) = n
-\end{cases} \\
-& \quad \textit{where} \; M = \mathit{analyse}(\textit{assg}) (a)
+\textsf{only}(|M|) & \forall u \in M . \, \#(M, u) = 1 \\
+\textsf{mult}(|M|) & \forall u \in M . \, \#(M, u) = n
+\end{cases}
 \end{align*}
@@
 $M$.
-
 Thus, the analysis maps source code to index schemes, which is then
@@
 
-If the LHS $\overline{e_1}$ contains non-$0$ neighbour offsets
+As a small semantic concession to an otherwise syntactic
+analysis: if the LHS $\overline{e_1}$ contains non-$0$ neighbour offsets
 then all RHS schemes are \emph{relativised} by
@@
 produces an analysis which is equivalent to the analysis of \texttt{a(i) = b(i-1)}.
-\dnote{Somewhere we need a discussion about syntactic vs semantic
-analysis...}
 
@@
     b(i, j) = (a(i, j) + x + y) / 5.0
-  end do
-end do
+end do; end do
 \end{minted}
@@
 \begin{align*}
+{\small{
 \{\texttt{a(i-1, j)}, \texttt{a(i+1, j)}, \texttt{a(i, j-1)},
-  \texttt{a(i, j+1)}, \texttt{a(i, j)}\}
-\end{align*}
-Subsequently the result of \textit{analyse} on line 4 yields the map:
-\begin{align*}
-\texttt{a} \mapsto \{\vtwoh{-1}{0}^1, \vtwoh{1}{0}^1,
-          \vtwoh{0}{-1}^1, \vtwoh{0}{1}^1, \vtwoh{0}{0}^1\}
-\end{align*}
-%
-where multiplicities are denoted by superscripts. The \emph{modelise}
-function then yields the following since all multiplicities in the model are one:
+  \texttt{a(i, j+1)}, \texttt{a(i, j)}\}}}
+\end{align*}
+The \emph{modelise} function then yields the following
+since all multiplicities in the model are one:
 %%
@@
 
-In checking mode, the access pattern of stencil computations in the
-source language is verified against any specification associated with
-them in the code. The model of
-\S~\ref{sec:semantics} guides this process.  Checking
+Checking verifies the access pattern of a stencil computation in the
+source language against its associated specifications. The model of
+(\S~\ref{sec:semantics}) guides this process.  Checking
 proceeds by generating a model from a specification and generating
 a model from the source code, comparing them for
-\emph{consistency}. The notion of consistency is built out of index scheme
-\emph{generalisation} (Sect.~\ref{subsec:schemes-and-gen}).
+\emph{consistency} base on index scheme
+\emph{generalisation} (Def.~\ref{def:generalisation}).
 
 \begin{definition}[Checking]
-Let $a$ be an array variable and let
-$S \in \textit{spec}$ be a specification associated to $a$
-via \texttt{!= stencil $S$ :: a}. A stencil kernel
- $v(\overline{e_1}) = e_2$ in the source language
+Let $a$ be an array variable and
+$S$ a specification associated to $a$
+via \texttt{!= stencil $S$ :: a}.
+
+A stencil kernel
+ $v(\overline{e_1}) = e_2$
 conforms to $S$ if the predicate $\textbf{check}_n(S, v(\overline{e_1}) = e_2, a)$ holds,
@@
 \end{align*}
-Thus, checking proceeds by using \textit{modelise}
-(\S~\ref{subsec:analyis}) to extract index schemes from source code
+Thus, checking applies \textit{modelise}
+(\S~\ref{subsec:analysis}) to extract index schemes from source code,
 augmented with multiplicitly for the whole set of schemes.
@@
 
+\begin{figure}[t]
+\vspace{-1em}
+\begin{align*}
+\hspace{-0.5em}
+\begin{array}{c}
+%%
+\trule{up}\dfrac{\forall u \in M, \exists v \in N .  \, u \preceq v}
+      {\cons{M}{\textsf{upper}(N)}}
+\,
+\trule{low}\dfrac{\forall v \in N, \exists u \in M . \, u \preceq v}
+      {\cons{M}{\textsf{lower}(N)}} \\[1.2em]
+%%
+\trule{exact}\dfrac{\cons{M}{\textsf{lower}(N)} \qquad \cons{M}{\textsf{upper}(N)}}
+  {\cons{M}{\textsf{exact}(N)}} \\[1.2em]
+%%
+\trule{both}\dfrac{\cons{M}{\textsf{lower}(M_1)} \qquad \cons{M}{\textsf{upper}(M_2)}}
+  {\cons{M}{\textsf{both}(M_1, M_2)}}\\[1.2em]
+%%
+\trule{$1$}
+\dfrac{\cons{M}{N}}{\consA{\textsf{once}(M)}{\textsf{once}(N)}}
+\;\;
+\trule{$*$}
+\dfrac{\cons{M}{N}}{\consA{\textsf{mult}(M)}{\textsf{mult}(N)}}
+  \\[1.2em]
+\trule{$1 < *$}
+\dfrac{\cons{M}{N}}{\consA{\textsf{once}(M)}{\textsf{mult}(N)}}
+\end{array}
+\end{align*}
+\vspace{-0.5em}
+\caption{Consistency of models (code vs. specification)}
+\label{fig:consistency}
+\vspace{-0.9em}
+\end{figure}
+
 \begin{definition}
 Consistency relates models of source code to models
-of specifications (which are bounded) via two predicates:
+of specifications (which are bounded) via predicates:
 \begin{align*}
@@
 dimensionality.  We denote sets of indexing schemes
-$\mathcal{P}((\domainVal{})^n)$ that come from a specification model
-by $M$ and from source code model by $N$.
-
-Rules (\emph{up}) and (\emph{low}) are the core rules relating models of
-specifications and models of code. The (up) rule
+$\mathcal{P}((\domainVal{})^n)$
+from source code model by $M$ and from a specification model by $M$.
+
+Rules (\emph{up}) and (\emph{low}) are the core; (\emph{up})
 considers the case where the specification is
 an \term{atMost} upper-bound approximation.
-An upper bound specification means that for every scheme in the code
-model, modelling a source-level index, $\forall u \in C$ there exists
-a generalising scheme in the spec model $\exists v \in S$, \ie{}, $u
+An upper-bound specification means that for every scheme in the code
+model, modelling a source-level index, $\forall u \in M$ there exists
+a generalising scheme in the spec model $\exists v \in N$, \ie{}, $u
 \preceq v$. That is, every source-level index scheme is generalised by the specification.
-The (down) rule provides the case of
+The (\emph{down}) rule provides the case of
 \term{atLeast} lower-bound specifications. Consistency is dual to the above:
-for every scheme in the specification model $\forall v \in S$ there
-exists a scheme in the code model $\exists u \in C$ such that $u
+for every scheme in the specification model $\forall v \in N$ there
+exists a scheme in the code model $\exists u \in M$ such that $u
 \preceq v$. That is, every specification scheme is generalised by some source
@@
 provides both a consistent lower and consistent upper bound on
-for the source code. Thus, for every scheme in the code model there
+the source code. Thus, for every scheme in the code model there
 is a scheme in the spec model that generalises it, and for every
-scheme in the spec model there is a scheme in the source model that
+scheme in the specifcation there is a scheme in the source model that
 specialises it.  The (\textit{both}) rule is similar, decomposing
-a specification which has both a lower and upper bound into
+a specification with both a lower and upper bound into
 consistency of these bounds.
@@
 Rules $(1)$ and $(*)$ extend the \consName{}
-predicate to $\consAName{} \subseteq
-\textsf{Mult}(\mathcal{P}(\domainVal{}^n))
-\times \textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$
-which inludes multiplicity information. Both rules
+predicate to $\consAName{}$ with
+multiplicity information. Both rules
 state that consistent models remain consistent if they
@@
 
-\begin{figure}[t]
-\begin{align*}
-\hspace{-0.5em}
-\begin{array}{c}
-%%
-\trule{up}\dfrac{\forall u \in M, \exists v \in N .  \, u \preceq v}
-      {\cons{M}{\textsf{upper}(N)}}
-\,
-\trule{low}\dfrac{\forall v \in N, \exists u \in M . \, u \preceq v}
-      {\cons{M}{\textsf{lower}(N)}} \\[1.5em]
-%%
-\trule{exact}\dfrac{\cons{M}{\textsf{lower}(N)} \qquad \cons{M}{\textsf{upper}(N)}}
-  {\cons{M}{\textsf{exact}(N)}} \\[1.5em]
-%%
-\trule{both}\dfrac{\cons{M}{\textsf{lower}(M_1)} \qquad \cons{M}{\textsf{upper}(M_2)}}
-  {\cons{M}{\textsf{both}(M_1, M_2)}}\\[1.5em]
-%%
-\trule{$1$}
-\dfrac{\cons{M}{N}}{\consA{\textsf{once}(M)}{\textsf{once}(N)}}
-\;\;
-\trule{$*$}
-\dfrac{\cons{M}{N}}{\consA{\textsf{mult}(M)}{\textsf{mult}(N)}}
-  \\[1.5em]
-\trule{$1 < *$}
-\dfrac{\cons{M}{N}}{\consA{\textsf{once}(M)}{\textsf{mult}(N)}}
-\end{array}
-\end{align*}
-\vspace{-0.5em}
-\caption{Consistency of models (code vs. specification)}
-\label{fig:consistency}
-\vspace{-0.9em}
-\end{figure}
-
 \begin{example}
-  We give an extended example of verifying a stencil computation
-  against its specification. We consider a more general, but still
-  consistent, version of the \textit{five-point} stencil specification
-  inferred previously in \S~\ref{subsec:inference}:
-%
-\begin{minted}[breakindent=8.1em]{fortran}
+  Consider the following five-point stencil kernel with a (quite general) specification:
+%
+\begin{minted}[breakindent=8.5em]{fortran}
 != region :: fivepoint = centered(depth=1,dim=1) + centered(depth=1,dim=2)
@@
 \begin{align*}
+\vspace{-0.3em}
  N_s= \{  \vtwo{-1}{\infty},
@@
   \textsf{once}(\textsf{exact} (N_s))
-\end{align*}
-%
-\item Obtain the code model via $\textit{modelise}$, giving:
-%
-\begin{align*}
+\vspace{-0.3em}
+\end{align*}
+%
+\item Generate code model $\textit{modelise}(l3, \texttt{b}) =
+  \textsf{once} (M_c)$ where:
+%
+\begin{align*}
+\vspace{-0.3em}
 M_c = & \{
@@
   \}
-\\
-\textit{modelise}(l3, \texttt{b}) = & \textsf{once} (M_c)
+\vspace{-0.3em}
 \end{align*}
@@
   \begin{align*}
-    & \consSub{2}{\textsf{up}(N_s)}{M_c} \; 
-    \wedge \; \consSub{2}{\textsf{low}(N_s)}{M_c}
+\vspace{-0.3em}
+    & \consSub{2}{\textsf{up}(N_s)}{M_c} \;\,
+    \wedge \;\, \consSub{2}{\textsf{low}(N_s)}{M_c}
+\vspace{-0.3em}
   \end{align*}
@@
   \end{align}
+\end{enumerate}
 Since the consistency predicate is satisfied, the array
@@
   with respect to the {\textcolor{darkgreen}{\texttt{fivepoint}}} specification.
-\end{enumerate}
-\vspace{-0.7em}
 \end{example}
@@
 Consider an erroneous user modification to $l3$ with an
-additional term \texttt{b(i+2,j+2)} such that $l3'$ is:
-%
-\begin{minted}[firstnumber=3]{fortran}
-a(i,j) = b(i,j-1) + b(i,j) + b(i,j+1) + b(i-1,j) + b(i+1,j) + b(i+2,j+2)
-\end{minted}
-%
-The checking procedure proceeds in the same way as the first three
-steps of the previous example, except that the result of
+additional term \texttt{b(i+2,j+2)}, call it $l3'$.% such that $l3'$ is:
+%
+%\begin{minted}[firstnumber=3]{fortran}
+%a(i,j) = b(i,j-1) + b(i,j) + b(i,j+1) + b(i-1,j) + b(i+1,j) + b(i+2,j+2)
+%\end{minted}
+
+Checking proceeds as in the previous example, except
 $\mathit{modelise}(l3',\texttt{b})$ generates an additional scheme
-$\vtwoh{2}{2}$. Since the spec model has not changed and the code
-model is has only the additional scheme $\vtwoh{2}{2}$, the generalisations for
+$\vtwoh{2}{2}$. The generalisations for
 the lower bound \eqref{eq:check-example-lower}
-of step 4 above still holds since its consistency condition
-universally quantifies over the \emph{spec model}. However, for
+of step 4 above still hold since its consistency condition
+universally quantifies over the \emph{specification model}. However, for
 the upper bound, the consistency condition universally quantifies
-over scheme $u$ of the \emph{code model} for which there exists
+over schemes $u$ of the \emph{code model} for which there must exist
 a scheme $v \in N_s$ such that $u \preceq_2 v$. Thus, we require a $v
 \in N_s$ such that $\vtwoh{2}{2} \preceq_2 v$ but no such generalising
-scheme existing in the spec model. Hence the model fails to hold as an upper bound.
-
-Note that if the specification were as follows:
-%
-\begin{minted}{fortran}
-!= stencil atLeast, readOnce, fivepoint :: b
-\end{minted}
-%
-then consistency holds as it no longer requires the failing upper-bound case:
-$\consSub{2}{\textsf{mult}(\textsf{up}(N_s))}{\textsf{mult}(M_c)}$.
- This makes sense semantically since the region ${\textcolor{darkgreen}{\texttt{fivepoint}}}$
-is indeed contained in the regions touched by $\texttt{b}$ in $l3'$.
+scheme exists in the spec model. Hence checking fails.
 \end{example}
-
-\dnote{Build this into the definitions earlier I think}
-One way the implementation may deviate from the model is elimination of all $\infty$
-offset vector that occurs when all indices are in non-neighbourhood form \eg{}
-\mintinline{fortran}{b(42,0)}. The only generalisation scheme that can unify
-with this pattern is the most general one, meaning it requires outright
-rejection of an otherwise fine stencil computation. See the following example,
-where not discarding all $\infty$ access pattern prevents verification:
-%
-\begin{minted}{fortran}
-!= stencil forward(dim=1,depth=1) :: a
-a(i) = a(i) + a(i+1) + a(1)
-\end{minted}
 
@@
 
-We illustrate each step of inference using the
- five-point stencil program from Example~\ref{exm:checking}.
+To aid adoption of our approach to legacy code, we provide
+an inference procedure for generating specifications from code, which
+are inserted automatically as comments.
+We illustrate inference using the
+ five-point stencil of
+ Example~\ref{exm:checking}.
 

### 2016-11-16 paper.tex

@@
 
-Indexing schemes give an abstract
-representation of array subscripts in the source language.
+Indexing schemes abstract array subscripts.
 Our model represents the meaning of stencil specifications

### 2016-11-16 paper.tex
trimming in 5

@@
 Let $U$ range over the finite maps computed by \textit{modelise}.  For
-each array variable $a \in \mathsf{dom}(U)$, the algorithm constructs next a set of
+each array variable $a \in \mathsf{dom}(U)$, the algorithm constructs a set of
 $n$-dimensional rectangles covering all contiguous groups of schemes
@@
   represented as a $2 \times n$ matrix) with values drawn from
-  $\domainVal{}$. This gives the co-ordinates of a lower-bound vertex
-  (first component) and the upper-bound (second component). We write
+  $\domainVal{}$. This gives the co-ordinates of lower-bound 
+  and upper-bound vertices. We write
   spans as $\vect{x} = [\vect{x}^L \; \vect{x}^U]$, where $\vect{x}^L$
@@
 $n$-dimensional cube of side length $1$. For the five-point stencil
-example there are five unit spans.
-Figure~\ref{fig:inference-steps-informal}(a) illustrates these.
+example there are five unit spans, illustrated in
+Figure~\ref{fig:inference-steps-informal}(a).
 
@@
 $j \in \{1, \ldots, n\}$ then $\vect{x}_j = \vect{y}_j$ and for one
-$j$ then $\vect{x}^U_1 + 1 = \vect{y}^U$. For example, spans
+$j$ then $\vect{x}^U_1 + 1 = \vect{y}^L$. For example, spans
 $\left[\vtwoh{-1}{0}\; \vtwoh{-1}{0}\right]$ and
@@
 \begin{definition}\label{def:span-coalesc}
-  Span coalescing $\bullet$ is defined for spans contiguous in the
+  Span coalescing $\bullet$ is partial, defined for spans contiguous in the
   first dimension:
@@
 \vect{x} \bullet \vect{y}
-= &
-\begin{cases}
-(\vect{x}^L, \vect{y}^U) & \vect{x}^U_1 + 1 = \vect{y}^L_1 \; \wedge \;
-\vect{x}_{2:n} = \vect{y}_{2:n}) \\
-\bot  & \textit{otherwise}
-\end{cases}
+= (\vect{x}^L, \vect{y}^U) \;\;\;\, \textit{if} \;\;\; \vect{x}^U_1 + 1 = \vect{y}^L_1 \; \wedge \;
+\vect{x}_{2:n} = \vect{y}_{2:n} %\\
+%\bot  & \textit{otherwise}
+%\end{cases}
 % $m$-dimensions into coalesced spans which vary in
@@
 Figure~\ref{fig:inference-steps-informal}(b).
-This procedure is iterated until a fixed point is reached. In this
-example this is reached in the first step.
+This procedure is iterated until a fixed point is reached (in one step
+in this example).
  Then any spans that are completely contained by any other span are deleted,
@@
 is mapped to a unit span
-by $\textsf{unitSpans}(M) = \{[u \; u] \mid u\!\in\!M\}$. For our example:
+by $\textsf{unitSpans}(M) = \{[u \; u] \mid u\!\in\!M\}$, \eg{}:
 %%
@@
 \caption{Illustration of grouping spans in
-specification inference}
+inference}
 \label{fig:inference-steps-informal}
@@
 %
-This corresponds to all rotations of the space.  For our example:
+This corresponds to all rotations of the space, \eg{}:
 %
@@
 where $\pi^2_1$ is the identity permutation and $\pi^2_2$ is the
-permutation flips the order of the two elements in the
+permutation that flips the order of the two elements in the
 vectors.
@@
 \noindent
-\textbf{2)\,} Sort each permutation set into an ordered list, by the ordering:
+\textbf{2)\,} Sort each permutation set into a list, by the ordering:
 \begin{equation*}
@@
 %
-giving the sorting function $\textsf{sort}(M)$. For our example:
-%%
-\begin{align*}
-\textsf{sort}(P(M_0)) =
+giving the sorting function $\textsf{sort}(M)$, \eg{}, $\textsf{sort}(P(M_0))$:
+%%
+\begin{align*}
 \{&(\pi^2_1, [
@@
 \noindent
-\textbf{3)\,} Reduce each list pairwise by $\bullet$ operation
+\textbf{3)\,} Reduce each list pairwise by $\bullet$
 (Def.~\ref{def:span-coalesc}) to coalesce contiguous regions,
-given by $\textsf{fold}\bullet$. For our example, this yields:
+given by $\textsf{fold}\bullet$. For our example:
 %
@@
 \stwo{0}{1}{0}{1}
-]), (\pi^2_2, [
-\stwo{0}{-1}{0}{-1},
-\stwo{-1}{0}{1}{0},
-\stwo{0}{1}{0}{1}])\}
+]), (\pi^2_2, \textit{same as for $\pi_1$})\}
 \end{align*}
@@
 %
-\textbf{5)\,} Filter by the region containment predicate $\containedin$, that
-  is, if any region is contained within another then remove the
+\textbf{5)\,} Filter by the region containment predicate
+$\containedin$; if any region is contained within another then remove the
   smaller, defined:
@@
 Applying \spanOp{} again yields the same
-result for our example; the fixed point is reached.
+result for our example.%; the fixed point is reached.
 
@@
 Def.~\ref{def:mult-and-approx}).  The algebra mirrors the shape
-of the region combinators for sum and product,
-with operators $\oplus$ and $\odot$ and a constructor for \specDNF{}:
-%%
-\begin{align*}
-\oplus & : \mathsf{Approx}(\specDNF) \rightarrow \mathsf{Approx}(\specDNF) \rightarrow
-  \mathsf{Approx}(\specDNF) \\
-\odot & : \mathsf{Approx}(\specDNF) \rightarrow \mathsf{Approx}(\specDNF) \rightarrow
-        \mathsf{Approx}(\specDNF) \\
-\mathsf{toSpec} & : \mathbb{N}_{>0} \rightarrow (\domainVal{} \times \domainVal{}) \rightarrow \mathsf{Approx}(\specDNF)
-\end{align*}
-%%
-The $\mathsf{toSpec}$ function maps a dimension identifier
+of the region operators $\term{+}$ and $\term{*}$ 
+with binary operators $\oplus$ and $\odot$ on
+$\mathsf{Approx}(\specDNF)$.
+A function $\mathsf{toSpec}$ maps a dimension identifier
 and a pair of lower and upper bound
 values (one-dimensional) to a specification with a one-dimensional region.
+%%
+\begin{align*}
+\mathsf{toSpec} & : \mathbb{N}_{>0} \rightarrow (\domainVal{} \times \domainVal{}) \rightarrow \mathsf{Approx}(\specDNF)
+\end{align*}
+%%
 An entire span is converted into an $n$-dimensional
@@
 Let $r_1, ..., r_n$ range over elements of the \textit{region}
-syntax (see Figure~\ref{fig:syntax}).
+syntax (see Fig.~\ref{fig:syntax}).
 The set \specDNF{} is a subset of syntax trees \textit{spec},
 containing specifications in disjunctive normal form over regions
-where the disjunct is \term{+} (since its model is in terms of set
-union) and the conjunct is \term{*} (since its model is a form of tensor).
-That is elements of \specDNF{} are of the form:
+where the disjunct is \term{+} and the conjunct is \term{*}; 
+ elements of \specDNF{} are of the form:
 %%
@@
 \begin{align*}
-\specDNF{} & ::= \tySum{\specDNF} \mid \epsilon \\
-\tySum{\specDNF} & ::= \tyProd{\specDNF} \; \term{+} \; \tySum{\specDNF} \mid
+\specDNF{} & ::= \tySum{\specDNF} \mid \epsilon \quad\;\;
+\tySum{\specDNF} ::= \tyProd{\specDNF} \; \term{+} \; \tySum{\specDNF} \mid
   \tyProd{\specDNF} \\
@@
 %%
-We include the empty element $\epsilon$ which is not present in the
-syntax.
 In the following, $S, T$ range over elements of
 \specDNF{} and $\tyProd{s}, \tyProd{t}$ range over elements
-of $\tyProd{\specDNF}$ (for example, $\tyProd{s} = r^1_{1} \; \term{*}
-\ldots \term{*} \; r^1_n$).
+of $\tyProd{\specDNF}$, \ie{} $\tyProd{s} = r^1_{1} \; \term{*}
+\ldots \term{*} \; r^1_n$.
 \end{definition}
@@
 %% POINT
-\exact{\stenReflS{d}} & \llower = 0 \, \wedge \, \lupper = 0\\[0.2em]
+\exact{\stenReflS{d}} & \llower = 0 \, \wedge \, \lupper = 0\\[0.02em]
 %% FWD
-\exact{\stenFwdS{\lupper}{d}}  & \llower = 0 \, \wedge \, \lupper > 0 \\[0.2em]
-\exact{\stenFwdSR{\lupper}{d}{\irreflS}}\!\!\!\! & \llower = 1 \, \wedge \, \lupper > 0 \\[0.2em]
+\exact{\stenFwdS{\lupper}{d}}  & \llower = 0 \, \wedge \, \lupper > 0 \\[0.02em]
+\exact{\stenFwdSR{\lupper}{d}{\irreflS}}\!\!\!\! & \llower = 1 \, \wedge \, \lupper > 0 \\[0.02em]
 %% BWD
-\exact{\stenBwdS{|\llower|}{d}}  & \llower < 0 \, \wedge \, \lupper = 0 \\[0.2em]
-\exact{\stenBwdSR{|\llower|}{d}{\irreflS}}\!\!\!\! & \llower < 0 \, \wedge \, \lupper = -1 \\[0.2em]
+\exact{\stenBwdS{|\llower|}{d}}  & \llower < 0 \, \wedge \, \lupper = 0 \\[0.02em]
+\exact{\stenBwdSR{|\llower|}{d}{\irreflS}}\!\!\!\! & \llower < 0 \, \wedge \, \lupper = -1 \\[0.02em]
 %% CEN
-\exact{\stenCenS{\lupper}{d}}& \llower < 0 \wedge \lupper > 0 \wedge |\llower| = \lupper \\[0.2em]
+\exact{\stenCenS{\lupper}{d}}& \llower < 0 \wedge \lupper > 0 \wedge |\llower| = \lupper \\[0.02em]
 \exactR(\stenBwdS{|\llower|}{d} & \llower < 0 \wedge \lupper > 0  \wedge |\llower| \neq \lupper \\
-\qquad\qquad \texttt{+} \, \stenFwdS{\lupper}{d}) & \\[0.2em]
+\qquad\qquad \texttt{+} \, \stenFwdS{\lupper}{d}) & \\[0.02em]
 % BOUND
-\upperB{\stenFwdS{\lupper}{d}} & \llower > 1 \\[0.2em]
+\upperB{\stenFwdS{\lupper}{d}} & \llower > 1 \\[0.02em]
 \upperB{\stenBwdS{|\llower|}{d}} \quad & \llower < (-1)
- \\[0.2em]
+ \\[0.02em]
 % ABSOLUTE REP
@@
 \begin{definition} The $\mathsf{toSpec}$ constructor is defined
-in Figure~\ref{fig:to-spec}.
-The first two cases map to \texttt{forward} stencils since the lower
-bound is $0$ or $1$, with the depth determined by the upper bound of
-the span. If the lower bound is $1$, then the
-$0$-offset point is missing so the region is given a
-\texttt{nonpointed} attribute. The second two cases are similar for
-\texttt{backward} but now the depth of the region is determined by the
-absolute of the lower bound. In the fifth-case, where the lower bound
+in Figure~\ref{fig:to-spec}, where the lower and upper bounds
+determine region constants; $\mathsf{toSpec} \, d \, (l, u)$ is always defined
+ if $l$ less than or equal to $u$.
+
+Positive lower and upper bound yield \term{forward} regions, and
+negative lower and upper bounds yield \term{backward} regions.
+The \texttt{nonpointed} attribute (abbreviated \term{np})
+is present when the lower bound is at $1$ or $-1$. 
+In the sixth-case, where the lower bound
 is negative and the upper bound is positive and they have the same
@@
 the depth as the common magnitude. If the magnitudes are not the same
-(sixth case) then separated \texttt{forward} and \texttt{backward}
+(seventh case) then separated \texttt{forward} and \texttt{backward}
 regions are summed, with their corresponding depths.
-
 If the lower bound is greater than 1
@@
 origin and thus cannot be represented in the specification language.
-Thus, an upper bound is instead returned showing the maximum extent
+Thus, an upper bound is produced showing the maximum extent
 of the region. The penultimate case is the dual, when the lower bound is less
 than $-1$, giving an upper bound on a \texttt{backward} specification.
-
-Finally, if either bound is $\infty$, representing a non
-neighbourhood index, then an empty specification is given
- with no regions, indicating that there is not even a
-pointed 0-offset access in this dimension.
-
-Note, $\mathsf{toSpec} \, d \, (l, u)$ is always defined
- if $l$ less than or equal to $u$.
 \end{definition}
@@
 is indeed in DNF-form, where $\hat{\odot}$ is essentially applying
- the (\trule{\textsc{dist}}) rule of our equational theory
+ the \trule{\textsc{dist}} rule of the equational theory
  (Fig.~\ref{fig:equations}). The operation is
@@
 \begin{definition}The $\oplus$ operation on $\textsf{Approx}(\specDNF)$
-is defined in a similar way to $\odot$, with the intermediate $\hat{\oplus}$ on
-$\specDNF$ given by:
+is defined similarly to $\odot$, with the intermediate $\hat{\oplus}$ on
+$\specDNF$:
 %%
@@
 
-\begin{lemma}[$\hat{\odot}$, $\hat{\oplus}$ soundness]$\forall S,
+\begin{lemma}%[$\hat{\odot}$, $\hat{\oplus}$ soundness]
+$\forall S,
   T \in \specDNF$ where $S\!\neq \epsilon \wedge T\!\neq \epsilon$:

### 2016-11-16 paper.tex
Replace use of refs with Cref

@@
 
+\usepackage{cleveref}
+\crefname{section}{\S}{\S\S}
+\Crefname{section}{\S}{\S\S}
+\crefname{subsection}{\S\S}{\S\S\S}
+\Crefname{subsection}{\S\S}{\S\S\S}
+\crefname{figure}{fig.}{}
+\Crefname{figure}{Fig.}{}
+\crefname{definition}{def.}{}
+\Crefname{definition}{Def.}{}
+
 \CustomVerbatimEnvironment{SpecVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.5cm,
@@
   that captures common forms of data access patterns
-  (\S~\ref{sec:lang});
+  (\cref{sec:lang});
 
 \item We detail checking and inference algorithms for our
-  specifications (\S~\ref{sec:analysis}), derived from a
+  specifications (\cref{sec:analysis}), derived from a
   denotational model for the specificaiton language
-  (\S~\ref{sec:semantics});
+  (\cref{sec:semantics});
 
@@
   corpus of numerical Fortran program
-  (\S~\ref{sec:evaluation}), totalling one million lines of code. Our tool
+  (\cref{sec:evaluation}), totalling one million lines of code. Our tool
   identifies and infers specifications for 35,000 stencil computations in the
@@
 \newcommand{\domainVal}{\mathbb{Z}_\infty}
-\section{Stencil specifications}
+\section{Stencil specifications}\label{sec:lang}
 
@@
 out-of-bounds access at the edges). We revisit this hypothesis
-in \S~\ref{sec:evaluation} with the inference of
+in \cref{sec:evaluation} with the inference of
 such regular stencil patterns on our corpus of numerical programs.
@@
 We give various definitions used in the paper then
-\S~\ref{sec:syntax} outlines the syntax of stencil specifications
-with an informal introduction. \S~\ref{sec:semantics}
+\cref{subsec:syntax} outlines the syntax of stencil specifications
+with an informal introduction. \cref{sec:semantics}
 gives a semantic model via a set-based interpretation which makes
-concrete the meaning of specifications. \S~\ref{sec:eqs} then provides
+concrete the meaning of specifications. \cref{sec:eqs} then provides
 a syntactic equational theory for specifications which is
@@
 Now we informally describe what it means for a stencil kernel to be consistent
-with a specification. This will be formalised in \S~\ref{sec:semantics} and
-algorithmically described in  and \S~\ref{sec:analysis}.
+with a specification. This will be formalised in \Cref{sec:semantics} and
+algorithmically described in  and \Cref{sec:analysis}.
 Specifications are associated to a stencil kernel and are declared for one or
@@
 \subsection{Specification syntax}
-\label{sec:syntax}
-
-Figure~\ref{fig:syntax} gives the syntax of stencil specifications, which is
+\label{subsec:syntax}
+
+\Cref{fig:syntax} gives the syntax of stencil specifications, which is
 detailed below. The entry point is the \nonterm{specification} production which
@@
 by \emph{approximation} and \emph{multiplicity} information
-(in \textit{spec} in Fig.~\ref{fig:syntax}).
+(in \textit{spec} in \Cref{fig:syntax}).
 
@@
 (2) it is used in the inference and checking algorithms
-(Section~\ref{sec:analysis}); (3) justifies an equational theory for
+(\Cref{sec:analysis}); (3) justifies an equational theory for
 specifications in the next section; and (4) can be used to guide
@@
   (\ie{}, the constant integer $c$ in an indexing expression of the
-  form $i \pm c$, Def.~\ref{def:neighbour}) and $\infty$ is a wild-card
+  form $i \pm c$, \Cref{def:neighbour}) and $\infty$ is a wild-card
   representing both constant indices and unspecified indices.
@@
  is given precisely by the following function \textsf{schematic},
- which forms part of checking and inference (\S\ref{sec:analysis}).
+ which forms part of checking and inference (\Cref{sec:analysis}).
 
@@
 $n$-dimensional indexing schemes. The interpretation is overloaded on
-\emph{regions} in Figure~\ref{fig:region-model} and on the top-level
-of a specification \textit{spec} in Figure~\ref{fig:spatial-model}.
+\emph{regions} in \Cref{fig:region-model} and on the top-level
+of a specification \textit{spec} in \Cref{fig:spatial-model}.
 
@@
 \noindent
-The first four equations of Fig.~\ref{fig:region-model}
+The first four equations of \Cref{fig:region-model}
 give the model of the region constants as sets of indexing
@@
 \noindent
-The final two equations of Fig.~\ref{fig:region-model} give the models
+The final two equations of \Cref{fig:region-model} give the models
 of $\term{+}$ and $\term{*}$. For $\term{+}$ this is the
@@
 $\interp{\textit{region}}_n \in \mathcal{P}(\domainVal{}^n)$.
-Figure~\ref{fig:spatial-model} defines the top-level model of the
+\Cref{fig:spatial-model} defines the top-level model of the
 \textit{spec} syntax, which includes the multiplicity and
@@
 $\textsf{Mult}$ type corresponds to the presence or absence of the
-\term{readOnce} modifier as shown in Figure~\ref{fig:spatial-model}.
+\term{readOnce} modifier as shown in \Cref{fig:spatial-model}.
 The $\textsf{Approx}$ type corresponds to the presence or absence of
@@
 used later to give a normal form for specifications
-akin to Disjunctive Normal Form (\S~\ref{sec:inf-step4}).
+akin to Disjunctive Normal Form (\cref{sec:inf-step4}).
 
@@
 specification, but it is also used extensively in the inference
-procedure (\S~\ref{subsec:inference}) to simplify specifications.
+procedure (\cref{subsec:inference}) to simplify specifications.
 
@@
 
-Figure~\ref{fig:inequations} defines a notion of sub-specifications
+\Cref{fig:inequations} defines a notion of sub-specifications
 via the relation $<:^r$ on \textit{region}
@@
 \end{equation*}
-where $\consAName$ implements consistency in checking (\S~\ref{subsec:checking}).
+where $\consAName$ implements consistency in checking (\cref{subsec:checking}).
 \end{theorem}
@@
 We outline here the procedures for checking conformance
-of source code against specifications (\S~\ref{subsec:checking})
-and for infering specifications from code (\S~\ref{subsec:inference}).
+of source code against specifications (\Cref{subsec:checking})
+and for infering specifications from code (\Cref{subsec:inference}).
 Both rely on a program analysis that converts array indexing
 expressions into sets of index schemes. We outline this analysis
-first (\S~\ref{subsec:analysis}). Note that the analysis procedure
+first (\cref{subsec:analysis}). Note that the analysis procedure
 can be made arbitrarily more complex and wide-ranging independent
@@
 to a multiset of indexing schemes for its subscripts, computed
-with \textsf{schematic} (Def.~\ref{def:schematic}).
+with \textsf{schematic} (\Cref{def:schematic}).
 Note \textsf{schematic} is undefined ($\bot$) if
@@
 source language against its associated specifications. The model of
-(\S~\ref{sec:semantics}) guides this process.  Checking
+(\Cref{sec:semantics}) guides this process.  Checking
 proceeds by generating a model from a specification and generating
@@
 \emph{consistency} base on index scheme
-\emph{generalisation} (Def.~\ref{def:generalisation}).
+\emph{generalisation} (\Cref{def:generalisation}).
 
@@
 Thus, checking applies \textit{modelise}
-(\S~\ref{subsec:analysis}) to extract index schemes from source code,
+(\Cref{subsec:analysis}) to extract index schemes from source code,
 augmented with multiplicitly for the whole set of schemes.
@@
 where $\consName$ relates models without multiplicity.
-Figure~\ref{fig:consistency} defines the two predicates, where for
+\Cref{fig:consistency} defines the two predicates, where for
 brevity we make implicit the subscript $n$ defining the
@@
 \item Generate the model from the specification as in
-  \S~\ref{sec:semantics}.
+  \cref{sec:semantics}.
 %
@@
 %
-\item By the definition of consistency (Fig.~\ref{fig:consistency})
+\item By the definition of consistency (\Cref{fig:consistency})
  $\textsf{once}(\textsf{exact}(N_s))$ is consistent with
@@
  five-point stencil of
- Example~\ref{exm:checking}.
+ \Cref{exm:checking}.
 
@@
 The first step performs the
-analysis of \S~\ref{subsec:analysis} to produce a
+analysis of \cref{subsec:analysis} to produce a
 model of the code in terms of indexing schemes.
@@
 example there are five unit spans, illustrated in
-Figure~\ref{fig:inference-steps-informal}(a).
+\Cref{fig:inference-steps-informal}(a).
 
@@
 generates all possible ways of coalescing regions. For our example,
-the unit spans in Figure~\ref{fig:inference-steps-informal}(a)
+the unit spans in \Cref{fig:inference-steps-informal}(a)
 are coalesced into the contiguous spans in
-Figure~\ref{fig:inference-steps-informal}(b).
+\Cref{fig:inference-steps-informal}(b).
 This procedure is iterated until a fixed point is reached (in one step
@@
 leaving a minimal set of spans (which may overlap, but none of which
-fully contains another), Figure~\ref{fig:inference-steps-informal}(c).
+fully contains another), \Cref{fig:inference-steps-informal}(c).
 The final spans are used to synthesise concrete stencil specification
-syntax (\S~\ref{sec:inf-step4}), where
+syntax (\cref{sec:inf-step4}), where
 each span becomes a product \texttt{*} of regions which are
@@
 \textbf{3)\,} Reduce each list pairwise by $\bullet$
-(Def.~\ref{def:span-coalesc}) to coalesce contiguous regions,
+(\Cref{def:span-coalesc}) to coalesce contiguous regions,
 given by $\textsf{fold}\bullet$. For our example:
@@
 in the $\textsf{Approx}$ type (see
-Def.~\ref{def:mult-and-approx}).  The algebra mirrors the shape
+\Cref{def:mult-and-approx}).  The algebra mirrors the shape
 of the region operators $\term{+}$ and $\term{*}$ 
@@
 Let $r_1, ..., r_n$ range over elements of the \textit{region}
-syntax (see Fig.~\ref{fig:syntax}).
+syntax (see \Cref{fig:syntax}).
 The set \specDNF{} is a subset of syntax trees \textit{spec},
@@
 \begin{definition} The $\mathsf{toSpec}$ constructor is defined
-in Figure~\ref{fig:to-spec}, where the lower and upper bounds
+in \Cref{fig:to-spec}, where the lower and upper bounds
 determine region constants; $\mathsf{toSpec} \, d \, (l, u)$ is always defined
@@
  the \trule{\textsc{dist}} rule of the equational theory
- (Fig.~\ref{fig:equations}). The operation is
+ (\Cref{fig:equations}). The operation is
 commutative and associative, and sound with respect to our
-model (see Lemma~\ref{lem:alg-soundness} below).
+model (see \Cref{lem:alg-soundness} below).
 The $\hat{\odot}$ operation is then lifted to all combinations of
@@
 \paragraph{Hypothesis 1: Stencil computations
-(as in Definition~\ref{def:stencil}) are a common programming pattern
+(as in \Cref{def:stencil}) are a common programming pattern
 in numerical code}
@@
 code for future edits. Even if they were to provide a syntactic representation,
-for complex stencils such as Navier-Stokes from \S~\ref{sec:intro}, it would be
+for complex stencils such as Navier-Stokes from \cref{sec:intro}, it would be
 as verbose as the code itself making it difficult for programmer to understand

### 2016-11-16 paper.tex
rev of the eval

@@
 
-We have processed nearly one million lines of Fortran code using our
-CamFort tool and examined the output to try and determine how useful
-and applicable it may be. The body of Fortran code is collected from
-many computational science projects for which we believe our tool will
-be potentially helpful.
+%We have processed nearly one million lines of Fortran code using our
+%CamFort tool and examined the output to determine how useful
+%and applicable it may be. Our corpus of Fortran code is collected from
+%many computational science projects for which we believe our tool will
+%be potentially helpful.
 
@@
 
-\subsection{Core hypotheses}
-
-\paragraph{Hypothesis 1: Stencil computations
-(as in \Cref{def:stencil}) are a common programming pattern
-in numerical code}
+We first examined how frequently stencil computations occur in our
+corpus of Fortran code. This was carried out by running CamFort in
+whole-code-base inference mode. We parsed 959,427 lines of Fortran
+code and found that 10\% (97,439) of statements have a left-hand side
+is an array subscript on neighbourhood indices. This supports the idea
+that stencil-like computations are common in scientific code.
+
+We would not expect to infer a stencil for each of these statements
+because our stencil language places significant restrictions on the
+types of array-subscript-statements that we will class as a
+stencil. In fact, CamFort was able to infer a stencil from 30\% of
+these statements. A single statement can produce multiple stencils and
+we ended up with 60,525 stencils. This shows that we can still express
+a large number of stencil uses within our high-level abstraction.
 
@@
 % Percentage of a code base that is taken up by stencils
-
-Out of 1.35 million lines of code that we fed into our program, and
-the 959,427 lines that we were able to parse, we identified 60,525
-cases that fit our criteria for stencil specification.
-
-\paragraph{Hypothesis 2: Most stencil computations
-have a regular pattern which can be captured with a simple
-language of contiguous regions}
+%Out of 1.35 million lines of code that we fed into our program, and
+%the 959,427 lines that we were able to parse, we identified 60,525
+%cases that fit our criteria for stencil specification.
+
+%\paragraph{Hypothesis 2: Most stencil computations
+%have a regular pattern which can be captured with a simple
+%language of contiguous regions}
 
@@
 
-In the analysed body of code we found that we were able to deduce
-some form of stencil specification from approximately 30\% of the
-potential stencils examined.
-
-%
-\subsection{Analysing the design of our specifications}
+%In the analysed body of code we found that we were able to deduce
+%some form of stencil specification from approximately 30\% of the
+%potential stencils examined.
+
+%
+%\subsection{Analysing the design of our specifications}
 
@@
 
-55,504 specifications involved only reflexive regions, of which 15,823
-only had specifications for some of the dimensions, while 39,681 were
-reflexive on all dimensions.
+The vast majority of stencils we found were relatively simple but we
+found significant numbers of more complex shapes. We grouped common
+patterns in to categories which we describe below.
+
+\textbf{All pointed} 55,504 of the stencils we found involved 
+only pointed regions. 39,681 of these were pointed in all
+dimensions. Common examples of this were pointwise transformations on
+data (such as scaling).
+
+\textbf{Single action} specifications consist of no more than one 
+forward, backward, or centered region and can be combined with any
+number of pointed regions. We identified 4,837 single-action
+specifications, of which 1,532 were single-action with a
+\texttt{nonpointed} modifier.
+
+\textbf{Multi-action} specifications consist of at least two uses 
+of forward, backward, or centered regions, and can be combined with
+any number of pointed regions. We identified 156 multi-action
+specifications out of which 146 had regions combined only with
+$\term{*}$.
+
+\textbf{Bounded} specifications were fairly unusual, with only 157
+\texttt{atMost} and 36 \texttt{atLeast}, the latter of which were
+always also bounded on both sides.
 
@@
 % singleActionIrr: 1532
-A single-action specification consists of no more than one forward,
-backward, or centered region and can be combined with any number of
-reflexive regions. We identified 4,837 single-action specifications,
-of which 1,532 were single-action with an \texttt{irreflexive}
-modifier.
 
@@
     % multiActionMulOnly: 146
-A multi-action specification consists of at least two uses of forward,
-backward, or centered regions, and can be combined with any number of
-reflexive regions. We identified 156 multi-action specifications out
-of which 146 had regions combined only with $\term{*}$.
 
@@
 
-Bounded specifications were fairly unusual, with only 157
-\texttt{atMost} and 36 \texttt{atLeast}, the latter of which were
-always also bounded on both sides.
     % atLeast: 36

### 2016-11-16 paper.tex
Punctuation correction
Clarify absorption of infinity
Simplify pairwise permutation expl. for infinity

@@
 \end{align*}
-Note the absorbing behaviour of $\infty$ with respect to
-multiplication.
+Note the absorbing behaviour of $\infty$ under multiplication above, \eg{} $2
+\times \infty = \infty$
 \end{example}
@@
 $\term{*}$ takes the tensor $\otimes$ of subterm models. The tensor is
-more involved; we give some informal intuition first.
+more involved. We give some informal intuition first.
 
@@
 Pairwise permutation generates all possible interleavings of two
-vectors. Some care is then taken over
-how this interacts with $\infty$: pairwise permutation
-corresponds to non-deterministic choice, in each dimension $i$,
-between two values $u_i$ and $v_i$, but if either
-is $\infty$ then the resulting pairwise permutation is
-filtered out if the dimension $i$ has \emph{any other scheme
-in the model where $u_i \neq \infty$}, \eg{}:
+vectors. Some care is then taken over how this interacts with $\infty$.
+In the presence of $\infty$, generate all interlavings as if $\infty$ is
+an integer, then filter out vectors with $\infty$ at any coordinate, \eg{}
 \begin{align*}

### 2016-11-16 paper.tex
Switch from conjunction to comma in tensor product definition
added examples to eval

--- 2016-11-16 paper.tex
@@
 
-\textbf{Single action} specifications consist of no more than one 
+\textbf{Single-action} specifications consist of no more than one 
 forward, backward, or centered region and can be combined with any
@@
 always also bounded on both sides.
+
+\begin{figure}[t]\begin{minted}[fontsize=\scriptsize,breakindent=0em,linenos=false,xleftmargin=0em,breakafter=)]{fortran}
+!=stencil readOnce,(forward(depth=1,dim=3,nonpointed))*(backward(depth=1,dim=1))*(backward(depth=1,dim=2))+(forward(depth=1,dim=3))*(backward(depth=1,dim=1))*(pointed(dim=2))+(forward(depth=1,dim=3))*(backward(depth=1,dim=1,nonpointed))*(backward(depth=1,dim=2,nonpointed))+(forward(depth=1,dim=3))*(backward(depth=1,dim=2))*(pointed(dim=1))+(backward(depth=1,dim=1))*(backward(depth=1,dim=2,nonpointed))*(pointed(dim=3))+(backward(depth=1,dim=1,nonpointed))*(backward(depth=1,dim=2))*(pointed(dim=3))::x
+\end{minted}
+\caption{A complex specification inferred from
+  \textbf{Unified Model}\label{fig:smagorinsky}}
+\vspace{-1em}
+\end{figure}
+
+The single- and multi-action classes represent more complex stencils
+with a real possibility for programmer error. For example, the Unified
+Model has an implementation of the Smagorinsky subgrid-scale model for
+calculating turbulence on which our inference yields 39 specs from 340
+lines of code. This is an impressive reduction given
+the complexity of the algorithm.  We show one example in
+Figure~\ref{fig:smagorinsky}. This specifies the access pattern to
+a 3-dimensional array (which we have renamed to be called
+\texttt{x}). The array is accessed 8 times in the computation.
+
+\subsection{Detecting errors in the 2-D Jacobi iteration}
+
+One common example of a stencil computation is two-dimensional
+Jacobi iteration that repeatedly goes through each cell in a matrix
+and computes the average value of the four adjacent cells. The kernel
+can be expressed in a line of Fortran code:
+\begin{minted}{fortran}
+  a(i,j) = (a(i-1,j)+a(i+1,j)+a(i,j+1)+a(i,j-1))/4
+\end{minted}
+and CamFort is able to infer a precise specification for this:
+\begin{minted}[breakafter=+:,breakindent=-0.6em,breaksymbolsep=0.4em,linenos=false,xleftmargin=-0.5em]{fortran}
+  != stencil pointed(dim=1)*centered(depth=1,dim=2,nonpointed)+pointed(dim=2)*centered(depth=1,dim=1,nonpointed)::a
+\end{minted}
+
+We examined whether programmer errors would be detected by replacing
+the array index offsets with $-1$, $0$, or $1$ and running our
+verification algorithm. Camfort correctly reported a verification
+failure in each of 6,537 permutations corresponding to an error.  We
+note that the iteration is computing the average of four adjacent
+cells so $24$ ($4!$) of the possible array index pertubations do not
+correspond to a programming error.
+
+% /um/trunk/src/atmosphere/diffusion_and_filtering/turb_smagorinsky.F90
+% Tried introducing indexing errors replacing one of the indices of cx_rho_instance with -1,0,+1
+% 42 cases were caught
+% 6 cases were not caught
 

### 2016-11-16 paper.tex

@@
   \Bigg\{x \; \Bigg| \;
-    \parbox{5.2cm}{$(u, v) \in M \times N $ \\[0.1em]
-                  $\;\; \wedge \;\, i \in \{1, \ldots, 2^n\} \, \wedge
-                  \, x = (u \bowtie v)_i$ \\[0.1em]
-                  $\;\; \wedge \;\, j \in (\effdims{M}{n} \cup
-                   \effdims{N}{n})$\\[0.1em]
-                   $\;\; \wedge \;\, x_j \neq \infty$
+    \parbox{5.2cm}{$(u, v) \in M \times N,$ \\[0.1em]
+                  $\;\; i \in \{1, \ldots, 2^n\}, \,
+                  x = (u \bowtie v)_i,$ \\[0.1em]
+                  $\;\; j \in (\effdims{M}{n} \cup
+                   \effdims{N}{n}),$\\[0.1em]
+                   $\;\; x_j \neq \infty$
                   } \Bigg\}

### 2016-11-16 paper.tex
Minor edit on infinity elimination of perms with infinity in constrained dimensions

@@
 The permuted schemes $x = (u \bowtie v)_i$ added to the model
-$M \otimes N$ are only those which have do not have $\infty$ in
+$M \otimes N$ are only those without $\infty$ in
 the constrained dimensions given by each model ($x_j \neq
-\infty$). This filters out all permutations which have
+\infty$). This eliminates all permutations with
 $\infty$ in a constrained dimension position.
@@
 
+There were a variety of reasons why we did not infer a
+stencil from every potential statement: \\
+\noindent 
+1) \textbf{Non-subset induction variables} occur when the
+induction variables on the RHS are not a subset of those in the LHS. These
+cases are not stencils by our definition. The degenerate case of this
+is to have only constant indices on the LHS. We see lots of examples
+of this in loops as accumulators \eg{} computing the sum over an array; \\
+\noindent
+2) \textbf{Derived induction variables} where the
+indexing variable (\mintinline{fortran}{x}) is derived from an
+induction variable (\mintinline{fortran}{i}) as in
+\mintinline{fortran}{x = len - i};  \\
+\noindent
+3) \textbf{Inconsistent induction dimensions} occur when
+an induction variable is used to specify more than one array dimension
+on the RHS or multiple induction variables are used for the same
+dimension on the RHS. These are common in matrix operations such as
+LU-decomposition with assignments such as
+\mintinline{fortran}{a(l) = a(l) - a(m) * b(l, m)}.
+
 %    numStencilSpecs: 60525
@@
 the complexity of the algorithm.  We show one example in
-Figure~\ref{fig:smagorinsky}. This specifies the access pattern to
+Figure~\cref{fig:smagorinsky}. This specifies the access pattern to
 a 3-dimensional array (which we have renamed to be called
@@
 
+We have presented a high-level language for the specification of
+stencils in numerical code. As opposed to existing approaches our
+language takes inspiration from the numerical literature and provides
+a concise, abstract description of the stencil.
+
+We evaluated our approach on top of CamFort, an open-source analysis
+tool for Fortran programs, running the inference process over a corpus
+of a million lines of Fortran code. We showed that our language is
+capable of capturing many real-world uses of stencils and showed in a
+case-study that it will detect (simulated) programmer errors.
+
+Our definition of a stencil computation meant that we generated no
+specification for a large number of iterated array accesses in the
+corpus, such as stencils that incorporates reductions. For
+future work, we will be looking into expanding the flexibility of 
+our inference approach to more of these examples.
+
 \bibliography{references}

### 2016-11-16 paper.tex
delete redundant textit
Be consistent in vector space domain notation

@@
 
-\paragraph{Model of \textit{spec}}
+\paragraph{Model of spec}
 
@@
 schemes, that is
-$\interp{\textit{region}}_n \in \mathcal{P}(\domainVal{}^n)$.
+$\interp{\textit{region}}_n \in \mathcal{P}((\domainVal{})^n)$.
 \Cref{fig:spatial-model} defines the top-level model of the
@@
 approximation modifiers. The domain of this interpretation
-is $\textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$
+is $\textsf{Mult}(\textsf{Approx}(\mathcal{P}((\domainVal{})^n)))$
 where $\textsf{Mult}$ and $\textsf{Approx}$ are labelled variants

### 2016-11-16 paper.tex
added some explanation of why we didn't infer things and a scrappy conclusion

--- 2016-11-16 paper.tex
@@
 The permuted schemes $x = (u \bowtie v)_i$ added to the model
-$M \otimes N$ are only those without $\infty$ in
+$M \otimes N$ are only those which have do not have $\infty$ in
 the constrained dimensions given by each model ($x_j \neq
-\infty$). This eliminates all permutations with
+\infty$). This filters out all permutations which have
 $\infty$ in a constrained dimension position.

### 2016-11-16 paper.tex
related work

@@
 stencil computations involves some ad hoc testing to ensure that no
-transcription mistakes have been made. This may involve visual
-inspection of the results, based on a rendered image. Typically such
+transcription mistakes have been made (\eg{}, by visual
+inspections, or comparison against an manufactured or 
+analytical solutionst~\cite{farrell2010automated}). Typically such
 testing is discarded once the code is seen to be correct.
@@
 
-Our verification tool does not target the
+Our verification tool does not target a
 class of bugs that can be detected automatically (\emph{push-button
@@
 
-Now we informally describe what it means for a stencil kernel to be consistent
-with a specification. This will be formalised in \Cref{sec:semantics} and
-algorithmically described in  and \Cref{sec:analysis}.
-Specifications are associated to a stencil kernel and are declared for one or
-more array variables $\bar{v}$ within that kernel.  A stencil specification
-defines a \emph{region} associated to an array variable $a$ against which a
-stencil kernel should be \emph{consistent}. The region defined by a
-specification is an $n$-dimensional rectilinear shape in a discrete space
-crossing the origin with respect to the induction variable. Consistency is then
-a two way requirement. All array subscripts on $a$
-contributing to the kernel should be consistent with the region. Conversely,
-all parts of the region should have an index corresponding it.
+For now we informally describe what it means for a stencil kernel to
+be consistent with a specification. This will be formalised in
+\Cref{sec:semantics} and algorithmically described in and
+\Cref{sec:analysis}.  Specifications are associated to a stencil
+kernel and are declared for one or more array variables $\bar{v}$
+within that kernel.  A stencil specification defines a \emph{region}
+associated to an array variable $a$ against which a stencil kernel
+should be \emph{consistent}. The region defined by a specification is
+an $n$-dimensional rectilinear shape in a discrete space crossing the
+origin with respect to the induction variable. Consistency is then a
+two way requirement. All array subscripts on $a$ contributing to the
+kernel should be consistent with the region. Conversely, all parts of
+the region should have an index corresponding it.
 
@@
 \end{align*}
-Note the absorbing behaviour of $\infty$ under multiplication above, \eg{} $2
-\times \infty = \infty$
+Note the absorbing behaviour of $\infty$ with respect to
+multiplication.
 \end{example}
@@
 $\term{*}$ takes the tensor $\otimes$ of subterm models. The tensor is
-more involved. We give some informal intuition first.
+more involved; we give some informal intuition first.
 
@@
 Pairwise permutation generates all possible interleavings of two
-vectors. Some care is then taken over how this interacts with $\infty$.
-In the presence of $\infty$, generate all interlavings as if $\infty$ is
-an integer, then filter out vectors with $\infty$ at any coordinate, \eg{}
+vectors. Some care is then taken over
+how this interacts with $\infty$: pairwise permutation
+corresponds to non-deterministic choice, in each dimension $i$,
+between two values $u_i$ and $v_i$, but if either
+is $\infty$ then the resulting pairwise permutation is
+filtered out if the dimension $i$ has \emph{any other scheme
+in the model where $u_i \neq \infty$}, \eg{}:
 \begin{align*}
@@
   \Bigg\{x \; \Bigg| \;
-    \parbox{5.2cm}{$(u, v) \in M \times N,$ \\[0.1em]
-                  $\;\; i \in \{1, \ldots, 2^n\}, \,
-                  x = (u \bowtie v)_i,$ \\[0.1em]
-                  $\;\; j \in (\effdims{M}{n} \cup
-                   \effdims{N}{n}),$\\[0.1em]
-                   $\;\; x_j \neq \infty$
+    \parbox{5.2cm}{$(u, v) \in M \times N $ \\[0.1em]
+                  $\;\; \wedge \;\, i \in \{1, \ldots, 2^n\} \, \wedge
+                  \, x = (u \bowtie v)_i$ \\[0.1em]
+                  $\;\; \wedge \;\, j \in (\effdims{M}{n} \cup
+                   \effdims{N}{n})$\\[0.1em]
+                   $\;\; \wedge \;\, x_j \neq \infty$
                   } \Bigg\}
@@
 
-There were a variety of reasons why we did not infer a
-stencil from every potential statement: \\
-\noindent 
-1) \textbf{Non-subset induction variables} occur when the
-induction variables on the RHS are not a subset of those in the LHS. These
-cases are not stencils by our definition. The degenerate case of this
-is to have only constant indices on the LHS. We see lots of examples
-of this in loops as accumulators \eg{} computing the sum over an array; \\
-\noindent
-2) \textbf{Derived induction variables} where the
-indexing variable (\mintinline{fortran}{x}) is derived from an
-induction variable (\mintinline{fortran}{i}) as in
-\mintinline{fortran}{x = len - i};  \\
-\noindent
-3) \textbf{Inconsistent induction dimensions} occur when
-an induction variable is used to specify more than one array dimension
-on the RHS or multiple induction variables are used for the same
-dimension on the RHS. These are common in matrix operations such as
-LU-decomposition with assignments such as
-\mintinline{fortran}{a(l) = a(l) - a(m) * b(l, m)}.
-
 %    numStencilSpecs: 60525
@@
 
-\textbf{Single-action} specifications consist of no more than one 
+\textbf{Single action} specifications consist of no more than one 
 forward, backward, or centered region and can be combined with any
@@
 
-\begin{figure}[t]\begin{minted}[fontsize=\scriptsize,breakindent=0em,linenos=false,xleftmargin=0em,breakafter=)]{fortran}
-!=stencil readOnce,(forward(depth=1,dim=3,nonpointed))*(backward(depth=1,dim=1))*(backward(depth=1,dim=2))+(forward(depth=1,dim=3))*(backward(depth=1,dim=1))*(pointed(dim=2))+(forward(depth=1,dim=3))*(backward(depth=1,dim=1,nonpointed))*(backward(depth=1,dim=2,nonpointed))+(forward(depth=1,dim=3))*(backward(depth=1,dim=2))*(pointed(dim=1))+(backward(depth=1,dim=1))*(backward(depth=1,dim=2,nonpointed))*(pointed(dim=3))+(backward(depth=1,dim=1,nonpointed))*(backward(depth=1,dim=2))*(pointed(dim=3))::x
-\end{minted}
-\caption{A complex specification inferred from
-  \textbf{Unified Model}\label{fig:smagorinsky}}
-\vspace{-1em}
-\end{figure}
-
-The single- and multi-action classes represent more complex stencils
-with a real possibility for programmer error. For example, the Unified
-Model has an implementation of the Smagorinsky subgrid-scale model for
-calculating turbulence on which our inference yields 39 specs from 340
-lines of code. This is an impressive reduction given
-the complexity of the algorithm.  We show one example in
-Figure~\cref{fig:smagorinsky}. This specifies the access pattern to
-a 3-dimensional array (which we have renamed to be called
-\texttt{x}). The array is accessed 8 times in the computation.
-
-\subsection{Detecting errors in the 2-D Jacobi iteration}
-
-One common example of a stencil computation is two-dimensional
-Jacobi iteration that repeatedly goes through each cell in a matrix
-and computes the average value of the four adjacent cells. The kernel
-can be expressed in a line of Fortran code:
-\begin{minted}{fortran}
-  a(i,j) = (a(i-1,j)+a(i+1,j)+a(i,j+1)+a(i,j-1))/4
-\end{minted}
-and CamFort is able to infer a precise specification for this:
-\begin{minted}[breakafter=+:,breakindent=-0.6em,breaksymbolsep=0.4em,linenos=false,xleftmargin=-0.5em]{fortran}
-  != stencil pointed(dim=1)*centered(depth=1,dim=2,nonpointed)+pointed(dim=2)*centered(depth=1,dim=1,nonpointed)::a
-\end{minted}
-
-We examined whether programmer errors would be detected by replacing
-the array index offsets with $-1$, $0$, or $1$ and running our
-verification algorithm. Camfort correctly reported a verification
-failure in each of 6,537 permutations corresponding to an error.  We
-note that the iteration is computing the average of four adjacent
-cells so $24$ ($4!$) of the possible array index pertubations do not
-correspond to a programming error.
-
-% /um/trunk/src/atmosphere/diffusion_and_filtering/turb_smagorinsky.F90
-% Tried introducing indexing errors replacing one of the indices of cx_rho_instance with -1,0,+1
-% 42 cases were caught
-% 6 cases were not caught
-
 % (/ 55504.0 60525) = 91.7%
@@
 
-\section{Related work}\label{sec:related-work}
-
-\citet{kamil2016verified} propose verified lifting to extract a mathematical
-representation of low-level, and potentially optimised, stencils in Fortran
-code. Their aim is to use the extracted predicate to pipe to the \textsf{Halide}~\citep{ragan2013halide} high-performance
-compiler and so they need to capture the full meaning of the stencil computation.
+\section{Related work and conclusions}
+\label{sec:discussion}
+
+Various deductive verification tools can express array indexing
+in their specifications, \eg{}, ACSL of \citet{baudin2008acsl}
+for C (see \eg{}~\citet[Example 3.4.1]{burghardt2010acsl}). Thus, a specification
+can be given for a stencil computation, 
+however, such a specification would exactly replicate the fine-grained
+indexing seen in the code and would simply be prone to low-level input
+errors. Our approach is much more abstract,
+capturing the shape of a specification in concise format different to
+the code.
+
+\citet{kamil2016verified} propose \emph{verified lifting} to extract a
+functionally-complete mathematical representation of low-level, and
+potentially optimised, stencils in Fortran code. This extracted
+predicate representation of a stencil is used to generate code
+for the \textsf{Halide} high-performance
+compiler~\citep{ragan2013halide}. Thus their approach must capture the
+full meaning of a stencil computation which requires significant
+program analysis. For example, they report that some degenerate
+stencil kernels take up to 17 hours to analyse and others require
+programmer intervention for correct invariants to be inferred.
 
@@
 
-CamFort differs from this significantly. The aim of CamFort is not a
-full representation of the stencil, but enough information to capture the access
-pattern. Hence, we do not require SMT solving or search of loop invariants. It
-suffices to do comparatively simple data-flow analysis.
-
-Due to the sheer amount of analysis involved, Kamil et al. report that some
-degenerate kernels take up to 17 hours to analyse and others require programmer
-intervention for correct invariants to be inferred. Since CamFort mostly does
-syntactic analysis, the inference procedure terminates quickly and never requires programmer intervention. The downside of our
-approach is that if optimisation causes the stencil computation to be heavily
-obfuscated, \textsf{STNG} would capture the access behaviour better.
-
-Most fundamentally, unlike CamFort, Kamil et al. does not provide a syntactic
-representation of their postcondition, so there is no way of future-proofing the
-code for future edits. Even if they were to provide a syntactic representation,
-for complex stencils such as Navier-Stokes from \cref{sec:intro}, it would be
-as verbose as the code itself making it difficult for programmer to understand
-the overall shape of the indexing behaviour.
+Our approach differs significantly. Rather than 
+full representation of stencils, we focus on specifying
+in a lightweight way just the spatial behaviour: a partial correctness
+technique. Thus, it suffices for us to perform a comparatively
+simple data-flow analysis which is efficient, scales linearly with
+code size, and does not require any user intervention.
+Whilst we do not perform deep semantic analysis of stencils,
+ the analysis part of our
+approach can be made arbitrarily more sophisticated independent of the rest of
+the work.
+% Hence, we do not require SMT solving or search of loop invariants. It
+%suffices to do comparatively simple data-flow analysis.
+%Since CamFort mostly does
+%syntactic analysis, the inference procedure terminates quickly and
+%never requires programmer intervention. 
+%The downside of our
+%approach is that if optimisation causes the stencil computation to be heavily
+%obfuscated, \textsf{STNG} would capture the access behaviour better.
+%
+Furthermore, Kamil \emph{et al.} do not provide a user-visible
+syntactic representation of their specifications, and they do
+not use their specifications for verification, \eg{}, to future-proof
+the code against later changes. Even if they were to provide a
+syntactic representation, for complex stencils such as Navier-Stokes
+from \cref{sec:intro}, it would be as verbose as the code itself,
+making it difficult for programmer to understand the overall shape of
+the indexing behaviour.
+
+% GPGPU
+Our work has some similarities with efforts to verify kernels
+written for General-Purpose Graphics Processing Unit (GPGPU)
+programming, such as in \citet{Blom:2014:SoCP}. %Stencils are a form of
+%kernel, and GPGPU programming can be viewed as a massively parallel
+%method of transforming a large matrix. 
+However, their focus is
+mainly on the synchronisation of kernels and the avoidance of
+data races, while we are interested in correctness 
+embedded within a more typical general-purpose programming
+language. %Other work on GPGPU computation, such as
+%\citet{Zhang:2012:CGO}, has focused primarily on generating
+%optimised code based on relatively simple specifications: to
+%provide performance while keeping the programmer's effort within
+%reason.
+
+% Sketching Stencils
+\citet{Solar-Lezama:2007:PLDI} give specifications of stencils using
+unoptimised ``reference'' stencils, coupled with partial
+implementations which are then completed by a code
+generation tool. %These kinds of specifications are just simple
+%implementations, so this tool is useful for hand-written, optimised
+%stencils. 
+The primary purpose of this tool is optimisation rather than
+correctness, and the language of specification is much different and
+more elaborate than in our approach.
+
+% Pochoir
+\citet{Tang:2011:SPAA} define a specification language for writing
+stencils embedded in C++ (with Cilk extensions) that are then compiled
+into fast parallel programs based on trapezoidal decompositions with
+hyperspace cuts. The Pochoir specification language is used for
+describing the kernel, boundary conditions and shape of the
+stencil. Programs are first compiled in C++ with a template library
+that checks whether the annotations are used correctly, and then a
+Pochoir preprocessor may be applied to generate the high-performance
+parallel versions using Cilk.  Pochoir
+is aimed primarily at assisting programmers who may be reluctant to
+implement the high-performance tricky cache-oblivious
+``hypertrapezoidal'' algorithms for stencils. 
+%that are significantly faster than the typical looping implementations
+%of stencil computations.
+Like much of the related work, the main goal 
+is optimisation rather than verifying program correctness.
+
+By contrast, the work of \citet{Abe:2013:IPDPSW} has looked at the
+correctness problem explicitly by bringing a form of model-checking to
+verify certain stencil computations. This is in the context of
+decomposing stencils for parallel computation on multiple processors
+ in partitioned global address
+space languages. %These are scenarios where an array is divided into
+%subarrays on multiple processors but each has the ability to access
+%the others' memory. The authors avoid the state explosion dilemma of
+%model-checking by relying upon the fact that accesses to the
+%``boundary elements'' will require a different method by virtue of the
+%non-locality of that memory. Since this access is performed in a
+%different manner, those can be identified automatically by static
+%analysis. To achieve this, 
+Abe \emph{et al.} provide a new language to
+be used for writing stencil computations. Much of the specification
+effort goes towards describing the nuts and bolts of distributing the
+computation over multiple processors. The code for the stencil kernel
+is generated from a relatively high-level specification. 
+In contrast, we are interested in integrating directly
+into existing, legacy codebases and established languages, bringing
+the benefits of verification more easily to scientific computing.%such as
+%older versions of Fortran.% We infer stencil specifications from
+%Fortran code and check annotations on stencil computations within
+%Fortran code.
 
 \section{Conclusion}\label{sec:conclusion}
-
-We have presented a high-level language for the specification of
-stencils in numerical code. As opposed to existing approaches our
-language takes inspiration from the numerical literature and provides
-a concise, abstract description of the stencil.
-
-We evaluated our approach on top of CamFort, an open-source analysis
-tool for Fortran programs, running the inference process over a corpus
-of a million lines of Fortran code. We showed that our language is
-capable of capturing many real-world uses of stencils and showed in a
-case-study that it will detect (simulated) programmer errors.
-
-Our definition of a stencil computation meant that we generated no
-specification for a large number of iterated array accesses in the
-corpus, such as stencils that incorporates reductions. For
-future work, we will be looking into expanding the flexibility of 
-our inference approach to more of these examples.
 

### 2016-11-16 paper.tex

@@
 always also bounded on both sides.
+
+\begin{figure}[t]\begin{minted}[fontsize=\scriptsize,breakindent=0em,linenos=false,xleftmargin=0em,breakafter=)]{fortran}
+!=stencil readOnce,(forward(depth=1,dim=3,nonpointed))*(backward(depth=1,dim=1))*(backward(depth=1,dim=2))+(forward(depth=1,dim=3))*(backward(depth=1,dim=1))*(pointed(dim=2))+(forward(depth=1,dim=3))*(backward(depth=1,dim=1,nonpointed))*(backward(depth=1,dim=2,nonpointed))+(forward(depth=1,dim=3))*(backward(depth=1,dim=2))*(pointed(dim=1))+(backward(depth=1,dim=1))*(backward(depth=1,dim=2,nonpointed))*(pointed(dim=3))+(backward(depth=1,dim=1,nonpointed))*(backward(depth=1,dim=2))*(pointed(dim=3))::x
+\end{minted}
+\caption{A complex specification inferred from
+  \textbf{Unified Model}\label{fig:smagorinsky}}
+\vspace{-1em}
+\end{figure}
+
+The single- and multi-action classes represent more complex stencils
+with a real possibility for programmer error. For example, the Unified
+Model has an implementation of the Smagorinsky subgrid-scale model for
+calculating turbulence on which our inference yields 39 specs from 340
+lines of code. This is an impressive reduction given
+the complexity of the algorithm.  We show one example in
+Figure~\cref{fig:smagorinsky}. This specifies the access pattern to
+a 3-dimensional array (which we have renamed to be called
+\texttt{x}). The array is accessed 8 times in the computation.
+
+\subsection{Detecting errors in the 2-D Jacobi iteration}
+
+One common example of a stencil computation is two-dimensional
+Jacobi iteration that repeatedly goes through each cell in a matrix
+and computes the average value of the four adjacent cells. The kernel
+can be expressed in a line of Fortran code:
+\begin{minted}{fortran}
+  a(i,j) = (a(i-1,j)+a(i+1,j)+a(i,j+1)+a(i,j-1))/4
+\end{minted}
+and CamFort is able to infer a precise specification for this:
+\begin{minted}[breakafter=+:,breakindent=-0.6em,breaksymbolsep=0.4em,linenos=false,xleftmargin=-0.5em]{fortran}
+  != stencil pointed(dim=1)*centered(depth=1,dim=2,nonpointed)+pointed(dim=2)*centered(depth=1,dim=1,nonpointed)::a
+\end{minted}
+
+We examined whether programmer errors would be detected by replacing
+the array index offsets with $-1$, $0$, or $1$ and running our
+verification algorithm. Camfort correctly reported a verification
+failure in each of 6,537 permutations corresponding to an error.  We
+note that the iteration is computing the average of four adjacent
+cells so $24$ ($4!$) of the possible array index pertubations do not
+correspond to a programming error.
+
+% /um/trunk/src/atmosphere/diffusion_and_filtering/turb_smagorinsky.F90
+% Tried introducing indexing errors replacing one of the indices of cx_rho_instance with -1,0,+1
+% 42 cases were caught
+% 6 cases were not caught
 

### 2016-11-16 paper.tex
Don't mention partial correctness. It means something else.
Reword part of Kamil discussion
Typo in nonpointed superscript in Eq. Theory.
add missing 'and'

@@
 where in each component one of $R$ or $S$ has the non-pointed
-attribute denoted by $R^{\term{np}}$ and $S^{\term{op}}$,
+attribute denoted by $R^{\term{np}}$ and $S^{\term{np}}$,
 the non-pointed attributes cancel to give $R \, \term{*} \, S$.
@@
 these equations). The $<:^r$ relation is congruent
-with respect to \term{+} and \term{*} $<:$ is congruent
+with respect to \term{+} and \term{*} and $<:$ is congruent
 with repect to \term{readOnce}, the equations of which we omit for brevity.
@@
 full representation of stencils, we focus on specifying
-in a lightweight way just the spatial behaviour: a partial correctness
-technique. Thus, it suffices for us to perform a comparatively
+just the spatial behaviour in a lightweight way.
+Thus, it suffices for us to perform a comparatively
 simple data-flow analysis which is efficient, scales linearly with

### 2016-11-16 paper.tex
various edits in section 3; reinstanting description of shorthand syntax

--- 2016-11-16 paper.tex
@@
 
-\item we provide an implementation of our approach as an extension to
+\item We provide an implementation of our approach as an extension to
   CamFort, an open-source program analysis tool for Fortran;
@@
 \nonterm{region} ::= & \nonterm{rvar} \; \\
-\multicolumn{2}{l}{\qquad\qquad \mid \; \stenRefl{\dimTy}} \\
-\multicolumn{2}{l}{\qquad\qquad \mid \;\stenFwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{nonpointed}]}} \\
-\multicolumn{2}{l}{\qquad\qquad \mid \; \stenBwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{nonpointed}]}} \\
-\multicolumn{2}{l}{\qquad\qquad \mid \; \stenCen{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{nonpointed}]}} \\
-\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{region} \, \term{+}
+\multicolumn{2}{l}{\qquad \mid \; \stenRefl{\dimTy}} \\
+\multicolumn{2}{l}{\qquad \mid \;\stenFwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{nonpointed}]}} \\
+\multicolumn{2}{l}{\qquad \mid \; \stenBwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{nonpointed}]}} \\
+\multicolumn{2}{l}{\qquad \mid \; \stenCen{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{nonpointed}]}} \\
+\multicolumn{2}{l}{\qquad \mid \; \nonterm{region} \, \term{+}
   \, \nonterm{region} \; \mid \; \nonterm{region} \; \term{*} \; \nonterm{region}} \\[0.25em]
@@
 declaration}.
-
 Regions comprise \emph{region constants} which are
@@
 following shows four specifications with four consistent stencil
-kernels reading from arrays \term{a}, \term{b}, \term{c} and \term{d} respectively:
+kernels reading from arrays \term{a}, \term{b}, \term{c} and \term{d}:
 %%
@@
 in the spcified code must be consistent with either of $r$ or $s$.
-Dually, for every component in the region $r$ there must be an index
-within it and every component in the region $s$ must have an index
-within it. For
+Dually, every part of region $r$ must have a
+corresponding index and similarly for $s$ independently. For
 example, the following gives the specification of a five-point
@@
 where there is a non-contiguous stencil pattern, which cannot be expressed
-precisely in our specification syntax. For example:
+precisely in our syntax. For example:
 %
@@
   representing both constant indices and unspecified indices.
-  For an index scheme $u$, $u_i$ denotes the $i^{th}$ element of $u$. We
+  For an index scheme $u$, the $i^{th}$ element is denoted $u_i$. We
   write $(\domainVal)^n$ for the set of all $n$-dimensional schemes.
@@
 The relationship between source language
-terms and indexing terms
+terms and index schemes
  is given precisely by the following function \textsf{schematic},
@@
 Let $\schemeOne$ be a partial function, parameterised by a set of
-induction variables $I$, mapping expression syntax to
+induction variables $I$, mapping source langauge expressions to
 $\domainVal$ values:
@@
 -c & e \equiv i - c \;\; \wedge \;\; i \in I \\
-\infty & \textit{$I \cap \textit{FreeVariables}(e) = \emptyset $} \\
-\bot   & \textit{otherwise}
+\infty & \textit{$I \cap \textit{FreeVariables}(e) = \emptyset $} %\\
+%\bot   & \textit{otherwise}
 \end{cases}
@@
 is mapped to $\infty \in \domainVal{}$. Otherwise, $\schemeOne(e)$
-is undefined (returns $\bot$).
+is undefined ($\bot$).
 
 The partial function $\schemeN$, also parameterised by a set of
-induction variables $I$, uses the above operation $\schemeOne$ to
- map from the syntax of indices to index schemes, \ie{} values of
-$(\domainVal)^n$. It is defined on the syntax of indices $\bar{e} = (e_1, ..., e_n)$ as:
-\begin{align*}
-\schemeN(e_1, ..., e_n) & =
+induction variables $I$, uses $\schemeOne$ to
+ map from the syntax of indices to index schemes
+$\in (\domainVal)^n$. It is defined on the syntax of indices $\bar{e} = (e_1, ..., e_n)$ as:
+\begin{align*}
+\schemeN(e_1, .., e_n) & =
 [\schemeOne(e_1) ... \schemeOne(e_n) ]
-\quad \textit{iff} \;\; \forall x . \schemeOne(e_x) \neq \bot
+\; \textit{iff} \; \forall x . \schemeOne(e_x) \neq \bot
 \end{align*}
@@
 Thus, if each expression in an index is a neighbourhood or constant
-index, then $\schemeN$ maps it to an indexing scheme.
+index, $\schemeN$ maps it to an indexing scheme.
 \end{definition}
@@
 (or conversely $v$ \emph{specialises} $w$),
-written $v \preceq_n w$, iff the following holds:
-%%
+written $v \preceq_n w$ defined:
 \[
@@
 %%
-Two schemes are related by the (reflexive and transitive) binary relation
-$\preceq_n$ over $(\mathbb{Z}_{\infty})^n \times (\mathbb{Z}_{\infty})^n$.
-The element $\infty$ generalises any index in the same dimension.
+where $\preceq_n$ is a reflexive, transitive relation on
+$(\mathbb{Z}_{\infty})^n \times (\mathbb{Z}_{\infty})^n$. Thus, the
+element $\infty$ generalises any index in the same dimension.
 %
@@
 
-Our terminology of \emph{indexing schemes} and
+Our terminology of \emph{index schemes} and
 \emph{generalisation} is by analogy
@@
 polymorphic types in ML~\cite{milner1978theory}. Our notion is
-much more rigid however, as our schemes do not bind names.
+much more rigid however, as index schemes do not bind names.
+
+\paragraph{Shorthand notation}
+
+We use abbreviations
+ \term{fwd}, \term{bwd}, \term{cen}, and \irreflS{}
+for \term{forward}, \term{backward},
+\term{centered}, and \irrefl{} respectively. Furthermore, 
+we quantify over the absence or presence of the \irrefl{}
+attribute on region constants as a boolean. For example,
+\term{fwd} is syntactic sugar defined:
+%%
+\begin{align*}
+\hspace{-0.5em}
+& \stenFwdSR{n}{d}{p}
+\; := \; \\[-0.4em]
+& \quad \begin{cases}
+\stenFwd{n}{d}{, \irrefl} & \textit{if} \;\, \neg \, p \\[-0.1em]
+\stenFwd{n}{d}{}  & \textit{if} \;\,  p
+\end{cases}
+\end{align*}
+%
+$\textit{or} \;\; \stenFwdS{n}{d} := \stenFwd{n}{d}{}$
+and similarly for \term{bwd} and \term{cen}.
 
@@
 
-The model is defined by the interpretation function $\interp{-}_n$
+An interpretation function $\interp{-}_n$ provides the model 
 (where $n$ is the maximum dimensionality of the specification being
@@
 of a specification \textit{spec} in \Cref{fig:spatial-model}.
-
-We first define some intermediate notations and definitions.
+Various intermediate notions are used.
 
@@
 \vtwoh{1}{\infty}$ and $\singleEntry{2}{3} = \vthreeh{\infty}{1}{\infty}$.
-Thus, a single-entry vector is an indexing scheme which generalises
+Thus, a single-entry vector is an index scheme generalising 
 all other schemes that have $1$ as their $r^{th}$ entry,
@@
 are $\infty$, \eg{}, $\zeroEntry{1}{2} = \vtwoh{0}{\infty}$.
-Similarly to the single-entry vector, zeroed single-entry vectors
+Similarly to single-entry vectors, zeroed single-entry vectors
 generalise all other schemes
@@
 The first four equations of \Cref{fig:region-model}
-give the model of the region constants as sets of indexing
+model region constants as sets of indexing
 schemes. We illustrate with an example.
@@
 as approximation $<:$ which we define here.
-For brevity here we use abbreviations $\term{fwd}$, $\term{bwd}$, and
-$\term{cen}$ for region constants.
+%For brevity here we use abbreviations $\term{fwd}$, $\term{bwd}$, and
+%$\term{cen}$ for region constants.
 

### 2016-11-16 paper.tex
See the supplement for the proof

@@
 \end{align*}
-Note the absorbing behaviour of $\infty$ under multiplication above, \eg{} $2
-\times \infty = \infty$
+Note, $\infty$ for multiplication above, \eg{} $2
+ \infty = \infty$.
 \end{example}
@@
 of $\term{+}$ and $\term{*}$. For $\term{+}$ this is the
-straightforward union of the models of subterms. The model of
-$\term{*}$ takes the tensor $\otimes$ of subterm models. The tensor is
+ union of the models of subterms. The model of
+$\term{*}$ takes a ``tensor'' $\otimes$ of subterm models. The tensor is
 more involved. We give some informal intuition first.
@@
 \begin{figure}
+\vspace{-1em}
 \input{spatial-model.tex}
-\caption{\textit{spec} model,
+\caption{%\textit{spec} model,
 $\interp{-}_n : \textit{spec} \rightarrow \textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$}
@@
 The intuition behind $\otimes$ is that it takes all possible pairs of
-indexing schemes, and treats each pair as lower and upper bounds of an
-$n$-dimensional rectangle. From these lower and upper bounds,
+index schemes, and treats each pair as lower and upper bounds of an
+$n$-dimensional rectangle. From these bounds,
 the remaining vertices of the $n$-dimensional rectangle are
@@
 \begin{align*}
-\{\vtwohs{1}{2}, \vtwohs{5}{6}\,\} \otimes \{\vtwohs{3}{4}\} =
-\{\vtwohs{1}{2}, \vtwohs{3}{2}, \vtwohs{1}{4}, \vtwohs{3}{4},
-   \vtwohs{3}{6}, \vtwohs{5}{4}, \vtwohs{5}{6}\,\}
+{\small{
+\{\vtwohs{1}{2},\vtwohs{5}{6}\,\}\!\otimes\!\{\vtwohs{3}{4}\}\!=\!\{\vtwohs{1}{2},\vtwohs{3}{2},\vtwohs{1}{4},\vtwohs{3}{4},
+   \vtwohs{3}{6},\vtwohs{5}{4},\vtwohs{5}{6}\,\}
+}}
 \end{align*}
@@
 vectors. Some care is then taken over how this interacts with $\infty$.
-In the presence of $\infty$, generate all interlavings as if $\infty$ is
+In the presence of $\infty$, generate all interleavings as if $\infty$ is
 an integer, then filter out vectors with $\infty$ at any coordinate, \eg{}
@@
 For an $n$-dimensional model $M$, its \emph{constrained dimensions}
-is the set of dimension identifiers for which at least one scheme $u \in M$ has
-a non-$\infty$ value in that dimension. That is:
+are the set of dimension identifiers where at least one scheme $u \in M$ has
+a non-$\infty$ value in that dimension:
 %
@@
 \begin{definition}The tensor $\otimes_n$ of $n$-dimensional
-models is defined:
+models is:
 \begin{alignat*}{2}
@@
 The $2^n$ unique choices for $\bowtie$ on $n$-vectors
-corresponds to taking all bit-strings of length $n$ and
+correspond to taking all bit-strings of length $n$ and
 selecting from $u$ for $1$ and $v$ for $0$. The pairwise permutation
@@
 and \textsf{Approx} are parametric labelled variant types
-with injection function given by their definition:
+with injections given by their definition:
 %
@@
 \end{equation*}
+(See the supplement for the proof)
 \end{theorem}
@@
 where $\consAName$ implements consistency in checking (\cref{subsec:checking}).
+(See the supplement for the proof)
 \end{theorem}

### 2016-11-16 paper.tex
more section 3 edits
Give the game away early in equational theory

@@
 
-Specifications have a theory of equality $\equiv$ as well
-as approximation $<:$ which we define here.
-%For brevity here we use abbreviations $\term{fwd}$, $\term{bwd}$, and
-%$\term{cen}$ for region constants.
+Specifications have a theory of equality, $\equiv$, providing axiomatic
+semantics for the language. This enables users to rewrite their specification
+and helps the inference procedure (\Cref{subsec:inference}) simplify
+specifications. Additionally, specifications have a theory of approximation,
+$<:$, that can be used by the users to weaken specifications for succintness in
+the source.
 
@@
 their regions.
-
-The equational theory provides not only an axiomatic semantics
-for the language, which the user can use to rewrite their
-specification, but it is also used extensively in the inference
-procedure (\cref{subsec:inference}) to simplify specifications.
 

### 2016-11-16 paper.tex
a few small edits in 5

--- 2016-11-16 paper.tex
@@
 \end{equation*}
-(See the supplement for the proof)
 \end{theorem}
@@
 where $\consAName$ implements consistency in checking (\cref{subsec:checking}).
-(See the supplement for the proof)
 \end{theorem}
@@
 \begin{figure}[t]
-\vspace{-0.8em}
+\vspace{-1em}
 \hspace{-0.5em}
@@
 \newcommand{\lupper}{u}
-{\scalebox{0.97}{
+{\scalebox{0.95}{
 \begin{minipage}{1\linewidth}
@@
 %% CEN
-\exact{\stenCenS{\lupper}{d}}& \llower < 0 \wedge \lupper > 0 \wedge |\llower| = \lupper \\[0.02em]
-\exactR(\stenBwdS{|\llower|}{d} & \llower < 0 \wedge \lupper > 0  \wedge |\llower| \neq \lupper \\
+\exact{\stenCenS{\lupper}{d}}& \!\!\!\llower < 0 \wedge \lupper > 0 \wedge |\llower| = \lupper \\[0.02em]
+\exactR(\stenBwdS{|\llower|}{d} & \!\!\!\llower < 0 \wedge \lupper > 0  \wedge |\llower| \neq \lupper \\
 \qquad\qquad \texttt{+} \, \stenFwdS{\lupper}{d}) & \\[0.02em]
@@
 in \Cref{fig:to-spec}, where the lower and upper bounds
-determine region constants; $\mathsf{toSpec} \, d \, (l, u)$ is always defined
+determine region constants; $\mathsf{toSpec} \, d \, (l, u)$ is defined
  if $l$ less than or equal to $u$.

### 2016-11-16 paper.tex

@@
 
-The equational theory provides not only an axiomatic semantics
-for the language, which the user can use to rewrite their
-specification, but it is also used extensively in the inference
-procedure (\cref{subsec:inference}) to simplify specifications.
-
 \begin{theorem}[Soundness of equational theory]

### 2016-11-16 paper.tex
section names and bib plain and some edits

@@
-\documentclass[10pt,preprint]{sigplanconf}
+\documentclass[10pt,preprint,numbers]{sigplanconf}
 
@@
 \usepackage{xypic}
-\usepackage{natbib}
+%\usepackage{natbib}
 
@@
 \Crefname{section}{\S}{\S\S}
-\crefname{subsection}{\S\S}{\S\S\S}
-\Crefname{subsection}{\S\S}{\S\S\S}
+\crefname{subsection}{\S}{\S\S\S}
+\Crefname{subsection}{\S}{\S\S\S}
 \crefname{figure}{fig.}{}
@@
 
-\bibliographystyle{abbrvnat}
+\bibliographystyle{plain}
 
@@
 \begin{definition}
-A \emph{zeroed single-entry vector} of size $n$, denoted
-$\zeroEntry{r}{n}$, is a vector where the $r^{th}$ entry is $0$ and all others
-are $\infty$, \eg{}, $\zeroEntry{1}{2} = \vtwoh{0}{\infty}$.
-Similarly to single-entry vectors, zeroed single-entry vectors
+A \emph{zeroed single-entry vector} of size $n$ denoted
+$\zeroEntry{r}{n}$ is a vector where the $r^{th}$ entry is $0$ and all others
+are $\infty$ \eg{}, $\zeroEntry{1}{2} = \vtwoh{0}{\infty}$.
+Akin to single-entry vectors, zeroed single-entry vectors
 generalise all other schemes
-that have $0$ as their $r^{th}$, representing indices at
-the ``origin'' in dimension $r$.
+that have $0$ as their $r^{th}$ entry, representing indices at
+the ``origin'' in $r$.
 \end{definition}
@@
 semantics for the language. This enables users to rewrite their specification
-and helps the inference procedure (\Cref{subsec:inference}) simplify
-specifications. Additionally, specifications have a theory of approximation,
+and provides a way to automatically simplify
+specifications after inference (\Cref{subsec:inference}).
+Additionally, specifications have a theory of approximation,
 $<:$, that can be used by the users to weaken specifications for succintness in

### 2016-11-16 paper.tex
various edits
removing DNF grammar
savin' space

@@
 \usepackage{xypic}
-%\usepackage{natbib}
+\usepackage{natbib}
 
@@
 where $\preceq_n$ is a reflexive, transitive relation on
-$(\mathbb{Z}_{\infty})^n \times (\mathbb{Z}_{\infty})^n$. Thus, the
-element $\infty$ generalises any index in the same dimension.
+index schemes $\domainVal^n$. Thus,
+ $\infty$ generalises any index in the same dimension.
 %
@@
 \schemeN{(1, j+1)} & = \vtwoh{\infty}{1} \\
-\schemeN{(3, j)} & = \vtwoh{\infty}{0} \\
 \schemeN{(i*2, j)} & = \bot
@@
 The notion of which dimensions $i$ have \emph{any scheme in the model
-where $u_i \neq \infty$} is represented by the \emph{constrained
-dimensions} intermediate function:
+where $u_i \neq \infty$} is represented by following  function:
 \begin{definition}%[Constrained dimensions]
@@
   \term{dim=}#2)}}
+%
+\vspace{-0.2em}
 \begin{align*}
@@
 
-\item (\textit{overlapping pointed}) \term{pointed} regions coalesce with other regions of the
-same dimension:
-%
+\item (\textit{overlapping pointed}) \term{pointed} regions coalesce
+  with other regions of the same dimension:
+%
+\vspace{-0.4em}
 \begin{align*}
@@
 \end{align*}
-%
+\vspace{-1.5em}
+
 \item \textit{(centered)} a forward and backward region of the same
   depth and dimension coalseces into a centered region:
+%
+\vspace{-0.4em}
 \begin{align*}
@@
 \end{align*}
-%%
+\vspace{-1.5em}
+
 \item $(\textsc{point}) \;\; (R \, \term{*} \, S^\irreflS) \, \term{+} \,
@@
 and for infering specifications from code (\Cref{subsec:inference}).
-Both rely on a program analysis that converts array indexing
-expressions into sets of index schemes. We outline this analysis
-first (\cref{subsec:analysis}). Note that the analysis procedure
+Both rely on a program analysis that converts array subscripts
+ into sets of index schemes. We outline this analysis
+first (\cref{subsec:analysis}). Note that the analysis 
 can be made arbitrarily more complex and wide-ranging independent
@@
 \begin{figure}[t]
-\vspace{-1em}
+\vspace{-0.5em}
 \begin{align*}
@@
 $.
-
-More formally, $\specDNF \subseteq (\textit{spec} \cup \{\epsilon\})$, defined:
-%%
-\begin{align*}
-\specDNF{} & ::= \tySum{\specDNF} \mid \epsilon \quad\;\;
-\tySum{\specDNF} ::= \tyProd{\specDNF} \; \term{+} \; \tySum{\specDNF} \mid
-  \tyProd{\specDNF} \\
-\tyProd{\specDNF} & ::= \textit{region} \; \term{*} \; \tyProd{\specDNF} \mid
-   \textit{region}
-\end{align*}
+%
+%More formally, $\specDNF \subseteq (\textit{spec} \cup \{\epsilon\})$, defined:
+%%%
+%\begin{align*}
+%\specDNF{} & ::= \tySum{\specDNF} \mid \epsilon \quad\;\;
+%\tySum{\specDNF} ::= \tyProd{\specDNF} \; \term{+} \; \tySum{\specDNF} \mid
+%  \tyProd{\specDNF} \\
+%\tyProd{\specDNF} & ::= \textit{region} \; \term{*} \; \tyProd{\specDNF} \mid
+%   \textit{region}
+%\end{align*}
 %%
 In the following, $S, T$ range over elements of
-\specDNF{} and $\tyProd{s}, \tyProd{t}$ range over elements
-of $\tyProd{\specDNF}$, \ie{} $\tyProd{s} = r^1_{1} \; \term{*}
+\specDNF{} and $\tyProd{s}, \tyProd{t}$ range over conjuncts
+ \eg{} $\tyProd{s} = r^1_{1} \; \term{*}
 \ldots \term{*} \; r^1_n$.
@@
 is indeed in DNF-form, where $\hat{\odot}$ is essentially applying
- the \trule{\textsc{dist}} rule of the equational theory
- (\Cref{fig:equations}). The operation is
+ the \trule{\textsc{dist}} rule of the equational theory. The operation is
 commutative and associative, and sound with respect to our
@@
 \section{Evaluation}
+\label{sec:evaluation}
 

### 2016-11-16 paper.tex
more space saving

@@
 \begin{definition}[Generalisation]
+\label{def:generalisation}
 Given two schemes $v, w$ of dimensionality $n$,
@@
 The \textit{analyse} function matches on
-line 4, with the following set for $\textsf{flowsTo}(\texttt{a(i, j) + x +
-  y})$:
-%
-\begin{align*}
-{\small{
-\{\texttt{a(i-1, j)}, \texttt{a(i+1, j)}, \texttt{a(i, j-1)},
-  \texttt{a(i, j+1)}, \texttt{a(i, j)}\}}}
-\end{align*}
-The \emph{modelise} function then yields the following
-since all multiplicities in the model are one:
+line 4 with the following set for $\textsf{flowsTo}(\texttt{a(i, j) + x +
+  y}) = $
+%
+%\begin{align*}
+{{ %{\small{
+$\{$\texttt{a(i-1, j)}, \texttt{a(i+1, j)}, \texttt{a(i, j-1)},
+  \texttt{a(i, j+1)}, \texttt{a(i, j)}$\}$}}. 
+%\end{align*}
+Thus, \emph{modelise} returns the following
+as all multiplicities are one:
 %%
@@
 %
-\begin{align*}
-\vspace{-0.3em}
+\vspace{-0.4em}
+\begin{align*}
  N_s= \{  \vtwo{-1}{\infty},
@@
   \textsf{once}(\textsf{exact} (N_s))
-\vspace{-0.3em}
-\end{align*}
-%
+\end{align*}
+\vspace{-1.5em}
+
 \item Generate code model $\textit{modelise}(l3, \texttt{b}) =
@@
 %
-\begin{align*}
-\vspace{-0.3em}
+\vspace{-0.7em}
+\begin{align*}
 M_c = & \{
@@
   \}
-\vspace{-0.3em}
-\end{align*}
+\end{align*}
+\vspace{-1em}
 %

### 2016-11-16 paper.tex
Typographic mistake

@@
 indexing seen in the code and would simply be prone to low-level input
-errors. Our approach is much more abstract,
-capturing the shape of a specification in concise format different to
-the code.
+errors. Our approach is much more abstract-- it does
+not aim to reify indexing in the specification, but is novel
+in that it instead
+provides abstract spatial descriptions which capture a large number
+of common patterns.
 
@@
 predicate representation of a stencil is used to generate code
-for the \textsf{Halide} high-performance
+for the \textsc{Halide} high-performance
 compiler~\citep{ragan2013halide}. Thus their approach must capture the
@@
 
-\section{Conclusion}\label{sec:conclusion}
-
-We have presented a high-level language for the specification of
-stencils in numerical code. As opposed to existing approaches our
-language takes inspiration from the numerical literature and provides
-a concise, abstract description of the stencil.
+\paragraph{Concluding remarks}
+
+There is an increasing awareness of the need for verification
+techniques in
+science~\cite{post2005computational,oberkampf2010verification,orchard2014computational}.
+Our specification language is an approach in this direction, providing
+a system of lightweight, high-level specification for numerical code.
+As opposed to existing approaches our language takes inspiration from
+the numerical literature and provides a concise, abstract description
+of the stencil.
 
@@
 future work, we will be looking into expanding the flexibility of 
-our inference approach to more of these examples.
+our analysis to include these examples.
 

### 2016-11-16 paper.tex
conclusion edit

@@
 predicate representation of a stencil is used to generate code
-for the \textsc{Halide} high-performance
+for the \textsf{Halide} high-performance
 compiler~\citep{ragan2013halide}. Thus their approach must capture the

### 2016-11-16 paper.tex
Restore origin of region names citation
Cite Cilk and solve author? problem by switching style to plainnat

--- 2016-11-16 paper.tex
@@
 
-\bibliographystyle{plain}
+\bibliographystyle{plainnat}
 
@@
 single dimension relative to the origin and are either \term{pointed},
-\term{forward}, \term{backward}, or \term{centered}.  Each region
+\term{forward}, \term{backward}, or \term{centered}. The region names are
+inspired by the scientific terminology, \eg{} the standard explicit method for
+approximating PDEs is known as the \emph{Forward Time, Centered Space} (FTCS)
+scheme~\citep{dawson1991finite}. Each region
 constant has a dimension identifier $d$ given as a positive natural number.
@@
 \citet{Tang:2011:SPAA} define a specification language for writing
-stencils embedded in C++ (with Cilk extensions) that are then compiled
+stencils embedded in C++ (with Cilk~\citep{blumofe1996cilk} extensions) that
+are then compiled
 into fast parallel programs based on trapezoidal decompositions with

### 2016-11-16 paper.tex
missing piece at end of inference

@@
 \hspace{-0.5em}
+{\scalebox{0.96}{
 \begin{minipage}{0.16\linewidth}
@@
 \end{align*}
-\end{minipage}
+\end{minipage}}}
 \caption{Illustration of grouping spans in
@@
 $\containedin$; if any region is contained within another then remove the
-  smaller, defined:
+  smaller:
 %
@@
 \noindent
-Thus our model validates the specification synthesis process which concludes the
-inference procedure. For our example, we infer and then synthesise the
+Thus, our model validates specification synthesis.
+%
+\subsection{Step 5: Simplification}
+The final step applies a simplification pass, orienting the
+equational theory $\equiv$ (\Cref{sec:eqs}) into rewrite rules
+based on some simple heuristics.
+
+For our example, we infer and then synthesise the
 5-point specification:

### 2016-11-16 paper.tex
added list of corpus packages

--- 2016-11-16 paper.tex
@@
 \begin{definition}[Generalisation]
-\label{def:generalisation}
 Given two schemes $v, w$ of dimensionality $n$,
@@
 where $\preceq_n$ is a reflexive, transitive relation on
-index schemes $\domainVal^n$. Thus,
- $\infty$ generalises any index in the same dimension.
+$(\mathbb{Z}_{\infty})^n \times (\mathbb{Z}_{\infty})^n$. Thus, the
+element $\infty$ generalises any index in the same dimension.
 %
@@
 \schemeN{(1, j+1)} & = \vtwoh{\infty}{1} \\
+\schemeN{(3, j)} & = \vtwoh{\infty}{0} \\
 \schemeN{(i*2, j)} & = \bot
@@
 The notion of which dimensions $i$ have \emph{any scheme in the model
-where $u_i \neq \infty$} is represented by following  function:
+where $u_i \neq \infty$} is represented by the \emph{constrained
+dimensions} intermediate function:
 \begin{definition}%[Constrained dimensions]
@@
 and for infering specifications from code (\Cref{subsec:inference}).
-Both rely on a program analysis that converts array subscripts
- into sets of index schemes. We outline this analysis
-first (\cref{subsec:analysis}). Note that the analysis 
+Both rely on a program analysis that converts array indexing
+expressions into sets of index schemes. We outline this analysis
+first (\cref{subsec:analysis}). Note that the analysis procedure
 can be made arbitrarily more complex and wide-ranging independent
@@
 The \textit{analyse} function matches on
-line 4 with the following set for $\textsf{flowsTo}(\texttt{a(i, j) + x +
-  y}) = $
-%
-%\begin{align*}
-{{ %{\small{
-$\{$\texttt{a(i-1, j)}, \texttt{a(i+1, j)}, \texttt{a(i, j-1)},
-  \texttt{a(i, j+1)}, \texttt{a(i, j)}$\}$}}. 
-%\end{align*}
-Thus, \emph{modelise} returns the following
-as all multiplicities are one:
+line 4, with the following set for $\textsf{flowsTo}(\texttt{a(i, j) + x +
+  y})$:
+%
+\begin{align*}
+{\small{
+\{\texttt{a(i-1, j)}, \texttt{a(i+1, j)}, \texttt{a(i, j-1)},
+  \texttt{a(i, j+1)}, \texttt{a(i, j)}\}}}
+\end{align*}
+The \emph{modelise} function then yields the following
+since all multiplicities in the model are one:
 %%
@@
 %
-\vspace{-0.4em}
-\begin{align*}
+\begin{align*}
+\vspace{-0.3em}
  N_s= \{  \vtwo{-1}{\infty},
@@
   \textsf{once}(\textsf{exact} (N_s))
-\end{align*}
-\vspace{-1.5em}
-
+\vspace{-0.3em}
+\end{align*}
+%
 \item Generate code model $\textit{modelise}(l3, \texttt{b}) =
@@
 %
-\vspace{-0.7em}
-\begin{align*}
+\begin{align*}
+\vspace{-0.3em}
 M_c = & \{
@@
   \}
-\end{align*}
-\vspace{-1em}
+\vspace{-0.3em}
+\end{align*}
 %
@@
 \section{Evaluation}
-\label{sec:evaluation}
 
@@
 % eigenvalue problems - https://github.com/opencollab/arpack-ng
+
+We built a corpus of around 1 million lines of Fortran code from a
+range of scientific computing packages: The Unified Model~\cite{um},
+E3MG~\cite{RePEc:aen:journl:2006se-a12}, BLAS~\cite{blas},
+Hybrid4~\cite{GBC:GBC635}, GEOS-Chem~\cite{geos-chem}, Navier (based
+on \cite{griebel1997numerical}), Computational Physics
+2~\cite{nicholas2006computational},
+ARPACK-NG\footnote{\url{https://github.com/opencollab/arpack-ng}},
+SPECFEM3D\footnote{\url{https://github.com/geodynamics/specfem3d}}.
 
@@
 indexing seen in the code and would simply be prone to low-level input
-errors. Our approach is much more abstract-- it does
-not aim to reify indexing in the specification, but is novel
-in that it instead
-provides abstract spatial descriptions which capture a large number
-of common patterns.
+errors. Our approach is much more abstract,
+capturing the shape of a specification in concise format different to
+the code.
 
@@
 
-\paragraph{Concluding remarks}
-
-There is an increasing awareness of the need for verification
-techniques in
-science~\cite{post2005computational,oberkampf2010verification,orchard2014computational}.
-Our specification language is an approach in this direction, providing
-a system of lightweight, high-level specification for numerical code.
-As opposed to existing approaches our language takes inspiration from
-the numerical literature and provides a concise, abstract description
-of the stencil.
+\section{Conclusion}\label{sec:conclusion}
+
+We have presented a high-level language for the specification of
+stencils in numerical code. As opposed to existing approaches our
+language takes inspiration from the numerical literature and provides
+a concise, abstract description of the stencil.
 
@@
 future work, we will be looking into expanding the flexibility of 
-our analysis to include these examples.
+our inference approach to more of these examples.
 

### 2016-11-16 paper.tex

--- 2016-11-16 paper.tex
@@
 
+We built a corpus of around 1 million lines of Fortran code from a
+range of scientific computing packages: The Unified Model~\cite{um},
+E3MG~\cite{RePEc:aen:journl:2006se-a12}, BLAS~\cite{blas},
+Hybrid4~\cite{GBC:GBC635}, GEOS-Chem~\cite{geos-chem}, Navier (based
+on \cite{griebel1997numerical}), Computational Physics
+2~\cite{nicholas2006computational},
+ARPACK-NG\footnote{\url{https://github.com/opencollab/arpack-ng}},
+SPECFEM3D\footnote{\url{https://github.com/geodynamics/specfem3d}}.
+
 We first examined how frequently stencil computations occur in our

### 2016-11-16 paper.tex
edit on evaluation

--- 2016-11-16 paper.tex
@@
 \end{minted}
-\caption{A complex specification inferred from
-  \textbf{Unified Model}\label{fig:smagorinsky}}
+\caption{Complex specification inferred from
+  \textbf{UM}\label{fig:smagorinsky}}
 \vspace{-1em}
@@
 lines of code. This is an impressive reduction given
-the complexity of the algorithm.  We show one example in
-Figure~\cref{fig:smagorinsky}. This specifies the access pattern to
+the complexity of the algorithm.  We show one example in~\cref{fig:smagorinsky}. This specifies the access pattern to
 a 3-dimensional array (which we have renamed to be called
@@
 and computes the average value of the four adjacent cells. The kernel
-can be expressed in a line of Fortran code:
+is given by:
 \begin{minted}{fortran}
@@
 \end{minted}
-and CamFort is able to infer a precise specification for this:
+and we infer a precise specification for this, as:
 \begin{minted}[breakafter=+:,breakindent=-0.6em,breaksymbolsep=0.4em,linenos=false,xleftmargin=-0.5em]{fortran}
@@
 \end{minted}
-
-We examined whether programmer errors would be detected by replacing
+%
+To test our implementation, 
+we examined whether programmer errors would be detected by replacing
 the array index offsets with $-1$, $0$, or $1$ and running our
-verification algorithm. Camfort correctly reported a verification
-failure in each of 6,537 permutations corresponding to an error.  We
-note that the iteration is computing the average of four adjacent
-cells so $24$ ($4!$) of the possible array index pertubations do not
-correspond to a programming error.
+verification algorithm. Our checking procedure correctly reported a verification
+failure in each of 6,537 permutations corresponding to an error.  
+The iteration computes the average of four adjacent
+cells so $24$ ($4$ factorial) of the possible array index pertubations
+are correct, all of which are accepted by checking.
 

### 2016-11-16 paper.tex
edits to last section

@@
 however, such a specification would exactly replicate the fine-grained
-indexing seen in the code and would simply be prone to low-level input
+indexing seen in the code and would be similarly prone to low-level input
 errors. Our approach is much more abstract-- it does
@@
 for the \textsc{Halide} high-performance
-compiler~\citep{ragan2013halide}. Thus their approach must capture the
+compiler~\citep{ragan2013halide}. Thus they must capture the
 full meaning of a stencil computation which requires significant
@@
 Furthermore, Kamil \emph{et al.} do not provide a user-visible
-syntactic representation of their specifications, and they do
-not use their specifications for verification, \eg{}, to future-proof
+syntactic representation of their specifications, and nor do
+they provide verification from specifications \eg{}, to future-proof
 the code against later changes. Even if they were to provide a
@@
 %analysis. To achieve this, 
-Abe \emph{et al.} provide a new language to
-be used for writing stencil computations. Much of the specification
-effort goes towards describing the nuts and bolts of distributing the
+Abe \emph{et al.} provide a new language for writing stencil computations. Much of the specification
+effort goes towards describing the distribution of the
 computation over multiple processors. The code for the stencil kernel

### 2016-11-16 paper.tex
moved footnotes to cites

--- 2016-11-16 paper.tex
@@
 
+We built a corpus of around 1 million lines of Fortran code from a
+range of scientific computing packages: The Unified Model~\cite{um},
+E3MG~\cite{RePEc:aen:journl:2006se-a12}, BLAS~\cite{blas},
+Hybrid4~\cite{GBC:GBC635}, GEOS-Chem~\cite{geos-chem}, Navier (based
+on \cite{griebel1997numerical}), Computational Physics
+2~\cite{nicholas2006computational},
+ARPACK-NG~\cite{arpackng}, and
+SPECFEM3D~\cite{specfem3d}.
+
 We first examined how frequently stencil computations occur in our
@@
 \end{minted}
-\caption{Complex specification inferred from
-  \textbf{UM}\label{fig:smagorinsky}}
+\caption{A complex specification inferred from
+  \textbf{Unified Model}\label{fig:smagorinsky}}
 \vspace{-1em}
@@
 lines of code. This is an impressive reduction given
-the complexity of the algorithm.  We show one example in~\cref{fig:smagorinsky}. This specifies the access pattern to
+the complexity of the algorithm.  We show one example in
+Figure~\cref{fig:smagorinsky}. This specifies the access pattern to
 a 3-dimensional array (which we have renamed to be called
@@
 and computes the average value of the four adjacent cells. The kernel
-is given by:
+can be expressed in a line of Fortran code:
 \begin{minted}{fortran}
@@
 \end{minted}
-and we infer a precise specification for this, as:
+and CamFort is able to infer a precise specification for this:
 \begin{minted}[breakafter=+:,breakindent=-0.6em,breaksymbolsep=0.4em,linenos=false,xleftmargin=-0.5em]{fortran}
@@
 \end{minted}
-%
-To test our implementation, 
-we examined whether programmer errors would be detected by replacing
+
+We examined whether programmer errors would be detected by replacing
 the array index offsets with $-1$, $0$, or $1$ and running our
-verification algorithm. Our checking procedure correctly reported a verification
-failure in each of 6,537 permutations corresponding to an error.  
-The iteration computes the average of four adjacent
-cells so $24$ ($4$ factorial) of the possible array index pertubations
-are correct, all of which are accepted by checking.
+verification algorithm. Camfort correctly reported a verification
+failure in each of 6,537 permutations corresponding to an error.  We
+note that the iteration is computing the average of four adjacent
+cells so $24$ ($4!$) of the possible array index pertubations do not
+correspond to a programming error.
 

### 2016-11-16 paper.tex

--- 2016-11-16 paper.tex
@@
 2~\cite{nicholas2006computational},
-ARPACK-NG\footnote{\url{https://github.com/opencollab/arpack-ng}},
-SPECFEM3D\footnote{\url{https://github.com/geodynamics/specfem3d}}.
+ARPACK-NG~\cite{arpackng}, and
+SPECFEM3D~\cite{specfem3d}.
 

### 2016-11-16 paper.tex
added missing references.bib entries and updated total stencils found in intro
typo

@@
   specifications (\cref{sec:analysis}), derived from a
-  denotational model for the specificaiton language
+  denotational model for the specification language
   (\cref{sec:semantics});
@@
   (\cref{sec:evaluation}), totalling one million lines of code. Our tool
-  identifies and infers specifications for 35,000 stencil computations in the
+  identifies and infers specifications for 60,000 stencil computations in the
   corpus. Approximately 10\% of the stencils we found are non-trivial,

### 2016-11-16 paper.tex
setion 2.1 edit
added paragraph to evaluation pointing out that inference mode is not the normal way to use the tool

@@
 
-We first examined how frequently stencil computations occur in our
-corpus of Fortran code. This was carried out by running CamFort in
-whole-code-base inference mode. We parsed 959,427 lines of Fortran
-code and found that 10\% (97,439) of statements have a left-hand side
-is an array subscript on neighbourhood indices. This supports the idea
-that stencil-like computations are common in scientific code.
-
-We would not expect to infer a stencil for each of these statements
-because our stencil language places significant restrictions on the
-types of array-subscript-statements that we will class as a
+We first examined how frequently stencil computations occur. We parsed
+959,427 lines of Fortran code and found that 10\% (97,439) of
+statements have a left-hand side is an array subscript on
+neighbourhood indices. This supports the idea that stencil-like
+computations are common in scientific code.
+
+We used CamFort in inference mode to identify stencils within the
+corpus. This assumes that the programs have been implemented
+correctly---normally one would use CamFort in verification mode to
+check the correctness of new code.
+
+We would not expect to infer a stencil for each array statement we
+found because our stencil language places significant restrictions on
+the types of array-subscript-statements that we will class as a
 stencil. In fact, CamFort was able to infer a stencil from 30\% of

### 2016-11-16 paper.tex

@@
   specifications (\cref{sec:analysis}), derived from a
-  denotational model for the specification language
+  denotational model for the specificaiton language
   (\cref{sec:semantics});
@@
   (\cref{sec:evaluation}), totalling one million lines of code. Our tool
-  identifies and infers specifications for 60,000 stencil computations in the
+  identifies and infers specifications for 35,000 stencil computations in the
   corpus. Approximately 10\% of the stencils we found are non-trivial,
@@
 
-\paragraph{Region constants}
+%\paragraph{Region constants}
 
@@
 approximating PDEs is known as the \emph{Forward Time, Centered Space} (FTCS)
-scheme~\citep{dawson1991finite}. Each region
+scheme~\citep{dawson1991finite}.
+
+Each region
 constant has a dimension identifier $d$ given as a positive natural number.

### 2016-11-16 paper.tex
Slightly curtail related work

@@
 % Pochoir
-\citet{Tang:2011:SPAA} define a specification language for writing
-stencils embedded in C++ (with Cilk~\citep{blumofe1996cilk} extensions) that
-are then compiled
-into fast parallel programs based on trapezoidal decompositions with
-hyperspace cuts. The Pochoir specification language is used for
-describing the kernel, boundary conditions and shape of the
-stencil. Programs are first compiled in C++ with a template library
-that checks whether the annotations are used correctly, and then a
-Pochoir preprocessor may be applied to generate the high-performance
-parallel versions using Cilk.  Pochoir
-is aimed primarily at assisting programmers who may be reluctant to
-implement the high-performance tricky cache-oblivious
-``hypertrapezoidal'' algorithms for stencils. 
-%that are significantly faster than the typical looping implementations
-%of stencil computations.
-Like much of the related work, the main goal 
-is optimisation rather than verifying program correctness.
+\citet{Tang:2011:SPAA} define a specification language for writing stencils
+embedded in C++ (with Cilk~\citep{blumofe1996cilk} extensions) that are then
+compiled into parallel programs based on trapezoidal decompositions with
+hyperspace cuts. The Pochoir specifications are used for describing the kernel,
+boundary conditions, and shape of the stencil. Pochoir is aimed at programmers
+reluctant to implement the high-performance cache-oblivious
+``hypertrapezoidal'' algorithms for stencils.  Like much of the related work,
+the goal is optimisation rather than verifying program correctness.
 

### 2016-11-16 paper.tex
Intro to definition is subsumed by the definition
Simplify index scheme index relationship explanation

@@
 Indexing schemes abstract array subscripts.
-Our model represents the meaning of stencil specifications
-as indexing schemes, which explain the range of possible
+Our model, consisting index schemes, explain the range of possible
 indexing behaviours which are consistent with the specification.
 The relationship between source language
-terms and index schemes
- is given precisely by the following function \textsf{schematic},
+and index schemes
+ is defined by \textsf{schematic} function,
  which forms part of checking and inference (\Cref{sec:analysis}).
@@
 
-The notion of which dimensions $i$ have \emph{any scheme in the model
-where $u_i \neq \infty$} is represented by following  function:
 \begin{definition}%[Constrained dimensions]

### 2016-11-16 paper.tex
Further simplify related work

@@
 in their specifications, \eg{}, ACSL of \citet{baudin2008acsl}
-for C (see \eg{}~\citet[Example 3.4.1]{burghardt2010acsl}). Thus, a specification
+for C (see \eg{}~\citet[Example 3.4.1]{burghardt2010acsl}). A specification
 can be given for a stencil computation, 
-however, such a specification would exactly replicate the fine-grained
-indexing seen in the code and would be similarly prone to low-level input
+however, such specification would replicate the fine-grained
+indexing seen in the code and would be similarly prone to indexing
 errors. Our approach is much more abstract-- it does
-not aim to reify indexing in the specification, but is novel
-in that it instead
-provides abstract spatial descriptions which capture a large number
+not aim to reify indexing in the specification, but
+provides simple spatial descriptions which capture a large number
 of common patterns.

### 2016-11-16 paper.tex
The reader need not be explained how Cartesian product works

@@
 \end{alignat*}
-The first guard takes the Cartesian product of
-the models and quantifies over all pairings $(u, v)$.
-The second guard binds $j$ which ranges over the constrained
+The second guard binds $j$ to the constrained
 dimensions for $M$ unioned with the constrained dimensions for $N$.

### 2016-11-16 paper.tex
Added commented-out paragraph about number of region operators in multi-action stencil specs. (You decide whether to uncomment).
various edits

@@
 rule out any accidental repetition of array subscripts.
-The notion is similar to that of \emph{linearity} in linear type
-systems~\cite{wadler1990linear}, where a value must be used
+The notion is similar to that of linear types~\cite{wadler1990linear}, where a value must be used
 exactly once. We opt for the more informative and easily understood name
@@
 
-Our terminology of \emph{index schemes} and
-\emph{generalisation} is by analogy
-with the \emph{type schemes} and \emph{type generalisation} of
-polymorphic types in ML~\cite{milner1978theory}. Our notion is
-much more rigid however, as index schemes do not bind names.
+%Our terminology of \emph{index schemes} and
+%\emph{generalisation} is by analogy
+%with the \emph{type schemes} and \emph{type generalisation} of
+%polymorphic types in ML~\cite{milner1978theory}. Our notion is
+%much more rigid however, as index schemes do not bind names.
 
@@
 \end{align*}
-Note, $\infty$ for multiplication above, \eg{} $2
+Note, $\infty$ is absorbing for multiplication above \eg{} $2
  \infty = \infty$.
@@
 \label{fig:spatial-model}
-\vspace{-0.75em}
+\vspace{-1em}
 \end{figure}
@@
 \end{equation*}
-(See the supplement for the proof)
+See the supplement for the proof.
 \end{theorem}
@@
 where $\consAName$ implements consistency in checking (\cref{subsec:checking}).
-(See the supplement for the proof)
+See the supplement for the proof.
 \end{theorem}
@@
 We demonstrate the analysis and modelise procedures over the
-following example program which computes the mean value
+following example kernel which computes the mean value
 of a five-point stencil:
@@
 \begin{minted}{fortran}
-do i = 1, (n-1)
-  do j = 1, (m-1)
 x = a(i-1, j)+a(i+1, j); y = a(i, j-1)+a(i, j+1)
 b(i, j) = (a(i, j) + x + y) / 5.0
-end do; end do
 \end{minted}
 %
-The \textit{analyse} function matches on
-line 4 with the following set for $\textsf{flowsTo}(\texttt{a(i, j) + x +
-  y}) = $
+%The \textit{analyse} function matches on
+%line 4 with the following set for $\textsf{flowsTo}(\texttt{a(i, j) + x +
+%  y}) = $
 %
 %\begin{align*}
-{{ %{\small{
-$\{$\texttt{a(i-1, j)}, \texttt{a(i+1, j)}, \texttt{a(i, j-1)},
-  \texttt{a(i, j+1)}, \texttt{a(i, j)}$\}$}}. 
+%{{ %{\small{
+%$\{$\texttt{a(i-1, j)}, \texttt{a(i+1, j)}, \texttt{a(i, j-1)},
+%  \texttt{a(i, j+1)}, \texttt{a(i, j)}$\}$}}. 
 %\end{align*}
-Thus, \emph{modelise} returns the following
-as all multiplicities are one:
+The  multiplicities are all one, and so \emph{modelise} yields:
 %%
@@
 \end{align*}
-\vspace{-1.5em}
+%\vspace{-1.6em}
 
@@
 %
-\vspace{-0.7em}
+\vspace{-0.9em}
 \begin{align*}
@@
 \end{align*}
-\vspace{-1em}
-%
+\vspace{-1.1em}
+
 \item By the definition of consistency (\Cref{fig:consistency})
@@
   $\textsf{once}(M_c)$ when:
-  \begin{align*}
-\vspace{-0.3em}
+\vspace{-0.6em}
+  \begin{align*}
     & \consSub{2}{\textsf{up}(N_s)}{M_c} \;\,
     \wedge \;\, \consSub{2}{\textsf{low}(N_s)}{M_c}
-\vspace{-0.3em}
-  \end{align*}
-%
+  \end{align*}
+%\vspace{-1.6em}
+
 \item This holds by the following sets of scheme
@@
 %%
-An entire span is converted into an $n$-dimensional
+A span is converted into an $n$-dimensional
 specification by applying $\mathsf{toSpec}$ to each
 pair of lower and upper bounds in a span, per dimension, and multiplying these
-specifications by $\odot$. The resulting specifications for
-each span are then summed by $\oplus$:
+by $\odot$. The resulting specifications for
+each span are summed by $\oplus$:
 \begin{equation*}
@@
 The $\hat{\odot}$ operation is then lifted to all combinations of
-\textsf{Approx} for $\odot$:
+\textsf{Approx} for $\odot$ in \Cref{fig:odotlift} where
+in the last case $\textit{inj}$ corresponds to the unary
+injections of \textsf{Approx}. For brevity we omit the cases where the arguments above are flipped
+as $\odot$ is commutative.
+\end{definition}
+
+\begin{figure}[t]
+\vspace{-1em}
+{\small{
 \begin{align*}
@@
 \end{align*}
-where in the last case $\textit{inj}$ corresponds to the unary
-injections of \textsf{Approx}. For brevity we omit the cases where the arguments above are flipped
-as $\odot$ is commutative.
-\end{definition}
+\vspace{-2em}
+}}
+\caption{Lifting $\hat{\odot}$ to approximations.}
+\label{fig:odotlift}
+\end{figure}
 
@@
 the syntactic sum $\term{+}$. Then $\oplus$ is the lifting of
-$\hat{\oplus}$ to \textsf{Approx} with the shape as $\odot$ in the previous
-definition.
+$\hat{\oplus}$ to \textsf{Approx} with the shape as $\odot$ in Figure~\ref{fig:odotlift}.
 \end{definition}
@@
 based on some simple heuristics.
-
 For our example, we infer and then synthesise the
-5-point specification:
-%%
-\begin{minted}[linenos=false]{fortran}
+final 5-point specification:
+%%
+\begin{minted}[linenos=false,breakindent=2em]{fortran}
 != stencil centered(depth=1, dim=1)*pointed(dim=2) + centered(depth=1, dim=2)*pointed(dim=1):: a
@@
 
-We built a corpus of around 1 million lines of Fortran code from a
+To study the effectiveness of our approach, 
+we built a corpus of around 1 million lines of Fortran code from a
 range of scientific computing packages: The Unified Model~\cite{um},
@@
 959,427 lines of Fortran code and found that 10\% (97,439) of
-statements have a left-hand side is an array subscript on
+statements have a left-hand side as an array subscript on
 neighbourhood indices. This supports the idea that stencil-like
@@
 
-We used CamFort in inference mode to identify stencils within the
-corpus. This assumes that the programs have been implemented
-correctly---normally one would use CamFort in verification mode to
+To used the inference procedure of the previous section to identify
+stencils within the corpus. This assumes that the programs have been implemented
+correctly---normally one would use our approach in verification mode to
 check the correctness of new code.
@@
 We would not expect to infer a stencil for each array statement we
-found because our stencil language places significant restrictions on
-the types of array-subscript-statements that we will class as a
-stencil. In fact, CamFort was able to infer a stencil from 30\% of
+found because our specification restricts 
+the array-subscript-statements that we class as a
+stencil. In fact, we was able to infer a stencil from 30\% of
 these statements. A single statement can produce multiple stencils and
@@
 
+% multiActionRegionOps1: 31
+% multiActionRegionOps2: 115
+% multiActionRegionOps8: 1
+% multiActionRegionOps14: 5
+% multiActionRegionOps17: 4
+
+% %%%%
+% Complexity-wise, out of these 156 multi-action specifications, 31 of
+% them used a single region operator, 115 used two region operators, 1
+% used eight region operators, 5 used fourteen region operators, and 4
+% used seventeen region operators. Some of these very complex stencils
+% were found in the atmospheric models of the UM.
+% %%%%
+
 \textbf{Bounded} specifications were fairly unusual, with only 157
@@
 in their specifications, \eg{}, ACSL of \citet{baudin2008acsl}
-for C (see \eg{}~\citet[Example 3.4.1]{burghardt2010acsl}). A specification
-can be given for a stencil computation, 
-however, such specification would replicate the fine-grained
-indexing seen in the code and would be similarly prone to indexing
+for C (\eg{}~\citet[Example 3.4.1]{burghardt2010acsl}). A specification
+can be given for a stencil computation but replicating the fine-grained
+indexing seen in the code and therefore similarly prone to indexing
 errors. Our approach is much more abstract-- it does
@@
 % GPGPU
-Our work has some similarities with efforts to verify kernels
-written for General-Purpose Graphics Processing Unit (GPGPU)
+Our work has similarities with efforts to verify kernels
+written for General-Purpose GPU
 programming, such as in \citet{Blom:2014:SoCP}. %Stencils are a form of
@@
 %reason.
-
+%
 % Sketching Stencils
@@
 unoptimised ``reference'' stencils, coupled with partial
-implementations which are then completed by a code
+implementations which are completed by a code
 generation tool. %These kinds of specifications are just simple
@@
 
-By contrast, the work of \citet{Abe:2013:IPDPSW} has looked at the
-correctness problem explicitly by bringing a form of model-checking to
-verify certain stencil computations. This is in the context of
-decomposing stencils for parallel computation on multiple processors
+By contrast, the work of \citet{Abe:2013:IPDPSW} studies
+correctness explicitly by bringing a form of model-checking to
+verify certain stencil computations in the context of
+parallel implementations 
  in partitioned global address
@@
 is generated from a relatively high-level specification. 
-In contrast, we are interested in integrating directly
+In contrast, we integrate directly
 into existing, legacy codebases and established languages, bringing
@@
 
-Our definition of a stencil computation meant that we generated no
-specification for a large number of iterated array accesses in the
-corpus, such as stencils that incorporates reductions. For
-future work, we will be looking into expanding the flexibility of 
+Our restrictive definition of stencils means we do note generate
+specifications for a large number of iterated array accesses in the
+corpus, \eg{} that incorporate reductions. For
+future work, we will look at expanding the flexibility of 
 our analysis to include these examples.

### 2016-11-16 paper.tex
intro consistency and fit into 12 pages

@@
   indices across multiple arrays and dimensions. This requires careful
-  testing, which is almost always done manually and then discarded
-  during development. In this paper, we propose an alternate approach:
+  testing, which is almost always done manually and then discarded.
+  In this paper, we propose an alternate approach:
   a specification language and automated tool for verifying the
@@
   on the hypothesis that stencil computations read from arrays with a
-  regular, fixed pattern which can be captured abstractly by a simple
-  set of combinators in our specification language. These
+  regular, fixed pattern which can be captured by a simple
+  set of abstract combinators. These
   specifications can be embedded as annotations in source code,
@@
   spatial specifications can be inferred, providing documentation and
-  aiding future software maintenance. We evaluate our language against
-  a corpus of numerical Fortran code, and show that the vast majority
-  of stencil computations indeed have a simple, regular, static shape
-  validating the design of our language.  We introduce the
-  language, checking and inference procedures, and its evaluation.
+  aiding future software maintenance. We evaluate our specification
+  language and verification tool against
+  a corpus of numerical Fortran code ($\sim$ 1 million lines) for
+  which we generate $\sim$60,000 specifications, showing the vast majority
+  of stencil computations indeed have a simple, regular, static shape. % We introduce the
+  %language, checking and inference procedures, and its evaluation.
 \end{abstract}
@@
 %
-This describes \texttt{a} is accessed with a symmetrical pattern
+That is, \texttt{a} is accessed with a symmetrical pattern
 (``centered'') to a depth of one in each direction in its first
-dimension.  The shape of the Navier-Stokes example
-is specified as:
+dimension.  The Navier-Stokes example has a shape
+specified as:
 %
@@
 first dimension and backward to a depth of $1$ in the second
-dimension. Note the relatively small size of the specification
-compared to the source code, and its high-level of abstraction.
+dimension. The specification is relatively small
+and much more abstract compared to the source code.
 
@@
 %  of software maintenance activities.
-\item The primary use case is that a programmer writes a specification first,
-  then writes the corresponding source code. Our tool then checks conformance between
+\item The primary use case is that a programmer writes a specification
+  first before the stencil code. Our tool then checks conformance between
   the two. This specify-and-check approach reduces testing effort and
@@
   combinators that \emph{do not involve any indexing expressions}, \eg{}
-  \texttt{a(i+1,j-1)}. This contrasts with other specification approaches. If a
-  user specifies, say, the Naiver-Stokes stencil above, in a deductive
-  verification language such as ACSL, they must write a specification involving
-  textually dense array-indexing expressions. Therefore, any low-level mistakes
+  \texttt{a(i+1,j-1)}. This contrasts with other specification
+  approaches, \eg{}, deductive verification tools such as ACSL, where
+  specifications must also be in terms of array-indexing expressions. Therefore, any low-level mistakes
   that could be made whilst programming complex indexing code could also be
-  made when writing its specification. Our specifications are more lightweight
+  made when writing its specification. Our specifications are more
+  abstract and lightweight
   with the aim of aiding adoption by scientists and preventing indexing errors.
@@
   (\cref{sec:evaluation}), totalling one million lines of code. Our tool
-  identifies and infers specifications for 35,000 stencil computations in the
-  corpus. Approximately 10\% of the stencils we found are non-trivial,
+  identifies and infers specifications for roughly 60,000 stencil computations in the
+  corpus. Approximately 5,000 of the stencils we found are non-trivial,
   corresponding to code which is a possible source of errors. This validates
@@
 
-\item We give a detailed verification case study of two
-  stencil computations commonly used in scientifc computing,
-  simulating programming errors. Our approach detects all
-  possible indexing errors for these computations.
+\item We give a verification case study of a
+  stencil computation commonly used in scientifc computing,
+  simulating programming errors (\Cref{sec:jacobi}). Our approach detects all
+  possible simulated programming errors for this problem.
 \end{itemize}
@@
 we built a corpus of around 1 million lines of Fortran code from a
-range of scientific computing packages: The Unified Model~\cite{um},
+range of scientific computing packages: The Unified Model (UM)~\cite{um},
 E3MG~\cite{RePEc:aen:journl:2006se-a12}, BLAS~\cite{blas},
@@
 computations are common in scientific code.
-
-To used the inference procedure of the previous section to identify
-stencils within the corpus. This assumes that the programs have been implemented
-correctly---normally one would use our approach in verification mode to
-check the correctness of new code.
+We then used the inference procedure of the previous section to
+generate specifications for stencils in the corpus to assess
+the design of the language. %  This assumes that the programs have been implemented
+%correctly---
+%Normally one would use our approach in verification mode to
+%check the correctness of new code.
 
 We would not expect to infer a stencil for each array statement we
-found because our specification restricts 
+found because our analysis restricts 
 the array-subscript-statements that we class as a
 stencil. In fact, we was able to infer a stencil from 30\% of
-these statements. A single statement can produce multiple stencils and
-we ended up with 60,525 stencils. This shows that we can still express
-a large number of stencil uses within our high-level abstraction.
+these statements. A single statement can involve multiple arrays and
+we ended up with 60,525 specifications. This shows that we can express
+a large number of stencil uses within our high-level abstraction and
+validates our initial hypothesis that many stencil computations have a
+regular shape.
 
@@
 \noindent
-2) \textbf{Derived induction variables} where the
-indexing variable (\mintinline{fortran}{x}) is derived from an
+2) \textbf{Derived induction variables} where an
+index (\mintinline{fortran}{x}) is derived from an
 induction variable (\mintinline{fortran}{i}) as in
@@
 
-The vast majority of stencils we found were relatively simple but we
+The majority of specifications generated were relatively simple but we
 found significant numbers of more complex shapes. We grouped common
-patterns in to categories which we describe below.
+patterns in to categories:
 
@@
 
-\textbf{Single-action} specifications consist of no more than one 
-forward, backward, or centered region and can be combined with any
-number of pointed regions. We identified 4,837 single-action
+\textbf{Single-action} specifications comprise one 
+forward, backward, or centered region contant combined via \term{+} 
+or \term{*} with any number of pointed regions. We identified 4,837 single-action
 specifications, of which 1,532 were single-action with a
@@
 
-\textbf{Multi-action} specifications consist of at least two uses 
-of forward, backward, or centered regions, and can be combined with
-any number of pointed regions. We identified 156 multi-action
-specifications out of which 146 had regions combined only with
-$\term{*}$.
+\textbf{Multi-action} specifications comprise at least two 
+ forward, backward, or centered regions, combined with
+any number of pointed regions. We identified 265 multi-action
+specifications out of which 207 had regions combined only with
+$\term{*}$ and 58 combined with a mix of $\term{*}$ and \term{+}.
 
@@
 \texttt{atMost} and 36 \texttt{atLeast}, the latter of which were
-always also bounded on both sides.
+always also paired with an upper bound.
 
@@
 lines of code. This is an impressive reduction given
-the complexity of the algorithm.  We show one example in~\cref{fig:smagorinsky}. This specifies the access pattern to
+the complexity of the algorithm.  We show one example in~\Cref{fig:smagorinsky}. This specifies the access pattern to
 a 3-dimensional array (which we have renamed to be called
-\texttt{x}). The array is accessed 8 times in the computation.
+\texttt{x}) arising from a kernel of 93 lines of code involving 142
+individual array subscripts. The specification was the most complex
+seen in our corpus, yet it still represents a significant abstraction
+to the spatial behaviour given size and complexity of the kernel it describes.
 
 \subsection{Detecting errors in the 2-D Jacobi iteration}
-
+\label{sec:jacobi}
 One common example of a stencil computation is two-dimensional

### 2016-11-16 paper.tex
camfort reference

@@
 \item We provide an implementation of our approach as an extension to
-  CamFort, an open-source program analysis tool for Fortran;
+  CamFort~\cite{camfort}, an open-source program analysis tool for Fortran;
 
@@
 
-We evaluated our approach on top of CamFort, an open-source analysis
-tool for Fortran programs, running the inference process over a corpus
+We evalauated our approach by extending CamFort, an open-source analysis
+tool for Fortran programs~\cite{camfort}, running the inference process over a corpus
 of a million lines of Fortran code. We showed that our language is

### 2016-11-16 paper.tex
spell checking

@@
 inspections, or comparison against an manufactured or 
-analytical solutionst~\cite{farrell2010automated}). Typically such
+analytical solutions~\cite{farrell2010automated}). Typically such
 testing is discarded once the code is seen to be correct.
@@
 of stencils. The shape of the
-Laplace example stencil is specifed in our language as:
+Laplace example stencil is specified in our language as:
 %
@@
   the two. This specify-and-check approach reduces testing effort and
-  future-proofs against bugs introduced during refactoring and maintanence.
+  future-proofs against bugs introduced during refactoring and maintenance.
 %
@@
 \item We give a verification case study of a
-  stencil computation commonly used in scientifc computing,
+  stencil computation commonly used in scientific computing,
   simulating programming errors (\Cref{sec:jacobi}). Our approach detects all
@@
   is interpreted as an induction variable only within the scope of the loop
-  body. Throuhgout, $i, j, k$ range over induction variables.
+  body. Throughout, $i, j, k$ range over induction variables.
 %TODO: consider mentioning derived, non-based induction variables
@@
 % specifying the shape of indexing expressions for
-%arrays $\bar{v}$ that are simultaenously permitted and expected within
+%arrays $\bar{v}$ that are simultaneously permitted and expected within
 %the code to which the specification is given.
@@
 
-A \term{forward} region of depth $n$ specifies a continguous
+A \term{forward} region of depth $n$ specifies a contiguous
 region in dimension $d$ starting at the origin. This corresponds
@@
 The region operators \term{+} and \term{*} respectively combine
-regions disjunctively and conjuctively.
-
-The conjuction of two regions $r \term{*} s$ means that any indices
+regions disjunctively and conjunctively.
+
+The conjunction of two regions $r \term{*} s$ means that any indices
 in the specified code must be consistent with both $r$ and $s$
-simultaenously. Dually, for every pair of components in the regions
+simultaneously. Dually, for every pair of components in the regions
 $r$ and $s$ there must be an index consistent with both.
@@
 The disjunction of two regions $r \term{+} s$ means that any indices
-in the spcified code must be consistent with either of $r$ or $s$.
+in the specified code must be consistent with either of $r$ or $s$.
 Dually, every part of region $r$ must have a
@@
 Let $\schemeOne$ be a partial function, parameterised by a set of
-induction variables $I$, mapping source langauge expressions to
+induction variables $I$, mapping source language expressions to
 $\domainVal$ values:
@@
 
-% r*s means indices must be consistent simultaenously with both r and s
+% r*s means indices must be consistent simultaneously with both r and s
 %           and there are indices which are generalised
@@
 Additionally, specifications have a theory of approximation,
-$<:$, that can be used by the users to weaken specifications for succintness in
+$<:$, that can be used by the users to weaken specifications for succinctness in
 the source.
@@
 \item \textit{(centered)} a forward and backward region of the same
-  depth and dimension coalseces into a centered region:
+  depth and dimension coalesces into a centered region:
 %
@@
 with respect to \term{+} and \term{*} and $<:$ is congruent
-with repect to \term{readOnce}, the equations of which we omit for brevity.
+with respect to \term{readOnce}, the equations of which we omit for brevity.
 
@@
 to be dropped. The \trule{\textsc{shrink}}
-and \trule{\textsc{grow}} rules turn the spatial approxiamation on
+and \trule{\textsc{grow}} rules turn the spatial approximation on
 regions into approximation on specifications via the \term{atLeast} and \term{atMost}
@@
 of source code against specifications (\Cref{subsec:checking})
-and for infering specifications from code (\Cref{subsec:inference}).
+and for inferring specifications from code (\Cref{subsec:inference}).
 Both rely on a program analysis that converts array subscripts
@@
 Thus, we focus on assignments to an array subscript where the LHS
-indices $\overline{e_1}$ are neighbourhood indidces.  For
+indices $\overline{e_1}$ are neighbourhood indices.  For
 all array subscripts that flow to the right-hand side of this
@@
 (\Cref{subsec:analysis}) to extract index schemes from source code,
-augmented with multiplicitly for the whole set of schemes.
+augmented with multiplicity for the whole set of schemes.
 This is compared with the model of the specification $\interp{S}_n$ by
@@
 is a scheme in the spec model that generalises it, and for every
-scheme in the specifcation there is a scheme in the source model that
+scheme in the specification there is a scheme in the source model that
 specialises it.  The (\textit{both}) rule is similar, decomposing
@@
 \textbf{Single-action} specifications comprise one 
-forward, backward, or centered region contant combined via \term{+} 
+forward, backward, or centered region constant combined via \term{+} 
 or \term{*} with any number of pointed regions. We identified 4,837 single-action
@@
 The iteration computes the average of four adjacent
-cells so $24$ ($4$ factorial) of the possible array index pertubations
+cells so $24$ ($4$ factorial) of the possible array index perturbations
 are correct, all of which are accepted by checking.
@@
 
-We evalauated our approach by extending CamFort, an open-source analysis
+We evaluated our approach by extending CamFort, an open-source analysis
 tool for Fortran programs~\cite{camfort}, running the inference process over a corpus

### 2016-11-16 paper.tex
FREQUENCY TABLE

@@
 Thus, our model validates specification synthesis.
-%
-\subsection{Step 5: Simplification}
+
 The final step applies a simplification pass, orienting the
@@
 %%
-\begin{minted}[linenos=false,breakindent=2em]{fortran}
+\begin{minted}[linenos=false,breakindent=2.4em,xleftmargin=0.2em]{fortran}
 != stencil centered(depth=1, dim=1)*pointed(dim=2) + centered(depth=1, dim=2)*pointed(dim=1):: a
@@
 found because our analysis restricts 
-the array-subscript-statements that we class as a
-stencil. In fact, we was able to infer a stencil from 30\% of
+the array-subscript-statements that we classify as a
+stencil. In fact, we were able to infer a stencil from 30\% of
 these statements. A single statement can involve multiple arrays and
 we ended up with 60,525 specifications. This shows that we can express
-a large number of stencil uses within our high-level abstraction and
+a large number of stencil shapes within our high-level abstraction and
 validates our initial hypothesis that many stencil computations have a
@@
 
-There were a variety of reasons why we did not infer a
-stencil from every potential statement: \\
+%    numStencilSpecs: 60525
+% 4338 files, linesTotal: 1357377, linesParsed: 959427
+% Number of stencils (tickAssign) / number of loops
+% Percentage of a code base that is taken up by stencils
+%Out of 1.35 million lines of code that we fed into our program, and
+%the 959,427 lines that we were able to parse, we identified 60,525
+%cases that fit our criteria for stencil specification.
+
+%\paragraph{Hypothesis 2: Most stencil computations
+%have a regular pattern which can be captured with a simple
+%language of contiguous regions}
+
+% Number of stencils we give a spec to (where a spec declaration to n
+% variables is counted n times) as percentage of number of stencils
+
+    % tickAssign: 97439
+    % tickAssignSuccess: 29402 % <- number of times tickAssign is accompanied by a successful stencil inference
+                               % counts only by +1 each time that tickAssign finds a stencil
+
+%In the analysed body of code we found that we were able to deduce
+%some form of stencil specification from approximately 30\% of the
+%potential stencils examined.
+
+%
+%\subsection{Analysing the design of our specifications}
+
+% 1. How may are just reflexive and how many some reflexive
+% justReflexive: 55504
+% someReflexive: 15823
+
+The majority of specifications generated were relatively simple but we
+found significant numbers of more complex shapes. We grouped common
+patterns in to categories:
+
+\textbf{All pointed} 55,504 of the stencils we found involved 
+only pointed regions. 39,681 of these were pointed in all
+dimensions. Common examples of this were pointwise transformations on
+data (such as scaling).
+
+\textbf{Single-action} specifications comprise one 
+forward, backward, or centered region constant combined via \term{+} 
+or \term{*} with any number of pointed regions. We identified 4,837 single-action
+specifications, of which 1,532 were single-action with a
+\texttt{nonpointed} modifier.
+
+\textbf{Multi-action} specifications comprise at least two 
+ forward, backward, or centered regions, combined with
+any number of pointed regions. We identified 265 multi-action
+specifications out of which 207 had regions combined only with
+$\term{*}$ and 58 combined with a mix of $\term{*}$ and \term{+}.
+
+% multiActionRegionOps1: 31
+% multiActionRegionOps2: 115
+% multiActionRegionOps8: 1
+% multiActionRegionOps14: 5
+% multiActionRegionOps17: 4
+
+% %%%%
+% Complexity-wise, out of these 156 multi-action specifications, 31 of
+% them used a single region operator, 115 used two region operators, 1
+% used eight region operators, 5 used fourteen region operators, and 4
+% used seventeen region operators. Some of these very complex stencils
+% were found in the atmospheric models of the UM.
+% %%%%
+
+\textbf{Bounded} specifications occured with 157
+\texttt{atMost} bounds and 36 \texttt{atLeast}, the latter of which were
+always also paired with an upper bound.
+
+\begin{figure}[t]\begin{minted}[fontsize=\scriptsize,breakindent=0em,linenos=false,xleftmargin=0em,breakafter=)]{fortran}
+!=stencil readOnce,(forward(depth=1,dim=3,nonpointed))*(backward(depth=1,dim=1))*(backward(depth=1,dim=2))+(forward(depth=1,dim=3))*(backward(depth=1,dim=1))*(pointed(dim=2))+(forward(depth=1,dim=3))*(backward(depth=1,dim=1,nonpointed))*(backward(depth=1,dim=2,nonpointed))+(forward(depth=1,dim=3))*(backward(depth=1,dim=2))*(pointed(dim=1))+(backward(depth=1,dim=1))*(backward(depth=1,dim=2,nonpointed))*(pointed(dim=3))+(backward(depth=1,dim=1,nonpointed))*(backward(depth=1,dim=2))*(pointed(dim=3))::x
+\end{minted}
+\caption{Complex specification inferred from
+  \textbf{UM}\label{fig:smagorinsky}}
+\vspace{-1em}
+\end{figure}
+
+The single- and multi-action classes represent more complex stencils
+with a real possibility for programmer error. As an extreme example, the Unified
+Model has an implementation of the Smagorinsky subgrid-scale model for
+calculating turbulence on which our inference yields 39 specs from 340
+lines of code. This is an large reduction given
+the complexity of the algorithm.  We show one example
+in~\Cref{fig:smagorinsky} which specifies the access pattern to
+a 3-dimensional array (which we have renamed to 
+\texttt{x}) arising from a kernel of 93 lines of code involving 142
+array subscripts. The specification was the most complex
+seen in our corpus, yet it still represents a significant abstraction
+of the spatial behaviour given size and complexity of the kernel it describes.
+
+We measured the frequency at which individual specifications involved 
+multiple occurences of \term{+} and \term{*}:
+
+\vspace{0.25em}
+\setlength{\tabcolsep}{0.57em}
+{\small{
+\hspace{-1em}\begin{tabular}{c|cccccccc}
+& 0 & 1 & 2 & 3 & 4 & 5 & 6 & $> 7$ \\ \hline
+\term{+} & 60,265 & 220 & 15 & 12 & 5 & 4 & 4 & 0 \\
+\term{*} & 31,393 & 15,151 & 13,457 & 424 & 41 & 42 & 5 & 12
+\end{tabular}}}
+\vspace{0.2em}
+
+\noindent
+\eg{}, roughly half the specifications did not involve $\term{*}$, 
+a quarter use one $\term{*}$ and just under a quarter
+use two $\term{*}$ operators.
+
+\paragraph{Limitations}
+
+There were various reasons why we did not infer 
+specifications on every looped array computation: \\
 \noindent 
@@
 
-%    numStencilSpecs: 60525
-% 4338 files, linesTotal: 1357377, linesParsed: 959427
-% Number of stencils (tickAssign) / number of loops
-% Percentage of a code base that is taken up by stencils
-%Out of 1.35 million lines of code that we fed into our program, and
-%the 959,427 lines that we were able to parse, we identified 60,525
-%cases that fit our criteria for stencil specification.
-
-%\paragraph{Hypothesis 2: Most stencil computations
-%have a regular pattern which can be captured with a simple
-%language of contiguous regions}
-
-% Number of stencils we give a spec to (where a spec declaration to n
-% variables is counted n times) as percentage of number of stencils
-
-    % tickAssign: 97439
-    % tickAssignSuccess: 29402 % <- number of times tickAssign is accompanied by a successful stencil inference
-                               % counts only by +1 each time that tickAssign finds a stencil
-
-%In the analysed body of code we found that we were able to deduce
-%some form of stencil specification from approximately 30\% of the
-%potential stencils examined.
-
-%
-%\subsection{Analysing the design of our specifications}
-
-% 1. How may are just reflexive and how many some reflexive
-% justReflexive: 55504
-% someReflexive: 15823
-
-The majority of specifications generated were relatively simple but we
-found significant numbers of more complex shapes. We grouped common
-patterns in to categories:
-
-\textbf{All pointed} 55,504 of the stencils we found involved 
-only pointed regions. 39,681 of these were pointed in all
-dimensions. Common examples of this were pointwise transformations on
-data (such as scaling).
-
-\textbf{Single-action} specifications comprise one 
-forward, backward, or centered region constant combined via \term{+} 
-or \term{*} with any number of pointed regions. We identified 4,837 single-action
-specifications, of which 1,532 were single-action with a
-\texttt{nonpointed} modifier.
-
-\textbf{Multi-action} specifications comprise at least two 
- forward, backward, or centered regions, combined with
-any number of pointed regions. We identified 265 multi-action
-specifications out of which 207 had regions combined only with
-$\term{*}$ and 58 combined with a mix of $\term{*}$ and \term{+}.
-
-% multiActionRegionOps1: 31
-% multiActionRegionOps2: 115
-% multiActionRegionOps8: 1
-% multiActionRegionOps14: 5
-% multiActionRegionOps17: 4
-
-% %%%%
-% Complexity-wise, out of these 156 multi-action specifications, 31 of
-% them used a single region operator, 115 used two region operators, 1
-% used eight region operators, 5 used fourteen region operators, and 4
-% used seventeen region operators. Some of these very complex stencils
-% were found in the atmospheric models of the UM.
-% %%%%
-
-\textbf{Bounded} specifications were fairly unusual, with only 157
-\texttt{atMost} and 36 \texttt{atLeast}, the latter of which were
-always also paired with an upper bound.
-
-\begin{figure}[t]\begin{minted}[fontsize=\scriptsize,breakindent=0em,linenos=false,xleftmargin=0em,breakafter=)]{fortran}
-!=stencil readOnce,(forward(depth=1,dim=3,nonpointed))*(backward(depth=1,dim=1))*(backward(depth=1,dim=2))+(forward(depth=1,dim=3))*(backward(depth=1,dim=1))*(pointed(dim=2))+(forward(depth=1,dim=3))*(backward(depth=1,dim=1,nonpointed))*(backward(depth=1,dim=2,nonpointed))+(forward(depth=1,dim=3))*(backward(depth=1,dim=2))*(pointed(dim=1))+(backward(depth=1,dim=1))*(backward(depth=1,dim=2,nonpointed))*(pointed(dim=3))+(backward(depth=1,dim=1,nonpointed))*(backward(depth=1,dim=2))*(pointed(dim=3))::x
-\end{minted}
-\caption{Complex specification inferred from
-  \textbf{UM}\label{fig:smagorinsky}}
-\vspace{-1em}
-\end{figure}
-
-The single- and multi-action classes represent more complex stencils
-with a real possibility for programmer error. For example, the Unified
-Model has an implementation of the Smagorinsky subgrid-scale model for
-calculating turbulence on which our inference yields 39 specs from 340
-lines of code. This is an impressive reduction given
-the complexity of the algorithm.  We show one example in~\Cref{fig:smagorinsky}. This specifies the access pattern to
-a 3-dimensional array (which we have renamed to be called
-\texttt{x}) arising from a kernel of 93 lines of code involving 142
-individual array subscripts. The specification was the most complex
-seen in our corpus, yet it still represents a significant abstraction
-to the spatial behaviour given size and complexity of the kernel it describes.
-
 \subsection{Detecting errors in the 2-D Jacobi iteration}
 \label{sec:jacobi}
-One common example of a stencil computation is two-dimensional
+One common example of a stencil computation is the two-dimensional
 Jacobi iteration that repeatedly goes through each cell in a matrix
@@
 \end{minted}
-and we infer a precise specification for this, as:
+We infer a precise specification of its shape as:
 \begin{minted}[breakafter=+:,breakindent=-0.6em,breaksymbolsep=0.4em,linenos=false,xleftmargin=-0.5em]{fortran}
@@
 cells so $24$ ($4$ factorial) of the possible array index perturbations
-are correct, all of which are accepted by checking.
+are correct, all of which are accepted by our checker.
 
@@
 for C (\eg{}~\citet[Example 3.4.1]{burghardt2010acsl}). A specification
-can be given for a stencil computation but replicating the fine-grained
-indexing seen in the code and therefore similarly prone to indexing
+can be given for a stencil computation but must use fine-grained
+indexing as in code and therefore is similarly prone to indexing
 errors. Our approach is much more abstract-- it does
@@
 mainly on the synchronisation of kernels and the avoidance of
-data races, while we are interested in correctness 
-embedded within a more typical general-purpose programming
-language. %Other work on GPGPU computation, such as
+data races., while we are interested in correctness 
+%embedded within a more typical general-purpose programming
+%language. %Other work on GPGPU computation, such as
 %\citet{Zhang:2012:CGO}, has focused primarily on generating
@@
 The primary purpose of this tool is optimisation rather than
-correctness, and the language of specification is much different and
-more elaborate than in our approach.
+correctness, and the language of specification is 
+more elaborate than ours.
 
@@
 compiled into parallel programs based on trapezoidal decompositions with
-hyperspace cuts. The Pochoir specifications are used for describing the kernel,
+hyperspace cuts. Pochoir specifications are used for describing the kernel,
 boundary conditions, and shape of the stencil. Pochoir is aimed at programmers
 reluctant to implement the high-performance cache-oblivious
-``hypertrapezoidal'' algorithms for stencils.  Like much of the related work,
+``hypertrapezoidal'' algorithms.  Like much of the related work,
 the goal is optimisation rather than verifying program correctness.
@@
 By contrast, the work of \citet{Abe:2013:IPDPSW} studies
-correctness explicitly by bringing a form of model-checking to
+correctness bringing a form of model-checking to
 verify certain stencil computations in the context of
-parallel implementations 
+parallelism 
  in partitioned global address

### 2016-11-16 paper.tex
some small edits near the end

@@
 an $n$-dimensional rectilinear shape in a discrete space crossing the
-origin with respect to the induction variable. Consistency is then a
+origin (which represents a collection of indices which are just 
+induction variables). Consistency is then a
 two way requirement. All array subscripts on $a$ contributing to the
@@
 
-Indexing schemes abstract array subscripts.
-Our model, consisting index schemes, explain the range of possible
+Index schemes abstract array subscripts.
+Our model consists of index schemes explaining the range of possible
 indexing behaviours which are consistent with the specification.
@@
 Thus, if each expression in an index is a neighbourhood or constant
-index, $\schemeN$ maps it to an indexing scheme.
+index, $\schemeN$ maps it to an index scheme.
 \end{definition}
@@
 \begin{example}
-Consider the following array subscript to indexing schemes
+Consider the following array subscript to index schemes
 conversions:
@@
   \textit{region}.} specifications to sets of
-$n$-dimensional indexing schemes. The interpretation is overloaded on
+$n$-dimensional index schemes. The interpretation is overloaded on
 \emph{regions} in \Cref{fig:region-model} and on the top-level
@@
 subscript on neighbourhood indices, a finite map is computed mapping
-array variables to \emph{multisets} of indexing schemes:
+array variables to \emph{multisets} of index schemes:
 %
@@
 statement, a finite map is constructed, mapping each array variable
-to a multiset of indexing schemes for its subscripts, computed
+to a multiset of index schemes for its subscripts, computed
 with \textsf{schematic} (\Cref{def:schematic}).
@@
 \textit{modelise} function which filters out those arrays
-which contain undefined indexing schemes and adds multiplicity
+which contain undefined index schemes and adds multiplicity
 information:
@@
 brevity we make implicit the subscript $n$ defining the
-dimensionality.  We denote sets of indexing schemes
+dimensionality.  We denote sets of index schemes
 $\mathcal{P}((\domainVal{})^n)$
@@
 analysis of \cref{subsec:analysis} to produce a
-model of the code in terms of indexing schemes.
+model of the code in terms of index schemes.
 The \emph{analyse} and \emph{modelise} procedures before yielded the finite map:
@@
 \noindent
-Each indexing scheme $u \in U(a)$ is converted to a \emph{unit} span
+Each index scheme $u \in U(a)$ is converted to a \emph{unit} span
 $u \mapsto [u \; u]$, defining the lower and upper bound of an

### 2016-11-16 paper.tex
three last typos

@@
 found significant numbers of more complex shapes. We grouped common
-patterns in to categories:
+patterns into categories:
 
@@
 calculating turbulence on which our inference yields 39 specs from 340
-lines of code. This is an large reduction given
+lines of code. This is a large reduction given
 the complexity of the algorithm.  We show one example
@@
 
-Our restrictive definition of stencils means we do note generate
+Our restrictive definition of stencils means we do not generate
 specifications for a large number of iterated array accesses in the

### 2017-03-11 array17.tex
Array 17 first draft

--- 2016-11-16 paper.tex
@@
 
-\title{Verifying Spatial Properties of Stencil Computations}
+\title{Verifying Spatial Properties of Stencil Computations using CamFort}
 %\titlebanner{DRAFT -- Do not distribute}
@@
 \begin{abstract}
-Stencil computations are a common programming pattern in numerical
-  code and typically form the core of graphics and scientific
-  computing applications. However, they are often complex and
-  unwieldy, involving substantial fine-grained manipulation of array
-  indices across multiple arrays and dimensions. This requires careful
-  testing, which is almost always done manually and then discarded.
-  In this paper, we propose an alternate approach:
-  a specification language and automated tool for verifying the
-  spatial behaviour of stencils. The specification language is based
-  on the hypothesis that stencil computations read from arrays with a
-  regular, fixed pattern which can be captured by a simple
-  set of abstract combinators. These
-  specifications can be embedded as annotations in source code,
-  against which code can be checked for conformance. For legacy code,
-  spatial specifications can be inferred, providing documentation and
-  aiding future software maintenance. We evaluate our specification
-  language and verification tool against
-  a corpus of numerical Fortran code ($\sim$ 1 million lines) for
-  which we generate $\sim$60,000 specifications, showing the vast majority
-  of stencil computations indeed have a simple, regular, static shape. % We introduce the
-  %language, checking and inference procedures, and its evaluation.
+  Stencil computations are a common programming pattern in numerical code and
+  typically form the core of graphics and scientific computing applications.
+  However, they are often complex and unwieldy, involving substantial
+  fine-grained manipulation of array indices across multiple arrays and
+  dimensions. This requires careful testing, which is almost always done
+  manually and then discarded. In this paper, we propose an alternate approach:
+  a specification language and automated tool for verifying the spatial
+  behaviour of stencils. The specification language is based on the hypothesis
+  that stencil computations read from arrays with a regular, fixed pattern
+  which can be captured by a simple set of abstract combinators. These
+  specifications can be embedded as annotations in source code, against which
+  code can be checked for conformance. For legacy code, spatial specifications
+  can be inferred, providing documentation and aiding future software
+  maintenance.
+
+  We evaluate our specification language and verification tool
+  against a corpus of numerical Fortran code ($\sim$ 1 million lines) for which
+  we generate $\sim$ 60,000 specifications, showing the vast majority of stencil
+  computations indeed have a simple, regular, static shape.
 \end{abstract}
@@
 
-\emph{Stencils} are a ubiquitous programming pattern, common in
-scientific and numerical computing applications. Informally, a stencil
-computation computes an array whose value at each index $i$ is
-calculated from a \emph{neighbourhood} of values around $i$ in some
-input array(s), \eg{}, the Game of Life, convolutions in image
-processing, approximations to differential equations. For example, the
-following computes the one-dimensional discrete Laplace transform (an
-approximation to a derivative) in Fortran:
+\emph{Stencils} are a ubiquitous programming pattern, common in scientific and
+numerical computing applications. Informally, a stencil computation computes an
+array whose value at each index $i$ is calculated from a \emph{neighbourhood}
+of values around $i$ in some input array(s), \eg{}, the Game of Life,
+convolutions in image processing, approximations to differential equations. For
+example, the following computes the one-dimensional discrete Laplace transform
+(an approximation to a derivative) in Fortran:
 %
@@
 %
-Values \texttt{b(i)} are calculated from a neighbourhood of elements
-around \texttt{i} in the input array \texttt{a}.  In this example, the
-access pattern is simple and easily understood. More complex
-stencil computations are more prone to errors from simple lexical
-mistakes. For example, the following is a small snippet from a
-Navier-Stokes fluid simulator (based on~\citet{griebel1997numerical})
- in which two arrays are read with
-different data access patterns, across two dimensions.
+Values \texttt{b(i)} are calculated from a neighbourhood of elements around
+\texttt{i} in the input array \texttt{a}. In this example, the access pattern
+is simple and easily understood. More complex stencil computations are more
+prone to errors from simple lexical mistakes. For example, the following is a
+small snippet from a Navier-Stokes fluid simulator (based
+on~\citet{griebel1997numerical}) in which two arrays are read with different
+data access patterns, across two dimensions.
 %%
@@
 %%
-The access pattern is much harder to understand than the
-one-dimensional Laplace.  The miasma of indexing expressions of the
-form \texttt{var(i$\pm$a, j$\pm$b)} is not only hard to read, but is
-prone to simple textual input mistakes, \eg{}, swapping \texttt{-} and
-\texttt{+}, missing an indexing term, or transforming the wrong
-variable \eg{} \texttt{(i+1,j)} instead of \texttt{(i,j+1)}.
-
-In practice, the typical development procedure for complex
-stencil computations involves some ad hoc testing to ensure that no
-transcription mistakes have been made (\eg{}, by visual
-inspections, or comparison against an manufactured or 
-analytical solutions~\cite{farrell2010automated}). Typically such
-testing is discarded once the code is seen to be correct.
-
-This is not the only information that is often discarded. The
-``shape'' of the indexing pattern is usually the result of choices made in the
-numerical-analysis procedure to convert some
-continuous equations into a discrete approximation. Rarely are these
-decisions captured in the source code, yet the shape of access is
-usually uniform and is has a perspicuous and concise description in
-numerical analysis literature \eg{}, ``\emph{centered in space, to a
-  depth of 1}'' referring to indexing terms \texttt{a(i)},
-\texttt{a(i-1)} and \texttt{a(i+1)}.
-
-To support the development of correct stencil computations, we propose
-a simple, abstract specification language for the data access pattern
-of stencils. The shape of the
-Laplace example stencil is specified in our language as:
+The access pattern is much harder to understand than the one-dimensional
+Laplace. The miasma of indexing expressions of the form \texttt{var(i$\pm$a,
+j$\pm$b)} is not only hard to read, but is prone to simple textual input
+mistakes, \eg{}, swapping \texttt{-} and \texttt{+}, missing an indexing term,
+or transforming the wrong variable \eg{} \texttt{(i+1,j)} instead of
+\texttt{(i,j+1)}.
+
+In practice, the development process for complex stencil computations involves
+ad hoc testing to ensure that no transcription mistakes have been made (\eg{},
+by visual inspections, or comparison against an manufactured or analytical
+solutions~\cite{farrell2010automated}). Typically such testing is discarded
+once the code is seen to be correct.
+
+This is not the only information that is often discarded. The ``shape'' of the
+indexing pattern is usually the result of choices made in the
+numerical-analysis procedure to convert some continuous equations into a
+discrete approximation. Rarely are these decisions captured in the source code,
+yet the shape of access is usually uniform and is has a perspicuous and concise
+description in numerical analysis literature \eg{}, ``\emph{centered in space,
+to a depth of 1}'' referring to indexing terms \texttt{a(i)}, \texttt{a(i-1)}
+and \texttt{a(i+1)}.
+
+To support the development of correct stencil computations, we propose a
+simple, abstract specification language for the data access pattern of
+stencils. The shape of the Laplace example stencil is specified in our language
+as:
 %
@@
 %
-That is, \texttt{a} is accessed with a symmetrical pattern
-(``centered'') to a depth of one in each direction in its first
-dimension.  The Navier-Stokes example has a shape
-specified as:
+That is, \texttt{a} is accessed with a symmetrical pattern (``centered'') to a
+depth of one in each direction in its first dimension. The Navier-Stokes
+example has a shape specified as:
 %
@@
 %
-The specification requires that, over the whole fragment, \texttt{u}
-is accessed with a centered pattern to depth of 1 in both dimensions
-(this is known as the \emph{five-point stencil}) and \texttt{v} is
-accessed in a neighbourhood bounded forwards to depth of $1$ in the
-first dimension and backward to a depth of $1$ in the second
-dimension. The specification is relatively small
-and much more abstract compared to the source code.
-
-We provide a verification tool which can check the spatial correctness
-of stencil computations against a specification. Furthermore, the tool
-can automatically infer specifications and insert these as comments
-into the source code. The main features and benefits of our approach
-are as follows:
+The specification requires that, over the whole fragment, \texttt{u} is
+accessed with a centered pattern to depth of 1 in both dimensions (this is
+known as the \emph{five-point stencil}) and \texttt{v} is accessed in a
+neighbourhood bounded forwards to depth of $1$ in the first dimension and
+backward to a depth of $1$ in the second dimension. The specification is
+relatively small and much more abstract compared to the source code.
+
+We provide a verification tool which can check the spatial correctness of
+stencil computations against a specification. Furthermore, the tool can
+automatically infer specifications and insert these as comments into the source
+code. The main features and benefits of our approach are as follows:
 %
 \begin{itemize}
-%\item The main use case is that a programmer writes a specification
-%  first (perhaps informed by their numerical analysis), then writes
-%  the corresponding source code, and lastly has the tool check its
-%  conformance to the specifications. This specify-and-check approach
-%  removes the usually ad hoc testing effort, and provides the
-%  opportunity for easy re-verification in the event of any refactoring
-%  of software maintenance activities.
-\item The primary use case is that a programmer writes a specification
-  first before the stencil code. Our tool then checks conformance between
-  the two. This specify-and-check approach reduces testing effort and
-  future-proofs against bugs introduced during refactoring and maintenance.
-%
-%\item The specification provides a concise description which can be
-%  preserved in the source code to provide documentation to future
-%  developers, contrasting with ad hoc tests which are usually
-%  discarded or specialised to the development environment. Separate from the
-%  verification procedure, the infer-and-synthesise mode of the tool
-%  provides a way to automatically generate additional useful
-%  documentation for future users.
-
-% Emphasis here is on documentation generation, so it makes little sense to
-% compare this aspect of CamFort to ad hoc testing. Apples and oranges.
-\item A specification concisely captures the access pattern of a
-  stencil and turns it into documentation. The inference procedure
-  efficiently produces specifications with no programmer intervention,
-  automatically inserting specifications at appropriate places in the
-  source code.
-%
-%\item Specifications are integrated into the code as comments for easy
-%  integration with existing tools, as well as documentation systems (\eg{}, we
-%  support Doxygen and FORD syntax).
+\item The primary use case is that a programmer writes a specification first
+  before the stencil code. Our tool then checks conformance between the two.
+  This specify-and-check approach reduces testing effort and future-proofs
+  against bugs introduced during refactoring and maintenance.
+%
+\item A specification concisely captures the access pattern of a stencil and
+  turns it into documentation. The inference procedure efficiently produces
+  specifications with no programmer intervention, automatically inserting
+  specifications at appropriate places in the source code.
 %
@@
   combinators that \emph{do not involve any indexing expressions}, \eg{}
-  \texttt{a(i+1,j-1)}. This contrasts with other specification
-  approaches, \eg{}, deductive verification tools such as ACSL, where
-  specifications must also be in terms of array-indexing expressions. Therefore, any low-level mistakes
-  that could be made whilst programming complex indexing code could also be
-  made when writing its specification. Our specifications are more
-  abstract and lightweight
-  with the aim of aiding adoption by scientists and preventing indexing errors.
+  \texttt{a(i+1,j-1)}. This contrasts with other specification approaches,
+  \eg{}, deductive verification tools such as ACSL, where specifications must
+  also be in terms of array-indexing expressions. Therefore, any low-level
+  mistakes that could be made whilst programming complex indexing code could
+  also be made when writing its specification. Our specifications are more
+  abstract and lightweight with the aim of aiding adoption by scientists and
+  preventing indexing errors.
 \end{itemize}
 
-Our verification tool does not target a
-class of bugs that can be detected automatically (\emph{push-button
-  verification}). Instead, stencil computation bugs must be
-identified relative to a specification of the intended access pattern.
-
-In this paper, we make the following contributions:
+Our verification tool does not target a class of bugs that can be detected
+automatically (\emph{push-button verification}). Instead, stencil computation
+bugs must be identified relative to a specification of the intended access
+pattern.
+
+In this paper, we solely focus on the definition and the usage of the language
+and the acoompanying verification tool and omit the underlying mathematical
+model and implementation details due to space considerations. We make the
+following contributions:
 %
 \begin{itemize}
-\item We introduce a specification language for stencils
-  that captures common forms of data access patterns
-  (\cref{sec:lang});
-
-\item We detail checking and inference algorithms for our
-  specifications (\cref{sec:analysis}), derived from a
-  denotational model for the specification language
-  (\cref{sec:semantics});
-
+\item We introduce a specification language for stencils that captures common
+  forms of data access patterns (\cref{sec:lang});
+%
 \item We provide an implementation of our approach as an extension to
   CamFort~\cite{camfort}, an open-source program analysis tool for Fortran;
-
-\item We report on a quantitative study of stencil computations on a
-  corpus of numerical Fortran program
-  (\cref{sec:evaluation}), totalling one million lines of code. Our tool
-  identifies and infers specifications for roughly 60,000 stencil computations in the
-  corpus. Approximately 5,000 of the stencils we found are non-trivial,
-  corresponding to code which is a possible source of errors. This validates
-  our hypothesis that the majority of stencil computations have a regular
-  shape, and validates the design of our language in its capability to capture
-  many core patterns.
-
-\item We give a verification case study of a
-  stencil computation commonly used in scientific computing,
-  simulating programming errors (\Cref{sec:jacobi}). Our approach detects all
-  possible simulated programming errors for this problem.
+%
+\item We report on a quantitative study of stencil computations on a corpus of
+  numerical Fortran program (\cref{sec:evaluation}), totalling one million
+  lines of code. Our tool identifies and infers specifications for roughly
+  60,000 stencil computations in the corpus. Approximately 5,000 of the
+  stencils we found are non-trivial, corresponding to code which is a possible
+  source of errors.
+%
+\item We give a verification case study of a stencil computation commonly used
+  in scientific computing, simulating programming errors (\Cref{sec:jacobi}).
+  Our approach detects all possible simulated programming errors for this
+  problem.
 \end{itemize}
@@
 
-Our specification language is based on the hypothesis
-that most forms of array access in numerical code have
-a fixed, statically-determined access pattern. For example, the
-\emph{five-point stencil} on a two-dimensional array reads from array
-indices $(i, j)$, $(i-1, j)$, $(i+1, j)$, $(i, j-1)$, and $(i, j+1)$
-for all $i, j$ within the inner boundary of the array (to avoid
-out-of-bounds access at the edges). We revisit this hypothesis
-in \cref{sec:evaluation} with the inference of
-such regular stencil patterns on our corpus of numerical programs.
-
-We give various definitions used in the paper then
-\cref{subsec:syntax} outlines the syntax of stencil specifications
-with an informal introduction. \cref{sec:semantics}
-gives a semantic model via a set-based interpretation which makes
-concrete the meaning of specifications. \cref{sec:eqs} then provides
-a syntactic equational theory for specifications which is
-sound with respect to the model.
-%sub-specification for which the
-%equality and sub-specification relation are proven sound.
+Our specification language is based on the hypothesis that most forms of array
+access in numerical code have a fixed, statically-determined access pattern.
+For example, the \emph{five-point stencil} on a two-dimensional array reads
+from array indices $(i, j)$, $(i-1, j)$, $(i+1, j)$, $(i, j-1)$, and $(i, j+1)$
+for all $i, j$ within the inner boundary of the array (to avoid out-of-bounds
+access at the edges). We revisit this hypothesis in \cref{sec:evaluation} with
+the inference of such regular stencil patterns on our corpus of numerical
+programs.
+
+We give various definitions used in the paper then \cref{subsec:syntax}
+outlines the syntax of stencil specifications. \cref{sec:eqs} then provides a
+syntactic equational theory for specifications which is sound with respect to
+the underlying model.
 
@@
 
+\texttt{BEGIN: NEEDS REWRITING}
 \begin{definition}[Stencil computations]
 \label{def:stencil}
-  Let $a$ be an array of dimensionality $n$ and $\overline{b}$ a
-  collection of arrays of arbitrary (possibly differing)
-  dimensionalities.
-  A \emph{stencil computation} comprises an iteration by a set $I$ of at
-  most $n$ induction variables
-  over a subset of the index space of $a$. Each iteration
-  determines the elements of $a$ by an
-  assignment $a(\bar{e}) = e_r$ where each $e \in
-  \bar{e}$ is either a neighbourhood or constant index.
-  For all $b \in \bar{b}$ and array subscripts $b(\bar{e}')$
-  that flow to $e_r$, each $e' \in \bar{e}'$ is
-  either a neighbourhood or absolute index.
+  Let $a$ be an array of dimensionality $n$ and $\overline{b}$ a collection of
+  arrays of arbitrary (possibly differing) dimensionalities.  A \emph{stencil
+  computation} comprises an iteration by a set $I$ of at most $n$ induction
+  variables over a subset of the index space of $a$. Each iteration determines
+  the elements of $a$ by an assignment $a(\bar{e}) = e_r$ where each $e \in
+  \bar{e}$ is either a neighbourhood or constant index.  For all $b \in
+  \bar{b}$ and array subscripts $b(\bar{e}')$ that flow to $e_r$, each $e' \in
+  \bar{e}'$ is either a neighbourhood or absolute index.
 
@@
 \end{definition}
-%
-% specifying the shape of indexing expressions for
-%arrays $\bar{v}$ that are simultaneously permitted and expected within
-%the code to which the specification is given.
-%
-%\marginpar{\dnote{Is it worth having this informal consistency here?}}
-
-For now we informally describe what it means for a stencil kernel to
-be consistent with a specification. This will be formalised in
-\Cref{sec:semantics} and algorithmically described in and
-\Cref{sec:analysis}.  Specifications are associated to a stencil
-kernel and are declared for one or more array variables $\bar{v}$
-within that kernel.  A stencil specification defines a \emph{region}
-associated to an array variable $a$ against which a stencil kernel
-should be \emph{consistent}. The region defined by a specification is
-an $n$-dimensional rectilinear shape in a discrete space crossing the
-origin (which represents a collection of indices which are just 
-induction variables). Consistency is then a
-two way requirement. All array subscripts on $a$ contributing to the
-kernel should be consistent with the region. Conversely, all parts of
-the region should have an index corresponding it.
+
+We informally describe what it means for a stencil kernel to be consistent with
+a specification. This will be algorithmically described in \Cref{sec:analysis}.
+Specifications are associated to a stencil kernel and are declared for one or
+more array variables $\bar{v}$ within that kernel.  A stencil specification
+defines a \emph{region} associated to an array variable $a$ against which a
+stencil kernel should be \emph{consistent}. The region defined by a
+specification is an $n$-dimensional rectilinear shape in a discrete space
+crossing the origin (which represents a collection of indices which are just
+induction variables). Consistency is then a two way requirement. All array
+subscripts on $a$ contributing to the kernel should be consistent with the
+region. Conversely, all parts of the region should have an index corresponding
+it.
+
+\texttt{END: NEEDS REWRITING}
 
@@
 
-\Cref{fig:syntax} gives the syntax of stencil specifications, which is
-detailed below. The entry point is the \nonterm{specification} production which
-splits into either a \emph{region declaration} or a \emph{specification
-declaration}.
-Regions comprise \emph{region constants} which are
-combined via region operators \term{+} and \term{*}.
+\Cref{fig:syntax} gives the syntax of stencil specifications, which is detailed
+below. The entry point is the \nonterm{specification} production which splits
+into either a \emph{region declaration} or a \emph{specification declaration}.
+Regions comprise \emph{region constants} which are combined via region
+operators \term{+} and \term{*}.
 
@@
 
-Region constants specify a finite contiguous region in a
-single dimension relative to the origin and are either \term{pointed},
-\term{forward}, \term{backward}, or \term{centered}. The region names are
-inspired by the scientific terminology, \eg{} the standard explicit method for
-approximating PDEs is known as the \emph{Forward Time, Centered Space} (FTCS)
+Region constants specify a finite contiguous region in a single dimension
+relative to the origin and are either \term{pointed}, \term{forward},
+\term{backward}, or \term{centered}. The region names are inspired by the
+scientific terminology, \eg{} the standard explicit method for approximating
+PDEs is known as the \emph{Forward Time, Centered Space} (FTCS)
 scheme~\citep{dawson1991finite}.
 
-Each region
-constant has a dimension identifier $d$ given as a positive natural number.
-Each constant except \term{pointed} has a depth
-parameter $n$ given as a positive natural number; \term{pointed}
-regions implicitly have a depth of $0$.
-
-A \term{forward} region of depth $n$ specifies a contiguous
-region in dimension $d$ starting at the origin. This corresponds
-to specifying neighbourhood indices in dimension $d$ ranging from $i$ to $i + n$
-for some induction variable $i$. Similarly, a
-\term{backward} region of depth $n$ corresponds to contiguous indices
-from $i$ to $i - n$ and \term{centered}
-of depth $n$ from $i - n$ to $i + n$. A \term{pointed}
-stencil specifies a neighbour index $i$. For example, the
-following shows four specifications with four consistent stencil
+Each region constant has a dimension identifier $d$ given as a positive natural
+number.  Each constant except \term{pointed} has a depth parameter $n$ given as
+a positive natural number; \term{pointed} regions implicitly have a depth of
+$0$.
+
+A \term{forward} region of depth $n$ specifies a contiguous region in dimension
+$d$ starting at the origin. This corresponds to specifying neighbourhood
+indices in dimension $d$ ranging from $i$ to $i + n$ for some induction
+variable $i$. Similarly, a \term{backward} region of depth $n$ corresponds to
+contiguous indices from $i$ to $i - n$ and \term{centered} of depth $n$ from $i
+- n$ to $i + n$. A \term{pointed} stencil specifies a neighbour index $i$. For
+example, the following shows four specifications with four consistent stencil
 kernels reading from arrays \term{a}, \term{b}, \term{c} and \term{d}:
@@
 \begin{minted}{fortran}
-!= stencil forward(depth=2, dim=1) :: a
-e(i, 0) = a(i, 0) + a(i+1, 0) + a(i+2, 0)
-!= stencil backward(depth=2, dim=1) :: b
-e(i, j) = b(i, j) + b(i-1, j) + b(i-2, j)
-!= stencil centered(depth=1, dim=2) :: c
+!= stencil backward(depth=2,dim=1) :: a
+e(i, 0) = a(i, 0) + a(i-1, 0) + a(i-2, 0)
+!= stencil forward(depth=2,dim=1)*pointed(dim=2) :: b
+e(i, j) = b(i, j) + b(i+1, j) + b(i+2, j)
+!= stencil centered(depth=1,dim=2) :: c
 e(i, j) = (c(j-1) + c(j) + c(j+1))/3.0
@@
 %%
-Not every dimension needs to be specified, \eg{},
-specifications on lines $1$, $3$, and $7$ leave some dimensions unspecified.
-
-The \term{forward}, \term{backward}, and \term{centered} regions may
-all have an additional attribute \term{nonpointed} which marks absence
-of the origin.  For example, the following is a
-\term{nonpointed} \term{backward} stencil
+Every dimension with neighbourhood indices need to be specified. The remaining
+non-conforming indices are not included in the specification, \eg{},
+specifications on lines $1$ and $7$ leave some dimensions unspecified since they
+do not conform to definition neighbourhood indices.
+
+The \term{forward}, \term{backward}, and \term{centered} regions may all have
+an additional attribute \term{nonpointed} which marks absence of the origin.
+For example, the following is a \term{nonpointed} \term{backward} stencil
 %
@@
 
-The region operators \term{+} and \term{*} respectively combine
-regions disjunctively and conjunctively.
-
-The conjunction of two regions $r \term{*} s$ means that any indices
-in the specified code must be consistent with both $r$ and $s$
-simultaneously. Dually, for every pair of components in the regions
-$r$ and $s$ there must be an index consistent with both.
-For example, the following \emph{nine-point stencil}
-has a specification given by the product of two \texttt{centered}
-regions in each dimension:
+The region operators \term{+} and \term{*} respectively combine regions
+permissively and restrictively.
+
+The restrictive combination of two regions, $r \term{*} s$, means that any
+indices in the specified code must be consistent with both $r$ and $s$
+simultaneously. Dually, for every pair of components in the regions $r$ and $s$
+there must be an index consistent with both. For example, the following
+\emph{nine-point stencil} has a specification given by the product of two
+\texttt{centered} regions in each dimension:
 %%
@@
 %%
-This pattern is common in image convolution applications.
-The specification ranges over the
-values that flow to the array subscript on the left-hand side,
-and so ranges over the intermediate assignments to \term{x},
-\term{y}, and \term{z}. Each index in the code is consistent
-with both specifications simultaneously, \eg{}, \texttt{a(i-1, j+1)}
-is within the centered region in dimension $1$ and the centered region
-in dimension $2$.
-
-The product $r \term{*} s$ can also be thought of as a bounding box
-over the two regions $r$ and $s$.
-
-The disjunction of two regions $r \term{+} s$ means that any indices
-in the specified code must be consistent with either of $r$ or $s$.
-Dually, every part of region $r$ must have a
-corresponding index and similarly for $s$ independently. For
-example, the following gives the specification of a five-point
-stencil which is the sum of two compound \texttt{pointed} and
+This pattern is common in image convolution applications. The specification
+ranges over the values that flow to the array subscript on the left-hand side,
+and so ranges over the intermediate assignments to \term{x}, \term{y}, and
+\term{z}. Each index in the code is consistent with both specifications
+simultaneously, \eg{}, \texttt{a(i-1, j+1)} is within the centered region in
+dimension $1$ and the centered region in dimension $2$.
+
+The product $r \term{*} s$ can also be thought of as a bounding box over the
+two regions $r$ and $s$.
+
+The permissive combination of two regions $r \term{+} s$ means that any indices
+in the specified code must be consistent with either of $r$ or $s$. Dually,
+every part of region $r$ must have a corresponding index and similarly for $s$
+independently. For example, the following gives the specification of a
+five-point stencil which is the sum of two compound \texttt{pointed} and
 \texttt{centered} regions in each dimension:
@@
 Here the left-hand side of \texttt{+} says that when the second dimension
-(induction variable $j$) is fixed at the origin, the first dimension
-(induction variable $i$) accesses the immediate vicinity of the origin
-(to depth of one). The right hand side of \texttt{+} is similar but the dimensions are reversed.
-This reflects the symmetry under rotation of the five-point stencil.
+(induction variable $j$) is fixed at the origin, the first dimension (induction
+variable $i$) accesses the immediate vicinity of the origin (to depth of one).
+The right hand side of \texttt{+} is similar but the dimensions are reversed.
+This reflects the symmetry under rotation of the five-point stencil. As
+presented in \cref{sec:evaluation}, this symmetry is common to stencil
+computations used in practice, which makes it easier to write specifications.
 
@@
 %
-This specification would be invalid if any of the
-array subscripts were repeated. This modifier provides a way to
-rule out any accidental repetition of array subscripts.
-The notion is similar to that of linear types~\cite{wadler1990linear}, where a value must be used
-exactly once. We opt for the more informative and easily understood name
-\texttt{readOnce}. This modifier is optional, so it need not
-be present even if the stencil is linear.
-
-In some cases, it is useful to give a lower and/or upper bound for a
-stencil. This can be done using either the \term{atMost} or
-\term{atLeast} modifiers. This is particularly useful in situations
-where there is a non-contiguous stencil pattern, which cannot be expressed
-precisely in our syntax. For example:
+This specification would be invalid if any of the array subscripts were
+repeated. This modifier provides a way to rule out any accidental repetition of
+array subscripts.  The notion is similar to that of linear
+types~\cite{wadler1990linear}, where a value must be used exactly once. We opt
+for the more informative and easily understood name \texttt{readOnce}. This
+modifier is optional, so it need not be present even if the stencil is linear.
+
+In some cases, it is useful to give a lower and/or upper bound for a stencil.
+This can be done using either the \term{atMost} or \term{atLeast} modifiers.
+This is particularly useful in situations where there is a non-contiguous
+stencil pattern, which cannot be expressed precisely in our syntax. For example:
 %
@@
 
-\section{A model for specifications}
-\label{sec:semantics}
-
-We define a simple set-theoretic denotational model of the semantics
-of our specification language. This has a number of purposes:
-(1) it serves to explain the meaning of our specifications
-(2) it is used in the inference and checking algorithms
-(\Cref{sec:analysis}); (3) justifies an equational theory for
-specifications in the next section; and (4) can be used to guide
-correct implementations. The model domain is over sets of vectors
-which we call \emph{index schemes}.
-
-\begin{definition}
-  An \emph{index scheme} is a vector of size $n$ (an $n$-vector) with
-  values drawn from $\domainVal = \mathbb{Z} \cup \{\infty\}$.  Values
-  in $\mathbb{Z}$ represent the offsets of neighbourhood indices
-  (\ie{}, the constant integer $c$ in an indexing expression of the
-  form $i \pm c$, \Cref{def:neighbour}) and $\infty$ is a wild-card
-  representing both constant indices and unspecified indices.
-  For an index scheme $u$, the $i^{th}$ element is denoted $u_i$. We
-  write $(\domainVal)^n$ for the set of all $n$-dimensional schemes.
-\end{definition}
-
-Index schemes abstract array subscripts.
-Our model consists of index schemes explaining the range of possible
-indexing behaviours which are consistent with the specification.
-The relationship between source language
-and index schemes
- is defined by \textsf{schematic} function,
- which forms part of checking and inference (\Cref{sec:analysis}).
-
-\newcommand{\schemeOne}{\textsf{s}_I}
-%\newcommand{\schemeOne}{\textsf{schema$^1_I$}}
-\newcommand{\schemeN}{\textsf{schematic}_I}
-%\newcommand{\schemeN}{\textsf{schema$^n_I$}}
-\begin{definition}%[Schematic]
-\label{def:schematic}
-Let $\schemeOne$ be a partial function, parameterised by a set of
-induction variables $I$, mapping source language expressions to
-$\domainVal$ values:
-%%
-\begin{align*}
-\schemeOne(e)
- & =  \begin{cases}
-c & e \equiv i + c \;\; \wedge \;\; i \in I \\
--c & e \equiv i - c \;\; \wedge \;\; i \in I \\
-\infty & \textit{$I \cap \textit{FreeVariables}(e) = \emptyset $} %\\
-%\bot   & \textit{otherwise}
-\end{cases}
-\end{align*}
-In the
-first two cases, expressions that are a constant offset of an
-induction variable (\ie{}, neighbourhood indices) are mapped to the
-offset $c$ or $-c$ $\in \domainVal$. In the third case, if the
-expression is not defined in terms of an induction variable, it
-is mapped to $\infty \in \domainVal{}$. Otherwise, $\schemeOne(e)$
-is undefined ($\bot$).
-
-The partial function $\schemeN$, also parameterised by a set of
-induction variables $I$, uses $\schemeOne$ to
- map from the syntax of indices to index schemes
-$\in (\domainVal)^n$. It is defined on the syntax of indices $\bar{e} = (e_1, ..., e_n)$ as:
-\begin{align*}
-\schemeN(e_1, .., e_n) & =
-[\schemeOne(e_1) ... \schemeOne(e_n) ]
-\; \textit{iff} \; \forall x . \schemeOne(e_x) \neq \bot
-\end{align*}
-%
-Thus, if each expression in an index is a neighbourhood or constant
-index, $\schemeN$ maps it to an index scheme.
-\end{definition}
-
-\begin{definition}[Generalisation]
-\label{def:generalisation}
-Given two schemes $v, w$ of dimensionality $n$,
-then $w$ is said to \emph{generalise} $v$
-(or conversely $v$ \emph{specialises} $w$),
-written $v \preceq_n w$ defined:
-\[
-v \preceq_n w \;\; \Leftrightarrow  \;\;
-  \forall i\!\in\!\{ 1, \ldots, n \} . \, (w_i = \infty \, \vee \, v_i = w_i)
-\]
-%%
-where $\preceq_n$ is a reflexive, transitive relation on
-index schemes $\domainVal^n$. Thus,
- $\infty$ generalises any index in the same dimension.
-%
-\end{definition}
-
-\begin{example}
-Consider the following array subscript to index schemes
-conversions:
-%
-\begin{align*}
-\begin{array}{ll}
-\schemeN{(i, j+1)} & =\vtwoh{0}{1} \\
-\schemeN{(1, j+1)} & = \vtwoh{\infty}{1} \\
-\schemeN{(i*2, j)} & = \bot
-\end{array}
-\end{align*}
-There is one generalisation between these
-schemes, namely that $\vtwoh{0}{1} \preceq_2 \vtwoh{\infty}{1}$.
-The last example is undefined as its first expression is a
-non-neighbour relative index.
-\end{example}
-
-%Our terminology of \emph{index schemes} and
-%\emph{generalisation} is by analogy
-%with the \emph{type schemes} and \emph{type generalisation} of
-%polymorphic types in ML~\cite{milner1978theory}. Our notion is
-%much more rigid however, as index schemes do not bind names.
-
-\paragraph{Shorthand notation}
-
-We use abbreviations
- \term{fwd}, \term{bwd}, \term{cen}, and \irreflS{}
-for \term{forward}, \term{backward},
-\term{centered}, and \irrefl{} respectively. Furthermore, 
-we quantify over the absence or presence of the \irrefl{}
-attribute on region constants as a boolean. For example,
-\term{fwd} is syntactic sugar defined:
-%%
-\begin{align*}
-\hspace{-0.5em}
-& \stenFwdSR{n}{d}{p}
-\; := \; \\[-0.4em]
-& \quad \begin{cases}
-\stenFwd{n}{d}{, \irrefl} & \textit{if} \;\, \neg \, p \\[-0.1em]
-\stenFwd{n}{d}{}  & \textit{if} \;\,  p
-\end{cases}
-\end{align*}
-%
-$\textit{or} \;\; \stenFwdS{n}{d} := \stenFwd{n}{d}{}$
-and similarly for \term{bwd} and \term{cen}.
-
-\subsection{Denotational model}
-
-An interpretation function $\interp{-}_n$ provides the model 
-(where $n$ is the maximum dimensionality of the specification being
-modelled) mapping closed\footnote{That is, we assume there are no
-  occurrences of \textit{rvar} in a specification being modelled.  Any
-  \emph{open} specification containing region variables can be made
-  closed by straightforward syntactic substitution with a (closed)
-  \textit{region}.} specifications to sets of
-$n$-dimensional index schemes. The interpretation is overloaded on
-\emph{regions} in \Cref{fig:region-model} and on the top-level
-of a specification \textit{spec} in \Cref{fig:spatial-model}.
-Various intermediate notions are used.
-
-\begin{definition}A \emph{single-entry vector} of size $n$, denoted
-$\singleEntry{r}{n}$, is a vector where the $r^{th}$ entry is $1$
-and all others are $\infty$, \eg{}, $\singleEntry{1}{2} =
-\vtwoh{1}{\infty}$ and $\singleEntry{2}{3} = \vthreeh{\infty}{1}{\infty}$.
-Thus, a single-entry vector is an index scheme generalising 
-all other schemes that have $1$ as their $r^{th}$ entry,
-describing a neighbour offset of $1$.
-\end{definition}
-
-\begin{definition}
-A \emph{zeroed single-entry vector} of size $n$ denoted
-$\zeroEntry{r}{n}$ is a vector where the $r^{th}$ entry is $0$ and all others
-are $\infty$ \eg{}, $\zeroEntry{1}{2} = \vtwoh{0}{\infty}$.
-Akin to single-entry vectors, zeroed single-entry vectors
-generalise all other schemes
-that have $0$ as their $r^{th}$ entry, representing indices at
-the ``origin'' in $r$.
-\end{definition}
-
-\begin{figure}[t]
-\vspace{-0.9em}
-\input{region-model.tex}
-%\textit{$a^m \in A$ means there are $n$ copies of $a$ in
-%  the multi-set $A$}
-\caption{Model of regions,
-$\interp{-}_n : \textit{region} \rightarrow \mathcal{P}(\domainVal{}^n)$}
-\label{fig:region-model}
-\vspace{-1em}
-\end{figure}
-
-\noindent
-The first four equations of \Cref{fig:region-model}
-model region constants as sets of indexing
-schemes. We illustrate with an example.
-
-\begin{example}
-For a $2$-dimensional stencil computation, then
-\begin{align*}
-& \interp{\stenCenS{2}{1}}_2 \\[-0.1em]
-& \qquad = \; \{i\singleEntry{1}{2} \,|\, i \in \{\text{-}2,..., 2\}\!\setminus\!0 \}
-  \cup  \{ \zeroEntry{1}{2} \,|\, \textit{true} \} \\[-0.2em]
-& \qquad = \; \{-2\singleEntry{1}{2}, \, -\singleEntry{1}{2}, \,
-  \singleEntry{1}{2}, \, 2\singleEntry{1}{2}\} \cup
-  \{\zeroEntry{1}{2}\} \\[-0.2em]
-& \qquad = \; \{\vtwoh{-2}{\infty}, \vtwoh{-1}{\infty}, \vtwoh{0}{\infty},
-\vtwoh{1}{\infty}, \vtwoh{2}{\infty}\}
-\end{align*}
-Note, $\infty$ is absorbing for multiplication above \eg{} $2
- \infty = \infty$.
-\end{example}
-\noindent
-The final two equations of \Cref{fig:region-model} give the models
-of $\term{+}$ and $\term{*}$. For $\term{+}$ this is the
- union of the models of subterms. The model of
-$\term{*}$ takes a ``tensor'' $\otimes$ of subterm models. The tensor is
-more involved. We give some informal intuition first.
-
-% specifications correspond to a set of schemes
-
-% r+s means an index must be generalised by a scheme in either r or s
-%       and every scheme in r must have an index which specialises it
-%       and every scheme in s must have an index which specialises it
-
-% r*s means an index must be generalised by a scheme in r and a scheme in s
-%       and every pairs of schemes in r and s must have an index which
-%           specialises both
-%
-
-% r+s means indices must be consistent with either r or s
-%           and there are indices why are generalised by r and s
-% (there are indices which are consistent with every part of r or s)
-% (every part of r and s independently have an index which is generalised by it)
-
-% r*s means indices must be consistent simultaneously with both r and s
-%           and there are indices which are generalised
-%
-
-\begin{figure}
-\vspace{-1em}
-\input{spatial-model.tex}
-\caption{%\textit{spec} model,
-$\interp{-}_n : \textit{spec} \rightarrow \textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$}
-\label{fig:spatial-model}
-\vspace{-1em}
-\end{figure}
-
-\paragraph{$\otimes$ of models}
-
-The intuition behind $\otimes$ is that it takes all possible pairs of
-index schemes, and treats each pair as lower and upper bounds of an
-$n$-dimensional rectangle. From these bounds,
-the remaining vertices of the $n$-dimensional rectangle are
-generated (like a bounding box) \eg{}:
-%%
-\begin{align*}
-{\small{
-\{\vtwohs{1}{2},\vtwohs{5}{6}\,\}\!\otimes\!\{\vtwohs{3}{4}\}\!=\!\{\vtwohs{1}{2},\vtwohs{3}{2},\vtwohs{1}{4},\vtwohs{3}{4},
-   \vtwohs{3}{6},\vtwohs{5}{4},\vtwohs{5}{6}\,\}
-}}
-\end{align*}
-%%
-The tensor takes a Cartesian product of the two models,
-giving all pairs of schemes which are combined using the
-\emph{pairwise permutation} written $\bowtie$ (defined later).
-Pairwise permutation generates all possible interleavings of two
-vectors. Some care is then taken over how this interacts with $\infty$.
-In the presence of $\infty$, generate all interleavings as if $\infty$ is
-an integer, then filter out vectors with $\infty$ at any coordinate, \eg{}
-\begin{align*}
-\{\vtwoh{1}{\infty}\} \otimes \{\vtwoh{3}{4}\} =
-\{\vtwoh{1}{4}, \vtwoh{3}{4}\}
-\end{align*}
-%
-where $\vtwoh{1}{\infty}$ and $\vtwoh{3}{\infty}$ have been filtered.
-
-\begin{definition}%[Constrained dimensions]
-For an $n$-dimensional model $M$, its \emph{constrained dimensions}
-are the set of dimension identifiers where at least one scheme $u \in M$ has
-a non-$\infty$ value in that dimension:
-%
-\begin{align*}
-\effdims{M}{n} =
-\bigcup_{u \in M} \{i \mid i \in \{1,\ldots,n\} \, \wedge \, u_i \neq
-  \infty\}
-\end{align*}
-For example, $\effdims{\{\vtwoh{0}{\infty},\vtwoh{\infty}{\infty}\}}{2}
- = \{1\}$.
-\end{definition}
-\begin{definition}The tensor $\otimes_n$ of $n$-dimensional
-models is:
-\begin{alignat*}{2}
-& M \otimes_n N =
-  \Bigg\{x \; \Bigg| \;
-    \parbox{5.2cm}{$(u, v) \in M \times N,$ \\[0.1em]
-                  $\;\; i \in \{1, \ldots, 2^n\}, \,
-                  x = (u \bowtie v)_i,$ \\[0.1em]
-                  $\;\; j \in (\effdims{M}{n} \cup
-                   \effdims{N}{n}),$\\[0.1em]
-                   $\;\; x_j \neq \infty$
-                  } \Bigg\}
-\end{alignat*}
-The second guard binds $j$ to the constrained
-dimensions for $M$ unioned with the constrained dimensions for $N$.
-That is, every $j$ corresponds to a dimension in which
-either $M$ or $N$ has a scheme $u$ where $u_j \neq \infty$,
-and thus is constrained.
-The \emph{pairwise permutation} $u \bowtie v$
-builds a $2^n \times n$ matrix
-of all possible $n$-vectors generated by
-non-deterministically picking for each $j^{th}$ entry either
-$u_j$ or $v_j$ \eg{}
-%
-\begin{equation*}
-\vthreeh{0}{1}{2} \bowtie \vthreeh{3}{4}{5} =
-\left[
-\setlength{\arraycolsep}{0.42em}
-\begin{array}{cccccccc}
-0 & 0 & 0 & 0 & 3 & 3 & 3 & 3 \\
-1 & 1 & 4 & 4 & 1 & 1 & 4 & 4 \\
-2 & 5 & 2 & 5 & 2 & 5 & 2 & 5
-\end{array}
-\right]
-\end{equation*}
-%
-The $2^n$ unique choices for $\bowtie$ on $n$-vectors
-correspond to taking all bit-strings of length $n$ and
-selecting from $u$ for $1$ and $v$ for $0$. The pairwise permutation
-matrix is thus defined in terms of the logical
-matrix $b$ where $b_{i,j}$ is the $j$-th bit of the integer
-$i - 1$ as:
-%
-\begin{align*}
-(u \bowtie v)_{i,j} & = b_{i,j} u_j + \neg b_{i,j} v_j
-\end{align*}
-%
-The permuted schemes $x = (u \bowtie v)_i$ added to the model
-$M \otimes N$ are only those without $\infty$ in
-the constrained dimensions given by each model ($x_j \neq
-\infty$). This eliminates all permutations with
-$\infty$ in a constrained dimension position.
-\end{definition}
-
-\paragraph{Model of spec}
-
-Regions are modelled as sets of index
-schemes, that is
-$\interp{\textit{region}}_n \in \mathcal{P}((\domainVal{})^n)$.
-\Cref{fig:spatial-model} defines the top-level model of the
-\textit{spec} syntax, which includes the multiplicity and
-approximation modifiers. The domain of this interpretation
-is $\textsf{Mult}(\textsf{Approx}(\mathcal{P}((\domainVal{})^n)))$
-where $\textsf{Mult}$ and $\textsf{Approx}$ are labelled variants
-labelling our set-based model with additional information.
-
-\begin{definition} \textsf{Mult}
-and \textsf{Approx} are parametric labelled variant types
-with injections given by their definition:
-%
-\begin{align*}
-\textsf{Mult} \, a & = \textsf{mult} \, a \,\mid\, \textsf{only} \, a \\
-\textsf{Approx} \, a & = \textsf{exact} \, a \,\mid\, \textsf{lower} \, a \,\mid\,
-\textsf{upper} \, a \,\mid\, \textsf{both} \, a \, a
-\end{align*}
-\eg{}, $\mathsf{lower}$ is an injection
-$\mathsf{lower} : a \rightarrow \mathsf{Approx} \, a$ etc. The
-$\textsf{Mult}$ type corresponds to the presence or absence of the
-\term{readOnce} modifier as shown in \Cref{fig:spatial-model}.
-The $\textsf{Approx}$ type corresponds to the presence or absence of
-the spatial approximation modifier, with \textsf{exact} when there is
-no such modifier and \textsf{lower} and \textsf{upper} for
-\term{atLeast} and \term{atMost}. Later, it will be useful to model
-pairs of lower and upper bounds provided by \textsf{both} but which is
-not used
-here.% These are used for labelled models, and other components later in.
-\label{def:mult-and-approx}
-\end{definition}
-
 \section{Equational theory}
@@
 semantics for the language. This enables users to rewrite their specification
-and provides a way to automatically simplify
-specifications after inference (\Cref{subsec:inference}).
-Additionally, specifications have a theory of approximation,
-$<:$, that can be used by the users to weaken specifications for succinctness in
-the source.
-
-The equality relation $\equiv$ is defined over \emph{region}
-syntax which gives the following axioms:
+and provides a way to automatically simplify specifications after inference.
+Here we only focus on the former and omit mathematical implications of theory.
+Additionally, specifications have a theory of approximation, $<:$, that can be
+used by the users to weaken specifications for succinctness in the source.
+
+The equality relation $\equiv$ is defined over \emph{region} syntax which gives
+the following axioms:
 %
@@
                        \; \texttt{*} \; T)$.
-Distributivity is
-used later to give a normal form for specifications
-akin to Disjunctive Normal Form (\cref{sec:inf-step4}).
 
@@
                          \, \term{*} \, S$
-meaning that for the sum of products of two regions $R$ and $S$,
-where in each component one of $R$ or $S$ has the non-pointed
-attribute denoted by $R^{\term{np}}$ and $S^{\term{np}}$,
-the non-pointed attributes cancel to give $R \, \term{*} \, S$.
-%
-\item $(\textsc{$\ast$Dim}) \;\; R^d \, \term{$\ast$} \, S^d \equiv R^d
-                             \, \term{+} \, S^d$
-%
-where $R^d$ and $S^d$ denote regions whose dimension is $d$.
-Thus a product of two regions on
-the same dimension is equal to their addition.
+meaning that for the sum of products of two regions $R$ and $S$, where in each
+component one of $R$ or $S$ has the non-pointed attribute denoted by
+$R^{\term{np}}$ and $S^{\term{np}}$, the non-pointed attributes cancel to give
+$R \, \term{*} \, S$.
 \end{itemize}
 %
-Stencil specifications in \textit{spec} are considered
-equal when they have the same modifiers and $\equiv$ equates
-their regions.
-
-\begin{theorem}[Soundness of equational theory]
-\begin{equation*}
-\forall S, T, n . \quad
-S \equiv T \; \Rightarrow \;
-\interp{S}_n = \interp{T}_n
-\end{equation*}
-See the supplement for the proof.
-\end{theorem}
-
-\Cref{fig:inequations} defines a notion of sub-specifications
-via the relation $<:^r$ on \textit{region}
-syntax and then $<:$ on \textit{spec} syntax inductively.
-Both relations are reflexive and transitive (we omit
-these equations). The $<:^r$ relation is congruent
-with respect to \term{+} and \term{*} and $<:$ is congruent
-with respect to \term{readOnce}, the equations of which we omit for brevity.
-
-The \trule{\textsc{eq}} rule connects region
-equations to approximation; \trule{\textsc{over}} explains
-the notion of spatial over-approximation via overlapping regions;
-\trule{\textsc{rep}} allows the \term{readOnly} modifier that
-to be dropped. The \trule{\textsc{shrink}}
-and \trule{\textsc{grow}} rules turn the spatial approximation on
-regions into approximation on specifications via the \term{atLeast} and \term{atMost}
-modifiers. Finally, \trule{\textsc{Gen$\ast$}} shows
-that specifications can be generalised by removing parts of a product
-$\ast$ which have disjoint dimensions to the rest of the region
-product (where $\mathsf{dims}$ denotes the set of dimensions specified
-in a region).
-
-\begin{theorem}[Approximation soundness]
-Code consistent with a specification $S$
- is consistent with a more general specification $T$,
-\ie{}, for all models $M$ of a piece of code:
-\begin{equation*}
-S <: T \; \wedge (\consA{M}{\interp{S}_n}) \;
-\Rightarrow \; (\consA{M}{\interp{T}_n})
-\end{equation*}
-where $\consAName$ implements consistency in checking (\cref{subsec:checking}).
-See the supplement for the proof.
-\end{theorem}
+Stencil specifications in \textit{spec} are considered equal when they have the
+same modifiers and $\equiv$ equates their regions.
+
+\Cref{fig:inequations} defines a notion of sub-specifications via the relation
+$<:^r$ on \textit{region} syntax and then $<:$ on \textit{spec} syntax
+inductively.  Both relations are reflexive and transitive (we omit these
+equations). The $<:^r$ relation is congruent with respect to \term{+} and
+\term{*} and $<:$ is congruent with respect to \term{readOnce}, the equations of
+which we omit for brevity.
+
+The \trule{\textsc{eq}} rule connects region equations to approximation;
+\trule{\textsc{over}} explains the notion of spatial over-approximation via
+overlapping regions; \trule{\textsc{rep}} allows the \term{readOnly} modifier
+that to be dropped. The \trule{\textsc{shrink}} and \trule{\textsc{grow}} rules
+turn the spatial approximation on regions into approximation on specifications
+via the \term{atLeast} and \term{atMost} modifiers. Finally,
+\trule{\textsc{Gen$\ast$}} shows that specifications can be generalised by
+removing parts of a product $\ast$ which have disjoint dimensions to the rest of
+the region product (where $\mathsf{dims}$ denotes the set of dimensions
+specified in a region).
 
@@
 
-\section{Analysis, checking, and inference}
-\label{sec:analysis}
-
-\noindent
-We outline here the procedures for checking conformance
-of source code against specifications (\Cref{subsec:checking})
-and for inferring specifications from code (\Cref{subsec:inference}).
-Both rely on a program analysis that converts array subscripts
- into sets of index schemes. We outline this analysis
-first (\cref{subsec:analysis}). Note that the analysis 
-can be made arbitrarily more complex and wide-ranging independent
-of the checking and inference procedures. At the moment, the analysis
-is largely \emph{syntactic}, with only a small amount of
-semantic interpretation of the code.
-
-\subsection{Analysis of array accesses}
-\label{subsec:analysis}
-
-\newcommand{\neigh}{\textsf{neigh}}
-
-The analysis builds on standard program analyses:
-%
-\textbf{(1)} basic blocks (CFG);
-\textbf{(2)} induction variables per basic block;
-\textbf{(3)} (interprocedural) data-flow analysis, providing a \emph{flows to}
-  graph (as shorthand, the function
-  $\mathsf{flowTo}$ is used, implicitly parameterised by this graph,
-  mapping an expression to the set of all expressions
-  with forwards data-flow to this expression, based on the transitive
-  closure of the flows graph);
-\textbf{(4)} type information per variable.% where we use the predicate
-%\arrayTy{} to classify variables of array type.
-
-The core analysis, captured by the function $\textit{analyse}$ below, then proceeds as follows.
-For each assignment statement \textit{assg} within a loop whose left-hand side is an array
-subscript on neighbourhood indices, a finite map is computed mapping
-array variables to \emph{multisets} of index schemes:
-%
-\begin{align*}
-\hspace{-0.5em}
-\begin{array}{lr}
-\textit{analyse}(a(\overline{e_1}) = e_2)
- := \qquad\qquad\qquad & \textit{where} \; \neigh(\overline{e_1}) \\[0.3em]
-\multicolumn{2}{l}{\qquad \bigcup\{a' \mapsto \{
-\textsf{schematic}(\bar{e})\} \mid a'(\bar{e}) \leftarrow \mathsf{flowsTo}(e_2)\}}
-\end{array}
-\end{align*}
-%
-The predicate $\neigh(\bar{e_1})$ classifies indices where each
-$e \in \bar{e_1}$ is either a
-neighbourhood index (based on
-the induction variables of the loop, implicit here)
-or a constant index.\footnote{If all indices $e \in \bar{e}$ are constants, then
-  $\neigh(\bar{e})$ does not hold. Instead $a(\bar{e}$) is
-  considered a constant by our analysis, rather
-than an array subscript.}% contributing to a stencil.}
-Thus, we focus on assignments to an array subscript where the LHS
-indices $\overline{e_1}$ are neighbourhood indices.  For
-all array subscripts that flow to the right-hand side of this
-statement, a finite map is constructed, mapping each array variable
-to a multiset of index schemes for its subscripts, computed
-with \textsf{schematic} (\Cref{def:schematic}).
-Note \textsf{schematic} is undefined ($\bot$) if
-$\bar{e}$ contains relative indices which are not neighbourhood
-indices. The \textit{analysis} function is used by the
-\textit{modelise} function which filters out those arrays
-which contain undefined index schemes and adds multiplicity
-information:
-%
-\begin{align*}
-& \mathit{modelise} (\textit{assg})
-= \textit{let} \;\; M = \mathit{analyse}(\textit{assg}) (a) \;\; \textit{in}
-\\
-& a \mapsto
-\begin{cases}
-\bot & \bot \in M \\
-\textsf{only}(|M|) & \forall u \in M . \, \#(M, u) = 1 \\
-\textsf{mult}(|M|) & \forall u \in M . \, \#(M, u) = n
-\end{cases}
-\end{align*}
-where $\#(M, u)$ denotes the multiplicity of an element $u$ of the
-multiset $M$ and $|M|$ denotes the set of elements in the multiset
-$M$.
-Thus, the analysis maps source code to index schemes, which is then
-augmented with multiplicity information by $\mathit{modelise}$,
-\ie{} values in $\mathsf{Mult}(\mathcal{P}((\domainVal{})^n))$.
-
-As a small semantic concession to an otherwise syntactic
-analysis: if the LHS $\overline{e_1}$ contains non-$0$ neighbour offsets
-then all RHS schemes are \emph{relativised} by
-$\textsf{schematic}(\overline{e_1})$. For example,
-if \texttt{a(i+1) = b(i)} then relativisation of the RHS by the LHS
-produces an analysis which is equivalent to the analysis of \texttt{a(i) = b(i-1)}.
-
-\begin{example}
-\label{exm:checking}
-We demonstrate the analysis and modelise procedures over the
-following example kernel which computes the mean value
-of a five-point stencil:
-%%
-\begin{minted}{fortran}
-x = a(i-1, j)+a(i+1, j); y = a(i, j-1)+a(i, j+1)
-b(i, j) = (a(i, j) + x + y) / 5.0
-\end{minted}
-%
-%The \textit{analyse} function matches on
-%line 4 with the following set for $\textsf{flowsTo}(\texttt{a(i, j) + x +
-%  y}) = $
-%
-%\begin{align*}
-%{{ %{\small{
-%$\{$\texttt{a(i-1, j)}, \texttt{a(i+1, j)}, \texttt{a(i, j-1)},
-%  \texttt{a(i, j+1)}, \texttt{a(i, j)}$\}$}}. 
-%\end{align*}
-The  multiplicities are all one, and so \emph{modelise} yields:
-%%
-\begin{align*}
-\texttt{a} \mapsto \mathsf{only}(\{\vtwoh{-1}{0}, \vtwoh{1}{0},
-          \vtwoh{0}{-1}, \vtwoh{0}{1}, \vtwoh{0}{0}\})
-\end{align*}
-%%
-\end{example}
-
-%Given an assignment $\textit{assg}$ to an array subscript in the source language
-%and an array variable $a$, then $\mathit{modelise}(\textit{assg}, a)$ performs
-%an analysis to find all array subscripts of $a$ converting them to
-%index schemes. For the moment, we leave this analysis abstract
-%as the function $\emph{analyse}$ in the following definition of $\emph{modelise}$:
-%
-
-\subsection{Specification checking}
-\label{subsec:checking}
-
-Checking verifies the access pattern of a stencil computation in the
-source language against its associated specifications. The model of
-(\Cref{sec:semantics}) guides this process.  Checking
-proceeds by generating a model from a specification and generating
-a model from the source code, comparing them for
-\emph{consistency} base on index scheme
-\emph{generalisation} (\Cref{def:generalisation}).
-
-\begin{definition}[Checking]
-Let $a$ be an array variable and
-$S$ a specification associated to $a$
-via \texttt{!= stencil $S$ :: a}.
-
-A stencil kernel
- $v(\overline{e_1}) = e_2$
-conforms to $S$ if the predicate $\textbf{check}_n(S, v(\overline{e_1}) = e_2, a)$ holds,
-defined:
-\begin{align*}
-& \textbf{check}_n(S, \textit{assg}, a) =
-  \consSub{n}{\textit{modelise}(\textit{assg})(a)}{\interp{S}_n}
-&
-\end{align*}
-Thus, checking applies \textit{modelise}
-(\Cref{subsec:analysis}) to extract index schemes from source code,
-augmented with multiplicity for the whole set of schemes.
-This is compared with the model of the specification $\interp{S}_n$ by
-the consistency predicate $\consAName$.
-\end{definition}
-
-\begin{figure}[t]
-\vspace{-0.5em}
-\begin{align*}
-\hspace{-0.5em}
-\begin{array}{c}
-%%
-\trule{up}\dfrac{\forall u \in M, \exists v \in N .  \, u \preceq v}
-      {\cons{M}{\textsf{upper}(N)}}
-\,
-\trule{low}\dfrac{\forall v \in N, \exists u \in M . \, u \preceq v}
-      {\cons{M}{\textsf{lower}(N)}} \\[1.2em]
-%%
-\trule{exact}\dfrac{\cons{M}{\textsf{lower}(N)} \qquad \cons{M}{\textsf{upper}(N)}}
-  {\cons{M}{\textsf{exact}(N)}} \\[1.2em]
-%%
-\trule{both}\dfrac{\cons{M}{\textsf{lower}(M_1)} \qquad \cons{M}{\textsf{upper}(M_2)}}
-  {\cons{M}{\textsf{both}(M_1, M_2)}}\\[1.2em]
-%%
-\trule{$1$}
-\dfrac{\cons{M}{N}}{\consA{\textsf{once}(M)}{\textsf{once}(N)}}
-\;\;
-\trule{$*$}
-\dfrac{\cons{M}{N}}{\consA{\textsf{mult}(M)}{\textsf{mult}(N)}}
-  \\[1.2em]
-\trule{$1 < *$}
-\dfrac{\cons{M}{N}}{\consA{\textsf{once}(M)}{\textsf{mult}(N)}}
-\end{array}
-\end{align*}
-\vspace{-0.5em}
-\caption{Consistency of models (code vs. specification)}
-\label{fig:consistency}
-\vspace{-0.9em}
-\end{figure}
-
-\begin{definition}
-Consistency relates models of source code to models
-of specifications (which are bounded) via predicates:
-\begin{align*}
-\consName_n & \subseteq \mathcal{P}(\domainVal{}^n)
-\times \textsf{Approx}(\mathcal{P}(\domainVal{}^n)) \\
-\consAName_n & \subseteq \textsf{Mult}(\mathcal{P}(\domainVal{}^n))
-\times \textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))
-\end{align*}
-%
-where $\consName$ relates models without multiplicity.
-\Cref{fig:consistency} defines the two predicates, where for
-brevity we make implicit the subscript $n$ defining the
-dimensionality.  We denote sets of index schemes
-$\mathcal{P}((\domainVal{})^n)$
-from source code model by $M$ and from a specification model by $M$.
-
-Rules (\emph{up}) and (\emph{low}) are the core; (\emph{up})
-considers the case where the specification is
-an \term{atMost} upper-bound approximation.
-An upper-bound specification means that for every scheme in the code
-model, modelling a source-level index, $\forall u \in M$ there exists
-a generalising scheme in the spec model $\exists v \in N$, \ie{}, $u
-\preceq v$. That is, every source-level index scheme is generalised by the specification.
-The (\emph{down}) rule provides the case of
-\term{atLeast} lower-bound specifications. Consistency is dual to the above:
-for every scheme in the specification model $\forall v \in N$ there
-exists a scheme in the code model $\exists u \in M$ such that $u
-\preceq v$. That is, every specification scheme is generalised by some source
-index scheme.
-
-The (\emph{up}) and (\emph{down}) rules are combined for an \textsf{exact}
-specification (\textit{exact}) such that a specification's model
-provides both a consistent lower and consistent upper bound on
-the source code. Thus, for every scheme in the code model there
-is a scheme in the spec model that generalises it, and for every
-scheme in the specification there is a scheme in the source model that
-specialises it.  The (\textit{both}) rule is similar, decomposing
-a specification with both a lower and upper bound into
-consistency of these bounds.
-
-Rules $(1)$ and $(*)$ extend the \consName{}
-predicate to $\consAName{}$ with
-multiplicity information. Both rules
-state that consistent models remain consistent if they
-have the same multiplicity. Relatedly, rule $(1 < *)$ ensures
-that absence of a \texttt{readOnce} modifier in the specification
-does not prohibit a set of unique stencil accesses in the
-code from being consistent.
-\end{definition}
-
-\begin{example}
-  Consider the following five-point stencil kernel with a (quite general) specification:
-%
-\begin{minted}[breakindent=8.5em]{fortran}
-!= region :: fivepoint = centered(depth=1,dim=1) + centered(depth=1,dim=2)
-!= stencil readOnce, fivepoint :: b
-a(i,j) = b(i,j-1)+b(i,j)+b(i,j+1)+b(i-1,j)+b(i+1,j)
-\end{minted}
-%
-We refer to line three as $l3$ for brevity,
-which is checked via
-$check(({\textcolor{darkgreen}{\texttt{readOnce, fivepoint}}}),l3,\texttt{b})$
-with steps:
-\begin{enumerate}
-%
-\item Generate the model from the specification as in
-  \cref{sec:semantics}.
-%
-\vspace{-0.4em}
-\begin{align*}
- N_s= \{  \vtwo{-1}{\infty},
-          \vtwo{0}{\infty},
-          \vtwo{1}{\infty},
-          \vtwo{\infty}{-1},
-          \vtwo{\infty}{0},
-        \vtwo{\infty}{1} \}  \\
-\interp{{\textcolor{darkgreen}{\texttt{readOnce, fivepoint}}}}_2 =
-  \textsf{once}(\textsf{exact} (N_s))
-\end{align*}
-%\vspace{-1.6em}
-
-\item Generate code model $\textit{modelise}(l3, \texttt{b}) =
-  \textsf{once} (M_c)$ where:
-%
-\vspace{-0.9em}
-\begin{align*}
-M_c = & \{
-    \vtwo{0}{-1},
-    \vtwo{0}{0},
-    \vtwo{0}{1},
-    \vtwo{-1}{0},
-    \vtwo{1}{0}
-  \}
-\end{align*}
-\vspace{-1.1em}
-
-\item By the definition of consistency (\Cref{fig:consistency})
- $\textsf{once}(\textsf{exact}(N_s))$ is consistent with
-  $\textsf{once}(M_c)$ when:
-\vspace{-0.6em}
-  \begin{align*}
-    & \consSub{2}{\textsf{up}(N_s)}{M_c} \;\,
-    \wedge \;\, \consSub{2}{\textsf{low}(N_s)}{M_c}
-  \end{align*}
-%\vspace{-1.6em}
-
-\item This holds by the following sets of scheme
-  generalisations: for the upper bound, $\forall u \in M_c,
-  \exists v \in N_s . u \preceq_2 v$:
-  \begin{align}
-    \hspace{-1em}
-    \begin{array}{l}
-   \hspace{-1.5em}
-     \vtwoh{0}{-1} \preceq_2 \vtwoh{0}{\infty} \quad
-     \vtwoh{0}{0} \preceq_2 \vtwoh{0}{\infty} \quad
-     \vtwoh{0}{1} \preceq_2 \vtwoh{0}{\infty} \\[0.5em]
-        \hspace{-1.5em}\vtwoh{-1}{0} \preceq_2 \vtwoh{\infty}{0} \quad
-     \vtwoh{1}{0} \preceq_2 \vtwoh{\infty}{0}
-     \end{array}
-  \label{eq:check-example-upper}
-  \end{align}
-  %
-  and for the lower bound: $\forall v \in N_s, \exists u \in M_c . u
-  \preceq_n v$:
-  \begin{align}
-    \hspace{-1em}
-    \begin{array}{l}
-     \hspace{-1.3em}\vtwoh{-1}{0} \preceq_2 \vtwoh{-1}{\infty} \quad
-     \vtwoh{0}{0} \preceq_2 \vtwoh{0}{\infty} \quad
-     \vtwoh{1}{0} \preceq_2 \vtwoh{1}{\infty}  \\
-     \hspace{-1.3em}\vtwoh{0}{-1}  \preceq_2 \vtwoh{\infty}{-1} \quad
-     \vtwoh{0}{0}  \preceq_2 \vtwoh{\infty}{0} \quad
-     \vtwoh{0}{1}  \preceq_2 \vtwoh{\infty}{1}
-     \end{array}
-    \label{eq:check-example-lower}
-  \end{align}
-\end{enumerate}
-Since the consistency predicate is satisfied, the array
-  accesses flowing to line three in the stencil computation are valid
-  with respect to the {\textcolor{darkgreen}{\texttt{fivepoint}}} specification.
-\end{example}
-\begin{example}
-Consider an erroneous user modification to $l3$ with an
-additional term \texttt{b(i+2,j+2)}, call it $l3'$.% such that $l3'$ is:
-%
-%\begin{minted}[firstnumber=3]{fortran}
-%a(i,j) = b(i,j-1) + b(i,j) + b(i,j+1) + b(i-1,j) + b(i+1,j) + b(i+2,j+2)
-%\end{minted}
-
-Checking proceeds as in the previous example, except
-$\mathit{modelise}(l3',\texttt{b})$ generates an additional scheme
-$\vtwoh{2}{2}$. The generalisations for
-the lower bound \eqref{eq:check-example-lower}
-of step 4 above still hold since its consistency condition
-universally quantifies over the \emph{specification model}. However, for
-the upper bound, the consistency condition universally quantifies
-over schemes $u$ of the \emph{code model} for which there must exist
-a scheme $v \in N_s$ such that $u \preceq_2 v$. Thus, we require a $v
-\in N_s$ such that $\vtwoh{2}{2} \preceq_2 v$ but no such generalising
-scheme exists in the spec model. Hence checking fails.
-\end{example}
-
-\subsection{Specification inference}
-\label{subsec:inference}
-
-To aid adoption of our approach to legacy code, we provide
-an inference procedure for generating specifications from code, which
-are inserted automatically as comments.
-We illustrate inference using the
- five-point stencil of
- \Cref{exm:checking}.
-
-\subsubsection{Step 1: Data access analysis}
-\label{sec:inf-step1}
-
-The first step performs the
-analysis of \cref{subsec:analysis} to produce a
-model of the code in terms of index schemes.
-The \emph{analyse} and \emph{modelise} procedures before yielded the finite map:
-%
-\begin{align*}
-\texttt{a} \mapsto \mathsf{only}(\{\vtwoh{-1}{0}, \vtwoh{1}{0},
-          \vtwoh{0}{-1}, \vtwoh{0}{1}, \vtwoh{0}{0}\})
-\end{align*}
-%
-
-\subsubsection{Step 2: Coalesce contiguous schemes into spans}
-\label{sec:inf-step3}
-
-Let $U$ range over the finite maps computed by \textit{modelise}.  For
-each array variable $a \in \mathsf{dom}(U)$, the algorithm constructs a set of
-$n$-dimensional rectangles covering all contiguous groups of schemes
-in $U(a)$.
-
-\newcommand{\spanOp}{\textsf{spans}}
-
-\begin{definition}[Spans]
-  A \emph{span} represents an $n$-dimensional rectangle
-  as a pair of $n$-dimensional vectors (or
-  represented as a $2 \times n$ matrix) with values drawn from
-  $\domainVal{}$. This gives the co-ordinates of lower-bound 
-  and upper-bound vertices. We write
-  spans as $\vect{x} = [\vect{x}^L \; \vect{x}^U]$, where $\vect{x}^L$
-  and $\vect{x}^U$ give the lower and upper bounds.  In
-  the following, $\vect{x}, \vect{y}$ range over spans.  For a span
-  $\vect{x}$, we write $\vect{x}_{i:j}$ for the sub-span comprising the
-  subvectors $[\vect{x}^L_{i:j} \; \vect{x}^U_{i:j}]$ from entry $i$
-  to entry $j$ (inclusive).
-\end{definition}
-
-\noindent
-Each index scheme $u \in U(a)$ is converted to a \emph{unit} span
-$u \mapsto [u \; u]$, defining the lower and upper bound of an
-$n$-dimensional cube of side length $1$. For the five-point stencil
-example there are five unit spans, illustrated in
-\Cref{fig:inference-steps-informal}(a).
-
-The next step repeatedly coalesces any contiguous spans until a fixed
-point is reached, resulting in a set of (possibly overlapping) spans
-covering the original space of schemes. Two $n$-dimensional spans
-$\vect{x}$ and $\vect{y}$ are contiguous if for all but one
-$j \in \{1, \ldots, n\}$ then $\vect{x}_j = \vect{y}_j$ and for one
-$j$ then $\vect{x}^U_1 + 1 = \vect{y}^L$. For example, spans
-$\left[\vtwoh{-1}{0}\; \vtwoh{-1}{0}\right]$ and
-$[\vtwoh{0}{0} \; \vtwoh{1}{0}]$ are contiguous in the first dimension
-and can be coalesced into the span $[\vtwoh{-1}{0} \; \vtwoh{1}{0}]$.
-
-\begin{definition}\label{def:span-coalesc}
-  Span coalescing $\bullet$ is partial, defined for spans contiguous in the
-  first dimension:
-%%
-\begin{align*}
-\vect{x} \bullet \vect{y}
-= (\vect{x}^L, \vect{y}^U) \;\;\;\, \textit{if} \;\;\; \vect{x}^U_1 + 1 = \vect{y}^L_1 \; \wedge \;
-\vect{x}_{2:n} = \vect{y}_{2:n} %\\
-%\bot  & \textit{otherwise}
-%\end{cases}
-% $m$-dimensions into coalesced spans which vary in
-% $m+1$-dimensions.
-\end{align*}
-%%
-\end{definition}
-\noindent
-To coalesce regions which are contiguous in other dimensions
-other than the first, we generate all rotations of the spans in
-$n$-dimensional space by permutation on spans, and fold together contiguous
-regions using $\bullet$ (above). This
-generates all possible ways of coalescing regions. For our example,
-the unit spans in \Cref{fig:inference-steps-informal}(a)
-are coalesced into the contiguous spans in
-\Cref{fig:inference-steps-informal}(b).
-This procedure is iterated until a fixed point is reached (in one step
-in this example).
- Then any spans that are completely contained by any other span are deleted,
-leaving a minimal set of spans (which may overlap, but none of which
-fully contains another), \Cref{fig:inference-steps-informal}(c).
-The final spans are used to synthesise concrete stencil specification
-syntax (\cref{sec:inf-step4}), where
-each span becomes a product \texttt{*} of regions which are
-combined by the sum \texttt{+}.
-
-More formally the algorithm proceeds as follows.
-For all $a \in \mathsf{dom}(U)$, let $M = U(a)$. Every scheme in $M$
-is mapped to a unit span
-by $\textsf{unitSpans}(M) = \{[u \; u] \mid u\!\in\!M\}$, \eg{}:
-%%
-\begin{align*}
-  N(U(\texttt{a})) = \stwo{0}{0}{0}{0} \stwo{1}{0}{1}{0} \stwo{-1}{0}{-1}{0} \stwo{0}{1}{0}{1} \stwo{0}{-1}{0}{-1}
-\end{align*}
-%%
-In the following we let $M_0 = N(U(\texttt{a}))$ for our example.
-
-Each set of unit spans is applied to the fixed point of the \spanOp{} function
-(defined shortly), which coalesces spans into contiguous regions. That is, we compute
- $(\mu \, \spanOp)$ on input $N(M)$, where \spanOp{} is defined by the steps:
-
-\begin{figure}[t]
-\vspace{-1em}
-\hspace{-0.5em}
-{\scalebox{0.96}{
-\begin{minipage}{0.16\linewidth}
-\begin{align*}
-\setlength{\arraycolsep}{0em}
-\begin{array}{c}
-\quad \\[0em]
-\begin{tikzpicture}[scale=0.8]
-%\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
-\node at (1.25,1.22) {\textnormal{\footnotesize{0,0}}};
-\draw[thick] (1,1) rectangle (1.5,1.5);
-\draw[thick] (1.5,1) rectangle (2,1.5);
-\draw[thick] (1,0.5) rectangle (1.5,1);
-\draw[thick] (0.5,1) rectangle (1,1.5);
-\draw[thick] (1,1.5) rectangle (1.5,2);
-\end{tikzpicture} \\[0em]
-(a)
-\end{array}
-\end{align*}
-\end{minipage} $\rightarrow$ \begin{minipage}{0.43\linewidth}
-\begin{align*}
-\setlength{\arraycolsep}{0em}
-\begin{array}{c}
-\begin{tikzpicture}[scale=0.72]
-%\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
-%\node at (1.25,1.22) {\textnormal{\small{0,0}}};
-\draw [black, thin] (0.35,0) to [round left paren ] (0.35,2);
-\draw[thick] (0.5,1) rectangle (2,1.5);
-\draw[thick] (1,0.40) rectangle (1.5,0.90);
-\draw[thick] (1,1.60) rectangle (1.5,2.10);
-\node at (2.3,1.22) {+};
-\draw[thick] (3.8,1) rectangle (4.3,1.5);
-\draw[thick] (2.6,1) rectangle (3.1,1.5);
-\draw[thick] (3.2,0.5) rectangle (3.7,2);
-\draw [black, thin] (4.4,0) to [round right paren ] (4.4,2);
-\end{tikzpicture} \\[-0.5em] (b)
-\end{array}
-\end{align*}
-\end{minipage}$\rightarrow$ \begin{minipage}{0.25\linewidth}
-\begin{align*}
-\setlength{\arraycolsep}{0em}
-\begin{array}{c}
-\begin{tikzpicture}[scale=0.72]
-%\node at (1.25,1.22) {\textnormal{\small{0,0}}};
-\draw [black, thin] (0.8,0) to [round left paren ] (0.8,2);
-\draw[thick] (1,0.5) rectangle (1.5,2);
-\node at (1.8,1.22) {+};
-\draw[thick] (2.1,1) rectangle (3.6,1.5);
-\draw [black, thin] (3.7,0) to [round right paren ] (3.7,2);
-%\node at (2.75,1.22) {\textnormal{\small{0,0}}};
-\end{tikzpicture} \\[-0.5em] (c)
-\end{array}
-\end{align*}
-\end{minipage}}}
-\caption{Illustration of grouping spans in
-inference}
-\label{fig:inference-steps-informal}
-\vspace{-0.8em}
-\end{figure}
-
-\noindent
-\textbf{1)\,} Compute all permutations on the column vectors in a span, \ie{},
-  $[\vect{x}^L \vect{x}^U] \mapsto [\pi\vect{x}^L \, \pi\vect{x}^U]$
-for a permutation $\pi$. For each permutation function $\pi^n_i$
-(the $i$-th permutation for vectors of size $n$) we pair the
-permutation function with the set of permuted spans so that
-the spans can be un-permuted later.
-%
-\begin{equation*}
-P(M) = \bigcup_{i \in n!} (\pi^n_{i} , \; \{[\pi^n_i
-\vect{x}^L \; \pi^n_i\vect{x}^U] \, \mid \, [\vect{x}^L \; \vect{x}^U]
-\leftarrow M\}
-\end{equation*}
-%
-This corresponds to all rotations of the space, \eg{}:
-%
-\begin{align*}
-P(M_0) =
-\{&(\pi^2_1, \{ \stwo{0}{0}{0}{0},
-\stwo{1}{0}{1}{0},
-\stwo{-1}{0}{-1}{0},
-\stwo{0}{1}{0}{1},
-\stwo{0}{-1}{0}{-1} \})
-\\
-&(\pi^2_2, \{
- \stwo{0}{0}{0}{0},
- \stwo{0}{1}{0}{1},
- \stwo{0}{-1}{0}{-1},
- \stwo{1}{0}{1}{0},
- \stwo{-1}{0}{-1}{0}\})\}
-\end{align*}
-%
-where $\pi^2_1$ is the identity permutation and $\pi^2_2$ is the
-permutation that flips the order of the two elements in the
-vectors.
-
-\noindent
-\textbf{2)\,} Sort each permutation set into a list, by the ordering:
-\begin{equation*}
-  \vect{x} \leq \vect{y} =
-      \exists i \, . \; \vect{x}^L_{i} \leq \vect{y}^L_{i} \; \wedge \;
-        (i = n \vee \vect{x}_{i+1:n} = \vect{y}_{i+1:n})
-\end{equation*}
-%
-giving the sorting function $\textsf{sort}(M)$, \eg{}, $\textsf{sort}(P(M_0))$:
-%%
-\begin{align*}
-\{&(\pi^2_1, [
-\stwo{0}{-1}{0}{-1},
-\stwo{-1}{0}{-1}{0},
-\stwo{0}{0}{0}{0},
-\stwo{1}{0}{1}{0},
-\stwo{0}{1}{0}{1}] )
-\\
-&(\pi^2_2, [
-\stwo{0}{-1}{0}{-1},
-\stwo{-1}{0}{-1}{0},
-\stwo{0}{0}{0}{0},
-\stwo{1}{0}{1}{0},
-\stwo{0}{1}{0}{1}])\}
-\end{align*}
-\noindent
-\textbf{3)\,} Reduce each list pairwise by $\bullet$
-(\Cref{def:span-coalesc}) to coalesce contiguous regions,
-given by $\textsf{fold}\bullet$. For our example:
-%
-\begin{align*}
-\{&(\pi^2_1, [
-\stwo{0}{-1}{0}{-1},
-\stwo{-1}{0}{1}{0},
-\stwo{0}{1}{0}{1}
-]), (\pi^2_2, \textit{same as for $\pi_1$})\}
-\end{align*}
-%
-\textbf{4)\,} Un-permute and union together, \ie{},
-%
-\[
-U(M) = \bigcup \{[\pi \vect{x}^L, \pi \vect{x}^U]
- \mid [\vect{x}^L, \vect{x}^U] \leftarrow S, (\pi, S) \leftarrow M\}
-\]
-For our example:
-%
-\begin{align*}
-U(M) =
-\{\stwo{0}{-1}{0}{-1},
-\stwo{-1}{0}{1}{0},
-\stwo{0}{1}{0}{1},
-\stwo{-1}{0}{-1}{0},
-\stwo{0}{-1}{0}{1},
-\stwo{1}{0}{1}{0}\}
-\end{align*}
-%
-\textbf{5)\,} Filter by the region containment predicate
-$\containedin$; if any region is contained within another then remove the
-  smaller:
-%
-\begin{align*}
-& \vect{x} \containedin \vect{y} = \vect{y}^L_1 \leq \vect{x}^L_1 \wedge \vect{x}^U_1 \leq \vect{y}^U_1
-  \wedge (\vect{x}^L_{2:n}, \vect{x}^U_{2:n}) \containedin
-  (\vect{y}^L_{2:n}, \vect{y}^U_{2:n})
-\end{align*}
-For our example this then yields the final result of:
-\begin{align*}
-\spanOp(M)
-= \{\stwo{-1}{0}{1}{0} \stwo{0}{-1}{0}{1}\}
-\end{align*}
-since $\stwo{0}{1}{0}{1} \sqsubseteq \stwo{0}{-1}{0}{1}$
-and $\stwo{1}{0}{1}{0} \sqsubseteq \stwo{-1}{0}{1}{0}$.
-Applying \spanOp{} again yields the same
-result for our example.%; the fixed point is reached.
-
-\subsubsection{Step 4: Synthesise specifications from spans}
-\label{sec:inf-step4}
-
-\newcommand{\oplusbig}{\operatornamewithlimits{\term{\Large{$\oplus$}}}}
-\newcommand{\odobig}{\operatornamewithlimits{\term{\Large{$\odot$}}}}
-\newcommand{\bplus}{\operatornamewithlimits{\term{\Large{+}}}}
-\newcommand{\tySum}[1]{#1^{\term{+}}}
-\newcommand{\tyProd}[1]{#1^{\term{*}}}
-\newcommand{\specDNF}{\textbf{spec}}
-
-Abstract syntax trees of the specification language
-are synthesised from the set $\spanOp(M)$ of coverings spans.
-This conversion is defined in terms
-of an algebra on \textit{spec} syntax in disjunctive-normal form
-(which we denote \specDNF{}) and which is wrapped
-in the $\textsf{Approx}$ type (see
-\Cref{def:mult-and-approx}).  The algebra mirrors the shape
-of the region operators $\term{+}$ and $\term{*}$ 
-with binary operators $\oplus$ and $\odot$ on
-$\mathsf{Approx}(\specDNF)$.
-A function $\mathsf{toSpec}$ maps a dimension identifier
-and a pair of lower and upper bound
-values (one-dimensional) to a specification with a one-dimensional region.
-%%
-\begin{align*}
-\mathsf{toSpec} & : \mathbb{N}_{>0} \rightarrow (\domainVal{} \times \domainVal{}) \rightarrow \mathsf{Approx}(\specDNF)
-\end{align*}
-%%
-A span is converted into an $n$-dimensional
-specification by applying $\mathsf{toSpec}$ to each
-pair of lower and upper bounds in a span, per dimension, and multiplying these
-by $\odot$. The resulting specifications for
-each span are summed by $\oplus$:
-\begin{equation*}
-\textsf{spec}(M) =
-\oplusbig\limits_{\vect{x} \in \spanOp(M)} \,
-\odobig\limits_{i \in \{1, \ldots, n\}} \, \mathsf{toSpec} \; i \; (\vect{x}^L_i, \vect{x}^U_i)
-\end{equation*}
-We unpack the definitions of this algebra.
-
-\begin{definition}[Specifications in DNF]
-Let $r_1, ..., r_n$ range over elements of the \textit{region}
-syntax (see \Cref{fig:syntax}).
-The set \specDNF{} is a subset of syntax trees \textit{spec},
-containing specifications in disjunctive normal form over regions
-where the disjunct is \term{+} and the conjunct is \term{*}; 
- elements of \specDNF{} are of the form:
-%%
-$
-(r^1_{1} \; \term{*} \ldots \term{*} \; r^1_n)\; \term{+} \; \ldots
-\term{+} \; (r^m_1 \, \term{*} \; \ldots \term{*} \; r^m_p)
-$.
-%
-%More formally, $\specDNF \subseteq (\textit{spec} \cup \{\epsilon\})$, defined:
-%%%
-%\begin{align*}
-%\specDNF{} & ::= \tySum{\specDNF} \mid \epsilon \quad\;\;
-%\tySum{\specDNF} ::= \tyProd{\specDNF} \; \term{+} \; \tySum{\specDNF} \mid
-%  \tyProd{\specDNF} \\
-%\tyProd{\specDNF} & ::= \textit{region} \; \term{*} \; \tyProd{\specDNF} \mid
-%   \textit{region}
-%\end{align*}
-%%
-In the following, $S, T$ range over elements of
-\specDNF{} and $\tyProd{s}, \tyProd{t}$ range over conjuncts
- \eg{} $\tyProd{s} = r^1_{1} \; \term{*}
-\ldots \term{*} \; r^1_n$.
-\end{definition}
-
-\newcommand{\exactR}{\mathsf{exact}\;}
-\newcommand{\exact}[1]{\exactR(#1)}
-\newcommand{\upperB}[1]{\mathsf{upper}\;(#1)}
-
-\begin{figure}[t]
-%\newcommand{\llower}{\vect{x}^L_i}
-%\newcommand{\lupper}{\vect{x}^U_i}
-\vspace{-0.7em}
-\newcommand{\llower}{l}
-\newcommand{\lupper}{u}
-{\scalebox{0.95}{
-\begin{minipage}{1\linewidth}
-\begin{align*}
-&\mathsf{toSpec} \; d \; (\llower, \lupper) = \\
-&\hspace{-0.4em}\begin{cases}
-%% POINT
-\exact{\stenReflS{d}} & \llower = 0 \, \wedge \, \lupper = 0\\[0.02em]
-%% FWD
-\exact{\stenFwdS{\lupper}{d}}  & \llower = 0 \, \wedge \, \lupper > 0 \\[0.02em]
-\exact{\stenFwdSR{\lupper}{d}{\irreflS}}\!\!\!\! & \llower = 1 \, \wedge \, \lupper > 0 \\[0.02em]
-%% BWD
-\exact{\stenBwdS{|\llower|}{d}}  & \llower < 0 \, \wedge \, \lupper = 0 \\[0.02em]
-\exact{\stenBwdSR{|\llower|}{d}{\irreflS}}\!\!\!\! & \llower < 0 \, \wedge \, \lupper = -1 \\[0.02em]
-%% CEN
-\exact{\stenCenS{\lupper}{d}}& \!\!\!\llower < 0 \wedge \lupper > 0 \wedge |\llower| = \lupper \\[0.02em]
-\exactR(\stenBwdS{|\llower|}{d} & \!\!\!\llower < 0 \wedge \lupper > 0  \wedge |\llower| \neq \lupper \\
-\qquad\qquad \texttt{+} \, \stenFwdS{\lupper}{d}) & \\[0.02em]
-% BOUND
-\upperB{\stenFwdS{\lupper}{d}} & \llower > 1 \\[0.02em]
-\upperB{\stenBwdS{|\llower|}{d}} \quad & \llower < (-1)
- \\[0.02em]
-% ABSOLUTE REP
-\exact{\epsilon} & \llower = \infty \, \vee \, \lupper = \infty
-  \end{cases}
-\end{align*}
-\end{minipage}
-}}
-\caption{\textsf{toSpec} constructor of $\mathsf{Approx}(\specDNF)$ values.}
-\label{fig:to-spec}
-\vspace{-0.8em}
-\end{figure}
-
-\begin{definition} The $\mathsf{toSpec}$ constructor is defined
-in \Cref{fig:to-spec}, where the lower and upper bounds
-determine region constants; $\mathsf{toSpec} \, d \, (l, u)$ is defined
- if $l$ less than or equal to $u$.
-
-Positive lower and upper bound yield \term{forward} regions, and
-negative lower and upper bounds yield \term{backward} regions.
-The \texttt{nonpointed} attribute (abbreviated \term{np})
-is present when the lower bound is at $1$ or $-1$. 
-In the sixth-case, where the lower bound
-is negative and the upper bound is positive and they have the same
-magnitude, then a \texttt{centered} region is constructed with
-the depth as the common magnitude. If the magnitudes are not the same
-(seventh case) then separated \texttt{forward} and \texttt{backward}
-regions are summed, with their corresponding depths.
-If the lower bound is greater than 1
-(which implies the upper bound is greater than 1 too) then
-this represents a region that is not at the origin (or next to) the
-origin and thus cannot be represented in the specification language.
-Thus, an upper bound is produced showing the maximum extent
-of the region. The penultimate case is the dual, when the lower bound is less
-than $-1$, giving an upper bound on a \texttt{backward} specification.
-\end{definition}
-
-\begin{definition} The $\odot$ operation on $\textsf{Approx}(\specDNF{})$
-  is defined in terms of the intermediate operator $\hat{\odot}$ on
-  \specDNF{}, defined:
-%%
-\begin{align*}
-S \,\hat{\odot}\, \epsilon = S \quad
-\epsilon \,\hat{\odot}\, S = S \qquad
-S \,\hat{\odot}\, T = \bplus_{(\tyProd{s}, \tyProd{t}) \in (S \times
-   T)} (\tyProd{s} \; \term{*} \; \tyProd{t})
-\end{align*}
-%%
-That is, the $\hat{\odot}$ product of two specifications $S$ and $T$ in DNF form
-is the $\term{+}$ sum of all pairwise $\term{*}$ products for every
-pair of $\tyProd{spec}$ drawn from $S$ and $T$. The result
-is indeed in DNF-form, where $\hat{\odot}$ is essentially applying
- the \trule{\textsc{dist}} rule of the equational theory. The operation is
-commutative and associative, and sound with respect to our
-model (see \Cref{lem:alg-soundness} below).
-The $\hat{\odot}$ operation is then lifted to all combinations of
-\textsf{Approx} for $\odot$ in \Cref{fig:odotlift} where
-in the last case $\textit{inj}$ corresponds to the unary
-injections of \textsf{Approx}. For brevity we omit the cases where the arguments above are flipped
-as $\odot$ is commutative.
-\end{definition}
-
-\begin{figure}[t]
-\vspace{-1em}
-{\small{
-\begin{align*}
-\textsf{lower}(S) \odot \textsf{exact}(T) = \; &
-       %\textsf{exact}(T) \odot \textsf{lower}(S) = \;
-       \textsf{both} \, (S \hat{\odot} T, \; T) \\
-\textsf{upper}(S) \odot \textsf{exact}(T) = \; &
-       %\textsf{exact}(T) \odot \textsf{upper}(S) =
-       \textsf{both} \, (T, \; S \hat{\odot} T) \\
-\textsf{lower}(S) \odot \textsf{both}(T_l,T_u) = \; &
-        %\textsf{both}(T_l,T_u) \odot \textsf{lower}(S) = \;
-        \textsf{both} \, (S \hat{\odot} T_l, \; T_u) \\
-\textsf{upper}(S) \odot \textsf{both}(T_l,T_u) = \; & \textsf{both} \,
-                                                      (T_l, \; S \hat{\odot} T_u) \\
-\textsf{both}(S_l,S_u) \odot
-\textsf{both}(T_l,T_u)  = \; & \textsf{both} (S_l \hat{\odot} T_l, \; S_u
-                         \hat{\odot} T_u) \\
-\textit{inj}(S) \odot \textit{inj}(T) = \; & \textit{inj}(S
-                                              \hat{\odot} T)
-\end{align*}
-\vspace{-2em}
-}}
-\caption{Lifting $\hat{\odot}$ to approximations.}
-\label{fig:odotlift}
-\end{figure}
-
-\begin{definition}The $\oplus$ operation on $\textsf{Approx}(\specDNF)$
-is defined similarly to $\odot$, with the intermediate $\hat{\oplus}$ on
-$\specDNF$:
-%%
-\begin{equation*}
-S \,\hat{\oplus}\, \epsilon = S \quad
-\epsilon \,\hat{\oplus}\, S = S \quad
-S \, \hat{\oplus} \, T = S \, \term{+} \, T
-\end{equation*}
-%%
-Thus, if neither specification is empty, their $\hat{\oplus}$ is just
-the syntactic sum $\term{+}$. Then $\oplus$ is the lifting of
-$\hat{\oplus}$ to \textsf{Approx} with the shape as $\odot$ in Figure~\ref{fig:odotlift}.
-\end{definition}
-
-\begin{lemma}%[$\hat{\odot}$, $\hat{\oplus}$ soundness]
-$\forall S,
-  T \in \specDNF$ where $S\!\neq \epsilon \wedge T\!\neq \epsilon$:
-\begin{align*}
-\interp{S \; \hat{\odot} \; T}_n = \interp{S}_n \otimes \interp{T}_n
-  \qquad
-\interp{S \; \hat{\oplus} \; T}_n = \interp{S}_n \cup \interp{T}_n
-\end{align*}
-\label{lem:alg-soundness}
-\vspace{-2em}
-\end{lemma}
-%
-\noindent
-Thus, our model validates specification synthesis.
-
-The final step applies a simplification pass, orienting the
-equational theory $\equiv$ (\Cref{sec:eqs}) into rewrite rules
-based on some simple heuristics.
-For our example, we infer and then synthesise the
-final 5-point specification:
-%%
-\begin{minted}[linenos=false,breakindent=2.4em,xleftmargin=0.2em]{fortran}
-!= stencil centered(depth=1, dim=1)*pointed(dim=2) + centered(depth=1, dim=2)*pointed(dim=1):: a
-\end{minted}
-%%
-
 \section{Evaluation}
 \label{sec:evaluation}
-
-%We have processed nearly one million lines of Fortran code using our
-%CamFort tool and examined the output to determine how useful
-%and applicable it may be. Our corpus of Fortran code is collected from
-%many computational science projects for which we believe our tool will
-%be potentially helpful.
-
-% Corpus details
-% specfem3d ``simulates acoustic (fluid), elastic (solid), coupled
-% acoustic/elastic, poroelastic or seismic wave propagation in any
-% type of conforming mesh of hexahedra (structured or not).''
-% https://github.com/geodynamics/specfem3d
-
-% arpack-ng - Fortran 77 routines designed to solve large scaled
-% eigenvalue problems - https://github.com/opencollab/arpack-ng
 
@@
 induction variables on the RHS are not a subset of those in the LHS. These
-cases are not stencils by our definition. The degenerate case of this
-is to have only constant indices on the LHS. We see lots of examples
-of this in loops as accumulators \eg{} computing the sum over an array; \\
+cases are not stencils by our definition. \\
 \noindent
@@
 %stencils.
-The primary purpose of this tool is optimisation rather than
-correctness, and the language of specification is 
-more elaborate than ours.
 
@@
 compiled into parallel programs based on trapezoidal decompositions with
-hyperspace cuts. Pochoir specifications are used for describing the kernel,
-boundary conditions, and shape of the stencil. Pochoir is aimed at programmers
-reluctant to implement the high-performance cache-oblivious
-``hypertrapezoidal'' algorithms.  Like much of the related work,
-the goal is optimisation rather than verifying program correctness.
+hyperspace cuts.
 

### 2017-03-13 oopsla17/paper.tex
oopsla stub

\documentclass[acmlarge,anonymous,review]{acmart}

\title{Understanding and Verifying the Shape of Stencil Computations}

\begin{document}
\maketitle

%\bibliographystyle{

\section{Introduction}

\section{An Empirical Study of Stencils in Scientific Fortran code}

\begin{enumerate}
\item Stencils are ubiquitious;
\item Most stencils read their data from source arrays with a static pattern
base on constant offsets from the mid-point;
\item Most stencils read their data from contiguous \emph{regions} in an array;
\item Most stencils read each index at most once.
\end{enumerate}

\section{A Specification Language for Shape}

\section{Analysis, Checking, and Inference}

\section{Evaluation}

\bibliography{references}

\end{document}

### 2017-03-14 Analyse.hs
analysis code initial work

@@ -1,0 +1,253 @@
+{-# LANGUAGE TupleSections #-}
+{-# LANGUAGE FlexibleContexts #-}
+{-# LANGUAGE DeriveDataTypeable #-}
+{-# LANGUAGE PatternGuards #-}
+{-# LANGUAGE ScopedTypeVariables #-}
+{-# LANGUAGE ConstraintKinds #-}
+
+module Main where
+
+import Camfort.Analysis.Simple
+import Camfort.Functionality
+import Camfort.Input
+
+import Control.Exception
+import Control.Monad
+import Data.List
+import Data.Monoid
+import Data.Typeable
+import System.Environment
+import Text.Printf
+import System.Directory
+
+import Control.Monad.State.Strict
+import Control.Monad.Reader
+import Control.Monad.Writer.Strict hiding (Product)
+
+import Camfort.Analysis.CommentAnnotator
+
+import Camfort.Specification.Stencils.InferenceBackend
+import Camfort.Specification.Stencils.InferenceFrontend
+import Camfort.Specification.Stencils.Syntax
+import Camfort.Specification.Stencils.Annotation
+import qualified Camfort.Specification.Stencils.Grammar as Gram
+import qualified Camfort.Specification.Stencils.Synthesis as Synth
+import Camfort.Analysis.Annotations
+import Camfort.Helpers.Vec
+import Camfort.Helpers (Filename, SourceText, collect)
+import Camfort.Input
+import qualified Camfort.Output as O
+
+import qualified Language.Fortran.AST as F
+import qualified Language.Fortran.Analysis as FA
+import qualified Language.Fortran.Analysis.Types as FAT
+import qualified Language.Fortran.Analysis.Renaming as FAR
+import qualified Language.Fortran.Analysis.BBlocks as FAB
+import qualified Language.Fortran.Analysis.DataFlow as FAD
+import qualified Language.Fortran.Util.Position as FU
+import qualified Language.Fortran.Util.SecondParameter as FUS
+
+import Data.Data
+import Data.Foldable
+import Data.Generics.Uniplate.Operations
+import Data.Graph.Inductive.Graph hiding (isEmpty)
+import qualified Data.Map as M
+import qualified Data.IntMap as IM
+import qualified Data.Set as S
+import Data.Maybe
+import Data.List
+import Data.Monoid
+import Debug.Trace
+
+main :: IO ()
+main = do
+    args <- getArgs
+    if (length args >= 1) then
[…]

### 2017-03-14 oopsla17/oopsla17.tex

\documentclass[acmlarge,review,anonymous]{acmart}\settopmatter{printfolios=true}

%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption

\usepackage{minted}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{syntax}
\usepackage{indentfirst}

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
\citestyle{acmauthoryear}   %% For author/year citations

\fvset{fontsize=\scriptsize}

\setlength{\grammarindent}{3em} % increase separation between LHS/RHS

\newcommand{\eg}{\emph{e.g.}}
\newcommand{\ie}{\emph{i.e.}}

\newcommand{\inbar}{\ \ \textbar\ \ }

\usepackage{cleveref}

\usepackage{xcolor}
\newcommand{\todo}[1]{\textcolor{blue}{#1}}

\begin{document}
\title{Understanding and Verifying the Shape of Stencil Computations}

\author{Mistral Contrastin}
\affiliation{
  \department{Computer Laboratory}
  \institution{University of Cambridge}
}
\email{Mistral.Contrastin@cl.cam.ac.uk}

\author{Andrew Rice}
\affiliation{
  \department{Computer Laboratory}
  \institution{University of Cambridge}
}
\email{Andrew.Rice@cl.cam.ac.uk}

\maketitle

\section{Introduction}

\section{An Empirical Study of Stencils in Scientific Fortran code}

\begin{enumerate}
\item Stencils are ubiquitious;
\item Most stencils read their data from source arrays with a static pattern
base on constant offsets from the mid-point;
\item Most stencils read their data from contiguous \emph{regions} in an array;
\item Most stencils read each index at most once.
\end{enumerate}

\section{A Specification Language for Shape}

\section{Analysis, Checking, and Inference}

\section{Evaluation}

\bibliography{references}

\end{document}

### 2017-03-14 oopsla17/analysis/Analyse.hs
analysis of stencil specs- moving on to looking at RHSes

@@ -74,39 +74,58 @@ applyAnalysisToDir dir = do
     let (dbg, results) = mconcat debugsAndResults
     putStrLn $ dbg
     putStrLn $ show results
-    
+
 
 -- Result type
 data Result = Result {
-   numArrayWrites :: Int,             -- n  where n >= p + m
-   numIVArrayWrites :: Int,           -- p
+   numArrayWrites           :: Int,   -- n  where n >= p + m
+   numIVArrayWrites         :: Int,   -- p
    numRelativeIVArrayWrites :: Int,   -- m
    numIVOrConstArrayWrites :: Int,    -- p'
+   numIVButInconsistentIVRHS :: Int,
+   numIVButNonNeighbourRHS   :: Int,
    numContiguousStencils :: Int,      -- c  where c >= d and p >= c
    numLinearStencils     :: Int,      -- d
    dimensionalityHist    :: [Int],    -- h1 where sum(h1) == c
    depthHist             :: [Int],    -- h2 where sum(h2) == c
+   numArraysReadHist     :: [Int],
    indexExprsHist        :: [Int],    -- h3
-   patternHist           :: (M.Map Int Int, M.Map (Int, Int) Int, M.Map (Int, Int, Int) Int) -- only for 1, 2 and 3 dimensional specs
+   patternHist           :: (M.Map Int Int
+                            , M.Map (Int, Int) Int
+                            , M.Map (Int, Int, Int) Int)
+                            -- only for 1, 2 and 3 dimensional specs
    } deriving Show
 
 
 instance Monoid Result where
-  mempty = Result 0 0 0 0 0 0 [] [] [] (M.empty, M.empty, M.empty)
-      
+  mempty = Result 0 0 0 0 0 0 0 0 [] [] [] [] (M.empty, M.empty, M.empty)
+
   mappend r1 r2 = Result
      { numArrayWrites = numArrayWrites r1 + numArrayWrites r2
      , numIVArrayWrites = numIVArrayWrites r1 + numIVArrayWrites r2
-     , numRelativeIVArrayWrites = numRelativeIVArrayWrites r1 + numRelativeIVArrayWrites r2
-     , numIVOrConstArrayWrites = numIVOrConstArrayWrites r1 + numIVOrConstArrayWrites r2
-     , numContiguousStencils = numContiguousStencils r1 + numContiguousStencils r2
+     , numRelativeIVArrayWrites = numRelativeIVArrayWrites r1
+                                + numRelativeIVArrayWrites r2
+
[…]

### 2017-03-15 Analyse.hs
full classification on stencils; more clear result output

@@ -4,6 +4,8 @@
 {-# LANGUAGE PatternGuards #-}
 {-# LANGUAGE ScopedTypeVariables #-}
 {-# LANGUAGE ConstraintKinds #-}
+{-# LANGUAGE FlexibleInstances #-}
+{-# LANGUAGE ImplicitParams #-}
 
 module Main where
 
@@ -17,7 +19,6 @@ import Data.List
 import Data.Monoid
 import Data.Typeable
 import System.Environment
-import Text.Printf
 import System.Directory
 
 import Control.Monad.State.Strict
@@ -54,6 +55,7 @@ import Data.Graph.Inductive.Graph hiding (isEmpty)
 import qualified Data.Map as M
 import qualified Data.IntMap as IM
 import qualified Data.Set as S
+import qualified Data.ByteString.Char8 as B
 import Data.Maybe
 import Data.List
 import Data.Monoid
@@ -70,78 +72,157 @@ main = do
 applyAnalysisToDir :: String -> IO ()
 applyAnalysisToDir dir = do
     files <- readParseSrcDir dir []
-    debugsAndResults <- mapM applyAnalysisToFile files
-    let (dbg, results) = mconcat debugsAndResults
+    let debugsAndResults = map applyAnalysisToFile files
+    let (dbg, results)   = mconcat debugsAndResults
     putStrLn $ dbg
-    putStrLn $ show results
+    putStrLn $ prettyResults results
 
+type ForContigAndNonContig a = (a, a)
 
--- Result type
+-- Results data type
 data Result = Result {
-   numArrayWrites           :: Int,   -- n  where n >= p + m
-   numIVArrayWrites         :: Int,   -- p
-   numRelativeIVArrayWrites :: Int,   -- m
-   numIVOrConstArrayWrites :: Int,    -- p'
-   numIVButInconsistentIVRHS :: Int,
-   numIVButNonNeighbourRHS   :: Int,
-   numContiguousStencils :: Int,      -- c  where c >= d and p >= c
-   numLinearStencils     :: Int,      -- d
-   dimensionalityHist    :: [Int],    -- h1 where sum(h1) == c
-   depthHist             :: [Int],    -- h2 where sum(h2) == c
-   numArraysReadHist     :: [Int],
-   indexExprsHist        :: [Int],    -- h3
-   patternHist           :: (M.Map Int Int
-                            , M.Map (Int, Int) Int
-                            , M.Map (Int, Int, Int) Int)
[…]

### 2017-03-15 oopsla17/analysis/Analyse.hs
parameterise debugging mode on the command line

@@ -64,17 +64,23 @@ import Debug.Trace
 main :: IO ()
 main = do
     args <- getArgs
-    if (length args >= 1) then
-      applyAnalysisToDir (head args)
+    if (length args == 1) then
+      applyAnalysisToDir (head args) False []
+    else if (length args > 1) then
+      if (args !! 1 == "-d") then
+         applyAnalysisToDir (head args) True (tail (tail args))
+      else
+         applyAnalysisToDir (head args) False (tail args)
     else
-      putStrLn $ "Please specify a directory on which to apply the analysis"
+      putStrLn $ "Please specify a directory on which to apply the analysis\
+                 \ followed by any number of file names to be excluded "
 
-applyAnalysisToDir :: String -> IO ()
-applyAnalysisToDir dir = do
-    files <- readParseSrcDir dir []
+applyAnalysisToDir :: String -> Bool -> [String] -> IO ()
+applyAnalysisToDir dir debug excludes = do
+    files <- readParseSrcDir dir excludes
     let debugsAndResults = map applyAnalysisToFile files
     let (dbg, results)   = mconcat debugsAndResults
-    putStrLn $ dbg
+    if debug then putStrLn $ dbg else return ()
     putStrLn $ prettyResults results
 
 type ForContigAndNonContig a = (a, a)
@@ -92,7 +98,7 @@ data Result = Result {
   , numStencils                :: Int   -- b
   , numRelativisedRHS          :: Int   -- R (where R = M)
   , numContiguousStencils      :: Int    -- c  where p >= c
-  , numNoOrigin                :: Int    -- c' where c >= c' 
+  , numNoOrigin                :: Int    -- c' where c >= c'
   , numSingNonContigStencils   :: Int
   , numContigLinearStencils    :: Int    -- d  where c >= d
   , dimensionalityHist         :: ForContigAndNonContig [Int]
@@ -190,7 +196,7 @@ instance Monoid Result where
      , numSingNonContigStencils = numSingNonContigStencils r1 + numSingNonContigStencils r2
 
      , numContigLinearStencils = numContigLinearStencils r1 + numContigLinearStencils r2
-     
+
      , dimensionalityHist = histZip (dimensionalityHist r1) (dimensionalityHist r2)
      , maxDepthHist = histZip (maxDepthHist r1) (maxDepthHist r2)
[…]

### 2017-03-16 Analyse.hs
result validator

@@ -79,9 +79,11 @@ applyAnalysisToDir :: String -> Bool -> [String] -> IO ()
 applyAnalysisToDir dir debug excludes = do
     files <- readParseSrcDir dir excludes
     let debugsAndResults = map applyAnalysisToFile files
-    let (dbg, results)   = mconcat debugsAndResults
+    let (dbg, result)   = mconcat debugsAndResults
     if debug then putStrLn $ dbg else return ()
-    putStrLn $ prettyResults results
+    putStrLn $ prettyResults result
+    valid <- resultValidation result
+    if valid then (putStrLn $ "Results were valid") else (putStrLn "Results were invalid")
 
 type ForContigAndNonContig a = (a, a)
 
@@ -115,13 +117,52 @@ data Result = Result {
                             -- only for 1, 2 and 3 dimensional specs
     } deriving Show
 
+resultValidation =
+      (\r -> (numArrayWrites r) `gte` (numNeighbourArrayWrites r))
+          `reason` "Array writes >= Neighbour Writes"
+
+ <**> (\r -> (numNeighbourArrayWrites r) `gte` ((numIVArrayWrites r) + (numConstArrayWrites r)))
+          `reason` "Neighour Writes >= (IV Writes + Neigh/Const Writes)"
+
+ <**> (\r -> (numNeighbourArrayWrites r) `gte` (numLHSButNonNeighbourRHS r))
+          `reason` "Neighbour Writes >= Non-neighour RHS"
+
+ <**> (\r -> (numNeighbourArrayWrites r) `gte` (numLHSButInconsistentIVRHS r))
+          `reason` "Neighbour Writes >= Inconsistent IV RHS"
+
+ <**> (\r -> ((numStencils r) + (numLHSButInconsistentIVRHS r) + (numLHSButNonNeighbourRHS r))
+           `eq` (numNeighbourArrayWrites r))
+          `reason` "Num stencils + RHS inconsistent IV + RHS non-neighbour = LHS Neighbour"
+
+ <**> (\r -> (numRelativisedRHS r)
+          `eq` ((numNeighbourArrayWrites r) - ((numConstArrayWrites r) + (numIVArrayWrites r))))
+          `reason` "Num relativised stencils = LHS Neighbour with some relative offset"
+
+ <**> (\r -> (numStencils r)
+            `gte` ((numContiguousStencils r) + (numSingNonContigStencils r)))
[…]

### 2017-03-16 Analyse.hs
performance tweak

@@ -368,7 +368,7 @@ perBlock b@(F.BlStatement ann span@(FU.SrcSpan lp up) _ stmnt) = do
              if (numNeighbourArrayWrites r > 0) then do
                  -- If the LHS is a neighbourhood index
                  (dbg, r') <- analyseRHS lhsIndices [b]
-                 return ("At: " ++ show span ++ "\n" ++ dbg, r `mappend` r')
+                 ("At: " ++ show span ++ "\n" ++ dbg) `trace` return ("At: " ++ show span ++ "\n" ++ dbg, r `mappend` r')
              else return ("", r)
            _ -> return mempty)
     tell (mconcat results)
@@ -443,7 +443,7 @@ classifyRHSsubscripts ivs rhses lhs =
     -- Num arrays read
     numArrays = toHist . length . M.keys $ rhses
     -- Index exprs
-    numIndexExprs = toHist . length . concat . M.elems $ rhses
+    numIndexExprs = toHist . Data.List.sum . map length . M.elems $ rhses
     -- Patterns
     patterns = mkPatterns . concat . M.elems $ rhsesO
     -- Work out if the stencil is linear or not
@@ -516,7 +516,6 @@ concatHist (x:y:xs) = (x `histZip` y) `histZip` (concatHist xs)
 contig, nonContig :: Monoid a => a -> (a, a)
 contig n = (n, mempty)
 nonContig n = (mempty, n)
-
 
 -- Predicate transformers on Maps
 anyMap p m = M.size (M.filter p m) > 0

### 2017-03-20 oopsla17/lattice-model.tex

\documentclass{article}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{stmaryrd}
\usepackage{indentfirst}
\usepackage{cleveref}

\theoremstyle{definition}
\newtheorem{defn}{Definition}

\theoremstyle{plain}
\newtheorem{lem}{Lemma}
\newtheorem{prop}{Proposition}

\newcommand{\eg}{\emph{e.g.}}

\newcommand{\zinf}{\textnormal{$\mathbb{Z}^\infty$}}
\newcommand{\interp}[1]{\llbracket\texttt{#1}\rrbracket}
\newcommand{\interv}[3]{\textnormal{$\lbrack#1,#2\rbrack^{#3}$}}
\newcommand{\vecgen}[3]{\textnormal{$\textit{vec-gen}_{#1}(#2,#3)$}}

\begin{document}

\section{Giving specifications meaning}

We construct a lattice and use this lattice to give meanings to regions. We
then give full meaning to specifications by wrapping regions in variant types
for multiplicity and approximation. Model defined in this section will later be
used for two purposes. First, we show soundness of the equational theory of
specifications with respect to the model. Second, we define inference and
consistency checking procedures of specifications with respect to the
interpretations of the stencils in the code and the specifications.

\subsection{Constructing the lattice model of regions}

\begin{defn}{(\zinf)}
  We define the set \zinf{} as $\mathbb{Z}$ extended with $\infty$ and
  $-\infty$. For any $n$ in \zinf{}, we have $-\infty \leq n \leq \infty$. The
  resulting set is a total order with top and bottom elements.
\end{defn}

\begin{defn}{(\emph{Offset space})}
  \emph{Offset space} is a subset of \zinf{} members. $\mathcal{P}(\zinf{})$
  couple with $\supseteq$ is the power set distributive lattice.
\end{defn}

\begin{defn}{(\emph{Offsets vector})}
  An \emph{offsets vector} is a $N$ dimensional vector of offset spaces denoted
  by $(\textit{offs}_1, \dots, \textit{offs}_N)$.

  The set of $N$ dimensional offsets vectors is $\mathcal{P}{(\zinf{})}^N$. We
  define the partial order $\leq$ on $N$ dimensional offsets vectors
  \textit{voffs} and \textit{voffs'} as follows:
%
  \begin{equation*}
    \textit{voffs} \leq \textit{voffs'} \triangleq
      \forall n. 1
        \leq i \leq N \implies \textit{voffs}_i \supseteq \textit{voffs'}_i
  \end{equation*}
\end{defn}

\begin{prop}{}
  $(\mathcal{P}{(\zinf{})}^N, \leq)$ is a distributive lattice.
\end{prop}
%
\begin{proof}
  Product of two distributive lattices is itself a distributive lattice when the
  partial order is defined as componentwise comparison in the underlying
  lattice.
\end{proof}

\begin{defn}{(\emph{\zinf{} interval})}
  A \emph{\zinf{} interval} is of the form \interv{a}{b}{c}, where $a$ and $b$
  are drawn from \zinf{} with $a \leq 0 \leq b$ and $c$ is drawn from $\{
  \textsf{p}, \textsf{np} \}$. We define it as follows:
%
  \begin{equation*}
    \interv{a}{b}{c} \triangleq
      \{ n \;|\; a \leq n \leq b \wedge (c = \textsf{np} \implies n \neq 0) \}
  \end{equation*}

  Set of all such intervals (sets) is $\textit{Interval}$.
\end{defn}

\begin{prop}{}
  $(\textit{Interval}, \supseteq)$ is a sublattice of
  $(\mathcal{P}(\zinf{}),\supseteq)$ with \interv{-\infty}{\infty}{\textsf{p}}
  and \interv{0}{0}{\textsf{np}} as its top and bottom elements.
\end{prop}
%
\begin{proof}
  We know $\textit{Interval} \subseteq \mathcal{P}(\zinf{})$ since intervals
  represents sets drawn from {\zinf{}}

  Let \interv{a}{b}{c} and \interv{d}{e}{f} be two arbitrary \zinf{} intervals.
  We need to show that their meet and join both lie in \textit{Interval}.
%
  \begin{align*}
    \interv{a}{b}{c} \wedge \interv{d}{e}{f} = &
      \{ n \;|\; a \leq n \leq b \wedge (c = \textsf{np} \implies n \neq 0) \}
      \;\cup \\
      & \{ n \;|\; d \leq n \leq e \wedge (f = \textsf{np} \implies n \neq 0) \}
      \\
    = & \{ n \;|\; \min{a,d} \leq n \leq \max{b,e} \wedge ((c =
      \textsf{np} \wedge f = \textsf{np}) \implies n \neq 0) \} \\
    = & \interv{\min{\{a,d\}}}{\max{\{b,e\}}}{k}
  \end{align*}
%
  where $k$ is \textsf{np} if both $c$ and $d$ are \textsf{np} and \textsf{p}
  otherwise. Minimum of lower bound is still negative or zero and maximum of
  upper bounds is still positive or zero, hence the meet lies in
  \textit{Interval}.
%
  \begin{align*}
    \interv{a}{b}{c} \vee \interv{d}{e}{f} = &
      \{ n \;|\; a \leq n \leq b \wedge (c = \textsf{np} \implies n \neq 0) \}
      \;\cap \\
      & \{ n \;|\; d \leq n \leq e \wedge (f = \textsf{np} \implies n \neq 0) \}
      \\
    = & \{ n \;|\; \max{a,d} \leq n \leq \min{b,e} \wedge ((c =
      \textsf{np} \vee f = \textsf{np}) \implies n \neq 0) \} \\
    = & \interv{\max{\{a,d\}}}{\min{\{b,e\}}}{k}
  \end{align*}
%
  where $k$ is \textsf{np} if either $c$ or $d$ are \textsf{np} and \textsf{p}
  otherwise. Maximum of lower bound is still negative or zero and minimum of
  upper bounds is still positive or zero, hence the join lies in
  \textit{Interval}.

  \interv{-\infty}{\infty}{\textsf{p}} is then \zinf{} and
  \interv{0}{0}{\textsf{np}} is $\emptyset$. Since these are in the sublattice
  and they are the bottom and top elements in the parent lattice, they are
  necessarily the bottom and top elements in the sublattice.
\end{proof}

\begin{prop}{}
  $(\textit{Interval},\supseteq)$ is a distributive lattice.
\end{prop}
%
\begin{proof}
  It is enough to show for arbitrary $\interv{a}{b}{c}$, $\interv{d}{e}{f}$,
  $\interv{g}{h}{i}$ in \textit{Interval}, $\interv{a}{b}{c} \vee
  (\interv{d}{e}{f} \wedge \interv{g}{h}{i}) = (\interv{a}{b}{c} \vee
  \interv{d}{e}{f}) \wedge (\interv{a}{b}{c} \vee \interv{g}{h}{i})$. When this
  holds the dual law also holds. The property immediately follows from
  definition due to distributivity of set intersection over union.
\end{proof}
%
\begin{defn}{(\emph{Vector interval})}
  A vector interval is an $N$ dimensional object which has \zinf{} intervals at
  each dimension. It is denoted with $(\textit{int}_1, \dots, \textit{int}_n)$,
  where $\textit{int}_i$ is a \zinf{} interval. If \textit{vint} is a vector
  interval, $\textit{vint}_i$ denotes the \zinf{} interval at the $i^{th}$
  dimension.

  The set of $N$ dimensional vector intervals is $\textit{Interval}^N$. We
  define a partial order $\leq$ on $N$ dimensional vector intervals
  \textit{vint} and \textit{vint'} as follows:
%
  \begin{equation*}
    \textit{vint} \leq \textit{vint}' \triangleq
      \forall n. 1
        \leq i \leq N \implies \textit{vint}_i \supseteq \textit{vint'}_i
  \end{equation*}
\end{defn}
%
\begin{prop}{}
  $(\textit{Interval}^N, \leq)$ is a distributive lattice.
\end{prop}
%
\begin{proof}
  It is a standard result that product of two distributive lattices is a
  distributive lattice itself when the partial order on the resulting lattice is
  defined as the componentwise comparison. This generalises to $N$ dimensions by
  induction.
\end{proof}

\begin{defn}{}
  \textsf{Mult} and \textsf{Approx} are parametric labelled variant types with
  injections given by their definition:
%
  \begin{align*}
    \textsf{Mult} \; a &
      \triangleq \textsf{mult} \; a \;\mid\; \textsf{only} \; a \\
    \textsf{Approx} \; a &
      \triangleq \textsf{exact} \; a \;\mid\; \textsf{lower} \; a \;\mid\;
        \textsf{upper} \; a \;\mid\; \textsf{both} \; a \; a
  \end{align*}
%
  \eg{}, \textsf{lower} is an injection $\mathsf{lower} : a \to \mathsf{Approx}
  \; a$ etc.

  These will be used in the following subsection to give meaning to the
  specification modifiers for approximation and multiplicity.
\end{defn}

\subsection{Denotational semantics for specifications}

We give some basic interpretations: $\interp{\texttt{n}} = n$ where $n$ is in
$\mathbb{Z}$, $\interp{pointed} = \llbracket \epsilon \rrbracket = \textsf{p}$,
$\interp{\texttt{nonpointed}} = \textsf{np}$. Typographically, cursive font is
used within specifications to indicate a partial textual capture.

Let $\textit{vec-gen}_N : \mathbb{N} \times \textit{Interval} \to
\textit{Interval}^N$ be the function generating an $N$ dimensional vector
interval such that \vecgen{N}{i}{\interv{a}{b}{c}} has
$\interv{-\infty}{\infty}{\textsf{p}}$ in all dimensions except the $i^{th}$
one. There, it has \interv{a}{b}{c}.

First, we give meaning to region constants (introduced in \Cref{}).
%
\begin{align*}
  \interp{\texttt{pointed(dim=\textit{i})}}_N & =
    \vecgen{N}{\interp{\textit{i}}}{\interv{0}{0}{\textsf{p}}}\\
%
  \interp{\texttt{centered(dim=\textit{i}, depth=\textit{k}, \textit{p})}}_N & =
    \vecgen{N}
           {\interp{\textit{i}}}
           {\interv{-\interp{\textit{k}}}
                   {\interp{\textit{k}}}
                   {\interp{\textit{p}}}
           } \\
%
  \interp{\texttt{forward(dim=\textit{i}, depth=\textit{k}, \textit{p})}}_N & =
    \vecgen{N}
           {\interp{\textit{i}}}
           {\interv{0}
                   {\interp{\textit{k}}}
                   {\interp{\textit{p}}}
           } \\
%
  \interp{\texttt{backward(dim=\textit{i}, depth=\textit{k}, \textit{p})}}_N & =
    \vecgen{N}
           {\interp{\textit{i}}}
           {\interv{-\interp{\textit{k}}}
                   {0}
                   {\interp{\textit{p}}}
           }
\end{align*}

Second, we provide the semantics for the permissive region operator, \texttt{+},
and the restrictive region operator, \texttt{*}, by associating them with meet
and join respectively. Let \texttt{\textit{R}} and \texttt{\textit{S}} be
regions.
%
\begin{align*}
  \interp{\texttt{\textit{R} + \textit{S}}}_N & =
    \interp{\texttt{\textit{R}}}_N \wedge \interp{\texttt{\textit{S}}}_N
&
  \interp{\texttt{\textit{R} * \textit{S}}}_N & =
    \interp{\texttt{\textit{R}}}_N \vee \interp{\texttt{\textit{S}}}_N
\end{align*}

Third, we need to mark presence of modifiers such as \texttt{readOnce} and
\texttt{atMost} as introduced in \Cref{}. In the context of
multiplicity modifiers we have the following semantics:
%
\begin{align*}
  \interp{readOnce} & = \textsf{mult} &
  \llbracket\epsilon\rrbracket & = \textsf{once}
\end{align*}

In the context of approximation modifiers, we have
%
\begin{align*}
  \interp{atLeast} & = \textsf{lower} &
  \interp{atMost} & = \textsf{upper} &
  \llbracket \epsilon \rrbracket = \textsf{exact}
\end{align*}

Finally, we can give full semantics of a stencil specification:
%
\begin{equation*}
  \interp{stencil \textit{Mult}, \textit{Appr}, \textit{Region}} =
    \interp{\textit{Mult}}{(\interp{\textit{Appr}}(\interp{\textit{Region}}))}
\end{equation*}

\section{Equational theory}

\section{Analysis, Consistency, Inference}

\subsection{Analysis}

\subsection{Consisteny}

\subsection{Inference}

\end{document}

### 2017-03-22 oopsla17/lattice-model.tex
what I got

@@
 \usepackage{cleveref}
+\usepackage{fixltx2e}
 
@@
 
-\newcommand{\zinf}{\textnormal{$\mathbb{Z}^\infty$}}
+\newcommand{\zinf}{\textnormal{$\mathbb{Z}_\infty$}}
 \newcommand{\interp}[1]{\llbracket\texttt{#1}\rrbracket}
@@
 \newcommand{\vecgen}[3]{\textnormal{$\textit{vec-gen}_{#1}(#2,#3)$}}
+\newcommand{\textcap}[1]{\texttt{\textit{#1}}}
 
@@
 \begin{defn}{(\emph{Offset space})}
-  \emph{Offset space} is a subset of \zinf{} members. $\mathcal{P}(\zinf{})$
-  couple with $\supseteq$ is the power set distributive lattice.
+  \emph{Offset space} is a subset of \zinf{} members. It is used to capture
+  offsets from induction variables.
+
+%  $\mathcal{P}(\zinf{})$ couple with $\supseteq$ is the power set distributive
+%  lattice.
 \end{defn}
-
-\begin{defn}{(\emph{Offsets vector})}
-  An \emph{offsets vector} is a $N$ dimensional vector of offset spaces denoted
-  by $(\textit{offs}_1, \dots, \textit{offs}_N)$.
-
-  The set of $N$ dimensional offsets vectors is $\mathcal{P}{(\zinf{})}^N$. We
-  define the partial order $\leq$ on $N$ dimensional offsets vectors
-  \textit{voffs} and \textit{voffs'} as follows:
-%
-  \begin{equation*}
-    \textit{voffs} \leq \textit{voffs'} \triangleq
-      \forall n. 1
-        \leq i \leq N \implies \textit{voffs}_i \supseteq \textit{voffs'}_i
-  \end{equation*}
-\end{defn}
-
-\begin{prop}{}
-  $(\mathcal{P}{(\zinf{})}^N, \leq)$ is a distributive lattice.
-\end{prop}
-%
-\begin{proof}
-  Product of two distributive lattices is itself a distributive lattice when the
-  partial order is defined as componentwise comparison in the underlying
-  lattice.
-\end{proof}
 
@@
   A \emph{\zinf{} interval} is of the form \interv{a}{b}{c}, where $a$ and $b$
-  are drawn from \zinf{} with $a \leq 0 \leq b$ and $c$ is drawn from $\{
-  \textsf{p}, \textsf{np} \}$. We define it as follows:
+  are drawn from \zinf{} with $a \leq 0 \leq b$ and $c$ is a Boolean value. We
+  define it as follows:
 %
@@
     \interv{a}{b}{c} \triangleq
-      \{ n \;|\; a \leq n \leq b \wedge (c = \textsf{np} \implies n \neq 0) \}
+      \{ n \;|\; a \leq n \leq b \wedge (\neg c \implies n \neq 0) \}
   \end{equation*}
@@
 \begin{prop}{}
-  $(\textit{Interval}, \supseteq)$ is a sublattice of
-  $(\mathcal{P}(\zinf{}),\supseteq)$ with \interv{-\infty}{\infty}{\textsf{p}}
-  and \interv{0}{0}{\textsf{np}} as its top and bottom elements.
+  We have the following identity for \zinf{} intervals:
+%
+  \begin{equation*}
+    \interv{a}{b}{c} \cap \interv{d}{e}{f} =
+      \interv{\max \{a,d\}}{\min \{b,e\}}{\neg c \vee \neg f}
+  \end{equation*}\label{def:zinf-intersect}
 \end{prop}
@@
 \begin{proof}
-  We know $\textit{Interval} \subseteq \mathcal{P}(\zinf{})$ since intervals
-  represents sets drawn from {\zinf{}}
-
-  Let \interv{a}{b}{c} and \interv{d}{e}{f} be two arbitrary \zinf{} intervals.
-  We need to show that their meet and join both lie in \textit{Interval}.
-%
   \begin{align*}
-    \interv{a}{b}{c} \wedge \interv{d}{e}{f} = &
-      \{ n \;|\; a \leq n \leq b \wedge (c = \textsf{np} \implies n \neq 0) \}
-      \;\cup \\
-      & \{ n \;|\; d \leq n \leq e \wedge (f = \textsf{np} \implies n \neq 0) \}
+    \interv{a}{b}{c} \cap \interv{d}{e}{f} = &
+      \; \{ n \;|\; a \leq n \leq b \wedge (\neg c \implies n \neq 0) \}
+      \;\cap \\
+      & \; \{ n \;|\; d \leq n \leq e \wedge (\neg f \implies n \neq 0) \}
       \\
-    = & \{ n \;|\; \min{a,d} \leq n \leq \max{b,e} \wedge ((c =
-      \textsf{np} \wedge f = \textsf{np}) \implies n \neq 0) \} \\
-    = & \interv{\min{\{a,d\}}}{\max{\{b,e\}}}{k}
+    = & \; \{ n \;|\; \max \{a,d\} \leq n \leq \min \{b,e\} \wedge (\neg c
+      \vee \neg f \implies n \neq 0) \} \\
+    = & \; \interv{\max{\{a,d\}}}{\min{\{b,e\}}}{\neg c \vee \neg f}
   \end{align*}
-%
-  where $k$ is \textsf{np} if both $c$ and $d$ are \textsf{np} and \textsf{p}
-  otherwise. Minimum of lower bound is still negative or zero and maximum of
-  upper bounds is still positive or zero, hence the meet lies in
-  \textit{Interval}.
-%
-  \begin{align*}
-    \interv{a}{b}{c} \vee \interv{d}{e}{f} = &
-      \{ n \;|\; a \leq n \leq b \wedge (c = \textsf{np} \implies n \neq 0) \}
-      \;\cap \\
-      & \{ n \;|\; d \leq n \leq e \wedge (f = \textsf{np} \implies n \neq 0) \}
-      \\
-    = & \{ n \;|\; \max{a,d} \leq n \leq \min{b,e} \wedge ((c =
-      \textsf{np} \vee f = \textsf{np}) \implies n \neq 0) \} \\
-    = & \interv{\max{\{a,d\}}}{\min{\{b,e\}}}{k}
-  \end{align*}
-%
-  where $k$ is \textsf{np} if either $c$ or $d$ are \textsf{np} and \textsf{p}
-  otherwise. Maximum of lower bound is still negative or zero and minimum of
-  upper bounds is still positive or zero, hence the join lies in
-  \textit{Interval}.
-
-  \interv{-\infty}{\infty}{\textsf{p}} is then \zinf{} and
-  \interv{0}{0}{\textsf{np}} is $\emptyset$. Since these are in the sublattice
-  and they are the bottom and top elements in the parent lattice, they are
-  necessarily the bottom and top elements in the sublattice.
 \end{proof}
 
-\begin{prop}{}
-  $(\textit{Interval},\supseteq)$ is a distributive lattice.
-\end{prop}
+%\begin{prop}{}
+%  $(\textit{Interval}, \supseteq)$ is a sublattice of
+%  $(\mathcal{P}(\zinf{}),\supseteq)$ with \interv{-\infty}{\infty}{\textsf{p}}
+%  and \interv{0}{0}{\textsf{np}} as its top and bottom elements.
+%\end{prop}
+%%
+%\begin{proof}
+%  We know $\textit{Interval} \subseteq \mathcal{P}(\zinf{})$ since intervals
+%  represents sets drawn from {\zinf{}}
 %
-\begin{proof}
-  It is enough to show for arbitrary $\interv{a}{b}{c}$, $\interv{d}{e}{f}$,
-  $\interv{g}{h}{i}$ in \textit{Interval}, $\interv{a}{b}{c} \vee
-  (\interv{d}{e}{f} \wedge \interv{g}{h}{i}) = (\interv{a}{b}{c} \vee
-  \interv{d}{e}{f}) \wedge (\interv{a}{b}{c} \vee \interv{g}{h}{i})$. When this
-  holds the dual law also holds. The property immediately follows from
-  definition due to distributivity of set intersection over union.
-\end{proof}
+%  Let \interv{a}{b}{c} and \interv{d}{e}{f} be two arbitrary \zinf{} intervals.
+%  We need to show that their meet and join both lie in \textit{Interval}.
+%%
+%  \begin{align*}
+%    \interv{a}{b}{c} \wedge \interv{d}{e}{f} = &
+%      \{ n \;|\; a \leq n \leq b \wedge (c = \textsf{np} \implies n \neq 0) \}
+%      \;\cup \\
+%      & \{ n \;|\; d \leq n \leq e \wedge (f = \textsf{np} \implies n \neq 0) \}
+%      \\
+%    = & \{ n \;|\; \min{a,d} \leq n \leq \max{b,e} \wedge ((c =
+%      \textsf{np} \wedge f = \textsf{np}) \implies n \neq 0) \} \\
+%    = & \interv{\min{\{a,d\}}}{\max{\{b,e\}}}{k}
+%  \end{align*}
+%%
+%  where $k$ is \textsf{np} if both $c$ and $d$ are \textsf{np} and \textsf{p}
+%  otherwise. Minimum of lower bound is still negative or zero and maximum of
+%  upper bounds is still positive or zero, hence the meet lies in
+%  \textit{Interval}.
+%%
+%  \begin{align*}
+%    \interv{a}{b}{c} \vee \interv{d}{e}{f} = &
+%      \{ n \;|\; a \leq n \leq b \wedge (c = \textsf{np} \implies n \neq 0) \}
+%      \;\cap \\
+%      & \{ n \;|\; d \leq n \leq e \wedge (f = \textsf{np} \implies n \neq 0) \}
+%      \\
+%    = & \{ n \;|\; \max{a,d} \leq n \leq \min{b,e} \wedge ((c =
+%      \textsf{np} \vee f = \textsf{np}) \implies n \neq 0) \} \\
+%    = & \interv{\max{\{a,d\}}}{\min{\{b,e\}}}{k}
+%  \end{align*}
+%%
+%  where $k$ is \textsf{np} if either $c$ or $d$ are \textsf{np} and \textsf{p}
+%  otherwise. Maximum of lower bound is still negative or zero and minimum of
+%  upper bounds is still positive or zero, hence the join lies in
+%  \textit{Interval}.
 %
+%  \interv{-\infty}{\infty}{\textsf{p}} is then \zinf{} and
+%  \interv{0}{0}{\textsf{np}} is $\emptyset$. Since these are in the sublattice
+%  and they are the bottom and top elements in the parent lattice, they are
+%  necessarily the bottom and top elements in the sublattice.
+%\end{proof}
+
+%\begin{prop}{}
+%  $(\textit{Interval},\supseteq)$ is a distributive lattice.
+%\end{prop}
+%%
+%\begin{proof}
+%  It is enough to show for arbitrary $\interv{a}{b}{c}$, $\interv{d}{e}{f}$,
+%  $\interv{g}{h}{i}$ in \textit{Interval}, $\interv{a}{b}{c} \vee
+%  (\interv{d}{e}{f} \wedge \interv{g}{h}{i}) = (\interv{a}{b}{c} \vee
+%  \interv{d}{e}{f}) \wedge (\interv{a}{b}{c} \vee \interv{g}{h}{i})$. When this
+%  holds the dual law also holds. The property immediately follows from
+%  definition due to distributivity of set intersection over union.
+%\end{proof}
+
+\begin{defn}{(\emph{Offsets vector})}
+  An \emph{offsets vector} is a $N$ dimensional vector of offset spaces denoted
+  by $(\textit{offs}_1, \dots, \textit{offs}_N)$ to mean $\textit{offs}_1 \times
+  \dots \times \textit{offs}_N$.
+
+  The set of $N$ dimensional offsets vectors is
+  $\mathcal{P}{(\zinf{}^{\mkern-19mu{N}})}$ and they trivially form a
+  distributive lattice under the superset relation.
+\end{defn}
+
 \begin{defn}{(\emph{Vector interval})}
@@
 
-  The set of $N$ dimensional vector intervals is $\textit{Interval}^N$. We
-  define a partial order $\leq$ on $N$ dimensional vector intervals
-  \textit{vint} and \textit{vint'} as follows:
+  The set of $N$ dimensional vector intervals is $\textit{Interval}^N$.
+\end{defn}
+
+\begin{lem}{}
+  Intersection distributes over vector intervals.
 %
   \begin{equation*}
-    \textit{vint} \leq \textit{vint}' \triangleq
-      \forall n. 1
-        \leq i \leq N \implies \textit{vint}_i \supseteq \textit{vint'}_i
-  \end{equation*}
-\end{defn}
+    (\textit{vint}_1, \dots, \textit{vint}_N) \;\cap\;
+      (\textit{vint'}_1, \dots, \textit{vint'}_N)
+      =
+    (\textit{vint}_1 \;\cap\; \textit{vint'}_1, \dots,
+     \textit{vint}_N \;\cap\; \textit{vint'}_N)
+  \end{equation*}\label{lem:vector-dist}
+\end{lem}
 %
 \begin{prop}{}
-  $(\textit{Interval}^N, \leq)$ is a distributive lattice.
+  We have the following identity for $N$ dimensional vector intervals:
+%
+  \begin{gather*}
+    (\interv{a_1}{b_1}{c_1}, \dots, \interv{a_N}{b_N}{c_N}) \; \cap \;
+      (\interv{a'_1}{b'_1}{c'_1}, \dots, \interv{a'_N}{b'_N}{c'_N}) \\
+    = \\
+    (\interv{\max \{a_1,a'_1\}}{\min \{b_1,b'_1\}}{\neg c_1 \vee \neg c'_1},
+      \dots, \interv{\max \{a_N,a'_N\}}{\min \{b_N,b'_N\}}{\neg c_N \vee \neg c'_N})
+  \end{gather*}
 \end{prop}
@@
 \begin{proof}
-  It is a standard result that product of two distributive lattices is a
-  distributive lattice itself when the partial order on the resulting lattice is
-  defined as the componentwise comparison. This generalises to $N$ dimensions by
-  induction.
+  This immediately follows from \Cref{lem:vector-dist} and
+  \Cref{def:zinf-intersect}.
+\end{proof}
+
+\begin{defn}{(\emph{Region space})}
+  Region space defines the set \textit{Region} in $N$ dimensions inductively as
+  follows:
+%
+  \begin{enumerate}
+    \item If $R$ is in $\textit{Interval}^N$, then $R$ is in \textit{Region}.
+    \item If $R$ and $S$ are in \textit{Region}, then so are $R \cap S$ and $R
+      \cup S$.
+  \end{enumerate}
+\end{defn}
+
+\begin{prop}{}
+  $(\textit{Region},\supseteq)$ is a distributive lattice.
+\end{prop}
+%
+\begin{proof}
+  Any sublattice of a distributive lattice is a distributive lattice. Power sets
+  coupled with partially ordered with superset or equals relation are
+  distributive lattices. \textit{Region} is a subset of
+  $\mathcal{P}{(\zinf{}^{\mkern-19mu{N}})}$ since all vector intervals are
+  subsets of this power set and union and intersection cannot add an element
+  that is not already in the powerset. Therefore, \textit{Region} is a
+  distributive sublattice of $\mathcal{P}{(\zinf{}^{\mkern-19mu{N}})}$.
 \end{proof}
@@
 We give some basic interpretations: $\interp{\texttt{n}} = n$ where $n$ is in
-$\mathbb{Z}$, $\interp{pointed} = \llbracket \epsilon \rrbracket = \textsf{p}$,
-$\interp{\texttt{nonpointed}} = \textsf{np}$. Typographically, cursive font is
-used within specifications to indicate a partial textual capture.
+$\mathbb{Z}$, $\interp{pointed} = \llbracket \epsilon \rrbracket =
+\textsf{true}$, $\interp{\texttt{nonpointed}} = \textsf{false}$.
+Typographically, cursive font is used within specifications to indicate a
+partial textual capture.
 
@@
   \interp{\texttt{pointed(dim=\textit{i})}}_N & =
-    \vecgen{N}{\interp{\textit{i}}}{\interv{0}{0}{\textsf{p}}}\\
+    \vecgen{N}{\interp{\textit{i}}}{\interv{0}{0}{\textsf{true}}}\\
 %
@@
 
-\section{Equational theory}
+\section{Equational \& approximation theories}
+
+We know introduce the equational and approximation theory of stencil
+specifications based on $\equiv$ and $<\:$.
+
+\subsection{Equivalences}
+
+\begin{description}
+  \item[Basic] \texttt{*} and \texttt{+} are both idempotent, commutative, and
+    associative;
+%
+  \item[Dist] \texttt{*} distributed over \texttt{+} and dually
+    \texttt{+} distributes over \texttt{*}, meaning if \textcap{R}, \textcap{S},
+    and \textcap{T} are regions, then we have the following equivalences:
+%
+    \begin{align*}
+      \texttt{\textcap{R}*(\textcap{S}+\textcap{T})} & \equiv
+        \texttt{(\textcap{R}*\textcap{S})+(\textcap{R}*\textcap{T})} &
+      \texttt{\textcap{R}+(\textcap{S}*\textcap{T})} & \equiv
+        \texttt{(\textcap{R}+\textcap{S})*(\textcap{R}+\textcap{T})}
+    \end{align*}
+%
+  \item[Overlapping pointed] If \textcap{R} is one of \texttt{forward},
+    \texttt{backward}, or \texttt{centered}, then we have
+    $\texttt{\textcap{R}(depth=\textcap{k},dim=\textcap{n},\_) +
+    pointed(dim=\textcap{n})} \equiv
+    \texttt{\textcap{R}(depth=\textcap{k},dim=\textcap{n})}$.
+%
+  \item[Centered] The region constants \texttt{forward} and \texttt{backward}
+    are two halves of \texttt{centered} specifications:
+%
+    \begin{gather*}
+      \texttt{centered(depth=\textcap{k},dim=\textit{n},\textcap{p1})} \\
+      \equiv \\
+      \texttt{forward(depth=\textcap{k},dim=\textit{n},\textcap{p2}) +
+        backward(depth=\textcap{k},dim=\textit{n},\textcap{p3})}
+    \end{gather*}
+%
+    Here \textcap{p1} is \texttt{nonpointed} if both \textcap{p1} and
+    \textcap{p3} are \texttt{nonpointed} and \texttt{pointed} otherwise.
+\end{description}
+
+\subsection{Approximations}
 

### 2017-03-23 oopsla17/analysis/Analyse.hs
some bonus functionality for assessing performance changes

@@ -65,7 +65,13 @@ main :: IO ()
 main = do
     args <- getArgs
     if (length args == 1) then
-      applyAnalysisToDir (head args) False []
+      if (head args == "PI") then
+        testLHSvars True
+      else 
+      if (head args == "PI2") then
+        testLHSvars False
+      else
+        applyAnalysisToDir (head args) False []
     else if (length args > 1) then
       if (args !! 1 == "-d") then
          applyAnalysisToDir (head args) True (tail (tail args))
@@ -74,6 +80,17 @@ main = do
     else
       putStrLn $ "Please specify a directory on which to apply the analysis\
                  \ followed by any number of file names to be excluded "
+
+testLHSvars flag = do
+    files <- readParseSrcDir "experiment" []
+    putStrLn $ show $ map (applyLHSvars flag) files
+
+applyLHSvars :: Bool -> (Filename, SourceText, F.ProgramFile A) -> [F.Name]
+applyLHSvars flag (filename, source, pf) =
+    if flag then FA.allLhsVars pf' else FA.allLhsVarsRobust pf' 
+  where
+    pf' = FA.initAnalysis $ pf
+
 
 applyAnalysisToDir :: String -> Bool -> [String] -> IO ()
 applyAnalysisToDir dir debug excludes = do

### 2017-03-25 oopsla17/lattice-model.tex
Complete equational theory proof

@@
 \usepackage{cleveref}
-\usepackage{fixltx2e}
 
@@
 \theoremstyle{plain}
+\newtheorem{thm}{Theorem}
 \newtheorem{lem}{Lemma}
 \newtheorem{prop}{Proposition}
+\newtheorem{case}{Case}
 
@@
 \newcommand{\zinf}{\textnormal{$\mathbb{Z}_\infty$}}
-\newcommand{\interp}[1]{\llbracket\texttt{#1}\rrbracket}
+\newcommand{\interp}[1]{\llbracket#1\rrbracket}
 \newcommand{\interv}[3]{\textnormal{$\lbrack#1,#2\rbrack^{#3}$}}
@@
 
-\begin{prop}{}
-  We have the following identity for \zinf{} intervals:
+\begin{lem}{}
+  We have the following dual identities for \zinf{} intervals:
 %
-  \begin{equation*}
-    \interv{a}{b}{c} \cap \interv{d}{e}{f} =
-      \interv{\max \{a,d\}}{\min \{b,e\}}{\neg c \vee \neg f}
-  \end{equation*}\label{def:zinf-intersect}
-\end{prop}
+  \begin{align*}
+    \interv{a}{b}{c} \cap \interv{d}{e}{f} & =
+      \interv{\max \{a,d\}}{\min \{b,e\}}{c \wedge f} \\
+    \interv{a}{b}{c} \cup \interv{d}{e}{f} & =
+      \interv{\min \{a,d\}}{\max \{b,e\}}{c \vee f}
+  \end{align*}\label{lem:zinf-identities}
+\end{lem}
 %
 \begin{proof}
+  We give the proof of the first identity and the second one is similar.
   \begin{align*}
@@
 
-%\begin{prop}{}
-%  $(\textit{Interval}, \supseteq)$ is a sublattice of
-%  $(\mathcal{P}(\zinf{}),\supseteq)$ with \interv{-\infty}{\infty}{\textsf{p}}
-%  and \interv{0}{0}{\textsf{np}} as its top and bottom elements.
-%\end{prop}
-%%
-%\begin{proof}
-%  We know $\textit{Interval} \subseteq \mathcal{P}(\zinf{})$ since intervals
-%  represents sets drawn from {\zinf{}}
-%
-%  Let \interv{a}{b}{c} and \interv{d}{e}{f} be two arbitrary \zinf{} intervals.
-%  We need to show that their meet and join both lie in \textit{Interval}.
-%%
-%  \begin{align*}
-%    \interv{a}{b}{c} \wedge \interv{d}{e}{f} = &
-%      \{ n \;|\; a \leq n \leq b \wedge (c = \textsf{np} \implies n \neq 0) \}
-%      \;\cup \\
-%      & \{ n \;|\; d \leq n \leq e \wedge (f = \textsf{np} \implies n \neq 0) \}
-%      \\
-%    = & \{ n \;|\; \min{a,d} \leq n \leq \max{b,e} \wedge ((c =
-%      \textsf{np} \wedge f = \textsf{np}) \implies n \neq 0) \} \\
-%    = & \interv{\min{\{a,d\}}}{\max{\{b,e\}}}{k}
-%  \end{align*}
-%%
-%  where $k$ is \textsf{np} if both $c$ and $d$ are \textsf{np} and \textsf{p}
-%  otherwise. Minimum of lower bound is still negative or zero and maximum of
-%  upper bounds is still positive or zero, hence the meet lies in
-%  \textit{Interval}.
-%%
-%  \begin{align*}
-%    \interv{a}{b}{c} \vee \interv{d}{e}{f} = &
-%      \{ n \;|\; a \leq n \leq b \wedge (c = \textsf{np} \implies n \neq 0) \}
-%      \;\cap \\
-%      & \{ n \;|\; d \leq n \leq e \wedge (f = \textsf{np} \implies n \neq 0) \}
-%      \\
-%    = & \{ n \;|\; \max{a,d} \leq n \leq \min{b,e} \wedge ((c =
-%      \textsf{np} \vee f = \textsf{np}) \implies n \neq 0) \} \\
-%    = & \interv{\max{\{a,d\}}}{\min{\{b,e\}}}{k}
-%  \end{align*}
-%%
-%  where $k$ is \textsf{np} if either $c$ or $d$ are \textsf{np} and \textsf{p}
-%  otherwise. Maximum of lower bound is still negative or zero and minimum of
-%  upper bounds is still positive or zero, hence the join lies in
-%  \textit{Interval}.
-%
-%  \interv{-\infty}{\infty}{\textsf{p}} is then \zinf{} and
-%  \interv{0}{0}{\textsf{np}} is $\emptyset$. Since these are in the sublattice
-%  and they are the bottom and top elements in the parent lattice, they are
-%  necessarily the bottom and top elements in the sublattice.
-%\end{proof}
-
-%\begin{prop}{}
-%  $(\textit{Interval},\supseteq)$ is a distributive lattice.
-%\end{prop}
-%%
-%\begin{proof}
-%  It is enough to show for arbitrary $\interv{a}{b}{c}$, $\interv{d}{e}{f}$,
-%  $\interv{g}{h}{i}$ in \textit{Interval}, $\interv{a}{b}{c} \vee
-%  (\interv{d}{e}{f} \wedge \interv{g}{h}{i}) = (\interv{a}{b}{c} \vee
-%  \interv{d}{e}{f}) \wedge (\interv{a}{b}{c} \vee \interv{g}{h}{i})$. When this
-%  holds the dual law also holds. The property immediately follows from
-%  definition due to distributivity of set intersection over union.
-%\end{proof}
-
 \begin{defn}{(\emph{Offsets vector})}
@@
 
+\begin{lem}{}
+  Intersection distributes over offsets vectors.
+%
+  \begin{equation*}
+    (\textit{vint}_1, \dots, \textit{vint}_N) \;\cap\;
+      (\textit{vint'}_1, \dots, \textit{vint'}_N)
+      =
+    (\textit{vint}_1 \;\cap\; \textit{vint'}_1, \dots,
+     \textit{vint}_N \;\cap\; \textit{vint'}_N)
+  \end{equation*}\label{lem:vector-intersect}
+\end{lem}
+
+\begin{lem}{}
+  Let $v$ and $v'$ be two $N$ dimensional offsets vectors such that $v_i$ is
+  equal to $v'_i$ for all $1 \leq i \leq N$ except $k$, then we have the
+  following:
+%
+  \begin{equation*}
+    (v_1,\dotsc, v_k,\dotsc, v_N) \; \cup \; (v'_1,\dotsc, v'_k, \dotsc, v_N)
+    =
+    (v_1,\dotsc, v_k \cup v'_k,\dotsc, v_N)
+  \end{equation*}\label{lem:vector-union}
+\end{lem}
+%
+\begin{proof}
+  \begin{align*}
+    v \; \cup \; v' &
+    = \{(x_1,\dotsc,x_n) \mid
+          \bigwedge_{1 \leq i \leq N } x_i \in v_i \}
+      \;\cup\;
+      \{(x_1,\dotsc,x_n) \mid
+          \bigwedge_{1 \leq i \leq N } x_i \in v'_i \} \\
+    & = \{(x_1,\dotsc,x_n) \mid
+          \bigwedge_{\substack{1 \leq i \leq N \\ i \neq k}} x_i \in v_i
+              \wedge x_k \in v_k \vee
+          \bigwedge_{\substack{1 \leq i \leq N \\ i \neq k}} x_i \in v'_i
+            \wedge x_k \in v'_k
+        \} \\
+    & = \{(x_1,\dotsc,x_n) \mid
+          \bigwedge_{\substack{1 \leq i \leq N \\ i \neq k}} x_i \in v_i
+          \wedge x_k \in v_k \cup v'_k
+        \} \\
+    & = (v_1,\dotsc, v_k \cup v'_k,\dotsc, v_N)
+  \end{align*}
+\end{proof}
+
 \begin{defn}{(\emph{Vector interval})}
@@
 
-  The set of $N$ dimensional vector intervals is $\textit{Interval}^N$.
+  The set of $N$ dimensional vector intervals is $\textit{Interval}^N$. Every
+  vector interval is also an offsets vector.
 \end{defn}
@@
 \begin{lem}{}
-  Intersection distributes over vector intervals.
-%
-  \begin{equation*}
-    (\textit{vint}_1, \dots, \textit{vint}_N) \;\cap\;
-      (\textit{vint'}_1, \dots, \textit{vint'}_N)
-      =
-    (\textit{vint}_1 \;\cap\; \textit{vint'}_1, \dots,
-     \textit{vint}_N \;\cap\; \textit{vint'}_N)
-  \end{equation*}\label{lem:vector-dist}
-\end{lem}
-%
-\begin{prop}{}
-  We have the following identity for $N$ dimensional vector intervals:
+  Intersection distributes $N$ dimensional vector intervals:
 %
@@
   \end{gather*}
-\end{prop}
+\end{lem}
 %
 \begin{proof}
-  This immediately follows from \Cref{lem:vector-dist} and
-  \Cref{def:zinf-intersect}.
+  This immediately follows from \Cref{lem:vector-intersect} and
+  \Cref{lem:zinf-identities}.
 \end{proof}
@@
 We give some basic interpretations: $\interp{\texttt{n}} = n$ where $n$ is in
-$\mathbb{Z}$, $\interp{pointed} = \llbracket \epsilon \rrbracket =
+$\mathbb{Z}$, $\interp{\texttt{pointed}} = \llbracket \epsilon \rrbracket =
 \textsf{true}$, $\interp{\texttt{nonpointed}} = \textsf{false}$.
@@
 \begin{align*}
-  \interp{\texttt{pointed(dim=\textit{i})}}_N & =
-    \vecgen{N}{\interp{\textit{i}}}{\interv{0}{0}{\textsf{true}}}\\
+  \interp{\texttt{pointed(dim=\textcap{i})}}_N & =
+    \vecgen{N}{\interp{\textcap{i}}}{\interv{0}{0}{\textsf{true}}}\\
 %
-  \interp{\texttt{centered(dim=\textit{i}, depth=\textit{k}, \textit{p})}}_N & =
-    \vecgen{N}
-           {\interp{\textit{i}}}
-           {\interv{-\interp{\textit{k}}}
-                   {\interp{\textit{k}}}
-                   {\interp{\textit{p}}}
+  \interp{\texttt{centered(dim=\textcap{i}, depth=\textcap{k}, \textcap{p})}}_N
+    & = \vecgen{N}
+           {\interp{\textcap{i}}}
+           {\interv{-\interp{\textcap{k}}}
+                   {\interp{\textcap{k}}}
+                   {\interp{\textcap{p}}}
            } \\
 %
-  \interp{\texttt{forward(dim=\textit{i}, depth=\textit{k}, \textit{p})}}_N & =
-    \vecgen{N}
-           {\interp{\textit{i}}}
+  \interp{\texttt{forward(dim=\textcap{i}, depth=\textcap{k}, \textcap{p})}}_N
+    & = \vecgen{N}
+           {\interp{\textcap{i}}}
            {\interv{0}
-                   {\interp{\textit{k}}}
-                   {\interp{\textit{p}}}
+                   {\interp{\textcap{k}}}
+                   {\interp{\textcap{p}}}
            } \\
 %
-  \interp{\texttt{backward(dim=\textit{i}, depth=\textit{k}, \textit{p})}}_N & =
-    \vecgen{N}
-           {\interp{\textit{i}}}
-           {\interv{-\interp{\textit{k}}}
+  \interp{\texttt{backward(dim=\textcap{i}, depth=\textcap{k}, \textcap{p})}}_N
+    & = \vecgen{N}
+           {\interp{\textcap{i}}}
+           {\interv{-\interp{\textcap{k}}}
                    {0}
-                   {\interp{\textit{p}}}
+                   {\interp{\textcap{p}}}
            }
@@
 \begin{align*}
-  \interp{\texttt{\textit{R} + \textit{S}}}_N & =
-    \interp{\texttt{\textit{R}}}_N \wedge \interp{\texttt{\textit{S}}}_N
+  \interp{\texttt{\textcap{R} + \textcap{S}}}_N & =
+    \interp{\textcap{R}}_N \wedge \interp{\textcap{S}}_N
 &
-  \interp{\texttt{\textit{R} * \textit{S}}}_N & =
-    \interp{\texttt{\textit{R}}}_N \vee \interp{\texttt{\textit{S}}}_N
+  \interp{\texttt{\textcap{R} * \textcap{S}}}_N & =
+    \interp{\textcap{R}}_N \vee \interp{\textcap{S}}_N
 \end{align*}
@@
 \begin{align*}
-  \interp{readOnce} & = \textsf{mult} &
+  \interp{\texttt{readOnce}} & = \textsf{mult} &
   \llbracket\epsilon\rrbracket & = \textsf{once}
@@
 \begin{align*}
-  \interp{atLeast} & = \textsf{lower} &
-  \interp{atMost} & = \textsf{upper} &
+  \interp{\texttt{atLeast}} & = \textsf{lower} &
+  \interp{\texttt{atMost}} & = \textsf{upper} &
   \llbracket \epsilon \rrbracket = \textsf{exact}
@@
 \begin{equation*}
-  \interp{stencil \textit{Mult}, \textit{Appr}, \textit{Region}} =
-    \interp{\textit{Mult}}{(\interp{\textit{Appr}}(\interp{\textit{Region}}))}
+  \interp{\texttt{stencil \textcap{Mult}, \textcap{Appr}, \textcap{Region}}} =
+    \interp{\textcap{Mult}}
+           {(\interp{\textcap{Appr}}
+                    {(\interp{\textcap{Region}})})}
 \end{equation*}
@@
 We know introduce the equational and approximation theory of stencil
-specifications based on $\equiv$ and $<\:$.
+specifications based on $\equiv$ and $\prec$.
 
@@
 %
-    Here \textcap{p1} is \texttt{nonpointed} if both \textcap{p1} and
+    Here \textcap{p1} is \texttt{nonpointed} if both \textcap{p2} and
     \textcap{p3} are \texttt{nonpointed} and \texttt{pointed} otherwise.
+%
+  \item[Point] Let \textcap{R} and \textcap{S} be region constants with two
+    variants a superscript \textit{f} to mean nonpointed and \textit{t} to mean
+    pointed and \textcap{dr} \& \textcap{ds} be the dimensions \textcap{R} and
+    \textcap{S} acts on. Then we have the following dual equivalences:
+%
+    \begin{align*}
+      (\textcap{R}^{\textit{t}} \texttt{*} \textcap{S}^{\textit{f}})
+      \;\texttt{+}\;
+      (\textcap{R}^{\textit{f}} \texttt{*} \textcap{S}^{\textit{t}})
+      \;\texttt{+}\;
+      \texttt{p(dim=\textcap{dr})*p(dim=\textcap{ds})} & =
+        \textcap{R}^{\textit{t}} \texttt{*} \textcap{S}^{\textit{t}} \\
+%
+      (\textcap{R}^{\textit{t}} \texttt{+} \textcap{S}^{\textit{f}}) \texttt{*}
+      (\textcap{R}^{\textit{f}} \texttt{+} \textcap{S}^{\textit{t}}) & =
+        \textcap{R}^{\textit{f}} \texttt{+} \textcap{S}^{\textit{f}}
+        \;\texttt{+}\;
+        \texttt{p(dim=\textcap{dr})*p(dim=\textcap{ds})}
+    \end{align*}
 \end{description}
+
+\begin{thm}{(Soundness)}
+  Equational theory is sound with respect to the lattice model. Let $R$
+  and $S$ be $N$ dimensional regions, then we have
+%
+  \begin{equation*}
+    R \equiv S \implies \interp{R}_N \equiv \interp{S}_N
+  \end{equation*}
+\end{thm}
+
+\begin{proof}
+  We show soundness for each of the equivalences.
+  \begin{description}
+    \item[Basic] These follow from basic properties of lattices.
+%
+    \item[Dist] This is immediate from the definition of distributive
+      lattice which the model is one.
+%
+    \item[Overlapping pointed] Let $\interp{\textcap{n}}$ be $n$. Counterparts
+      of individual regions for some Boolean $p$ are given below:
+%
+      \begin{align*}
+        \interp{\texttt{\textcap{R}(depth=\textcap{k},dim=\textcap{n},\_)}}_N
+        & = \vecgen{N}{n}{\interv{-k}{k}{p}} \\
+%
+        \interp{\texttt{pointed(dim=\textcap{n})}}_N
+        &= \vecgen{N}{n}{\interv{0}{0}{\textsf{true}}} \\
+%
+        \interp{\texttt{\textcap{R}(depth=\textcap{k},dim=\textcap{n})}}_N
+        &= \vecgen{N}{n}{\interv{-k}{k}{\textsf{true}}}
+      \end{align*}
+
+      Using \Cref{lem:vector-union}, it is sufficient to show
+      $ \interv{-k}{k}{p} \;\cup\; \interv{0}{0}{\textsf{true}} =
+        \interv{-k}{k}{\textsf{true}} $ holds. This immediately follows from
+      \Cref{lem:zinf-identities}.
+%
+    \item[Centered] Let $\interp{\textcap{n}}$, $\interp{\textcap{k}}$,
+      $\interp{\textcap{p1}}$, $\interp{\textcap{p2}}$, and
+      $\interp{\textcap{p3}}$ be $n$, $k$, $p1$, $p2$, and $p3$ respectively. We
+      also know that $p1$ is $p2 \vee p3$ from \textbf{Centered} detinition.
+      Then we get the following interpretation for region constants involved in
+      the equivalence:
+%
+      \begin{align*}
+        \interp{\texttt{centered(depth=\textcap{k},dim=\textit{n},\textcap{p1})}}_N
+          &= \vecgen{N}{n}{\interv{-k}{k}{p2 \vee p3}} \\
+%
+        \interp{\texttt{forward(depth=\textcap{k},dim=\textit{n},\textcap{p2})}}_N
+          &= \vecgen{N}{n}{\interv{0}{k}{p2} }\\
+%
+        \interp{\texttt{backward(depth=\textcap{k},dim=\textit{n},\textcap{p3})}}_N
+          &= \vecgen{N}{n}{\interv{0}{k}{p3}}
+      \end{align*}
+%
+      Using \Cref{lem:vector-union}, it is sufficient to show $\interv{-k}{k}{p2
+      \vee p3} = \interv{0}{k}{p2} \cup \interv{-k}{0}{p3}$ holds. This
+      immediately follows from \Cref{lem:zinf-identities}.
+
+    \item[Point] Let $\interp{\textcap{R}^{\textnormal{\textit{p}}}}_N$ and
+      $\interp{\textcap{S}^{\textnormal{\textit{p'}}}}_N$ be
+      $\vecgen{N}{k}{\interv{a}{b}{p}}$ and $\vecgen{N}{l}{\interv{c}{d}{p'}}$
+      respectively for some $a$ and $c$ less than or equal to 0, for some $c$
+      and $d$ greater than or equal to 0, for some indices $k$ and $l$ be
+      $\interp{\textcap{dr}}$ and $\interp{\textcap{ds}}$, which are not
+      necessarily distinct and $p$ and $p'$ are \textsf{true} if pointed and
+      \textsf{false} otherwise.
+%
+      \begin{align*}
+        \hspace{2em&\hspace{-2em}}
+        \interp
+          {{(\textcap{R}^{\textit{t}} * \textcap{S}^{\textit{f}})}
+           \;\texttt{+}\;
+           {(\textcap{R}^{\textit{f}} * \textcap{S}^{\textit{t}})}
+           \;\texttt{+}\;
+           \texttt{pointed(dim=\textcap{dr})*pointed(dim=\textcap{ds})}}_N \\
+        & = {(\interp{\textcap{R}^{\textit{t}}}_N \cap
+              \interp{\textcap{S}^{\textit{f}}}_N)} \cup
+            {(\interp{\textcap{R}^{\textit{f}}}_N \cap
+              \interp{\textcap{S}^{\textit{t}}}_N)} \;\cup \\
+        & \quad\,\, {(\vecgen{N}{\interp{ds}}{\interv{0}{0}{\textsf{true}}}
+                      \;\cup\;
+                      \vecgen{N}{\interp{dr}}{\interv{0}{0}{\textsf{true}}})} \\
+        & = \{ x \mid
+            \bigwedge_{\substack{1 \leq i \leq N \\ i \neq k \neq l}} x_i \in
+              \interv{\infty}{\infty}{\textsf{true}}
+            \wedge a \leq x_k \leq b
+            \wedge {(c \leq x_l \leq d \wedge x_l \neq 0)} \} \;\cup \\
+        & \quad\,\, \{ x \mid
+            \bigwedge_{\substack{1 \leq i \leq N \\ i \neq k \neq l}} x_i \in
+              \interv{\infty}{\infty}{\textsf{true}}
+            \wedge {(a \leq x_k \leq b \wedge x_k \neq 0)}
+            \wedge c \leq x_l \leq d \} \;\cup \\
+        & \quad\,\, \{ x \mid
+            \bigwedge_{\substack{1 \leq i \leq N \\ i \neq k \neq l}} x_i \in
+              \interv{\infty}{\infty}{\textsf{true}}
+            \wedge (x_k = 0 \wedge x_l = 0) \} \\
+        & = \{ x \mid
+            \bigwedge_{\substack{1 \leq i \leq N \\ i \neq k \neq l}} x_i \in
+              \interv{\infty}{\infty}{\textsf{true}}
+            \wedge a \leq x_k \leq b
+          \wedge c \leq x_l \leq d \wedge {(x_k \neq 0 \vee x_l \neq 0)} \}
+          \;\cup \\
+        & \quad\,\, \{ x \mid
+            \bigwedge_{\substack{1 \leq i \leq N \\ i \neq k \neq l}} x_i \in
+              \interv{\infty}{\infty}{\textsf{true}}
+            \wedge (x_k = 0 \wedge x_l = 0) \} \\
+        & = \{ x \mid
+            \bigwedge_{\substack{1 \leq i \leq N \\ i \neq k \neq l}} x_i \in
+              \interv{\infty}{\infty}{\textsf{true}}
+            \wedge a \leq x_k \leq b
+            \wedge c \leq x_l \leq d \} \\
+        & = \textcap{R}^{\textit{t}} \texttt{*} \textcap{S}^{\textit{t}}
+      \end{align*}
+  \end{description}
+\end{proof}
 

### 2017-03-25 oopsla17/lattice-model.tex
forgotten part of the proof

@@
 
+We define an equivalence relation, $\equiv$, on regions as follows:
+
 \begin{description}
@@
 \begin{proof}
-  We show soundness for each of the equivalences.
+  Both $\equiv$ and $=$ are equivalence relations. We only need to show that
+  each property associated with specifications equivalence holds as an equality
+  in the model.
+%
   \begin{description}
@@
 
-    \item[Point] Let $\interp{\textcap{R}^{\textnormal{\textit{p}}}}_N$ and
+    \item[Point] We will prove the first equivalence and then use that to prove
+      the second.
+%
+      Let $\interp{\textcap{R}^{\textnormal{\textit{p}}}}_N$ and
       $\interp{\textcap{S}^{\textnormal{\textit{p'}}}}_N$ be
@@
             \wedge c \leq x_l \leq d \} \\
-        & = \textcap{R}^{\textit{t}} \texttt{*} \textcap{S}^{\textit{t}}
+        & = \interp{\textcap{R}^{\textit{t}} \texttt{*} \textcap{S}^{\textit{t}}}_N
+      \end{align*}
+%
+      Now we prove the second equivalence:
+      \begin{align*}
+        \hspace{2em}&\hspace{-2em} \interp{
+          (\textcap{R}^{\textit{t}} \texttt{+} \textcap{S}^{\textit{f}})
+          \texttt{*}
+          (\textcap{R}^{\textit{f}} \texttt{+} \textcap{S}^{\textit{t}})
+        }_N \\
+        &= \interp{
+             (\textcap{R}^{\textit{t}} \texttt{*} \textcap{R}^{\textit{f}})
+             \texttt{+}
+             (\textcap{S}^{\textit{t}} \texttt{*} \textcap{S}^{\textit{f}})
+             \texttt{+}
+             (\textcap{R}^{\textit{t}} \texttt{*} \textcap{S}^{\textit{t}})
+             \texttt{+}
+             (\textcap{R}^{\textit{f}} \texttt{*} \textcap{S}^{\textit{f}})
+           }_N \\
+        &= \interp{
+             \textcap{R}^{\textit{f}}
+             \texttt{+}
+             \textcap{S}^{\textit{f}}
+             \texttt{+}
+             (\textcap{R}^{\textit{t}} \texttt{*} \textcap{S}^{\textit{t}})
+           }_N \\
+        &= \interp
+             {\textcap{R}^{\textit{f}}
+              \texttt{+}
+              \textcap{S}^{\textit{f}}
+              \texttt{+}
+              {(\textcap{R}^{\textit{t}} \texttt{*} \textcap{S}^{\textit{f}})}
+               \;\texttt{+}\;
+               {(\textcap{R}^{\textit{f}} \texttt{*} \textcap{S}^{\textit{t}})}
+               \;\texttt{+}\;
+               \texttt{pointed(dim=\textcap{dr})*pointed(dim=\textcap{ds})}}_N \\
+        &= \interp{\textcap{R}^{\textit{f}} \texttt{+} \textcap{S}^{\textit{f}}
+                   \;\texttt{+}\;
+                   \texttt{pointed(dim=\textcap{dr})*pointed(dim=\textcap{ds})}}_N
       \end{align*}

### 2017-03-26 oopsla17/lattice-model.tex
Add few macros, add some goals for inference procedure

@@
 
+% Essential packages
 \usepackage{amsmath}
@@
 \usepackage{cleveref}
+\usepackage{xifthen}
 
+% Non-essential
+\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
+
+% Essential declarations
 \theoremstyle{definition}
@@
 
+% Auxillary declaration
 \newcommand{\eg}{\emph{e.g.}}
 
+% Essential macros
 \newcommand{\zinf}{\textnormal{$\mathbb{Z}_\infty$}}
-\newcommand{\interp}[1]{\llbracket#1\rrbracket}
+\newcommand{\interp}[1]{\llbracket {#1} \rrbracket}
 \newcommand{\interv}[3]{\textnormal{$\lbrack#1,#2\rbrack^{#3}$}}
@@
 \newcommand{\textcap}[1]{\texttt{\textit{#1}}}
+
+% Stencil macro
+%
+% Optional argument: s when we want the short form, empty otherwise.
+%
+% Mandatory arguments:
+% 1st: c/b/f/p for centered, backward, forward, pointed
+% 2st: dimension
+% 3st: depth
+% 4st: if pointed/nonpointed needs to be specified leave it blank if omitted.
+%
+% Example: \stencil[s]{f}{1}{2}{} = fwd(dim=1,depth=2)
+% Example: \stencil{b}{1}{2}{pointed} = backward(dim=1,depth=2,pointed)
+\newcommand{\stencil}[5][]{%
+  \ifthenelse{\equal{#2}{f}}{%
+    \ifthenelse{\equal{#1}{s}}{%
+      \texttt{fwd}%
+    }{\ifthenelse{\isempty{#1}}{%
+        \texttt{forward}%
+      }{INVALID OPTIONAL}%
+    }%
+  }{\ifthenelse{\equal{#2}{b}}{%
+      \ifthenelse{\equal{#1}{s}}{%
+          \texttt{bwd}%
+      }{\ifthenelse{\isempty{#1}}{%
+          \texttt{backward}%
+        }{INVALID OPTIONAL}%
+      }%
+  }{\ifthenelse{\equal{#2}{c}}{%
+      \ifthenelse{\equal{#1}{s}}{%
+          \texttt{ctd}%
+      }{\ifthenelse{\isempty{#1}}{%
+          \texttt{centered}%
+        }{INVALID OPTIONAL}%
+      }%
+  }{\ifthenelse{\equal{#2}{p}}{%
+      \ifthenelse{\equal{#1}{s}}{%
+          \texttt{ptd}%
+      }{\ifthenelse{\isempty{#1}}{%
+          \texttt{pointed}%
+        }{INVALID OPTIONAL}%
+      }%
+  }{}}}}%
+  \texttt{(dim=\textnormal{#3}%
+    \ifthenelse{\isempty{#4}}{}{,depth=\textnormal{#4}}%
+    \ifthenelse{\isempty{#5}}{}{,\textnormal{#5}}%
+  )}%
+}
+
+% Non-essential macro
+\definecolor{unused}{RGB}{211,167,192}
+\newcommand{\unused}[1]{{\color{unused} #1}}
 
@@
 
-\begin{lem}{}
+\begin{lem}{}\label{lem:zinf-identities}
   We have the following dual identities for \zinf{} intervals:
@@
       \interv{\min \{a,d\}}{\max \{b,e\}}{c \vee f}
-  \end{align*}\label{lem:zinf-identities}
+  \end{align*}
 \end{lem}
@@
 
-\begin{lem}{}
+\unused{
+\begin{lem}{}\label{lem:vector-intersect}
   Intersection distributes over offsets vectors.
@@
      \textit{vint}_N \;\cap\; \textit{vint'}_N)
-  \end{equation*}\label{lem:vector-intersect}
+  \end{equation*}
 \end{lem}
+%
+\begin{proof}
+  TODO. Trivial. Do if needed.
+\end{proof}
+}
 
-\begin{lem}{}
+\begin{lem}{}\label{lem:vector-union}
   Let $v$ and $v'$ be two $N$ dimensional offsets vectors such that $v_i$ is
@@
     (v_1,\dotsc, v_k \cup v'_k,\dotsc, v_N)
-  \end{equation*}\label{lem:vector-union}
+  \end{equation*}
 \end{lem}
@@
 
+\unused{
   \begin{lem}{}
@@
   \end{proof}
+}
 
@@
 Typographically, cursive font is used within specifications to indicate a
-partial textual capture.
+partial textual capture. For convenience we use \textit{italic} serif font
+inside specifications rather than interpreting its textual capture.  These
+coalesces syntactic and semantic representation but avoids clutter in the
+presentation.
 
@@
 \begin{align*}
-  \interp{\texttt{pointed(dim=\textcap{i})}}_N & =
-    \vecgen{N}{\interp{\textcap{i}}}{\interv{0}{0}{\textsf{true}}}\\
+  \interp{\stencil{p}{$n$}{}{}}_N & =
+    \vecgen{N}{\interp{i}}{\interv{0}{0}{\textsf{true}}}\\
 %
-  \interp{\texttt{centered(dim=\textcap{i}, depth=\textcap{k}, \textcap{p})}}_N
-    & = \vecgen{N}
-           {\interp{\textcap{i}}}
-           {\interv{-\interp{\textcap{k}}}
-                   {\interp{\textcap{k}}}
-                   {\interp{\textcap{p}}}
-           } \\
+  \interp{\stencil{c}{$i$}{$k$}{\textcap{p}}}_N & =
+    \vecgen{N}{i}{\interv{-k}{k}{\interp{\textcap{p}}}} \\
 %
-  \interp{\texttt{forward(dim=\textcap{i}, depth=\textcap{k}, \textcap{p})}}_N
-    & = \vecgen{N}
-           {\interp{\textcap{i}}}
-           {\interv{0}
-                   {\interp{\textcap{k}}}
-                   {\interp{\textcap{p}}}
-           } \\
+  \interp{\stencil{f}{$i$}{$k$}{\textcap{p}}}_N & =
+    \vecgen{N}{\textcap{i}}{\interv{0}{k}{\interp{\textcap{p}}}} \\
 %
-  \interp{\texttt{backward(dim=\textcap{i}, depth=\textcap{k}, \textcap{p})}}_N
-    & = \vecgen{N}
-           {\interp{\textcap{i}}}
-           {\interv{-\interp{\textcap{k}}}
-                   {0}
-                   {\interp{\textcap{p}}}
-           }
+  \interp{\stencil{b}{$i$}{$k$}{\textcap{p}}}_N & =
+  \vecgen{N}{\textcap{i}}{\interv{-k}{0}{\interp{\textcap{p}}}}
 \end{align*}
@@
 We know introduce the equational and approximation theory of stencil
-specifications based on $\equiv$ and $\prec$.
+specifications based on $\equiv$ and $\preceq$.
 
@@
     \texttt{backward}, or \texttt{centered}, then we have
-    $\texttt{\textcap{R}(depth=\textcap{k},dim=\textcap{n},\_) +
-    pointed(dim=\textcap{n})} \equiv
-    \texttt{\textcap{R}(depth=\textcap{k},dim=\textcap{n})}$.
+    $\texttt{\textcap{R}(depth=$k$,dim=$n$,\_) + \stencil[s]{p}{$n$}{}{}} \equiv
+    \texttt{\textcap{R}(depth=$k$,dim=$n$)}$.
 %
@@
 %
-    \begin{gather*}
-      \texttt{centered(depth=\textcap{k},dim=\textit{n},\textcap{p1})} \\
-      \equiv \\
-      \texttt{forward(depth=\textcap{k},dim=\textit{n},\textcap{p2}) +
-        backward(depth=\textcap{k},dim=\textit{n},\textcap{p3})}
-    \end{gather*}
+    \begin{equation*}
+      \stencil[s]{c}{$n$}{$k$}{\textcap{p1}} \equiv
+        \stencil[s]{f}{$n$}{$k$}{\textcap{p2}} \texttt{+}
+        \stencil[s]{f}{$n$}{$k$}{\textcap{p3}}
+    \end{equation*}
 %
@@
 \subsection{Approximations}
+We define a partial of order of approximations, $\preceq$. This relation
+provides us with the following:
+
+\begin{description}
+  \item[Centered]
+  \item[Point] Let \textcap{R} be a region constant with two variants: a
+    superscript $f$ to mean nonpointed and $t$ to mean pointed. Then we have
+    $\textcap{R}^t \preceq \textcap{R}^f$.
+\end{description}
 
@@
 
+\begin{thm}{(Completeness of inference)}
+  Inference procedure can infer a specification for every stencil computation.
+\end{thm}
+%
+\begin{proof}
+  Would be nice to have.
+\end{proof}
+
+\begin{thm}{(Completeness of exact inference)}
+  Inference procedure can infer a specification for every stencil computation
+  with array indexing behaviour forms a superposition of hyperrectangles removed
+  at most one unit from the origin in any of its dimension.
+\end{thm}
+%
+\begin{proof}
+  Would be nice to have.
+\end{proof}
+
+\begin{thm}{(Optimality of inference)}
+  For exact inference, inference procedure always infers specifications with
+  fewest number of region constants.
+\end{thm}
+%
+\begin{proof}
+  Would be nice to have.
+\end{proof}
+
 \end{document}

### 2017-03-26 oopsla17/lattice-model.tex
Add remark on general optimality

@@
 \newtheorem{case}{Case}
+
+\theoremstyle{remark}
+\newtheorem{remark}{Remark}
 
@@
 
-\begin{thm}{(Completeness of inference)}
+\begin{thm}{(Completeness of inference)}\label{thm:inf-completeness}
   Inference procedure can infer a specification for every stencil computation.
@@
 
-\begin{thm}{(Optimality of inference)}
+\begin{thm}{(Optimality of inference)}\label{thm:inf-optimality}
   For exact inference, inference procedure always infers specifications with
@@
 
+\begin{remark}{}
+  Optimality of general inference is trivial. We only need a single region
+  constant covering accesses in any one of the stencil's dimensions and that
+  would provide an upper bound for the stencil.
+\end{remark}
+
 \end{document}

### 2017-03-27 oopsla17/lattice-model.tex
theory of approximation

@@
       }%
-  }{}}}}%
+  }{\ifthenelse{\isempty{#1}}{%
+      \textcap{#2}%
+    }{NO OPTIONAL FOR CUSTOM}%
+  }}}}%
   \texttt{(dim=\textnormal{#3}%
@@
 \newcommand{\unused}[1]{{\color{unused} #1}}
+
+\definecolor{version}{RGB}{163,193,173}
+\newcommand{\longv}[1]{{\color{version} Long version:} #1}
+\newcommand{\shortv}[1]{{\color{version} Short version:} #1}
 
@@
 %
+  \item[Subsumption] If $S$ and $R$ are regions with $S \preceq R$, then
+    $S \texttt{+} R \equiv S$ and $S \texttt{*} R \equiv R$.
+%
   \item[Dist] \texttt{*} distributed over \texttt{+} and dually
@@
     \texttt{backward}, or \texttt{centered}, then we have
-    $\texttt{\textcap{R}(depth=$k$,dim=$n$,\_) + \stencil[s]{p}{$n$}{}{}} \equiv
-    \texttt{\textcap{R}(depth=$k$,dim=$n$)}$.
+    $\stencil{\textcap{R}}{$n$}{$k$}{\texttt{nonpointed}} \;\texttt{+}\;
+     \stencil[s]{p}{$n$}{}{} \equiv
+     \stencil{\textcap{R}}{$n$}{$k$}{}$.
 %
@@
         \stencil[s]{f}{$n$}{$k$}{\textcap{p2}} \texttt{+}
-        \stencil[s]{f}{$n$}{$k$}{\textcap{p3}}
+        \stencil[s]{b}{$n$}{$k$}{\textcap{p3}}
     \end{equation*}
@@
       \;\texttt{+}\;
-      \texttt{p(dim=\textcap{dr})*p(dim=\textcap{ds})} & =
+      \stencil[s]{p}{\textit{dr}}{}{}*\stencil[s]{p}{\textit{ds}}{}{} & =
         \textcap{R}^{\textit{t}} \texttt{*} \textcap{S}^{\textit{t}} \\
@@
         \;\texttt{+}\;
-        \texttt{p(dim=\textcap{dr})*p(dim=\textcap{ds})}
+        \stencil[s]{p}{\textit{dr}}{}{}*\stencil[s]{p}{\textit{ds}}{}{}
     \end{align*}
@@
 
-\begin{thm}{(Soundness)}
+\begin{thm}{(Equational soundness)}
   Equational theory is sound with respect to the lattice model. Let $R$
@@
   \begin{description}
-    \item[Basic] These follow from basic properties of lattices.
+    \item[Basic \& Subsumption] These follow from basic properties of lattices.
 %
@@
 %
-    \item[Overlapping pointed] Let $\interp{\textcap{n}}$ be $n$. Counterparts
-      of individual regions for some Boolean $p$ are given below:
+    \item[Overlapping pointed] Counterparts of individual regions for some
+      Boolean $p$ are given below:
 %
       \begin{align*}
-        \interp{\texttt{\textcap{R}(depth=\textcap{k},dim=\textcap{n},\_)}}_N
-        & = \vecgen{N}{n}{\interv{-k}{k}{p}} \\
+        \interp{\stencil{\textcap{R}}{$n$}{$k$}{\texttt{nonpointed}}}_N
+        & = \vecgen{N}{n}{\interv{-k}{k}{\textsf{false}}} \\
 %
-        \interp{\texttt{pointed(dim=\textcap{n})}}_N
+        \interp{\stencil{p}{$n$}{}{}}_N
         &= \vecgen{N}{n}{\interv{0}{0}{\textsf{true}}} \\
 %
-        \interp{\texttt{\textcap{R}(depth=\textcap{k},dim=\textcap{n})}}_N
+        \interp{\stencil{\textcap{R}}{$n$}{$k$}{}}_N
         &= \vecgen{N}{n}{\interv{-k}{k}{\textsf{true}}}
@@
 %
-    \item[Centered] Let $\interp{\textcap{n}}$, $\interp{\textcap{k}}$,
-      $\interp{\textcap{p1}}$, $\interp{\textcap{p2}}$, and
-      $\interp{\textcap{p3}}$ be $n$, $k$, $p1$, $p2$, and $p3$ respectively. We
+    \item[Centered] Let $\interp{\textcap{p1}}$, $\interp{\textcap{p2}}$, and
+      $\interp{\textcap{p3}}$ be $p1$, $p2$, and $p3$ respectively. We
       also know that $p1$ is $p2 \vee p3$ from \textbf{Centered} detinition.
@@
       \begin{align*}
-        \interp{\texttt{centered(depth=\textcap{k},dim=\textit{n},\textcap{p1})}}_N
+        \interp{\stencil{c}{$k$}{$n$}{$p1$}}_N
           &= \vecgen{N}{n}{\interv{-k}{k}{p2 \vee p3}} \\
 %
-        \interp{\texttt{forward(depth=\textcap{k},dim=\textit{n},\textcap{p2})}}_N
+        \interp{\stencil{f}{$k$}{$n$}{$p2$}}_N
           &= \vecgen{N}{n}{\interv{0}{k}{p2} }\\
 %
-        \interp{\texttt{backward(depth=\textcap{k},dim=\textit{n},\textcap{p3})}}_N
+        \interp{\stencil{b}{$k$}{$n$}{$p3$}}_N
           &= \vecgen{N}{n}{\interv{0}{k}{p3}}
@@
 \begin{description}
-  \item[Centered]
-  \item[Point] Let \textcap{R} be a region constant with two variants: a
-    superscript $f$ to mean nonpointed and $t$ to mean pointed. Then we have
-    $\textcap{R}^t \preceq \textcap{R}^f$.
+  \item[Equivalence] If $S$ and $R$ are regions and $S \equiv R$, then we have
+    $S \preceq R$.
+%
+  \item[Combined] If $S$ and $R$ are regions, then we have
+    $S \texttt{+} R \preceq S$ and $S \preceq S \texttt{*} R$.
+%
+  \item[Depth] Let $k$ and $l$ be in positive integers and $k \leq l$, $n$ some
+    fixed dimension, and \textcap{p} either \texttt{pointed} or
+    \texttt{nonpointed}. Further let \textcap{R} be one of \texttt{centered},
+    \texttt{forward}, and \texttt{backward}. We then have
+%
+    \begin{equation*}
+      \stencil{R}{$n$}{$l$}{\textcap{p}} \preceq \stencil{R}{$n$}{$k$}{\textcap{p}}
+    \end{equation*}
+%
 \end{description}
+
+\begin{thm}{(Approximation soundness)}
+  Theory of approximation is sound with respect to the lattice model. Let $R$
+  and $S$ be $N$ dimensional regions, then we have
+%
+  \begin{equation*}
+    R \preceq S \implies \interp{R}_N \supseteq \interp{S}_N
+  \end{equation*}
+\end{thm}
+%
+\begin{proof}
+  Both $\preceq$ and $\supseteq$ are partial orders on sets. We only need to
+  show that each inequality holds in the model.
+%
+  \begin{description}
+    \item[Equivalence] Region equavilance is modeled with set equality and the
+      partial order on regions is modeled with superset relation. Partial orders
+      are reflexive and the model of eqivalence agrees with that of the partial
+      order, so this rule holds.
+%
+    \item[Combined] \texttt{*} and \texttt{+} maps to join and meet operations
+      in the lattice. All joins and meets in a lattice satisfy these
+      inequalities.
+
+    \item[Depth] When $m$ is $k$ and $l$ in
+      $\stencil{\textcap{R}}{$n$}{$m$}{\textcap{p}}$, we obtain the
+      interpretations $\vecgen{N}{n}{\textit{int}_k}$ and
+      $\vecgen{N}{n}{\textit{int}_l}$ for some $N$ as interpretations of
+      regions. Call these \textit{ks} and \textit{ls} respectively. One way to
+      show $\textit{ls} \supseteq \textit{ks}$ is to show that the intersection
+      of \textit{ls} and \textit{ks} is \textit{ks}.
+
+      To show that the intersection is \textit{ks}, we use
+      \Cref{lem:vector-intersect} which reduces the problem to showing that the
+      intersection of $\textit{int}_k$ and $\textit{int}_l$ is the former since
+      at every point except $n$ the intervals agree, hence the intersection
+      remains the same. By assumption we have $k \leq l$, so the maximum of the
+      lower bounds is $-k$ and the minimum of lower bounds is $k$ this is same
+      as interval $\textit{int}_k$. Note that pointedness is the same for both
+      intervals and depths are positive integers, so the pointedness is
+      preserved under intersection. This shows the intersection of \textit{ls}
+      and \textit{ks} is indeed \textit{ks} and concludes the proof.
+  \end{description}
+\end{proof}
+
+We present few inequalities that can be derived from the axioms and are useful
+when writing specifications without inference:
+
+\begin{prop}{(Centered approximation)}
+  For any dimension $n$, depth $k$, and pointed attribute $p$,
+  we have
+%
+  \begin{align*}
+    \stencil{c}{$n$}{$k$}{\textcap{p}} & \preceq
+      \stencil{f}{$n$}{$k$}{\textcap{p}} \\
+%
+    \stencil{c}{$n$}{$k$}{\textcap{p}} & \preceq
+      \stencil{b}{$n$}{$k$}{\textcap{p}}
+  \end{align*}
+\end{prop}
+%
+\begin{proof}
+  \shortv{%
+    Both inequalities are proved by breaking \texttt{centered} using
+    (\textbf{Centered}) followed by the inequality (\textbf{Combined}).
+  }
+
+  \longv{%
+    Proof of the first inequality:
+    \begin{align*}
+      \stencil[s]{c}{$n$}{$k$}{\textcap{p}}
+        & \equiv \stencil[s]{f}{$n$}{$k$}{\textcap{p}}
+                 \;\texttt{+}\;
+                 \stencil[s]{b}{$n$}{$k$}{\textcap{p}}
+          & (\textbf{Centered}) \\
+        & \preceq \stencil[s]{f}{$n$}{$k$}{\textcap{p}}
+          & (\textbf{Combined})
+    \end{align*}
+
+    Proof of the second inequality:
+    \begin{align*}
+      \stencil[s]{c}{$n$}{$k$}{\textcap{p}}
+        & \equiv \stencil[s]{f}{$n$}{$k$}{\textcap{p}}
+                 \;\texttt{+}\;
+                 \stencil[s]{b}{$n$}{$k$}{\textcap{p}}
+          & (\textbf{Centered}) \\
+        & \preceq \stencil[s]{b}{$n$}{$k$}{\textcap{p}}
+          & (\textbf{Combined})
+    \end{align*}
+  }
+\end{proof}
+
+\begin{prop}{(Point approximation)}
+  Let \textcap{R} be one of \texttt{forward}, \texttt{backward}, and
+  \texttt{centered}, $n$ a fixed dimension, and $k$ a fixed depth, then we have
+%
+  \begin{align*}
+    \stencil{R}{$n$}{$k$}{} & \preceq \stencil{p}{$n$}{}{} \\
+%
+    \stencil{R}{$n$}{$k$}{} & \preceq
+      \stencil{R}{$n$}{$k$}{\texttt{nonpointed}}
+  \end{align*}
+\end{prop}
+%
+\begin{proof}
+  \shortv{%
+    Both inequalities are proved by breaking \textcap{R} using
+    (\textbf{Overlapping pointed}) followed by the inequality
+    (\textbf{Combined}).
+  }
+
+  \longv{%
+    Proof of the first inequality:
+    \begin{align*}
+      \stencil{R}{$n$}{$k$}{}
+        & \equiv \stencil{R}{$n$}{$k$}{\texttt{nonpointed}}
+                 \;\texttt{+}\;
+                 \stencil[s]{p}{$n$}{}{}
+        & (\textbf{Overlapping pointed}) \\
+      & \preceq \stencil[s]{p}{$n$}{}{} & (\textbf{Combined})
+    \end{align*}
+
+    Proof of the second inequality:
+    \begin{align*}
+      \stencil{R}{$n$}{$k$}{}
+        & \equiv \stencil{R}{$n$}{$k$}{\texttt{nonpointed}}
+                 \;\texttt{+}\;
+                 \stencil[s]{p}{$n$}{}{}
+        & (\textbf{Overlapping pointed}) \\
+      & \preceq \stencil{R}{$n$}{$k$}{\texttt{nonpointed}}
+        & (\textbf{Combined})
+    \end{align*}
+  }
+\end{proof}
 

### 2017-03-27 oopsla17/lattice-model.tex
Add lemma proof + stylistic fixes
Modify region space to region directly

@@
 \newcommand{\zinf}{\textnormal{$\mathbb{Z}_\infty$}}
-\newcommand{\interp}[1]{\llbracket {#1} \rrbracket}
+\newcommand{\interp}[1]{\llbracket{} {#1} \rrbracket{}}
 \newcommand{\interv}[3]{\textnormal{$\lbrack#1,#2\rbrack^{#3}$}}
@@
     \ifthenelse{\isempty{#4}}{}{,depth=\textnormal{#4}}%
-    \ifthenelse{\isempty{#5}}{}{,\textnormal{#5}}%
-  )}%
+    \ifthenelse{\isempty{#5}}{}{,\textnormal{#5}})}%
 }
@@
 
-\unused{
 \begin{lem}{}\label{lem:vector-intersect}
@@
 \begin{proof}
-  TODO. Trivial. Do if needed.
+  \begin{align*}
+    v \; \cap \; v' &
+    = \{x \mid \bigwedge_{1 \leq i \leq N } x_i \in v_i \}
+      \;\cap\;
+      \{x \mid \bigwedge_{1 \leq i \leq N } x_i \in v'_i \} \\
+    & = \{x \mid \bigwedge_{1 \leq i \leq N } (x_i \in v_i \wedge x_i \in v'_i) \} \\
+    & = (v_1 \cap v'_1,\dotsc,v_N \cap v'_N)
+  \end{align*}
 \end{proof}
-}
 
@@
     v \; \cup \; v' &
-    = \{(x_1,\dotsc,x_n) \mid
+    = \{x \mid
           \bigwedge_{1 \leq i \leq N } x_i \in v_i \}
       \;\cup\;
-      \{(x_1,\dotsc,x_n) \mid
+      \{x \mid
           \bigwedge_{1 \leq i \leq N } x_i \in v'_i \} \\
-    & = \{(x_1,\dotsc,x_n) \mid
+    & = \{x \mid
           \bigwedge_{\substack{1 \leq i \leq N \\ i \neq k}} x_i \in v_i
@@
         \} \\
-    & = \{(x_1,\dotsc,x_n) \mid
+    & = \{x \mid
           \bigwedge_{\substack{1 \leq i \leq N \\ i \neq k}} x_i \in v_i
@@
 
-\unused{
-  \begin{lem}{}
-    Intersection distributes $N$ dimensional vector intervals:
-  %
-    \begin{gather*}
-      (\interv{a_1}{b_1}{c_1}, \dots, \interv{a_N}{b_N}{c_N}) \; \cap \;
-        (\interv{a'_1}{b'_1}{c'_1}, \dots, \interv{a'_N}{b'_N}{c'_N}) \\
-      = \\
-      (\interv{\max \{a_1,a'_1\}}{\min \{b_1,b'_1\}}{\neg c_1 \vee \neg c'_1},
-        \dots, \interv{\max \{a_N,a'_N\}}{\min \{b_N,b'_N\}}{\neg c_N \vee \neg c'_N})
-    \end{gather*}
-  \end{lem}
-  %
-  \begin{proof}
-    This immediately follows from \Cref{lem:vector-intersect} and
-    \Cref{lem:zinf-identities}.
-  \end{proof}
-}
-
-\begin{defn}{(\emph{Region space})}
-  Region space defines the set \textit{Region} in $N$ dimensions inductively as
-  follows:
+\begin{defn}{(\emph{Region})}
+  \textit{Region} is the set of $N$ dimensional vector intervals defined as the
+  smallest set satisfying the following:
 %
@@
       \begin{align*}
-        \hspace{2em}&\hspace{-2em} \interp{
+        \hspace{2em}&\hspace{-2em} \interp{%
             (\textcap{R}^{\textit{t}} \texttt{+} \textcap{S}^{\textit{f}})
@@
           }_N \\
-        &= \interp{
+        &= \interp{%
              (\textcap{R}^{\textit{t}} \texttt{*} \textcap{R}^{\textit{f}})
@@
            }_N \\
-        &= \interp{
+        &= \interp{%
              \textcap{R}^{\textit{f}}
@@
       at every point except $n$ the intervals agree, hence the intersection
-      remains the same. By assumption we have $k \leq l$, so the maximum of the
+      remains the same. Then we use \Cref{lem:zinf-identities} to calculate the
+      intersection. By assumption we have $k \leq l$, so the maximum of the
       lower bounds is $-k$ and the minimum of lower bounds is $k$ this is same

### 2017-03-27 oopsla17/lattice-model.tex
Flip the lattice

@@
 \begin{prop}{}
-  $(\textit{Region},\supseteq)$ is a distributive lattice.
+  $(\textit{Region},\subseteq)$ is a distributive lattice.
 \end{prop}
@@
   Any sublattice of a distributive lattice is a distributive lattice. Power sets
-  coupled with partially ordered with superset or equals relation are
-  distributive lattices. \textit{Region} is a subset of
-  $\mathcal{P}{(\zinf{}^{\mkern-19mu{N}})}$ since all vector intervals are
-  subsets of this power set and union and intersection cannot add an element
-  that is not already in the powerset. Therefore, \textit{Region} is a
-  distributive sublattice of $\mathcal{P}{(\zinf{}^{\mkern-19mu{N}})}$.
+  partially ordered with subset or equal to relation are distributive lattices.
+  \textit{Region} is a subset of $\mathcal{P}{(\zinf{}^{\mkern-19mu{N}})}$ since
+  all vector intervals are subsets of this power set and union and intersection
+  cannot add an element that is not already in the powerset. Inductive
+  definition also ensures that \textit{Region} is closed under these operations.
+  Therefore, \textit{Region} is a distributive sublattice of
+  $(\mathcal{P}{(\zinf{}^{\mkern-19mu{N}})}, \subseteq)$.
 \end{proof}
@@
 We give some basic interpretations: $\interp{\texttt{n}} = n$ where $n$ is in
-$\mathbb{Z}$, $\interp{\texttt{pointed}} = \llbracket \epsilon \rrbracket =
+$\mathbb{Z}^+$, $\interp{\texttt{pointed}} = \llbracket \epsilon \rrbracket =
 \textsf{true}$, $\interp{\texttt{nonpointed}} = \textsf{false}$.
-Typographically, cursive font is used within specifications to indicate a
-partial textual capture. For convenience we use \textit{italic} serif font
-inside specifications rather than interpreting its textual capture.  These
+Typographically, cursive monotype font is used within specifications to indicate
+a partial textual capture. For convenience we use \textit{italic} serif font
+inside specifications rather than interpreting its textual capture. This
 coalesces syntactic and semantic representation but avoids clutter in the
@@
 
-Let $\textit{vec-gen}_N : \mathbb{N} \times \textit{Interval} \to
+Let $\textit{vec-gen}_N : \mathbb{N}^+ \times \textit{Interval} \to
 \textit{Interval}^N$ be the function generating an $N$ dimensional vector
 interval such that \vecgen{N}{i}{\interv{a}{b}{c}} has
-$\interv{-\infty}{\infty}{\textsf{p}}$ in all dimensions except the $i^{th}$
+$\interv{-\infty}{\infty}{\textsf{true}}$ in all dimensions except the $i^{th}$
 one. There, it has \interv{a}{b}{c}.
@@
 \begin{align*}
-  \interp{\stencil{p}{$n$}{}{}}_N & =
-    \vecgen{N}{\interp{i}}{\interv{0}{0}{\textsf{true}}}\\
+  \interp{\stencil{p}{$i$}{}{}}_N & =
+    \vecgen{N}{i}{\interv{0}{0}{\textsf{true}}}\\
 %
@@
   \interp{\stencil{f}{$i$}{$k$}{\textcap{p}}}_N & =
-    \vecgen{N}{\textcap{i}}{\interv{0}{k}{\interp{\textcap{p}}}} \\
+    \vecgen{N}{i}{\interv{0}{k}{\interp{\textcap{p}}}} \\
 %
   \interp{\stencil{b}{$i$}{$k$}{\textcap{p}}}_N & =
-  \vecgen{N}{\textcap{i}}{\interv{-k}{0}{\interp{\textcap{p}}}}
+  \vecgen{N}{i}{\interv{-k}{0}{\interp{\textcap{p}}}}
 \end{align*}
@@
 Second, we provide the semantics for the permissive region operator, \texttt{+},
-and the restrictive region operator, \texttt{*}, by associating them with meet
-and join respectively. Let \texttt{\textit{R}} and \texttt{\textit{S}} be
-regions.
+and the restrictive region operator, \texttt{*}, by associating them with join
+and meet respectively. Let \textcap{R} and \textcap{S} be regions.
 %
@@
   \interp{\texttt{\textcap{R} + \textcap{S}}}_N & =
-    \interp{\textcap{R}}_N \wedge \interp{\textcap{S}}_N
+    \interp{\textcap{R}}_N \vee \interp{\textcap{S}}_N
 &
   \interp{\texttt{\textcap{R} * \textcap{S}}}_N & =
-    \interp{\textcap{R}}_N \vee \interp{\textcap{S}}_N
+    \interp{\textcap{R}}_N \wedge \interp{\textcap{S}}_N
 \end{align*}
@@
 We know introduce the equational and approximation theory of stencil
-specifications based on $\equiv$ and $\preceq$.
+specifications based on $\equiv$ and $\preceq$ in the following two subsections.
 
@@
   \item[Subsumption] If $S$ and $R$ are regions with $S \preceq R$, then
-    $S \texttt{+} R \equiv S$ and $S \texttt{*} R \equiv R$.
+    $S \texttt{+} R \equiv R$ and $S \texttt{*} R \equiv S$.
 %
@@
     \texttt{+} distributes over \texttt{*}, meaning if \textcap{R}, \textcap{S},
-    and \textcap{T} are regions, then we have the following equivalences:
+    and \textcap{T} are regions, then we have the following dual equivalences:
 %
@@
   \begin{equation*}
-    R \equiv S \implies \interp{R}_N \equiv \interp{S}_N
+    R \equiv S \implies \interp{R}_N = \interp{S}_N
   \end{equation*}
@@
       $\interp{\textcap{p3}}$ be $p1$, $p2$, and $p3$ respectively. We
-      also know that $p1$ is $p2 \vee p3$ from \textbf{Centered} detinition.
-      Then we get the following interpretation for region constants involved in
-      the equivalence:
+      know that $p1$ is $p2 \vee p3$ from (\textbf{Centered}) detinition. Then
+      we get the following interpretation for region constants involved in the
+      equivalence:
 %
@@
   \item[Combined] If $S$ and $R$ are regions, then we have
-    $S \texttt{+} R \preceq S$ and $S \preceq S \texttt{*} R$.
+    $S \preceq S \texttt{+} R$ and $S \texttt{*} R \preceq S$.
 %
@@
     fixed dimension, and \textcap{p} either \texttt{pointed} or
-    \texttt{nonpointed}. Further let \textcap{R} be one of \texttt{centered},
+    \texttt{nonpointed}. Further, let \textcap{R} be one of \texttt{centered},
     \texttt{forward}, and \texttt{backward}. We then have
@@
     \begin{equation*}
-      \stencil{R}{$n$}{$l$}{\textcap{p}} \preceq \stencil{R}{$n$}{$k$}{\textcap{p}}
+      \stencil{R}{$n$}{$k$}{\textcap{p}} \preceq \stencil{R}{$n$}{$l$}{\textcap{p}}
     \end{equation*}
@@
   \begin{equation*}
-    R \preceq S \implies \interp{R}_N \supseteq \interp{S}_N
+    R \preceq S \implies \interp{R}_N \subseteq \interp{S}_N
   \end{equation*}
@@
 \begin{proof}
-  Both $\preceq$ and $\supseteq$ are partial orders on sets. We only need to
+  Both $\preceq$ and $\subseteq$ are partial orders on sets. We only need to
   show that each inequality holds in the model.
@@
     \item[Equivalence] Region equavilance is modeled with set equality and the
-      partial order on regions is modeled with superset relation. Partial orders
+      partial order on regions is modeled with subset relation. Partial orders
       are reflexive and the model of eqivalence agrees with that of the partial
@@
 %
-    \item[Combined] \texttt{*} and \texttt{+} maps to join and meet operations
-      in the lattice. All joins and meets in a lattice satisfy these
+    \item[Combined] \texttt{*} and \texttt{+} maps to meet and join operations
+      in the lattice. All meets and joins in a lattice satisfy these
       inequalities.
@@
       regions. Call these \textit{ks} and \textit{ls} respectively. One way to
-      show $\textit{ls} \supseteq \textit{ks}$ is to show that the intersection
-      of \textit{ls} and \textit{ks} is \textit{ks}.
+      show $\textit{ks} \subseteq \textit{ls}$ is to show that the intersection
+      of \textit{ks} and \textit{ls} is \textit{ks}.
 
@@
       at every point except $n$ the intervals agree, hence the intersection
-      remains the same. Then we use \Cref{lem:zinf-identities} to calculate the
-      intersection. By assumption we have $k \leq l$, so the maximum of the
-      lower bounds is $-k$ and the minimum of lower bounds is $k$ this is same
-      as interval $\textit{int}_k$. Note that pointedness is the same for both
-      intervals and depths are positive integers, so the pointedness is
-      preserved under intersection. This shows the intersection of \textit{ls}
-      and \textit{ks} is indeed \textit{ks} and concludes the proof.
+      remains the same at these dimensions. Then we use
+      \Cref{lem:zinf-identities} to calculate the intersection. By assumption we
+      have $k \leq l$, so the maximum of the lower bounds is $-k$ and the
+      minimum of lower bounds is $k$ this is same as interval $\textit{int}_k$.
+      Note that pointedness is the same for both intervals and depths are
+      positive integers, so the pointedness is preserved under intersection.
+      This shows the intersection of \textit{ls} and \textit{ks} is indeed
+      \textit{ks} and concludes the proof.
   \end{description}
@@
   \begin{align*}
-    \stencil{c}{$n$}{$k$}{\textcap{p}} & \preceq
-      \stencil{f}{$n$}{$k$}{\textcap{p}} \\
+    \stencil{f}{$n$}{$k$}{\textcap{p}} & \preceq
+      \stencil{c}{$n$}{$k$}{\textcap{p}} \\
 %
-    \stencil{c}{$n$}{$k$}{\textcap{p}} & \preceq
-      \stencil{b}{$n$}{$k$}{\textcap{p}}
+    \stencil{b}{$n$}{$k$}{\textcap{p}} & \preceq
+      \stencil{c}{$n$}{$k$}{\textcap{p}}
   \end{align*}
@@
   \shortv{%
-    Both inequalities are proved by breaking \texttt{centered} using
-    (\textbf{Centered}) followed by the inequality (\textbf{Combined}).
+    Both inequalities are proved by combining the missing half of centered by
+    using the inequality (\textbf{Combined}) and then merging them into a single
+    region constant using (\textbf{Centered}).
   }
@@
     \begin{align*}
-      \stencil[s]{c}{$n$}{$k$}{\textcap{p}}
-        & \equiv \stencil[s]{f}{$n$}{$k$}{\textcap{p}}
+      \stencil[s]{f}{$n$}{$k$}{\textcap{p}}
+        & \preceq
+          \stencil[s]{f}{$n$}{$k$}{\textcap{p}}
           \;\texttt{+}\;
           \stencil[s]{b}{$n$}{$k$}{\textcap{p}}
-          & (\textbf{Centered}) \\
-        & \preceq \stencil[s]{f}{$n$}{$k$}{\textcap{p}}
-          & (\textbf{Combined})
+        & (\textbf{Combined}) \\
+      & \equiv \stencil[s]{c}{$n$}{$k$}{\textcap{p}} & (\textbf{Centered})
     \end{align*}
@@
     \begin{align*}
-      \stencil[s]{c}{$n$}{$k$}{\textcap{p}}
-        & \equiv \stencil[s]{f}{$n$}{$k$}{\textcap{p}}
+      \stencil[s]{b}{$n$}{$k$}{\textcap{p}}
+        & \preceq
+          \stencil[s]{f}{$n$}{$k$}{\textcap{p}}
           \;\texttt{+}\;
           \stencil[s]{b}{$n$}{$k$}{\textcap{p}}
-          & (\textbf{Centered}) \\
-        & \preceq \stencil[s]{b}{$n$}{$k$}{\textcap{p}}
-          & (\textbf{Combined})
+        & (\textbf{Combined}) \\
+      & \equiv \stencil[s]{c}{$n$}{$k$}{\textcap{p}} & (\textbf{Centered})
     \end{align*}
@@
   \begin{align*}
-    \stencil{R}{$n$}{$k$}{} & \preceq \stencil{p}{$n$}{}{} \\
+    \stencil{p}{$n$}{}{} & \preceq \stencil{R}{$n$}{$k$}{} \\
 %
-    \stencil{R}{$n$}{$k$}{} & \preceq
-      \stencil{R}{$n$}{$k$}{\texttt{nonpointed}}
+    \stencil{R}{$n$}{$k$}{\texttt{nonpointed}} & \preceq
+      \stencil{R}{$n$}{$k$}{}
   \end{align*}
@@
     \begin{align*}
-      \stencil{R}{$n$}{$k$}{}
-        & \equiv \stencil{R}{$n$}{$k$}{\texttt{nonpointed}}
+      \stencil[s]{p}{$n$}{}{}
+        & \preceq \stencil{R}{$n$}{$k$}{\texttt{nonpointed}}
                   \;\texttt{+}\;
                   \stencil[s]{p}{$n$}{}{}
-        & (\textbf{Overlapping pointed}) \\
-      & \preceq \stencil[s]{p}{$n$}{}{} & (\textbf{Combined})
+        & (\textbf{Combined}) \\
+      & \equiv \stencil{R}{$n$}{$k$}{} & (\textbf{Overlapping pointed})
     \end{align*}
@@
       \stencil{R}{$n$}{$k$}{}
-        & \equiv \stencil{R}{$n$}{$k$}{\texttt{nonpointed}}
+        & \preceq \stencil{R}{$n$}{$k$}{\texttt{nonpointed}}
                   \;\texttt{+}\;
                   \stencil[s]{p}{$n$}{}{}
-        & (\textbf{Overlapping pointed}) \\
-      & \preceq \stencil{R}{$n$}{$k$}{\texttt{nonpointed}}
-        & (\textbf{Combined})
+        & (\textbf{Combined}) \\
+      & \equiv \stencil{R}{$n$}{$k$}{} & (\textbf{Overlapping pointed})
     \end{align*}

### 2017-03-27 oopsla17/lattice-model.tex
Consistency checking

@@
 
-\subsection{Consisteny}
+\subsection{Consistency}
+
+Once the specification is synthesised into objects in the model and stencil
+under examination is analysed, consistency checking itself is remarkably simple.
+Two stage checking is outlined below.
+
+\begin{enumerate}
+  \item Multiplicity of the model is checked against the stencil. The only way
+    this fails is if the original specification involved \texttt{readOnce}, but
+    some indexing occur multiple times in the stencil.  Note that absence of
+    \texttt{readOnce} is interpreted liberally, and if the stencil indices
+    happen to occur uniquely, this does not constitute a failure.
+  \item Definition of vector intervals is completely unwinded into sets of
+    offset tuples and depending the approximation modifier (or lack thereof)
+    we compare these tuples to offsets of the stencil computation. Lower and
+    upper bounds are checked with subset and superset relations respectively,
+    while absence of \texttt{atMost} and \texttt{atLeast} results in exact
+    checking with set equality.
+\end{enumerate}
+
+Here is the mathematical definition of consistency checking:
+\begin{align*}
+  \textit{consistent}(\textit{ix}, \textit{model}) & = \begin{cases}
+    \textsf{false} & \textit{model} = \textsf{once}(x) \wedge ix = \textsf{mult}(y) \\
+    \textit{consistent'}(\textit{peel}(ix), \textit{peel}(model)) & \textit{otherwise}
+  \end{cases} \\
+%
+  \textit{consistent'}(\textit{ix}, \textit{model}) & = \begin{cases}
+    m = ix & \textit{model} = \textsf{exact}(m) \\
+    m \supseteq ix & \textit{model} = \textsf{upper}(m) \\
+    m \subseteq ix & \textit{model} = \textsf{lower}(m)
+  \end{cases}
+\end{align*}
 

### 2017-03-28 oopsla17/lattice-model.tex
Take Point out of equational axioms
It is also not a useful proposition so take it out of there as well.

@@
     \textcap{p3} are \texttt{nonpointed} and \texttt{pointed} otherwise.
-%
-  \item[Point] Let \textcap{R} and \textcap{S} be region constants with two
-    variants a superscript \textit{f} to mean nonpointed and \textit{t} to mean
-    pointed and \textcap{dr} \& \textcap{ds} be the dimensions \textcap{R} and
-    \textcap{S} acts on. Then we have the following dual equivalences:
-%
-    \begin{align*}
-      (\textcap{R}^{\textit{t}} \texttt{*} \textcap{S}^{\textit{f}})
-      \;\texttt{+}\;
-      (\textcap{R}^{\textit{f}} \texttt{*} \textcap{S}^{\textit{t}})
-      \;\texttt{+}\;
-      \stencil[s]{p}{\textit{dr}}{}{}*\stencil[s]{p}{\textit{ds}}{}{} & =
-        \textcap{R}^{\textit{t}} \texttt{*} \textcap{S}^{\textit{t}} \\
-%
-      (\textcap{R}^{\textit{t}} \texttt{+} \textcap{S}^{\textit{f}}) \texttt{*}
-      (\textcap{R}^{\textit{f}} \texttt{+} \textcap{S}^{\textit{t}}) & =
-        \textcap{R}^{\textit{f}} \texttt{+} \textcap{S}^{\textit{f}}
-        \;\texttt{+}\;
-        \stencil[s]{p}{\textit{dr}}{}{}*\stencil[s]{p}{\textit{ds}}{}{}
-    \end{align*}
 \end{description}
@@
       immediately follows from \Cref{lem:zinf-identities}.
-
-    \item[Point] We will prove the first equivalence and then use that to prove
-      the second.
-%
-      Let $\interp{\textcap{R}^{\textnormal{\textit{p}}}}_N$ and
-      $\interp{\textcap{S}^{\textnormal{\textit{p'}}}}_N$ be
-      $\vecgen{N}{k}{\interv{a}{b}{p}}$ and $\vecgen{N}{l}{\interv{c}{d}{p'}}$
-      respectively for some $a$ and $c$ less than or equal to 0, for some $c$
-      and $d$ greater than or equal to 0, for some indices $k$ and $l$ be
-      $\interp{\textcap{dr}}$ and $\interp{\textcap{ds}}$, which are not
-      necessarily distinct and $p$ and $p'$ are \textsf{true} if pointed and
-      \textsf{false} otherwise.
-%
-      \begin{align*}
-        \hspace{2em&\hspace{-2em}}
-        \interp
-          {{(\textcap{R}^{\textit{t}} * \textcap{S}^{\textit{f}})}
-           \;\texttt{+}\;
-           {(\textcap{R}^{\textit{f}} * \textcap{S}^{\textit{t}})}
-           \;\texttt{+}\;
-           \texttt{pointed(dim=\textcap{dr})*pointed(dim=\textcap{ds})}}_N \\
-        & = {(\interp{\textcap{R}^{\textit{t}}}_N \cap
-              \interp{\textcap{S}^{\textit{f}}}_N)} \cup
-            {(\interp{\textcap{R}^{\textit{f}}}_N \cap
-              \interp{\textcap{S}^{\textit{t}}}_N)} \;\cup \\
-        & \quad\,\, {(\vecgen{N}{\interp{ds}}{\interv{0}{0}{\textsf{true}}}
-                      \;\cup\;
-                      \vecgen{N}{\interp{dr}}{\interv{0}{0}{\textsf{true}}})} \\
-        & = \{ x \mid
-            \bigwedge_{\substack{1 \leq i \leq N \\ i \neq k \neq l}} x_i \in
-              \interv{\infty}{\infty}{\textsf{true}}
-            \wedge a \leq x_k \leq b
-            \wedge {(c \leq x_l \leq d \wedge x_l \neq 0)} \} \;\cup \\
-        & \quad\,\, \{ x \mid
-            \bigwedge_{\substack{1 \leq i \leq N \\ i \neq k \neq l}} x_i \in
-              \interv{\infty}{\infty}{\textsf{true}}
-            \wedge {(a \leq x_k \leq b \wedge x_k \neq 0)}
-            \wedge c \leq x_l \leq d \} \;\cup \\
-        & \quad\,\, \{ x \mid
-            \bigwedge_{\substack{1 \leq i \leq N \\ i \neq k \neq l}} x_i \in
-              \interv{\infty}{\infty}{\textsf{true}}
-            \wedge (x_k = 0 \wedge x_l = 0) \} \\
-        & = \{ x \mid
-            \bigwedge_{\substack{1 \leq i \leq N \\ i \neq k \neq l}} x_i \in
-              \interv{\infty}{\infty}{\textsf{true}}
-            \wedge a \leq x_k \leq b
-          \wedge c \leq x_l \leq d \wedge {(x_k \neq 0 \vee x_l \neq 0)} \}
-          \;\cup \\
-        & \quad\,\, \{ x \mid
-            \bigwedge_{\substack{1 \leq i \leq N \\ i \neq k \neq l}} x_i \in
-              \interv{\infty}{\infty}{\textsf{true}}
-            \wedge (x_k = 0 \wedge x_l = 0) \} \\
-        & = \{ x \mid
-            \bigwedge_{\substack{1 \leq i \leq N \\ i \neq k \neq l}} x_i \in
-              \interv{\infty}{\infty}{\textsf{true}}
-            \wedge a \leq x_k \leq b
-            \wedge c \leq x_l \leq d \} \\
-        & = \interp{\textcap{R}^{\textit{t}} \texttt{*} \textcap{S}^{\textit{t}}}_N
-      \end{align*}
-%
-      Now we prove the second equivalence:
-      \begin{align*}
-        \hspace{2em}&\hspace{-2em} \interp{%
-            (\textcap{R}^{\textit{t}} \texttt{+} \textcap{S}^{\textit{f}})
-            \texttt{*}
-            (\textcap{R}^{\textit{f}} \texttt{+} \textcap{S}^{\textit{t}})
-          }_N \\
-        &= \interp{%
-             (\textcap{R}^{\textit{t}} \texttt{*} \textcap{R}^{\textit{f}})
-             \texttt{+}
-             (\textcap{S}^{\textit{t}} \texttt{*} \textcap{S}^{\textit{f}})
-             \texttt{+}
-             (\textcap{R}^{\textit{t}} \texttt{*} \textcap{S}^{\textit{t}})
-             \texttt{+}
-             (\textcap{R}^{\textit{f}} \texttt{*} \textcap{S}^{\textit{f}})
-           }_N \\
-        &= \interp{%
-             \textcap{R}^{\textit{f}}
-             \texttt{+}
-             \textcap{S}^{\textit{f}}
-             \texttt{+}
-             (\textcap{R}^{\textit{t}} \texttt{*} \textcap{S}^{\textit{t}})
-           }_N \\
-        &= \interp
-             {\textcap{R}^{\textit{f}}
-              \texttt{+}
-              \textcap{S}^{\textit{f}}
-              \texttt{+}
-              {(\textcap{R}^{\textit{t}} \texttt{*} \textcap{S}^{\textit{f}})}
-               \;\texttt{+}\;
-               {(\textcap{R}^{\textit{f}} \texttt{*} \textcap{S}^{\textit{t}})}
-               \;\texttt{+}\;
-               \texttt{pointed(dim=\textcap{dr})*pointed(dim=\textcap{ds})}}_N \\
-        &= \interp{\textcap{R}^{\textit{f}} \texttt{+} \textcap{S}^{\textit{f}}
-                   \;\texttt{+}\;
-                   \texttt{pointed(dim=\textcap{dr})*pointed(dim=\textcap{ds})}}_N
-      \end{align*}
   \end{description}

### 2017-03-29 oopsla17/lattice-model.tex
Factor macros used in lattice model out
Remove redundant comment
change to cartesian product and pi for projection notation

@@
 \usepackage{amsthm}
-\usepackage{stmaryrd}
 \usepackage{indentfirst}
 \usepackage{cleveref}
-\usepackage{xifthen}
-
-% Non-essential
-\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
+\usepackage{stencilmacros}
 
@@
 \newtheorem{remark}{Remark}
-
-% Auxillary declaration
-\newcommand{\eg}{\emph{e.g.}}
-
-% Essential macros
-\newcommand{\zinf}{\textnormal{$\mathbb{Z}_\infty$}}
-\newcommand{\interp}[1]{\llbracket{} {#1} \rrbracket{}}
-\newcommand{\interv}[3]{\textnormal{$\lbrack#1,#2\rbrack^{#3}$}}
-\newcommand{\vecgen}[3]{\textnormal{$\textit{vec-gen}_{#1}(#2,#3)$}}
-\newcommand{\textcap}[1]{\texttt{\textit{#1}}}
-
-% Stencil macro
-%
-% Optional argument: s when we want the short form, empty otherwise.
-%
-% Mandatory arguments:
-% 1st: c/b/f/p for centered, backward, forward, pointed
-% 2st: dimension
-% 3st: depth
-% 4st: if pointed/nonpointed needs to be specified leave it blank if omitted.
-%
-% Example: \stencil[s]{f}{1}{2}{} = fwd(dim=1,depth=2)
-% Example: \stencil{b}{1}{2}{pointed} = backward(dim=1,depth=2,pointed)
-\newcommand{\stencil}[5][]{%
-  \ifthenelse{\equal{#2}{f}}{%
-    \ifthenelse{\equal{#1}{s}}{%
-      \texttt{fwd}%
-    }{\ifthenelse{\isempty{#1}}{%
-        \texttt{forward}%
-      }{INVALID OPTIONAL}%
-    }%
-  }{\ifthenelse{\equal{#2}{b}}{%
-      \ifthenelse{\equal{#1}{s}}{%
-          \texttt{bwd}%
-      }{\ifthenelse{\isempty{#1}}{%
-          \texttt{backward}%
-        }{INVALID OPTIONAL}%
-      }%
-  }{\ifthenelse{\equal{#2}{c}}{%
-      \ifthenelse{\equal{#1}{s}}{%
-          \texttt{ctd}%
-      }{\ifthenelse{\isempty{#1}}{%
-          \texttt{centered}%
-        }{INVALID OPTIONAL}%
-      }%
-  }{\ifthenelse{\equal{#2}{p}}{%
-      \ifthenelse{\equal{#1}{s}}{%
-          \texttt{ptd}%
-      }{\ifthenelse{\isempty{#1}}{%
-          \texttt{pointed}%
-        }{INVALID OPTIONAL}%
-      }%
-  }{\ifthenelse{\isempty{#1}}{%
-      \textcap{#2}%
-    }{NO OPTIONAL FOR CUSTOM}%
-  }}}}%
-  \texttt{(dim=\textnormal{#3}%
-    \ifthenelse{\isempty{#4}}{}{,depth=\textnormal{#4}}%
-    \ifthenelse{\isempty{#5}}{}{,\textnormal{#5}})}%
-}
-
-% Non-essential macro
-\definecolor{unused}{RGB}{211,167,192}
-\newcommand{\unused}[1]{{\color{unused} #1}}
-
-\definecolor{version}{RGB}{163,193,173}
-\newcommand{\longv}[1]{{\color{version} Long version:} #1}
-\newcommand{\shortv}[1]{{\color{version} Short version:} #1}
 
@@
   offsets from induction variables.
-
-%  $\mathcal{P}(\zinf{})$ couple with $\supseteq$ is the power set distributive
-%  lattice.
 \end{defn}
@@
 \begin{defn}{(\emph{Offsets vector})}
-  An \emph{offsets vector} is a $N$ dimensional vector of offset spaces denoted
-  by $(\textit{offs}_1, \dots, \textit{offs}_N)$ to mean $\textit{offs}_1 \times
-  \dots \times \textit{offs}_N$.
+  An $N$ dimensional \emph{offsets vector} is a Cartesian product of offset
+  spaces denoted with $\textit{offs}_1 \times \dots \times \textit{offs}_N$.
+  The function $\pi_i : \zinf{}^{\mkern-19mu{N}} \to \zinf{}$ retrieves the
+  $i^{th}$ set of offsets.
 
@@
 \begin{lem}{}\label{lem:vector-intersect}
-  Intersection distributes over offsets vectors.
+  Intersection distributes over offsets vectors. If $v$ and $v'$ are $N$
+  dimensional ofssets vectors, then
 %
   \begin{equation*}
-    (\textit{vint}_1, \dots, \textit{vint}_N) \;\cap\;
-      (\textit{vint'}_1, \dots, \textit{vint'}_N)
-      =
-    (\textit{vint}_1 \;\cap\; \textit{vint'}_1, \dots,
-     \textit{vint}_N \;\cap\; \textit{vint'}_N)
+    v \cap v' = \prod_{i = 1}^{N} \pi_i(v) \cap \pi_i(v')
   \end{equation*}
@@
     & = \{x \mid \bigwedge_{1 \leq i \leq N } (x_i \in v_i \wedge x_i \in v'_i) \} \\
-    & = (v_1 \cap v'_1,\dotsc,v_N \cap v'_N)
+    & = \prod_{i = 1}^{N} \pi_i(v) \cap \pi_i(v')
   \end{align*}
@@
 \begin{lem}{}\label{lem:vector-union}
-  Let $v$ and $v'$ be two $N$ dimensional offsets vectors such that $v_i$ is
-  equal to $v'_i$ for all $1 \leq i \leq N$ except $k$, then we have the
+  Let $v$ and $v'$ be two $N$ dimensional offsets vectors such that $\pi_i(v)$
+  is equal to $\pi_i(v')$ for all $1 \leq i \leq N$ except $k$, then we have the
   following:
@@
   \begin{equation*}
-    (v_1,\dotsc, v_k,\dotsc, v_N) \; \cup \; (v'_1,\dotsc, v'_k, \dotsc, v_N)
+    v \; \cup \; v'
     =
-    (v_1,\dotsc, v_k \cup v'_k,\dotsc, v_N)
+    \pi_1(v) \times \cdots \times
+    (\pi_k(v) \cup \pi_k(v')) \times \cdots \times
+    \pi_N(v)
   \end{equation*}
@@
         \} \\
-    & = (v_1,\dotsc, v_k \cup v'_k,\dotsc, v_N)
+    & = \pi_1(v) \times \cdots \times
+        (\pi_k(v) \cup \pi_k(v')) \times \cdots \times
+        \pi_N(v)
   \end{align*}
@@
   A vector interval is an $N$ dimensional object which has \zinf{} intervals at
-  each dimension. It is denoted with $(\textit{int}_1, \dots, \textit{int}_n)$,
-  where $\textit{int}_i$ is a \zinf{} interval. If \textit{vint} is a vector
-  interval, $\textit{vint}_i$ denotes the \zinf{} interval at the $i^{th}$
-  dimension.
+  each dimension. A vector interval is also an ofssets vector.
 
-  The set of $N$ dimensional vector intervals is $\textit{Interval}^N$. Every
-  vector interval is also an offsets vector.
+  The set of all $N$ dimensional vector intervals is $\textit{Interval}^N$.
 \end{defn}
@@
 \textsf{true}$, $\interp{\texttt{nonpointed}} = \textsf{false}$.
-Typographically, cursive monotype font is used within specifications to indicate
-a partial textual capture. For convenience we use \textit{italic} serif font
-inside specifications rather than interpreting its textual capture. This
-coalesces syntactic and semantic representation but avoids clutter in the
+Typographically, \textcap{italic monotype font} is used within specifications
+to indicate a partial textual capture. For convenience, we use \textit{italic
+serif font} inside specifications rather than interpreting its textual capture.
+This coalesces syntactic and semantic representation but avoids clutter in the
 presentation.

### 2017-03-29 oopsla17/lattice-model.tex
More notes
switch lattice model to acmart
Minor fix

@@
-\documentclass{article}
+\documentclass[acmlarge,review]{acmart}
 
@@
 \begin{document}
+
+\title{Lattice model}
+\author{Mistral Contrastin}
 
@@
     \texttt{backward}, or \texttt{centered}, then we have
-    $\stencil{\textcap{R}}{$n$}{$k$}{\texttt{nonpointed}} \;\texttt{+}\;
+%
+    \begin{equation*}
+      \stencil{\textcap{R}}{$n$}{$k$}{\texttt{nonpointed}} \;\texttt{+}\;
       \stencil[s]{p}{$n$}{}{} \equiv
-     \stencil{\textcap{R}}{$n$}{$k$}{}$.
+      \stencil{\textcap{R}}{$n$}{$k$}{}
+    \end{equation*}
 %
@@
 %
-    \begin{equation*}
+    \begin{align*}
       \stencil[s]{c}{$n$}{$k$}{\textcap{p1}} \equiv
@@
         \stencil[s]{b}{$n$}{$k$}{\textcap{p3}}
-    \end{equation*}
+    \end{align*}
 %
@@
 
+Here helper function $\textit{peel} : \textsf{Approx} a \to a$ removes the
+approximation label.
+
+\todo{Here we simply show \textit{consistency'} as set equality/subset/superset,
+but since sets are infinite we might need to elaborate based on representation.}
+
 \subsection{Inference}

### 2017-03-29 oopsla17/lattice-model.tex
Add forgotten subscript
expand stencil abbreviation that now fits
Add omitted modifier in proof
fix dot

@@
   An $N$ dimensional \emph{offsets vector} is a Cartesian product of offset
-  spaces denoted with $\textit{offs}_1 \times \dots \times \textit{offs}_N$.
+  spaces denoted with $\textit{offs}_1 \times \cdots \times \textit{offs}_N$.
   The function $\pi_i : \zinf{}^{\mkern-19mu{N}} \to \zinf{}$ retrieves the
@@
 \begin{equation*}
-  \interp{\texttt{stencil \textcap{Mult}, \textcap{Appr}, \textcap{Region}}} =
+  \interp{\texttt{stencil \textcap{Mult}, \textcap{Appr}, \textcap{Region}}}_N =
     \interp{\textcap{Mult}}
            {(\interp{\textcap{Appr}}
-                    {(\interp{\textcap{Region}})})}
+                    {(\interp{\textcap{Region}}_N)})}
 \end{equation*}
@@
       \stencil{\textcap{R}}{$n$}{$k$}{\texttt{nonpointed}} \;\texttt{+}\;
-      \stencil[s]{p}{$n$}{}{} \equiv
+      \stencil{p}{$n$}{}{} \equiv
       \stencil{\textcap{R}}{$n$}{$k$}{}
@@
     \begin{align*}
-      \stencil{R}{$n$}{$k$}{}
+      \stencil{R}{$n$}{$k$}{\textcap{nonpointed}}
         & \preceq \stencil{R}{$n$}{$k$}{\texttt{nonpointed}}

### 2017-03-30 oopsla17/lattice-model.tex

@@
 \begin{defn}{(\emph{Region})}
-  \textit{Region} is the set of $N$ dimensional vector intervals defined as the
-  smallest set satisfying the following:
+  \region{N} is the set of $N$ dimensional vector intervals defined as
+  the smallest set satisfying the following:
 %
   \begin{enumerate}
-    \item If $R$ is in $\textit{Interval}^N$, then $R$ is in \textit{Region}.
-    \item If $R$ and $S$ are in \textit{Region}, then so are $R \cap S$ and $R
-      \cup S$.
+    \item If $R$ is in $\textit{Interval}^N$, then $R$ is in \region{N}.
+    \item If $R$ and $S$ are in \region{N}, then so are $R \cap S$ and
+      $R \cup S$.
   \end{enumerate}
@@
 \begin{prop}{}
-  $(\textit{Region},\subseteq)$ is a distributive lattice.
+  $(\region{N},\subseteq)$ is a distributive lattice.
 \end{prop}
@@
   partially ordered with subset or equal to relation are distributive lattices.
-  \textit{Region} is a subset of $\mathcal{P}{(\zinf{}^{\mkern-19mu{N}})}$ since
-  all vector intervals are subsets of this power set and union and intersection
+  \region{N} is a subset of $\mathcal{P}{(\zinf{}^{\mkern-19mu{N}})}$ since all
+  vector intervals are subsets of this power set and union and intersection
   cannot add an element that is not already in the powerset.  Inductive
-  definition also ensures that \textit{Region} is closed under these operations.
-  Therefore, \textit{Region} is a distributive sublattice of
+  definition also ensures that \region{N} is closed under these operations.
+  Therefore, \region{N} is a distributive sublattice of
   $(\mathcal{P}{(\zinf{}^{\mkern-19mu{N}})}, \subseteq)$.
@@
 
+\begin{defn}{(Extended origin)}
+  In the $N$ dimensional region space, ${(\interv{-1}{1}{\textsf{true}})}^N$ is
+  said to be the extended origin.
+\end{defn}
+
+\begin{defn}{(Hearing)}
+  A point, $p$, in \region{N} is said to \emph{talk to} the extended origin if
+%
+  \begin{enumerate}
+    \item $p$ is in the extended origin;
+    \item there are distinct indices $i$ and $j$ (between $1$ and $N$) such that
+      $p_i$ is $-1$ to $1$ and there is a $p'$ that talks to the extended origin
+      such that $p_j$ is ${p'}_j - 1$ or ${p'}_j + 1$ and for all other indices
+      $k$, $p_k$ is ${p'}_k$;
+    \item all immediate neighbours of $p$ that are no farther away from the
+      extended origin than $p$ can talk to the extended origin.
+  \end{enumerate}
+
+  A set of points talk to the extended origin if all points in the set talk to
+  the extended origin.
+\end{defn}
+
+\begin{lem}
+  The following three statements are equivalent for a set of points $P \subseteq
+  \mathbb{Z}^N$:
+%
+  \begin{enumerate}
+    \item $P$ defines a superposition of hyper-rectangles each of which crosses
+      the extended origin;
+    \item $P$ talks to the extended origin
+    \item $P$ when regarded as offsets from induction variables forms a
+      neighbourhood stencil;
+    \item $P$ lies in \region{N}.
+  \end{enumerate}
+\end{lem}
+
 \begin{thm}{(Completeness of inference)}\label{thm:inf-completeness}

### 2017-03-30 oopsla17/lattice-model.tex
Add exact set and related definitions

@@
 
-  Set of all such intervals (sets) is $\textit{Interval}$.
+  Set of all such intervals (sets) is $\textit{Interval}$. If the superscript is
+  leftout it is treated as \textsf{true}.
 \end{defn}
@@
   distributive lattice under the superset relation.
+\end{defn}
+
+\begin{defn}{(\emph{Point offsets vector})}
+  A \emph{point offsets vector}, \textit{poff}, is a an offsets vector such that
+  for all indices $i$ $\pi_i(\textit{poff})$, we either have a singleton set or
+  the infinite interval $\interv{-\infty}{\infty}{\textsf{true}}$.
 \end{defn}
@@
 \begin{defn}{(Extended origin)}
-  In the $N$ dimensional region space, ${(\interv{-1}{1}{\textsf{true}})}^N$ is
-  said to be the extended origin.
+  In the $N$ dimensional region space, ${(\interv{-1}{1}{})}^N$ is said to be
+  the extended origin denoted $\textit{EO}_N$.
 \end{defn}
 
-\begin{defn}{(Hearing)}
-  A point, $p$, in \region{N} is said to \emph{talk to} the extended origin if
+\begin{defn}{(Origin distance)}
+  We measure the \emph{origin distance} of a point offsets vector from the
+  extended origin as the minimum number of translations by $\pm 1$ unit in one
+  dimension at a time to reach any point in the extended origin. Since point
+  offsets vector may represent infinite number of points, we measure the
+  distance from the closest one.
+\end{defn}
 %
-  \begin{enumerate}
-    \item $p$ is in the extended origin;
-    \item there are distinct indices $i$ and $j$ (between $1$ and $N$) such that
-      $p_i$ is $-1$ to $1$ and there is a $p'$ that talks to the extended origin
-      such that $p_j$ is ${p'}_j - 1$ or ${p'}_j + 1$ and for all other indices
-      $k$, $p_k$ is ${p'}_k$;
-    \item all immediate neighbours of $p$ that are no farther away from the
-      extended origin than $p$ can talk to the extended origin.
-  \end{enumerate}
+\begin{prop}{}
+  The closed form of origin distance of an $N$ dimensional point offsets vector
+  $p$ is the following:
+%
+  \begin{align*}
+    \textit{dist}_i(p) & = \begin{cases}
+      0 & \pi_i(p) = \interv{-\infty}{\infty}{} \\
+      |\mathsf{a}| - 1 & \pi_i(p) = \{\mathsf{a}\} \\
+    \end{cases} &
+%
+    |p| & = \sum_{i=1}^{N} dist_i(p)
+  \end{align*}
+\end{prop}
+%
+\begin{proof}
+\end{proof}
 
-  A set of points talk to the extended origin if all points in the set talk to
-  the extended origin.
+\begin{defn}{(Close neighbourhood)}
+  A point offsets vector $p'$ is said to be a close neighbour of point offsets
+  vector $p$ if for every index $i$ except $j$, $\pi_i(p)$ agrees with
+  $\pi_i(p')$ but at $j$, $\pi_j(p')$ is such that $|p'| + 1 = |p|$.
+
+  This necessarily means point vectors within the extended origin and the
+  degenerate point offsets vector with \interv{-\infty}{\infty}{} does not have
+  close neighbours.
+\end{defn}
+%
+\begin{prop}
+  The unique close neighbour at index $i$ of an $N$ dimensional point offsets
+  vector, if it exists, is given by the following partial function:
+%
+  \begin{align*}
+    \mathit{close\_neighbour}(p,i) & = \prod^{N}_{j=1} \mathit{shift}(p,i,j) &
+%
+    \mathit{shift}(p,i,j) & =
+      \begin{cases}
+        \{a - 1\} & i = j \wedge \pi_i(p) = \{\mathsf{a}\} \wedge \mathsf{a} > 0\\
+        \{a + 1\} & i = j \wedge \pi_i(p) = \{\mathsf{a}\} \wedge \mathsf{a} < 0\\
+        \pi_j(p)  & otherwise
+      \end{cases}
+  \end{align*}
+\end{prop}
+%
+\begin{proof}
+\end{proof}
+
+\begin{defn}{(Exact set)}
+  The \emph{exact set} at origin distance $d$, $\mathit{ES}_{P,d}$, of a group
+  of $N$ dimensional points $P$ is defined below:
+%
+  \begin{equation*}
+    \mathit{ES}_{P,d} =
+      \{ p \in P \mid \forall i.\
+         p_i \notin \{-1,0,1\} \implies
+         \exists p' \in \mathit{ES}_{P,d-1}.\ p' =
+         \mathit{close\_neighbour}(p,i) \}
+  \end{equation*}
+%
+  The overall \emph{exact set}, $\mathit{ES}_P$ is then $\mathit{ES}_{P,n}$ such
+  tat $\mathit{ES}_{P,n} = \mathit{ES}_{P,n+1}$.
 \end{defn}
@@
       the extended origin;
-    \item $P$ talks to the extended origin
+    \item $P$ is the same as its exact set $\mathit{ES}_P$;
     \item $P$ when regarded as offsets from induction variables forms a
       neighbourhood stencil;
-    \item $P$ lies in \region{N}.
+    \item $P$ is a subset of \region{N}.
   \end{enumerate}

### 2017-03-31 oopsla17/oopsla17.tex
a bit of intro context and appendix

@@
 \usepackage{indentfirst}
+\usepackage{cleveref}
+\usepackage{xcolor}
+\usepackage{siunitx} % For pretty-printing numeric values and SI units
+                     % of measure. e.g., the tabular column type S is
+                     % used to print nice-looking tables of numbers.
+\sisetup{ % defaults
+  group-separator={,},
+  group-minimum-digits={3},
+  output-decimal-marker={.},
+  table-format = 6
+}
 
@@
 
+% Writing macros
 \newcommand{\eg}{\emph{e.g.}}
@@
 
-\usepackage{cleveref}
+% Meta macros
+\newcommand{\todo}[1]{\textcolor{blue}{#1}}
 
-\usepackage{xcolor}
-\newcommand{\todo}[1]{\textcolor{blue}{#1}}
+% For evluation and empirical study
+% Defines the commands (headers)
+\input{resultsCmds}
+% renews the commands with actual data
+%\input{results}
 
@@
 
+\paragraph{Context of this research}
+
+There is an increasing awareness of the need for program verification
+techniques in
+science~\cite{post2005computational,oberkampf2010verification,orchard2014computational}.
+Whilst there are various kinds of automated and semi-automated
+verification tools available, they see little use in science. This is
+partly due to a lack of training and a lack of awareness. But its also
+due to a lack of targetting. This paper is part of a line of research
+to provide lightweight verification tools targetting
+programming patterns common in computational science code, 
+that are easy to use by scientists, and that integrate with common
+practises and workflows.
+
 \section{An Empirical Study of Stencils in Scientific Fortran code}
+
+\subsection{Our corpus}
+%
+\noindent
+We put together a corpus of around 1 million lines of Fortran code from a
+\numPackages{} scientific computing packages ranging in size and scope: The Unified Model (UM)~\cite{um},
+E3MG~\cite{RePEc:aen:journl:2006se-a12}, BLAS~\cite{blas},
+Hybrid4~\cite{GBC:GBC635}, GEOS-Chem~\cite{geos-chem}, Navier (based
+on \cite{griebel1997numerical}), Computational Physics
+2~\cite{nicholas2006computational},
+ARPACK-NG~\cite{arpackng}, and
+SPECFEM3D~\cite{specfem3d}. Appendix~\ref{app:corpus} provides
+further detail on these packages.
 
@@
 
+\appendix
+% http://2017.splashcon.org/track/splash-2017-OOPSLA#Instructions-for-Authors
+% "There is no page limit for bibliographic references and appendices,
+% and, therefore, for the overall submission."
+\input{appendix}
+
 \end{document}

### 2017-03-31 oopsla17/lattice-model.tex
stylistic fixes

@@
   A \emph{\zinf{} interval} is of the form \interv{a}{b}{c}, where $a$ and $b$
-  are drawn from \zinf{} with $a \leq 0 \leq b$ and $c$ is a Boolean value. We
+  are drawn from \zinf{} with $a \leq 0 \leq b$, $a$ is $-\infty$ \emph{iff} $b$
+  is $\infty$, and $c$ is drawn from $\{ \textsf{true}, \textsf{false} \}$. We
   define it as follows:
@@
       \vee \neg f \implies n \neq 0) \} \\
-    = & \; \interv{\max{\{a,d\}}}{\min{\{b,e\}}}{\neg c \vee \neg f}
+    = & \; \interv{\max{\{a,d\}}}{\min{\{b,e\}}}{c \wedge f}
   \end{align*}
@@
 \begin{lem}{}\label{lem:vector-intersect}
-  Intersection distributes over offsets vectors. If $v$ and $v'$ are $N$
-  dimensional ofssets vectors, then
+  Intersection distributes over offsets vectors. If $\mathit{off}$ and
+  $\mathit{off'}$ are $N$ dimensional ofssets vectors, then
 %
   \begin{equation*}
-    v \cap v' = \prod_{i = 1}^{N} \pi_i(v) \cap \pi_i(v')
+    \mathit{off} \cap \mathit{off'} =
+      \prod_{i = 1}^{N} \pi_i(\mathit{off}) \cap \pi_i(\mathit{off'})
   \end{equation*}
@@
   \begin{align*}
-    v \; \cap \; v' &
-    = \{x \mid \bigwedge_{1 \leq i \leq N } x_i \in v_i \}
+    \mathit{off} \; \cap \; \mathit{off'} &
+    = \{x \mid \bigwedge_{1 \leq i \leq N } x_i \in \pi_i(\mathit{off}) \}
       \;\cap\;
-      \{x \mid \bigwedge_{1 \leq i \leq N } x_i \in v'_i \} \\
-    & = \{x \mid \bigwedge_{1 \leq i \leq N } (x_i \in v_i \wedge x_i \in v'_i) \} \\
-    & = \prod_{i = 1}^{N} \pi_i(v) \cap \pi_i(v')
+      \{x \mid \bigwedge_{1 \leq i \leq N } x_i \in \pi_i(\mathit{off'}) \} \\
+    & = \{x \mid \bigwedge_{1 \leq i \leq N }
+      (x_i \in \pi_i(\mathit{off}) \wedge x_i \in \pi_i(\mathit{off'})) \} \\
+    & = \prod_{i = 1}^{N} \pi_i(\mathit{off}) \cap \pi_i(\mathit{off'})
   \end{align*}
@@
 \begin{lem}{}\label{lem:vector-union}
-  Let $v$ and $v'$ be two $N$ dimensional offsets vectors such that $\pi_i(v)$
-  is equal to $\pi_i(v')$ for all $1 \leq i \leq N$ except $k$, then we have the
-  following:
+  Let $\mathit{off}$ and $\mathit{off'}$ be two $N$ dimensional offsets vectors
+  such that $\pi_i(\mathit{off})$ is equal to $\pi_i(\mathit{off'})$ for all $1
+  \leq i \leq N$ except possibly at $k$, then we have the following:
 %
   \begin{equation*}
-    v \; \cup \; v'
+    \mathit{off} \; \cup \; \mathit{off'}
     =
-    \pi_1(v) \times \cdots \times
-    (\pi_k(v) \cup \pi_k(v')) \times \cdots \times
-    \pi_N(v)
+    \pi_1(\mathit{off}) \times \cdots \times
+    (\pi_k(\mathit{off}) \cup \pi_k(\mathit{off'})) \times \cdots \times
+    \pi_N(\mathit{off})
   \end{equation*}
@@
   \begin{align*}
-    v \; \cup \; v' &
+    \mathit{off} \; \cup \; \mathit{off'} &
     = \{x \mid
-          \bigwedge_{1 \leq i \leq N } x_i \in v_i \}
+      \bigwedge_{1 \leq i \leq N } x_i \in \pi_i(\mathit{off}) \}
       \;\cup\;
       \{x \mid
-          \bigwedge_{1 \leq i \leq N } x_i \in v'_i \} \\
+          \bigwedge_{1 \leq i \leq N } x_i \in \pi_i(\mathit{off'}) \} \\
     & = \{x \mid
-          \bigwedge_{\substack{1 \leq i \leq N \\ i \neq k}} x_i \in v_i
-              \wedge x_k \in v_k \vee
-          \bigwedge_{\substack{1 \leq i \leq N \\ i \neq k}} x_i \in v'_i
-            \wedge x_k \in v'_k
+          \bigwedge_{\substack{1 \leq i \leq N \\ i \neq k}}
+            x_i \in \pi_i(\mathit{off}) \wedge x_k \in \pi_k(\mathit{off}) \vee
+          \bigwedge_{\substack{1 \leq i \leq N \\ i \neq k}} x_i \in
+            \pi_i(\mathit{off'}) \wedge x_k \in \pi_k(\mathit{off'})
         \} \\
     & = \{x \mid
-          \bigwedge_{\substack{1 \leq i \leq N \\ i \neq k}} x_i \in v_i
-          \wedge x_k \in v_k \cup v'_k
+          \bigwedge_{\substack{1 \leq i \leq N \\ i \neq k}} x_i \in
+            \pi_i(\mathit{off}) \wedge
+            x_k \in \pi_k(\mathit{off}) \cup \pi_k(\mathit{off'})
         \} \\
-    & = \pi_1(v) \times \cdots \times
-        (\pi_k(v) \cup \pi_k(v')) \times \cdots \times
-        \pi_N(v)
+        & = \pi_1(\mathit{off}) \times \cdots \times
+        (\pi_k(\mathit{off}) \cup \pi_k(\mathit{off'})) \times \cdots \times
+        \pi_N(\mathit{off})
   \end{align*}
@@
 \textsf{true}$, $\interp{\texttt{nonpointed}} = \textsf{false}$.
-Typographically, \textcap{italic monotype font} is used within specifications
+Typographically, \textcap{bold monotype font} is used within specifications
 to indicate a partial textual capture. For convenience, we use \textit{italic
@@
       \stencil[s]{p}{$n$}{}{}
-        & \preceq \stencil{R}{$n$}{$k$}{\texttt{nonpointed}}
+        & \preceq \stencil{R}{$n$}{$k$}{nonpointed}
                   \;\texttt{+}\;
@@
     \begin{align*}
-      \stencil{R}{$n$}{$k$}{\textcap{nonpointed}}
-        & \preceq \stencil{R}{$n$}{$k$}{\texttt{nonpointed}}
+      \stencil{R}{$n$}{$k$}{nonpointed}
+        & \preceq \stencil{R}{$n$}{$k$}{nonpointed}
                   \;\texttt{+}\;

### 2017-03-31 oopsla17/lattice-model.tex
Fix overall exact set def
Add soundness condition and fix some typesetting

@@
   are drawn from \zinf{} with $a \leq 0 \leq b$, $a$ is $-\infty$ \emph{iff} $b$
-  is $\infty$, and $c$ is drawn from $\{ \textsf{true}, \textsf{false} \}$. We
+  is $\infty$, and $c$ is drawn from $\{ \mathsf{true}, \mathsf{false} \}$. We
   define it as follows:
@@
   Set of all such intervals (sets) is $\textit{Interval}$. If the superscript is
-  leftout it is treated as \textsf{true}.
+  leftout it is treated as $\mathsf{true}$.
 \end{defn}
@@
   for all indices $i$ $\pi_i(\textit{poff})$, we either have a singleton set or
-  the infinite interval $\interv{-\infty}{\infty}{\textsf{true}}$.
+  the infinite interval $\interv{-\infty}{\infty}{\mathsf{true}}$.
 \end{defn}
@@
 \begin{defn}{}
-  \textsf{Mult} and \textsf{Approx} are parametric labelled variant types with
-  injections given by their definition:
+  $\mathsf{Mult}$ and $\mathsf{Approx}$ are parametric labelled variant types
+  with injections given by their definition:
 %
   \begin{align*}
-    \textsf{Mult} \; a &
-      \triangleq \textsf{mult} \; a \;\mid\; \textsf{only} \; a \\
-    \textsf{Approx} \; a &
-      \triangleq \textsf{exact} \; a \;\mid\; \textsf{lower} \; a \;\mid\;
-        \textsf{upper} \; a \;\mid\; \textsf{both} \; a \; a
+    \mathsf{Mult} \; a &
+      \triangleq \mathsf{mult} \; a \;\mid\; \mathsf{only} \; a \\
+    \mathsf{Approx} \; a &
+      \triangleq \mathsf{exact} \; a \;\mid\; \mathsf{lower} \; a \;\mid\;
+        \mathsf{upper} \; a \;\mid\; \mathsf{both} \; a \; a
   \end{align*}
 %
-  \eg{}, \textsf{lower} is an injection $\mathsf{lower} : a \to \mathsf{Approx}
+  \eg{}, $\mathsf{lower}$ is an injection $\mathsf{lower} : a \to \mathsf{Approx}
   \; a$ etc.
@@
 $\mathbb{Z}^+$, $\interp{\texttt{pointed}} = \llbracket \epsilon \rrbracket =
-\textsf{true}$, $\interp{\texttt{nonpointed}} = \textsf{false}$.
+\mathsf{true}$, $\interp{\texttt{nonpointed}} = \mathsf{false}$.
 Typographically, \textcap{bold monotype font} is used within specifications
@@
 interval such that \vecgen{N}{i}{\interv{a}{b}{c}} has
-$\interv{-\infty}{\infty}{\textsf{true}}$ in all dimensions except the $i^{th}$
+$\interv{-\infty}{\infty}{\mathsf{true}}$ in all dimensions except the $i^{th}$
 one. There, it has \interv{a}{b}{c}.
@@
   \interp{\stencil{p}{$i$}{}{}}_N & =
-    \vecgen{N}{i}{\interv{0}{0}{\textsf{true}}}\\
+    \vecgen{N}{i}{\interv{0}{0}{\mathsf{true}}}\\
 %
@@
 \begin{align*}
-  \interp{\texttt{readOnce}} & = \textsf{mult} &
-  \llbracket\epsilon\rrbracket & = \textsf{once}
+  \interp{\texttt{readOnce}} & = \mathsf{mult} &
+  \llbracket\epsilon\rrbracket & = \mathsf{once}
 \end{align*}
@@
 \begin{align*}
-  \interp{\texttt{atLeast}} & = \textsf{lower} &
-  \interp{\texttt{atMost}} & = \textsf{upper} &
-  \llbracket \epsilon \rrbracket = \textsf{exact}
+  \interp{\texttt{atLeast}} & = \mathsf{lower} &
+  \interp{\texttt{atMost}} & = \mathsf{upper} &
+  \llbracket \epsilon \rrbracket = \mathsf{exact}
 \end{align*}
@@
         \interp{\stencil{\textcap{R}}{$n$}{$k$}{\texttt{nonpointed}}}_N
-        & = \vecgen{N}{n}{\interv{-k}{k}{\textsf{false}}} \\
+        & = \vecgen{N}{n}{\interv{-k}{k}{\mathsf{false}}} \\
 %
         \interp{\stencil{p}{$n$}{}{}}_N
-        &= \vecgen{N}{n}{\interv{0}{0}{\textsf{true}}} \\
+        &= \vecgen{N}{n}{\interv{0}{0}{\mathsf{true}}} \\
 %
         \interp{\stencil{\textcap{R}}{$n$}{$k$}{}}_N
-        &= \vecgen{N}{n}{\interv{-k}{k}{\textsf{true}}}
+        &= \vecgen{N}{n}{\interv{-k}{k}{\mathsf{true}}}
       \end{align*}
@@
       Using \Cref{lem:vector-union}, it is sufficient to show
-      $ \interv{-k}{k}{p} \;\cup\; \interv{0}{0}{\textsf{true}} =
-        \interv{-k}{k}{\textsf{true}} $ holds. This immediately follows from
+      $ \interv{-k}{k}{p} \;\cup\; \interv{0}{0}{\mathsf{true}} =
+        \interv{-k}{k}{\mathsf{true}} $ holds. This immediately follows from
       \Cref{lem:zinf-identities}.
@@
   \textit{consistent}(\textit{ix}, \textit{model}) & = \begin{cases}
-    \textsf{false} & \textit{model} = \textsf{once}(x) \wedge ix = \textsf{mult}(y) \\
+    \mathsf{false} & \textit{model} = \mathsf{once}(x) \wedge ix = \mathsf{mult}(y) \\
     \textit{consistent'}(\textit{peel}(ix), \textit{peel}(model)) & \textit{otherwise}
@@
   \textit{consistent'}(\textit{ix}, \textit{model}) & = \begin{cases}
-    m = ix & \textit{model} = \textsf{exact}(m) \\
-    m \supseteq ix & \textit{model} = \textsf{upper}(m) \\
-    m \subseteq ix & \textit{model} = \textsf{lower}(m)
+    m = ix & \textit{model} = \mathsf{exact}(m) \\
+    m \supseteq ix & \textit{model} = \mathsf{upper}(m) \\
+    m \subseteq ix & \textit{model} = \mathsf{lower}(m)
   \end{cases}
@@
 
-Here helper function $\textit{peel} : \textsf{Approx} a \to a$ removes the
+Here helper function $\textit{peel} : \mathsf{Approx} a \to a$ removes the
 approximation label.
@@
 %
-  The overall \emph{exact set}, $\mathit{ES}_P$ is then $\mathit{ES}_{P,n}$ such
-  tat $\mathit{ES}_{P,n} = \mathit{ES}_{P,n+1}$.
+  The overall \emph{exact set}, $\mathit{ES}_P$ is then
+  $\bigcup_{1 \leq d \leq D} \mathit{ES}_{P,d}$, where $D$ is the maximum origin
+  distance in $P$.
 \end{defn}
@@
 \end{lem}
+
+\begin{thm}{(Soundness of exact shape inference)}
+  For a set of point offsets vectors $P$, the inference procedure for
+  neighbourhood stencils is sound if the following holds:
+%
+  \begin{equation*}
+    \mathit{consistent}(\mathsf{mult}(P),\mathsf{mult}(\mathit{infer}(P)))
+  \end{equation*}
+\end{thm}
 

### 2017-03-31 oopsla17/appendix.tex

% http://2017.splashcon.org/track/splash-2017-OOPSLA#Instructions-for-Authors
% "There is no page limit for bibliographic references and appendices,
% and, therefore, for the overall submission."

\section{Details of the corpus data set}
\label{app:corpus}

\paragraph{Software Corpus}
Table~\ref{tab:corpus} shows summary statistics of the software
packages used in our evaluation, all of which are written in Fortran
90 or Fortran 77. In total we analysed \SI{\overallLoC} lines of code
from \numPackages{} packages, of which we successfully parsed
\SI{\overalllinesParsed} lines. The ``Number of files'' column shows
how many files in each corpus that we were able to analyse with
CamFort. The most common reasons for CamFort rejecting a file were
either use of a C-preprocessor, or illegal use of language features
from a modern Fortran variant.

\begin{enumerate}
\item \textbf{The Unified Model}~\cite{um} is a weather
  forecasting and climate modelling tool developed by the Met Office
  in the United Kingdom. It is used by research organisations and
  meteorological services around the world. We use the development
  branch (trunk) of the model. The code base in closed source but
  institutional licenses are available for research purposes. The Met
  Office runs a comprehensive code quality system incorporating
  dedicated committers (we counted 11) for particular parts of the
  model. We counted 120 additional contributors whose submissions are
  reviewed and tested before being accepted into the code base.

\item \textbf{E3MG} (An Energy-Environment-Economy (E3) Model at the Global
Level) is a macroeconomic model used for assessment of environmental
policy~\cite{RePEc:aen:journl:2006se-a12}. This was developed by
Cambridge Econometrics, an independent consultancy company.

\item \textbf{BLAS}~\cite{blas} (Basic Linear Algebra Subprograms) is
  a popular library providing efficient and portable routines for
  vector and matrix operations. These routines feature in many other
  libraries (including LAPACK). We used version 3.6.0. We chose to
  include this package for breadth, as it provides general numerical
  functions rather than a specialised scientific model.

\item \textbf{Hybrid4} is a vegetation and biomass model for
  simulating carbon, water and nitrogen flows~\cite{GBC:GBC635}.

\item \textbf{GEOS-Chem}~\cite{geos-chem} is a three-dimensional model
  of tropospheric chemistry developed at Harvard and used by $\sim$70
  universities and research institutions world-wide. We use v.10-01.

\item \textbf{Navier} is a small numerical simulation, giving a
  discrete approximation to the two-dimensional Navier-Stokes fluid
  equations, based on the book of~\citet{griebel1997numerical}.

\item \textbf{CP} consists of the example code from the second edition
  of the book ``Computational
  Physics''~\cite{nicholas2006computational} introducing numerical
  techniques and their application to modern physics problems such as
  fields, waves, statistical mechanics and quantum mechanics.

\item \textbf{ARPACK-NG}~\cite{arpackng} \todo{add a sentence of detail}

\item \textbf{SPECFEM3D}~\cite{specfem3d} \todo{add a little detail}
\end{enumerate}

### 2017-03-31 oopsla17/analysis/Analyse.hs
various work on the core definitions

@@ -65,12 +65,12 @@ main :: IO ()
 main = do
     args <- getArgs
     if (length args == 1) then
-      if (head args == "PI") then
+      {- if (head args == "PI") then
         testLHSvars True
       else 
       if (head args == "PI2") then
         testLHSvars False
-      else
+      else -}
         applyAnalysisToDir (head args) False []
     else if (length args > 1) then
       if (args !! 1 == "-d") then
@@ -81,16 +81,18 @@ main = do
       putStrLn $ "Please specify a directory on which to apply the analysis\
                  \ followed by any number of file names to be excluded "
 
+{-
 testLHSvars flag = do
     files <- readParseSrcDir "experiment" []
     putStrLn $ show $ map (applyLHSvars flag) files
+
 
 applyLHSvars :: Bool -> (Filename, SourceText, F.ProgramFile A) -> [F.Name]
 applyLHSvars flag (filename, source, pf) =
     if flag then FA.allLhsVars pf' else FA.allLhsVarsRobust pf' 
   where
     pf' = FA.initAnalysis $ pf
-
+-}
 
 applyAnalysisToDir :: String -> Bool -> [String] -> IO ()
 applyAnalysisToDir dir debug excludes = do

### 2017-04-01 oopsla17/lattice-model.tex
type setting fix
Fix projection signature

@@
 
-\begin{defn}{(\zinf)}
-  We define the set \zinf{} as $\mathbb{Z}$ extended with $\infty$ and
-  $-\infty$. For any $n$ in \zinf{}, we have $-\infty \leq n \leq \infty$. The
-  resulting set is a total order with top and bottom elements.
+\begin{defn}{(\emph{Points})}
+  We define the set \zinf{} of (one-dimensional) \emph{points} as
+  $\mathbb{Z}$ extended with $\infty$ and $-\infty$. For any $n$ in
+  \zinf{}, we have $-\infty \leq n \leq \infty$. The resulting set is
+  a total order with top and bottom elements.
 \end{defn}
 
-\begin{defn}{(\emph{Offset space})}
-  \emph{Offset space} is a subset of \zinf{} members. It is used to capture
-  offsets from induction variables.
-\end{defn}
-
-\begin{defn}{(\emph{\zinf{} interval})}
-  A \emph{\zinf{} interval} is of the form \interv{a}{b}{c}, where $a$ and $b$
+\begin{defn}{(\emph{Point interval})}
+  A \emph{point interval} is a closed interval $\zinf$ which may
+  contain a ``hole'' at the origin, written \interv{a}{b}{c} where $a$ and $b$
   are drawn from \zinf{} with $a \leq 0 \leq b$, $a$ is $-\infty$ \emph{iff} $b$
-  is $\infty$, and $c$ is drawn from $\{ \mathsf{true}, \mathsf{false} \}$. We
+  is $\infty$, and $c$ is drawn from $\mathbb{B} = \{ \mathsf{true}, \mathsf{false} \}$. We
   define it as follows:
+\marginpar{We might add the restriction to $[-\infty, \infty]$ later,
+  but okay for now.}
 %
@@
 
-\begin{defn}{(\emph{Offsets vector})}
-  An $N$ dimensional \emph{offsets vector} is a Cartesian product of offset
-  spaces denoted with $\textit{offs}_1 \times \cdots \times \textit{offs}_N$.
-  The function $\pi_i : \zinf{}^{\mkern-19mu{N}} \to \zinf{}$ retrieves the
-  $i^{th}$ set of offsets.
+\begin{defn}{(\emph{Offsets and offset spaces})}
+  An \emph{offset} is a Cartesian product (vector) of $\zinf$ values representing an
+  $N$-dimensional point. The set of $N$-dimensional offsets is
+  written $\zinf^N$.  An $N$-dimensional \emph{offset space} is a set of
+  offsets, i.e. a member of $\mathcal{P}(\zinf^N)$.
 
-  The set of $N$ dimensional offsets vectors is
-  $\mathcal{P}{(\zinf{}^{\mkern-19mu{N}})}$ and they trivially form a
-  distributive lattice under the superset relation.
+An $N$-dimensional offset space can be \emph{projected} into the set
+of points in the $i^{\text{th}}$ dimension by $\pi_i : \mathcal{P}(\zinf{})^{\mkern-19mu{N}} \to
+  \mathcal{P}(\zinf{})$ where $i \in \mathsf{Fin}(N)$. We refer to this projected set as the \emph{$i^{th}$
+    component of an offset space}. 
 \end{defn}
 
-\begin{defn}{(\emph{Point offsets vector})}
-  A \emph{point offsets vector}, \textit{poff}, is a an offsets vector such that
-  for all indices $i$ $\pi_i(\textit{poff})$, we either have a singleton set or
-  the infinite interval $\interv{-\infty}{\infty}{\mathsf{true}}$.
+\begin{proposition}
+The set of $N$-dimensional offset spaces
+$\mathcal{P}{(\zinf{}^{\mkern-19mu{N}})}$ trivially forms a
+  distributive lattice under the superset relation.
+\end{proposition}
+
+\begin{remark}
+Note that $\mathcal{P}(\zinf^N)$ is isomorphic to a sets of point
+vectors $\mathcal{P}(\zinf)^N$. Therefore, projection $\pi$ on 
+offset spaces is defined by lifting normal Cartesian product
+projection to sets.
+\end{remark}
+
+%\begin{defn}{(\emph{Offsets vector}, $\mathcal{P}(\zinf^N)$)}
+%  An $N$ dimensional \emph{offset space vector} is a Cartesian product of offset
+%  spaces denoted with $\textit{offs}_1 \times \cdots \times
+%  \textit{offs}_N$.
+%  The function $\pi_i : \mathcal{P}(\zinf{})^{\mkern-19mu{N}} \to
+%  \mathcal{P}(\zinf{})$ retrieves the $i^{th}$ offset space.%
+%
+%  The set of $N$ dimensional offsets vectors is
+%  $\mathcal{P}{(\zinf{}^{\mkern-19mu{N}})}$ and they trivially form a
+%  distributive lattice under the superset relation.
+%\end{defn}
+
+\begin{defn}{(\emph{Subscript space})}
+  An $N$-dimensional offset space $\textit{off}$ is a \emph{subscript space} if it
+  satisfies:
+%
+\begin{equation*}
+\forall i \in \textsf{Fin}(N) . \quad 
+\pi_i (\textit{off}) = \{p\} \wedge p \in \mathbb{Z} 
+\;\;
+\vee
+\;\;
+\pi_i (\textit{off}) = \interv{-\infty}{\infty}{\mathsf{true}}
+\end{equation*}
+%
+That is, the $i^{th}$ component of the space is either a singleton 
+in $\mathbb{Z}$ or the infinite interval.
+
+Subscript spaces model array subscript expressions in our source
+langauge where each subscripted expression is either a 
+constant translation of an induction variable within a loop or some
+other expression which is constant with respect to the loop e.g. \texttt{a(i + 1, 0)}.
 \end{defn}
@@
 \begin{lem}{}\label{lem:vector-intersect}
-  Intersection distributes over offsets vectors. If $\mathit{off}$ and
-  $\mathit{off'}$ are $N$ dimensional ofssets vectors, then
+  Intersection distributes over offset spaces. If $\mathit{off}$ and
+  $\mathit{off'}$ are $N$-dimensional offset spaces, then
 %
@@
 \begin{lem}{}\label{lem:vector-union}
-  Let $\mathit{off}$ and $\mathit{off'}$ be two $N$ dimensional offsets vectors
-  such that $\pi_i(\mathit{off})$ is equal to $\pi_i(\mathit{off'})$ for all $1
+  Let $\mathit{off}$ and $\mathit{off'}$ be two $N$ dimensional offset
+  spaces such that $\pi_i(\mathit{off})$ is equal to $\pi_i(\mathit{off'})$ for all $1
   \leq i \leq N$ except possibly at $k$, then we have the following:
@@
 
-\begin{defn}{(\emph{Vector interval})}
-  A vector interval is an $N$ dimensional object which has \zinf{} intervals at
-  each dimension. A vector interval is also an ofssets vector.
-
-  The set of all $N$ dimensional vector intervals is $\textit{Interval}^N$.
+\begin{defn}{(\emph{Offset interval})}
+An offset interval is a finite Cartesian product of point intervals.
+We referer to an interval vector of size $N$ as being $N$-dimensional, and
+denoted by the set $\textit{Interval}^N$.
 \end{defn}
@@
 \begin{defn}{(\emph{Region})} 
-  \region{N} is the set of $N$ dimensional vector intervals defined as
+  A region \region{N} is a set of $N$-dimensional offset intervals defined as
   the smallest set satisfying the following:
@@
 
-Here helper function $\textit{peel} : \mathsf{Approx} a \to a$ removes the
+Here helper function $\textit{peel} : \mathsf{Approx} \; a \to a$ removes the
 approximation label.
@@
 
-\begin{defn}{(Extended origin)}
-  In the $N$ dimensional region space, ${(\interv{-1}{1}{})}^N$ is said to be
-  the extended origin denoted $\textit{EO}_N$.
+\begin{defn}{(Origin)}
+In $N$-dimensions, the \emph{origin} $\mathbf{0}_N$ is the singleton set
+$\{0^N\}$, or equivalently the vector of unit intervals $[0, 0]^N$.
 \end{defn}
 
-\begin{defn}{(Origin distance)}
-  We measure the \emph{origin distance} of a point offsets vector from the
-  extended origin as the minimum number of translations by $\pm 1$ unit in one
-  dimension at a time to reach any point in the extended origin. Since point
+\begin{defn}{(Extended origin)}
+In $N$-dimensions, the \emph{extended origin} is the
+interval vector $[-1, 1]^N$ which is a subspace of \region{N}. 
+\end{defn}
+
+\begin{defn}{(Extended origin distance)}
+  We measure the \emph{extended origin distance} of an offset from the
+  extended origin as the minimum number of translations by $\pm 1$ in one
+  dimension at a time, to reach an offset in the extended
+  origin.
+
+This notion extends to Since an 
   offsets vector may represent infinite number of points, we measure the

### 2017-04-01 oopsla17/lattice-model.tex
attempts at clarity

@@
 
-\begin{defn}{(\emph{Points})}
-  We define the set \zinf{} of (one-dimensional) \emph{points} as
+\begin{defn}
+  We define the set \zinf{} as
   $\mathbb{Z}$ extended with $\infty$ and $-\infty$. For any $n$ in
@@
 
-\begin{defn}{(\emph{Point interval})}
-  A \emph{point interval} is a closed interval $\zinf$ which may
+\begin{defn}
+  We define an extended notion of closed interval on $\zinf$ which may
   contain a ``hole'' at the origin, written \interv{a}{b}{c} where $a$ and $b$
@@
 
-\begin{defn}{(\emph{Offsets and offset spaces})}
-  An \emph{offset} is a Cartesian product (vector) of $\zinf$ values representing an
+\begin{defn}{(\emph{Points})}
+  A \emph{point} is a Cartesian product (vector) of $\zinf$ values representing an
   $N$-dimensional point. The set of $N$-dimensional offsets is
-  written $\zinf^N$.  An $N$-dimensional \emph{offset space} is a set of
-  offsets, i.e. a member of $\mathcal{P}(\zinf^N)$.
+  written $\zinf^N$. We denote sets of $N$-dimensionl points
+by the set $\mathcal{P}(\zinf^N)$.
 
-An $N$-dimensional offset space can be \emph{projected} into the set
-of points in the $i^{\text{th}}$ dimension by $\pi_i : \mathcal{P}(\zinf{})^{\mkern-19mu{N}} \to
-  \mathcal{P}(\zinf{})$ where $i \in \mathsf{Fin}(N)$. We refer to this projected set as the \emph{$i^{th}$
-    component of an offset space}. 
+A set of $N$-dimensional points \emph{projected} into the set
+of 1-dimensional points in the $i^{\text{th}}$ dimension by $\pi_i : \mathcal{P}(\zinf{})^{\mkern-19mu{N}} \to
+  \mathcal{P}(\zinf{})$ where $i \in \mathsf{Fin}(N)$. We refer to this projected set as the \emph{$i^{th}$ components}.
 \end{defn}
@@
 \begin{proposition}
-The set of $N$-dimensional offset spaces
-$\mathcal{P}{(\zinf{}^{\mkern-19mu{N}})}$ trivially forms a
+The set $\mathcal{P}{(\zinf{}^{{N}})}$ trivially forms a
   distributive lattice under the superset relation.
@@
 
-\begin{defn}{(\emph{Subscript space})}
-  An $N$-dimensional offset space $\textit{off}$ is a \emph{subscript space} if it
+\begin{defn}{(\emph{Subscript})}
+  A set of $N$-dimensional points $S$ is called a \emph{subscript} if it
   satisfies:
@@
 \forall i \in \textsf{Fin}(N) . \quad 
-\pi_i (\textit{off}) = \{p\} \wedge p \in \mathbb{Z} 
+\pi_i (S) = \{p\} \wedge p \in \mathbb{Z} 
 \;\;
@@
 \;\;
-\pi_i (\textit{off}) = \interv{-\infty}{\infty}{\mathsf{true}}
+\pi_i (X) = \mathbb{Z} = \interv{-\infty}{\infty}{\mathsf{true}}
 \end{equation*}
 %
-That is, the $i^{th}$ component of the space is either a singleton 
+That is, the $i^{th}$ component of the set is either a singleton 
 in $\mathbb{Z}$ or the infinite interval.
 
+\todo{We can move this explanation.}
 Subscript spaces model array subscript expressions in our source
@@
 \begin{lem}{}\label{lem:vector-intersect}
-  Intersection distributes over offset spaces. If $\mathit{off}$ and
-  $\mathit{off'}$ are $N$-dimensional offset spaces, then
+  Intersection distributes over sets of points. If $\mathit{off}$ and
+  $\mathit{off'}$ are sets of $N$-dimensional points, then
 %
@@
 \begin{lem}{}\label{lem:vector-union}
-  Let $\mathit{off}$ and $\mathit{off'}$ be two $N$ dimensional offset
-  spaces such that $\pi_i(\mathit{off})$ is equal to $\pi_i(\mathit{off'})$ for all $1
+  Let $\mathit{off}$ and $\mathit{off'}$ be two sets of 
+  points such that $\pi_i(\mathit{off})$ is equal to $\pi_i(\mathit{off'})$ for all $1
   \leq i \leq N$ except possibly at $k$, then we have the following:
@@
 
-\begin{defn}{(\emph{Offset interval})}
-An offset interval is a finite Cartesian product of point intervals.
-We referer to an interval vector of size $N$ as being $N$-dimensional, and
-denoted by the set $\textit{Interval}^N$.
+\begin{defn}{(\emph{Point interval})}
+An point interval is a finite Cartesian product of intervals on \zinf, 
+denoted by the set $\textit{Interval}^N$ for a product of $N$ intervals.
 \end{defn}
@@
 \begin{defn}{(\emph{Region})} 
-  A region \region{N} is a set of $N$-dimensional offset intervals defined as
+  A region \region{N} is a set of $N$-dimensional point intervals defined as
   the smallest set satisfying the following:

### 2017-04-01 oopsla17/lattice-model.tex
fix

@@
 \begin{defn}{(\emph{Points})}
-  A \emph{point} is a Cartesian product (vector) of $\zinf$ values representing an
-  $N$-dimensional point. The set of $N$-dimensional offsets is
-  written $\zinf^N$. We denote sets of $N$-dimensionl points
-by the set $\mathcal{P}(\zinf^N)$.
+  A $N$-dimenaional \emph{point} is a finite Cartesian product
+  (vector) of $N$ values drawn from $\zinf$.
 
-A set of $N$-dimensional points \emph{projected} into the set
-of 1-dimensional points in the $i^{\text{th}}$ dimension by $\pi_i : \mathcal{P}(\zinf{})^{\mkern-19mu{N}} \to
-  \mathcal{P}(\zinf{})$ where $i \in \mathsf{Fin}(N)$. We refer to this projected set as the \emph{$i^{th}$ components}.
+  The set of $N$-dimensional points is written $\zinf^N$ and the set
+  of sets of $N$-dimensionl points is written $\mathcal{P}(\zinf^N)$.
+
+A set of $N$-dimensional points can be \emph{projected} to the set of
+1-dimensional points in the $i^{\text{th}}$ dimension by
+$\pi_i : \mathcal{P}(\zinf{})^{\mkern-19mu{N}} \to
+\mathcal{P}(\zinf{})$
+where $i \in \mathsf{Fin}(N)$. We refer to this projected set as the
+\emph{$i^{th}$ components}.
 \end{defn}

### 2017-04-01 oopsla17/lattice-model.tex
prop environment for proposition 1; change to isomorphic type for projections

@@
 1-dimensional points in the $i^{\text{th}}$ dimension by
-$\pi_i : \mathcal{P}(\zinf{})^{\mkern-19mu{N}} \to
+$\pi_i : \mathcal{P}(\zinf{}^{\mkern0mu{N}}) \to
 \mathcal{P}(\zinf{})$
@@
 
-\begin{proposition}
+\begin{prop}
 The set $\mathcal{P}{(\zinf{}^{{N}})}$ trivially forms a
   distributive lattice under the superset relation.
-\end{proposition}
+\end{prop}
 

### 2017-04-01 oopsla17/lattice-model.tex
various comments/questions in TODOs

@@
 \;\;
-\pi_i (X) = \mathbb{Z} = \interv{-\infty}{\infty}{\mathsf{true}}
+\pi_i (S) = \interv{-\infty}{\infty}{\mathsf{true}}
 \end{equation*}
@@
   partially ordered with subset or equal to relation are distributive lattices.
-  \region{N} is a subset of $\mathcal{P}{(\zinf{}^{\mkern-19mu{N}})}$ since all
+  \region{N} is a subset of $\mathcal{P}{(\zinf{}^{\mkern0mu{N}})}$ since all
   vector intervals are subsets of this power set and union and intersection
@@
   Therefore, \region{N} is a distributive sublattice of
-  $(\mathcal{P}{(\zinf{}^{\mkern-19mu{N}})}, \subseteq)$.
+  $(\mathcal{P}{(\zinf{}^{\mkern0mu{N}})}, \subseteq)$.
 \end{proof}
@@
 In $N$-dimensions, the \emph{extended origin} is the
-interval vector $[-1, 1]^N$ which is a subspace of \region{N}. 
+interval $[-1, 1]^N$ which is a subset of \region{N}. 
 \end{defn}
@@
 \begin{defn}{(Extended origin distance)}
-  We measure the \emph{extended origin distance} of an offset from the
+  We measure the \emph{extended origin distance} of a point from the
   extended origin as the minimum number of translations by $\pm 1$ in one
-  dimension at a time, to reach an offset in the extended
+  dimension at a time, to reach a point in the extended
   origin.
 
-This notion extends to Since an 
-  offsets vector may represent infinite number of points, we measure the
+This notion extends to sets of points. Since an 
+   there may be an infinite number of points, we measure the
   distance from the closest one.
+\todo{Mistral to check: I may have broken this notion}
 \end{defn}
@@
 \begin{prop}{}
-  The closed form of origin distance of an $N$ dimensional point offsets vector
-  $p$ is the following:
+  The closed form of extend origin distance for a set of $N$-dimensional
+  points $P$ is written as $|P|$ and is defined:
 %
   \begin{align*}
-    \textit{dist}_i(p) & = \begin{cases}
-      0 & \pi_i(p) = \interv{-\infty}{\infty}{} \\
-      |\mathsf{a}| - 1 & \pi_i(p) = \{\mathsf{a}\} \\
-    \end{cases} &
-%
-    |p| & = \sum_{i=1}^{N} dist_i(p)
+|P| & = \sum_{i=1}^{N} dist_i(P)
+ & 
+    \textit{dist}_i(P) & = \begin{cases}
+      0 & \pi_i(P) = \interv{-\infty}{\infty}{} \\
+      \mathsf{a} - 1 & \pi_i(P) = \{\mathsf{a}\} \\
+    \end{cases}
   \end{align*}
@@
 \begin{proof}
+\todo{Not sure what we are going to prove here? That it is a metric
+space or something?}
 \end{proof}
 
-\begin{defn}{(Close neighbourhood)}
-  A point offsets vector $p'$ is said to be a close neighbour of point offsets
-  vector $p$ if for every index $i$ except $j$, $\pi_i(p)$ agrees with
-  $\pi_i(p')$ but at $j$, $\pi_j(p')$ is such that $|p'| + 1 = |p|$.
+\todo{We used to call the below property contiguity, and I think this is a natural
+notion- does ``close neighbour'' have some standard use or is
+it okay to call this contiguity?}
+\begin{defn}{(Close neighbour)}
+ A set of points $P$ is said to be a \emph{close neighbour at index $j$} of a set
+   $P'$ if for every index $i$ except $j$, $\pi_i(P) = \pi_i(P')$ 
+ but at $j$ then $\pi_j(P) \neq \pi_i(P')$ and
+ $\textit{dist}_j(P) = \textit{dist}_i(P') + 1$.
 
+\todo{Not sure how to rewrite the following as I have a bit of hard
+time parsing it.}
   This necessarily means point vectors within the extended origin and the
@@
 \begin{prop}
-  The unique close neighbour at index $i$ of an $N$ dimensional point offsets
-  vector, if it exists, is given by the following partial function:
+  The unique close neighbour at index $i$ of an $N$-dimensional set of
+  points, if it exists, is given by the following partial function:
 %
@@
 \begin{proof}
+\todo{Hm- I guess uniqueness follows from this being a function rather
+  than a relation? Seems fine no problem to prove- }
 \end{proof}
 
+\todo{On the following, add some intutition first- what is this and
+  what is it for?}
 \begin{defn}{(Exact set)}
@@
 \end{defn}
+\todo{Ah, I think I am starting to understand, but at 
+$\mathit{ES}_{P,1}$ then we end up saying there exists a 
+$p'$ in $\textit{ES}_{P,0}$, but we don't calculate that set...
+so how does this end?}
 

### 2017-04-03 oopsla17/lattice-model.tex
Add base case to Exact Set
Elaborate how subset and superset relations are computed

@@
 \begin{align*}
-  \textit{consistent}(\textit{ix}, \textit{model}) & = \begin{cases}
-    \mathsf{false} & \textit{model} = \mathsf{once}(x) \wedge ix = \mathsf{mult}(y) \\
-    \textit{consistent'}(\textit{peel}(ix), \textit{peel}(model)) & \textit{otherwise}
+  \mathit{consistent}(\mathit{ix}, \mathit{model}) & = \begin{cases}
+    \mathsf{false} & \mathit{model} = \mathsf{once}(x) \wedge ix = \mathsf{mult}(y) \\
+    \mathit{consistent'}(\mathit{peel}(ix), \mathit{peel}(model)) & \textit{otherwise}
   \end{cases} \\
 %
-  \textit{consistent'}(\textit{ix}, \textit{model}) & = \begin{cases}
-    m = ix & \textit{model} = \mathsf{exact}(m) \\
-    m \supseteq ix & \textit{model} = \mathsf{upper}(m) \\
-    m \subseteq ix & \textit{model} = \mathsf{lower}(m)
+  \mathit{consistent'}(\mathit{ix}, \mathit{model}) & = \begin{cases}
+    m = \mathit{ix} & \mathit{model} = \mathsf{exact}(m) \\
+    m \supseteq \mathit{ix} & \mathit{model} = \mathsf{upper}(m) \\
+    m \subseteq \mathit{ix} & \mathit{model} = \mathsf{lower}(m)
   \end{cases}
@@
 
-Here helper function $\textit{peel} : \mathsf{Approx} \; a \to a$ removes the
+Here helper function $\mathit{peel} : \mathsf{Approx} \; a \to a$ removes the
 approximation label.
 
-\todo{Here we simply show \textit{consistency'} as set equality/subset/superset,
-but since sets are infinite we might need to elaborate based on representation.}
+Due to use of infinite intervals, the model and index sets are potentially
+infinite meaning simple procedures used to compute equality, subset, and
+superset relations for finite sets are no avail. The subset and superset
+relations are handled by computing the intersection of $m$ and $\mathit{ix}$
+using \Cref{lem:vector-intersect}, then checking equality with $m$ and
+$\mathit{ix}$ respectively.
+
+\todo{Here we simply show \textit{consistency'} as set equality, but since sets
+are infinite we might need to elaborate based on representation.}
 
@@
 %
-  \begin{equation*}
-    \mathit{ES}_{P,d} =
+  \begin{align*}
+    \mathit{ES}_{P,0} & = \{ p \in P \mid \forall \i.\ \pi_i(p) \in \interv{-1}{1}{} \} \\
+    \mathit{ES}_{P,d} & =
       \{ p \in P \mid \forall i.\
-         p_i \notin \{-1,0,1\} \implies
+      p_i \notin [-1,1] \implies
          \exists p' \in \mathit{ES}_{P,d-1}.\ p' =
          \mathit{close\_neighbour}(p,i) \}
-  \end{equation*}
+  \end{align*}
 %
@@
 \end{defn}
-\todo{Ah, I think I am starting to understand, but at 
-$\mathit{ES}_{P,1}$ then we end up saying there exists a 
-$p'$ in $\textit{ES}_{P,0}$, but we don't calculate that set...
-so how does this end?}
 

### 2017-04-03 oopsla17/lattice-model.tex
Remove trailing whitespace
fix mixups in lattice model subsection

@@
 
-\begin{defn}{(\emph{Points})}
-  A $N$-dimenaional \emph{point} is a finite Cartesian product
-  (vector) of $N$ values drawn from $\zinf$.
+\begin{defn}{(\emph{Offset subspace})}
+  An $N$-dimenaional \emph{offset subspace} is a finite Cartesian product of $N$
+  subsets of $\zinf{}$. The set of $N$-dimensional offset subspaces is written
+  ${\mathcal{P}(\zinf{})}^N$.
 
-  The set of $N$-dimensional points is written $\zinf^N$ and the set
-  of sets of $N$-dimensionl points is written $\mathcal{P}(\zinf^N)$.
-
-A set of $N$-dimensional points can be \emph{projected} to the set of
-1-dimensional points in the $i^{\text{th}}$ dimension by
-$\pi_i : \mathcal{P}(\zinf{}^{\mkern0mu{N}}) \to
-\mathcal{P}(\zinf{})$
-where $i \in \mathsf{Fin}(N)$. We refer to this projected set as the
-\emph{$i^{th}$ components}.
+  A offset spaces can be \emph{projected} in the $i^{\textit{th}}$ dimension by
+  $\pi_i : {\mathcal{P}(\zinf{})}^N \to \mathcal{P}(\zinf{})$ where $i \in
+  \mathsf{Fin}(N)$.
 \end{defn}
 
-\begin{prop}
-The set $\mathcal{P}{(\zinf{}^{{N}})}$ trivially forms a
-  distributive lattice under the superset relation.
-\end{prop}
-
-\begin{remark}
-Note that $\mathcal{P}(\zinf^N)$ is isomorphic to a sets of point
-vectors $\mathcal{P}(\zinf)^N$. Therefore, projection $\pi$ on 
-offset spaces is defined by lifting normal Cartesian product
-projection to sets.
-\end{remark}
-
-%\begin{defn}{(\emph{Offsets vector}, $\mathcal{P}(\zinf^N)$)}
-%  An $N$ dimensional \emph{offset space vector} is a Cartesian product of offset
-%  spaces denoted with $\textit{offs}_1 \times \cdots \times
-%  \textit{offs}_N$.
-%  The function $\pi_i : \mathcal{P}(\zinf{})^{\mkern-19mu{N}} \to
-%  \mathcal{P}(\zinf{})$ retrieves the $i^{th}$ offset space.%
-%
-%  The set of $N$ dimensional offsets vectors is
-%  $\mathcal{P}{(\zinf{}^{\mkern-19mu{N}})}$ and they trivially form a
-%  distributive lattice under the superset relation.
-%\end{defn}
-
 \begin{defn}{(\emph{Subscript})}
-  A set of $N$-dimensional points $S$ is called a \emph{subscript} if it
+  An $N$-dimensional offset subspace $S$ is called a \emph{subscript} if it
   satisfies:
@@
   \begin{equation*}
-\forall i \in \textsf{Fin}(N) . \quad 
+    \forall i \in \textsf{Fin}(N).\
       \pi_i (S) = \{p\} \wedge p \in \mathbb{Z}
-\;\;
       \vee
-\;\;
       \pi_i (S) = \interv{-\infty}{\infty}{\mathsf{true}}
@@
 
-\todo{We can move this explanation.}
-Subscript spaces model array subscript expressions in our source
-langauge where each subscripted expression is either a 
-constant translation of an induction variable within a loop or some
-other expression which is constant with respect to the loop e.g. \texttt{a(i + 1, 0)}.
+  \todo{%
+    We can move this explanation.
+
+    Subscript spaces model array subscript expressions in our source langauge
+    where each subscripted expression is either a constant translation of an
+    induction variable within a loop or some other expression which is constant
+    with respect to the loop e.g. \texttt{a(i+1, 0)}.
+  }
 \end{defn}
@@
 
-\begin{defn}{(\emph{Point interval})}
-An point interval is a finite Cartesian product of intervals on \zinf, 
-denoted by the set $\textit{Interval}^N$ for a product of $N$ intervals.
+\begin{defn}{(\emph{Interval subspace})}
+  An interval subspace is a finite Cartesian product of intervals on \zinf{},
+  denoted by the set $\textit{Interval}^N$ for a product of $N$ intervals. An
+  interval subspace is an offset subspace.
 \end{defn}
@@
   Any sublattice of a distributive lattice is a distributive lattice. Power sets
-  partially ordered with subset or equal to relation are distributive lattices.
-  \region{N} is a subset of $\mathcal{P}{(\zinf{}^{\mkern0mu{N}})}$ since all
-  vector intervals are subsets of this power set and union and intersection
-  cannot add an element that is not already in the powerset.  Inductive
-  definition also ensures that \region{N} is closed under these operations.
-  Therefore, \region{N} is a distributive sublattice of
-  $(\mathcal{P}{(\zinf{}^{\mkern0mu{N}})}, \subseteq)$.
+  partially ordered with subset relation are distributive lattices. \region{N}
+  is a subset of $\mathcal{P}{(\zinf{}^N)}$ since all interval subspaces are
+  subsets of this power set and; union and intersection cannot add an element
+  that is not already in the powerset. Inductive definition also ensures that
+  \region{N} is closed under these operations. Therefore, \region{N} is a
+  distributive sublattice of $(\mathcal{P}{(\zinf{}^N)}, \subseteq)$.
 \end{proof}

### 2017-04-03 oopsla17/lattice-model.tex
Fix markup in inference

@@
   In $N$-dimensions, the \emph{origin} $\mathbf{0}_N$ is the singleton set
-$\{0^N\}$, or equivalently the vector of unit intervals $[0, 0]^N$.
+  $\{0^N\}$, or equivalently the vector of unit intervals $\interv{0}{0}{N}$.
 \end{defn}
@@
 \begin{defn}{(Extended origin)}
-In $N$-dimensions, the \emph{extended origin} is the
-interval $[-1, 1]^N$ which is a subset of \region{N}.
+  In $N$-dimensions, the \emph{extended origin} is the interval
+  $\interv{-1}{1}{N}$ which is a subset of \region{N}.
 \end{defn}
@@
 
-This notion extends to sets of points. Since an
-   there may be an infinite number of points, we measure the
-  distance from the closest one.
+  This notion extends to sets of points. Since an there may be an infinite
+  number of points, we measure the distance from the closest one.
+
   \todo{Mistral to check: I may have broken this notion}
@@
 \begin{proof}
-\todo{Not sure what we are going to prove here? That it is a metric
-space or something?}
+  \todo{%
+    Not sure what we are going to prove here? That it is a metric space or
+    something?
+  }
 \end{proof}
 
-\todo{We used to call the below property contiguity, and I think this is a natural
-notion- does ``close neighbour'' have some standard use or is
-it okay to call this contiguity?}
+\todo{%
+  We used to call the below property contiguity, and I think this is a natural
+  notion- does ``close neighbour'' have some standard use or is it okay to call
+  this contiguity?
+}
 \begin{defn}{(Close neighbour)}
- A set of points $P$ is said to be a \emph{close neighbour at index $j$} of a set
-   $P'$ if for every index $i$ except $j$, $\pi_i(P) = \pi_i(P')$
- but at $j$ then $\pi_j(P) \neq \pi_i(P')$ and
- $\textit{dist}_j(P) = \textit{dist}_i(P') + 1$.
+  A set of points $P$ is said to be a \emph{close neighbour at index $j$} of a
+  set $P'$ if for every index $i$ except $j$, $\pi_i(P) = \pi_i(P')$ but at $j$
+  then $\pi_j(P) \neq \pi_i(P')$ and $\textit{dist}_j(P) = \textit{dist}_i(P') +
+  1$.
 
-\todo{Not sure how to rewrite the following as I have a bit of hard
-time parsing it.}
+\todo{%
+  Not sure how to rewrite the following as I have a bit of hard time parsing it.
+}
+
   This necessarily means point vectors within the extended origin and the

### 2017-04-03 oopsla17/lattice-model.tex
Changes to origin definitions

@@
 
+\todo{%
+  Definition of origin is not needed.
+
   \begin{defn}{(Origin)}
@@
   \end{defn}
+}
 
@@
   In $N$-dimensions, the \emph{extended origin} is the interval
-  $\interv{-1}{1}{N}$ which is a subset of \region{N}.
+  $\interv{-1}{1}{N}$ contained in \region{N}.
 \end{defn}

### 2017-04-03 oopsla17/lattice-model.tex
Reintroduce deleted absolute value
Fix terminology

@@
 \begin{prop}{}
-  The closed form of extend origin distance for a set of $N$-dimensional
-  points $P$ is written as $|P|$ and is defined:
+  The closed form of extend origin distance for an $N$-dimensional subscript $S$
+  is denoted with $|S|$ and is defined as:
 %
   \begin{align*}
-    |P| & = \sum_{i=1}^{N} dist_i(P)
+    |S| & = \sum_{i=1}^{N} dist_i(S)
     &
-    \textit{dist}_i(P) & = \begin{cases}
-      0 & \pi_i(P) = \interv{-\infty}{\infty}{} \\
-      \mathsf{a} - 1 & \pi_i(P) = \{\mathsf{a}\} \\
+    \textit{dist}_i(S) & = \begin{cases}
+      0 & \pi_i(S) = \interv{-\infty}{\infty}{} \\
+      |\mathsf{a}| - 1 & \pi_i(S) = \{\mathsf{a}\} \\
     \end{cases}
@@
     something?
+
+    Response: The definition says extended origin distance is the smallest
+    number of translations to reach the extended origin, we need to prove that
+    this formula indeed gives that.
   }
@@
   this contiguity?
+
+  Response: how about adjecency?
 }
-\begin{defn}{(Close neighbour)}
-  A set of points $P$ is said to be a \emph{close neighbour at index $j$} of a
-  set $P'$ if for every index $i$ except $j$, $\pi_i(P) = \pi_i(P')$ but at $j$
-  then $\pi_j(P) \neq \pi_i(P')$ and $\textit{dist}_j(P) = \textit{dist}_i(P') +
-  1$.
+\begin{defn}{(Adjacent)}
+  A subscript $S$ is said to be \emph{adjacent at index $j$} of another
+  subscript $S'$ if for every index $i$ except $j$, $\pi_i(S) = \pi_i(S')$ but
+  at $j$ then $\pi_j(S) \neq \pi_i(S')$ such that $|S| = |S'| + 1$.
 
@@
 \begin{prop}
-  The unique close neighbour at index $i$ of an $N$-dimensional set of
-  points, if it exists, is given by the following partial function:
+  The unique adjacent at index $i$ of a subscript, if it exists, is given by the
+  following partial function:
 %
   \begin{align*}
-    \mathit{close\_neighbour}(p,i) & = \prod^{N}_{j=1} \mathit{shift}(p,i,j) &
+    \mathit{adjacent}(S,i) & = \prod^{N}_{j=1} \mathit{shift}(S,i,j) &
 %
-    \mathit{shift}(p,i,j) & =
+    \mathit{shift}(S,i,j) & =
       \begin{cases}
-        \{a - 1\} & i = j \wedge \pi_i(p) = \{\mathsf{a}\} \wedge \mathsf{a} > 0\\
-        \{a + 1\} & i = j \wedge \pi_i(p) = \{\mathsf{a}\} \wedge \mathsf{a} < 0\\
-        \pi_j(p)  & otherwise
+        \{a - 1\} & i = j \wedge \pi_i(S) = \{\mathsf{a}\} \wedge \mathsf{a} > 0\\
+        \{a + 1\} & i = j \wedge \pi_i(S) = \{\mathsf{a}\} \wedge \mathsf{a} < 0\\
+        \pi_j(S)  & otherwise
       \end{cases}

### 2017-04-03 oopsla17/lattice-model.tex
Remove todo about close neighbour
Fix terminology in extended origin

@@
 \begin{defn}{(Extended origin distance)}
-  We measure the \emph{extended origin distance} of a point from the
-  extended origin as the minimum number of translations by $\pm 1$ in one
-  dimension at a time, to reach a point in the extended
-  origin.
+  We measure the \emph{extended origin distance} of a point from the extended
+  origin as the minimum number of translations by $\pm 1$ in one dimension at a
+  time, to reach a point in the extended origin.
 
-  This notion extends to sets of points. Since an there may be an infinite
-  number of points, we measure the distance from the closest one.
-
-  \todo{Mistral to check: I may have broken this notion}
+  This notion extends to subscripts. Since there may be an infinite number of
+  points, we measure the distance from the closest one.
 \end{defn}
@@
 
-\todo{%
-  We used to call the below property contiguity, and I think this is a natural
-  notion- does ``close neighbour'' have some standard use or is it okay to call
-  this contiguity?
-
-  Response: how about adjecency?
-}
 \begin{defn}{(Adjacent)}

### 2017-04-03 oopsla17/lattice-model.tex
Remove definition of unused origin
Fix extended origin distance

@@
 
-\begin{defn}{(Extended origin)}
-  In $N$-dimensions, the \emph{extended origin} is the interval
-  $\interv{-1}{1}{N}$ contained in \region{N}.
-\end{defn}
-
 \begin{defn}{(Extended origin distance)}
@@
   The closed form of extend origin distance for an $N$-dimensional subscript $S$
-  is denoted with $|S|$ and is defined as:
+  is denoted with $\mathit{dist}(S)$ and is defined as:
 %
   \begin{align*}
-    |S| & = \sum_{i=1}^{N} dist_i(S)
+    \mathit{dist}(S) & = \sum_{i=1}^{N} \mathit{dist}_i(S)
     &
-    \textit{dist}_i(S) & = \begin{cases}
+    \mathit{dist}_i(S) & = \begin{cases}
       0 & \pi_i(S) = \interv{-\infty}{\infty}{} \\
@@
 \begin{proof}
-  \todo{%
-    Not sure what we are going to prove here? That it is a metric space or
-    something?
-
-    Response: The definition says extended origin distance is the smallest
-    number of translations to reach the extended origin, we need to prove that
-    this formula indeed gives that.
-  }
+  Now we prove that function $\mathit{dist}$ indeed gives the lowest number of
+  unit translations to reach the extended origin.
 \end{proof}

### 2017-04-03 oopsla17/lattice-model.tex
Fix reference to adjacency and index

@@
   subscript $S'$ if for every index $i$ except $j$, $\pi_i(S) = \pi_i(S')$ but
-  at $j$ then $\pi_j(S) \neq \pi_i(S')$ such that $|S| = |S'| + 1$.
-
-\todo{%
-  Not sure how to rewrite the following as I have a bit of hard time parsing it.
-}
-
-  This necessarily means point vectors within the extended origin and the
-  degenerate point offsets vector with \interv{-\infty}{\infty}{} does not have
-  close neighbours.
+  at $j$ then $\pi_j(S) \neq \pi_j(S')$ such that $|S| = |S'| + 1$.
 \end{defn}
@@
          \exists p' \in \mathit{ES}_{P,d-1}.\ p' =
-         \mathit{close\_neighbour}(p,i) \}
+         \mathit{adjacent}(p,i) \}
   \end{align*}

### 2017-04-03 oopsla17/lattice-model.tex
Remove origin redfinition again
Connecting lemma

@@
 
-\todo{%
-  Definition of origin is not needed.
-
-  \begin{defn}{(Origin)}
-    In $N$-dimensions, the \emph{origin} $\mathbf{0}_N$ is the singleton set
-    $\{0^N\}$, or equivalently the vector of unit intervals $\interv{0}{0}{N}$.
-  \end{defn}
-}
-
 \begin{defn}{(Extended origin distance)}
@@
 
-\begin{lem}
+\begin{lem}{(\emph{Connecting})}
   The following three statements are equivalent for a set of points $P \subseteq

### 2017-04-03 oopsla17/lattice-model.tex
Improve soundness theorem
Improve interval def

@@
 \begin{defn}
-  We define an extended notion of closed interval on $\zinf$ which may
-  contain a ``hole'' at the origin, written \interv{a}{b}{c} where $a$ and $b$
-  are drawn from \zinf{} with $a \leq 0 \leq b$, $a$ is $-\infty$ \emph{iff} $b$
-  is $\infty$, and $c$ is drawn from $\mathbb{B} = \{ \mathsf{true}, \mathsf{false} \}$. We
-  define it as follows:
-\marginpar{We might add the restriction to $[-\infty, \infty]$ later,
-  but okay for now.}
+  We define an extended notion of closed interval on $\zinf{}$ which may contain
+  a \emph{hole} at the origin, written \interv{a}{b}{c} where $a$ and $b$ are
+  drawn from \zinf{} with $a \leq 0 \leq b$, $a$ is $-\infty$ \emph{iff} $b$ is
+  $\infty$, and $c$ is drawn from $\mathbb{B} = \{ \mathsf{true}, \mathsf{false}
+  \}$. We define it as follows:
 %
@@
     \interv{a}{b}{c} \triangleq
-      \{ n \;|\; a \leq n \leq b \wedge (\neg c \implies n \neq 0) \}
+      \{ n \mid a \leq n \leq b \wedge (\neg c \implies n \neq 0) \}
   \end{equation*}
@@
   \begin{equation*}
-    \mathit{consistent}(\mathsf{mult}(P),\mathsf{mult}(\mathit{infer}(P)))
+    \forall P. \mathit{peel}(\mathit{infer}(P)) = m \implies
+      \mathit{consistent'}(P,m)
   \end{equation*}

### 2017-04-03 oopsla17/lattice-model.tex
typo
Get rid of predicate Fin

@@
 \begin{defn}{(\emph{Offset subspace})}
-  An $N$-dimenaional \emph{offset subspace} is a finite Cartesian product of $N$
+  An $N$-dimensional \emph{offset subspace} is a finite Cartesian product of $N$
   subsets of $\zinf{}$. The set of $N$-dimensional offset subspaces is written
@@
   A offset spaces can be \emph{projected} in the $i^{\textit{th}}$ dimension by
-  $\pi_i : {\mathcal{P}(\zinf{})}^N \to \mathcal{P}(\zinf{})$ where $i \in
-  \mathsf{Fin}(N)$.
+  $\pi_i : {\mathcal{P}(\zinf{})}^N \to \mathcal{P}(\zinf{})$ where $1 \leq i
+  \leq N$.
 \end{defn}
@@
   \begin{equation*}
-    \forall i \in \textsf{Fin}(N).\
+    \forall i.\ 1 \leq i \leq N \implies
       \pi_i (S) = \{p\} \wedge p \in \mathbb{Z}

### 2017-04-05 oopsla17/analysis/Analyse.hs
towards deeper analysis of array usage in loops

@@ -65,12 +65,6 @@ main :: IO ()
 main = do
     args <- getArgs
     if (length args == 1) then
-      {- if (head args == "PI") then
-        testLHSvars True
-      else 
-      if (head args == "PI2") then
-        testLHSvars False
-      else -}
         applyAnalysisToDir (head args) False []
     else if (length args > 1) then
       if (args !! 1 == "-d") then
@@ -80,19 +74,6 @@ main = do
     else
       putStrLn $ "Please specify a directory on which to apply the analysis\
                  \ followed by any number of file names to be excluded "
-
-{-
-testLHSvars flag = do
-    files <- readParseSrcDir "experiment" []
-    putStrLn $ show $ map (applyLHSvars flag) files
-
-
-applyLHSvars :: Bool -> (Filename, SourceText, F.ProgramFile A) -> [F.Name]
-applyLHSvars flag (filename, source, pf) =
-    if flag then FA.allLhsVars pf' else FA.allLhsVarsRobust pf' 
-  where
-    pf' = FA.initAnalysis $ pf
--}
 
 applyAnalysisToDir :: String -> Bool -> [String] -> IO ()
 applyAnalysisToDir dir debug excludes = do
@@ -106,6 +87,18 @@ applyAnalysisToDir dir debug excludes = do
 
 type ForContigAndNonContig a = (a, a)
 
+-- Categories
+data Kind = Read | Stencil
+data Contig = NonContig | Contig
+
+type Cat = (Kind, Contig)
+
+isStencil (Stencil, _) = True
+isStencil _            = False
+
+isContig (_, Contig) = True
+isContig _ = False
+
 -- Results data type
 data Result = Result {
     numLines                   :: Int   -- Number of lines parsed
@@ -116,7 +109,38 @@ data Result = Result {
   , numLHSButNonNeighbourRHS   :: Int  -- Q1
   , numLHSButInconsistentIVRHS :: Int  -- Q2 where Q >= Q1 >= Q2
   ------------------------------------
-  , numStencils                :: Int   -- b
+  -- Special interest
+  ------------------------------------
+  , numSingNonContigStencils :: Int
+  , numStencilRelativisedRHS :: Int
+  ------------------------------------
+  , numOverall  :: M.Map Cat Int
+  , numLinear   :: M.Map Cat Int
+  , numNoOrigin :: M.Map Cat Int
[…]

### 2017-04-06 Analyse.hs
capture more data about different kinds of array access

@@ -88,8 +88,8 @@ applyAnalysisToDir dir debug excludes = do
 type ForContigAndNonContig a = (a, a)
 
 -- Categories
-data Kind = Read | Stencil
-data Contig = NonContig | Contig
+data Kind = Read | Stencil deriving (Eq, Show, Ord)
+data Contig = NonContig | Contig deriving (Eq, Show, Ord)
 
 type Cat = (Kind, Contig)
 
@@ -123,22 +123,22 @@ data Result = Result {
   , histMaxDepth       :: M.Map Cat [Int]
   , histNumArraysRead  :: M.Map Cat [Int]
   , histNumIndexExprs  :: M.Map Cat [Int]
-  , histPatterns       :: M.Map (M.Map Int Int
+  , histPatterns       :: M.Map Cat (M.Map Int Int
                              , M.Map (Int, Int) Int
                              , M.Map (Int, Int, Int) Int)
 
   } deriving Show
 
-showCats :: Show a => M.Map Cat a -> String -> String
-showCats map prefix =
+showCats :: (Num a, Show a) => String -> M.Map Cat a -> String
+showCats prefix amap =
      showCat Stencil " stencils"
   ++ showCat Read " reads"
   where
     showCat cat msg =
       let prefix' = prefix ++ msg
-      in rline prefix' $ (map ! (cat, Contig) + map ! (cat, NonContig))
-      ++ rline (replicate (length prefix') ' ' ++ " (contiguous)") $ map ! (cat, Contig)
-      ++ rline (replicate (length prefix') ' ' ++ " (non-contiguous)") $ map ! (cat, NonContig)
+      in rline prefix' (amap M.! (cat, Contig) + amap M.! (cat, NonContig))
+      ++ (rline ((replicate (length prefix') ' ') ++ " (contiguous)") $ amap M.! (cat, Contig))
+      ++ (rline ((replicate (length prefix') ' ') ++ " (non-contiguous)") $ amap M.! (cat, NonContig))
 
 {-, numStencils                :: Int   -- b
   , numRelativisedRHS          :: Int   -- R (where R = M)
@@ -174,18 +174,19 @@ resultValidation =
  <**> (\r -> (numNeighbourArrayWrites r) `gte` (numLHSButInconsistentIVRHS r))
           `reason` "Neighbour Writes >= Inconsistent IV RHS"
 
- <**> (\r -> ((numStencils r) + (numLHSButInconsistentIVRHS r) + (numLHSButNonNeighbourRHS r))
[…]

### 2017-04-07 Analyse.hs
more indepth analysis of array reads

@@ -1,3 +1,20 @@
+{-
+
+ Analysis of array usage
+
+ TODO
+   * Detect affine indices (see classifyRHSsubscripts)
+   * Figure out whether to bin the different kinds of shape (maybe up to 3D)
+            * Tag patterns with either rectilinear,
+                                       contiguous composite of rectilinear,
+                                       contiguous composite of rectilinear (over ex. origin)
+                                       other
+   * Collapse first few 'classifyRHSSubscripts'
+
+
+
+-}
+
 {-# LANGUAGE TupleSections #-}
 {-# LANGUAGE FlexibleContexts #-}
 {-# LANGUAGE DeriveDataTypeable #-}
@@ -34,7 +51,7 @@ import Camfort.Specification.Stencils.Annotation
 import qualified Camfort.Specification.Stencils.Grammar as Gram
 import qualified Camfort.Specification.Stencils.Synthesis as Synth
 import Camfort.Analysis.Annotations
-import Camfort.Helpers.Vec
+import Camfort.Helpers.Vec hiding (length)
 import Camfort.Helpers (Filename, SourceText, collect)
 import Camfort.Input
 import qualified Camfort.Output as O
@@ -61,261 +78,34 @@ import Data.List
 import Data.Monoid
 import Debug.Trace
 
+import Results
+
 main :: IO ()
 main = do
     args <- getArgs
-    if (length args == 1) then
-        applyAnalysisToDir (head args) False []
-    else if (length args > 1) then
-      if (args !! 1 == "-d") then
-         applyAnalysisToDir (head args) True (tail (tail args))
-      else
-         applyAnalysisToDir (head args) False (tail args)
-    else
-      putStrLn $ "Please specify a directory on which to apply the analysis\
-                 \ followed by any number of file names to be excluded "
+    case args of
+      [dir]
+        -> applyAnalysisToDir (head args) False False []
+      dir:args@(a1:a2:_)
+        -> applyAnalysisToDir dir (a1 == "-d" || a2 == "-d")
+                                  (a1 == "-h" || a2 == "-h") (filterFlags args)
+      dir:args@(a1:_)
+        -> applyAnalysisToDir dir (a1 == "-d") (a1 == "-h") (filterFlags args)
[…]

### 2017-04-07 oopsla17/analysis/Results.hs
results data structures/functions

@@ -1,0 +1,243 @@
+{-# LANGUAGE FlexibleInstances #-}
+
+module Results where
+
+import Control.Monad
+import qualified Data.Map as M
+import Data.List
+
+import Camfort.Specification.Stencils.InferenceFrontend
+
+-- Categories
+data Kind = Read | Stencil deriving (Eq, Show, Ord)
+data Contig = NonContig | Contig deriving (Eq, Show, Ord)
+
+data LHS = Assign | O Class
+data Class = Subscript | Affine [(Int, Int)] | Neigh [Neighbour]
+
+
+
+
+type Cat = (Kind, Contig)
+
+isStencil (Stencil, _) = True
+isStencil _            = False
+
+isContig (_, Contig) = True
+isContig _ = False
+
+-- Results data type
+data Result = Result {
+    numLines                   :: Int   -- Number of lines parsed
+  , numArrayWrites             :: Int   -- N  where N >= P + M
+  , numIVArrayWrites           :: Int   -- P
+  , numNeighbourArrayWrites    :: Int   -- M
+  , numConstArrayWrites        :: Int   -- Q where M >= Q and M >= P
+  , numLHSButNonNeighbourRHS   :: Int  -- Q1
+  , numLHSButInconsistentIVRHS :: Int  -- Q2 where Q >= Q1 >= Q2
+  ------------------------------------
+  -- Special interest
+  ------------------------------------
+  , numSingNonContigStencils :: Int
+  , numStencilRelativisedRHS :: Int
+  ------------------------------------
+  , numOverall  :: M.Map Cat Int
+  , numLinear   :: M.Map Cat Int
+  , numNoOrigin :: M.Map Cat Int
+  ------------------------------------
+  -- Histograms
+  , histLengthOfDataflow :: M.Map Cat [Int]
+  , histDimensionality   :: M.Map Cat [Int]
+  , histMaxDepth         :: M.Map Cat [Int]
+  , histNumArraysRead    :: M.Map Cat [Int]
+  , histNumIndexExprs    :: M.Map Cat [Int]
+  , histPatterns         :: M.Map Cat (M.Map Int Int
+                                     , M.Map (Int, Int) Int
+                                     , M.Map (Int, Int, Int) Int)
+
+  } deriving Show
+
+
+-- Results form a monoid
+instance Monoid Result where
+  mempty = Result 0 0 0  0 0 0 0
+                  0 0
[…]

### 2017-04-07 Analyse.hs
more fine grained subscript classification (wip)

@@ -79,6 +79,7 @@ import Data.Monoid
 import Debug.Trace
 
 import Results
+import Indices
 
 main :: IO ()
 main = do
@@ -178,7 +179,9 @@ stencilAnalyse nameMap pf@(F.ProgramFile mi cm_pus blocks) =
 
 
 
-classifyIx :: FAD.InductionVarMapByASTBlock -> [F.Index (FA.Analysis Annotation)] -> Class
+classifyIx :: FAD.InductionVarMapByASTBlock
+           -> [F.Index (FA.Analysis Annotation)]
+           -> Class
 classifyIx ivs ix =
   case neighbourIndex ivs ix of
     Nothing ->
@@ -186,10 +189,6 @@ classifyIx ivs ix =
         Nothing -> Subscript
         Just afs -> Affine afs
     Just n -> Neigh n
-
-
-affineIndex :: FAD.InductionVarMapByASTBlock -> [F.Index (FA.Analysis Annotation)] -> Maybe [(Int, Int)]
-affineIndex ivs ix = undefined
 
 -- Used to classify an index
 

### 2017-04-07 oopsla17/analysis/Indices.hs
more fine grained subscript classification (wip)

@@ -1,0 +1,100 @@
+module Indices where
+
+import Camfort.Specification.Stencils.InferenceBackend
+import Camfort.Specification.Stencils.InferenceFrontend
+import Camfort.Specification.Stencils.Syntax
+import Camfort.Specification.Stencils.Annotation
+import qualified Camfort.Specification.Stencils.Grammar as Gram
+import qualified Camfort.Specification.Stencils.Synthesis as Synth
+import Camfort.Analysis.Annotations
+import Camfort.Helpers.Vec hiding (length)
+import Camfort.Helpers (Filename, SourceText, collect)
+import Camfort.Input
+import qualified Camfort.Output as O
+
+import qualified Language.Fortran.AST as F
+import qualified Language.Fortran.Analysis as FA
+import qualified Language.Fortran.Analysis.Types as FAT
+import qualified Language.Fortran.Analysis.Renaming as FAR
+import qualified Language.Fortran.Analysis.BBlocks as FAB
+import qualified Language.Fortran.Analysis.DataFlow as FAD
+import qualified Language.Fortran.Util.Position as FU
+import qualified Language.Fortran.Util.SecondParameter as FUS
+
+import qualified Data.Set as S
+import qualified Data.IntMap as IM
+import Data.Foldable
+import Data.Maybe
+import Results
+
+affineIndex :: FAD.InductionVarMapByASTBlock
+            -> [F.Index (FA.Analysis Annotation)]
+            -> Maybe [(Int, String, Int)]
+affineIndex ivs ix = mapM (ixToAffine ivs) ix
+
+ixToAffine ::  FAD.InductionVarMapByASTBlock
+            -> (F.Index (FA.Analysis Annotation))
+            -> Maybe (Int, String, Int)
+ixToAffine ivmap f@(F.IxSingle _ _ _ exp) =
+    matchAffine ivsList exp
+  where
+    insl = FA.insLabel . F.getAnnotation $ f
+    insl' = fromJust insl
+    ivsList = S.toList $ fromMaybe S.empty $ IM.lookup insl' ivmap
+
+
+matchAffine :: [Variable]
+            -> F.Expression (FA.Analysis Annotation)
+            -> Maybe (Int, String, Int)
+
+matchAffine ivs (F.ExpBinary _ _ F.Addition e e') =
[…]

### 2017-04-07 oopsla17/analysis/Results.hs

@@ -12,10 +12,15 @@ import Camfort.Specification.Stencils.InferenceFrontend
 data Kind = Read | Stencil deriving (Eq, Show, Ord)
 data Contig = NonContig | Contig deriving (Eq, Show, Ord)
 
+
+-- LHS assign, RHS subscript, RHS affine, RHS neigh
+-- LHS subscript, RHS subscript, RHS affine, RHS neigh
+-- LHS affine, RHS
+
+-- lhs neight
+
 data LHS = Assign | O Class
-data Class = Subscript | Affine [(Int, Int)] | Neigh [Neighbour]
-
-
+data Class = Subscript | Affine [(Int, String, Int)] | Neigh [Neighbour]
 
 
 type Cat = (Kind, Contig)

### 2017-04-08 oopsla17/analysis/Analyse.hs
towards a much more in-depth and fine-grained multi-variate analysis

@@ -178,18 +178,7 @@ stencilAnalyse nameMap pf@(F.ProgramFile mi cm_pus blocks) =
     dm    = FAD.genDefMap bm
 
 
-
-classifyIx :: FAD.InductionVarMapByASTBlock
-           -> [F.Index (FA.Analysis Annotation)]
-           -> Class
-classifyIx ivs ix =
-  case neighbourIndex ivs ix of
-    Nothing ->
-      case affineIndex ivs ix of
-        Nothing -> Subscript
-        Just afs -> Affine afs
-    Just n -> Neigh n
-
+{-
 -- Used to classify an index
 
 classifyLHS ixs | any ((==) NonNeighbour) ixs =
@@ -208,6 +197,8 @@ classifyLHS ixs | any (\i -> case i of Constant _ -> True; _ -> False) ixs =
 classifyLHS ixs | isOrigin ixs =
     -- All induction variables
     mempty { numArrayWrites = 1, numNeighbourArrayWrites = 1, numIVArrayWrites = 1 }
+-}
+
 
 -- Traverse Blocks in the AST and infer stencil specifications
 -- Note that this should be applied by `descendBiM`, as it does its
@@ -238,25 +229,14 @@ perBlockInnerDo b@(F.BlStatement ann span@(FU.SrcSpan lp up) _ stmnt) = do
 
     else do
       results <- forM lhses $ \lhs -> do
-           -- A potential stencil
-           case isArraySubscript lhs of
-             Just subs -> do
-               let lhsIndices = map (ixToNeighbour ivs) subs
-               let r = classifyLHS lhsIndices
-               if (numNeighbourArrayWrites r > 0) then do
-                   -- If the LHS is a neighbourhood index
-                   (dbg, rhses, dflowLen) <- analyseRHS [b]
-                   let r' = classifyRHSsubscripts Stencil rhses lhsIndices dflowLen
-                   return ("At: " ++ show span ++ "\n" ++ dbg, r `mappend` r')
-               else return ("", r)
-
-             -- Not a stencil, find out if its an array read
-             Nothing  -> do
-               (dbg, rhses, dflowLen) <- analyseRHS [b]
-               let r = classifyRHSsubscripts Read rhses [] dflowLen
-               return ("Read at: " ++ show span ++ "\n" ++ dbg, r)
+         (dbg, rhses, dflowLen) <- analyseRHS [b]
+         let (dbg', cat, result) = classify ivs lhs rhses
[…]

### 2017-04-08 oopsla17/analysis/Results.hs

@@ -1,4 +1,6 @@
 {-# LANGUAGE FlexibleInstances #-}
+{-# LANGUAGE GADTs, DataKinds, PolyKinds, KindSignatures #-}
+{-# LANGUAGE StandaloneDeriving #-}
 
 module Results where
 
@@ -8,47 +10,63 @@ import Data.List
 
 import Camfort.Specification.Stencils.InferenceFrontend
 
--- Categories
-data Kind = Read | Stencil deriving (Eq, Show, Ord)
-data Contig = NonContig | Contig deriving (Eq, Show, Ord)
+-- ## Categorisations:
 
+--   * RHS shape/position/contiguity
+data Shape    = Orthotope | SumOfOrthotope | Other
+    deriving (Eq, Show, Ord)
 
--- LHS assign, RHS subscript, RHS affine, RHS neigh
--- LHS subscript, RHS subscript, RHS affine, RHS neigh
--- LHS affine, RHS
+data Position = OverOrigin | StraddleOrigin | Elsewhere
+    deriving (Eq, Show, Ord)
 
--- lhs neight
+data Contig   = SingleNonContig | NonContig | Contig
+    deriving (Eq, Show, Ord)
 
-data LHS = Assign | O Class
+data Reuse    = Linear | NonLinear
+    deriving (Eq, Show, Ord)
+
+data Physicality s where
+     PL :: Physicality LHS
+     PR :: (Shape, Position, Contig, Reuse) -> Physicality RHS
+
+deriving instance Eq (Physicality s)
+deriving instance Ord (Physicality s)
+deriving instance Show (Physicality s)
+
+--   * Categorisation of indices (on LHS or RHS)
+data Side = LHS | RHS
+
+data Form (s :: Side) where
+    Vars        :: Form LHS
+    Subscripts  :: Form s
+    Affines     :: [(Int, String)] -> Physicality s -> Form s
+    Neighbours  :: [String]        -> Physicality s -> Form s
+    IVs         :: Form LHS
+
+deriving instance Eq (Form s)
+deriving instance Ord (Form s)
+deriving instance Show (Form s)
+
+type Relativised = Bool
+
+--   * Consistency between sides
+data Consistency = Consistent  Relativised
+                 | Permutation Relativised
+                 | LHSsubset   Relativised
+                 | LHSsuperset Relativised
+                 | Inconsistent
+      deriving (Eq, Show, Ord)
+
+-- Overall categorisation
[…]

### 2017-04-08 Indices.hs
towards a much more in-depth and fine-grained multi-variate analysis

@@ -23,9 +23,29 @@ import qualified Language.Fortran.Util.SecondParameter as FUS
 
 import qualified Data.Set as S
 import qualified Data.IntMap as IM
+import qualified Data.Map as M
 import Data.Foldable
 import Data.Maybe
 import Results
+
+
+-- Main function for classifying array statements
+classify :: FAD.InductionVarMapByASTBlock
+         -> F.Expression (FA.Analysis A)
+         -> M.Map Variable [[F.Index (FA.Analysis A)]]
+         -> (String, Cat, Result)
+classify ivs lhs rhses = (debug, cat, undefined)
+  where
+    debug = ""
+    result = undefined
+    cat = (formLHS, formRHS, consistency)
+    formLHS = undefined
+    formRHS = undefined
+    consistency = undefined
+    lhsClassifiers =
+      case isArraySubscript lhs of
+        Nothing   -> Vars
+        Just subs -> undefined -- classifyIx ivs subs
 
 affineIndex :: FAD.InductionVarMapByASTBlock
             -> [F.Index (FA.Analysis Annotation)]
@@ -76,25 +96,74 @@ matchMult ivs (F.ExpBinary _ _ F.Multiplication
                 e@(F.ExpValue _ _ (F.ValVariable _)))
     | FA.varName e `elem` ivs = Just (read a, FA.varName e)
 
+-- Allow a bare induction variable, since `matchAffine` is called
+-- as part of `affineIndex`, which is only ever called after `neighbourIndex`.
+-- This accounts for indices which are a mixture of affine and neighbour
+matchMult ivs e@(F.ExpValue _ _ (F.ValVariable {}))
+    | FA.varName e `elem` ivs = Just (1, FA.varName e)
+
 matchMult ivs _ = Nothing
 
 matchConst :: [Variable]
-           -> F.Expression (FA.Analysis Annotation)
+           -> F.Expression (FA.Analysis A)
            -> Maybe Int
 matchConst ivs (F.ExpValue _ _ (F.ValInteger val)) = Just $ read val
 matchConst ivs _                                   = Nothing
 
-consistentIV :: Class -> [Class] -> Maybe (Either [String] [(Int, String)])
-consistentIV lhs rhses = undefined
+classifyIx :: FAD.InductionVarMapByASTBlock
+           -> [F.Index (FA.Analysis A)]
+           -> Class
[…]

### 2017-04-09 Indices.hs
adding more of the classification functionality, and restructuring results slightly

@@ -1,3 +1,5 @@
+{-# LANGUAGE FlexibleInstances #-}
+
 module Indices where
 
 import Camfort.Specification.Stencils.InferenceBackend
@@ -38,14 +40,74 @@ classify ivs lhs rhses = (debug, cat, undefined)
   where
     debug = ""
     result = undefined
-    cat = (formLHS, formRHS, consistency)
-    formLHS = undefined
-    formRHS = undefined
-    consistency = undefined
-    lhsClassifiers =
+    --cat = (formLHS, formRHS, consistency)
+    --formLHS = undefined
+    --formRHS = undefined
+    --consistency = undefined
+    rhses' = concat . M.elems $ rhses
+    cat =
       case isArraySubscript lhs of
-        Nothing   -> Vars
-        Just subs -> undefined -- classifyIx ivs subs
+         Nothing   -> (Vars, undefined, undefined)
+         Just subs -> case classifyIx ivs subs of
+                        Subscript -> (Subscripts, undefined, undefined)
+                        Affine as -> classifyRHSLHSAffine ivs as rhses'
+                        Neigh ns  -> classifyRHSLHSNeighbour ivs ns rhses'
+
+classifyRHSLHSNeighbour ivs lhsAs rhses =
+   case mapM (neighbourIndex ivs) rhses of
+      Nothing -> (Neighbours PL, Neighbours (PR Nothing), Inconsistent)
+      Just rhsAs ->
+           (Neighbours PL, Neighbours (PR (Just undefined)) , cons)
+        where (cons, rel) = consistency lhsAs rhsAs
+  where justVars = fromJust . mapM neighbourVar
+        lhsASRep = justVars lhsAs
+
+classifyRHSLHSAffine ivs lhsAs rhses =
+    case mapM (affineIndex ivs) rhses of
+      Nothing -> (Affines PL, Affines (PR Nothing), Inconsistent)
+      Just rhsAs ->
+         (Affines PL, Affines physicality, cons)
+           -- TOOD: probably need to do something with the
+           -- relative offsets...
+           where (cons, rel) = consistency lhsAs rhsAs
+                 physicality = error "TOOD"
+  where lhsAsRep = map trunc lhsAs
+        trunc (a, i, b) = (a, i)
+
[…]

### 2017-04-09 oopsla17/analysis/Results.hs
adding more of the classification functionality, and restructuring results slightly

@@ -27,7 +27,7 @@ data Reuse    = Linear | NonLinear
 
 data Physicality s where
      PL :: Physicality LHS
-     PR :: (Shape, Position, Contig, Reuse) -> Physicality RHS
+     PR :: Maybe (Shape, Position, Contig, Reuse) -> Physicality RHS
 
 deriving instance Eq (Physicality s)
 deriving instance Ord (Physicality s)
@@ -39,8 +39,8 @@ data Side = LHS | RHS
 data Form (s :: Side) where
     Vars        :: Form LHS
     Subscripts  :: Form s
-    Affines     :: [(Int, String)] -> Physicality s -> Form s
-    Neighbours  :: [String]        -> Physicality s -> Form s
+    Affines     :: Physicality s -> Form s
+    Neighbours  :: Physicality s -> Form s
     IVs         :: Form LHS
 
 deriving instance Eq (Form s)
@@ -57,11 +57,23 @@ data Consistency = Consistent  Relativised
                  | Inconsistent
       deriving (Eq, Show, Ord)
 
+setRelativised :: Consistency -> Bool -> Consistency
+setRelativised (Consistent _)  r = Consistent r
+setRelativised (Permutation _) r = Permutation r
+setRelativised (LHSsubset   _) r = LHSsubset r
+setRelativised (LHSsuperset _) r = LHSsuperset r
+setRelativised Inconsistent    _ = Inconsistent
+
+joinConsistency :: Consistency -> Consistency -> Consistency
+joinConsistency (Consistent r)  (Consistent s)  = Consistent (r && s)
+joinConsistency (Permutation r) (Permutation s) = Permutation (r && s)
+joinConsistency (LHSsubset r)   (LHSsubset s)   = LHSsubset (r && s)
+joinConsistency (LHSsuperset r) (LHSsuperset s) = LHSsuperset (r && s)
+joinConsistency Inconsistent    Inconsistent    = Inconsistent
+joinConsistency _               _               = Inconsistent
+
 -- Overall categorisation
 type Cat = (Form LHS, Form RHS, Consistency)
-
--- ## Classification on subscripts
-data Class = Subscript | Affine [(Int, String, Int)] | Neigh [Neighbour]
 
 -- Results data type
 data Result = Result {
@@ -74,6 +86,7 @@ data Result = Result {
   , histMaxDepth         :: M.Map Cat [Int]
   , histNumArraysRead    :: M.Map Cat [Int]
[…]

### 2017-04-10 oopsla17/lattice-model.tex
Add equivalence from the connecting lemma
Fix Exact set definition

@@
   The \emph{exact set} at origin distance $d$, $\mathit{ES}_{P,d}$, of a group
-  of $N$ dimensional points $P$ is defined below:
+  of $N$ dimensional subscripts $S$ is defined below:
 %
   \begin{align*}
-    \mathit{ES}_{P,0} & = \{ p \in P \mid \forall \i.\ \pi_i(p) \in \interv{-1}{1}{} \} \\
-    \mathit{ES}_{P,d} & =
-      \{ p \in P \mid \forall i.\
-      p_i \notin [-1,1] \implies
-         \exists p' \in \mathit{ES}_{P,d-1}.\ p' =
-         \mathit{adjacent}(p,i) \}
+    \mathit{ES}_{S,0} = &
+      \{ s \in S \mid \forall \i.\ \pi_i(s) \in \{\{-1\}, \{0\}, \{1\}\} \} \\
+    \mathit{ES}_{S,d} = &
+      \{ s \in S \mid \forall i.\
+      \pi_i(s) \notin \{\{-1\}, \{0\}, \{1\}\} \implies
+      \exists s' \in \mathit{ES}_{S,d-1}.\ s' =
+         \mathit{adjacent}(s,i) \}
   \end{align*}
 %
-  The overall \emph{exact set}, $\mathit{ES}_P$ is then
-  $\bigcup_{1 \leq d \leq D} \mathit{ES}_{P,d}$, where $D$ is the maximum origin
-  distance in $P$.
+  The overall \emph{exact set}, $\mathit{ES}_S$ is then
+  $\bigcup_{0 \leq d \leq D} \mathit{ES}_{S,d}$, where $D$ is the maximum origin
+  distance in $S$.
 \end{defn}
@@
   \begin{enumerate}
-    \item $P$ defines a superposition of hyper-rectangles each of which crosses
-      the extended origin;
     \item $P$ is the same as its exact set $\mathit{ES}_P$;

### 2017-04-10 oopsla17/lattice-model.tex
Fix distance definition for the origin
Fix the connecting lemma

@@
     \mathit{dist}_i(S) & = \begin{cases}
-      0 & \pi_i(S) = \interv{-\infty}{\infty}{} \\
+      0 & \pi_i(S) \in \{ \interv{-\infty}{\infty}{}, \{ 0 \}\} \\
       |\mathsf{a}| - 1 & \pi_i(S) = \{\mathsf{a}\} \\
@@
 \begin{lem}{(\emph{Connecting})}
-  The following three statements are equivalent for a set of points $P \subseteq
-  \mathbb{Z}^N$:
+  The following three statements are equivalent for a set of $N$-dimensional
+  subscripts $S$:
 %
   \begin{enumerate}
-    \item $P$ is the same as its exact set $\mathit{ES}_P$;
-    \item $P$ when regarded as offsets from induction variables forms a
+    \item $S$ is the same as its exact set $\mathit{ES}_S$;
+    \item $S$ when regarded as offsets from induction variables forms a
       neighbourhood stencil;
-    \item $P$ is a subset of \region{N}.
+    \item $S$ is in \region{N}.
   \end{enumerate}

### 2017-04-10 oopsla17/analysis/Analyse.hs
analyse in almost complete form

@@ -3,13 +3,12 @@
  Analysis of array usage
 
  TODO
-   * Detect affine indices (see classifyRHSsubscripts)
+   * capture single non-contig
    * Figure out whether to bin the different kinds of shape (maybe up to 3D)
             * Tag patterns with either rectilinear,
                                        contiguous composite of rectilinear,
                                        contiguous composite of rectilinear (over ex. origin)
                                        other
-   * Collapse first few 'classifyRHSSubscripts'
 
 
 
@@ -87,11 +86,14 @@ main = do
     case args of
       [dir]
         -> applyAnalysisToDir (head args) False False []
+
       dir:args@(a1:a2:_)
         -> applyAnalysisToDir dir (a1 == "-d" || a2 == "-d")
                                   (a1 == "-h" || a2 == "-h") (filterFlags args)
+
       dir:args@(a1:_)
         -> applyAnalysisToDir dir (a1 == "-d") (a1 == "-h") (filterFlags args)
+
       _ -> putStrLn $ "Please specify a directory on which to apply the\
                      \ analysis followed by any number of file names\
                      \ to be excluded."
@@ -177,29 +179,6 @@ stencilAnalyse nameMap pf@(F.ProgramFile mi cm_pus blocks) =
     -- get map of variable name ==> { defining AST-Block-IDs }
     dm    = FAD.genDefMap bm
 
-
-{-
--- Used to classify an index
-
-classifyLHS ixs | any ((==) NonNeighbour) ixs =
-    -- Any non-neighour indexing
-    mempty { numArrayWrites = 1 }
-
-classifyLHS ixs | all (\i -> case i of Neighbour _ _ -> True; _ -> False) ixs
-            && any (\i -> case i of Neighbour _ o -> o /= 0; _ -> False) ixs =
-    -- All neighbour with some relative
-    mempty { numArrayWrites = 1, numNeighbourArrayWrites = 1 }
-
-classifyLHS ixs | any (\i -> case i of Constant _ -> True; _ -> False) ixs =
-    -- All relative or constant
-    mempty { numArrayWrites = 1, numNeighbourArrayWrites = 1, numConstArrayWrites = 1 }
-
-classifyLHS ixs | isOrigin ixs =
-    -- All induction variables
[…]

### 2017-04-10 oopsla17/analysis/Indices.hs

@@ -1,4 +1,7 @@
 {-# LANGUAGE FlexibleInstances #-}
+{-# LANGUAGE MultiParamTypeClasses #-}
+{-# LANGUAGE TypeFamilies #-}
+{-# LANGUAGE FlexibleContexts #-}
 
 module Indices where
 
@@ -27,7 +30,10 @@ import qualified Data.Set as S
 import qualified Data.IntMap as IM
 import qualified Data.Map as M
 import Data.Foldable
+import Data.List
 import Data.Maybe
+
+import Neighbour
 import Results
 
 
@@ -35,51 +41,116 @@ import Results
 classify :: FAD.InductionVarMapByASTBlock
          -> F.Expression (FA.Analysis A)
          -> M.Map Variable [[F.Index (FA.Analysis A)]]
-         -> (String, Cat, Result)
-classify ivs lhs rhses = (debug, cat, undefined)
+         -> (String, Result)
+classify ivs lhs rhses = (debug, resultN)
   where
     debug = ""
-    result = undefined
-    --cat = (formLHS, formRHS, consistency)
-    --formLHS = undefined
-    --formRHS = undefined
-    --consistency = undefined
-    rhses' = concat . M.elems $ rhses
-    cat =
-      case isArraySubscript lhs of
-         Nothing   -> (Vars, undefined, undefined)
-         Just subs -> case classifyIx ivs subs of
-                        Subscript -> (Subscripts, undefined, undefined)
-                        Affine as -> classifyRHSLHSAffine ivs as rhses'
-                        Neigh ns  -> classifyRHSLHSNeighbour ivs ns rhses'
 
-classifyRHSLHSNeighbour ivs lhsAs rhses =
-   case mapM (neighbourIndex ivs) rhses of
-      Nothing -> (Neighbours PL, Neighbours (PR Nothing), Inconsistent)
-      Just rhsAs ->
-           (Neighbours PL, Neighbours (PR (Just undefined)) , cons)
-        where (cons, rel) = consistency lhsAs rhsAs
-  where justVars = fromJust . mapM neighbourVar
-        lhsASRep = justVars lhsAs
+    -- Num arrays read
+    numArrays = length . M.keys $ rhses
 
-classifyRHSLHSAffine ivs lhsAs rhses =
-    case mapM (affineIndex ivs) rhses of
-      Nothing -> (Affines PL, Affines (PR Nothing), Inconsistent)
-      Just rhsAs ->
-         (Affines PL, Affines physicality, cons)
[…]

### 2017-04-10 oopsla17/analysis/Neighbour.hs

@@ -1,0 +1,95 @@
+{-# LANGUAGE GADTs#-}
+
+-- Various helpers for manipulating the 'Neighbour' data type and analysisng
+-- the shape and other spatial properties of groups of neighbours
+
+module Neighbour where
+
+import Results
+
+import Data.List
+import Camfort.Helpers.Vec
+import Camfort.Specification.Stencils.Syntax
+import Camfort.Specification.Stencils.InferenceBackend
+import Camfort.Specification.Stencils.InferenceFrontend
+
+
+
+shapeAndPosition :: [[Neighbour]] -> (Shape, Position)
+shapeAndPosition ivs =
+  case mapM (mapM neighbourToOffset) ivs of
+    Nothing -> (Other, Elsewhere)
+    Just intIvs ->
+      case fromLists intIvs of
+        VL intIvs ->
+          classifyIntervalRegions (inferMinimalVectorRegions intIvs)
+
+classifyIntervalRegions :: [Span (Vec n Int)] -> (Shape, Position)
+classifyIntervalRegions [x] =
+    (Orthotope, intervalToPosition x)
+classifyIntervalRegions xs =
+    (SumOfOrthotope, foldr1 joinPosition . map intervalToPosition $ xs)
+
+intervalToPosition = foldr1 joinPosition . map positionInterval . toList . transposeVecInterval
+
+transposeVecInterval :: Span (Vec n Int) -> Vec n (Span Int)
+transposeVecInterval (Nil, Nil) = Nil
+transposeVecInterval (Cons l ls, Cons u us) = Cons (l, u) intervalVec
+  where
+    intervalVec = transposeVecInterval (ls, us)
+
+
+positionInterval :: (Int, Int) -> Position
+positionInterval (l, u)
+  | l == -absoluteRep || u == absoluteRep = Elsewhere
+  | l <= 0 && u >= 0                      = OverOrigin
+  | l < 0  && u == (-1)                   = StraddleOrigin
+  | l == 1 && u > 0                       = StraddleOrigin
+  | otherwise                             = Elsewhere
+
+-- Contiguous stencil (need not include the origin)
+contiguous :: [[Neighbour]] -> Bool
+contiguous xs = contiguity' xs
+  where
+    contiguity' [] = True
+    contiguity' (y : ys) | isOrigin y = contiguity' ys
[…]

### 2017-04-10 oopsla17/analysis/Results.hs

@@ -14,20 +14,35 @@ import Camfort.Specification.Stencils.InferenceFrontend
 
 --   * RHS shape/position/contiguity
 data Shape    = Orthotope | SumOfOrthotope | Other
+              -- I think 'Other' shouldn't actually occur, but until I have a proof
+              -- I'll leave this hear to catch 'unwiedly' things
     deriving (Eq, Show, Ord)
 
 data Position = OverOrigin | StraddleOrigin | Elsewhere
     deriving (Eq, Show, Ord)
 
+joinPosition :: Position -> Position -> Position
+joinPosition OverOrigin OverOrigin = OverOrigin
+joinPosition StraddleOrigin StraddleOrigin = StraddleOrigin
+joinPosition _ _ = Elsewhere
+
 data Contig   = SingleNonContig | NonContig | Contig
     deriving (Eq, Show, Ord)
+
+boolToContig :: Bool -> Contig
+boolToContig True = Contig
+boolToContig False = NonContig
 
 data Reuse    = Linear | NonLinear
     deriving (Eq, Show, Ord)
 
+boolToReuse :: Bool -> Reuse
+boolToReuse True = Linear
+boolToReuse False = NonLinear
+
 data Physicality s where
-     PL :: Physicality LHS
-     PR :: Maybe (Shape, Position, Contig, Reuse) -> Physicality RHS
+     L :: Physicality LHS
+     R :: (Shape, Position, Contig, Reuse) -> Physicality RHS
 
 deriving instance Eq (Physicality s)
 deriving instance Ord (Physicality s)
@@ -55,7 +70,17 @@ data Consistency = Consistent  Relativised
                  | LHSsubset   Relativised
                  | LHSsuperset Relativised
                  | Inconsistent
-      deriving (Eq, Show, Ord)
+      deriving (Eq, Ord)
+
+instance Show Consistency where
+  show (Consistent r)  = "Consistent" ++ showR r
+  show (Permutation r) = "Permutation" ++ showR r
+  show (LHSsubset r)   = "LHSsubset" ++ showR r
+  show (LHSsuperset r) = "LHSsuperset" ++ showR r
+  show Inconsistent    = "Inconsistent"
+
+showR True = " relativised"
+showR False = ""
 
 setRelativised :: Consistency -> Bool -> Consistency
 setRelativised (Consistent _)  r = Consistent r
@@ -81,110 +106,105 @@ data Result = Result {
   , counts                :: M.Map Cat Int
[…]

### 2017-04-10 oopsla17/lattice-model.tex

@@
 \begin{thm}{(Soundness of exact shape inference)}
-  For a set of point offsets vectors $P$, the inference procedure for
+  For a set of subscripts $S$ of same dimension, the inference procedure for
   neighbourhood stencils is sound if the following holds:
@@
   \begin{equation*}
-    \forall P. \mathit{peel}(\mathit{infer}(P)) = m \implies
-      \mathit{consistent'}(P,m)
+    \forall S. \mathit{peel}(\mathit{infer}(S)) = m \implies
+      \mathit{consistent'}(S,m)
   \end{equation*}

### 2017-04-10 oopsla17/analysis/tmp/tmp.f90
testing file

@@ -1,0 +1,20 @@
+program tmp
+  implicit none
+  integer :: i
+  real, dimension(10) :: a, b, c
+
+  do i = 1, 10
+     a(i) = b(i+1) + b(i) + c(i-1)
+  end do
+
+!  do i = 1, 10
+!     b(i*2) = a(i*2) + a(i*2+1)
+!     z = a(i)
+!     b(i) = z + a(i+1)
+!     x = a(i)
+!     z = a(i+1)
+!     w = a(i-1)
+!     y = x + z + z + w + a(i-1) + a(i-2)
+!     w = a(i*2)
+!  end do
+end program

### 2017-04-10 oopsla17/analysis/Neighbour.hs
various bug fixes; adding serialisation of results

@@ -32,6 +32,10 @@ classifyIntervalRegions xs =
 
 intervalToPosition = foldr1 joinPosition . map positionInterval . toList . transposeVecInterval
 
+toList :: Vec n a -> [a]
+toList Nil = []
+toList (Cons x xs) = x : (toList xs)
+
 transposeVecInterval :: Span (Vec n Int) -> Vec n (Span Int)
 transposeVecInterval (Nil, Nil) = Nil
 transposeVecInterval (Cons l ls, Cons u us) = Cons (l, u) intervalVec

### 2017-04-10 oopsla17/analysis/Results.hs

@@ -16,10 +16,10 @@ import Camfort.Specification.Stencils.InferenceFrontend
 data Shape    = Orthotope | SumOfOrthotope | Other
               -- I think 'Other' shouldn't actually occur, but until I have a proof
               -- I'll leave this hear to catch 'unwiedly' things
-    deriving (Eq, Show, Ord)
+    deriving (Eq, Show, Ord, Read)
 
 data Position = OverOrigin | StraddleOrigin | Elsewhere
-    deriving (Eq, Show, Ord)
+    deriving (Eq, Show, Ord, Read)
 
 joinPosition :: Position -> Position -> Position
 joinPosition OverOrigin OverOrigin = OverOrigin
@@ -27,14 +27,14 @@ joinPosition StraddleOrigin StraddleOrigin = StraddleOrigin
 joinPosition _ _ = Elsewhere
 
 data Contig   = SingleNonContig | NonContig | Contig
-    deriving (Eq, Show, Ord)
+    deriving (Eq, Show, Ord, Read)
 
 boolToContig :: Bool -> Contig
 boolToContig True = Contig
 boolToContig False = NonContig
 
 data Reuse    = Linear | NonLinear
-    deriving (Eq, Show, Ord)
+    deriving (Eq, Show, Ord, Read)
 
 boolToReuse :: Bool -> Reuse
 boolToReuse True = Linear
@@ -47,6 +47,14 @@ data Physicality s where
 deriving instance Eq (Physicality s)
 deriving instance Ord (Physicality s)
 deriving instance Show (Physicality s)
+
+instance Read (Physicality LHS) where
+  readsPrec n ('L':xs) = [(L, xs)]
+  readsPrec n _        = []
+
+instance Read (Physicality RHS) where
+  readsPrec n ('R':' ':xs) = map (\(n, rest) -> (R n, rest)) $ readsPrec n xs
+  readsPrec n _            = []
 
 --   * Categorisation of indices (on LHS or RHS)
 data Side = LHS | RHS
@@ -62,6 +70,39 @@ deriving instance Eq (Form s)
 deriving instance Ord (Form s)
 deriving instance Show (Form s)
 
+instance Read (Form LHS) where
+  readsPrec n xs | "Vars" `isPrefixOf` xs
+    = [(Vars, drop (length "Vars") xs)]
+  readsPrec n xs | "Subscripts" `isPrefixOf` xs
[…]

### 2017-04-10 oopsla17/analysis/tmp/tmp.f90

@@ -4,17 +4,18 @@ program tmp
   real, dimension(10) :: a, b, c
 
   do i = 1, 10
-     a(i) = b(i+1) + b(i) + c(i-1)
+!     a(i) = b(i+1) + b(i) + c(i-1)
   end do
 
-!  do i = 1, 10
-!     b(i*2) = a(i*2) + a(i*2+1)
-!     z = a(i)
-!     b(i) = z + a(i+1)
-!     x = a(i)
-!     z = a(i+1)
-!     w = a(i-1)
-!     y = x + z + z + w + a(i-1) + a(i-2)
-!     w = a(i*2)
-!  end do
+
+  do i = 1, 10
+     b(i*2) = a(i*2) + a(i*2+1)     
+     z = a(i+1)
+     b(i) = z + a(i)
+     x = a(i)
+     z = a(i+1)
+     w = a(i-1)
+     y = x + z + z + w + a(i-1) + a(i-2)
+     w = a(i*2)
+  end do
 end program

### 2017-04-10 Analyse.hs
various bug fixes; adding serialisation of results

@@ -5,12 +5,6 @@
  TODO
    * capture single non-contig
    * Figure out whether to bin the different kinds of shape (maybe up to 3D)
-            * Tag patterns with either rectilinear,
-                                       contiguous composite of rectilinear,
-                                       contiguous composite of rectilinear (over ex. origin)
-                                       other
-
-
 
 -}
 
@@ -89,24 +83,25 @@ main = do
 
       dir:args@(a1:a2:_)
         -> applyAnalysisToDir dir (a1 == "-d" || a2 == "-d")
-                                  (a1 == "-h" || a2 == "-h") (filterFlags args)
+                                  (a1 == "-b" || a2 == "-b") (filterFlags args)
 
       dir:args@(a1:_)
-        -> applyAnalysisToDir dir (a1 == "-d") (a1 == "-h") (filterFlags args)
+        -> applyAnalysisToDir dir (a1 == "-d") (a1 == "-b") (filterFlags args)
 
       _ -> putStrLn $ "Please specify a directory on which to apply the\
                      \ analysis followed by any number of file names\
                      \ to be excluded."
   where
-    filterFlags = filter (\arg -> arg /= "-d" && arg /= "-h")
+    filterFlags = filter (\arg -> arg /= "-d" && arg /= "-b")
 
 applyAnalysisToDir :: String -> Bool -> Bool -> [String] -> IO ()
-applyAnalysisToDir dir debug histograms excludes = do
+applyAnalysisToDir dir debug bins excludes = do
     files <- readParseSrcDir dir excludes
     let debugsAndResults = map applyAnalysisToFile files
-    let (dbg, result)   = mconcat debugsAndResults
+    let (dbg, result)    = mconcat debugsAndResults
+    writeFile (dir ++ ".stencils-analysis") (show result)
     if debug then putStrLn $ dbg else return ()
-    putStrLn $ prettyResults result histograms
+    putStrLn $ prettyResults result bins
     valid <- resultValidation result
     if valid then (putStrLn $ "Results were valid") else (putStrLn "Results were invalid")
 
@@ -210,14 +205,14 @@ perBlockInnerDo b@(F.BlStatement ann span@(FU.SrcSpan lp up) _ stmnt) = do
       results <- forM lhses $ \lhs -> do
          (dbg, rhses, dflowLen) <- analyseRHS [b]
          let (dbg', result) = classify ivs lhs rhses
[…]

### 2017-04-10 Indices.hs

@@ -32,6 +32,7 @@ import qualified Data.Map as M
 import Data.Foldable
 import Data.List
 import Data.Maybe
+import Debug.Trace
 
 import Neighbour
 import Results
@@ -67,7 +68,7 @@ classifyArrayCode ivs lhs rhses =
 
     result  = case affineScales of
                 Nothing -> result1
-                Just scales -> result1 { histAffineScale = scales }
+                Just scales -> result1 { histAffineScale = concatHist (map toHist scales) }
     ------------
     lhsRep = case isArraySubscript lhs of
                   Nothing -> Nothing
@@ -89,7 +90,7 @@ classifyArrayCode ivs lhs rhses =
                 Subscript -> Subscripts
                 Affine _  -> Affines . R    $ rhsPhysical
                 Neigh  _  -> Neighbours . R $ rhsPhysical
-    rhsPhysical = (shape, position, boolToContig contiguity, boolToReuse linear)
+    rhsPhysical = (shape, position, boolToContig contiguity, boolToReuse (not nonLinear))
     ------------
     neighbourisedRhs = case rhsRepRel of
                          Subscript -> []
@@ -97,7 +98,7 @@ classifyArrayCode ivs lhs rhses =
                          Neigh  ns -> ns
     (shape, position) = shapeAndPosition neighbourisedRhs
     -- Work out if the pattern is linear or not
-    (_, linear) = hasDuplicates neighbourisedRhs
+    (_, nonLinear) = hasDuplicates neighbourisedRhs
     -- Contiguity
     contiguity = contiguous neighbourisedRhs
     -- Calculate the max depth (from relativised)
@@ -116,7 +117,7 @@ classifyArrayCode ivs lhs rhses =
     maximum0 xs = maximum xs
 
     affineScales = case rhsRep of
-                      Affine as -> Just . map (\(a, _, _) -> a) . concat $ as
+                      Affine as -> Just . nub . map (\(a, _, _) -> a) . concat $ as
                       _         -> Nothing
 
 checkConsistency :: (Eq t, Relativise t, Basis t, Eq (Base t)) => [t] -> [[t]] -> (Consistency, [[t]])
@@ -200,13 +201,17 @@ matchAffine :: [Variable]
[…]

### 2017-04-10 Indices.hs
pattern collection feature

@@ -66,9 +66,15 @@ classifyArrayCode ivs lhs rhses =
                      , histNumIndexExprs  = mkHist cat . toHist . length $ rhses
                      , counts             = mkHist cat 1 }
 
-    result  = case affineScales of
+    result2  = case affineScales of
                 Nothing -> result1
                 Just scales -> result1 { histAffineScale = concatHist (map toHist scales) }
+    -- pattern bins
+    result = case maximum . map length $ rhses of
+               1 -> result2 { patternBin1D = M.fromList [(to1D pattern, 1)] }
+               2 -> result2 { patternBin2D = M.fromList [(to2D pattern, 1)] }
+               3 -> result2 { patternBin3D = M.fromList [(to3D pattern, 1)] }
+               _ -> result2
     ------------
     lhsRep = case isArraySubscript lhs of
                   Nothing -> Nothing
@@ -119,6 +125,9 @@ classifyArrayCode ivs lhs rhses =
     affineScales = case rhsRep of
                       Affine as -> Just . nub . map (\(a, _, _) -> a) . concat $ as
                       _         -> Nothing
+    -- capture pattern bin
+    pattern = sort . nub $ rhsOffsets
+
 
 checkConsistency :: (Eq t, Relativise t, Basis t, Eq (Base t)) => [t] -> [[t]] -> (Consistency, [[t]])
 checkConsistency lhs rhses = (cons `setRelativised` (rel /= rhses), rel)

### 2017-04-10 oopsla17/analysis/Results.hs
pattern collection feature

@@ -175,6 +175,19 @@ data Result = Result {
   , patternBin3D         :: M.Map [(Int, Int, Int)] Int
   } deriving (Show, Read)
 
+-- pre-condition: list of 1-dimensional offsets
+to1D :: [[Int]] -> [Int]
+to1D [] = []
+to1D ([x]:xs) = x : to1D xs
+-- pre-condition: list of 2-dimensional offsets
+to2D :: [[Int]] -> [(Int, Int)]
+to2D [] = []
+to2D ([x, y]:xs) = (x, y) : to2D xs
+-- pre-condition: list of 3-dimensional offsets
+to3D :: [[Int]] -> [(Int, Int, Int)]
+to3D [] = []
+to3D ([x, y, z]:xs) = (x, y, z) : to3D xs
+
 
 -- Results form a monoid
 instance Monoid Result where
@@ -245,7 +258,7 @@ rline' msg dat = "   " ++ msg ++ ":" ++ (replicate (90 - (length msg)) ' ') ++ d
 
 binView msg bin =
       rline msg (length . M.keys $ bin)
-   ++ (concatMap (\(pat, count) -> "   " ++ show count ++ " of " ++ show pat)
+   ++ (concatMap (\(pat, count) -> "      " ++ show count ++ " of " ++ show pat ++ "\n")
         (M.assocs bin))
 
 mapView msg map =

### 2017-04-11 oopsla17/analysis/Analyse.hs
allow the analysis to be restarted on partial data; extended categorisation of neighbours

@@ -70,6 +70,8 @@ import Data.Maybe
 import Data.List
 import Data.Monoid
 import Debug.Trace
+import System.Directory
+import qualified System.IO.Strict as Strict
 
 import Results
 import Indices
@@ -77,16 +79,22 @@ import Indices
 main :: IO ()
 main = do
     args <- getArgs
-    case args of
+    let (restart, args') = case args of
+                             (x:y:args') | x == "RESTART" -> (Just y, args')
+                             _                            -> (Nothing, args)
+    case args' of
       [dir]
-        -> applyAnalysisToDir (head args) False False []
+        -> applyAnalysisToDir restart dir False False []
 
       dir:args@(a1:a2:_)
-        -> applyAnalysisToDir dir (a1 == "-d" || a2 == "-d")
-                                  (a1 == "-b" || a2 == "-b") (filterFlags args)
+        -> applyAnalysisToDir restart dir (a1 == "-d" || a2 == "-d")
+                                          (a1 == "-b" || a2 == "-b")
+                                          (filterFlags args)
 
       dir:args@(a1:_)
-        -> applyAnalysisToDir dir (a1 == "-d") (a1 == "-b") (filterFlags args)
+        -> applyAnalysisToDir restart dir (a1 == "-d")
+                                          (a1 == "-b")
+                                          (filterFlags args)
 
       _ -> putStrLn $ "Please specify a directory on which to apply the\
                      \ analysis followed by any number of file names\
@@ -94,25 +102,42 @@ main = do
   where
     filterFlags = filter (\arg -> arg /= "-d" && arg /= "-b")
 
-applyAnalysisToDir :: String -> Bool -> Bool -> [String] -> IO ()
-applyAnalysisToDir dir debug bins excludes = do
+applyAnalysisToDir :: Maybe String -> String -> Bool -> Bool -> [String] -> IO ()
+applyAnalysisToDir restart dir debug bins excludes = do
     files <- readParseSrcDir dir excludes
-    let debugsAndResults = map applyAnalysisToFile files
-    let (dbg, result)    = mconcat debugsAndResults
-    writeFile (dir ++ ".stencils-analysis") (show result)
+    let resultsFile = dir ++ ".stencil-analysis"
+
+    -- See if we have a 'restarts' file to include
+    result0 <- case restart of
+                 Just restartFile -> do
[…]

### 2017-04-11 oopsla17/analysis/Indices.hs

@@ -61,7 +61,8 @@ classifyArrayCode ivs lhs rhses =
     result
   where
     cat    = (lhsForm, rhsForm, consistency)
-    result1 = mempty { histMaxDepth = M.fromList [(cat, toHist maxDepth)]
+    result1 = mempty { histMaxDepth = M.fromList [(cat, toHist maxDepth),
+                                                  (cat, toHist minDepth)]
                      , histDimensionality = mkHist cat dim
                      , histNumIndexExprs  = mkHist cat . toHist . length $ rhses
                      , counts             = mkHist cat 1 }
@@ -90,12 +91,12 @@ classifyArrayCode ivs lhs rhses =
     lhsForm = case lhsRep of
                 Nothing -> Vars
                 Just Subscript  -> Subscripts
-                Just (Affine _) -> Affines L
-                Just (Neigh  _) -> Neighbours L
+                Just (Affine as) -> if allIVs as then IVs else Affines (hasConstants as) L
+                Just (Neigh  ns) -> if allIVs ns then IVs else Neighbours (hasConstants ns) L
     rhsForm = case rhsRep of
                 Subscript -> Subscripts
-                Affine _  -> Affines . R    $ rhsPhysical
-                Neigh  _  -> Neighbours . R $ rhsPhysical
+                Affine as  -> Affines (hasConstants as) (R rhsPhysical)
+                Neigh  ns  -> Neighbours (hasConstants ns) (R rhsPhysical)
     rhsPhysical = (shape, position, boolToContig contiguity, boolToReuse (not nonLinear))
     ------------
     neighbourisedRhs = case rhsRepRel of
@@ -113,23 +114,33 @@ classifyArrayCode ivs lhs rhses =
     -- Dimensionality
     dim = concatHist . map toHist . nub . map length $ rhses
 
-    rhsOffsets = map (filter ((/=) absoluteRep)
-                       . fromJust . mapM neighbourToOffset) neighbourisedRhs
+    rhsOffsets = map (fromJust . mapM neighbourToOffset) neighbourisedRhs
 
-    maxDepth = maximum0 . map maximum0 $ rhsOffsets
+    -- Max and min depth
[…]

### 2017-04-11 oopsla17/analysis/Results.hs

@@ -9,6 +9,7 @@ import qualified Data.Map as M
 import Data.List
 
 import Camfort.Specification.Stencils.InferenceFrontend
+import Camfort.Specification.Stencils.Syntax (absoluteRep)
 
 -- ## Categorisations:
 
@@ -59,11 +60,27 @@ instance Read (Physicality RHS) where
 --   * Categorisation of indices (on LHS or RHS)
 data Side = LHS | RHS
 
+data HasConstants = WithConsts | Normal
+   deriving (Eq, Ord)
+
+instance Show HasConstants where
+   show WithConsts = "+consts"
+   show Normal     = ""
+
+instance Read HasConstants where
+   readsPrec n xs | "+consts" `isPrefixOf` xs
+       = [(WithConsts, drop (length "+consts") xs)]
+                  | otherwise = [(Normal, trim xs)]
+
+boolToHasConstants :: Bool -> HasConstants
+boolToHasConstants True = WithConsts
+boolToHasConstants False = Normal
+
 data Form (s :: Side) where
     Vars        :: Form LHS
     Subscripts  :: Form s
-    Affines     :: Physicality s -> Form s
-    Neighbours  :: Physicality s -> Form s
+    Affines     :: HasConstants -> Physicality s -> Form s
+    Neighbours  :: HasConstants -> Physicality s -> Form s
     IVs         :: Form LHS
 
 deriving instance Eq (Form s)
@@ -77,12 +94,14 @@ instance Read (Form LHS) where
     = [(Subscripts, drop (length "Subscripts") xs)]
 
   readsPrec n xs | "Affines" `isPrefixOf` xs
-    = readsPrec n (drop (length "Affines ") xs)
-          >>= (\(pl, rest) -> [(Affines pl, rest)])
+    = do (consts, rest) <- readsPrec n (drop (length "Affines ") xs)
+         (pl, rest)     <- readsPrec n rest
+         return (Affines consts pl, rest)
 
   readsPrec n xs | "Neighbours" `isPrefixOf` xs
-    = readsPrec n (drop (length "Neighbours ") xs)
-         >>= (\(pl, rest) -> [(Neighbours pl, rest)])
+    = do (consts, rest) <- readsPrec n (drop (length "Neighbours ") xs)
+         (pl, rest)     <- readsPrec n rest
[…]

### 2017-04-11 Analyse.hs
support for index range expressions

@@ -210,6 +210,17 @@ perBlock b@(F.BlDo aan span lab cname lab' mDoSpec body tlab) = do
    mapM perBlockInnerDo (reverse body)
    return b
 
+perBlock b@(F.BlStatement ann span@(FU.SrcSpan lp up) _ stmnt) = do
+  -- Detect subscript range expressions outside loops
+  let lhses = [lhs | (F.StExpressionAssign _ _ lhs _)
+                         <- universe stmnt :: [F.Statement (FA.Analysis A)]]
+  -- Find subscript ranges involved
+  let subscriptRanges = [r | r@(F.IxRange {}) <- universeBi lhses :: [F.Index (FA.Analysis A)]]
+  if (null subscriptRanges)
+    then return b
+     -- Use the normal 'inner do' mode if this looks like a ranged-based expression
+    else perBlockInnerDo b
+
 perBlock b = do
     -- Go inside child blocks
     b' <- descendM (descendBiM perBlock) b

### 2017-04-11 oopsla17/analysis/Indices.hs
support for index range expressions

@@ -51,7 +51,9 @@ classify ivs lhs rhses = (debug, resultN)
     numArrays = length . M.keys $ rhses
 
     resultN = result { histNumArraysRead = toHist numArrays }
-    result = foldr1 mappend (map (classifyArrayCode ivs lhs) (M.elems rhses))
+    result = if null $ M.elems rhses
+             then mempty
+             else foldr1 mappend (map (classifyArrayCode ivs lhs) (M.elems rhses))
 
 (><) :: (a -> b) -> (c -> d) -> (a, c) -> (b, d)
 f >< g = \(x, y) -> (f x, g y)
@@ -117,7 +119,9 @@ classifyArrayCode ivs lhs rhses =
     rhsOffsets = map (fromJust . mapM neighbourToOffset) neighbourisedRhs
 
     -- Max and min depth
-    (maxDepth, minDepth) = (maximum' >< minimum') . unzip . map (maxMin . filter (/= absoluteRep)) $ rhsOffsets
+    (maxDepth, minDepth) = (maximum' >< minimum')
+                         . unzip
+                         . map (maxMin . filter (/= absoluteRep)) $ rhsOffsets
     maxMin x = (maximum' x, minimum' x)
 
     maximum' :: (Ord a, Bounded a) => [a] -> a

### 2017-04-11 tmp/tmp.f90
change to strictness to try to stop restarts file being junked by lazy write

@@ -1,10 +1,13 @@
 program tmp
   implicit none
   integer :: i
+  integer :: j
   real, dimension(10) :: a, b, c
 
+  j = a
+  
   do i = 1, 10
-!     a(i) = b(i+1) + b(i) + c(i-1)
+     a(i) = b(i+1) + b(i) + c(i-1)
   end do
 
 
@@ -18,4 +21,11 @@ program tmp
      y = x + z + z + w + a(i-1) + a(i-2)
      w = a(i*2)
   end do
+
+  contains
+
+    integer function foo(x)
+      integer :: x
+      foo = x
+    end function foo
 end program

### 2017-04-11 oopsla17/analysis/Analyse.hs
change to strictness to try to stop restarts file being junked by lazy write

@@ -72,6 +72,7 @@ import Data.Monoid
 import Debug.Trace
 import System.Directory
 import qualified System.IO.Strict as Strict
+import Control.DeepSeq
 
 import Results
 import Indices
@@ -128,7 +129,7 @@ applyAnalysisToFile :: String
                     -> IO (String, Result)
 applyAnalysisToFile dir (filename, source, pf) (dbg0, result0) = do
     -- Write results so far to the restart file
-    writeFile (dir ++ ".restart") (show (result0 `mappend` result'))
+    dbg' `deepseq` writeFile (dir ++ ".restart") (show (result0 `mappend` result'))
     -- Return results and debugs
     return (dbg0 ++ dbg', result0 `mappend` result')
   where

### 2017-04-11 oopsla17/oopsla17.tex
adding language grammar to the paper
abstract'

@@
 
+\usepackage{stencilmacros}
+
 %% Bibliography style
@@
 
-% Writing macros
-\newcommand{\eg}{\emph{e.g.}}
-\newcommand{\ie}{\emph{i.e.}}
-
 \newcommand{\inbar}{\ \ \textbar\ \ }
-
-% Meta macros
-\newcommand{\todo}[1]{\textcolor{blue}{#1}}
 
@@
 
+\author{Dominic Orchard}
+\affiliation{
+  \department{School of Computing}
+  \institution{University of Kent}
+}
+\email{d.a.orchard@kent.ac.uk}
+
 \author{Mistral Contrastin}
@@
 \email{Mistral.Contrastin@cl.cam.ac.uk}
+
+\author{Matthew Danish}
+\affiliation{
+  \department{Computer Laboratory}
+  \institution{University of Cambridge}
+}
+\email{Matthew.Danish@cl.cam.ac.uk}
 
@@
 \maketitle
+
+\begin{abstract}
+  Arrays computations are the workhorse of numerical modelling and
+  computational science.
+  Many practitioners are aware of need to ensure program
+  correctness, yet very few of the techniques from the programming
+  research community are applied by scientists. We aim to change that
+  by providing lightweight verification tools to scientists.
+  We focus on one error-prone aspect of numerical code: the 
+  data access pattern of arrays. 
+  Firstly, we report on a code analysis study on twelve real-world
+  computational science code base, identifying common idioms of array
+  usage and their spatial properties. This provides much needed data
+  on array programming idioms common in science, spanning roughly 1.5
+  million lines of code.
+  From this data, we 
+  designed a lightweight declarative specification language capturing
+  the majority of access patterns for arrays. We details its semantics
+  and algorithms for checking and inference.
+  We detail the design and implementation of a verification tool for
+  our specification language, which both checks and infers
+  specifications.
+  We evaluate our tool on the our previous corpus of scientific code
+  and give verification case studies of bugs that could be detected
+  with our approach. We find roughly 60,000 targets for specification, showing the vast majority
+  of stencil computations indeed have a simple, regular, static shape.
+\end{abstract}
+
+ %S ubsequently, Fortran continues to be a
+%  popular language choice, due to its focus on arrays and its high
+%  performance.
+
+%  In this paper, we propose an alternate approach:
+%  a specification language and automated tool for verifying the
+%  spatial behaviour of stencils. The specification language is based
+%  on the hypothesis that stencil computations read from arrays with a
+%  regular, fixed pattern which can be captured by a simple
+%  set of abstract combinators. These
+%  specifications can be embedded as annotations in source code,
+%  against which code can be checked for conformance. For legacy code,
+%  spatial specifications can be inferred, providing documentation and
+%  aiding future software maintenance. We evaluate our specification
+%  language and verification tool against
+%  a corpus of numerical Fortran code ($\sim$ 1 million lines) for
+%  which we generate $\sim$60,000 specifications, showing the vast majority
+%  of stencil computations indeed have a simple, regular, static shape.
+%\end{abstract}
 
@@
 
-\section{An Empirical Study of Stencils in Scientific Fortran code}
+\subsection{Terminology and notation}
 
-\subsection{Our corpus}
+\begin{definition}[Induction variables]
+  An integer variable is a \emph{base induction variable} if it is the control
+  variable of a ``\texttt{for}'' loop, incremented by $1$ per iteration. A variable
+  is interpreted as an induction variable only within the scope of the loop
+  body. Throughout, $i, j, k$ range over induction variables.
+
+  A \emph{derived induction variable} is an expression of the form $a
+  i + b$, \ie{}, an affine expression on an induction variable $i$.
+\end{definition}
+
+\begin{definition}[Array subscripts and indices]
+  An \emph{array subscript}, denoted $a(\bar{e})$, is an expression
+  which reads from an array $a$ at an \emph{index} specified by a
+  comma-separated sequence of integer expressions denoted $\bar{e}$ or
+  in expanded form as $(e_1, \ldots, e_n)$. An index $e_i$ is called
+  \emph{relative} if the expression involves an induction variable.
+  An \emph{absolute index} is a integer expression which is constant
+  relative to the enclosing loop.
+\end{definition}
+
+\begin{definition}[Neighbourhood index]
+\label{def:neighbour}
+  For an array subscript $a(\bar{e})$ an index $e \in \bar{e}$ is a
+  \emph{neighbourhood index} if $e$ is of the form $e \equiv i$, $e \equiv i +
+  c$, or $e \equiv i - c$, where $c$ is an integer constant. That is, a
+  neighbourhood index is a constant translation of an induction variable. (The
+  relation $\equiv$ identifies terms up-to commutativity of $+$ and the inverse
+  relationship of $+$ and $-$ \eg{}, $(-b) + i \equiv i - b$).
+\end{definition}
+
+\section{An Empirical Study of Array Computations in Scientific Fortran code}
+
+\begin{enumerate}
+\item Stencils are ubiquitious;
+\item Most stencils read their data from source arrays with a static pattern
+base on constant offsets from the mid-point;
+\item Most stencils read their data from contiguous \emph{regions} in an array;
+\item Most stencils read each index at most once.
+\end{enumerate}
+
+\subsection{Study}
 %
@@
 
-\begin{enumerate}
-\item Stencils are ubiquitious;
-\item Most stencils read their data from source arrays with a static pattern
-base on constant offsets from the mid-point;
-\item Most stencils read their data from contiguous \emph{regions} in an array;
-\item Most stencils read each index at most once.
-\end{enumerate}
+\section{A Specification Language for Shape}
 
-\section{A Specification Language for Shape}
+\begin{figure}[t]
+\begin{align*}
+\def\arraystretch{1}
+\setlength{\arraycolsep}{0.2em}
+\newcommand{\dimTy}{$\mathbb{N}_{>0}$}
+\newcommand{\depthRange}{$\mathbb{N}_{>0}$}
+\begin{array}{rl}
+\nonterm{specification} ::= & \nonterm{regionDec} \mid \nonterm{specDec} \\
+\nonterm{specDec} ::= & \term{stencil} \; \nonterm{spec} \;
+                        \texttt{::} \; v \\
+\nonterm{regionDec} ::= &  \texttt{region} \; \texttt{::} \; \nonterm{rvar} \; \texttt{=} \;
+                         \nonterm{region}\\[0.4em]
+%\nonterm{spec} ::= & \nonterm{spatial} \mid \nonterm{temporal}
+%\\[1em]
+\nonterm{spec} ::= & [\nonterm{mult},] \; [\nonterm{approx},] \; \nonterm{region} \\
+\nonterm{mult} ::= & \term{readOnce} \\
+\nonterm{approx} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
+\nonterm{region} ::= & \nonterm{rvar} \; \mid \;
+                       \stencil{p}{\dimTy}{}{} \; \\
+& {\qquad\; \mid \; \stencil{f}{\dimTy}{\depthRange}{\;[, \nonpointed]}} \\
+& {\qquad\; \mid \; \stencil{b}{\dimTy}{\depthRange}{\;[, \nonpointed]}} \\
+& {\qquad\; \mid \; \stencil{c}{\dimTy}{\depthRange}{\;[, \nonpointed]}} \\
+& {\qquad\; \mid \; \nonterm{region} \, \term{+}
+  \, \nonterm{region} \; \mid \; \nonterm{region} \; \term{*} \; \nonterm{region}} \\[0.25em]
+%\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{rvar}}
+%\\[0.5em]
+%\nonterm{temporal} ::= \; & \term{dependency} \; (v \; \{ , v \}) [, \texttt{mutual}]
+%  \\[0.5em]
+\nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+\\[-1em]
+\end{array}
+\end{align*}
+\caption{Specification syntax (EBNF grammar)}
+\label{fig:syntax}
+\end{figure}
+
+\subsection{Specification syntax}
+\label{subsec:syntax}
+
+\Cref{fig:syntax} gives the syntax of stencil specifications, which is
+detailed below. The entry point is the \nonterm{specification} production which
+splits into either a \emph{region declaration} or a \emph{specification
+declaration}.
+Regions comprise \emph{region constants} which are
+combined via region operators \term{+} and \term{*}.
 

### 2017-04-11 oopsla17/analysis/Results.hs
more robustness to weird indexing scheme in analysis

@@ -251,7 +251,7 @@ toHist n = (replicate n 0) ++ [1]
 
 -- 'zip' together a list of histograms
 concatHist :: [[Int]] -> [Int]
-concatHist = foldr1 histZip
+concatHist = foldr histZip []
 
 -- Singleton histogram
 mkHist :: Cat -> a -> M.Map Cat a
@@ -301,7 +301,7 @@ instance HistogramShow Int where
 
 instance HistogramShow [Int] where
   show'     = hview
-  histTotal = foldr1 histZip
+  histTotal = foldr histZip []
 
 hline' cat dat =
   rline' ((replicate 5 ' ') ++ (show cat)) dat

### 2017-04-11 Indices.hs
more robustness to weird indexing scheme in analysis

@@ -219,7 +219,16 @@ ixToAffine ivmap f@(F.IxSingle _ _ _ exp) =
     insl = FA.insLabel . F.getAnnotation $ f
     insl' = fromJust insl
     ivsList = S.toList $ fromMaybe S.empty $ IM.lookup insl' ivmap
-
+ixToAffine ivmap (F.IxRange _ _ Nothing Nothing Nothing) =
+    return (1, "", 0)
+ixToAffine ivmap (F.IxRange _ _ Nothing Nothing
+                  (Just (F.ExpValue _ _ (F.ValInteger "1")))) =
+    return (1, "", 0)
+ixToAffine ivmap (F.IxRange {}) =
+   -- Dealing with general array ranges requires much more semantic interpretation
+   -- so for the moment, we don't deal with it
+    "NON-HANDLED-RANGE" `trace` Nothing
+ixToAffine _ _ = Nothing
 
 matchAffine :: [Variable]
             -> F.Expression (FA.Analysis Annotation)

### 2017-04-11 Neighbour.hs

@@ -25,12 +25,15 @@ shapeAndPosition ivs =
           classifyIntervalRegions (inferMinimalVectorRegions intIvs)
 
 classifyIntervalRegions :: [Span (Vec n Int)] -> (Shape, Position)
+classifyIntervalRegions [] = error "classifyIntervalRegions []"
 classifyIntervalRegions [x] =
     (Orthotope, intervalToPosition x)
 classifyIntervalRegions xs =
     (SumOfOrthotope, foldr1 joinPosition . map intervalToPosition $ xs)
 
-intervalToPosition = foldr1 joinPosition . map positionInterval . toList . transposeVecInterval
+intervalToPosition = foldr1 joinPosition . reportEmptyError . map positionInterval . toList . transposeVecInterval
+reportEmptyError [] = error "intervalToPosition []"
+reportEmptyError xs = xs
 
 toList :: Vec n a -> [a]
 toList Nil = []

### 2017-04-11 Analyse.hs
bug fixes; and dealing with different kinds of control flow in the right way

@@ -61,6 +61,7 @@ import qualified Language.Fortran.Util.SecondParameter as FUS
 import Data.Data
 import Data.Foldable
 import Data.Generics.Uniplate.Operations
+import qualified Data.Generics.Str as Str
 import Data.Graph.Inductive.Graph hiding (isEmpty)
 import qualified Data.Map as M
 import qualified Data.IntMap as IM
@@ -69,6 +70,8 @@ import qualified Data.ByteString.Char8 as B
 import Data.Maybe
 import Data.List
 import Data.Monoid
+import Data.Traversable
+
 import Debug.Trace
 import System.Directory
 import qualified System.IO.Strict as Strict
@@ -165,14 +168,15 @@ stencilAnalyse nameMap pf@(F.ProgramFile mi cm_pus blocks) =
     (dbg1 ++ dbg2, result1 `mappend` result2)
   where
     (_, (dbg1, result1)) = runWriter (transformBiM perPU cm_pus)
-    (_, (dbg2, result2)) = runAnalysis ivMap flTo nameMap (descendBiM perBlock blocks)
+    (_, (dbg2, result2)) = runAnalysis ivMap flTo nameMap
+                             (descendBiM (perBlock False) blocks)
 
     -- Run inference per program unit, placing the flowsmap in scope
     perPU :: F.ProgramUnit (FA.Analysis A)
           -> Writer (String, Result) (F.ProgramUnit (FA.Analysis A))
 
     perPU pu | Just _ <- FA.bBlocks $ F.getAnnotation pu = do
-              let pum = descendBiM perBlock pu
+              let pum = descendBiM (perBlock False) pu
               let (pu', log) = runAnalysis ivMap flTo nameMap pum
               tell log
               return pu'
@@ -203,15 +207,15 @@ stencilAnalyse nameMap pf@(F.ProgramFile mi cm_pus blocks) =
 -- Traverse Blocks in the AST and infer stencil specifications
 -- Note that this should be applied by `descendBiM`, as it does its
 -- own inner traversal and relies on the order in which blocks are visited
-perBlock :: F.Block (FA.Analysis A) -> Analysis (F.Block (FA.Analysis A))
+perBlock :: Bool -> F.Block (FA.Analysis A) -> Analysis (F.Block (FA.Analysis A))
 
-perBlock b@(F.BlDo aan span lab cname lab' mDoSpec body tlab) = do
[…]

### 2017-04-11 Indices.hs

@@ -45,13 +45,11 @@ classify :: FAD.InductionVarMapByASTBlock
          -> (String, Result)
 classify ivs lhs rhses = (debug, resultN)
   where
-    debug = ""
-
     -- Num arrays read
     numArrays = length . M.keys $ rhses
 
     resultN = result { histNumArraysRead = toHist numArrays }
-    result = if null $ M.elems rhses
+    (debug, result) = if null $ M.elems rhses
              then mempty
              else foldr1 mappend (map (classifyArrayCode ivs lhs) (M.elems rhses))
 
@@ -60,7 +58,7 @@ f >< g = \(x, y) -> (f x, g y)
 
 -- LHS are neighbours
 classifyArrayCode ivs lhs rhses =
-    result
+    (show cat ++ "\n", result)
   where
     cat    = (lhsForm, rhsForm, consistency)
     result1 = mempty { histMaxDepth = M.fromList [(cat, toHist maxDepth),
@@ -121,15 +119,16 @@ classifyArrayCode ivs lhs rhses =
     -- Max and min depth
     (maxDepth, minDepth) = (maximum' >< minimum')
                          . unzip
-                         . map (maxMin . filter (/= absoluteRep)) $ rhsOffsets
+                         . map (maxMin . filter notAbsolute) $ rhsOffsets
     maxMin x = (maximum' x, minimum' x)
+    notAbsolute x = x /= absoluteRep && x /= -absoluteRep
 
-    maximum' :: (Ord a, Bounded a) => [a] -> a
-    maximum' [] = minBound
+    maximum' :: [Int] -> Int
+    maximum' [] = 0
     maximum' xs = maximum xs
 
-    minimum' :: (Ord a, Bounded a) => [a] -> a
-    minimum' [] = maxBound
+    minimum' :: [Int] -> Int
+    minimum' [] = 0
     minimum' xs = minimum xs
 
     -- Affine scalar multiples

### 2017-04-11 Results.hs

@@ -25,6 +25,8 @@ data Position = OverOrigin | StraddleOrigin | Elsewhere
 joinPosition :: Position -> Position -> Position
 joinPosition OverOrigin OverOrigin = OverOrigin
 joinPosition StraddleOrigin StraddleOrigin = StraddleOrigin
+joinPosition OverOrigin StraddleOrigin = OverOrigin
+joinPosition StraddleOrigin OverOrigin = OverOrigin
 joinPosition _ _ = Elsewhere
 
 data Contig   = SingleNonContig | NonContig | Contig

### 2017-04-11 oopsla17/analysis/Analyse.hs
categorisation of results based on what will likely be camfortable

@@ -117,8 +117,8 @@ applyAnalysisToDir restart dir debug bins excludes = do
                         resultsString <- Strict.readFile (restartFile ++ ".restart")
                         return $ ((read resultsString) :: Result)
                  Nothing -> mempty
-
-    (dbg, result) <- foldrM (applyAnalysisToFile dir) ("", result0) files
+    result1 <- return $ result0 { dirs = [dir] }
+    (dbg, result) <- foldrM (applyAnalysisToFile dir) ("", result1) files
 
     writeFile resultsFile (show result)
     if debug then putStrLn $ dbg else return ()

### 2017-04-11 oopsla17/analysis/Indices.hs

@@ -37,7 +37,6 @@ import Debug.Trace
 import Neighbour
 import Results
 
-
 -- Main function for classifying array statements
 classify :: FAD.InductionVarMapByASTBlock
          -> F.Expression (FA.Analysis A)
@@ -53,16 +52,13 @@ classify ivs lhs rhses = (debug, resultN)
              then mempty
              else foldr1 mappend (map (classifyArrayCode ivs lhs) (M.elems rhses))
 
-(><) :: (a -> b) -> (c -> d) -> (a, c) -> (b, d)
-f >< g = \(x, y) -> (f x, g y)
-
 -- LHS are neighbours
 classifyArrayCode ivs lhs rhses =
     (show cat ++ "\n", result)
   where
     cat    = (lhsForm, rhsForm, consistency)
-    result1 = mempty { histMaxDepth = M.fromList [(cat, toHist maxDepth),
-                                                  (cat, toHist minDepth)]
+    result1 = mempty { histMaxDepth = M.fromList [(cat, toHist maxDepth `histZip`
+                                                          toHist minDepth)]
                      , histDimensionality = mkHist cat dim
                      , histNumIndexExprs  = mkHist cat . toHist . length $ rhses
                      , counts             = mkHist cat 1 }

### 2017-04-11 oopsla17/analysis/Results.hs

@@ -181,9 +181,49 @@ joinConsistency _               _               = Inconsistent
 -- Overall categorisation
 type Cat = (Form LHS, Form RHS, Consistency)
 
+-- Predicate on whether a particular category is something
+-- we think we might target with CamFort
+class Camfortable t where
+  camfortable :: t -> Bool
+
+instance Camfortable (Form LHS, Form RHS, Consistency) where
+  camfortable (l, r, c) = camfortable l && camfortable r && camfortable c
+
+instance Camfortable Consistency where
+  camfortable (Consistent _)  = True
+  camfortable (LHSsuperset _) = True
+  camfortable _               = False
+
+instance Camfortable (Form LHS) where
+  camfortable Vars = True
+  camfortable (Neighbours _ L) = True
+  camfortable IVs = True
+  camfortable _   = False
+
+instance Camfortable (Form RHS) where
+  camfortable (Neighbours _ p) = camfortable p
+
+instance Camfortable (Physicality p) where
+  camfortable L     = True
+  camfortable (R p) = camfortable p
+
+instance Camfortable (Shape, Position, Contig, Reuse) where
+  camfortable (s, p, _, _) = camfortable s && camfortable p
+
+instance Camfortable Shape where
+  camfortable Orthotope = True
+  camfortable SumOfOrthotope = True
+  camfortable _              = False
+
+instance Camfortable Position where
+  camfortable OverOrigin     = True
+  camfortable StraddleOrigin = True
+  camfortable _              = False
+
 -- Results data type
 data Result = Result {
-    numLines              :: Int
+    dirs                  :: [String]
+  , numLines              :: Int
   , counts                :: M.Map Cat Int
   ------------------------------------
   -- Histograms
@@ -197,6 +237,15 @@ data Result = Result {
   , patternBin2D         :: M.Map [(Int,Int)] Int
   , patternBin3D         :: M.Map [(Int, Int, Int)] Int
   } deriving (Show, Read)
+
+camfortableResult :: HistogramShow t => M.Map Cat t -> M.Map String t
+camfortableResult =
+       M.fromList
[…]

### 2017-04-11 Results.hs
pattern match coverage bug fixed

@@ -202,6 +202,7 @@ instance Camfortable (Form LHS) where
 
 instance Camfortable (Form RHS) where
   camfortable (Neighbours _ p) = camfortable p
+  camfortable _                = False
 
 instance Camfortable (Physicality p) where
   camfortable L     = True

### 2017-04-11 oopsla17/oopsla17.tex
bug fixes; and dealing with different kinds of control flow in the right way
title and abstract tweak

@@
 
-\begin{document}
-\title{Understanding and Verifying the Shape of Stencil Computations}
+\title{Verifying Spatial Properties of Array Computations}
 
@@
 
-\maketitle
-
 \begin{abstract}
   Arrays computations are the workhorse of numerical modelling and
-  computational science.
-  Many practitioners are aware of need to ensure program
-  correctness, yet very few of the techniques from the programming
-  research community are applied by scientists. We aim to change that
-  by providing lightweight verification tools to scientists.
-  We focus on one error-prone aspect of numerical code: the 
-  data access pattern of arrays. 
-  Firstly, we report on a code analysis study on twelve real-world
-  computational science code base, identifying common idioms of array
-  usage and their spatial properties. This provides much needed data
-  on array programming idioms common in science, spanning roughly 1.5
-  million lines of code.
-  From this data, we 
-  designed a lightweight declarative specification language capturing
-  the majority of access patterns for arrays. We details its semantics
-  and algorithms for checking and inference.
-  We detail the design and implementation of a verification tool for
+  computational science. However, low-level manipulation of array
+  indices is a source of program error.  Many practitioners are aware
+  of the need to ensure program correctness, yet very few of the
+  techniques from the programming research community are applied by
+  scientists. We aim to change that by providing targetted lightweight
+  verification techniques for scientific code.  We focus on the all
+  too common mistake of array offset errors as a generalisation of
+  off-by-one errors.  Firstly, we report on a code analysis study on
+  twelve real-world computational science code base, identifying
+  common idioms of array usage and their spatial properties. This
+  provides much needed data on array programming idioms common in
+  scientific code.  From this data, we designed a lightweight
+  declarative specification language capturing the majority of array
+  access patterns via a small set of combinators. We detail a semantic
+  model, and the design and implementation of a verification tool for
   our specification language, which both checks and infers
-  specifications.
-  We evaluate our tool on the our previous corpus of scientific code
-  and give verification case studies of bugs that could be detected
-  with our approach. We find roughly 60,000 targets for specification, showing the vast majority
-  of stencil computations indeed have a simple, regular, static shape.
+  specifications.  We evaluate our tool on our corpus of scientific
+  code and give verification case studies of bug fixes that are
+  detected by our approach. We found roughly 60,000 targets for
+  specification across roughly 1.2 million lines of code, showing the
+  vast majority of array computations read from arrays in a pattern with
+  a simple, regular, static shape.
 \end{abstract}
 
- %S ubsequently, Fortran continues to be a
-%  popular language choice, due to its focus on arrays and its high
-%  performance.
+%  We focus on one error-prone aspect of numerical code: the 
+%  data access pattern of arrays.
 
-%  In this paper, we propose an alternate approach:
-%  a specification language and automated tool for verifying the
-%  spatial behaviour of stencils. The specification language is based
-%  on the hypothesis that stencil computations read from arrays with a
-%  regular, fixed pattern which can be captured by a simple
-%  set of abstract combinators. These
-%  specifications can be embedded as annotations in source code,
-%  against which code can be checked for conformance. For legacy code,
-%  spatial specifications can be inferred, providing documentation and
-%  aiding future software maintenance. We evaluate our specification
-%  language and verification tool against
-%  a corpus of numerical Fortran code ($\sim$ 1 million lines) for
-%  which we generate $\sim$60,000 specifications, showing the vast majority
-%  of stencil computations indeed have a simple, regular, static shape.
-%\end{abstract}
+\begin{document}
+\maketitle
 
@@
 
+\begin{itemize}
+% (Vars, ...)
+\item An assignment to a variable, with an RHS comprising some array
+  computation, tends to correspond to a reduction (e.g., calculating
+the max value in an array).
+
+\end{itemize}
+
 \section{A Specification Language for Shape}

### 2017-04-11 Analyse.hs
two small bug fixes to do with restarting

@@ -117,14 +117,14 @@ applyAnalysisToDir restart dir debug bins excludes = do
                         resultsString <- Strict.readFile (restartFile ++ ".restart")
                         return $ ((read resultsString) :: Result)
                  Nothing -> mempty
-    result1 <- return $ result0 { dirs = [dir] }
+    let result1 = result0 `mappend` mempty { dirs = [dir] }
     (dbg, result) <- foldrM (applyAnalysisToFile dir) ("", result1) files
 
     writeFile resultsFile (show result)
     if debug then putStrLn $ dbg else return ()
     putStrLn $ prettyResults result bins
-    valid <- resultValidation result
-    if valid then (putStrLn $ "Results were valid") else (putStrLn "Results were invalid")
+    --valid <- resultValidation result
+    --if valid then (putStrLn $ "Results were valid") else (putStrLn "Results were invalid")
 
 applyAnalysisToFile :: String
                     -> (Filename, SourceText, F.ProgramFile A)

### 2017-04-11 Results.hs

@@ -142,7 +142,7 @@ instance Read Consistency where
     ++ consistencyRead "Permutation" Permutation 0 xs
     ++ consistencyRead "LHSsubset"   LHSsubset 0 xs
     ++ consistencyRead "LHSsuperset" LHSsuperset 0 xs
-    ++ consistencyRead "Inconsitent" (\b -> Inconsistent) 0 xs
+    ++ consistencyRead "Inconsistent" (\b -> Inconsistent) 0 xs
 
 consistencyRead consString cons n xs =
     if consString `isPrefixOf` xs

### 2017-04-12 Results.hs
filling in details of the language and the model

@@ -71,7 +71,7 @@ instance Show HasConstants where
 
 instance Read HasConstants where
    readsPrec n xs | "+consts" `isPrefixOf` xs
-       = [(WithConsts, drop (length "+consts") xs)]
+       = [(WithConsts, drop (length "+consts ") xs)]
                   | otherwise = [(Normal, trim xs)]
 
 boolToHasConstants :: Bool -> HasConstants

### 2017-04-12 oopsla17/oopsla17.tex
filling in details of the language and the model

@@
 
-\usepackage{minted}
 \usepackage{hyperref}
@@
 
+\usepackage{minted}
+\usepackage{etoolbox}
+
 \usepackage{stencilmacros}
@@
 
-\fvset{fontsize=\scriptsize}
+\fvset{
+  linenos=true,
+  fontsize=\footnotesize,
+  breaklines=true,
+  breakafter=w),
+  xleftmargin=3\parindent,
+  numbersep=1em
+}
+
+\usemintedstyle{vs}
 
@@
 % renews the commands with actual data
-%\input{results}
+% \input{results}
+
+% Essential declarations
+\theoremstyle{definition}
+\newtheorem{defn}{Definition}
+
+\theoremstyle{plain}
+\newtheorem{thm}{Theorem}
+\newtheorem{lem}{Lemma}
+\newtheorem{prop}{Proposition}
+\newtheorem{case}{Case}
+
+\definecolor{darkgreen}{rgb}{0.0,0.5,0.0}
+\definecolor{darkpurple}{rgb}{0.6,0.0,0.6}
+\definecolor{orange}{rgb}{0.8,0.4,0.0}
+\newcommand{\dnote}[1]{\textcolor{darkpurple}{Dom: #1}}
+\newcommand{\mnote}[1]{\textcolor{darkgreen}{Mistral: #1}}
+\newcommand{\anote}[1]{\textcolor{red}{Andy: #1}}
 
@@
 
+Region constants specify a finite interval in a 
+single dimension relative to the origin and are either \term{pointed},
+\term{forward}, \term{backward}, or \term{centered}. The region names are
+inspired by numerical analysis terminology, \eg{} the standard explicit method for
+approximating PDEs is known as the \emph{Forward Time, Centered Space} (FTCS)
+scheme~\citep{dawson1991finite}.
+
+Each region
+constant has a dimension identifier $d$ given as a positive natural number.
+Each constant except \term{pointed} has a depth
+parameter $n$ given as a positive natural number; \term{pointed}
+regions implicitly have a depth of $0$.
+
+A \term{forward} region of depth $n$ specifies a contiguous
+region in dimension $d$ starting at the origin. This corresponds
+to specifying neighbourhood indices in dimension $d$ ranging from $i$ to $i + n$
+for some induction variable $i$. Similarly, a
+\term{backward} region of depth $n$ corresponds to contiguous indices
+from $i$ to $i - n$ and \term{centered}
+of depth $n$ from $i - n$ to $i + n$. A \term{pointed}
+stencil specifies a neighbour index $i$. For example, the
+following shows four specifications with four consistent stencil
+kernels reading from arrays \term{a}, \term{b}, \term{c} and \term{d}:
+%%
+\begin{minted}{fortran}
+!= stencil forward(depth=2, dim=1) :: a
+e(i, 0) = a(i, 0) + a(i+1, 0) + a(i+2, 0)
+
+!= stencil backward(depth=2, dim=1) :: b
+e(i, j) = b(i, j) + b(i-1, j) + b(i-2, j)
+
+!= stencil centered(depth=1, dim=1) :: c
+e(i, j) = (c(j-1) + c(j) + c(j+1))/3.0
+
+!= stencil pointed(dim=3) :: d
+e(i, j) = d(0, 0, i)
+\end{minted}
+%%
+Not every dimension needs to be specified, \eg{},
+specifications on lines $1$, $4$, and $10$ leave some dimensions unspecified.
+The \term{forward}, \term{backward}, and \term{centered} regions may
+all have an additional attribute \term{nonpointed} which marks absence
+of the origin.  For example, the following is a
+\term{nonpointed} \term{backward} stencil
+%
+\begin{minted}{fortran}
+!= stencil backward(depth=2, dim=1, nonpointed) :: a
+b(i) = a(i-1) + 10*a(i-2)
+\end{minted}
+%
+Not every dimension needs to be specified, \eg{},
+specifications on lines $1$, $3$, and $7$ leave some dimensions
+unspecified which leaves indices in these dimensions unconstrained.
+
+The \term{forward}, \term{backward}, and \term{centered} regions may
+all have an additional attribute \term{nonpointed} which marks absence
+of the origin.  For example, the following is a
+\term{nonpointed} \term{backward} stencil
+%
+\begin{minted}{fortran}
+!= stencil backward(depth=2, dim=1, nonpointed) :: a
+b(i) = a(i-1) + 10*a(i-2)
+\end{minted}
+
+\paragraph{Combining regions}
+
+The region operators \term{+} and \term{*} respectively combine
+regions by union and intersection. The intersection of two regions
+$r \term{*} s$ means that any indices in the specified code must be
+consistent with both $r$ and $s$ simultaneously.
+ Dually, for the union of two regions
+ $r \term{+} s$ means that indices in the specified code must be
+ in consistent with one of $r$ or $s$, or both.
+For example, the following \emph{nine-point stencil}
+has a specification given by the product of two \texttt{centered}
+regions in each dimension:
+%%
+\begin{minted}[breakindent=2.9em]{fortran}
+x = a(i, j)   + a(i-1, j)   + a(i+1, j)
+y = a(i, j-1) + a(i-1, j-1) + a(i+1, j-1)
+z = a(i, j+1) + a(i-1, j+1) + a(i+1, j+1)
+!= stencil centered(depth=1, dim=1) * centered(depth=1, dim=2) :: a
+b(i, j) = (x + y + z) / 9.0
+\end{minted}
+%
+%This pattern is common in image convolution applications.
+The specification ranges over the
+values that flow to the array subscript on the left-hand side,
+and so ranges over the intermediate assignments to \term{x},
+\term{y}, and \term{z}. Each index in the code is consistent
+with both specifications simultaneously, \eg{}, \texttt{a(i-1, j+1)}
+is within the centered region in dimension $1$ and the centered region
+in dimension $2$.
+
+The union of two regions $r \term{+} s$ means that any indices
+in the specified code must be consistent with either of $r$ or $s$.
+For example, the following gives the specification of a five-point
+stencil which is the sum of two compound \texttt{pointed} and
+\texttt{centered} regions in each dimension:
+%
+\begin{minted}{fortran}
+!= stencil centered(depth=1, dim=1)*pointed(dim=2) + centered(depth=1, dim=2)*pointed(dim=1) :: a
+b(i,j) = -4*a(i,j) + a(i-1,j) + a(i+1,j) + a(i,j-1) + a(i,j+1)
+\end{minted}
+%%
+Here the left-hand side of \texttt{+} says that when the second dimension
+(induction variable $j$) is fixed at the origin, the first dimension
+(induction variable $i$) accesses the immediate vicinity of the origin
+(to depth of one). The right hand side of \texttt{+} is similar but the dimensions are reversed.
+This reflects the symmetry under rotation of the five-point stencil.
+
+\paragraph{Region declarations and variables}
+
+Region specifications can be assigned to region variables via
+region declarations. For example, the shape of a
+``\emph{Roberts cross}'' edge-detection convolution~\cite{davis1975survey}
+can be stated:
+%%
+\begin{minted}{fortran}
+!= region :: r1 = forward(depth=1, dim=1)
+!= region :: r2 = forward(depth=1, dim=2)
+!= region :: robertsCross = r1*r2
+!= stencil robertsCross :: a
+\end{minted}
+This is useful for common patterns, such as the five-point
+pattern, as the regions can be defined once and reused.
+%%
+\paragraph{Modifiers}
+%%
+Region specifications can be modified
+by \emph{approximation} and \emph{multiplicity} information
+(in \textit{spec} in \Cref{fig:syntax}).
+The \texttt{readOnce} modifier enforces that no index appears more
+than once (that is, its multiplicity is one). For example, all of
+the previous examples could have \texttt{readOnce} added:
+%
+\begin{minted}{fortran}
+!= stencil readOnce, backward(depth=2, dim=1) :: a
+b(i+1) = a(i) + a(i-1) + a(i-2)
+\end{minted}
+%
+This specification would be invalid if any of the
+array subscripts were repeated. This modifier provides a way to
+rule out any accidental repetition of array subscripts.
+The notion is similar to that of linear types~\cite{wadler1990linear}, where a value must be used
+exactly once. We opt for the more informative and easily understood name
+\texttt{readOnce}. This modifier is optional, so it need not
+be present even if the stencil is linear.
+
+In some cases, it is useful to give a lower and/or upper bound for a
+stencil. This can be done using either the \term{atMost} or
+\term{atLeast} modifiers. This is particularly useful in situations
+where there is a non-contiguous stencil pattern, which cannot be expressed
+precisely in our syntax. For example:
+%
+\begin{minted}{fortran}
+!= stencil atLeast, pointed(dim=1)         :: a
+!= stencil atMost, forward(depth=4, dim=1) :: a
+b(i) = a(i) + a(i+4)
+\end{minted}
+
+\subsection{Semantic model}
+\label{sec:semantics}
+
+\noindent
+We define a simple set-theoretic denotational model of the semantics
+of our specification language. This has a number of purposes:
+(1) it serves to explain the meaning of our specifications
+(2) it is used in the inference and checking algorithms
+(\Cref{sec:analysis}); (3) justifies an equational theory for
+specifications in the next section; and (4) can be used to guide
+correct implementations.
+
+As an initial informal intuition, consider the following simple
+stencil computation.
+%
+\begin{minted}{fortran}
+do i = 1, n
+  b(i) = a(i,0) + a(i+1,0)
+end do
+\end{minted}
+%
+This is modelled by a vector of integer sets: $\langle{\{0, 1\},
+  \mathbb{Z}\}}\rangle$ describing that in the first dimension
+the array is read from at offsets of $0$ and $1$ from an induction
+variable, and in the second dimension the array reads are
+unconstrained. \dnote{Don't like this explanation.}
+
+we model neighbourhood array subscripts
+in the context of a loop by vectors of sets
+
+The model domain is over vectors of sets 
+which we call \emph{index schemes}.
+
+\begin{defn}[Extended integers]
+  We define the set \zinf{} as
+  $\mathbb{Z}$ extended with $\infty$ and $-\infty$. For any $a$ in
+  \zinf{}, we have $-\infty \leq a \leq \infty$. The resulting set is
+  a total order with top and bottom elements by $-\infty$ and
+  $\infty$.
+\end{defn}
+
+\begin{defn}[Index scheme]
+  An \emph{index scheme} is a vector of extended integer sets,
+  that is, an index scheme of size $n$ is a value
+  $\mathcal{P}(\zinf)^{n}$.
+\end{defn}
+
+\begin{defn}
+The function $\textit{schematic}_I$ parameterised by
+a set of induction variables $I$, which maps
+from an array subscript into a $n$-dimensional (closed) interval:
+%
+\begin{align*}
+\textit{schematic}_I(\bar{e}) & =
+\prod_{i \in |\bar{e}|} \textit{interval}(e_i)
+\qquad
+\textit{interval}_I(e) = \begin{cases}
+[c, c] & e = i \pm c \, \wedge \, i \in I \\
+[-\infty, \infty] & FV(e) \cap I = \emptyset
+\end{cases}
+\end{align*}
+Note that \textit{inveral} may be undefined (it is a partial function)
+in which case \textit{schematic} is undefined if any of its
+components are undefined.
+\end{defn}
+
 \section{Analysis, Checking, and Inference}

### 2017-04-12 oopsla17/analysis/Analyse.hs

@@ -80,62 +80,104 @@ import Control.DeepSeq
 import Results
 import Indices
 
+data Mode = SingleFile | ViewMode | NormalMode
+
 main :: IO ()
 main = do
     args <- getArgs
-    let (restart, args') = case args of
-                             (x:y:args') | x == "RESTART" -> (Just y, args')
-                             _                            -> (Nothing, args)
-    case args' of
-      [dir]
-        -> applyAnalysisToDir restart dir False False []
+    if null args then do
+       putStrLn "Example usages:"
+       putStrLn "    stencil-analysis dir-or-file [excluded-files]"
+       putStrLn "    stencil-analysis [-b] [-d] dir-or-file [excluded-files]"
+       putStrLn " (where -b means to print pattern bins, and -d is debug mode)"
+       putStrLn "    stencil-analysis RESTART rfile [-b] [-d] dir-or-file [excluded-files]"
+       putStrLn " (restart the analysis with rfile.restart)"
+       putStrLn "    stencil-analysis SINGLE rfile [-b] [-d] dir-or-file [excluded-files]"
+       putStrLn " (restart the analysis with rfile.restart and suprise the final file)"
+       putStrLn "    stencil-analysis VIEW rfile"
+    else do
+       let (restart, args', mode) =
+            case args of
+               (x:y:args') | x == "RESTART" -> (Just y, args', NormalMode)
+                           | x == "SINGLE"  -> (Just y, args', SingleFile)
+                           | x == "VIEW"    -> (Just y, [],    ViewMode)
+                           | otherwise      -> (Nothing, args, NormalMode)
+       case args' of
+          [] -> applyAnalysisToDir restart mode "" False False []
+          [dir]
+             -> applyAnalysisToDir restart mode dir False False []
 
-      dir:args@(a1:a2:_)
-        -> applyAnalysisToDir restart dir (a1 == "-d" || a2 == "-d")
+          dir:args@(a1:a2:_)
+             -> applyAnalysisToDir restart mode dir (a1 == "-d" || a2 == "-d")
                                           (a1 == "-b" || a2 == "-b")
                                           (filterFlags args)
 
-      dir:args@(a1:_)
-        -> applyAnalysisToDir restart dir (a1 == "-d")
+          dir:args@(a1:_)
+             -> applyAnalysisToDir restart mode dir (a1 == "-d")
                                           (a1 == "-b")
                                           (filterFlags args)
 
[…]

### 2017-04-12 oopsla17/oopsla17.tex
add call to sloccount in VIEW mode

@@
 further detail on these packages.
+
+We used Wheeler's \emph{SLOCCount} to get a count of the physical
+source lines (excluding comments and blank
+lines)~\cite{wheeler2001sloccount}; including them our corpus
+totals
 
@@
 
-\subsection{Semantic model}
-\label{sec:semantics}
+\section{Semantic model}
+\label{sec:model}
 
 \noindent
-We define a simple set-theoretic denotational model of the semantics
-of our specification language. This has a number of purposes:
+We define a simple set-theoretic model of array access patterns
+which serves as both a denotational model of the semantics
+of our specification language and an abstract interpretation
+domain for source code. This model has a number of uses: 
 (1) it serves to explain the meaning of our specifications
@@
 
-As an initial informal intuition, consider the following simple
-stencil computation.
+The model is defined over vectors of sets of integers which we call
+\emph{index schemes}.  As an initial informal example, consider the
+following simple stencil computation:
 %
@@
 %
-This is modelled by a vector of integer sets: $\langle{\{0, 1\},
-  \mathbb{Z}\}}\rangle$ describing that in the first dimension
-the array is read from at offsets of $0$ and $1$ from an induction
-variable, and in the second dimension the array reads are
-unconstrained. \dnote{Don't like this explanation.}
+The access pattern on array \term{a}, relative
+to induction variable \term{i}, is captured
+by a vector of length 2 containing integer sets $\langle{\{0, 1\},
+  \mathbb{Z}\}}\rangle$. This describes that, in the first dimension,
+the array is read at offsets of $0$ and $1$ from an induction
+variable. In the second dimension, the index is unconstrained
+as it is a constant.
 
-we model neighbourhood array subscripts
-in the context of a loop by vectors of sets
+Index schemes form a lattice which provides a rich set of equations
+and properties, which we exploit. We first set up the 
+domain of the model (\S\ref{sec:domain}), using it to define a semantics
+for our specification language (\S\ref{sec:semantics}) and then as
+the target for an abstract interpretation on imperative
+code (\S\ref{sec:fromcode}).
 
-The model domain is over vectors of sets 
-which we call \emph{index schemes}.
+\subsection{Lattice model of regions}
+\label{sec:domain}
 
@@
   \zinf{}, we have $-\infty \leq a \leq \infty$. The resulting set is
-  a total order with top and bottom elements by $-\infty$ and
-  $\infty$.
+  a total order with top and bottom elements $-\infty$ and
+  $\infty$ respectively.
 \end{defn}
@@
 \begin{defn}[Index scheme]
-  An \emph{index scheme} is a vector of extended integer sets,
-  that is, an index scheme of size $n$ is a value
-  $\mathcal{P}(\zinf)^{n}$.
+  An $N$-dimensional \emph{index scheme} is a vector of length $N$ of
+  extended integer sets, \ie{}, a member of $\mathcal{P}(\zinf)^{N}$.
+  An alternate (equivalent) view is of a $N$-times finite Cartesian
+  product on subsets of $\zinf$.  We use $S, T, U$ to denote index
+  schemes.
+
+Index schemes can be \emph{projected} in the $i^{\textit{th}}$ dimension by
+  $\pi_i : {\mathcal{P}(\zinf{})}^N \to \mathcal{P}(\zinf{})$ where $1 \leq i
+  \leq N$. For an index scheme $I$, we refer to $\pi_i(S)$ as the
+  \emph{$i^{th}$ component} of $S$.
 \end{defn}
+
+%\begin{theorem}
+%$N$-dimensional index schemes form a lattice where join and m
+%\end{theorem}
+
+\begin{lem}{}\label{lem:vector-intersect}
+  Intersection distributes over $N$-dimensional index schemes. 
+  That is, for index schemes $I, J \in \mathcal{P}(\zinf)^N$ then:
+%
+  \begin{equation*}
+    I \cap J = \prod_{i = 1}^{N} \pi_i(I) \cap \pi_i(J)
+  \end{equation*}
+\end{lem}
+
+ Union does not distribute over $N$-dimensional index schemes--
+ however, a more restricted property holds.
+
+\begin{lem}{}\label{lem:vector-union}
+Let $I, J$ be $N$-dimensional index schemes such that
+$\pi_i(I) = \pi_i(J)$ for all $1 \leq i \leq N$ apart 
+from some dimension $k$ where $\pi_k(I) \neq \pi_k(J)$,
+then:
+%
+  \begin{equation*}
+    I \; \cup \; J
+    =
+    \pi_1(I) \times \cdots \times
+    (\pi_k(I) \cup \pi_k(J)) \times \cdots \times
+    \pi_N(I)
+  \end{equation*}
+%
+\end{lem}
+%
+\paragraph{Intervals}
+
+We define an extended notion of closed interval on
+
+\begin{defn}{(\emph{Interval subspace})}
+  An interval subspace is a finite Cartesian product of intervals on \zinf{},
+  denoted by the set $\textit{Interval}^N$ for a product of $N$ intervals. An
+  interval subspace is an offset subspace.
+\end{defn}
+
+\subsection{Denotational semantics for specifications}
+\label{sec:semantics}
+
+\subsection{Abstract interpretation}
+\label{sec:fromcode}
 

### 2017-04-12 oopsla17/analysis/Analyse.hs

@@ -71,6 +71,8 @@ import Data.Maybe
 import Data.List
 import Data.Monoid
 import Data.Traversable
+import System.Process
+
 
 import Debug.Trace
 import System.Directory
@@ -147,7 +149,11 @@ applyAnalysisToDir restart mode dir debug bins excludes = do
     if debug then putStrLn $ dbg else return ()
 
     case mode of
-        ViewMode -> putStrLn $ prettyResults result True
+        ViewMode -> do
+               putStrLn $ prettyResults result True
+               -- Use sloccount to give a total of the physical lines
+               sloccount <- readProcess "sloccount" (dirs result) ""
+               putStrLn sloccount
 
         NormalMode -> do
                let resultsFile = dir ++ ".stencil-analysis"

### 2017-04-12 oopsla17/oopsla17.tex
moving the model into the main text and adapting

@@
 \usepackage{etoolbox}
+
+\usepackage{thmtools}
+\usepackage{thm-restate}
 
@@
 source lines (excluding comments and blank
-lines)~\cite{wheeler2001sloccount}; including them our corpus
-totals
+lines)~\cite{wheeler2001sloccount}; including non-physical lines
+ our corpus totals \dnote{insert}.
 
@@
 
-\begin{defn}[Extended integers]
-  We define the set \zinf{} as
-  $\mathbb{Z}$ extended with $\infty$ and $-\infty$. For any $a$ in
-  \zinf{}, we have $-\infty \leq a \leq \infty$. The resulting set is
-  a total order with top and bottom elements $-\infty$ and
-  $\infty$ respectively.
-\end{defn}
+%\begin{defn}[Extended integers]
+%  We define the set \zinf{} as
+%  $\mathbb{Z}$ extended with $\infty$ and $-\infty$. For any $a$ in
+%  \zinf{}, we have $-\infty \leq a \leq \infty$. The resulting set is
+%  a total order with top and bottom elements $-\infty$ and
+%  $\infty$ respectively.
+%\end{defn}
 
@@
   An $N$-dimensional \emph{index scheme} is a vector of length $N$ of
-  extended integer sets, \ie{}, a member of $\mathcal{P}(\zinf)^{N}$.
+  extended integer sets, \ie{}, a member of $\mathcal{P}(\bz)^{N}$.
   An alternate (equivalent) view is of a $N$-times finite Cartesian
-  product on subsets of $\zinf$.  We use $S, T, U$ to denote index
+  product on subsets of $\bz$.  We use $S, T, U$ to denote index
   schemes.
@@
 Index schemes can be \emph{projected} in the $i^{\textit{th}}$ dimension by
-  $\pi_i : {\mathcal{P}(\zinf{})}^N \to \mathcal{P}(\zinf{})$ where $1 \leq i
+  $\pi_i : {\mathcal{P}(\bz{})}^N \to \mathcal{P}(\bz{})$ where $1 \leq i
   \leq N$. For an index scheme $I$, we refer to $\pi_i(S)$ as the
@@
 
-\begin{lem}{}\label{lem:vector-intersect}
+\begin{restatable}{lem}{vectorIntersect}
+\label{lem:vector-intersect}
   Intersection distributes over $N$-dimensional index schemes. 
-  That is, for index schemes $I, J \in \mathcal{P}(\zinf)^N$ then:
+  That is, for index schemes $S, T \in \mathcal{P}(\bz)^N$
 %
   \begin{equation*}
-    I \cap J = \prod_{i = 1}^{N} \pi_i(I) \cap \pi_i(J)
+    S \cap T = \prod_{i = 1}^{N} \pi_i(S) \cap \pi_i(T)
   \end{equation*}
-\end{lem}
+\end{restatable}
 
@@
 
-\begin{lem}{}\label{lem:vector-union}
-Let $I, J$ be $N$-dimensional index schemes such that
-$\pi_i(I) = \pi_i(J)$ for all $1 \leq i \leq N$ apart 
-from some dimension $k$ where $\pi_k(I) \neq \pi_k(J)$,
+\begin{restatable}{lem}{vectorUnion}
+\label{lem:vector-union}
+Let $S, T$ be $N$-dimensional index schemes such that
+$\pi_i(S) = \pi_i(T)$ for all $1 \leq i \leq N$ apart 
+from some dimension $k$ where $\pi_k(S) \neq \pi_k(T)$,
 then:
@@
   \begin{equation*}
-    I \; \cup \; J
+    S \; \cup \; T
     =
-    \pi_1(I) \times \cdots \times
-    (\pi_k(I) \cup \pi_k(J)) \times \cdots \times
-    \pi_N(I)
+    \pi_1(S) \times \cdots \times
+    (\pi_k(S) \cup \pi_k(T)) \times \cdots \times
+    \pi_N(S)
   \end{equation*}
 %
-\end{lem}
+\end{restatable}
 %
-\paragraph{Intervals}
 
-We define an extended notion of closed interval on
+\begin{defn}
+  We define an extended notion of closed interval on $\bz{}$ which may contain
+  a \emph{hole} at the origin, written \interv{a}{b}{c} where $a$ and $b$ are
+  drawn from \bz{} with $a \leq 0 \leq b$
+   and $c$ is drawn from $\mathbb{B} = \{ \mathsf{true}, \mathsf{false}
+  \}$. Intervals are interpreted as sets as follows:
+%
+  \begin{equation*}
+    \interv{a}{b}{c} \triangleq
+      \{ n \mid a \leq n \leq b \wedge (\neg c \implies n \neq 0) \}
+  \end{equation*}
+%
+  We denote the set of all such intervals (sets) as
+   $\textit{Interval}$. If the superscript is
+  omitted it is treated as $\mathsf{true}$.
+\end{defn}
+
+\begin{restatable}{lem}{intervalIdentities}
+ \label{lem:interval-identities}
+  We have the following dual identities for \bz{} intervals:
+%
+  \begin{align*}
+    \interv{a}{b}{c} \cap \interv{d}{e}{f} & =
+      \interv{\max \{a,d\}}{\min \{b,e\}}{c \wedge f} \\
+    \interv{a}{b}{c} \cup \interv{d}{e}{f} & =
+      \interv{\min \{a,d\}}{\max \{b,e\}}{c \vee f}
+  \end{align*}
+\end{restatable}
+%
 
 \begin{defn}{(\emph{Interval subspace})}
-  An interval subspace is a finite Cartesian product of intervals on \zinf{},
-  denoted by the set $\textit{Interval}^N$ for a product of $N$ intervals. An
-  interval subspace is an offset subspace.
+  An interval subspace is a finite Cartesian product of intervals on
+  \bz{} or \bz itself, 
+  denoted by the set $\textit{Interval}^N$ for a product of $N$ intervals.
 \end{defn}
@@
 [c, c] & e = i \pm c \, \wedge \, i \in I \\
-[-\infty, \infty] & FV(e) \cap I = \emptyset
+\bz & FV(e) \cap I = \emptyset
 \end{cases}

### 2017-04-12 oopsla17/proofs.tex
moving the model into the main text and adapting

--- 2017-04-10 oopsla17/lattice-model.tex
@@
-\documentclass[acmlarge,review]{acmart}
+\newcommand{\off}{S}
+\newcommand{\offP}{T}
 
-% Essential packages
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{amsthm}
-\usepackage{indentfirst}
-\usepackage{cleveref}
-\usepackage{stencilmacros}
+\vectorIntersect*
 
-% Essential declarations
-\theoremstyle{definition}
-\newtheorem{defn}{Definition}
+\begin{proof}
+  \begin{align*}
+    \off \; \cap \; \offP &
+    = \{x \mid \bigwedge_{1 \leq i \leq N } x_i \in \pi_i(\off) \}
+      \;\cap\;
+      \{x \mid \bigwedge_{1 \leq i \leq N } x_i \in \pi_i(\offP) \} \\
+    & = \{x \mid \bigwedge_{1 \leq i \leq N }
+      (x_i \in \pi_i(\off) \wedge x_i \in \pi_i(\offP)) \} \\
+    & = \prod_{i = 1}^{N} \pi_i(\off) \cap \pi_i(\offP)
+  \end{align*}
+\end{proof}
 
-\theoremstyle{plain}
-\newtheorem{thm}{Theorem}
-\newtheorem{lem}{Lemma}
-\newtheorem{prop}{Proposition}
-\newtheorem{case}{Case}
+\vectorUnion*
 
-\theoremstyle{remark}
-\newtheorem{remark}{Remark}
+\begin{proof}
+  \begin{align*}
+    \off \; \cup \; \offP &
+    = \{x \mid
+      \bigwedge_{1 \leq i \leq N } x_i \in \pi_i(\off) \}
+      \;\cup\;
+      \{x \mid
+          \bigwedge_{1 \leq i \leq N } x_i \in \pi_i(\offP) \} \\
+    & = \{x \mid
+          \bigwedge_{\substack{1 \leq i \leq N \\ i \neq k}}
+            x_i \in \pi_i(\off) \wedge x_k \in \pi_k(\off) \vee
+          \bigwedge_{\substack{1 \leq i \leq N \\ i \neq k}} x_i \in
+            \pi_i(\offP) \wedge x_k \in \pi_k(\offP)
+        \} \\
+    & = \{x \mid
+          \bigwedge_{\substack{1 \leq i \leq N \\ i \neq k}} x_i \in
+            \pi_i(\off) \wedge
+            x_k \in \pi_k(\off) \cup \pi_k(\offP)
+        \} \\
+        & = \pi_1(\off) \times \cdots \times
+        (\pi_k(\off) \cup \pi_k(\offP)) \times \cdots \times
+        \pi_N(\off)
+  \end{align*}
+\end{proof}
 
-\begin{document}
+\intervalIdentities*
 
-\title{Lattice model}
-\author{Mistral Contrastin}
-
-\section{Giving specifications meaning}
-
-We construct a lattice and use this lattice to give meanings to regions. We
-then give full meaning to specifications by wrapping regions in variant types
-for multiplicity and approximation. Model defined in this section will later be
-used for two purposes. First, we show soundness of the equational theory of
-specifications with respect to the model. Second, we define inference and
-consistency checking procedures of specifications with respect to the
-interpretations of the stencils in the code and the specifications.
-
-\subsection{Constructing the lattice model of regions}
-
-\begin{defn}
-  We define the set \zinf{} as
-  $\mathbb{Z}$ extended with $\infty$ and $-\infty$. For any $n$ in
-  \zinf{}, we have $-\infty \leq n \leq \infty$. The resulting set is
-  a total order with top and bottom elements.
-\end{defn}
-
-\begin{defn}
-  We define an extended notion of closed interval on $\zinf{}$ which may contain
-  a \emph{hole} at the origin, written \interv{a}{b}{c} where $a$ and $b$ are
-  drawn from \zinf{} with $a \leq 0 \leq b$, $a$ is $-\infty$ \emph{iff} $b$ is
-  $\infty$, and $c$ is drawn from $\mathbb{B} = \{ \mathsf{true}, \mathsf{false}
-  \}$. We define it as follows:
-%
-  \begin{equation*}
-    \interv{a}{b}{c} \triangleq
-      \{ n \mid a \leq n \leq b \wedge (\neg c \implies n \neq 0) \}
-  \end{equation*}
-
-  Set of all such intervals (sets) is $\textit{Interval}$. If the superscript is
-  leftout it is treated as $\mathsf{true}$.
-\end{defn}
-
-\begin{lem}{}\label{lem:zinf-identities}
-  We have the following dual identities for \zinf{} intervals:
-%
-  \begin{align*}
-    \interv{a}{b}{c} \cap \interv{d}{e}{f} & =
-      \interv{\max \{a,d\}}{\min \{b,e\}}{c \wedge f} \\
-    \interv{a}{b}{c} \cup \interv{d}{e}{f} & =
-      \interv{\min \{a,d\}}{\max \{b,e\}}{c \vee f}
-  \end{align*}
-\end{lem}
-%
 \begin{proof}
@@
 \end{proof}
-
-\begin{defn}{(\emph{Offset subspace})}
-  An $N$-dimensional \emph{offset subspace} is a finite Cartesian product of $N$
-  subsets of $\zinf{}$. The set of $N$-dimensional offset subspaces is written
-  ${\mathcal{P}(\zinf{})}^N$.
-
-  A offset spaces can be \emph{projected} in the $i^{\textit{th}}$ dimension by
-  $\pi_i : {\mathcal{P}(\zinf{})}^N \to \mathcal{P}(\zinf{})$ where $1 \leq i
-  \leq N$.
-\end{defn}
-
-\begin{defn}{(\emph{Subscript})}
-  An $N$-dimensional offset subspace $S$ is called a \emph{subscript} if it
-  satisfies:
-%
-  \begin{equation*}
-    \forall i.\ 1 \leq i \leq N \implies
-      \pi_i (S) = \{p\} \wedge p \in \mathbb{Z}
-      \vee
-      \pi_i (S) = \interv{-\infty}{\infty}{\mathsf{true}}
-  \end{equation*}
-  %
-  That is, the $i^{th}$ component of the set is either a singleton
-  in $\mathbb{Z}$ or the infinite interval.
-
-  \todo{%
-    We can move this explanation.
-
-    Subscript spaces model array subscript expressions in our source langauge
-    where each subscripted expression is either a constant translation of an
-    induction variable within a loop or some other expression which is constant
-    with respect to the loop e.g. \texttt{a(i+1, 0)}.
-  }
-\end{defn}
-
-\begin{lem}{}\label{lem:vector-intersect}
-  Intersection distributes over sets of points. If $\mathit{off}$ and
-  $\mathit{off'}$ are sets of $N$-dimensional points, then
-%
-  \begin{equation*}
-    \mathit{off} \cap \mathit{off'} =
-      \prod_{i = 1}^{N} \pi_i(\mathit{off}) \cap \pi_i(\mathit{off'})
-  \end{equation*}
-\end{lem}
-%
-\begin{proof}
-  \begin{align*}
-    \mathit{off} \; \cap \; \mathit{off'} &
-    = \{x \mid \bigwedge_{1 \leq i \leq N } x_i \in \pi_i(\mathit{off}) \}
-      \;\cap\;
-      \{x \mid \bigwedge_{1 \leq i \leq N } x_i \in \pi_i(\mathit{off'}) \} \\
-    & = \{x \mid \bigwedge_{1 \leq i \leq N }
-      (x_i \in \pi_i(\mathit{off}) \wedge x_i \in \pi_i(\mathit{off'})) \} \\
-    & = \prod_{i = 1}^{N} \pi_i(\mathit{off}) \cap \pi_i(\mathit{off'})
-  \end{align*}
-\end{proof}
-
-\begin{lem}{}\label{lem:vector-union}
-  Let $\mathit{off}$ and $\mathit{off'}$ be two sets of
-  points such that $\pi_i(\mathit{off})$ is equal to $\pi_i(\mathit{off'})$ for all $1
-  \leq i \leq N$ except possibly at $k$, then we have the following:
-%
-  \begin{equation*}
-    \mathit{off} \; \cup \; \mathit{off'}
-    =
-    \pi_1(\mathit{off}) \times \cdots \times
-    (\pi_k(\mathit{off}) \cup \pi_k(\mathit{off'})) \times \cdots \times
-    \pi_N(\mathit{off})
-  \end{equation*}
-\end{lem}
-%
-\begin{proof}
-  \begin{align*}
-    \mathit{off} \; \cup \; \mathit{off'} &
-    = \{x \mid
-      \bigwedge_{1 \leq i \leq N } x_i \in \pi_i(\mathit{off}) \}
-      \;\cup\;
-      \{x \mid
-          \bigwedge_{1 \leq i \leq N } x_i \in \pi_i(\mathit{off'}) \} \\
-    & = \{x \mid
-          \bigwedge_{\substack{1 \leq i \leq N \\ i \neq k}}
-            x_i \in \pi_i(\mathit{off}) \wedge x_k \in \pi_k(\mathit{off}) \vee
-          \bigwedge_{\substack{1 \leq i \leq N \\ i \neq k}} x_i \in
-            \pi_i(\mathit{off'}) \wedge x_k \in \pi_k(\mathit{off'})
-        \} \\
-    & = \{x \mid
-          \bigwedge_{\substack{1 \leq i \leq N \\ i \neq k}} x_i \in
-            \pi_i(\mathit{off}) \wedge
-            x_k \in \pi_k(\mathit{off}) \cup \pi_k(\mathit{off'})
-        \} \\
-        & = \pi_1(\mathit{off}) \times \cdots \times
-        (\pi_k(\mathit{off}) \cup \pi_k(\mathit{off'})) \times \cdots \times
-        \pi_N(\mathit{off})
-  \end{align*}
-\end{proof}
-
-\begin{defn}{(\emph{Interval subspace})}
-  An interval subspace is a finite Cartesian product of intervals on \zinf{},
-  denoted by the set $\textit{Interval}^N$ for a product of $N$ intervals. An
-  interval subspace is an offset subspace.
-\end{defn}
-
-\begin{defn}{(\emph{Region})}
-  A region \region{N} is a set of $N$-dimensional point intervals defined as
-  the smallest set satisfying the following:
-%
-  \begin{enumerate}
-    \item If $R$ is in $\textit{Interval}^N$, then $R$ is in \region{N}.
-    \item If $R$ and $S$ are in \region{N}, then so are $R \cap S$ and
-      $R \cup S$.
-  \end{enumerate}
-\end{defn}
-
-\begin{prop}{}
-  $(\region{N},\subseteq)$ is a distributive lattice.
-\end{prop}
-%
-\begin{proof}
-  Any sublattice of a distributive lattice is a distributive lattice. Power sets
-  partially ordered with subset relation are distributive lattices. \region{N}
-  is a subset of $\mathcal{P}{(\zinf{}^N)}$ since all interval subspaces are
-  subsets of this power set and; union and intersection cannot add an element
-  that is not already in the powerset. Inductive definition also ensures that
-  \region{N} is closed under these operations. Therefore, \region{N} is a
-  distributive sublattice of $(\mathcal{P}{(\zinf{}^N)}, \subseteq)$.
-\end{proof}
-
-\begin{defn}{}
-  $\mathsf{Mult}$ and $\mathsf{Approx}$ are parametric labelled variant types
-  with injections given by their definition:
-%
-  \begin{align*}
-    \mathsf{Mult} \; a &
-      \triangleq \mathsf{mult} \; a \;\mid\; \mathsf{only} \; a \\
-    \mathsf{Approx} \; a &
-      \triangleq \mathsf{exact} \; a \;\mid\; \mathsf{lower} \; a \;\mid\;
-        \mathsf{upper} \; a \;\mid\; \mathsf{both} \; a \; a
-  \end{align*}
-%
-  \eg{}, $\mathsf{lower}$ is an injection $\mathsf{lower} : a \to \mathsf{Approx}
-  \; a$ etc.
-
-  These will be used in the following subsection to give meaning to the
-  specification modifiers for approximation and multiplicity.
-\end{defn}
-
-\subsection{Denotational semantics for specifications}
-
-We give some basic interpretations: $\interp{\texttt{n}} = n$ where $n$ is in
-$\mathbb{Z}^+$, $\interp{\texttt{pointed}} = \llbracket \epsilon \rrbracket =
-\mathsf{true}$, $\interp{\texttt{nonpointed}} = \mathsf{false}$.
-Typographically, \textcap{bold monotype font} is used within specifications
-to indicate a partial textual capture. For convenience, we use \textit{italic
-serif font} inside specifications rather than interpreting its textual capture.
-This coalesces syntactic and semantic representation but avoids clutter in the
-presentation.
-
-Let $\textit{vec-gen}_N : \mathbb{N}^+ \times \textit{Interval} \to
-\textit{Interval}^N$ be the function generating an $N$ dimensional vector
-interval such that \vecgen{N}{i}{\interv{a}{b}{c}} has
-$\interv{-\infty}{\infty}{\mathsf{true}}$ in all dimensions except the $i^{th}$
-one. There, it has \interv{a}{b}{c}.
-
-First, we give meaning to region constants (introduced in \Cref{}).
-%
-\begin{align*}
-  \interp{\stencil{p}{$i$}{}{}}_N & =
-    \vecgen{N}{i}{\interv{0}{0}{\mathsf{true}}}\\
-%
-  \interp{\stencil{c}{$i$}{$k$}{\textcap{p}}}_N & =
-    \vecgen{N}{i}{\interv{-k}{k}{\interp{\textcap{p}}}} \\
-%
-  \interp{\stencil{f}{$i$}{$k$}{\textcap{p}}}_N & =
-    \vecgen{N}{i}{\interv{0}{k}{\interp{\textcap{p}}}} \\
-%
-  \interp{\stencil{b}{$i$}{$k$}{\textcap{p}}}_N & =
-  \vecgen{N}{i}{\interv{-k}{0}{\interp{\textcap{p}}}}
-\end{align*}
-
-Second, we provide the semantics for the permissive region operator, \texttt{+},
-and the restrictive region operator, \texttt{*}, by associating them with join
-and meet respectively. Let \textcap{R} and \textcap{S} be regions.
-%
-\begin{align*}
-  \interp{\texttt{\textcap{R} + \textcap{S}}}_N & =
-    \interp{\textcap{R}}_N \vee \interp{\textcap{S}}_N
-&
-  \interp{\texttt{\textcap{R} * \textcap{S}}}_N & =
-    \interp{\textcap{R}}_N \wedge \interp{\textcap{S}}_N
-\end{align*}
-
-Third, we need to mark presence of modifiers such as \texttt{readOnce} and
-\texttt{atMost} as introduced in \Cref{}. In the context of
-multiplicity modifiers we have the following semantics:
-%
-\begin{align*}
-  \interp{\texttt{readOnce}} & = \mathsf{mult} &
-  \llbracket\epsilon\rrbracket & = \mathsf{once}
-\end{align*}
-
-In the context of approximation modifiers, we have
-%
-\begin{align*}
-  \interp{\texttt{atLeast}} & = \mathsf{lower} &
-  \interp{\texttt{atMost}} & = \mathsf{upper} &
-  \llbracket \epsilon \rrbracket = \mathsf{exact}
-\end{align*}
-
-Finally, we can give full semantics of a stencil specification:
-%
-\begin{equation*}
-  \interp{\texttt{stencil \textcap{Mult}, \textcap{Appr}, \textcap{Region}}}_N =
-    \interp{\textcap{Mult}}
-           {(\interp{\textcap{Appr}}
-                    {(\interp{\textcap{Region}}_N)})}
-\end{equation*}
-
-\section{Equational \& approximation theories}
-
-We know introduce the equational and approximation theory of stencil
-specifications based on $\equiv$ and $\preceq$ in the following two subsections.
-
-\subsection{Equivalences}
-
-We define an equivalence relation, $\equiv$, on regions as follows:
-
-\begin{description}
-  \item[Basic] \texttt{*} and \texttt{+} are both idempotent, commutative, and
-    associative;
-%
-  \item[Subsumption] If $S$ and $R$ are regions with $S \preceq R$, then
-    $S \texttt{+} R \equiv R$ and $S \texttt{*} R \equiv S$.
-%
-  \item[Dist] \texttt{*} distributed over \texttt{+} and dually
-    \texttt{+} distributes over \texttt{*}, meaning if \textcap{R}, \textcap{S},
-    and \textcap{T} are regions, then we have the following dual equivalences:
-%
-    \begin{align*}
-      \texttt{\textcap{R}*(\textcap{S}+\textcap{T})} & \equiv
-        \texttt{(\textcap{R}*\textcap{S})+(\textcap{R}*\textcap{T})} &
-      \texttt{\textcap{R}+(\textcap{S}*\textcap{T})} & \equiv
-        \texttt{(\textcap{R}+\textcap{S})*(\textcap{R}+\textcap{T})}
-    \end{align*}
-%
-  \item[Overlapping pointed] If \textcap{R} is one of \texttt{forward},
-    \texttt{backward}, or \texttt{centered}, then we have
-%
-    \begin{equation*}
-      \stencil{\textcap{R}}{$n$}{$k$}{\texttt{nonpointed}} \;\texttt{+}\;
-      \stencil{p}{$n$}{}{} \equiv
-      \stencil{\textcap{R}}{$n$}{$k$}{}
-    \end{equation*}
-%
-  \item[Centered] The region constants \texttt{forward} and \texttt{backward}
-    are two halves of \texttt{centered} specifications:
-%
-    \begin{align*}
-      \stencil[s]{c}{$n$}{$k$}{\textcap{p1}} \equiv
-        \stencil[s]{f}{$n$}{$k$}{\textcap{p2}} \texttt{+}
-        \stencil[s]{b}{$n$}{$k$}{\textcap{p3}}
-    \end{align*}
-%
-    Here \textcap{p1} is \texttt{nonpointed} if both \textcap{p2} and
-    \textcap{p3} are \texttt{nonpointed} and \texttt{pointed} otherwise.
-\end{description}
-
-\begin{thm}{(Equational soundness)}
-  Equational theory is sound with respect to the lattice model. Let $R$
-  and $S$ be $N$ dimensional regions, then we have
-%
-  \begin{equation*}
-    R \equiv S \implies \interp{R}_N = \interp{S}_N
-  \end{equation*}
-\end{thm}
-
-\begin{proof}
-  Both $\equiv$ and $=$ are equivalence relations. We only need to show that
-  each property associated with specifications equivalence holds as an equality
-  in the model.
-%
-  \begin{description}
-    \item[Basic \& Subsumption] These follow from basic properties of lattices.
-%
-    \item[Dist] This is immediate from the definition of distributive
-      lattice which the model is one.
-%
-    \item[Overlapping pointed] Counterparts of individual regions for some
-      Boolean $p$ are given below:
-%
-      \begin{align*}
-        \interp{\stencil{\textcap{R}}{$n$}{$k$}{\texttt{nonpointed}}}_N
-        & = \vecgen{N}{n}{\interv{-k}{k}{\mathsf{false}}} \\
-%
-        \interp{\stencil{p}{$n$}{}{}}_N
-        &= \vecgen{N}{n}{\interv{0}{0}{\mathsf{true}}} \\
-%
-        \interp{\stencil{\textcap{R}}{$n$}{$k$}{}}_N
-        &= \vecgen{N}{n}{\interv{-k}{k}{\mathsf{true}}}
-      \end{align*}
-
-      Using \Cref{lem:vector-union}, it is sufficient to show
-      $ \interv{-k}{k}{p} \;\cup\; \interv{0}{0}{\mathsf{true}} =
-        \interv{-k}{k}{\mathsf{true}} $ holds. This immediately follows from
-      \Cref{lem:zinf-identities}.
-%
-    \item[Centered] Let $\interp{\textcap{p1}}$, $\interp{\textcap{p2}}$, and
-      $\interp{\textcap{p3}}$ be $p1$, $p2$, and $p3$ respectively. We
-      know that $p1$ is $p2 \vee p3$ from (\textbf{Centered}) detinition. Then
-      we get the following interpretation for region constants involved in the
-      equivalence:
-%
-      \begin{align*}
-        \interp{\stencil{c}{$k$}{$n$}{$p1$}}_N
-          &= \vecgen{N}{n}{\interv{-k}{k}{p2 \vee p3}} \\
-%
-        \interp{\stencil{f}{$k$}{$n$}{$p2$}}_N
-          &= \vecgen{N}{n}{\interv{0}{k}{p2} }\\
-%
-        \interp{\stencil{b}{$k$}{$n$}{$p3$}}_N
-          &= \vecgen{N}{n}{\interv{0}{k}{p3}}
-      \end{align*}
-%
-      Using \Cref{lem:vector-union}, it is sufficient to show $\interv{-k}{k}{p2
-      \vee p3} = \interv{0}{k}{p2} \cup \interv{-k}{0}{p3}$ holds. This
-      immediately follows from \Cref{lem:zinf-identities}.
-  \end{description}
-\end{proof}
-
-\subsection{Approximations}
-We define a partial of order of approximations, $\preceq$. This relation
-provides us with the following:
-
-\begin{description}
-  \item[Equivalence] If $S$ and $R$ are regions and $S \equiv R$, then we have
-    $S \preceq R$.
-%
-  \item[Combined] If $S$ and $R$ are regions, then we have
-    $S \preceq S \texttt{+} R$ and $S \texttt{*} R \preceq S$.
-%
-  \item[Depth] Let $k$ and $l$ be in positive integers and $k \leq l$, $n$ some
-    fixed dimension, and \textcap{p} either \texttt{pointed} or
-    \texttt{nonpointed}. Further, let \textcap{R} be one of \texttt{centered},
-    \texttt{forward}, and \texttt{backward}. We then have
-%
-    \begin{equation*}
-      \stencil{R}{$n$}{$k$}{\textcap{p}} \preceq \stencil{R}{$n$}{$l$}{\textcap{p}}
-    \end{equation*}
-%
-\end{description}
-
-\begin{thm}{(Approximation soundness)}
-  Theory of approximation is sound with respect to the lattice model. Let $R$
-  and $S$ be $N$ dimensional regions, then we have
-%
-  \begin{equation*}
-    R \preceq S \implies \interp{R}_N \subseteq \interp{S}_N
-  \end{equation*}
-\end{thm}
-%
-\begin{proof}
-  Both $\preceq$ and $\subseteq$ are partial orders on sets. We only need to
-  show that each inequality holds in the model.
-%
-  \begin{description}
-    \item[Equivalence] Region equavilance is modeled with set equality and the
-      partial order on regions is modeled with subset relation. Partial orders
-      are reflexive and the model of eqivalence agrees with that of the partial
-      order, so this rule holds.
-%
-    \item[Combined] \texttt{*} and \texttt{+} maps to meet and join operations
-      in the lattice. All meets and joins in a lattice satisfy these
-      inequalities.
-
-    \item[Depth] When $m$ is $k$ and $l$ in
-      $\stencil{\textcap{R}}{$n$}{$m$}{\textcap{p}}$, we obtain the
-      interpretations $\vecgen{N}{n}{\textit{int}_k}$ and
-      $\vecgen{N}{n}{\textit{int}_l}$ for some $N$ as interpretations of
-      regions. Call these \textit{ks} and \textit{ls} respectively. One way to
-      show $\textit{ks} \subseteq \textit{ls}$ is to show that the intersection
-      of \textit{ks} and \textit{ls} is \textit{ks}.
-
-      To show that the intersection is \textit{ks}, we use
-      \Cref{lem:vector-intersect} which reduces the problem to showing that the
-      intersection of $\textit{int}_k$ and $\textit{int}_l$ is the former since
-      at every point except $n$ the intervals agree, hence the intersection
-      remains the same at these dimensions. Then we use
-      \Cref{lem:zinf-identities} to calculate the intersection. By assumption we
-      have $k \leq l$, so the maximum of the lower bounds is $-k$ and the
-      minimum of lower bounds is $k$ this is same as interval $\textit{int}_k$.
-      Note that pointedness is the same for both intervals and depths are
-      positive integers, so the pointedness is preserved under intersection.
-      This shows the intersection of \textit{ls} and \textit{ks} is indeed
-      \textit{ks} and concludes the proof.
-  \end{description}
-\end{proof}
-
-We present few inequalities that can be derived from the axioms and are useful
-when writing specifications without inference:
-
-\begin{prop}{(Centered approximation)}
-  For any dimension $n$, depth $k$, and pointed attribute $p$,
-  we have
-%
-  \begin{align*}
-    \stencil{f}{$n$}{$k$}{\textcap{p}} & \preceq
-      \stencil{c}{$n$}{$k$}{\textcap{p}} \\
-%
-    \stencil{b}{$n$}{$k$}{\textcap{p}} & \preceq
-      \stencil{c}{$n$}{$k$}{\textcap{p}}
-  \end{align*}
-\end{prop}
-%
-\begin{proof}
-  \shortv{%
-    Both inequalities are proved by combining the missing half of centered by
-    using the inequality (\textbf{Combined}) and then merging them into a single
-    region constant using (\textbf{Centered}).
-  }
-
-  \longv{%
-    Proof of the first inequality:
-    \begin{align*}
-      \stencil[s]{f}{$n$}{$k$}{\textcap{p}}
-        & \preceq
-          \stencil[s]{f}{$n$}{$k$}{\textcap{p}}
-          \;\texttt{+}\;
-          \stencil[s]{b}{$n$}{$k$}{\textcap{p}}
-        & (\textbf{Combined}) \\
-      & \equiv \stencil[s]{c}{$n$}{$k$}{\textcap{p}} & (\textbf{Centered})
-    \end{align*}
-
-    Proof of the second inequality:
-    \begin{align*}
-      \stencil[s]{b}{$n$}{$k$}{\textcap{p}}
-        & \preceq
-          \stencil[s]{f}{$n$}{$k$}{\textcap{p}}
-          \;\texttt{+}\;
-          \stencil[s]{b}{$n$}{$k$}{\textcap{p}}
-        & (\textbf{Combined}) \\
-      & \equiv \stencil[s]{c}{$n$}{$k$}{\textcap{p}} & (\textbf{Centered})
-    \end{align*}
-  }
-\end{proof}
-
-\begin{prop}{(Point approximation)}
-  Let \textcap{R} be one of \texttt{forward}, \texttt{backward}, and
-  \texttt{centered}, $n$ a fixed dimension, and $k$ a fixed depth, then we have
-%
-  \begin{align*}
-    \stencil{p}{$n$}{}{} & \preceq \stencil{R}{$n$}{$k$}{} \\
-%
-    \stencil{R}{$n$}{$k$}{\texttt{nonpointed}} & \preceq
-      \stencil{R}{$n$}{$k$}{}
-  \end{align*}
-\end{prop}
-%
-\begin{proof}
-  \shortv{%
-    Both inequalities are proved by breaking \textcap{R} using
-    (\textbf{Overlapping pointed}) followed by the inequality
-    (\textbf{Combined}).
-  }
-
-  \longv{%
-    Proof of the first inequality:
-    \begin{align*}
-      \stencil[s]{p}{$n$}{}{}
-        & \preceq \stencil{R}{$n$}{$k$}{nonpointed}
-                  \;\texttt{+}\;
-                  \stencil[s]{p}{$n$}{}{}
-        & (\textbf{Combined}) \\
-      & \equiv \stencil{R}{$n$}{$k$}{} & (\textbf{Overlapping pointed})
-    \end{align*}
-
-    Proof of the second inequality:
-    \begin{align*}
-      \stencil{R}{$n$}{$k$}{nonpointed}
-        & \preceq \stencil{R}{$n$}{$k$}{nonpointed}
-                  \;\texttt{+}\;
-                  \stencil[s]{p}{$n$}{}{}
-        & (\textbf{Combined}) \\
-      & \equiv \stencil{R}{$n$}{$k$}{} & (\textbf{Overlapping pointed})
-    \end{align*}
-  }
-\end{proof}
-
-\section{Analysis, Consistency, Inference}
-
-\subsection{Analysis}
-
-\subsection{Consistency}
-
-Once the specification is synthesised into objects in the model and stencil
-under examination is analysed, consistency checking itself is remarkably simple.
-Two stage checking is outlined below.
-
-\begin{enumerate}
-  \item Multiplicity of the model is checked against the stencil. The only way
-    this fails is if the original specification involved \texttt{readOnce}, but
-    some indexing occur multiple times in the stencil.  Note that absence of
-    \texttt{readOnce} is interpreted liberally, and if the stencil indices
-    happen to occur uniquely, this does not constitute a failure.
-  \item Definition of vector intervals is completely unwinded into sets of
-    offset tuples and depending the approximation modifier (or lack thereof)
-    we compare these tuples to offsets of the stencil computation. Lower and
-    upper bounds are checked with subset and superset relations respectively,
-    while absence of \texttt{atMost} and \texttt{atLeast} results in exact
-    checking with set equality.
-\end{enumerate}
-
-Here is the mathematical definition of consistency checking:
-\begin{align*}
-  \mathit{consistent}(\mathit{ix}, \mathit{model}) & = \begin{cases}
-    \mathsf{false} & \mathit{model} = \mathsf{once}(x) \wedge ix = \mathsf{mult}(y) \\
-    \mathit{consistent'}(\mathit{peel}(ix), \mathit{peel}(model)) & \textit{otherwise}
-  \end{cases} \\
-%
-  \mathit{consistent'}(\mathit{ix}, \mathit{model}) & = \begin{cases}
-    m = \mathit{ix} & \mathit{model} = \mathsf{exact}(m) \\
-    m \supseteq \mathit{ix} & \mathit{model} = \mathsf{upper}(m) \\
-    m \subseteq \mathit{ix} & \mathit{model} = \mathsf{lower}(m)
-  \end{cases}
-\end{align*}
-
-Here helper function $\mathit{peel} : \mathsf{Approx} \; a \to a$ removes the
-approximation label.
-
-Due to use of infinite intervals, the model and index sets are potentially
-infinite meaning simple procedures used to compute equality, subset, and
-superset relations for finite sets are no avail. The subset and superset
-relations are handled by computing the intersection of $m$ and $\mathit{ix}$
-using \Cref{lem:vector-intersect}, then checking equality with $m$ and
-$\mathit{ix}$ respectively.
-
-\todo{Here we simply show \textit{consistency'} as set equality, but since sets
-are infinite we might need to elaborate based on representation.}
-
-\subsection{Inference}
-
-\begin{defn}{(Extended origin distance)}
-  We measure the \emph{extended origin distance} of a point from the extended
-  origin as the minimum number of translations by $\pm 1$ in one dimension at a
-  time, to reach a point in the extended origin.
-
-  This notion extends to subscripts. Since there may be an infinite number of
-  points, we measure the distance from the closest one.
-\end{defn}
-%
-\begin{prop}{}
-  The closed form of extend origin distance for an $N$-dimensional subscript $S$
-  is denoted with $\mathit{dist}(S)$ and is defined as:
-%
-  \begin{align*}
-    \mathit{dist}(S) & = \sum_{i=1}^{N} \mathit{dist}_i(S)
-    &
-    \mathit{dist}_i(S) & = \begin{cases}
-      0 & \pi_i(S) = \interv{-\infty}{\infty}{} \\
-      |\mathsf{a}| - 1 & \pi_i(S) = \{\mathsf{a}\} \\
-    \end{cases}
-  \end{align*}
-\end{prop}
-%
-\begin{proof}
-  Now we prove that function $\mathit{dist}$ indeed gives the lowest number of
-  unit translations to reach the extended origin.
-\end{proof}
-
-\begin{defn}{(Adjacent)}
-  A subscript $S$ is said to be \emph{adjacent at index $j$} of another
-  subscript $S'$ if for every index $i$ except $j$, $\pi_i(S) = \pi_i(S')$ but
-  at $j$ then $\pi_j(S) \neq \pi_j(S')$ such that $|S| = |S'| + 1$.
-\end{defn}
-%
-\begin{prop}
-  The unique adjacent at index $i$ of a subscript, if it exists, is given by the
-  following partial function:
-%
-  \begin{align*}
-    \mathit{adjacent}(S,i) & = \prod^{N}_{j=1} \mathit{shift}(S,i,j) &
-%
-    \mathit{shift}(S,i,j) & =
-      \begin{cases}
-        \{a - 1\} & i = j \wedge \pi_i(S) = \{\mathsf{a}\} \wedge \mathsf{a} > 0\\
-        \{a + 1\} & i = j \wedge \pi_i(S) = \{\mathsf{a}\} \wedge \mathsf{a} < 0\\
-        \pi_j(S)  & otherwise
-      \end{cases}
-  \end{align*}
-\end{prop}
-%
-\begin{proof}
-\todo{Hm- I guess uniqueness follows from this being a function rather
-  than a relation? Seems fine no problem to prove- }
-\end{proof}
-
-\todo{On the following, add some intutition first- what is this and
-  what is it for?}
-\begin{defn}{(Exact set)}
-  The \emph{exact set} at origin distance $d$, $\mathit{ES}_{P,d}$, of a group
-  of $N$ dimensional subscripts $S$ is defined below:
-%
-  \begin{align*}
-    \mathit{ES}_{S,0} = &
-      \{ s \in S \mid \forall \i.\ \pi_i(s) \in \{\{-1\}, \{0\}, \{1\}\} \} \\
-    \mathit{ES}_{S,d} = &
-      \{ s \in S \mid \forall i.\
-      \pi_i(s) \notin \{\{-1\}, \{0\}, \{1\}\} \implies
-      \exists s' \in \mathit{ES}_{S,d-1}.\ s' =
-         \mathit{adjacent}(s,i) \}
-  \end{align*}
-%
-  The overall \emph{exact set}, $\mathit{ES}_S$ is then
-  $\bigcup_{0 \leq d \leq D} \mathit{ES}_{S,d}$, where $D$ is the maximum origin
-  distance in $S$.
-\end{defn}
-
-\begin{lem}{(\emph{Connecting})}
-  The following three statements are equivalent for a set of points $P \subseteq
-  \mathbb{Z}^N$:
-%
-  \begin{enumerate}
-    \item $P$ is the same as its exact set $\mathit{ES}_P$;
-    \item $P$ when regarded as offsets from induction variables forms a
-      neighbourhood stencil;
-    \item $P$ is a subset of \region{N}.
-  \end{enumerate}
-\end{lem}
-
-\begin{thm}{(Soundness of exact shape inference)}
-  For a set of point offsets vectors $P$, the inference procedure for
-  neighbourhood stencils is sound if the following holds:
-%
-  \begin{equation*}
-    \forall P. \mathit{peel}(\mathit{infer}(P)) = m \implies
-      \mathit{consistent'}(P,m)
-  \end{equation*}
-\end{thm}
-
-\begin{thm}{(Completeness of inference)}\label{thm:inf-completeness}
-  Inference procedure can infer a specification for every stencil computation.
-\end{thm}
-%
-\begin{proof}
-  Would be nice to have.
-\end{proof}
-
-\begin{thm}{(Completeness of exact inference)}
-  Inference procedure can infer a specification for every stencil computation
-  with array indexing behaviour forms a superposition of hyperrectangles removed
-  at most one unit from the origin in any of its dimension.
-\end{thm}
-%
-\begin{proof}
-  Would be nice to have.
-\end{proof}
-
-\begin{thm}{(Optimality of inference)}\label{thm:inf-optimality}
-  For exact inference, inference procedure always infers specifications with
-  fewest number of region constants.
-\end{thm}
-%
-\begin{proof}
-  Would be nice to have.
-\end{proof}
-
-\begin{remark}{}
-  Optimality of general inference is trivial. We only need a single region
-  constant covering accesses in any one of the stencil's dimensions and that
-  would provide an upper bound for the stencil.
-\end{remark}
-
-\end{document}

### 2017-04-13 oopsla17/appendix.tex
moving the model into the main text and adapting
corpus data'

@@
 \label{app:corpus}
+
+\begin{table}
+\centering
+\footnotesize
+\begin{tabular}{lS[table-format=4]S[table-format=4]S[table-format=4]S[table-format=4]S[table-format=4]S[table-format=4]}
+\hline
+  & \multicolumn{2}{c}{Number of files}
+  & \multicolumn{2}{c}{Lines of physical code}
+  & \multicolumn{2}{c}{Lines of raw code} \\
+%%
+  Package   & {Total} & {Parsed}& {\hspace{2em}Total} & {Parsed} & {\hspace{2em}Total} & {Parsed}  \\ % braces prevent headers from being formatted as numbers
+  \hline
+%%%
+  ARPACK-NG & \SSArpackFiles & \SSArpackFilesParsed & \SSArpackLoC &
+\SSArpackLoCParsed & \SSArpackLoCRaw & \SSArpackLoCParsedRaw \\
+%%%
+  BLAS & \SSBlasFiles & \SSBlasFilesParsed &
+\SSBlasLoC & \SSBlasLoCParsed & \SSBlasLoCRaw & \SSBlasLoCParsedRaw \\
+%%%
+  Cliffs & \SSCliffsFiles & \SSCliffsFilesParsed & \SSCliffsLoC &
+\SSCliffsLoCParsed & \SSCliffsLoCRaw & \SSCliffsLoCParsedRaw \\
+%%%
+  CP & \SSCPFiles & \SSCPFilesParsed & \SSCPLoC & \SSCPLoCParsed &
+\SSCPLoCRaw & \SSCPLoCParsedRaw \\
+%%%
+  E3MG & \SSEThreeMEFiles & \SSEThreeMEFilesParsed & \SSEThreeMELoC &
+\SSEThreeMELoCParsed & \SSEThreeMELoCRaw & \SSEThreeMELoCParsedRaw \\
+%%%
+  GEOS-Chem &
+\SSgeosChemFiles & \SSgeosChemFilesParsed & \SSgeosChemLoC &
+\SSgeosChemLoCParsed & \SSgeosChemLoCRaw & \SSgeosChemLoCParsedRaw \\
+%%%
+  Hybrid4 & \SShybridFiles & \SShybridFilesParsed & \SShybridLoC &
+\SShybridLoCParsed & \SShybridLoCRaw & \SShybridLoCParsedRaw \\
+%%%
+  Navier & \SSnavierFiles & \SSnavierFilesParsed & \SSnavierLoC &
+\SSnavierLoCParsed & \SSnavierLoCRaw & \SSnavierLoCParsedRaw \\
+%%
+  Specfem3D & \SSspecfemFiles & \SSspecfemFilesParsed & \SSspecfemLoC &
+\SSspecfemLoCParsed & \SSspecfemLoCRaw & \SSspecfemLoCParsedRaw \\
+%%
+  UM &
+\SSumFiles & \SSumFilesParsed & \SSumLoC & \SSumLoCParsed &
+\SSumLoCRaw & \SSumLoCParsedRaw \\ \hline
+%%% TOTAL
+  \textbf{Total} & \SSFiles &
+\SSFilesParsed & \SSLoC & \SSLoCParsed & \SSLoCRaw & \SSLoCParsedRaw
+\end{tabular}
+\caption{Summary of software packages used for
+  evaluation\label{tab:corpus}}
+\vspace{-1em}
+\end{table}
 
@@
 \end{enumerate}
+
+\section{Proofs and extended definitions for the model}
+\label{sec:proofs}
+
+\input{proofs}

### 2017-04-13 oopsla17/oopsla17.tex
including more of the model

@@
 
+\errorcontextlines 10000
 \usepackage{stencilmacros}
@@
 \input{resultsCmds}
+\input{studyResults}
 % renews the commands with actual data
@@
   extended integer sets, \ie{}, a member of $\mathcal{P}(\bz)^{N}$.
-  An alternate (equivalent) view is of a $N$-times finite Cartesian
+  An equivalent view of these vectors is as an $N$-times finite Cartesian
   product on subsets of $\bz$.  We use $S, T, U$ to denote index
@@
 \end{defn}
-
-%\begin{theorem}
-%$N$-dimensional index schemes form a lattice where join and m
-%\end{theorem}
 
@@
   \begin{equation*}
-    S \; \cup \; T
+    S \cup  T
     =
@@
   \end{equation*}
+%  \begin{equation*}
+%    S \cup T
+%    =
+%    (\prod_{i=1}^{k-1} \pi_i(S)) \times (\pi_k(S) \cup \pi_k(T)) \times \prod_{k+1}^{N} \pi_i(S)
+%  \end{equation*}
 %
@@
 
-\begin{defn}
+\begin{defn}[Intervals with an optional hole]
   We define an extended notion of closed interval on $\bz{}$ which may contain
@@
 
-\begin{defn}{(\emph{Interval subspace})}
-  An interval subspace is a finite Cartesian product of intervals on
-  \bz{} or \bz itself, 
+We define two subsets of index scheme: \emph{subscript} schemes and
+\emph{interval} schemes:
+
+\begin{defn}[Interval scheme]
+  An interval scheme is a finite Cartesian product of intervals on
+  \bz{} or \bz{} itself, 
   denoted by the set $\textit{Interval}^N$ for a product of $N$ intervals.
 \end{defn}
+
+\begin{defn}[Subscript scheme]
+  A $N$-dimensional \emph{subscript scheme} is an $N$-dimensional
+  interval scheme where:
+  %
+  \begin{equation*}
+    \forall i.\ 1 \leq i \leq N \implies
+      \pi_i (S) = [p, p] 
+      \; \vee \;
+      \pi_i (S) = \bz{}
+    \end{equation*}
+    That is, the $i^{th}$ component of the set is either a singleton
+  in $\mathbb{Z}$ or the infinite interval.
+\end{defn}
+%
+\dnote{May want a predicate as well}
+
+\begin{defn}{(\emph{Region})}
+  A region \region{N} is an $N$-dimensional index scheme
+  (\ie{}, $\region{N} \subseteq \mathcal{P}(\bz)^N)$ defined as the smallest set satisfying the following:
+%
+  \begin{enumerate}
+    \item If $R$ is in $\textit{Interval}^N$, then $R$ is in \region{N}.
+    \item If $R$ and $S$ are in \region{N}, then so are $R \cap S$ and
+      $R \cup S$.
+  \end{enumerate}
+\end{defn}
+
+\begin{prop}{}
+  $(\region{N},\subseteq)$ is a distributive lattice.
+\end{prop}
+%
+\begin{proof}(sketch)
+%  Any sublattice of a distributive lattice is a distributive lattice. Power sets
+%  partially ordered with subset relation are distributive lattices. \region{N}
+%  is a subset of $\mathcal{P}{(\bz{})^N}$ since all interval schemes are
+%  subsets of this power set and; union and intersection cannot add an element
+%  that is not already in the powerset. The inductive definition ensures that
+%  \region{N} is closed under these operations. Therefore, \region{N} is a
+%  distributive sublattice of $(\mathcal{P}{(\bz{})^N}, \subseteq)$.
+\end{proof}
 

### 2017-04-13 Analyse.hs
functionality for combining intermediate files from the shell

@@ -82,7 +82,8 @@ import Control.DeepSeq
 import Results
 import Indices
 
-data Mode = SingleFile | ViewMode | NormalMode
+data Mode = SingleFile | ViewMode | NormalMode | CombineMode
+  deriving (Eq)
 
 main :: IO ()
 main = do
@@ -103,8 +104,13 @@ main = do
                (x:y:args') | x == "RESTART" -> (Just y, args', NormalMode)
                            | x == "SINGLE"  -> (Just y, args', SingleFile)
                            | x == "VIEW"    -> (Just y, [],    ViewMode)
+                           | x == "COMBINE"  -> (Nothing, (y:args'), CombineMode)
                            | otherwise      -> (Nothing, args, NormalMode)
-       case args' of
+               _ -> (Nothing, args, NormalMode)
+       if mode == CombineMode then
+         combineMode args'
+       else
+         case args' of
           [] -> applyAnalysisToDir restart mode "" False False []
           [dir]
              -> applyAnalysisToDir restart mode dir False False []
@@ -124,6 +130,18 @@ main = do
                        \ to be excluded."
    where
       filterFlags = filter (\arg -> arg /= "-d" && arg /= "-b")
+
+combineMode :: [String] -> IO ()
+combineMode restarts = do
+  files <- mapM readFile restarts
+  let (results :: [Result]) = map read files
+  let result = foldr mappend mempty results
+  writeFile "combined.restart" (show result)
+  putStrLn $ prettyResults result True
+  sloccount <- readProcess "sloccount" (map quotesWhenNeeded $ dirs result) ""
+  putStrLn sloccount
+  putStrLn $ "Files/dirs counted:" ++ (show . length . dirs $ result)
+  putStrLn $ "Raw count (parsed):" ++ (show . numLines $ result)
 
 applyAnalysisToDir :: Maybe String -> Mode -> String -> Bool -> Bool -> [String] -> IO ()
 applyAnalysisToDir restart mode dir debug bins excludes = do
@@ -152,8 +170,10 @@ applyAnalysisToDir restart mode dir debug bins excludes = do
         ViewMode -> do
                putStrLn $ prettyResults result True
                -- Use sloccount to give a total of the physical lines
-               sloccount <- readProcess "sloccount" (dirs result) ""
+               sloccount <- readProcess "sloccount" (map quotesWhenNeeded $ dirs result) ""
                putStrLn sloccount
[…]

### 2017-04-13 oopsla17/oopsla17.tex
more work on the model and more results

@@
 
-\begin{prop}{}
-  $(\region{N},\subseteq)$ is a distributive lattice.
-\end{prop}
+\begin{restatable}{prop}{regionLattice}
+  \label{prop:regionLattice}
+  $(\region{N},\cup,\cap,\subseteq)$ is a bounded distributive lattice with top
+  $\top = \bz{}^N$ and bottom $\bot = \emptyset$.
+\end{restatable}
 %
-\begin{proof}(sketch)
-%  Any sublattice of a distributive lattice is a distributive lattice. Power sets
-%  partially ordered with subset relation are distributive lattices. \region{N}
-%  is a subset of $\mathcal{P}{(\bz{})^N}$ since all interval schemes are
-%  subsets of this power set and; union and intersection cannot add an element
-%  that is not already in the powerset. The inductive definition ensures that
-%  \region{N} is closed under these operations. Therefore, \region{N} is a
-%  distributive sublattice of $(\mathcal{P}{(\bz{})^N}, \subseteq)$.
+\begin{proof}
+  Straightforward, since $\cup$ and $\cap$ are on sets (and therefore commutative,
+  associative, and absorptive with each
+  other). Furthermore, $\mathbb{Z}^N \cap R = R$ and $\empty \cup R =
+  R$ giving the bounds. Furthermore, the lattice is distributive since
+ union distributes over intersection and vice versa
 \end{proof}
+
+\begin{defn}{}
+  $\mathsf{Mult}$ and $\mathsf{Approx}$ are parametric labelled variant types
+  with injections given by their definition:
+%
+  \begin{align*}
+    \mathsf{Mult} \;\; a \;\; &
+      \triangleq \mathsf{mult} \; a \;\mid\; \mathsf{only} \; a \\
+    \mathsf{Approx} \;\; a \;\; &
+      \triangleq \mathsf{exact} \; a \;\mid\; \mathsf{lower} \; a \;\mid\;
+        \mathsf{upper} \; a
+  \end{align*}
+%
+  \eg{}, $\mathsf{lower}$ is an injection $\mathsf{lower} : a \to \mathsf{Approx}
+  \; a$ etc.
+  These will be used in the following subsection to give meaning to the
+  specification modifiers for approximation and multiplicity.
+\end{defn}
 
@@
 \label{sec:semantics}
+
+\noindent
+An interpretation function $\interp{-}_N$ provides a semantics to
+specifications as a member of $\textsf{Mult} (\textsf{Approx} (\mathcal{P}{\bz}^N)))$
+(where $n$ is the maximum dimensionality of the specification being
+modelled) mapping closed\footnote{That is, we assume there are no
+  occurrences of \textit{rvar} in a specification being modelled.  Any
+  \emph{open} specification containing region variables can be made
+  closed by straightforward syntactic substitution with a (closed)
+  \textit{region}.} specifications to sets of
+$n$-dimensional index schemes. The interpretation is overloaded on
+\emph{regions} in \Cref{fig:region-model} and on the top-level
+of a specification \textit{spec} in \Cref{fig:spatial-model}.
+Various intermediate notions are used.
 

### 2017-04-13 oopsla17/lattice-model.tex
semantic model for specs

@@
 
-Third, we need to mark presence of modifiers such as \texttt{readOnce} and
-\texttt{atMost} as introduced in \Cref{}. In the context of
-multiplicity modifiers we have the following semantics:
-%
-\begin{align*}
-  \interp{\texttt{readOnce}} & = \mathsf{mult} &
-  \llbracket\epsilon\rrbracket & = \mathsf{once}
-\end{align*}
-
-In the context of approximation modifiers, we have
-%
-\begin{align*}
-  \interp{\texttt{atLeast}} & = \mathsf{lower} &
-  \interp{\texttt{atMost}} & = \mathsf{upper} &
-  \llbracket \epsilon \rrbracket = \mathsf{exact}
-\end{align*}
+We mark the presence of modifiers such as \texttt{readOnce} and
+\texttt{atMost} as introduced in \Cref{}. Multiplicity
+modifiers are interpreted as injections in the $\mathsf{Mult}$
+variant by $\interpMult{}$ in Fig~\ref{fig:semantics}.
+In a similar way, approximation modifiers are interpreted
+as injections into the $\mathsf{Approx}$ variant by $\interpApprox{}$.
 
@@
 \begin{equation*}
-  \interp{\texttt{stencil \textcap{Mult}, \textcap{Appr}, \textcap{Region}}}_N =
-    \interp{\textcap{Mult}}
-           {(\interp{\textcap{Appr}}
-                    {(\interp{\textcap{Region}}_N)})}
+  \interp{\texttt{stencil \textcap{mult}, \textcap{approx}, \textcap{region}}}_N =
+    \interpMult{\textcap{mult}}
+           {(\interpMax{\textcap{approx}}
+                    {(\interp{\textcap{region}}_N)})}
 \end{equation*}

### 2017-04-13 oopsla17/oopsla17.tex
semantic model for specs

@@
 \noindent
-An interpretation function $\interp{-}_N$ provides a semantics to
-specifications as a member of $\textsf{Mult} (\textsf{Approx} (\mathcal{P}{\bz}^N)))$
-(where $n$ is the maximum dimensionality of the specification being
-modelled) mapping closed\footnote{That is, we assume there are no
-  occurrences of \textit{rvar} in a specification being modelled.  Any
-  \emph{open} specification containing region variables can be made
-  closed by straightforward syntactic substitution with a (closed)
-  \textit{region}.} specifications to sets of
-$n$-dimensional index schemes. The interpretation is overloaded on
-\emph{regions} in \Cref{fig:region-model} and on the top-level
-of a specification \textit{spec} in \Cref{fig:spatial-model}.
-Various intermediate notions are used.
+An interpretation function $\interp{-}_N$ maps closed\footnote{That
+  is, we assume there are no occurrences of \textit{rvar} in a
+  specification being modelled.  Any \emph{open} specification
+  containing region variables can be made closed by straightforward
+  syntactic substitution with a (closed) \textit{region}.}
+specifications to sets of $N$-dimensional index schemes with modifier
+information, \ie{} specifications are mapped to
+$\textsf{Mult} (\textsf{Approx} (\mathcal{P}({\bz})^N))$ where $N$ is
+the maximum dimensionality of the specification being modelled.
+
+The interpretation is overloaded on \emph{regions} in
+\Cref{subfig:region-model}. Various intermediate notions are used.
+
+\begin{defn}
+Let $\textit{promote}_N : \mathbb{N}^+ \times \textit{Interval} \to
+\textit{Interval}^N$ be a function generating an $N$-dimensional 
+interval such that $\pi_i(\vecgen{N}{i}{\interv{a}{b}{c}}) = 
+\interv{a}{b}{c}$ and $\pi_j(\vecgen{N}{i}{\interv{a}{b}{c}}) = 
+\bz{}$ in all other dimensions $j$. 
+\end{defn}
+
+%
+\begin{figure}[!t]
+\begin{subfigure}[t]{0.5\textwidth}
+\begin{align*}
+\interp{-}_N & : \textit{region} \rightarrow \mathcal{P}(\bz)^N \\
+  \interp{\stencil{p}{$i$}{}{}}_N & =
+    \vecgen{N}{i}{\interv{0}{0}{\mathsf{true}}}\\
+%
+  \interp{\stencil{c}{$i$}{$k$}{\textcap{p}}}_N & =
+    \vecgen{N}{i}{\interv{-k}{k}{\interp{\textcap{p}}}} \\
+%
+  \interp{\stencil{f}{$i$}{$k$}{\textcap{p}}}_N & =
+    \vecgen{N}{i}{\interv{0}{k}{\interp{\textcap{p}}}} \\
+%
+  \interp{\stencil{b}{$i$}{$k$}{\textcap{p}}}_N & =
+  \vecgen{N}{i}{\interv{-k}{0}{\interp{\textcap{p}}}}
+\\
+  \interp{\texttt{\textcap{r} + \textcap{s}}}_N & =
+    \interp{\textcap{r}}_N \vee \interp{\textcap{s}}_N
+\\
+  \interp{\texttt{\textcap{r} * \textcap{s}}}_N & =
+    \interp{\textcap{r}}_N \wedge \interp{\textcap{s}}_N \\[-1em]
+\end{align*}
+\caption{Interpretation of regions}
+\label{subfig:region-model}
+\end{subfigure}
+\hspace{1em}
+\begin{subfigure}[t]{0.4\textwidth}
+\begin{align*}
+\interpApprox{-} & : \textit{approx} \rightarrow (A \rightarrow
+  \textsf{Approx} \, A) \\
+\interpApprox{\texttt{atLeast}} & = \mathsf{lower} \\
+  \interpApprox{\texttt{atMost}} & = \mathsf{upper} \\
+  \interpApprox{\epsilon} & = \mathsf{exact} \\ \\
+  \interpMult{-} & : \textit{mult} \rightarrow (A \rightarrow
+  \textsf{Mult} \, A) \\
+  \interpMult{\texttt{readOnce}} & = \mathsf{once} \\
+  \interpMult{\epsilon} & = \mathsf{mult}
+\end{align*}
+\caption{Interpretation of modifiers}
+\label{subfig:modifier-model}
+\end{subfigure}
+\label{fig:semantics}
+\caption{Semantic model of specifications}
+\end{figure}
+
+The first four equations of \Cref{subfig:region-model} model region 
+contants. The final two equations model the \term{+} and \term{*}
+operators in terms of the join (union) and meet (intersection)
+of interval schemes. Thus regions are modelled as members
+of $\mathcal{P}(\bz)^N$.
+
+We mark in our model the presence of modifiers such as
+\texttt{readOnce} and \texttt{atMost} as introduced in \Cref{}.
+Approximation modifiers are interpreted as injections into the
+$\mathsf{Approx}$ variant by $\interpApprox{}$ in
+\Cref{subfig:modifier-model}.  The $\textsf{Approx}$ type corresponds
+to the presence or absence of the spatial approximation modifier, with
+\textsf{exact} when there is no such modifier and \textsf{lower} and
+\textsf{upper} for \term{atLeast} and \term{atMost}. In a similar way,
+multiplicity modifiers are interpreted as injections in the
+$\mathsf{Mult}$ variant by $\interpMult{}$, corresponding to the to
+the presence or absence of the \term{readOnce} modifier as shown in
+\Cref{subfig:modifier-model}.
+
+\begin{defn}[Semantics of specifications]
+The intermediate interpretations of \Cref{fig:semantics}
+are composed to give a model for the top-level specification
+syntax as:
+%
+\begin{equation*}
+  \interp{\texttt{stencil \textcap{mult}, \textcap{approx}, \textcap{region}}}_N =
+    \interpMult{\textcap{mult}} \;
+           {(\interpApprox{\textcap{approx}} \;
+                    {\interp{\textcap{region}}_N)}}
+\end{equation*}
+\end{defn}
+%
+\begin{thm}
+The semantic model $\interp{-}_N$ of $N$-dimensional specifications
+is sound with respects to the equational theory of the language,
+that is:
+%
+\begin{equation*}
+\forall S, T, N . \quad
+S \equiv T \; \Rightarrow \;
+\interp{S}_N = \interp{T}_N
+\end{equation*}
+%
+We define the equational theory in \Cref{sec:equational-theory}.
+\end{thm}
 

### 2017-04-13 oopsla17/oopsla17.tex
expanding hypotheses

@@
 
+Hypotheses:
 \begin{enumerate}
-\item Stencils are ubiquitious;
-\item Most stencils read their data from source arrays with a static pattern
-base on constant offsets from the mid-point;
-\item Most stencils read their data from contiguous \emph{regions} in an array;
-\item Most stencils read each index at most once.
+\item Loops over arrays mainly read from arrays with a static pattern
+based on constant offsets from (base or dervied) induction variables;
+
+\item Most loop-array computations of the previous form read
+from a arrays with a contiguous pattern;
+
+\item Many array computations are \emph{stencil computations},
+with a static access pattern, writing
+to an array at an index based on a constant offset from induction
+variables.
+
+\item Many array computations read from a particular index just once
 \end{enumerate}

### 2017-04-14 Study.hs
separate executable for running analyses over the raw data

@@ -1,0 +1,181 @@
+{-# LANGUAGE GADTs #-}
+{-# LANGUAGE DataKinds #-}
+
+import qualified Data.Map as M
+import Results
+import Data.List
+import System.Environment
+import Text.Printf
+
+main = do
+  args <- getArgs
+  case args of
+    [] -> putStrLn "Specify a .restart file to analyse"
+    [x] -> putStrLn "Specify an analysis you want to run, e.g., hyps"
+    (file:study:_) -> do
+      resultsString <- readFile file
+      let result = ((read resultsString) :: Result)
+      case study of
+        "hyps" -> putStrLn $ hyps result
+
+regroup :: (Ord c, HistogramShow t) => (Cat -> c) -> M.Map Cat t -> M.Map c t
+regroup classifier =
+  M.fromListWith histZip . map (\(k, v) -> (classifier k, v)) . M.assocs
+
+mapViewT msg map =
+       "   " ++ msg ++ ":\n"
+    ++ rline' ((replicate 5 ' ') ++ "Total") (show' total)
+    ++ concatMap (\(cat, dat) -> hline' cat (showPad dat ++ "("
+                                 ++ (twoDP $ (cast(dat)/cast(total))*100) ++ "%)")) (M.assocs map)
+  where
+    showPad dat = show dat ++ replicate (10 - (length (show dat))) ' '
+    twoDP = printf "%.2f"
+    cast :: Int -> Float
+    cast = fromInteger . toInteger
+    total = histTotal $ M.elems map
+
+hyps r =
+     mapViewT "Hypothesis 1" (countWrapper hypothesis1 r)
+  ++ mapViewT "Hypothesis 1 finer" (countWrapper hypothesis1finer r)
+  ++ mapViewT "Hypothesis 2" (countWrapper hypothesis2 r)
+  ++ mapViewT "Hypothesis 3" (countWrapper hypothesis3 r)
+  ++ mapViewT "Hypothesis 4" (countWrapper (interpret4 . hypothesis4) r)
+  ++ mapViewT "Hypothesis 4 finer" (countWrapper hypothesis4finer r)
+  ++ mapViewT "Hypothesis 5" (countWrapper hypothesis5 r)
+
+countWrapper :: Ord c => (Cat -> c) -> Result -> M.Map c Int
+countWrapper classifier = regroup classifier . counts
+
+-- Hypothesis 1 : Loops over arrays mainly read from arrays with a
+-- static pattern based on constant offsets from (base or dervied) induction variables;
+
[…]

### 2017-04-14 Indices.hs
a couple of bug fixes for singletons and adding 'all constant' categorisation

@@ -87,11 +87,17 @@ classifyArrayCode ivs lhs rhses =
     lhsForm = case lhsRep of
                 Nothing -> Vars
                 Just Subscript  -> Subscripts
-                Just (Affine as) -> if allIVs as then IVs else Affines (hasConstants as) L
-                Just (Neigh  ns) -> if allIVs ns then IVs else Neighbours (hasConstants ns) L
+                Just (Affine as) | allIVs as -> IVs
+                Just (Affine as) | allConstants as -> AllConsts
+                Just (Affine as) -> Affines (hasConstants as) L
+                Just (Neigh  ns) | allIVs ns -> IVs
+                Just (Neigh  ns) | allConstants ns -> AllConsts
+                Just (Neigh ns) -> Neighbours (hasConstants ns) L
     rhsForm = case rhsRep of
                 Subscript -> Subscripts
+                Affine as | allConstants as -> AllConsts
                 Affine as  -> Affines (hasConstants as) (R rhsPhysical)
+                Neigh ns  | allConstants ns -> AllConsts
                 Neigh  ns  -> Neighbours (hasConstants ns) (R rhsPhysical)
     rhsPhysical = (shape, position, boolToContig contiguity, boolToReuse (not nonLinear))
     ------------
@@ -288,10 +294,12 @@ hasConstants = boolToHasConstants . hasConstantsFlag
 class Subscripts t where
   allIVs :: [t] -> Bool
   hasConstantsFlag :: [t] -> Bool
+  allConstants :: [t] -> Bool
 
 instance Subscripts a => Subscripts [a] where
   allIVs       = and . map allIVs
   hasConstantsFlag = or  . map hasConstantsFlag
+  allConstants = and . map allConstants
 
 instance Subscripts (Int, String, Int) where
   allIVs [] = True
@@ -302,6 +310,10 @@ instance Subscripts (Int, String, Int) where
   hasConstantsFlag ((0, i, 0):xs) = True
   hasConstantsFlag (_:xs) = hasConstantsFlag xs
 
+  allConstants [] = True
+  allConstants ((0, i, 0):xs) = allConstants xs
+  allConstants (_:xs) = False
+
 instance Subscripts Neighbour where
   allIVs [] = True
   allIVs ((Neighbour i 0):xs) = allIVs xs
@@ -310,3 +322,7 @@ instance Subscripts Neighbour where
   hasConstantsFlag [] = False
[…]

### 2017-04-14 Neighbour.hs

@@ -48,7 +48,7 @@ transposeVecInterval (Cons l ls, Cons u us) = Cons (l, u) intervalVec
 
 positionInterval :: (Int, Int) -> Position
 positionInterval (l, u)
-  | l == -absoluteRep || u == absoluteRep = Elsewhere
+  | l == -absoluteRep || u == absoluteRep = OverOrigin
   | l <= 0 && u >= 0                      = OverOrigin
   | l < 0  && u == (-1)                   = StraddleOrigin
   | l == 1 && u > 0                       = StraddleOrigin
@@ -59,6 +59,7 @@ contiguous :: [[Neighbour]] -> Bool
 contiguous xs = contiguity' xs
   where
     contiguity' [] = True
+    contiguity' [x] = True
     contiguity' (y : ys) | isOrigin y = contiguity' ys
     contiguity' (y : ys) | nextToOrigin y = contiguity' ys
     contiguity' (y : ys) | hasNeighbouringIx y (xs \\ [y]) = contiguity' ys

### 2017-04-14 Results.hs

@@ -78,9 +78,14 @@ boolToHasConstants :: Bool -> HasConstants
 boolToHasConstants True = WithConsts
 boolToHasConstants False = Normal
 
+hasConstantsToBool :: HasConstants -> Bool
+hasConstantsToBool WithConsts = True
+hasConstantsToBool Normal = False
+
 data Form (s :: Side) where
     Vars        :: Form LHS
     Subscripts  :: Form s
+    AllConsts   :: Form s
     Affines     :: HasConstants -> Physicality s -> Form s
     Neighbours  :: HasConstants -> Physicality s -> Form s
     IVs         :: Form LHS
@@ -344,8 +349,10 @@ binView msg bin =
 
 mapView msg map =
        "   " ++ msg ++ ":\n"
-    ++ rline' ((replicate 5 ' ') ++ "Total") (show' . histTotal $ M.elems map)
+    ++ rline' ((replicate 5 ' ') ++ "Total") (show' total)
     ++ concatMap (\(cat, dat) -> hline' cat (show' dat)) (M.assocs map)
+  where
+    total = histTotal $ M.elems map
 
 mapView' msg map =
        "   " ++ msg ++ ":\n"

### 2017-04-14 oopsla17/oopsla17.tex
some editing in results

@@
 
+\item Most loop-array computations of the previous form read
+from arrays with a pattern that includes the immediate
+neighbours (i.e., offset 1 from the induction variable);
+
 \item Many array computations are \emph{stencil computations},
@@
 
-\subsection{Study}
+\subsection{Methodology}
 %
 \noindent
-We put together a corpus of around 1 million lines of Fortran code from a
-\numPackages{} scientific computing packages ranging in size and scope: The Unified Model (UM)~\cite{um},
-E3MG~\cite{RePEc:aen:journl:2006se-a12}, BLAS~\cite{blas},
-Hybrid4~\cite{GBC:GBC635}, GEOS-Chem~\cite{geos-chem}, Navier (based
-on \cite{griebel1997numerical}), Computational Physics
-2~\cite{nicholas2006computational},
-ARPACK-NG~\cite{arpackng}, and
-SPECFEM3D~\cite{specfem3d}. Appendix~\ref{app:corpus} provides
-further detail on these packages.
+We constructed a corpus of around 1.4 million lines of Fortran code
+(2.4 million including comments and white space) from a \numPackages{}
+scientific computing packages ranging in size and scope: The Unified
+Model (UM)~\cite{um}, E3MG~\cite{RePEc:aen:journl:2006se-a12},
+BLAS~\cite{blas}, Hybrid4~\cite{GBC:GBC635},
+GEOS-Chem~\cite{geos-chem}, Navier (based on
+\cite{griebel1997numerical}), Computational Physics
+2 (CP2) ~\cite{nicholas2006computational}, ARPACK-NG~\cite{arpackng}, and
+SPECFEM3D~\cite{specfem3d}. \dnote{Include the rest}
+Appendix~\ref{app:corpus} provides further detail on these packages.
+Half of the packages come from research groups who 
+are our project partners (\ie{}, they were
+not selected carefully to skew results). BLAS was selected as it is
+a common library in numerical code; similarly CP2 was selected as 
+it presents a set of standard numerical analysis algorithms from a
+popular computational physics textbook.
 
@@
 the max value in an array).
+\end{itemize}
 
-\end{itemize}
+\subsection{Results}
+
+\paragraph{Addressing the hypotheses}
+
+\begin{enumerate}
+\item Loops over arrays mainly read from arrays with a static pattern
+based on constant offsets from (base or dervied) induction variables;
+\end{enumerate}
+
+%\begin{tabular}
+%Neighbourhood or Constant Affine RHS &  Total 
+%\end{tabular}
 

### 2017-04-14 Results.hs
bug fix in analysis file deserialiser

@@ -97,6 +97,10 @@ deriving instance Show (Form s)
 instance Read (Form LHS) where
   readsPrec n xs | "Vars" `isPrefixOf` xs
     = [(Vars, drop (length "Vars") xs)]
+
+  readsPrec n xs | "AllConsts" `isPrefixOf` xs
+    = [(Vars, drop (length "AllConsts") xs)]
+
   readsPrec n xs | "Subscripts" `isPrefixOf` xs
     = [(Subscripts, drop (length "Subscripts") xs)]
 

### 2017-04-14 Analyse.hs
tweak to debug mode

@@ -164,7 +164,7 @@ applyAnalysisToDir restart mode dir debug bins excludes = do
                   let result1 = result0 `mappend` mempty { dirs = [dir] }
                   foldrM (applyAnalysisToFile (mode, restart) dir) ("", result1) files
 
-    if debug then putStrLn $ dbg else return ()
+    if (debug || mode == SingleFile) then putStrLn $ dbg else return ()
 
     case mode of
         ViewMode -> do

### 2017-04-14 oopsla17/analysis/Results.hs
another bug fix in deserialiser

@@ -99,7 +99,7 @@ instance Read (Form LHS) where
     = [(Vars, drop (length "Vars") xs)]
 
   readsPrec n xs | "AllConsts" `isPrefixOf` xs
-    = [(Vars, drop (length "AllConsts") xs)]
+    = [(AllConsts, drop (length "AllConsts") xs)]
 
   readsPrec n xs | "Subscripts" `isPrefixOf` xs
     = [(Subscripts, drop (length "Subscripts") xs)]

### 2017-04-14 oopsla17/analysis/Results.hs
bug fix to deserialiser

@@ -123,6 +123,9 @@ instance Read (Form RHS) where
   readsPrec n xs | "Subscripts" `isPrefixOf` xs
     = [(Subscripts, drop (length "Subscripts") xs)]
 
+  readsPrec n xs | "AllConsts" `isPrefixOf` xs
+    = [(AllConsts, drop (length "AllConsts") xs)]
+
   readsPrec n xs | "Affines" `isPrefixOf` xs
     = do (consts, '(':rest) <- readsPrec n (drop (length "Affines ") xs)
          (pr, ')':rest)     <- readsPrec n rest

### 2017-04-14 oopsla17/oopsla17.tex
describing our methodology

--- 2017-04-13 oopsla17/oopsla17.tex
@@
 \usepackage{xcolor}
+\usepackage{multirow}
 \usepackage{siunitx} % For pretty-printing numeric values and SI units
@@
 \usepackage{etoolbox}
-
+\usepackage{enumitem}
 \usepackage{thmtools}
 \usepackage{thm-restate}
+
+\newcommand{\fortran}[1]{\mintinline{fortran}{#1}}
 
@@
 \section{An Empirical Study of Array Computations in Scientific Fortran code}
-
-\begin{enumerate}
-\item Stencils are ubiquitious;
-\item Most stencils read their data from source arrays with a static pattern
-base on constant offsets from the mid-point;
-\item Most stencils read their data from contiguous \emph{regions} in an array;
-\item Most stencils read each index at most once.
-\end{enumerate}
-
-\subsection{Study}
-%
-\noindent
-We put together a corpus of around 1 million lines of Fortran code from a
-\numPackages{} scientific computing packages ranging in size and scope: The Unified Model (UM)~\cite{um},
-E3MG~\cite{RePEc:aen:journl:2006se-a12}, BLAS~\cite{blas},
-Hybrid4~\cite{GBC:GBC635}, GEOS-Chem~\cite{geos-chem}, Navier (based
-on \cite{griebel1997numerical}), Computational Physics
-2~\cite{nicholas2006computational},
-ARPACK-NG~\cite{arpackng}, and
-SPECFEM3D~\cite{specfem3d}. Appendix~\ref{app:corpus} provides
-further detail on these packages.
-
-We used Wheeler's \emph{SLOCCount} to get a count of the physical
-source lines (excluding comments and blank
-lines)~\cite{wheeler2001sloccount}; including non-physical lines
- our corpus totals \dnote{insert}.
-
-\begin{itemize}
-% (Vars, ...)
-\item An assignment to a variable, with an RHS comprising some array
-  computation, tends to correspond to a reduction (e.g., calculating
-the max value in an array).
-
-\end{itemize}
+\label{sec:study}
+\input{study}
 

### 2017-04-14 oopsla17/study.tex
describing our methodology

At the beginning of this project, we had the following hypotheses:
\begin{enumerate}
\item Loops over arrays mainly read from arrays with a static pattern
based on constant offsets from (base or dervied) induction variables;

\item Most loop-array computations of the previous form read
from a arrays with a \emph{contiguous} pattern, \eg{}:
%
\begin{minted}{fortran}
do i = 1, n
  b(n) = (a(i) + a(i+1)) / 2.0
\end{minted}
%
\item Most loop-array computations of the previous form read
from arrays with a pattern that includes the immediate
neighbours (offsets of 1 from the induction variables);

\item Many array computations are \emph{stencil computations},
with a static access pattern as described in Hypothesis 2, writing
to an array at an index based on a (possibly constant offset from induction
variables, \eg{}, the classic five-point stencil;

\item Many array computations read from each particular index just once.
\end{enumerate}
%
From these hypotheses, we conjectured that the common programming
patterns captured above could be specified declaratively with a small
set of combinators capture the shape of data access patterns based on
overlapping (hyper)rectangles. We performed a large scale source-code
analysis to validate these hypotheses and guide the design of our
language. We believe these results are also of wider value.
\dnote{Explain why: what else do they tell us?}

\subsection{Methodology}
%
\noindent
We constructed a corpus of \numPackages{} scientific computing
packages written in Fortran ranging in size and scope: The Unified
Model (UM)~\cite{um}, E3MG~\cite{RePEc:aen:journl:2006se-a12},
BLAS~\cite{blas}, Hybrid4~\cite{GBC:GBC635},
GEOS-Chem~\cite{geos-chem}, Navier (based on
\cite{griebel1997numerical}), Computational Physics 2 (CP2)
~\cite{nicholas2006computational}, ARPACK-NG~\cite{arpackng}, and
SPECFEM3D~\cite{specfem3d}. \dnote{Include the rest} This covers
approximately 1.4 million lines of physical Fortran code (2.4 million
including comments and white space).  Appendix~\ref{app:corpus}
provides further detail on these packages and their sizes, with UM and
GEOS-Chem the largest at $\approx$630kloc and $\approx$450kloc respectively. We used
Wheeler's \emph{SLOCCount} to get a count of the physical source lines
(excluding comments and blank lines)~\cite{wheeler2001sloccount}.

Half of the packages come from active research teams who are our
project partners (\ie{}, they were not selected carefully to skew
results). BLAS was selected as it is a common numerical library;
similarly CP2 was selected as it gives standard numerical
analysis algorithms from a popular computational physics textbook.
The smallest project Navier ($\approx$500loc) is based on a standard textbook on
computational fluid dynamics.

We built a code analysis tool based on CamFort, an open-source analysis
tool for Fortran programs~\cite{camfort}. Fortran source files are
parsed to an AST and various standard control-flow and data-flow
analyses are computed, including of key importance for us: induction
variable identification, and reaching definitions. The resulting AST
is traversed top-down and assignment statements inside loops are
analysed and classified.

\paragraph{Classifications}

We focus on assignment statements where
the right-hand side reads from one or more arrays. Left-hand
sides and right-hand sides are classified in the following way:

\begin{tabular}{l|l|l}
  Shorthand & Classification (of subscripts) & Example \\ \hline
 (\textsf{vars}) & Variables (just left-hand side) & \fortran{x = ...} \\
 (\textsf{const}) & Constants & \fortran{a(0, 1)} \\
 (\textsf{IVs}) & Induction variables & \fortran{a(i, j)} \\
 (\textsf{neigh}) & Neighbourhood offsets (of the form
                                                  $i \pm c$) & \fortran{a(i, j-1)} \\
 (\textsf{neigh+c}) & Neighbourhood offsets and constants &
                                                            \fortran{b(i, 0, j+1)} \\
 (\textsf{aff}) & Affine offsets (of the form $a * i \pm b$) & 
                                     \fortran{x(2*i+1,j)} \\
 (\textsf{aff+c}) & Affine offsets and constants & \fortran{x(i+1, 0, 3*j+2)}
\end{tabular} \\[1em]

\noindent
The classification is computed from all information that flows into an
assignment, for example, the following decomposes a one-dimensional
three-point pattern across multiple intermediate assignments:
%%
\begin{minted}{fortran}
do i = 1, n
  x = a(i)
  y = a(i+1)
  b(i) = (a(i-1) + x + y)/3.0
end do
\end{minted}
%%x
Our analysis recognises this as one array computation (rather than three), starting at
line 4, and reading from subscripts \fortran{a(i)}, \fortran{a(i+1)},
\fortran{a(i-1)}. We thus traverse the body of a loop in reverse,
using reaching-definitions to calculate of the array reads that flow
to a particular assignment. Since \fortran{x} and \fortran{y} flow to
the line 4, line 2 and line 3 are not classified, though multiple
assignments can flow to multiple classified statements.

We further sub-classify sets of array subscripts
on the right-hand side based on their spatial properties: \\

\begin{tabular}{l||l|p{0.28\linewidth}|l}
  Property   & Shorthand & Classifications (of RHS pattern) & Example \\ \hline
%%%%%%
\multirow{2}{*}{Contiguity} & (\textsf{contig}) & Contiguous  &
                                                                \fortran{a(i) + a(i+1) + a(i+2)} \\
  & (\textsf{disjoint}) & Non-contiguous & \fortran{a(i) + a(i+2)} \\ \hline
%%%%%%
\multirow{2}{*}{Shape} & (\textsf{rect}) & (Hyper)rectangle &
 \fortran{a(i,j) + a(i+1,j) + a(i,j+1) + a(i+1,j+1)} \\
           & (\textsf{sumRect}) & Composed (hyper)rectangles &
  \fortran{a(i,j) + a(i,j-1) + a(i-1,j) + a(i+1,j) + a(i,j+1)} \\ \hline
%%%%%%
  \multirow{2}{*}{Reuse} & (\textsf{readonce}) & Unique subscripts
                                         & \eg{} \fortran{b(i) = a(i) + a(i+1)} \\
  & (\textsf{mult}) & Repeated subscripts &
\eg{}  \fortran{b(i) = a(i)  + a(i)}
\end{tabular} \\[1em]

\noindent
To assess the hypothesis that 
arrays are mostly read in a pattern that includes the immediate neighbours
to the ``origin'', we categorised the position of subscript patterns for each
right-hand side, in each dimension: \\

\begin{tabular}{l||l|p{0.33\linewidth}|l}
  Property   & Shorthand & Classification (indices per
                           dimension) & Example \\ \hline
%%%%%%
  \multirow{2}{*}{Positioning} & (\textsf{origin}) & Includes origin
& \fortran{a(i)} or \fortran{a(i+1, j)} (in $2^{\textit{nd}}$
  dimension) \\
             & (\textsf{straddle}) & Within distance 1 of origin &
\fortran{a(i+1), a(i-1)} \\
             & (\textsf{away}) & Away from the origin
                                          & \fortran{a(i+2), a(i+3)}
\end{tabular} \\[1em]

\noindent
Finally, we categorised the relationship between the left-hand side
and right-hand side, in terms of the use of induction variables.
The two sides are \emph{consistent} if the same induction variables
appear in each side, used for the same dimension. This is
weakened to a \emph{permutation} if the roles of the induction
variables changes. This is weakened further if the left-hand side
induction variables are either a 
a subset or superset of the induction variables on the right-hand
side. Otherwise, the two sides are seen as inconsistent:

\begin{tabular}{l|l|l}
  (\textsf{const}) & Consistent & \fortran{a(i, j) = b(i, j) +
                                  b(i+1,j+1)} \\
  (\textsf{perm} & Permutation & \fortran{a(i, j) = c(j, i)} or \fortran{a(i,
                                                               0) =
                                 b(0, i)} \\
  (\textsf{LHSsub} & LHS subset & \fortran{a(i) = b(i, j) + b(i, j-1)}
  \\
  (\textsf{LHSsup} & LHS superset & \fortran{a(i, j) = b(i)} \\
  (\textsf{inconst} & Inconsistent & \fortran{a(i) = b(j)}
\end{tabular} \\[1em]
%

%\begin{itemize}
% (Vars, ...)
%\item An assignment to a variable, with an RHS comprising some array
%  computation, tends to correspond to a reduction (e.g., calculating
%the max value in an array).
%\end{itemize}

\subsection{Results}

\paragraph{Addressing the hypotheses}

\begin{enumerate}
\item Loops over arrays mainly read from arrays with a static pattern
based on constant offsets from (base or dervied) induction variables;
\end{enumerate}

%\begin{tabular}
%Neighbourhood or Constant Affine RHS &  Total 
%\end{tabular}

### 2017-04-14 oopsla17/analysis/Analyse.hs
add a slightly different view mode, and updated data

@@ -82,7 +82,7 @@ import Control.DeepSeq
 import Results
 import Indices
 
-data Mode = SingleFile | ViewMode | NormalMode | CombineMode
+data Mode = SingleFile | ViewMode | ViewMode2 | NormalMode | CombineMode
   deriving (Eq)
 
 main :: IO ()
@@ -104,6 +104,7 @@ main = do
                (x:y:args') | x == "RESTART" -> (Just y, args', NormalMode)
                            | x == "SINGLE"  -> (Just y, args', SingleFile)
                            | x == "VIEW"    -> (Just y, [],    ViewMode)
+                           | x == "VIEW2"   -> (Just y, [],    ViewMode2)
                            | x == "COMBINE"  -> (Nothing, (y:args'), CombineMode)
                            | otherwise      -> (Nothing, args, NormalMode)
                _ -> (Nothing, args, NormalMode)
@@ -160,6 +161,8 @@ applyAnalysisToDir restart mode dir debug bins excludes = do
       case mode of
          ViewMode
             -> return ("", result0)
+         ViewMode2
+            -> return ("", result0)
          _  -> do files <- readParseSrcDir dir excludes
                   let result1 = result0 `mappend` mempty { dirs = [dir] }
                   foldrM (applyAnalysisToFile (mode, restart) dir) ("", result1) files
@@ -172,6 +175,12 @@ applyAnalysisToDir restart mode dir debug bins excludes = do
                -- Use sloccount to give a total of the physical lines
                sloccount <- readProcess "sloccount" (map quotesWhenNeeded $ dirs result) ""
                putStrLn sloccount
+               putStrLn $ "Files/dirs counted:" ++ (show . length . dirs $ result)
+               putStrLn $ "Raw count (parsed):" ++ (show . numLines $ result)
+
+        ViewMode2 -> do
+               putStrLn $ prettyResults result True
+               -- Use sloccount to give a total of the physical lines
                putStrLn $ "Files/dirs counted:" ++ (show . length . dirs $ result)
                putStrLn $ "Raw count (parsed):" ++ (show . numLines $ result)
 

### 2017-04-14 oopsla17/oopsla17.tex
Switch to better def style

@@
 \usepackage{amssymb}
+\usepackage{amsthm}
 \usepackage{syntax}
@@
 
-% Essential declarations
-\theoremstyle{definition}
+\definecolor{darkgreen}{rgb}{0.0,0.5,0.0}
+\definecolor{darkpurple}{rgb}{0.6,0.0,0.6}
+\definecolor{orange}{rgb}{0.8,0.4,0.0}
+\newcommand{\dnote}[1]{\textcolor{darkpurple}{Dom: #1}}
+\newcommand{\mnote}[1]{\textcolor{darkgreen}{Mistral: #1}}
+\newcommand{\anote}[1]{\textcolor{red}{Andy: #1}}
+
+\theoremstyle{defn}
 \newtheorem{defn}{Definition}
@@
 
-\definecolor{darkgreen}{rgb}{0.0,0.5,0.0}
-\definecolor{darkpurple}{rgb}{0.6,0.0,0.6}
-\definecolor{orange}{rgb}{0.8,0.4,0.0}
-\newcommand{\dnote}[1]{\textcolor{darkpurple}{Dom: #1}}
-\newcommand{\mnote}[1]{\textcolor{darkgreen}{Mistral: #1}}
-\newcommand{\anote}[1]{\textcolor{red}{Andy: #1}}
+\theoremstyle{remark}
+\newtheorem{remark}{Remark}
 
@@
 
-\begin{definition}[Induction variables]
+\begin{defn}[Induction variables]
   An integer variable is a \emph{base induction variable} if it is the control
@@
   i + b$, \ie{}, an affine expression on an induction variable $i$.
-\end{definition}
+\end{defn}
 
-\begin{definition}[Array subscripts and indices]
+\begin{defn}[Array subscripts and indices]
   An \emph{array subscript}, denoted $a(\bar{e})$, is an expression
@@
   relative to the enclosing loop.
-\end{definition}
+\end{defn}
 
-\begin{definition}[Neighbourhood index]
+\begin{defn}[Neighbourhood index]
 \label{def:neighbour}
@@
   relationship of $+$ and $-$ \eg{}, $(-b) + i \equiv i - b$).
-\end{definition}
+\end{defn}
 
@@
 
-\begin{defn}{(\emph{Region})}
+\begin{defn}[Region]
   A region \region{N} is an $N$-dimensional index scheme
@@
 
-\begin{defn}{}
+\begin{defn}
   $\mathsf{Mult}$ and $\mathsf{Approx}$ are parametric labelled variant types

### 2017-04-14 oopsla17/oopsla17.tex
Further explain derived induction vars

@@
 
-  A \emph{derived induction variable} is an expression of the form $a
-  i + b$, \ie{}, an affine expression on an induction variable $i$.
+  A \emph{derived induction variable} is an expression of the form $a \times i
+  + b$, where $a$ and $b$ are constant expressions, \ie{}, an affine expression
+  on an induction variable $i$.
 \end{defn}

### 2017-04-14 oopsla17/oopsla17.tex
Fix style of definitions
Specify dimensionality in subscript

@@
 
-\theoremstyle{defn}
+\theoremstyle{definition}
 \newtheorem{defn}{Definition}
@@
 \begin{defn}[Array subscripts and indices]
-  An \emph{array subscript}, denoted $a(\bar{e})$, is an expression
-  which reads from an array $a$ at an \emph{index} specified by a
-  comma-separated sequence of integer expressions denoted $\bar{e}$ or
-  in expanded form as $(e_1, \ldots, e_n)$. An index $e_i$ is called
-  \emph{relative} if the expression involves an induction variable.
-  An \emph{absolute index} is a integer expression which is constant
-  relative to the enclosing loop.
+  An \emph{array subscript}, denoted $a(\bar{e})$, is an expression which reads
+  from an $N$-dimensional array $a$ at an \emph{index} specified by a
+  comma-separated sequence of integer expressions denoted $\bar{e}$ or in
+  expanded form as $(e_1, \ldots, e_N)$. An index $e_i$ is called
+  \emph{relative} if the expression involves an induction variable. An
+  \emph{absolute index} is a integer expression which is constant relative to
+  the enclosing loop.
 \end{defn}

### 2017-04-14 oopsla17/study.tex
Typo
first results for hypothesis 1

@@
   Shorthand & Classification (of subscripts) & Example \\ \hline
- (\textsf{vars}) & Variables (just left-hand side) & \fortran{x = ...} \\
+ (\textsf{vars}) & Variables (just LHS) & \fortran{x = ...}
+  \\
+ (\textsf{IVs}) & Induction variables (just LHS) & \fortran{a(i, j)} \\
  (\textsf{const}) & Constants & \fortran{a(0, 1)} \\
- (\textsf{IVs}) & Induction variables & \fortran{a(i, j)} \\
  (\textsf{neigh}) & Neighbourhood offsets (of the form
@@
  (\textsf{aff}) & Affine offsets (of the form $a * i \pm b$) &
-                                     \fortran{x(2*i+1,j)} \\
- (\textsf{aff+c}) & Affine offsets and constants & \fortran{x(i+1, 0, 3*j+2)}
+                                     \fortran{a(2*i+1,j)} \\
+ (\textsf{aff+c}) & Affine offsets and constants & \fortran{a(i+1, 0,
+                                                   3*j+2)} \\
+ (\textsf{other}) & Subscript expression not included above &
+\fortran{x(f(i))}, \fortran{a(i `mod` 2)}
 \end{tabular} \\[1em]
@@
 \noindent
+Note that the \textsf{vars} and \textsf{IVs} are only give to
+left-hand sides.
 The classification is computed from all information that flows into an
@@
 variables changes. This is weakened further if the left-hand side
-induction variables are either a 
+induction variables are either
 a subset or superset of the induction variables on the right-hand
@@
 
-\paragraph{Addressing the hypotheses}
-
-\begin{enumerate}
-\item Loops over arrays mainly read from arrays with a static pattern
-based on constant offsets from (base or dervied) induction variables;
-\end{enumerate}
+%\begin{enumerate}
+%\item Loops over arrays mainly read from arrays with a static pattern
+%based on constant offsets from (base or dervied) induction variables;
+%\end{enumerate}
 
@@
 %\end{tabular}
+
+\noindent
+We identified $108,773$ instances of assignments within loops in
+which an array subscript flows to the right-hand side. We refer
+to each one of these as an \emph{array computation}.
+We performed the classification described in the methodology,
+and then grouped the data by based only on the right-hand side
+classification:
+%%
+\begin{center}
+\begin{tabular}{lrrcc}
+\textbf{RHS classification} & \textbf{Number} & \textbf{\%} &
+\textit{grouping 1} & \textit{grouping 2} \\ \hline
+Affine                          & 34        & 0.03\%  &
+                                                            \multirow{4}{*}{92.61\%}
+  \\ \cline{1-3}\cline{5-5}
+All constant                    & 12160     & 11.18\%  \\ \cline{1-3}\cline{5-5}
+Neighbourhood                   & 65334     & 60.06\% &
+                                                          & \multirow{2}{*}{81.4\%}
+                                                           \\ \cline{1-3}
+Neighbourhood + constants       & 23209     & 21.34\%  \\ \hline
+Other                           & 8036      & 7.39\%  \\ \hline \hline
+Total                           & 108773    &  \\
+\end{tabular}
+\end{center}
+
+\noindent
+Since, affine, consant, and neighbour are all static patterns, we
+group them (\textit{grouping 1}) to see that $92.61\%$ of the array computations have
+a static pattern that are either constants or constant of offsets
+of base/derived induction variables, thus confirming
+\emph{hypothesis 1}. Note that affine (derived induction variables) are very
+rare. Overwhelmingly the main category is neighbourhood offsets
+(\emph{grouping 2})
+(possibly mixed with constants in some dimensions), at $81.4\%$ in
+total. The ``affine + constants'' class is not represented at all.

### 2017-04-14 Study.hs
first results for hypothesis 1

@@ -1,5 +1,7 @@
 {-# LANGUAGE GADTs #-}
 {-# LANGUAGE DataKinds #-}
+{-# LANGUAGE FlexibleInstances #-}
+{-# LANGUAGE UndecidableInstances #-}
 
 import qualified Data.Map as M
 import Results
@@ -22,11 +24,25 @@ regroup :: (Ord c, HistogramShow t) => (Cat -> c) -> M.Map Cat t -> M.Map c t
 regroup classifier =
   M.fromListWith histZip . map (\(k, v) -> (classifier k, v)) . M.assocs
 
+resultLine cat dat = showA cat
+           ++ (replicate (35 - (length (show cat))) ' ')
+           ++ " & " ++ dat ++ "  \\\\\n"
+
+class ShowA s where
+  showA :: s -> String
+
+instance {-# OVERLAPS #-} ShowA String where
+  showA = id
+
+instance {-# OVERLAPPABLE #-} Show a => ShowA a where
+  showA = show
+
 mapViewT msg map =
        "   " ++ msg ++ ":\n"
-    ++ rline' ((replicate 5 ' ') ++ "Total") (show' total)
-    ++ concatMap (\(cat, dat) -> hline' cat (showPad dat ++ "("
-                                 ++ (twoDP $ (cast(dat)/cast(total))*100) ++ "%)")) (M.assocs map)
+    ++ concatMap (\(cat, dat) -> resultLine cat (showPad dat ++ "& ("
+    ++ (twoDP $ (cast(dat)/cast(total))*100) ++ "\\%)")) (M.assocs map)
+    ++ resultLine "Total" (showPad total ++ "&")
+    ++ "\n"
   where
     showPad dat = show dat ++ replicate (10 - (length (show dat))) ' '
     twoDP = printf "%.2f"
@@ -61,6 +77,7 @@ hypothesis1finer (_, rhs, _) =
   case rhs of
     Affines    c (R (s, _, _, _)) | s /= Other -> "Affine RHS" ++ hasConst c
     Neighbours c (R (s, _, _, _)) | s /= Other -> "Neigh RHS" ++ hasConst c
+    AllConsts                                  -> "All consts"
     _                                          -> "Other"
 
 hasConst WithConsts = " + constants"
@@ -101,16 +118,19 @@ hypothesis3 (_, rhs, _) =
 -- variables.
 
[…]

### 2017-04-15 oopsla17/analysis/Neighbour.hs
change to the way positioning is classified

@@ -8,11 +8,11 @@ module Neighbour where
 import Results
 
 import Data.List
-import Camfort.Helpers.Vec
+import Camfort.Helpers.Vec hiding (toList)
 import Camfort.Specification.Stencils.Syntax
 import Camfort.Specification.Stencils.InferenceBackend
 import Camfort.Specification.Stencils.InferenceFrontend
-
+import Data.Maybe
 
 
 shapeAndPosition :: [[Neighbour]] -> (Shape, Position)
@@ -22,22 +22,77 @@ shapeAndPosition ivs =
     Just intIvs ->
       case fromLists intIvs of
         VL intIvs ->
-          classifyIntervalRegions (inferMinimalVectorRegions intIvs)
+            case regions of
+              [x] -> (Orthotope, positioning)
+              xs  -> (SumOfOrthotope, positioning)
+          where
+            regions = inferMinimalVectorRegions intIvs
+            positioning = if (any originVec intIvs)
+                          then OverOrigin
+                          else intervalsToPosition regions
 
-classifyIntervalRegions :: [Span (Vec n Int)] -> (Shape, Position)
-classifyIntervalRegions [] = error "classifyIntervalRegions []"
-classifyIntervalRegions [x] =
-    (Orthotope, intervalToPosition x)
-classifyIntervalRegions xs =
-    (SumOfOrthotope, foldr1 joinPosition . map intervalToPosition $ xs)
-
-intervalToPosition = foldr1 joinPosition . reportEmptyError . map positionInterval . toList . transposeVecInterval
+intervalsToPosition :: [Span (Vec n Int)] -> Position
+intervalsToPosition vs =
+    fromMaybe OverOrigin
+  . foldVecMeet
+  . foldr1 meetVecPosition
+  . map (fmap positionInterval . transposeVecInterval)
+  $ vs
 reportEmptyError [] = error "intervalToPosition []"
 reportEmptyError xs = xs
+
+originVec :: Vec n Int -> Bool
+originVec Nil         = True
+originVec (Cons 0 vs) = originVec vs
+originVec _           = False
+
+originSpan :: Vec n Int -> Vec n Int -> Bool
+originSpan Nil Nil                 = True
+originSpan (Cons 0 xs) (Cons 0 ys) = originSpan xs ys
+originSpan _ _                     = False
 
 toList :: Vec n a -> [a]
 toList Nil = []
[…]

### 2017-04-15 oopsla17/oopsla17.tex
Fix model introduction

@@
 \noindent
-We define a simple set-theoretic model of array access patterns
-which serves as both a denotational model of the semantics
-of our specification language and an abstract interpretation
-domain for source code. This model has a number of uses: 
-(1) it serves to explain the meaning of our specifications
-(2) it is used in the inference and checking algorithms
-(\Cref{sec:analysis}); (3) justifies an equational theory for
-specifications in the next section; and (4) can be used to guide
-correct implementations.
+We define a lattice model of array access patterns which serves as both a
+denotational model of the semantics of our specification language and an
+abstract interpretation domain for source code. This model (1) serves to
+explain the meaning of our specifications; (2) is used in the inference and
+checking algorithms (\Cref{sec:analysis}); (3) justifies an equational theory
+for specifications in the next section; (4) is used to optimise specifications
+using lattice identities; and (5) can be used to guide correct implementations.
 

### 2017-04-15 oopsla17/oopsla17.tex
Add conventions to index scheme & improve

@@
   An $N$-dimensional \emph{index scheme} is a vector of length $N$ of
-  extended integer sets, \ie{}, a member of $\mathcal{P}(\bz)^{N}$.
-  An equivalent view of these vectors is as an $N$-times finite Cartesian
-  product on subsets of $\bz$.  We use $S, T, U$ to denote index
-  schemes.
+  integer sets, \ie{}, a member of $\mathcal{P}(\bz)^{N}$.
+%  An equivalent view of these vectors is as an $N$-times finite Cartesian
+%  product on subsets of $\bz$.
+  We use $S, T, U$ to denote index schemes. Henceforth, we assume index schemes
+  are all $N$-dimensional for some $N$.
 
   Index schemes can be \emph{projected} in the $i^{\textit{th}}$ dimension by
-  $\pi_i : {\mathcal{P}(\bz{})}^N \to \mathcal{P}(\bz{})$ where $1 \leq i
-  \leq N$. For an index scheme $I$, we refer to $\pi_i(S)$ as the
-  \emph{$i^{th}$ component} of $S$.
+  $\pi_i : {\mathcal{P}(\bz{})}^N \to \mathcal{P}(\bz{})$. For an index scheme
+  $I$, we refer to $\pi_i(S)$ as the \emph{$i^{th}$ component} of $S$. We assume
+  that $i$, when used for projection, always lies between $1$ and $N$.
 \end{defn}

### 2017-04-15 oopsla17/oopsla17.tex
Tidy unnion and intersection lemmas

@@
 \label{lem:vector-intersect}
-  Intersection distributes over $N$-dimensional index schemes. 
-  That is, for index schemes $S, T \in \mathcal{P}(\bz)^N$
+  Intersection distributes over index schemes. That is, for index schemes $S, T
+  \in \mathcal{P}(\bz)^N$
 %
@@
 
- Union does not distribute over $N$-dimensional index schemes--
- however, a more restricted property holds.
+Union does not distribute over index schemes, however, a more restricted
+property holds.
 
@@
 \label{lem:vector-union}
-Let $S, T$ be $N$-dimensional index schemes such that
-$\pi_i(S) = \pi_i(T)$ for all $1 \leq i \leq N$ apart 
-from some dimension $k$ where $\pi_k(S) \neq \pi_k(T)$,
-then:
+$S$ and $T$ are index schemes such that $\pi_i(S) = \pi_i(T)$ for all $1 \leq i
+\leq N$ apart from some dimension $k$, then:
 %
@@
   \end{equation*}
-%  \begin{equation*}
-%    S \cup T
-%    =
-%    (\prod_{i=1}^{k-1} \pi_i(S)) \times (\pi_k(S) \cup \pi_k(T)) \times \prod_{k+1}^{N} \pi_i(S)
-%  \end{equation*}
-%
 \end{restatable}

### 2017-04-15 oopsla17/study.tex
switch to si unit format in table

@@
 \item Most loop-array computations of the previous form read
-from a arrays with a \emph{contiguous} pattern, \eg{}:
+from arrays with a \emph{contiguous} pattern, \eg{}:
 %
@@
 
-%\begin{enumerate}
-%\item Loops over arrays mainly read from arrays with a static pattern
-%based on constant offsets from (base or dervied) induction variables;
-%\end{enumerate}
-
-%\begin{tabular}
-%Neighbourhood or Constant Affine RHS &  Total
-%\end{tabular}
-
-\noindent
+We first consider the initial hypotheses in turn again the collected data.
+%
+\begin{enumerate}
+\item Loops over arrays mainly read from arrays with a static pattern
+based on constant offsets from (base or dervied) induction variables;
+\end{enumerate}
+%
 We identified $108,773$ instances of assignments within loops in
@@
 to each one of these as an \emph{array computation}.
-We performed the classification described in the methodology,
-and then grouped the data by based only on the right-hand side
-classification:
+We performed the classification described above
+and grouped the data by the right-hand side classification only:
 %%
 \begin{center}
-\begin{tabular}{lrrcc}
-\textbf{RHS classification} & \textbf{Number} & \textbf{\%} &
+\begin{tabular}{lS[table-format=4]S[table-format=2.2]cc}
+\textbf{RHS classification} & \textbf{Number} & \% &
 \textit{grouping 1} & \textit{grouping 2} \\ \hline
@@
 \end{center}
-
+%
 \noindent
 Since, affine, consant, and neighbour are all static patterns, we
-group them (\textit{grouping 1}) to see that $92.61\%$ of the array computations have
-a static pattern that are either constants or constant of offsets
-of base/derived induction variables, thus confirming
-\emph{hypothesis 1}. Note that affine (derived induction variables) are very
-rare. Overwhelmingly the main category is neighbourhood offsets
-(\emph{grouping 2})
-(possibly mixed with constants in some dimensions), at $81.4\%$ in
-total. The ``affine + constants'' class is not represented at all.
+group them (\textit{grouping 1}) to see that $92.61\%$ of the array
+computations (\numexpr(100,737))
+ have a static pattern that are either constants or
+constant offsets of base/derived induction variables, confirming
+\emph{hypothesis 1}. Note that affine (derived induction variables)
+are rare. Overwhelmingly, the main category is neighbourhood
+offsets (\emph{grouping 2}) (possibly mixed with constants in some
+dimensions), at $81.4\%$ in total. The ``affine + constants'' class is
+not represented at all.
+%
+\begin{enumerate}[resume]
+\item Most loop-array computations of the previous form read
+from arrays with a \emph{contiguous} pattern, \eg{}:
+%
+\begin{minted}{fortran}
+do i = 1, n
+  b(n) = (a(i) + a(i+1)) / 2.0
+\end{minted}
+%
+\item Most loop-array computations of the previous form read
+from arrays with a pattern that includes the immediate
+neighbours (offsets of 1 from the induction variables);
+\end{enumerate}
+%
+Focussing on grouping1 
+\end{enumerate}

### 2017-04-15 oopsla17/study.tex
fixing broken tex in study.tex

@@
 group them (\textit{grouping 1}) to see that $92.61\%$ of the array
-computations (\numexpr(100,737))
- have a static pattern that are either constants or
+computations (100,737) have a static pattern that are either constants or
 constant offsets of base/derived induction variables, confirming
@@
 %
-Focussing on grouping1 
-\end{enumerate}
+%Focussing on grouping 1.

### 2017-04-15 oopsla17/oopsla17.tex
Return of the infinite interval

@@
 %
-
 \begin{defn}[Intervals with an optional hole]
-  We define an extended notion of closed interval on $\bz{}$ which may contain
-  a \emph{hole} at the origin, written \interv{a}{b}{c} where $a$ and $b$ are
-  drawn from \bz{} with $a \leq 0 \leq b$
-   and $c$ is drawn from $\mathbb{B} = \{ \mathsf{true}, \mathsf{false}
-  \}$. Intervals are interpreted as sets as follows:
+  We define an extended notion of closed interval on $\bz{}$ which may contain a
+  \emph{hole} at the origin, written \interv{a}{b}{c} where $a$ and $b$ are
+  drawn from \bz{} with $a \leq 0 \leq b$ and $c$ is drawn from $\mathbb{B} = \{
+  \mathsf{true}, \mathsf{false} \}$. Intervals are interpreted as sets as
+  follows:
 %
@@
   \end{equation*}
+
+  We also add the distinguished interval $[-\infty, \infty]$, which is simply an
+  alias for \bz{}, but this notation prevents handling infinite interval
+  separately in the following definitions, theorems, and proofs. Here, $-\infty$
+  and $\infty$ behaves like top and bottom elements to \bz{} respectively.
+
+  We denote the set of all such intervals (sets) as $\textit{Interval}$. If the
+  superscript to the interval is omitted it is treated as $\mathsf{true}$.
+\end{defn}
 %
-  We denote the set of all such intervals (sets) as
-   $\textit{Interval}$. If the superscript is
-  omitted it is treated as $\mathsf{true}$.
-\end{defn}
-
 \begin{restatable}{lem}{intervalIdentities}
@@
 
-We define two subsets of index scheme: \emph{subscript} schemes and
-\emph{interval} schemes:
+We define two specialisations (and subset spaces) of index scheme:
+\emph{subscript scheme} and \emph{interval scheme}:
 

### 2017-04-15 oopsla17/oopsla17.tex
Restore singletons in subscripts scheme

@@
 \begin{defn}[Interval scheme]
-  An interval scheme is a finite Cartesian product of intervals on
-  \bz{} or \bz{} itself, 
+  An interval scheme is a finite Cartesian product of intervals on \bz{},
   denoted by the set $\textit{Interval}^N$ for a product of $N$ intervals.
@@
 \begin{defn}[Subscript scheme]
-  A $N$-dimensional \emph{subscript scheme} is an $N$-dimensional
-  interval scheme where:
+  A \emph{subscript scheme} is an index scheme scheme where:
   %
@@
     \forall i.\ 1 \leq i \leq N \implies
-      \pi_i (S) = [p, p] 
+      \pi_i (S) = \{ p \}
       \; \vee \;
-      \pi_i (S) = \bz{}
+      \pi_i (S) = [-\infty, \infty]
     \end{equation*}
-    That is, the $i^{th}$ component of the set is either a singleton
-  in $\mathbb{Z}$ or the infinite interval.
+%
+  That is, the $i^{th}$ component of the set is either a singleton in \bz{} or
+  the infinite interval.
 \end{defn}

### 2017-04-15 oopsla17/oopsla17.tex
Fix region definition

@@
 \begin{defn}[Region]
-  A region \region{N} is an $N$-dimensional index scheme
-  (\ie{}, $\region{N} \subseteq \mathcal{P}(\bz)^N)$ defined as the smallest set satisfying the following:
+  A region is an index scheme and \region{N} is the set of all regions (\ie{},
+  $\region{N} \subseteq \mathcal{P}(\bz)^N)$. The set of all regions is  defined
+  as the smallest set satisfying the following:
 %
@@
 \end{defn}
-
+%
 \begin{restatable}{prop}{regionLattice}

### 2017-04-15 oopsla17/oopsla17.tex
Fix distributive bounded lattice proof of Region

@@
 \begin{proof}
-  Straightforward, since $\cup$ and $\cap$ are on sets (and therefore commutative,
-  associative, and absorptive with each
-  other). Furthermore, $\mathbb{Z}^N \cap R = R$ and $\empty \cup R =
-  R$ giving the bounds. Furthermore, the lattice is distributive since
- union distributes over intersection and vice versa
+  Straightforward, the join and meet are mapped to $\cup$ and $\cap$, the set is
+  inductively designed to be closed under these operations. Union and
+  intersection are associative, commutative, and absroptive under closed sets
+  and then they are also for \region{N}. This is enough to show that, it is a
+  lattice.
+
+  Further, we have $\bz{}^N \cap R = R$ and $\emptyset \cup R = R$ with
+  $\bz{}^N$ and $\emptyset$ belonging to \region{N}. This makes the lattice a
+  bounded one.
+
+  Finally, the lattice is distributive since union distributes over intersection
+  and vice versa when the set is closed under these operations.
 \end{proof}

### 2017-04-15 oopsla17/oopsla17.tex
Fix denotational semantics intro

@@
 information, \ie{} specifications are mapped to
-$\textsf{Mult} (\textsf{Approx} (\mathcal{P}({\bz})^N))$ where $N$ is
-the maximum dimensionality of the specification being modelled.
+$\textsf{Mult} (\textsf{Approx} (\region{N}))$.
 

### 2017-04-15 oopsla17/study.tex
hypothesis 2 and 3 data and discussion

@@
-At the beginning of this project, we had the following hypotheses:
+%\setlength{\tabcolsep}{8pt}
+
+At the beginning of this project, we made the following hypotheses
+about array programming idioms:
 \begin{enumerate}
@@
 %
-From these hypotheses, we conjectured that the common programming
-patterns captured above could be specified declaratively with a small
-set of combinators capture the shape of data access patterns based on
+From these hypotheses, we conjecture that the above programming idioms
+can be specified declaratively via a small
+set of combinators, capturing data access patterns based on
 overlapping (hyper)rectangles. We performed a large scale source-code
 analysis to validate these hypotheses and guide the design of our
-language. We believe these results are also of wider value.
-\dnote{Explain why: what else do they tell us?}
+language. These results are also of wider value to others, such as
+designers of array/stencil DSLs or autotuners.
 
@@
 
-Half of the packages come from active research teams who are our
-project partners (\ie{}, they were not selected carefully to skew
-results). BLAS was selected as it is a common numerical library;
+Half the packages come from active research teams who are our
+project partners (\ie{}, they were not selected carefully to
+fit our hypotheses). BLAS was selected as it is a common numerical library;
 similarly CP2 was selected as it gives standard numerical
@@
 
-We built a code analysis tool based on CamFort, an open-source analysis
-tool for Fortran programs~\cite{camfort}. Fortran source files are
-parsed to an AST and various standard control-flow and data-flow
-analyses are computed, including of key importance for us: induction
-variable identification, and reaching definitions. The resulting AST
-is traversed top-down and assignment statements inside loops are
-analysed and classified.
+We built a code analysis tool based on CamFort, an open-source Fortran
+analyser~\cite{camfort}. Fortran source files are parsed to an AST and
+ standard control-flow and data-flow analyses are computed,
+including of key importance for us: induction variable identification
+and reaching definitions. The resulting AST is traversed top-down and
+assignment statements inside loops are analysed and classified.
 
@@
 
-We focus on assignment statements where
-the right-hand side reads from one or more arrays. Left-hand
-sides and right-hand sides are classified in the following way:
+We focus on assignment statements in loops where the right-hand side
+expression reads from one or more arrays.  Left-hand sides and
+right-hand sides are classified in the following way:
 
-\begin{tabular}{l|l|l}
-  Shorthand & Classification (of subscripts) & Example \\ \hline
- (\textsf{vars}) & Variables (just LHS) & \fortran{x = ...}
+\begin{tabular}{l|l|l|l}
+  Shorthand & Classification (of subscripts) & Applies to & Example \\ \hline
+ (\textsf{vars}) & Variables & LHS & \fortran{x = ...}
   \\
- (\textsf{IVs}) & Induction variables (just LHS) & \fortran{a(i, j)} \\
- (\textsf{const}) & Constants & \fortran{a(0, 1)} \\
+ (\textsf{IVs}) & Induction variables & LHS & \fortran{a(i, j)} \\
  (\textsf{neigh}) & Neighbourhood offsets (of the form
-                                                  $i \pm c$) & \fortran{a(i, j-1)} \\
- (\textsf{neigh+c}) & Neighbourhood offsets and constants &
+                                                  $i \pm c$) & LHS/RHS
+                                                               & \fortran{a(i, j-1)} \\
+ (\textsf{neigh+c}) & Neighbourhood offsets and constants & LHS/RHS &
                                                             \fortran{b(i, 0, j+1)} \\
- (\textsf{aff}) & Affine offsets (of the form $a * i \pm b$) &
+ (\textsf{aff}) & Affine offsets (of the form $a * i \pm b$) & LHS/RHS &
                                      \fortran{a(2*i+1,j)} \\
- (\textsf{aff+c}) & Affine offsets and constants & \fortran{a(i+1, 0,
+ (\textsf{aff+c}) & Affine offsets and constants & LHS/RHS & \fortran{a(i+1, 0,
                                                    3*j+2)} \\
- (\textsf{other}) & Subscript expression not included above &
-\fortran{x(f(i))}, \fortran{a(i `mod` 2)}
+ (\textsf{other}) & Subscript expression not included above & LHS/RHS &
+\fortran{x(f(i))}, \fortran{a(i*i)}
 \end{tabular} \\[1em]
@@
 \noindent
-Note that the \textsf{vars} and \textsf{IVs} are only give to
+Note \textsf{vars} and \textsf{IVs} classifications apply only to
 left-hand sides.
@@
 \noindent
-Finally, we categorised the relationship between the left-hand side
+Finally, we categorise the relationship between the left-hand side
 and right-hand side, in terms of the use of induction variables.
@@
 
-We first consider the initial hypotheses in turn again the collected data.
+\noindent
+We revist each hypothesis and show the related data and conclusions.
 %
@@
 \begin{center}
-\begin{tabular}{lS[table-format=4]S[table-format=2.2]cc}
+\begin{tabular}{lS[table-format=4]@{\extracolsep{5pt}}S[table-format=2.2]c}
 \textbf{RHS classification} & \textbf{Number} & \% &
-\textit{grouping 1} & \textit{grouping 2} \\ \hline
-Affine                          & 34        & 0.03\%  &
-                                                            \multirow{4}{*}{92.61\%}
-  \\ \cline{1-3}\cline{5-5}
-All constant                    & 12160     & 11.18\%  \\ \cline{1-3}\cline{5-5}
-Neighbourhood                   & 65334     & 60.06\% &
-                                                          & \multirow{2}{*}{81.4\%}
-                                                           \\ \cline{1-3}
+\textit{grouping 1} \\ \hline
+Affine                          & 34        & 0.03\%
+& \multirow{3}{*}{81.43\%} \\ \cline{1-3}
+Neighbourhood                   & 65334     & 60.06 \%  \\ \cline{1-3}
 Neighbourhood + constants       & 23209     & 21.34\%  \\ \hline
-Other                           & 8036      & 7.39\%  \\ \hline \hline
+Other                           & 20196     & 18.57 \%  \\ \hline \hline
 Total                           & 108773    &  \\
@@
 \noindent
-Since, affine, consant, and neighbour are all static patterns, we
-group them (\textit{grouping 1}) to see that $92.61\%$ of the array
-computations (100,737) have a static pattern that are either constants or
+Since, affine and neighbour patterns are static, we
+group them (\textit{grouping 1}) to see that $81.43\%$ of the array
+computations (88,577) have a static pattern that are either constants or
 constant offsets of base/derived induction variables, confirming
@@
 are rare. Overwhelmingly, the main category is neighbourhood
-offsets (\emph{grouping 2}) (possibly mixed with constants in some
-dimensions), at $81.4\%$ in total. The ``affine + constants'' class is
+offsets (possibly mixed with constants in some
+dimensions). The ``affine + constants'' class is
 not represented at all.
@@
 %
+We grouped the data based on subclassifications of \emph{grouping 1}
+in terms of contiguity and positioning.
 %Focussing on grouping 1.
+\begin{center}
+\begin{tabular}{lS[table-format=4]S[table-format=2.2]S[table-format=2.2]lS[table-format=4]S[table-format=2.2]S[table-format=2.2]}
+\textbf{RHS classification} & \textbf{Number} & \% {(of whole)}
+                                              & \% {(of group)}
+&\textbf{Position} & \textbf{Number} & \% {(whole)} & \% {(group)} \\ \hline
+Affine, contiguous
+& 31 & 0.03\% & 0.03\% \\ \hline
+%\multirow{3}{*}{Affine, contiguous}
+% & %\multirow{3}{*}{31}
+% & %\multirow{3}{*}{0.03\%}
+% & \textsf{away} & 4 & 0.00\% \\
+%& 31 & 0.03\% & \textsf{origin} & 15 & 0.01\% \\
+%&&& \textsf{straddle} & 12 & 0.01\% \\ \hline
+Affine, non-contiguous
+ & 3
+ & 0.00\% & 0.00\% \\ \hline
+\multirow{3}{*}{Neighbour, contiguous}
+ & %\multirow{3}{*}{83689}
+ & %\multirow{3}{*}{83.08\%}
+& & \textsf{away} & 152 & 0.14\% & 0.17 \% \\
+& 83689 & 76.94\% & 94.48\% & \textsf{origin} & 81309 & 74.75\% & 91.79\% \\
+&&&& \textsf{straddle} & 2228 & 2.05\% & 2.52\% \\ \hline
+Neighbour, non-contiguous
+ & 4854
+ & 4.46\% & 5.48\%  \\ \hline \hline
+\textbf{Total}  & 88577    &  \\
+\end{tabular}
+\end{center}
+%%
+Thus, $76.94\%$ of array computations have only contiguous
+neighbour indices (which may include constants in some indices)
+in a contiguous pattern on the right hand side, with $\76.80\%$
+either at the origin ($74.75\%$) or offset by a distance of 1 from the origin
+($2.05\%$). This confirms both hypotheses 2 and 3. Only 152 instances
+were found of index patterns not immediate next to or over the origin.

### 2017-04-15 oopsla17/analysis/Study.hs

@@ -46,8 +46,8 @@ instance {-# OVERLAPPABLE #-} Show a => ShowA a where
 
 mapViewT msg map =
        "   " ++ msg ++ ":\n"
-    ++ concatMap (\(cat, dat) -> resultLine cat (showPad dat ++ "& ("
-    ++ (twoDP $ (cast(dat)/cast(total))*100) ++ "\\%)")) (M.assocs map)
+    ++ concatMap (\(cat, dat) -> resultLine cat (showPad dat ++ "& "
+    ++ (twoDP $ (cast(dat)/cast(total))*100) ++ "\\%")) (M.assocs map)
     ++ resultLine "Total" (showPad total ++ "&")
     ++ "\n"
   where
@@ -60,8 +60,11 @@ mapViewT msg map =
 hyps r =
      mapViewT "Hypothesis 1" (countWrapper hypothesis1 r)
   ++ mapViewT "Hypothesis 1 finer" (countWrapper hypothesis1finer r)
+  ++ mapViewT "Hypothesis 1 finer - all consts" (countWrapper hypothesis1finerA r)
   ++ mapViewT "Hypothesis 2" (countWrapperFilter hypothesis1filter hypothesis2 r)
+  ++ mapViewT "Hypothesis 2 - unfiltered" (countWrapper hypothesis2 r)
   ++ mapViewT "Hypothesis 3" (countWrapperFilter hypothesis1filter hypothesis3 r)
+  ++ mapViewT "Hypothesis 3 - unfiltered" (countWrapper hypothesis3 r)
   ++ mapViewT "Hypothesis 4" (countWrapper (interpret4 . hypothesis4) r)
   ++ mapViewT "Hypothesis 4 finer" (countWrapper hypothesis4finer r)
   ++ mapViewT "Hypothesis 5" (countWrapper hypothesis5 r)
@@ -94,15 +97,28 @@ hypothesis1finer (_, rhs, _) =
     AllConsts                                  -> "All consts"
     _                                          -> "Other"
 
+hypothesis1finerA :: (Form LHS, Form RHS, Consistency) -> String
+hypothesis1finerA (_, rhs, _) =
+  case rhs of
+    Affines    c (R (s, _, _, _)) | s /= Other -> "Affine RHS" ++ hasConst c
+    Neighbours c (R (s, _, _, _)) | s /= Other -> "Neigh RHS" ++ hasConst c
+    _                                          -> "Other"
+
 hasConst WithConsts = " + constants"
 hasConst Normal = " only"
 
[…]

### 2017-04-15 oopsla17/oopsla17.tex
Tidy up promote definition for interval schemes
Consistently use RegionN as codomain of spatial interpretation
Better semantics for array subscripts

@@
 
-\begin{defn}[Array subscripts and indices]
+\begin{defn}[Array subscripts and indices]\label{def:array-subs}
   An \emph{array subscript}, denoted $a(\bar{e})$, is an expression which reads
@@
   Let $\textit{promote}_N : \mathbb{N}^+ \times \textit{Interval} \to
-\textit{Interval}^N$ be a function generating an $N$-dimensional 
-interval such that $\pi_i(\vecgen{N}{i}{\interv{a}{b}{c}}) = 
-\interv{a}{b}{c}$ and $\pi_j(\vecgen{N}{i}{\interv{a}{b}{c}}) = 
-\bz{}$ in all other dimensions $j$. 
+  \textit{Interval}^N$ be a function generating an interval scheme such that if
+  $v$ is $\vecgen{N}{i}{\interv{a}{b}{c}}$, then $\pi_i(v) = \interv{a}{b}{c}$
+  and $\pi_j(v) = \bz{}$ in all other dimensions $j$.
 \end{defn}
@@
 \begin{align*}
-\interp{-}_N & : \textit{region} \rightarrow \mathcal{P}(\bz)^N \\
+  \interp{-}_N & : \textit{region} \rightarrow \region{N} \\
+%
   \interp{\stencil{p}{$i$}{}{}}_N & =
@@
 of interval schemes. Thus regions are modelled as members
-of $\mathcal{P}(\bz)^N$.
+of $\region{N}$.
 
@@
 
-\subsection{Abstract interpretation}
+\subsection{Denotational semantics for array subscripts}
 \label{sec:fromcode}
@@
 \begin{defn}
-The function $\textit{schematic}_I$ parameterised by
-a set of induction variables $I$, which maps
-from an array subscript into a $n$-dimensional (closed) interval:
+  Recall array subscript terms of the form $a(\bar{e})$ from
+  \Cref{defarray-subs}. We interpret these terms with
+  $\interp{-}^{\mathit{aexp}} : \textit{array-exp} \to \mathcal{P}(\bz{})^N$.
 %
   \begin{align*}
-\textit{schematic}_I(\bar{e}) & =
-\prod_{i \in |\bar{e}|} \textit{interval}(e_i)
-\qquad
-\textit{interval}_I(e) = \begin{cases}
-[c, c] & e = i \pm c \, \wedge \, i \in I \\
-\bz & FV(e) \cap I = \emptyset
+    \interp{a(\bar{e})}^{\mathit{aexp}} =
+      \prod_{1 \leq i \leq N} \mathit{subscript}(\bar{e}_i) & &
+  %
+    \textit{subscript}(e) = \begin{cases}
+      \{ c \} & e \equiv i \pm c \\
+      \bz & e \; \mbox{is constant}
     \end{cases}
   \end{align*}
-Note that \textit{inveral} may be undefined (it is a partial function)
-in which case \textit{schematic} is undefined if any of its
-components are undefined.
 \end{defn}

### 2017-04-15 oopsla17/study.tex
Unfuck the text file

@@
 neighbour indices (which may include constants in some indices)
-in a contiguous pattern on the right hand side, with $\76.80\%$
+in a contiguous pattern on the right hand side, with $76.80\%$
 either at the origin ($74.75\%$) or offset by a distance of 1 from the origin

### 2017-04-15 oopsla17/oopsla17.tex
Typo
Emphasise partial nature of array term interpretation

@@
 
-\begin{defn}[Neighbourhood index]
-\label{def:neighbour}
+\begin{defn}[Neighbourhood index]\label{def:neighbour-ix}
   For an array subscript $a(\bar{e})$ an index $e \in \bar{e}$ is a
@@
   Recall array subscript terms of the form $a(\bar{e})$ from
-  \Cref{defarray-subs}. We interpret these terms with
-  $\interp{-}^{\mathit{aexp}} : \textit{array-exp} \to \mathcal{P}(\bz{})^N$.
+  \Cref{def:array-subs}. We interpret these terms with the partial
+  interpretation $\interp{-}^{\mathit{aterm}} : \textit{array-term}
+  \prightarrow{} \mathcal{P}(\bz{})^N$. The interpretation is defined when
+  all indices are either constant or neighbourhood indices as defined in
+  \Cref{def:neighbour-ix}.
 %
   \begin{align*}
-    \interp{a(\bar{e})}^{\mathit{aexp}} =
+    \interp{a(\bar{e})}^{\mathit{aterm}} =
       \prod_{1 \leq i \leq N} \mathit{subscript}(\bar{e}_i) & &

### 2017-04-15 oopsla17/study.tex
Notes on hypothesis 2

@@
 not represented at all.
+\mnote{(1) State that constant here includes variables that are constant with
+respect to the current loop scope.}
+\mnote{(2) Mention what sort of things are
+included in other.}
+\mnote{(3) Add a remark about lack of affine indices to be
+perhaps programmers relying on compiler more regarding optimisations than hand
+optimising. We can later use this to justify our inability to handle affine
+indices.}
 %

### 2017-04-15 oopsla17/study.tex
improved table 2

--- 2017-04-15 oopsla17/study.tex
@@
 
-\item Many array computations are \emph{stencil computations},
-with a static access pattern as described in Hypothesis 2, writing
-to an array at an index based on a (possibly constant offset from induction
-variables, \eg{}, the classic five-point stencil;
+\item Many array computations are \emph{stencil computations}: an
+  assignment whose right-hand side comprises array subscripts with
+  neighbourhood indices and whose left-hand side is an
+  neighbourhood-indexed array subscript (\eg{}, the classic five-point
+  stencil);
 
@@
 \begin{center}
-\begin{tabular}{lS[table-format=4]S[table-format=2.2]S[table-format=2.2]lS[table-format=4]S[table-format=2.2]S[table-format=2.2]}
+\begin{tabular}{lS[table-format=4]S[table-format=2.2]lS[table-format=4]S[table-format=2.2]}
 \textbf{RHS classification} & \textbf{Number} & \% {(of whole)}
-                                              & \% {(of group)}
-&\textbf{Position} & \textbf{Number} & \% {(whole)} & \% {(group)} \\ \hline
+&\textbf{Position} & \textbf{Number} & \% {(of whole)} \\ \hline
 Affine, contiguous
-& 31 & 0.03\% & 0.03\% \\ \hline
+& 31 & 0.03\% \\ \hline
 %\multirow{3}{*}{Affine, contiguous}
@@
  & 3
- & 0.00\% & 0.00\% \\ \hline
+ & 0.00\% & \\ \hline
 \multirow{3}{*}{Neighbour, contiguous}
@@
  & %\multirow{3}{*}{83.08\%}
-& & \textsf{away} & 152 & 0.14\% & 0.17 \% \\
-& 83689 & 76.94\% & 94.48\% & \textsf{origin} & 81309 & 74.75\% & 91.79\% \\
-&&&& \textsf{straddle} & 2228 & 2.05\% & 2.52\% \\ \hline
+ & \textsf{away} & 152 & 0.14\% \\
+& 83689 & 76.94\% & \textsf{origin} & 81309 & 74.75\% \\
+&&& \textsf{straddle} & 2228 & 2.05\% \\ \hline
 Neighbour, non-contiguous
  & 4854
- & 4.46\% & 5.48\%  \\ \hline \hline
+ & 4.46\% \\ \hline
+Other & 20196 & 18.57\%  \\
+ \hline
 \textbf{Total}  & 88577    &  \\
@@
 %%
-Thus, $76.94\%$ of array computations have only contiguous
-neighbour indices (which may include constants in some indices)
-in a contiguous pattern on the right hand side, with $76.80\%$
-either at the origin ($74.75\%$) or offset by a distance of 1 from the origin
+Thus, $76.94\%$ of array computations have only contiguous neighbour
+indices (which may include constants in some indices) in a contiguous
+pattern on the right hand side, with $76.80\%$ either at the origin
+($74.75\%$) or offset by a distance of 1 from the origin
 ($2.05\%$). This confirms both hypotheses 2 and 3. Only 152 instances
-were found of index patterns not immediate next to or over the origin.
+of index patterns not immediately next to or over the origin were
+found.  There is a non trivial amount of non-contiguous arrays access,
+but it is a much smaller proportion of the whole.
+
+\begin{enumerate}[resume]
+\item Many array computations are \emph{stencil computations}:
+an assignment whose
+ right-hand side comprises array subscripts with neighbourhood indices
+and whose left-hand side is an neighbourhood-indexed array subscript
+(\eg{}, the classic five-point stencil);
+\end{enumerate}

### 2017-04-16 oopsla17/study.tex
hypothesis 4 and 5 results

@@
                                   b(i+1,j+1)} \\
-  (\textsf{perm} & Permutation & \fortran{a(i, j) = c(j, i)} or \fortran{a(i,
+  (\textsf{perm}) & Permutation & \fortran{a(i, j) = c(j, i)} or \fortran{a(i,
                                                                0) =
                                  b(0, i)} \\
-  (\textsf{LHSsub} & LHS subset & \fortran{a(i) = b(i, j) + b(i, j-1)}
+  (\textsf{LHSsub}) & LHS subset & \fortran{a(i) = b(i, j) + b(i, j-1)}
   \\
-  (\textsf{LHSsup} & LHS superset & \fortran{a(i, j) = b(i)} \\
-  (\textsf{inconst} & Inconsistent & \fortran{a(i) = b(j)}
+  (\textsf{LHSsup}) & LHS superset & \fortran{a(i, j) = b(i)} \\
+  (\textsf{inconst}) & Inconsistent & \fortran{a(i) = b(j)}
 \end{tabular} \\[1em]
@@
 \begin{center}
-\begin{tabular}{lS[table-format=4]@{\extracolsep{5pt}}S[table-format=2.2]c}
+\begin{tabular}{l|S[table-format=4]@{\extracolsep{5pt}}S[table-format=2.2]c}
 \textbf{RHS classification} & \textbf{Number} & \% &
@@
 Other                           & 20196     & 18.57 \%  \\ \hline \hline
-Total                           & 108773    &  \\
+\textbf{Total}                  & 108773    &  \\
 \end{tabular}
@@
 \begin{center}
-\begin{tabular}{lS[table-format=4]S[table-format=2.2]lS[table-format=4]S[table-format=2.2]}
+\begin{tabular}{l|S[table-format=4]S[table-format=2.2]lS[table-format=4]S[table-format=2.2]}
 \textbf{RHS classification} & \textbf{Number} & \% {(of whole)}
@@
 Other & 20196 & 18.57\%  \\
- \hline
+ \hline \hline
 \textbf{Total}  & 88577    &  \\
@@
 \end{enumerate}
+%
+We subclassified the data based on the right-hand side
+and left-hand side categories
+being either a variable, or induction-variable indexed,
+neighbour-indexed or affined-indexed array. We additionally
+included an orthogonal classification of whether
+left-hand and right-hand sides are consistent (have a common subset
+of induction variables) or not.
+%
+\begin{center}
+\begin{tabular}{ll|S[table-format=4]S[table-format=2.2]S[table-format=4]S[table-format=2.2]c}
+\multicolumn{2}{l|}{\textbf{LHS and RHS classification}}
+& \textbf{\# inconsistent}
+& \% {(whole)}
+& \textbf{\# consistent}
+& \% {(whole)}
+& \textit{stencils} \\ \hline
+%
+LHS Vars & RHS neigh & 0 & 0.00\% & 4591      & 4.22\%
+ & \multirow{7}{*}{66.01\%} \\\cline{1-6}
+%
+%
+%
+LHS IVs & RHS neigh   &  4560      & 4.19\%  & 50861     & 46.76\% \\\cline{1-6}
+%
+%
+LHS neigh & RHS neigh & 2358      & 2.17\% & 16336     & 15.02\%   \\\cline{1-6}
+%
+LHS Vars & RHS affines & 0 & 0.00\%  & 4         & 0.00\% \\\cline{1-6}
+%
+LHS IVs, & RHS affines & 7 & 0.01\% & 0 & 0.00\%  \\\cline{1-6}
+%
+LHS affine & RHS affines & 0 & 0.00\% &  6         & 0.01\% \\\hline
+%
+LHS neigh & RHS affines & 8         & 0.01\% & 0 & 0.00\%  \\\hline
+%%
+LHS affine & RHS neigh  & 23        & 0.02\% & 0 & 0.00\% \\\hline \hline
+%
+other       &                      & 30019     & 27.60\%  & &  \\\hline%\hline
+%Total        &                      & 108773    &
+%  LHS var, & RHS affine             & 4         & 0.00\%
+% & \multirow{7}{*}{72.37\%} \\\cline{1-3}
+%  LHS var, & RHS neighbour          & 4591      & 4.22\%  \\\cline{1-3}
+%  LHS IVs, &RHS affine             & 7         & 0.01\%  \\\cline{1-3}
+%  LHS IVs, &RHS neighbour          & 55421     & 50.95\% \\\cline{1-3}
+%  LHS neighbour, &RHS neighbour    & 18694     & 17.19\% \\\cline{1-3}
+%  LHS affine, &RHS affine          & 6         & 0.01\%  \\\hline
+%  LHS affine, &RHS neighbour       & 23        & 0.02\%  \\\hline
+%  LHS neighbour, &RHS affine       & 8         & 0.01\%  \\\hline
+%  other   & other                   & 30019     & 27.60\%  %\\\hline\hline
+%  Total                           & 108773    &  \\
+\end{tabular}
+\end{center}
+%
+The most common category has an l.h.s array subscript
+indexed by induction variables, and a right-hand side comprising
+neighbourhood offsets. Next most common is to have neighbourhood
+offsets on the left-hand and right-hand sides. Notably
+there are two categories always seen as inconsistent:
+``LHS affine, RHS neighbour'' and ``LHS neighbour, RHS affine''
+since they comprise different indexing schemes
+on each side of the assignment, e.g. $a(i + 1) = b(2*i + 1)$. These
+appear very rarely. Thus, we find that $66.01\%$ of array computations
+have a static pattern on the left and right-hand side that is either
+affine or a constant translation. Thus, the data support ours
+hypothesis that stencis are common. Based on this grouping, we
+considered the two subclassifications of different kinds of
+left-right-hand side relationship and contiguity:
+%%
+\begin{center}
+\begin{minipage}{1\linewidth}
+\begin{tabular}{l|S[table-format=4]S[table-format=2.2]}
+\textbf{Contiguity} & \textbf{Number} & \% {(whole)} \\\hline
+Contiguous stencil         & 68246     & 62.74\%  \\\hline
+Non-contiguous stencil     & 3552      & 3.27\%  \\\hline
+other                      & 36975     & 33.99\% \\\hline
+%\textbf{Total}             & 108773    &  \\
+\end{tabular}
+\hspace{1em}
+\begin{tabular}{l|S[table-format=4]S[table-format=2.2]S[table-format=4]S[table-format=2.2]}
+\textbf{Consistency} & \textbf{Number} & \% & \textbf{relativised \#} & \% \\\hline
+Direct                            & 47448     & 43.62\% & 636       & 0.58\%  \\\hline
+LHSsubset                         & 9812      & 9.02\% & 65        & 0.06\%  \\\hline
+LHSsuperset                       & 10110     & 9.29\%  & 84        & 0.08\%  \\\hline
+Permutation                       & 3604      & 3.31\%  & 39        & 0.04\%  \\\hline
+other                             & 36975     & 33.99\% \\\hline
+\end{tabular}
+\end{minipage}
+\end{center}
+%%
+Most stencils are contiguous, but there is a non trivial
+amount of non-contiguity. Most stencils have matching induction
+variables on the left and right-hand sides but again there
+are non-trivial amounts of other kinds of relationship.
+%%
+\begin{enumerate}[resume]
+\item Many array computations read from each particular index just
+  once.
+\end{enumerate}
+%
+We classified right-hand sides which are neighbourhood offsets (or
+affine) and subclassified this by whether array subscripts were
+unique or repeated:
+
+\begin{center}
+\begin{tabular}{l|S[table-format=4]S[table-format=2.2]S[table-format=4]}
+\textbf{RHS form and reuse} & \textbf{Number} & \% \\ \hline
+RHS neighbour/affine unique indices      & 78349     & 72.03\%  \\\hline
+RHS neighbour/affine repeated indices   & 10228     & 9.40\%  \\\hline
+other                             & 20196     & 18.57\%  \\\hline\hline
+Total                             & 108773    &
+\end{tabular}
+\end{center}
+%
+Thus, hypothesis 5 is confirmed, the majority of array computations
+have access patterns with linear usage of array subscripts.

### 2017-04-16 Study.hs
hypothesis 4 and 5 results

@@ -65,8 +65,16 @@ hyps r =
   ++ mapViewT "Hypothesis 2 - unfiltered" (countWrapper hypothesis2 r)
   ++ mapViewT "Hypothesis 3" (countWrapperFilter hypothesis1filter hypothesis3 r)
   ++ mapViewT "Hypothesis 3 - unfiltered" (countWrapper hypothesis3 r)
-  ++ mapViewT "Hypothesis 4" (countWrapper (interpret4 . hypothesis4) r)
-  ++ mapViewT "Hypothesis 4 finer" (countWrapper hypothesis4finer r)
+
+--  ++ mapViewT "Hypothesis 4" (countWrapper (interpret4 . hypothesis4) r)
+--  ++ mapViewT "Hypothesis 4 finer" (countWrapper hypothesis4finer r)
+
+  ++ mapViewT "Hypothesis 4A" (countWrapper hypothesis4A r)
+  ++ mapViewT "Hypothesis 4A" (countWrapper hypothesis4AInconsistents r)
+  ++ mapViewT "Hypothesis 4B" (countWrapper hypothesis4B r)
+  ++ mapViewT "Hypothesis 4 contig" (countWrapper hypothesis4contig r)
+  ++ mapViewT "Hypothesis 4 consistency" (countWrapper hypothesis4consistency r)
+
   ++ mapViewT "Hypothesis 5" (countWrapper hypothesis5 r)
 
 countWrapper :: Ord c => (Cat -> c) -> Result -> M.Map c Int
@@ -173,6 +181,93 @@ interpret4 (340, b, c) = "lhs(allconst), RHS(neigh, b, contig,I" ++ showe c ++ "
 
 showe _ = ""
 
+hypothesis4A :: (Form LHS, Form RHS, Consistency) -> String
+hypothesis4A (Subscripts, _, _) = "other"
+hypothesis4A (AllConsts, _, _) = "other"
+hypothesis4A (lhs, rhs, const) =
+  if classRhs rhs == "other" then
+    "other"
+  else
+    "LHS " ++ classLhs ++ ", RHS " ++ (classRhs rhs)
+  where
+    classConst = show const
+    classLhs =
+      case lhs of
+        Neighbours _ _ -> "neigh"
+        Affines    _ _ -> "affine"
+        _              -> show lhs
+    classRhs
+      (Affines c (R (s, p, con, _))) | s /= Other = "affines"
+    classRhs
[…]

### 2017-04-16 oopsla17/oopsla17.tex
Add checking to the paper

@@
 
+\subsection{Analysis}
+
+\subsection{Checking}
+
+In the model, we can regard both array indices and specifications as groups of
+points in $N$-dimensional space. Notion of consistency is then intuitively
+whether these two (potentially infinite) sets of points in space match or not.
+This leads to a simple set based consistency function defined as
+%
+\begin{align*}
+  \mathit{consistent} & :
+    \mathsf{Mult}(\mathcal{P}(\mathcal{P}(\bz{})^N)) \times
+    \mathsf{Mult}(\mathsf{Approx}(\region{N})) \to \mathbb{B} & \\
+%
+  \mathit{consistent}(\mathit{ix}, \mathit{model}) & = \begin{cases}
+    \mathsf{false} & \mathit{model} = \mathsf{once}(x) \wedge ix = \mathsf{mult}(y) \\
+    \mathit{consistent'}(\mathit{peel}(ix), \mathit{peel}(model)) & \textit{otherwise}
+  \end{cases} \\
+%
+  \mathit{consistent'} & :
+    \mathcal{P}(\mathcal{P}(\bz{})^N) \times
+    \mathsf{Approx}(\region{N}) \to \mathbb{B} & \\
+%
+  \mathit{consistent'}(\mathit{ix}, \mathit{model}) & = \begin{cases}
+    m = \mathit{ix} & \mathit{model} = \mathsf{exact}(m) \\
+    m \supseteq \mathit{ix} & \mathit{model} = \mathsf{upper}(m) \\
+    m \subseteq \mathit{ix} & \mathit{model} = \mathsf{lower}(m)
+  \end{cases}
+\end{align*}
+%
+The first function, $\mathit{consistent}$, checks whether the linearity of the
+specification matches that of the indices, \ie{} if the specification allows
+indices to be repeated or not.Then it delegates to $\mathit{consistent'}$
+function which checks if the points observed in the array expression matches
+the space defined by the specification. Lower bounds, marked by \texttt{atLeast}
+modifier, require the space defined by the specification to remain inside the
+set of indices, while an upper bound, market by \texttt{atMost} modifier,
+requires the opposite: enclosure. In the absence of such modifiers, we expect
+the space defined by observed indices correspond exactly those defined by the
+region in the specification, hence requiring set equality.
+
+Reader will realise that since index schemes allows infinite intervals in the
+product, set equality cannot be done by exhaustively checking the points of one
+are inside the other. Instead, we compile the region into interval constraints
+and subscript schemes into membership constraints and pass it to
+\textsc{Z3}~\citep{de2008z3} SMT solver to see if they are equal. The query
+belongs to to quantifier free linear arithmetic, which is a decidable theory.
+Performance of consistency checking using SMT solver is irrelevant as the number
+of interval constraints and the number of indices that can reasonably appear in
+an assignment is too small restricting the size of the formula being checked.
+
+\mnote{It would be great if we had data on number of indexing expressions that
+occur (counting the ones that flow in) and also the length of specifications
+(which I think we do generate). This would justify my efficiency claims}.
+
+\subsection{Inference}
+
 \section{Evaluation}

### 2017-04-16 oopsla17/oopsla17.tex
Explain union normal form

@@
 
+\subsection{Union Normal Form}\label{subsec:union-normal-form}
+
+We need to address the problem of representation in the model both for observed
+array terms converted into subscripts schemes and for specifications converted
+into interval schemes. In each case, the objects are sets and union and
+intersection are the combinators. However, they are often infinite sets due to
+infinite interval and constant indices being converted to \bz{}. This calls for
+a finite (and ideally compact) representation for indices.
+
+First recall \Cref{lem:vector-intersect}, which states that intersection of
+indexing schemes is still an indexing scheme. When combined with
+\Cref{lem:interval-identities}, we can extend this closure to interval schemes.
+Subscripts schemes do not enjoy the same propery as intersection of two
+singleton sets might easily be empty, which disqualifies it from being a
+subscript index. In the following section we will not rely on intersection of
+subscript indices, so that is okay.
+
+Union is, however, not as flexible as intersection. \Cref{lem:vector-union} does
+not form a closure for indexing schemes in general. For this reason, we
+represent collection of all forms of indexing schemes as unevaluated unions.
+Since the model forms a distributive lattice as established in
+\Cref{prop:region-lattice}, the intersections can be pushed in using the
+distributive law leaving only unions. We further exploit associativity of union
+to put unions into a cons tree making the unions effectively a non-empty list.
+We do not, however, attempt to canonicalise the representation.
+
 \subsection{Analysis}

### 2017-04-16 oopsla17/algorithms.tex
start of putting in analysis and inference

--- 2016-11-16 paper.tex
@@
-\documentclass[10pt,preprint,numbers]{sigplanconf}
-
-\usepackage{amssymb}
-\usepackage{amsmath}
-\usepackage{amsthm}
-\usepackage{stmaryrd}
-\usepackage{color}
-\usepackage{graphics}
-\usepackage{fancyvrb}
-\usepackage{subfigure}
-\usepackage{multirow}
-\usepackage{minted}
-\usepackage{breqn}
-\usepackage{enumitem}
-%\usepackage{subcaption}
-\usepackage{siunitx} % For pretty-printing numeric values and SI units
-                     % of measure. e.g., the tabular column type S is
-                     % used to print nice-looking tables of numbers.
-\sisetup{ % defaults
-  group-separator={,},
-  group-minimum-digits={3},
-  output-decimal-marker={.},
-  table-format = 6
-}
-
-\usepackage{tikz}
-\usepackage{xypic}
-\usepackage{natbib}
-
-\usetikzlibrary{calc}
-
-%% http://tex.stackexchange.com/questions/55068/is-there-a-tikz-equivalent-to-the-pstricks-ncbar-command
-\tikzset{
-    ncbar angle/.initial=90,
-    ncbar/.style={
-        to path=(\tikztostart)
-        -- ($(\tikztostart)!#1!\pgfkeysvalueof{/tikz/ncbar angle}:(\tikztotarget)$)
-        -- ($(\tikztotarget)!($(\tikztostart)!#1!\pgfkeysvalueof{/tikz/ncbar angle}:(\tikztotarget)$)!\pgfkeysvalueof{/tikz/ncbar angle}:(\tikztostart)$)
-        -- (\tikztotarget)
-    },
-    ncbar/.default=0.5cm,
-}
-
-\tikzset{round left paren/.style={ncbar=0.4cm,out=110,in=-110}}
-\tikzset{round right paren/.style={ncbar=0.4cm,out=70,in=-70}}
-
-\fvset{
-  linenos=true,
-  fontsize=\footnotesize,
-  breaklines=true,
- breakafter=w),
-  xleftmargin=\parindent,
-  numbersep=1em
-}
-
-\usemintedstyle{vs}
-
-\definecolor{darkblue}{rgb}{0.0,0.0,0.5}
-\definecolor{darkgreen}{rgb}{0.0,0.4,0.0}
-\definecolor{darkdarkgreen}{rgb}{0.0,0.35,0.0}
-
-\usepackage{hyperref}
-\hypersetup{
-    unicode=false,          % non-Latin characters in Acrobat's bookmarks
-    pdftoolbar=true,        % show Acrobat toolbar?
-    pdfmenubar=true,        % show Acrobat menu?
-    pdffitwindow=false,      % page fit to window when opened
-    pdftitle={},    % title
-    pdfauthor={}
-    pdfsubject={},   % subject of the document
-    pdfnewwindow=true,      % links in new window
-    pdfkeywords={keywords}, % list of keywords
-    colorlinks=true,       % false: boxed links; true: colored links
-    linkcolor=darkblue,          % color of internal links
-    citecolor=darkblue,        % color of links to bibliography
-    filecolor=green,      % color of file links
-    urlcolor=blue,          % color of external links
-}
-
-\usepackage{cleveref}
-\crefname{section}{\S}{\S\S}
-\Crefname{section}{\S}{\S\S}
-\crefname{subsection}{\S}{\S\S\S}
-\Crefname{subsection}{\S}{\S\S\S}
-\crefname{figure}{fig.}{}
-\Crefname{figure}{Fig.}{}
-\crefname{definition}{def.}{}
-\Crefname{definition}{Def.}{}
-
-\CustomVerbatimEnvironment{SpecVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.5cm,
-xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.98,numbersep=0.9em}
-\CustomVerbatimEnvironment{ExmVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.5cm,
-xrightmargin=0.2cm,baselinestretch=0.98,numbers=left,numbersep=0.9em,commandchars=\\\{\}}
-\CustomVerbatimEnvironment{IVerbatim}{Verbatim}{fontsize=\relsize{-1},xleftmargin=0.5cm,
-xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.98,numbersep=0.9em}
-
-\definecolor{darkgreen}{rgb}{0.0,0.5,0.0}
-\definecolor{darkpurple}{rgb}{0.6,0.0,0.6}
-\definecolor{orange}{rgb}{0.8,0.4,0.0}
-\definecolor{darkorange}{rgb}{0.5,0.2,0.0}
-\definecolor{marco}{rgb}{0.0,0.3,0.5}
-\definecolor{gray}{rgb}{0.2,0.2,0.2}
-
-\newcommand\mdef{\stackrel{\mathclap{\normalfont\mbox{def}}}{=}}
-
-\newcommand{\bn}{\mathbb{N}}
-
-\newcommand{\todo}[1]{\textcolor{blue}{Dom: #1}}
-\newcommand{\dnote}[1]{\textcolor{darkpurple}{Dom: #1}}
-\newcommand{\mnote}[1]{\textcolor{darkgreen}{Mistral: #1}}
-\newcommand{\anote}[1]{\textcolor{red}{Andy: #1}}
-
-\newcounter{block}
-
-\newtheorem{lemma}[block]{Lemma}
-\newtheorem{proposition}[block]{Proposition}
-
-\theoremstyle{definition}
-
-\newtheorem{theorem}[block]{Theorem}
-\newtheorem{remark}[block]{Remark}
-\newtheorem{example}[block]{Example}
-\newtheorem{definition}[block]{Definition}
-
-% Writing macros
-\newcommand{\ie}{\emph{i.e.}}
-\newcommand{\eg}{\emph{e.g.}}
-
-\newcommand{\dimId}{\texttt{dim}}
-
-% Semantics related
-\newcommand{\interp}[1]{\llbracket{#1}\rrbracket}
-
-% Syntax macros
-\newcommand{\nonterm}[1]{\textit{#1}}
-\newcommand{\term}[1]{\texttt{#1}}
-
-\newcommand{\stenRefl}[1]{\term{pointed} \, (\term{dim=}#1)}
-\newcommand{\stenFwd}[3]{\term{forward} \, (\term{depth=}#1,
-  \term{dim=}#2{#3})}
-\newcommand{\stenBwd}[3]{\term{backward} \, (\term{depth=}#1,
-  \term{dim=}#2{#3})}
-\newcommand{\stenCen}[3]{\term{centered} \, (\term{depth=}#1,
-  \term{dim=}#2{#3})}
-\newcommand{\irrefl}{\texttt{nonpointed}}
-
-\newcommand{\stenReflS}[1]{\term{pointed} \, (\term{dim=}#1)}
-\newcommand{\stenFwdS}[2]{\term{fwd} \, (\term{depth=}#1,
-  \term{dim=}#2)}
-\newcommand{\stenBwdS}[2]{\term{bwd} \, (\term{depth=}#1,
-  \term{dim=}#2)}
-\newcommand{\stenCenS}[2]{\term{cen} \, (\term{depth=}#1,
-  \term{dim=}#2)}
-\newcommand{\irreflS}{\texttt{np}}
-
-\newcommand{\stenFwdSR}[3]{\term{fwd} (\term{depth=}#1,
-  \term{dim=}#2, #3)}
-\newcommand{\stenBwdSR}[3]{\term{bwd} (\term{depth=}#1,
-  \term{dim=}#2, #3)}
-\newcommand{\stenCenSR}[3]{\term{cen} (\term{depth=}#1,
-  \term{dim=}#2, #3)}
-\newcommand{\stenReflSR}[1]{\term{point} (\term{dim=}#1)}
-
-% SYNTAX OPERATIONS AND PREDICATES
-%\newcommand{\neigh}{\textsf{neigh}}
-\newcommand{\arrayTy}{\textsf{array}}
-\newcommand{\rhsExp}{\textsf{rhsExp}}
-\newcommand{\var}{\textsf{var}}
-
-%% VECTOR NOTATIONS
-\newcommand{\vect}[1]{\textbf{#1}}
-
-\newcommand{\vtwohs}[2]{\setlength{\arraycolsep}{0em}
-\left[\begin{array}{cc}#1 \, & \, #2\end{array}\right]\!}
-
-\newcommand{\vtwoh}[2]{\setlength{\arraycolsep}{0em}
-\left[\begin{array}{cc}#1 \; & \; #2\end{array}\right]}
-
-\newcommand{\vthreeh}[3]{\setlength{\arraycolsep}{0em}
-\left[\begin{array}{ccc}#1 \; & \; #2 \; & \; #3\end{array}\right]}
-\newcommand{\vtwo}[2]{\setlength{\arraycolsep}{0em}
-\left[\begin{array}{l}#1\\#2\end{array}\right]}
-\newcommand{\vthree}[3]{\setlength{\arraycolsep}{0em}
-\left[\begin{array}{l}$#1$\\$#2$\\$#3$\end{array}\right]}
-\newcommand{\stwo}[4]
-%{\vtwo{#1}{#2}\!\vtwo{#3}{#4}}
-{\setlength{\arraycolsep}{0.1em}
-\left[\begin{array}{rr}$#1$ & $#3$\\$#2$ & $#4$\end{array}\right]}
-
-\newcommand{\singleEntry}[2]{\textbf{J}_{#2}^{#1}}
-\newcommand{\zeroEntry}[2]{\textbf{K}_{#2}^{#1}}
-
-%% OPERATIONS ON SPANS and VECTORS
-\newcommand{\containedin}{\sqsubseteq}
-
-%% MODEL
-\newcommand{\effdims}[2]{\mathit{constr}(#1)_{#2}}
-
-\newcommand{\trule}[1]{{\footnotesize{(\text{#1})}}}
-%\include{results}
-
-%\newcommand{\cons}[3]{\mathbf{C}_{#3}(#1, #2)}
-\newcommand{\consName}{\textbf{agree\textprime}}
-\newcommand{\cons}[2]{#1 \,\, \consName{} \,\, #2}
-\newcommand{\consAName}{\textbf{agree}}
-\newcommand{\consA}[2]{#1 \,\, \consAName{} \,\, #2}
-\newcommand{\consSub}[3]{#2 \,\, \consAName{}_{#1} \,\,#3}
-
-\title{Verifying Spatial Properties of Stencil Computations}
-%\titlebanner{DRAFT -- Do not distribute}
-\authorinfo{}{}{}
-%\preprintfooter{DO NOT DISTRIBUTE}
-
-\begin{document}
-\maketitle
-
-\bibliographystyle{plainnat}
-
-\begin{abstract}
-Stencil computations are a common programming pattern in numerical
-  code and typically form the core of graphics and scientific
-  computing applications. However, they are often complex and
-  unwieldy, involving substantial fine-grained manipulation of array
-  indices across multiple arrays and dimensions. This requires careful
-  testing, which is almost always done manually and then discarded.
-  In this paper, we propose an alternate approach:
-  a specification language and automated tool for verifying the
-  spatial behaviour of stencils. The specification language is based
-  on the hypothesis that stencil computations read from arrays with a
-  regular, fixed pattern which can be captured by a simple
-  set of abstract combinators. These
-  specifications can be embedded as annotations in source code,
-  against which code can be checked for conformance. For legacy code,
-  spatial specifications can be inferred, providing documentation and
-  aiding future software maintenance. We evaluate our specification
-  language and verification tool against
-  a corpus of numerical Fortran code ($\sim$ 1 million lines) for
-  which we generate $\sim$60,000 specifications, showing the vast majority
-  of stencil computations indeed have a simple, regular, static shape. % We introduce the
-  %language, checking and inference procedures, and its evaluation.
-\end{abstract}
-
-%\keywords{program verification, specification, stencils}
-
-\section{Introduction}\label{sec:intro}
-
-\emph{Stencils} are a ubiquitous programming pattern, common in
-scientific and numerical computing applications. Informally, a stencil
-computation computes an array whose value at each index $i$ is
-calculated from a \emph{neighbourhood} of values around $i$ in some
-input array(s), \eg{}, the Game of Life, convolutions in image
-processing, approximations to differential equations. For example, the
-following computes the one-dimensional discrete Laplace transform (an
-approximation to a derivative) in Fortran:
-%
-\begin{minted}{fortran}
-do i = 1, (n-1)
-      b(i) = a(i-1) - 2*a(i) + a(i+1)
-end do
-\end{minted}
-%
-Values \texttt{b(i)} are calculated from a neighbourhood of elements
-around \texttt{i} in the input array \texttt{a}.  In this example, the
-access pattern is simple and easily understood. More complex
-stencil computations are more prone to errors from simple lexical
-mistakes. For example, the following is a small snippet from a
-Navier-Stokes fluid simulator (based on~\citet{griebel1997numerical})
- in which two arrays are read with
-different data access patterns, across two dimensions.
-%%
-\begin{minted}[firstnumber=20,xleftmargin=2em]{fortran}
-du2dx = ((u(i,j)+u(i+1,j))*(u(i,j)+u(i+1,j))+   &
-  gamma*abs(u(i,j)+u(i+1,j))*(u(i,j)-u(i+1,j))- &
-  (u(i-1,j)+u(i,j))*(u(i-1,j)+u(i,j))-          &
-  gamma*abs(u(i-1,j)+u(i,j))*(u(i-1,j)-u(i,j))) &
-  /(4.0*delx)
-
-duvdy = ((v(i,j)+v(i+1,j))*(u(i,j)+u(i,j+1))+   &
-  gamma*abs(v(i,j)+v(i+1,j))*(u(i,j)-u(i,j+1))- &
-  (v(i,j-1)+v(i+1,j-1))*(u(i,j-1)+u(i,j))-      &
-  gamma*abs(v(i,j-1)+v(i+1,j-1))*(u(i,j-1)-     &
-  u(i,j))) / (4.0*dely)
-
-laplu = (u(i+1,j)-2.0*u(i,j)+u(i-1,j))/delx/delx+ &
-  (u(i,j+1)-2.0*u(i,j)+u(i,j-1))/dely/dely
-
-f(i,j) = u(i,j)+del_t*(laplu/Re-du2dx-duvdy)
-\end{minted}
-%%
-The access pattern is much harder to understand than the
-one-dimensional Laplace.  The miasma of indexing expressions of the
-form \texttt{var(i$\pm$a, j$\pm$b)} is not only hard to read, but is
-prone to simple textual input mistakes, \eg{}, swapping \texttt{-} and
-\texttt{+}, missing an indexing term, or transforming the wrong
-variable \eg{} \texttt{(i+1,j)} instead of \texttt{(i,j+1)}.
-
-In practice, the typical development procedure for complex
-stencil computations involves some ad hoc testing to ensure that no
-transcription mistakes have been made (\eg{}, by visual
-inspections, or comparison against an manufactured or 
-analytical solutions~\cite{farrell2010automated}). Typically such
-testing is discarded once the code is seen to be correct.
-
-This is not the only information that is often discarded. The
-``shape'' of the indexing pattern is usually the result of choices made in the
-numerical-analysis procedure to convert some
-continuous equations into a discrete approximation. Rarely are these
-decisions captured in the source code, yet the shape of access is
-usually uniform and is has a perspicuous and concise description in
-numerical analysis literature \eg{}, ``\emph{centered in space, to a
-  depth of 1}'' referring to indexing terms \texttt{a(i)},
-\texttt{a(i-1)} and \texttt{a(i+1)}.
-
-To support the development of correct stencil computations, we propose
-a simple, abstract specification language for the data access pattern
-of stencils. The shape of the
-Laplace example stencil is specified in our language as:
-%
-\begin{minted}[linenos=false]{fortran}
-!=  stencil centered(depth=1, dim=1) :: a
-\end{minted}
-%
-That is, \texttt{a} is accessed with a symmetrical pattern
-(``centered'') to a depth of one in each direction in its first
-dimension.  The Navier-Stokes example has a shape
-specified as:
-%
-\begin{minted}[linenos=false,xleftmargin=0em,xrightmargin=0.8em,breakindent=1.9em]{fortran}
-!= stencil centered(depth=1,dim=1)*pointed(dim=2)     + centered(depth=1,dim=2)*pointed(dim=1) :: u
-
-!= stencil forward(depth=1,dim=1)                          * backward(depth=1,dim=2) :: v
-\end{minted}
-%
-The specification requires that, over the whole fragment, \texttt{u}
-is accessed with a centered pattern to depth of 1 in both dimensions
-(this is known as the \emph{five-point stencil}) and \texttt{v} is
-accessed in a neighbourhood bounded forwards to depth of $1$ in the
-first dimension and backward to a depth of $1$ in the second
-dimension. The specification is relatively small
-and much more abstract compared to the source code.
-
-We provide a verification tool which can check the spatial correctness
-of stencil computations against a specification. Furthermore, the tool
-can automatically infer specifications and insert these as comments
-into the source code. The main features and benefits of our approach
-are as follows:
-%
-\begin{itemize}
-%\item The main use case is that a programmer writes a specification
-%  first (perhaps informed by their numerical analysis), then writes
-%  the corresponding source code, and lastly has the tool check its
-%  conformance to the specifications. This specify-and-check approach
-%  removes the usually ad hoc testing effort, and provides the
-%  opportunity for easy re-verification in the event of any refactoring
-%  of software maintenance activities.
-\item The primary use case is that a programmer writes a specification
-  first before the stencil code. Our tool then checks conformance between
-  the two. This specify-and-check approach reduces testing effort and
-  future-proofs against bugs introduced during refactoring and maintenance.
-%
-%\item The specification provides a concise description which can be
-%  preserved in the source code to provide documentation to future
-%  developers, contrasting with ad hoc tests which are usually
-%  discarded or specialised to the development environment. Separate from the
-%  verification procedure, the infer-and-synthesise mode of the tool
-%  provides a way to automatically generate additional useful
-%  documentation for future users.
-
-% Emphasis here is on documentation generation, so it makes little sense to
-% compare this aspect of CamFort to ad hoc testing. Apples and oranges.
-\item A specification concisely captures the access pattern of a
-  stencil and turns it into documentation. The inference procedure
-  efficiently produces specifications with no programmer intervention,
-  automatically inserting specifications at appropriate places in the
-  source code.
-%
-%\item Specifications are integrated into the code as comments for easy
-%  integration with existing tools, as well as documentation systems (\eg{}, we
-%  support Doxygen and FORD syntax).
-%
-\item Our specification format is deliberately abstract, with a small number of
-  combinators that \emph{do not involve any indexing expressions}, \eg{}
-  \texttt{a(i+1,j-1)}. This contrasts with other specification
-  approaches, \eg{}, deductive verification tools such as ACSL, where
-  specifications must also be in terms of array-indexing expressions. Therefore, any low-level mistakes
-  that could be made whilst programming complex indexing code could also be
-  made when writing its specification. Our specifications are more
-  abstract and lightweight
-  with the aim of aiding adoption by scientists and preventing indexing errors.
-\end{itemize}
-
-Our verification tool does not target a
-class of bugs that can be detected automatically (\emph{push-button
-  verification}). Instead, stencil computation bugs must be
-identified relative to a specification of the intended access pattern.
-
-In this paper, we make the following contributions:
-%
-\begin{itemize}
-\item We introduce a specification language for stencils
-  that captures common forms of data access patterns
-  (\cref{sec:lang});
-
-\item We detail checking and inference algorithms for our
-  specifications (\cref{sec:analysis}), derived from a
-  denotational model for the specification language
-  (\cref{sec:semantics});
-
-\item We provide an implementation of our approach as an extension to
-  CamFort~\cite{camfort}, an open-source program analysis tool for Fortran;
-
-\item We report on a quantitative study of stencil computations on a
-  corpus of numerical Fortran program
-  (\cref{sec:evaluation}), totalling one million lines of code. Our tool
-  identifies and infers specifications for roughly 60,000 stencil computations in the
-  corpus. Approximately 5,000 of the stencils we found are non-trivial,
-  corresponding to code which is a possible source of errors. This validates
-  our hypothesis that the majority of stencil computations have a regular
-  shape, and validates the design of our language in its capability to capture
-  many core patterns.
-
-\item We give a verification case study of a
-  stencil computation commonly used in scientific computing,
-  simulating programming errors (\Cref{sec:jacobi}). Our approach detects all
-  possible simulated programming errors for this problem.
-\end{itemize}
-%
-
-\newcommand{\domainVal}{\mathbb{Z}_\infty}
-\section{Stencil specifications}\label{sec:lang}
-
-Our specification language is based on the hypothesis
-that most forms of array access in numerical code have
-a fixed, statically-determined access pattern. For example, the
-\emph{five-point stencil} on a two-dimensional array reads from array
-indices $(i, j)$, $(i-1, j)$, $(i+1, j)$, $(i, j-1)$, and $(i, j+1)$
-for all $i, j$ within the inner boundary of the array (to avoid
-out-of-bounds access at the edges). We revisit this hypothesis
-in \cref{sec:evaluation} with the inference of
-such regular stencil patterns on our corpus of numerical programs.
-
-We give various definitions used in the paper then
-\cref{subsec:syntax} outlines the syntax of stencil specifications
-with an informal introduction. \cref{sec:semantics}
-gives a semantic model via a set-based interpretation which makes
-concrete the meaning of specifications. \cref{sec:eqs} then provides
-a syntactic equational theory for specifications which is
-sound with respect to the model.
-%sub-specification for which the
-%equality and sub-specification relation are proven sound.
-
-\paragraph{Notation}
-%\label{sec:notation}
-
-\renewcommand*{\arraystretch}{0.8}
-For the target language, $e$ ranges over expressions (may be impure) and $v$
-over its (imperative) variables. %By
-%``variable'', we mean in the imperative sense (named binders to memory cells).
-
-\begin{definition}[Induction variable]
-  An integer variable is an \emph{induction variable} if it is the control
-  variable of a ``\texttt{for}'' loop incremented by $1$ per iteration. A variable
-  is interpreted as an induction variable only within the scope of the loop
-  body. Throughout, $i, j, k$ range over induction variables.
-%TODO: consider mentioning derived, non-based induction variables
-% as we did before.
-\end{definition}
-
-\begin{definition}[Array subscripts and indices]
-  An \emph{array subscript}, denoted $a(\bar{e})$, is an expression
-  which reads from an array $a$ at an \emph{index} specified by a
-  comma-separated sequence of integer expressions denoted $\bar{e}$ or
-  in expanded form as $(e_1, \ldots, e_n)$. An index $e_i$
-  is called \emph{relative} if the expression involves an induction
-  variable.
-  An \emph{absolute index} is a constant integer expression relative to the
-  enclosing loop.
-\end{definition}
-
-\begin{definition}[Neighbourhood index]
-\label{def:neighbour}
-  For an array subscript $a(\bar{e})$ an index $e \in \bar{e}$ is a
-  \emph{neighbourhood index} if $e$ is of the form $e \equiv i$, $e \equiv i +
-  c$, or $e \equiv i - c$, where $c$ is an integer constant. That is, a
-  neighbourhood index is a constant translation of an induction variable. (The
-  relation $\equiv$ identifies terms up-to commutativity of $+$ and the inverse
-  relationship of $+$ and $-$ \eg{}, $(-b) + i \equiv i - b$).
-\end{definition}
-
-\begin{definition}[Stencil computations]
-\label{def:stencil}
-  Let $a$ be an array of dimensionality $n$ and $\overline{b}$ a
-  collection of arrays of arbitrary (possibly differing)
-  dimensionalities.
-  A \emph{stencil computation} comprises an iteration by a set $I$ of at
-  most $n$ induction variables
-  over a subset of the index space of $a$. Each iteration
-  determines the elements of $a$ by an
-  assignment $a(\bar{e}) = e_r$ where each $e \in
-  \bar{e}$ is either a neighbourhood or constant index.
-  For all $b \in \bar{b}$ and array subscripts $b(\bar{e}')$
-  that flow to $e_r$, each $e' \in \bar{e}'$ is
-  either a neighbourhood or absolute index.
-
-  We refer to the assignment $a(\bar{e}) = e_r$ and all the associated
-  statements with dataflow to $e_r$ as the \emph{stencil kernel}.
-\end{definition}
-%
-% specifying the shape of indexing expressions for
-%arrays $\bar{v}$ that are simultaneously permitted and expected within
-%the code to which the specification is given.
-%
-%\marginpar{\dnote{Is it worth having this informal consistency here?}}
-
-For now we informally describe what it means for a stencil kernel to
-be consistent with a specification. This will be formalised in
-\Cref{sec:semantics} and algorithmically described in and
-\Cref{sec:analysis}.  Specifications are associated to a stencil
-kernel and are declared for one or more array variables $\bar{v}$
-within that kernel.  A stencil specification defines a \emph{region}
-associated to an array variable $a$ against which a stencil kernel
-should be \emph{consistent}. The region defined by a specification is
-an $n$-dimensional rectilinear shape in a discrete space crossing the
-origin with respect to the induction variable. Consistency is then a
-two way requirement. All array subscripts on $a$ contributing to the
-kernel should be consistent with the region. Conversely, all parts of
-the region should have an index corresponding it.
-
-\begin{figure}[t]
-\vspace{-0.9em}
-\begin{align*}
-\def\arraystretch{1}
-\setlength{\arraycolsep}{0.2em}
-\newcommand{\dimTy}{\mathbb{N}_{>0}}
-\begin{array}{rl}
-\nonterm{specification} ::= & \nonterm{regionDec} \mid \nonterm{specDec} \\
-\nonterm{specDec} ::= & \term{stencil} \; \nonterm{spec} \;
-                        \texttt{::} \; v \\
-\nonterm{regionDec} ::= &  \texttt{region} \; \texttt{::} \; \nonterm{rvar} \; \texttt{=} \;
-                         \nonterm{region}\\[0.4em]
-%\nonterm{spec} ::= & \nonterm{spatial} \mid \nonterm{temporal}
-%\\[1em]
-\nonterm{spec} ::= & [\nonterm{mult},] \; [\nonterm{approx},] \; \nonterm{region} \\
-\nonterm{mult} ::= & \term{readOnce} \\
-\nonterm{approx} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
-\nonterm{region} ::= & \nonterm{rvar} \; \\
-\multicolumn{2}{l}{\qquad \mid \; \stenRefl{\dimTy}} \\
-\multicolumn{2}{l}{\qquad \mid \;\stenFwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{nonpointed}]}} \\
-\multicolumn{2}{l}{\qquad \mid \; \stenBwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{nonpointed}]}} \\
-\multicolumn{2}{l}{\qquad \mid \; \stenCen{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{nonpointed}]}} \\
-\multicolumn{2}{l}{\qquad \mid \; \nonterm{region} \, \term{+}
-  \, \nonterm{region} \; \mid \; \nonterm{region} \; \term{*} \; \nonterm{region}} \\[0.25em]
-%\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{rvar}}
-%\\[0.5em]
-%\nonterm{temporal} ::= \; & \term{dependency} \; (v \; \{ , v \}) [, \texttt{mutual}]
-%  \\[0.5em]
-\nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+\\[-1em]
-\end{array}
-\end{align*}
-\caption{Specification syntax (EBNF grammar)}
-\label{fig:syntax}
-\vspace{-0.8em}
-\end{figure}
-
-\subsection{Specification syntax}
-\label{subsec:syntax}
-
-\Cref{fig:syntax} gives the syntax of stencil specifications, which is
-detailed below. The entry point is the \nonterm{specification} production which
-splits into either a \emph{region declaration} or a \emph{specification
-declaration}.
-Regions comprise \emph{region constants} which are
-combined via region operators \term{+} and \term{*}.
-
-%\paragraph{Region constants}
-
-Region constants specify a finite contiguous region in a
-single dimension relative to the origin and are either \term{pointed},
-\term{forward}, \term{backward}, or \term{centered}. The region names are
-inspired by the scientific terminology, \eg{} the standard explicit method for
-approximating PDEs is known as the \emph{Forward Time, Centered Space} (FTCS)
-scheme~\citep{dawson1991finite}.
-
-Each region
-constant has a dimension identifier $d$ given as a positive natural number.
-Each constant except \term{pointed} has a depth
-parameter $n$ given as a positive natural number; \term{pointed}
-regions implicitly have a depth of $0$.
-
-A \term{forward} region of depth $n$ specifies a contiguous
-region in dimension $d$ starting at the origin. This corresponds
-to specifying neighbourhood indices in dimension $d$ ranging from $i$ to $i + n$
-for some induction variable $i$. Similarly, a
-\term{backward} region of depth $n$ corresponds to contiguous indices
-from $i$ to $i - n$ and \term{centered}
-of depth $n$ from $i - n$ to $i + n$. A \term{pointed}
-stencil specifies a neighbour index $i$. For example, the
-following shows four specifications with four consistent stencil
-kernels reading from arrays \term{a}, \term{b}, \term{c} and \term{d}:
-%%
-\begin{minted}{fortran}
-!= stencil forward(depth=2, dim=1) :: a
-e(i, 0) = a(i, 0) + a(i+1, 0) + a(i+2, 0)
-!= stencil backward(depth=2, dim=1) :: b
-e(i, j) = b(i, j) + b(i-1, j) + b(i-2, j)
-!= stencil centered(depth=1, dim=2) :: c
-e(i, j) = (c(j-1) + c(j) + c(j+1))/3.0
-!= stencil pointed(dim=3) :: d
-e(i, j) = d(0, 0, i)
-\end{minted}
-%%
-Not every dimension needs to be specified, \eg{},
-specifications on lines $1$, $3$, and $7$ leave some dimensions unspecified.
-
-The \term{forward}, \term{backward}, and \term{centered} regions may
-all have an additional attribute \term{nonpointed} which marks absence
-of the origin.  For example, the following is a
-\term{nonpointed} \term{backward} stencil
-%
-\begin{minted}{fortran}
-!= stencil backward(depth=2, dim=1, nonpointed) :: a
-b(i) = a(i-1) + 10*a(i-2)
-\end{minted}
-
-\paragraph{Combining regions}
-
-% r+s means that indices must be consistent with either r or s
-% r*s means that indices must be consistent with both r and s at the
-% same time
-
-% specifications correspond to a set of schemes
-
-% r+s means an index must be generalised by a scheme in either r or s
-%       and every scheme in r must have an index which specialises it
-%       and every scheme in s must have an index which specialises it
-
-% r*s means an index must be generalised by a scheme in r and a scheme in s
-%       and every pairs of schemes in r and s must have an index which
-%           specialises both
-%
-
-The region operators \term{+} and \term{*} respectively combine
-regions disjunctively and conjunctively.
-
-The conjunction of two regions $r \term{*} s$ means that any indices
-in the specified code must be consistent with both $r$ and $s$
-simultaneously. Dually, for every pair of components in the regions
-$r$ and $s$ there must be an index consistent with both.
-For example, the following \emph{nine-point stencil}
-has a specification given by the product of two \texttt{centered}
-regions in each dimension:
-%%
-\begin{minted}[breakindent=2.9em]{fortran}
-x = a(i, j)   + a(i-1, j)   + a(i+1, j)
-y = a(i, j-1) + a(i-1, j-1) + a(i+1, j-1)
-z = a(i, j+1) + a(i-1, j+1) + a(i+1, j+1)
-!= stencil centered(depth=1, dim=1) * centered(depth=1, dim=2) :: a
-b(i, j) = (x + y + z) / 9.0
-\end{minted}
-%%
-This pattern is common in image convolution applications.
-The specification ranges over the
-values that flow to the array subscript on the left-hand side,
-and so ranges over the intermediate assignments to \term{x},
-\term{y}, and \term{z}. Each index in the code is consistent
-with both specifications simultaneously, \eg{}, \texttt{a(i-1, j+1)}
-is within the centered region in dimension $1$ and the centered region
-in dimension $2$.
-
-The product $r \term{*} s$ can also be thought of as a bounding box
-over the two regions $r$ and $s$.
-
-The disjunction of two regions $r \term{+} s$ means that any indices
-in the specified code must be consistent with either of $r$ or $s$.
-Dually, every part of region $r$ must have a
-corresponding index and similarly for $s$ independently. For
-example, the following gives the specification of a five-point
-stencil which is the sum of two compound \texttt{pointed} and
-\texttt{centered} regions in each dimension:
-%
-\begin{minted}{fortran}
-!= stencil centered(depth=1, dim=1)*pointed(dim=2)    + centered(depth=1, dim=2)*pointed(dim=1) :: a
-b(i,j) = -4*a(i,j) + a(i-1,j) + a(i+1,j) &
-                   + a(i,j-1) + a(i,j+1)
-\end{minted}
-%%
-Here the left-hand side of \texttt{+} says that when the second dimension
-(induction variable $j$) is fixed at the origin, the first dimension
-(induction variable $i$) accesses the immediate vicinity of the origin
-(to depth of one). The right hand side of \texttt{+} is similar but the dimensions are reversed.
-This reflects the symmetry under rotation of the five-point stencil.
-
-\paragraph{Region declarations and variables}
-
-Region specifications can be assigned to region variables via
-region declarations. For example, the shape of a
-``\emph{Roberts cross}'' edge-detection convolution~\cite{davis1975survey}
-can be stated:
-%%
-\begin{minted}{fortran}
-!= region :: r1 = forward(depth=1, dim=1)
-!= region :: r2 = forward(depth=1, dim=2)
-!= region :: robertsCross = r1*r2
-!= stencil robertsCross :: a
-\end{minted}
-This is useful for common stencil patterns, such as the five-point
-pattern, as the region can be defined once and reused.
-%%
-\paragraph{Modifiers}
-%%
-Region specifications can be modified
-by \emph{approximation} and \emph{multiplicity} information
-(in \textit{spec} in \Cref{fig:syntax}).
-
-The \texttt{readOnce} modifier enforces that no index appears more
-than once (that is, its multiplicity is one). For example, all of
-the previous examples could have \texttt{readOnce} added:
-%
-\begin{minted}{fortran}
-!= stencil readOnce, backward(depth=2, dim=1) :: a
-b(i+1) = a(i) + a(i-1) + a(i-2)
-\end{minted}
-%
-This specification would be invalid if any of the
-array subscripts were repeated. This modifier provides a way to
-rule out any accidental repetition of array subscripts.
-The notion is similar to that of linear types~\cite{wadler1990linear}, where a value must be used
-exactly once. We opt for the more informative and easily understood name
-\texttt{readOnce}. This modifier is optional, so it need not
-be present even if the stencil is linear.
-
-In some cases, it is useful to give a lower and/or upper bound for a
-stencil. This can be done using either the \term{atMost} or
-\term{atLeast} modifiers. This is particularly useful in situations
-where there is a non-contiguous stencil pattern, which cannot be expressed
-precisely in our syntax. For example:
-%
-\begin{minted}{fortran}
-!= stencil atLeast, pointed(dim=1)         :: a
-!= stencil atMost, forward(depth=4, dim=1) :: a
-b(i) = a(i) + a(i+4)
-\end{minted}
-
-\section{A model for specifications}
-\label{sec:semantics}
-
-We define a simple set-theoretic denotational model of the semantics
-of our specification language. This has a number of purposes:
-(1) it serves to explain the meaning of our specifications
-(2) it is used in the inference and checking algorithms
-(\Cref{sec:analysis}); (3) justifies an equational theory for
-specifications in the next section; and (4) can be used to guide
-correct implementations. The model domain is over sets of vectors
-which we call \emph{index schemes}.
-
-\begin{definition}
-  An \emph{index scheme} is a vector of size $n$ (an $n$-vector) with
-  values drawn from $\domainVal = \mathbb{Z} \cup \{\infty\}$.  Values
-  in $\mathbb{Z}$ represent the offsets of neighbourhood indices
-  (\ie{}, the constant integer $c$ in an indexing expression of the
-  form $i \pm c$, \Cref{def:neighbour}) and $\infty$ is a wild-card
-  representing both constant indices and unspecified indices.
-  For an index scheme $u$, the $i^{th}$ element is denoted $u_i$. We
-  write $(\domainVal)^n$ for the set of all $n$-dimensional schemes.
-\end{definition}
-
-Indexing schemes abstract array subscripts.
-Our model, consisting index schemes, explain the range of possible
-indexing behaviours which are consistent with the specification.
-The relationship between source language
-and index schemes
- is defined by \textsf{schematic} function,
- which forms part of checking and inference (\Cref{sec:analysis}).
-
-\newcommand{\schemeOne}{\textsf{s}_I}
-%\newcommand{\schemeOne}{\textsf{schema$^1_I$}}
-\newcommand{\schemeN}{\textsf{schematic}_I}
-%\newcommand{\schemeN}{\textsf{schema$^n_I$}}
-\begin{definition}%[Schematic]
-\label{def:schematic}
-Let $\schemeOne$ be a partial function, parameterised by a set of
-induction variables $I$, mapping source language expressions to
-$\domainVal$ values:
-%%
-\begin{align*}
-\schemeOne(e)
- & =  \begin{cases}
-c & e \equiv i + c \;\; \wedge \;\; i \in I \\
--c & e \equiv i - c \;\; \wedge \;\; i \in I \\
-\infty & \textit{$I \cap \textit{FreeVariables}(e) = \emptyset $} %\\
-%\bot   & \textit{otherwise}
-\end{cases}
-\end{align*}
-In the
-first two cases, expressions that are a constant offset of an
-induction variable (\ie{}, neighbourhood indices) are mapped to the
-offset $c$ or $-c$ $\in \domainVal$. In the third case, if the
-expression is not defined in terms of an induction variable, it
-is mapped to $\infty \in \domainVal{}$. Otherwise, $\schemeOne(e)$
-is undefined ($\bot$).
-
-The partial function $\schemeN$, also parameterised by a set of
-induction variables $I$, uses $\schemeOne$ to
- map from the syntax of indices to index schemes
-$\in (\domainVal)^n$. It is defined on the syntax of indices $\bar{e} = (e_1, ..., e_n)$ as:
-\begin{align*}
-\schemeN(e_1, .., e_n) & =
-[\schemeOne(e_1) ... \schemeOne(e_n) ]
-\; \textit{iff} \; \forall x . \schemeOne(e_x) \neq \bot
-\end{align*}
-%
-Thus, if each expression in an index is a neighbourhood or constant
-index, $\schemeN$ maps it to an indexing scheme.
-\end{definition}
-
-\begin{definition}[Generalisation]
-\label{def:generalisation}
-Given two schemes $v, w$ of dimensionality $n$,
-then $w$ is said to \emph{generalise} $v$
-(or conversely $v$ \emph{specialises} $w$),
-written $v \preceq_n w$ defined:
-\[
-v \preceq_n w \;\; \Leftrightarrow  \;\;
-  \forall i\!\in\!\{ 1, \ldots, n \} . \, (w_i = \infty \, \vee \, v_i = w_i)
-\]
-%%
-where $\preceq_n$ is a reflexive, transitive relation on
-index schemes $\domainVal^n$. Thus,
- $\infty$ generalises any index in the same dimension.
-%
-\end{definition}
-
-\begin{example}
-Consider the following array subscript to indexing schemes
-conversions:
-%
-\begin{align*}
-\begin{array}{ll}
-\schemeN{(i, j+1)} & =\vtwoh{0}{1} \\
-\schemeN{(1, j+1)} & = \vtwoh{\infty}{1} \\
-\schemeN{(i*2, j)} & = \bot
-\end{array}
-\end{align*}
-There is one generalisation between these
-schemes, namely that $\vtwoh{0}{1} \preceq_2 \vtwoh{\infty}{1}$.
-The last example is undefined as its first expression is a
-non-neighbour relative index.
-\end{example}
-
-%Our terminology of \emph{index schemes} and
-%\emph{generalisation} is by analogy
-%with the \emph{type schemes} and \emph{type generalisation} of
-%polymorphic types in ML~\cite{milner1978theory}. Our notion is
-%much more rigid however, as index schemes do not bind names.
-
-\paragraph{Shorthand notation}
-
-We use abbreviations
- \term{fwd}, \term{bwd}, \term{cen}, and \irreflS{}
-for \term{forward}, \term{backward},
-\term{centered}, and \irrefl{} respectively. Furthermore, 
-we quantify over the absence or presence of the \irrefl{}
-attribute on region constants as a boolean. For example,
-\term{fwd} is syntactic sugar defined:
-%%
-\begin{align*}
-\hspace{-0.5em}
-& \stenFwdSR{n}{d}{p}
-\; := \; \\[-0.4em]
-& \quad \begin{cases}
-\stenFwd{n}{d}{, \irrefl} & \textit{if} \;\, \neg \, p \\[-0.1em]
-\stenFwd{n}{d}{}  & \textit{if} \;\,  p
-\end{cases}
-\end{align*}
-%
-$\textit{or} \;\; \stenFwdS{n}{d} := \stenFwd{n}{d}{}$
-and similarly for \term{bwd} and \term{cen}.
-
-\subsection{Denotational model}
-
-An interpretation function $\interp{-}_n$ provides the model 
-(where $n$ is the maximum dimensionality of the specification being
-modelled) mapping closed\footnote{That is, we assume there are no
-  occurrences of \textit{rvar} in a specification being modelled.  Any
-  \emph{open} specification containing region variables can be made
-  closed by straightforward syntactic substitution with a (closed)
-  \textit{region}.} specifications to sets of
-$n$-dimensional indexing schemes. The interpretation is overloaded on
-\emph{regions} in \Cref{fig:region-model} and on the top-level
-of a specification \textit{spec} in \Cref{fig:spatial-model}.
-Various intermediate notions are used.
-
-\begin{definition}A \emph{single-entry vector} of size $n$, denoted
-$\singleEntry{r}{n}$, is a vector where the $r^{th}$ entry is $1$
-and all others are $\infty$, \eg{}, $\singleEntry{1}{2} =
-\vtwoh{1}{\infty}$ and $\singleEntry{2}{3} = \vthreeh{\infty}{1}{\infty}$.
-Thus, a single-entry vector is an index scheme generalising 
-all other schemes that have $1$ as their $r^{th}$ entry,
-describing a neighbour offset of $1$.
-\end{definition}
-
-\begin{definition}
-A \emph{zeroed single-entry vector} of size $n$ denoted
-$\zeroEntry{r}{n}$ is a vector where the $r^{th}$ entry is $0$ and all others
-are $\infty$ \eg{}, $\zeroEntry{1}{2} = \vtwoh{0}{\infty}$.
-Akin to single-entry vectors, zeroed single-entry vectors
-generalise all other schemes
-that have $0$ as their $r^{th}$ entry, representing indices at
-the ``origin'' in $r$.
-\end{definition}
-
-\begin{figure}[t]
-\vspace{-0.9em}
-\input{region-model.tex}
-%\textit{$a^m \in A$ means there are $n$ copies of $a$ in
-%  the multi-set $A$}
-\caption{Model of regions,
-$\interp{-}_n : \textit{region} \rightarrow \mathcal{P}(\domainVal{}^n)$}
-\label{fig:region-model}
-\vspace{-1em}
-\end{figure}
-
-\noindent
-The first four equations of \Cref{fig:region-model}
-model region constants as sets of indexing
-schemes. We illustrate with an example.
-
-\begin{example}
-For a $2$-dimensional stencil computation, then
-\begin{align*}
-& \interp{\stenCenS{2}{1}}_2 \\[-0.1em]
-& \qquad = \; \{i\singleEntry{1}{2} \,|\, i \in \{\text{-}2,..., 2\}\!\setminus\!0 \}
-  \cup  \{ \zeroEntry{1}{2} \,|\, \textit{true} \} \\[-0.2em]
-& \qquad = \; \{-2\singleEntry{1}{2}, \, -\singleEntry{1}{2}, \,
-  \singleEntry{1}{2}, \, 2\singleEntry{1}{2}\} \cup
-  \{\zeroEntry{1}{2}\} \\[-0.2em]
-& \qquad = \; \{\vtwoh{-2}{\infty}, \vtwoh{-1}{\infty}, \vtwoh{0}{\infty},
-\vtwoh{1}{\infty}, \vtwoh{2}{\infty}\}
-\end{align*}
-Note, $\infty$ is absorbing for multiplication above \eg{} $2
- \infty = \infty$.
-\end{example}
-\noindent
-The final two equations of \Cref{fig:region-model} give the models
-of $\term{+}$ and $\term{*}$. For $\term{+}$ this is the
- union of the models of subterms. The model of
-$\term{*}$ takes a ``tensor'' $\otimes$ of subterm models. The tensor is
-more involved. We give some informal intuition first.
-
-% specifications correspond to a set of schemes
-
-% r+s means an index must be generalised by a scheme in either r or s
-%       and every scheme in r must have an index which specialises it
-%       and every scheme in s must have an index which specialises it
-
-% r*s means an index must be generalised by a scheme in r and a scheme in s
-%       and every pairs of schemes in r and s must have an index which
-%           specialises both
-%
-
-% r+s means indices must be consistent with either r or s
-%           and there are indices why are generalised by r and s
-% (there are indices which are consistent with every part of r or s)
-% (every part of r and s independently have an index which is generalised by it)
-
-% r*s means indices must be consistent simultaneously with both r and s
-%           and there are indices which are generalised
-%
-
-\begin{figure}
-\vspace{-1em}
-\input{spatial-model.tex}
-\caption{%\textit{spec} model,
-$\interp{-}_n : \textit{spec} \rightarrow \textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$}
-\label{fig:spatial-model}
-\vspace{-1em}
-\end{figure}
-
-\paragraph{$\otimes$ of models}
-
-The intuition behind $\otimes$ is that it takes all possible pairs of
-index schemes, and treats each pair as lower and upper bounds of an
-$n$-dimensional rectangle. From these bounds,
-the remaining vertices of the $n$-dimensional rectangle are
-generated (like a bounding box) \eg{}:
-%%
-\begin{align*}
-{\small{
-\{\vtwohs{1}{2},\vtwohs{5}{6}\,\}\!\otimes\!\{\vtwohs{3}{4}\}\!=\!\{\vtwohs{1}{2},\vtwohs{3}{2},\vtwohs{1}{4},\vtwohs{3}{4},
-   \vtwohs{3}{6},\vtwohs{5}{4},\vtwohs{5}{6}\,\}
-}}
-\end{align*}
-%%
-The tensor takes a Cartesian product of the two models,
-giving all pairs of schemes which are combined using the
-\emph{pairwise permutation} written $\bowtie$ (defined later).
-Pairwise permutation generates all possible interleavings of two
-vectors. Some care is then taken over how this interacts with $\infty$.
-In the presence of $\infty$, generate all interleavings as if $\infty$ is
-an integer, then filter out vectors with $\infty$ at any coordinate, \eg{}
-\begin{align*}
-\{\vtwoh{1}{\infty}\} \otimes \{\vtwoh{3}{4}\} =
-\{\vtwoh{1}{4}, \vtwoh{3}{4}\}
-\end{align*}
-%
-where $\vtwoh{1}{\infty}$ and $\vtwoh{3}{\infty}$ have been filtered.
-
-\begin{definition}%[Constrained dimensions]
-For an $n$-dimensional model $M$, its \emph{constrained dimensions}
-are the set of dimension identifiers where at least one scheme $u \in M$ has
-a non-$\infty$ value in that dimension:
-%
-\begin{align*}
-\effdims{M}{n} =
-\bigcup_{u \in M} \{i \mid i \in \{1,\ldots,n\} \, \wedge \, u_i \neq
-  \infty\}
-\end{align*}
-For example, $\effdims{\{\vtwoh{0}{\infty},\vtwoh{\infty}{\infty}\}}{2}
- = \{1\}$.
-\end{definition}
-\begin{definition}The tensor $\otimes_n$ of $n$-dimensional
-models is:
-\begin{alignat*}{2}
-& M \otimes_n N =
-  \Bigg\{x \; \Bigg| \;
-    \parbox{5.2cm}{$(u, v) \in M \times N,$ \\[0.1em]
-                  $\;\; i \in \{1, \ldots, 2^n\}, \,
-                  x = (u \bowtie v)_i,$ \\[0.1em]
-                  $\;\; j \in (\effdims{M}{n} \cup
-                   \effdims{N}{n}),$\\[0.1em]
-                   $\;\; x_j \neq \infty$
-                  } \Bigg\}
-\end{alignat*}
-The second guard binds $j$ to the constrained
-dimensions for $M$ unioned with the constrained dimensions for $N$.
-That is, every $j$ corresponds to a dimension in which
-either $M$ or $N$ has a scheme $u$ where $u_j \neq \infty$,
-and thus is constrained.
-The \emph{pairwise permutation} $u \bowtie v$
-builds a $2^n \times n$ matrix
-of all possible $n$-vectors generated by
-non-deterministically picking for each $j^{th}$ entry either
-$u_j$ or $v_j$ \eg{}
-%
 \begin{equation*}
-\vthreeh{0}{1}{2} \bowtie \vthreeh{3}{4}{5} =
-\left[
-\setlength{\arraycolsep}{0.42em}
-\begin{array}{cccccccc}
-0 & 0 & 0 & 0 & 3 & 3 & 3 & 3 \\
-1 & 1 & 4 & 4 & 1 & 1 & 4 & 4 \\
-2 & 5 & 2 & 5 & 2 & 5 & 2 & 5
-\end{array}
-\right]
+\texttt{a} \mapsto \{[0,0]\times[0,0], \;\; [-1,0]\times[0,0], \;\;
+[0,0]\times[-1,0], \;\; [0,0]\times[1,0], \;\; [0,1]\times[0,0]\}
 \end{equation*}
 %
-The $2^n$ unique choices for $\bowtie$ on $n$-vectors
-correspond to taking all bit-strings of length $n$ and
-selecting from $u$ for $1$ and $v$ for $0$. The pairwise permutation
-matrix is thus defined in terms of the logical
-matrix $b$ where $b_{i,j}$ is the $j$-th bit of the integer
-$i - 1$ as:
-%
-\begin{align*}
-(u \bowtie v)_{i,j} & = b_{i,j} u_j + \neg b_{i,j} v_j
-\end{align*}
-%
-The permuted schemes $x = (u \bowtie v)_i$ added to the model
-$M \otimes N$ are only those without $\infty$ in
-the constrained dimensions given by each model ($x_j \neq
-\infty$). This eliminates all permutations with
-$\infty$ in a constrained dimension position.
-\end{definition}
-
-\paragraph{Model of spec}
-
-Regions are modelled as sets of index
-schemes, that is
-$\interp{\textit{region}}_n \in \mathcal{P}((\domainVal{})^n)$.
-\Cref{fig:spatial-model} defines the top-level model of the
-\textit{spec} syntax, which includes the multiplicity and
-approximation modifiers. The domain of this interpretation
-is $\textsf{Mult}(\textsf{Approx}(\mathcal{P}((\domainVal{})^n)))$
-where $\textsf{Mult}$ and $\textsf{Approx}$ are labelled variants
-labelling our set-based model with additional information.
-
-\begin{definition} \textsf{Mult}
-and \textsf{Approx} are parametric labelled variant types
-with injections given by their definition:
-%
-\begin{align*}
-\textsf{Mult} \, a & = \textsf{mult} \, a \,\mid\, \textsf{only} \, a \\
-\textsf{Approx} \, a & = \textsf{exact} \, a \,\mid\, \textsf{lower} \, a \,\mid\,
-\textsf{upper} \, a \,\mid\, \textsf{both} \, a \, a
-\end{align*}
-\eg{}, $\mathsf{lower}$ is an injection
-$\mathsf{lower} : a \rightarrow \mathsf{Approx} \, a$ etc. The
-$\textsf{Mult}$ type corresponds to the presence or absence of the
-\term{readOnce} modifier as shown in \Cref{fig:spatial-model}.
-The $\textsf{Approx}$ type corresponds to the presence or absence of
-the spatial approximation modifier, with \textsf{exact} when there is
-no such modifier and \textsf{lower} and \textsf{upper} for
-\term{atLeast} and \term{atMost}. Later, it will be useful to model
-pairs of lower and upper bounds provided by \textsf{both} but which is
-not used
-here.% These are used for labelled models, and other components later in.
-\label{def:mult-and-approx}
-\end{definition}
-
-\section{Equational theory}
-\label{sec:eqs}
-
-Specifications have a theory of equality, $\equiv$, providing axiomatic
-semantics for the language. This enables users to rewrite their specification
-and provides a way to automatically simplify
-specifications after inference (\Cref{subsec:inference}).
-Additionally, specifications have a theory of approximation,
-$<:$, that can be used by the users to weaken specifications for succinctness in
-the source.
-
-The equality relation $\equiv$ is defined over \emph{region}
-syntax which gives the following axioms:
-%
-\begin{itemize}
-\item $\term{+}$ is idempotent, commutative, and associative;
-\item $\term{*}$ is commutative, associative, and distributes over
-$\term{+}$, \ie{},  $R \; \texttt{*} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
-                       \texttt{*} \; S) \; \texttt{+} \; (R
-                       \; \texttt{*} \; T)$.
-Distributivity is
-used later to give a normal form for specifications
-akin to Disjunctive Normal Form (\cref{sec:inf-step4}).
-
-\item (\emph{overlapping}) Given two region constants $R_1$ and $R_2$
-on the same dimension, then some combinations of region constants
-can be coalesced into a single region constant $R_3$ if $R_1$ completely
-overlaps $R_2$. There are five possible instantiations of the
-following rule $R_1 + R_2 \equiv R_3$:
-%
-\newcommand{\stenGR}[4]{\small{\textit{#4}(\term{depth=}#1,
-  \term{dim=}#2, #3)}}
-\newcommand{\stenG}[3]{\small{\textit{#3}(\term{depth=}#1,
-  \term{dim=}#2)}}
-%
-\vspace{-0.2em}
-\begin{align*}
-\hspace{-1em} & \stenGR{n}{d}{p \vee q}{$R_3$} \\[-0.4em]
-\hspace{-1em} \equiv \; & \stenGR{n}{d}{p}{$R_1$} \; \texttt{+} \; \stenGR{m}{d}{q}{$R_2$}
-\end{align*}
-where $n \geq m$ and $p$ and $q$ range
-over booleans representing the absence (true) or presence (false)
-of \irrefl{} which are combined via disjunction.
-We denote the possibilities as $\term{fwd+fwd} \equiv \term{fwd}$, $\term{bwd+bwd} \equiv \term{bwd}$,
-$\term{cen+cen} \equiv \term{cen}$, $\term{cen+fwd} \equiv \term{cen}$ and
-$\term{cen+bwd} \equiv \term{cen}$. %Commutative of $\term{+}$
-%provides symmetric variants.
-
-\item (\textit{overlapping pointed}) \term{pointed} regions coalesce
-  with other regions of the same dimension:
-%
-\vspace{-0.4em}
-\begin{align*}
-& \stenG{n}{d}{region} \\[-0.4em]
-\equiv \; & \stenGR{n}{d}{p}{region} \; \texttt{+} \;
-                          \stenReflS{d}
-\end{align*}
-\vspace{-1.5em}
-
-\item \textit{(centered)} a forward and backward region of the same
-  depth and dimension coalesces into a centered region:
-%
-\vspace{-0.4em}
-\begin{align*}
-\hspace{-1em} & {\small{\stenCen{n}{d}{, p \vee q}}} \\[-0.4em]
-\hspace{-1em} \equiv \; & {\small{\stenFwdSR{n}{d}{p} \; \texttt{+} \;
-                          \stenBwdSR{n}{d}{q}}}
-\end{align*}
-\vspace{-1.5em}
-
-\item $(\textsc{point}) \;\; (R \, \term{*} \, S^\irreflS) \, \term{+} \,
-                         (R^\irreflS \, \term{*} \, S) \; \equiv \; R
-                         \, \term{*} \, S$
-meaning that for the sum of products of two regions $R$ and $S$,
-where in each component one of $R$ or $S$ has the non-pointed
-attribute denoted by $R^{\term{np}}$ and $S^{\term{np}}$,
-the non-pointed attributes cancel to give $R \, \term{*} \, S$.
-%
-\item $(\textsc{$\ast$Dim}) \;\; R^d \, \term{$\ast$} \, S^d \equiv R^d
-                             \, \term{+} \, S^d$
-%
-where $R^d$ and $S^d$ denote regions whose dimension is $d$.
-Thus a product of two regions on
-the same dimension is equal to their addition.
-\end{itemize}
-%
-Stencil specifications in \textit{spec} are considered
-equal when they have the same modifiers and $\equiv$ equates
-their regions.
-
-\begin{theorem}[Soundness of equational theory]
-\begin{equation*}
-\forall S, T, n . \quad
-S \equiv T \; \Rightarrow \;
-\interp{S}_n = \interp{T}_n
-\end{equation*}
-See the supplement for the proof.
-\end{theorem}
-
-\Cref{fig:inequations} defines a notion of sub-specifications
-via the relation $<:^r$ on \textit{region}
-syntax and then $<:$ on \textit{spec} syntax inductively.
-Both relations are reflexive and transitive (we omit
-these equations). The $<:^r$ relation is congruent
-with respect to \term{+} and \term{*} and $<:$ is congruent
-with respect to \term{readOnce}, the equations of which we omit for brevity.
-
-The \trule{\textsc{eq}} rule connects region
-equations to approximation; \trule{\textsc{over}} explains
-the notion of spatial over-approximation via overlapping regions;
-\trule{\textsc{rep}} allows the \term{readOnly} modifier that
-to be dropped. The \trule{\textsc{shrink}}
-and \trule{\textsc{grow}} rules turn the spatial approximation on
-regions into approximation on specifications via the \term{atLeast} and \term{atMost}
-modifiers. Finally, \trule{\textsc{Gen$\ast$}} shows
-that specifications can be generalised by removing parts of a product
-$\ast$ which have disjoint dimensions to the rest of the region
-product (where $\mathsf{dims}$ denotes the set of dimensions specified
-in a region).
-
-\begin{theorem}[Approximation soundness]
-Code consistent with a specification $S$
- is consistent with a more general specification $T$,
-\ie{}, for all models $M$ of a piece of code:
-\begin{equation*}
-S <: T \; \wedge (\consA{M}{\interp{S}_n}) \;
-\Rightarrow \; (\consA{M}{\interp{T}_n})
-\end{equation*}
-where $\consAName$ implements consistency in checking (\cref{subsec:checking}).
-See the supplement for the proof.
-\end{theorem}
-
-\begin{figure}[t]
-\vspace{-0.5em}
-\begin{align*}
-\hspace{-0.5em}
-{\small{
-\begin{array}{c}
-%%
-%% INEQUATIONS ON REGION
-%%
-%\framebox{$\textit{region} <: \textit{region}'$} \\[0.85em]
-%
-%\dfrac{}{R <:^r R}\trule{\textsc{refl}} \qquad \dfrac{P <:^r R \quad R <:^r Q}{P <:^r
-%  Q}\trule{\textsc{trans}}
-%\qquad
-%\\[1.5em]
-\vspace{-0.4em}
-\setlength{\arraycolsep}{0.1em}
-\dfrac{\hspace{3em} m \leq n \qquad\qquad r \implies s \hspace{3em}}
-{\begin{array}{rl}
-\stenFwdSR{m}{d}{r} & <:^r \stenFwdSR{n}{d}{s}
-%& \;\, ds \subseteq es \wedge n \leq m
-\\
-\wedge \; \stenBwdSR{m}{d}{r} & <:^r \stenBwdSR{n}{d}{s}
-%& \;\, ds \subseteq es \wedge n \leq m
-\\
-\wedge \; \stenCenSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s}
-%& \;\, ds \subseteq es \wedge n \leq m
-\\
-\wedge \; \stenBwdSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s} \\
-\wedge \; \stenFwdSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s} \\[1.2em]
-\end{array}}\trule{\textsc{over}} \\[-0.2em]
-%\hspace{-0.5em}
-%\dfrac{R_1 <:^r S_1 \quad R_2 <:^r S_2}
-%      {R_1 \, \texttt{+} \, R_2 <:^r S_1 \, \texttt{+} \, S_2}
-%\trule{\textsc{cong}\texttt{+}}
-%\dfrac{R_1 <:^r S_1 \quad R_2 <:^r S_2}
-%      {R_1 \, \texttt{*} \, R_2 <:^r S_1 \, \texttt{*} \, S_2}
-%\trule{\textsc{cong}\texttt{*}} \\[1.5em]
-%%
-%% INEQUATIONS ON SPEC
-%%
-%\framebox{$\textit{spec} <: \textit{spec}'$} \\[0.8em]
-%
-%
-\dfrac{R <:^r S}{\term{atLeast}  S <: \term{atLeast}
-  R}\trule{\texttt{L}}
-\;\;
-\dfrac{R <:^r S}{\term{atMost}  R <: \term{atMost}
-  S}\trule{\texttt{M}} \\[1.5em]
-\dfrac{R <:^r S}{R <: \term{atMost} \, S}\trule{\textsc{grow}} \,
-\dfrac{R <:^r S}{S <: \term{atLeast} \, R}\trule{\textsc{shrink}} \,
-\dfrac{R \equiv S}{R <:^r S}\trule{\textsc{eq}}
- \\[1.5em]
-%%
-\dfrac{\mathsf{dims}(R) \cap \mathsf{dims}(S) = \emptyset}{R \ast S <: R}
-\trule{\textsc{Gen$\ast$}}
-\quad
-%%
-%\\[1.5em]
-%\dfrac{R <: S}{\term{readOnce} \, R <: \term{readOnce} \,
-%  S}\trule{\textsc{cong}\texttt{R}}
-%\;\;
-\;
-\dfrac{}{\term{readOnce} \, R <: R}\trule{\textsc{rep}}
-%\dfrac{}{R <: \term{atMost} R}
-%\quad
-%\dfrac{}{R <: \term{atLeast} R}\\[1.5em]
-\end{array}}}
-\end{align*}
-\vspace{-0.5em}
-\caption{Approximation $<:^r$ on \textit{regions} and
-$<:$ on \textit{specs} (omitting reflexivity, transitivity, and congruences). }
-\label{fig:inequations}
-\vspace{-0.5em}
-\end{figure}
-
-\section{Analysis, checking, and inference}
-\label{sec:analysis}
-
-\noindent
 We outline here the procedures for checking conformance
@@
 
-The core analysis, captured by the function $\textit{analyse}$ below, then proceeds as follows.
-For each assignment statement \textit{assg} within a loop whose left-hand side is an array
-subscript on neighbourhood indices, a finite map is computed mapping
-array variables to \emph{multisets} of indexing schemes:
-%
-\begin{align*}
-\hspace{-0.5em}
-\begin{array}{lr}
-\textit{analyse}(a(\overline{e_1}) = e_2)
- := \qquad\qquad\qquad & \textit{where} \; \neigh(\overline{e_1}) \\[0.3em]
-\multicolumn{2}{l}{\qquad \bigcup\{a' \mapsto \{
-\textsf{schematic}(\bar{e})\} \mid a'(\bar{e}) \leftarrow \mathsf{flowsTo}(e_2)\}}
-\end{array}
-\end{align*}
-%
-The predicate $\neigh(\bar{e_1})$ classifies indices where each
-$e \in \bar{e_1}$ is either a
-neighbourhood index (based on
-the induction variables of the loop, implicit here)
-or a constant index.\footnote{If all indices $e \in \bar{e}$ are constants, then
-  $\neigh(\bar{e})$ does not hold. Instead $a(\bar{e}$) is
-  considered a constant by our analysis, rather
-than an array subscript.}% contributing to a stencil.}
-Thus, we focus on assignments to an array subscript where the LHS
-indices $\overline{e_1}$ are neighbourhood indices.  For
-all array subscripts that flow to the right-hand side of this
-statement, a finite map is constructed, mapping each array variable
-to a multiset of indexing schemes for its subscripts, computed
-with \textsf{schematic} (\Cref{def:schematic}).
-Note \textsf{schematic} is undefined ($\bot$) if
-$\bar{e}$ contains relative indices which are not neighbourhood
-indices. The \textit{analysis} function is used by the
-\textit{modelise} function which filters out those arrays
-which contain undefined indexing schemes and adds multiplicity
-information:
-%
-\begin{align*}
-& \mathit{modelise} (\textit{assg})
-= \textit{let} \;\; M = \mathit{analyse}(\textit{assg}) (a) \;\; \textit{in}
-\\
-& a \mapsto
-\begin{cases}
-\bot & \bot \in M \\
-\textsf{only}(|M|) & \forall u \in M . \, \#(M, u) = 1 \\
-\textsf{mult}(|M|) & \forall u \in M . \, \#(M, u) = n
-\end{cases}
-\end{align*}
-where $\#(M, u)$ denotes the multiplicity of an element $u$ of the
-multiset $M$ and $|M|$ denotes the set of elements in the multiset
-$M$.
-Thus, the analysis maps source code to index schemes, which is then
-augmented with multiplicity information by $\mathit{modelise}$,
-\ie{} values in $\mathsf{Mult}(\mathcal{P}((\domainVal{})^n))$.
-
-As a small semantic concession to an otherwise syntactic
-analysis: if the LHS $\overline{e_1}$ contains non-$0$ neighbour offsets
-then all RHS schemes are \emph{relativised} by
-$\textsf{schematic}(\overline{e_1})$. For example,
-if \texttt{a(i+1) = b(i)} then relativisation of the RHS by the LHS
-produces an analysis which is equivalent to the analysis of \texttt{a(i) = b(i-1)}.
-
-\begin{example}
-\label{exm:checking}
-We demonstrate the analysis and modelise procedures over the
-following example kernel which computes the mean value
-of a five-point stencil:
-%%
-\begin{minted}{fortran}
-x = a(i-1, j)+a(i+1, j); y = a(i, j-1)+a(i, j+1)
-b(i, j) = (a(i, j) + x + y) / 5.0
-\end{minted}
-%
-%The \textit{analyse} function matches on
-%line 4 with the following set for $\textsf{flowsTo}(\texttt{a(i, j) + x +
-%  y}) = $
-%
-%\begin{align*}
-%{{ %{\small{
-%$\{$\texttt{a(i-1, j)}, \texttt{a(i+1, j)}, \texttt{a(i, j-1)},
-%  \texttt{a(i, j+1)}, \texttt{a(i, j)}$\}$}}. 
-%\end{align*}
-The  multiplicities are all one, and so \emph{modelise} yields:
-%%
-\begin{align*}
-\texttt{a} \mapsto \mathsf{only}(\{\vtwoh{-1}{0}, \vtwoh{1}{0},
-          \vtwoh{0}{-1}, \vtwoh{0}{1}, \vtwoh{0}{0}\})
-\end{align*}
-%%
-\end{example}
-
-%Given an assignment $\textit{assg}$ to an array subscript in the source language
-%and an array variable $a$, then $\mathit{modelise}(\textit{assg}, a)$ performs
-%an analysis to find all array subscripts of $a$ converting them to
-%index schemes. For the moment, we leave this analysis abstract
-%as the function $\emph{analyse}$ in the following definition of $\emph{modelise}$:
-%
-
-\subsection{Specification checking}
+\subsection{Checking code against specifications}
 \label{subsec:checking}
 
-Checking verifies the access pattern of a stencil computation in the
-source language against its associated specifications. The model of
-(\Cref{sec:semantics}) guides this process.  Checking
-proceeds by generating a model from a specification and generating
-a model from the source code, comparing them for
-\emph{consistency} base on index scheme
-\emph{generalisation} (\Cref{def:generalisation}).
-
-\begin{definition}[Checking]
-Let $a$ be an array variable and
-$S$ a specification associated to $a$
-via \texttt{!= stencil $S$ :: a}.
-
-A stencil kernel
- $v(\overline{e_1}) = e_2$
-conforms to $S$ if the predicate $\textbf{check}_n(S, v(\overline{e_1}) = e_2, a)$ holds,
-defined:
-\begin{align*}
-& \textbf{check}_n(S, \textit{assg}, a) =
-  \consSub{n}{\textit{modelise}(\textit{assg})(a)}{\interp{S}_n}
-&
-\end{align*}
-Thus, checking applies \textit{modelise}
-(\Cref{subsec:analysis}) to extract index schemes from source code,
-augmented with multiplicity for the whole set of schemes.
-This is compared with the model of the specification $\interp{S}_n$ by
-the consistency predicate $\consAName$.
-\end{definition}
-
-\begin{figure}[t]
-\vspace{-0.5em}
-\begin{align*}
-\hspace{-0.5em}
-\begin{array}{c}
-%%
-\trule{up}\dfrac{\forall u \in M, \exists v \in N .  \, u \preceq v}
-      {\cons{M}{\textsf{upper}(N)}}
-\,
-\trule{low}\dfrac{\forall v \in N, \exists u \in M . \, u \preceq v}
-      {\cons{M}{\textsf{lower}(N)}} \\[1.2em]
-%%
-\trule{exact}\dfrac{\cons{M}{\textsf{lower}(N)} \qquad \cons{M}{\textsf{upper}(N)}}
-  {\cons{M}{\textsf{exact}(N)}} \\[1.2em]
-%%
-\trule{both}\dfrac{\cons{M}{\textsf{lower}(M_1)} \qquad \cons{M}{\textsf{upper}(M_2)}}
-  {\cons{M}{\textsf{both}(M_1, M_2)}}\\[1.2em]
-%%
-\trule{$1$}
-\dfrac{\cons{M}{N}}{\consA{\textsf{once}(M)}{\textsf{once}(N)}}
-\;\;
-\trule{$*$}
-\dfrac{\cons{M}{N}}{\consA{\textsf{mult}(M)}{\textsf{mult}(N)}}
-  \\[1.2em]
-\trule{$1 < *$}
-\dfrac{\cons{M}{N}}{\consA{\textsf{once}(M)}{\textsf{mult}(N)}}
-\end{array}
-\end{align*}
-\vspace{-0.5em}
-\caption{Consistency of models (code vs. specification)}
-\label{fig:consistency}
-\vspace{-0.9em}
-\end{figure}
-
-\begin{definition}
-Consistency relates models of source code to models
-of specifications (which are bounded) via predicates:
-\begin{align*}
-\consName_n & \subseteq \mathcal{P}(\domainVal{}^n)
-\times \textsf{Approx}(\mathcal{P}(\domainVal{}^n)) \\
-\consAName_n & \subseteq \textsf{Mult}(\mathcal{P}(\domainVal{}^n))
-\times \textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))
-\end{align*}
-%
-where $\consName$ relates models without multiplicity.
-\Cref{fig:consistency} defines the two predicates, where for
-brevity we make implicit the subscript $n$ defining the
-dimensionality.  We denote sets of indexing schemes
-$\mathcal{P}((\domainVal{})^n)$
-from source code model by $M$ and from a specification model by $M$.
-
-Rules (\emph{up}) and (\emph{low}) are the core; (\emph{up})
-considers the case where the specification is
-an \term{atMost} upper-bound approximation.
-An upper-bound specification means that for every scheme in the code
-model, modelling a source-level index, $\forall u \in M$ there exists
-a generalising scheme in the spec model $\exists v \in N$, \ie{}, $u
-\preceq v$. That is, every source-level index scheme is generalised by the specification.
-The (\emph{down}) rule provides the case of
-\term{atLeast} lower-bound specifications. Consistency is dual to the above:
-for every scheme in the specification model $\forall v \in N$ there
-exists a scheme in the code model $\exists u \in M$ such that $u
-\preceq v$. That is, every specification scheme is generalised by some source
-index scheme.
-
-The (\emph{up}) and (\emph{down}) rules are combined for an \textsf{exact}
-specification (\textit{exact}) such that a specification's model
-provides both a consistent lower and consistent upper bound on
-the source code. Thus, for every scheme in the code model there
-is a scheme in the spec model that generalises it, and for every
-scheme in the specification there is a scheme in the source model that
-specialises it.  The (\textit{both}) rule is similar, decomposing
-a specification with both a lower and upper bound into
-consistency of these bounds.
-
-Rules $(1)$ and $(*)$ extend the \consName{}
-predicate to $\consAName{}$ with
-multiplicity information. Both rules
-state that consistent models remain consistent if they
-have the same multiplicity. Relatedly, rule $(1 < *)$ ensures
-that absence of a \texttt{readOnce} modifier in the specification
-does not prohibit a set of unique stencil accesses in the
-code from being consistent.
-\end{definition}
-
-\begin{example}
-  Consider the following five-point stencil kernel with a (quite general) specification:
-%
-\begin{minted}[breakindent=8.5em]{fortran}
-!= region :: fivepoint = centered(depth=1,dim=1) + centered(depth=1,dim=2)
-!= stencil readOnce, fivepoint :: b
-a(i,j) = b(i,j-1)+b(i,j)+b(i,j+1)+b(i-1,j)+b(i+1,j)
-\end{minted}
-%
-We refer to line three as $l3$ for brevity,
-which is checked via
-$check(({\textcolor{darkgreen}{\texttt{readOnce, fivepoint}}}),l3,\texttt{b})$
-with steps:
-\begin{enumerate}
-%
-\item Generate the model from the specification as in
-  \cref{sec:semantics}.
-%
-\vspace{-0.4em}
-\begin{align*}
- N_s= \{  \vtwo{-1}{\infty},
-          \vtwo{0}{\infty},
-          \vtwo{1}{\infty},
-          \vtwo{\infty}{-1},
-          \vtwo{\infty}{0},
-        \vtwo{\infty}{1} \}  \\
-\interp{{\textcolor{darkgreen}{\texttt{readOnce, fivepoint}}}}_2 =
-  \textsf{once}(\textsf{exact} (N_s))
-\end{align*}
-%\vspace{-1.6em}
-
-\item Generate code model $\textit{modelise}(l3, \texttt{b}) =
-  \textsf{once} (M_c)$ where:
-%
-\vspace{-0.9em}
-\begin{align*}
-M_c = & \{
-    \vtwo{0}{-1},
-    \vtwo{0}{0},
-    \vtwo{0}{1},
-    \vtwo{-1}{0},
-    \vtwo{1}{0}
-  \}
-\end{align*}
-\vspace{-1.1em}
-
-\item By the definition of consistency (\Cref{fig:consistency})
- $\textsf{once}(\textsf{exact}(N_s))$ is consistent with
-  $\textsf{once}(M_c)$ when:
-\vspace{-0.6em}
-  \begin{align*}
-    & \consSub{2}{\textsf{up}(N_s)}{M_c} \;\,
-    \wedge \;\, \consSub{2}{\textsf{low}(N_s)}{M_c}
-  \end{align*}
-%\vspace{-1.6em}
-
-\item This holds by the following sets of scheme
-  generalisations: for the upper bound, $\forall u \in M_c,
-  \exists v \in N_s . u \preceq_2 v$:
-  \begin{align}
-    \hspace{-1em}
-    \begin{array}{l}
-   \hspace{-1.5em}
-     \vtwoh{0}{-1} \preceq_2 \vtwoh{0}{\infty} \quad
-     \vtwoh{0}{0} \preceq_2 \vtwoh{0}{\infty} \quad
-     \vtwoh{0}{1} \preceq_2 \vtwoh{0}{\infty} \\[0.5em]
-        \hspace{-1.5em}\vtwoh{-1}{0} \preceq_2 \vtwoh{\infty}{0} \quad
-     \vtwoh{1}{0} \preceq_2 \vtwoh{\infty}{0}
-     \end{array}
-  \label{eq:check-example-upper}
-  \end{align}
-  %
-  and for the lower bound: $\forall v \in N_s, \exists u \in M_c . u
-  \preceq_n v$:
-  \begin{align}
-    \hspace{-1em}
-    \begin{array}{l}
-     \hspace{-1.3em}\vtwoh{-1}{0} \preceq_2 \vtwoh{-1}{\infty} \quad
-     \vtwoh{0}{0} \preceq_2 \vtwoh{0}{\infty} \quad
-     \vtwoh{1}{0} \preceq_2 \vtwoh{1}{\infty}  \\
-     \hspace{-1.3em}\vtwoh{0}{-1}  \preceq_2 \vtwoh{\infty}{-1} \quad
-     \vtwoh{0}{0}  \preceq_2 \vtwoh{\infty}{0} \quad
-     \vtwoh{0}{1}  \preceq_2 \vtwoh{\infty}{1}
-     \end{array}
-    \label{eq:check-example-lower}
-  \end{align}
-\end{enumerate}
-Since the consistency predicate is satisfied, the array
-  accesses flowing to line three in the stencil computation are valid
-  with respect to the {\textcolor{darkgreen}{\texttt{fivepoint}}} specification.
-\end{example}
-\begin{example}
-Consider an erroneous user modification to $l3$ with an
-additional term \texttt{b(i+2,j+2)}, call it $l3'$.% such that $l3'$ is:
-%
-%\begin{minted}[firstnumber=3]{fortran}
-%a(i,j) = b(i,j-1) + b(i,j) + b(i,j+1) + b(i-1,j) + b(i+1,j) + b(i+2,j+2)
-%\end{minted}
-
-Checking proceeds as in the previous example, except
-$\mathit{modelise}(l3',\texttt{b})$ generates an additional scheme
-$\vtwoh{2}{2}$. The generalisations for
-the lower bound \eqref{eq:check-example-lower}
-of step 4 above still hold since its consistency condition
-universally quantifies over the \emph{specification model}. However, for
-the upper bound, the consistency condition universally quantifies
-over schemes $u$ of the \emph{code model} for which there must exist
-a scheme $v \in N_s$ such that $u \preceq_2 v$. Thus, we require a $v
-\in N_s$ such that $\vtwoh{2}{2} \preceq_2 v$ but no such generalising
-scheme exists in the spec model. Hence checking fails.
-\end{example}
-
-\subsection{Specification inference}
+\subsection{Infering specification automatically}
 \label{subsec:inference}
 
-To aid adoption of our approach to legacy code, we provide
-an inference procedure for generating specifications from code, which
-are inserted automatically as comments.
-We illustrate inference using the
- five-point stencil of
- \Cref{exm:checking}.
+We provide an inference procedure for generating specifications from
+code, which are inserted automatically as comments. This helps support
+maintenance of legacy code, and aids understanding of our
+specification language. We illustrate inference using the
+five-point stencil of \Cref{exm:checking}.
 
-\subsubsection{Step 1: Data access analysis}
-\label{sec:inf-step1}
+The preceding source code analysis converts the concrete syntax of array
+subscripts into a subscript space in $\mathcal{P}(\bz)^N$.
+Let $U$ range over finite maps from array variables to sets of
+of subscript schemes.
 
-The first step performs the
-analysis of \cref{subsec:analysis} to produce a
-model of the code in terms of indexing schemes.
-The \emph{analyse} and \emph{modelise} procedures before yielded the finite map:
-%
-\begin{align*}
-\texttt{a} \mapsto \mathsf{only}(\{\vtwoh{-1}{0}, \vtwoh{1}{0},
-          \vtwoh{0}{-1}, \vtwoh{0}{1}, \vtwoh{0}{0}\})
-\end{align*}
-%
+ The
+inference then attempts to construct a $\emph{Region}_N$ from this.
 
-\subsubsection{Step 2: Coalesce contiguous schemes into spans}
-\label{sec:inf-step3}
-
-Let $U$ range over the finite maps computed by \textit{modelise}.  For
+ For
 each array variable $a \in \mathsf{dom}(U)$, the algorithm constructs a set of
@@
 
-\newcommand{\spanOp}{\textsf{spans}}
+\begin{enumerate}[leftmargin=1.5em]
+A covering of (possibly overlapping) regions is computed iteratively
+by the following steps until a fixed-point is reached:
 
-\begin{definition}[Spans]
-  A \emph{span} represents an $n$-dimensional rectangle
-  as a pair of $n$-dimensional vectors (or
-  represented as a $2 \times n$ matrix) with values drawn from
-  $\domainVal{}$. This gives the co-ordinates of lower-bound 
-  and upper-bound vertices. We write
-  spans as $\vect{x} = [\vect{x}^L \; \vect{x}^U]$, where $\vect{x}^L$
-  and $\vect{x}^U$ give the lower and upper bounds.  In
-  the following, $\vect{x}, \vect{y}$ range over spans.  For a span
-  $\vect{x}$, we write $\vect{x}_{i:j}$ for the sub-span comprising the
-  subvectors $[\vect{x}^L_{i:j} \; \vect{x}^U_{i:j}]$ from entry $i$
-  to entry $j$ (inclusive).
+\begin{enumerate}
+
+\item \textbf{Coalesce intervals}
+Intervals which are adjacent (contiguous in one dimension) are
+coalesced.
+
+\begin{defn}{(Adjacent)}
+  A subscript $S$ is said to be \emph{adjacent at index $j$} of another
+  subscript $S'$ if for every index $i$ except $j$, $\pi_i(S) = \pi_i(S')$ but
+  at $j$ then $\pi_j(S) \neq \pi_j(S')$ such that $|S| = |S'| + 1$.
+\end{defn}
+%
+
+\begin{definition}[Contiguous]
+
 \end{definition}
 
-\noindent
-Each indexing scheme $u \in U(a)$ is converted to a \emph{unit} span
-$u \mapsto [u \; u]$, defining the lower and upper bound of an
-$n$-dimensional cube of side length $1$. For the five-point stencil
-example there are five unit spans, illustrated in
-\Cref{fig:inference-steps-informal}(a).
+\item \textbf{Remove completely overlapped intervals}
 
-The next step repeatedly coalesces any contiguous spans until a fixed
-point is reached, resulting in a set of (possibly overlapping) spans
-covering the original space of schemes. Two $n$-dimensional spans
-$\vect{x}$ and $\vect{y}$ are contiguous if for all but one
-$j \in \{1, \ldots, n\}$ then $\vect{x}_j = \vect{y}_j$ and for one
-$j$ then $\vect{x}^U_1 + 1 = \vect{y}^L$. For example, spans
-$\left[\vtwoh{-1}{0}\; \vtwoh{-1}{0}\right]$ and
-$[\vtwoh{0}{0} \; \vtwoh{1}{0}]$ are contiguous in the first dimension
-and can be coalesced into the span $[\vtwoh{-1}{0} \; \vtwoh{1}{0}]$.
-
-\begin{definition}\label{def:span-coalesc}
-  Span coalescing $\bullet$ is partial, defined for spans contiguous in the
-  first dimension:
-%%
-\begin{align*}
-\vect{x} \bullet \vect{y}
-= (\vect{x}^L, \vect{y}^U) \;\;\;\, \textit{if} \;\;\; \vect{x}^U_1 + 1 = \vect{y}^L_1 \; \wedge \;
-\vect{x}_{2:n} = \vect{y}_{2:n} %\\
-%\bot  & \textit{otherwise}
-%\end{cases}
-% $m$-dimensions into coalesced spans which vary in
-% $m+1$-dimensions.
-\end{align*}
-%%
-\end{definition}
-\noindent
-To coalesce regions which are contiguous in other dimensions
-other than the first, we generate all rotations of the spans in
-$n$-dimensional space by permutation on spans, and fold together contiguous
-regions using $\bullet$ (above). This
-generates all possible ways of coalescing regions. For our example,
-the unit spans in \Cref{fig:inference-steps-informal}(a)
-are coalesced into the contiguous spans in
-\Cref{fig:inference-steps-informal}(b).
-This procedure is iterated until a fixed point is reached (in one step
-in this example).
- Then any spans that are completely contained by any other span are deleted,
-leaving a minimal set of spans (which may overlap, but none of which
-fully contains another), \Cref{fig:inference-steps-informal}(c).
-The final spans are used to synthesise concrete stencil specification
-syntax (\cref{sec:inf-step4}), where
-each span becomes a product \texttt{*} of regions which are
-combined by the sum \texttt{+}.
-
-More formally the algorithm proceeds as follows.
-For all $a \in \mathsf{dom}(U)$, let $M = U(a)$. Every scheme in $M$
-is mapped to a unit span
-by $\textsf{unitSpans}(M) = \{[u \; u] \mid u\!\in\!M\}$, \eg{}:
-%%
-\begin{align*}
-  N(U(\texttt{a})) = \stwo{0}{0}{0}{0} \stwo{1}{0}{1}{0} \stwo{-1}{0}{-1}{0} \stwo{0}{1}{0}{1} \stwo{0}{-1}{0}{-1}
-\end{align*}
-%%
-In the following we let $M_0 = N(U(\texttt{a}))$ for our example.
-
-Each set of unit spans is applied to the fixed point of the \spanOp{} function
-(defined shortly), which coalesces spans into contiguous regions. That is, we compute
- $(\mu \, \spanOp)$ on input $N(M)$, where \spanOp{} is defined by the steps:
-
-\begin{figure}[t]
-\vspace{-1em}
-\hspace{-0.5em}
-{\scalebox{0.96}{
-\begin{minipage}{0.16\linewidth}
-\begin{align*}
-\setlength{\arraycolsep}{0em}
-\begin{array}{c}
-\quad \\[0em]
-\begin{tikzpicture}[scale=0.8]
-%\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
-\node at (1.25,1.22) {\textnormal{\footnotesize{0,0}}};
-\draw[thick] (1,1) rectangle (1.5,1.5);
-\draw[thick] (1.5,1) rectangle (2,1.5);
-\draw[thick] (1,0.5) rectangle (1.5,1);
-\draw[thick] (0.5,1) rectangle (1,1.5);
-\draw[thick] (1,1.5) rectangle (1.5,2);
-\end{tikzpicture} \\[0em]
-(a)
-\end{array}
-\end{align*}
-\end{minipage} $\rightarrow$ \begin{minipage}{0.43\linewidth}
-\begin{align*}
-\setlength{\arraycolsep}{0em}
-\begin{array}{c}
-\begin{tikzpicture}[scale=0.72]
-%\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
-%\node at (1.25,1.22) {\textnormal{\small{0,0}}};
-\draw [black, thin] (0.35,0) to [round left paren ] (0.35,2);
-\draw[thick] (0.5,1) rectangle (2,1.5);
-\draw[thick] (1,0.40) rectangle (1.5,0.90);
-\draw[thick] (1,1.60) rectangle (1.5,2.10);
-\node at (2.3,1.22) {+};
-\draw[thick] (3.8,1) rectangle (4.3,1.5);
-\draw[thick] (2.6,1) rectangle (3.1,1.5);
-\draw[thick] (3.2,0.5) rectangle (3.7,2);
-\draw [black, thin] (4.4,0) to [round right paren ] (4.4,2);
-\end{tikzpicture} \\[-0.5em] (b)
-\end{array}
-\end{align*}
-\end{minipage}$\rightarrow$ \begin{minipage}{0.25\linewidth}
-\begin{align*}
-\setlength{\arraycolsep}{0em}
-\begin{array}{c}
-\begin{tikzpicture}[scale=0.72]
-%\node at (1.25,1.22) {\textnormal{\small{0,0}}};
-\draw [black, thin] (0.8,0) to [round left paren ] (0.8,2);
-\draw[thick] (1,0.5) rectangle (1.5,2);
-\node at (1.8,1.22) {+};
-\draw[thick] (2.1,1) rectangle (3.6,1.5);
-\draw [black, thin] (3.7,0) to [round right paren ] (3.7,2);
-%\node at (2.75,1.22) {\textnormal{\small{0,0}}};
-\end{tikzpicture} \\[-0.5em] (c)
-\end{array}
-\end{align*}
-\end{minipage}}}
-\caption{Illustration of grouping spans in
-inference}
-\label{fig:inference-steps-informal}
-\vspace{-0.8em}
-\end{figure}
-
-\noindent
-\textbf{1)\,} Compute all permutations on the column vectors in a span, \ie{},
-  $[\vect{x}^L \vect{x}^U] \mapsto [\pi\vect{x}^L \, \pi\vect{x}^U]$
-for a permutation $\pi$. For each permutation function $\pi^n_i$
-(the $i$-th permutation for vectors of size $n$) we pair the
-permutation function with the set of permuted spans so that
-the spans can be un-permuted later.
-%
-\begin{equation*}
-P(M) = \bigcup_{i \in n!} (\pi^n_{i} , \; \{[\pi^n_i
-\vect{x}^L \; \pi^n_i\vect{x}^U] \, \mid \, [\vect{x}^L \; \vect{x}^U]
-\leftarrow M\}
-\end{equation*}
-%
-This corresponds to all rotations of the space, \eg{}:
-%
-\begin{align*}
-P(M_0) =
-\{&(\pi^2_1, \{ \stwo{0}{0}{0}{0},
-\stwo{1}{0}{1}{0},
-\stwo{-1}{0}{-1}{0},
-\stwo{0}{1}{0}{1},
-\stwo{0}{-1}{0}{-1} \})
-\\
-&(\pi^2_2, \{
- \stwo{0}{0}{0}{0},
- \stwo{0}{1}{0}{1},
- \stwo{0}{-1}{0}{-1},
- \stwo{1}{0}{1}{0},
- \stwo{-1}{0}{-1}{0}\})\}
-\end{align*}
-%
-where $\pi^2_1$ is the identity permutation and $\pi^2_2$ is the
-permutation that flips the order of the two elements in the
-vectors.
-
-\noindent
-\textbf{2)\,} Sort each permutation set into a list, by the ordering:
-\begin{equation*}
-  \vect{x} \leq \vect{y} =
-      \exists i \, . \; \vect{x}^L_{i} \leq \vect{y}^L_{i} \; \wedge \;
-        (i = n \vee \vect{x}_{i+1:n} = \vect{y}_{i+1:n})
-\end{equation*}
-%
-giving the sorting function $\textsf{sort}(M)$, \eg{}, $\textsf{sort}(P(M_0))$:
-%%
-\begin{align*}
-\{&(\pi^2_1, [
-\stwo{0}{-1}{0}{-1},
-\stwo{-1}{0}{-1}{0},
-\stwo{0}{0}{0}{0},
-\stwo{1}{0}{1}{0},
-\stwo{0}{1}{0}{1}] )
-\\
-&(\pi^2_2, [
-\stwo{0}{-1}{0}{-1},
-\stwo{-1}{0}{-1}{0},
-\stwo{0}{0}{0}{0},
-\stwo{1}{0}{1}{0},
-\stwo{0}{1}{0}{1}])\}
-\end{align*}
-\noindent
-\textbf{3)\,} Reduce each list pairwise by $\bullet$
-(\Cref{def:span-coalesc}) to coalesce contiguous regions,
-given by $\textsf{fold}\bullet$. For our example:
-%
-\begin{align*}
-\{&(\pi^2_1, [
-\stwo{0}{-1}{0}{-1},
-\stwo{-1}{0}{1}{0},
-\stwo{0}{1}{0}{1}
-]), (\pi^2_2, \textit{same as for $\pi_1$})\}
-\end{align*}
-%
-\textbf{4)\,} Un-permute and union together, \ie{},
-%
-\[
-U(M) = \bigcup \{[\pi \vect{x}^L, \pi \vect{x}^U]
- \mid [\vect{x}^L, \vect{x}^U] \leftarrow S, (\pi, S) \leftarrow M\}
-\]
-For our example:
-%
-\begin{align*}
-U(M) =
-\{\stwo{0}{-1}{0}{-1},
-\stwo{-1}{0}{1}{0},
-\stwo{0}{1}{0}{1},
-\stwo{-1}{0}{-1}{0},
-\stwo{0}{-1}{0}{1},
-\stwo{1}{0}{1}{0}\}
-\end{align*}
-%
-\textbf{5)\,} Filter by the region containment predicate
-$\containedin$; if any region is contained within another then remove the
-  smaller:
-%
-\begin{align*}
-& \vect{x} \containedin \vect{y} = \vect{y}^L_1 \leq \vect{x}^L_1 \wedge \vect{x}^U_1 \leq \vect{y}^U_1
-  \wedge (\vect{x}^L_{2:n}, \vect{x}^U_{2:n}) \containedin
-  (\vect{y}^L_{2:n}, \vect{y}^U_{2:n})
-\end{align*}
-For our example this then yields the final result of:
-\begin{align*}
-\spanOp(M)
-= \{\stwo{-1}{0}{1}{0} \stwo{0}{-1}{0}{1}\}
-\end{align*}
-since $\stwo{0}{1}{0}{1} \sqsubseteq \stwo{0}{-1}{0}{1}$
-and $\stwo{1}{0}{1}{0} \sqsubseteq \stwo{-1}{0}{1}{0}$.
-Applying \spanOp{} again yields the same
-result for our example.%; the fixed point is reached.
-
-\subsubsection{Step 4: Synthesise specifications from spans}
-\label{sec:inf-step4}
-
-\newcommand{\oplusbig}{\operatornamewithlimits{\term{\Large{$\oplus$}}}}
-\newcommand{\odobig}{\operatornamewithlimits{\term{\Large{$\odot$}}}}
-\newcommand{\bplus}{\operatornamewithlimits{\term{\Large{+}}}}
-\newcommand{\tySum}[1]{#1^{\term{+}}}
-\newcommand{\tyProd}[1]{#1^{\term{*}}}
-\newcommand{\specDNF}{\textbf{spec}}
-
-Abstract syntax trees of the specification language
-are synthesised from the set $\spanOp(M)$ of coverings spans.
-This conversion is defined in terms
-of an algebra on \textit{spec} syntax in disjunctive-normal form
-(which we denote \specDNF{}) and which is wrapped
-in the $\textsf{Approx}$ type (see
-\Cref{def:mult-and-approx}).  The algebra mirrors the shape
-of the region operators $\term{+}$ and $\term{*}$ 
-with binary operators $\oplus$ and $\odot$ on
-$\mathsf{Approx}(\specDNF)$.
-A function $\mathsf{toSpec}$ maps a dimension identifier
-and a pair of lower and upper bound
-values (one-dimensional) to a specification with a one-dimensional region.
-%%
-\begin{align*}
-\mathsf{toSpec} & : \mathbb{N}_{>0} \rightarrow (\domainVal{} \times \domainVal{}) \rightarrow \mathsf{Approx}(\specDNF)
-\end{align*}
-%%
-A span is converted into an $n$-dimensional
-specification by applying $\mathsf{toSpec}$ to each
-pair of lower and upper bounds in a span, per dimension, and multiplying these
-by $\odot$. The resulting specifications for
-each span are summed by $\oplus$:
-\begin{equation*}
-\textsf{spec}(M) =
-\oplusbig\limits_{\vect{x} \in \spanOp(M)} \,
-\odobig\limits_{i \in \{1, \ldots, n\}} \, \mathsf{toSpec} \; i \; (\vect{x}^L_i, \vect{x}^U_i)
-\end{equation*}
-We unpack the definitions of this algebra.
-
-\begin{definition}[Specifications in DNF]
-Let $r_1, ..., r_n$ range over elements of the \textit{region}
-syntax (see \Cref{fig:syntax}).
-The set \specDNF{} is a subset of syntax trees \textit{spec},
-containing specifications in disjunctive normal form over regions
-where the disjunct is \term{+} and the conjunct is \term{*}; 
- elements of \specDNF{} are of the form:
-%%
-$
-(r^1_{1} \; \term{*} \ldots \term{*} \; r^1_n)\; \term{+} \; \ldots
-\term{+} \; (r^m_1 \, \term{*} \; \ldots \term{*} \; r^m_p)
-$.
-%
-%More formally, $\specDNF \subseteq (\textit{spec} \cup \{\epsilon\})$, defined:
-%%%
-%\begin{align*}
-%\specDNF{} & ::= \tySum{\specDNF} \mid \epsilon \quad\;\;
-%\tySum{\specDNF} ::= \tyProd{\specDNF} \; \term{+} \; \tySum{\specDNF} \mid
-%  \tyProd{\specDNF} \\
-%\tyProd{\specDNF} & ::= \textit{region} \; \term{*} \; \tyProd{\specDNF} \mid
-%   \textit{region}
-%\end{align*}
-%%
-In the following, $S, T$ range over elements of
-\specDNF{} and $\tyProd{s}, \tyProd{t}$ range over conjuncts
- \eg{} $\tyProd{s} = r^1_{1} \; \term{*}
-\ldots \term{*} \; r^1_n$.
-\end{definition}
-
-\newcommand{\exactR}{\mathsf{exact}\;}
-\newcommand{\exact}[1]{\exactR(#1)}
-\newcommand{\upperB}[1]{\mathsf{upper}\;(#1)}
-
-\begin{figure}[t]
-%\newcommand{\llower}{\vect{x}^L_i}
-%\newcommand{\lupper}{\vect{x}^U_i}
-\vspace{-0.7em}
-\newcommand{\llower}{l}
-\newcommand{\lupper}{u}
-{\scalebox{0.95}{
-\begin{minipage}{1\linewidth}
-\begin{align*}
-&\mathsf{toSpec} \; d \; (\llower, \lupper) = \\
-&\hspace{-0.4em}\begin{cases}
-%% POINT
-\exact{\stenReflS{d}} & \llower = 0 \, \wedge \, \lupper = 0\\[0.02em]
-%% FWD
-\exact{\stenFwdS{\lupper}{d}}  & \llower = 0 \, \wedge \, \lupper > 0 \\[0.02em]
-\exact{\stenFwdSR{\lupper}{d}{\irreflS}}\!\!\!\! & \llower = 1 \, \wedge \, \lupper > 0 \\[0.02em]
-%% BWD
-\exact{\stenBwdS{|\llower|}{d}}  & \llower < 0 \, \wedge \, \lupper = 0 \\[0.02em]
-\exact{\stenBwdSR{|\llower|}{d}{\irreflS}}\!\!\!\! & \llower < 0 \, \wedge \, \lupper = -1 \\[0.02em]
-%% CEN
-\exact{\stenCenS{\lupper}{d}}& \!\!\!\llower < 0 \wedge \lupper > 0 \wedge |\llower| = \lupper \\[0.02em]
-\exactR(\stenBwdS{|\llower|}{d} & \!\!\!\llower < 0 \wedge \lupper > 0  \wedge |\llower| \neq \lupper \\
-\qquad\qquad \texttt{+} \, \stenFwdS{\lupper}{d}) & \\[0.02em]
-% BOUND
-\upperB{\stenFwdS{\lupper}{d}} & \llower > 1 \\[0.02em]
-\upperB{\stenBwdS{|\llower|}{d}} \quad & \llower < (-1)
- \\[0.02em]
-% ABSOLUTE REP
-\exact{\epsilon} & \llower = \infty \, \vee \, \lupper = \infty
-  \end{cases}
-\end{align*}
-\end{minipage}
-}}
-\caption{\textsf{toSpec} constructor of $\mathsf{Approx}(\specDNF)$ values.}
-\label{fig:to-spec}
-\vspace{-0.8em}
-\end{figure}
-
-\begin{definition} The $\mathsf{toSpec}$ constructor is defined
-in \Cref{fig:to-spec}, where the lower and upper bounds
-determine region constants; $\mathsf{toSpec} \, d \, (l, u)$ is defined
- if $l$ less than or equal to $u$.
-
-Positive lower and upper bound yield \term{forward} regions, and
-negative lower and upper bounds yield \term{backward} regions.
-The \texttt{nonpointed} attribute (abbreviated \term{np})
-is present when the lower bound is at $1$ or $-1$. 
-In the sixth-case, where the lower bound
-is negative and the upper bound is positive and they have the same
-magnitude, then a \texttt{centered} region is constructed with
-the depth as the common magnitude. If the magnitudes are not the same
-(seventh case) then separated \texttt{forward} and \texttt{backward}
-regions are summed, with their corresponding depths.
-If the lower bound is greater than 1
-(which implies the upper bound is greater than 1 too) then
-this represents a region that is not at the origin (or next to) the
-origin and thus cannot be represented in the specification language.
-Thus, an upper bound is produced showing the maximum extent
-of the region. The penultimate case is the dual, when the lower bound is less
-than $-1$, giving an upper bound on a \texttt{backward} specification.
-\end{definition}
-
-\begin{definition} The $\odot$ operation on $\textsf{Approx}(\specDNF{})$
-  is defined in terms of the intermediate operator $\hat{\odot}$ on
-  \specDNF{}, defined:
-%%
-\begin{align*}
-S \,\hat{\odot}\, \epsilon = S \quad
-\epsilon \,\hat{\odot}\, S = S \qquad
-S \,\hat{\odot}\, T = \bplus_{(\tyProd{s}, \tyProd{t}) \in (S \times
-   T)} (\tyProd{s} \; \term{*} \; \tyProd{t})
-\end{align*}
-%%
-That is, the $\hat{\odot}$ product of two specifications $S$ and $T$ in DNF form
-is the $\term{+}$ sum of all pairwise $\term{*}$ products for every
-pair of $\tyProd{spec}$ drawn from $S$ and $T$. The result
-is indeed in DNF-form, where $\hat{\odot}$ is essentially applying
- the \trule{\textsc{dist}} rule of the equational theory. The operation is
-commutative and associative, and sound with respect to our
-model (see \Cref{lem:alg-soundness} below).
-The $\hat{\odot}$ operation is then lifted to all combinations of
-\textsf{Approx} for $\odot$ in \Cref{fig:odotlift} where
-in the last case $\textit{inj}$ corresponds to the unary
-injections of \textsf{Approx}. For brevity we omit the cases where the arguments above are flipped
-as $\odot$ is commutative.
-\end{definition}
-
-\begin{figure}[t]
-\vspace{-1em}
-{\small{
-\begin{align*}
-\textsf{lower}(S) \odot \textsf{exact}(T) = \; &
-       %\textsf{exact}(T) \odot \textsf{lower}(S) = \;
-       \textsf{both} \, (S \hat{\odot} T, \; T) \\
-\textsf{upper}(S) \odot \textsf{exact}(T) = \; &
-       %\textsf{exact}(T) \odot \textsf{upper}(S) =
-       \textsf{both} \, (T, \; S \hat{\odot} T) \\
-\textsf{lower}(S) \odot \textsf{both}(T_l,T_u) = \; &
-        %\textsf{both}(T_l,T_u) \odot \textsf{lower}(S) = \;
-        \textsf{both} \, (S \hat{\odot} T_l, \; T_u) \\
-\textsf{upper}(S) \odot \textsf{both}(T_l,T_u) = \; & \textsf{both} \,
-                                                      (T_l, \; S \hat{\odot} T_u) \\
-\textsf{both}(S_l,S_u) \odot
-\textsf{both}(T_l,T_u)  = \; & \textsf{both} (S_l \hat{\odot} T_l, \; S_u
-                         \hat{\odot} T_u) \\
-\textit{inj}(S) \odot \textit{inj}(T) = \; & \textit{inj}(S
-                                              \hat{\odot} T)
-\end{align*}
-\vspace{-2em}
-}}
-\caption{Lifting $\hat{\odot}$ to approximations.}
-\label{fig:odotlift}
-\end{figure}
-
-\begin{definition}The $\oplus$ operation on $\textsf{Approx}(\specDNF)$
-is defined similarly to $\odot$, with the intermediate $\hat{\oplus}$ on
-$\specDNF$:
-%%
-\begin{equation*}
-S \,\hat{\oplus}\, \epsilon = S \quad
-\epsilon \,\hat{\oplus}\, S = S \quad
-S \, \hat{\oplus} \, T = S \, \term{+} \, T
-\end{equation*}
-%%
-Thus, if neither specification is empty, their $\hat{\oplus}$ is just
-the syntactic sum $\term{+}$. Then $\oplus$ is the lifting of
-$\hat{\oplus}$ to \textsf{Approx} with the shape as $\odot$ in Figure~\ref{fig:odotlift}.
-\end{definition}
-
-\begin{lemma}%[$\hat{\odot}$, $\hat{\oplus}$ soundness]
-$\forall S,
-  T \in \specDNF$ where $S\!\neq \epsilon \wedge T\!\neq \epsilon$:
-\begin{align*}
-\interp{S \; \hat{\odot} \; T}_n = \interp{S}_n \otimes \interp{T}_n
-  \qquad
-\interp{S \; \hat{\oplus} \; T}_n = \interp{S}_n \cup \interp{T}_n
-\end{align*}
-\label{lem:alg-soundness}
-\vspace{-2em}
-\end{lemma}
-%
-\noindent
-Thus, our model validates specification synthesis.
-
-The final step applies a simplification pass, orienting the
-equational theory $\equiv$ (\Cref{sec:eqs}) into rewrite rules
-based on some simple heuristics.
-For our example, we infer and then synthesise the
-final 5-point specification:
-%%
-\begin{minted}[linenos=false,breakindent=2.4em,xleftmargin=0.2em]{fortran}
-!= stencil centered(depth=1, dim=1)*pointed(dim=2) + centered(depth=1, dim=2)*pointed(dim=1):: a
-\end{minted}
-%%
-
-\section{Evaluation}
-\label{sec:evaluation}
-
-%We have processed nearly one million lines of Fortran code using our
-%CamFort tool and examined the output to determine how useful
-%and applicable it may be. Our corpus of Fortran code is collected from
-%many computational science projects for which we believe our tool will
-%be potentially helpful.
-
-% Corpus details
-% specfem3d ``simulates acoustic (fluid), elastic (solid), coupled
-% acoustic/elastic, poroelastic or seismic wave propagation in any
-% type of conforming mesh of hexahedra (structured or not).''
-% https://github.com/geodynamics/specfem3d
-
-% arpack-ng - Fortran 77 routines designed to solve large scaled
-% eigenvalue problems - https://github.com/opencollab/arpack-ng
-
-To study the effectiveness of our approach, 
-we built a corpus of around 1 million lines of Fortran code from a
-range of scientific computing packages: The Unified Model (UM)~\cite{um},
-E3MG~\cite{RePEc:aen:journl:2006se-a12}, BLAS~\cite{blas},
-Hybrid4~\cite{GBC:GBC635}, GEOS-Chem~\cite{geos-chem}, Navier (based
-on \cite{griebel1997numerical}), Computational Physics
-2~\cite{nicholas2006computational},
-ARPACK-NG~\cite{arpackng}, and
-SPECFEM3D~\cite{specfem3d}.
-
-We first examined how frequently stencil computations occur. We parsed
-959,427 lines of Fortran code and found that 10\% (97,439) of
-statements have a left-hand side as an array subscript on
-neighbourhood indices. This supports the idea that stencil-like
-computations are common in scientific code.
-We then used the inference procedure of the previous section to
-generate specifications for stencils in the corpus to assess
-the design of the language. %  This assumes that the programs have been implemented
-%correctly---
-%Normally one would use our approach in verification mode to
-%check the correctness of new code.
-
-We would not expect to infer a stencil for each array statement we
-found because our analysis restricts 
-the array-subscript-statements that we classify as a
-stencil. In fact, we were able to infer a stencil from 30\% of
-these statements. A single statement can involve multiple arrays and
-we ended up with 60,525 specifications. This shows that we can express
-a large number of stencil shapes within our high-level abstraction and
-validates our initial hypothesis that many stencil computations have a
-regular shape.
-
-%    numStencilSpecs: 60525
-% 4338 files, linesTotal: 1357377, linesParsed: 959427
-% Number of stencils (tickAssign) / number of loops
-% Percentage of a code base that is taken up by stencils
-%Out of 1.35 million lines of code that we fed into our program, and
-%the 959,427 lines that we were able to parse, we identified 60,525
-%cases that fit our criteria for stencil specification.
-
-%\paragraph{Hypothesis 2: Most stencil computations
-%have a regular pattern which can be captured with a simple
-%language of contiguous regions}
-
-% Number of stencils we give a spec to (where a spec declaration to n
-% variables is counted n times) as percentage of number of stencils
-
-    % tickAssign: 97439
-    % tickAssignSuccess: 29402 % <- number of times tickAssign is accompanied by a successful stencil inference
-                               % counts only by +1 each time that tickAssign finds a stencil
-
-%In the analysed body of code we found that we were able to deduce
-%some form of stencil specification from approximately 30\% of the
-%potential stencils examined.
-
-%
-%\subsection{Analysing the design of our specifications}
-
-% 1. How may are just reflexive and how many some reflexive
-% justReflexive: 55504
-% someReflexive: 15823
-
-The majority of specifications generated were relatively simple but we
-found significant numbers of more complex shapes. We grouped common
-patterns in to categories:
-
-\textbf{All pointed} 55,504 of the stencils we found involved 
-only pointed regions. 39,681 of these were pointed in all
-dimensions. Common examples of this were pointwise transformations on
-data (such as scaling).
-
-\textbf{Single-action} specifications comprise one 
-forward, backward, or centered region constant combined via \term{+} 
-or \term{*} with any number of pointed regions. We identified 4,837 single-action
-specifications, of which 1,532 were single-action with a
-\texttt{nonpointed} modifier.
-
-\textbf{Multi-action} specifications comprise at least two 
- forward, backward, or centered regions, combined with
-any number of pointed regions. We identified 265 multi-action
-specifications out of which 207 had regions combined only with
-$\term{*}$ and 58 combined with a mix of $\term{*}$ and \term{+}.
-
-% multiActionRegionOps1: 31
-% multiActionRegionOps2: 115
-% multiActionRegionOps8: 1
-% multiActionRegionOps14: 5
-% multiActionRegionOps17: 4
-
-% %%%%
-% Complexity-wise, out of these 156 multi-action specifications, 31 of
-% them used a single region operator, 115 used two region operators, 1
-% used eight region operators, 5 used fourteen region operators, and 4
-% used seventeen region operators. Some of these very complex stencils
-% were found in the atmospheric models of the UM.
-% %%%%
-
-\textbf{Bounded} specifications occured with 157
-\texttt{atMost} bounds and 36 \texttt{atLeast}, the latter of which were
-always also paired with an upper bound.
-
-\begin{figure}[t]\begin{minted}[fontsize=\scriptsize,breakindent=0em,linenos=false,xleftmargin=0em,breakafter=)]{fortran}
-!=stencil readOnce,(forward(depth=1,dim=3,nonpointed))*(backward(depth=1,dim=1))*(backward(depth=1,dim=2))+(forward(depth=1,dim=3))*(backward(depth=1,dim=1))*(pointed(dim=2))+(forward(depth=1,dim=3))*(backward(depth=1,dim=1,nonpointed))*(backward(depth=1,dim=2,nonpointed))+(forward(depth=1,dim=3))*(backward(depth=1,dim=2))*(pointed(dim=1))+(backward(depth=1,dim=1))*(backward(depth=1,dim=2,nonpointed))*(pointed(dim=3))+(backward(depth=1,dim=1,nonpointed))*(backward(depth=1,dim=2))*(pointed(dim=3))::x
-\end{minted}
-\caption{Complex specification inferred from
-  \textbf{UM}\label{fig:smagorinsky}}
-\vspace{-1em}
-\end{figure}
-
-The single- and multi-action classes represent more complex stencils
-with a real possibility for programmer error. As an extreme example, the Unified
-Model has an implementation of the Smagorinsky subgrid-scale model for
-calculating turbulence on which our inference yields 39 specs from 340
-lines of code. This is an large reduction given
-the complexity of the algorithm.  We show one example
-in~\Cref{fig:smagorinsky} which specifies the access pattern to
-a 3-dimensional array (which we have renamed to 
-\texttt{x}) arising from a kernel of 93 lines of code involving 142
-array subscripts. The specification was the most complex
-seen in our corpus, yet it still represents a significant abstraction
-of the spatial behaviour given size and complexity of the kernel it describes.
-
-We measured the frequency at which individual specifications involved 
-multiple occurences of \term{+} and \term{*}:
-
-\vspace{0.25em}
-\setlength{\tabcolsep}{0.57em}
-{\small{
-\hspace{-1em}\begin{tabular}{c|cccccccc}
-& 0 & 1 & 2 & 3 & 4 & 5 & 6 & $> 7$ \\ \hline
-\term{+} & 60,265 & 220 & 15 & 12 & 5 & 4 & 4 & 0 \\
-\term{*} & 31,393 & 15,151 & 13,457 & 424 & 41 & 42 & 5 & 12
-\end{tabular}}}
-\vspace{0.2em}
-
-\noindent
-\eg{}, roughly half the specifications did not involve $\term{*}$, 
-a quarter use one $\term{*}$ and just under a quarter
-use two $\term{*}$ operators.
-
-\paragraph{Limitations}
-
-There were various reasons why we did not infer 
-specifications on every looped array computation: \\
-\noindent 
-1) \textbf{Non-subset induction variables} occur when the
-induction variables on the RHS are not a subset of those in the LHS. These
-cases are not stencils by our definition. The degenerate case of this
-is to have only constant indices on the LHS. We see lots of examples
-of this in loops as accumulators \eg{} computing the sum over an array; \\
-\noindent
-2) \textbf{Derived induction variables} where an
-index (\mintinline{fortran}{x}) is derived from an
-induction variable (\mintinline{fortran}{i}) as in
-\mintinline{fortran}{x = len - i};  \\
-\noindent
-3) \textbf{Inconsistent induction dimensions} occur when
-an induction variable is used to specify more than one array dimension
-on the RHS or multiple induction variables are used for the same
-dimension on the RHS. These are common in matrix operations such as
-LU-decomposition with assignments such as
-\mintinline{fortran}{a(l) = a(l) - a(m) * b(l, m)}.
-
-\subsection{Detecting errors in the 2-D Jacobi iteration}
-\label{sec:jacobi}
-One common example of a stencil computation is the two-dimensional
-Jacobi iteration that repeatedly goes through each cell in a matrix
-and computes the average value of the four adjacent cells. The kernel
-is given by:
-\begin{minted}{fortran}
-  a(i,j) = (a(i-1,j)+a(i+1,j)+a(i,j+1)+a(i,j-1))/4
-\end{minted}
-We infer a precise specification of its shape as:
-\begin{minted}[breakafter=+:,breakindent=-0.6em,breaksymbolsep=0.4em,linenos=false,xleftmargin=-0.5em]{fortran}
-  != stencil pointed(dim=1)*centered(depth=1,dim=2,nonpointed)+pointed(dim=2)*centered(depth=1,dim=1,nonpointed)::a
-\end{minted}
-%
-To test our implementation, 
-we examined whether programmer errors would be detected by replacing
-the array index offsets with $-1$, $0$, or $1$ and running our
-verification algorithm. Our checking procedure correctly reported a verification
-failure in each of 6,537 permutations corresponding to an error.  
-The iteration computes the average of four adjacent
-cells so $24$ ($4$ factorial) of the possible array index perturbations
-are correct, all of which are accepted by our checker.
-
-% /um/trunk/src/atmosphere/diffusion_and_filtering/turb_smagorinsky.F90
-% Tried introducing indexing errors replacing one of the indices of cx_rho_instance with -1,0,+1
-% 42 cases were caught
-% 6 cases were not caught
-
-% (/ 55504.0 60525) = 91.7%
-% (/ 15823.0 60525) = 26.1%
-% 91.7\% of the specifications involved only reflexive regions. 26.1\%
-
-% 2. Break down of "interesting specifications" (contain more than
-% just reflexive)
-%   - Single-action, histogram on depths
-%   - Single-action + irreflexives, histogram on depths
-
-% singleAction: 4837
-% singleActionIrr: 1532
-
-%   - Multi-action combined only by *
-%   - Multi-action combined also by +, histogram of tree depths
-
-    % multiAction: 156
-    % multiActionMulOnly: 146
-
-%   - Bounded specs (how many lower, upper, or both)
-
-    % atLeast: 36
-    % atMost: 157
-    % boundedBoth: 36
-
-% Present not just percentages but concrete numbers
-
-\section{Related work and conclusions}
-\label{sec:discussion}
-
-Various deductive verification tools can express array indexing
-in their specifications, \eg{}, ACSL of \citet{baudin2008acsl}
-for C (\eg{}~\citet[Example 3.4.1]{burghardt2010acsl}). A specification
-can be given for a stencil computation but must use fine-grained
-indexing as in code and therefore is similarly prone to indexing
-errors. Our approach is much more abstract-- it does
-not aim to reify indexing in the specification, but
-provides simple spatial descriptions which capture a large number
-of common patterns.
-
-\citet{kamil2016verified} propose \emph{verified lifting} to extract a
-functionally-complete mathematical representation of low-level, and
-potentially optimised, stencils in Fortran code. This extracted
-predicate representation of a stencil is used to generate code
-for the \textsc{Halide} high-performance
-compiler~\citep{ragan2013halide}. Thus they must capture the
-full meaning of a stencil computation which requires significant
-program analysis. For example, they report that some degenerate
-stencil kernels take up to 17 hours to analyse and others require
-programmer intervention for correct invariants to be inferred.
-
-% ACR-I think we could skip this in the interests of space
-%This led them to develop \textsf{STNG}, a loop invariant and postcondition
-%finder through syntax-guided synthesis. Using this approach they restrict the
-%search space of postconditions to predicates of a certain form, one that is
-%compatible with \textsf{Halide}. They then look at concrete loop iterations and
-%form an hypothesis about the invariant, which is later confirmed or put back
-%into the system using a SMT solver.
-
-Our approach differs significantly. Rather than 
-full representation of stencils, we focus on specifying
-just the spatial behaviour in a lightweight way.
-Thus, it suffices for us to perform a comparatively
-simple data-flow analysis which is efficient, scales linearly with
-code size, and does not require any user intervention.
-Whilst we do not perform deep semantic analysis of stencils,
- the analysis part of our
-approach can be made arbitrarily more sophisticated independent of the rest of
-the work.
-% Hence, we do not require SMT solving or search of loop invariants. It
-%suffices to do comparatively simple data-flow analysis.
-%Since CamFort mostly does
-%syntactic analysis, the inference procedure terminates quickly and
-%never requires programmer intervention. 
-%The downside of our
-%approach is that if optimisation causes the stencil computation to be heavily
-%obfuscated, \textsf{STNG} would capture the access behaviour better.
-%
-Furthermore, Kamil \emph{et al.} do not provide a user-visible
-syntactic representation of their specifications, and nor do
-they provide verification from specifications \eg{}, to future-proof
-the code against later changes. Even if they were to provide a
-syntactic representation, for complex stencils such as Navier-Stokes
-from \cref{sec:intro}, it would be as verbose as the code itself,
-making it difficult for programmer to understand the overall shape of
-the indexing behaviour.
-
-% GPGPU
-Our work has similarities with efforts to verify kernels
-written for General-Purpose GPU
-programming, such as in \citet{Blom:2014:SoCP}. %Stencils are a form of
-%kernel, and GPGPU programming can be viewed as a massively parallel
-%method of transforming a large matrix. 
-However, their focus is
-mainly on the synchronisation of kernels and the avoidance of
-data races., while we are interested in correctness 
-%embedded within a more typical general-purpose programming
-%language. %Other work on GPGPU computation, such as
-%\citet{Zhang:2012:CGO}, has focused primarily on generating
-%optimised code based on relatively simple specifications: to
-%provide performance while keeping the programmer's effort within
-%reason.
-%
-% Sketching Stencils
-\citet{Solar-Lezama:2007:PLDI} give specifications of stencils using
-unoptimised ``reference'' stencils, coupled with partial
-implementations which are completed by a code
-generation tool. %These kinds of specifications are just simple
-%implementations, so this tool is useful for hand-written, optimised
-%stencils. 
-The primary purpose of this tool is optimisation rather than
-correctness, and the language of specification is 
-more elaborate than ours.
-
-% Pochoir
-\citet{Tang:2011:SPAA} define a specification language for writing stencils
-embedded in C++ (with Cilk~\citep{blumofe1996cilk} extensions) that are then
-compiled into parallel programs based on trapezoidal decompositions with
-hyperspace cuts. Pochoir specifications are used for describing the kernel,
-boundary conditions, and shape of the stencil. Pochoir is aimed at programmers
-reluctant to implement the high-performance cache-oblivious
-``hypertrapezoidal'' algorithms.  Like much of the related work,
-the goal is optimisation rather than verifying program correctness.
-
-By contrast, the work of \citet{Abe:2013:IPDPSW} studies
-correctness bringing a form of model-checking to
-verify certain stencil computations in the context of
-parallelism 
- in partitioned global address
-space languages. %These are scenarios where an array is divided into
-%subarrays on multiple processors but each has the ability to access
-%the others' memory. The authors avoid the state explosion dilemma of
-%model-checking by relying upon the fact that accesses to the
-%``boundary elements'' will require a different method by virtue of the
-%non-locality of that memory. Since this access is performed in a
-%different manner, those can be identified automatically by static
-%analysis. To achieve this, 
-Abe \emph{et al.} provide a new language for writing stencil computations. Much of the specification
-effort goes towards describing the distribution of the
-computation over multiple processors. The code for the stencil kernel
-is generated from a relatively high-level specification. 
-In contrast, we integrate directly
-into existing, legacy codebases and established languages, bringing
-the benefits of verification more easily to scientific computing.%such as
-%older versions of Fortran.% We infer stencil specifications from
-%Fortran code and check annotations on stencil computations within
-%Fortran code.
-
-\paragraph{Concluding remarks}
-
-There is an increasing awareness of the need for verification
-techniques in
-science~\cite{post2005computational,oberkampf2010verification,orchard2014computational}.
-Our specification language is an approach in this direction, providing
-a system of lightweight, high-level specification for numerical code.
-As opposed to existing approaches our language takes inspiration from
-the numerical literature and provides a concise, abstract description
-of the stencil.
-
-We evaluated our approach by extending CamFort, an open-source analysis
-tool for Fortran programs~\cite{camfort}, running the inference process over a corpus
-of a million lines of Fortran code. We showed that our language is
-capable of capturing many real-world uses of stencils and showed in a
-case-study that it will detect (simulated) programmer errors.
-
-Our restrictive definition of stencils means we do note generate
-specifications for a large number of iterated array accesses in the
-corpus, \eg{} that incorporate reductions. For
-future work, we will look at expanding the flexibility of 
-our analysis to include these examples.
-
-\bibliography{references}
-
-\end{document}
-
-%%  LocalWords:  refactoring affine parameterised nonpointed atMost
-%%  LocalWords:  centered readOnce atLeast discretisation Equational
-%%  LocalWords:  equational disjunction denotational dimensionality
-%%  LocalWords:  interprocedural Fortran CamFort preprocessor
-%%  LocalWords:  committers
+\item \textbf{Covering}

### 2017-04-16 oopsla17/algorithms.tex
moving checking/analysis into separate file

@@
 semantic interpretation of the code.
+
+\subsection{Union Normal Form}\label{subsec:union-normal-form}
+
+We need to address the problem of representation in the model both for observed
+array terms converted into subscripts schemes and for specifications converted
+into interval schemes. In each case, the objects are sets and union and
+intersection are the combinators. However, they are often infinite sets due to
+infinite interval and constant indices being converted to \bz{}. This calls for
+a finite (and ideally compact) representation for indices.
+
+First recall \Cref{lem:vector-intersect}, which states that intersection of
+indexing schemes is still an indexing scheme. When combined with
+\Cref{lem:interval-identities}, we can extend this closure to interval schemes.
+Subscripts schemes do not enjoy the same propery as intersection of two
+singleton sets might easily be empty, which disqualifies it from being a
+subscript index. In the following section we will not rely on intersection of
+subscript indices, so that is okay.
+
+Union is, however, not as flexible as intersection. \Cref{lem:vector-union} does
+not form a closure for indexing schemes in general. For this reason, we
+represent collection of all forms of indexing schemes as unevaluated unions.
+Since the model forms a distributive lattice as established in
+\Cref{prop:region-lattice}, the intersections can be pushed in using the
+distributive law leaving only unions. We further exploit associativity of union
+to put unions into a cons tree making the unions effectively a non-empty list.
+We do not, however, attempt to canonicalise the representation.
 
@@
 \label{subsec:checking}
+
+In the model, we can regard both array indices and specifications as groups of
+points in $N$-dimensional space. Notion of consistency is then intuitively
+whether these two (potentially infinite) sets of points in space match or not.
+This leads to a simple set based consistency function defined as
+%
+\begin{align*}
+  \mathit{consistent} & :
+    \mathsf{Mult}(\mathcal{P}(\mathcal{P}(\bz{})^N)) \times
+    \mathsf{Mult}(\mathsf{Approx}(\region{N})) \to \mathbb{B} & \\
+%
+  \mathit{consistent}(\mathit{ix}, \mathit{model}) & = \begin{cases}
+    \mathsf{false} & \mathit{model} = \mathsf{once}(x) \wedge ix = \mathsf{mult}(y) \\
+    \mathit{consistent'}(\mathit{peel}(ix), \mathit{peel}(model)) & \textit{otherwise}
+  \end{cases} \\
+%
+  \mathit{consistent'} & :
+    \mathcal{P}(\mathcal{P}(\bz{})^N) \times
+    \mathsf{Approx}(\region{N}) \to \mathbb{B} & \\
+%
+  \mathit{consistent'}(\mathit{ix}, \mathit{model}) & = \begin{cases}
+    m = \mathit{ix} & \mathit{model} = \mathsf{exact}(m) \\
+    m \supseteq \mathit{ix} & \mathit{model} = \mathsf{upper}(m) \\
+    m \subseteq \mathit{ix} & \mathit{model} = \mathsf{lower}(m)
+  \end{cases}
+\end{align*}
+%
+The first function, $\mathit{consistent}$, checks whether the linearity of the
+specification matches that of the indices, \ie{} if the specification allows
+indices to be repeated or not.Then it delegates to $\mathit{consistent'}$
+function which checks if the points observed in the array expression matches
+the space defined by the specification. Lower bounds, marked by \texttt{atLeast}
+modifier, require the space defined by the specification to remain inside the
+set of indices, while an upper bound, market by \texttt{atMost} modifier,
+requires the opposite: enclosure. In the absence of such modifiers, we expect
+the space defined by observed indices correspond exactly those defined by the
+region in the specification, hence requiring set equality.
+
+Reader will realise that since index schemes allows infinite intervals in the
+product, set equality cannot be done by exhaustively checking the points of one
+are inside the other. Instead, we compile the region into interval constraints
+and subscript schemes into membership constraints and pass it to
+\textsc{Z3}~\citep{de2008z3} SMT solver to see if they are equal. The query
+belongs to to quantifier free linear arithmetic, which is a decidable theory.
+Performance of consistency checking using SMT solver is irrelevant as the number
+of interval constraints and the number of indices that can reasonably appear in
+an assignment is too small restricting the size of the formula being checked.
+
+\mnote{It would be great if we had data on number of indexing expressions that
+occur (counting the ones that flow in) and also the length of specifications
+(which I think we do generate). This would justify my efficiency claims}.
 

### 2017-04-16 oopsla17/algorithms.tex
fix enumerate

@@
 \begin{enumerate}[leftmargin=1.5em]
+\item
 A covering of (possibly overlapping) regions is computed iteratively
@@
 \begin{enumerate}
-
 \item \textbf{Coalesce intervals}
@@
 
-\begin{defn}{(Adjacent)}
+\begin{defn}{[Contiguous]}
+  An index scheme
   A subscript $S$ is said to be \emph{adjacent at index $j$} of another
@@
 %
-
-\begin{definition}[Contiguous]
-
-\end{definition}
-
 \item \textbf{Remove completely overlapped intervals}
-
 \item \textbf{Covering}
+\end{enumerate}
+\end{enumerate}

### 2017-04-17 oopsla17/algorithms.tex
on inference part 1

@@
-\begin{equation*}
-\texttt{a} \mapsto \{[0,0]\times[0,0], \;\; [-1,0]\times[0,0], \;\;
-[0,0]\times[-1,0], \;\; [0,0]\times[1,0], \;\; [0,1]\times[0,0]\}
-\end{equation*}
 %
@@
 \label{subsec:inference}
-
+%
 We provide an inference procedure for generating specifications from
@@
 
-The preceding source code analysis converts the concrete syntax of array
-subscripts into a subscript space in $\mathcal{P}(\bz)^N$.
-Let $U$ range over finite maps from array variables to sets of
-of subscript schemes.
+The preceding source code analysis converts the concrete syntax of
+array subscripts into a set of subscript schemes in
+$\mathcal{P}(\bz)^N$. These subscripts are then coalesced interatively
+into sets of index schemes. We denote sets of index schemes by
+$\mathcal{S}$, and initial set of subscript schemes from the analysis
+as $\mathcal{S}_0$. For our example:
+%
+\begin{equation*}
+\mathcal{S}_0 = \{\{0\} \times \{0\}, \{-1\} \times \{0\},
+\{0\} \times \{-1\}, \{0\} \times \{1\}, \{1\} \times \{0\}\}
+\end{equation*}
 
- The
-inference then attempts to construct a $\emph{Region}_N$ from this.
-
- For
-each array variable $a \in \mathsf{dom}(U)$, the algorithm constructs a set of
-$n$-dimensional rectangles covering all contiguous groups of schemes
-in $U(a)$.
+%For each array variable $a \in \mathsf{dom}(U)$, the algorithm
+%constructs a set of $n$-dimensional rectangles covering all contiguous
+%groups of schemes in $U(a)$.
 
 \begin{enumerate}[leftmargin=1.5em]
-\item
-A covering of (possibly overlapping) regions is computed iteratively
+\item (\textbf{Covering})
+A covering of (possibly overlapping) the subscript space
+is calculated as a value of $Region_N$. This is computed iteratively
 by the following steps until a fixed-point is reached:
 
-\begin{enumerate}
-\item \textbf{Coalesce intervals}
-Intervals which are adjacent (contiguous in one dimension) are
+\begin{enumerate}[leftmargin=0.5em]
+\item (\textbf{Coalesce intervals})
+%
+Index schemes which are \emph{contiguous} in one dimension are
 coalesced.
 
+\newcommand{\contig}[2]{\mathsf{contig}(#1, #2)}
 \begin{defn}{[Contiguous]}
-  An index scheme
-  A subscript $S$ is said to be \emph{adjacent at index $j$} of another
-  subscript $S'$ if for every index $i$ except $j$, $\pi_i(S) = \pi_i(S')$ but
-  at $j$ then $\pi_j(S) \neq \pi_j(S')$ such that $|S| = |S'| + 1$.
+Two index schemes $S$ and $T$ are \emph{contiguous}
+written $\contig{S}{T}$,
+iff $\pi_i(S) = \pi_i(T)$
+for all $1 \leq i \leq N$ apart from some dimension $k$ such that
+$\pi_k(S) = [a, b]$ and $\pi_k(T) = [b+1, c]$.
 \end{defn}
 %
-\item \textbf{Remove completely overlapped intervals}
-\item \textbf{Covering}
+A coalesced set of index schemes is computed by:%
+\begin{align*}
+\mathsf{coalesce}(\mathcal{S}) = \; &
+\{\,S \cup T \mid
+\forall S \in \mathcal{S}, \;
+\forall T \in \mathcal{S}. \;
+\contig{S}{T} \} \\
+\cup \; &
+\{ \,\;\;\; S \;\;\; \mid \forall S \in \mathcal{S},
+\; \not \exists T \in \mathcal{S} . \; \contig{S}{T}\}
+\end{align*}
+%
+That is, the new set of index schemes $\mathsf{coalesc}(\mathcal{S})$
+comprises the union of the set of coalesced pairs of contiguous
+schemes and the set of index schemes for which there is no other
+contiguous scheme.
+\item (\textbf{Remove redundant schemes})
+Any indexing schemes which are subsumed by another are filtered out:
+%
+\begin{align*}
+\textsf{nonOverlap}(\mathcal{S}) =
+\{S \mid \forall S \in \mathcal{S}, \; \not \exists T \in (\mathcal{S} -
+  \{S\}), \; \forall i . \; \pi_i(S) \subseteq \pi_i(T) \}
+\end{align*}
+%
+\dnote{I'm starting to suspect that this step is actually redundant
+  now- but I will need to do some math to check}
+%\item \textbf{Covering}
 \end{enumerate}
+The least fixed-point of $\textsf{nonOverlap} \circ \textsf{coalesce}$
+is computed, providing a minimal indexing scheme covering the space.
+\dnote{Proof?}
+
+For our example:
+%
+\begin{align*}
+\textsf{coalesce}(\mathcal{S}_0) & =
+\{[-1,0] \times [0,0],\,[0,0] \times [-1, 0],\,[0,0] \times [0,1],\,[0,1]
+  \times [0,0]\} \\
+%\textsf{nonOverlap}(\textsf{coalesce}(\mathcal{S}_0)) & =
+%                                                        \textsf{coalesce}(\mathcal{S}_0)
+%  \\
+\textsf{coalesce}^2(\mathcal{S}_0) & =
+\{[-1, 1] \times [0, 0],\,[0, 0] \times [-1, 1]\}
+\end{align*}
+%
+The fixed-pointed is reached within two applications of
+\textsf{coalesce} for our example.
+\dnote{We probably need to say something about classifying index
+schemes are \textit{Region$_N$}?}
+\item (\textbf{Index schemes to syntax}) \dnote{Mistral todo}
 \end{enumerate}

### 2017-04-17 oopsla17/oopsla17.tex
moving checking/analysis into separate file
on inference part 1

--- 2017-04-15 oopsla17/oopsla17.tex
@@
 
-  We also add the distinguished interval $[-\infty, \infty]$, which is simply an
+  We also add the distinguished interval $\interv{-\infty}{\infty}{}$, which is simply an
   alias for \bz{}, but this notation prevents handling infinite interval
@@
       \; \vee \;
-      \pi_i (S) = [-\infty, \infty]
+      \pi_i (S) = \interv{-\infty}{\infty}{}
     \end{equation*}
@@
 \section{Analysis, Checking, and Inference}
+\label{sec:algorithms}
+\input{algorithms}
 
 \section{Evaluation}
+\label{sec:evaluation}
 

### 2017-04-17 oopsla17/introduction.tex
fix to coalesce

--- 2017-04-17 oopsla17/oopsla17.tex
@@
-\documentclass[acmlarge,review,anonymous]{acmart}\settopmatter{printfolios=true}
-
-%% Some recommended packages.
-\usepackage{booktabs}   %% For formal tables:
-                        %% http://ctan.org/pkg/booktabs
-\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
-                        %% http://ctan.org/pkg/subcaption
-
-\usepackage{hyperref}
-\usepackage{natbib}
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{amsthm}
-\usepackage{syntax}
-\usepackage{indentfirst}
-\usepackage{cleveref}
-\usepackage{xcolor}
-\usepackage{multirow}
-\usepackage{siunitx} % For pretty-printing numeric values and SI units
-                     % of measure. e.g., the tabular column type S is
-                     % used to print nice-looking tables of numbers.
-\sisetup{ % defaults
-  group-separator={,},
-  group-minimum-digits={3},
-  output-decimal-marker={.},
-  table-format = 6
-}
-
-\usepackage{minted}
-\usepackage{etoolbox}
-\usepackage{enumitem}
-\usepackage{thmtools}
-\usepackage{thm-restate}
-
-\newcommand{\fortran}[1]{\mintinline{fortran}{#1}}
-
-\errorcontextlines 10000
-\usepackage{stencilmacros}
-
-%% Bibliography style
-\bibliographystyle{ACM-Reference-Format}
-%% Citation style
-%% Note: author/year citations are required for papers published as an
-%% issue of PACMPL.
-\citestyle{acmauthoryear}   %% For author/year citations
-
-\fvset{
-  linenos=true,
-  fontsize=\footnotesize,
-  breaklines=true,
-  breakafter=w),
-  xleftmargin=3\parindent,
-  numbersep=1em
-}
-
-\usemintedstyle{vs}
-
-\setlength{\grammarindent}{3em} % increase separation between LHS/RHS
-
-\newcommand{\inbar}{\ \ \textbar\ \ }
-
-% For evluation and empirical study
-% Defines the commands (headers)
-\input{resultsCmds}
-\input{studyResults}
-% renews the commands with actual data
-% \input{results}
-
-\definecolor{darkgreen}{rgb}{0.0,0.5,0.0}
-\definecolor{darkpurple}{rgb}{0.6,0.0,0.6}
-\definecolor{orange}{rgb}{0.8,0.4,0.0}
-\newcommand{\dnote}[1]{\textcolor{darkpurple}{Dom: #1}}
-\newcommand{\mnote}[1]{\textcolor{darkgreen}{Mistral: #1}}
-\newcommand{\anote}[1]{\textcolor{red}{Andy: #1}}
-
-\theoremstyle{definition}
-\newtheorem{defn}{Definition}
-
-\theoremstyle{plain}
-\newtheorem{thm}{Theorem}
-\newtheorem{lem}{Lemma}
-\newtheorem{prop}{Proposition}
-\newtheorem{case}{Case}
-
-\theoremstyle{remark}
-\newtheorem{remark}{Remark}
-
-\title{Verifying Spatial Properties of Array Computations}
-
-\author{Dominic Orchard}
-\affiliation{
-  \department{School of Computing}
-  \institution{University of Kent}
-}
-\email{d.a.orchard@kent.ac.uk}
-
-\author{Mistral Contrastin}
-\affiliation{
-  \department{Computer Laboratory}
-  \institution{University of Cambridge}
-}
-\email{Mistral.Contrastin@cl.cam.ac.uk}
-
-\author{Matthew Danish}
-\affiliation{
-  \department{Computer Laboratory}
-  \institution{University of Cambridge}
-}
-\email{Matthew.Danish@cl.cam.ac.uk}
-
-\author{Andrew Rice}
-\affiliation{
-  \department{Computer Laboratory}
-  \institution{University of Cambridge}
-}
-\email{Andrew.Rice@cl.cam.ac.uk}
-
-\begin{abstract}
-  Arrays computations are the workhorse of numerical modelling and
-  computational science. However, low-level manipulation of array
-  indices is a source of program error.  Many practitioners are aware
-  of the need to ensure program correctness, yet very few of the
-  techniques from the programming research community are applied by
-  scientists. We aim to change that by providing targetted lightweight
-  verification techniques for scientific code.  We focus on the all
-  too common mistake of array offset errors as a generalisation of
-  off-by-one errors.  Firstly, we report on a code analysis study on
-  twelve real-world computational science code base, identifying
-  common idioms of array usage and their spatial properties. This
-  provides much needed data on array programming idioms common in
-  scientific code.  From this data, we designed a lightweight
-  declarative specification language capturing the majority of array
-  access patterns via a small set of combinators. We detail a semantic
-  model, and the design and implementation of a verification tool for
-  our specification language, which both checks and infers
-  specifications.  We evaluate our tool on our corpus of scientific
-  code and give verification case studies of bug fixes that are
-  detected by our approach. We found roughly 60,000 targets for
-  specification across roughly 1.2 million lines of code, showing the
-  vast majority of array computations read from arrays in a pattern with
-  a simple, regular, static shape.
-\end{abstract}
-
-%  We focus on one error-prone aspect of numerical code: the 
-%  data access pattern of arrays.
-
-\begin{document}
-\maketitle
-
-\section{Introduction}
-
 \paragraph{Context of this research}
@@
 
-\section{An Empirical Study of Array Computations in Scientific Fortran code}
-\label{sec:study}
-\input{study}
-
-\section{A Specification Language for Shape}
-
-\begin{figure}[t]
-\begin{align*}
-\def\arraystretch{1}
-\setlength{\arraycolsep}{0.2em}
-\newcommand{\dimTy}{$\mathbb{N}_{>0}$}
-\newcommand{\depthRange}{$\mathbb{N}_{>0}$}
-\begin{array}{rl}
-\nonterm{specification} ::= & \nonterm{regionDec} \mid \nonterm{specDec} \\
-\nonterm{specDec} ::= & \term{stencil} \; \nonterm{spec} \;
-                        \texttt{::} \; v \\
-\nonterm{regionDec} ::= &  \texttt{region} \; \texttt{::} \; \nonterm{rvar} \; \texttt{=} \;
-                         \nonterm{region}\\[0.4em]
-%\nonterm{spec} ::= & \nonterm{spatial} \mid \nonterm{temporal}
-%\\[1em]
-\nonterm{spec} ::= & [\nonterm{mult},] \; [\nonterm{approx},] \; \nonterm{region} \\
-\nonterm{mult} ::= & \term{readOnce} \\
-\nonterm{approx} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
-\nonterm{region} ::= & \nonterm{rvar} \; \mid \;
-                       \stencil{p}{\dimTy}{}{} \; \\
-& {\qquad\; \mid \; \stencil{f}{\dimTy}{\depthRange}{\;[, \nonpointed]}} \\
-& {\qquad\; \mid \; \stencil{b}{\dimTy}{\depthRange}{\;[, \nonpointed]}} \\
-& {\qquad\; \mid \; \stencil{c}{\dimTy}{\depthRange}{\;[, \nonpointed]}} \\
-& {\qquad\; \mid \; \nonterm{region} \, \term{+}
-  \, \nonterm{region} \; \mid \; \nonterm{region} \; \term{*} \; \nonterm{region}} \\[0.25em]
-%\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{rvar}}
-%\\[0.5em]
-%\nonterm{temporal} ::= \; & \term{dependency} \; (v \; \{ , v \}) [, \texttt{mutual}]
-%  \\[0.5em]
-\nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+\\[-1em]
-\end{array}
-\end{align*}
-\caption{Specification syntax (EBNF grammar)}
-\label{fig:syntax}
-\end{figure}
-
-\subsection{Specification syntax}
-\label{subsec:syntax}
-
-\Cref{fig:syntax} gives the syntax of stencil specifications, which is
-detailed below. The entry point is the \nonterm{specification} production which
-splits into either a \emph{region declaration} or a \emph{specification
-declaration}.
-Regions comprise \emph{region constants} which are
-combined via region operators \term{+} and \term{*}.
-
-Region constants specify a finite interval in a 
-single dimension relative to the origin and are either \term{pointed},
-\term{forward}, \term{backward}, or \term{centered}. The region names are
-inspired by numerical analysis terminology, \eg{} the standard explicit method for
-approximating PDEs is known as the \emph{Forward Time, Centered Space} (FTCS)
-scheme~\citep{dawson1991finite}.
-
-Each region
-constant has a dimension identifier $d$ given as a positive natural number.
-Each constant except \term{pointed} has a depth
-parameter $n$ given as a positive natural number; \term{pointed}
-regions implicitly have a depth of $0$.
-
-A \term{forward} region of depth $n$ specifies a contiguous
-region in dimension $d$ starting at the origin. This corresponds
-to specifying neighbourhood indices in dimension $d$ ranging from $i$ to $i + n$
-for some induction variable $i$. Similarly, a
-\term{backward} region of depth $n$ corresponds to contiguous indices
-from $i$ to $i - n$ and \term{centered}
-of depth $n$ from $i - n$ to $i + n$. A \term{pointed}
-stencil specifies a neighbour index $i$. For example, the
-following shows four specifications with four consistent stencil
-kernels reading from arrays \term{a}, \term{b}, \term{c} and \term{d}:
-%%
-\begin{minted}{fortran}
-!= stencil forward(depth=2, dim=1) :: a
-e(i, 0) = a(i, 0) + a(i+1, 0) + a(i+2, 0)
-
-!= stencil backward(depth=2, dim=1) :: b
-e(i, j) = b(i, j) + b(i-1, j) + b(i-2, j)
-
-!= stencil centered(depth=1, dim=1) :: c
-e(i, j) = (c(j-1) + c(j) + c(j+1))/3.0
-
-!= stencil pointed(dim=3) :: d
-e(i, j) = d(0, 0, i)
-\end{minted}
-%%
-Not every dimension needs to be specified, \eg{},
-specifications on lines $1$, $4$, and $10$ leave some dimensions unspecified.
-The \term{forward}, \term{backward}, and \term{centered} regions may
-all have an additional attribute \term{nonpointed} which marks absence
-of the origin.  For example, the following is a
-\term{nonpointed} \term{backward} stencil
-%
-\begin{minted}{fortran}
-!= stencil backward(depth=2, dim=1, nonpointed) :: a
-b(i) = a(i-1) + 10*a(i-2)
-\end{minted}
-%
-Not every dimension needs to be specified, \eg{},
-specifications on lines $1$, $3$, and $7$ leave some dimensions
-unspecified which leaves indices in these dimensions unconstrained.
-
-The \term{forward}, \term{backward}, and \term{centered} regions may
-all have an additional attribute \term{nonpointed} which marks absence
-of the origin.  For example, the following is a
-\term{nonpointed} \term{backward} stencil
-%
-\begin{minted}{fortran}
-!= stencil backward(depth=2, dim=1, nonpointed) :: a
-b(i) = a(i-1) + 10*a(i-2)
-\end{minted}
-
-\paragraph{Combining regions}
-
-The region operators \term{+} and \term{*} respectively combine
-regions by union and intersection. The intersection of two regions
-$r \term{*} s$ means that any indices in the specified code must be
-consistent with both $r$ and $s$ simultaneously.
- Dually, for the union of two regions
- $r \term{+} s$ means that indices in the specified code must be
- in consistent with one of $r$ or $s$, or both.
-For example, the following \emph{nine-point stencil}
-has a specification given by the product of two \texttt{centered}
-regions in each dimension:
-%%
-\begin{minted}[breakindent=2.9em]{fortran}
-x = a(i, j)   + a(i-1, j)   + a(i+1, j)
-y = a(i, j-1) + a(i-1, j-1) + a(i+1, j-1)
-z = a(i, j+1) + a(i-1, j+1) + a(i+1, j+1)
-!= stencil centered(depth=1, dim=1) * centered(depth=1, dim=2) :: a
-b(i, j) = (x + y + z) / 9.0
-\end{minted}
-%
-%This pattern is common in image convolution applications.
-The specification ranges over the
-values that flow to the array subscript on the left-hand side,
-and so ranges over the intermediate assignments to \term{x},
-\term{y}, and \term{z}. Each index in the code is consistent
-with both specifications simultaneously, \eg{}, \texttt{a(i-1, j+1)}
-is within the centered region in dimension $1$ and the centered region
-in dimension $2$.
-
-The union of two regions $r \term{+} s$ means that any indices
-in the specified code must be consistent with either of $r$ or $s$.
-For example, the following gives the specification of a five-point
-stencil which is the sum of two compound \texttt{pointed} and
-\texttt{centered} regions in each dimension:
-%
-\begin{minted}{fortran}
-!= stencil centered(depth=1, dim=1)*pointed(dim=2) + centered(depth=1, dim=2)*pointed(dim=1) :: a
-b(i,j) = -4*a(i,j) + a(i-1,j) + a(i+1,j) + a(i,j-1) + a(i,j+1)
-\end{minted}
-%%
-Here the left-hand side of \texttt{+} says that when the second dimension
-(induction variable $j$) is fixed at the origin, the first dimension
-(induction variable $i$) accesses the immediate vicinity of the origin
-(to depth of one). The right hand side of \texttt{+} is similar but the dimensions are reversed.
-This reflects the symmetry under rotation of the five-point stencil.
-
-\paragraph{Region declarations and variables}
-
-Region specifications can be assigned to region variables via
-region declarations. For example, the shape of a
-``\emph{Roberts cross}'' edge-detection convolution~\cite{davis1975survey}
-can be stated:
-%%
-\begin{minted}{fortran}
-!= region :: r1 = forward(depth=1, dim=1)
-!= region :: r2 = forward(depth=1, dim=2)
-!= region :: robertsCross = r1*r2
-!= stencil robertsCross :: a
-\end{minted}
-This is useful for common patterns, such as the five-point
-pattern, as the regions can be defined once and reused.
-%%
-\paragraph{Modifiers}
-%%
-Region specifications can be modified
-by \emph{approximation} and \emph{multiplicity} information
-(in \textit{spec} in \Cref{fig:syntax}).
-The \texttt{readOnce} modifier enforces that no index appears more
-than once (that is, its multiplicity is one). For example, all of
-the previous examples could have \texttt{readOnce} added:
-%
-\begin{minted}{fortran}
-!= stencil readOnce, backward(depth=2, dim=1) :: a
-b(i+1) = a(i) + a(i-1) + a(i-2)
-\end{minted}
-%
-This specification would be invalid if any of the
-array subscripts were repeated. This modifier provides a way to
-rule out any accidental repetition of array subscripts.
-The notion is similar to that of linear types~\cite{wadler1990linear}, where a value must be used
-exactly once. We opt for the more informative and easily understood name
-\texttt{readOnce}. This modifier is optional, so it need not
-be present even if the stencil is linear.
-
-In some cases, it is useful to give a lower and/or upper bound for a
-stencil. This can be done using either the \term{atMost} or
-\term{atLeast} modifiers. This is particularly useful in situations
-where there is a non-contiguous stencil pattern, which cannot be expressed
-precisely in our syntax. For example:
-%
-\begin{minted}{fortran}
-!= stencil atLeast, pointed(dim=1)         :: a
-!= stencil atMost, forward(depth=4, dim=1) :: a
-b(i) = a(i) + a(i+4)
-\end{minted}
-
-\section{Semantic model}
-\label{sec:model}
-
-\noindent
-We define a lattice model of array access patterns which serves as both a
-denotational model of the semantics of our specification language and an
-abstract interpretation domain for source code. This model (1) serves to
-explain the meaning of our specifications; (2) is used in the inference and
-checking algorithms (\Cref{sec:analysis}); (3) justifies an equational theory
-for specifications in the next section; (4) is used to optimise specifications
-using lattice identities; and (5) can be used to guide correct implementations.
-
-The model is defined over vectors of sets of integers which we call
-\emph{index schemes}.  As an initial informal example, consider the
-following simple stencil computation:
-%
-\begin{minted}{fortran}
-do i = 1, n
-  b(i) = a(i,0) + a(i+1,0)
-end do
-\end{minted}
-%
-The access pattern on array \term{a}, relative
-to induction variable \term{i}, is captured
-by a vector of length 2 containing integer sets $\langle{\{0, 1\},
-  \mathbb{Z}\}}\rangle$. This describes that, in the first dimension,
-the array is read at offsets of $0$ and $1$ from an induction
-variable. In the second dimension, the index is unconstrained
-as it is a constant.
-
-Index schemes form a lattice which provides a rich set of equations
-and properties, which we exploit. We first set up the 
-domain of the model (\S\ref{sec:domain}), using it to define a semantics
-for our specification language (\S\ref{sec:semantics}) and then as
-the target for an abstract interpretation on imperative
-code (\S\ref{sec:fromcode}).
-
-\subsection{Lattice model of regions}
-\label{sec:domain}
-
-%\begin{defn}[Extended integers]
-%  We define the set \zinf{} as
-%  $\mathbb{Z}$ extended with $\infty$ and $-\infty$. For any $a$ in
-%  \zinf{}, we have $-\infty \leq a \leq \infty$. The resulting set is
-%  a total order with top and bottom elements $-\infty$ and
-%  $\infty$ respectively.
-%\end{defn}
-
-\begin{defn}[Index scheme]
-  An $N$-dimensional \emph{index scheme} is a vector of length $N$ of
-  integer sets, \ie{}, a member of $\mathcal{P}(\bz)^{N}$.
-%  An equivalent view of these vectors is as an $N$-times finite Cartesian
-%  product on subsets of $\bz$.
-  We use $S, T, U$ to denote index schemes. Henceforth, we assume index schemes
-  are all $N$-dimensional for some $N$.
-
-  Index schemes can be \emph{projected} in the $i^{\textit{th}}$ dimension by
-  $\pi_i : {\mathcal{P}(\bz{})}^N \to \mathcal{P}(\bz{})$. For an index scheme
-  $I$, we refer to $\pi_i(S)$ as the \emph{$i^{th}$ component} of $S$. We assume
-  that $i$, when used for projection, always lies between $1$ and $N$.
-\end{defn}
-
-\begin{restatable}{lem}{vectorIntersect}
-\label{lem:vector-intersect}
-  Intersection distributes over index schemes. That is, for index schemes $S, T
-  \in \mathcal{P}(\bz)^N$
-%
-  \begin{equation*}
-    S \cap T = \prod_{i = 1}^{N} \pi_i(S) \cap \pi_i(T)
-  \end{equation*}
-\end{restatable}
-
-Union does not distribute over index schemes, however, a more restricted
-property holds.
-
-\begin{restatable}{lem}{vectorUnion}
-\label{lem:vector-union}
-$S$ and $T$ are index schemes such that $\pi_i(S) = \pi_i(T)$ for all $1 \leq i
-\leq N$ apart from some dimension $k$, then:
-%
-  \begin{equation*}
-    S \cup  T
-    =
-    \pi_1(S) \times \cdots \times
-    (\pi_k(S) \cup \pi_k(T)) \times \cdots \times
-    \pi_N(S)
-  \end{equation*}
-\end{restatable}
-%
-\begin{defn}[Intervals with an optional hole]
-  We define an extended notion of closed interval on $\bz{}$ which may contain a
-  \emph{hole} at the origin, written \interv{a}{b}{c} where $a$ and $b$ are
-  drawn from \bz{} with $a \leq 0 \leq b$ and $c$ is drawn from $\mathbb{B} = \{
-  \mathsf{true}, \mathsf{false} \}$. Intervals are interpreted as sets as
-  follows:
-%
-  \begin{equation*}
-    \interv{a}{b}{c} \triangleq
-      \{ n \mid a \leq n \leq b \wedge (\neg c \implies n \neq 0) \}
-  \end{equation*}
-
-  We also add the distinguished interval $\interv{-\infty}{\infty}{}$, which is simply an
-  alias for \bz{}, but this notation prevents handling infinite interval
-  separately in the following definitions, theorems, and proofs. Here, $-\infty$
-  and $\infty$ behaves like top and bottom elements to \bz{} respectively.
-
-  We denote the set of all such intervals (sets) as $\textit{Interval}$. If the
-  superscript to the interval is omitted it is treated as $\mathsf{true}$.
-\end{defn}
-%
-\begin{restatable}{lem}{intervalIdentities}
- \label{lem:interval-identities}
-  We have the following dual identities for \bz{} intervals:
-%
-  \begin{align*}
-    \interv{a}{b}{c} \cap \interv{d}{e}{f} & =
-      \interv{\max \{a,d\}}{\min \{b,e\}}{c \wedge f} \\
-    \interv{a}{b}{c} \cup \interv{d}{e}{f} & =
-      \interv{\min \{a,d\}}{\max \{b,e\}}{c \vee f}
-  \end{align*}
-\end{restatable}
-%
-
-We define two specialisations (and subset spaces) of index scheme:
-\emph{subscript scheme} and \emph{interval scheme}:
-
-\begin{defn}[Interval scheme]
-  An interval scheme is a finite Cartesian product of intervals on \bz{},
-  denoted by the set $\textit{Interval}^N$ for a product of $N$ intervals.
-\end{defn}
-
-\begin{defn}[Subscript scheme]
-  A \emph{subscript scheme} is an index scheme scheme where:
-  %
-  \begin{equation*}
-    \forall i.\ 1 \leq i \leq N \implies
-      \pi_i (S) = \{ p \}
-      \; \vee \;
-      \pi_i (S) = \interv{-\infty}{\infty}{}
-    \end{equation*}
-%
-  That is, the $i^{th}$ component of the set is either a singleton in \bz{} or
-  the infinite interval.
-\end{defn}
-%
-\dnote{May want a predicate as well}
-
-\begin{defn}[Region]
-  A region is an index scheme and \region{N} is the set of all regions (\ie{},
-  $\region{N} \subseteq \mathcal{P}(\bz)^N)$. The set of all regions is  defined
-  as the smallest set satisfying the following:
-%
-  \begin{enumerate}
-    \item If $R$ is in $\textit{Interval}^N$, then $R$ is in \region{N}.
-    \item If $R$ and $S$ are in \region{N}, then so are $R \cap S$ and
-      $R \cup S$.
-  \end{enumerate}
-\end{defn}
-%
-\begin{restatable}{prop}{regionLattice}
-  \label{prop:regionLattice}
-  $(\region{N},\cup,\cap,\subseteq)$ is a bounded distributive lattice with top
-  $\top = \bz{}^N$ and bottom $\bot = \emptyset$.
-\end{restatable}
-%
-\begin{proof}
-  Straightforward, the join and meet are mapped to $\cup$ and $\cap$, the set is
-  inductively designed to be closed under these operations. Union and
-  intersection are associative, commutative, and absroptive under closed sets
-  and then they are also for \region{N}. This is enough to show that, it is a
-  lattice.
-
-  Further, we have $\bz{}^N \cap R = R$ and $\emptyset \cup R = R$ with
-  $\bz{}^N$ and $\emptyset$ belonging to \region{N}. This makes the lattice a
-  bounded one.
-
-  Finally, the lattice is distributive since union distributes over intersection
-  and vice versa when the set is closed under these operations.
-\end{proof}
-
-\begin{defn}
-  $\mathsf{Mult}$ and $\mathsf{Approx}$ are parametric labelled variant types
-  with injections given by their definition:
-%
-  \begin{align*}
-    \mathsf{Mult} \;\; a \;\; &
-      \triangleq \mathsf{mult} \; a \;\mid\; \mathsf{only} \; a \\
-    \mathsf{Approx} \;\; a \;\; &
-      \triangleq \mathsf{exact} \; a \;\mid\; \mathsf{lower} \; a \;\mid\;
-        \mathsf{upper} \; a
-  \end{align*}
-%
-  \eg{}, $\mathsf{lower}$ is an injection $\mathsf{lower} : a \to \mathsf{Approx}
-  \; a$ etc.
-  These will be used in the following subsection to give meaning to the
-  specification modifiers for approximation and multiplicity.
-\end{defn}
-
-\subsection{Denotational semantics for specifications}
-\label{sec:semantics}
-
-\noindent
-An interpretation function $\interp{-}_N$ maps closed\footnote{That
-  is, we assume there are no occurrences of \textit{rvar} in a
-  specification being modelled.  Any \emph{open} specification
-  containing region variables can be made closed by straightforward
-  syntactic substitution with a (closed) \textit{region}.}
-specifications to sets of $N$-dimensional index schemes with modifier
-information, \ie{} specifications are mapped to
-$\textsf{Mult} (\textsf{Approx} (\region{N}))$.
-
-The interpretation is overloaded on \emph{regions} in
-\Cref{subfig:region-model}. Various intermediate notions are used.
-
-\begin{defn}
-  Let $\textit{promote}_N : \mathbb{N}^+ \times \textit{Interval} \to
-  \textit{Interval}^N$ be a function generating an interval scheme such that if
-  $v$ is $\vecgen{N}{i}{\interv{a}{b}{c}}$, then $\pi_i(v) = \interv{a}{b}{c}$
-  and $\pi_j(v) = \bz{}$ in all other dimensions $j$.
-\end{defn}
-
-%
-\begin{figure}[!t]
-\begin{subfigure}[t]{0.5\textwidth}
-\begin{align*}
-  \interp{-}_N & : \textit{region} \rightarrow \region{N} \\
-%
-  \interp{\stencil{p}{$i$}{}{}}_N & =
-    \vecgen{N}{i}{\interv{0}{0}{\mathsf{true}}}\\
-%
-  \interp{\stencil{c}{$i$}{$k$}{\textcap{p}}}_N & =
-    \vecgen{N}{i}{\interv{-k}{k}{\interp{\textcap{p}}}} \\
-%
-  \interp{\stencil{f}{$i$}{$k$}{\textcap{p}}}_N & =
-    \vecgen{N}{i}{\interv{0}{k}{\interp{\textcap{p}}}} \\
-%
-  \interp{\stencil{b}{$i$}{$k$}{\textcap{p}}}_N & =
-  \vecgen{N}{i}{\interv{-k}{0}{\interp{\textcap{p}}}}
-\\
-  \interp{\texttt{\textcap{r} + \textcap{s}}}_N & =
-    \interp{\textcap{r}}_N \vee \interp{\textcap{s}}_N
-\\
-  \interp{\texttt{\textcap{r} * \textcap{s}}}_N & =
-    \interp{\textcap{r}}_N \wedge \interp{\textcap{s}}_N \\[-1em]
-\end{align*}
-\caption{Interpretation of regions}
-\label{subfig:region-model}
-\end{subfigure}
-\hspace{1em}
-\begin{subfigure}[t]{0.4\textwidth}
-\begin{align*}
-\interpApprox{-} & : \textit{approx} \rightarrow (A \rightarrow
-  \textsf{Approx} \, A) \\
-\interpApprox{\texttt{atLeast}} & = \mathsf{lower} \\
-  \interpApprox{\texttt{atMost}} & = \mathsf{upper} \\
-  \interpApprox{\epsilon} & = \mathsf{exact} \\ \\
-  \interpMult{-} & : \textit{mult} \rightarrow (A \rightarrow
-  \textsf{Mult} \, A) \\
-  \interpMult{\texttt{readOnce}} & = \mathsf{once} \\
-  \interpMult{\epsilon} & = \mathsf{mult}
-\end{align*}
-\caption{Interpretation of modifiers}
-\label{subfig:modifier-model}
-\end{subfigure}
-\label{fig:semantics}
-\caption{Semantic model of specifications}
-\end{figure}
-
-The first four equations of \Cref{subfig:region-model} model region 
-contants. The final two equations model the \term{+} and \term{*}
-operators in terms of the join (union) and meet (intersection)
-of interval schemes. Thus regions are modelled as members
-of $\region{N}$.
-
-We mark in our model the presence of modifiers such as
-\texttt{readOnce} and \texttt{atMost} as introduced in \Cref{}.
-Approximation modifiers are interpreted as injections into the
-$\mathsf{Approx}$ variant by $\interpApprox{}$ in
-\Cref{subfig:modifier-model}.  The $\textsf{Approx}$ type corresponds
-to the presence or absence of the spatial approximation modifier, with
-\textsf{exact} when there is no such modifier and \textsf{lower} and
-\textsf{upper} for \term{atLeast} and \term{atMost}. In a similar way,
-multiplicity modifiers are interpreted as injections in the
-$\mathsf{Mult}$ variant by $\interpMult{}$, corresponding to the to
-the presence or absence of the \term{readOnce} modifier as shown in
-\Cref{subfig:modifier-model}.
-
-\begin{defn}[Semantics of specifications]
-The intermediate interpretations of \Cref{fig:semantics}
-are composed to give a model for the top-level specification
-syntax as:
-%
-\begin{equation*}
-  \interp{\texttt{stencil \textcap{mult}, \textcap{approx}, \textcap{region}}}_N =
-    \interpMult{\textcap{mult}} \;
-           {(\interpApprox{\textcap{approx}} \;
-                    {\interp{\textcap{region}}_N)}}
-\end{equation*}
-\end{defn}
-%
-\begin{thm}
-The semantic model $\interp{-}_N$ of $N$-dimensional specifications
-is sound with respects to the equational theory of the language,
-that is:
-%
-\begin{equation*}
-\forall S, T, N . \quad
-S \equiv T \; \Rightarrow \;
-\interp{S}_N = \interp{T}_N
-\end{equation*}
-%
-We define the equational theory in \Cref{sec:equational-theory}.
-\end{thm}
-
-\subsection{Denotational semantics for array subscripts}
-\label{sec:fromcode}
-
-\begin{defn}
-  Recall array subscript terms of the form $a(\bar{e})$ from
-  \Cref{def:array-subs}. We interpret these terms with the partial
-  interpretation $\interp{-}^{\mathit{aterm}} : \textit{array-term}
-  \prightarrow{} \mathcal{P}(\bz{})^N$. The interpretation is defined when
-  all indices are either constant or neighbourhood indices as defined in
-  \Cref{def:neighbour-ix}.
-%
-  \begin{align*}
-    \interp{a(\bar{e})}^{\mathit{aterm}} =
-      \prod_{1 \leq i \leq N} \mathit{subscript}(\bar{e}_i) & &
-  %
-    \textit{subscript}(e) = \begin{cases}
-      \{ c \} & e \equiv i \pm c \\
-      \bz & e \; \mbox{is constant}
-    \end{cases}
-  \end{align*}
-\end{defn}
-
-\section{Analysis, Checking, and Inference}
-\label{sec:algorithms}
-\input{algorithms}
-
-\section{Evaluation}
-\label{sec:evaluation}
-
-\bibliography{references}
-
-\appendix
-% http://2017.splashcon.org/track/splash-2017-OOPSLA#Instructions-for-Authors
-% "There is no page limit for bibliographic references and appendices,
-% and, therefore, for the overall submission."
-\input{appendix}
-
-\end{document}
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: t
+%%% End:

### 2017-04-17 oopsla17/oopsla17.tex
fix to coalesce

@@
 \section{Introduction}
-
-\paragraph{Context of this research}
-
-There is an increasing awareness of the need for program verification
-techniques in
-science~\cite{post2005computational,oberkampf2010verification,orchard2014computational}.
-Whilst there are various kinds of automated and semi-automated
-verification tools available, they see little use in science. This is
-partly due to a lack of training and a lack of awareness. But its also
-due to a lack of targetting. This paper is part of a line of research
-to provide lightweight verification tools targetting
-programming patterns common in computational science code, 
-that are easy to use by scientists, and that integrate with common
-practises and workflows.
-
-\subsection{Terminology and notation}
-
-\begin{defn}[Induction variables]
-  An integer variable is a \emph{base induction variable} if it is the control
-  variable of a ``\texttt{for}'' loop, incremented by $1$ per iteration. A variable
-  is interpreted as an induction variable only within the scope of the loop
-  body. Throughout, $i, j, k$ range over induction variables.
-
-  A \emph{derived induction variable} is an expression of the form $a \times i
-  + b$, where $a$ and $b$ are constant expressions, \ie{}, an affine expression
-  on an induction variable $i$.
-\end{defn}
-
-\begin{defn}[Array subscripts and indices]\label{def:array-subs}
-  An \emph{array subscript}, denoted $a(\bar{e})$, is an expression which reads
-  from an $N$-dimensional array $a$ at an \emph{index} specified by a
-  comma-separated sequence of integer expressions denoted $\bar{e}$ or in
-  expanded form as $(e_1, \ldots, e_N)$. An index $e_i$ is called
-  \emph{relative} if the expression involves an induction variable. An
-  \emph{absolute index} is a integer expression which is constant relative to
-  the enclosing loop.
-\end{defn}
-
-\begin{defn}[Neighbourhood index]\label{def:neighbour-ix}
-  For an array subscript $a(\bar{e})$ an index $e \in \bar{e}$ is a
-  \emph{neighbourhood index} if $e$ is of the form $e \equiv i$, $e \equiv i +
-  c$, or $e \equiv i - c$, where $c$ is an integer constant. That is, a
-  neighbourhood index is a constant translation of an induction variable. (The
-  relation $\equiv$ identifies terms up-to commutativity of $+$ and the inverse
-  relationship of $+$ and $-$ \eg{}, $(-b) + i \equiv i - b$).
-\end{defn}
+\label{sec:introduction}
+\input{introduction}
 

### 2017-04-17 oopsla17/oopsla17.tex
introduction

@@
 \begin{abstract}
-  Arrays computations are the workhorse of numerical modelling and
-  computational science. However, low-level manipulation of array
+  Arrays computations are at the core of numerical modelling and
+  computational science applications. However, low-level manipulation of array
   indices is a source of program error.  Many practitioners are aware
@@
   detected by our approach. We found roughly 60,000 targets for
-  specification across roughly 1.2 million lines of code, showing the
+  specification across roughly 1.4 million lines of code, showing the
   vast majority of array computations read from arrays in a pattern with
@@
 
-\section{A Specification Language for Shape}
+\section{A Specification Language for the Shape of Array Patterns}
+\label{sec:lang}
 
@@
 
+\subsection{Verification case studies}
+\label{sec:case-studies}
+
+\subsubsection{Catching bugs}
+
+\subsubsection{Aiding refactoring}
+
+\subsubsection{Stress-testing Camfort}
+
 \bibliography{references}

### 2017-04-17 oopsla17/algorithms.tex
corrected coalescing
fix to coalesce
Tidy coalesceTop

@@
 \begin{align*}
-\mathsf{coalesce}(\mathcal{S}) = \; &
+\mathsf{coalesce1}(S, \mathcal{S}) & =
  \{\,S \cup T \mid
-\forall S \in \mathcal{S}, \;
-\forall T \in \mathcal{S}. \;
+T \in \mathcal{S} \, \wedge \,
 \contig{S}{T} \} \\
-\cup \; &
-\{ \,\;\;\; S \;\;\; \mid \forall S \in \mathcal{S},
-\; \not \exists T \in \mathcal{S} . \; \contig{S}{T}\}
+\mathsf{coalesce}(\mathcal{S}) & =
+\bigcup_{S \in \mathcal{S}} \mathsf{coalesce}(
+\mathsf{coalesce1}(S, \mathcal{S})
+\cup (\mathcal{S} - \{S\})) \\
+\mathsf{coalesceTop}(\mathcal{S}) & =
+ \mathsf{coalesce}(\mathcal{S})
+\; \cup \;
+\{ S \in \mathcal{S} \mid \neg \exists T \in \mathcal{S} . \; \contig{S}{T}\}
 \end{align*}
+\dnote{May need to prove that this is total}
+%\begin{align*}
+%\mathsf{coalesce}(\mathcal{S}) = \; &
+%\{\,S \cup T \mid
+%\forall S \in \mathcal{S}, \;
+%\forall T \in \mathcal{S}. \;
+%\contig{S}{T} \} \\
+%\cup \; &
+%\{ \,\;\;\; S \;\;\; \mid \forall S \in \mathcal{S},
+%\; \not \exists T \in \mathcal{S} . \; \contig{S}{T}\}
+%\end{align*}
 %
-That is, the new set of index schemes $\mathsf{coalesc}(\mathcal{S})$
+That is, the new set of index schemes $\mathsf{coalesce}(\mathcal{S})$
 comprises the union of the set of coalesced pairs of contiguous

### 2017-04-17 oopsla17/introduction.tex

@@
-\paragraph{Context of this research}
+\noindent
+In the sciences, complex models are now almost always expressed as
+computer programs. But how can a scientist have confidence that the
+implementation of their model is as they intended? There is an
+increasing awareness of the need for program verification in science
+and the possibility of using (semi-)automated
+tools~\cite{post2005computational,oberkampf2010verification,orchard2014computational}.
+However, whilst program verification approaches are slowly maturing in
+computer science they see little use in the natural and physical
+sciences. This is partly due to a lack of training and awareness, but
+also a lack of tools targeted at the needs of scientists. This
+paper is part of a line of research on providing lightweight, easy-to-ue
+verification tools targeted at common programming patterns in science, 
+motivated by analysis of real code.
 
-There is an increasing awareness of the need for program verification
-techniques in
-science~\cite{post2005computational,oberkampf2010verification,orchard2014computational}.
-Whilst there are various kinds of automated and semi-automated
-verification tools available, they see little use in science. This is
-partly due to a lack of training and a lack of awareness. But its also
-due to a lack of targetting. This paper is part of a line of research
-to provide lightweight verification tools targetting
-programming patterns common in computational science code, 
-that are easy to use by scientists, and that integrate with common
-practises and workflows.
+We focus on one common concept: \emph{arrays}, the core data structure
+in numerical modelling code, typically representing discrete
+approximations of physical space or data sets. A common programming
+pattern, sometimes referred to as the \emph{structured grid}
+pattern~\cite{Asanovic2006}, traverses the index space of one or more
+arrays via a loop, computing elements of another array or reducing the
+elements to a single value. For example, the following Fortran code
+computes the one-dimensional discrete Laplace transform approximating
+a derivative:
+%
+\begin{minted}{fortran}
+do i = 1, (n-1)
+      b(i) = a(i-1) - 2*a(i) + a(i+1)
+end do
+\end{minted}
+%
+This is a \emph{stencil} computation, a programming idiom
+where the elements of an array at each index $i$ from a
+\emph{neighbourhood} of values around $i$ in some input array(s). This
+pattern is common in scientific, graphical, and numerical
+code, \eg{}, convolutions in image processing, approximations
+to differential equations, cellular automata.
 
-\subsection{Terminology and notation}
+Such array computations are prone to error in their indexing
+terms. For example, a logical off-by-one-error might manifest itself
+as writing $\texttt{a(i)}$ instead of $\texttt{a(i-1)}$ (we revisit an
+example we found of this in \Cref{sec:case-studies}).
+Errors also arise by simple lexical mistakes when large amounts of
+fine-grained indexing are involved in a single expression. For
+example, the following snippet from a Navier-Stokes
+fluid model~\citet{griebel1997numerical} has two arrays are
+read with different data access patterns, across two dimensions, with
+dense index-manipulation code:
+%%
+\begin{minted}[firstnumber=20,xleftmargin=2em]{fortran}
+du2dx = ((u(i,j)+u(i+1,j))*(u(i,j)+u(i+1,j))+gamma*abs(u(i,j)+u(i+1,j))*(u(i,j)-u(i+1,j))- &
+  (u(i-1,j)+u(i,j))*(u(i-1,j)+u(i,j))-gamma*abs(u(i-1,j)+u(i,j))*(u(i-1,j)-u(i,j))) /(4.0*delx)
+
+duvdy = ((v(i,j)+v(i+1,j))*(u(i,j)+u(i,j+1))+gamma*abs(v(i,j)+v(i+1,j))*(u(i,j)-u(i,j+1))- &
+  (v(i,j-1)+v(i+1,j-1))*(u(i,j-1)+u(i,j))-gamma*abs(v(i,j-1)+v(i+1,j-1))*(u(i,j-1)- u(i,j))) / (4.0*dely)
+
+laplu = (u(i+1,j)-2.0*u(i,j)+u(i-1,j))/delx/delx+(u(i,j+1)-2.0*u(i,j)+u(i,j-1))/dely/dely
+f(i,j) = u(i,j)+del_t*(laplu/Re-du2dx-duvdy)
+\end{minted}
+%
+This miasma of indexing expressions is hard to read and
+prone to simple textual input mistakes, \eg{}, swapping \texttt{-} and
+\texttt{+}, missing an indexing term, or transforming the wrong
+variable \eg{} \texttt{(i+1,j)} instead of \texttt{(i,j+1)}.
+
+In practice, the typical development procedure for complex stencil
+computations involves some ad hoc testing to ensure that no
+mistakes have been made \eg{}, by visual inspections on data,
+or comparison against manufactured or analytical
+solutions~\cite{farrell2010automated}. Such testing is often
+discarded once the code is shown correct. This is not the only
+information that is discarded. The shape of the indexing pattern
+is usually the result of choices made in the numerical-analysis
+procedure to discretise some continuous equations. Rarely are these decisions captured in the source code,
+yet the derived shape is usually uniform with a 
+clear and concise description \eg{}, 
+\emph{centered space, of depth 1} referring to indexing
+terms \texttt{a(i)}, \texttt{a(i-1)} and
+\texttt{a(i+1)}~\cite{recktenwald2004finite}.
+
+To support correct array programming, we propose a simple,
+abstract specification language for the data access pattern of
+array-loop computations. This provides a way to rule out bugs due to
+indexing errors. The language design is informed by an initial
+empirical study of array computations in a corpus real-world
+scientific code base, totally 1.4 million lines of
+code (\Cref{sec:study}). We confirm our initial hypotheses of the
+ubiquity of looped array computations, but also that they have a
+common form, reading from arrays in a fixed neighbourhood of
+contiguous elements with a simple static pattern. From this we
+designed a simple set of combinators to express common array patterns (\Cref{sec:lang}).
+As an example, the Laplace program is specified in our language by:
+%
+\begin{minted}[linenos=false]{fortran}
+!=  stencil centered(depth=1, dim=1) :: a
+\end{minted}
+%
+That is, \texttt{a} is accessed with a symmetrical pattern in its
+first dimension (``centered'') to a depth of one in each direction and
+its result contributes to an array
+write.  The Navier-Stokes example has two specifications:
+%
+\begin{minted}[linenos=false,xleftmargin=0em,xrightmargin=0.8em,breakindent=1.9em]{fortran}
+!= stencil centered(depth=1,dim=1)*pointed(dim=2) + centered(depth=1,dim=2)*pointed(dim=1) :: u
+!= stencil forward(depth=1,dim=1)*backward(depth=1,dim=2) :: v
+\end{minted}
+%
+The specifications requires that \texttt{u}
+is accessed with a centered pattern to depth of 1 in both dimensions
+(this is known as the \emph{five-point stencil}) and \texttt{v} is
+accessed in a neighbourhood bounded forwards to depth of $1$ in the
+first dimension and backward to a depth of $1$ in the second
+dimension. The specification is relatively small and abstract compared
+with the code, with a small number of combinators that \emph{do not
+  involve any indexing expressions}, \eg{} \texttt{a(i+1,j-1)}. This
+contrasts with other specification approaches, \eg{}, deductive
+verification tools such as ACSL~\citet{baudin2008acsl}, where
+specifications about array computations must also be expressed in
+terms of array-indexing expressions. Thus, any low-level mistakes that
+could be made whilst programming complex indexing code could also be
+made when writing its specification. Our specifications are 
+abstract to avoid the error-prone nature of index manipulation; our
+specifications are more lightweight to aid its adoption in science.
+
+We implemented a verification tool for our specification language as
+an extension to CamFort~\cite{camfort}, an open-source program
+analysis tool for Fortran. Specifications are associated to code via
+comments, against which the tool checks the code for conformance. This
+specify-and-check approach reduces testing effort and future-proofs
+against bugs introduced during refactoring and maintenance. A
+specification also concisely captures the array access pattern 
+and provides documentation. We also provide an inference procedure
+which efficiently produces specifications with no programmer
+intervention, automatically inserting specifications at appropriate
+places in the source code. This aids adoption of our approach to
+existing legacy code base.
+
+The checking and inference algorithms (\Cref{sec:algorithms}) are
+derived from a model of array access patterns, which is also used to
+give a denotational model of our specification language
+(\Cref{sec:semantics}).
+
+Using the inference-mode of the tool, we applied our tool to our
+original corpus (\Cref{sec:evaluation}). Our tool identifies and
+infers specifications for roughly 60,000 array computations in the
+corpus. Approximately 5,000 of the computations we found are non-trivial,
+corresponding to code which is a possible source of errors. This
+validates the design of our language in its capability to capture many
+core patterns.
+
+Our approach is not targeting a class of bugs that can be detected
+automatically (\emph{push-button verification}). Instead, array
+indexing bugs must be identified relative to a specification of the
+intended access pattern.  Nevertheless, we report on instances of code
+revisions from the commit histories of our corpus where a
+correct specification would have either spotted an error (which was
+latter correct) or would have assisted in an array refactoring task
+(\Cref{sec:case-studies}).
+
+\paragraph{Terminology and notation}
+
+We fix some terminology used throughout to describe
+syntactic constructs.
 
 \begin{defn}[Induction variables]
-  An integer variable is a \emph{base induction variable} if it is the control
-  variable of a ``\texttt{for}'' loop, incremented by $1$ per iteration. A variable
-  is interpreted as an induction variable only within the scope of the loop
-  body. Throughout, $i, j, k$ range over induction variables.
+  An integer variable is a \emph{base induction variable} if it is the
+  control variable of a ``\texttt{for}'' loop (\texttt{do} in
+  Fortran), incremented by $1$ per iteration. A variable is
+  interpreted as an induction variable only within the scope of the
+  loop body. Throughout, $i, j, k$ range over induction variables.
 
-  A \emph{derived induction variable} is an expression of the form $a \times i
-  + b$, where $a$ and $b$ are constant expressions, \ie{}, an affine expression
-  on an induction variable $i$.
+  A \emph{derived induction variable} is an expression of the form
+  $a \ast i + b$, where $a$ and $b$ are constant expressions, \ie{},
+  an affine expression on an induction variable $i$.
 \end{defn}
@@
 \begin{defn}[Array subscripts and indices]\label{def:array-subs}
-  An \emph{array subscript}, denoted $a(\bar{e})$, is an expression which reads
-  from an $N$-dimensional array $a$ at an \emph{index} specified by a
-  comma-separated sequence of integer expressions denoted $\bar{e}$ or in
-  expanded form as $(e_1, \ldots, e_N)$. An index $e_i$ is called
-  \emph{relative} if the expression involves an induction variable. An
-  \emph{absolute index} is a integer expression which is constant relative to
-  the enclosing loop.
+  An \emph{array subscript}, denoted $a(\bar{e})$, is an expression
+  which indicates the element of an $N$-dimensional array $a$ at the
+  \emph{index} $\bar{e}$, specified by a comma-separated sequence of
+  integer expressions, in expanded form as $(e_1, \ldots, e_N)$. An
+  index $e_i$ is called \emph{relative} if the expression involves an
+  induction variable. An \emph{absolute index} is a integer expression
+  which is constant relative to the enclosing loop.
 \end{defn}

### 2017-04-17 oopsla17/algorithms.tex
Typo in nonOverlap
Add formatting break algorithm into two

@@
 
-\begin{enumerate}[leftmargin=1.5em]
-\item (\textbf{Covering})
+We explain the algorithm in two parts. First, we show how \emph{adjacent} index
+schemes are coalesced into smaller index schemes while remaining in union normal
+form. Then, we explain how this smaller unioned indexing schemes are converted
+to specifications.
+
+\subsubsection{Covering}
 A covering of (possibly overlapping) the subscript space
@@
 
-\begin{enumerate}[leftmargin=0.5em]
-\item (\textbf{Coalesce intervals})
+\begin{description}[leftmargin=0.5em]
+
+\item[Coalesce intervals]
 %
@@
 contiguous scheme.
-\item (\textbf{Remove redundant schemes})
+
+\item[Remove redundant schemes]
 Any indexing schemes which are subsumed by another are filtered out:
@@
 \textsf{nonOverlap}(\mathcal{S}) =
-\{S \mid \forall S \in \mathcal{S}, \; \not \exists T \in (\mathcal{S} -
+\{S \mid \forall S \in \mathcal{S}, \; \neg \exists T \in (\mathcal{S} -
   \{S\}), \; \forall i . \; \pi_i(S) \subseteq \pi_i(T) \}
@@
 %\item \textbf{Covering}
-\end{enumerate}
+\end{description}
+
 The least fixed-point of $\textsf{nonOverlap} \circ \textsf{coalesce}$
@@
 schemes are \textit{Region$_N$}?}
-\item (\textbf{Index schemes to syntax}) \dnote{Mistral todo}
-\end{enumerate}
+
+\subsubsection{Index schemes to syntax}

### 2017-04-17 oopsla17/lang.tex

--- 2017-04-17 oopsla17/oopsla17.tex
@@
-\documentclass[acmlarge,review,anonymous]{acmart}\settopmatter{printfolios=true}
-
-%% Some recommended packages.
-\usepackage{booktabs}   %% For formal tables:
-                        %% http://ctan.org/pkg/booktabs
-\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
-                        %% http://ctan.org/pkg/subcaption
-
-\usepackage{hyperref}
-\usepackage{natbib}
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{amsthm}
-\usepackage{syntax}
-\usepackage{indentfirst}
-\usepackage{cleveref}
-\usepackage{xcolor}
-\usepackage{multirow}
-\usepackage{siunitx} % For pretty-printing numeric values and SI units
-                     % of measure. e.g., the tabular column type S is
-                     % used to print nice-looking tables of numbers.
-\sisetup{ % defaults
-  group-separator={,},
-  group-minimum-digits={3},
-  output-decimal-marker={.},
-  table-format = 6
-}
-
-\usepackage{minted}
-\usepackage{etoolbox}
-\usepackage{enumitem}
-\usepackage{thmtools}
-\usepackage{thm-restate}
-
-\newcommand{\fortran}[1]{\mintinline{fortran}{#1}}
-
-\errorcontextlines 10000
-\usepackage{stencilmacros}
-
-%% Bibliography style
-\bibliographystyle{ACM-Reference-Format}
-%% Citation style
-%% Note: author/year citations are required for papers published as an
-%% issue of PACMPL.
-\citestyle{acmauthoryear}   %% For author/year citations
-
-\fvset{
-  linenos=true,
-  fontsize=\footnotesize,
-  breaklines=true,
-  breakafter=w),
-  xleftmargin=3\parindent,
-  numbersep=1em
-}
-
-\usemintedstyle{vs}
-
-\setlength{\grammarindent}{3em} % increase separation between LHS/RHS
-
-\newcommand{\inbar}{\ \ \textbar\ \ }
-
-% For evluation and empirical study
-% Defines the commands (headers)
-\input{resultsCmds}
-\input{studyResults}
-% renews the commands with actual data
-% \input{results}
-
-\definecolor{darkgreen}{rgb}{0.0,0.5,0.0}
-\definecolor{darkpurple}{rgb}{0.6,0.0,0.6}
-\definecolor{orange}{rgb}{0.8,0.4,0.0}
-\newcommand{\dnote}[1]{\textcolor{darkpurple}{Dom: #1}}
-\newcommand{\mnote}[1]{\textcolor{darkgreen}{Mistral: #1}}
-\newcommand{\anote}[1]{\textcolor{red}{Andy: #1}}
-
-\theoremstyle{definition}
-\newtheorem{defn}{Definition}
-
-\theoremstyle{plain}
-\newtheorem{thm}{Theorem}
-\newtheorem{lem}{Lemma}
-\newtheorem{prop}{Proposition}
-\newtheorem{case}{Case}
-
-\theoremstyle{remark}
-\newtheorem{remark}{Remark}
-
-\title{Verifying Spatial Properties of Array Computations}
-
-\author{Dominic Orchard}
-\affiliation{
-  \department{School of Computing}
-  \institution{University of Kent}
-}
-\email{d.a.orchard@kent.ac.uk}
-
-\author{Mistral Contrastin}
-\affiliation{
-  \department{Computer Laboratory}
-  \institution{University of Cambridge}
-}
-\email{Mistral.Contrastin@cl.cam.ac.uk}
-
-\author{Matthew Danish}
-\affiliation{
-  \department{Computer Laboratory}
-  \institution{University of Cambridge}
-}
-\email{Matthew.Danish@cl.cam.ac.uk}
-
-\author{Andrew Rice}
-\affiliation{
-  \department{Computer Laboratory}
-  \institution{University of Cambridge}
-}
-\email{Andrew.Rice@cl.cam.ac.uk}
-
-\begin{abstract}
-  Arrays computations are the workhorse of numerical modelling and
-  computational science. However, low-level manipulation of array
-  indices is a source of program error.  Many practitioners are aware
-  of the need to ensure program correctness, yet very few of the
-  techniques from the programming research community are applied by
-  scientists. We aim to change that by providing targetted lightweight
-  verification techniques for scientific code.  We focus on the all
-  too common mistake of array offset errors as a generalisation of
-  off-by-one errors.  Firstly, we report on a code analysis study on
-  twelve real-world computational science code base, identifying
-  common idioms of array usage and their spatial properties. This
-  provides much needed data on array programming idioms common in
-  scientific code.  From this data, we designed a lightweight
-  declarative specification language capturing the majority of array
-  access patterns via a small set of combinators. We detail a semantic
-  model, and the design and implementation of a verification tool for
-  our specification language, which both checks and infers
-  specifications.  We evaluate our tool on our corpus of scientific
-  code and give verification case studies of bug fixes that are
-  detected by our approach. We found roughly 60,000 targets for
-  specification across roughly 1.2 million lines of code, showing the
-  vast majority of array computations read from arrays in a pattern with
-  a simple, regular, static shape.
-\end{abstract}
-
-%  We focus on one error-prone aspect of numerical code: the 
-%  data access pattern of arrays.
-
-\begin{document}
-\maketitle
-
-\section{Introduction}
-\label{sec:introduction}
-\input{introduction}
-
-\section{An Empirical Study of Array Computations in Scientific Fortran code}
-\label{sec:study}
-\input{study}
-
-\section{A Specification Language for Shape}
-
 \begin{figure}[t]
@@
 
-\section{Semantic model}
-\label{sec:model}
-
-\noindent
-We define a lattice model of array access patterns which serves as both a
-denotational model of the semantics of our specification language and an
-abstract interpretation domain for source code. This model (1) serves to
-explain the meaning of our specifications; (2) is used in the inference and
-checking algorithms (\Cref{sec:analysis}); (3) justifies an equational theory
-for specifications in the next section; (4) is used to optimise specifications
-using lattice identities; and (5) can be used to guide correct implementations.
-
-The model is defined over vectors of sets of integers which we call
-\emph{index schemes}.  As an initial informal example, consider the
-following simple stencil computation:
-%
-\begin{minted}{fortran}
-do i = 1, n
-  b(i) = a(i,0) + a(i+1,0)
-end do
-\end{minted}
-%
-The access pattern on array \term{a}, relative
-to induction variable \term{i}, is captured
-by a vector of length 2 containing integer sets $\langle{\{0, 1\},
-  \mathbb{Z}\}}\rangle$. This describes that, in the first dimension,
-the array is read at offsets of $0$ and $1$ from an induction
-variable. In the second dimension, the index is unconstrained
-as it is a constant.
-
-Index schemes form a lattice which provides a rich set of equations
-and properties, which we exploit. We first set up the 
-domain of the model (\S\ref{sec:domain}), using it to define a semantics
-for our specification language (\S\ref{sec:semantics}) and then as
-the target for an abstract interpretation on imperative
-code (\S\ref{sec:fromcode}).
-
-\subsection{Lattice model of regions}
-\label{sec:domain}
-
-%\begin{defn}[Extended integers]
-%  We define the set \zinf{} as
-%  $\mathbb{Z}$ extended with $\infty$ and $-\infty$. For any $a$ in
-%  \zinf{}, we have $-\infty \leq a \leq \infty$. The resulting set is
-%  a total order with top and bottom elements $-\infty$ and
-%  $\infty$ respectively.
-%\end{defn}
-
-\begin{defn}[Index scheme]
-  An $N$-dimensional \emph{index scheme} is a vector of length $N$ of
-  integer sets, \ie{}, a member of $\mathcal{P}(\bz)^{N}$.
-%  An equivalent view of these vectors is as an $N$-times finite Cartesian
-%  product on subsets of $\bz$.
-  We use $S, T, U$ to denote index schemes. Henceforth, we assume index schemes
-  are all $N$-dimensional for some $N$.
-
-  Index schemes can be \emph{projected} in the $i^{\textit{th}}$ dimension by
-  $\pi_i : {\mathcal{P}(\bz{})}^N \to \mathcal{P}(\bz{})$. For an index scheme
-  $I$, we refer to $\pi_i(S)$ as the \emph{$i^{th}$ component} of $S$. We assume
-  that $i$, when used for projection, always lies between $1$ and $N$.
-\end{defn}
-
-\begin{restatable}{lem}{vectorIntersect}
-\label{lem:vector-intersect}
-  Intersection distributes over index schemes. That is, for index schemes $S, T
-  \in \mathcal{P}(\bz)^N$
-%
-  \begin{equation*}
-    S \cap T = \prod_{i = 1}^{N} \pi_i(S) \cap \pi_i(T)
-  \end{equation*}
-\end{restatable}
-
-Union does not distribute over index schemes, however, a more restricted
-property holds.
-
-\begin{restatable}{lem}{vectorUnion}
-\label{lem:vector-union}
-$S$ and $T$ are index schemes such that $\pi_i(S) = \pi_i(T)$ for all $1 \leq i
-\leq N$ apart from some dimension $k$, then:
-%
-  \begin{equation*}
-    S \cup  T
-    =
-    \pi_1(S) \times \cdots \times
-    (\pi_k(S) \cup \pi_k(T)) \times \cdots \times
-    \pi_N(S)
-  \end{equation*}
-\end{restatable}
-%
-\begin{defn}[Intervals with an optional hole]
-  We define an extended notion of closed interval on $\bz{}$ which may contain a
-  \emph{hole} at the origin, written \interv{a}{b}{c} where $a$ and $b$ are
-  drawn from \bz{} with $a \leq 0 \leq b$ and $c$ is drawn from $\mathbb{B} = \{
-  \mathsf{true}, \mathsf{false} \}$. Intervals are interpreted as sets as
-  follows:
-%
-  \begin{equation*}
-    \interv{a}{b}{c} \triangleq
-      \{ n \mid a \leq n \leq b \wedge (\neg c \implies n \neq 0) \}
-  \end{equation*}
-
-  We also add the distinguished interval $\interv{-\infty}{\infty}{}$, which is simply an
-  alias for \bz{}, but this notation prevents handling infinite interval
-  separately in the following definitions, theorems, and proofs. Here, $-\infty$
-  and $\infty$ behaves like top and bottom elements to \bz{} respectively.
-
-  We denote the set of all such intervals (sets) as $\textit{Interval}$. If the
-  superscript to the interval is omitted it is treated as $\mathsf{true}$.
-\end{defn}
-%
-\begin{restatable}{lem}{intervalIdentities}
- \label{lem:interval-identities}
-  We have the following dual identities for \bz{} intervals:
-%
-  \begin{align*}
-    \interv{a}{b}{c} \cap \interv{d}{e}{f} & =
-      \interv{\max \{a,d\}}{\min \{b,e\}}{c \wedge f} \\
-    \interv{a}{b}{c} \cup \interv{d}{e}{f} & =
-      \interv{\min \{a,d\}}{\max \{b,e\}}{c \vee f}
-  \end{align*}
-\end{restatable}
-%
-
-We define two specialisations (and subset spaces) of index scheme:
-\emph{subscript scheme} and \emph{interval scheme}:
-
-\begin{defn}[Interval scheme]
-  An interval scheme is a finite Cartesian product of intervals on \bz{},
-  denoted by the set $\textit{Interval}^N$ for a product of $N$ intervals.
-\end{defn}
-
-\begin{defn}[Subscript scheme]
-  A \emph{subscript scheme} is an index scheme scheme where:
-  %
-  \begin{equation*}
-    \forall i.\ 1 \leq i \leq N \implies
-      \pi_i (S) = \{ p \}
-      \; \vee \;
-      \pi_i (S) = \interv{-\infty}{\infty}{}
-    \end{equation*}
-%
-  That is, the $i^{th}$ component of the set is either a singleton in \bz{} or
-  the infinite interval.
-\end{defn}
-%
-\dnote{May want a predicate as well}
-
-\begin{defn}[Region]
-  A region is an index scheme and \region{N} is the set of all regions (\ie{},
-  $\region{N} \subseteq \mathcal{P}(\bz)^N)$. The set of all regions is  defined
-  as the smallest set satisfying the following:
-%
-  \begin{enumerate}
-    \item If $R$ is in $\textit{Interval}^N$, then $R$ is in \region{N}.
-    \item If $R$ and $S$ are in \region{N}, then so are $R \cap S$ and
-      $R \cup S$.
-  \end{enumerate}
-\end{defn}
-%
-\begin{restatable}{prop}{regionLattice}
-  \label{prop:regionLattice}
-  $(\region{N},\cup,\cap,\subseteq)$ is a bounded distributive lattice with top
-  $\top = \bz{}^N$ and bottom $\bot = \emptyset$.
-\end{restatable}
-%
-\begin{proof}
-  Straightforward, the join and meet are mapped to $\cup$ and $\cap$, the set is
-  inductively designed to be closed under these operations. Union and
-  intersection are associative, commutative, and absroptive under closed sets
-  and then they are also for \region{N}. This is enough to show that, it is a
-  lattice.
-
-  Further, we have $\bz{}^N \cap R = R$ and $\emptyset \cup R = R$ with
-  $\bz{}^N$ and $\emptyset$ belonging to \region{N}. This makes the lattice a
-  bounded one.
-
-  Finally, the lattice is distributive since union distributes over intersection
-  and vice versa when the set is closed under these operations.
-\end{proof}
-
-\begin{defn}
-  $\mathsf{Mult}$ and $\mathsf{Approx}$ are parametric labelled variant types
-  with injections given by their definition:
-%
-  \begin{align*}
-    \mathsf{Mult} \;\; a \;\; &
-      \triangleq \mathsf{mult} \; a \;\mid\; \mathsf{only} \; a \\
-    \mathsf{Approx} \;\; a \;\; &
-      \triangleq \mathsf{exact} \; a \;\mid\; \mathsf{lower} \; a \;\mid\;
-        \mathsf{upper} \; a
-  \end{align*}
-%
-  \eg{}, $\mathsf{lower}$ is an injection $\mathsf{lower} : a \to \mathsf{Approx}
-  \; a$ etc.
-  These will be used in the following subsection to give meaning to the
-  specification modifiers for approximation and multiplicity.
-\end{defn}
-
-\subsection{Denotational semantics for specifications}
-\label{sec:semantics}
-
-\noindent
-An interpretation function $\interp{-}_N$ maps closed\footnote{That
-  is, we assume there are no occurrences of \textit{rvar} in a
-  specification being modelled.  Any \emph{open} specification
-  containing region variables can be made closed by straightforward
-  syntactic substitution with a (closed) \textit{region}.}
-specifications to sets of $N$-dimensional index schemes with modifier
-information, \ie{} specifications are mapped to
-$\textsf{Mult} (\textsf{Approx} (\region{N}))$.
-
-The interpretation is overloaded on \emph{regions} in
-\Cref{subfig:region-model}. Various intermediate notions are used.
-
-\begin{defn}
-  Let $\textit{promote}_N : \mathbb{N}^+ \times \textit{Interval} \to
-  \textit{Interval}^N$ be a function generating an interval scheme such that if
-  $v$ is $\vecgen{N}{i}{\interv{a}{b}{c}}$, then $\pi_i(v) = \interv{a}{b}{c}$
-  and $\pi_j(v) = \bz{}$ in all other dimensions $j$.
-\end{defn}
-
-%
-\begin{figure}[!t]
-\begin{subfigure}[t]{0.5\textwidth}
-\begin{align*}
-  \interp{-}_N & : \textit{region} \rightarrow \region{N} \\
-%
-  \interp{\stencil{p}{$i$}{}{}}_N & =
-    \vecgen{N}{i}{\interv{0}{0}{\mathsf{true}}}\\
-%
-  \interp{\stencil{c}{$i$}{$k$}{\textcap{p}}}_N & =
-    \vecgen{N}{i}{\interv{-k}{k}{\interp{\textcap{p}}}} \\
-%
-  \interp{\stencil{f}{$i$}{$k$}{\textcap{p}}}_N & =
-    \vecgen{N}{i}{\interv{0}{k}{\interp{\textcap{p}}}} \\
-%
-  \interp{\stencil{b}{$i$}{$k$}{\textcap{p}}}_N & =
-  \vecgen{N}{i}{\interv{-k}{0}{\interp{\textcap{p}}}}
-\\
-  \interp{\texttt{\textcap{r} + \textcap{s}}}_N & =
-    \interp{\textcap{r}}_N \vee \interp{\textcap{s}}_N
-\\
-  \interp{\texttt{\textcap{r} * \textcap{s}}}_N & =
-    \interp{\textcap{r}}_N \wedge \interp{\textcap{s}}_N \\[-1em]
-\end{align*}
-\caption{Interpretation of regions}
-\label{subfig:region-model}
-\end{subfigure}
-\hspace{1em}
-\begin{subfigure}[t]{0.4\textwidth}
-\begin{align*}
-\interpApprox{-} & : \textit{approx} \rightarrow (A \rightarrow
-  \textsf{Approx} \, A) \\
-\interpApprox{\texttt{atLeast}} & = \mathsf{lower} \\
-  \interpApprox{\texttt{atMost}} & = \mathsf{upper} \\
-  \interpApprox{\epsilon} & = \mathsf{exact} \\ \\
-  \interpMult{-} & : \textit{mult} \rightarrow (A \rightarrow
-  \textsf{Mult} \, A) \\
-  \interpMult{\texttt{readOnce}} & = \mathsf{once} \\
-  \interpMult{\epsilon} & = \mathsf{mult}
-\end{align*}
-\caption{Interpretation of modifiers}
-\label{subfig:modifier-model}
-\end{subfigure}
-\label{fig:semantics}
-\caption{Semantic model of specifications}
-\end{figure}
-
-The first four equations of \Cref{subfig:region-model} model region 
-contants. The final two equations model the \term{+} and \term{*}
-operators in terms of the join (union) and meet (intersection)
-of interval schemes. Thus regions are modelled as members
-of $\region{N}$.
-
-We mark in our model the presence of modifiers such as
-\texttt{readOnce} and \texttt{atMost} as introduced in \Cref{}.
-Approximation modifiers are interpreted as injections into the
-$\mathsf{Approx}$ variant by $\interpApprox{}$ in
-\Cref{subfig:modifier-model}.  The $\textsf{Approx}$ type corresponds
-to the presence or absence of the spatial approximation modifier, with
-\textsf{exact} when there is no such modifier and \textsf{lower} and
-\textsf{upper} for \term{atLeast} and \term{atMost}. In a similar way,
-multiplicity modifiers are interpreted as injections in the
-$\mathsf{Mult}$ variant by $\interpMult{}$, corresponding to the to
-the presence or absence of the \term{readOnce} modifier as shown in
-\Cref{subfig:modifier-model}.
-
-\begin{defn}[Semantics of specifications]
-The intermediate interpretations of \Cref{fig:semantics}
-are composed to give a model for the top-level specification
-syntax as:
-%
-\begin{equation*}
-  \interp{\texttt{stencil \textcap{mult}, \textcap{approx}, \textcap{region}}}_N =
-    \interpMult{\textcap{mult}} \;
-           {(\interpApprox{\textcap{approx}} \;
-                    {\interp{\textcap{region}}_N)}}
-\end{equation*}
-\end{defn}
-%
-\begin{thm}
-The semantic model $\interp{-}_N$ of $N$-dimensional specifications
-is sound with respects to the equational theory of the language,
-that is:
-%
-\begin{equation*}
-\forall S, T, N . \quad
-S \equiv T \; \Rightarrow \;
-\interp{S}_N = \interp{T}_N
-\end{equation*}
-%
-We define the equational theory in \Cref{sec:equational-theory}.
-\end{thm}
-
-\subsection{Denotational semantics for array subscripts}
-\label{sec:fromcode}
-
-\begin{defn}
-  Recall array subscript terms of the form $a(\bar{e})$ from
-  \Cref{def:array-subs}. We interpret these terms with the partial
-  interpretation $\interp{-}^{\mathit{aterm}} : \textit{array-term}
-  \prightarrow{} \mathcal{P}(\bz{})^N$. The interpretation is defined when
-  all indices are either constant or neighbourhood indices as defined in
-  \Cref{def:neighbour-ix}.
-%
-  \begin{align*}
-    \interp{a(\bar{e})}^{\mathit{aterm}} =
-      \prod_{1 \leq i \leq N} \mathit{subscript}(\bar{e}_i) & &
-  %
-    \textit{subscript}(e) = \begin{cases}
-      \{ c \} & e \equiv i \pm c \\
-      \bz & e \; \mbox{is constant}
-    \end{cases}
-  \end{align*}
-\end{defn}
-
-\section{Analysis, Checking, and Inference}
-\label{sec:algorithms}
-\input{algorithms}
-
-\section{Evaluation}
-\label{sec:evaluation}
-
-\bibliography{references}
-
-\appendix
-% http://2017.splashcon.org/track/splash-2017-OOPSLA#Instructions-for-Authors
-% "There is no page limit for bibliographic references and appendices,
-% and, therefore, for the overall submission."
-\input{appendix}
-
-\end{document}
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: t
+%%% End:

### 2017-04-17 oopsla17/model.tex

--- 2017-04-17 oopsla17/oopsla17.tex
@@
-\documentclass[acmlarge,review,anonymous]{acmart}\settopmatter{printfolios=true}
-
-%% Some recommended packages.
-\usepackage{booktabs}   %% For formal tables:
-                        %% http://ctan.org/pkg/booktabs
-\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
-                        %% http://ctan.org/pkg/subcaption
-
-\usepackage{hyperref}
-\usepackage{natbib}
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{amsthm}
-\usepackage{syntax}
-\usepackage{indentfirst}
-\usepackage{cleveref}
-\usepackage{xcolor}
-\usepackage{multirow}
-\usepackage{siunitx} % For pretty-printing numeric values and SI units
-                     % of measure. e.g., the tabular column type S is
-                     % used to print nice-looking tables of numbers.
-\sisetup{ % defaults
-  group-separator={,},
-  group-minimum-digits={3},
-  output-decimal-marker={.},
-  table-format = 6
-}
-
-\usepackage{minted}
-\usepackage{etoolbox}
-\usepackage{enumitem}
-\usepackage{thmtools}
-\usepackage{thm-restate}
-
-\newcommand{\fortran}[1]{\mintinline{fortran}{#1}}
-
-\errorcontextlines 10000
-\usepackage{stencilmacros}
-
-%% Bibliography style
-\bibliographystyle{ACM-Reference-Format}
-%% Citation style
-%% Note: author/year citations are required for papers published as an
-%% issue of PACMPL.
-\citestyle{acmauthoryear}   %% For author/year citations
-
-\fvset{
-  linenos=true,
-  fontsize=\footnotesize,
-  breaklines=true,
-  breakafter=w),
-  xleftmargin=3\parindent,
-  numbersep=1em
-}
-
-\usemintedstyle{vs}
-
-\setlength{\grammarindent}{3em} % increase separation between LHS/RHS
-
-\newcommand{\inbar}{\ \ \textbar\ \ }
-
-% For evluation and empirical study
-% Defines the commands (headers)
-\input{resultsCmds}
-\input{studyResults}
-% renews the commands with actual data
-% \input{results}
-
-\definecolor{darkgreen}{rgb}{0.0,0.5,0.0}
-\definecolor{darkpurple}{rgb}{0.6,0.0,0.6}
-\definecolor{orange}{rgb}{0.8,0.4,0.0}
-\newcommand{\dnote}[1]{\textcolor{darkpurple}{Dom: #1}}
-\newcommand{\mnote}[1]{\textcolor{darkgreen}{Mistral: #1}}
-\newcommand{\anote}[1]{\textcolor{red}{Andy: #1}}
-
-\theoremstyle{definition}
-\newtheorem{defn}{Definition}
-
-\theoremstyle{plain}
-\newtheorem{thm}{Theorem}
-\newtheorem{lem}{Lemma}
-\newtheorem{prop}{Proposition}
-\newtheorem{case}{Case}
-
-\theoremstyle{remark}
-\newtheorem{remark}{Remark}
-
-\title{Verifying Spatial Properties of Array Computations}
-
-\author{Dominic Orchard}
-\affiliation{
-  \department{School of Computing}
-  \institution{University of Kent}
-}
-\email{d.a.orchard@kent.ac.uk}
-
-\author{Mistral Contrastin}
-\affiliation{
-  \department{Computer Laboratory}
-  \institution{University of Cambridge}
-}
-\email{Mistral.Contrastin@cl.cam.ac.uk}
-
-\author{Matthew Danish}
-\affiliation{
-  \department{Computer Laboratory}
-  \institution{University of Cambridge}
-}
-\email{Matthew.Danish@cl.cam.ac.uk}
-
-\author{Andrew Rice}
-\affiliation{
-  \department{Computer Laboratory}
-  \institution{University of Cambridge}
-}
-\email{Andrew.Rice@cl.cam.ac.uk}
-
-\begin{abstract}
-  Arrays computations are the workhorse of numerical modelling and
-  computational science. However, low-level manipulation of array
-  indices is a source of program error.  Many practitioners are aware
-  of the need to ensure program correctness, yet very few of the
-  techniques from the programming research community are applied by
-  scientists. We aim to change that by providing targetted lightweight
-  verification techniques for scientific code.  We focus on the all
-  too common mistake of array offset errors as a generalisation of
-  off-by-one errors.  Firstly, we report on a code analysis study on
-  twelve real-world computational science code base, identifying
-  common idioms of array usage and their spatial properties. This
-  provides much needed data on array programming idioms common in
-  scientific code.  From this data, we designed a lightweight
-  declarative specification language capturing the majority of array
-  access patterns via a small set of combinators. We detail a semantic
-  model, and the design and implementation of a verification tool for
-  our specification language, which both checks and infers
-  specifications.  We evaluate our tool on our corpus of scientific
-  code and give verification case studies of bug fixes that are
-  detected by our approach. We found roughly 60,000 targets for
-  specification across roughly 1.2 million lines of code, showing the
-  vast majority of array computations read from arrays in a pattern with
-  a simple, regular, static shape.
-\end{abstract}
-
-%  We focus on one error-prone aspect of numerical code: the 
-%  data access pattern of arrays.
-
-\begin{document}
-\maketitle
-
-\section{Introduction}
-\label{sec:introduction}
-\input{introduction}
-
-\section{An Empirical Study of Array Computations in Scientific Fortran code}
-\label{sec:study}
-\input{study}
-
-\section{A Specification Language for Shape}
-
-\begin{figure}[t]
-\begin{align*}
-\def\arraystretch{1}
-\setlength{\arraycolsep}{0.2em}
-\newcommand{\dimTy}{$\mathbb{N}_{>0}$}
-\newcommand{\depthRange}{$\mathbb{N}_{>0}$}
-\begin{array}{rl}
-\nonterm{specification} ::= & \nonterm{regionDec} \mid \nonterm{specDec} \\
-\nonterm{specDec} ::= & \term{stencil} \; \nonterm{spec} \;
-                        \texttt{::} \; v \\
-\nonterm{regionDec} ::= &  \texttt{region} \; \texttt{::} \; \nonterm{rvar} \; \texttt{=} \;
-                         \nonterm{region}\\[0.4em]
-%\nonterm{spec} ::= & \nonterm{spatial} \mid \nonterm{temporal}
-%\\[1em]
-\nonterm{spec} ::= & [\nonterm{mult},] \; [\nonterm{approx},] \; \nonterm{region} \\
-\nonterm{mult} ::= & \term{readOnce} \\
-\nonterm{approx} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
-\nonterm{region} ::= & \nonterm{rvar} \; \mid \;
-                       \stencil{p}{\dimTy}{}{} \; \\
-& {\qquad\; \mid \; \stencil{f}{\dimTy}{\depthRange}{\;[, \nonpointed]}} \\
-& {\qquad\; \mid \; \stencil{b}{\dimTy}{\depthRange}{\;[, \nonpointed]}} \\
-& {\qquad\; \mid \; \stencil{c}{\dimTy}{\depthRange}{\;[, \nonpointed]}} \\
-& {\qquad\; \mid \; \nonterm{region} \, \term{+}
-  \, \nonterm{region} \; \mid \; \nonterm{region} \; \term{*} \; \nonterm{region}} \\[0.25em]
-%\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{rvar}}
-%\\[0.5em]
-%\nonterm{temporal} ::= \; & \term{dependency} \; (v \; \{ , v \}) [, \texttt{mutual}]
-%  \\[0.5em]
-\nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+\\[-1em]
-\end{array}
-\end{align*}
-\caption{Specification syntax (EBNF grammar)}
-\label{fig:syntax}
-\end{figure}
-
-\subsection{Specification syntax}
-\label{subsec:syntax}
-
-\Cref{fig:syntax} gives the syntax of stencil specifications, which is
-detailed below. The entry point is the \nonterm{specification} production which
-splits into either a \emph{region declaration} or a \emph{specification
-declaration}.
-Regions comprise \emph{region constants} which are
-combined via region operators \term{+} and \term{*}.
-
-Region constants specify a finite interval in a 
-single dimension relative to the origin and are either \term{pointed},
-\term{forward}, \term{backward}, or \term{centered}. The region names are
-inspired by numerical analysis terminology, \eg{} the standard explicit method for
-approximating PDEs is known as the \emph{Forward Time, Centered Space} (FTCS)
-scheme~\citep{dawson1991finite}.
-
-Each region
-constant has a dimension identifier $d$ given as a positive natural number.
-Each constant except \term{pointed} has a depth
-parameter $n$ given as a positive natural number; \term{pointed}
-regions implicitly have a depth of $0$.
-
-A \term{forward} region of depth $n$ specifies a contiguous
-region in dimension $d$ starting at the origin. This corresponds
-to specifying neighbourhood indices in dimension $d$ ranging from $i$ to $i + n$
-for some induction variable $i$. Similarly, a
-\term{backward} region of depth $n$ corresponds to contiguous indices
-from $i$ to $i - n$ and \term{centered}
-of depth $n$ from $i - n$ to $i + n$. A \term{pointed}
-stencil specifies a neighbour index $i$. For example, the
-following shows four specifications with four consistent stencil
-kernels reading from arrays \term{a}, \term{b}, \term{c} and \term{d}:
-%%
-\begin{minted}{fortran}
-!= stencil forward(depth=2, dim=1) :: a
-e(i, 0) = a(i, 0) + a(i+1, 0) + a(i+2, 0)
-
-!= stencil backward(depth=2, dim=1) :: b
-e(i, j) = b(i, j) + b(i-1, j) + b(i-2, j)
-
-!= stencil centered(depth=1, dim=1) :: c
-e(i, j) = (c(j-1) + c(j) + c(j+1))/3.0
-
-!= stencil pointed(dim=3) :: d
-e(i, j) = d(0, 0, i)
-\end{minted}
-%%
-Not every dimension needs to be specified, \eg{},
-specifications on lines $1$, $4$, and $10$ leave some dimensions unspecified.
-The \term{forward}, \term{backward}, and \term{centered} regions may
-all have an additional attribute \term{nonpointed} which marks absence
-of the origin.  For example, the following is a
-\term{nonpointed} \term{backward} stencil
-%
-\begin{minted}{fortran}
-!= stencil backward(depth=2, dim=1, nonpointed) :: a
-b(i) = a(i-1) + 10*a(i-2)
-\end{minted}
-%
-Not every dimension needs to be specified, \eg{},
-specifications on lines $1$, $3$, and $7$ leave some dimensions
-unspecified which leaves indices in these dimensions unconstrained.
-
-The \term{forward}, \term{backward}, and \term{centered} regions may
-all have an additional attribute \term{nonpointed} which marks absence
-of the origin.  For example, the following is a
-\term{nonpointed} \term{backward} stencil
-%
-\begin{minted}{fortran}
-!= stencil backward(depth=2, dim=1, nonpointed) :: a
-b(i) = a(i-1) + 10*a(i-2)
-\end{minted}
-
-\paragraph{Combining regions}
-
-The region operators \term{+} and \term{*} respectively combine
-regions by union and intersection. The intersection of two regions
-$r \term{*} s$ means that any indices in the specified code must be
-consistent with both $r$ and $s$ simultaneously.
- Dually, for the union of two regions
- $r \term{+} s$ means that indices in the specified code must be
- in consistent with one of $r$ or $s$, or both.
-For example, the following \emph{nine-point stencil}
-has a specification given by the product of two \texttt{centered}
-regions in each dimension:
-%%
-\begin{minted}[breakindent=2.9em]{fortran}
-x = a(i, j)   + a(i-1, j)   + a(i+1, j)
-y = a(i, j-1) + a(i-1, j-1) + a(i+1, j-1)
-z = a(i, j+1) + a(i-1, j+1) + a(i+1, j+1)
-!= stencil centered(depth=1, dim=1) * centered(depth=1, dim=2) :: a
-b(i, j) = (x + y + z) / 9.0
-\end{minted}
-%
-%This pattern is common in image convolution applications.
-The specification ranges over the
-values that flow to the array subscript on the left-hand side,
-and so ranges over the intermediate assignments to \term{x},
-\term{y}, and \term{z}. Each index in the code is consistent
-with both specifications simultaneously, \eg{}, \texttt{a(i-1, j+1)}
-is within the centered region in dimension $1$ and the centered region
-in dimension $2$.
-
-The union of two regions $r \term{+} s$ means that any indices
-in the specified code must be consistent with either of $r$ or $s$.
-For example, the following gives the specification of a five-point
-stencil which is the sum of two compound \texttt{pointed} and
-\texttt{centered} regions in each dimension:
-%
-\begin{minted}{fortran}
-!= stencil centered(depth=1, dim=1)*pointed(dim=2) + centered(depth=1, dim=2)*pointed(dim=1) :: a
-b(i,j) = -4*a(i,j) + a(i-1,j) + a(i+1,j) + a(i,j-1) + a(i,j+1)
-\end{minted}
-%%
-Here the left-hand side of \texttt{+} says that when the second dimension
-(induction variable $j$) is fixed at the origin, the first dimension
-(induction variable $i$) accesses the immediate vicinity of the origin
-(to depth of one). The right hand side of \texttt{+} is similar but the dimensions are reversed.
-This reflects the symmetry under rotation of the five-point stencil.
-
-\paragraph{Region declarations and variables}
-
-Region specifications can be assigned to region variables via
-region declarations. For example, the shape of a
-``\emph{Roberts cross}'' edge-detection convolution~\cite{davis1975survey}
-can be stated:
-%%
-\begin{minted}{fortran}
-!= region :: r1 = forward(depth=1, dim=1)
-!= region :: r2 = forward(depth=1, dim=2)
-!= region :: robertsCross = r1*r2
-!= stencil robertsCross :: a
-\end{minted}
-This is useful for common patterns, such as the five-point
-pattern, as the regions can be defined once and reused.
-%%
-\paragraph{Modifiers}
-%%
-Region specifications can be modified
-by \emph{approximation} and \emph{multiplicity} information
-(in \textit{spec} in \Cref{fig:syntax}).
-The \texttt{readOnce} modifier enforces that no index appears more
-than once (that is, its multiplicity is one). For example, all of
-the previous examples could have \texttt{readOnce} added:
-%
-\begin{minted}{fortran}
-!= stencil readOnce, backward(depth=2, dim=1) :: a
-b(i+1) = a(i) + a(i-1) + a(i-2)
-\end{minted}
-%
-This specification would be invalid if any of the
-array subscripts were repeated. This modifier provides a way to
-rule out any accidental repetition of array subscripts.
-The notion is similar to that of linear types~\cite{wadler1990linear}, where a value must be used
-exactly once. We opt for the more informative and easily understood name
-\texttt{readOnce}. This modifier is optional, so it need not
-be present even if the stencil is linear.
-
-In some cases, it is useful to give a lower and/or upper bound for a
-stencil. This can be done using either the \term{atMost} or
-\term{atLeast} modifiers. This is particularly useful in situations
-where there is a non-contiguous stencil pattern, which cannot be expressed
-precisely in our syntax. For example:
-%
-\begin{minted}{fortran}
-!= stencil atLeast, pointed(dim=1)         :: a
-!= stencil atMost, forward(depth=4, dim=1) :: a
-b(i) = a(i) + a(i+4)
-\end{minted}
-
-\section{Semantic model}
-\label{sec:model}
-
 \noindent
@@
 
-\section{Analysis, Checking, and Inference}
-\label{sec:algorithms}
-\input{algorithms}
-
-\section{Evaluation}
-\label{sec:evaluation}
-
-\bibliography{references}
-
-\appendix
-% http://2017.splashcon.org/track/splash-2017-OOPSLA#Instructions-for-Authors
-% "There is no page limit for bibliographic references and appendices,
-% and, therefore, for the overall submission."
-\input{appendix}
-
-\end{document}
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: t
+%%% End:

### 2017-04-17 oopsla17/oopsla17.tex

@@
 \label{sec:lang}
-
-\begin{figure}[t]
-\begin{align*}
-\def\arraystretch{1}
-\setlength{\arraycolsep}{0.2em}
-\newcommand{\dimTy}{$\mathbb{N}_{>0}$}
-\newcommand{\depthRange}{$\mathbb{N}_{>0}$}
-\begin{array}{rl}
-\nonterm{specification} ::= & \nonterm{regionDec} \mid \nonterm{specDec} \\
-\nonterm{specDec} ::= & \term{stencil} \; \nonterm{spec} \;
-                        \texttt{::} \; v \\
-\nonterm{regionDec} ::= &  \texttt{region} \; \texttt{::} \; \nonterm{rvar} \; \texttt{=} \;
-                         \nonterm{region}\\[0.4em]
-%\nonterm{spec} ::= & \nonterm{spatial} \mid \nonterm{temporal}
-%\\[1em]
-\nonterm{spec} ::= & [\nonterm{mult},] \; [\nonterm{approx},] \; \nonterm{region} \\
-\nonterm{mult} ::= & \term{readOnce} \\
-\nonterm{approx} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
-\nonterm{region} ::= & \nonterm{rvar} \; \mid \;
-                       \stencil{p}{\dimTy}{}{} \; \\
-& {\qquad\; \mid \; \stencil{f}{\dimTy}{\depthRange}{\;[, \nonpointed]}} \\
-& {\qquad\; \mid \; \stencil{b}{\dimTy}{\depthRange}{\;[, \nonpointed]}} \\
-& {\qquad\; \mid \; \stencil{c}{\dimTy}{\depthRange}{\;[, \nonpointed]}} \\
-& {\qquad\; \mid \; \nonterm{region} \, \term{+}
-  \, \nonterm{region} \; \mid \; \nonterm{region} \; \term{*} \; \nonterm{region}} \\[0.25em]
-%\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{rvar}}
-%\\[0.5em]
-%\nonterm{temporal} ::= \; & \term{dependency} \; (v \; \{ , v \}) [, \texttt{mutual}]
-%  \\[0.5em]
-\nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+\\[-1em]
-\end{array}
-\end{align*}
-\caption{Specification syntax (EBNF grammar)}
-\label{fig:syntax}
-\end{figure}
-
-\subsection{Specification syntax}
-\label{subsec:syntax}
-
-\Cref{fig:syntax} gives the syntax of stencil specifications, which is
-detailed below. The entry point is the \nonterm{specification} production which
-splits into either a \emph{region declaration} or a \emph{specification
-declaration}.
-Regions comprise \emph{region constants} which are
-combined via region operators \term{+} and \term{*}.
-
-Region constants specify a finite interval in a 
-single dimension relative to the origin and are either \term{pointed},
-\term{forward}, \term{backward}, or \term{centered}. The region names are
-inspired by numerical analysis terminology, \eg{} the standard explicit method for
-approximating PDEs is known as the \emph{Forward Time, Centered Space} (FTCS)
-scheme~\citep{dawson1991finite}.
-
-Each region
-constant has a dimension identifier $d$ given as a positive natural number.
-Each constant except \term{pointed} has a depth
-parameter $n$ given as a positive natural number; \term{pointed}
-regions implicitly have a depth of $0$.
-
-A \term{forward} region of depth $n$ specifies a contiguous
-region in dimension $d$ starting at the origin. This corresponds
-to specifying neighbourhood indices in dimension $d$ ranging from $i$ to $i + n$
-for some induction variable $i$. Similarly, a
-\term{backward} region of depth $n$ corresponds to contiguous indices
-from $i$ to $i - n$ and \term{centered}
-of depth $n$ from $i - n$ to $i + n$. A \term{pointed}
-stencil specifies a neighbour index $i$. For example, the
-following shows four specifications with four consistent stencil
-kernels reading from arrays \term{a}, \term{b}, \term{c} and \term{d}:
-%%
-\begin{minted}{fortran}
-!= stencil forward(depth=2, dim=1) :: a
-e(i, 0) = a(i, 0) + a(i+1, 0) + a(i+2, 0)
-
-!= stencil backward(depth=2, dim=1) :: b
-e(i, j) = b(i, j) + b(i-1, j) + b(i-2, j)
-
-!= stencil centered(depth=1, dim=1) :: c
-e(i, j) = (c(j-1) + c(j) + c(j+1))/3.0
-
-!= stencil pointed(dim=3) :: d
-e(i, j) = d(0, 0, i)
-\end{minted}
-%%
-Not every dimension needs to be specified, \eg{},
-specifications on lines $1$, $4$, and $10$ leave some dimensions unspecified.
-The \term{forward}, \term{backward}, and \term{centered} regions may
-all have an additional attribute \term{nonpointed} which marks absence
-of the origin.  For example, the following is a
-\term{nonpointed} \term{backward} stencil
-%
-\begin{minted}{fortran}
-!= stencil backward(depth=2, dim=1, nonpointed) :: a
-b(i) = a(i-1) + 10*a(i-2)
-\end{minted}
-%
-Not every dimension needs to be specified, \eg{},
-specifications on lines $1$, $3$, and $7$ leave some dimensions
-unspecified which leaves indices in these dimensions unconstrained.
-
-The \term{forward}, \term{backward}, and \term{centered} regions may
-all have an additional attribute \term{nonpointed} which marks absence
-of the origin.  For example, the following is a
-\term{nonpointed} \term{backward} stencil
-%
-\begin{minted}{fortran}
-!= stencil backward(depth=2, dim=1, nonpointed) :: a
-b(i) = a(i-1) + 10*a(i-2)
-\end{minted}
-
-\paragraph{Combining regions}
-
-The region operators \term{+} and \term{*} respectively combine
-regions by union and intersection. The intersection of two regions
-$r \term{*} s$ means that any indices in the specified code must be
-consistent with both $r$ and $s$ simultaneously.
- Dually, for the union of two regions
- $r \term{+} s$ means that indices in the specified code must be
- in consistent with one of $r$ or $s$, or both.
-For example, the following \emph{nine-point stencil}
-has a specification given by the product of two \texttt{centered}
-regions in each dimension:
-%%
-\begin{minted}[breakindent=2.9em]{fortran}
-x = a(i, j)   + a(i-1, j)   + a(i+1, j)
-y = a(i, j-1) + a(i-1, j-1) + a(i+1, j-1)
-z = a(i, j+1) + a(i-1, j+1) + a(i+1, j+1)
-!= stencil centered(depth=1, dim=1) * centered(depth=1, dim=2) :: a
-b(i, j) = (x + y + z) / 9.0
-\end{minted}
-%
-%This pattern is common in image convolution applications.
-The specification ranges over the
-values that flow to the array subscript on the left-hand side,
-and so ranges over the intermediate assignments to \term{x},
-\term{y}, and \term{z}. Each index in the code is consistent
-with both specifications simultaneously, \eg{}, \texttt{a(i-1, j+1)}
-is within the centered region in dimension $1$ and the centered region
-in dimension $2$.
-
-The union of two regions $r \term{+} s$ means that any indices
-in the specified code must be consistent with either of $r$ or $s$.
-For example, the following gives the specification of a five-point
-stencil which is the sum of two compound \texttt{pointed} and
-\texttt{centered} regions in each dimension:
-%
-\begin{minted}{fortran}
-!= stencil centered(depth=1, dim=1)*pointed(dim=2) + centered(depth=1, dim=2)*pointed(dim=1) :: a
-b(i,j) = -4*a(i,j) + a(i-1,j) + a(i+1,j) + a(i,j-1) + a(i,j+1)
-\end{minted}
-%%
-Here the left-hand side of \texttt{+} says that when the second dimension
-(induction variable $j$) is fixed at the origin, the first dimension
-(induction variable $i$) accesses the immediate vicinity of the origin
-(to depth of one). The right hand side of \texttt{+} is similar but the dimensions are reversed.
-This reflects the symmetry under rotation of the five-point stencil.
-
-\paragraph{Region declarations and variables}
-
-Region specifications can be assigned to region variables via
-region declarations. For example, the shape of a
-``\emph{Roberts cross}'' edge-detection convolution~\cite{davis1975survey}
-can be stated:
-%%
-\begin{minted}{fortran}
-!= region :: r1 = forward(depth=1, dim=1)
-!= region :: r2 = forward(depth=1, dim=2)
-!= region :: robertsCross = r1*r2
-!= stencil robertsCross :: a
-\end{minted}
-This is useful for common patterns, such as the five-point
-pattern, as the regions can be defined once and reused.
-%%
-\paragraph{Modifiers}
-%%
-Region specifications can be modified
-by \emph{approximation} and \emph{multiplicity} information
-(in \textit{spec} in \Cref{fig:syntax}).
-The \texttt{readOnce} modifier enforces that no index appears more
-than once (that is, its multiplicity is one). For example, all of
-the previous examples could have \texttt{readOnce} added:
-%
-\begin{minted}{fortran}
-!= stencil readOnce, backward(depth=2, dim=1) :: a
-b(i+1) = a(i) + a(i-1) + a(i-2)
-\end{minted}
-%
-This specification would be invalid if any of the
-array subscripts were repeated. This modifier provides a way to
-rule out any accidental repetition of array subscripts.
-The notion is similar to that of linear types~\cite{wadler1990linear}, where a value must be used
-exactly once. We opt for the more informative and easily understood name
-\texttt{readOnce}. This modifier is optional, so it need not
-be present even if the stencil is linear.
-
-In some cases, it is useful to give a lower and/or upper bound for a
-stencil. This can be done using either the \term{atMost} or
-\term{atLeast} modifiers. This is particularly useful in situations
-where there is a non-contiguous stencil pattern, which cannot be expressed
-precisely in our syntax. For example:
-%
-\begin{minted}{fortran}
-!= stencil atLeast, pointed(dim=1)         :: a
-!= stencil atMost, forward(depth=4, dim=1) :: a
-b(i) = a(i) + a(i+4)
-\end{minted}
+\input{lang}
 
@@
 \label{sec:model}
-
-\noindent
-We define a lattice model of array access patterns which serves as both a
-denotational model of the semantics of our specification language and an
-abstract interpretation domain for source code. This model (1) serves to
-explain the meaning of our specifications; (2) is used in the inference and
-checking algorithms (\Cref{sec:analysis}); (3) justifies an equational theory
-for specifications in the next section; (4) is used to optimise specifications
-using lattice identities; and (5) can be used to guide correct implementations.
-
-The model is defined over vectors of sets of integers which we call
-\emph{index schemes}.  As an initial informal example, consider the
-following simple stencil computation:
-%
-\begin{minted}{fortran}
-do i = 1, n
-  b(i) = a(i,0) + a(i+1,0)
-end do
-\end{minted}
-%
-The access pattern on array \term{a}, relative
-to induction variable \term{i}, is captured
-by a vector of length 2 containing integer sets $\langle{\{0, 1\},
-  \mathbb{Z}\}}\rangle$. This describes that, in the first dimension,
-the array is read at offsets of $0$ and $1$ from an induction
-variable. In the second dimension, the index is unconstrained
-as it is a constant.
-
-Index schemes form a lattice which provides a rich set of equations
-and properties, which we exploit. We first set up the 
-domain of the model (\S\ref{sec:domain}), using it to define a semantics
-for our specification language (\S\ref{sec:semantics}) and then as
-the target for an abstract interpretation on imperative
-code (\S\ref{sec:fromcode}).
-
-\subsection{Lattice model of regions}
-\label{sec:domain}
-
-%\begin{defn}[Extended integers]
-%  We define the set \zinf{} as
-%  $\mathbb{Z}$ extended with $\infty$ and $-\infty$. For any $a$ in
-%  \zinf{}, we have $-\infty \leq a \leq \infty$. The resulting set is
-%  a total order with top and bottom elements $-\infty$ and
-%  $\infty$ respectively.
-%\end{defn}
-
-\begin{defn}[Index scheme]
-  An $N$-dimensional \emph{index scheme} is a vector of length $N$ of
-  integer sets, \ie{}, a member of $\mathcal{P}(\bz)^{N}$.
-%  An equivalent view of these vectors is as an $N$-times finite Cartesian
-%  product on subsets of $\bz$.
-  We use $S, T, U$ to denote index schemes. Henceforth, we assume index schemes
-  are all $N$-dimensional for some $N$.
-
-  Index schemes can be \emph{projected} in the $i^{\textit{th}}$ dimension by
-  $\pi_i : {\mathcal{P}(\bz{})}^N \to \mathcal{P}(\bz{})$. For an index scheme
-  $I$, we refer to $\pi_i(S)$ as the \emph{$i^{th}$ component} of $S$. We assume
-  that $i$, when used for projection, always lies between $1$ and $N$.
-\end{defn}
-
-\begin{restatable}{lem}{vectorIntersect}
-\label{lem:vector-intersect}
-  Intersection distributes over index schemes. That is, for index schemes $S, T
-  \in \mathcal{P}(\bz)^N$
-%
-  \begin{equation*}
-    S \cap T = \prod_{i = 1}^{N} \pi_i(S) \cap \pi_i(T)
-  \end{equation*}
-\end{restatable}
-
-Union does not distribute over index schemes, however, a more restricted
-property holds.
-
-\begin{restatable}{lem}{vectorUnion}
-\label{lem:vector-union}
-$S$ and $T$ are index schemes such that $\pi_i(S) = \pi_i(T)$ for all $1 \leq i
-\leq N$ apart from some dimension $k$, then:
-%
-  \begin{equation*}
-    S \cup  T
-    =
-    \pi_1(S) \times \cdots \times
-    (\pi_k(S) \cup \pi_k(T)) \times \cdots \times
-    \pi_N(S)
-  \end{equation*}
-\end{restatable}
-%
-\begin{defn}[Intervals with an optional hole]
-  We define an extended notion of closed interval on $\bz{}$ which may contain a
-  \emph{hole} at the origin, written \interv{a}{b}{c} where $a$ and $b$ are
-  drawn from \bz{} with $a \leq 0 \leq b$ and $c$ is drawn from $\mathbb{B} = \{
-  \mathsf{true}, \mathsf{false} \}$. Intervals are interpreted as sets as
-  follows:
-%
-  \begin{equation*}
-    \interv{a}{b}{c} \triangleq
-      \{ n \mid a \leq n \leq b \wedge (\neg c \implies n \neq 0) \}
-  \end{equation*}
-
-  We also add the distinguished interval $\interv{-\infty}{\infty}{}$, which is simply an
-  alias for \bz{}, but this notation prevents handling infinite interval
-  separately in the following definitions, theorems, and proofs. Here, $-\infty$
-  and $\infty$ behaves like top and bottom elements to \bz{} respectively.
-
-  We denote the set of all such intervals (sets) as $\textit{Interval}$. If the
-  superscript to the interval is omitted it is treated as $\mathsf{true}$.
-\end{defn}
-%
-\begin{restatable}{lem}{intervalIdentities}
- \label{lem:interval-identities}
-  We have the following dual identities for \bz{} intervals:
-%
-  \begin{align*}
-    \interv{a}{b}{c} \cap \interv{d}{e}{f} & =
-      \interv{\max \{a,d\}}{\min \{b,e\}}{c \wedge f} \\
-    \interv{a}{b}{c} \cup \interv{d}{e}{f} & =
-      \interv{\min \{a,d\}}{\max \{b,e\}}{c \vee f}
-  \end{align*}
-\end{restatable}
-%
-
-We define two specialisations (and subset spaces) of index scheme:
-\emph{subscript scheme} and \emph{interval scheme}:
-
-\begin{defn}[Interval scheme]
-  An interval scheme is a finite Cartesian product of intervals on \bz{},
-  denoted by the set $\textit{Interval}^N$ for a product of $N$ intervals.
-\end{defn}
-
-\begin{defn}[Subscript scheme]
-  A \emph{subscript scheme} is an index scheme scheme where:
-  %
-  \begin{equation*}
-    \forall i.\ 1 \leq i \leq N \implies
-      \pi_i (S) = \{ p \}
-      \; \vee \;
-      \pi_i (S) = \interv{-\infty}{\infty}{}
-    \end{equation*}
-%
-  That is, the $i^{th}$ component of the set is either a singleton in \bz{} or
-  the infinite interval.
-\end{defn}
-%
-\dnote{May want a predicate as well}
-
-\begin{defn}[Region]
-  A region is an index scheme and \region{N} is the set of all regions (\ie{},
-  $\region{N} \subseteq \mathcal{P}(\bz)^N)$. The set of all regions is  defined
-  as the smallest set satisfying the following:
-%
-  \begin{enumerate}
-    \item If $R$ is in $\textit{Interval}^N$, then $R$ is in \region{N}.
-    \item If $R$ and $S$ are in \region{N}, then so are $R \cap S$ and
-      $R \cup S$.
-  \end{enumerate}
-\end{defn}
-%
-\begin{restatable}{prop}{regionLattice}
-  \label{prop:regionLattice}
-  $(\region{N},\cup,\cap,\subseteq)$ is a bounded distributive lattice with top
-  $\top = \bz{}^N$ and bottom $\bot = \emptyset$.
-\end{restatable}
-%
-\begin{proof}
-  Straightforward, the join and meet are mapped to $\cup$ and $\cap$, the set is
-  inductively designed to be closed under these operations. Union and
-  intersection are associative, commutative, and absroptive under closed sets
-  and then they are also for \region{N}. This is enough to show that, it is a
-  lattice.
-
-  Further, we have $\bz{}^N \cap R = R$ and $\emptyset \cup R = R$ with
-  $\bz{}^N$ and $\emptyset$ belonging to \region{N}. This makes the lattice a
-  bounded one.
-
-  Finally, the lattice is distributive since union distributes over intersection
-  and vice versa when the set is closed under these operations.
-\end{proof}
-
-\begin{defn}
-  $\mathsf{Mult}$ and $\mathsf{Approx}$ are parametric labelled variant types
-  with injections given by their definition:
-%
-  \begin{align*}
-    \mathsf{Mult} \;\; a \;\; &
-      \triangleq \mathsf{mult} \; a \;\mid\; \mathsf{only} \; a \\
-    \mathsf{Approx} \;\; a \;\; &
-      \triangleq \mathsf{exact} \; a \;\mid\; \mathsf{lower} \; a \;\mid\;
-        \mathsf{upper} \; a
-  \end{align*}
-%
-  \eg{}, $\mathsf{lower}$ is an injection $\mathsf{lower} : a \to \mathsf{Approx}
-  \; a$ etc.
-  These will be used in the following subsection to give meaning to the
-  specification modifiers for approximation and multiplicity.
-\end{defn}
-
-\subsection{Denotational semantics for specifications}
-\label{sec:semantics}
-
-\noindent
-An interpretation function $\interp{-}_N$ maps closed\footnote{That
-  is, we assume there are no occurrences of \textit{rvar} in a
-  specification being modelled.  Any \emph{open} specification
-  containing region variables can be made closed by straightforward
-  syntactic substitution with a (closed) \textit{region}.}
-specifications to sets of $N$-dimensional index schemes with modifier
-information, \ie{} specifications are mapped to
-$\textsf{Mult} (\textsf{Approx} (\region{N}))$.
-
-The interpretation is overloaded on \emph{regions} in
-\Cref{subfig:region-model}. Various intermediate notions are used.
-
-\begin{defn}
-  Let $\textit{promote}_N : \mathbb{N}^+ \times \textit{Interval} \to
-  \textit{Interval}^N$ be a function generating an interval scheme such that if
-  $v$ is $\vecgen{N}{i}{\interv{a}{b}{c}}$, then $\pi_i(v) = \interv{a}{b}{c}$
-  and $\pi_j(v) = \bz{}$ in all other dimensions $j$.
-\end{defn}
-
-%
-\begin{figure}[!t]
-\begin{subfigure}[t]{0.5\textwidth}
-\begin{align*}
-  \interp{-}_N & : \textit{region} \rightarrow \region{N} \\
-%
-  \interp{\stencil{p}{$i$}{}{}}_N & =
-    \vecgen{N}{i}{\interv{0}{0}{\mathsf{true}}}\\
-%
-  \interp{\stencil{c}{$i$}{$k$}{\textcap{p}}}_N & =
-    \vecgen{N}{i}{\interv{-k}{k}{\interp{\textcap{p}}}} \\
-%
-  \interp{\stencil{f}{$i$}{$k$}{\textcap{p}}}_N & =
-    \vecgen{N}{i}{\interv{0}{k}{\interp{\textcap{p}}}} \\
-%
-  \interp{\stencil{b}{$i$}{$k$}{\textcap{p}}}_N & =
-  \vecgen{N}{i}{\interv{-k}{0}{\interp{\textcap{p}}}}
-\\
-  \interp{\texttt{\textcap{r} + \textcap{s}}}_N & =
-    \interp{\textcap{r}}_N \vee \interp{\textcap{s}}_N
-\\
-  \interp{\texttt{\textcap{r} * \textcap{s}}}_N & =
-    \interp{\textcap{r}}_N \wedge \interp{\textcap{s}}_N \\[-1em]
-\end{align*}
-\caption{Interpretation of regions}
-\label{subfig:region-model}
-\end{subfigure}
-\hspace{1em}
-\begin{subfigure}[t]{0.4\textwidth}
-\begin{align*}
-\interpApprox{-} & : \textit{approx} \rightarrow (A \rightarrow
-  \textsf{Approx} \, A) \\
-\interpApprox{\texttt{atLeast}} & = \mathsf{lower} \\
-  \interpApprox{\texttt{atMost}} & = \mathsf{upper} \\
-  \interpApprox{\epsilon} & = \mathsf{exact} \\ \\
-  \interpMult{-} & : \textit{mult} \rightarrow (A \rightarrow
-  \textsf{Mult} \, A) \\
-  \interpMult{\texttt{readOnce}} & = \mathsf{once} \\
-  \interpMult{\epsilon} & = \mathsf{mult}
-\end{align*}
-\caption{Interpretation of modifiers}
-\label{subfig:modifier-model}
-\end{subfigure}
-\label{fig:semantics}
-\caption{Semantic model of specifications}
-\end{figure}
-
-The first four equations of \Cref{subfig:region-model} model region 
-contants. The final two equations model the \term{+} and \term{*}
-operators in terms of the join (union) and meet (intersection)
-of interval schemes. Thus regions are modelled as members
-of $\region{N}$.
-
-We mark in our model the presence of modifiers such as
-\texttt{readOnce} and \texttt{atMost} as introduced in \Cref{}.
-Approximation modifiers are interpreted as injections into the
-$\mathsf{Approx}$ variant by $\interpApprox{}$ in
-\Cref{subfig:modifier-model}.  The $\textsf{Approx}$ type corresponds
-to the presence or absence of the spatial approximation modifier, with
-\textsf{exact} when there is no such modifier and \textsf{lower} and
-\textsf{upper} for \term{atLeast} and \term{atMost}. In a similar way,
-multiplicity modifiers are interpreted as injections in the
-$\mathsf{Mult}$ variant by $\interpMult{}$, corresponding to the to
-the presence or absence of the \term{readOnce} modifier as shown in
-\Cref{subfig:modifier-model}.
-
-\begin{defn}[Semantics of specifications]
-The intermediate interpretations of \Cref{fig:semantics}
-are composed to give a model for the top-level specification
-syntax as:
-%
-\begin{equation*}
-  \interp{\texttt{stencil \textcap{mult}, \textcap{approx}, \textcap{region}}}_N =
-    \interpMult{\textcap{mult}} \;
-           {(\interpApprox{\textcap{approx}} \;
-                    {\interp{\textcap{region}}_N)}}
-\end{equation*}
-\end{defn}
-%
-\begin{thm}
-The semantic model $\interp{-}_N$ of $N$-dimensional specifications
-is sound with respects to the equational theory of the language,
-that is:
-%
-\begin{equation*}
-\forall S, T, N . \quad
-S \equiv T \; \Rightarrow \;
-\interp{S}_N = \interp{T}_N
-\end{equation*}
-%
-We define the equational theory in \Cref{sec:equational-theory}.
-\end{thm}
-
-\subsection{Denotational semantics for array subscripts}
-\label{sec:fromcode}
-
-\begin{defn}
-  Recall array subscript terms of the form $a(\bar{e})$ from
-  \Cref{def:array-subs}. We interpret these terms with the partial
-  interpretation $\interp{-}^{\mathit{aterm}} : \textit{array-term}
-  \prightarrow{} \mathcal{P}(\bz{})^N$. The interpretation is defined when
-  all indices are either constant or neighbourhood indices as defined in
-  \Cref{def:neighbour-ix}.
-%
-  \begin{align*}
-    \interp{a(\bar{e})}^{\mathit{aterm}} =
-      \prod_{1 \leq i \leq N} \mathit{subscript}(\bar{e}_i) & &
-  %
-    \textit{subscript}(e) = \begin{cases}
-      \{ c \} & e \equiv i \pm c \\
-      \bz & e \; \mbox{is constant}
-    \end{cases}
-  \end{align*}
-\end{defn}
+\input{model}
 

### 2017-04-17 oopsla17/algorithms.tex
Improve on union normal form explanation

@@
 
-We need to address the problem of representation in the model both for observed
-array terms converted into subscripts schemes and for specifications converted
-into interval schemes. In each case, the objects are sets and union and
-intersection are the combinators. However, they are often infinite sets due to
-infinite interval and constant indices being converted to \bz{}. This calls for
-a finite (and ideally compact) representation for indices.
+We need to address the problem of representation in the model for index schemes.
+In each case, the objects are sets and union and intersection are the
+combinators. However, they are often infinite sets due to \bz{} in Cartesian
+product. This happens when constant indices of array terms are converted to
+\bz{} in subscript schemes and when infinite interval is generated in interval
+schemes. This calls for a finite (and ideally compact) representation for
+indices.
 
@@
 singleton sets might easily be empty, which disqualifies it from being a
-subscript index. In the following section we will not rely on intersection of
-subscript indices, so that is okay.
+subscript index. In the following sections we will not rely on intersection of
+subscript indices, so that is not a problem.
 
-Union is, however, not as flexible as intersection. \Cref{lem:vector-union} does
+Union, however, is not as flexible as intersection. \Cref{lem:vector-union} does
 not form a closure for indexing schemes in general. For this reason, we
@@
 to put unions into a cons tree making the unions effectively a non-empty list.
-We do not, however, attempt to canonicalise the representation.
+We do not, however, attempt to canonicalise the representation. Later on
+\Cref{lem:vector-union} and index scheme subset are used to reduce the size of
+union normal form.
 

### 2017-04-17 oopsla17/algorithms.tex
Fix the subset rel of Region and use macro for region

@@
   \mathit{consistent} & :
-    \mathsf{Mult}(\mathcal{P}(\mathcal{P}(\bz{})^N)) \times
+    \mathsf{Mult}(\mathcal{P}(\bz{}^N) \times
     \mathsf{Mult}(\mathsf{Approx}(\region{N})) \to \mathbb{B} & \\
@@
 %
-  \mathit{consistent'} & :
-    \mathcal{P}(\mathcal{P}(\bz{})^N) \times
+  \textit{consistent'} & :
+    \mathcal{P}(\bz{}^N) \times
     \mathsf{Approx}(\region{N}) \to \mathbb{B} & \\
 %
-  \mathit{consistent'}(\mathit{ix}, \mathit{model}) & = \begin{cases}
+  \textit{consistent'}(\mathit{ix}, \mathit{model}) & = \begin{cases}
     m = \mathit{ix} & \mathit{model} = \mathsf{exact}(m) \\
@@
 A covering of (possibly overlapping) the subscript space
-is calculated as a value of $Region_N$. This is computed iteratively
+is calculated as a value of \region{N}. This is computed iteratively
 by the following steps until a fixed-point is reached:
@@
 \dnote{We probably need to say something about classifying index
-schemes are \textit{Region$_N$}?}
+schemes are \region{N}?}
 

### 2017-04-17 oopsla17/algorithms.tex
simplifying coalescing
Fix typo
simplified coalescing- back to the previous definition

@@
 %
+\noindent
 We provide an inference procedure for generating specifications from
@@
 array subscripts into a set of subscript schemes in
-$\mathcal{P}(\bz)^N$. These subscripts are then coalesced interatively
+$\mathcal{P}(\bz)^N$. These subscripts are then coalesced iteratively
 into sets of index schemes. We denote sets of index schemes by
@@
 \end{equation*}
-
+%
 %For each array variable $a \in \mathsf{dom}(U)$, the algorithm
@@
 %groups of schemes in $U(a)$.
-
-We explain the algorithm in two parts. First, we show how \emph{adjacent} index
-schemes are coalesced into smaller index schemes while remaining in union normal
-form. Then, we explain how this smaller unioned indexing schemes are converted
-to specifications.
+%
+We explain the algorithm in two parts. First, we show how
+\emph{contiguous} index schemes are coalesced into a smaller set of
+index schemes while remaining in union normal form. Then, we explain
+how this smaller unioned indexing schemes are converted to
+specifications.
 
 \subsubsection{Covering}
-A covering of (possibly overlapping) the subscript space
-is calculated as a value of \region{N}. This is computed iteratively
-by the following steps until a fixed-point is reached:
-
-\begin{description}[leftmargin=0.5em]
-
-\item[Coalesce intervals]
-%
-Index schemes which are \emph{contiguous} in one dimension are
-coalesced.
+A covering of (possibly overlapping) intervals is calculated as a
+value of \region{N} by coalescing \emph{contiguous} index schemes
+until a fixed-point is reached.
 
 \newcommand{\contig}[2]{\mathsf{contig}(#1, #2)}
-\begin{defn}{[Contiguous]}
-Two index schemes $S$ and $T$ are \emph{contiguous}
-written $\contig{S}{T}$,
-iff $\pi_i(S) = \pi_i(T)$
-for all $1 \leq i \leq N$ apart from some dimension $k$ such that
-$\pi_k(S) = [a, b]$ and $\pi_k(T) = [b+1, c]$.
+\begin{defn}[Contiguity]
+  \label{def:contiguity}
+  Two index schemes $S$ and $T$ are \emph{contiguous} written
+  $\contig{S}{T}$, iff $\pi_i(S) = \pi_i(T)$ for all $1 \leq i \leq N$
+  apart from some dimension $k$ such that $\pi_k(S) = [a, b]$ and
+  $\pi_k(T) = [b+1, c]$.
 \end{defn}
@@
 A coalesced set of index schemes is computed by:%
+%
 \begin{align*}
-\mathsf{coalesce1}(S, \mathcal{S}) & =
- \{\,S \cup T \mid
-T \in \mathcal{S} \, \wedge \,
-\contig{S}{T} \} \\
-\mathsf{coalesce}(\mathcal{S}) & =
-\bigcup_{S \in \mathcal{S}} \mathsf{coalesce}(
-\mathsf{coalesce1}(S, \mathcal{S})
-\cup (\mathcal{S} - \{S\})) \\
-\mathsf{coalesceTop}(\mathcal{S}) & =
- \mathsf{coalesce}(\mathcal{S})
-\; \cup \;
-\{ S \in \mathcal{S} \mid \neg \exists T \in \mathcal{S} . \; \contig{S}{T}\}
+  \mathsf{coalesce}(\mathcal{S}) = \; & \;\;
+  \{\,S \cup T \mid S \in \mathcal{S}, T \in \mathcal{S} \, \wedge \,
+                                        \contig{S}{T} \}
+  \;\;\; \cup \;\;\;
+  \{\,S \in \mathcal{S} \mid \neg \exists T \in \mathcal{S} . \; \contig{S}{T}\}
 \end{align*}
-\dnote{May need to prove that this is total}
-%\begin{align*}
-%\mathsf{coalesce}(\mathcal{S}) = \; &
-%\{\,S \cup T \mid
-%\forall S \in \mathcal{S}, \;
-%\forall T \in \mathcal{S}. \;
-%\contig{S}{T} \} \\
-%\cup \; &
-%\{ \,\;\;\; S \;\;\; \mid \forall S \in \mathcal{S},
-%\; \not \exists T \in \mathcal{S} . \; \contig{S}{T}\}
-%\end{align*}
 %
-That is, the new set of index schemes $\mathsf{coalesce}(\mathcal{S})$
+Thus, $\mathsf{coalesce}(\mathcal{S})$
 comprises the union of the set of coalesced pairs of contiguous
@@
 contiguous scheme.
-
-\item[Remove redundant schemes]
-Any indexing schemes which are subsumed by another are filtered out:
 %
-\begin{align*}
-\textsf{nonOverlap}(\mathcal{S}) =
-\{S \mid \forall S \in \mathcal{S}, \; \neg \exists T \in (\mathcal{S} -
-  \{S\}), \; \forall i . \; \pi_i(S) \subseteq \pi_i(T) \}
-\end{align*}
+\begin{lemma}[Closure]
+  For a set of index schemes $\mathcal{S}$, then
+  $\mathsf{coalesce}(\mathcal{S})$ is well-defined and is a set of
+  index schemes.
+\end{lemma}
+\begin{proof}
+  By \Cref{lem:vector-union}, the union of two index schemes is an
+  index scheme if all components are equal except at most one
+  component. Contiguity (\Cref{def:contiguity}) matches
+  this precondition, thus the first part of the union in
+  \textsf{coalesce} produces a set of index schemes. The second part
+  comprises a subset of the original index schemes.
+\end{proof}
 %
-\dnote{I'm starting to suspect that this step is actually redundant
-  now- but I will need to do some math to check}
-%\item \textbf{Covering}
-\end{description}
-
-The least fixed-point of $\textsf{nonOverlap} \circ \textsf{coalesce}$
-is computed, providing a minimal indexing scheme covering the space.
-\dnote{Proof?}
-
-For our example:
+For our example, the fixed-pointed of \textsf{coalesce} is reached within two steps:
 %
@@
   \textsf{coalesce}(\mathcal{S}_0) & =
-\{[-1,0] \times [0,0],\,[0,0] \times [-1, 0],\,[0,0] \times [0,1],\,[0,1]
-  \times [0,0]\} \\
-%\textsf{nonOverlap}(\textsf{coalesce}(\mathcal{S}_0)) & =
-%                                                        \textsf{coalesce}(\mathcal{S}_0)
-%  \\
+  \{[-1,0] \times [0,0],\,[0,0] \times [-1, 0],\,[0,0] \times
+    [0,1],\,[0,1] \times [0,0]\} \\
   \textsf{coalesce}^2(\mathcal{S}_0) & =
-\{[-1, 1] \times [0, 0],\,[0, 0] \times [-1, 1]\}
+  \{[-1, 1] \times [0, 0],\,[0, 0] \times [-1, 1]\} = \textsf{coalesce}^3(\mathcal{S}_0)
 \end{align*}
 %
-The fixed-pointed is reached within two applications of
-\textsf{coalesce} for our example.
-\dnote{We probably need to say something about classifying index
-schemes are \region{N}?}
+\todo{Prove that this is in $Region_N$}
 

### 2017-04-17 oopsla17/algorithms.tex
Rewriting consistency bits and adding complexity explanation

--- 2017-04-17 oopsla17/algorithms.tex
@@
   \mathit{consistent} & :
-    \mathsf{Mult}(\mathcal{P}(\mathcal{P}(\bz{})^N)) \times
+    \mathsf{Mult}(\mathcal{P}(\bz{}^N)) \times
     \mathsf{Mult}(\mathsf{Approx}(\region{N})) \to \mathbb{B} & \\
@@
   \mathit{consistent'} & :
-    \mathcal{P}(\mathcal{P}(\bz{})^N) \times
+    \mathcal{P}(\bz{}^N) \times
     \mathsf{Approx}(\region{N}) \to \mathbb{B} & \\
@@
 specification matches that of the indices, \ie{} if the specification allows
-indices to be repeated or not.Then it delegates to $\mathit{consistent'}$
-function which checks if the points observed in the array expression matches
-the space defined by the specification. Lower bounds, marked by \texttt{atLeast}
+indices to be repeated or not. Then it delegates to $\mathit{consistent'}$
+function to check if the points observed in the array terms match the space
+defined by the specification. Lower bounds, marked by \texttt{atLeast}
 modifier, require the space defined by the specification to remain inside the
@@
 
-Reader will realise that since index schemes allows infinite intervals in the
-product, set equality cannot be done by exhaustively checking the points of one
-are inside the other. Instead, we compile the region into interval constraints
-and subscript schemes into membership constraints and pass it to
-\textsc{Z3}~\citep{de2008z3} SMT solver to see if they are equal. The query
-belongs to to quantifier free linear arithmetic, which is a decidable theory.
-Performance of consistency checking using SMT solver is irrelevant as the number
-of interval constraints and the number of indices that can reasonably appear in
-an assignment is too small restricting the size of the formula being checked.
+As explained in \Cref{subsec:union-normal-form}, the sets being compared are
+potentially infinite and set comparisons cannot be done by exhaustively
+comparing the points of one with the other. Instead, we compile the region into
+interval constraints and subscript schemes into membership constraints and pass
+it to \textsc{Z3}~\citep{de2008z3} SMT solver to see if they are equal. The
+query belongs to to quantifier free linear arithmetic, which is a decidable
+theory.
 
-\mnote{It would be great if we had data on number of indexing expressions that
-occur (counting the ones that flow in) and also the length of specifications
-(which I think we do generate). This would justify my efficiency claims}.
+Although satisfiability is super-exponential with the length of the formula in
+the worst case~\cite{fischer1974super}. Performance of consistency checking
+using satisfiability is irrelevant as the the length of the formula is linearly
+related with two factors: (1) dimensionality times number of regions conntected
+with \texttt{+} and (2) dimensionality times the number of array terms flowing
+into an assignment assignment. In \Cref{sec:study}, we establish dimensonality
+99\% of all array variables that we can assign a specification is X, and 99\% of
+number of array terms that flows into such an assignment is Y. In
+\Cref{sec:evaluation}, we show the number of regions that are connected with
+\texttt{+} is Z in 99\% of the cases.
+
+\mnote{DOMINIC fills in X, Y, and Z}
 

### 2017-04-17 oopsla17/algorithms.tex

--- 2017-04-17 oopsla17/algorithms.tex
@@
   \mathit{consistent} & :
-    \mathsf{Mult}(\mathcal{P}(\bz{}^N) \times
+    \mathsf{Mult}(\mathcal{P}(\bz{}^N)) \times
     \mathsf{Mult}(\mathsf{Approx}(\region{N})) \to \mathbb{B} & \\
@@
 %
-  \textit{consistent'} & :
+  \mathit{consistent'} & :
     \mathcal{P}(\bz{}^N) \times
@@
 %
-  \textit{consistent'}(\mathit{ix}, \mathit{model}) & = \begin{cases}
+  \mathit{consistent'}(\mathit{ix}, \mathit{model}) & = \begin{cases}
     m = \mathit{ix} & \mathit{model} = \mathsf{exact}(m) \\

### 2017-04-17 oopsla17/algorithms.tex
typeset functions correctly

@@
   graph (as shorthand, the function
-  $\mathsf{flowTo}$ is used, implicitly parameterised by this graph,
+  $\mathit{flowTo}$ is used, implicitly parameterised by this graph,
   mapping an expression to the set of all expressions
@@
 
-\newcommand{\contig}[2]{\mathsf{contig}(#1, #2)}
+\newcommand{\contig}[2]{\mathit{contig}(#1, #2)}
 \begin{defn}[Contiguity]
@@
 \begin{align*}
-  \mathsf{coalesce}(\mathcal{S}) = \; & \;\;
+  \mathit{coalesce}(\mathcal{S}) = \; & \;\;
   \{\,S \cup T \mid S \in \mathcal{S}, T \in \mathcal{S} \, \wedge \,
@@
 %
-Thus, $\mathsf{coalesce}(\mathcal{S})$
+Thus, $\mathit{coalesce}(\mathcal{S})$
 comprises the union of the set of coalesced pairs of contiguous
@@
   For a set of index schemes $\mathcal{S}$, then
-  $\mathsf{coalesce}(\mathcal{S})$ is well-defined and is a set of
+  $\mathit{coalesce}(\mathcal{S})$ is well-defined and is a set of
   index schemes.
@@
   this precondition, thus the first part of the union in
-  \textsf{coalesce} produces a set of index schemes. The second part
+  $\mathit{coalesce}$ produces a set of index schemes. The second part
   comprises a subset of the original index schemes.
@@
 %
-For our example, the fixed-pointed of \textsf{coalesce} is reached within two steps:
+For our example, the fixed-pointed of $\mathit{coalesce}$ is reached within two
+steps:
 %
 \begin{align*}
-  \textsf{coalesce}(\mathcal{S}_0) & =
+  \mathit{coalesce}(\mathcal{S}_0) & =
   \{[-1,0] \times [0,0],\,[0,0] \times [-1, 0],\,[0,0] \times
     [0,1],\,[0,1] \times [0,0]\} \\
-  \textsf{coalesce}^2(\mathcal{S}_0) & =
-  \{[-1, 1] \times [0, 0],\,[0, 0] \times [-1, 1]\} = \textsf{coalesce}^3(\mathcal{S}_0)
+  \mathit{coalesce}^2(\mathcal{S}_0) & =
+  \{[-1, 1] \times [0, 0],\,[0, 0] \times [-1, 1]\} = \mathit{coalesce}^3(\mathcal{S}_0)
 \end{align*}

### 2017-04-17 oopsla17/algorithms.tex
edits to algorithms

--- 2017-04-17 oopsla17/algorithms.tex
@@
 \label{subsec:checking}
-
-In the model, we can regard both array indices and specifications as groups of
-points in $N$-dimensional space. Notion of consistency is then intuitively
-whether these two (potentially infinite) sets of points in space match or not.
+%
+In the model, we can regard both array indices and specifications as sets of
+points in $N$-dimensional space. The notion of consistency is then intuitively
+whether these two (potentially infinite) sets of points match or not.
 This leads to a simple set based consistency function defined as
@@
   \mathit{consistent}(\mathit{ix}, \mathit{model}) & = \begin{cases}
-    \mathsf{false} & \mathit{model} = \mathsf{once}(x) \wedge ix = \mathsf{mult}(y) \\
+    \mathsf{false} & \mathit{model} = \mathsf{once}(x) \wedge ix =
+    \mathsf{mult}(y) \\
+    \mathsf{false} & \mathit{model} = \mathsf{mult}(x) \wedge ix = \mathsf{once}(y) \\
     \mathit{consistent'}(\mathit{peel}(ix), \mathit{peel}(model)) & \textit{otherwise}
@@
 function to check if the points observed in the array terms match the space
-defined by the specification. Lower bounds, marked by \texttt{atLeast}
+defined by the specification. Lower bounds, marked by the \texttt{atLeast}
 modifier, require the space defined by the specification to remain inside the
-set of indices, while an upper bound, market by \texttt{atMost} modifier,
+set of indices, while an upper bound, market by \texttt{atMost},
 requires the opposite: enclosure. In the absence of such modifiers, we expect
-the space defined by observed indices correspond exactly those defined by the
+the space defined by observed indices corresponds exactly with those defined by the
 region in the specification, hence requiring set equality.
@@
 interval constraints and subscript schemes into membership constraints and pass
-it to \textsc{Z3}~\citep{de2008z3} SMT solver to see if they are equal. The
-query belongs to to quantifier free linear arithmetic, which is a decidable
-theory.
+these to the \textsc{Z3} SMT solver~\citep{de2008z3} to see if they are equal. The
+query is expressed in quantifier free linear arithmetic, which is decidable.
 
 Although satisfiability is super-exponential with the length of the formula in
-the worst case~\cite{fischer1974super}. Performance of consistency checking
-using satisfiability is irrelevant as the the length of the formula is linearly
-related with two factors: (1) dimensionality times number of regions conntected
+the worst case~\cite{fischer1974super}, performance of consistency checking
+using satisfiability is fast in practice as the length of the formula is linearly
+related with two factors: (1) dimensionality times number of regions connected
 with \texttt{+} and (2) dimensionality times the number of array terms flowing
@@
 
+\begin{align*}
+  \mathsf{coalesce1}(\mathcal{S}) & =
+                                    \{\,S \cup T \mid S \in \mathcal{S}, 
+                                    T \in \mathcal{S} \, \wedge \,
+                                    \contig{S}{T} \} \\
+  \mathsf{coalesce}(\mathcal{S}) & =
+                                   \mathsf{coalesce}(\mathsf{coalesce1}(\mathcal{S}) \cup \mathcal{S})
+                                   \; \cup \;
+                                   \{ S \in \mathcal{S} \mid \neg \exists T \in \mathcal{S} . \; \contig{S}{T}\}
+\end{align*}
+
 \subsubsection{Index schemes to syntax}

### 2017-04-17 oopsla17/introduction.tex
cleanup

@@
-\noindent
+s\noindent
 In the sciences, complex models are now almost always expressed as
@@
 
-We fix some terminology used throughout to describe
-syntactic constructs.
+We fix some terminology for syntactic constructs in our target language.
 

### 2017-04-17 oopsla17/study.tex
fix enumerate
tidying on the study section

@@
 
+\noindent
 At the beginning of this project, we made the following hypotheses
 about array programming idioms:
-\begin{enumerate}
-\item Loops over arrays mainly read from arrays with a static pattern
-based on constant offsets from (base or dervied) induction variables;
+\begin{enumerate}[leftmargin=2em]
+\item Computations involving loops over arrays are common, where an
+  arrays are read from with a static pattern based on constant offsets
+  from (base or dervied) induction variables;
 
@@
 \begin{minted}{fortran}
-do i = 1, n
-  b(n) = (a(i) + a(i+1)) / 2.0
+  b(n) = a(i-1) + a(i) + a(i+1)
 \end{minted}
@@
 \item Most loop-array computations of the previous form read
-from arrays with a pattern that includes the immediate
-neighbours (offsets of 1 from the induction variables);
+from arrays with a pattern that includes the origin index
+or its immediate neighbours (offsets of 1 from the induction variables);
 
@@
   neighbourhood indices and whose left-hand side is an
-  neighbourhood-indexed array subscript (\eg{}, the classic five-point
-  stencil);
+  neighbourhood-indexed array subscript (\eg{}, the Laplace operator);
 
@@
 %%
-LHS affine & RHS neigh  & 23        & 0.02\% & 0 & 0.00\% \\\hline \hline
+LHS affine & RHS neigh  & 23        & 0.02\% & 0 & 0.00\% \\\hline
 %
@@
 \begin{center}
-\begin{minipage}{1\linewidth}
+\begin{minipage}{0.48\linewidth}
 \begin{tabular}{l|S[table-format=4]S[table-format=2.2]}
@@
 \end{tabular}
-\hspace{1em}
+\vspace{2.4em}
+\end{minipage}
+\begin{minipage}{0.5\linewidth}
 \begin{tabular}{l|S[table-format=4]S[table-format=2.2]S[table-format=4]S[table-format=2.2]}

### 2017-04-17 oopsla17/algorithms.tex
Second part of inference

--- 2017-04-17 oopsla17/algorithms.tex
@@
 \subsubsection{Index schemes to syntax}
+
+Now we convert coalesced indexing schemes into specifications. This happens in
+three stages. First, we see if there is an all infinite indexing scheme. Second,
+we determine if the indexing schemes are interval schemes so that they can be
+represented exactly and if not how it can be altered into interval schemes to be
+represented as approximations. Third, we split up the interval schemes so that
+the region constants can express them. Finally, we convert the interval schemes
+into a joins of meets of region constants.
+
+If there is all infinite indexing scheme, then we cannot represent it
+syntactically. This happens when use of constant indices in one or more array
+terms causes the offsets to be subsumed. The simplest case is when there is an
+array term with all of its indices constant.
+
+In our example, there are no constant indices, so we proceed.
+
+After handling the trivial case, correct approximation is chosen. Recall
+\Cref{def:interval-scheme}, an indexing scheme is an interval scheme if each of
+the sets in products are holed intervals. If that is the case, we trivially
+convert the sets in each dimension of indexing schemes to holed interval
+representation and proceed to the next stage and output an exact specification
+at the end. If not, then the output will definitely have an upper bound and
+potentially a lower bound.
+
+Upper bound is established by \emph{elongating} index schemes in multiple
+dimensions such that they become interval schemes. The elongation function is
+given as
+%
+\begin{align*}
+  \mathit{elongate}([a,b]) = \begin{cases}
+    \interv{0}{b}{\textsf{false}} & a > 1 \\
+    \interv{a}{0}{\textsf{false}} & b < -1
+  \end{cases}
+\end{align*}
+
+Since index schemes of the previous stage are all contiguous and form closed
+intervals, 0 lying between the lower and upper bounds of intervals is sufficient
+to form an interval scheme. If any dimension of an index scheme is elongated,
+then the whole index scheme is elongated. In the initial set of index schemes,
+all schemes that were already interval schemes form a lower bound, while those
+schemes and the new interval schemes generated by elongation form the upper
+bound. The reason a lower bound may not be generated is all indexing schemes
+with might need elongation.
+
+The index schemes in the example are all interval schemes, so we proceed to the
+next stage noting the output will be exact.
+
+There is no direct translation from an interval scheme such as $\interv{-2}{1}{}
+\times \interv{-1}{1}{}$ to region constants. The reason is region constants
+\texttt{forward} and \texttt{backward} require upper and lower bounds to be 0
+respectively and \texttt{centered} requires the upper bound to be the negation
+of the lower bound. The solution is to use \Cref{lem:vector-union} to break up
+interval schemes into two producing $\interv{-2}{0}{} \times \interv{-1}{1}{}$
+and $\interv{0}{1}{} \times \interv{-1}{1}{}$.
+
+The extended example does not have such representation problem, so the interval
+schemes are left as they are.
+
+The final stage of inference is simply mapping interval schemes into region
+constants. Each interval scheme is connected with \texttt{+} and each dimension
+of the interval scheme is connected with \texttt{*}. The intervals at each
+dimension are mapped to region constants using the bijection defined in
+\Cref{sec:semantics}.
+
+In the example, $[-1, 1] \times [0, 0]$ is mapped to \texttt{\stencil{c}{1}{1}{}
+* \stencil{p}{2}{}{}} and $[0, 0] \times [-1, 1]$ is mapped to
+\texttt{\stencil{p}{1}{}{} * \stencil{c}{2}{1}{}}, producing the overall
+specification:
+%
+\begin{equation*}
+  \texttt{stencil readOnce,} \;
+  \stencil[s]{c}{1}{1}{} \texttt{*} \stencil[s]{p}{2}{}{} \; \texttt{+} \;
+  \stencil[s]{p}{1}{}{}  \texttt{*} \stencil[s]{c}{2}{1}{}
+\end{equation*}

### 2017-04-17 oopsla17/study.tex

--- 2017-04-16 oopsla17/study.tex
@@
 
-At the beginning of this project, we made the following hypotheses
-about array programming idioms:
-\begin{enumerate}
+\noindent
+We start with the following hypotheses about array programming idioms
+in scientific code, formed from our observations and conversations
+with scientists:
+%
+\begin{enumerate}[leftmargin=2em]
 \item Loops over arrays mainly read from arrays with a static pattern
@@
 \begin{minted}{fortran}
-do i = 1, n
   b(n) = (a(i) + a(i+1)) / 2.0
@@
 
-\item Many array computations read from each particular index just once.
+\item Many array computations read from each particular index just
+once. The significance of this hypothesis is that, if true, it would
+be useful to provide a specification of this so accidental repetition
+of subscripts in complex code can be detected.
 \end{enumerate}
@@
 %%%%%%
-\multirow{2}{*}{Shape} & (\textsf{rect}) & (Hyper)rectangle &
- \fortran{a(i,j) + a(i+1,j) + a(i,j+1) + a(i+1,j+1)} \\
-           & (\textsf{sumRect}) & Composed (hyper)rectangles &
-  \fortran{a(i,j) + a(i,j-1) + a(i-1,j) + a(i+1,j) + a(i,j+1)} \\ \hline
+%\multirow{2}{*}{Shape} & (\textsf{rect}) & (Hyper)rectangle &
+% \fortran{a(i,j) + a(i+1,j) + a(i,j+1) + a(i+1,j+1)} \\
+%           & (\textsf{sumRect}) & Composed (hyper)rectangles &
+%  \fortran{a(i,j) + a(i,j-1) + a(i-1,j) + a(i+1,j) + a(i,j+1)} \\ \hline
 %%%%%%

### 2017-04-17 oopsla17/study.tex

--- 2017-04-17 oopsla17/study.tex
@@
 \noindent
-At the beginning of this project, we made the following hypotheses
-about array programming idioms:
+We start with the following hypotheses about array programming idioms
+in scientific code, formed from our observations and conversations
+with scientists:
 \begin{enumerate}[leftmargin=2em]
@@
 
-\item Most loop-array computations of the previous form read
-from arrays with a \emph{contiguous} pattern, \eg{}:
+\item Most loop-array computations of the previous form read from
+  arrays with a \emph{contiguous} pattern, \eg{}:
 %
@@
 %
-\item Most loop-array computations of the previous form read
-from arrays with a pattern that includes the origin index
-or its immediate neighbours (offsets of 1 from the induction variables);
+\item Most loop-array computations of the previous form read from
+  arrays with a pattern that includes the origin index or its
+  immediate neighbours (offsets of 1 from the induction variables);
 
@@
 
-\item Many array computations read from each particular index just once.
+\item Many array computations read from each particular array index just
+  once. The significance of this hypothesis is that, if true, it would
+  be useful to provide a specification of index uniqueness, so accidental
+  repetition of subscripts in complex code can be detected.
 \end{enumerate}
 %
-From these hypotheses, we conjecture that the above programming idioms
-can be specified declaratively via a small
-set of combinators, capturing data access patterns based on
-overlapping (hyper)rectangles. We performed a large scale source-code
-analysis to validate these hypotheses and guide the design of our
-language. These results are also of wider value to others, such as
-designers of array/stencil DSLs or autotuners.
+From these hypotheses, we conjecture that the spatial properties of
+the above programming idioms can be specified declaratively via a
+small set of combinators, capturing data access patterns.% based on
+%overlapping (hyper)rectangles.
+We performed a large scale source-code analysis to validate these
+hypotheses and guide the design of our language. These results are
+potentially of interest to others, such as designers of array/stencil DSLs
+or autotuners.
 
@@
 \noindent
-We constructed a corpus of \numPackages{} scientific computing
-packages written in Fortran ranging in size and scope: The Unified
-Model (UM)~\cite{um}, E3MG~\cite{RePEc:aen:journl:2006se-a12},
-BLAS~\cite{blas}, Hybrid4~\cite{GBC:GBC635},
-GEOS-Chem~\cite{geos-chem}, Navier (based on
-\cite{griebel1997numerical}), Computational Physics 2 (CP2)
-~\cite{nicholas2006computational}, ARPACK-NG~\cite{arpackng}, and
-SPECFEM3D~\cite{specfem3d}. \dnote{Include the rest} This covers
-approximately 1.4 million lines of physical Fortran code (2.4 million
-including comments and white space).  Appendix~\ref{app:corpus}
-provides further detail on these packages and their sizes, with UM and
-GEOS-Chem the largest at $\approx$630kloc and $\approx$450kloc respectively. We used
-Wheeler's \emph{SLOCCount} to get a count of the physical source lines
+We constructed a corpus of eleven scientific computing packages
+written in Fortran ranging in size and scope:
+%
+\begin{enumerate}[leftmargin=1.5em]
+\item The Unified Model (UM) developed by the UK Met Office
+for weather modelling~\cite{um};
+\item E3ME, a mixed economic/energy impact predication
+  model~\cite{RePEc:aen:journl:2006se-a12}
+
+\item Hybrid4, a global scale ecosystem model~\cite{GBC:GBC635};
+
+\item GEOS-Chem, troposhperic chemstry model~\cite{geos-chem};
+
+\item Navier, a small-size Navier-Stokes fluid model (based
+  on \cite{griebel1997numerical};
+
+\item Computational Physics (CP), programs from a popular
+ textbook~\cite{giordano1997computational};
+
+\item BLAS, a common linear-algebra
+  library used in modelling~\cite{blas};
+
+\item ARPACK-NG, an open-source library for solving large-scale
+eigenvalue problems~\cite{arpackng};
+
+\item SPECFEM3D, global seismic wave models~\cite{specfem3d};
+
+\item MUDPACK, a general multi-grid solver for elliptical
+partial differentials~\cite{MUD};
+
+\item Cliffs, a tsunami model~\cite{tolkova2014land}
+\end{enumerate}
+%
+These cover approximately 1.4 million lines of physical Fortran code
+(2.4 million including comments and white space).  The UM and
+GEOS-Chem packages are the largest at $\approx$630kloc and
+$\approx$450kloc respectively.  Appendix~\ref{app:corpus} provides
+further detail on these packages and their sizes.  We used Wheeler's
+\emph{SLOCCount} to get a count of the physical source lines
 (excluding comments and blank lines)~\cite{wheeler2001sloccount}.
-
-Half the packages come from active research teams who are our
-project partners (\ie{}, they were not selected carefully to
-fit our hypotheses). BLAS was selected as it is a common numerical library;
-similarly CP2 was selected as it gives standard numerical
-analysis algorithms from a popular computational physics textbook.
-The smallest project Navier ($\approx$500loc) is based on a standard textbook on
-computational fluid dynamics.
+A third of the packages come from active research teams who are our project
+partners (\ie{}, they were not selected carefully to fit our
+hypotheses).
 
@@
 
-\paragraph{Classifications}
-
-We focus on assignment statements in loops where the right-hand side
-expression reads from one or more arrays.  Left-hand sides and
-right-hand sides are classified in the following way:
-
-\begin{tabular}{l|l|l|l}
-  Shorthand & Classification (of subscripts) & Applies to & Example \\ \hline
- (\textsf{vars}) & Variables & LHS & \fortran{x = ...}
+%\paragraph{Classifications}
+Left-hand sides and right-hand sides of assignments are classified
+based on their array subscripting expressions:
+%
+\begin{center}
+\begin{tabular}{l|l|l}
+ % Shorthand &
+ Classification (of subscripts) & Applies to & Example \\ \hline
+ %(\textsf{vars}) &
+ Just a variable & LHS & \fortran{x = ...}
   \\
- (\textsf{IVs}) & Induction variables & LHS & \fortran{a(i, j)} \\
- (\textsf{neigh}) & Neighbourhood offsets (of the form
+ %(\textsf{IVs}) &
+ Induction variables & LHS & \fortran{a(i, j)} \\
+ %(\textsf{neigh}) &
+ Neighbourhood offsets (of the form
                                                   $i \pm c$) & LHS/RHS
                                                                & \fortran{a(i, j-1)} \\
- (\textsf{neigh+c}) & Neighbourhood offsets and constants & LHS/RHS &
+ %(\textsf{neigh+c}) &
+ Neighbourhood offsets and absolutes & LHS/RHS &
                                                             \fortran{b(i, 0, j+1)} \\
- (\textsf{aff}) & Affine offsets (of the form $a * i \pm b$) & LHS/RHS &
+ %(\textsf{aff}) &
+ Affine offsets (of the form $a * i \pm b$) & LHS/RHS &
                                      \fortran{a(2*i+1,j)} \\
- (\textsf{aff+c}) & Affine offsets and constants & LHS/RHS & \fortran{a(i+1, 0,
+ %(\textsf{aff+c}) &
+ Affine offsets and absolutes & LHS/RHS & \fortran{a(i+1, 0,
                                                    3*j+2)} \\
- (\textsf{other}) & Subscript expression not included above & LHS/RHS &
-\fortran{x(f(i))}, \fortran{a(i*i)}
-\end{tabular} \\[1em]
-
-\noindent
-Note \textsf{vars} and \textsf{IVs} classifications apply only to
-left-hand sides.
-The classification is computed from all information that flows into an
-assignment, for example, the following decomposes a one-dimensional
+ %(\textsf{other}) &
+ Subscript expression not included above & LHS/RHS &
+\fortran{x(f(i))}, \fortran{a(i*i)}, \fortran{a(0,1)}
+\end{tabular}
+\end{center}
+%
+%Note \textsf{vars} and \textsf{IVs} classifications apply only to
+%left-hand sides.
+Classifications are computed from all information that flows into an
+assignment. For example, the following decomposes a one-dimensional
 three-point pattern across multiple intermediate assignments:
@@
 \fortran{a(i-1)}. We thus traverse the body of a loop in reverse,
-using reaching-definitions to calculate of the array reads that flow
+using reaching-definitions to calculate the array reads that flow
 to a particular assignment. Since \fortran{x} and \fortran{y} flow to
-the line 4, line 2 and line 3 are not classified, though multiple
+the line 4, line 2 and line 3 are not classified separately, though multiple
 assignments can flow to multiple classified statements.
@@
 We further sub-classify sets of array subscripts
-on the right-hand side based on their spatial properties: \\
+on the right-hand side based on their spatial properties:
 
+\begin{center}
 \begin{tabular}{l||l|p{0.28\linewidth}|l}
@@
 %%%%%%
-\multirow{2}{*}{Shape} & (\textsf{rect}) & (Hyper)rectangle &
- \fortran{a(i,j) + a(i+1,j) + a(i,j+1) + a(i+1,j+1)} \\
-           & (\textsf{sumRect}) & Composed (hyper)rectangles &
-  \fortran{a(i,j) + a(i,j-1) + a(i-1,j) + a(i+1,j) + a(i,j+1)} \\ \hline
+%\multirow{2}{*}{Shape} & (\textsf{rect}) & (Hyper)rectangle &
+% \fortran{a(i,j) + a(i+1,j) + a(i,j+1) + a(i+1,j+1)} \\
+%           & (\textsf{sumRect}) & Composed (hyper)rectangles &
+%  \fortran{a(i,j) + a(i,j-1) + a(i-1,j) + a(i+1,j) + a(i,j+1)} \\ \hline
 %%%%%%
   \multirow{2}{*}{Reuse} & (\textsf{readonce}) & Unique subscripts
-                                         & \eg{} \fortran{b(i) = a(i) + a(i+1)} \\
+                                         & \fortran{b(i) = a(i) + a(i+1)} \\
   & (\textsf{mult}) & Repeated subscripts &
-\eg{}  \fortran{b(i) = a(i)  + a(i)}
-\end{tabular} \\[1em]
-
-\noindent
-To assess the hypothesis that
-arrays are mostly read in a pattern that includes the immediate neighbours
-to the ``origin'', we categorised the position of subscript patterns for each
-right-hand side, in each dimension: \\
-
-\begin{tabular}{l||l|p{0.33\linewidth}|l}
-  Property   & Shorthand & Classification (indices per
-                           dimension) & Example \\ \hline
-%%%%%%
+ \fortran{b(i) = a(i)  + a(i)} \\ \hline
   \multirow{2}{*}{Positioning} & (\textsf{origin}) & Includes origin
@@
              & (\textsf{straddle}) & Within distance 1 of origin &
-\fortran{a(i+1), a(i-1)} \\
+\fortran{a(i+1,j) + a(i-1,j)} \\
              & (\textsf{away}) & Away from the origin
                                           & \fortran{a(i+2), a(i+3)}
-\end{tabular} \\[1em]
-
-\noindent
+\end{tabular}
+\end{center}
+%
+%To assess the hypothesis that
+%arrays are mostly read in a pattern that includes the immediate neighbours
+%to the ``origin'', we categorised the position of subscript patterns for each
+%right-hand side, in each dimension: \\
+%
+%\noindent
+%
 Finally, we categorise the relationship between the left-hand side
@@
 side. Otherwise, the two sides are seen as inconsistent:
-
+%
+\begin{center}
 \begin{tabular}{l|l|l}
+  Shorthand & Classification & Example \\ \hline
   (\textsf{const}) & Consistent & \fortran{a(i, j) = b(i, j) +
@@
   (\textsf{inconst}) & Inconsistent & \fortran{a(i) = b(j)}
-\end{tabular} \\[1em]
+\end{tabular}
+\end{center}
 %
@@
 \noindent
-We revist each hypothesis and show the related data and conclusions.
+We revisit each hypothesis and show the related data and conclusions.
 %
 \begin{enumerate}
-\item Loops over arrays mainly read from arrays with a static pattern
-based on constant offsets from (base or dervied) induction variables;
+\item Computations involving loops over arrays are common, where an
+  arrays are read from with a static pattern based on constant offsets
+  from (base or dervied) induction variables;
 \end{enumerate}
@@
 Neighbourhood                   & 65334     & 60.06 \%  \\ \cline{1-3}
-Neighbourhood + constants       & 23209     & 21.34\%  \\ \hline
+Neighbourhood + absolute       & 23209     & 21.34\%  \\ \hline
 Other                           & 20196     & 18.57 \%  \\ \hline \hline
@@
 group them (\textit{grouping 1}) to see that $81.43\%$ of the array
-computations (88,577) have a static pattern that are either constants or
-constant offsets of base/derived induction variables, confirming
+computations (88,577) have a static pattern that comprises
+constant offsets of base/derived induction variables (potential with
+some absolute indices), confirming
 \emph{hypothesis 1}. Note that affine (derived induction variables)
@@
 offsets (possibly mixed with constants in some
-dimensions). The ``affine + constants'' class is
+dimensions). The ``affine + absolute'' class is
 not represented at all.

### 2017-04-17 oopsla17/model.tex
Fix the subset rel of Region and use macro for region

@@
 
-\begin{defn}[Interval scheme]
+\begin{defn}[Interval scheme]\label{def:interval-scheme}
   An interval scheme is a finite Cartesian product of intervals on \bz{},
@@
 %
-\dnote{May want a predicate as well}
-
 \begin{defn}[Region]
   A region is an index scheme and \region{N} is the set of all regions (\ie{},
-  $\region{N} \subseteq \mathcal{P}(\bz)^N)$. The set of all regions is  defined
+  $\region{N} \subseteq \bz^N)$. The set of all regions is  defined
   as the smallest set satisfying the following:

### 2017-04-17 oopsla17/study.tex
edits to the study, intro and appendix

@@
   assignment whose right-hand side comprises array subscripts with
-  neighbourhood indices and whose left-hand side is an
+  neighbourhood indices and whose left-hand side is a
   neighbourhood-indexed array subscript (\eg{}, the Laplace operator);
@@
 %
-
-%\begin{itemize}
-% (Vars, ...)
-%\item An assignment to a variable, with an RHS comprising some array
-%  computation, tends to correspond to a reduction (e.g., calculating
-%the max value in an array).
-%\end{itemize}
+If an assignment is not inconsistent, but one of the other four
+categories, we refer to it as *consistent.
 
@@
 Other                           & 20196     & 18.57 \%  \\ \hline \hline
-\textbf{Total}                  & 108773    &  \\
+\textbf{Total}                  & 108773    & 100\% \\
 \end{tabular}
@@
 \noindent
-Since, affine and neighbour patterns are static, we
-group them (\textit{grouping 1}) to see that $81.43\%$ of the array
-computations (88,577) have a static pattern that comprises
-constant offsets of base/derived induction variables (potential with
-some absolute indices), confirming
-\emph{hypothesis 1}. Note that affine (derived induction variables)
-are rare. Overwhelmingly, the main category is neighbourhood
-offsets (possibly mixed with constants in some
-dimensions). The ``affine + absolute'' class is
-not represented at all.
-\mnote{(1) State that constant here includes variables that are constant with
-respect to the current loop scope.}
-\mnote{(2) Mention what sort of things are
-included in other.}
-\mnote{(3) Add a remark about lack of affine indices to be
-perhaps programmers relying on compiler more regarding optimisations than hand
-optimising. We can later use this to justify our inability to handle affine
-indices.}
-%
+Since, affine and neighbour patterns are static, we group them
+(\textit{grouping 1}) to see that $81.43\%$ of the array computations
+(88,577) have a static pattern that comprises constant offsets of
+base/derived induction variables (potential with some absolute
+indices), confirming \emph{hypothesis 1}. Note that affine subscripts
+are rare, perhaps due to programmers relying on compilers for
+optimisation rather than hand optimising (which tends to lead to
+affine indices).  Overwhelmingly, the main category is neighbourhood
+offsets (possibly mixed with constants in some dimensions). The
+``affine + absolute'' class is not represented at all. \marginpar{\mnote{(2)
+  Mention what sort of things are included in other.}}
+
 \begin{enumerate}[resume]
 \item Most loop-array computations of the previous form read
-from arrays with a \emph{contiguous} pattern, \eg{}:
-%
-\begin{minted}{fortran}
-do i = 1, n
-  b(n) = (a(i) + a(i+1)) / 2.0
-\end{minted}
+from arrays with a \emph{contiguous} pattern.
 %
@@
  \hline \hline
-\textbf{Total}  & 88577    &  \\
+\textbf{Total}  & 108773    & 100\% \\
 \end{tabular}
@@
 %%
-Thus, $76.94\%$ of array computations have only contiguous neighbour
-indices (which may include constants in some indices) in a contiguous
-pattern on the right hand side, with $76.80\%$ either at the origin
+Thus, $76.94\%$ of array computations have only neighbour
+indices (which may include some absolute indices) in a contiguous
+pattern on the right hand side, with $76.80\%$ either including the origin
 ($74.75\%$) or offset by a distance of 1 from the origin
@@
  right-hand side comprises array subscripts with neighbourhood indices
-and whose left-hand side is an neighbourhood-indexed array subscript
-(\eg{}, the classic five-point stencil);
+and whose left-hand side is a neighbourhood array subscript.
 \end{enumerate}
@@
 neighbour-indexed or affined-indexed array. We additionally
-included an orthogonal classification of whether
-left-hand and right-hand sides are consistent (have a common subset
-of induction variables) or not.
+included the orthogonal classification of whether
+left- and right-hand sides are inconsistent or
+otherwise (have a common subset of induction variables).
 %
@@
 & \% {(whole)}
-& \textbf{\# consistent}
+& \textbf{\# *consistent}
 & \% {(whole)}
@@
 have a static pattern on the left and right-hand side that is either
-affine or a constant translation. Thus, the data support ours
+affine or a constant translation. Thus, the data supports ours
 hypothesis that stencis are common. Based on this grouping, we
@@
 \begin{center}
-\begin{minipage}{0.48\linewidth}
+\begin{minipage}{0.42\linewidth}
+\hspace{-1.6em}
 \begin{tabular}{l|S[table-format=4]S[table-format=2.2]}
@@
 %%
-Most stencils are contiguous, but there is a non trivial
+Most stencils are contiguous, but there is a non-trivial
 amount of non-contiguity. Most stencils have matching induction
 variables on the left and right-hand sides but again there
-are non-trivial amounts of other kinds of relationship.
+are non-trivial amounts of the other kinds of relationship.
 %%
@@
 \textbf{RHS form and reuse} & \textbf{Number} & \% \\ \hline
-RHS neighbour/affine unique indices      & 78349     & 72.03\%  \\\hline
-RHS neighbour/affine repeated indices   & 10228     & 9.40\%  \\\hline
+RHS neighbour/affine unique subscripts      & 78349     & 72.03\%  \\\hline
+RHS neighbour/affine repeated subscripts   & 10228     & 9.40\%  \\\hline
 other                             & 20196     & 18.57\%  \\\hline\hline

### 2017-04-17 oopsla17/analysis/Results.hs

@@ -202,18 +202,19 @@ instance Camfortable (Form LHS, Form RHS, Consistency) where
   camfortable (l, r, c) = camfortable l && camfortable r && camfortable c
 
 instance Camfortable Consistency where
-  camfortable (Consistent _)  = True
-  camfortable (LHSsuperset _) = True
-  camfortable _               = False
+  camfortable Inconsistent  = False
+  camfortable _             = True
 
 instance Camfortable (Form LHS) where
   camfortable Vars = True
   camfortable (Neighbours _ L) = True
+  camfortable (Affines _ L)    = True
   camfortable IVs = True
   camfortable _   = False
 
 instance Camfortable (Form RHS) where
   camfortable (Neighbours _ p) = camfortable p
+  camfortable (Affines _ p) = camfortable p
   camfortable _                = False
 
 instance Camfortable (Physicality p) where

### 2017-04-17 oopsla17/analysis/Results.hs
count of number of 'camfortable' things in the histogram

@@ -346,6 +346,12 @@ prettyResultsCamfort r =
  ++ mapView' "Dimensionality" (camfortableResult . histDimensionality $ r)
  ++ mapView' "Max depth" (camfortableResult . histMaxDepth $ r)
  ++ mapView' "Number of indexing expressions" (camfortableResult . histNumIndexExprs $ r)
+ ++ "\nDims result total: " ++ show (tally $ histDimensionality r)
+ ++ "\nMax depth result total: " ++ show (tally $ histMaxDepth r)
+ ++ "\nNumber of indexing expr total: " ++ show (tally $ histNumIndexExprs r)
+  where
+    tally h = sum $ ((M.!) (camfortableResult h) "Camfort")
+
 
 rline msg num = "   " ++ msg ++ ":" ++ (replicate (90 - (length msg)) ' ') ++ (show num) ++ "\n"
 rline' msg dat = "   " ++ msg ++ ":" ++ (replicate (90 - (length msg)) ' ') ++ dat ++ "\n"

### 2017-04-17 oopsla17/algorithms.tex
Add the missing step in inference

--- 2017-04-17 oopsla17/algorithms.tex
@@
 The final stage of inference is simply mapping interval schemes into region
-constants. Each interval scheme is connected with \texttt{+} and each dimension
-of the interval scheme is connected with \texttt{*}. The intervals at each
-dimension are mapped to region constants using the bijection defined in
-\Cref{sec:semantics}.
+constants. Each interval scheme is connected with \texttt{+}. Individual schemes
+are decomposed using \Cref{lem:vector-intersect} such that each of the resulting
+index schemes have all but one dimensions set to \interv{-\infty}{\infty}{}.
+Each of these interval schemes are mapped to region constants using
+\Cref{subfig:region-model} and connected with \texttt{*}.
 
-In the example, $[-1, 1] \times [0, 0]$ is mapped to \texttt{\stencil{c}{1}{1}{}
-* \stencil{p}{2}{}{}} and $[0, 0] \times [-1, 1]$ is mapped to
-\texttt{\stencil{p}{1}{}{} * \stencil{c}{2}{1}{}}, producing the overall
-specification:
+In the example, $[-1, 1] \times [0, 0]$ is decomposed into $\interv{-1}{1}{}
+\times \interv{-\infty}{\infty}{}$ and $\interv{-\infty}{\infty}{} \times
+\interv{0}{0}{}$. These are then mapped to \texttt{\stencil{c}{1}{1}{}
+* \stencil{p}{2}{}{}}. Decomposition of $[0, 0] \times [-1, 1]$ is similar. When
+combined we obtain the following specification:
 %

### 2017-04-17 oopsla17/appendix.tex
more work on the model and more results

@@
 %%%
-  E3MG & \SSEThreeMEFiles & \SSEThreeMEFilesParsed & \SSEThreeMELoC &
+  E3ME & \SSEThreeMEFiles & \SSEThreeMEFilesParsed & \SSEThreeMELoC &
 \SSEThreeMELoCParsed & \SSEThreeMELoCRaw & \SSEThreeMELoCParsedRaw \\
@@
 %%%
+  mudpack & \SSmudpackFiles & \SSmudpackFilesParsed & \SSmudpackLoC &
+\SSmudpackLoCParsed & \SSmudpackLoCRaw & \SSmudpackLoCParsedRaw \\
+%%
   Navier & \SSnavierFiles & \SSnavierFilesParsed & \SSnavierLoC &
@@
   of the book ``Computational
-  Physics''~\cite{nicholas2006computational} introducing numerical
+  Physics''~\cite{giordano1997computational} introducing numerical
   techniques and their application to modern physics problems such as
@@
 
-\item \textbf{ARPACK-NG}~\cite{arpackng} \todo{add a sentence of detail}
+\item \textbf{ARPACK-NG} is an open-source library for solving large-scale
+eigenvalue problems~\cite{arpackng}.
 
-\item \textbf{SPECFEM3D}~\cite{specfem3d} \todo{add a little detail}
+\item \textbf{SPECFEM3D} is a global seismic wave models
+  ~\cite{specfem3d}.
+
+\item \textbf{MUDPACK} is a general multi-grid solver for elliptical
+partial differentials~\cite{MUD};
+
+\item \textbf{Cliffs} is a parallelised tsunami
+  model~\cite{tolkova2014land}
 \end{enumerate}

### 2017-04-17 oopsla17/resultsToTex.py
create evaluation section and begin to import results

@@ -1,0 +1,58 @@
+#!/usr/bin/python
+
+import sys
+import re
+from collections import defaultdict
+
+def sanitize(s):
+    s = s.strip()
+    s = re.sub("_","",s)
+    for (i,j)  in enumerate(["zero","one","two","three","four","five","six","seven","eight","nine","ten"]):
+        s = re.sub("%d" % i,j,s)
+    return s
+
+files = {}
+
+corpora = defaultdict(lambda:{})
+keys = {}
+with open(sys.argv[1]) as f:
+    for l in f:
+        if l.strip():
+            if l[0:7] == "overall":
+                corpus = "overall"
+            elif l[0:6] == "corpus":
+                corpus = sanitize(l.split(":")[1])
+            elif l.strip().startswith("numStencilSpecs"):
+                pass
+            else:
+                (key,value) = map(lambda x:x.strip(),l.split(":"))
+                key = sanitize(key);
+                keys[key] = 1
+                corpora[corpus][key] = 1
+                print "\\newcommand{\\%s}{%s}" % (corpus+key,value)
+                files[sanitize(corpus+key)] = value
+
+for corpus in corpora.keys():
+    numFiles = int(files.get(corpus+"parseOk", 0)) + int(files.get(corpus+"lexOrParseFailed", 0))
+    print "\\newcommand{\\%s}{%s}" % (corpus+"Files",numFiles)
+    numLoC = files[corpus+"linesTotal"]
+    print "\\newcommand{\\%s}{%s}" % (corpus+"LoC",numLoC)
+
+
+(totalTickAssign, totalTickAssignSuccess, totalLoC) = (0,0,0)
+for corpus in corpora.keys():
+    if corpus != "overall" and corpus != "ethreemgmodernise":
+        for key in keys.keys():
+            if not corpora[corpus].has_key(key):
+                print "\\newcommand{\\%s}{0}" % (corpus+key)
+        loc = files[corpus+"linesParsed"]
+        tickAssign = files[corpus+"tickAssign"]
+        tickAssignSuccess = files[corpus+"tickAssignSuccess"]
+        totalTickAssign += int(tickAssign)
+        totalTickAssignSuccess += int(tickAssignSuccess)
+        totalLoC += int(loc)
+        print "\\newcommand{\\%s}{%.2g}" % (corpus+"tickAssignPercent", float(tickAssign) / float(loc) * 100)
[…]

### 2017-04-17 oopsla17/resultsToTex.py
put resultsToTex into the build-chain

@@ -22,7 +22,7 @@ with open(sys.argv[1]) as f:
                 corpus = "overall"
             elif l[0:6] == "corpus":
                 corpus = sanitize(l.split(":")[1])
-            elif l.strip().startswith("numStencilSpecs"):
+            elif l.strip().startswith("numStencilSpecs ("):
                 pass
             else:
                 (key,value) = map(lambda x:x.strip(),l.split(":"))
@@ -56,3 +56,4 @@ for corpus in corpora.keys():
 
 print "\\newcommand{\\%s}{%.2g}" % ("overalltickAssignPercent", float(totalTickAssign) / float(totalLoC) * 100)
 print "\\newcommand{\\%s}{%.2g}" % ("overalltickAssignSuccessPercent", float(totalTickAssignSuccess) / float(totalLoC) * 100)
+print "\\newcommand{\\%s}{%.2g}" % ("overalltickAssignSuccessPercentOfTickAssign", float(totalTickAssignSuccess) / float(totalTickAssign) * 100)

### 2017-04-17 oopsla17/evaluation.tex
next eval paragraph

--- 2017-03-11 array17.tex
@@
-\documentclass[10pt,preprint,numbers]{sigplanconf}
-
-\usepackage{amssymb}
-\usepackage{amsmath}
-\usepackage{amsthm}
-\usepackage{stmaryrd}
-\usepackage{color}
-\usepackage{graphics}
-\usepackage{fancyvrb}
-\usepackage{subfigure}
-\usepackage{multirow}
-\usepackage{minted}
-\usepackage{breqn}
-\usepackage{enumitem}
-%\usepackage{subcaption}
-\usepackage{siunitx} % For pretty-printing numeric values and SI units
-                     % of measure. e.g., the tabular column type S is
-                     % used to print nice-looking tables of numbers.
-\sisetup{ % defaults
-  group-separator={,},
-  group-minimum-digits={3},
-  output-decimal-marker={.},
-  table-format = 6
-}
-
-\usepackage{tikz}
-\usepackage{xypic}
-\usepackage{natbib}
-
-\usetikzlibrary{calc}
-
-%% http://tex.stackexchange.com/questions/55068/is-there-a-tikz-equivalent-to-the-pstricks-ncbar-command
-\tikzset{
-    ncbar angle/.initial=90,
-    ncbar/.style={
-        to path=(\tikztostart)
-        -- ($(\tikztostart)!#1!\pgfkeysvalueof{/tikz/ncbar angle}:(\tikztotarget)$)
-        -- ($(\tikztotarget)!($(\tikztostart)!#1!\pgfkeysvalueof{/tikz/ncbar angle}:(\tikztotarget)$)!\pgfkeysvalueof{/tikz/ncbar angle}:(\tikztostart)$)
-        -- (\tikztotarget)
-    },
-    ncbar/.default=0.5cm,
-}
-
-\tikzset{round left paren/.style={ncbar=0.4cm,out=110,in=-110}}
-\tikzset{round right paren/.style={ncbar=0.4cm,out=70,in=-70}}
-
-\fvset{
-  linenos=true,
-  fontsize=\footnotesize,
-  breaklines=true,
- breakafter=w),
-  xleftmargin=\parindent,
-  numbersep=1em
-}
-
-\usemintedstyle{vs}
-
-\definecolor{darkblue}{rgb}{0.0,0.0,0.5}
-\definecolor{darkgreen}{rgb}{0.0,0.4,0.0}
-\definecolor{darkdarkgreen}{rgb}{0.0,0.35,0.0}
-
-\usepackage{hyperref}
-\hypersetup{
-    unicode=false,          % non-Latin characters in Acrobat's bookmarks
-    pdftoolbar=true,        % show Acrobat toolbar?
-    pdfmenubar=true,        % show Acrobat menu?
-    pdffitwindow=false,      % page fit to window when opened
-    pdftitle={},    % title
-    pdfauthor={}
-    pdfsubject={},   % subject of the document
-    pdfnewwindow=true,      % links in new window
-    pdfkeywords={keywords}, % list of keywords
-    colorlinks=true,       % false: boxed links; true: colored links
-    linkcolor=darkblue,          % color of internal links
-    citecolor=darkblue,        % color of links to bibliography
-    filecolor=green,      % color of file links
-    urlcolor=blue,          % color of external links
-}
-
-\usepackage{cleveref}
-\crefname{section}{\S}{\S\S}
-\Crefname{section}{\S}{\S\S}
-\crefname{subsection}{\S}{\S\S\S}
-\Crefname{subsection}{\S}{\S\S\S}
-\crefname{figure}{fig.}{}
-\Crefname{figure}{Fig.}{}
-\crefname{definition}{def.}{}
-\Crefname{definition}{Def.}{}
-
-\CustomVerbatimEnvironment{SpecVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.5cm,
-xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.98,numbersep=0.9em}
-\CustomVerbatimEnvironment{ExmVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.5cm,
-xrightmargin=0.2cm,baselinestretch=0.98,numbers=left,numbersep=0.9em,commandchars=\\\{\}}
-\CustomVerbatimEnvironment{IVerbatim}{Verbatim}{fontsize=\relsize{-1},xleftmargin=0.5cm,
-xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.98,numbersep=0.9em}
-
-\definecolor{darkgreen}{rgb}{0.0,0.5,0.0}
-\definecolor{darkpurple}{rgb}{0.6,0.0,0.6}
-\definecolor{orange}{rgb}{0.8,0.4,0.0}
-\definecolor{darkorange}{rgb}{0.5,0.2,0.0}
-\definecolor{marco}{rgb}{0.0,0.3,0.5}
-\definecolor{gray}{rgb}{0.2,0.2,0.2}
-
-\newcommand\mdef{\stackrel{\mathclap{\normalfont\mbox{def}}}{=}}
-
-\newcommand{\bn}{\mathbb{N}}
-
-\newcommand{\todo}[1]{\textcolor{blue}{Dom: #1}}
-\newcommand{\dnote}[1]{\textcolor{darkpurple}{Dom: #1}}
-\newcommand{\mnote}[1]{\textcolor{darkgreen}{Mistral: #1}}
-\newcommand{\anote}[1]{\textcolor{red}{Andy: #1}}
-
-\newcounter{block}
-
-\newtheorem{lemma}[block]{Lemma}
-\newtheorem{proposition}[block]{Proposition}
-
-\theoremstyle{definition}
-
-\newtheorem{theorem}[block]{Theorem}
-\newtheorem{remark}[block]{Remark}
-\newtheorem{example}[block]{Example}
-\newtheorem{definition}[block]{Definition}
-
-% Writing macros
-\newcommand{\ie}{\emph{i.e.}}
-\newcommand{\eg}{\emph{e.g.}}
-
-\newcommand{\dimId}{\texttt{dim}}
-
-% Semantics related
-\newcommand{\interp}[1]{\llbracket{#1}\rrbracket}
-
-% Syntax macros
-\newcommand{\nonterm}[1]{\textit{#1}}
-\newcommand{\term}[1]{\texttt{#1}}
-
-\newcommand{\stenRefl}[1]{\term{pointed} \, (\term{dim=}#1)}
-\newcommand{\stenFwd}[3]{\term{forward} \, (\term{depth=}#1,
-  \term{dim=}#2{#3})}
-\newcommand{\stenBwd}[3]{\term{backward} \, (\term{depth=}#1,
-  \term{dim=}#2{#3})}
-\newcommand{\stenCen}[3]{\term{centered} \, (\term{depth=}#1,
-  \term{dim=}#2{#3})}
-\newcommand{\irrefl}{\texttt{nonpointed}}
-
-\newcommand{\stenReflS}[1]{\term{pointed} \, (\term{dim=}#1)}
-\newcommand{\stenFwdS}[2]{\term{fwd} \, (\term{depth=}#1,
-  \term{dim=}#2)}
-\newcommand{\stenBwdS}[2]{\term{bwd} \, (\term{depth=}#1,
-  \term{dim=}#2)}
-\newcommand{\stenCenS}[2]{\term{cen} \, (\term{depth=}#1,
-  \term{dim=}#2)}
-\newcommand{\irreflS}{\texttt{np}}
-
-\newcommand{\stenFwdSR}[3]{\term{fwd} (\term{depth=}#1,
-  \term{dim=}#2, #3)}
-\newcommand{\stenBwdSR}[3]{\term{bwd} (\term{depth=}#1,
-  \term{dim=}#2, #3)}
-\newcommand{\stenCenSR}[3]{\term{cen} (\term{depth=}#1,
-  \term{dim=}#2, #3)}
-\newcommand{\stenReflSR}[1]{\term{point} (\term{dim=}#1)}
-
-% SYNTAX OPERATIONS AND PREDICATES
-%\newcommand{\neigh}{\textsf{neigh}}
-\newcommand{\arrayTy}{\textsf{array}}
-\newcommand{\rhsExp}{\textsf{rhsExp}}
-\newcommand{\var}{\textsf{var}}
-
-%% VECTOR NOTATIONS
-\newcommand{\vect}[1]{\textbf{#1}}
-
-\newcommand{\vtwohs}[2]{\setlength{\arraycolsep}{0em}
-\left[\begin{array}{cc}#1 \, & \, #2\end{array}\right]\!}
-
-\newcommand{\vtwoh}[2]{\setlength{\arraycolsep}{0em}
-\left[\begin{array}{cc}#1 \; & \; #2\end{array}\right]}
-
-\newcommand{\vthreeh}[3]{\setlength{\arraycolsep}{0em}
-\left[\begin{array}{ccc}#1 \; & \; #2 \; & \; #3\end{array}\right]}
-\newcommand{\vtwo}[2]{\setlength{\arraycolsep}{0em}
-\left[\begin{array}{l}#1\\#2\end{array}\right]}
-\newcommand{\vthree}[3]{\setlength{\arraycolsep}{0em}
-\left[\begin{array}{l}$#1$\\$#2$\\$#3$\end{array}\right]}
-\newcommand{\stwo}[4]
-%{\vtwo{#1}{#2}\!\vtwo{#3}{#4}}
-{\setlength{\arraycolsep}{0.1em}
-\left[\begin{array}{rr}$#1$ & $#3$\\$#2$ & $#4$\end{array}\right]}
-
-\newcommand{\singleEntry}[2]{\textbf{J}_{#2}^{#1}}
-\newcommand{\zeroEntry}[2]{\textbf{K}_{#2}^{#1}}
-
-%% OPERATIONS ON SPANS and VECTORS
-\newcommand{\containedin}{\sqsubseteq}
-
-%% MODEL
-\newcommand{\effdims}[2]{\mathit{constr}(#1)_{#2}}
-
-\newcommand{\trule}[1]{{\footnotesize{(\text{#1})}}}
-%\include{results}
-
-%\newcommand{\cons}[3]{\mathbf{C}_{#3}(#1, #2)}
-\newcommand{\consName}{\textbf{agree\textprime}}
-\newcommand{\cons}[2]{#1 \,\, \consName{} \,\, #2}
-\newcommand{\consAName}{\textbf{agree}}
-\newcommand{\consA}[2]{#1 \,\, \consAName{} \,\, #2}
-\newcommand{\consSub}[3]{#2 \,\, \consAName{}_{#1} \,\,#3}
-
-\title{Verifying Spatial Properties of Stencil Computations using CamFort}
-%\titlebanner{DRAFT -- Do not distribute}
-\authorinfo{}{}{}
-%\preprintfooter{DO NOT DISTRIBUTE}
-
-\begin{document}
-\maketitle
-
-\bibliographystyle{plainnat}
-
-\begin{abstract}
-  Stencil computations are a common programming pattern in numerical code and
-  typically form the core of graphics and scientific computing applications.
-  However, they are often complex and unwieldy, involving substantial
-  fine-grained manipulation of array indices across multiple arrays and
-  dimensions. This requires careful testing, which is almost always done
-  manually and then discarded. In this paper, we propose an alternate approach:
-  a specification language and automated tool for verifying the spatial
-  behaviour of stencils. The specification language is based on the hypothesis
-  that stencil computations read from arrays with a regular, fixed pattern
-  which can be captured by a simple set of abstract combinators. These
-  specifications can be embedded as annotations in source code, against which
-  code can be checked for conformance. For legacy code, spatial specifications
-  can be inferred, providing documentation and aiding future software
-  maintenance.
-
-  We evaluate our specification language and verification tool
-  against a corpus of numerical Fortran code ($\sim$ 1 million lines) for which
-  we generate $\sim$ 60,000 specifications, showing the vast majority of stencil
-  computations indeed have a simple, regular, static shape.
-\end{abstract}
-
-%\keywords{program verification, specification, stencils}
-
-\section{Introduction}\label{sec:intro}
-
-\emph{Stencils} are a ubiquitous programming pattern, common in scientific and
-numerical computing applications. Informally, a stencil computation computes an
-array whose value at each index $i$ is calculated from a \emph{neighbourhood}
-of values around $i$ in some input array(s), \eg{}, the Game of Life,
-convolutions in image processing, approximations to differential equations. For
-example, the following computes the one-dimensional discrete Laplace transform
-(an approximation to a derivative) in Fortran:
-%
-\begin{minted}{fortran}
-do i = 1, (n-1)
-      b(i) = a(i-1) - 2*a(i) + a(i+1)
-end do
-\end{minted}
-%
-Values \texttt{b(i)} are calculated from a neighbourhood of elements around
-\texttt{i} in the input array \texttt{a}. In this example, the access pattern
-is simple and easily understood. More complex stencil computations are more
-prone to errors from simple lexical mistakes. For example, the following is a
-small snippet from a Navier-Stokes fluid simulator (based
-on~\citet{griebel1997numerical}) in which two arrays are read with different
-data access patterns, across two dimensions.
-%%
-\begin{minted}[firstnumber=20,xleftmargin=2em]{fortran}
-du2dx = ((u(i,j)+u(i+1,j))*(u(i,j)+u(i+1,j))+   &
-  gamma*abs(u(i,j)+u(i+1,j))*(u(i,j)-u(i+1,j))- &
-  (u(i-1,j)+u(i,j))*(u(i-1,j)+u(i,j))-          &
-  gamma*abs(u(i-1,j)+u(i,j))*(u(i-1,j)-u(i,j))) &
-  /(4.0*delx)
-
-duvdy = ((v(i,j)+v(i+1,j))*(u(i,j)+u(i,j+1))+   &
-  gamma*abs(v(i,j)+v(i+1,j))*(u(i,j)-u(i,j+1))- &
-  (v(i,j-1)+v(i+1,j-1))*(u(i,j-1)+u(i,j))-      &
-  gamma*abs(v(i,j-1)+v(i+1,j-1))*(u(i,j-1)-     &
-  u(i,j))) / (4.0*dely)
-
-laplu = (u(i+1,j)-2.0*u(i,j)+u(i-1,j))/delx/delx+ &
-  (u(i,j+1)-2.0*u(i,j)+u(i,j-1))/dely/dely
-
-f(i,j) = u(i,j)+del_t*(laplu/Re-du2dx-duvdy)
-\end{minted}
-%%
-The access pattern is much harder to understand than the one-dimensional
-Laplace. The miasma of indexing expressions of the form \texttt{var(i$\pm$a,
-j$\pm$b)} is not only hard to read, but is prone to simple textual input
-mistakes, \eg{}, swapping \texttt{-} and \texttt{+}, missing an indexing term,
-or transforming the wrong variable \eg{} \texttt{(i+1,j)} instead of
-\texttt{(i,j+1)}.
-
-In practice, the development process for complex stencil computations involves
-ad hoc testing to ensure that no transcription mistakes have been made (\eg{},
-by visual inspections, or comparison against an manufactured or analytical
-solutions~\cite{farrell2010automated}). Typically such testing is discarded
-once the code is seen to be correct.
-
-This is not the only information that is often discarded. The ``shape'' of the
-indexing pattern is usually the result of choices made in the
-numerical-analysis procedure to convert some continuous equations into a
-discrete approximation. Rarely are these decisions captured in the source code,
-yet the shape of access is usually uniform and is has a perspicuous and concise
-description in numerical analysis literature \eg{}, ``\emph{centered in space,
-to a depth of 1}'' referring to indexing terms \texttt{a(i)}, \texttt{a(i-1)}
-and \texttt{a(i+1)}.
-
-To support the development of correct stencil computations, we propose a
-simple, abstract specification language for the data access pattern of
-stencils. The shape of the Laplace example stencil is specified in our language
-as:
-%
-\begin{minted}[linenos=false]{fortran}
-!=  stencil centered(depth=1, dim=1) :: a
-\end{minted}
-%
-That is, \texttt{a} is accessed with a symmetrical pattern (``centered'') to a
-depth of one in each direction in its first dimension. The Navier-Stokes
-example has a shape specified as:
-%
-\begin{minted}[linenos=false,xleftmargin=0em,xrightmargin=0.8em,breakindent=1.9em]{fortran}
-!= stencil centered(depth=1,dim=1)*pointed(dim=2)     + centered(depth=1,dim=2)*pointed(dim=1) :: u
-
-!= stencil forward(depth=1,dim=1)                          * backward(depth=1,dim=2) :: v
-\end{minted}
-%
-The specification requires that, over the whole fragment, \texttt{u} is
-accessed with a centered pattern to depth of 1 in both dimensions (this is
-known as the \emph{five-point stencil}) and \texttt{v} is accessed in a
-neighbourhood bounded forwards to depth of $1$ in the first dimension and
-backward to a depth of $1$ in the second dimension. The specification is
-relatively small and much more abstract compared to the source code.
-
-We provide a verification tool which can check the spatial correctness of
-stencil computations against a specification. Furthermore, the tool can
-automatically infer specifications and insert these as comments into the source
-code. The main features and benefits of our approach are as follows:
-%
-\begin{itemize}
-\item The primary use case is that a programmer writes a specification first
-  before the stencil code. Our tool then checks conformance between the two.
-  This specify-and-check approach reduces testing effort and future-proofs
-  against bugs introduced during refactoring and maintenance.
-%
-\item A specification concisely captures the access pattern of a stencil and
-  turns it into documentation. The inference procedure efficiently produces
-  specifications with no programmer intervention, automatically inserting
-  specifications at appropriate places in the source code.
-%
-\item Our specification format is deliberately abstract, with a small number of
-  combinators that \emph{do not involve any indexing expressions}, \eg{}
-  \texttt{a(i+1,j-1)}. This contrasts with other specification approaches,
-  \eg{}, deductive verification tools such as ACSL, where specifications must
-  also be in terms of array-indexing expressions. Therefore, any low-level
-  mistakes that could be made whilst programming complex indexing code could
-  also be made when writing its specification. Our specifications are more
-  abstract and lightweight with the aim of aiding adoption by scientists and
-  preventing indexing errors.
-\end{itemize}
-
-Our verification tool does not target a class of bugs that can be detected
-automatically (\emph{push-button verification}). Instead, stencil computation
-bugs must be identified relative to a specification of the intended access
-pattern.
-
-In this paper, we solely focus on the definition and the usage of the language
-and the acoompanying verification tool and omit the underlying mathematical
-model and implementation details due to space considerations. We make the
-following contributions:
-%
-\begin{itemize}
-\item We introduce a specification language for stencils that captures common
-  forms of data access patterns (\cref{sec:lang});
-%
-\item We provide an implementation of our approach as an extension to
-  CamFort~\cite{camfort}, an open-source program analysis tool for Fortran;
-%
-\item We report on a quantitative study of stencil computations on a corpus of
-  numerical Fortran program (\cref{sec:evaluation}), totalling one million
-  lines of code. Our tool identifies and infers specifications for roughly
-  60,000 stencil computations in the corpus. Approximately 5,000 of the
-  stencils we found are non-trivial, corresponding to code which is a possible
-  source of errors.
-%
-\item We give a verification case study of a stencil computation commonly used
-  in scientific computing, simulating programming errors (\Cref{sec:jacobi}).
-  Our approach detects all possible simulated programming errors for this
-  problem.
-\end{itemize}
-%
-
-\newcommand{\domainVal}{\mathbb{Z}_\infty}
-\section{Stencil specifications}\label{sec:lang}
-
-Our specification language is based on the hypothesis that most forms of array
-access in numerical code have a fixed, statically-determined access pattern.
-For example, the \emph{five-point stencil} on a two-dimensional array reads
-from array indices $(i, j)$, $(i-1, j)$, $(i+1, j)$, $(i, j-1)$, and $(i, j+1)$
-for all $i, j$ within the inner boundary of the array (to avoid out-of-bounds
-access at the edges). We revisit this hypothesis in \cref{sec:evaluation} with
-the inference of such regular stencil patterns on our corpus of numerical
-programs.
-
-We give various definitions used in the paper then \cref{subsec:syntax}
-outlines the syntax of stencil specifications. \cref{sec:eqs} then provides a
-syntactic equational theory for specifications which is sound with respect to
-the underlying model.
-
-\paragraph{Notation}
-%\label{sec:notation}
-
-\renewcommand*{\arraystretch}{0.8}
-For the target language, $e$ ranges over expressions (may be impure) and $v$
-over its (imperative) variables. %By
-%``variable'', we mean in the imperative sense (named binders to memory cells).
-
-\begin{definition}[Induction variable]
-  An integer variable is an \emph{induction variable} if it is the control
-  variable of a ``\texttt{for}'' loop incremented by $1$ per iteration. A variable
-  is interpreted as an induction variable only within the scope of the loop
-  body. Throughout, $i, j, k$ range over induction variables.
-%TODO: consider mentioning derived, non-based induction variables
-% as we did before.
-\end{definition}
-
-\begin{definition}[Array subscripts and indices]
-  An \emph{array subscript}, denoted $a(\bar{e})$, is an expression
-  which reads from an array $a$ at an \emph{index} specified by a
-  comma-separated sequence of integer expressions denoted $\bar{e}$ or
-  in expanded form as $(e_1, \ldots, e_n)$. An index $e_i$
-  is called \emph{relative} if the expression involves an induction
-  variable.
-  An \emph{absolute index} is a constant integer expression relative to the
-  enclosing loop.
-\end{definition}
-
-\begin{definition}[Neighbourhood index]
-\label{def:neighbour}
-  For an array subscript $a(\bar{e})$ an index $e \in \bar{e}$ is a
-  \emph{neighbourhood index} if $e$ is of the form $e \equiv i$, $e \equiv i +
-  c$, or $e \equiv i - c$, where $c$ is an integer constant. That is, a
-  neighbourhood index is a constant translation of an induction variable. (The
-  relation $\equiv$ identifies terms up-to commutativity of $+$ and the inverse
-  relationship of $+$ and $-$ \eg{}, $(-b) + i \equiv i - b$).
-\end{definition}
-
-\texttt{BEGIN: NEEDS REWRITING}
-\begin{definition}[Stencil computations]
-\label{def:stencil}
-  Let $a$ be an array of dimensionality $n$ and $\overline{b}$ a collection of
-  arrays of arbitrary (possibly differing) dimensionalities.  A \emph{stencil
-  computation} comprises an iteration by a set $I$ of at most $n$ induction
-  variables over a subset of the index space of $a$. Each iteration determines
-  the elements of $a$ by an assignment $a(\bar{e}) = e_r$ where each $e \in
-  \bar{e}$ is either a neighbourhood or constant index.  For all $b \in
-  \bar{b}$ and array subscripts $b(\bar{e}')$ that flow to $e_r$, each $e' \in
-  \bar{e}'$ is either a neighbourhood or absolute index.
-
-  We refer to the assignment $a(\bar{e}) = e_r$ and all the associated
-  statements with dataflow to $e_r$ as the \emph{stencil kernel}.
-\end{definition}
-
-We informally describe what it means for a stencil kernel to be consistent with
-a specification. This will be algorithmically described in \Cref{sec:analysis}.
-Specifications are associated to a stencil kernel and are declared for one or
-more array variables $\bar{v}$ within that kernel.  A stencil specification
-defines a \emph{region} associated to an array variable $a$ against which a
-stencil kernel should be \emph{consistent}. The region defined by a
-specification is an $n$-dimensional rectilinear shape in a discrete space
-crossing the origin (which represents a collection of indices which are just
-induction variables). Consistency is then a two way requirement. All array
-subscripts on $a$ contributing to the kernel should be consistent with the
-region. Conversely, all parts of the region should have an index corresponding
-it.
-
-\texttt{END: NEEDS REWRITING}
-
-\begin{figure}[t]
-\vspace{-0.9em}
-\begin{align*}
-\def\arraystretch{1}
-\setlength{\arraycolsep}{0.2em}
-\newcommand{\dimTy}{\mathbb{N}_{>0}}
-\begin{array}{rl}
-\nonterm{specification} ::= & \nonterm{regionDec} \mid \nonterm{specDec} \\
-\nonterm{specDec} ::= & \term{stencil} \; \nonterm{spec} \;
-                        \texttt{::} \; v \\
-\nonterm{regionDec} ::= &  \texttt{region} \; \texttt{::} \; \nonterm{rvar} \; \texttt{=} \;
-                         \nonterm{region}\\[0.4em]
-%\nonterm{spec} ::= & \nonterm{spatial} \mid \nonterm{temporal}
-%\\[1em]
-\nonterm{spec} ::= & [\nonterm{mult},] \; [\nonterm{approx},] \; \nonterm{region} \\
-\nonterm{mult} ::= & \term{readOnce} \\
-\nonterm{approx} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
-\nonterm{region} ::= & \nonterm{rvar} \; \\
-\multicolumn{2}{l}{\qquad \mid \; \stenRefl{\dimTy}} \\
-\multicolumn{2}{l}{\qquad \mid \;\stenFwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{nonpointed}]}} \\
-\multicolumn{2}{l}{\qquad \mid \; \stenBwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{nonpointed}]}} \\
-\multicolumn{2}{l}{\qquad \mid \; \stenCen{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{nonpointed}]}} \\
-\multicolumn{2}{l}{\qquad \mid \; \nonterm{region} \, \term{+}
-  \, \nonterm{region} \; \mid \; \nonterm{region} \; \term{*} \; \nonterm{region}} \\[0.25em]
-%\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{rvar}}
-%\\[0.5em]
-%\nonterm{temporal} ::= \; & \term{dependency} \; (v \; \{ , v \}) [, \texttt{mutual}]
-%  \\[0.5em]
-\nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+\\[-1em]
-\end{array}
-\end{align*}
-\caption{Specification syntax (EBNF grammar)}
-\label{fig:syntax}
-\vspace{-0.8em}
-\end{figure}
-
-\subsection{Specification syntax}
-\label{subsec:syntax}
-
-\Cref{fig:syntax} gives the syntax of stencil specifications, which is detailed
-below. The entry point is the \nonterm{specification} production which splits
-into either a \emph{region declaration} or a \emph{specification declaration}.
-Regions comprise \emph{region constants} which are combined via region
-operators \term{+} and \term{*}.
-
-%\paragraph{Region constants}
-
-Region constants specify a finite contiguous region in a single dimension
-relative to the origin and are either \term{pointed}, \term{forward},
-\term{backward}, or \term{centered}. The region names are inspired by the
-scientific terminology, \eg{} the standard explicit method for approximating
-PDEs is known as the \emph{Forward Time, Centered Space} (FTCS)
-scheme~\citep{dawson1991finite}.
-
-Each region constant has a dimension identifier $d$ given as a positive natural
-number.  Each constant except \term{pointed} has a depth parameter $n$ given as
-a positive natural number; \term{pointed} regions implicitly have a depth of
-$0$.
-
-A \term{forward} region of depth $n$ specifies a contiguous region in dimension
-$d$ starting at the origin. This corresponds to specifying neighbourhood
-indices in dimension $d$ ranging from $i$ to $i + n$ for some induction
-variable $i$. Similarly, a \term{backward} region of depth $n$ corresponds to
-contiguous indices from $i$ to $i - n$ and \term{centered} of depth $n$ from $i
-- n$ to $i + n$. A \term{pointed} stencil specifies a neighbour index $i$. For
-example, the following shows four specifications with four consistent stencil
-kernels reading from arrays \term{a}, \term{b}, \term{c} and \term{d}:
-%%
-\begin{minted}{fortran}
-!= stencil backward(depth=2,dim=1) :: a
-e(i, 0) = a(i, 0) + a(i-1, 0) + a(i-2, 0)
-!= stencil forward(depth=2,dim=1)*pointed(dim=2) :: b
-e(i, j) = b(i, j) + b(i+1, j) + b(i+2, j)
-!= stencil centered(depth=1,dim=2) :: c
-e(i, j) = (c(j-1) + c(j) + c(j+1))/3.0
-!= stencil pointed(dim=3) :: d
-e(i, j) = d(0, 0, i)
-\end{minted}
-%%
-Every dimension with neighbourhood indices need to be specified. The remaining
-non-conforming indices are not included in the specification, \eg{},
-specifications on lines $1$ and $7$ leave some dimensions unspecified since they
-do not conform to definition neighbourhood indices.
-
-The \term{forward}, \term{backward}, and \term{centered} regions may all have
-an additional attribute \term{nonpointed} which marks absence of the origin.
-For example, the following is a \term{nonpointed} \term{backward} stencil
-%
-\begin{minted}{fortran}
-!= stencil backward(depth=2, dim=1, nonpointed) :: a
-b(i) = a(i-1) + 10*a(i-2)
-\end{minted}
-
-\paragraph{Combining regions}
-
-% r+s means that indices must be consistent with either r or s
-% r*s means that indices must be consistent with both r and s at the
-% same time
-
-% specifications correspond to a set of schemes
-
-% r+s means an index must be generalised by a scheme in either r or s
-%       and every scheme in r must have an index which specialises it
-%       and every scheme in s must have an index which specialises it
-
-% r*s means an index must be generalised by a scheme in r and a scheme in s
-%       and every pairs of schemes in r and s must have an index which
-%           specialises both
-%
-
-The region operators \term{+} and \term{*} respectively combine regions
-permissively and restrictively.
-
-The restrictive combination of two regions, $r \term{*} s$, means that any
-indices in the specified code must be consistent with both $r$ and $s$
-simultaneously. Dually, for every pair of components in the regions $r$ and $s$
-there must be an index consistent with both. For example, the following
-\emph{nine-point stencil} has a specification given by the product of two
-\texttt{centered} regions in each dimension:
-%%
-\begin{minted}[breakindent=2.9em]{fortran}
-x = a(i, j)   + a(i-1, j)   + a(i+1, j)
-y = a(i, j-1) + a(i-1, j-1) + a(i+1, j-1)
-z = a(i, j+1) + a(i-1, j+1) + a(i+1, j+1)
-!= stencil centered(depth=1, dim=1) * centered(depth=1, dim=2) :: a
-b(i, j) = (x + y + z) / 9.0
-\end{minted}
-%%
-This pattern is common in image convolution applications. The specification
-ranges over the values that flow to the array subscript on the left-hand side,
-and so ranges over the intermediate assignments to \term{x}, \term{y}, and
-\term{z}. Each index in the code is consistent with both specifications
-simultaneously, \eg{}, \texttt{a(i-1, j+1)} is within the centered region in
-dimension $1$ and the centered region in dimension $2$.
-
-The product $r \term{*} s$ can also be thought of as a bounding box over the
-two regions $r$ and $s$.
-
-The permissive combination of two regions $r \term{+} s$ means that any indices
-in the specified code must be consistent with either of $r$ or $s$. Dually,
-every part of region $r$ must have a corresponding index and similarly for $s$
-independently. For example, the following gives the specification of a
-five-point stencil which is the sum of two compound \texttt{pointed} and
-\texttt{centered} regions in each dimension:
-%
-\begin{minted}{fortran}
-!= stencil centered(depth=1, dim=1)*pointed(dim=2)    + centered(depth=1, dim=2)*pointed(dim=1) :: a
-b(i,j) = -4*a(i,j) + a(i-1,j) + a(i+1,j) &
-                   + a(i,j-1) + a(i,j+1)
-\end{minted}
-%%
-Here the left-hand side of \texttt{+} says that when the second dimension
-(induction variable $j$) is fixed at the origin, the first dimension (induction
-variable $i$) accesses the immediate vicinity of the origin (to depth of one).
-The right hand side of \texttt{+} is similar but the dimensions are reversed.
-This reflects the symmetry under rotation of the five-point stencil. As
-presented in \cref{sec:evaluation}, this symmetry is common to stencil
-computations used in practice, which makes it easier to write specifications.
-
-\paragraph{Region declarations and variables}
-
-Region specifications can be assigned to region variables via
-region declarations. For example, the shape of a
-``\emph{Roberts cross}'' edge-detection convolution~\cite{davis1975survey}
-can be stated:
-%%
-\begin{minted}{fortran}
-!= region :: r1 = forward(depth=1, dim=1)
-!= region :: r2 = forward(depth=1, dim=2)
-!= region :: robertsCross = r1*r2
-!= stencil robertsCross :: a
-\end{minted}
-This is useful for common stencil patterns, such as the five-point
-pattern, as the region can be defined once and reused.
-%%
-\paragraph{Modifiers}
-%%
-Region specifications can be modified
-by \emph{approximation} and \emph{multiplicity} information
-(in \textit{spec} in \Cref{fig:syntax}).
-
-The \texttt{readOnce} modifier enforces that no index appears more
-than once (that is, its multiplicity is one). For example, all of
-the previous examples could have \texttt{readOnce} added:
-%
-\begin{minted}{fortran}
-!= stencil readOnce, backward(depth=2, dim=1) :: a
-b(i+1) = a(i) + a(i-1) + a(i-2)
-\end{minted}
-%
-This specification would be invalid if any of the array subscripts were
-repeated. This modifier provides a way to rule out any accidental repetition of
-array subscripts.  The notion is similar to that of linear
-types~\cite{wadler1990linear}, where a value must be used exactly once. We opt
-for the more informative and easily understood name \texttt{readOnce}. This
-modifier is optional, so it need not be present even if the stencil is linear.
-
-In some cases, it is useful to give a lower and/or upper bound for a stencil.
-This can be done using either the \term{atMost} or \term{atLeast} modifiers.
-This is particularly useful in situations where there is a non-contiguous
-stencil pattern, which cannot be expressed precisely in our syntax. For example:
-%
-\begin{minted}{fortran}
-!= stencil atLeast, pointed(dim=1)         :: a
-!= stencil atMost, forward(depth=4, dim=1) :: a
-b(i) = a(i) + a(i+4)
-\end{minted}
-
-\section{Equational theory}
-\label{sec:eqs}
-
-Specifications have a theory of equality, $\equiv$, providing axiomatic
-semantics for the language. This enables users to rewrite their specification
-and provides a way to automatically simplify specifications after inference.
-Here we only focus on the former and omit mathematical implications of theory.
-Additionally, specifications have a theory of approximation, $<:$, that can be
-used by the users to weaken specifications for succinctness in the source.
-
-The equality relation $\equiv$ is defined over \emph{region} syntax which gives
-the following axioms:
-%
-\begin{itemize}
-\item $\term{+}$ is idempotent, commutative, and associative;
-\item $\term{*}$ is commutative, associative, and distributes over
-$\term{+}$, \ie{},  $R \; \texttt{*} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
-                       \texttt{*} \; S) \; \texttt{+} \; (R
-                       \; \texttt{*} \; T)$.
-
-\item (\emph{overlapping}) Given two region constants $R_1$ and $R_2$
-on the same dimension, then some combinations of region constants
-can be coalesced into a single region constant $R_3$ if $R_1$ completely
-overlaps $R_2$. There are five possible instantiations of the
-following rule $R_1 + R_2 \equiv R_3$:
-%
-\newcommand{\stenGR}[4]{\small{\textit{#4}(\term{depth=}#1,
-  \term{dim=}#2, #3)}}
-\newcommand{\stenG}[3]{\small{\textit{#3}(\term{depth=}#1,
-  \term{dim=}#2)}}
-%
-\vspace{-0.2em}
-\begin{align*}
-\hspace{-1em} & \stenGR{n}{d}{p \vee q}{$R_3$} \\[-0.4em]
-\hspace{-1em} \equiv \; & \stenGR{n}{d}{p}{$R_1$} \; \texttt{+} \; \stenGR{m}{d}{q}{$R_2$}
-\end{align*}
-where $n \geq m$ and $p$ and $q$ range
-over booleans representing the absence (true) or presence (false)
-of \irrefl{} which are combined via disjunction.
-We denote the possibilities as $\term{fwd+fwd} \equiv \term{fwd}$, $\term{bwd+bwd} \equiv \term{bwd}$,
-$\term{cen+cen} \equiv \term{cen}$, $\term{cen+fwd} \equiv \term{cen}$ and
-$\term{cen+bwd} \equiv \term{cen}$. %Commutative of $\term{+}$
-%provides symmetric variants.
-
-\item (\textit{overlapping pointed}) \term{pointed} regions coalesce
-  with other regions of the same dimension:
-%
-\vspace{-0.4em}
-\begin{align*}
-& \stenG{n}{d}{region} \\[-0.4em]
-\equiv \; & \stenGR{n}{d}{p}{region} \; \texttt{+} \;
-                          \stenReflS{d}
-\end{align*}
-\vspace{-1.5em}
-
-\item \textit{(centered)} a forward and backward region of the same
-  depth and dimension coalesces into a centered region:
-%
-\vspace{-0.4em}
-\begin{align*}
-\hspace{-1em} & {\small{\stenCen{n}{d}{, p \vee q}}} \\[-0.4em]
-\hspace{-1em} \equiv \; & {\small{\stenFwdSR{n}{d}{p} \; \texttt{+} \;
-                          \stenBwdSR{n}{d}{q}}}
-\end{align*}
-\vspace{-1.5em}
-
-\item $(\textsc{point}) \;\; (R \, \term{*} \, S^\irreflS) \, \term{+} \,
-                         (R^\irreflS \, \term{*} \, S) \; \equiv \; R
-                         \, \term{*} \, S$
-meaning that for the sum of products of two regions $R$ and $S$, where in each
-component one of $R$ or $S$ has the non-pointed attribute denoted by
-$R^{\term{np}}$ and $S^{\term{np}}$, the non-pointed attributes cancel to give
-$R \, \term{*} \, S$.
-\end{itemize}
-%
-Stencil specifications in \textit{spec} are considered equal when they have the
-same modifiers and $\equiv$ equates their regions.
-
-\Cref{fig:inequations} defines a notion of sub-specifications via the relation
-$<:^r$ on \textit{region} syntax and then $<:$ on \textit{spec} syntax
-inductively.  Both relations are reflexive and transitive (we omit these
-equations). The $<:^r$ relation is congruent with respect to \term{+} and
-\term{*} and $<:$ is congruent with respect to \term{readOnce}, the equations of
-which we omit for brevity.
-
-The \trule{\textsc{eq}} rule connects region equations to approximation;
-\trule{\textsc{over}} explains the notion of spatial over-approximation via
-overlapping regions; \trule{\textsc{rep}} allows the \term{readOnly} modifier
-that to be dropped. The \trule{\textsc{shrink}} and \trule{\textsc{grow}} rules
-turn the spatial approximation on regions into approximation on specifications
-via the \term{atLeast} and \term{atMost} modifiers. Finally,
-\trule{\textsc{Gen$\ast$}} shows that specifications can be generalised by
-removing parts of a product $\ast$ which have disjoint dimensions to the rest of
-the region product (where $\mathsf{dims}$ denotes the set of dimensions
-specified in a region).
-
-\begin{figure}[t]
-\vspace{-0.5em}
-\begin{align*}
-\hspace{-0.5em}
-{\small{
-\begin{array}{c}
-%%
-%% INEQUATIONS ON REGION
-%%
-%\framebox{$\textit{region} <: \textit{region}'$} \\[0.85em]
-%
-%\dfrac{}{R <:^r R}\trule{\textsc{refl}} \qquad \dfrac{P <:^r R \quad R <:^r Q}{P <:^r
-%  Q}\trule{\textsc{trans}}
-%\qquad
-%\\[1.5em]
-\vspace{-0.4em}
-\setlength{\arraycolsep}{0.1em}
-\dfrac{\hspace{3em} m \leq n \qquad\qquad r \implies s \hspace{3em}}
-{\begin{array}{rl}
-\stenFwdSR{m}{d}{r} & <:^r \stenFwdSR{n}{d}{s}
-%& \;\, ds \subseteq es \wedge n \leq m
-\\
-\wedge \; \stenBwdSR{m}{d}{r} & <:^r \stenBwdSR{n}{d}{s}
-%& \;\, ds \subseteq es \wedge n \leq m
-\\
-\wedge \; \stenCenSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s}
-%& \;\, ds \subseteq es \wedge n \leq m
-\\
-\wedge \; \stenBwdSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s} \\
-\wedge \; \stenFwdSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s} \\[1.2em]
-\end{array}}\trule{\textsc{over}} \\[-0.2em]
-%\hspace{-0.5em}
-%\dfrac{R_1 <:^r S_1 \quad R_2 <:^r S_2}
-%      {R_1 \, \texttt{+} \, R_2 <:^r S_1 \, \texttt{+} \, S_2}
-%\trule{\textsc{cong}\texttt{+}}
-%\dfrac{R_1 <:^r S_1 \quad R_2 <:^r S_2}
-%      {R_1 \, \texttt{*} \, R_2 <:^r S_1 \, \texttt{*} \, S_2}
-%\trule{\textsc{cong}\texttt{*}} \\[1.5em]
-%%
-%% INEQUATIONS ON SPEC
-%%
-%\framebox{$\textit{spec} <: \textit{spec}'$} \\[0.8em]
-%
-%
-\dfrac{R <:^r S}{\term{atLeast}  S <: \term{atLeast}
-  R}\trule{\texttt{L}}
-\;\;
-\dfrac{R <:^r S}{\term{atMost}  R <: \term{atMost}
-  S}\trule{\texttt{M}} \\[1.5em]
-\dfrac{R <:^r S}{R <: \term{atMost} \, S}\trule{\textsc{grow}} \,
-\dfrac{R <:^r S}{S <: \term{atLeast} \, R}\trule{\textsc{shrink}} \,
-\dfrac{R \equiv S}{R <:^r S}\trule{\textsc{eq}}
- \\[1.5em]
-%%
-\dfrac{\mathsf{dims}(R) \cap \mathsf{dims}(S) = \emptyset}{R \ast S <: R}
-\trule{\textsc{Gen$\ast$}}
-\quad
-%%
-%\\[1.5em]
-%\dfrac{R <: S}{\term{readOnce} \, R <: \term{readOnce} \,
-%  S}\trule{\textsc{cong}\texttt{R}}
-%\;\;
-\;
-\dfrac{}{\term{readOnce} \, R <: R}\trule{\textsc{rep}}
-%\dfrac{}{R <: \term{atMost} R}
-%\quad
-%\dfrac{}{R <: \term{atLeast} R}\\[1.5em]
-\end{array}}}
-\end{align*}
-\vspace{-0.5em}
-\caption{Approximation $<:^r$ on \textit{regions} and
-$<:$ on \textit{specs} (omitting reflexivity, transitivity, and congruences). }
-\label{fig:inequations}
-\vspace{-0.5em}
-\end{figure}
-
-\section{Evaluation}
-\label{sec:evaluation}
+\input{results}
 
 To study the effectiveness of our approach, 
-we built a corpus of around 1 million lines of Fortran code from a
+we built a corpus of over 1 million lines of Fortran code from a
 range of scientific computing packages: The Unified Model (UM)~\cite{um},
-E3MG~\cite{RePEc:aen:journl:2006se-a12}, BLAS~\cite{blas},
+E3ME~\cite{RePEc:aen:journl:2006se-a12}, BLAS~\cite{blas},
 Hybrid4~\cite{GBC:GBC635}, GEOS-Chem~\cite{geos-chem}, Navier (based
 on \cite{griebel1997numerical}), Computational Physics
-2~\cite{nicholas2006computational},
-ARPACK-NG~\cite{arpackng}, and
+ed. 1~\cite{giordano1997computational},
+ARPACK-NG~\cite{arpackng}, MUDPACK~\cite{MUD}, Cliffs~\cite{Cliffs}, and
 SPECFEM3D~\cite{specfem3d}.
@@
 We first examined how frequently stencil computations occur. We parsed
-959,427 lines of Fortran code and found that 10\% (97,439) of
-statements have a left-hand side as an array subscript on
-neighbourhood indices. This supports the idea that stencil-like
-computations are common in scientific code.
-We then used the inference procedure of the previous section to
-generate specifications for stencils in the corpus to assess
-the design of the language. %  This assumes that the programs have been implemented
-%correctly---
-%Normally one would use our approach in verification mode to
-%check the correctness of new code.
+\num{\overalllinesParsed} lines of Fortran code and found that \overalltickAssignPercent\%
+(\num{\overalltickAssign}) of statements have a left-hand side as an
+array subscript on neighbourhood indices. This supports the idea that
+stencil-like computations are common in scientific code.  We then used
+the inference procedure of the previous section to generate
+specifications for stencils in the corpus to assess the design of the
+language.
 
 We would not expect to infer a stencil for each array statement we
-found because our analysis restricts
-the array-subscript-statements that we classify as a
-stencil. In fact, we were able to infer a stencil from 30\% of
-these statements. A single statement can involve multiple arrays and
-we ended up with 60,525 specifications. This shows that we can express
-a large number of stencil shapes within our high-level abstraction and
-validates our initial hypothesis that many stencil computations have a
-regular shape.
-
-%    numStencilSpecs: 60525
-% 4338 files, linesTotal: 1357377, linesParsed: 959427
-% Number of stencils (tickAssign) / number of loops
-% Percentage of a code base that is taken up by stencils
-%Out of 1.35 million lines of code that we fed into our program, and
-%the 959,427 lines that we were able to parse, we identified 60,525
-%cases that fit our criteria for stencil specification.
-
-%\paragraph{Hypothesis 2: Most stencil computations
-%have a regular pattern which can be captured with a simple
-%language of contiguous regions}
-
-% Number of stencils we give a spec to (where a spec declaration to n
-% variables is counted n times) as percentage of number of stencils
-
-    % tickAssign: 97439
-    % tickAssignSuccess: 29402 % <- number of times tickAssign is accompanied by a successful stencil inference
-                               % counts only by +1 each time that tickAssign finds a stencil
-
-%In the analysed body of code we found that we were able to deduce
-%some form of stencil specification from approximately 30\% of the
-%potential stencils examined.
-
-%
-%\subsection{Analysing the design of our specifications}
-
-% 1. How may are just reflexive and how many some reflexive
-% justReflexive: 55504
-% someReflexive: 15823
-
-The majority of specifications generated were relatively simple but we
-found significant numbers of more complex shapes. We grouped common
-patterns into categories:
-
-\textbf{All pointed} 55,504 of the stencils we found involved
-only pointed regions. 39,681 of these were pointed in all
-dimensions. Common examples of this were pointwise transformations on
-data (such as scaling).
-
-\textbf{Single-action} specifications comprise one
-forward, backward, or centered region constant combined via \term{+}
-or \term{*} with any number of pointed regions. We identified 4,837 single-action
-specifications, of which 1,532 were single-action with a
-\texttt{nonpointed} modifier.
-
-\textbf{Multi-action} specifications comprise at least two
- forward, backward, or centered regions, combined with
-any number of pointed regions. We identified 265 multi-action
-specifications out of which 207 had regions combined only with
-$\term{*}$ and 58 combined with a mix of $\term{*}$ and \term{+}.
-
-% multiActionRegionOps1: 31
-% multiActionRegionOps2: 115
-% multiActionRegionOps8: 1
-% multiActionRegionOps14: 5
-% multiActionRegionOps17: 4
-
-% %%%%
-% Complexity-wise, out of these 156 multi-action specifications, 31 of
-% them used a single region operator, 115 used two region operators, 1
-% used eight region operators, 5 used fourteen region operators, and 4
-% used seventeen region operators. Some of these very complex stencils
-% were found in the atmospheric models of the UM.
-% %%%%
-
-\textbf{Bounded} specifications occured with 157
-\texttt{atMost} bounds and 36 \texttt{atLeast}, the latter of which were
-always also paired with an upper bound.
-
-\begin{figure}[t]\begin{minted}[fontsize=\scriptsize,breakindent=0em,linenos=false,xleftmargin=0em,breakafter=)]{fortran}
-!=stencil readOnce,(forward(depth=1,dim=3,nonpointed))*(backward(depth=1,dim=1))*(backward(depth=1,dim=2))+(forward(depth=1,dim=3))*(backward(depth=1,dim=1))*(pointed(dim=2))+(forward(depth=1,dim=3))*(backward(depth=1,dim=1,nonpointed))*(backward(depth=1,dim=2,nonpointed))+(forward(depth=1,dim=3))*(backward(depth=1,dim=2))*(pointed(dim=1))+(backward(depth=1,dim=1))*(backward(depth=1,dim=2,nonpointed))*(pointed(dim=3))+(backward(depth=1,dim=1,nonpointed))*(backward(depth=1,dim=2))*(pointed(dim=3))::x
-\end{minted}
-\caption{Complex specification inferred from
-  \textbf{UM}\label{fig:smagorinsky}}
-\vspace{-1em}
-\end{figure}
-
-The single- and multi-action classes represent more complex stencils
-with a real possibility for programmer error. As an extreme example, the Unified
-Model has an implementation of the Smagorinsky subgrid-scale model for
-calculating turbulence on which our inference yields 39 specs from 340
-lines of code. This is a large reduction given
-the complexity of the algorithm.  We show one example
-in~\Cref{fig:smagorinsky} which specifies the access pattern to
-a 3-dimensional array (which we have renamed to
-\texttt{x}) arising from a kernel of 93 lines of code involving 142
-array subscripts. The specification was the most complex
-seen in our corpus, yet it still represents a significant abstraction
-of the spatial behaviour given size and complexity of the kernel it describes.
-
-We measured the frequency at which individual specifications involved
-multiple occurences of \term{+} and \term{*}:
-
-\vspace{0.25em}
-\setlength{\tabcolsep}{0.57em}
-{\small{
-\hspace{-1em}\begin{tabular}{c|cccccccc}
-& 0 & 1 & 2 & 3 & 4 & 5 & 6 & $> 7$ \\ \hline
-\term{+} & 60,265 & 220 & 15 & 12 & 5 & 4 & 4 & 0 \\
-\term{*} & 31,393 & 15,151 & 13,457 & 424 & 41 & 42 & 5 & 12
-\end{tabular}}}
-\vspace{0.2em}
-
-\noindent
-\eg{}, roughly half the specifications did not involve $\term{*}$,
-a quarter use one $\term{*}$ and just under a quarter
-use two $\term{*}$ operators.
-
-\paragraph{Limitations}
-
-There were various reasons why we did not infer
-specifications on every looped array computation: \\
-\noindent
-1) \textbf{Non-subset induction variables} occur when the
-induction variables on the RHS are not a subset of those in the LHS. These
-cases are not stencils by our definition. \\
-\noindent
-2) \textbf{Derived induction variables} where an
-index (\mintinline{fortran}{x}) is derived from an
-induction variable (\mintinline{fortran}{i}) as in
-\mintinline{fortran}{x = len - i};  \\
-\noindent
-3) \textbf{Inconsistent induction dimensions} occur when
-an induction variable is used to specify more than one array dimension
-on the RHS or multiple induction variables are used for the same
-dimension on the RHS. These are common in matrix operations such as
-LU-decomposition with assignments such as
-\mintinline{fortran}{a(l) = a(l) - a(m) * b(l, m)}.
-
-\subsection{Detecting errors in the 2-D Jacobi iteration}
-\label{sec:jacobi}
-One common example of a stencil computation is the two-dimensional
-Jacobi iteration that repeatedly goes through each cell in a matrix
-and computes the average value of the four adjacent cells. The kernel
-is given by:
-\begin{minted}{fortran}
-  a(i,j) = (a(i-1,j)+a(i+1,j)+a(i,j+1)+a(i,j-1))/4
-\end{minted}
-We infer a precise specification of its shape as:
-\begin{minted}[breakafter=+:,breakindent=-0.6em,breaksymbolsep=0.4em,linenos=false,xleftmargin=-0.5em]{fortran}
-  != stencil pointed(dim=1)*centered(depth=1,dim=2,nonpointed)+pointed(dim=2)*centered(depth=1,dim=1,nonpointed)::a
-\end{minted}
-%
-To test our implementation,
-we examined whether programmer errors would be detected by replacing
-the array index offsets with $-1$, $0$, or $1$ and running our
-verification algorithm. Our checking procedure correctly reported a verification
-failure in each of 6,537 permutations corresponding to an error.
-The iteration computes the average of four adjacent
-cells so $24$ ($4$ factorial) of the possible array index perturbations
-are correct, all of which are accepted by our checker.
-
-% /um/trunk/src/atmosphere/diffusion_and_filtering/turb_smagorinsky.F90
-% Tried introducing indexing errors replacing one of the indices of cx_rho_instance with -1,0,+1
-% 42 cases were caught
-% 6 cases were not caught
-
-% (/ 55504.0 60525) = 91.7%
-% (/ 15823.0 60525) = 26.1%
-% 91.7\% of the specifications involved only reflexive regions. 26.1\%
-
-% 2. Break down of "interesting specifications" (contain more than
-% just reflexive)
-%   - Single-action, histogram on depths
-%   - Single-action + irreflexives, histogram on depths
-
-% singleAction: 4837
-% singleActionIrr: 1532
-
-%   - Multi-action combined only by *
-%   - Multi-action combined also by +, histogram of tree depths
-
-    % multiAction: 156
-    % multiActionMulOnly: 146
-
-%   - Bounded specs (how many lower, upper, or both)
-
-    % atLeast: 36
-    % atMost: 157
-    % boundedBoth: 36
-
-% Present not just percentages but concrete numbers
-
-\section{Related work and conclusions}
-\label{sec:discussion}
-
-Various deductive verification tools can express array indexing
-in their specifications, \eg{}, ACSL of \citet{baudin2008acsl}
-for C (\eg{}~\citet[Example 3.4.1]{burghardt2010acsl}). A specification
-can be given for a stencil computation but must use fine-grained
-indexing as in code and therefore is similarly prone to indexing
-errors. Our approach is much more abstract-- it does
-not aim to reify indexing in the specification, but
-provides simple spatial descriptions which capture a large number
-of common patterns.
-
-\citet{kamil2016verified} propose \emph{verified lifting} to extract a
-functionally-complete mathematical representation of low-level, and
-potentially optimised, stencils in Fortran code. This extracted
-predicate representation of a stencil is used to generate code
-for the \textsc{Halide} high-performance
-compiler~\citep{ragan2013halide}. Thus they must capture the
-full meaning of a stencil computation which requires significant
-program analysis. For example, they report that some degenerate
-stencil kernels take up to 17 hours to analyse and others require
-programmer intervention for correct invariants to be inferred.
-
-% ACR-I think we could skip this in the interests of space
-%This led them to develop \textsf{STNG}, a loop invariant and postcondition
-%finder through syntax-guided synthesis. Using this approach they restrict the
-%search space of postconditions to predicates of a certain form, one that is
-%compatible with \textsf{Halide}. They then look at concrete loop iterations and
-%form an hypothesis about the invariant, which is later confirmed or put back
-%into the system using a SMT solver.
-
-Our approach differs significantly. Rather than
-full representation of stencils, we focus on specifying
-just the spatial behaviour in a lightweight way.
-Thus, it suffices for us to perform a comparatively
-simple data-flow analysis which is efficient, scales linearly with
-code size, and does not require any user intervention.
-Whilst we do not perform deep semantic analysis of stencils,
- the analysis part of our
-approach can be made arbitrarily more sophisticated independent of the rest of
-the work.
-% Hence, we do not require SMT solving or search of loop invariants. It
-%suffices to do comparatively simple data-flow analysis.
-%Since CamFort mostly does
-%syntactic analysis, the inference procedure terminates quickly and
-%never requires programmer intervention.
-%The downside of our
-%approach is that if optimisation causes the stencil computation to be heavily
-%obfuscated, \textsf{STNG} would capture the access behaviour better.
-%
-Furthermore, Kamil \emph{et al.} do not provide a user-visible
-syntactic representation of their specifications, and nor do
-they provide verification from specifications \eg{}, to future-proof
-the code against later changes. Even if they were to provide a
-syntactic representation, for complex stencils such as Navier-Stokes
-from \cref{sec:intro}, it would be as verbose as the code itself,
-making it difficult for programmer to understand the overall shape of
-the indexing behaviour.
-
-% GPGPU
-Our work has similarities with efforts to verify kernels
-written for General-Purpose GPU
-programming, such as in \citet{Blom:2014:SoCP}. %Stencils are a form of
-%kernel, and GPGPU programming can be viewed as a massively parallel
-%method of transforming a large matrix.
-However, their focus is
-mainly on the synchronisation of kernels and the avoidance of
-data races., while we are interested in correctness
-%embedded within a more typical general-purpose programming
-%language. %Other work on GPGPU computation, such as
-%\citet{Zhang:2012:CGO}, has focused primarily on generating
-%optimised code based on relatively simple specifications: to
-%provide performance while keeping the programmer's effort within
-%reason.
-%
-% Sketching Stencils
-\citet{Solar-Lezama:2007:PLDI} give specifications of stencils using
-unoptimised ``reference'' stencils, coupled with partial
-implementations which are completed by a code
-generation tool. %These kinds of specifications are just simple
-%implementations, so this tool is useful for hand-written, optimised
-%stencils.
-
-% Pochoir
-\citet{Tang:2011:SPAA} define a specification language for writing stencils
-embedded in C++ (with Cilk~\citep{blumofe1996cilk} extensions) that are then
-compiled into parallel programs based on trapezoidal decompositions with
-hyperspace cuts.
-
-By contrast, the work of \citet{Abe:2013:IPDPSW} studies
-correctness bringing a form of model-checking to
-verify certain stencil computations in the context of
-parallelism
- in partitioned global address
-space languages. %These are scenarios where an array is divided into
-%subarrays on multiple processors but each has the ability to access
-%the others' memory. The authors avoid the state explosion dilemma of
-%model-checking by relying upon the fact that accesses to the
-%``boundary elements'' will require a different method by virtue of the
-%non-locality of that memory. Since this access is performed in a
-%different manner, those can be identified automatically by static
-%analysis. To achieve this,
-Abe \emph{et al.} provide a new language for writing stencil computations. Much of the specification
-effort goes towards describing the distribution of the
-computation over multiple processors. The code for the stencil kernel
-is generated from a relatively high-level specification.
-In contrast, we integrate directly
-into existing, legacy codebases and established languages, bringing
-the benefits of verification more easily to scientific computing.%such as
-%older versions of Fortran.% We infer stencil specifications from
-%Fortran code and check annotations on stencil computations within
-%Fortran code.
-
-\paragraph{Concluding remarks}
-
-There is an increasing awareness of the need for verification
-techniques in
-science~\cite{post2005computational,oberkampf2010verification,orchard2014computational}.
-Our specification language is an approach in this direction, providing
-a system of lightweight, high-level specification for numerical code.
-As opposed to existing approaches our language takes inspiration from
-the numerical literature and provides a concise, abstract description
-of the stencil.
-
-We evaluated our approach by extending CamFort, an open-source analysis
-tool for Fortran programs~\cite{camfort}, running the inference process over a corpus
-of a million lines of Fortran code. We showed that our language is
-capable of capturing many real-world uses of stencils and showed in a
-case-study that it will detect (simulated) programmer errors.
-
-Our restrictive definition of stencils means we do not generate
-specifications for a large number of iterated array accesses in the
-corpus, \eg{} that incorporate reductions. For
-future work, we will look at expanding the flexibility of
-our analysis to include these examples.
-
-\bibliography{references}
-
-\end{document}
-
-%%  LocalWords:  refactoring affine parameterised nonpointed atMost
-%%  LocalWords:  centered readOnce atLeast discretisation Equational
-%%  LocalWords:  equational disjunction denotational dimensionality
-%%  LocalWords:  interprocedural Fortran CamFort preprocessor
-%%  LocalWords:  committers
+found because our analysis restricts the array-subscript-statements
+that we classify as a stencil. In fact, we were able to infer a
+stencil from \overalltickAssignSuccessPercentOfTickAssign\% of these
+statements. A single statement can involve multiple arrays and we
+ended up with \num{\overallnumStencilSpecs} specifications. This shows
+that we can express a large number of stencil shapes within our
+high-level abstraction and validates our initial hypothesis that many
+stencil computations have a regular shape.

### 2017-04-17 oopsla17/model.tex
Equational and approximation theories

@@
 \end{defn}
-%
-\begin{thm}
-The semantic model $\interp{-}_N$ of $N$-dimensional specifications
-is sound with respects to the equational theory of the language,
-that is:
-%
-\begin{equation*}
-\forall S, T, N . \quad
-S \equiv T \; \Rightarrow \;
-\interp{S}_N = \interp{T}_N
-\end{equation*}
-%
-We define the equational theory in \Cref{sec:equational-theory}.
-\end{thm}
 
@@
 
+\subsection{Soundness of model with respect to equational and approximation theories}
+
+\begin{restatable}[Equational soundness]{thm}{equationalSoundness}
+  Lattice model is sound with respect to the equational theory. Let $R$ and $S$
+  be $N$ dimensional regions, then we have
+%
+  \begin{equation*}
+    \forall R, S, N. \; R \equiv S \implies \interp{R}_N = \interp{S}_N
+  \end{equation*}
+\end{restatable}
+
+\begin{restatable}[Approximation soundness]{thm}{approxSoundness}
+  Lattice model is sound with respect to the theory of approximation. Let $R$
+  and $S$ be $N$ dimensional regions, then we have
+%
+  \begin{equation*}
+    \forall R, S, N. \; R \preceq S \implies \interp{R}_N \subseteq \interp{S}_N
+  \end{equation*}
+\end{restatable}
+
 %%% Local Variables:

### 2017-04-17 oopsla17/proofs.tex
Equational and approximation theories

@@
 \end{proof}
+
+\centeredApprox*
+
+\begin{proof}
+  Proof of the first inequality:
+  \begin{align*}
+    \stencil[s]{f}{$n$}{$k$}{\textcap{p}}
+      & \preceq
+        \stencil[s]{f}{$n$}{$k$}{\textcap{p}}
+        \;\texttt{+}\;
+        \stencil[s]{b}{$n$}{$k$}{\textcap{p}}
+      & (\textbf{Combined}) \\
+    & \equiv \stencil[s]{c}{$n$}{$k$}{\textcap{p}} & (\textbf{Centered})
+  \end{align*}
+
+  Proof of the second inequality:
+  \begin{align*}
+    \stencil[s]{b}{$n$}{$k$}{\textcap{p}}
+      & \preceq
+        \stencil[s]{f}{$n$}{$k$}{\textcap{p}}
+        \;\texttt{+}\;
+        \stencil[s]{b}{$n$}{$k$}{\textcap{p}}
+      & (\textbf{Combined}) \\
+    & \equiv \stencil[s]{c}{$n$}{$k$}{\textcap{p}} & (\textbf{Centered})
+  \end{align*}
+\end{proof}
+
+\pointApprox*
+
+\begin{proof}
+  Proof of the first inequality:
+  \begin{align*}
+    \stencil[s]{p}{$n$}{}{}
+      & \preceq \stencil{R}{$n$}{$k$}{nonpointed}
+                \;\texttt{+}\;
+                \stencil[s]{p}{$n$}{}{}
+      & (\textbf{Combined}) \\
+    & \equiv \stencil{R}{$n$}{$k$}{} & (\textbf{Overlapping pointed})
+  \end{align*}
+
+  Proof of the second inequality:
+  \begin{align*}
+    \stencil{R}{$n$}{$k$}{nonpointed}
+      & \preceq \stencil{R}{$n$}{$k$}{nonpointed}
+                \;\texttt{+}\;
+                \stencil[s]{p}{$n$}{}{}
+      & (\textbf{Combined}) \\
+    & \equiv \stencil{R}{$n$}{$k$}{} & (\textbf{Overlapping pointed})
+  \end{align*}
+\end{proof}
+
+\equationalSoundness*
+
+\begin{proof}
+  Both $\equiv$ and $=$ are equivalence relations. We only need to show that
+  each property associated with specifications equivalence holds as an equality
+  in the model.
+%
+  \begin{description}
+    \item[Basic \& Subsumption] These follow from basic properties of lattices.
+%
+    \item[Dist] This is immediate from the definition of distributive
+      lattice which the model is one.
+%
+    \item[Overlapping pointed] Counterparts of individual regions for some
+      Boolean $p$ are given below:
+%
+      \begin{align*}
+        \interp{\stencil{\textcap{R}}{$n$}{$k$}{\texttt{nonpointed}}}_N
+        & = \vecgen{N}{n}{\interv{-k}{k}{\mathsf{false}}} \\
+%
+        \interp{\stencil{p}{$n$}{}{}}_N
+        &= \vecgen{N}{n}{\interv{0}{0}{\mathsf{true}}} \\
+%
+        \interp{\stencil{\textcap{R}}{$n$}{$k$}{}}_N
+        &= \vecgen{N}{n}{\interv{-k}{k}{\mathsf{true}}}
+      \end{align*}
+
+      Using \Cref{lem:vector-union}, it is sufficient to show
+      $ \interv{-k}{k}{p} \;\cup\; \interv{0}{0}{\mathsf{true}} =
+        \interv{-k}{k}{\mathsf{true}} $ holds. This immediately follows from
+      \Cref{lem:zinf-identities}.
+%
+    \item[Centered] Let $\interp{\textcap{p1}}$, $\interp{\textcap{p2}}$, and
+      $\interp{\textcap{p3}}$ be $p1$, $p2$, and $p3$ respectively. We
+      know that $p1$ is $p2 \vee p3$ from (\textbf{Centered}) detinition. Then
+      we get the following interpretation for region constants involved in the
+      equivalence:
+%
+      \begin{align*}
+        \interp{\stencil{c}{$k$}{$n$}{$p1$}}_N
+          &= \vecgen{N}{n}{\interv{-k}{k}{p2 \vee p3}} \\
+%
+        \interp{\stencil{f}{$k$}{$n$}{$p2$}}_N
+          &= \vecgen{N}{n}{\interv{0}{k}{p2} }\\
+%
+        \interp{\stencil{b}{$k$}{$n$}{$p3$}}_N
+          &= \vecgen{N}{n}{\interv{0}{k}{p3}}
+      \end{align*}
+%
+      Using \Cref{lem:vector-union}, it is sufficient to show $\interv{-k}{k}{p2
+      \vee p3} = \interv{0}{k}{p2} \cup \interv{-k}{0}{p3}$ holds. This
+      immediately follows from \Cref{lem:zinf-identities}.
+  \end{description}
+\end{proof}
+
+\approxSoundness*
+
+\begin{proof}
+  Both $\preceq$ and $\subseteq$ are partial orders on sets. We only need to
+  show that each inequality holds in the model.
+%
+  \begin{description}
+    \item[Equivalence] Region equavilance is modeled with set equality and the
+      partial order on regions is modeled with subset relation. Partial orders
+      are reflexive and the model of eqivalence agrees with that of the partial
+      order, so this rule holds.
+%
+    \item[Combined] \texttt{*} and \texttt{+} maps to meet and join operations
+      in the lattice. All meets and joins in a lattice satisfy these
+      inequalities.
+
+    \item[Depth] When $m$ is $k$ and $l$ in
+      $\stencil{\textcap{R}}{$n$}{$m$}{\textcap{p}}$, we obtain the
+      interpretations $\vecgen{N}{n}{\textit{int}_k}$ and
+      $\vecgen{N}{n}{\textit{int}_l}$ for some $N$ as interpretations of
+      regions. Call these \textit{ks} and \textit{ls} respectively. One way to
+      show $\textit{ks} \subseteq \textit{ls}$ is to show that the intersection
+      of \textit{ks} and \textit{ls} is \textit{ks}.
+
+      To show that the intersection is \textit{ks}, we use
+      \Cref{lem:vector-intersect} which reduces the problem to showing that the
+      intersection of $\textit{int}_k$ and $\textit{int}_l$ is the former since
+      at every point except $n$ the intervals agree, hence the intersection
+      remains the same at these dimensions. Then we use
+      \Cref{lem:zinf-identities} to calculate the intersection. By assumption we
+      have $k \leq l$, so the maximum of the lower bounds is $-k$ and the
+      minimum of lower bounds is $k$ this is same as interval $\textit{int}_k$.
+      Note that pointedness is the same for both intervals and depths are
+      positive integers, so the pointedness is preserved under intersection.
+      This shows the intersection of \textit{ls} and \textit{ks} is indeed
+      \textit{ks} and concludes the proof.
+  \end{description}
+\end{proof}

### 2017-04-17 oopsla17/evaluation.tex

@@
 language.
-
-We would not expect to infer a stencil for each array statement we
-found because our analysis restricts the array-subscript-statements
-that we classify as a stencil. In fact, we were able to infer a
-stencil from \overalltickAssignSuccessPercentOfTickAssign\% of these
-statements. A single statement can involve multiple arrays and we
-ended up with \num{\overallnumStencilSpecs} specifications. This shows
-that we can express a large number of stencil shapes within our
-high-level abstraction and validates our initial hypothesis that many
-stencil computations have a regular shape.

### 2017-04-17 oopsla17/related-work.tex
Add related work back in

--- 2016-11-16 paper.tex
@@
-\documentclass[10pt,preprint,numbers]{sigplanconf}
-
-\usepackage{amssymb}
-\usepackage{amsmath}
-\usepackage{amsthm}
-\usepackage{stmaryrd}
-\usepackage{color}
-\usepackage{graphics}
-\usepackage{fancyvrb}
-\usepackage{subfigure}
-\usepackage{multirow}
-\usepackage{minted}
-\usepackage{breqn}
-\usepackage{enumitem}
-%\usepackage{subcaption}
-\usepackage{siunitx} % For pretty-printing numeric values and SI units
-                     % of measure. e.g., the tabular column type S is
-                     % used to print nice-looking tables of numbers.
-\sisetup{ % defaults
-  group-separator={,},
-  group-minimum-digits={3},
-  output-decimal-marker={.},
-  table-format = 6
-}
-
-\usepackage{tikz}
-\usepackage{xypic}
-\usepackage{natbib}
-
-\usetikzlibrary{calc}
-
-%% http://tex.stackexchange.com/questions/55068/is-there-a-tikz-equivalent-to-the-pstricks-ncbar-command
-\tikzset{
-    ncbar angle/.initial=90,
-    ncbar/.style={
-        to path=(\tikztostart)
-        -- ($(\tikztostart)!#1!\pgfkeysvalueof{/tikz/ncbar angle}:(\tikztotarget)$)
-        -- ($(\tikztotarget)!($(\tikztostart)!#1!\pgfkeysvalueof{/tikz/ncbar angle}:(\tikztotarget)$)!\pgfkeysvalueof{/tikz/ncbar angle}:(\tikztostart)$)
-        -- (\tikztotarget)
-    },
-    ncbar/.default=0.5cm,
-}
-
-\tikzset{round left paren/.style={ncbar=0.4cm,out=110,in=-110}}
-\tikzset{round right paren/.style={ncbar=0.4cm,out=70,in=-70}}
-
-\fvset{
-  linenos=true,
-  fontsize=\footnotesize,
-  breaklines=true,
- breakafter=w),
-  xleftmargin=\parindent,
-  numbersep=1em
-}
-
-\usemintedstyle{vs}
-
-\definecolor{darkblue}{rgb}{0.0,0.0,0.5}
-\definecolor{darkgreen}{rgb}{0.0,0.4,0.0}
-\definecolor{darkdarkgreen}{rgb}{0.0,0.35,0.0}
-
-\usepackage{hyperref}
-\hypersetup{
-    unicode=false,          % non-Latin characters in Acrobat's bookmarks
-    pdftoolbar=true,        % show Acrobat toolbar?
-    pdfmenubar=true,        % show Acrobat menu?
-    pdffitwindow=false,      % page fit to window when opened
-    pdftitle={},    % title
-    pdfauthor={}
-    pdfsubject={},   % subject of the document
-    pdfnewwindow=true,      % links in new window
-    pdfkeywords={keywords}, % list of keywords
-    colorlinks=true,       % false: boxed links; true: colored links
-    linkcolor=darkblue,          % color of internal links
-    citecolor=darkblue,        % color of links to bibliography
-    filecolor=green,      % color of file links
-    urlcolor=blue,          % color of external links
-}
-
-\usepackage{cleveref}
-\crefname{section}{\S}{\S\S}
-\Crefname{section}{\S}{\S\S}
-\crefname{subsection}{\S}{\S\S\S}
-\Crefname{subsection}{\S}{\S\S\S}
-\crefname{figure}{fig.}{}
-\Crefname{figure}{Fig.}{}
-\crefname{definition}{def.}{}
-\Crefname{definition}{Def.}{}
-
-\CustomVerbatimEnvironment{SpecVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.5cm,
-xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.98,numbersep=0.9em}
-\CustomVerbatimEnvironment{ExmVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.5cm,
-xrightmargin=0.2cm,baselinestretch=0.98,numbers=left,numbersep=0.9em,commandchars=\\\{\}}
-\CustomVerbatimEnvironment{IVerbatim}{Verbatim}{fontsize=\relsize{-1},xleftmargin=0.5cm,
-xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.98,numbersep=0.9em}
-
-\definecolor{darkgreen}{rgb}{0.0,0.5,0.0}
-\definecolor{darkpurple}{rgb}{0.6,0.0,0.6}
-\definecolor{orange}{rgb}{0.8,0.4,0.0}
-\definecolor{darkorange}{rgb}{0.5,0.2,0.0}
-\definecolor{marco}{rgb}{0.0,0.3,0.5}
-\definecolor{gray}{rgb}{0.2,0.2,0.2}
-
-\newcommand\mdef{\stackrel{\mathclap{\normalfont\mbox{def}}}{=}}
-
-\newcommand{\bn}{\mathbb{N}}
-
-\newcommand{\todo}[1]{\textcolor{blue}{Dom: #1}}
-\newcommand{\dnote}[1]{\textcolor{darkpurple}{Dom: #1}}
-\newcommand{\mnote}[1]{\textcolor{darkgreen}{Mistral: #1}}
-\newcommand{\anote}[1]{\textcolor{red}{Andy: #1}}
-
-\newcounter{block}
-
-\newtheorem{lemma}[block]{Lemma}
-\newtheorem{proposition}[block]{Proposition}
-
-\theoremstyle{definition}
-
-\newtheorem{theorem}[block]{Theorem}
-\newtheorem{remark}[block]{Remark}
-\newtheorem{example}[block]{Example}
-\newtheorem{definition}[block]{Definition}
-
-% Writing macros
-\newcommand{\ie}{\emph{i.e.}}
-\newcommand{\eg}{\emph{e.g.}}
-
-\newcommand{\dimId}{\texttt{dim}}
-
-% Semantics related
-\newcommand{\interp}[1]{\llbracket{#1}\rrbracket}
-
-% Syntax macros
-\newcommand{\nonterm}[1]{\textit{#1}}
-\newcommand{\term}[1]{\texttt{#1}}
-
-\newcommand{\stenRefl}[1]{\term{pointed} \, (\term{dim=}#1)}
-\newcommand{\stenFwd}[3]{\term{forward} \, (\term{depth=}#1,
-  \term{dim=}#2{#3})}
-\newcommand{\stenBwd}[3]{\term{backward} \, (\term{depth=}#1,
-  \term{dim=}#2{#3})}
-\newcommand{\stenCen}[3]{\term{centered} \, (\term{depth=}#1,
-  \term{dim=}#2{#3})}
-\newcommand{\irrefl}{\texttt{nonpointed}}
-
-\newcommand{\stenReflS}[1]{\term{pointed} \, (\term{dim=}#1)}
-\newcommand{\stenFwdS}[2]{\term{fwd} \, (\term{depth=}#1,
-  \term{dim=}#2)}
-\newcommand{\stenBwdS}[2]{\term{bwd} \, (\term{depth=}#1,
-  \term{dim=}#2)}
-\newcommand{\stenCenS}[2]{\term{cen} \, (\term{depth=}#1,
-  \term{dim=}#2)}
-\newcommand{\irreflS}{\texttt{np}}
-
-\newcommand{\stenFwdSR}[3]{\term{fwd} (\term{depth=}#1,
-  \term{dim=}#2, #3)}
-\newcommand{\stenBwdSR}[3]{\term{bwd} (\term{depth=}#1,
-  \term{dim=}#2, #3)}
-\newcommand{\stenCenSR}[3]{\term{cen} (\term{depth=}#1,
-  \term{dim=}#2, #3)}
-\newcommand{\stenReflSR}[1]{\term{point} (\term{dim=}#1)}
-
-% SYNTAX OPERATIONS AND PREDICATES
-%\newcommand{\neigh}{\textsf{neigh}}
-\newcommand{\arrayTy}{\textsf{array}}
-\newcommand{\rhsExp}{\textsf{rhsExp}}
-\newcommand{\var}{\textsf{var}}
-
-%% VECTOR NOTATIONS
-\newcommand{\vect}[1]{\textbf{#1}}
-
-\newcommand{\vtwohs}[2]{\setlength{\arraycolsep}{0em}
-\left[\begin{array}{cc}#1 \, & \, #2\end{array}\right]\!}
-
-\newcommand{\vtwoh}[2]{\setlength{\arraycolsep}{0em}
-\left[\begin{array}{cc}#1 \; & \; #2\end{array}\right]}
-
-\newcommand{\vthreeh}[3]{\setlength{\arraycolsep}{0em}
-\left[\begin{array}{ccc}#1 \; & \; #2 \; & \; #3\end{array}\right]}
-\newcommand{\vtwo}[2]{\setlength{\arraycolsep}{0em}
-\left[\begin{array}{l}#1\\#2\end{array}\right]}
-\newcommand{\vthree}[3]{\setlength{\arraycolsep}{0em}
-\left[\begin{array}{l}$#1$\\$#2$\\$#3$\end{array}\right]}
-\newcommand{\stwo}[4]
-%{\vtwo{#1}{#2}\!\vtwo{#3}{#4}}
-{\setlength{\arraycolsep}{0.1em}
-\left[\begin{array}{rr}$#1$ & $#3$\\$#2$ & $#4$\end{array}\right]}
-
-\newcommand{\singleEntry}[2]{\textbf{J}_{#2}^{#1}}
-\newcommand{\zeroEntry}[2]{\textbf{K}_{#2}^{#1}}
-
-%% OPERATIONS ON SPANS and VECTORS
-\newcommand{\containedin}{\sqsubseteq}
-
-%% MODEL
-\newcommand{\effdims}[2]{\mathit{constr}(#1)_{#2}}
-
-\newcommand{\trule}[1]{{\footnotesize{(\text{#1})}}}
-%\include{results}
-
-%\newcommand{\cons}[3]{\mathbf{C}_{#3}(#1, #2)}
-\newcommand{\consName}{\textbf{agree\textprime}}
-\newcommand{\cons}[2]{#1 \,\, \consName{} \,\, #2}
-\newcommand{\consAName}{\textbf{agree}}
-\newcommand{\consA}[2]{#1 \,\, \consAName{} \,\, #2}
-\newcommand{\consSub}[3]{#2 \,\, \consAName{}_{#1} \,\,#3}
-
-\title{Verifying Spatial Properties of Stencil Computations}
-%\titlebanner{DRAFT -- Do not distribute}
-\authorinfo{}{}{}
-%\preprintfooter{DO NOT DISTRIBUTE}
-
-\begin{document}
-\maketitle
-
-\bibliographystyle{plainnat}
-
-\begin{abstract}
-Stencil computations are a common programming pattern in numerical
-  code and typically form the core of graphics and scientific
-  computing applications. However, they are often complex and
-  unwieldy, involving substantial fine-grained manipulation of array
-  indices across multiple arrays and dimensions. This requires careful
-  testing, which is almost always done manually and then discarded.
-  In this paper, we propose an alternate approach:
-  a specification language and automated tool for verifying the
-  spatial behaviour of stencils. The specification language is based
-  on the hypothesis that stencil computations read from arrays with a
-  regular, fixed pattern which can be captured by a simple
-  set of abstract combinators. These
-  specifications can be embedded as annotations in source code,
-  against which code can be checked for conformance. For legacy code,
-  spatial specifications can be inferred, providing documentation and
-  aiding future software maintenance. We evaluate our specification
-  language and verification tool against
-  a corpus of numerical Fortran code ($\sim$ 1 million lines) for
-  which we generate $\sim$60,000 specifications, showing the vast majority
-  of stencil computations indeed have a simple, regular, static shape. % We introduce the
-  %language, checking and inference procedures, and its evaluation.
-\end{abstract}
-
-%\keywords{program verification, specification, stencils}
-
-\section{Introduction}\label{sec:intro}
-
-\emph{Stencils} are a ubiquitous programming pattern, common in
-scientific and numerical computing applications. Informally, a stencil
-computation computes an array whose value at each index $i$ is
-calculated from a \emph{neighbourhood} of values around $i$ in some
-input array(s), \eg{}, the Game of Life, convolutions in image
-processing, approximations to differential equations. For example, the
-following computes the one-dimensional discrete Laplace transform (an
-approximation to a derivative) in Fortran:
-%
-\begin{minted}{fortran}
-do i = 1, (n-1)
-      b(i) = a(i-1) - 2*a(i) + a(i+1)
-end do
-\end{minted}
-%
-Values \texttt{b(i)} are calculated from a neighbourhood of elements
-around \texttt{i} in the input array \texttt{a}.  In this example, the
-access pattern is simple and easily understood. More complex
-stencil computations are more prone to errors from simple lexical
-mistakes. For example, the following is a small snippet from a
-Navier-Stokes fluid simulator (based on~\citet{griebel1997numerical})
- in which two arrays are read with
-different data access patterns, across two dimensions.
-%%
-\begin{minted}[firstnumber=20,xleftmargin=2em]{fortran}
-du2dx = ((u(i,j)+u(i+1,j))*(u(i,j)+u(i+1,j))+   &
-  gamma*abs(u(i,j)+u(i+1,j))*(u(i,j)-u(i+1,j))- &
-  (u(i-1,j)+u(i,j))*(u(i-1,j)+u(i,j))-          &
-  gamma*abs(u(i-1,j)+u(i,j))*(u(i-1,j)-u(i,j))) &
-  /(4.0*delx)
-
-duvdy = ((v(i,j)+v(i+1,j))*(u(i,j)+u(i,j+1))+   &
-  gamma*abs(v(i,j)+v(i+1,j))*(u(i,j)-u(i,j+1))- &
-  (v(i,j-1)+v(i+1,j-1))*(u(i,j-1)+u(i,j))-      &
-  gamma*abs(v(i,j-1)+v(i+1,j-1))*(u(i,j-1)-     &
-  u(i,j))) / (4.0*dely)
-
-laplu = (u(i+1,j)-2.0*u(i,j)+u(i-1,j))/delx/delx+ &
-  (u(i,j+1)-2.0*u(i,j)+u(i,j-1))/dely/dely
-
-f(i,j) = u(i,j)+del_t*(laplu/Re-du2dx-duvdy)
-\end{minted}
-%%
-The access pattern is much harder to understand than the
-one-dimensional Laplace.  The miasma of indexing expressions of the
-form \texttt{var(i$\pm$a, j$\pm$b)} is not only hard to read, but is
-prone to simple textual input mistakes, \eg{}, swapping \texttt{-} and
-\texttt{+}, missing an indexing term, or transforming the wrong
-variable \eg{} \texttt{(i+1,j)} instead of \texttt{(i,j+1)}.
-
-In practice, the typical development procedure for complex
-stencil computations involves some ad hoc testing to ensure that no
-transcription mistakes have been made (\eg{}, by visual
-inspections, or comparison against an manufactured or 
-analytical solutions~\cite{farrell2010automated}). Typically such
-testing is discarded once the code is seen to be correct.
-
-This is not the only information that is often discarded. The
-``shape'' of the indexing pattern is usually the result of choices made in the
-numerical-analysis procedure to convert some
-continuous equations into a discrete approximation. Rarely are these
-decisions captured in the source code, yet the shape of access is
-usually uniform and is has a perspicuous and concise description in
-numerical analysis literature \eg{}, ``\emph{centered in space, to a
-  depth of 1}'' referring to indexing terms \texttt{a(i)},
-\texttt{a(i-1)} and \texttt{a(i+1)}.
-
-To support the development of correct stencil computations, we propose
-a simple, abstract specification language for the data access pattern
-of stencils. The shape of the
-Laplace example stencil is specified in our language as:
-%
-\begin{minted}[linenos=false]{fortran}
-!=  stencil centered(depth=1, dim=1) :: a
-\end{minted}
-%
-That is, \texttt{a} is accessed with a symmetrical pattern
-(``centered'') to a depth of one in each direction in its first
-dimension.  The Navier-Stokes example has a shape
-specified as:
-%
-\begin{minted}[linenos=false,xleftmargin=0em,xrightmargin=0.8em,breakindent=1.9em]{fortran}
-!= stencil centered(depth=1,dim=1)*pointed(dim=2)     + centered(depth=1,dim=2)*pointed(dim=1) :: u
-
-!= stencil forward(depth=1,dim=1)                          * backward(depth=1,dim=2) :: v
-\end{minted}
-%
-The specification requires that, over the whole fragment, \texttt{u}
-is accessed with a centered pattern to depth of 1 in both dimensions
-(this is known as the \emph{five-point stencil}) and \texttt{v} is
-accessed in a neighbourhood bounded forwards to depth of $1$ in the
-first dimension and backward to a depth of $1$ in the second
-dimension. The specification is relatively small
-and much more abstract compared to the source code.
-
-We provide a verification tool which can check the spatial correctness
-of stencil computations against a specification. Furthermore, the tool
-can automatically infer specifications and insert these as comments
-into the source code. The main features and benefits of our approach
-are as follows:
-%
-\begin{itemize}
-%\item The main use case is that a programmer writes a specification
-%  first (perhaps informed by their numerical analysis), then writes
-%  the corresponding source code, and lastly has the tool check its
-%  conformance to the specifications. This specify-and-check approach
-%  removes the usually ad hoc testing effort, and provides the
-%  opportunity for easy re-verification in the event of any refactoring
-%  of software maintenance activities.
-\item The primary use case is that a programmer writes a specification
-  first before the stencil code. Our tool then checks conformance between
-  the two. This specify-and-check approach reduces testing effort and
-  future-proofs against bugs introduced during refactoring and maintenance.
-%
-%\item The specification provides a concise description which can be
-%  preserved in the source code to provide documentation to future
-%  developers, contrasting with ad hoc tests which are usually
-%  discarded or specialised to the development environment. Separate from the
-%  verification procedure, the infer-and-synthesise mode of the tool
-%  provides a way to automatically generate additional useful
-%  documentation for future users.
-
-% Emphasis here is on documentation generation, so it makes little sense to
-% compare this aspect of CamFort to ad hoc testing. Apples and oranges.
-\item A specification concisely captures the access pattern of a
-  stencil and turns it into documentation. The inference procedure
-  efficiently produces specifications with no programmer intervention,
-  automatically inserting specifications at appropriate places in the
-  source code.
-%
-%\item Specifications are integrated into the code as comments for easy
-%  integration with existing tools, as well as documentation systems (\eg{}, we
-%  support Doxygen and FORD syntax).
-%
-\item Our specification format is deliberately abstract, with a small number of
-  combinators that \emph{do not involve any indexing expressions}, \eg{}
-  \texttt{a(i+1,j-1)}. This contrasts with other specification
-  approaches, \eg{}, deductive verification tools such as ACSL, where
-  specifications must also be in terms of array-indexing expressions. Therefore, any low-level mistakes
-  that could be made whilst programming complex indexing code could also be
-  made when writing its specification. Our specifications are more
-  abstract and lightweight
-  with the aim of aiding adoption by scientists and preventing indexing errors.
-\end{itemize}
-
-Our verification tool does not target a
-class of bugs that can be detected automatically (\emph{push-button
-  verification}). Instead, stencil computation bugs must be
-identified relative to a specification of the intended access pattern.
-
-In this paper, we make the following contributions:
-%
-\begin{itemize}
-\item We introduce a specification language for stencils
-  that captures common forms of data access patterns
-  (\cref{sec:lang});
-
-\item We detail checking and inference algorithms for our
-  specifications (\cref{sec:analysis}), derived from a
-  denotational model for the specification language
-  (\cref{sec:semantics});
-
-\item We provide an implementation of our approach as an extension to
-  CamFort~\cite{camfort}, an open-source program analysis tool for Fortran;
-
-\item We report on a quantitative study of stencil computations on a
-  corpus of numerical Fortran program
-  (\cref{sec:evaluation}), totalling one million lines of code. Our tool
-  identifies and infers specifications for roughly 60,000 stencil computations in the
-  corpus. Approximately 5,000 of the stencils we found are non-trivial,
-  corresponding to code which is a possible source of errors. This validates
-  our hypothesis that the majority of stencil computations have a regular
-  shape, and validates the design of our language in its capability to capture
-  many core patterns.
-
-\item We give a verification case study of a
-  stencil computation commonly used in scientific computing,
-  simulating programming errors (\Cref{sec:jacobi}). Our approach detects all
-  possible simulated programming errors for this problem.
-\end{itemize}
-%
-
-\newcommand{\domainVal}{\mathbb{Z}_\infty}
-\section{Stencil specifications}\label{sec:lang}
-
-Our specification language is based on the hypothesis
-that most forms of array access in numerical code have
-a fixed, statically-determined access pattern. For example, the
-\emph{five-point stencil} on a two-dimensional array reads from array
-indices $(i, j)$, $(i-1, j)$, $(i+1, j)$, $(i, j-1)$, and $(i, j+1)$
-for all $i, j$ within the inner boundary of the array (to avoid
-out-of-bounds access at the edges). We revisit this hypothesis
-in \cref{sec:evaluation} with the inference of
-such regular stencil patterns on our corpus of numerical programs.
-
-We give various definitions used in the paper then
-\cref{subsec:syntax} outlines the syntax of stencil specifications
-with an informal introduction. \cref{sec:semantics}
-gives a semantic model via a set-based interpretation which makes
-concrete the meaning of specifications. \cref{sec:eqs} then provides
-a syntactic equational theory for specifications which is
-sound with respect to the model.
-%sub-specification for which the
-%equality and sub-specification relation are proven sound.
-
-\paragraph{Notation}
-%\label{sec:notation}
-
-\renewcommand*{\arraystretch}{0.8}
-For the target language, $e$ ranges over expressions (may be impure) and $v$
-over its (imperative) variables. %By
-%``variable'', we mean in the imperative sense (named binders to memory cells).
-
-\begin{definition}[Induction variable]
-  An integer variable is an \emph{induction variable} if it is the control
-  variable of a ``\texttt{for}'' loop incremented by $1$ per iteration. A variable
-  is interpreted as an induction variable only within the scope of the loop
-  body. Throughout, $i, j, k$ range over induction variables.
-%TODO: consider mentioning derived, non-based induction variables
-% as we did before.
-\end{definition}
-
-\begin{definition}[Array subscripts and indices]
-  An \emph{array subscript}, denoted $a(\bar{e})$, is an expression
-  which reads from an array $a$ at an \emph{index} specified by a
-  comma-separated sequence of integer expressions denoted $\bar{e}$ or
-  in expanded form as $(e_1, \ldots, e_n)$. An index $e_i$
-  is called \emph{relative} if the expression involves an induction
-  variable.
-  An \emph{absolute index} is a constant integer expression relative to the
-  enclosing loop.
-\end{definition}
-
-\begin{definition}[Neighbourhood index]
-\label{def:neighbour}
-  For an array subscript $a(\bar{e})$ an index $e \in \bar{e}$ is a
-  \emph{neighbourhood index} if $e$ is of the form $e \equiv i$, $e \equiv i +
-  c$, or $e \equiv i - c$, where $c$ is an integer constant. That is, a
-  neighbourhood index is a constant translation of an induction variable. (The
-  relation $\equiv$ identifies terms up-to commutativity of $+$ and the inverse
-  relationship of $+$ and $-$ \eg{}, $(-b) + i \equiv i - b$).
-\end{definition}
-
-\begin{definition}[Stencil computations]
-\label{def:stencil}
-  Let $a$ be an array of dimensionality $n$ and $\overline{b}$ a
-  collection of arrays of arbitrary (possibly differing)
-  dimensionalities.
-  A \emph{stencil computation} comprises an iteration by a set $I$ of at
-  most $n$ induction variables
-  over a subset of the index space of $a$. Each iteration
-  determines the elements of $a$ by an
-  assignment $a(\bar{e}) = e_r$ where each $e \in
-  \bar{e}$ is either a neighbourhood or constant index.
-  For all $b \in \bar{b}$ and array subscripts $b(\bar{e}')$
-  that flow to $e_r$, each $e' \in \bar{e}'$ is
-  either a neighbourhood or absolute index.
-
-  We refer to the assignment $a(\bar{e}) = e_r$ and all the associated
-  statements with dataflow to $e_r$ as the \emph{stencil kernel}.
-\end{definition}
-%
-% specifying the shape of indexing expressions for
-%arrays $\bar{v}$ that are simultaneously permitted and expected within
-%the code to which the specification is given.
-%
-%\marginpar{\dnote{Is it worth having this informal consistency here?}}
-
-For now we informally describe what it means for a stencil kernel to
-be consistent with a specification. This will be formalised in
-\Cref{sec:semantics} and algorithmically described in and
-\Cref{sec:analysis}.  Specifications are associated to a stencil
-kernel and are declared for one or more array variables $\bar{v}$
-within that kernel.  A stencil specification defines a \emph{region}
-associated to an array variable $a$ against which a stencil kernel
-should be \emph{consistent}. The region defined by a specification is
-an $n$-dimensional rectilinear shape in a discrete space crossing the
-origin with respect to the induction variable. Consistency is then a
-two way requirement. All array subscripts on $a$ contributing to the
-kernel should be consistent with the region. Conversely, all parts of
-the region should have an index corresponding it.
-
-\begin{figure}[t]
-\vspace{-0.9em}
-\begin{align*}
-\def\arraystretch{1}
-\setlength{\arraycolsep}{0.2em}
-\newcommand{\dimTy}{\mathbb{N}_{>0}}
-\begin{array}{rl}
-\nonterm{specification} ::= & \nonterm{regionDec} \mid \nonterm{specDec} \\
-\nonterm{specDec} ::= & \term{stencil} \; \nonterm{spec} \;
-                        \texttt{::} \; v \\
-\nonterm{regionDec} ::= &  \texttt{region} \; \texttt{::} \; \nonterm{rvar} \; \texttt{=} \;
-                         \nonterm{region}\\[0.4em]
-%\nonterm{spec} ::= & \nonterm{spatial} \mid \nonterm{temporal}
-%\\[1em]
-\nonterm{spec} ::= & [\nonterm{mult},] \; [\nonterm{approx},] \; \nonterm{region} \\
-\nonterm{mult} ::= & \term{readOnce} \\
-\nonterm{approx} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
-\nonterm{region} ::= & \nonterm{rvar} \; \\
-\multicolumn{2}{l}{\qquad \mid \; \stenRefl{\dimTy}} \\
-\multicolumn{2}{l}{\qquad \mid \;\stenFwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{nonpointed}]}} \\
-\multicolumn{2}{l}{\qquad \mid \; \stenBwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{nonpointed}]}} \\
-\multicolumn{2}{l}{\qquad \mid \; \stenCen{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{nonpointed}]}} \\
-\multicolumn{2}{l}{\qquad \mid \; \nonterm{region} \, \term{+}
-  \, \nonterm{region} \; \mid \; \nonterm{region} \; \term{*} \; \nonterm{region}} \\[0.25em]
-%\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{rvar}}
-%\\[0.5em]
-%\nonterm{temporal} ::= \; & \term{dependency} \; (v \; \{ , v \}) [, \texttt{mutual}]
-%  \\[0.5em]
-\nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+\\[-1em]
-\end{array}
-\end{align*}
-\caption{Specification syntax (EBNF grammar)}
-\label{fig:syntax}
-\vspace{-0.8em}
-\end{figure}
-
-\subsection{Specification syntax}
-\label{subsec:syntax}
-
-\Cref{fig:syntax} gives the syntax of stencil specifications, which is
-detailed below. The entry point is the \nonterm{specification} production which
-splits into either a \emph{region declaration} or a \emph{specification
-declaration}.
-Regions comprise \emph{region constants} which are
-combined via region operators \term{+} and \term{*}.
-
-%\paragraph{Region constants}
-
-Region constants specify a finite contiguous region in a
-single dimension relative to the origin and are either \term{pointed},
-\term{forward}, \term{backward}, or \term{centered}. The region names are
-inspired by the scientific terminology, \eg{} the standard explicit method for
-approximating PDEs is known as the \emph{Forward Time, Centered Space} (FTCS)
-scheme~\citep{dawson1991finite}.
-
-Each region
-constant has a dimension identifier $d$ given as a positive natural number.
-Each constant except \term{pointed} has a depth
-parameter $n$ given as a positive natural number; \term{pointed}
-regions implicitly have a depth of $0$.
-
-A \term{forward} region of depth $n$ specifies a contiguous
-region in dimension $d$ starting at the origin. This corresponds
-to specifying neighbourhood indices in dimension $d$ ranging from $i$ to $i + n$
-for some induction variable $i$. Similarly, a
-\term{backward} region of depth $n$ corresponds to contiguous indices
-from $i$ to $i - n$ and \term{centered}
-of depth $n$ from $i - n$ to $i + n$. A \term{pointed}
-stencil specifies a neighbour index $i$. For example, the
-following shows four specifications with four consistent stencil
-kernels reading from arrays \term{a}, \term{b}, \term{c} and \term{d}:
-%%
-\begin{minted}{fortran}
-!= stencil forward(depth=2, dim=1) :: a
-e(i, 0) = a(i, 0) + a(i+1, 0) + a(i+2, 0)
-!= stencil backward(depth=2, dim=1) :: b
-e(i, j) = b(i, j) + b(i-1, j) + b(i-2, j)
-!= stencil centered(depth=1, dim=2) :: c
-e(i, j) = (c(j-1) + c(j) + c(j+1))/3.0
-!= stencil pointed(dim=3) :: d
-e(i, j) = d(0, 0, i)
-\end{minted}
-%%
-Not every dimension needs to be specified, \eg{},
-specifications on lines $1$, $3$, and $7$ leave some dimensions unspecified.
-
-The \term{forward}, \term{backward}, and \term{centered} regions may
-all have an additional attribute \term{nonpointed} which marks absence
-of the origin.  For example, the following is a
-\term{nonpointed} \term{backward} stencil
-%
-\begin{minted}{fortran}
-!= stencil backward(depth=2, dim=1, nonpointed) :: a
-b(i) = a(i-1) + 10*a(i-2)
-\end{minted}
-
-\paragraph{Combining regions}
-
-% r+s means that indices must be consistent with either r or s
-% r*s means that indices must be consistent with both r and s at the
-% same time
-
-% specifications correspond to a set of schemes
-
-% r+s means an index must be generalised by a scheme in either r or s
-%       and every scheme in r must have an index which specialises it
-%       and every scheme in s must have an index which specialises it
-
-% r*s means an index must be generalised by a scheme in r and a scheme in s
-%       and every pairs of schemes in r and s must have an index which
-%           specialises both
-%
-
-The region operators \term{+} and \term{*} respectively combine
-regions disjunctively and conjunctively.
-
-The conjunction of two regions $r \term{*} s$ means that any indices
-in the specified code must be consistent with both $r$ and $s$
-simultaneously. Dually, for every pair of components in the regions
-$r$ and $s$ there must be an index consistent with both.
-For example, the following \emph{nine-point stencil}
-has a specification given by the product of two \texttt{centered}
-regions in each dimension:
-%%
-\begin{minted}[breakindent=2.9em]{fortran}
-x = a(i, j)   + a(i-1, j)   + a(i+1, j)
-y = a(i, j-1) + a(i-1, j-1) + a(i+1, j-1)
-z = a(i, j+1) + a(i-1, j+1) + a(i+1, j+1)
-!= stencil centered(depth=1, dim=1) * centered(depth=1, dim=2) :: a
-b(i, j) = (x + y + z) / 9.0
-\end{minted}
-%%
-This pattern is common in image convolution applications.
-The specification ranges over the
-values that flow to the array subscript on the left-hand side,
-and so ranges over the intermediate assignments to \term{x},
-\term{y}, and \term{z}. Each index in the code is consistent
-with both specifications simultaneously, \eg{}, \texttt{a(i-1, j+1)}
-is within the centered region in dimension $1$ and the centered region
-in dimension $2$.
-
-The product $r \term{*} s$ can also be thought of as a bounding box
-over the two regions $r$ and $s$.
-
-The disjunction of two regions $r \term{+} s$ means that any indices
-in the specified code must be consistent with either of $r$ or $s$.
-Dually, every part of region $r$ must have a
-corresponding index and similarly for $s$ independently. For
-example, the following gives the specification of a five-point
-stencil which is the sum of two compound \texttt{pointed} and
-\texttt{centered} regions in each dimension:
-%
-\begin{minted}{fortran}
-!= stencil centered(depth=1, dim=1)*pointed(dim=2)    + centered(depth=1, dim=2)*pointed(dim=1) :: a
-b(i,j) = -4*a(i,j) + a(i-1,j) + a(i+1,j) &
-                   + a(i,j-1) + a(i,j+1)
-\end{minted}
-%%
-Here the left-hand side of \texttt{+} says that when the second dimension
-(induction variable $j$) is fixed at the origin, the first dimension
-(induction variable $i$) accesses the immediate vicinity of the origin
-(to depth of one). The right hand side of \texttt{+} is similar but the dimensions are reversed.
-This reflects the symmetry under rotation of the five-point stencil.
-
-\paragraph{Region declarations and variables}
-
-Region specifications can be assigned to region variables via
-region declarations. For example, the shape of a
-``\emph{Roberts cross}'' edge-detection convolution~\cite{davis1975survey}
-can be stated:
-%%
-\begin{minted}{fortran}
-!= region :: r1 = forward(depth=1, dim=1)
-!= region :: r2 = forward(depth=1, dim=2)
-!= region :: robertsCross = r1*r2
-!= stencil robertsCross :: a
-\end{minted}
-This is useful for common stencil patterns, such as the five-point
-pattern, as the region can be defined once and reused.
-%%
-\paragraph{Modifiers}
-%%
-Region specifications can be modified
-by \emph{approximation} and \emph{multiplicity} information
-(in \textit{spec} in \Cref{fig:syntax}).
-
-The \texttt{readOnce} modifier enforces that no index appears more
-than once (that is, its multiplicity is one). For example, all of
-the previous examples could have \texttt{readOnce} added:
-%
-\begin{minted}{fortran}
-!= stencil readOnce, backward(depth=2, dim=1) :: a
-b(i+1) = a(i) + a(i-1) + a(i-2)
-\end{minted}
-%
-This specification would be invalid if any of the
-array subscripts were repeated. This modifier provides a way to
-rule out any accidental repetition of array subscripts.
-The notion is similar to that of linear types~\cite{wadler1990linear}, where a value must be used
-exactly once. We opt for the more informative and easily understood name
-\texttt{readOnce}. This modifier is optional, so it need not
-be present even if the stencil is linear.
-
-In some cases, it is useful to give a lower and/or upper bound for a
-stencil. This can be done using either the \term{atMost} or
-\term{atLeast} modifiers. This is particularly useful in situations
-where there is a non-contiguous stencil pattern, which cannot be expressed
-precisely in our syntax. For example:
-%
-\begin{minted}{fortran}
-!= stencil atLeast, pointed(dim=1)         :: a
-!= stencil atMost, forward(depth=4, dim=1) :: a
-b(i) = a(i) + a(i+4)
-\end{minted}
-
-\section{A model for specifications}
-\label{sec:semantics}
-
-We define a simple set-theoretic denotational model of the semantics
-of our specification language. This has a number of purposes:
-(1) it serves to explain the meaning of our specifications
-(2) it is used in the inference and checking algorithms
-(\Cref{sec:analysis}); (3) justifies an equational theory for
-specifications in the next section; and (4) can be used to guide
-correct implementations. The model domain is over sets of vectors
-which we call \emph{index schemes}.
-
-\begin{definition}
-  An \emph{index scheme} is a vector of size $n$ (an $n$-vector) with
-  values drawn from $\domainVal = \mathbb{Z} \cup \{\infty\}$.  Values
-  in $\mathbb{Z}$ represent the offsets of neighbourhood indices
-  (\ie{}, the constant integer $c$ in an indexing expression of the
-  form $i \pm c$, \Cref{def:neighbour}) and $\infty$ is a wild-card
-  representing both constant indices and unspecified indices.
-  For an index scheme $u$, the $i^{th}$ element is denoted $u_i$. We
-  write $(\domainVal)^n$ for the set of all $n$-dimensional schemes.
-\end{definition}
-
-Indexing schemes abstract array subscripts.
-Our model, consisting index schemes, explain the range of possible
-indexing behaviours which are consistent with the specification.
-The relationship between source language
-and index schemes
- is defined by \textsf{schematic} function,
- which forms part of checking and inference (\Cref{sec:analysis}).
-
-\newcommand{\schemeOne}{\textsf{s}_I}
-%\newcommand{\schemeOne}{\textsf{schema$^1_I$}}
-\newcommand{\schemeN}{\textsf{schematic}_I}
-%\newcommand{\schemeN}{\textsf{schema$^n_I$}}
-\begin{definition}%[Schematic]
-\label{def:schematic}
-Let $\schemeOne$ be a partial function, parameterised by a set of
-induction variables $I$, mapping source language expressions to
-$\domainVal$ values:
-%%
-\begin{align*}
-\schemeOne(e)
- & =  \begin{cases}
-c & e \equiv i + c \;\; \wedge \;\; i \in I \\
--c & e \equiv i - c \;\; \wedge \;\; i \in I \\
-\infty & \textit{$I \cap \textit{FreeVariables}(e) = \emptyset $} %\\
-%\bot   & \textit{otherwise}
-\end{cases}
-\end{align*}
-In the
-first two cases, expressions that are a constant offset of an
-induction variable (\ie{}, neighbourhood indices) are mapped to the
-offset $c$ or $-c$ $\in \domainVal$. In the third case, if the
-expression is not defined in terms of an induction variable, it
-is mapped to $\infty \in \domainVal{}$. Otherwise, $\schemeOne(e)$
-is undefined ($\bot$).
-
-The partial function $\schemeN$, also parameterised by a set of
-induction variables $I$, uses $\schemeOne$ to
- map from the syntax of indices to index schemes
-$\in (\domainVal)^n$. It is defined on the syntax of indices $\bar{e} = (e_1, ..., e_n)$ as:
-\begin{align*}
-\schemeN(e_1, .., e_n) & =
-[\schemeOne(e_1) ... \schemeOne(e_n) ]
-\; \textit{iff} \; \forall x . \schemeOne(e_x) \neq \bot
-\end{align*}
-%
-Thus, if each expression in an index is a neighbourhood or constant
-index, $\schemeN$ maps it to an indexing scheme.
-\end{definition}
-
-\begin{definition}[Generalisation]
-\label{def:generalisation}
-Given two schemes $v, w$ of dimensionality $n$,
-then $w$ is said to \emph{generalise} $v$
-(or conversely $v$ \emph{specialises} $w$),
-written $v \preceq_n w$ defined:
-\[
-v \preceq_n w \;\; \Leftrightarrow  \;\;
-  \forall i\!\in\!\{ 1, \ldots, n \} . \, (w_i = \infty \, \vee \, v_i = w_i)
-\]
-%%
-where $\preceq_n$ is a reflexive, transitive relation on
-index schemes $\domainVal^n$. Thus,
- $\infty$ generalises any index in the same dimension.
-%
-\end{definition}
-
-\begin{example}
-Consider the following array subscript to indexing schemes
-conversions:
-%
-\begin{align*}
-\begin{array}{ll}
-\schemeN{(i, j+1)} & =\vtwoh{0}{1} \\
-\schemeN{(1, j+1)} & = \vtwoh{\infty}{1} \\
-\schemeN{(i*2, j)} & = \bot
-\end{array}
-\end{align*}
-There is one generalisation between these
-schemes, namely that $\vtwoh{0}{1} \preceq_2 \vtwoh{\infty}{1}$.
-The last example is undefined as its first expression is a
-non-neighbour relative index.
-\end{example}
-
-%Our terminology of \emph{index schemes} and
-%\emph{generalisation} is by analogy
-%with the \emph{type schemes} and \emph{type generalisation} of
-%polymorphic types in ML~\cite{milner1978theory}. Our notion is
-%much more rigid however, as index schemes do not bind names.
-
-\paragraph{Shorthand notation}
-
-We use abbreviations
- \term{fwd}, \term{bwd}, \term{cen}, and \irreflS{}
-for \term{forward}, \term{backward},
-\term{centered}, and \irrefl{} respectively. Furthermore, 
-we quantify over the absence or presence of the \irrefl{}
-attribute on region constants as a boolean. For example,
-\term{fwd} is syntactic sugar defined:
-%%
-\begin{align*}
-\hspace{-0.5em}
-& \stenFwdSR{n}{d}{p}
-\; := \; \\[-0.4em]
-& \quad \begin{cases}
-\stenFwd{n}{d}{, \irrefl} & \textit{if} \;\, \neg \, p \\[-0.1em]
-\stenFwd{n}{d}{}  & \textit{if} \;\,  p
-\end{cases}
-\end{align*}
-%
-$\textit{or} \;\; \stenFwdS{n}{d} := \stenFwd{n}{d}{}$
-and similarly for \term{bwd} and \term{cen}.
-
-\subsection{Denotational model}
-
-An interpretation function $\interp{-}_n$ provides the model 
-(where $n$ is the maximum dimensionality of the specification being
-modelled) mapping closed\footnote{That is, we assume there are no
-  occurrences of \textit{rvar} in a specification being modelled.  Any
-  \emph{open} specification containing region variables can be made
-  closed by straightforward syntactic substitution with a (closed)
-  \textit{region}.} specifications to sets of
-$n$-dimensional indexing schemes. The interpretation is overloaded on
-\emph{regions} in \Cref{fig:region-model} and on the top-level
-of a specification \textit{spec} in \Cref{fig:spatial-model}.
-Various intermediate notions are used.
-
-\begin{definition}A \emph{single-entry vector} of size $n$, denoted
-$\singleEntry{r}{n}$, is a vector where the $r^{th}$ entry is $1$
-and all others are $\infty$, \eg{}, $\singleEntry{1}{2} =
-\vtwoh{1}{\infty}$ and $\singleEntry{2}{3} = \vthreeh{\infty}{1}{\infty}$.
-Thus, a single-entry vector is an index scheme generalising 
-all other schemes that have $1$ as their $r^{th}$ entry,
-describing a neighbour offset of $1$.
-\end{definition}
-
-\begin{definition}
-A \emph{zeroed single-entry vector} of size $n$ denoted
-$\zeroEntry{r}{n}$ is a vector where the $r^{th}$ entry is $0$ and all others
-are $\infty$ \eg{}, $\zeroEntry{1}{2} = \vtwoh{0}{\infty}$.
-Akin to single-entry vectors, zeroed single-entry vectors
-generalise all other schemes
-that have $0$ as their $r^{th}$ entry, representing indices at
-the ``origin'' in $r$.
-\end{definition}
-
-\begin{figure}[t]
-\vspace{-0.9em}
-\input{region-model.tex}
-%\textit{$a^m \in A$ means there are $n$ copies of $a$ in
-%  the multi-set $A$}
-\caption{Model of regions,
-$\interp{-}_n : \textit{region} \rightarrow \mathcal{P}(\domainVal{}^n)$}
-\label{fig:region-model}
-\vspace{-1em}
-\end{figure}
-
-\noindent
-The first four equations of \Cref{fig:region-model}
-model region constants as sets of indexing
-schemes. We illustrate with an example.
-
-\begin{example}
-For a $2$-dimensional stencil computation, then
-\begin{align*}
-& \interp{\stenCenS{2}{1}}_2 \\[-0.1em]
-& \qquad = \; \{i\singleEntry{1}{2} \,|\, i \in \{\text{-}2,..., 2\}\!\setminus\!0 \}
-  \cup  \{ \zeroEntry{1}{2} \,|\, \textit{true} \} \\[-0.2em]
-& \qquad = \; \{-2\singleEntry{1}{2}, \, -\singleEntry{1}{2}, \,
-  \singleEntry{1}{2}, \, 2\singleEntry{1}{2}\} \cup
-  \{\zeroEntry{1}{2}\} \\[-0.2em]
-& \qquad = \; \{\vtwoh{-2}{\infty}, \vtwoh{-1}{\infty}, \vtwoh{0}{\infty},
-\vtwoh{1}{\infty}, \vtwoh{2}{\infty}\}
-\end{align*}
-Note, $\infty$ is absorbing for multiplication above \eg{} $2
- \infty = \infty$.
-\end{example}
-\noindent
-The final two equations of \Cref{fig:region-model} give the models
-of $\term{+}$ and $\term{*}$. For $\term{+}$ this is the
- union of the models of subterms. The model of
-$\term{*}$ takes a ``tensor'' $\otimes$ of subterm models. The tensor is
-more involved. We give some informal intuition first.
-
-% specifications correspond to a set of schemes
-
-% r+s means an index must be generalised by a scheme in either r or s
-%       and every scheme in r must have an index which specialises it
-%       and every scheme in s must have an index which specialises it
-
-% r*s means an index must be generalised by a scheme in r and a scheme in s
-%       and every pairs of schemes in r and s must have an index which
-%           specialises both
-%
-
-% r+s means indices must be consistent with either r or s
-%           and there are indices why are generalised by r and s
-% (there are indices which are consistent with every part of r or s)
-% (every part of r and s independently have an index which is generalised by it)
-
-% r*s means indices must be consistent simultaneously with both r and s
-%           and there are indices which are generalised
-%
-
-\begin{figure}
-\vspace{-1em}
-\input{spatial-model.tex}
-\caption{%\textit{spec} model,
-$\interp{-}_n : \textit{spec} \rightarrow \textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$}
-\label{fig:spatial-model}
-\vspace{-1em}
-\end{figure}
-
-\paragraph{$\otimes$ of models}
-
-The intuition behind $\otimes$ is that it takes all possible pairs of
-index schemes, and treats each pair as lower and upper bounds of an
-$n$-dimensional rectangle. From these bounds,
-the remaining vertices of the $n$-dimensional rectangle are
-generated (like a bounding box) \eg{}:
-%%
-\begin{align*}
-{\small{
-\{\vtwohs{1}{2},\vtwohs{5}{6}\,\}\!\otimes\!\{\vtwohs{3}{4}\}\!=\!\{\vtwohs{1}{2},\vtwohs{3}{2},\vtwohs{1}{4},\vtwohs{3}{4},
-   \vtwohs{3}{6},\vtwohs{5}{4},\vtwohs{5}{6}\,\}
-}}
-\end{align*}
-%%
-The tensor takes a Cartesian product of the two models,
-giving all pairs of schemes which are combined using the
-\emph{pairwise permutation} written $\bowtie$ (defined later).
-Pairwise permutation generates all possible interleavings of two
-vectors. Some care is then taken over how this interacts with $\infty$.
-In the presence of $\infty$, generate all interleavings as if $\infty$ is
-an integer, then filter out vectors with $\infty$ at any coordinate, \eg{}
-\begin{align*}
-\{\vtwoh{1}{\infty}\} \otimes \{\vtwoh{3}{4}\} =
-\{\vtwoh{1}{4}, \vtwoh{3}{4}\}
-\end{align*}
-%
-where $\vtwoh{1}{\infty}$ and $\vtwoh{3}{\infty}$ have been filtered.
-
-\begin{definition}%[Constrained dimensions]
-For an $n$-dimensional model $M$, its \emph{constrained dimensions}
-are the set of dimension identifiers where at least one scheme $u \in M$ has
-a non-$\infty$ value in that dimension:
-%
-\begin{align*}
-\effdims{M}{n} =
-\bigcup_{u \in M} \{i \mid i \in \{1,\ldots,n\} \, \wedge \, u_i \neq
-  \infty\}
-\end{align*}
-For example, $\effdims{\{\vtwoh{0}{\infty},\vtwoh{\infty}{\infty}\}}{2}
- = \{1\}$.
-\end{definition}
-\begin{definition}The tensor $\otimes_n$ of $n$-dimensional
-models is:
-\begin{alignat*}{2}
-& M \otimes_n N =
-  \Bigg\{x \; \Bigg| \;
-    \parbox{5.2cm}{$(u, v) \in M \times N,$ \\[0.1em]
-                  $\;\; i \in \{1, \ldots, 2^n\}, \,
-                  x = (u \bowtie v)_i,$ \\[0.1em]
-                  $\;\; j \in (\effdims{M}{n} \cup
-                   \effdims{N}{n}),$\\[0.1em]
-                   $\;\; x_j \neq \infty$
-                  } \Bigg\}
-\end{alignat*}
-The second guard binds $j$ to the constrained
-dimensions for $M$ unioned with the constrained dimensions for $N$.
-That is, every $j$ corresponds to a dimension in which
-either $M$ or $N$ has a scheme $u$ where $u_j \neq \infty$,
-and thus is constrained.
-The \emph{pairwise permutation} $u \bowtie v$
-builds a $2^n \times n$ matrix
-of all possible $n$-vectors generated by
-non-deterministically picking for each $j^{th}$ entry either
-$u_j$ or $v_j$ \eg{}
-%
-\begin{equation*}
-\vthreeh{0}{1}{2} \bowtie \vthreeh{3}{4}{5} =
-\left[
-\setlength{\arraycolsep}{0.42em}
-\begin{array}{cccccccc}
-0 & 0 & 0 & 0 & 3 & 3 & 3 & 3 \\
-1 & 1 & 4 & 4 & 1 & 1 & 4 & 4 \\
-2 & 5 & 2 & 5 & 2 & 5 & 2 & 5
-\end{array}
-\right]
-\end{equation*}
-%
-The $2^n$ unique choices for $\bowtie$ on $n$-vectors
-correspond to taking all bit-strings of length $n$ and
-selecting from $u$ for $1$ and $v$ for $0$. The pairwise permutation
-matrix is thus defined in terms of the logical
-matrix $b$ where $b_{i,j}$ is the $j$-th bit of the integer
-$i - 1$ as:
-%
-\begin{align*}
-(u \bowtie v)_{i,j} & = b_{i,j} u_j + \neg b_{i,j} v_j
-\end{align*}
-%
-The permuted schemes $x = (u \bowtie v)_i$ added to the model
-$M \otimes N$ are only those without $\infty$ in
-the constrained dimensions given by each model ($x_j \neq
-\infty$). This eliminates all permutations with
-$\infty$ in a constrained dimension position.
-\end{definition}
-
-\paragraph{Model of spec}
-
-Regions are modelled as sets of index
-schemes, that is
-$\interp{\textit{region}}_n \in \mathcal{P}((\domainVal{})^n)$.
-\Cref{fig:spatial-model} defines the top-level model of the
-\textit{spec} syntax, which includes the multiplicity and
-approximation modifiers. The domain of this interpretation
-is $\textsf{Mult}(\textsf{Approx}(\mathcal{P}((\domainVal{})^n)))$
-where $\textsf{Mult}$ and $\textsf{Approx}$ are labelled variants
-labelling our set-based model with additional information.
-
-\begin{definition} \textsf{Mult}
-and \textsf{Approx} are parametric labelled variant types
-with injections given by their definition:
-%
-\begin{align*}
-\textsf{Mult} \, a & = \textsf{mult} \, a \,\mid\, \textsf{only} \, a \\
-\textsf{Approx} \, a & = \textsf{exact} \, a \,\mid\, \textsf{lower} \, a \,\mid\,
-\textsf{upper} \, a \,\mid\, \textsf{both} \, a \, a
-\end{align*}
-\eg{}, $\mathsf{lower}$ is an injection
-$\mathsf{lower} : a \rightarrow \mathsf{Approx} \, a$ etc. The
-$\textsf{Mult}$ type corresponds to the presence or absence of the
-\term{readOnce} modifier as shown in \Cref{fig:spatial-model}.
-The $\textsf{Approx}$ type corresponds to the presence or absence of
-the spatial approximation modifier, with \textsf{exact} when there is
-no such modifier and \textsf{lower} and \textsf{upper} for
-\term{atLeast} and \term{atMost}. Later, it will be useful to model
-pairs of lower and upper bounds provided by \textsf{both} but which is
-not used
-here.% These are used for labelled models, and other components later in.
-\label{def:mult-and-approx}
-\end{definition}
-
-\section{Equational theory}
-\label{sec:eqs}
-
-Specifications have a theory of equality, $\equiv$, providing axiomatic
-semantics for the language. This enables users to rewrite their specification
-and provides a way to automatically simplify
-specifications after inference (\Cref{subsec:inference}).
-Additionally, specifications have a theory of approximation,
-$<:$, that can be used by the users to weaken specifications for succinctness in
-the source.
-
-The equality relation $\equiv$ is defined over \emph{region}
-syntax which gives the following axioms:
-%
-\begin{itemize}
-\item $\term{+}$ is idempotent, commutative, and associative;
-\item $\term{*}$ is commutative, associative, and distributes over
-$\term{+}$, \ie{},  $R \; \texttt{*} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
-                       \texttt{*} \; S) \; \texttt{+} \; (R
-                       \; \texttt{*} \; T)$.
-Distributivity is
-used later to give a normal form for specifications
-akin to Disjunctive Normal Form (\cref{sec:inf-step4}).
-
-\item (\emph{overlapping}) Given two region constants $R_1$ and $R_2$
-on the same dimension, then some combinations of region constants
-can be coalesced into a single region constant $R_3$ if $R_1$ completely
-overlaps $R_2$. There are five possible instantiations of the
-following rule $R_1 + R_2 \equiv R_3$:
-%
-\newcommand{\stenGR}[4]{\small{\textit{#4}(\term{depth=}#1,
-  \term{dim=}#2, #3)}}
-\newcommand{\stenG}[3]{\small{\textit{#3}(\term{depth=}#1,
-  \term{dim=}#2)}}
-%
-\vspace{-0.2em}
-\begin{align*}
-\hspace{-1em} & \stenGR{n}{d}{p \vee q}{$R_3$} \\[-0.4em]
-\hspace{-1em} \equiv \; & \stenGR{n}{d}{p}{$R_1$} \; \texttt{+} \; \stenGR{m}{d}{q}{$R_2$}
-\end{align*}
-where $n \geq m$ and $p$ and $q$ range
-over booleans representing the absence (true) or presence (false)
-of \irrefl{} which are combined via disjunction.
-We denote the possibilities as $\term{fwd+fwd} \equiv \term{fwd}$, $\term{bwd+bwd} \equiv \term{bwd}$,
-$\term{cen+cen} \equiv \term{cen}$, $\term{cen+fwd} \equiv \term{cen}$ and
-$\term{cen+bwd} \equiv \term{cen}$. %Commutative of $\term{+}$
-%provides symmetric variants.
-
-\item (\textit{overlapping pointed}) \term{pointed} regions coalesce
-  with other regions of the same dimension:
-%
-\vspace{-0.4em}
-\begin{align*}
-& \stenG{n}{d}{region} \\[-0.4em]
-\equiv \; & \stenGR{n}{d}{p}{region} \; \texttt{+} \;
-                          \stenReflS{d}
-\end{align*}
-\vspace{-1.5em}
-
-\item \textit{(centered)} a forward and backward region of the same
-  depth and dimension coalesces into a centered region:
-%
-\vspace{-0.4em}
-\begin{align*}
-\hspace{-1em} & {\small{\stenCen{n}{d}{, p \vee q}}} \\[-0.4em]
-\hspace{-1em} \equiv \; & {\small{\stenFwdSR{n}{d}{p} \; \texttt{+} \;
-                          \stenBwdSR{n}{d}{q}}}
-\end{align*}
-\vspace{-1.5em}
-
-\item $(\textsc{point}) \;\; (R \, \term{*} \, S^\irreflS) \, \term{+} \,
-                         (R^\irreflS \, \term{*} \, S) \; \equiv \; R
-                         \, \term{*} \, S$
-meaning that for the sum of products of two regions $R$ and $S$,
-where in each component one of $R$ or $S$ has the non-pointed
-attribute denoted by $R^{\term{np}}$ and $S^{\term{np}}$,
-the non-pointed attributes cancel to give $R \, \term{*} \, S$.
-%
-\item $(\textsc{$\ast$Dim}) \;\; R^d \, \term{$\ast$} \, S^d \equiv R^d
-                             \, \term{+} \, S^d$
-%
-where $R^d$ and $S^d$ denote regions whose dimension is $d$.
-Thus a product of two regions on
-the same dimension is equal to their addition.
-\end{itemize}
-%
-Stencil specifications in \textit{spec} are considered
-equal when they have the same modifiers and $\equiv$ equates
-their regions.
-
-\begin{theorem}[Soundness of equational theory]
-\begin{equation*}
-\forall S, T, n . \quad
-S \equiv T \; \Rightarrow \;
-\interp{S}_n = \interp{T}_n
-\end{equation*}
-See the supplement for the proof.
-\end{theorem}
-
-\Cref{fig:inequations} defines a notion of sub-specifications
-via the relation $<:^r$ on \textit{region}
-syntax and then $<:$ on \textit{spec} syntax inductively.
-Both relations are reflexive and transitive (we omit
-these equations). The $<:^r$ relation is congruent
-with respect to \term{+} and \term{*} and $<:$ is congruent
-with respect to \term{readOnce}, the equations of which we omit for brevity.
-
-The \trule{\textsc{eq}} rule connects region
-equations to approximation; \trule{\textsc{over}} explains
-the notion of spatial over-approximation via overlapping regions;
-\trule{\textsc{rep}} allows the \term{readOnly} modifier that
-to be dropped. The \trule{\textsc{shrink}}
-and \trule{\textsc{grow}} rules turn the spatial approximation on
-regions into approximation on specifications via the \term{atLeast} and \term{atMost}
-modifiers. Finally, \trule{\textsc{Gen$\ast$}} shows
-that specifications can be generalised by removing parts of a product
-$\ast$ which have disjoint dimensions to the rest of the region
-product (where $\mathsf{dims}$ denotes the set of dimensions specified
-in a region).
-
-\begin{theorem}[Approximation soundness]
-Code consistent with a specification $S$
- is consistent with a more general specification $T$,
-\ie{}, for all models $M$ of a piece of code:
-\begin{equation*}
-S <: T \; \wedge (\consA{M}{\interp{S}_n}) \;
-\Rightarrow \; (\consA{M}{\interp{T}_n})
-\end{equation*}
-where $\consAName$ implements consistency in checking (\cref{subsec:checking}).
-See the supplement for the proof.
-\end{theorem}
-
-\begin{figure}[t]
-\vspace{-0.5em}
-\begin{align*}
-\hspace{-0.5em}
-{\small{
-\begin{array}{c}
-%%
-%% INEQUATIONS ON REGION
-%%
-%\framebox{$\textit{region} <: \textit{region}'$} \\[0.85em]
-%
-%\dfrac{}{R <:^r R}\trule{\textsc{refl}} \qquad \dfrac{P <:^r R \quad R <:^r Q}{P <:^r
-%  Q}\trule{\textsc{trans}}
-%\qquad
-%\\[1.5em]
-\vspace{-0.4em}
-\setlength{\arraycolsep}{0.1em}
-\dfrac{\hspace{3em} m \leq n \qquad\qquad r \implies s \hspace{3em}}
-{\begin{array}{rl}
-\stenFwdSR{m}{d}{r} & <:^r \stenFwdSR{n}{d}{s}
-%& \;\, ds \subseteq es \wedge n \leq m
-\\
-\wedge \; \stenBwdSR{m}{d}{r} & <:^r \stenBwdSR{n}{d}{s}
-%& \;\, ds \subseteq es \wedge n \leq m
-\\
-\wedge \; \stenCenSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s}
-%& \;\, ds \subseteq es \wedge n \leq m
-\\
-\wedge \; \stenBwdSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s} \\
-\wedge \; \stenFwdSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s} \\[1.2em]
-\end{array}}\trule{\textsc{over}} \\[-0.2em]
-%\hspace{-0.5em}
-%\dfrac{R_1 <:^r S_1 \quad R_2 <:^r S_2}
-%      {R_1 \, \texttt{+} \, R_2 <:^r S_1 \, \texttt{+} \, S_2}
-%\trule{\textsc{cong}\texttt{+}}
-%\dfrac{R_1 <:^r S_1 \quad R_2 <:^r S_2}
-%      {R_1 \, \texttt{*} \, R_2 <:^r S_1 \, \texttt{*} \, S_2}
-%\trule{\textsc{cong}\texttt{*}} \\[1.5em]
-%%
-%% INEQUATIONS ON SPEC
-%%
-%\framebox{$\textit{spec} <: \textit{spec}'$} \\[0.8em]
-%
-%
-\dfrac{R <:^r S}{\term{atLeast}  S <: \term{atLeast}
-  R}\trule{\texttt{L}}
-\;\;
-\dfrac{R <:^r S}{\term{atMost}  R <: \term{atMost}
-  S}\trule{\texttt{M}} \\[1.5em]
-\dfrac{R <:^r S}{R <: \term{atMost} \, S}\trule{\textsc{grow}} \,
-\dfrac{R <:^r S}{S <: \term{atLeast} \, R}\trule{\textsc{shrink}} \,
-\dfrac{R \equiv S}{R <:^r S}\trule{\textsc{eq}}
- \\[1.5em]
-%%
-\dfrac{\mathsf{dims}(R) \cap \mathsf{dims}(S) = \emptyset}{R \ast S <: R}
-\trule{\textsc{Gen$\ast$}}
-\quad
-%%
-%\\[1.5em]
-%\dfrac{R <: S}{\term{readOnce} \, R <: \term{readOnce} \,
-%  S}\trule{\textsc{cong}\texttt{R}}
-%\;\;
-\;
-\dfrac{}{\term{readOnce} \, R <: R}\trule{\textsc{rep}}
-%\dfrac{}{R <: \term{atMost} R}
-%\quad
-%\dfrac{}{R <: \term{atLeast} R}\\[1.5em]
-\end{array}}}
-\end{align*}
-\vspace{-0.5em}
-\caption{Approximation $<:^r$ on \textit{regions} and
-$<:$ on \textit{specs} (omitting reflexivity, transitivity, and congruences). }
-\label{fig:inequations}
-\vspace{-0.5em}
-\end{figure}
-
-\section{Analysis, checking, and inference}
-\label{sec:analysis}
-
-\noindent
-We outline here the procedures for checking conformance
-of source code against specifications (\Cref{subsec:checking})
-and for inferring specifications from code (\Cref{subsec:inference}).
-Both rely on a program analysis that converts array subscripts
- into sets of index schemes. We outline this analysis
-first (\cref{subsec:analysis}). Note that the analysis 
-can be made arbitrarily more complex and wide-ranging independent
-of the checking and inference procedures. At the moment, the analysis
-is largely \emph{syntactic}, with only a small amount of
-semantic interpretation of the code.
-
-\subsection{Analysis of array accesses}
-\label{subsec:analysis}
-
-\newcommand{\neigh}{\textsf{neigh}}
-
-The analysis builds on standard program analyses:
-%
-\textbf{(1)} basic blocks (CFG);
-\textbf{(2)} induction variables per basic block;
-\textbf{(3)} (interprocedural) data-flow analysis, providing a \emph{flows to}
-  graph (as shorthand, the function
-  $\mathsf{flowTo}$ is used, implicitly parameterised by this graph,
-  mapping an expression to the set of all expressions
-  with forwards data-flow to this expression, based on the transitive
-  closure of the flows graph);
-\textbf{(4)} type information per variable.% where we use the predicate
-%\arrayTy{} to classify variables of array type.
-
-The core analysis, captured by the function $\textit{analyse}$ below, then proceeds as follows.
-For each assignment statement \textit{assg} within a loop whose left-hand side is an array
-subscript on neighbourhood indices, a finite map is computed mapping
-array variables to \emph{multisets} of indexing schemes:
-%
-\begin{align*}
-\hspace{-0.5em}
-\begin{array}{lr}
-\textit{analyse}(a(\overline{e_1}) = e_2)
- := \qquad\qquad\qquad & \textit{where} \; \neigh(\overline{e_1}) \\[0.3em]
-\multicolumn{2}{l}{\qquad \bigcup\{a' \mapsto \{
-\textsf{schematic}(\bar{e})\} \mid a'(\bar{e}) \leftarrow \mathsf{flowsTo}(e_2)\}}
-\end{array}
-\end{align*}
-%
-The predicate $\neigh(\bar{e_1})$ classifies indices where each
-$e \in \bar{e_1}$ is either a
-neighbourhood index (based on
-the induction variables of the loop, implicit here)
-or a constant index.\footnote{If all indices $e \in \bar{e}$ are constants, then
-  $\neigh(\bar{e})$ does not hold. Instead $a(\bar{e}$) is
-  considered a constant by our analysis, rather
-than an array subscript.}% contributing to a stencil.}
-Thus, we focus on assignments to an array subscript where the LHS
-indices $\overline{e_1}$ are neighbourhood indices.  For
-all array subscripts that flow to the right-hand side of this
-statement, a finite map is constructed, mapping each array variable
-to a multiset of indexing schemes for its subscripts, computed
-with \textsf{schematic} (\Cref{def:schematic}).
-Note \textsf{schematic} is undefined ($\bot$) if
-$\bar{e}$ contains relative indices which are not neighbourhood
-indices. The \textit{analysis} function is used by the
-\textit{modelise} function which filters out those arrays
-which contain undefined indexing schemes and adds multiplicity
-information:
-%
-\begin{align*}
-& \mathit{modelise} (\textit{assg})
-= \textit{let} \;\; M = \mathit{analyse}(\textit{assg}) (a) \;\; \textit{in}
-\\
-& a \mapsto
-\begin{cases}
-\bot & \bot \in M \\
-\textsf{only}(|M|) & \forall u \in M . \, \#(M, u) = 1 \\
-\textsf{mult}(|M|) & \forall u \in M . \, \#(M, u) = n
-\end{cases}
-\end{align*}
-where $\#(M, u)$ denotes the multiplicity of an element $u$ of the
-multiset $M$ and $|M|$ denotes the set of elements in the multiset
-$M$.
-Thus, the analysis maps source code to index schemes, which is then
-augmented with multiplicity information by $\mathit{modelise}$,
-\ie{} values in $\mathsf{Mult}(\mathcal{P}((\domainVal{})^n))$.
-
-As a small semantic concession to an otherwise syntactic
-analysis: if the LHS $\overline{e_1}$ contains non-$0$ neighbour offsets
-then all RHS schemes are \emph{relativised} by
-$\textsf{schematic}(\overline{e_1})$. For example,
-if \texttt{a(i+1) = b(i)} then relativisation of the RHS by the LHS
-produces an analysis which is equivalent to the analysis of \texttt{a(i) = b(i-1)}.
-
-\begin{example}
-\label{exm:checking}
-We demonstrate the analysis and modelise procedures over the
-following example kernel which computes the mean value
-of a five-point stencil:
-%%
-\begin{minted}{fortran}
-x = a(i-1, j)+a(i+1, j); y = a(i, j-1)+a(i, j+1)
-b(i, j) = (a(i, j) + x + y) / 5.0
-\end{minted}
-%
-%The \textit{analyse} function matches on
-%line 4 with the following set for $\textsf{flowsTo}(\texttt{a(i, j) + x +
-%  y}) = $
-%
-%\begin{align*}
-%{{ %{\small{
-%$\{$\texttt{a(i-1, j)}, \texttt{a(i+1, j)}, \texttt{a(i, j-1)},
-%  \texttt{a(i, j+1)}, \texttt{a(i, j)}$\}$}}. 
-%\end{align*}
-The  multiplicities are all one, and so \emph{modelise} yields:
-%%
-\begin{align*}
-\texttt{a} \mapsto \mathsf{only}(\{\vtwoh{-1}{0}, \vtwoh{1}{0},
-          \vtwoh{0}{-1}, \vtwoh{0}{1}, \vtwoh{0}{0}\})
-\end{align*}
-%%
-\end{example}
-
-%Given an assignment $\textit{assg}$ to an array subscript in the source language
-%and an array variable $a$, then $\mathit{modelise}(\textit{assg}, a)$ performs
-%an analysis to find all array subscripts of $a$ converting them to
-%index schemes. For the moment, we leave this analysis abstract
-%as the function $\emph{analyse}$ in the following definition of $\emph{modelise}$:
-%
-
-\subsection{Specification checking}
-\label{subsec:checking}
-
-Checking verifies the access pattern of a stencil computation in the
-source language against its associated specifications. The model of
-(\Cref{sec:semantics}) guides this process.  Checking
-proceeds by generating a model from a specification and generating
-a model from the source code, comparing them for
-\emph{consistency} base on index scheme
-\emph{generalisation} (\Cref{def:generalisation}).
-
-\begin{definition}[Checking]
-Let $a$ be an array variable and
-$S$ a specification associated to $a$
-via \texttt{!= stencil $S$ :: a}.
-
-A stencil kernel
- $v(\overline{e_1}) = e_2$
-conforms to $S$ if the predicate $\textbf{check}_n(S, v(\overline{e_1}) = e_2, a)$ holds,
-defined:
-\begin{align*}
-& \textbf{check}_n(S, \textit{assg}, a) =
-  \consSub{n}{\textit{modelise}(\textit{assg})(a)}{\interp{S}_n}
-&
-\end{align*}
-Thus, checking applies \textit{modelise}
-(\Cref{subsec:analysis}) to extract index schemes from source code,
-augmented with multiplicity for the whole set of schemes.
-This is compared with the model of the specification $\interp{S}_n$ by
-the consistency predicate $\consAName$.
-\end{definition}
-
-\begin{figure}[t]
-\vspace{-0.5em}
-\begin{align*}
-\hspace{-0.5em}
-\begin{array}{c}
-%%
-\trule{up}\dfrac{\forall u \in M, \exists v \in N .  \, u \preceq v}
-      {\cons{M}{\textsf{upper}(N)}}
-\,
-\trule{low}\dfrac{\forall v \in N, \exists u \in M . \, u \preceq v}
-      {\cons{M}{\textsf{lower}(N)}} \\[1.2em]
-%%
-\trule{exact}\dfrac{\cons{M}{\textsf{lower}(N)} \qquad \cons{M}{\textsf{upper}(N)}}
-  {\cons{M}{\textsf{exact}(N)}} \\[1.2em]
-%%
-\trule{both}\dfrac{\cons{M}{\textsf{lower}(M_1)} \qquad \cons{M}{\textsf{upper}(M_2)}}
-  {\cons{M}{\textsf{both}(M_1, M_2)}}\\[1.2em]
-%%
-\trule{$1$}
-\dfrac{\cons{M}{N}}{\consA{\textsf{once}(M)}{\textsf{once}(N)}}
-\;\;
-\trule{$*$}
-\dfrac{\cons{M}{N}}{\consA{\textsf{mult}(M)}{\textsf{mult}(N)}}
-  \\[1.2em]
-\trule{$1 < *$}
-\dfrac{\cons{M}{N}}{\consA{\textsf{once}(M)}{\textsf{mult}(N)}}
-\end{array}
-\end{align*}
-\vspace{-0.5em}
-\caption{Consistency of models (code vs. specification)}
-\label{fig:consistency}
-\vspace{-0.9em}
-\end{figure}
-
-\begin{definition}
-Consistency relates models of source code to models
-of specifications (which are bounded) via predicates:
-\begin{align*}
-\consName_n & \subseteq \mathcal{P}(\domainVal{}^n)
-\times \textsf{Approx}(\mathcal{P}(\domainVal{}^n)) \\
-\consAName_n & \subseteq \textsf{Mult}(\mathcal{P}(\domainVal{}^n))
-\times \textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))
-\end{align*}
-%
-where $\consName$ relates models without multiplicity.
-\Cref{fig:consistency} defines the two predicates, where for
-brevity we make implicit the subscript $n$ defining the
-dimensionality.  We denote sets of indexing schemes
-$\mathcal{P}((\domainVal{})^n)$
-from source code model by $M$ and from a specification model by $M$.
-
-Rules (\emph{up}) and (\emph{low}) are the core; (\emph{up})
-considers the case where the specification is
-an \term{atMost} upper-bound approximation.
-An upper-bound specification means that for every scheme in the code
-model, modelling a source-level index, $\forall u \in M$ there exists
-a generalising scheme in the spec model $\exists v \in N$, \ie{}, $u
-\preceq v$. That is, every source-level index scheme is generalised by the specification.
-The (\emph{down}) rule provides the case of
-\term{atLeast} lower-bound specifications. Consistency is dual to the above:
-for every scheme in the specification model $\forall v \in N$ there
-exists a scheme in the code model $\exists u \in M$ such that $u
-\preceq v$. That is, every specification scheme is generalised by some source
-index scheme.
-
-The (\emph{up}) and (\emph{down}) rules are combined for an \textsf{exact}
-specification (\textit{exact}) such that a specification's model
-provides both a consistent lower and consistent upper bound on
-the source code. Thus, for every scheme in the code model there
-is a scheme in the spec model that generalises it, and for every
-scheme in the specification there is a scheme in the source model that
-specialises it.  The (\textit{both}) rule is similar, decomposing
-a specification with both a lower and upper bound into
-consistency of these bounds.
-
-Rules $(1)$ and $(*)$ extend the \consName{}
-predicate to $\consAName{}$ with
-multiplicity information. Both rules
-state that consistent models remain consistent if they
-have the same multiplicity. Relatedly, rule $(1 < *)$ ensures
-that absence of a \texttt{readOnce} modifier in the specification
-does not prohibit a set of unique stencil accesses in the
-code from being consistent.
-\end{definition}
-
-\begin{example}
-  Consider the following five-point stencil kernel with a (quite general) specification:
-%
-\begin{minted}[breakindent=8.5em]{fortran}
-!= region :: fivepoint = centered(depth=1,dim=1) + centered(depth=1,dim=2)
-!= stencil readOnce, fivepoint :: b
-a(i,j) = b(i,j-1)+b(i,j)+b(i,j+1)+b(i-1,j)+b(i+1,j)
-\end{minted}
-%
-We refer to line three as $l3$ for brevity,
-which is checked via
-$check(({\textcolor{darkgreen}{\texttt{readOnce, fivepoint}}}),l3,\texttt{b})$
-with steps:
-\begin{enumerate}
-%
-\item Generate the model from the specification as in
-  \cref{sec:semantics}.
-%
-\vspace{-0.4em}
-\begin{align*}
- N_s= \{  \vtwo{-1}{\infty},
-          \vtwo{0}{\infty},
-          \vtwo{1}{\infty},
-          \vtwo{\infty}{-1},
-          \vtwo{\infty}{0},
-        \vtwo{\infty}{1} \}  \\
-\interp{{\textcolor{darkgreen}{\texttt{readOnce, fivepoint}}}}_2 =
-  \textsf{once}(\textsf{exact} (N_s))
-\end{align*}
-%\vspace{-1.6em}
-
-\item Generate code model $\textit{modelise}(l3, \texttt{b}) =
-  \textsf{once} (M_c)$ where:
-%
-\vspace{-0.9em}
-\begin{align*}
-M_c = & \{
-    \vtwo{0}{-1},
-    \vtwo{0}{0},
-    \vtwo{0}{1},
-    \vtwo{-1}{0},
-    \vtwo{1}{0}
-  \}
-\end{align*}
-\vspace{-1.1em}
-
-\item By the definition of consistency (\Cref{fig:consistency})
- $\textsf{once}(\textsf{exact}(N_s))$ is consistent with
-  $\textsf{once}(M_c)$ when:
-\vspace{-0.6em}
-  \begin{align*}
-    & \consSub{2}{\textsf{up}(N_s)}{M_c} \;\,
-    \wedge \;\, \consSub{2}{\textsf{low}(N_s)}{M_c}
-  \end{align*}
-%\vspace{-1.6em}
-
-\item This holds by the following sets of scheme
-  generalisations: for the upper bound, $\forall u \in M_c,
-  \exists v \in N_s . u \preceq_2 v$:
-  \begin{align}
-    \hspace{-1em}
-    \begin{array}{l}
-   \hspace{-1.5em}
-     \vtwoh{0}{-1} \preceq_2 \vtwoh{0}{\infty} \quad
-     \vtwoh{0}{0} \preceq_2 \vtwoh{0}{\infty} \quad
-     \vtwoh{0}{1} \preceq_2 \vtwoh{0}{\infty} \\[0.5em]
-        \hspace{-1.5em}\vtwoh{-1}{0} \preceq_2 \vtwoh{\infty}{0} \quad
-     \vtwoh{1}{0} \preceq_2 \vtwoh{\infty}{0}
-     \end{array}
-  \label{eq:check-example-upper}
-  \end{align}
-  %
-  and for the lower bound: $\forall v \in N_s, \exists u \in M_c . u
-  \preceq_n v$:
-  \begin{align}
-    \hspace{-1em}
-    \begin{array}{l}
-     \hspace{-1.3em}\vtwoh{-1}{0} \preceq_2 \vtwoh{-1}{\infty} \quad
-     \vtwoh{0}{0} \preceq_2 \vtwoh{0}{\infty} \quad
-     \vtwoh{1}{0} \preceq_2 \vtwoh{1}{\infty}  \\
-     \hspace{-1.3em}\vtwoh{0}{-1}  \preceq_2 \vtwoh{\infty}{-1} \quad
-     \vtwoh{0}{0}  \preceq_2 \vtwoh{\infty}{0} \quad
-     \vtwoh{0}{1}  \preceq_2 \vtwoh{\infty}{1}
-     \end{array}
-    \label{eq:check-example-lower}
-  \end{align}
-\end{enumerate}
-Since the consistency predicate is satisfied, the array
-  accesses flowing to line three in the stencil computation are valid
-  with respect to the {\textcolor{darkgreen}{\texttt{fivepoint}}} specification.
-\end{example}
-\begin{example}
-Consider an erroneous user modification to $l3$ with an
-additional term \texttt{b(i+2,j+2)}, call it $l3'$.% such that $l3'$ is:
-%
-%\begin{minted}[firstnumber=3]{fortran}
-%a(i,j) = b(i,j-1) + b(i,j) + b(i,j+1) + b(i-1,j) + b(i+1,j) + b(i+2,j+2)
-%\end{minted}
-
-Checking proceeds as in the previous example, except
-$\mathit{modelise}(l3',\texttt{b})$ generates an additional scheme
-$\vtwoh{2}{2}$. The generalisations for
-the lower bound \eqref{eq:check-example-lower}
-of step 4 above still hold since its consistency condition
-universally quantifies over the \emph{specification model}. However, for
-the upper bound, the consistency condition universally quantifies
-over schemes $u$ of the \emph{code model} for which there must exist
-a scheme $v \in N_s$ such that $u \preceq_2 v$. Thus, we require a $v
-\in N_s$ such that $\vtwoh{2}{2} \preceq_2 v$ but no such generalising
-scheme exists in the spec model. Hence checking fails.
-\end{example}
-
-\subsection{Specification inference}
-\label{subsec:inference}
-
-To aid adoption of our approach to legacy code, we provide
-an inference procedure for generating specifications from code, which
-are inserted automatically as comments.
-We illustrate inference using the
- five-point stencil of
- \Cref{exm:checking}.
-
-\subsubsection{Step 1: Data access analysis}
-\label{sec:inf-step1}
-
-The first step performs the
-analysis of \cref{subsec:analysis} to produce a
-model of the code in terms of indexing schemes.
-The \emph{analyse} and \emph{modelise} procedures before yielded the finite map:
-%
-\begin{align*}
-\texttt{a} \mapsto \mathsf{only}(\{\vtwoh{-1}{0}, \vtwoh{1}{0},
-          \vtwoh{0}{-1}, \vtwoh{0}{1}, \vtwoh{0}{0}\})
-\end{align*}
-%
-
-\subsubsection{Step 2: Coalesce contiguous schemes into spans}
-\label{sec:inf-step3}
-
-Let $U$ range over the finite maps computed by \textit{modelise}.  For
-each array variable $a \in \mathsf{dom}(U)$, the algorithm constructs a set of
-$n$-dimensional rectangles covering all contiguous groups of schemes
-in $U(a)$.
-
-\newcommand{\spanOp}{\textsf{spans}}
-
-\begin{definition}[Spans]
-  A \emph{span} represents an $n$-dimensional rectangle
-  as a pair of $n$-dimensional vectors (or
-  represented as a $2 \times n$ matrix) with values drawn from
-  $\domainVal{}$. This gives the co-ordinates of lower-bound 
-  and upper-bound vertices. We write
-  spans as $\vect{x} = [\vect{x}^L \; \vect{x}^U]$, where $\vect{x}^L$
-  and $\vect{x}^U$ give the lower and upper bounds.  In
-  the following, $\vect{x}, \vect{y}$ range over spans.  For a span
-  $\vect{x}$, we write $\vect{x}_{i:j}$ for the sub-span comprising the
-  subvectors $[\vect{x}^L_{i:j} \; \vect{x}^U_{i:j}]$ from entry $i$
-  to entry $j$ (inclusive).
-\end{definition}
-
-\noindent
-Each indexing scheme $u \in U(a)$ is converted to a \emph{unit} span
-$u \mapsto [u \; u]$, defining the lower and upper bound of an
-$n$-dimensional cube of side length $1$. For the five-point stencil
-example there are five unit spans, illustrated in
-\Cref{fig:inference-steps-informal}(a).
-
-The next step repeatedly coalesces any contiguous spans until a fixed
-point is reached, resulting in a set of (possibly overlapping) spans
-covering the original space of schemes. Two $n$-dimensional spans
-$\vect{x}$ and $\vect{y}$ are contiguous if for all but one
-$j \in \{1, \ldots, n\}$ then $\vect{x}_j = \vect{y}_j$ and for one
-$j$ then $\vect{x}^U_1 + 1 = \vect{y}^L$. For example, spans
-$\left[\vtwoh{-1}{0}\; \vtwoh{-1}{0}\right]$ and
-$[\vtwoh{0}{0} \; \vtwoh{1}{0}]$ are contiguous in the first dimension
-and can be coalesced into the span $[\vtwoh{-1}{0} \; \vtwoh{1}{0}]$.
-
-\begin{definition}\label{def:span-coalesc}
-  Span coalescing $\bullet$ is partial, defined for spans contiguous in the
-  first dimension:
-%%
-\begin{align*}
-\vect{x} \bullet \vect{y}
-= (\vect{x}^L, \vect{y}^U) \;\;\;\, \textit{if} \;\;\; \vect{x}^U_1 + 1 = \vect{y}^L_1 \; \wedge \;
-\vect{x}_{2:n} = \vect{y}_{2:n} %\\
-%\bot  & \textit{otherwise}
-%\end{cases}
-% $m$-dimensions into coalesced spans which vary in
-% $m+1$-dimensions.
-\end{align*}
-%%
-\end{definition}
-\noindent
-To coalesce regions which are contiguous in other dimensions
-other than the first, we generate all rotations of the spans in
-$n$-dimensional space by permutation on spans, and fold together contiguous
-regions using $\bullet$ (above). This
-generates all possible ways of coalescing regions. For our example,
-the unit spans in \Cref{fig:inference-steps-informal}(a)
-are coalesced into the contiguous spans in
-\Cref{fig:inference-steps-informal}(b).
-This procedure is iterated until a fixed point is reached (in one step
-in this example).
- Then any spans that are completely contained by any other span are deleted,
-leaving a minimal set of spans (which may overlap, but none of which
-fully contains another), \Cref{fig:inference-steps-informal}(c).
-The final spans are used to synthesise concrete stencil specification
-syntax (\cref{sec:inf-step4}), where
-each span becomes a product \texttt{*} of regions which are
-combined by the sum \texttt{+}.
-
-More formally the algorithm proceeds as follows.
-For all $a \in \mathsf{dom}(U)$, let $M = U(a)$. Every scheme in $M$
-is mapped to a unit span
-by $\textsf{unitSpans}(M) = \{[u \; u] \mid u\!\in\!M\}$, \eg{}:
-%%
-\begin{align*}
-  N(U(\texttt{a})) = \stwo{0}{0}{0}{0} \stwo{1}{0}{1}{0} \stwo{-1}{0}{-1}{0} \stwo{0}{1}{0}{1} \stwo{0}{-1}{0}{-1}
-\end{align*}
-%%
-In the following we let $M_0 = N(U(\texttt{a}))$ for our example.
-
-Each set of unit spans is applied to the fixed point of the \spanOp{} function
-(defined shortly), which coalesces spans into contiguous regions. That is, we compute
- $(\mu \, \spanOp)$ on input $N(M)$, where \spanOp{} is defined by the steps:
-
-\begin{figure}[t]
-\vspace{-1em}
-\hspace{-0.5em}
-{\scalebox{0.96}{
-\begin{minipage}{0.16\linewidth}
-\begin{align*}
-\setlength{\arraycolsep}{0em}
-\begin{array}{c}
-\quad \\[0em]
-\begin{tikzpicture}[scale=0.8]
-%\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
-\node at (1.25,1.22) {\textnormal{\footnotesize{0,0}}};
-\draw[thick] (1,1) rectangle (1.5,1.5);
-\draw[thick] (1.5,1) rectangle (2,1.5);
-\draw[thick] (1,0.5) rectangle (1.5,1);
-\draw[thick] (0.5,1) rectangle (1,1.5);
-\draw[thick] (1,1.5) rectangle (1.5,2);
-\end{tikzpicture} \\[0em]
-(a)
-\end{array}
-\end{align*}
-\end{minipage} $\rightarrow$ \begin{minipage}{0.43\linewidth}
-\begin{align*}
-\setlength{\arraycolsep}{0em}
-\begin{array}{c}
-\begin{tikzpicture}[scale=0.72]
-%\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
-%\node at (1.25,1.22) {\textnormal{\small{0,0}}};
-\draw [black, thin] (0.35,0) to [round left paren ] (0.35,2);
-\draw[thick] (0.5,1) rectangle (2,1.5);
-\draw[thick] (1,0.40) rectangle (1.5,0.90);
-\draw[thick] (1,1.60) rectangle (1.5,2.10);
-\node at (2.3,1.22) {+};
-\draw[thick] (3.8,1) rectangle (4.3,1.5);
-\draw[thick] (2.6,1) rectangle (3.1,1.5);
-\draw[thick] (3.2,0.5) rectangle (3.7,2);
-\draw [black, thin] (4.4,0) to [round right paren ] (4.4,2);
-\end{tikzpicture} \\[-0.5em] (b)
-\end{array}
-\end{align*}
-\end{minipage}$\rightarrow$ \begin{minipage}{0.25\linewidth}
-\begin{align*}
-\setlength{\arraycolsep}{0em}
-\begin{array}{c}
-\begin{tikzpicture}[scale=0.72]
-%\node at (1.25,1.22) {\textnormal{\small{0,0}}};
-\draw [black, thin] (0.8,0) to [round left paren ] (0.8,2);
-\draw[thick] (1,0.5) rectangle (1.5,2);
-\node at (1.8,1.22) {+};
-\draw[thick] (2.1,1) rectangle (3.6,1.5);
-\draw [black, thin] (3.7,0) to [round right paren ] (3.7,2);
-%\node at (2.75,1.22) {\textnormal{\small{0,0}}};
-\end{tikzpicture} \\[-0.5em] (c)
-\end{array}
-\end{align*}
-\end{minipage}}}
-\caption{Illustration of grouping spans in
-inference}
-\label{fig:inference-steps-informal}
-\vspace{-0.8em}
-\end{figure}
-
-\noindent
-\textbf{1)\,} Compute all permutations on the column vectors in a span, \ie{},
-  $[\vect{x}^L \vect{x}^U] \mapsto [\pi\vect{x}^L \, \pi\vect{x}^U]$
-for a permutation $\pi$. For each permutation function $\pi^n_i$
-(the $i$-th permutation for vectors of size $n$) we pair the
-permutation function with the set of permuted spans so that
-the spans can be un-permuted later.
-%
-\begin{equation*}
-P(M) = \bigcup_{i \in n!} (\pi^n_{i} , \; \{[\pi^n_i
-\vect{x}^L \; \pi^n_i\vect{x}^U] \, \mid \, [\vect{x}^L \; \vect{x}^U]
-\leftarrow M\}
-\end{equation*}
-%
-This corresponds to all rotations of the space, \eg{}:
-%
-\begin{align*}
-P(M_0) =
-\{&(\pi^2_1, \{ \stwo{0}{0}{0}{0},
-\stwo{1}{0}{1}{0},
-\stwo{-1}{0}{-1}{0},
-\stwo{0}{1}{0}{1},
-\stwo{0}{-1}{0}{-1} \})
-\\
-&(\pi^2_2, \{
- \stwo{0}{0}{0}{0},
- \stwo{0}{1}{0}{1},
- \stwo{0}{-1}{0}{-1},
- \stwo{1}{0}{1}{0},
- \stwo{-1}{0}{-1}{0}\})\}
-\end{align*}
-%
-where $\pi^2_1$ is the identity permutation and $\pi^2_2$ is the
-permutation that flips the order of the two elements in the
-vectors.
-
-\noindent
-\textbf{2)\,} Sort each permutation set into a list, by the ordering:
-\begin{equation*}
-  \vect{x} \leq \vect{y} =
-      \exists i \, . \; \vect{x}^L_{i} \leq \vect{y}^L_{i} \; \wedge \;
-        (i = n \vee \vect{x}_{i+1:n} = \vect{y}_{i+1:n})
-\end{equation*}
-%
-giving the sorting function $\textsf{sort}(M)$, \eg{}, $\textsf{sort}(P(M_0))$:
-%%
-\begin{align*}
-\{&(\pi^2_1, [
-\stwo{0}{-1}{0}{-1},
-\stwo{-1}{0}{-1}{0},
-\stwo{0}{0}{0}{0},
-\stwo{1}{0}{1}{0},
-\stwo{0}{1}{0}{1}] )
-\\
-&(\pi^2_2, [
-\stwo{0}{-1}{0}{-1},
-\stwo{-1}{0}{-1}{0},
-\stwo{0}{0}{0}{0},
-\stwo{1}{0}{1}{0},
-\stwo{0}{1}{0}{1}])\}
-\end{align*}
-\noindent
-\textbf{3)\,} Reduce each list pairwise by $\bullet$
-(\Cref{def:span-coalesc}) to coalesce contiguous regions,
-given by $\textsf{fold}\bullet$. For our example:
-%
-\begin{align*}
-\{&(\pi^2_1, [
-\stwo{0}{-1}{0}{-1},
-\stwo{-1}{0}{1}{0},
-\stwo{0}{1}{0}{1}
-]), (\pi^2_2, \textit{same as for $\pi_1$})\}
-\end{align*}
-%
-\textbf{4)\,} Un-permute and union together, \ie{},
-%
-\[
-U(M) = \bigcup \{[\pi \vect{x}^L, \pi \vect{x}^U]
- \mid [\vect{x}^L, \vect{x}^U] \leftarrow S, (\pi, S) \leftarrow M\}
-\]
-For our example:
-%
-\begin{align*}
-U(M) =
-\{\stwo{0}{-1}{0}{-1},
-\stwo{-1}{0}{1}{0},
-\stwo{0}{1}{0}{1},
-\stwo{-1}{0}{-1}{0},
-\stwo{0}{-1}{0}{1},
-\stwo{1}{0}{1}{0}\}
-\end{align*}
-%
-\textbf{5)\,} Filter by the region containment predicate
-$\containedin$; if any region is contained within another then remove the
-  smaller:
-%
-\begin{align*}
-& \vect{x} \containedin \vect{y} = \vect{y}^L_1 \leq \vect{x}^L_1 \wedge \vect{x}^U_1 \leq \vect{y}^U_1
-  \wedge (\vect{x}^L_{2:n}, \vect{x}^U_{2:n}) \containedin
-  (\vect{y}^L_{2:n}, \vect{y}^U_{2:n})
-\end{align*}
-For our example this then yields the final result of:
-\begin{align*}
-\spanOp(M)
-= \{\stwo{-1}{0}{1}{0} \stwo{0}{-1}{0}{1}\}
-\end{align*}
-since $\stwo{0}{1}{0}{1} \sqsubseteq \stwo{0}{-1}{0}{1}$
-and $\stwo{1}{0}{1}{0} \sqsubseteq \stwo{-1}{0}{1}{0}$.
-Applying \spanOp{} again yields the same
-result for our example.%; the fixed point is reached.
-
-\subsubsection{Step 4: Synthesise specifications from spans}
-\label{sec:inf-step4}
-
-\newcommand{\oplusbig}{\operatornamewithlimits{\term{\Large{$\oplus$}}}}
-\newcommand{\odobig}{\operatornamewithlimits{\term{\Large{$\odot$}}}}
-\newcommand{\bplus}{\operatornamewithlimits{\term{\Large{+}}}}
-\newcommand{\tySum}[1]{#1^{\term{+}}}
-\newcommand{\tyProd}[1]{#1^{\term{*}}}
-\newcommand{\specDNF}{\textbf{spec}}
-
-Abstract syntax trees of the specification language
-are synthesised from the set $\spanOp(M)$ of coverings spans.
-This conversion is defined in terms
-of an algebra on \textit{spec} syntax in disjunctive-normal form
-(which we denote \specDNF{}) and which is wrapped
-in the $\textsf{Approx}$ type (see
-\Cref{def:mult-and-approx}).  The algebra mirrors the shape
-of the region operators $\term{+}$ and $\term{*}$ 
-with binary operators $\oplus$ and $\odot$ on
-$\mathsf{Approx}(\specDNF)$.
-A function $\mathsf{toSpec}$ maps a dimension identifier
-and a pair of lower and upper bound
-values (one-dimensional) to a specification with a one-dimensional region.
-%%
-\begin{align*}
-\mathsf{toSpec} & : \mathbb{N}_{>0} \rightarrow (\domainVal{} \times \domainVal{}) \rightarrow \mathsf{Approx}(\specDNF)
-\end{align*}
-%%
-A span is converted into an $n$-dimensional
-specification by applying $\mathsf{toSpec}$ to each
-pair of lower and upper bounds in a span, per dimension, and multiplying these
-by $\odot$. The resulting specifications for
-each span are summed by $\oplus$:
-\begin{equation*}
-\textsf{spec}(M) =
-\oplusbig\limits_{\vect{x} \in \spanOp(M)} \,
-\odobig\limits_{i \in \{1, \ldots, n\}} \, \mathsf{toSpec} \; i \; (\vect{x}^L_i, \vect{x}^U_i)
-\end{equation*}
-We unpack the definitions of this algebra.
-
-\begin{definition}[Specifications in DNF]
-Let $r_1, ..., r_n$ range over elements of the \textit{region}
-syntax (see \Cref{fig:syntax}).
-The set \specDNF{} is a subset of syntax trees \textit{spec},
-containing specifications in disjunctive normal form over regions
-where the disjunct is \term{+} and the conjunct is \term{*}; 
- elements of \specDNF{} are of the form:
-%%
-$
-(r^1_{1} \; \term{*} \ldots \term{*} \; r^1_n)\; \term{+} \; \ldots
-\term{+} \; (r^m_1 \, \term{*} \; \ldots \term{*} \; r^m_p)
-$.
-%
-%More formally, $\specDNF \subseteq (\textit{spec} \cup \{\epsilon\})$, defined:
-%%%
-%\begin{align*}
-%\specDNF{} & ::= \tySum{\specDNF} \mid \epsilon \quad\;\;
-%\tySum{\specDNF} ::= \tyProd{\specDNF} \; \term{+} \; \tySum{\specDNF} \mid
-%  \tyProd{\specDNF} \\
-%\tyProd{\specDNF} & ::= \textit{region} \; \term{*} \; \tyProd{\specDNF} \mid
-%   \textit{region}
-%\end{align*}
-%%
-In the following, $S, T$ range over elements of
-\specDNF{} and $\tyProd{s}, \tyProd{t}$ range over conjuncts
- \eg{} $\tyProd{s} = r^1_{1} \; \term{*}
-\ldots \term{*} \; r^1_n$.
-\end{definition}
-
-\newcommand{\exactR}{\mathsf{exact}\;}
-\newcommand{\exact}[1]{\exactR(#1)}
-\newcommand{\upperB}[1]{\mathsf{upper}\;(#1)}
-
-\begin{figure}[t]
-%\newcommand{\llower}{\vect{x}^L_i}
-%\newcommand{\lupper}{\vect{x}^U_i}
-\vspace{-0.7em}
-\newcommand{\llower}{l}
-\newcommand{\lupper}{u}
-{\scalebox{0.95}{
-\begin{minipage}{1\linewidth}
-\begin{align*}
-&\mathsf{toSpec} \; d \; (\llower, \lupper) = \\
-&\hspace{-0.4em}\begin{cases}
-%% POINT
-\exact{\stenReflS{d}} & \llower = 0 \, \wedge \, \lupper = 0\\[0.02em]
-%% FWD
-\exact{\stenFwdS{\lupper}{d}}  & \llower = 0 \, \wedge \, \lupper > 0 \\[0.02em]
-\exact{\stenFwdSR{\lupper}{d}{\irreflS}}\!\!\!\! & \llower = 1 \, \wedge \, \lupper > 0 \\[0.02em]
-%% BWD
-\exact{\stenBwdS{|\llower|}{d}}  & \llower < 0 \, \wedge \, \lupper = 0 \\[0.02em]
-\exact{\stenBwdSR{|\llower|}{d}{\irreflS}}\!\!\!\! & \llower < 0 \, \wedge \, \lupper = -1 \\[0.02em]
-%% CEN
-\exact{\stenCenS{\lupper}{d}}& \!\!\!\llower < 0 \wedge \lupper > 0 \wedge |\llower| = \lupper \\[0.02em]
-\exactR(\stenBwdS{|\llower|}{d} & \!\!\!\llower < 0 \wedge \lupper > 0  \wedge |\llower| \neq \lupper \\
-\qquad\qquad \texttt{+} \, \stenFwdS{\lupper}{d}) & \\[0.02em]
-% BOUND
-\upperB{\stenFwdS{\lupper}{d}} & \llower > 1 \\[0.02em]
-\upperB{\stenBwdS{|\llower|}{d}} \quad & \llower < (-1)
- \\[0.02em]
-% ABSOLUTE REP
-\exact{\epsilon} & \llower = \infty \, \vee \, \lupper = \infty
-  \end{cases}
-\end{align*}
-\end{minipage}
-}}
-\caption{\textsf{toSpec} constructor of $\mathsf{Approx}(\specDNF)$ values.}
-\label{fig:to-spec}
-\vspace{-0.8em}
-\end{figure}
-
-\begin{definition} The $\mathsf{toSpec}$ constructor is defined
-in \Cref{fig:to-spec}, where the lower and upper bounds
-determine region constants; $\mathsf{toSpec} \, d \, (l, u)$ is defined
- if $l$ less than or equal to $u$.
-
-Positive lower and upper bound yield \term{forward} regions, and
-negative lower and upper bounds yield \term{backward} regions.
-The \texttt{nonpointed} attribute (abbreviated \term{np})
-is present when the lower bound is at $1$ or $-1$. 
-In the sixth-case, where the lower bound
-is negative and the upper bound is positive and they have the same
-magnitude, then a \texttt{centered} region is constructed with
-the depth as the common magnitude. If the magnitudes are not the same
-(seventh case) then separated \texttt{forward} and \texttt{backward}
-regions are summed, with their corresponding depths.
-If the lower bound is greater than 1
-(which implies the upper bound is greater than 1 too) then
-this represents a region that is not at the origin (or next to) the
-origin and thus cannot be represented in the specification language.
-Thus, an upper bound is produced showing the maximum extent
-of the region. The penultimate case is the dual, when the lower bound is less
-than $-1$, giving an upper bound on a \texttt{backward} specification.
-\end{definition}
-
-\begin{definition} The $\odot$ operation on $\textsf{Approx}(\specDNF{})$
-  is defined in terms of the intermediate operator $\hat{\odot}$ on
-  \specDNF{}, defined:
-%%
-\begin{align*}
-S \,\hat{\odot}\, \epsilon = S \quad
-\epsilon \,\hat{\odot}\, S = S \qquad
-S \,\hat{\odot}\, T = \bplus_{(\tyProd{s}, \tyProd{t}) \in (S \times
-   T)} (\tyProd{s} \; \term{*} \; \tyProd{t})
-\end{align*}
-%%
-That is, the $\hat{\odot}$ product of two specifications $S$ and $T$ in DNF form
-is the $\term{+}$ sum of all pairwise $\term{*}$ products for every
-pair of $\tyProd{spec}$ drawn from $S$ and $T$. The result
-is indeed in DNF-form, where $\hat{\odot}$ is essentially applying
- the \trule{\textsc{dist}} rule of the equational theory. The operation is
-commutative and associative, and sound with respect to our
-model (see \Cref{lem:alg-soundness} below).
-The $\hat{\odot}$ operation is then lifted to all combinations of
-\textsf{Approx} for $\odot$ in \Cref{fig:odotlift} where
-in the last case $\textit{inj}$ corresponds to the unary
-injections of \textsf{Approx}. For brevity we omit the cases where the arguments above are flipped
-as $\odot$ is commutative.
-\end{definition}
-
-\begin{figure}[t]
-\vspace{-1em}
-{\small{
-\begin{align*}
-\textsf{lower}(S) \odot \textsf{exact}(T) = \; &
-       %\textsf{exact}(T) \odot \textsf{lower}(S) = \;
-       \textsf{both} \, (S \hat{\odot} T, \; T) \\
-\textsf{upper}(S) \odot \textsf{exact}(T) = \; &
-       %\textsf{exact}(T) \odot \textsf{upper}(S) =
-       \textsf{both} \, (T, \; S \hat{\odot} T) \\
-\textsf{lower}(S) \odot \textsf{both}(T_l,T_u) = \; &
-        %\textsf{both}(T_l,T_u) \odot \textsf{lower}(S) = \;
-        \textsf{both} \, (S \hat{\odot} T_l, \; T_u) \\
-\textsf{upper}(S) \odot \textsf{both}(T_l,T_u) = \; & \textsf{both} \,
-                                                      (T_l, \; S \hat{\odot} T_u) \\
-\textsf{both}(S_l,S_u) \odot
-\textsf{both}(T_l,T_u)  = \; & \textsf{both} (S_l \hat{\odot} T_l, \; S_u
-                         \hat{\odot} T_u) \\
-\textit{inj}(S) \odot \textit{inj}(T) = \; & \textit{inj}(S
-                                              \hat{\odot} T)
-\end{align*}
-\vspace{-2em}
-}}
-\caption{Lifting $\hat{\odot}$ to approximations.}
-\label{fig:odotlift}
-\end{figure}
-
-\begin{definition}The $\oplus$ operation on $\textsf{Approx}(\specDNF)$
-is defined similarly to $\odot$, with the intermediate $\hat{\oplus}$ on
-$\specDNF$:
-%%
-\begin{equation*}
-S \,\hat{\oplus}\, \epsilon = S \quad
-\epsilon \,\hat{\oplus}\, S = S \quad
-S \, \hat{\oplus} \, T = S \, \term{+} \, T
-\end{equation*}
-%%
-Thus, if neither specification is empty, their $\hat{\oplus}$ is just
-the syntactic sum $\term{+}$. Then $\oplus$ is the lifting of
-$\hat{\oplus}$ to \textsf{Approx} with the shape as $\odot$ in Figure~\ref{fig:odotlift}.
-\end{definition}
-
-\begin{lemma}%[$\hat{\odot}$, $\hat{\oplus}$ soundness]
-$\forall S,
-  T \in \specDNF$ where $S\!\neq \epsilon \wedge T\!\neq \epsilon$:
-\begin{align*}
-\interp{S \; \hat{\odot} \; T}_n = \interp{S}_n \otimes \interp{T}_n
-  \qquad
-\interp{S \; \hat{\oplus} \; T}_n = \interp{S}_n \cup \interp{T}_n
-\end{align*}
-\label{lem:alg-soundness}
-\vspace{-2em}
-\end{lemma}
-%
-\noindent
-Thus, our model validates specification synthesis.
-
-The final step applies a simplification pass, orienting the
-equational theory $\equiv$ (\Cref{sec:eqs}) into rewrite rules
-based on some simple heuristics.
-For our example, we infer and then synthesise the
-final 5-point specification:
-%%
-\begin{minted}[linenos=false,breakindent=2.4em,xleftmargin=0.2em]{fortran}
-!= stencil centered(depth=1, dim=1)*pointed(dim=2) + centered(depth=1, dim=2)*pointed(dim=1):: a
-\end{minted}
-%%
-
-\section{Evaluation}
-\label{sec:evaluation}
-
-%We have processed nearly one million lines of Fortran code using our
-%CamFort tool and examined the output to determine how useful
-%and applicable it may be. Our corpus of Fortran code is collected from
-%many computational science projects for which we believe our tool will
-%be potentially helpful.
-
-% Corpus details
-% specfem3d ``simulates acoustic (fluid), elastic (solid), coupled
-% acoustic/elastic, poroelastic or seismic wave propagation in any
-% type of conforming mesh of hexahedra (structured or not).''
-% https://github.com/geodynamics/specfem3d
-
-% arpack-ng - Fortran 77 routines designed to solve large scaled
-% eigenvalue problems - https://github.com/opencollab/arpack-ng
-
-To study the effectiveness of our approach, 
-we built a corpus of around 1 million lines of Fortran code from a
-range of scientific computing packages: The Unified Model (UM)~\cite{um},
-E3MG~\cite{RePEc:aen:journl:2006se-a12}, BLAS~\cite{blas},
-Hybrid4~\cite{GBC:GBC635}, GEOS-Chem~\cite{geos-chem}, Navier (based
-on \cite{griebel1997numerical}), Computational Physics
-2~\cite{nicholas2006computational},
-ARPACK-NG~\cite{arpackng}, and
-SPECFEM3D~\cite{specfem3d}.
-
-We first examined how frequently stencil computations occur. We parsed
-959,427 lines of Fortran code and found that 10\% (97,439) of
-statements have a left-hand side as an array subscript on
-neighbourhood indices. This supports the idea that stencil-like
-computations are common in scientific code.
-We then used the inference procedure of the previous section to
-generate specifications for stencils in the corpus to assess
-the design of the language. %  This assumes that the programs have been implemented
-%correctly---
-%Normally one would use our approach in verification mode to
-%check the correctness of new code.
-
-We would not expect to infer a stencil for each array statement we
-found because our analysis restricts 
-the array-subscript-statements that we classify as a
-stencil. In fact, we were able to infer a stencil from 30\% of
-these statements. A single statement can involve multiple arrays and
-we ended up with 60,525 specifications. This shows that we can express
-a large number of stencil shapes within our high-level abstraction and
-validates our initial hypothesis that many stencil computations have a
-regular shape.
-
-%    numStencilSpecs: 60525
-% 4338 files, linesTotal: 1357377, linesParsed: 959427
-% Number of stencils (tickAssign) / number of loops
-% Percentage of a code base that is taken up by stencils
-%Out of 1.35 million lines of code that we fed into our program, and
-%the 959,427 lines that we were able to parse, we identified 60,525
-%cases that fit our criteria for stencil specification.
-
-%\paragraph{Hypothesis 2: Most stencil computations
-%have a regular pattern which can be captured with a simple
-%language of contiguous regions}
-
-% Number of stencils we give a spec to (where a spec declaration to n
-% variables is counted n times) as percentage of number of stencils
-
-    % tickAssign: 97439
-    % tickAssignSuccess: 29402 % <- number of times tickAssign is accompanied by a successful stencil inference
-                               % counts only by +1 each time that tickAssign finds a stencil
-
-%In the analysed body of code we found that we were able to deduce
-%some form of stencil specification from approximately 30\% of the
-%potential stencils examined.
-
-%
-%\subsection{Analysing the design of our specifications}
-
-% 1. How may are just reflexive and how many some reflexive
-% justReflexive: 55504
-% someReflexive: 15823
-
-The majority of specifications generated were relatively simple but we
-found significant numbers of more complex shapes. We grouped common
-patterns in to categories:
-
-\textbf{All pointed} 55,504 of the stencils we found involved 
-only pointed regions. 39,681 of these were pointed in all
-dimensions. Common examples of this were pointwise transformations on
-data (such as scaling).
-
-\textbf{Single-action} specifications comprise one 
-forward, backward, or centered region constant combined via \term{+} 
-or \term{*} with any number of pointed regions. We identified 4,837 single-action
-specifications, of which 1,532 were single-action with a
-\texttt{nonpointed} modifier.
-
-\textbf{Multi-action} specifications comprise at least two 
- forward, backward, or centered regions, combined with
-any number of pointed regions. We identified 265 multi-action
-specifications out of which 207 had regions combined only with
-$\term{*}$ and 58 combined with a mix of $\term{*}$ and \term{+}.
-
-% multiActionRegionOps1: 31
-% multiActionRegionOps2: 115
-% multiActionRegionOps8: 1
-% multiActionRegionOps14: 5
-% multiActionRegionOps17: 4
-
-% %%%%
-% Complexity-wise, out of these 156 multi-action specifications, 31 of
-% them used a single region operator, 115 used two region operators, 1
-% used eight region operators, 5 used fourteen region operators, and 4
-% used seventeen region operators. Some of these very complex stencils
-% were found in the atmospheric models of the UM.
-% %%%%
-
-\textbf{Bounded} specifications occured with 157
-\texttt{atMost} bounds and 36 \texttt{atLeast}, the latter of which were
-always also paired with an upper bound.
-
-\begin{figure}[t]\begin{minted}[fontsize=\scriptsize,breakindent=0em,linenos=false,xleftmargin=0em,breakafter=)]{fortran}
-!=stencil readOnce,(forward(depth=1,dim=3,nonpointed))*(backward(depth=1,dim=1))*(backward(depth=1,dim=2))+(forward(depth=1,dim=3))*(backward(depth=1,dim=1))*(pointed(dim=2))+(forward(depth=1,dim=3))*(backward(depth=1,dim=1,nonpointed))*(backward(depth=1,dim=2,nonpointed))+(forward(depth=1,dim=3))*(backward(depth=1,dim=2))*(pointed(dim=1))+(backward(depth=1,dim=1))*(backward(depth=1,dim=2,nonpointed))*(pointed(dim=3))+(backward(depth=1,dim=1,nonpointed))*(backward(depth=1,dim=2))*(pointed(dim=3))::x
-\end{minted}
-\caption{Complex specification inferred from
-  \textbf{UM}\label{fig:smagorinsky}}
-\vspace{-1em}
-\end{figure}
-
-The single- and multi-action classes represent more complex stencils
-with a real possibility for programmer error. As an extreme example, the Unified
-Model has an implementation of the Smagorinsky subgrid-scale model for
-calculating turbulence on which our inference yields 39 specs from 340
-lines of code. This is an large reduction given
-the complexity of the algorithm.  We show one example
-in~\Cref{fig:smagorinsky} which specifies the access pattern to
-a 3-dimensional array (which we have renamed to 
-\texttt{x}) arising from a kernel of 93 lines of code involving 142
-array subscripts. The specification was the most complex
-seen in our corpus, yet it still represents a significant abstraction
-of the spatial behaviour given size and complexity of the kernel it describes.
-
-We measured the frequency at which individual specifications involved 
-multiple occurences of \term{+} and \term{*}:
-
-\vspace{0.25em}
-\setlength{\tabcolsep}{0.57em}
-{\small{
-\hspace{-1em}\begin{tabular}{c|cccccccc}
-& 0 & 1 & 2 & 3 & 4 & 5 & 6 & $> 7$ \\ \hline
-\term{+} & 60,265 & 220 & 15 & 12 & 5 & 4 & 4 & 0 \\
-\term{*} & 31,393 & 15,151 & 13,457 & 424 & 41 & 42 & 5 & 12
-\end{tabular}}}
-\vspace{0.2em}
-
-\noindent
-\eg{}, roughly half the specifications did not involve $\term{*}$, 
-a quarter use one $\term{*}$ and just under a quarter
-use two $\term{*}$ operators.
-
-\paragraph{Limitations}
-
-There were various reasons why we did not infer 
-specifications on every looped array computation: \\
-\noindent 
-1) \textbf{Non-subset induction variables} occur when the
-induction variables on the RHS are not a subset of those in the LHS. These
-cases are not stencils by our definition. The degenerate case of this
-is to have only constant indices on the LHS. We see lots of examples
-of this in loops as accumulators \eg{} computing the sum over an array; \\
-\noindent
-2) \textbf{Derived induction variables} where an
-index (\mintinline{fortran}{x}) is derived from an
-induction variable (\mintinline{fortran}{i}) as in
-\mintinline{fortran}{x = len - i};  \\
-\noindent
-3) \textbf{Inconsistent induction dimensions} occur when
-an induction variable is used to specify more than one array dimension
-on the RHS or multiple induction variables are used for the same
-dimension on the RHS. These are common in matrix operations such as
-LU-decomposition with assignments such as
-\mintinline{fortran}{a(l) = a(l) - a(m) * b(l, m)}.
-
-\subsection{Detecting errors in the 2-D Jacobi iteration}
-\label{sec:jacobi}
-One common example of a stencil computation is the two-dimensional
-Jacobi iteration that repeatedly goes through each cell in a matrix
-and computes the average value of the four adjacent cells. The kernel
-is given by:
-\begin{minted}{fortran}
-  a(i,j) = (a(i-1,j)+a(i+1,j)+a(i,j+1)+a(i,j-1))/4
-\end{minted}
-We infer a precise specification of its shape as:
-\begin{minted}[breakafter=+:,breakindent=-0.6em,breaksymbolsep=0.4em,linenos=false,xleftmargin=-0.5em]{fortran}
-  != stencil pointed(dim=1)*centered(depth=1,dim=2,nonpointed)+pointed(dim=2)*centered(depth=1,dim=1,nonpointed)::a
-\end{minted}
-%
-To test our implementation, 
-we examined whether programmer errors would be detected by replacing
-the array index offsets with $-1$, $0$, or $1$ and running our
-verification algorithm. Our checking procedure correctly reported a verification
-failure in each of 6,537 permutations corresponding to an error.  
-The iteration computes the average of four adjacent
-cells so $24$ ($4$ factorial) of the possible array index perturbations
-are correct, all of which are accepted by our checker.
-
-% /um/trunk/src/atmosphere/diffusion_and_filtering/turb_smagorinsky.F90
-% Tried introducing indexing errors replacing one of the indices of cx_rho_instance with -1,0,+1
-% 42 cases were caught
-% 6 cases were not caught
-
-% (/ 55504.0 60525) = 91.7%
-% (/ 15823.0 60525) = 26.1%
-% 91.7\% of the specifications involved only reflexive regions. 26.1\%
-
-% 2. Break down of "interesting specifications" (contain more than
-% just reflexive)
-%   - Single-action, histogram on depths
-%   - Single-action + irreflexives, histogram on depths
-
-% singleAction: 4837
-% singleActionIrr: 1532
-
-%   - Multi-action combined only by *
-%   - Multi-action combined also by +, histogram of tree depths
-
-    % multiAction: 156
-    % multiActionMulOnly: 146
-
-%   - Bounded specs (how many lower, upper, or both)
-
-    % atLeast: 36
-    % atMost: 157
-    % boundedBoth: 36
-
-% Present not just percentages but concrete numbers
-
-\section{Related work and conclusions}
-\label{sec:discussion}
-
-Various deductive verification tools can express array indexing
-in their specifications, \eg{}, ACSL of \citet{baudin2008acsl}
-for C (\eg{}~\citet[Example 3.4.1]{burghardt2010acsl}). A specification
-can be given for a stencil computation but must use fine-grained
-indexing as in code and therefore is similarly prone to indexing
-errors. Our approach is much more abstract-- it does
-not aim to reify indexing in the specification, but
-provides simple spatial descriptions which capture a large number
-of common patterns.
+Various deductive verification tools can express array indexing in their
+specifications, \eg{}, ACSL of \citet{baudin2008acsl} for C
+(\eg{}~\citet[Example 3.4.1]{burghardt2010acsl}). A specification can be given
+for a stencil computation but must use fine-grained indexing as in code and
+therefore is similarly prone to indexing errors. Our approach is much more
+abstract-- it does not aim to reify indexing in the specification, but
+provides simple spatial descriptions which capture a large number of common
+patterns.
 
 \citet{kamil2016verified} propose \emph{verified lifting} to extract a
-functionally-complete mathematical representation of low-level, and
-potentially optimised, stencils in Fortran code. This extracted
-predicate representation of a stencil is used to generate code
-for the \textsc{Halide} high-performance
-compiler~\citep{ragan2013halide}. Thus they must capture the
-full meaning of a stencil computation which requires significant
-program analysis. For example, they report that some degenerate
-stencil kernels take up to 17 hours to analyse and others require
-programmer intervention for correct invariants to be inferred.
+functionally-complete mathematical representation of low-level, and potentially
+optimised, stencils in Fortran code. This extracted predicate representation of
+a stencil is used to generate code for the \textsc{Halide} high-performance
+compiler~\citep{ragan2013halide}. Thus they must capture the full meaning of a
+stencil computation which requires significant program analysis. For example,
+they report that some degenerate stencil kernels take up to 17 hours to analyse
+and others require programmer intervention for correct invariants to be
+inferred.
 
@@
 
-Our approach differs significantly. Rather than 
-full representation of stencils, we focus on specifying
-just the spatial behaviour in a lightweight way.
-Thus, it suffices for us to perform a comparatively
-simple data-flow analysis which is efficient, scales linearly with
-code size, and does not require any user intervention.
-Whilst we do not perform deep semantic analysis of stencils,
- the analysis part of our
-approach can be made arbitrarily more sophisticated independent of the rest of
-the work.
+Our approach differs significantly. Rather than full representation of stencils,
+we focus on specifying just the spatial behaviour in a lightweight way.  Thus,
+it suffices for us to perform a comparatively simple data-flow analysis which is
+efficient, scales linearly with code size, and does not require any user
+intervention.  Whilst we do not perform deep semantic analysis of stencils, the
+analysis part of our approach can be made arbitrarily more sophisticated
+independent of the rest of the work.
 % Hence, we do not require SMT solving or search of loop invariants. It
@@
 %
-Furthermore, Kamil \emph{et al.} do not provide a user-visible
-syntactic representation of their specifications, and nor do
-they provide verification from specifications \eg{}, to future-proof
-the code against later changes. Even if they were to provide a
-syntactic representation, for complex stencils such as Navier-Stokes
-from \cref{sec:intro}, it would be as verbose as the code itself,
-making it difficult for programmer to understand the overall shape of
+Furthermore, Kamil \emph{et al.} do not provide a user-visible syntactic
+representation of their specifications, and nor do they provide verification
+from specifications \eg{}, to future-proof the code against later changes. Even
+if they were to provide a syntactic representation, for complex stencils such as
+Navier-Stokes from \Cref{sec:introduction}, it would be as verbose as the code
+itself, making it difficult for programmer to understand the overall shape of
 the indexing behaviour.
@@
 % GPGPU
-Our work has similarities with efforts to verify kernels
-written for General-Purpose GPU
-programming, such as in \citet{Blom:2014:SoCP}. %Stencils are a form of
+Our work has similarities with efforts to verify kernels written for
+General-Purpose GPU programming, such as in \citet{Blom:2014:SoCP}.
+%Stencils are a form of
 %kernel, and GPGPU programming can be viewed as a massively parallel
 %method of transforming a large matrix.
-However, their focus is
-mainly on the synchronisation of kernels and the avoidance of
-data races., while we are interested in correctness 
+However, their focus is mainly on the synchronisation of kernels and the
+avoidance of data races., while we are interested in correctness
 %embedded within a more typical general-purpose programming
@@
 % Sketching Stencils
-\citet{Solar-Lezama:2007:PLDI} give specifications of stencils using
-unoptimised ``reference'' stencils, coupled with partial
-implementations which are completed by a code
-generation tool. %These kinds of specifications are just simple
+\citet{Solar-Lezama:2007:PLDI} give specifications of stencils using unoptimised
+``reference'' stencils, coupled with partial implementations which are completed
+by a code generation tool. %These kinds of specifications are just simple
 %implementations, so this tool is useful for hand-written, optimised
 %stencils.
-The primary purpose of this tool is optimisation rather than
-correctness, and the language of specification is 
-more elaborate than ours.
+The primary purpose of this tool is optimisation rather than correctness, and
+the language of specification is more elaborate than ours.
 
@@
 boundary conditions, and shape of the stencil. Pochoir is aimed at programmers
-reluctant to implement the high-performance cache-oblivious
-``hypertrapezoidal'' algorithms.  Like much of the related work,
-the goal is optimisation rather than verifying program correctness.
+reluctant to implement the high-performance cache-oblivious ``hypertrapezoidal''
+algorithms.  Like much of the related work, the goal is optimisation rather than
+verifying program correctness.
 
-By contrast, the work of \citet{Abe:2013:IPDPSW} studies
-correctness bringing a form of model-checking to
-verify certain stencil computations in the context of
-parallelism 
- in partitioned global address
+By contrast, the work of \citet{Abe:2013:IPDPSW} studies correctness bringing a
+form of model-checking to verify certain stencil computations in the context of
+parallelism in partitioned global address
 space languages. %These are scenarios where an array is divided into
@@
 %analysis. To achieve this,
-Abe \emph{et al.} provide a new language for writing stencil computations. Much of the specification
-effort goes towards describing the distribution of the
-computation over multiple processors. The code for the stencil kernel
-is generated from a relatively high-level specification. 
-In contrast, we integrate directly
-into existing, legacy codebases and established languages, bringing
+Abe \emph{et al.} provide a new language for writing stencil computations. Much
+of the specification effort goes towards describing the distribution of the
+computation over multiple processors. The code for the stencil kernel is
+generated from a relatively high-level specification.  In contrast, we integrate
+directly into existing, legacy codebases and established languages, bringing
 the benefits of verification more easily to scientific computing.%such as
@@
 %Fortran code.
-
-\paragraph{Concluding remarks}
-
-There is an increasing awareness of the need for verification
-techniques in
-science~\cite{post2005computational,oberkampf2010verification,orchard2014computational}.
-Our specification language is an approach in this direction, providing
-a system of lightweight, high-level specification for numerical code.
-As opposed to existing approaches our language takes inspiration from
-the numerical literature and provides a concise, abstract description
-of the stencil.
-
-We evaluated our approach by extending CamFort, an open-source analysis
-tool for Fortran programs~\cite{camfort}, running the inference process over a corpus
-of a million lines of Fortran code. We showed that our language is
-capable of capturing many real-world uses of stencils and showed in a
-case-study that it will detect (simulated) programmer errors.
-
-Our restrictive definition of stencils means we do note generate
-specifications for a large number of iterated array accesses in the
-corpus, \eg{} that incorporate reductions. For
-future work, we will look at expanding the flexibility of 
-our analysis to include these examples.
-
-\bibliography{references}
-
-\end{document}
-
-%%  LocalWords:  refactoring affine parameterised nonpointed atMost
-%%  LocalWords:  centered readOnce atLeast discretisation Equational
-%%  LocalWords:  equational disjunction denotational dimensionality
-%%  LocalWords:  interprocedural Fortran CamFort preprocessor
-%%  LocalWords:  committers

### 2017-04-17 oopsla17/evaluation.tex
next section of eval

--- 2017-04-17 oopsla17/evaluation.tex
@@
 stencil computations have a regular shape.
+
+The majority of specifications generated were relatively simple but we
+found significant numbers of more complex shapes. We grouped common
+patterns into categories:
+
+\textbf{All pointed} \num{\overalljustPointed} of the stencils we
+found involved only pointed regions.
+% 39,681 of these were pointed in all dimensions.
+Common examples of this were pointwise transformations on data (such
+as scaling).
+
+\textbf{Single-action} specifications comprise one 
+forward, backward, or centered region constant combined via \term{+} 
+or \term{*} with any number of pointed regions. We identified \num{\overallsingleAction} single-action
+specifications, of which \num{\overallsingleActionIrr} were single-action with a
+\texttt{nonpointed} modifier.
+
+\textbf{Multi-action} specifications comprise at least two 
+ forward, backward, or centered regions, combined with
+any number of pointed regions. We identified \num{\overallmultiAction} multi-action
+specifications out of which \num{\overallmultiActionMulOnly} had regions combined only with
+$\term{*}$ and \num{\overallmultiActionPlusAndMul} combined with a mix of $\term{*}$ and \term{+}.

### 2017-04-17 oopsla17/resultsToTex.py

@@ -57,3 +57,4 @@ for corpus in corpora.keys():
 print "\\newcommand{\\%s}{%.2g}" % ("overalltickAssignPercent", float(totalTickAssign) / float(totalLoC) * 100)
 print "\\newcommand{\\%s}{%.2g}" % ("overalltickAssignSuccessPercent", float(totalTickAssignSuccess) / float(totalLoC) * 100)
 print "\\newcommand{\\%s}{%.2g}" % ("overalltickAssignSuccessPercentOfTickAssign", float(totalTickAssignSuccess) / float(totalTickAssign) * 100)
+print "\\newcommand{\\%s}{%d}" % ("overallmultiActionPlusAndMul", int(files["overallmultiAction"]) - int(files["overallmultiActionMulOnly"]))

### 2017-04-17 Results.hs
stencil-analysis: better view on histrograms

@@ -222,7 +222,7 @@ instance Camfortable (Physicality p) where
   camfortable (R p) = camfortable p
 
 instance Camfortable (Shape, Position, Contig, Reuse) where
-  camfortable (s, p, _, _) = camfortable s && camfortable p
+  camfortable (s, _, _, _) = s /= Other && camfortable s
 
 instance Camfortable Shape where
   camfortable Orthotope = True
@@ -368,6 +368,12 @@ mapView msg map =
   where
     total = histTotal $ M.elems map
 
+mapViewTotal msg map =
+       "   " ++ msg ++ ":\n"
+    ++ rline' ((replicate 5 ' ') ++ "Total") (show' total)
+  where
+    total = histTotal $ M.elems map
+
 mapView' msg map =
        "   " ++ msg ++ ":\n"
     ++ concatMap (\(cat, dat) -> hline' cat (show' dat)) (M.assocs map)
@@ -393,6 +399,15 @@ instance HistogramShow [Int] where
   histZip (x:xs) (y:ys) = (x+y):(histZip xs ys)
 
   histEmpty = []
+
+
+gnuplotHist :: Int -> [Int] -> String
+gnuplotHist k xs = concat . mapUpTo k gnuplotHist $ zip [0..] xs
+  where
+    mapUpTo k f [] = []
+    mapUpTo k f ((n, v):xs) | k == n = [f (n, sum (v : map snd xs))]
+    mapUpTo k f (x:xs) = f x : (mapUpTo k f xs)
+    gnuplotHist (n, v) = show n ++ " " ++ show n ++ " " ++ show v ++ "\n"
 
 
 hline' cat dat =

### 2017-04-17 oopsla17/analysis/Study.hs
stencil-analysis: better view on histrograms

@@ -3,6 +3,8 @@
 {-# LANGUAGE FlexibleInstances #-}
 {-# LANGUAGE UndecidableInstances #-}
 
+{-# LANGUAGE ScopedTypeVariables #-}
+
 import qualified Data.Map as M
 import Results
 import Data.List
@@ -14,11 +16,21 @@ main = do
   case args of
     [] -> putStrLn "Specify a .restart file to analyse"
     [x] -> putStrLn "Specify an analysis you want to run, e.g., hyps"
-    (file:study:_) -> do
+    (file:study:args) -> do
       resultsString <- readFile file
       let result = ((read resultsString) :: Result)
       case study of
         "hyps" -> putStrLn $ hyps result
+        "histsplot" -> do
+          let camfort h = ((M.!) (camfortableResult h) "Camfort")
+          writeFile "indexExprs.dat" (gnuplotHist (read . head $ args) .  camfort . histNumIndexExprs $ result)
+          writeFile "dimensionality.dat" (gnuplotHist (read . head $ args) . camfort . histDimensionality $ result)
+        "hists" -> do
+          putStrLn $ mapViewTotal "Dimensionality" (histDimensionality result)
+             ++ mapViewTotal "Max depth" (histMaxDepth result)
+             ++ mapViewTotal "Number of indexing expressions" (histNumIndexExprs result)
+             ++ rline' "Length of dataflow path"  (hview . histLengthOfDataflow $ result)
+
 
 regroup :: (Ord c, HistogramShow t) => (Cat -> c) -> M.Map Cat t -> M.Map c t
 regroup classifier =
@@ -155,8 +167,8 @@ hypothesis3 (_, rhs, _) =
   case rhs of
     Affines    _ (R (s, p, Contig, _)) | s /= Other -> "rhs(aff,contig," ++ positionString p ++ ")"
     Neighbours _ (R (s, p, Contig, _)) | s /= Other -> "rhs(neigh,contig," ++ positionString p ++ ")"
-    Affines    _ (R (s, _, _, _))      | s /= Other -> "rhs(affine,nonContig)"
-    Neighbours _ (R (s, _, _, _))      | s /= Other -> "rhs(neigh,nonContig)"
+    Affines    _ (R (s, p, _, _))      | s /= Other -> "rhs(affine,nonContig," ++ positionString p ++ ")"
+    Neighbours _ (R (s, p, _, _))      | s /= Other -> "rhs(neigh,nonContig," ++ positionString p ++ ")"
[…]

### 2017-04-17 oopsla17/model.tex
editing the model and lang

--- 2017-04-17 oopsla17/model.tex
@@
 \noindent
-We define a lattice model of array access patterns which serves as both a
-denotational model of the semantics of our specification language and an
-abstract interpretation domain for source code. This model (1) serves to
-explain the meaning of our specifications; (2) is used in the inference and
-checking algorithms (\Cref{sec:analysis}); (3) justifies an equational theory
-for specifications in the next section; (4) is used to optimise specifications
-using lattice identities; and (5) can be used to guide correct implementations.
+We define a lattice model of array access patterns which serves as
+a semantic model of our specification language and model of array
+access patterns in source code. This model (1) used to explain the
+meaning of our specifications; (2) as a basis for the inference and
+checking algorithms (\Cref{sec:analysis}); (3) justifies an equational
+theory for specifications in the next section; (4) is used to optimise
+specifications using lattice identities; and (5) can be used to guide
+correct implementations.
 
@@
 \emph{index schemes}.  As an initial informal example, consider the
-following simple stencil computation:
-%
-\begin{minted}{fortran}
-do i = 1, n
-  b(i) = a(i,0) + a(i+1,0)
-end do
-\end{minted}
-%
-The access pattern on array \term{a}, relative
+a stencil computation with loop body \fortran{y(i) = x(i,0) + x(i+1,0)}.
+The access pattern on array \term{x}, relative
 to induction variable \term{i}, is captured
@@
 for our specification language (\S\ref{sec:semantics}) and then as
-the target for an abstract interpretation on imperative
-code (\S\ref{sec:fromcode}).
+the target for an interpretation on imperative code (\S\ref{sec:fromcode}).
+We state various results in this section, for which the proofs are
+provided in \Cref{sec:proofs}.
 
@@
   integer sets, \ie{}, a member of $\mathcal{P}(\bz)^{N}$.
-%  An equivalent view of these vectors is as an $N$-times finite Cartesian
-%  product on subsets of $\bz$.
+  An equivalent view of these vectors is as an $N$-times Cartesian
+  product on subsets of $\bz$.
   We use $S, T, U$ to denote index schemes. Henceforth, we assume index schemes
-  are all $N$-dimensional for some $N$.
+  are all $N$-dimensional\footnote{The particular
+    dimensionality is derived from array types when the model is
+    used.} for some $N$.
 
@@
 \begin{restatable}{lem}{vectorUnion}
-\label{lem:vector-union}
-$S$ and $T$ are index schemes such that $\pi_i(S) = \pi_i(T)$ for all $1 \leq i
+\label{lem:vector-union} If
+$S$ and $T$ are index schemes where $\pi_i(S) = \pi_i(T)$ for all $1 \leq i
 \leq N$ apart from some dimension $k$, then:
@@
   \end{equation*}
-
+  %
+  If the
+  superscript to the interval is omitted it is treated as $\mathsf{true}$.
   We also add the distinguished interval $\interv{-\infty}{\infty}{}$, which is simply an
   alias for \bz{}, but this notation prevents handling infinite interval
-  separately in the following definitions, theorems, and proofs. Here, $-\infty$
-  and $\infty$ behaves like top and bottom elements to \bz{} respectively.
+  separately in the following definitions, lemmas, and proofs. Here, $-\infty$
+  and $\infty$ behave as top and bottom elements to \bz{} respectively.
 
-  We denote the set of all such intervals (sets) as $\textit{Interval}$. If the
-  superscript to the interval is omitted it is treated as $\mathsf{true}$.
+  We denote the set of all such intervals as $\textit{Interval}$. 
 \end{defn}
@@
 
-We define two specialisations (and subset spaces) of index scheme:
-\emph{subscript scheme} and \emph{interval scheme}:
+We define two specialisations of index schemes:
+\emph{subscript scheme} and \emph{interval scheme}.
 
 \begin{defn}[Interval scheme]\label{def:interval-scheme}
-  An interval scheme is a finite Cartesian product of intervals on \bz{},
-  denoted by the set $\textit{Interval}^N$ for a product of $N$ intervals.
+  An interval scheme is an $N$-length vector ($N$-ways Cartesian
+  product) of intervals on \bz{}, denoted by the set
+  $\textit{Interval}^N$ for a product of $N$ intervals.
 \end{defn}
@@
 %
-\begin{proof}
-  Straightforward, the join and meet are mapped to $\cup$ and $\cap$, the set is
-  inductively designed to be closed under these operations. Union and
-  intersection are associative, commutative, and absroptive under closed sets
-  and then they are also for \region{N}. This is enough to show that, it is a
-  lattice.
-
-  Further, we have $\bz{}^N \cap R = R$ and $\emptyset \cup R = R$ with
-  $\bz{}^N$ and $\emptyset$ belonging to \region{N}. This makes the lattice a
-  bounded one.
-
-  Finally, the lattice is distributive since union distributes over intersection
-  and vice versa when the set is closed under these operations.
-\end{proof}
+The set of regions $\region{N}$ is target of our specification
+language model, and of the code analysis. We further wrap this in a
+labelled variant which provides information on multiplicity and approximation.
 
@@
   \begin{align*}
-    \mathsf{Mult} \;\; a \;\; &
-      \triangleq \mathsf{mult} \; a \;\mid\; \mathsf{only} \; a \\
-    \mathsf{Approx} \;\; a \;\; &
-      \triangleq \mathsf{exact} \; a \;\mid\; \mathsf{lower} \; a \;\mid\;
+    \mathsf{Mult} \;\; a \;\,
+      \triangleq \;\, \mathsf{mult} \; a \;\mid\; \mathsf{only} \; a \qquad\quad
+    \mathsf{Approx} \;\; a \;\,
+      \triangleq \;\, \mathsf{exact} \; a \;\mid\; \mathsf{lower} \; a \;\mid\;
         \mathsf{upper} \; a
@@
   \; a$ etc.
-  These will be used in the following subsection to give meaning to the
-  specification modifiers for approximation and multiplicity.
+%  These will be used in the following subsection to give meaning to the
+%  specification modifiers for approximation and multiplicity.
 \end{defn}
@@
 \begin{figure}[!t]
-\begin{subfigure}[t]{0.5\textwidth}
+\begin{subfigure}[t]{0.4\textwidth}
 \begin{align*}
@@
   %
-  \interp{\stencil{p}{$i$}{}{}}_N & =
-    \vecgen{N}{i}{\interv{0}{0}{\mathsf{true}}}\\
-%
-  \interp{\stencil{c}{$i$}{$k$}{\textcap{p}}}_N & =
-    \vecgen{N}{i}{\interv{-k}{k}{\interp{\textcap{p}}}} \\
-%
-  \interp{\stencil{f}{$i$}{$k$}{\textcap{p}}}_N & =
-    \vecgen{N}{i}{\interv{0}{k}{\interp{\textcap{p}}}} \\
-%
-  \interp{\stencil{b}{$i$}{$k$}{\textcap{p}}}_N & =
-  \vecgen{N}{i}{\interv{-k}{0}{\interp{\textcap{p}}}}
-\\
    \interp{\texttt{\textcap{r} + \textcap{s}}}_N & =
@@
   \interp{\texttt{\textcap{r} * \textcap{s}}}_N & =
-    \interp{\textcap{r}}_N \wedge \interp{\textcap{s}}_N \\[-1em]
+   \interp{\textcap{r}}_N \wedge \interp{\textcap{s}}_N \\
+%%
+  \interp{\texttt{\textcap{rconst}}}_N & =
+    \vecgen{N}{i}{\interp{\texttt{\textcap{rconst}}}^{\mathsf{c}}}
+  \\[1em]
+  %%
+  \interp{-}^{\mathsf{c}} & : \textit{rconst} \rightarrow
+                            \region{} \\
+  \interp{\stencil{p}{$i$}{}{}}^{\mathsf{c}} & =
+    {\interv{0}{0}{\mathsf{true}}}\\
+%
+  \interp{\stencil{c}{$i$}{$k$}{\textcap{p}}}^{\mathsf{c}} & =
+    {\interv{-k}{k}{\interp{\textcap{p}}}} \\
+%
+  \interp{\stencil{f}{$i$}{$k$}{\textcap{p}}}^{\mathsf{c}} & =
+    {\interv{0}{k}{\interp{\textcap{p}}}} \\
+%
+  \interp{\stencil{b}{$i$}{$k$}{\textcap{p}}}^{\mathsf{c}} & =
+    {\interv{-k}{0}{\interp{\textcap{p}}}}
 \end{align*}
@@
 \hspace{1em}
-\begin{subfigure}[t]{0.4\textwidth}
+\begin{subfigure}[t]{0.3\textwidth}
 \begin{align*}
@@
 operators in terms of the join (union) and meet (intersection)
-of interval schemes. Thus regions are modelled as members
+of interval schemes. Thus the syntax of regions is modelled as members 
 of $\region{N}$.

### 2017-04-17 oopsla17/proofs.tex
editing the model and lang

--- 2017-04-12 oopsla17/proofs.tex
@@
 \end{proof}
+
+\regionLattice*
+
+\begin{proof}
+  Straightforward, the join and meet are mapped to $\cup$ and $\cap$, the set is
+  inductively designed to be closed under these operations. Union and
+  intersection are associative, commutative, and absroptive under closed sets
+  and then they are also for \region{N}. This is enough to show that, it is a
+  lattice.
+
+  Further, we have $\bz{}^N \cap R = R$ and $\emptyset \cup R = R$ with
+  $\bz{}^N$ and $\emptyset$ belonging to \region{N}. This makes the lattice a
+  bounded one.
+
+  Finally, the lattice is distributive since union distributes over intersection
+  and vice versa when the set is closed under these operations.
+\end{proof}

### 2017-04-17 oopsla17/proofs.tex

--- 2017-04-17 oopsla17/proofs.tex
@@
   \end{align*}
+\end{proof}
+
+\regionLattice*
+
+\begin{proof}
+  Straightforward, the join and meet are mapped to $\cup$ and $\cap$, the set is
+  inductively designed to be closed under these operations. Union and
+  intersection are associative, commutative, and absroptive under closed sets
+  and then they are also for \region{N}. This is enough to show that, it is a
+  lattice.
+
+  Further, we have $\bz{}^N \cap R = R$ and $\emptyset \cup R = R$ with
+  $\bz{}^N$ and $\emptyset$ belonging to \region{N}. This makes the lattice a
+  bounded one.
+
+  Finally, the lattice is distributive since union distributes over intersection
+  and vice versa when the set is closed under these operations.
 \end{proof}

### 2017-04-17 oopsla17/theory.tex

--- 2017-04-13 oopsla17/lattice-model.tex
@@
-\documentclass[acmlarge,review]{acmart}
-
-% Essential packages
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{amsthm}
-\usepackage{indentfirst}
-\usepackage{cleveref}
-\usepackage{stencilmacros}
-
-% Essential declarations
-\theoremstyle{definition}
-\newtheorem{defn}{Definition}
-
-\theoremstyle{plain}
-\newtheorem{thm}{Theorem}
-\newtheorem{lem}{Lemma}
-\newtheorem{prop}{Proposition}
-\newtheorem{case}{Case}
-
-\theoremstyle{remark}
-\newtheorem{remark}{Remark}
-
-\begin{document}
-
-\title{Lattice model}
-\author{Mistral Contrastin}
-
-\section{Giving specifications meaning}
-
-We construct a lattice and use this lattice to give meanings to regions. We
-then give full meaning to specifications by wrapping regions in variant types
-for multiplicity and approximation. Model defined in this section will later be
-used for two purposes. First, we show soundness of the equational theory of
-specifications with respect to the model. Second, we define inference and
-consistency checking procedures of specifications with respect to the
-interpretations of the stencils in the code and the specifications.
-
-\subsection{Constructing the lattice model of regions}
-
-\begin{defn}
-  We define the set \zinf{} as
-  $\mathbb{Z}$ extended with $\infty$ and $-\infty$. For any $n$ in
-  \zinf{}, we have $-\infty \leq n \leq \infty$. The resulting set is
-  a total order with top and bottom elements.
-\end{defn}
-
-\begin{defn}
-  We define an extended notion of closed interval on $\zinf{}$ which may contain
-  a \emph{hole} at the origin, written \interv{a}{b}{c} where $a$ and $b$ are
-  drawn from \zinf{} with $a \leq 0 \leq b$, $a$ is $-\infty$ \emph{iff} $b$ is
-  $\infty$, and $c$ is drawn from $\mathbb{B} = \{ \mathsf{true}, \mathsf{false}
-  \}$. We define it as follows:
-%
-  \begin{equation*}
-    \interv{a}{b}{c} \triangleq
-      \{ n \mid a \leq n \leq b \wedge (\neg c \implies n \neq 0) \}
-  \end{equation*}
-
-  Set of all such intervals (sets) is $\textit{Interval}$. If the superscript is
-  leftout it is treated as $\mathsf{true}$.
-\end{defn}
-
-\begin{lem}{}\label{lem:zinf-identities}
-  We have the following dual identities for \zinf{} intervals:
-%
-  \begin{align*}
-    \interv{a}{b}{c} \cap \interv{d}{e}{f} & =
-      \interv{\max \{a,d\}}{\min \{b,e\}}{c \wedge f} \\
-    \interv{a}{b}{c} \cup \interv{d}{e}{f} & =
-      \interv{\min \{a,d\}}{\max \{b,e\}}{c \vee f}
-  \end{align*}
-\end{lem}
-%
-\begin{proof}
-  We give the proof of the first identity and the second one is similar.
-  \begin{align*}
-    \interv{a}{b}{c} \cap \interv{d}{e}{f} = &
-      \; \{ n \;|\; a \leq n \leq b \wedge (\neg c \implies n \neq 0) \}
-      \;\cap \\
-      & \; \{ n \;|\; d \leq n \leq e \wedge (\neg f \implies n \neq 0) \}
-      \\
-    = & \; \{ n \;|\; \max \{a,d\} \leq n \leq \min \{b,e\} \wedge (\neg c
-      \vee \neg f \implies n \neq 0) \} \\
-    = & \; \interv{\max{\{a,d\}}}{\min{\{b,e\}}}{c \wedge f}
-  \end{align*}
-\end{proof}
-
-\begin{defn}{(\emph{Offset subspace})}
-  An $N$-dimensional \emph{offset subspace} is a finite Cartesian product of $N$
-  subsets of $\zinf{}$. The set of $N$-dimensional offset subspaces is written
-  ${\mathcal{P}(\zinf{})}^N$.
-
-  A offset spaces can be \emph{projected} in the $i^{\textit{th}}$ dimension by
-  $\pi_i : {\mathcal{P}(\zinf{})}^N \to \mathcal{P}(\zinf{})$ where $1 \leq i
-  \leq N$.
-\end{defn}
-
-\begin{defn}{(\emph{Subscript})}
-  An $N$-dimensional offset subspace $S$ is called a \emph{subscript} if it
-  satisfies:
-%
-  \begin{equation*}
-    \forall i.\ 1 \leq i \leq N \implies
-      \pi_i (S) = \{p\} \wedge p \in \mathbb{Z}
-      \vee
-      \pi_i (S) = \interv{-\infty}{\infty}{\mathsf{true}}
-  \end{equation*}
-  %
-  That is, the $i^{th}$ component of the set is either a singleton
-  in $\mathbb{Z}$ or the infinite interval.
-
-  \todo{%
-    We can move this explanation.
-
-    Subscript spaces model array subscript expressions in our source langauge
-    where each subscripted expression is either a constant translation of an
-    induction variable within a loop or some other expression which is constant
-    with respect to the loop e.g. \texttt{a(i+1, 0)}.
-  }
-\end{defn}
-
-\begin{lem}{}\label{lem:vector-intersect}
-  Intersection distributes over sets of points. If $\mathit{off}$ and
-  $\mathit{off'}$ are sets of $N$-dimensional points, then
-%
-  \begin{equation*}
-    \mathit{off} \cap \mathit{off'} =
-      \prod_{i = 1}^{N} \pi_i(\mathit{off}) \cap \pi_i(\mathit{off'})
-  \end{equation*}
-\end{lem}
-%
-\begin{proof}
-  \begin{align*}
-    \mathit{off} \; \cap \; \mathit{off'} &
-    = \{x \mid \bigwedge_{1 \leq i \leq N } x_i \in \pi_i(\mathit{off}) \}
-      \;\cap\;
-      \{x \mid \bigwedge_{1 \leq i \leq N } x_i \in \pi_i(\mathit{off'}) \} \\
-    & = \{x \mid \bigwedge_{1 \leq i \leq N }
-      (x_i \in \pi_i(\mathit{off}) \wedge x_i \in \pi_i(\mathit{off'})) \} \\
-    & = \prod_{i = 1}^{N} \pi_i(\mathit{off}) \cap \pi_i(\mathit{off'})
-  \end{align*}
-\end{proof}
-
-\begin{lem}{}\label{lem:vector-union}
-  Let $\mathit{off}$ and $\mathit{off'}$ be two sets of
-  points such that $\pi_i(\mathit{off})$ is equal to $\pi_i(\mathit{off'})$ for all $1
-  \leq i \leq N$ except possibly at $k$, then we have the following:
-%
-  \begin{equation*}
-    \mathit{off} \; \cup \; \mathit{off'}
-    =
-    \pi_1(\mathit{off}) \times \cdots \times
-    (\pi_k(\mathit{off}) \cup \pi_k(\mathit{off'})) \times \cdots \times
-    \pi_N(\mathit{off})
-  \end{equation*}
-\end{lem}
-%
-\begin{proof}
-  \begin{align*}
-    \mathit{off} \; \cup \; \mathit{off'} &
-    = \{x \mid
-      \bigwedge_{1 \leq i \leq N } x_i \in \pi_i(\mathit{off}) \}
-      \;\cup\;
-      \{x \mid
-          \bigwedge_{1 \leq i \leq N } x_i \in \pi_i(\mathit{off'}) \} \\
-    & = \{x \mid
-          \bigwedge_{\substack{1 \leq i \leq N \\ i \neq k}}
-            x_i \in \pi_i(\mathit{off}) \wedge x_k \in \pi_k(\mathit{off}) \vee
-          \bigwedge_{\substack{1 \leq i \leq N \\ i \neq k}} x_i \in
-            \pi_i(\mathit{off'}) \wedge x_k \in \pi_k(\mathit{off'})
-        \} \\
-    & = \{x \mid
-          \bigwedge_{\substack{1 \leq i \leq N \\ i \neq k}} x_i \in
-            \pi_i(\mathit{off}) \wedge
-            x_k \in \pi_k(\mathit{off}) \cup \pi_k(\mathit{off'})
-        \} \\
-        & = \pi_1(\mathit{off}) \times \cdots \times
-        (\pi_k(\mathit{off}) \cup \pi_k(\mathit{off'})) \times \cdots \times
-        \pi_N(\mathit{off})
-  \end{align*}
-\end{proof}
-
-\begin{defn}{(\emph{Interval subspace})}
-  An interval subspace is a finite Cartesian product of intervals on \zinf{},
-  denoted by the set $\textit{Interval}^N$ for a product of $N$ intervals. An
-  interval subspace is an offset subspace.
-\end{defn}
-
-\begin{defn}{(\emph{Region})}
-  A region \region{N} is a set of $N$-dimensional point intervals defined as
-  the smallest set satisfying the following:
-%
-  \begin{enumerate}
-    \item If $R$ is in $\textit{Interval}^N$, then $R$ is in \region{N}.
-    \item If $R$ and $S$ are in \region{N}, then so are $R \cap S$ and
-      $R \cup S$.
-  \end{enumerate}
-\end{defn}
-
-\begin{prop}{}
-  $(\region{N},\subseteq)$ is a distributive lattice.
-\end{prop}
-%
-\begin{proof}
-  Any sublattice of a distributive lattice is a distributive lattice. Power sets
-  partially ordered with subset relation are distributive lattices. \region{N}
-  is a subset of $\mathcal{P}{(\zinf{}^N)}$ since all interval subspaces are
-  subsets of this power set and; union and intersection cannot add an element
-  that is not already in the powerset. Inductive definition also ensures that
-  \region{N} is closed under these operations. Therefore, \region{N} is a
-  distributive sublattice of $(\mathcal{P}{(\zinf{}^N)}, \subseteq)$.
-\end{proof}
-
-\begin{defn}{}
-  $\mathsf{Mult}$ and $\mathsf{Approx}$ are parametric labelled variant types
-  with injections given by their definition:
-%
-  \begin{align*}
-    \mathsf{Mult} \; a &
-      \triangleq \mathsf{mult} \; a \;\mid\; \mathsf{only} \; a \\
-    \mathsf{Approx} \; a &
-      \triangleq \mathsf{exact} \; a \;\mid\; \mathsf{lower} \; a \;\mid\;
-        \mathsf{upper} \; a \;\mid\; \mathsf{both} \; a \; a
-  \end{align*}
-%
-  \eg{}, $\mathsf{lower}$ is an injection $\mathsf{lower} : a \to \mathsf{Approx}
-  \; a$ etc.
-
-  These will be used in the following subsection to give meaning to the
-  specification modifiers for approximation and multiplicity.
-\end{defn}
-
-\subsection{Denotational semantics for specifications}
-
-We give some basic interpretations: $\interp{\texttt{n}} = n$ where $n$ is in
-$\mathbb{Z}^+$, $\interp{\texttt{pointed}} = \llbracket \epsilon \rrbracket =
-\mathsf{true}$, $\interp{\texttt{nonpointed}} = \mathsf{false}$.
-Typographically, \textcap{bold monotype font} is used within specifications
-to indicate a partial textual capture. For convenience, we use \textit{italic
-serif font} inside specifications rather than interpreting its textual capture.
-This coalesces syntactic and semantic representation but avoids clutter in the
-presentation.
-
-Let $\textit{vec-gen}_N : \mathbb{N}^+ \times \textit{Interval} \to
-\textit{Interval}^N$ be the function generating an $N$ dimensional vector
-interval such that \vecgen{N}{i}{\interv{a}{b}{c}} has
-$\interv{-\infty}{\infty}{\mathsf{true}}$ in all dimensions except the $i^{th}$
-one. There, it has \interv{a}{b}{c}.
-
-First, we give meaning to region constants (introduced in \Cref{}).
-%
-\begin{align*}
-  \interp{\stencil{p}{$i$}{}{}}_N & =
-    \vecgen{N}{i}{\interv{0}{0}{\mathsf{true}}}\\
-%
-  \interp{\stencil{c}{$i$}{$k$}{\textcap{p}}}_N & =
-    \vecgen{N}{i}{\interv{-k}{k}{\interp{\textcap{p}}}} \\
-%
-  \interp{\stencil{f}{$i$}{$k$}{\textcap{p}}}_N & =
-    \vecgen{N}{i}{\interv{0}{k}{\interp{\textcap{p}}}} \\
-%
-  \interp{\stencil{b}{$i$}{$k$}{\textcap{p}}}_N & =
-  \vecgen{N}{i}{\interv{-k}{0}{\interp{\textcap{p}}}}
-\end{align*}
-
-Second, we provide the semantics for the permissive region operator, \texttt{+},
-and the restrictive region operator, \texttt{*}, by associating them with join
-and meet respectively. Let \textcap{R} and \textcap{S} be regions.
-%
-\begin{align*}
-  \interp{\texttt{\textcap{R} + \textcap{S}}}_N & =
-    \interp{\textcap{R}}_N \vee \interp{\textcap{S}}_N
-&
-  \interp{\texttt{\textcap{R} * \textcap{S}}}_N & =
-    \interp{\textcap{R}}_N \wedge \interp{\textcap{S}}_N
-\end{align*}
-
-We mark the presence of modifiers such as \texttt{readOnce} and
-\texttt{atMost} as introduced in \Cref{}. Multiplicity
-modifiers are interpreted as injections in the $\mathsf{Mult}$
-variant by $\interpMult{}$ in Fig~\ref{fig:semantics}.
-In a similar way, approximation modifiers are interpreted
-as injections into the $\mathsf{Approx}$ variant by $\interpApprox{}$.
-
-Finally, we can give full semantics of a stencil specification:
-%
-\begin{equation*}
-  \interp{\texttt{stencil \textcap{mult}, \textcap{approx}, \textcap{region}}}_N =
-    \interpMult{\textcap{mult}}
-           {(\interpMax{\textcap{approx}}
-                    {(\interp{\textcap{region}}_N)})}
-\end{equation*}
-
-\section{Equational \& approximation theories}
-
-We know introduce the equational and approximation theory of stencil
+We now introduce the equational and approximation theory of stencil
 specifications based on $\equiv$ and $\preceq$ in the following two subsections.
@@
 
-We define an equivalence relation, $\equiv$, on regions as follows:
+We define an equivalence relation, $\equiv$. The purpose of this relation is to
+allow programmers to write specifications with greater flexibility. It allows
+specifications to be written in various levels of compactness allowing
+optimising for space and better documentation. The relation is defined on
+regions as follows:
 
@@
 
-\begin{thm}{(Equational soundness)}
-  Equational theory is sound with respect to the lattice model. Let $R$
-  and $S$ be $N$ dimensional regions, then we have
-%
-  \begin{equation*}
-    R \equiv S \implies \interp{R}_N = \interp{S}_N
-  \end{equation*}
-\end{thm}
+\subsection{Approximations}
 
-\begin{proof}
-  Both $\equiv$ and $=$ are equivalence relations. We only need to show that
-  each property associated with specifications equivalence holds as an equality
-  in the model.
-%
-  \begin{description}
-    \item[Basic \& Subsumption] These follow from basic properties of lattices.
-%
-    \item[Dist] This is immediate from the definition of distributive
-      lattice which the model is one.
-%
-    \item[Overlapping pointed] Counterparts of individual regions for some
-      Boolean $p$ are given below:
-%
-      \begin{align*}
-        \interp{\stencil{\textcap{R}}{$n$}{$k$}{\texttt{nonpointed}}}_N
-        & = \vecgen{N}{n}{\interv{-k}{k}{\mathsf{false}}} \\
-%
-        \interp{\stencil{p}{$n$}{}{}}_N
-        &= \vecgen{N}{n}{\interv{0}{0}{\mathsf{true}}} \\
-%
-        \interp{\stencil{\textcap{R}}{$n$}{$k$}{}}_N
-        &= \vecgen{N}{n}{\interv{-k}{k}{\mathsf{true}}}
-      \end{align*}
-
-      Using \Cref{lem:vector-union}, it is sufficient to show
-      $ \interv{-k}{k}{p} \;\cup\; \interv{0}{0}{\mathsf{true}} =
-        \interv{-k}{k}{\mathsf{true}} $ holds. This immediately follows from
-      \Cref{lem:zinf-identities}.
-%
-    \item[Centered] Let $\interp{\textcap{p1}}$, $\interp{\textcap{p2}}$, and
-      $\interp{\textcap{p3}}$ be $p1$, $p2$, and $p3$ respectively. We
-      know that $p1$ is $p2 \vee p3$ from (\textbf{Centered}) detinition. Then
-      we get the following interpretation for region constants involved in the
-      equivalence:
-%
-      \begin{align*}
-        \interp{\stencil{c}{$k$}{$n$}{$p1$}}_N
-          &= \vecgen{N}{n}{\interv{-k}{k}{p2 \vee p3}} \\
-%
-        \interp{\stencil{f}{$k$}{$n$}{$p2$}}_N
-          &= \vecgen{N}{n}{\interv{0}{k}{p2} }\\
-%
-        \interp{\stencil{b}{$k$}{$n$}{$p3$}}_N
-          &= \vecgen{N}{n}{\interv{0}{k}{p3}}
-      \end{align*}
-%
-      Using \Cref{lem:vector-union}, it is sufficient to show $\interv{-k}{k}{p2
-      \vee p3} = \interv{0}{k}{p2} \cup \interv{-k}{0}{p3}$ holds. This
-      immediately follows from \Cref{lem:zinf-identities}.
-  \end{description}
-\end{proof}
-
-\subsection{Approximations}
-We define a partial of order of approximations, $\preceq$. This relation
-provides us with the following:
+We define a partial of order of approximations, $\preceq$. The relation is used
+in equational theory and provides means of writing more compact lower and upper
+bound specifications. The relation is defined as follows:
 
@@
 
-\begin{thm}{(Approximation soundness)}
-  Theory of approximation is sound with respect to the lattice model. Let $R$
-  and $S$ be $N$ dimensional regions, then we have
-%
-  \begin{equation*}
-    R \preceq S \implies \interp{R}_N \subseteq \interp{S}_N
-  \end{equation*}
-\end{thm}
-%
-\begin{proof}
-  Both $\preceq$ and $\subseteq$ are partial orders on sets. We only need to
-  show that each inequality holds in the model.
-%
-  \begin{description}
-    \item[Equivalence] Region equavilance is modeled with set equality and the
-      partial order on regions is modeled with subset relation. Partial orders
-      are reflexive and the model of eqivalence agrees with that of the partial
-      order, so this rule holds.
-%
-    \item[Combined] \texttt{*} and \texttt{+} maps to meet and join operations
-      in the lattice. All meets and joins in a lattice satisfy these
-      inequalities.
-
-    \item[Depth] When $m$ is $k$ and $l$ in
-      $\stencil{\textcap{R}}{$n$}{$m$}{\textcap{p}}$, we obtain the
-      interpretations $\vecgen{N}{n}{\textit{int}_k}$ and
-      $\vecgen{N}{n}{\textit{int}_l}$ for some $N$ as interpretations of
-      regions. Call these \textit{ks} and \textit{ls} respectively. One way to
-      show $\textit{ks} \subseteq \textit{ls}$ is to show that the intersection
-      of \textit{ks} and \textit{ls} is \textit{ks}.
-
-      To show that the intersection is \textit{ks}, we use
-      \Cref{lem:vector-intersect} which reduces the problem to showing that the
-      intersection of $\textit{int}_k$ and $\textit{int}_l$ is the former since
-      at every point except $n$ the intervals agree, hence the intersection
-      remains the same at these dimensions. Then we use
-      \Cref{lem:zinf-identities} to calculate the intersection. By assumption we
-      have $k \leq l$, so the maximum of the lower bounds is $-k$ and the
-      minimum of lower bounds is $k$ this is same as interval $\textit{int}_k$.
-      Note that pointedness is the same for both intervals and depths are
-      positive integers, so the pointedness is preserved under intersection.
-      This shows the intersection of \textit{ls} and \textit{ks} is indeed
-      \textit{ks} and concludes the proof.
-  \end{description}
-\end{proof}
-
 We present few inequalities that can be derived from the axioms and are useful
@@
 
-\begin{prop}{(Centered approximation)}
+\begin{restatable}[Centered approximation]{prop}{centeredApprox}
   For any dimension $n$, depth $k$, and pointed attribute $p$,
@@
   \end{align*}
-\end{prop}
-%
-\begin{proof}
-  \shortv{%
-    Both inequalities are proved by combining the missing half of centered by
-    using the inequality (\textbf{Combined}) and then merging them into a single
-    region constant using (\textbf{Centered}).
-  }
+\end{restatable}
 
-  \longv{%
-    Proof of the first inequality:
-    \begin{align*}
-      \stencil[s]{f}{$n$}{$k$}{\textcap{p}}
-        & \preceq
-          \stencil[s]{f}{$n$}{$k$}{\textcap{p}}
-          \;\texttt{+}\;
-          \stencil[s]{b}{$n$}{$k$}{\textcap{p}}
-        & (\textbf{Combined}) \\
-      & \equiv \stencil[s]{c}{$n$}{$k$}{\textcap{p}} & (\textbf{Centered})
-    \end{align*}
-
-    Proof of the second inequality:
-    \begin{align*}
-      \stencil[s]{b}{$n$}{$k$}{\textcap{p}}
-        & \preceq
-          \stencil[s]{f}{$n$}{$k$}{\textcap{p}}
-          \;\texttt{+}\;
-          \stencil[s]{b}{$n$}{$k$}{\textcap{p}}
-        & (\textbf{Combined}) \\
-      & \equiv \stencil[s]{c}{$n$}{$k$}{\textcap{p}} & (\textbf{Centered})
-    \end{align*}
-  }
-\end{proof}
-
-\begin{prop}{(Point approximation)}
+\begin{restatable}[Point approximation]{prop}{pointApprox}
   Let \textcap{R} be one of \texttt{forward}, \texttt{backward}, and
@@
   \end{align*}
-\end{prop}
-%
-\begin{proof}
-  \shortv{%
-    Both inequalities are proved by breaking \textcap{R} using
-    (\textbf{Overlapping pointed}) followed by the inequality
-    (\textbf{Combined}).
-  }
-
-  \longv{%
-    Proof of the first inequality:
-    \begin{align*}
-      \stencil[s]{p}{$n$}{}{}
-        & \preceq \stencil{R}{$n$}{$k$}{nonpointed}
-                  \;\texttt{+}\;
-                  \stencil[s]{p}{$n$}{}{}
-        & (\textbf{Combined}) \\
-      & \equiv \stencil{R}{$n$}{$k$}{} & (\textbf{Overlapping pointed})
-    \end{align*}
-
-    Proof of the second inequality:
-    \begin{align*}
-      \stencil{R}{$n$}{$k$}{nonpointed}
-        & \preceq \stencil{R}{$n$}{$k$}{nonpointed}
-                  \;\texttt{+}\;
-                  \stencil[s]{p}{$n$}{}{}
-        & (\textbf{Combined}) \\
-      & \equiv \stencil{R}{$n$}{$k$}{} & (\textbf{Overlapping pointed})
-    \end{align*}
-  }
-\end{proof}
-
-\section{Analysis, Consistency, Inference}
-
-\subsection{Analysis}
-
-\subsection{Consistency}
-
-Once the specification is synthesised into objects in the model and stencil
-under examination is analysed, consistency checking itself is remarkably simple.
-Two stage checking is outlined below.
-
-\begin{enumerate}
-  \item Multiplicity of the model is checked against the stencil. The only way
-    this fails is if the original specification involved \texttt{readOnce}, but
-    some indexing occur multiple times in the stencil.  Note that absence of
-    \texttt{readOnce} is interpreted liberally, and if the stencil indices
-    happen to occur uniquely, this does not constitute a failure.
-  \item Definition of vector intervals is completely unwinded into sets of
-    offset tuples and depending the approximation modifier (or lack thereof)
-    we compare these tuples to offsets of the stencil computation. Lower and
-    upper bounds are checked with subset and superset relations respectively,
-    while absence of \texttt{atMost} and \texttt{atLeast} results in exact
-    checking with set equality.
-\end{enumerate}
-
-Here is the mathematical definition of consistency checking:
-\begin{align*}
-  \mathit{consistent}(\mathit{ix}, \mathit{model}) & = \begin{cases}
-    \mathsf{false} & \mathit{model} = \mathsf{once}(x) \wedge ix = \mathsf{mult}(y) \\
-    \mathit{consistent'}(\mathit{peel}(ix), \mathit{peel}(model)) & \textit{otherwise}
-  \end{cases} \\
-%
-  \mathit{consistent'}(\mathit{ix}, \mathit{model}) & = \begin{cases}
-    m = \mathit{ix} & \mathit{model} = \mathsf{exact}(m) \\
-    m \supseteq \mathit{ix} & \mathit{model} = \mathsf{upper}(m) \\
-    m \subseteq \mathit{ix} & \mathit{model} = \mathsf{lower}(m)
-  \end{cases}
-\end{align*}
-
-Here helper function $\mathit{peel} : \mathsf{Approx} \; a \to a$ removes the
-approximation label.
-
-Due to use of infinite intervals, the model and index sets are potentially
-infinite meaning simple procedures used to compute equality, subset, and
-superset relations for finite sets are no avail. The subset and superset
-relations are handled by computing the intersection of $m$ and $\mathit{ix}$
-using \Cref{lem:vector-intersect}, then checking equality with $m$ and
-$\mathit{ix}$ respectively.
-
-\todo{Here we simply show \textit{consistency'} as set equality, but since sets
-are infinite we might need to elaborate based on representation.}
-
-\subsection{Inference}
-
-\begin{defn}{(Extended origin distance)}
-  We measure the \emph{extended origin distance} of a point from the extended
-  origin as the minimum number of translations by $\pm 1$ in one dimension at a
-  time, to reach a point in the extended origin.
-
-  This notion extends to subscripts. Since there may be an infinite number of
-  points, we measure the distance from the closest one.
-\end{defn}
-%
-\begin{prop}{}
-  The closed form of extend origin distance for an $N$-dimensional subscript $S$
-  is denoted with $\mathit{dist}(S)$ and is defined as:
-%
-  \begin{align*}
-    \mathit{dist}(S) & = \sum_{i=1}^{N} \mathit{dist}_i(S)
-    &
-    \mathit{dist}_i(S) & = \begin{cases}
-      0 & \pi_i(S) \in \{ \interv{-\infty}{\infty}{}, \{ 0 \}\} \\
-      |\mathsf{a}| - 1 & \pi_i(S) = \{\mathsf{a}\} \\
-    \end{cases}
-  \end{align*}
-\end{prop}
-%
-\begin{proof}
-  Now we prove that function $\mathit{dist}$ indeed gives the lowest number of
-  unit translations to reach the extended origin.
-\end{proof}
-
-\begin{defn}{(Adjacent)}
-  A subscript $S$ is said to be \emph{adjacent at index $j$} of another
-  subscript $S'$ if for every index $i$ except $j$, $\pi_i(S) = \pi_i(S')$ but
-  at $j$ then $\pi_j(S) \neq \pi_j(S')$ such that $|S| = |S'| + 1$.
-\end{defn}
-%
-\begin{prop}
-  The unique adjacent at index $i$ of a subscript, if it exists, is given by the
-  following partial function:
-%
-  \begin{align*}
-    \mathit{adjacent}(S,i) & = \prod^{N}_{j=1} \mathit{shift}(S,i,j) &
-%
-    \mathit{shift}(S,i,j) & =
-      \begin{cases}
-        \{a - 1\} & i = j \wedge \pi_i(S) = \{\mathsf{a}\} \wedge \mathsf{a} > 0\\
-        \{a + 1\} & i = j \wedge \pi_i(S) = \{\mathsf{a}\} \wedge \mathsf{a} < 0\\
-        \pi_j(S)  & otherwise
-      \end{cases}
-  \end{align*}
-\end{prop}
-%
-\begin{proof}
-\todo{Hm- I guess uniqueness follows from this being a function rather
-  than a relation? Seems fine no problem to prove- }
-\end{proof}
-
-\todo{On the following, add some intutition first- what is this and
-  what is it for?}
-\begin{defn}{(Exact set)}
-  The \emph{exact set} at origin distance $d$, $\mathit{ES}_{P,d}$, of a group
-  of $N$ dimensional subscripts $S$ is defined below:
-%
-  \begin{align*}
-    \mathit{ES}_{S,0} = &
-      \{ s \in S \mid \forall \i.\ \pi_i(s) \in \{\{-1\}, \{0\}, \{1\}\} \} \\
-    \mathit{ES}_{S,d} = &
-      \{ s \in S \mid \forall i.\
-      \pi_i(s) \notin \{\{-1\}, \{0\}, \{1\}\} \implies
-      \exists s' \in \mathit{ES}_{S,d-1}.\ s' =
-         \mathit{adjacent}(s,i) \}
-  \end{align*}
-%
-  The overall \emph{exact set}, $\mathit{ES}_S$ is then
-  $\bigcup_{0 \leq d \leq D} \mathit{ES}_{S,d}$, where $D$ is the maximum origin
-  distance in $S$.
-\end{defn}
-
-\begin{lem}{(\emph{Connecting})}
-  The following three statements are equivalent for a set of $N$-dimensional
-  subscripts $S$:
-%
-  \begin{enumerate}
-    \item $S$ is the same as its exact set $\mathit{ES}_S$;
-    \item $S$ when regarded as offsets from induction variables forms a
-      neighbourhood stencil;
-    \item $S$ is in \region{N}.
-  \end{enumerate}
-\end{lem}
-
-\begin{thm}{(Soundness of exact shape inference)}
-  For a set of subscripts $S$ of same dimension, the inference procedure for
-  neighbourhood stencils is sound if the following holds:
-%
-  \begin{equation*}
-    \forall S. \mathit{peel}(\mathit{infer}(S)) = m \implies
-      \mathit{consistent'}(S,m)
-  \end{equation*}
-\end{thm}
-
-\begin{thm}{(Completeness of inference)}\label{thm:inf-completeness}
-  Inference procedure can infer a specification for every stencil computation.
-\end{thm}
-%
-\begin{proof}
-  Would be nice to have.
-\end{proof}
-
-\begin{thm}{(Completeness of exact inference)}
-  Inference procedure can infer a specification for every stencil computation
-  with array indexing behaviour forms a superposition of hyperrectangles removed
-  at most one unit from the origin in any of its dimension.
-\end{thm}
-%
-\begin{proof}
-  Would be nice to have.
-\end{proof}
-
-\begin{thm}{(Optimality of inference)}\label{thm:inf-optimality}
-  For exact inference, inference procedure always infers specifications with
-  fewest number of region constants.
-\end{thm}
-%
-\begin{proof}
-  Would be nice to have.
-\end{proof}
-
-\begin{remark}{}
-  Optimality of general inference is trivial. We only need a single region
-  constant covering accesses in any one of the stencil's dimensions and that
-  would provide an upper bound for the stencil.
-\end{remark}
-
-\end{document}
+\end{restatable}

### 2017-04-17 oopsla17/evaluation.tex
bounded specs
added a table of multiple use of mulOp; included smagorinsky example commented because it seems to be no longer the case under our new model...
limitations & jacobi

@@
 $\term{*}$ and \num{\overallmultiActionPlusAndMul} combined with a mix of $\term{*}$ and \term{+}.
+
+\textbf{Bounded} specifications occured with \num{\overallatMost}
+\texttt{atMost} bounds and \num{\overallatLeast} \texttt{atLeast}, the latter of which were
+always also paired with an upper bound.
+
+% \begin{figure}[t]\begin{minted}[fontsize=\scriptsize,breakindent=0em,linenos=false,xleftmargin=0em,breakafter=)]{fortran}
+% !=stencil readOnce,(forward(depth=1,dim=3,nonpointed))*(backward(depth=1,dim=1))*(backward(depth=1,dim=2))+(forward(depth=1,dim=3))*(backward(depth=1,dim=1))*(pointed(dim=2))+(forward(depth=1,dim=3))*(backward(depth=1,dim=1,nonpointed))*(backward(depth=1,dim=2,nonpointed))+(forward(depth=1,dim=3))*(backward(depth=1,dim=2))*(pointed(dim=1))+(backward(depth=1,dim=1))*(backward(depth=1,dim=2,nonpointed))*(pointed(dim=3))+(backward(depth=1,dim=1,nonpointed))*(backward(depth=1,dim=2))*(pointed(dim=3))::x
+% \end{minted}
+% \caption{Complex specification inferred from
+%   \textbf{UM}\label{fig:smagorinsky}}
+% \vspace{-1em}
+% \end{figure}
+
+% The single- and multi-action classes represent more complex stencils
+% with a real possibility for programmer error. As an extreme example, the Unified
+% Model has an implementation of the Smagorinsky subgrid-scale model for
+% calculating turbulence on which our inference yields 39 specs from 340
+% lines of code. This is a large reduction given
+% the complexity of the algorithm.  We show one example
+% in~\Cref{fig:smagorinsky} which specifies the access pattern to
+% a 3-dimensional array (which we have renamed to
+% \texttt{x}) arising from a kernel of 93 lines of code involving 142
+% array subscripts. The specification was the most complex
+% seen in our corpus, yet it still represents a significant abstraction
+% of the spatial behaviour given size and complexity of the kernel it describes.
+
+We measured the frequency at which individual specifications involved
+multiple occurences of \term{*}
+ % and \term{+}:
+\begin{center}
+\vspace{0.25em}
+\setlength{\tabcolsep}{0.57em}
+{\small{
+\hspace{-1em}\begin{tabular}{c|cccccccc}
+& 0 & 1 & 2 & 3 & 4 & 5 & 6  \\ \hline
+\term{*} & \num{\overallmulOpszero} & \num{\overallmulOpsone} & \num{\overallmulOpstwo} & \num{\overallmulOpsthree} & \num{\overallmulOpsfour} & \num{0} & \num{\overallmulOpssix}
+\end{tabular}}}
+\vspace{0.2em}
+\end{center}
+\noindent
+% \eg{}, roughly half the specifications did not involve $\term{*}$,
+% a quarter use one $\term{*}$ and just under a quarter
+% use two $\term{*}$ operators.
+
+\paragraph{Limitations}
+
+There were various reasons why we did not infer
+specifications on every looped array computation: \\
+\noindent
+1) \textbf{Non-subset induction variables} occur when the
+induction variables on the RHS are not a subset of those in the LHS. These
+cases are not stencils by our definition. The degenerate case of this
+is to have only constant indices on the LHS. We see lots of examples
+of this in loops as accumulators \eg{} computing the sum over an array; \\
+\noindent
+2) \textbf{Derived induction variables} where an
+index (\mintinline{fortran}{x}) is derived from an
+induction variable (\mintinline{fortran}{i}) as in
+\mintinline{fortran}{x = len - i};  \\
+\noindent
+3) \textbf{Inconsistent induction dimensions} occur when
+an induction variable is used to specify more than one array dimension
+on the RHS or multiple induction variables are used for the same
+dimension on the RHS. These are common in matrix operations such as
+LU-decomposition with assignments such as
+\mintinline{fortran}{a(l) = a(l) - a(m) * b(l, m)}.
+
+\subsection{Detecting errors in the 2-D Jacobi iteration}
+\label{sec:jacobi}
+One common example of a stencil computation is the two-dimensional
+Jacobi iteration that repeatedly goes through each cell in a matrix
+and computes the average value of the four adjacent cells. The kernel
+is given by:
+\begin{minted}{fortran}
+  a(i,j) = (a(i-1,j)+a(i+1,j)+a(i,j+1)+a(i,j-1))/4
+\end{minted}
+We infer a precise specification of its shape as:
+\begin{minted}[breakafter=+:,breakindent=-0.6em,breaksymbolsep=0.4em,linenos=false,xleftmargin=-0.5em]{fortran}
+  != stencil pointed(dim=1)*centered(depth=1,dim=2,nonpointed)+pointed(dim=2)*centered(depth=1,dim=1,nonpointed)::a
+\end{minted}
+%
+To test our implementation,
+we examined whether programmer errors would be detected by replacing
+the array index offsets with $-1$, $0$, or $1$ and running our
+verification algorithm. Our checking procedure correctly reported a verification
+failure in each of 6,537 permutations corresponding to an error.
+The iteration computes the average of four adjacent
+cells so $24$ ($4$ factorial) of the possible array index perturbations
+are correct, all of which are accepted by our checker.

### 2017-04-18 oopsla17/evaluation.tex
verification case studies

@@
 \input{results}
+
+\newcommand\regname[1]{\texttt{#1}}
+\newcommand\pointed{\regname{pointed}}
+\newcommand\forward{\regname{forward}}
 
@@
 \textbf{All pointed} \num{\overalljustPointed} of the stencils we
-found involved only pointed regions.
+found involved only \pointed\ regions.
 % 39,681 of these were pointed in all dimensions.
@@
 forward, backward, or centered region constant combined via \term{+}
-or \term{*} with any number of pointed regions. We identified \num{\overallsingleAction} single-action
+or \term{*} with any number of \pointed\ regions. We identified \num{\overallsingleAction} single-action
 specifications, of which \num{\overallsingleActionIrr} were single-action with a
@@
  forward, backward, or centered regions, combined with
-any number of pointed regions. We identified \num{\overallmultiAction} multi-action
+any number of \pointed\ regions. We identified \num{\overallmultiAction} multi-action
 specifications out of which \num{\overallmultiActionMulOnly} had regions combined only with
@@
 are correct, all of which are accepted by our checker.
+
+\subsection{Verification case studies}
+\label{sec:case-studies}
+
+\subsubsection{Catching bugs}
+
+One problem we face while examining the millions of line of source
+code is that we do not have the original programmer's intention at
+hand. But sometimes we have the next best thing, the log entry in the
+revision control system. Scanning through the UM we identified an
+example of an off-by-1 error that could have been avoided through the
+use of stencil specification. The buggy code was identified as
+\pointed\ in all dimensions:
+
+\begin{minted}[breakafter=+:,breakindent=-0.6em,breaksymbolsep=0.4em,linenos=false,xleftmargin=-0.5em]{fortran}
+  !=stencil readOnce, (pointed(dim=1))*(pointed(dim=2))*(pointed(dim=3))
+\end{minted}
+
+But the programmer made it clear in the comments after the bug fix
+that the intention had been \forward\ in the third dimension:
+
+\begin{minted}[breakafter=+:,breakindent=-0.6em,breaksymbolsep=0.4em,linenos=false,xleftmargin=-0.5em]{fortran}
+  !=stencil readOnce, (forward(depth=1, dim=3, nonpointed))*(pointed(dim=1))*(pointed(dim=2))
+\end{minted}
+
+\subsubsection{Aiding refactoring}
+
+One of the initial goals of CamFort was to provide tools that enable
+optimisation and refactoring without changing behaviour, and the
+stencil specification feature was designed with that in mind. But
+while we were perusing real-world source bases, we realised that there
+were also many cases where CamFort could help with refactoring
+precisely because they required changes in behaviour. For example, a
+relatively common change observed in the logs of revision control for
+various projects is the refactoring of array indices. Either
+re-ordering, adding or deleting indices. Any of these, if not
+perfectly propagated throughout all uses, could result in unexpected
+outcomes. However, CamFort will pick up the difference during its
+stencil specification checking. Ideally, an experienced programmer may
+take advantage of the region variable feature of CamFort in order to
+minimise the number of specifications that need changing and gain the
+full time-saving advantage. But even if not, a quick search for the
+specifications can have them updated quickly and then any code that is
+not updated will trigger an error when checked.
+
+% \subsubsection{Stress-testing CamFort}

### 2017-04-18 oopsla17/algorithms.tex
Fix use of interval
editing work on the model, and theories

@@
-%
+\noindent
 We outline here the procedures for checking conformance
@@
  into sets of index schemes. We outline this analysis
-first (\cref{subsec:analysis}). Note that the analysis
+first (\Cref{subsec:analysis}). Note that the analysis
 can be made arbitrarily more complex and wide-ranging independent
@@
 
+\noindent
 We need to address the problem of representation in the model for index schemes.
@@
 
-In the example, $[-1, 1] \times [0, 0]$ is decomposed into $\interv{-1}{1}{}
-\times \interv{-\infty}{\infty}{}$ and $\interv{-\infty}{\infty}{} \times
-\interv{0}{0}{}$. These are then mapped to \texttt{\stencil{c}{1}{1}{}
-* \stencil{p}{2}{}{}}. Decomposition of $[0, 0] \times [-1, 1]$ is similar. When
-combined we obtain the following specification:
+In the example, $\interv{-1}{1}{} \times \interv{0}{0}{}$ is decomposed into
+$\interv{-1}{1}{} \times \interv{-\infty}{\infty}{}$ and
+$\interv{-\infty}{\infty}{} \times \interv{0}{0}{}$. These are then mapped to
+\texttt{\stencil{c}{1}{1}{} * \stencil{p}{2}{}{}}. Decomposition of
+$\interv{0}{0}{} \times \interv{-1}{1}{}$ is similar. When combined we obtain the
+following specification:
 %

### 2017-04-18 oopsla17/lang.tex
editing the model and lang
editing work on the model, and theories

@@
 \nonterm{region} ::= & \nonterm{rvar} \; \mid \;
-                       \stencil{p}{\dimTy}{}{} \; \\
-& {\qquad\; \mid \; \stencil{f}{\dimTy}{\depthRange}{\;[, \nonpointed]}} \\
-& {\qquad\; \mid \; \stencil{b}{\dimTy}{\depthRange}{\;[, \nonpointed]}} \\
-& {\qquad\; \mid \; \stencil{c}{\dimTy}{\depthRange}{\;[, \nonpointed]}} \\
-& {\qquad\; \mid \; \nonterm{region} \, \term{+}
-  \, \nonterm{region} \; \mid \; \nonterm{region} \; \term{*} \; \nonterm{region}} \\[0.25em]
-%\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{rvar}}
-%\\[0.5em]
-%\nonterm{temporal} ::= \; & \term{dependency} \; (v \; \{ , v \}) [, \texttt{mutual}]
-%  \\[0.5em]
+   \nonterm{rconst} \; \mid \; \nonterm{region} \, \term{+}
+  \, \nonterm{region} \; \mid \; \nonterm{region} \; \term{*} \;
+    \nonterm{region} \\
+\nonterm{rconst} ::= & \stencil{p}{\dimTy}{}{} \; \\
+\mid \; & \stencil{f}{\dimTy}{\depthRange}{\;[, \nonpointed]} \\
+\mid \; & \stencil{b}{\dimTy}{\depthRange}{\;[, \nonpointed]} \\
+\mid \; & \stencil{c}{\dimTy}{\depthRange}{\;[, \nonpointed]} \\[0.25em]
 \nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+\\[-1em]
@@
 
+\noindent
 \Cref{fig:syntax} gives the syntax of stencil specifications, which is
-detailed below. The entry point is the \nonterm{specification} production which
-splits into either a \emph{region declaration} or a \emph{specification
-declaration}.
-Regions comprise \emph{region constants} which are
-combined via region operators \term{+} and \term{*}.
+detailed below. The entry point is the \nonterm{specification}
+production which splits into either a \emph{region declaration} or a
+\emph{specification declaration}.  Regions comprise \emph{region
+  constants} which are combined via
+region operators \term{+} and \term{*}.
 
-Region constants specify a finite interval in a 
-single dimension relative to the origin and are either \term{pointed},
+Region constants (non-terminal \textit{rconst}) specify a finite interval in a 
+single dimension starting at the origin and are either \term{pointed},
 \term{forward}, \term{backward}, or \term{centered}. The region names are
@@
 scheme~\citep{dawson1991finite}.
-
 Each region
@@
 
-Region specifications can be assigned to region variables via
-region declarations. For example, the shape of a
-``\emph{Roberts cross}'' edge-detection convolution~\cite{davis1975survey}
-can be stated:
+Region specifications can be assigned to region variables
+(\emph{rvar}) via region declarations. For example, the shape of a
+``\emph{Roberts cross}'' edge-detection
+convolution~\cite{davis1975survey} can be stated:
 %%

### 2017-04-18 oopsla17/model.tex
editing work on the model, and theories

@@
 We define a lattice model of array access patterns which serves as
-a semantic model of our specification language and model of array
+a semantic model of our specification language and a model of array
 access patterns in source code. This model (1) used to explain the
 meaning of our specifications; (2) as a basis for the inference and
-checking algorithms (\Cref{sec:analysis}); (3) justifies an equational
+checking algorithms (\Cref{subsec:analysis}); (3) justifies an equational
 theory for specifications in the next section; (4) is used to optimise
@@
 \begin{defn}
+  \label{defn:modifiers}
   $\mathsf{Mult}$ and $\mathsf{Approx}$ are parametric labelled variant types
@@
 \begin{figure}[!t]
+ \label{fig:semantics}
 \begin{subfigure}[t]{0.4\textwidth}
@@
   \interp{-}^{\mathsf{c}} & : \textit{rconst} \rightarrow
-                            \region{} \\
+                            \textit{Interval} \\
   \interp{\stencil{p}{$i$}{}{}}^{\mathsf{c}} & =
@@
 \end{subfigure}
-\label{fig:semantics}
 \caption{Semantic model of specifications}
@@
 
-The first four equations of \Cref{subfig:region-model} model region 
-contants. The final two equations model the \term{+} and \term{*}
-operators in terms of the join (union) and meet (intersection)
-of interval schemes. Thus the syntax of regions is modelled as members 
-of $\region{N}$.
+The intermediate interpretation $\interp{-}^{\mathsf{c}}$ of
+\Cref{subfig:region-model} models region contants. This is lifted by
+$\textit{promote}_N$ to interpret region constants in the last
+equation of $\interp{-}_N$. The \term{+} and \term{*} operators are
+modelled in terms of the join (union) and meet (intersection) of
+interval schemes. Thus the syntax of regions is modelled as members of
+$\region{N}$.
 
 We mark in our model the presence of modifiers such as
-\texttt{readOnce} and \texttt{atMost} as introduced in \Cref{}.
-Approximation modifiers are interpreted as injections into the
-$\mathsf{Approx}$ variant by $\interpApprox{}$ in
+\texttt{readOnce} and \texttt{atMost} as introduced in
+\Cref{defn:modifiers}.  Approximation modifiers are interpreted as
+injections into the $\mathsf{Approx}$ variant by $\interpApprox{}$ in
 \Cref{subfig:modifier-model}.  The $\textsf{Approx}$ type corresponds
@@
 
-\subsection{Denotational semantics for array subscripts}
+\begin{restatable}[Equational soundness]{thm}{equationalSoundness}
+  The lattice model is sound with respect to the equational theory. Let $R$ and $S$
+  be $N$-dimensional region terms, then we have
+%
+  \begin{equation*}
+    \forall R, S, N. \; R \equiv S \implies \interp{R}_N = \interp{S}_N
+  \end{equation*}
+\end{restatable}
+
+\begin{restatable}[Approximation soundness]{thm}{approxSoundness}
+  The lattice model is sound with respect to the theory of approximation. Let $R$
+  and $S$ be $N$ dimensional regions, then we have
+%
+  \begin{equation*}
+    \forall R, S, N. \; R \preceq S \implies \interp{R}_N \subseteq \interp{S}_N
+  \end{equation*}
+\end{restatable}
+%
+Note that the model is not complete with respect to equations or
+approximations since the specification language has no model of the
+bottom element of the lattice.
+
+\subsection{Interpreting array subscripts}
 \label{sec:fromcode}
@@
 \end{defn}
-
-\subsection{Soundness of model with respect to equational and approximation theories}
-
-\begin{restatable}[Equational soundness]{thm}{equationalSoundness}
-  Lattice model is sound with respect to the equational theory. Let $R$ and $S$
-  be $N$ dimensional regions, then we have
 %
-  \begin{equation*}
-    \forall R, S, N. \; R \equiv S \implies \interp{R}_N = \interp{S}_N
-  \end{equation*}
-\end{restatable}
-
-\begin{restatable}[Approximation soundness]{thm}{approxSoundness}
-  Lattice model is sound with respect to the theory of approximation. Let $R$
-  and $S$ be $N$ dimensional regions, then we have
-%
-  \begin{equation*}
-    \forall R, S, N. \; R \preceq S \implies \interp{R}_N \subseteq \interp{S}_N
-  \end{equation*}
-\end{restatable}
+Note that this produce subscript schemes (a form of index scheme), but
+they are not members of \textit{Interval} or $\textit{Region}_N$. In
+the next section, we use this interpretation on array subscripts in
+the static analysis producer preceding checking and inference.
 

### 2017-04-18 oopsla17/related-work.tex
editing work on the model, and theories

@@
+\noindent
 Various deductive verification tools can express array indexing in their
@@
 
-Our approach differs significantly. Rather than full representation of stencils,
-we focus on specifying just the spatial behaviour in a lightweight way.  Thus,
-it suffices for us to perform a comparatively simple data-flow analysis which is
-efficient, scales linearly with code size, and does not require any user
-intervention.  Whilst we do not perform deep semantic analysis of stencils, the
-analysis part of our approach can be made arbitrarily more sophisticated
-independent of the rest of the work.
+Our approach differs significantly. Rather than full representation of
+array computations, we focus on specifying just the spatial behaviour
+in a lightweight way.  Thus, it suffices for us to perform a
+comparatively simple data-flow analysis which is efficient, scales
+linearly with code size, and does not require any user intervention.
+Whilst we do not perform deep semantic analysis of stencils, the
+analysis part of our approach can be made arbitrarily more
+sophisticated independent of the rest of the work.
 % Hence, we do not require SMT solving or search of loop invariants. It

### 2017-04-18 oopsla17/evaluation.tex
refactor example

@@
 \begin{minted}[breakafter=+:,breakindent=-0.6em,breaksymbolsep=0.4em,linenos=false,xleftmargin=-0.5em]{fortran}
-  !=stencil readOnce, (pointed(dim=1))*(pointed(dim=2))*(pointed(dim=3))
+  != stencil readOnce, pointed(dim=1)*pointed(dim=2)*pointed(dim=3)
 \end{minted}
@@
 \begin{minted}[breakafter=+:,breakindent=-0.6em,breaksymbolsep=0.4em,linenos=false,xleftmargin=-0.5em]{fortran}
-  !=stencil readOnce, (forward(depth=1, dim=3, nonpointed))*(pointed(dim=1))*(pointed(dim=2))
+  != stencil readOnce, forward(depth=1, dim=3, nonpointed)*pointed(dim=1)*pointed(dim=2)
 \end{minted}
@@
 relatively common change observed in the logs of revision control for
-various projects is the refactoring of array indices. Either
-re-ordering, adding or deleting indices. Any of these, if not
+various projects is the refactoring of array dimensions. Either
+re-ordering, adding or deleting dimensions. Any of these, if not
 perfectly propagated throughout all uses, could result in unexpected
@@
 
+An example of a refactored region adding a single dimension to an array, adapted from the UM:
+\begin{minted}[breakafter=+:,breakindent=-0.6em,breaksymbolsep=0.4em,linenos=false,xleftmargin=-0.5em]{fortran}
+  ! before
+  != region :: r = readOnce, pointed(dim=1)
+  ! after
+  != region :: r = readOnce, pointed(dim=1)*pointed(dim=2)
+\end{minted}
+
 % \subsubsection{Stress-testing CamFort}

### 2017-04-18 oopsla17/evaluation.tex
new run and tickLHSvar wording

@@
 found because our analysis restricts the array-subscript-statements
-that we classify as a stencil. In fact, we were able to infer a
-stencil from \overalltickAssignSuccessPercentOfTickAssign\% of these
-statements. A single statement can involve multiple arrays and we
-ended up with \num{\overallnumStencilSpecs} specifications. This shows
-that we can express a large number of stencil shapes within our
-high-level abstraction and validates our initial hypothesis that many
-stencil computations have a regular shape.
+that we classify as a stencil. For example, \num{\overalltickLHSvar}
+potential stencils were discarded because the left-hand side was not
+an array access. Overall, we were able to infer a stencil from
+\overalltickAssignSuccessPercentOfTickAssign\% of the statements
+involving an array assignment. A single statement can involve multiple
+arrays and we ended up with \num{\overallnumStencilSpecs}
+specifications. This shows that we can express a large number of
+stencil shapes within our high-level abstraction and validates our
+initial hypothesis that many stencil computations have a regular
+shape.
 

### 2017-04-18 oopsla17/introduction.tex
edits to the study, intro and appendix
fixing word 1 of introduction from sIn to In!

--- 2017-04-17 oopsla17/introduction.tex
@@
 
+\begin{defn}[Origin]\label{def:origin}
+  An array subscript $a(\bar{e})$ has an \emph{origin index} if
+  all $e \in \bar{e}$ are induction variables, \eg{}, \fortran{a(i,
+    j)}.
+\end{defn}
+
 \begin{defn}[Neighbourhood index]\label{def:neighbour-ix}

### 2017-04-18 oopsla17/introduction.tex
totally fixing a typo

@@
 indexing errors. The language design is informed by an initial
-empirical study of array computations in a corpus real-world
-scientific code base, totally 1.4 million lines of
+empirical study of array computations in a corpus of a real-world
+scientific code base, totaling 1.4 million lines of
 code (\Cref{sec:study}). We confirm our initial hypotheses of the

### 2017-04-18 oopsla17/introduction.tex
'non-trivial' count

@@
 original corpus (\Cref{sec:evaluation}). Our tool identifies and
-infers specifications for roughly 60,000 array computations in the
-corpus. Approximately 5,000 of the computations we found are non-trivial,
-corresponding to code which is a possible source of errors. This
+infers specifications for \num{\overallnumStencilSpecs} array computations in the
+corpus. Of those computations we found, \num{\overallnonTrivial} are non-trivial,
+corresponding to code that we think has a higher potential for errors. This
 validates the design of our language in its capability to capture many

### 2017-04-18 oopsla17/study.tex
editing the model and lang
replace literal with macro

@@
 the above programming idioms can be specified declaratively via a
-small set of combinators, capturing data access patterns.% based on
+small set of combinators, capturing data access patterns. % based on
 %overlapping (hyper)rectangles.
@@
  standard control-flow and data-flow analyses are computed,
-including of key importance for us: induction variable identification
+including some of key importance for us: induction variable identification
 and reaching definitions. The resulting AST is traversed top-down and
@@
 %
-We identified $108,773$ instances of assignments within loops in
+We identified \num{\overalltickAssign} instances of assignments within loops in
 which an array subscript flows to the right-hand side. We refer
@@
 %
-other       &                      & 30019     & 27.60\%  & &  \\\hline%\hline
+other       &                      & 30019     & 27.60\%  & &
+  \\\hline\hline
+ Total & & 36975 & & 71798 & 
 %Total        &                      & 108773    &

### 2017-04-18 oopsla17/study.tex
revert use of macro
use LHS instead of l.h.s.

@@
 %
-We identified \num{\overalltickAssign} instances of assignments within loops in
+We identified \num{108773} instances of assignments within loops in
 which an array subscript flows to the right-hand side. We refer
@@
 %
-The most common category has an l.h.s array subscript
+The most common category has an LHS array subscript
 indexed by induction variables, and a right-hand side comprising

### 2017-04-18 Results.hs
various edits from sections 2-6

@@ -410,6 +410,9 @@ gnuplotHist k xs = concat . mapUpTo k gnuplotHist $ zip [0..] xs
     gnuplotHist (n, v) = show n ++ " " ++ show n ++ " " ++ show v ++ "\n"
 
 
+cropHistogram :: Int -> [Int] -> [Int]
+cropHistogram n xs = take n xs ++ [sum (drop n xs)]
+
 hline' cat dat =
   rline' ((replicate 5 ' ') ++ (show cat)) dat
 

### 2017-04-18 oopsla17/algorithms.tex
various edits from sections 2-6

@@
 
-\subsection{Union Normal Form}\label{subsec:union-normal-form}
-
-\noindent
-We need to address the problem of representation in the model for index schemes.
-In each case, the objects are sets and union and intersection are the
-combinators. However, they are often infinite sets due to \bz{} in Cartesian
-product. This happens when constant indices of array terms are converted to
-\bz{} in subscript schemes and when infinite interval is generated in interval
-schemes. This calls for a finite (and ideally compact) representation for
-indices.
-
-First recall \Cref{lem:vector-intersect}, which states that intersection of
-indexing schemes is still an indexing scheme. When combined with
-\Cref{lem:interval-identities}, we can extend this closure to interval schemes.
-Subscripts schemes do not enjoy the same propery as intersection of two
-singleton sets might easily be empty, which disqualifies it from being a
-subscript index. In the following sections we will not rely on intersection of
-subscript indices, so that is not a problem.
-
-Union, however, is not as flexible as intersection. \Cref{lem:vector-union} does
-not form a closure for indexing schemes in general. For this reason, we
-represent collection of all forms of indexing schemes as unevaluated unions.
-Since the model forms a distributive lattice as established in
-\Cref{prop:region-lattice}, the intersections can be pushed in using the
-distributive law leaving only unions. We further exploit associativity of union
-to put unions into a cons tree making the unions effectively a non-empty list.
-We do not, however, attempt to canonicalise the representation. Later on
-\Cref{lem:vector-union} and index scheme subset are used to reduce the size of
-union normal form.
+\begin{example}
+\label{exm:checking}
+We demonstrate the analysis and modelise procedures over the
+following example kernel which computes the mean value
+of a five-point stencil:
+%%
+\begin{minted}{fortran}
+x = a(i-1, j)+a(i+1, j); y = a(i, j-1)+a(i, j+1)
+b(i, j) = (a(i, j) + x + y) / 5.0
+\end{minted}
+\end{example}
 
@@
 \newcommand{\neigh}{\textsf{neigh}}
-
+\noindent
 The analysis builds on standard program analyses:
@@
 %\arrayTy{} to classify variables of array type.
+
+The analyse then traverse the control-flow graph of a program and,
+traverses statements inside of loops in reverse. The right-hand side
+of an assignment statement is classified on whether it has array
+subscripts which are neighbourhood offsets (or potential absolute in
+some dimensions). These are then converted into our model domain by
+the interpretation $\interp{-}^{\mathit{aterm}}$
+(~\Cref{{defn:subscript}) and grouped into a finite
+  map from array variables to set of subscript schemes.
+
+We denote sets of index/subscripts schemes by
+$\mathcal{S}$, For our example, the initial set of subscript schemes from the analysis
+is given by $\mathcal{S}_0$ as:
+%
+\begin{equation*}
+\mathcal{S}_0 = \{\{0\} \times \{0\}, \{-1\} \times \{0\},
+\{0\} \times \{-1\}, \{0\} \times \{1\}, \{1\} \times \{0\}\}
+\end{equation*}
 
@@
 with \texttt{+} and (2) dimensionality times the number of array terms flowing
-into an assignment assignment. In \Cref{sec:study}, we establish dimensonality
-99\% of all array variables that we can assign a specification is X, and 99\% of
-number of array terms that flows into such an assignment is Y. In
+into an assignment assignment. In \Cref{subsec:additional-data}, we
+established that  99.6\% of array computations have 
+dimensionality of at most four, and 97\% of array computations
+involve at most 4 array subscript terms. In
 \Cref{sec:evaluation}, we show the number of regions that are connected with
@@
 $\mathcal{P}(\bz)^N$. These subscripts are then coalesced iteratively
-into sets of index schemes. We denote sets of index schemes by
-$\mathcal{S}$, and initial set of subscript schemes from the analysis
-as $\mathcal{S}_0$. For our example:
-%
-\begin{equation*}
-\mathcal{S}_0 = \{\{0\} \times \{0\}, \{-1\} \times \{0\},
-\{0\} \times \{-1\}, \{0\} \times \{1\}, \{1\} \times \{0\}\}
-\end{equation*}
+into sets of index schemes.
+
 %

### 2017-04-18 oopsla17/evaluation.tex
'non-trivial' count
various edits from sections 2-6

--- 2017-04-18 oopsla17/evaluation.tex
@@
 optimisation and refactoring without changing behaviour, and the
-stencil specification feature was designed with that in mind. But
+array specification feature was designed with that in mind. But
 while we were perusing real-world source bases, we realised that there
@@
 relatively common change observed in the logs of revision control for
-various projects is the refactoring of array indices. Either
-re-ordering, adding or deleting indices. Any of these, if not
+our corpus is the refactoring of array indices. Either
+re-ordering (swapping dimensions), or adding or deleting indices. Any of these, if not
 perfectly propagated throughout all uses, could result in unexpected

### 2017-04-18 oopsla17/introduction.tex
various edits from sections 2-6

--- 2017-04-18 oopsla17/introduction.tex
@@
 
-Our approach is not targeting a class of bugs that can be detected
+Our approach does not target a class of bugs that can be detected
 automatically (\emph{push-button verification}). Instead, array
@@
 revisions from the commit histories of our corpus where a
-correct specification would have either spotted an error (which was
-latter correct) or would have assisted in an array refactoring task
+correct specification would have spotted an error (which was
+latter corrected) or would have assisted in refactoring 
 (\Cref{sec:case-studies}).

### 2017-04-18 oopsla17/lang.tex
various edits from sections 2-6

@@
 \nonterm{specDec} ::= & \term{stencil} \; \nonterm{spec} \;
-                        \texttt{::} \; v \\
+                        \texttt{::} \; v \; \mid \; \term{access} \;
+                        \nonterm{spec} \; \texttt{::} \; v \\
 \nonterm{regionDec} ::= &  \texttt{region} \; \texttt{::} \; \nonterm{rvar} \; \texttt{=} \;
@@
 
-\subsection{Specification syntax}
-\label{subsec:syntax}
+%\subsection{Specification syntax}
+%\label{subsec:syntax}
 
@@
 != stencil readOnce, backward(depth=2, dim=1) :: a
-b(i+1) = a(i) + a(i-1) + a(i-2)
+b(i) = a(i) + a(i-1) + a(i-2)
 \end{minted}
@@
 
+\paragraph{Relativisation}
+
+The data access pattern of a stencil is understood relative to the
+subscript on the left-hand side. If the left-hand side of a stencil
+has a neighbourhood subscript with non-zero offsets, then the
+right-hand side is understood relative to this shifted origin. For example:
+%%
+\begin{minted}{fortran}
+do i = 1, n
+   != stencil (backward(depth=2, dim=1, nonpointed))*(pointed(dim=2)) :: b
+   a(i+1, j) = b(i, j) + b(i-1, j)
+\end{minted}
+%%
+Note that the specification for dimension 1 is to a depth of 2, since
+the left-hand side is shifted forward by 1.
+
+\paragraph{Access specifications}
+
+Specifications can also be given to array computations which are not
+stencils (\ie{}, do not have a left-hand side which is an array
+subscript). For example, for a reduction:
+%%
+\begin{minted}{fortran}
+do i = 1, n
+  != access, pointed(dim=1) :: a
+  r = max(a(i), r)
+end do
+\end{minted}
+%%
+The main difference between \fortran{stencil} and \fortran{access} is
+that there is no left-hand side with which to check consistency or
+relativise against.
+
 %%% Local Variables:

### 2017-04-18 oopsla17/model.tex
various edits from sections 2-6

@@
 \begin{defn}
+  \label{defn:subscript}
   Recall array subscript terms of the form $a(\bar{e})$ from
@@
       \{ c \} & e \equiv i \pm c \\
-      \bz & e \; \mbox{is constant}
+      \bz & e \; \mbox{is absolute}
     \end{cases}
@@
 
+\subsection{Union Normal Form}\label{subsec:union-normal-form}
+
+\noindent
+We lastly address a problem of \emph{representation} for index schemes
+in the model. Index schemes may incorporate the infinite set \bz{} in
+their components, \eg{} when a dimension is unconstrained or when an
+absolute index of array term is interpreted. In order to provide a
+finite checking procedure, we require a finite and compact
+representation for indices which accounts for this.
+
+Recall \Cref{lem:vector-intersect} states that the
+intersection of indexing schemes is still an indexing scheme. When
+combined with \Cref{lem:interval-identities} (union and intersection
+on holed intervals), we can extend this closure to interval schemes.
+Subscripts schemes do not enjoy the same propery as intersection of
+two singleton sets might easily be empty, which disqualifies it from
+being a subscript index. In the following sections we will not take
+intersection of subscript indices.
+
+Union, however, is not as flexible as
+intersection. \Cref{lem:vector-union} does not form a closure for
+indexing schemes in general. For this reason, we represent sets of
+indexing schemes as unevaluated union terms, or \emph{union normal
+  form}.  Since the model forms a distributive lattice as established
+in \Cref{prop:region-lattice}, the intersections can be pushed inwards
+using the distributive law leaving only unions at the outer-level. We
+further exploit associativity of union to put unions into a cons-tree
+making the unions effectively a non-empty list.  We do not, however,
+attempt to canonicalise the representation. Later on
+\Cref{lem:vector-union} and index scheme subset are used to reduce the
+size of union normal form.
+
 %%% Local Variables:

### 2017-04-18 oopsla17/study.tex
various edits from sections 2-6

--- 2017-04-17 oopsla17/study.tex
@@
 \begin{center}
-\begin{tabular}{l|l|l}
-  Shorthand & Classification & Example \\ \hline
-  (\textsf{const}) & Consistent & \fortran{a(i, j) = b(i, j) +
+\begin{tabular}{l|l}
+  %Shorthand &
+  Classification & Example \\ \hline
+  %(\textsf{const}) &
+  Consistent & \fortran{a(i, j) = b(i, j) +
                                   b(i+1,j+1)} \\
-  (\textsf{perm}) & Permutation & \fortran{a(i, j) = c(j, i)} or \fortran{a(i,
+  %(\textsf{perm})
+  Permutation & \fortran{a(i, j) = c(j, i)} or \fortran{a(i,
                                                                0) =
                                  b(0, i)} \\
-  (\textsf{LHSsub}) & LHS subset & \fortran{a(i) = b(i, j) + b(i, j-1)}
+  %(\textsf{LHSsub}) &
+  LHS subset & \fortran{a(i) = b(i, j) + b(i, j-1)}
   \\
-  (\textsf{LHSsup}) & LHS superset & \fortran{a(i, j) = b(i)} \\
-  (\textsf{inconst}) & Inconsistent & \fortran{a(i) = b(j)}
+  %(\textsf{LHSsup}) &
+  LHS superset & \fortran{a(i, j) = b(i)} \\
+  %(\textsf{inconst}) &
+  Inconsistent & \fortran{a(i) = b(j)}
 \end{tabular}
@@
 %
-If an assignment is not inconsistent, but one of the other four
+If an assignment is not inconsistent, but is one of the other four
 categories, we refer to it as *consistent.
@@
 Other                           & 20196     & 18.57 \%  \\ \hline \hline
-\textbf{Total}                  & 108773    & 100\% \\
+\textbf{Total}                  & 108773    & 100.00\% \\
 \end{tabular}
@@
 offsets (possibly mixed with constants in some dimensions). The
-``affine + absolute'' class is not represented at all. \marginpar{\mnote{(2)
-  Mention what sort of things are included in other.}}
+``affine + absolute'' class is not represented at all. %\marginpar{\mnote{(2)
+%  Mention what sort of things are included in other.}}
 
@@
  \hline \hline
-\textbf{Total}  & 108773    & 100\% \\
+\textbf{Total}  & 108773    & 100.00\% \\
 \end{tabular}
@@
 %
-We subclassified the data based on the right-hand side
-and left-hand side categories
+We subclassified our data by r.h.s
+and l.h.s categories
 being either a variable, or induction-variable indexed,
@@
 LHS Vars & RHS neigh & 0 & 0.00\% & 4591      & 4.22\%
- & \multirow{7}{*}{66.01\%} \\\cline{1-6}
+         & \\\hline
+  LHS Vars & RHS affines & 0 & 0.00\%  & 4         & 0.00\% & 
+  \\\hline
 %
@@
 %
-LHS IVs & RHS neigh   &  4560      & 4.19\%  & 50861     & 46.76\% \\\cline{1-6}
+LHS IVs & RHS neigh   &  4560      & 4.19\%  & 50861     & 46.76\% &
+                                                                     \multirow{4}{*}{61.79\%}
+  \\\cline{1-6}
 %
@@
 %
-LHS Vars & RHS affines & 0 & 0.00\%  & 4         & 0.00\% \\\cline{1-6}
 %
@@
 %
-other       &                      & 30019     & 27.60\%  & &  \\\hline%\hline
+other       &                      & 30019     & 27.60\%  & &
+  \\\hline\hline
+ Total & & 36975 & 34.00\% & 71798 & 66.01\%
 %Total        &                      & 108773    &
@@
 have a static pattern on the left and right-hand side that is either
-affine or a constant translation. Thus, the data supports ours
+a neighbour or affine, and $61.79\%$ we classify as \emph{stencils},
+writing to an array subscript on the left. Thus, the data supports ours
 hypothesis that stencis are common. Based on this grouping, we
@@
 %%
-Most stencils are contiguous, but there is a non-trivial
-amount of non-contiguity. Most stencils have matching induction
+The additional column in the right table (relativised) counts the
+number of specifications where the left-hand side has neighbourhood
+offsets, \eg{}, \fortran{a(i+1) = b(i)}.
+
+Most stencils are contiguous, but there is some
+non-contiguity. Most stencils have matching induction
 variables on the left and right-hand sides but again there
@@
 %
-We classified right-hand sides which are neighbourhood offsets (or
-affine) and subclassified this by whether array subscripts were
-unique or repeated:
-
+We classified right-hand sides comprising neighbourhood (or
+affine) subscripts and subclassified this by whether array subscript terms were
+all unique or whether there were some repeats.
+%
 \begin{center}
@@
 RHS neighbour/affine repeated subscripts   & 10228     & 9.40\%  \\\hline
-other                             & 20196     & 18.57\%  \\\hline\hline
-Total                             & 108773    &
+other                             & 20196     & 18.57\%  %\\\hline\hline
+%Total                             & 108773    &
 \end{tabular}
@@
 %
-Thus, hypothesis 5 is confirmed, the majority of array computations
-have access patterns with linear usage of array subscripts.
+Thus most array computations have access patterns with
+unique use of array subscripts, confirming hypothesis 5.
+
+\subsection{Additional properties of array computations}
+\label{subsec:additional-data}
+
+\noindent
+For each of the $108,773$ array computations, we recorded four
+additional properties: the max dimensionality of array
+subscripts involved, %the maximum depth of neighbourhood offsets,
+the number of subscript terms involved in the computation, and the
+length of the dataflow path (\ie{}, how many intermediate assignments
+contribute to the array computation). The following gives the data
+as histograms:
+
+\setlength{\tabcolsep}{0.25em}
+\begin{center}
+\begin{tabular}{l|ccccccccccc}
+  \textbf{Dimensionality} & 0 & 1     & 2     & 3     & 4    & 5   & 6 & 7 & 8 & 9 & 10 \\\hline
+  Count                   & 0 & 43,227 & 32,830 & 24,621 & 7,608 & 467 & 13 & 5 & 0 & 0 & 3
+  \end{tabular}
+\end{center}
+
+%\begin{center} \begin{tabular}{l|ccccccccccccccccccccccc}
+%   \textbf{Max. offset depth}: & 0      & 1    & 2   & 3  & 4  & 5
+%    & 6  & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 & \ldots & 21 & 22 &
+%    \ldots & 45 & \ldots & 100 \\\hline
+%    Count & 207,512 & 9,530 & 327 & 57 & 55 & 7 & 11 & 0 & 0 & 1 & 11 &
+%    1 & 8  & 4  & 0  & 10 & \ldots & 1  & 1  & \ldots & 8  & \ldots & 2
+%   \end{tabular}
+% \end{center}
+
+   \begin{center}     \begin{tabular}{l|ccccccccccccccccccccc}
+   \textbf{\# subscripts}: & 0 & 1     & 2     & 3    &
+                                                                       4
+     & 5   & 6   & 7   & 8   & 9   & 10  & 11 & 12 & 13 & 14 & 15 & 16
+     & 17 & 18 & 19 & 20$\leq$ \\\hline
+   Count & 0 & 84,690 & 14,377 & 4,234 & 2,367 & 881 & 607 & 192 & 457 & 284 & 207 & 45 & 72 & 33 & 16 & 41 & 40 & 27 & 15 & 4  & 184
+   \end{tabular}
+\end{center}
+
+ \begin{center}
+\begin{tabular}{l|ccccccccccccccccccccc}
+  \textbf{Dataflow length}: & 0      & 1     & 2    & 3
+  & 4    & 5 & 6    & 7   & 8   & 9   & 10  & 11  & 12  & 13  & 14
+  & 15$\leq$ \\\hline
+  Count & 106,712 & 13,323 & 7,091 & 3,898 & 3,031 & 1,695 & 1,333 & 965 & 792 & 704 & 482 & 430 & 339 & 310 & 270 & 2544
+\end{tabular}
+\end{center}
+%
+The histogram of ``number of subscripts'' shows that most
+computations are relatively simple, with only 1-3 array subscripts.
+However, some are complex, with 184 with over 20 indexing terms. We had one example with 96
+contributing array subscripts. Thus, we see that there are many cases
+in practice with complex patterns, that we aim to provide support for
+with our specification language.
+
+The rest of the data here is of use later when considering the
+performance of our checking procedure (\Cref{subsec:checking}).
+
+\subsection{From experiment to language design}
+
+\noindent
+From the data,
+we make the following design decisions:
+%
+\begin{itemize}[leftmargin=1.5em]
+\item The core support should be for neighbourhood offsets on
+right-hand side of assignments in loops ($81.43\%$
+of our array computations);
+
+\item Access patterns are mostly contiguous ($76.94\%$) and cross or
+straddle the origin ($75.75\%$), thus we can represent patterns by
+\emph{finite intervals} that meet at, or are next to, the origin.
+These intervals will be given declaratively and abstractly, avoiding
+replicating indexing terms at the specification level.
+
+\item We will support only consistent left and right-hand sides
+(where there are common induction variables on each side) ($66.01\%$)
+
+\item Stencils will be the primary focus $(61.79\%)$ but we will also
+allow specifications on assignments where the l.h.s is a variable
+(\eg{}, in a reduction) ($4.22\%$).
+
+\item There is some non-contiguity ($4.46\%$) which we will specify
+by finite intervals given as approximations that cover a non-contiguous pattern.
+
+\item Affine indexing is very rare $<0.03\%$ so we do not consider it.
+\end{itemize}
+%

### 2017-04-18 oopsla17/analysis/Study.hs

@@ -28,8 +28,8 @@ main = do
         "hists" -> do
           putStrLn $ mapViewTotal "Dimensionality" (histDimensionality result)
              ++ mapViewTotal "Max depth" (histMaxDepth result)
-             ++ mapViewTotal "Number of indexing expressions" (histNumIndexExprs result)
-             ++ rline' "Length of dataflow path"  (hview . histLengthOfDataflow $ result)
+             ++ mapViewTotal "Number of indexing expressions" (M.map (cropHistogram 20) $ histNumIndexExprs result)
+             ++ rline' "Length of dataflow path"  (hview . cropHistogram 20 . histLengthOfDataflow $ result)
 
 
 regroup :: (Ord c, HistogramShow t) => (Cat -> c) -> M.Map Cat t -> M.Map c t

### 2017-04-18 oopsla17/model.tex
broke ref sorted

--- 2017-04-17 oopsla17/model.tex
@@
 meaning of our specifications; (2) as a basis for the inference and
-checking algorithms (\Cref{sec:analysis}); (3) justifies an equational
+checking algorithms (\Cref{sec:algorithms}); (3) justifies an equational
 theory for specifications in the next section; (4) is used to optimise

### 2017-04-18 oopsla17/algorithms.tex
fix typos

--- 2017-04-17 oopsla17/algorithms.tex
@@
 \Cref{lem:interval-identities}, we can extend this closure to interval schemes.
-Subscripts schemes do not enjoy the same propery as intersection of two
+Subscripts schemes do not enjoy the same property as intersection of two
 singleton sets might easily be empty, which disqualifies it from being a

### 2017-04-18 oopsla17/introduction.tex
add missing label

--- 2017-04-18 oopsla17/introduction.tex
@@
 %
+\label{exm:checking}
 The specifications requires that \texttt{u}

### 2017-04-18 oopsla17/introduction.tex
minor rephrasings in intro

--- 2017-04-18 oopsla17/introduction.tex
@@
 example, the following snippet from a Navier-Stokes
-fluid model~\citet{griebel1997numerical} has two arrays are
+fluid model~\cite{griebel1997numerical} has two arrays which are
 read with different data access patterns, across two dimensions, with
@@
 abstract specification language for the data access pattern of
-array-loop computations. This provides a way to rule out bugs due to
-indexing errors. The language design is informed by an initial
+array-loop computations. This provides a way to prevent
+indexing errors and also to capture some of the original high-level intent of the algorithm. The language design is informed by an initial
 empirical study of array computations in a corpus of a real-world
@@
 %
-The specifications requires that \texttt{u}
+These specifications record that \texttt{u}
 is accessed with a centered pattern to depth of 1 in both dimensions
@@
 an extension to CamFort~\cite{camfort}, an open-source program
-analysis tool for Fortran. Specifications are associated to code via
+analysis tool for Fortran. Specifications are associated with code via
 comments, against which the tool checks the code for conformance. This
@@
 
-We fix some terminology for syntactic constructs in our target language.
+In the remainder of this paper we apply the following terminology for syntactic constructs in our target language.
 

### 2017-04-18 oopsla17/model.tex
fix typos

@@
 \begin{restatable}{prop}{regionLattice}
-  \label{prop:regionLattice}
+  \label{prop:region-lattice}
   $(\region{N},\cup,\cap,\subseteq)$ is a bounded distributive lattice with top
@@
 The first four equations of \Cref{subfig:region-model} model region 
-contants. The final two equations model the \term{+} and \term{*}
+constants. The final two equations model the \term{+} and \term{*}
 operators in terms of the join (union) and meet (intersection)
@@
 We mark in our model the presence of modifiers such as
-\texttt{readOnce} and \texttt{atMost} as introduced in \Cref{}.
+\texttt{readOnce} and \texttt{atMost} as introduced in \Cref{sec:lang}.
 Approximation modifiers are interpreted as injections into the

### 2017-04-18 oopsla17/algorithms.tex

--- 2017-04-18 oopsla17/algorithms.tex
@@
 \label{exm:checking}
-We demonstrate the analysis and modelise procedures over the
-following example kernel which computes the mean value
-of a five-point stencil:
+We demonstrate analysis, checking, and inference on the 
+five-point stencil example:
 %%
 \begin{minted}{fortran}
-x = a(i-1, j)+a(i+1, j); y = a(i, j-1)+a(i, j+1)
-b(i, j) = (a(i, j) + x + y) / 5.0
+b(i, j) = (a(i, j) + a(i-1, j) +a (i+1, j) + a(i, j-1) + a(i, j + 1)) / 5.0
 \end{minted}
@@
 
-The analyse then traverse the control-flow graph of a program and,
-traverses statements inside of loops in reverse. The right-hand side
-of an assignment statement is classified on whether it has array
-subscripts which are neighbourhood offsets (or potential absolute in
-some dimensions). These are then converted into our model domain by
-the interpretation $\interp{-}^{\mathit{aterm}}$
-(~\Cref{{defn:subscript}) and grouped into a finite
-  map from array variables to set of subscript schemes.
+The analysis traverses the control-flow graph of a program top-down
+and traverses statements inside of loops bottom-up. The right-hand
+side of any assignment statement in a loop is classified based on
+whether it has array subscripts flowing to it which are neighbourhood
+offsets (or potential absolute in some dimensions). These are then
+converted into our model domain by the interpretation
+$\interp{-}^{\mathit{aterm}}$ (\Cref{defn:subscript}) and grouped
+into a finite map from array variables to set of subscript schemes.
+We denote sets of index/subscripts schemes by
+$\mathcal{S}$.
 
-We denote sets of index/subscripts schemes by
-$\mathcal{S}$, For our example, the initial set of subscript schemes from the analysis
-is given by $\mathcal{S}_0$ as:
+For our example, the set of subscript schemes from the analysis is:
 %
@@
 \end{equation*}
+%
+This set of index schemes is then augmented with multiplicity
+information (\textsf{only} or \textsf{mult}) depending on whether
+subscripts are unique or not in the analysed statement.
+Thus, for each assignment, the analysis generates a map from array
+variables to values in $\mathsf{Mult}(\mathcal{P}(\bz)^N)$.
+
+Any assignment statement from which array subscripts flow to the
+current assignment is marked as visited such that the main analysis
+does not classify it as the root of an array computation.  Thus,
+combined with the bottom-up traversal inside of loops, we assign
+specifications only to the leaves of dataflow paths in a loop.
 
@@
 \label{subsec:checking}
-%
-In the model, we can regard both array indices and specifications as sets of
-points in $N$-dimensional space. The notion of consistency is then intuitively
-whether these two (potentially infinite) sets of points match or not.
-This leads to a simple set based consistency function defined as
+
+\noindent
+Checking verifies the access pattern of an array computation in the
+source language against any associated specifications. Checking
+proceeds by generating a model from a specification and generating a
+model from the source code (above), and comparing them for
+consistency.  Since the model of \Cref{sec:model} interprets both
+array indices and specifications as sets of points in $N$-dimensional
+space. The notion of consistency is then intuitively whether these two
+(potentially infinite) sets of points are equal. This leads to a
+simple notion of consistency, where $\mathit{consistent}(M_C, M_S)$
+tests the consistency of a modal $M_C$ of source code against
+a model $M_S$ of a specification:
 %
@@
 As explained in \Cref{subsec:union-normal-form}, the sets being compared are
-potentially infinite and set comparisons cannot be done by exhaustively
+potentially infinite thus set comparison cannot be done by exhaustively
 comparing the points of one with the other. Instead, we compile the region into
@@
 \texttt{+} is Z in 99\% of the cases.
-
-\mnote{DOMINIC fills in X, Y, and Z}
 

### 2017-04-18 oopsla17/evaluation.tex

--- 2017-04-18 oopsla17/evaluation.tex
@@
-\input{results}
-
 \newcommand\regname[1]{\texttt{#1}}
@@
 relatively common change observed in the logs of revision control for
-various projects is the refactoring of array dimensions. Either
+our corpus is the refactoring of array dimensions. Either
 re-ordering, adding or deleting dimensions. Any of these, if not

### 2017-04-18 oopsla17/introduction.tex
match alignment of stencil examples

--- 2017-04-18 oopsla17/introduction.tex
@@
-s\noindent
+\noindent
 In the sciences, complex models are now almost always expressed as

### 2017-04-18 oopsla17/model.tex

--- 2017-04-18 oopsla17/model.tex
@@
 Recall \Cref{lem:vector-intersect} states that the
-intersection of indexing schemes is still an indexing scheme. When
+intersection of index schemes is an index scheme. When
 combined with \Cref{lem:interval-identities} (union and intersection
-on holed intervals), we can extend this closure to interval schemes.
-Subscripts schemes do not enjoy the same propery as intersection of
-two singleton sets might easily be empty, which disqualifies it from
-being a subscript index. In the following sections we will not take
+on holed intervals), we can extend this closure to \emph{interval} schemes.
+Subscripts schemes do not enjoy the same closure propery as the intersection of
+two singleton sets may be empty, disqualifying the result from 
+being a subscript index. In the following sections, we do not take
 intersection of subscript indices.
 
-Union, however, is not as flexible as
+Union is not as flexible as
 intersection. \Cref{lem:vector-union} does not form a closure for

### 2017-04-18 oopsla17/study.tex

@@
 %
-We identified $108,773$ instances of assignments within loops in
+We identified \num{108773} instances of assignments within loops in
 which an array subscript flows to the right-hand side. We refer

### 2017-04-18 oopsla17/introduction.tex

--- 2017-04-18 oopsla17/introduction.tex
@@
 %
-\begin{minted}[linenos=false]{fortran}
+\begin{minted}[linenos=false,xleftmargin=0em,xrightmargin=0.8em,breakindent=1.9em]{fortran}
 != stencil centered(depth=1, dim=1) :: a

### 2017-04-18 oopsla17/evaluation.tex
reworking eval a bit, adding back + histo

@@
 
-\textbf{Bounded} specifications occured with \num{\overallatMost}
-\texttt{atMost} bounds and \num{\overallatLeast} \texttt{atLeast}, the latter of which were
-always also paired with an upper bound.
-
 % \begin{figure}[t]\begin{minted}[fontsize=\scriptsize,breakindent=0em,linenos=false,xleftmargin=0em,breakafter=)]{fortran}
@@
 We measured the frequency at which individual specifications involved
-multiple occurences of \term{*}
- % and \term{+}:
+multiple occurrences of \term{*}
+and \term{+}:
 \begin{center}
@@
 & 0 & 1 & 2 & 3 & 4 & 5 & 6  \\ \hline
-\term{*} & \num{\overallmulOpszero} & \num{\overallmulOpsone} & \num{\overallmulOpstwo} & \num{\overallmulOpsthree} & \num{\overallmulOpsfour} & \num{0} & \num{\overallmulOpssix}
+\term{*} & \num{\overallmulOpszero} & \num{\overallmulOpsone} & \num{\overallmulOpstwo} & \num{\overallmulOpsthree} & \num{\overallmulOpsfour} &  & \num{\overallmulOpssix} \\
+\term{+} & \num{\overallplusOpszero} & \num{\overallplusOpsone} & \num{\overallplusOpstwo}
 \end{tabular}}}
@@
 \end{center}
-\noindent
+
+\textbf{Bounded} specifications occured with \num{\overallatMost}
+\texttt{atMost} bounds and \num{\overallatLeast} \texttt{atLeast}, the latter of which were
+always also paired with an upper bound.
+
 % \eg{}, roughly half the specifications did not involve $\term{*}$,
@@
 
-\paragraph{Limitations}
+\subsection{Limitations}
 
 There were various reasons why we did not infer
-specifications on every looped array computation: \\
-\noindent
-1) \textbf{Non-subset induction variables} occur when the
+specifications on every looped array computation:
+\begin{itemize}
+\item \textbf{Non-subset induction variables} occur when the
 induction variables on the RHS are not a subset of those in the LHS. These
@@
 is to have only constant indices on the LHS. We see lots of examples
-of this in loops as accumulators \eg{} computing the sum over an array; \\
-\noindent
-2) \textbf{Derived induction variables} where an
+of this in loops as accumulators \eg{} computing the sum over an array;
+
+\item \textbf{Derived induction variables} where an
 index (\mintinline{fortran}{x}) is derived from an
 induction variable (\mintinline{fortran}{i}) as in
-\mintinline{fortran}{x = len - i};  \\
-\noindent
-3) \textbf{Inconsistent induction dimensions} occur when
+\mintinline{fortran}{x = len - i};
+
+\item \textbf{Inconsistent induction dimensions} occur when
 an induction variable is used to specify more than one array dimension
@@
 \mintinline{fortran}{a(l) = a(l) - a(m) * b(l, m)}.
+\end{itemize}
 

### 2017-04-18 oopsla17/study.tex
minor rephrasing of study

--- 2017-04-18 oopsla17/study.tex
@@
 \begin{enumerate}[leftmargin=2em]
-\item Computations involving loops over arrays are common, where an
-  arrays are read from with a static pattern based on constant offsets
+\item Computations involving loops over arrays are common, where
+  arrays are read according to a static pattern based on constant offsets
   from (base or dervied) induction variables;
@@
 %
-These cover approximately 1.4 million lines of physical Fortran code
+These cover approximately 1.4 million lines of Fortran code
 (2.4 million including comments and white space).  The UM and
@@
 %left-hand sides.
-Classifications are computed from all information that flows into an
+Classifications are computed from all the information that flows into an
 assignment. For example, the following decomposes a one-dimensional
@@
 %%x
-Our analysis recognises this as one array computation (rather than three), starting at
+Our analysis recognises this as a single array computation (rather than three), starting at
 line 4, and reading from subscripts \fortran{a(i)}, \fortran{a(i+1)},
-\fortran{a(i-1)}. We thus traverse the body of a loop in reverse,
+\fortran{a(i-1)}. To compute this we traverse the loop body in reverse,
 using reaching-definitions to calculate the array reads that flow
 to a particular assignment. Since \fortran{x} and \fortran{y} flow to
-the line 4, line 2 and line 3 are not classified separately, though multiple
+line 4, the intermediate statements on line 2 and line 3 are not classified separately, though multiple
 assignments can flow to multiple classified statements.
@@
 \textbf{RHS classification} & \textbf{Number} & \% &
-\textit{grouping 1} \\ \hline
+\textit{Grouping 1} \\ \hline
 Affine                          & 34        & 0.03\%
@@
 Since, affine and neighbour patterns are static, we group them
-(\textit{grouping 1}) to see that $81.43\%$ of the array computations
+(\textit{Grouping 1}) to see that $81.43\%$ of the array computations
 (88,577) have a static pattern that comprises constant offsets of
 base/derived induction variables (potential with some absolute
-indices), confirming \emph{hypothesis 1}. Note that affine subscripts
+indices), confirming \emph{Hypothesis 1}. Note that affine subscripts
 are rare, perhaps due to programmers relying on compilers for
@@
 %
-We grouped the data based on subclassifications of \emph{grouping 1}
+We grouped the data based on subclassifications of \emph{Grouping 1}
 in terms of contiguity and positioning.
@@
 ($74.75\%$) or offset by a distance of 1 from the origin
-($2.05\%$). This confirms both hypotheses 2 and 3. Only 152 instances
-of index patterns not immediately next to or over the origin were
+($2.05\%$). This confirms both Hypothesis 2 and Hypothesis 3. Only 152 instances
+of index patterns not immediately next-to or over the origin were
 found.  There is a non trivial amount of non-contiguous arrays access,
@@
 appear very rarely. Thus, we find that $66.01\%$ of array computations
-have a static pattern on the left and right-hand side that is either
-affine or a constant translation. Thus, the data supports ours
-hypothesis that stencis are common. Based on this grouping, we
-considered the two subclassifications of different kinds of
+have a static pattern on the left- and right-hand side that is either
+affine or a constant translation. Thus, the data supports our
+hypothesis that stencils are common. Based on this grouping, we
+considered two subclassifications of different kinds of
 left-right-hand side relationship and contiguity:
@@
 %
-Thus, hypothesis 5 is confirmed, the majority of array computations
+Thus, Hypothesis 5 is confirmed, the majority of array computations
 have access patterns with linear usage of array subscripts.
+
+We used these results to inform the design of our specification
+language. The objective is to capture the salient aspects of the
+programming patterns we identified above whilst remaining independent
+of their implementation (and associated programming errors).

### 2017-04-18 oopsla17/theory.tex

@@
-We now introduce the equational and approximation theory of stencil
-specifications based on $\equiv$ and $\preceq$ in the following two subsections.
+\noindent
+Our region specifications are subject to an equational theory
+$\equiv$, which explains which region specifications are equivalent,
+and a mutually-defined approximation theory $\preceq$ for
+over- and under- approximation on regions.
 
@@
 
-We define an equivalence relation, $\equiv$. The purpose of this relation is to
-allow programmers to write specifications with greater flexibility. It allows
-specifications to be written in various levels of compactness allowing
-optimising for space and better documentation. The relation is defined on
-regions as follows:
+\noindent
+We define an equivalence relation, $\equiv$. The purpose of this
+relation is to allow programmers to write specifications with greater
+flexibility. It allows specifications to be written in various levels
+of compactness allowing for space optimisation, or greater clarity of
+documentation. The relation is defined on regions as follows:
 
@@
 %
-  \item[Dist] \texttt{*} distributed over \texttt{+} and dually
+  \item[Distributioneq] \texttt{*} distributed over \texttt{+} and dually
     \texttt{+} distributes over \texttt{*}, meaning if \textcap{R}, \textcap{S},
@@
   \item[Overlapping pointed] If \textcap{R} is one of \texttt{forward},
-    \texttt{backward}, or \texttt{centered}, then we have
+    \texttt{backward}, or \texttt{centered}, then we have the following:
 %
@@
 \subsection{Approximations}
-
+%
 We define a partial of order of approximations, $\preceq$. The relation is used
@@
 \end{description}
-
-We present few inequalities that can be derived from the axioms and are useful
+%
+We present some derivable inequalities that are useful
 when writing specifications without inference:

### 2017-04-18 oopsla17/model.tex
fix model.tex

@@
 meaning of our specifications; (2) as a basis for the inference and
-checking algorithms (\Cref{subsec:analysis}); (3) justifies an equational
+checking algorithms (\Cref{sec:algorithms}); (3) justifies an equational
 theory for specifications in the next section; (4) is used to optimise
@@
 We mark in our model the presence of modifiers such as
-\texttt{readOnce} and \texttt{atMost} as introduced in
-\Cref{defn:modifiers}.  Approximation modifiers are interpreted as
+\texttt{readOnce} and \texttt{atMost}.  Approximation modifiers are interpreted as
 injections into the $\mathsf{Approx}$ variant by $\interpApprox{}$ in

### 2017-04-18 oopsla17/proofs.tex

@@
 \newcommand{\offP}{T}
+\newcommand\MID{\;\mathrel{\Big|}\;}
 
@@
     \off \; \cap \; \offP &
-    = \{x \mid \bigwedge_{1 \leq i \leq N } x_i \in \pi_i(\off) \}
-      \;\cap\;
-      \{x \mid \bigwedge_{1 \leq i \leq N } x_i \in \pi_i(\offP) \} \\
-    & = \{x \mid \bigwedge_{1 \leq i \leq N }
-      (x_i \in \pi_i(\off) \wedge x_i \in \pi_i(\offP)) \} \\
+    = \left\{x \MID \bigwedge_{1 \leq i \leq N } x_i \in \pi_i(\off) \right\}
+      \;\bigcap\;
+      \left\{x \MID \bigwedge_{1 \leq i \leq N } x_i \in \pi_i(\offP) \right\} \\
+    & = \left\{x \MID \bigwedge_{1 \leq i \leq N }
+      (x_i \in \pi_i(\off) \wedge x_i \in \pi_i(\offP)) \right\} \\
     & = \prod_{i = 1}^{N} \pi_i(\off) \cap \pi_i(\offP)
@@
     \off \; \cup \; \offP &
-    = \{x \mid
-      \bigwedge_{1 \leq i \leq N } x_i \in \pi_i(\off) \}
-      \;\cup\;
-      \{x \mid
-          \bigwedge_{1 \leq i \leq N } x_i \in \pi_i(\offP) \} \\
-    & = \{x \mid
+    = \left\{x \MID
+      \bigwedge_{1 \leq i \leq N } x_i \in \pi_i(\off) \right\}
+      \;\bigcup\;
+      \left\{x \MID
+          \bigwedge_{1 \leq i \leq N } x_i \in \pi_i(\offP) \right\} \\
+    & = \left\{x \MID
           \bigwedge_{\substack{1 \leq i \leq N \\ i \neq k}}
@@
             \pi_i(\offP) \wedge x_k \in \pi_k(\offP)
-        \} \\
-    & = \{x \mid
+        \right\} \\
+    & = \left\{x \MID
           \bigwedge_{\substack{1 \leq i \leq N \\ i \neq k}} x_i \in
@@
             x_k \in \pi_k(\off) \cup \pi_k(\offP)
-        \} \\
+        \right\} \\
         & = \pi_1(\off) \times \cdots \times
@@
     \interv{a}{b}{c} \cap \interv{d}{e}{f} = &
-      \; \{ n \;|\; a \leq n \leq b \wedge (\neg c \implies n \neq 0) \}
+      \; \left\{ n \MID a \leq n \leq b \wedge (\neg c \implies n \neq 0) \right\}
       \;\cap \\
-      & \; \{ n \;|\; d \leq n \leq e \wedge (\neg f \implies n \neq 0) \}
+      & \; \left\{ n \MID d \leq n \leq e \wedge (\neg f \implies n \neq 0) \right\}
       \\
-    = & \; \{ n \;|\; \max \{a,d\} \leq n \leq \min \{b,e\} \wedge (\neg c
-      \vee \neg f \implies n \neq 0) \} \\
-    = & \; \interv{\max{\{a,d\}}}{\min{\{b,e\}}}{c \wedge f}
+    = & \; \left\{ n \MID \max \left\{a,d\right\} \leq n \leq \min \left\{b,e\right\} \wedge (\neg c
+      \vee \neg f \implies n \neq 0) \right\} \\
+    = & \; \interv{\max{\left\{a,d\right\}}}{\min{\left\{b,e\right\}}}{c \wedge f}
   \end{align*}
@@
         \interv{-k}{k}{\mathsf{true}} $ holds. This immediately follows from
-      \Cref{lem:zinf-identities}.
+      \Cref{lem:interval-identities}.
 %
@@
       \vee p3} = \interv{0}{k}{p2} \cup \interv{-k}{0}{p3}$ holds. This
-      immediately follows from \Cref{lem:zinf-identities}.
+      immediately follows from \Cref{lem:interval-identities}.
   \end{description}
@@
   \begin{description}
-    \item[Equivalence] Region equavilance is modeled with set equality and the
+    \item[Equivalence] Region equivalence is modeled with set equality and the
       partial order on regions is modeled with subset relation. Partial orders
@@
       remains the same at these dimensions. Then we use
-      \Cref{lem:zinf-identities} to calculate the intersection. By assumption we
+      \Cref{lem:interval-identities} to calculate the intersection. By assumption we
       have $k \leq l$, so the maximum of the lower bounds is $-k$ and the

### 2017-04-18 oopsla17/model.tex
fix model.tex

@@
 \texttt{readOnce} and \texttt{atMost}.  Approximation modifiers are interpreted as
-injections into the $\mathsf{Approx}$ variant by $\interpApprox{}$ in
+injections into the $\mathsf{Approx}$ variant by $\interpApprox{-}$ in
 \Cref{subfig:modifier-model}.  The $\textsf{Approx}$ type corresponds
@@
 multiplicity modifiers are interpreted as injections in the
-$\mathsf{Mult}$ variant by $\interpMult{}$, corresponding to the to
+$\mathsf{Mult}$ variant by $\interpMult{-}$, corresponding to
 the presence or absence of the \term{readOnce} modifier as shown in

### 2017-04-18 oopsla17/proofs.tex
editing checking and inference

--- 2017-04-17 oopsla17/proofs.tex
@@
 \end{proof}
+
+\closureInference*
+
+\begin{proof}
+  By \Cref{lem:vector-union}, the union of two index schemes is an
+  index scheme if all components are equal except at most one
+  component. Contiguity (\Cref{def:contiguity}) matches
+  this precondition, thus the first part of the union in
+  $\mathit{coalesce}$ produces a set of index schemes. The second part
+  comprises a subset of the original index schemes.
+\end{proof}

### 2017-04-18 oopsla17/evaluation.tex

@@
 
-To study the effectiveness of our approach,
-we built a corpus of over 1 million lines of Fortran code from a
-range of scientific computing packages: The Unified Model (UM)~\cite{um},
-E3ME~\cite{RePEc:aen:journl:2006se-a12}, BLAS~\cite{blas},
-Hybrid4~\cite{GBC:GBC635}, GEOS-Chem~\cite{geos-chem}, Navier (based
-on \cite{griebel1997numerical}), Computational Physics
-ed. 1~\cite{giordano1997computational},
-ARPACK-NG~\cite{arpackng}, MUDPACK~\cite{MUD}, Cliffs~\cite{Cliffs}, and
-SPECFEM3D~\cite{specfem3d}.
-
-We first examined how frequently stencil computations occur. We parsed
-\num{\overalllinesParsed} lines of Fortran code and found that \overalltickAssignPercent\%
-(\num{\overalltickAssign}) of statements have a left-hand side as an
-array subscript on neighbourhood indices. This supports the idea that
-stencil-like computations are common in scientific code.  We then used
-the inference procedure of the previous section to generate
-specifications for stencils in the corpus to assess the design of the
-language.
+To study the effectiveness of our approach, we applied our tool to the
+more than one million lines of source code contained within the corpus
+that we have assembled (\Cref{sec:study}). We first examined how
+frequently stencil computations occur. We parsed
+\num{\overalllinesParsed} lines of Fortran code and found that
+\overalltickAssignPercent\% (\num{\overalltickAssign}) of statements
+have a left-hand side as an array subscript on neighbourhood
+indices. This supports the idea that stencil-like computations are
+common in scientific code.  We then used the inference procedure of
+the previous section to generate specifications for stencils in the
+corpus to assess the design of the language.
 
@@
 \begin{minted}[breakafter=+:,breakindent=-0.6em,breaksymbolsep=0.4em,linenos=false,xleftmargin=-0.5em]{fortran}
-  != stencil pointed(dim=1)*centered(depth=1,dim=2,nonpointed)+pointed(dim=2)*centered(depth=1,dim=1,nonpointed)::a
+  != stencil pointed(dim=1)*centered(depth=1,dim=2,nonpointed)+pointed(dim=2)*centered(depth=1,dim=1,nonpointed)
 \end{minted}

### 2017-04-18 oopsla17/lang.tex

@@
 %
-Not every dimension needs to be specified, \eg{},
-specifications on lines $1$, $3$, and $7$ leave some dimensions
-unspecified which leaves indices in these dimensions unconstrained.
-
-The \term{forward}, \term{backward}, and \term{centered} regions may
-all have an additional attribute \term{nonpointed} which marks absence
-of the origin.  For example, the following is a
-\term{nonpointed} \term{backward} stencil
-%
-\begin{minted}{fortran}
-!= stencil backward(depth=2, dim=1, nonpointed) :: a
-b(i) = a(i-1) + 10*a(i-2)
-\end{minted}
-
 \paragraph{Combining regions}
@@
  $r \term{+} s$ means that indices in the specified code must be
- in consistent with one of $r$ or $s$, or both.
+ inconsistent with one of $r$ or $s$, or both.
 For example, the following \emph{nine-point stencil}

### 2017-04-18 oopsla17/study.tex

--- 2017-04-18 oopsla17/study.tex
@@
 appear very rarely. Thus, we find that $66.01\%$ of array computations
-have a static pattern on the left and right-hand side that is either
+have a static pattern on the left- and right-hand side that is either
 a neighbour or affine, and $61.79\%$ we classify as \emph{stencils},
-writing to an array subscript on the left. Thus, the data supports ours
+writing to an array subscript on the left. Thus, the data supports our
 hypothesis that stencis are common. Based on this grouping, we
-considered the two subclassifications of different kinds of
+considered two subclassifications of different kinds of
 left-right-hand side relationship and contiguity:
@@
 %
+
 Thus most array computations have access patterns with
-unique use of array subscripts, confirming hypothesis 5.
+unique use of array subscripts, confirming Hypothesis 5.
 
@@
 \noindent
-From the data,
-we make the following design decisions:
+We used these results to inform the design of our specification
+language. The objective was to capture the salient aspects of the
+programming patterns we identified above whilst remaining independent
+of their implementation (and associated programming errors). We made the following design decisions:
 %

### 2017-04-18 oopsla17/algorithms.tex

@@
 As explained in \Cref{subsec:union-normal-form}, the sets being compared are
-potentially infinite thus set comparison cannot be done by exhaustively
-comparing the points of one with the other. Instead, we compile the region into
+potentially infinite thus equality cannot be computed by exhaustively
+comparing elements in each set. Instead, we compile the region into
 interval constraints and subscript schemes into membership constraints and pass
@@
 involve at most 4 array subscript terms. In
-\Cref{sec:evaluation}, we show the number of regions that are connected with
-\texttt{+} is Z in 99\% of the cases.
+\Cref{sec:evaluation}, we show the number of regions that are
+connected with no more than two \texttt{+} operations in 100\% of the cases.
 
-\subsection{Infering specification automatically}
+\subsection{Inferring specification automatically}
 \label{subsec:inference}
@@
 We provide an inference procedure for generating specifications from
-code, which are inserted automatically as comments. This helps support
-maintenance of legacy code, and aids understanding of our
-specification language. We illustrate inference using the
-five-point stencil of \Cref{exm:checking}.
+code which are inserted automatically as comments. This supports
+maintenance of legacy code base, and aids understanding of our
+specification language.
 
 The preceding source code analysis converts the concrete syntax of
-array subscripts into a set of subscript schemes in
-$\mathcal{P}(\bz)^N$. These subscripts are then coalesced iteratively
-into sets of index schemes.
-
-%
-%For each array variable $a \in \mathsf{dom}(U)$, the algorithm
-%constructs a set of $n$-dimensional rectangles covering all contiguous
-%groups of schemes in $U(a)$.
-%
-We explain the algorithm in two parts. First, we show how
-\emph{contiguous} index schemes are coalesced into a smaller set of
-index schemes while remaining in union normal form. Then, we explain
-how this smaller unioned indexing schemes are converted to
-specifications.
+array subscripts into a sets of subscript schemes. Inference then has
+two parts. First, ``adjacent'' index schemes are coalesced into a
+smaller sets of index schemes, remaining in union normal
+form. Secondly, the resulting union normal form is converted to the
+specification syntax.
 
 \subsubsection{Covering}
-A covering of (possibly overlapping) intervals is calculated as a
-value of \region{N} by coalescing \emph{contiguous} index schemes
+A covering of (possibly overlapping) intervals is calculated
+by coalescing \emph{contiguous} index schemes
 until a fixed-point is reached.
 
-\newcommand{\contig}[2]{\mathit{contig}(#1, #2)}
-\begin{defn}[Contiguity]
+\newcommand{\contig}[2]{\mathit{adjacent}(#1, #2)}
+\begin{defn}[Adjacent]
   \label{def:contiguity}
-  Two index schemes $S$ and $T$ are \emph{contiguous} written
+  Two index schemes $S$ and $T$ are \emph{adjacent} written
   $\contig{S}{T}$, iff $\pi_i(S) = \pi_i(T)$ for all $1 \leq i \leq N$
@@
 %
-A coalesced set of index schemes is computed by:%
+Given a set of indexing scheme $\mathcal{S}$ and a particular index scheme $S$
+we generate the set of coalescing of $S$ with adjacent index schemes:
+%
+\begin{equation*}
+\mathit{coalesce}(S, \mathcal{S}) = 
+  \{\,S \cup T \mid T \in \mathcal{S} \, \wedge \,
+  \contig{S}{T} \}
+\end{equation*}
+%
+This is then used by the following recursive procedure:
 %
 \begin{align*}
-  \mathit{coalesce}(\mathcal{S}) = \; & \;\;
-  \{\,S \cup T \mid S \in \mathcal{S}, T \in \mathcal{S} \, \wedge \,
-                                        \contig{S}{T} \}
-  \;\;\; \cup \;\;\;
-  \{\,S \in \mathcal{S} \mid \neg \exists T \in \mathcal{S} . \; \contig{S}{T}\}
+  \mathit{coalesceStep}(\mathcal{S}) =
+  \exists T \in \mathcal{S} . 
+  \begin{cases}
+    \{T\} \cup \mathit{coalesceStep}(\mathcal{S} - \{T\}) &
+    \mathit{coalesce}(T, \mathcal{S}) = \emptyset \\
+    \mathit{coalesceStep}(\mathit{coalesce}(T, \mathcal{S}) \cup
+                                        \mathcal{S} - \{T\}) & \textit{otherwise}
+  \end{cases}
 \end{align*}
 %
-Thus, $\mathit{coalesce}(\mathcal{S})$
-comprises the union of the set of coalesced pairs of contiguous
-schemes and the set of index schemes for which there is no other
-contiguous scheme.
+This give a specification rather than an implementation. Our
+implementation represents sets by a list, and so
+$\exists T \in \mathcal{S}$ above corresponds to deconstructing
+the list into its head element $S$ (picking an element from the set).
+If $T$ has no adjacent index schemes, then coalescing is attempted on
+the rest of the set, and $T$ is returned in the result. Otherwise,
+the set of coalesced index scheme is computed and this is passed to
+a recursive called to $\mathit{coalesceStep}$, along with the rest of
+the elements).
+
+The fixed-point of $\textit{coalesceStep}(\mathcal{S})$ is computed to
+give a covering over the initial subscript space.
 %
-\begin{lemma}[Closure]
+\begin{restatable}{lem}{closureInference}
+  \label{lem:closuer-inference}
   For a set of index schemes $\mathcal{S}$, then
-  $\mathit{coalesce}(\mathcal{S})$ is well-defined and is a set of
-  index schemes.
-\end{lemma}
-\begin{proof}
-  By \Cref{lem:vector-union}, the union of two index schemes is an
-  index scheme if all components are equal except at most one
-  component. Contiguity (\Cref{def:contiguity}) matches
-  this precondition, thus the first part of the union in
-  $\mathit{coalesce}$ produces a set of index schemes. The second part
-  comprises a subset of the original index schemes.
-\end{proof}
+  $\mathit{coalesceSet}(\mathcal{S})$ is well-defined and is a set of
+  index schemes in Union Normal Form.
+\end{restatable}
 %
-For our example, the fixed-pointed of $\mathit{coalesce}$ is reached within two
+For our example, the fixed-pointed of $\mathit{coalesceStep}$ is reached within two
 steps:
@@
 \begin{align*}
-  \mathit{coalesce}(\mathcal{S}_0) & =
+  \mathit{coalesceStep}(\mathcal{S}_0) & =
   \{[-1,0] \times [0,0],\,[0,0] \times [-1, 0],\,[0,0] \times
     [0,1],\,[0,1] \times [0,0]\} \\
-  \mathit{coalesce}^2(\mathcal{S}_0) & =
-  \{[-1, 1] \times [0, 0],\,[0, 0] \times [-1, 1]\} = \mathit{coalesce}^3(\mathcal{S}_0)
+  \mathit{coalesceStep}^2(\mathcal{S}_0) & =
+  \{[-1, 1] \times [0, 0],\,[0, 0] \times [-1, 1]\} = \mathit{coalesceStep}^3(\mathcal{S}_0)
 \end{align*}
 %
-\todo{Prove that this is in $Region_N$}
+%\todo{Prove that this is in $Region_N$}
 
@@
 
-Now we convert coalesced indexing schemes into specifications. This happens in
-three stages. First, we see if there is an all infinite indexing scheme. Second,
-we determine if the indexing schemes are interval schemes so that they can be
-represented exactly and if not how it can be altered into interval schemes to be
-represented as approximations. Third, we split up the interval schemes so that
-the region constants can express them. Finally, we convert the interval schemes
-into a joins of meets of region constants.
+\newcommand{\finalSet}{\mathcal{S}_\omega}
+Let the covering indexing scheme from the fixed point of
+$\mathit{coalesceStep}$ on input $\mathcal{S}_0$ be written as
+$\finalSet{}$.  Next, $\finalSet{}$ the is translated into specification
+syntax. This happens in four stages.
+
+\begin{enumerate}[leftmargin=1.5em]
+  \item We check whether $\finalSet{}$
+  is the top indexing scheme $\bz{}^N$. This occurs if absolute indices
+  occur in each of the dimensions in any of the array subscripts,
+  leading to an unconstrained access pattern, which is not represented
+  in our syntax, ending inference.
+
+  \item Otherwise, we determine if the indexing schemes are interval
+  schemes so that they can be represented exactly. It not, they are
+  altered into interval schemes to be represented as approximations.
+
+  \item We split the interval schemes so that the region constants can
+  express them.
+
+  \item Finally, we convert the interval schemes into a joins of meets
+  of region constants.
+\end{enumerate}
 

### 2017-04-18 oopsla17/study.tex
remove spurious comma
some editing in study

@@
 %
-\begin{enumerate}[leftmargin=1.5em]
+\begin{enumerate}[leftmargin=1.76em]
 \item The Unified Model (UM) developed by the UK Met Office
@@
 \begin{center}
-\begin{tabular}{l|l}
+\begin{tabular}{l|l|l}
   %Shorthand &
-  Classification & Example \\ \hline
+  Classification & Sub Classification & Example \\ \hline
   %(\textsf{const}) &
-  Consistent & \fortran{a(i, j) = b(i, j) +
+  Consistent & Direct & \fortran{a(i, j) = b(i, j) +
                                   b(i+1,j+1)} \\
   %(\textsf{perm})
-  Permutation & \fortran{a(i, j) = c(j, i)} or \fortran{a(i,
+  & Permutation & \fortran{a(i, j) = c(j, i)} or \fortran{a(i,
                                                                0) =
@@
   %(\textsf{LHSsub}) &
-  LHS subset & \fortran{a(i) = b(i, j) + b(i, j-1)}
+  & LHS subset & \fortran{a(i) = b(i, j) + b(i, j-1)}
   \\
   %(\textsf{LHSsup}) &
-  LHS superset & \fortran{a(i, j) = b(i)} \\
+  & LHS superset & \fortran{a(i, j) = b(i)} \\ \hline
   %(\textsf{inconst}) &
-  Inconsistent & \fortran{a(i) = b(j)}
+  \multicolumn{2}{l|}{Inconsistent} & \fortran{a(i) = b(j)}
 \end{tabular}
@@
 %
-If an assignment is not inconsistent, but is one of the other four
-categories, we refer to it as *consistent.
 
@@
 & \% {(whole)}
-& \textbf{\# *consistent}
+& \textbf{\# Consistent}
 & \% {(whole)}
@@
 The most common category has an LHS array subscript
-indexed by induction variables, and a right-hand side comprising
+indexed by induction variables, and an RHS comprising
 neighbourhood offsets. Next most common is to have neighbourhood
@@
 appear very rarely. Thus, we find that $66.01\%$ of array computations
-have a static pattern on the left- and right-hand side that is either
-a neighbour or affine, and $61.79\%$ we classify as \emph{stencils},
+have a static pattern on the LHS and RHS that is either
+a neighbourhood or affine, and $61.79\%$ are \emph{stencils},
 writing to an array subscript on the left. Thus, the data supports our
@@
 %
-
 Thus most array computations have access patterns with

### 2017-04-18 oopsla17/theory.tex

@@
 %
-  \item[Distributioneq] \texttt{*} distributed over \texttt{+} and dually
+  \item[Distributioneq] \texttt{*} distributes over \texttt{+} and dually
     \texttt{+} distributes over \texttt{*}, meaning if \textcap{R}, \textcap{S},
@@
 %
-We define a partial of order of approximations, $\preceq$. The relation is used
-in equational theory and provides means of writing more compact lower and upper
+We define a partial of order of approximations, $\preceq$. This relation is used
+in the equational theory and provides a means of writing a more compact lower and upper
 bound specifications. The relation is defined as follows:
@@
 We present some derivable inequalities that are useful
-when writing specifications without inference:
+when writing specifications:
 

### 2017-04-18 oopsla17/model.tex
typos

@@
 a semantic model of our specification language and a model of array
-access patterns in source code. This model (1) used to explain the
-meaning of our specifications; (2) as a basis for the inference and
+access patterns in source code. This model (1) is used to explain the
+meaning of our specifications; (2) provides a basis for the inference and
 checking algorithms (\Cref{sec:algorithms}); (3) justifies an equational
@@
 \emph{index schemes}.  As an initial informal example, consider the
-a stencil computation with loop body \fortran{y(i) = x(i,0) + x(i+1,0)}.
+ stencil kernel \fortran{y(i) = x(i,0) + x(i+1,0)}.
 The access pattern on array \term{x}, relative
 to induction variable \term{i}, is captured
-by a vector of length 2 containing integer sets $\langle{\{0, 1\},
+by a vector containing two integer sets: $\langle{\{0, 1\},
   \mathbb{Z}\}}\rangle$. This describes that, in the first dimension,
@@
 \begin{defn}[Index scheme]
-  An $N$-dimensional \emph{index scheme} is a vector of length $N$ of
+  An $N$-dimensional \emph{index scheme} is a vector of containing $N$
   integer sets, \ie{}, a member of $\mathcal{P}(\bz)^{N}$.
@@
   If the
-  superscript to the interval is omitted it is treated as $\mathsf{true}$.
+  superscript to the interval is omitted it is treated as $\mathsf{true}$ (no hole).
   We also add the distinguished interval $\interv{-\infty}{\infty}{}$, which is simply an
-  alias for \bz{}, but this notation prevents handling infinite interval
-  separately in the following definitions, lemmas, and proofs. Here, $-\infty$
+  alias for \bz{}, but this notation avoids separate handling of the infinite interval
+  in the following definitions, lemmas, and proofs. Here, $-\infty$
   and $\infty$ behave as top and bottom elements to \bz{} respectively.
@@
 We define two specialisations of index schemes:
-\emph{subscript scheme} and \emph{interval scheme}.
+the \emph{subscript scheme} and the \emph{interval scheme}.
 
@@
 %
-The set of regions $\region{N}$ is target of our specification
+The set of regions $\region{N}$ is the target of our specification
 language model, and of the code analysis. We further wrap this in a
@@
 \textsf{exact} when there is no such modifier and \textsf{lower} and
-\textsf{upper} for \term{atLeast} and \term{atMost}. In a similar way,
+\textsf{upper} for \term{atLeast} and \term{atMost} respectively. In a similar way,
 multiplicity modifiers are interpreted as injections in the
@@
 %
-Note that this produce subscript schemes (a form of index scheme), but
-they are not members of \textit{Interval} or $\textit{Region}_N$. In
+Note that this produces subscript schemes (a form of index scheme), but
+these are not members of \textit{Interval} or $\textit{Region}_N$. In
 the next section, we use this interpretation on array subscripts in
@@
 
-Recall \Cref{lem:vector-intersect} states that the
+Recall that \Cref{lem:vector-intersect} states that the
 intersection of index schemes is an index scheme. When
@@
 attempt to canonicalise the representation. Later on
-\Cref{lem:vector-union} and index scheme subset are used to reduce the
+\Cref{lem:vector-union} and the index scheme subset are used to reduce the
 size of union normal form.

### 2017-04-18 oopsla17/oopsla17.tex
create evaluation section and begin to import results
Equational and approximation theories
verification case studies
editing work on the model, and theories
conclusions stub
'non-trivial' count

@@
-\documentclass[acmlarge,review,anonymous]{acmart}\settopmatter{printfolios=true}
+\documentclass[acmlarge,review,anonymous,10pt]{acmart}\settopmatter{printfolios=true}
 
@@
   code and give verification case studies of bug fixes that are
-  detected by our approach. We found roughly 60,000 targets for
-  specification across roughly 1.4 million lines of code, showing the
+  detected by our approach. We found roughly 80,000 targets for
+  specification across roughly 1.4 million lines of code, showing that the
   vast majority of array computations read from arrays in a pattern with
@@
 \maketitle
+
+\input{results}
 
@@
 
-\section{Semantic model}
+\section{Equational \& Approximation Theories}
+\label{sec:theory}
+\input{theory}
+
+\section{Semantic Model}
 \label{sec:model}
@@
 \label{sec:evaluation}
+\input{evaluation}
 
-\subsection{Verification case studies}
-\label{sec:case-studies}
+\section{Related Work}
+\label{sec:related-work}
+\input{related-work}
 
-\subsubsection{Catching bugs}
-
-\subsubsection{Aiding refactoring}
-
-\subsubsection{Stress-testing Camfort}
+\section{Conclusions}
+\label{sec:conclusions}
+\input{conclusions}
 

### 2017-04-18 oopsla17/model.tex
fix typo

@@
 \begin{defn}[Index scheme]
-  An $N$-dimensional \emph{index scheme} is a vector of containing $N$
+  An $N$-dimensional \emph{index scheme} is a vector containing $N$
   integer sets, \ie{}, a member of $\mathcal{P}(\bz)^{N}$.

### 2017-04-18 oopsla17/algorithms.tex
recursive procedure

@@
 the set of coalesced index scheme is computed and this is passed to
-a recursive called to $\mathit{coalesceStep}$, along with the rest of
+a recursive procedure called to $\mathit{coalesceStep}$, along with the rest of
 the elements).

### 2017-04-18 oopsla17/algorithms.tex
spurious 'the'
typos

--- 2017-04-18 oopsla17/algorithms.tex
@@
 indices to be repeated or not. Then it delegates to $\mathit{consistent'}$
-function to check if the points observed in the array terms match the space
+to check if the points observed in the array terms match the space
 defined by the specification. Lower bounds, marked by the \texttt{atLeast}
@@
 these to the \textsc{Z3} SMT solver~\citep{de2008z3} to see if they are equal. The
-query is expressed in quantifier free linear arithmetic, which is decidable.
+query is expressed in quantifier-free linear arithmetic, which is decidable.
 
 Although satisfiability is super-exponential with the length of the formula in
-the worst case~\cite{fischer1974super}, performance of consistency checking
+the worst case~\cite{fischer1974super}, the performance of consistency checking
 using satisfiability is fast in practice as the length of the formula is linearly
-related with two factors: (1) dimensionality times number of regions connected
+related with two factors: (1) dimensionality times the number of regions connected
 with \texttt{+} and (2) dimensionality times the number of array terms flowing
-into an assignment assignment. In \Cref{subsec:additional-data}, we
+into an assignment. In \Cref{subsec:additional-data}, we
 established that  99.6\% of array computations have 
@@
 code which are inserted automatically as comments. This supports
-maintenance of legacy code base, and aids understanding of our
+maintenance of a legacy code base, and aids adoption of the
 specification language.
@@
 The preceding source code analysis converts the concrete syntax of
-array subscripts into a sets of subscript schemes. Inference then has
+array subscripts into sets of subscript schemes. Inference then has
 two parts. First, ``adjacent'' index schemes are coalesced into a
@@
 %
-Given a set of indexing scheme $\mathcal{S}$ and a particular index scheme $S$
-we generate the set of coalescing of $S$ with adjacent index schemes:
+Given a set of indexing schemes $\mathcal{S}$ and a particular index scheme $S$
+we generate a set from coalescing $S$ with adjacent index schemes:
 %
@@
 %
-This give a specification rather than an implementation. Our
+This gives a specification rather than an implementation. Our
 implementation represents sets by a list, and so
@@
   \label{lem:closuer-inference}
-  For a set of index schemes $\mathcal{S}$, then
+  For a set of index schemes $\mathcal{S}$,
   $\mathit{coalesceSet}(\mathcal{S})$ is well-defined and is a set of
@@
 $\mathit{coalesceStep}$ on input $\mathcal{S}_0$ be written as
-$\finalSet{}$.  Next, $\finalSet{}$ the is translated into specification
+$\finalSet{}$.  Next, $\finalSet{}$ is translated into specification
 syntax. This happens in four stages.
@@
 
-If there is all infinite indexing scheme, then we cannot represent it
+If the result is an all infinite indexing scheme, then we cannot represent it
 syntactically. This happens when use of constant indices in one or more array
@@
 
-After handling the trivial case, correct approximation is chosen. Recall
+After handling the trivial case, the correct approximation is chosen. Recall
 \Cref{def:interval-scheme}, an indexing scheme is an interval scheme if each of
 the sets in products are holed intervals. If that is the case, we trivially
-convert the sets in each dimension of indexing schemes to holed interval
+convert the sets in each dimension of the indexing schemes to a holed interval
 representation and proceed to the next stage and output an exact specification
@@
 
-Upper bound is established by \emph{elongating} index schemes in multiple
+The upper bound is established by \emph{elongating} index schemes in multiple
 dimensions such that they become interval schemes. The elongation function is
@@
 
-Since index schemes of the previous stage are all contiguous and form closed
+Since the index schemes from the previous stage are all contiguous and form closed
 intervals, 0 lying between the lower and upper bounds of intervals is sufficient
@@
 schemes and the new interval schemes generated by elongation form the upper
-bound. The reason a lower bound may not be generated is all indexing schemes
-with might need elongation.
+bound. The reason a lower bound may not be generated is that all indexing schemes
+might need elongation.
 
@@
 respectively and \texttt{centered} requires the upper bound to be the negation
-of the lower bound. The solution is to use \Cref{lem:vector-union} to break up
-interval schemes into two producing $\interv{-2}{0}{} \times \interv{-1}{1}{}$
+of the lower bound. The solution is to use \Cref{lem:vector-union} to break this
+interval scheme into two producing $\interv{-2}{0}{} \times \interv{-1}{1}{}$
 and $\interv{0}{1}{} \times \interv{-1}{1}{}$.
@@
 $\interv{-\infty}{\infty}{} \times \interv{0}{0}{}$. These are then mapped to
-\texttt{\stencil{c}{1}{1}{} * \stencil{p}{2}{}{}}. Decomposition of
+\texttt{\stencil{c}{1}{1}{} * \stencil{p}{2}{}{}}. The decomposition of
 $\interv{0}{0}{} \times \interv{-1}{1}{}$ is similar. When combined we obtain the

### 2017-04-18 oopsla17/evaluation.tex
rephrased the UM code mining section

@@
 
-We measured the frequency at which individual specifications involved
-multiple occurrences of \term{*}
-and \term{+}:
+We measured the frequencies of occurrences of the operators \term{*} and \term{+} within stencil specifications inferred from our corpus:
 \begin{center}
@@
 \hspace{-1em}\begin{tabular}{c|cccccccc}
-& 0 & 1 & 2 & 3 & 4 & 5 & 6  \\ \hline
+Occurrences: & 0 & 1 & 2 & 3 & 4 & 5 & 6  \\ \hline
 \term{*} & \num{\overallmulOpszero} & \num{\overallmulOpsone} & \num{\overallmulOpstwo} & \num{\overallmulOpsthree} & \num{\overallmulOpsfour} &  & \num{\overallmulOpssix} \\
@@
 
-One problem we face while examining the millions of line of source
-code is that we do not have the original programmer's intention at
-hand. But sometimes we have the next best thing, the log entry in the
-revision control system. Scanning through the UM we identified an
-example of an off-by-1 error that could have been avoided through the
-use of stencil specification. The buggy code was identified as
-\pointed\ in all dimensions:
+Detecting whether an error could have been prevented by a stencil
+specification is difficult because it requires knowledge of the
+programmer's intent. However, this is occasionally captured in commit
+messages accompanying a code change. For example in the Unified Model
+we identified an example of an off-by-1 error that could have been
+avoided through the use of stencil specification. The buggy code was
+identified as \pointed\ in all dimensions:
 
@@
 while we were perusing real-world source bases, we realised that there
-were also many cases where CamFort could help with refactoring
+were also many cases where CamFort could help with refactorings
 precisely because they required changes in behaviour. For example, a
@@
 outcomes. However, CamFort will pick up the difference during its
-stencil specification checking. Ideally, an experienced programmer may
+stencil specification checking. Ideally, an experienced programmer might
 take advantage of the region variable feature of CamFort in order to

### 2017-04-18 oopsla17/evaluation.tex
non-subset IVs not a problem

@@
 \begin{itemize}
-\item \textbf{Non-subset induction variables} occur when the
-induction variables on the RHS are not a subset of those in the LHS. These
-cases are not stencils by our definition. The degenerate case of this
-is to have only constant indices on the LHS. We see lots of examples
-of this in loops as accumulators \eg{} computing the sum over an array;
-
 \item \textbf{Derived induction variables} where an

### 2017-04-18 oopsla17/algorithms.tex
second part of inference

--- 2017-04-18 oopsla17/algorithms.tex
@@
 
-The preceding source code analysis converts the concrete syntax of
+The program analysis converts the concrete syntax of
 array subscripts into a sets of subscript schemes. Inference then has
@@
 the set of coalesced index scheme is computed and this is passed to
-a recursive procedure called to $\mathit{coalesceStep}$, along with the rest of
+a recursive called to $\mathit{coalesceStep}$, along with the rest of
 the elements).
@@
   For a set of index schemes $\mathcal{S}$, then
-  $\mathit{coalesceSet}(\mathcal{S})$ is well-defined and is a set of
+  $\mathit{coalesceSet}(\mathcal{S})$ is a set of
   index schemes in Union Normal Form.
@@
 $\mathit{coalesceStep}$ on input $\mathcal{S}_0$ be written as
-$\finalSet{}$.  Next, $\finalSet{}$ the is translated into specification
-syntax. This happens in four stages.
+$\finalSet{}$.  Next, $\finalSet{}$ is translated into specification
+syntax. This happens in three stages.
 
-\begin{enumerate}[leftmargin=1.5em]
+\begin{enumerate}[leftmargin=2em]
   \item We check whether $\finalSet{}$
   is the top indexing scheme $\bz{}^N$. This occurs if absolute indices
-  occur in each of the dimensions in any of the array subscripts,
+  appear in each of the dimension across the array subscripts,
   leading to an unconstrained access pattern, which is not represented
@@
   \item Otherwise, we determine if the indexing schemes are interval
-  schemes so that they can be represented exactly. It not, they are
-  altered into interval schemes to be represented as approximations.
+  schemes (\Cref{def:interval-scheme}) so that they can be represented
+  exactly by our syntax. It not, they are altered into interval
+  schemes to be represented as approximations.
 
-  \item We split the interval schemes so that the region constants can
-  express them.
+  \item Interval schemes into a joins of meets
+  of region constants, where some intervals are split into separate
+  region constants.
+\end{enumerate}
+%
+An index scheme is an interval scheme (\Cref{def:interval-scheme}) if
+each of vector can be represented as a vector of holed intervals
+$\interv{a}{b}{c}$.  If so, we convert each index scheme to an
+interval scheme as a vector of holed interval representation.
 
-  \item Finally, we convert the interval schemes into a joins of meets
-  of region constants.
-\end{enumerate}
-
-If there is all infinite indexing scheme, then we cannot represent it
-syntactically. This happens when use of constant indices in one or more array
-terms causes the offsets to be subsumed. The simplest case is when there is an
-array term with all of its indices constant.
-
-In our example, there are no constant indices, so we proceed.
-
-After handling the trivial case, correct approximation is chosen. Recall
-\Cref{def:interval-scheme}, an indexing scheme is an interval scheme if each of
-the sets in products are holed intervals. If that is the case, we trivially
-convert the sets in each dimension of indexing schemes to holed interval
-representation and proceed to the next stage and output an exact specification
-at the end. If not, then the output will definitely have an upper bound and
-potentially a lower bound.
-
-Upper bound is established by \emph{elongating} index schemes in multiple
-dimensions such that they become interval schemes. The elongation function is
-given as
+Since index schemes of the previous stage are all contiguous and form
+closed intervals, a lower bound $\leq 0$ and an upper bound $\geq 0$
+is sufficient to form an interval scheme.  Otherwise, an approximation
+specification is generated. An upper bound is established by
+\emph{elongating} any index schemes in multiple dimensions such that
+they become interval schemes. This is given by the function:
 %
@@
 \end{align*}
+%
+If any dimension of an index scheme is elongated, then the whole index
+scheme is elongated. In $\finalSet$, all index schemes that are
+representable as interval schemes form a lower bound, whilst those
+that do not produce an upper bound by their elongation. A lower bound
+may not be generated if all indexing schemes are not interval schemes.
+In the case of our example, the resulting $\finalSet$ comprises a set
+of interval schemes, thus we can generate an exact specification.
 
-Since index schemes of the previous stage are all contiguous and form closed
-intervals, 0 lying between the lower and upper bounds of intervals is sufficient
-to form an interval scheme. If any dimension of an index scheme is elongated,
-then the whole index scheme is elongated. In the initial set of index schemes,
-all schemes that were already interval schemes form a lower bound, while those
-schemes and the new interval schemes generated by elongation form the upper
-bound. The reason a lower bound may not be generated is all indexing schemes
-with might need elongation.
+\paragraph{Translation from interval schemes}
 
-The index schemes in the example are all interval schemes, so we proceed to the
-next stage noting the output will be exact.
+Recall the model of region constants in \Cref{subfig:region-model}
+where $\interp{-}$ maps to holed intervals which have either $0$ as
+the lower-bound or upper-bound for \fortran{forward} and
+\fortran{backward} respectively, or both $0$ for \fortran{pointed}, or
+$-k$ and $k$ lower and upper bounds for \fortran{centered}. We can
+invert this map to generate region constants from holed
+intervals. However, some intervals computed from the above steps
+may not the constraints of this interpretation,
+\eg{}, $\interv{-2}{1}{} \times \interv{-1}{1}{}$.  By
+\Cref{lem:vector-union}, we can split interval schemes into two into
+the relevant form, for example, producing
+$\interv{-2}{0}{} \times \interv{-1}{1}{}$ and
+$\interv{0}{1}{} \times \interv{-1}{1}{}$. 
+The following function
+$\textit{split}$ iterates over the components of an index scheme,
+splitting apart interval schemes when necessary:
+%
+\begin{align*}
+  %\textit{split}(\emptyset) = \emptyset \quad
+  \textit{split}(\interv{a}{b}{c} \times S) =
+   \begin{cases}
+  \{ \interv{a}{0}{c} \times T \mid T \in \textit{split}(S) \}
+  \cup
+  \{ \interv{0}{b}{c} \times T \mid T \in \textit{split}(S) \}
+  & a < 0 \vee b > 0 \vee (|a| \neq b) \\
+  \{ \interv{a}{b}{c} \times T \mid T \in \textit{split}(S) \}
+\end{cases}
+\end{align*}
+%
+This forms a set of interval schemes to which $\interp{-}^{-1}$ can be
+applied. In the case of our example, no splitting is necessary.
 
-There is no direct translation from an interval scheme such as $\interv{-2}{1}{}
-\times \interv{-1}{1}{}$ to region constants. The reason is region constants
-\texttt{forward} and \texttt{backward} require upper and lower bounds to be 0
-respectively and \texttt{centered} requires the upper bound to be the negation
-of the lower bound. The solution is to use \Cref{lem:vector-union} to break up
-interval schemes into two producing $\interv{-2}{0}{} \times \interv{-1}{1}{}$
-and $\interv{0}{1}{} \times \interv{-1}{1}{}$.
-
-The extended example does not have such representation problem, so the interval
-schemes are left as they are.
-
-The final stage of inference is simply mapping interval schemes into region
-constants. Each interval scheme is connected with \texttt{+}. Individual schemes
-are decomposed using \Cref{lem:vector-intersect} such that each of the resulting
-index schemes have all but one dimensions set to \interv{-\infty}{\infty}{}.
+The final stage of inference is simply mapping interval schemes into
+region constants and combining these. Each translated interval scheme
+is connected with \texttt{+}. Individual schemes are decomposed using
+\Cref{lem:vector-intersect} such that each of the resulting index
+schemes have all but one dimensions set to \interv{-\infty}{\infty}{}.
 Each of these interval schemes are mapped to region constants using
 \Cref{subfig:region-model} and connected with \texttt{*}.
-
+%
+\begin{align*}
+\textit{convert}_N(\mathcal{S}) = \sum_{S \in S} \prod_{1 \leq i \leq N} \interp{\pi_i(S)}^{-1}
+\end{align*}
+%
 In the example, $\interv{-1}{1}{} \times \interv{0}{0}{}$ is decomposed into
@@
 \texttt{\stencil{c}{1}{1}{} * \stencil{p}{2}{}{}}. Decomposition of
-$\interv{0}{0}{} \times \interv{-1}{1}{}$ is similar. When combined we obtain the
-following specification:
+$\interv{0}{0}{} \times \interv{-1}{1}{}$ is similar. When combined we
+obtain the following specification:
 %

### 2017-04-18 oopsla17/introduction.tex
second part of inference

@@
 read with different data access patterns, across two dimensions, with
-dense index-manipulation code:
+dense index-manipulation:
 %%

### 2017-04-18 oopsla17/lang.tex
second part of inference

@@
 values that flow to the array subscript on the left-hand side,
-and so ranges over the intermediate assignments to \term{x},
+and so ranges over the assignments to \term{x},
 \term{y}, and \term{z}. Each index in the code is consistent
-with both specifications simultaneously, \eg{}, \texttt{a(i-1, j+1)}
+with both parts of the specification, \eg{}, \texttt{a(i-1, j+1)}
 is within the centered region in dimension $1$ and the centered region

### 2017-04-18 oopsla17/algorithms.tex

@@
 The program analysis converts the concrete syntax of
-array subscripts into a sets of subscript schemes. Inference then has
+array subscripts into sets of subscript schemes. Inference then has
 two parts. First, ``adjacent'' index schemes are coalesced into a
@@
 syntax. This happens in three stages.
-
+%
 \begin{enumerate}[leftmargin=2em]
@@
 closed intervals, a lower bound $\leq 0$ and an upper bound $\geq 0$
-is sufficient to form an interval scheme.  Otherwise, an approximation
-specification is generated. An upper bound is established by
+is sufficient to form an interval scheme.  Otherwise, an approximate
+specification is generated.
+
+An upper bound is established by
 \emph{elongating} any index schemes in multiple dimensions such that
-they become interval schemes. This is given by the function:
+they become interval schemes. The elongation function is
+given as
 %
@@
 schemes have all but one dimensions set to \interv{-\infty}{\infty}{}.
+
 Each of these interval schemes are mapped to region constants using
@@
 $\interv{-\infty}{\infty}{} \times \interv{0}{0}{}$. These are then mapped to
-\texttt{\stencil{c}{1}{1}{} * \stencil{p}{2}{}{}}. Decomposition of
-$\interv{0}{0}{} \times \interv{-1}{1}{}$ is similar. When combined we
-obtain the following specification:
+\texttt{\stencil{c}{1}{1}{} * \stencil{p}{2}{}{}}. The decomposition of
+$\interv{0}{0}{} \times \interv{-1}{1}{}$ is similar. When combined we obtain the
+following specification:
 %

### 2017-04-18 oopsla17/related-work.tex

@@
 therefore is similarly prone to indexing errors. Our approach is much more
-abstract-- it does not aim to reify indexing in the specification, but
+abstract---it does not aim to reify indexing in the specification, but
 provides simple spatial descriptions which capture a large number of common

### 2017-04-18 oopsla17/algorithms.tex
edits to evaluation to make it logically more consistent with the study

@@
 %
-The first function, $\mathit{consistent}$, checks whether the linearity of the
+The $\mathit{consistent}$ function checks whether linearity of the
 specification matches that of the indices, \ie{} if the specification allows
-indices to be repeated or not. Then it delegates to $\mathit{consistent'}$
+indices to be repeated or not. It then delegates to $\mathit{consistent'}$
 to check if the points observed in the array terms match the space
-defined by the specification. Lower bounds, marked by the \texttt{atLeast}
-modifier, require the space defined by the specification to remain inside the
-set of indices, while an upper bound, market by \texttt{atMost},
+defined by the specification. Lower bounds, marked \texttt{atLeast},
+require the space defined by the specification to remain inside the
+set of indices, while an upper bound, marked by \texttt{atMost},
 requires the opposite: enclosure. In the absence of such modifiers, we expect
@@
 
-Although satisfiability is super-exponential with the length of the formula in
+Although satisfiability is super-exponential in the length of the formula in
 the worst case~\cite{fischer1974super}, the performance of consistency checking
 using satisfiability is fast in practice as the length of the formula is linearly
-related with two factors: (1) dimensionality times the number of regions connected
+related with two factors: (1) dimensionality times the number of regions composed
 with \texttt{+} and (2) dimensionality times the number of array terms flowing
@@
 established that 99.6\% of array computations have 
-dimensionality of at most four, and 97\% of array computations
+dimensionality at most four and 97\% of array computations
 involve at most 4 array subscript terms. In
-\Cref{sec:evaluation}, we show the number of regions that are
-connected with no more than two \texttt{+} operations in 100\% of the cases.
+\Cref{sec:evaluation}, we show that all specifications in the corpus
+comprises no more than two \texttt{+} operations. Thus, our approach
+using Z3 is practical and efficient in all but corner cases.
 
@@
 An index scheme is an interval scheme (\Cref{def:interval-scheme}) if
-each of vector can be represented as a vector of holed intervals
-$\interv{a}{b}{c}$.  If so, we convert each index scheme to an
-interval scheme as a vector of holed interval representation.
-
-Since index schemes of the previous stage are all contiguous and form
-closed intervals, a lower bound $\leq 0$ and an upper bound $\geq 0$
-is sufficient to form an interval scheme.  Otherwise, an approximate
-specification is generated.
+each vector can be represented as a vector of holed intervals
+$\interv{a}{b}{c}$.  Since index schemes of the previous stage are all
+contiguous and form closed intervals, a lower bound $\leq 0$ and an
+upper bound $\geq 0$ is sufficient to form an interval scheme.
+Otherwise, an approximate specification is generated.
 
@@
 %
-This forms a set of interval schemes to which $\interp{-}^{-1}$ can be
-applied. In the case of our example, no splitting is necessary.
+This forms a set of interval schemes to which $\interp{-}^{-1}$
+(\Cref{subfig:region-model}) is well-defined for each component.
 
-The final stage of inference is simply mapping interval schemes into
-region constants and combining these. Each translated interval scheme
-is connected with \texttt{+}. Individual schemes are decomposed using
-\Cref{lem:vector-intersect} such that each of the resulting index
-schemes have all but one dimensions set to \interv{-\infty}{\infty}{}.
+The final stage of inference, thus maps the component
+of every interval scheme to a region constant via $\interp{-}^{-1}$,
+combining each component with \texttt{*} can combining the
+resulting regions by \texttt{+}, that is:
+%
+%interval schemes to region constants and combines them. %Individual
+%schemes are decomposed using \Cref{lem:vector-intersect} such that
+%each of the resulting index schemes have all but one dimensions set to
+%\interv{-\infty}{\infty}{}.
+%Each of these interval scheme are mapped to region constants using
+%\Cref{subfig:region-model} and connected with \texttt{*}.
+%
+\begin{align}
+\textit{convert}_N(\mathcal{S}) = \sum_{S \in \mathcal{S}} \; \prod_{\pi_i(S) \neq
+  \interv{-\infty}{\infty}{}} \interp{\pi_i(S)}^{-1}
+  \label{eq:convert}
+\end{align}
+%
+Where summation is by the syntactic \term{+} and product is by the
+syntactic \term{*}
 
-Each of these interval schemes are mapped to region constants using
-\Cref{subfig:region-model} and connected with \texttt{*}.
-%
-\begin{align*}
-\textit{convert}_N(\mathcal{S}) = \sum_{S \in S} \prod_{1 \leq i \leq N} \interp{\pi_i(S)}^{-1}
-\end{align*}
-%
-In the example, $\interv{-1}{1}{} \times \interv{0}{0}{}$ is decomposed into
-$\interv{-1}{1}{} \times \interv{-\infty}{\infty}{}$ and
-$\interv{-\infty}{\infty}{} \times \interv{0}{0}{}$. These are then mapped to
-\texttt{\stencil{c}{1}{1}{} * \stencil{p}{2}{}{}}. The decomposition of
-$\interv{0}{0}{} \times \interv{-1}{1}{}$ is similar. When combined we obtain the
-following specification:
+In the example, $\interv{-1}{1}{} \times \interv{0}{0}{}$ is mapped to
+\texttt{\stencil{c}{1}{1}{} * \stencil{p}{2}{}{}}, and
+$\interv{0}{0}{} \times \interv{-1}{1}{}$ is similar. When combined by
+\term{+} we obtain the following specification:
 %
@@
 \end{equation*}
+\vspace{-1em}
+\begin{restatable}[Inference soundness]{lem}{inferenceSoundness}
+  For all region specs $R$, then $\textit{infer}(\interp{R}_N)_N \equiv R$
+\end{restatable}

### 2017-04-18 oopsla17/evaluation.tex
edits to evaluation to make it logically more consistent with the study

@@
 
-To study the effectiveness of our approach, we applied our tool to the
-more than one million lines of source code contained within the corpus
-that we have assembled (\Cref{sec:study}). We first examined how
-frequently stencil computations occur. We parsed
-\num{\overalllinesParsed} lines of Fortran code and found that
-\overalltickAssignPercent\% (\num{\overalltickAssign}) of statements
-have a left-hand side as an array subscript on neighbourhood
-indices. This supports the idea that stencil-like computations are
-common in scientific code.  We then used the inference procedure of
-the previous section to generate specifications for stencils in the
-corpus to assess the design of the language.
-
-We would not expect to infer a stencil for each array statement we
-found because our analysis restricts the array-subscript-statements
-that we classify as a stencil. For example, \num{\overalltickLHSvar}
-potential stencils were discarded because the left-hand side was not
-an array access. Overall, we were able to infer a stencil from
-\overalltickAssignSuccessPercentOfTickAssign\% of the statements
-involving an array assignment. A single statement can involve multiple
-arrays and we ended up with \num{\overallnumStencilSpecs}
-specifications. This shows that we can express a large number of
-stencil shapes within our high-level abstraction and validates our
-initial hypothesis that many stencil computations have a regular
-shape.
+\noindent
+To study the effectiveness of our approach, we applied our tool back
+onto the corpus from \Cref{sec:study}.  We used the inference
+procedure of the previous section to generate array specifications
+for the corpus to assess the design of the language.
+%
+%We first examined how
+%frequently stencil computations occur. We parsed
+%\num{\overalllinesParsed} lines of Fortran code and found that
+%\overalltickAssignPercent\% (\num{\overalltickAssign}) of statements
+%have a left-hand side as an array subscript on neighbourhood
+%indices. This supports the idea that stencil-like computations are
+%common in scientific code.  
+%
+We did not expect to infer specifications for each array computation
+we found because our analysis restricts the
+array-subscript-statements, along the lines of~\Cref{sec:study}.
+%A
+%single statement can involve multiple arrays and therefore can have
+%multiple specifications associated with it.
+Overall we ended up with
+\num{\overallnumStencilSpecs} specifications of which
+\num{\overalltickLHSvar} were with a left-hand side variable rather
+than a subscript (an \term{access} specification), leaving
+\num{\overallStencilsSpecs} \term{stencil} specifications.
+%
+%For example, \num{\overalltickLHSvar}
+%potential stencils were discarded because the left-hand side was not
+%an array access. Overall, we were able to infer a stencil from
+%\overalltickAssignSuccessPercentOfTickAssign\% of the statements
+%involving an array assignment. A single statement can involve multiple
+%arrays and.
+This shows that we can express a large number of
+array access patterns within our high-level abstractions and validates our
+initial design choices informed by our data (~\Cref{subsec:choices}).
 
@@
 
-\textbf{All pointed} \num{\overalljustPointed} of the stencils we
-found involved only \pointed\ regions.
+\textbf{All pointed} \num{\overalljustPointed} of the specifications we
+generated involved only \pointed\ regions.
 % 39,681 of these were pointed in all dimensions.
@@
 
-We measured the frequencies of occurrences of the operators \term{*} and \term{+} within stencil specifications inferred from our corpus:
+We measured the frequencies of occurrences of the operators \term{*}
+and \term{+} within the inferred specifications:
 \begin{center}
@@
 
+\noindent
 There were various reasons why we did not infer
@@
 \begin{itemize}
-\item \textbf{Derived induction variables} where an
-index (\mintinline{fortran}{x}) is derived from an
-induction variable (\mintinline{fortran}{i}) as in
-\mintinline{fortran}{x = len - i};
+\item \textbf{Computed induction variables} where an
+index (\mintinline{fortran}{x}) is computed from an
+induction variable via an intermediate definition, \eg{}, 
+\mintinline{fortran}{x = i + 1}, or using functions that
+we could analyse or classify;
 
-\item \textbf{Inconsistent induction dimensions} occur when
-an induction variable is used to specify more than one array dimension
+\item \textbf{Inconsistent LHS/RHS} occur when
+induction variable is used to specify more than one array dimension
 on the RHS or multiple induction variables are used for the same
@@
 \label{sec:jacobi}
+
+\noindent
 One common example of a stencil computation is the two-dimensional
@@
 \end{minted}
+%
+We found a similar bug in 
+Cliffs\footnote{\url{https://github.com/Delta-function/cliffs-src/commit/e2ca312cfc5398287c7ca8594e977658617c2540}}
+where a subscript \fortran{u(i)} is corrected to \fortran{u(i-1)} which
+could have been detected by an initial specification \fortran{backward(depth = 1,
+  dim = 1, nonpointed)}.
 

### 2017-04-18 oopsla17/proofs.tex
edits to evaluation to make it logically more consistent with the study

@@
 \end{proof}
+
+\inferenceSoundness*
+
+\begin{proof}
+  This follows by induction on the definition of regions syntax
+  $R$. and by definitions of inference and the semantics sharing the
+  same underlying model and using inverse mappings between each other.
+
+  For constants, we map to
+  $\vecgen{N}{i}{\interp{\texttt{\textcap{rconst}}}}$
+  which by \textit{convert}, in \eqref{eq:convert} is mapped
+  back to $\textcap{rconst}$ since \textit{convert} is defined
+  in terms of the inverse mapping $\interp{-}^{-1}$.
+
+  For $\interp{r \term{+} s}_N = \interp{r}_N \vee \interp{s}$
+  we map to a Union Normal Form of interval spaces which is convert
+  back directly by \textit{convert}. Similarly for $\term{*}$.
+\end{proof}

### 2017-04-18 oopsla17/algorithms.tex
space saving

@@
 tests the consistency of a modal $M_C$ of source code against
-a model $M_S$ of a specification:
-%
+a model $M_S$ of a specification, where
+$\mathit{consistent} :
+    \mathsf{Mult}(\mathcal{P}(\bz{}^N)) \times
+    \mathsf{Mult}(\mathsf{Approx}(\region{N})) \to \mathbb{B}$ is defined:
+
 \begin{align*}
-  \mathit{consistent} & :
-    \mathsf{Mult}(\mathcal{P}(\bz{}^N)) \times
-    \mathsf{Mult}(\mathsf{Approx}(\region{N})) \to \mathbb{B} & \\
-%
   \mathit{consistent}(\mathit{ix}, \mathit{model}) & = \begin{cases}
@@
   \end{cases} \\
-%
+\end{align*}
+\begin{align*}
   \mathit{consistent'} & :
@@
   \{ \interv{a}{b}{c} \times T \mid T \in \textit{split}(S) \}
+  & \textit{otherwise}
 \end{cases}
@@
 \end{equation*}
-\vspace{-1em}
+\vspace{-2em}
 \begin{restatable}[Inference soundness]{lem}{inferenceSoundness}

### 2017-04-18 oopsla17/introduction.tex
space saving

@@
   (u(i-1,j)+u(i,j))*(u(i-1,j)+u(i,j))-gamma*abs(u(i-1,j)+u(i,j))*(u(i-1,j)-u(i,j))) /(4.0*delx)
-
 duvdy = ((v(i,j)+v(i+1,j))*(u(i,j)+u(i,j+1))+gamma*abs(v(i,j)+v(i+1,j))*(u(i,j)-u(i,j+1))- &

### 2017-04-18 oopsla17/study.tex
edits to evaluation to make it logically more consistent with the study
space saving

@@
   arrays are read according to a static pattern based on constant offsets
-  from (base or dervied) induction variables;
+  from (base or derived) induction variables;
 
@@
 assignment statements inside loops are analysed and classified.
+CamFort implements standards compliant parsers, and so is not able
+to parse all of the corpus, some of which contains non-standard code.
+Of the 1.4 million lines, over 1 million was parsed
+(see~Appendix~\ref{app:corpus} for exact numbers).
+
+Classifications are computed from all the information that flows into an
+assignment. For example, the following decomposes a one-dimensional
+three-point pattern across multiple intermediate assignments:
+%%
+\begin{minted}{fortran}
+do i = 1, n
+  x = a(i)
+  y = a(i+1)
+  b(i) = (a(i-1) + x + y)/3.0
+end do
+\end{minted}
+%%x
+Our analysis recognises this as a single array computation (rather
+than three), starting at line 4, and reading from subscripts
+\fortran{a(i)}, \fortran{a(i+1)}, \fortran{a(i-1)}. We
+traverse loop bodies bottom-up, using reaching-definitions to
+calculate the array subscripts that flow to an assignment. Since
+\fortran{x} and \fortran{y} flow to line 4, the intermediate
+statements on line 2 and line 3 are not classified separately, though
+multiple assignments can flow to multiple classified statements.
 
@@
 %left-hand sides.
-Classifications are computed from all the information that flows into an
-assignment. For example, the following decomposes a one-dimensional
-three-point pattern across multiple intermediate assignments:
-%%
-\begin{minted}{fortran}
-do i = 1, n
-  x = a(i)
-  y = a(i+1)
-  b(i) = (a(i-1) + x + y)/3.0
-end do
-\end{minted}
-%%x
-Our analysis recognises this as a single array computation (rather than three), starting at
-line 4, and reading from subscripts \fortran{a(i)}, \fortran{a(i+1)},
-\fortran{a(i-1)}. To compute this we traverse the loop body in reverse,
-using reaching-definitions to calculate the array reads that flow
-to a particular assignment. Since \fortran{x} and \fortran{y} flow to
-line 4, the intermediate statements on line 2 and line 3 are not classified separately, though multiple
-assignments can flow to multiple classified statements.
 
@@
 %
-Finally, we categorise the relationship between the left-hand side
-and right-hand side, in terms of the use of induction variables.
-The two sides are \emph{consistent} if the same induction variables
+Finally, we classify the relationship between the LHS
+and RHS in terms of the use of induction variables.
+The two sides are \emph{directly consistent} if the same induction variables
 appear in each side, used for the same dimension. This is
 weakened to a \emph{permutation} if the roles of the induction
-variables changes. This is weakened further if the left-hand side
+variables changes. This is weakened further if the LHS
 induction variables are either
-a subset or superset of the induction variables on the right-hand
-side. Otherwise, the two sides are seen as inconsistent:
+a subset or superset of the induction variables on the RHS.
+Otherwise, the two sides are seen as inconsistent:
 %
@@
   arrays are read from with a static pattern based on constant offsets
-  from (base or dervied) induction variables;
+  from (base or derived) induction variables;
 \end{enumerate}
@@
 (\textit{Grouping 1}) to see that $81.43\%$ of the array computations
-(88,577) have a static pattern that comprises constant offsets of
-base/derived induction variables (potential with some absolute
+(88,577) have a static pattern that neighbourhood or affine subscript
+expressions (potential with some absolute
 indices), confirming \emph{Hypothesis 1}. Note that affine subscripts
@@
 %
-LHS IVs, & RHS affines & 7 & 0.01\% & 0 & 0.00\%  \\\cline{1-6}
+LHS IVs & RHS affines & 7 & 0.01\% & 0 & 0.00\%  \\\cline{1-6}
 %
@@
 writing to an array subscript on the left. Thus, the data supports our
-hypothesis that stencis are common. Based on this grouping, we
+hypothesis that stencils are common. Based on this grouping, we
 considered two subclassifications of different kinds of
@@
 \subsection{From experiment to language design}
+\label{subsec:choices}
 

### 2017-04-18 oopsla17/conclusions.tex
various edits, shrinking, and conclusion

\textbf{Concluding remarks} \hspace{0.25em}
We proposed a novel solution to a common programming task in
numerical and scientific computing. Our approach to language design is
also relatively uncommon-- informed by a quantitative study of
existing code. The resulting language is flexible and expressive,
capturing roughly 80,000 array computations across a million lines of
Fortran with simple, short, abstract specifications. %As
%opposed to existing approaches our language takes inspiration from the
%numerical literature, is derived from a quantitative analysis of
%real-world acode, and provides a concise, abstract description of the
%stencil.
We are now in the
process of applying our tool in collaboration with
the authors' of some of our corpus packages.
Future work is to expand the range of our program analysis, including
options for further semantic analysis and capturing a wider set of
patterns.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2017-04-18 oopsla17/lattice-model.tex
fixing something I broke in the lattice model last night
various edits, shrinking, and conclusion

@@
 
-    Subscript spaces model array subscript expressions in our source langauge
+    Subscript spaces model array subscript expressions in our source language
     where each subscripted expression is either a constant translation of an
@@
     \interpMult{\textcap{mult}}
-           {(\interpMax{\textcap{approx}}
+           {(\interpApprox{\textcap{approx}}
                     {(\interp{\textcap{region}}_N)})}
@@
       $\interp{\textcap{p3}}$ be $p1$, $p2$, and $p3$ respectively. We
-      know that $p1$ is $p2 \vee p3$ from (\textbf{Centered}) detinition. Then
+      know that $p1$ is $p2 \vee p3$ from (\textbf{Centered}) definition. Then
       we get the following interpretation for region constants involved in the
@@
   \begin{description}
-    \item[Equivalence] Region equavilance is modeled with set equality and the
+    \item[Equivalence] Region equivalence is modeled with set equality and the
       partial order on regions is modeled with subset relation. Partial orders
-      are reflexive and the model of eqivalence agrees with that of the partial
+      are reflexive and the model of equivalence agrees with that of the partial
       order, so this rule holds.
@@
     happen to occur uniquely, this does not constitute a failure.
-  \item Definition of vector intervals is completely unwinded into sets of
+  \item Definition of vector intervals is completely expanded into sets of
     offset tuples and depending the approximation modifier (or lack thereof)
@@
 
-\todo{On the following, add some intutition first- what is this and
+\todo{On the following, add some intuition first- what is this and
   what is it for?}

### 2017-04-18 oopsla17/model.tex
second part of inference
various edits, shrinking, and conclusion

@@
   \; a$ etc.
-%  These will be used in the following subsection to give meaning to the
+st%  These will be used in the following subsection to give meaning to the
 %  specification modifiers for approximation and multiplicity.
@@
   \interp{\texttt{\textcap{rconst}}}_N & =
-    \vecgen{N}{i}{\interp{\texttt{\textcap{rconst}}}^{\mathsf{c}}}
+    \vecgen{N}{i}{\interp{\texttt{\textcap{rconst}}}}
   \\[1em]
   %%
-  \interp{-}^{\mathsf{c}} & : \textit{rconst} \rightarrow
+  \interp{-} & : \textit{rconst} \rightarrow
                             \textit{Interval} \\
-  \interp{\stencil{p}{$i$}{}{}}^{\mathsf{c}} & =
+  \interp{\stencil{p}{$i$}{}{}} & =
     {\interv{0}{0}{\mathsf{true}}}\\
 %
-  \interp{\stencil{c}{$i$}{$k$}{\textcap{p}}}^{\mathsf{c}} & =
+  \interp{\stencil{c}{$i$}{$k$}{\textcap{p}}} & =
     {\interv{-k}{k}{\interp{\textcap{p}}}} \\
 %
-  \interp{\stencil{f}{$i$}{$k$}{\textcap{p}}}^{\mathsf{c}} & =
+  \interp{\stencil{f}{$i$}{$k$}{\textcap{p}}} & =
     {\interv{0}{k}{\interp{\textcap{p}}}} \\
 %
-  \interp{\stencil{b}{$i$}{$k$}{\textcap{p}}}^{\mathsf{c}} & =
+  \interp{\stencil{b}{$i$}{$k$}{\textcap{p}}} & =
     {\interv{-k}{0}{\interp{\textcap{p}}}}
@@
 
-The intermediate interpretation $\interp{-}^{\mathsf{c}}$ of
-\Cref{subfig:region-model} models region contants. This is lifted by
+The intermediate interpretation $\interp{-}$ of
+\Cref{subfig:region-model} models region constants. This is lifted by
 $\textit{promote}_N$ to interpret region constants in the last
@@
 on holed intervals), we can extend this closure to \emph{interval} schemes.
-Subscripts schemes do not enjoy the same closure propery as the intersection of
+Subscripts schemes do not enjoy the same closure property as the intersection of
 two singleton sets may be empty, disqualifying the result from 

### 2017-04-18 oopsla17/proofs.tex
various edits, shrinking, and conclusion

@@
   inductively designed to be closed under these operations. Union and
-  intersection are associative, commutative, and absroptive under closed sets
+  intersection are associative, commutative, and absorptive under closed sets
   and then they are also for \region{N}. This is enough to show that, it is a
@@
       partial order on regions is modeled with subset relation. Partial orders
-      are reflexive and the model of eqivalence agrees with that of the partial
+      are reflexive and the model of equivalence agrees with that of the partial
       order, so this rule holds.

### 2017-04-18 oopsla17/related-work.tex
various edits, shrinking, and conclusion

@@
 and others require programmer intervention for correct invariants to be
-inferred.
+inferred. By contrast, it takes roughly 1.5 hours on commodity
+hardware (3.2Ghz Intel Core i5, 16 Gb of RAM) to infer and generate
+stencil specifications for our entire corpus.
 
@@
 algorithms.  Like much of the related work, the goal is optimisation rather than
-verifying program correctness.
+correctness.
 

### 2017-04-18 oopsla17/introduction.tex
various edits, shrinking, and conclusion
section 1 and 2 edits/nits

@@
 sciences. This is partly due to a lack of training and awareness, but
-also a lack of tools targeted at the needs of scientists. This
-paper is part of a line of research on providing lightweight, easy-to-ue
+also a lack of tools targeted at the needs of scientists. We aim to
+change that. This
+paper is part of a line of research providing lightweight, easy-to-use
 verification tools targeted at common programming patterns in science,
@@
 %
-This is a \emph{stencil} computation, a programming idiom
-where the elements of an array at each index $i$ from a
-\emph{neighbourhood} of values around $i$ in some input array(s). This
-pattern is common in scientific, graphical, and numerical
+This is a \emph{stencil} computation, an idiom
+where elements of an array at each index $i$ are computed from a
+\emph{neighbourhood} of values around $i$ in some input array(s). Stencils
+are common in scientific, graphical, and numerical
 code, \eg{}, convolutions in image processing, approximations
@@
 terms. For example, a logical off-by-one-error might manifest itself
-as writing $\texttt{a(i)}$ instead of $\texttt{a(i-1)}$ (we revisit an
-example we found of this in \Cref{sec:case-studies}).
+as writing $\texttt{a(i)}$ instead of $\texttt{a(i-1)}$ (we revisit
+examples we found of this in \Cref{sec:case-studies}).
 Errors also arise by simple lexical mistakes when large amounts of
@@
 common form, reading from arrays in a fixed neighbourhood of
-contiguous elements with a simple static pattern. From this we
+contiguous elements with simple static patterns. From this we
 designed a simple set of combinators to express common array patterns (\Cref{sec:lang}).
-As an example, the Laplace program is specified in our language by:
+As an example, the one-dimensional Laplace program is specified in our language by:
 %
@@
 made when writing its specification. Our specifications are
-abstract to avoid the error-prone nature of index manipulation; our
-specifications are more lightweight to aid its adoption in science.
+abstract to avoid the error-prone nature of index manipulation and
+are lightweight to aid their adoption by scientists.
 
@@
 corpus. Of those computations we found, \num{\overallnonTrivial} are non-trivial,
-corresponding to code that we think has a higher potential for errors. This
+corresponding to code that has a higher potential for errors. This
 validates the design of our language in its capability to capture many
@@
 
-In the remainder of this paper we apply the following terminology for syntactic constructs in our target language.
+We use the following terminology for syntactic constructs in our target language.
 
@@
   induction variable. An \emph{absolute index} is a integer expression
-  which is constant relative to the enclosing loop.
+  which is constant with respect to the enclosing loop.
 \end{defn}
@@
   For an array subscript $a(\bar{e})$ an index $e \in \bar{e}$ is a
-  \emph{neighbourhood index} if $e$ is of the form $e \equiv i$, $e \equiv i +
-  c$, or $e \equiv i - c$, where $c$ is an integer constant. That is, a
-  neighbourhood index is a constant translation of an induction variable. (The
-  relation $\equiv$ identifies terms up-to commutativity of $+$ and the inverse
+  \emph{neighbourhood index} (or \emph{neighbour}) if $e$ is of the
+  form $e \equiv i$, $e \equiv i + c$, or $e \equiv i - c$, where $c$
+  is an integer constant. That is, a neighbourhood index is a constant
+  translation of an induction variable. (The relation $\equiv$
+  identifies terms up-to commutativity of $+$ and the inverse
   relationship of $+$ and $-$ \eg{}, $(-b) + i \equiv i - b$).

### 2017-04-18 oopsla17/oopsla17.tex
various edits, shrinking, and conclusion
section 1 and 2 edits/nits

@@
 
-% For evluation and empirical study
+% For evaluation and empirical study
 % Defines the commands (headers)
@@
   off-by-one errors.  Firstly, we report on a code analysis study on
-  twelve real-world computational science code base, identifying
+  eleven real-world computational science code base, identifying
   common idioms of array usage and their spatial properties. This
@@
 
-\section{Related Work}
+\section{Related Work and Conclusions}
 \label{sec:related-work}
 \input{related-work}
-
-\section{Conclusions}
-\label{sec:conclusions}
 \input{conclusions}

### 2017-04-18 oopsla17/study.tex
various edits, shrinking, and conclusion
section 1 and 2 edits/nits

@@
 
-\item Navier, a small-size Navier-Stokes fluid model (based
+\item Navier, a small-size Navier-Stokes fluid model, based
   on \cite{griebel1997numerical};
@@
 The two sides are \emph{directly consistent} if the same induction variables
-appear in each side, used for the same dimension. This is
+appear in each side, used for the same dimensions. This is
 weakened to a \emph{permutation} if the roles of the induction
@@
 (\textit{Grouping 1}) to see that $81.43\%$ of the array computations
-(88,577) have a static pattern that neighbourhood or affine subscript
-expressions (potential with some absolute
+(88,577) have a static pattern comprising neighbourhood or affine subscript
+expressions (potentially with some absolute
 indices), confirming \emph{Hypothesis 1}. Note that affine subscripts
@@
 affine indices).  Overwhelmingly, the main category is neighbourhood
-offsets (possibly mixed with constants in some dimensions). The
+offsets (possibly mixed with absolute indices in some dimensions). The
 ``affine + absolute'' class is not represented at all. %\marginpar{\mnote{(2)
@@
 &&& \textsf{straddle} & 2228 & 2.05\% \\ \hline
-Neighbour, non-contiguous
- & 4854
- & 4.46\% \\ \hline
+\multirow{3}{*}{Neighbour, non-contiguous}
+& & & \textsf{away}         & 2         & 0.00\%  \\
+& 4854 & 4.46\% & \textsf{origin}       & 4836      & 4.45\%  \\
+& & & \textsf{straddle}     & 16        & 0.01\%  \\ \hline
+%Neighbour, non-contiguous
+% & 4854
+% & 4.46\% \\ \hline
 Other & 20196 & 18.57\%  \\
@@
 found.  There is a non trivial amount of non-contiguous arrays access,
-but it is a much smaller proportion of the whole.
+but it is a much smaller proportion of the whole, and mostly contains
+the origin.
 
@@
 %
-We subclassified our data by r.h.s
-and l.h.s categories
+We subclassified our data by RHS
+and LHS categories
 being either a variable, or induction-variable indexed,
-neighbour-indexed or affined-indexed array. We additionally
+neighbour-indexed or affine-indexed array. We additionally
 included the orthogonal classification of whether
@@
 & \% {(whole)}
-& \textbf{\# Consistent}
+& \textbf{\# consistent}
 & \% {(whole)}
@@
 %
-The most common category has an LHS array subscript
+The most common category is an LHS array subscript
 indexed by induction variables, and an RHS comprising
@@
 offsets, \eg{}, \fortran{a(i+1) = b(i)}.
-
 Most stencils are contiguous, but there is some
@@
 \noindent
-For each of the $108,773$ array computations, we recorded four
-additional properties: the max dimensionality of array
+For each of the $108,773$ array computations, we recorded three
+additional properties: the dimensionality of array
 subscripts involved, %the maximum depth of neighbourhood offsets,
-the number of subscript terms involved in the computation, and the
+the number of subscript terms contributing to the computation, and the
 length of the dataflow path (\ie{}, how many intermediate assignments
-contribute to the array computation). The following gives the data
-as histograms:
+contribute to the array computation):%The following gives the data
+%as histograms:
 
@@
 %
-The histogram of ``number of subscripts'' shows that most
-computations are relatively simple, with only 1-3 array subscripts.
-However, some are complex, with 184 with over 20 indexing terms. We had one example with 96
-contributing array subscripts. Thus, we see that there are many cases
-in practice with complex patterns, that we aim to provide support for
-with our specification language.
+The data for ``number of subscripts'' shows that most computations are
+relatively simple, with only 1-3 array subscripts.  However, some are
+complex, with 184 with over 20 indexing terms. We had one example with
+96 contributing array subscripts. Thus, we see that there are many
+cases in practice with complex patterns. We aim to provide
+support for such code with our specification language.
 
@@
 \item The core support should be for neighbourhood offsets on
-right-hand side of assignments in loops ($81.43\%$
-of our array computations);
+  right-hand side of assignments in loops ($81.43\%$ of our array
+  computations);
 
-\item Access patterns are mostly contiguous ($76.94\%$) and cross or
-straddle the origin ($75.75\%$), thus we can represent patterns by
-\emph{finite intervals} that meet at, or are next to, the origin.
-These intervals will be given declaratively and abstractly, avoiding
-replicating indexing terms at the specification level.
+\item Access patterns are mostly contiguous ($76.94\%$ of all array
+  computations) and cross or straddle the origin ($75.75\%$), thus we
+  can represent patterns by \emph{finite intervals} that meet at, or
+  are next to, the origin.  These intervals will be given
+  declaratively and abstractly, avoiding replicating indexing terms at
+  the specification level.
 
-\item We will support only consistent left and right-hand sides
-(where there are common induction variables on each side) ($66.01\%$)
+\item We will support only consistent left and right-hand sides (where
+  there are common induction variables on each side) ($66.01\%$)
 
 \item Stencils will be the primary focus $(61.79\%)$ but we will also
-allow specifications on assignments where the l.h.s is a variable
+  allow specifications on assignments where the LHS is a variable
   (\eg{}, in a reduction) ($4.22\%$).
 
-\item There is some non-contiguity ($4.46\%$) which we will specify
-by finite intervals given as approximations that cover a non-contiguous pattern.
+\item There is some non-contiguity ($4.46\%$) which we will specify by
+  finite intervals given as approximations that cover a non-contiguous
+  pattern.
 
@@
 \end{itemize}
-%

### 2017-04-18 oopsla17/lang.tex
Access syntax typo
Section 3 consistency with the new semantics

@@
 != stencil backward(depth=2, dim=1) :: b
-e(i, j) = b(i, j) + b(i-1, j) + b(i-2, j)
+f(i) = b(i) + b(i-1) + b(i-2)
 
@@
 %%
-Not every dimension needs to be specified, \eg{},
-specifications on lines $1$, $4$, and $10$ leave some dimensions unspecified.
+If a dimension contains only absolute index terms then it is left
+unspecified.
+
 The \term{forward}, \term{backward}, and \term{centered} regions may
 all have an additional attribute \term{nonpointed} which marks absence
-of the origin.  For example, the following is a
+of the origin, corresponding to those array computations classified
+as \textsf{straddle} in the previous section.  For example, the following is a
 \term{nonpointed} \term{backward} stencil
@@
 \paragraph{Combining regions}
-
 The region operators \term{+} and \term{*} respectively combine
@@
  $r \term{+} s$ means that indices in the specified code must be
- inconsistent with one of $r$ or $s$, or both.
+ consistent with one of $r$ or $s$, or both.
 For example, the following \emph{nine-point stencil}
@@
 (in \textit{spec} in \Cref{fig:syntax}).
-The \texttt{readOnce} modifier enforces that no index appears more
+The \texttt{readOnce} modifier enforces that no subscript appears more
 than once (that is, its multiplicity is one). For example, all of
@@
 The notion is similar to that of linear types~\cite{wadler1990linear}, where a value must be used
-exactly once. We opt for the more informative and easily understood name
-\texttt{readOnce}. This modifier is optional, so it need not
-be present even if the stencil is linear.
+exactly once. We opt for a more informative and easily understood name
+\texttt{readOnce}.
 
@@
 do i = 1, n
-  != access, pointed(dim=1) :: a
+  != access pointed(dim=1) :: a
   r = max(a(i), r)

### 2017-04-18 oopsla17/study.tex
Section 3 consistency with the new semantics

@@
 \item Access patterns are mostly contiguous ($76.94\%$ of all array
-  computations) and cross or straddle the origin ($75.75\%$), thus we
-  can represent patterns by \emph{finite intervals} that meet at, or
+  computations) and cross or straddle the origin ($75.75\%$). We thus
+  represent patterns by \emph{finite intervals} that meet at, or
   are next to, the origin.  These intervals will be given
@@
 \item We will support only consistent left and right-hand sides (where
-  there are common induction variables on each side) ($66.01\%$)
+  there are common induction variables on each side) ($66.01\%$ of all
+  array computations).
 

### 2017-04-18 oopsla17/algorithms.tex
Reword lower bound condition
various edits in section 4-6

@@
 \begin{minted}{fortran}
-b(i, j) = (a(i, j) + a(i-1, j) +a (i+1, j) + a(i, j-1) + a(i, j + 1)) / 5.0
+b(i, j) = (a(i, j) + a(i-1, j) + a(i+1, j) + a(i, j-1) + a(i, j+1)) / 5.0
 \end{minted}
@@
 
-\subsection{Analysis of array accesses}
+\subsection{Static analysis of array accesses}
 \label{subsec:analysis}
@@
 %
-\textbf{(1)} basic blocks (CFG);
-\textbf{(2)} induction variables per basic block;
-\textbf{(3)} (interprocedural) data-flow analysis, providing a \emph{flows to}
-  graph (as shorthand, the function
-  $\mathit{flowTo}$ is used, implicitly parameterised by this graph,
-  mapping an expression to the set of all expressions
-  with forwards data-flow to this expression, based on the transitive
-  closure of the flows graph);
-\textbf{(4)} type information per variable.% where we use the predicate
-%\arrayTy{} to classify variables of array type.
-
-The analysis traverses the control-flow graph of a program top-down
-and traverses statements inside of loops bottom-up. The right-hand
-side of any assignment statement in a loop is classified based on
-whether it has array subscripts flowing to it which are neighbourhood
-offsets (or potential absolute in some dimensions). These are then
-converted into our model domain by the interpretation
-$\interp{-}^{\mathit{aterm}}$ (\Cref{defn:subscript}) and grouped
-into a finite map from array variables to set of subscript schemes.
-We denote sets of index/subscripts schemes by
-$\mathcal{S}$.
+\textbf{(1)} basic blocks (CFG); \textbf{(2)} induction variables per
+basic block; \textbf{(3)} (interprocedural) data-flow analysis,
+providing a \emph{flows to} graph (reaching definitions); and
+\textbf{(4)} type information per variable.  The analysis traverses
+the control-flow graph of a program top-down and traverses statements
+inside of loops bottom-up. The right-hand side of any assignment
+statement in a loop is classified based on whether it has array
+subscripts flowing to it which are neighbourhood offsets (or a
+combination of neighbourhood and absolute in some dimensions). These
+are then converted into our model domain by the interpretation
+$\interp{-}^{\mathit{aterm}}$ (\Cref{defn:subscript}) and grouped into
+a finite map from array variables to set of subscript schemes.  We
+denote sets of index/subscripts schemes by $\mathcal{S}$.
 
@@
 current assignment is marked as visited such that the main analysis
-does not classify it as the root of an array computation.  Thus,
-combined with the bottom-up traversal inside of loops, we assign
-specifications only to the leaves of dataflow paths in a loop.
+does not classify it as the root of an array computation.  In the
+inference procedure, we assign specifications only to these array
+computation roots.
 
@@
 array indices and specifications as sets of points in $N$-dimensional
-space. The notion of consistency is then intuitively whether these two
+space, the notion of consistency is then intuitively whether these two
 (potentially infinite) sets of points are equal. This leads to a
-simple notion of consistency, where $\mathit{consistent}(M_C, M_S)$
-tests the consistency of a modal $M_C$ of source code against
+simple notion of consistency $\mathit{consistent}(M_C, M_S)$ which
+tests the consistency of a model $M_C$ of source code against
 a model $M_S$ of a specification, where
@@
 \begin{align*}
-  \mathit{consistent}(\mathit{ix}, \mathit{model}) & = \begin{cases}
+  \mathit{consistent}(\mathit{ixs}, \mathit{model}) & = \begin{cases}
     \mathsf{false} & \mathit{model} = \mathsf{once}(x) \wedge ix =
@@
     \mathit{consistent'}(\mathit{peel}(ix), \mathit{peel}(model)) & \textit{otherwise}
-  \end{cases} \\
+  \end{cases}
 \end{align*}
+
+with the intermediate
+$\mathit{consistent'} : \mathcal{P}(\bz{}^N) \times
+\mathsf{Approx}(\region{N}) \to \mathbb{B}$ defined as:
+
 \begin{align*}
-  \mathit{consistent'} & :
-    \mathcal{P}(\bz{}^N) \times
-    \mathsf{Approx}(\region{N}) \to \mathbb{B} & \\
-%
-  \mathit{consistent'}(\mathit{ix}, \mathit{model}) & = \begin{cases}
-    m = \mathit{ix} & \mathit{model} = \mathsf{exact}(m) \\
-    m \supseteq \mathit{ix} & \mathit{model} = \mathsf{upper}(m) \\
-    m \subseteq \mathit{ix} & \mathit{model} = \mathsf{lower}(m)
+  \mathit{consistent'}(\mathit{ixs}, \mathit{model}) & = \begin{cases}
+    m = \mathit{ixs} & \mathit{model} = \mathsf{exact}(m) \\
+    m \supseteq \mathit{ixs} & \mathit{model} = \mathsf{upper}(m) \\
+    m \subseteq \mathit{ixs} & \mathit{model} = \mathsf{lower}(m)
   \end{cases}
@@
 We provide an inference procedure for generating specifications from
-code which are inserted automatically as comments. This supports
-maintenance of a legacy code base, and aids adoption of the
-specification language.
+code which are inserted automatically as comments for the root
+array computations in a loop. This supports maintenance of a legacy
+code base, and aids adoption of the specification language.
 
@@
 A covering of (possibly overlapping) intervals is calculated
-by coalescing \emph{contiguous} index schemes
+by coalescing \emph{adjacent} index schemes
 until a fixed-point is reached.
@@
 the set of coalesced index scheme is computed and this is passed to
-a recursive procedure called to $\mathit{coalesceStep}$, along with the rest of
+a recursive procedure called $\mathit{coalesceStep}$, along with the rest of
 the elements).
@@
   $\mathit{coalesceSet}(\mathcal{S})$ is a set of
-  index schemes in Union Normal Form.
+  index schemes in union normal form.
 \end{restatable}
@@
 
-  \item Interval schemes into a joins of meets
-  of region constants, where some intervals are split into separate
+  \item Interval schemes are mapped into a joins of meets
+  of region constants, where some intervals split into multiple separate
   region constants.
@@
 $\interv{a}{b}{c}$.  Since index schemes of the previous stage are all
-contiguous and form closed intervals, a lower bound $\leq 0$ and an
+adjacent and form closed intervals (in the usual mathematical sense)
+then an interval with a lower bound $\leq 0$ and an
 upper bound $\geq 0$ is sufficient to form an interval scheme.
@@
 that do not produce an upper bound by their elongation. A lower bound
-may not be generated if all indexing schemes are not interval schemes.
+may not be generated if none of the indexing schemes are interval schemes.
 In the case of our example, the resulting $\finalSet$ comprises a set
@@
 intervals. However, some intervals computed from the above steps
-may not the constraints of this interpretation,
+may not match the constraints of this interpretation,
 \eg{}, $\interv{-2}{1}{} \times \interv{-1}{1}{}$.  By
@@
 
-The final stage of inference, thus maps the component
+The final stage of inference thus maps the component
 of every interval scheme to a region constant via $\interp{-}^{-1}$,
@@
 Where summation is by the syntactic \term{+} and product is by the
-syntactic \term{*}
+syntactic \term{*}.
 

### 2017-04-18 oopsla17/model.tex
various edits in section 4-6

@@
 meaning of our specifications; (2) provides a basis for the inference and
-checking algorithms (\Cref{sec:algorithms}); (3) justifies an equational
-theory for specifications in the next section; (4) is used to optimise
+checking algorithms in \Cref{sec:algorithms}; (3) justifies the equational
+theory for specifications; (4) is used to optimise
 specifications using lattice identities; and (5) can be used to guide
@@
 variable. In the second dimension, the index is unconstrained
-as it is a constant.
+as they are constants.
 
 Index schemes form a lattice which provides a rich set of equations
-and properties, which we exploit. We first set up the 
-domain of the model (\S\ref{sec:domain}), using it to define a semantics
-for our specification language (\S\ref{sec:semantics}) and then as
-the target for an interpretation on imperative code (\S\ref{sec:fromcode}).
-We state various results in this section, for which the proofs are
-provided in \Cref{sec:proofs}.
+and properties, which we exploit. We first set up the domain of the
+model (\S\ref{sec:domain}), using it to define a semantics for our
+specification language (\S\ref{sec:semantics}) and then as the target
+for an interpretation of the sytnax of array subscripts in code
+(\S\ref{sec:fromcode}).  We state various results in this section, for
+which the proofs are provided in \Cref{sec:proofs}.
 
@@
   $\pi_i : {\mathcal{P}(\bz{})}^N \to \mathcal{P}(\bz{})$. For an index scheme
-  $I$, we refer to $\pi_i(S)$ as the \emph{$i^{th}$ component} of $S$. We assume
+  $S$, we refer to $\pi_i(S)$ as the \emph{$i^{th}$ component} of $S$. We assume
   that $i$, when used for projection, always lies between $1$ and $N$.
@@
 \begin{defn}[Intervals with an optional hole]
+\label{defn:my-interval}
   We define an extended notion of closed interval on $\bz{}$ which may contain a
@@
   An interval scheme is an $N$-length vector ($N$-ways Cartesian
-  product) of intervals on \bz{}, denoted by the set
-  $\textit{Interval}^N$ for a product of $N$ intervals.
+  product) of intervals on \bz{} (as in \Cref{defn:my-interval}),
+  denoted by the set $\textit{Interval}^N$ for a product of $N$
+  intervals.
 \end{defn}
@@
 %
-  \eg{}, $\mathsf{lower}$ is an injection $\mathsf{lower} : a \to \mathsf{Approx}
-  \; a$ etc.
-st%  These will be used in the following subsection to give meaning to the
+  \eg{}, $\mathsf{lower}$ is an injection into $\mathsf{Approx}$, of
+  type $\mathsf{lower} : a \to \mathsf{Approx} \; a$.
+%  These will be used in the following subsection to give meaning to the
 %  specification modifiers for approximation and multiplicity.
@@
    \interp{\texttt{\textcap{r} + \textcap{s}}}_N & =
-    \interp{\textcap{r}}_N \vee \interp{\textcap{s}}_N
+    \interp{\textcap{r}}_N \cup \interp{\textcap{s}}_N
 \\
   \interp{\texttt{\textcap{r} * \textcap{s}}}_N & =
-   \interp{\textcap{r}}_N \wedge \interp{\textcap{s}}_N \\
+   \interp{\textcap{r}}_N \cap \interp{\textcap{s}}_N \\
 %%
@@
 modelled in terms of the join (union) and meet (intersection) of
-interval schemes. Thus the syntax of regions is modelled as members of
+regions. Thus the syntax of regions is modelled by members of
 $\region{N}$.
@@
   The lattice model is sound with respect to the theory of approximation. Let $R$
-  and $S$ be $N$ dimensional regions, then we have
+  and $S$ be $N$-dimensional regions, then we have
 %
@@
 the next section, we use this interpretation on array subscripts in
-the static analysis producer preceding checking and inference.
+the static analysis preceding checking and inference.
 
@@
 
-Union is not as flexible as
-intersection. \Cref{lem:vector-union} does not form a closure for
-indexing schemes in general. For this reason, we represent sets of
-indexing schemes as unevaluated union terms, or \emph{union normal
-  form}.  Since the model forms a distributive lattice as established
-in \Cref{prop:region-lattice}, the intersections can be pushed inwards
-using the distributive law leaving only unions at the outer-level. We
-further exploit associativity of union to put unions into a cons-tree
-making the unions effectively a non-empty list.  We do not, however,
-attempt to canonicalise the representation. Later on
-\Cref{lem:vector-union} and the index scheme subset are used to reduce the
-size of union normal form.
+Union is not as flexible as intersection; it is not closed over
+indexing schemes, only on indexing schemes which are equal in all
+components but one (\Cref{lem:vector-union}). For this reason, we
+represent sets of indexing schemes as unevaluated union terms, or
+\emph{union normal form}.  Since the model forms a distributive
+lattice as established in \Cref{prop:region-lattice}, intersections
+can be pushed inwards using the distributive law leaving only unions
+at the outer-level. We further exploit associativity of union to put
+unions into a cons-tree effectively representing union normal form as
+a non-empty list.  We do not, however, attempt to canonicalise the
+representation. However, at the end of the inference procedure,
+\Cref{lem:vector-union} and index scheme subset are used to reduce the
+size of union normal form as a way of simplifying specifications.
 

### 2017-04-18 oopsla17/theory.tex
Typo in equational theory
various edits in section 4-6

@@
 %
-  \item[Distributioneq] \texttt{*} distributes over \texttt{+} and dually
+  \item[Distribution] \texttt{*} distributes over \texttt{+} and dually
     \texttt{+} distributes over \texttt{*}, meaning if \textcap{R}, \textcap{S},
@@
     Here \textcap{p1} is \texttt{nonpointed} if both \textcap{p2} and
-    \textcap{p3} are \texttt{nonpointed} and \texttt{pointed} otherwise.
+    \textcap{p3} are \texttt{nonpointed}, otherwise \textcap{p1} is
+    \texttt{pointed}.
 \end{description}
@@
 \subsection{Approximations}
-%
+
+\noindent
 We define a partial of order of approximations, $\preceq$. This relation is used
-in the equational theory and provides a means of writing a more compact lower and upper
+in the equational theory and provides a means of writing more compact lower and upper
 bound specifications. The relation is defined as follows:
@@
 \end{restatable}
+%
+If an array computation conforms to a specification $R$ and
+$R \preceq S$ then it conforms also to $\texttt{atMost} \, S$.

### 2017-04-18 oopsla17/evaluation.tex
various edits, shrinking, and conclusion
adding number back in
submitted version

@@
+\newcommand{\overallStencilsSpecs}{77118}
 \newcommand\regname[1]{\texttt{#1}}
@@
 This shows that we can express a large number of
-array access patterns within our high-level abstractions and validates our
-initial design choices informed by our data (~\Cref{subsec:choices}).
+array access patterns within our high-level abstractions and it validates our
+initial design choices informed by our data (\Cref{subsec:choices}).
 
@@
 Common examples of this were pointwise transformations on data (such
-as scaling).
+as scaling). The remaining \num{\overallnonTrivial} we consider to
+be non-trivial, \ie{} with a higher potential for programming error.
 
@@
 
+\textbf{ReadOnce} specification occurred \num{\overallreadOnce} times, which
+correlates with the high proportion of unique-subscript expressions
+in the initial study.
+
 % \eg{}, roughly half the specifications did not involve $\term{*}$,
@@
 \mintinline{fortran}{x = i + 1}, or using functions that
-we could analyse or classify;
+we could not analyse or classify;
 
@@
 avoided through the use of stencil specification. The buggy code was
-identified as \pointed\ in all dimensions:
+identified as \pointed\ in all dimensions by our inference:
 
@@
 \end{minted}
-
+%
 But the programmer made it clear in the comments after the bug fix
-that the intention had been \forward\ in the third dimension:
-
+that the intention had been \forward\ in the third dimension, \ie{}, a
+specification:
+%
 \begin{minted}[breakafter=+:,breakindent=-0.6em,breaksymbolsep=0.4em,linenos=false,xleftmargin=-0.5em]{fortran}
@@
 where a subscript \fortran{u(i)} is corrected to \fortran{u(i-1)} which
-could have been detected by an initial specification \fortran{backward(depth = 1,
-  dim = 1, nonpointed)}.
+could have been detected by an initial specification of \fortran{!=
+  stencil backward(depth=1, dim=1, nonpointed)}.
 
